### 2016-10-26 draft.tex
ok

\documentclass[a4paper, 11pt]{article}

\usepackage{natbib}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{appendix}

\title{TITLE}
\author{Radu IOSIF \& Xiao XU \\
	$from$ CNRS VERIMAG, FRANCE}
\date{$DATE$}

% ====================
\begin{document}

% Title
\maketitle

% Begin Section{Alternating Data Automata}	
\section{Alternating Data Automata}

	% Begin Subsection{Th(D)}
	\subsection{$Th(\mathcal{D})$}

	Given a possibly infinite data domain $\mathcal{D}$, we denote by $Th(\mathcal{D}) = \langle \mathcal{D}, f_1, f_2, ..., f_m \rangle$ the set of syntactically correct first-order formulae with function symbols $f_1, f_2, ..., f_m$.
	% End Subsection{Th(D)}

	% Begin Subsection{B^Q}
	\subsection{$\mathcal{B}^{\mathcal{Q}}$}

	Denote by the symbol $\mathcal{B}$ the two-element Boolean algebra $\mathcal{B} = (\{ 0, 1 \}, \lor, \land, \lnot, 0, 1)$.\\

	Let $\mathcal{Q}$ be a set, then $\mathcal{B}^{\mathcal{Q}}$ is the set of all mappings from $\mathcal{Q}$ to $\mathcal{B}$.
	% End Subsection{B^Q}

	% Begin Subsection{Definition}
	\subsection{Definition}

	An $Alternating\ Data\ Automaton$ (ADA) is a tuple $\mathcal{A} = \langle \mathcal{D}, \mathcal{X}, \Sigma, \mathcal{Q}, \bm{i}, \mathcal{F}, \bm{g} \rangle$ where \cite{HoFL}:
	\begin{itemize}
		\item $\mathcal{D}$ is an initial data domain;
		\item $\mathcal{X} = \{ x_1, x_2, ..., x_n \}$ is a set of variables;
		\item $\Sigma$ is a finite alphabet of input events;
		\item $\mathcal{Q}$ is a finite set of states;
		\item $\bm{i} \in \mathcal{Q}$ is an initial state;
		\item $\mathcal{F} \subseteq \mathcal{Q}$ is a set of final states;
		\item $\bm{g}$ is a function of $\mathcal{Q}$ into the set of all functions of $\Sigma$ into functions $\mathcal{D}^{\mathcal{X}} \times \mathcal{D}^{\mathcal{X}} \to ( \mathcal{B}^{\mathcal{Q}} \to \mathcal{B} )$.\\
	\end{itemize}

	We define $\bm{f} \in \mathcal{B}^{\mathcal{Q}}$ by the condition: $\bm{f}(q) = 1$ iff $q \in \mathcal{F}$.
	% End Subsection{Definition}

	% Begin Subsection{Function g}
	\subsection{Function $\bm{g}$}

	For each $q \in \mathcal{Q}$, $a \in \Sigma$, $v \in \mathcal{D}^{\mathcal{X}}$, $v' \in \mathcal{D}^{\mathcal{X}}$, $u \in \mathcal{B}^{\mathcal{Q}}$, $\bm{g}(q)(a)(v, v')(u)$ is a Boolean combination of $u(q_t)$ and $\phi_t(v, v')$ where $q_t \in \mathcal{Q}$ is the successor of $q$ with symbol $a$ and $\phi_t \in Th(\mathcal{D})$.\\

	In the cases where $v'$ is not used, we write $\bm{g}(q)(a)(v)(u)$ instead of $\bm{g}(q)(a)(v, v')(u)$.\\

	For example, we define an ADA $\mathcal{A} = \langle \mathcal{D}_{\mathcal{A}}, \mathcal{X}_{\mathcal{A}}, \Sigma_{\mathcal{A}}, \mathcal{Q}_{\mathcal{A}}, \bm{i}_{\mathcal{A}}, \mathcal{F}_{\mathcal{A}}, \bm{g}_{\mathcal{A}} \rangle$, where:
	\begin{itemize}
		\item $\mathcal{D}_{\mathcal{A}} = \mathcal{N}$;
		\item $\mathcal{X}_{\mathcal{A}} = \{ x \}$;
		\item $\Sigma_{\mathcal{A}} = \{ a, b \}$;
		\item $\mathcal{Q}_{\mathcal{A}} = \{ q_0, q_1, q_2, q_3 \}$;
		\item $\bm{i}_{\mathcal{A}} = q_0$;
		\item $\mathcal{F}_{\mathcal{A}} = \{ q_0 \}$;
		\item $\bm{g}_{\mathcal{A}}$ is given by:
		\begin{tabular}{|c|p{5cm}|l|}
		\hline  & $a$ & $b$\\
		\hline $q_0$ & $u(q_1) \land (x > 3 \land x' = x + 1) \lor u(q_2) \land (x < 3 \land x' = x + 1) \lor u(q_3) \land (x < 1 \land x' = x + 1)$ & $0$\\
		\hline $q_1$ & 0 & $u(q_0) \land x < 5$\\
		\hline $q_2$ & 0 & $u(q_0) \land x = 3$\\
		\hline $q_3$ & 0 & $u(q_1) \land x = 0$\\
		\hline
		\end{tabular}
	\end{itemize}

	\begin{align*}
		&\qquad\ \bm{g}(q_0)(a)(2, 3)(\bm{f})\\
		&= \quad \bm{f}(q_1) \land (2 > 3 \land 3 = 2 + 1) \lor \bm{f}(q_2) \land (2 < 3 \land 3 = 2 + 1) \lor \bm{f}(q_3) \land (2 < 1 \land 3 = 2 + 1)\\
		&= \quad 0 \land 0 \lor 0 \land 1 \lor 0 \land 0\\
		&= \quad 0 \lor 0 \lor 0\\
		&= \quad 0 \\
		\\
		&\qquad\ \bm{g}(q_1)(b)(4)(\bm{f})\\
		&=  \quad \bm{f}(q_0) \land 4 < 5\\
		&=  \quad 1 \land 1\\
		&=  \quad 1
	\end{align*}
	% End Subsection{Function g}

	% Begin Subsection{Extension of Function g}
	\subsection{Extension of Function $\bm{g}$}

	Now we extend $\bm{g}$ to $\mathcal{Q} \to ((\mathcal{D}^{\mathcal{X}} \times \Sigma)^* \times \mathcal{B}^{\mathcal{Q}} \to \mathcal{B})$. Giving $q \in \mathcal{Q}$, $w \in (\mathcal{D}^{\mathcal{X}} \times \Sigma)^*$, $u \in \mathcal{B}^{\mathcal{Q}}$, we have:
	\begin{itemize}
		\item If $w = \lambda$, then $\bm{g}(q)(w, u) = u(q)$;
		\item If $w = \langle v, a \rangle$ with $v \in \mathcal{D}^{\mathcal{X}}$ and $a \in \Sigma$, then $\bm{g}(q)(w, u) = \bm{g}(q)(a)(v)(u)$;
		\item If $w = \langle v, a \rangle \langle v', b \rangle w'$ with $v, v' \in \mathcal{D}^{\mathcal{X}}$ and $a, b \in \Sigma$ and $w' \in (\mathcal{D}^{\mathcal{X}} \times \Sigma)^*$, then $\bm{g}(q)(w, u) = \bm{g}(q)(a)(v, v')(u)_{[u(q_t) / \bm{g}(q_t)(\langle v', b \rangle w', u)]}$ where $q_t \in \mathcal{Q}$ is the successor of $q$ with symbol $a$.
	\end{itemize}
	% End Subsection{Extension of Function g}

	% Begin Subsection{Acceptance of a Word}
	\subsection{Acceptance of a Word}

	Let $\mathcal{A} = \langle \mathcal{D}, \mathcal{X}, \Sigma, \mathcal{Q}, \bm{i}, \mathcal{F}, \bm{g} \rangle$	be an ADA, a word $w \in (\mathcal{D}^{\mathcal{X}} \times \Sigma)^*$ is accepted by $\mathcal{A}$ iff $\bm{g}(\bm{i})(w, \bm{f})$.\\

	Let's take the example in last page, and we try the following words:
	\begin{align*}
		&\qquad\ \bm{g}(q_0)(\langle v_0, a \rangle \langle v_1, b \rangle, \bm{f})\\
		&= \quad \bm{g}(q_1)(\langle v_1, b \rangle, \bm{f}) \land (v_0 > 3 \land v_1 = v_0 + 1)\\
			&\quad \lor\ \bm{g}(q_2)(\langle v_1, b \rangle, \bm{f}) \land (v_0 < 3 \land v_1 = v_0 + 1)\\
			&\quad \lor\ \bm{g}(q_3)(\langle v_1, b \rangle, \bm{f}) \land (v_0 < 1 \land v_1 = v_0 + 1)\\
		&= \quad \bm{f}(q_0) \land v_1 < 5 \land v_0 > 3 \land v_1 = v_0 + 1\\
			&\quad \lor\ \bm{f}(q_0) \land v_1 = 3 \land v_0 < 3 \land v_1 = v_0 + 1\\
			&\quad \lor\ \bm{f}(q_1) \land v_1 = 0 \land v_0 < 1 \land v_1 = v_0 + 1\\
		&= \quad 1 \land v_1 < 5 \land v_0 > 3 \land v_1 = v_0 + 1\\
			&\quad \lor\ 1 \land v_1 = 3 \land v_0 < 3 \land v_1 = v_0 + 1\\
			&\quad \lor\ 0 \land v_1 = 0 \land v_0 < 1 \land v_1 = v_0 + 1\\
		&= \quad v_1 < 5 \land v_0 > 3 \land v_1 = v_0 + 1 \lor v_1 = 3 \land v_0 < 3 \land v_1 = v_0 + 1\\
		&= \quad v_1 = 3 \land v_0 < 3 \land v_1 = v_0 + 1\\
		&= \quad v_0 = 2 \land v_1 = 3
	\end{align*}

	Therefore, the automaton $\mathcal{A}$ accepts the word $\langle 2, a \rangle \langle 3, b \rangle$.
	% End Subsection{Acceptance of a Word}

	% Begin Subsection{Language}
	\subsection{Language}

	The language accepted by an alternating automaton $\mathcal{A} = \langle \mathcal{D}, \mathcal{X}, \Sigma, \mathcal{Q}, \bm{i}, \mathcal{F}, \bm{g} \rangle$ is the set $\mathcal{L}(\mathcal{A}) = \{ w \in (\mathcal{D}^{\mathcal{X}} \times \Sigma)^*\ |\ \bm{g}(\bm{i})(w, \bm{f}) = 1 \}$.
	% End Subsection{Language}

	% Begin Subsection{Complementation}
	\subsection{Complementation}

	Let $\mathcal{A} = \langle \mathcal{D}, \mathcal{X}, \Sigma, \mathcal{Q}, \bm{i}, \mathcal{F}, \bm{g} \rangle$ be an ADA, we can now construct another ADA $\mathcal{\overline{A}} = \langle \mathcal{D}', \mathcal{X}', \Sigma', \mathcal{Q}', \bm{i}', \mathcal{F}', \bm{g}' \rangle$ such that $\mathcal{L}(\mathcal{\overline{A}}) = \overline{\mathcal{L}(\mathcal{A})}$:
	\begin{itemize}
		\item $\mathcal{D}' = \mathcal{D}$;
		\item $\mathcal{X}' = \mathcal{X}$;
		\item $\Sigma' = \Sigma$;
		\item $\mathcal{Q}' = \mathcal{Q}$;
		\item $\bm{i}' = \bm{i}$;
		\item $\mathcal{F}' = \mathcal{Q} - \mathcal{F}$ therefore for each state $q \in \mathcal{Q}$, $\bm{f}'(q) = 1$ iff $q \in \mathcal{F}'$;
		\item For each $q \in \mathcal{Q}$, $a \in \Sigma$, $v \in \mathcal{D}^{\mathcal{X}}$, $v' \in \mathcal{D}^{\mathcal{X}}$, $u \in \mathcal{B}^{\mathcal{Q}}$, $u' \in \mathcal{B}^{\mathcal{Q}}$ and $u'(q) = \overline{u(q)}$, we have $\bm{g}'(q)(a)(v, v')(u') = \overline{\bm{g}(q)(a)(v, v')(u)_{[u(q_t) / \overline{u'(q_t)}]}}$ where $q_t \in \mathcal{Q}$ is the successor of $q$ with symbol $a$.
		\end{itemize}

		Hence, for the extension of function $\bm{g}'$, we have:
		\begin{itemize}
			\item If $w = \lambda$, then:\\
			$\bm{g}'(q)(w, u') = u'(q)$;
			\item If $w = \langle v, a \rangle$ with $v \in \mathcal{D}^{\mathcal{X}}$ and $a \in \Sigma$, then:\\
			$\bm{g}'(q)(w, u') = \bm{g}'(q)(a)(v)(u') = \overline{\bm{g}(q)(a)(v)(u)_{[u(q_t) / \overline{u'(q_t)}]}}$ where $q_t \in \mathcal{Q}$ is a successor of $q$ with symbol $a$;
			\item If $w = \langle v, a \rangle \langle v', b \rangle w'$ with $v \in \mathcal{D}^{\mathcal{X}}$, $v' \in \mathcal{D}^{\mathcal{X}}$, $a \in \Sigma$, $b \in \Sigma$ and $w' \in (\mathcal{D}^{\mathcal{X}} \times \Sigma)^*$, then:\\
			$\bm{g}'(q)(w, u') = \bm{g}'(q)(a)(v, v')(u')_{[u'(q_t) / \bm{g}'(q_t)(\langle v', b \rangle w', u')]} = \overline{\bm{g}(q)(a)(v, v')(u)_{[u(q_t) / \overline{\bm{g}'(q_t)(\langle v', b \rangle w', u')}]}}$ where $q_t \in \mathcal{Q}$ is the successor of $q$ with symbol $a$.
		\end{itemize}

		Let's still take the example in previous pages, now we construct the $\mathcal{\overline{A}} = \langle \mathcal{D}', \mathcal{X}', \Sigma', \mathcal{Q}', \bm{i}', \mathcal{F}', \bm{g}' \rangle$ such that $\mathcal{L}(\mathcal{\overline{A}}) = \overline{\mathcal{L}(\mathcal{A})}$:
		\begin{itemize}
			\item $\mathcal{D}' = \mathcal{N}$;
			\item $\mathcal{X}' = \{ x \}$;
			\item $\Sigma' = \{ a, b \}$;
			\item $\mathcal{Q}' = \{ q_0, q_1, q_2, q_3 \}$;
			\item $\bm{i}' = q_0$;
			\item $\mathcal{F}' = \{ q_1, q_2, q_3 \}$;
			\item $\bm{g}'$ is given by:
			\begin{tabular}{|c|p{5cm}|l|}
				\hline  & $a$ & $b$\\
				\hline $q_0$ & $(u(q_1) \lor (x \leq 3 \lor x' \neq x + 1)) \land (u(q_2) \lor (x \geq 3 \lor x' \neq x + 1)) \land (u(q_3) \lor (x \geq 1 \lor x' \neq x + 1))$ & $1$\\
				\hline $q_1$ & 1 & $u(q_0) \lor x \geq 5$\\
				\hline $q_2$ & 1 & $u(q_0) \lor x \neq 3$\\
				\hline $q_3$ & 1 & $u(q_1) \lor x \neq 0$\\
				\hline
			\end{tabular}
		\end{itemize}

		\begin{align*}
			&\qquad\ \bm{g}'(q_0)(\langle v_0, a \rangle \langle v_1, b \rangle, \bm{f}')\\
			&= \quad (\bm{g}'(q_1)(\langle v_1, b \rangle, \bm{f}') \lor (v_0 \leq 3 \lor v_1 \neq v_0 + 1))\\
				&\quad \land\ (\bm{g}'(q_2)(\langle v_1, b \rangle, \bm{f}') \lor (v_0 \geq 3 \lor v_1 \neq v_0 + 1))\\
				&\quad \land\ (\bm{g}'(q_3)(\langle v_1, b \rangle, \bm{f}') \lor (v_0 \geq 1 \lor v_1 \neq v_0 + 1))\\
			&= \quad (\bm{f}'(q_0) \lor v_1 \geq 5 \lor v_0 \leq 3 \lor v_1 \neq v_0 + 1)\\
				&\quad \land\ (\bm{f}'(q_0) \lor v_1 \neq 3 \lor v_0 \geq 3 \lor v_1 \neq v_0 + 1)\\
				&\quad \land\ (\bm{f}'(q_1) \lor v_1 \neq 0 \lor v_0 \geq 1 \lor v_1 \neq v_0 + 1)\\
			&= \quad (v_1 \geq 5 \lor v_0 \leq 3 \lor v_1 \neq v_0 + 1) \land (v_1 \neq 3 \lor v_0 \geq 3 \lor v_1 \neq v_0 + 1)\\
			&= \quad v_1 \geq 5 \lor v_0 \leq 3 \land v_1 \neq 3 \lor v_0 = 3 \lor v_1 \neq v_0 + 1\\
			&= \quad v_0 \neq 2 \lor v_1 \neq 3
		\end{align*}

		Therefore, the automaton $\mathcal{\overline{A}}$ does not accept the word $\langle 2, a \rangle \langle 3, b \rangle$.\\

		$\bm{g}'(q_0)(\lambda, \bm{f}') = \bm{f}'(q_0) = 0$\\

		Therefore, the automaton $\mathcal{\overline{A}}$ accepts the empty word.

	% End Subsection{Complementation}

% End Section{Alternating Data Automata}

% References
\bibliographystyle{plain}
\bibliography{references}

% Appendix
\appendix

	% Begin Section{Length of a Word}
	\section{Length of a Word}

	The length of a word $w$ is defined as below:\\
	$length(w) = k$ iff $w \in (\mathcal{D}^{\mathcal{X}} \times \Sigma)^k$\\

	Hence, $length(\lambda) = 0$.
	% End Section{Length of a Word}

	% Begin Section{Proof of the Correctness of Complementation}
	\section{Proof of the Correctness of Complementation}

	It is same to prove $\bm{g}(\bm{i})(w, \bm{f}) = \overline{\bm{g}'(\bm{i})(w, \bm{f}')}$ with $w \in (\mathcal{D}^{\mathcal{X}} \times \Sigma)^*$:
	\begin{itemize}
		\item If $w = \lambda$, then for each $q \in \mathcal{Q}$:
		\begin{align*}
			&\qquad\ \bm{g}(q)(w, \bm{f})\\
			&= \quad \bm{f}(q)\\
			&= \quad q \in \mathcal{F}\\
			&= \quad \overline{q \in \mathcal{F}'}\\
			&= \quad \overline{\bm{f}'(q)}\\
			&= \quad \overline{\bm{g}'(q)(w, \bm{f}')}
		\end{align*}
		\item If $w = \langle v, a \rangle$ with $v \in \mathcal{D}^{\mathcal{X}}$ and $a \in \Sigma$, then for each $q \in \mathcal{Q}$:
		\begin{align*}
			&\qquad\ \bm{g}(q)(w, \bm{f})\\
			&= \quad \bm{g}(q)(a)(v)(\bm{f})\\
			&= \quad \bm{g}(q)(a)(v)(\bm{f})_{[\bm{f}(q_t) / \overline{\bm{f}'(q_t)}]} \quad \diamondsuit\\
			&= \quad \overline{\overline{\bm{g}(q)(a)(v)(\bm{f})_{[\bm{f}(q_t) / \overline{\bm{f}'(q_t)}]}}}\\
			&= \quad \overline{\bm{g}'(q)(a)(v)(\bm{f}')}\\
			&= \quad \overline{\bm{g}'(q)(w, \bm{f}')}
		\end{align*}
		$\diamondsuit$ $q_t \in \mathcal{Q}$ is the successor of $q$ with symbol $a$.\\
	\end{itemize}

	Now, let's $\bm{suppose}$ that for each $k$-length non-empty($k \geq 1$) word $w' = \langle v', a' \rangle w''$ with $v' \in \mathcal{D}^{\mathcal{X}}$ and $a' \in \Sigma$ and $w'' \in (\mathcal{D}^{\mathcal{X}} \times \Sigma)^{k-1}$, we always have $\bm{g}(q)(w', \bm{f}) = \overline{\bm{g}'(q)(w', \bm{f}')}$, then for any $(k+1)$-length word $w = \langle v, a \rangle w' = \langle v, a \rangle \langle v', a' \rangle w''$ with $v \in \mathcal{D}^{\mathcal{X}}$ and $a \in \Sigma$, we can have:
	\begin{align*}
		&\qquad\ \bm{g}(q)(w, \bm{f})\\
		&= \quad \bm{g}(q)(\langle v, a \rangle \langle v', a' \rangle w'', \bm{f})\\
		&= \quad \bm{g}(q)(a)(v, v')(\bm{f})_{[\bm{f}(q_t) / \bm{g}(q_t)(\langle v', a' \rangle w'', \bm{f})]} \quad \clubsuit\\
		&= \quad \bm{g}(q)(a)(v, v')(\bm{f})_{[\bm{f}(q_t) / \bm{g}(q_t)(w', \bm{f})]}\\
		&= \quad \bm{g}(q)(a)(v, v')(\bm{f})_{[\bm{f}(q_t) / \overline{\bm{g}'(q_t)(w', \bm{f}')}]}\\
		&= \quad \overline{\overline{\bm{g}(q)(a)(v, v')(\bm{f})_{[\bm{f}(q_t) / \overline{\bm{g}'(q_t)(w', \bm{f}')}]}}}\\
		&= \quad \overline{\bm{g}'(q)(a)(v, v')(\bm{f}')_{[\bm{f}'(q_t) / \bm{g}'(q_t)(w', \bm{f}')]}}\\
		&= \quad \overline{\bm{g}'(q)(w, \bm{f}')}
	\end{align*}
	$\clubsuit$ $q_t \in \mathcal{Q}$ is the successor of $q$ with symbol $a$.\\

	We already have $\bm{g}(q)(w, \bm{f}) = \overline{\bm{g}'(q)(w, \bm{f}')}$ when the length is 1. Therefore, we can have it for all the length $k \geq 1$.\\

	Hence, $\bm{g}(\bm{i})(w, \bm{f}) = \overline{\bm{g}'(\bm{i})(w, \bm{f}')}$ with $w \in (\mathcal{D}^{\mathcal{X}} \times \Sigma)^*$.
	% End Section{Proof of the Correctness of Complementation}

\end{document}
% ====================

### 2017-05-01 draft.tex
first version

@@
-\documentclass[a4paper, 11pt]{article}
+\documentclass{llncs}
 
-\usepackage{natbib}
-\usepackage{bm}
+\usepackage[usenames,dvipsnames]{color}
+
+\usepackage{latexsym}
+\usepackage{amsxtra} 
+\usepackage{amssymb}
 \usepackage{amsmath}
-\usepackage{appendix}
+\usepackage{pslatex}
+\usepackage{epsfig}
+\usepackage{wrapfig}
+\usepackage{paralist}
+\usepackage{graphics}
+\usepackage{stmaryrd}
+\usepackage{txfonts}
+\usepackage{framed}
+\usepackage{makecell}
+\usepackage{url}
+\usepackage{tikz}
+\usetikzlibrary{automata,positioning, calc}
+\usepackage[inline,shortlabels]{enumitem}
 
-\title{TITLE}
-\author{Radu IOSIF \& Xiao XU \\
-	$from$ CNRS VERIMAG, FRANCE}
-\date{$DATE$}
+\usepackage{proof}
 
-% ====================
+\usepackage{algorithm}
+\usepackage{algorithmicx}
+\usepackage[noend]{algpseudocode}
+
+\usepackage[draft]{commenting}
+
+\pagestyle{plain}
+
+\include{commands}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-% Title
+\title{The Impact of Alternation}
+
+\author{Radu Iosif \and Xiao Xu}
+
+\institute{Verimag/CNRS/Universit\'e de Grenoble Alpes}
+
 \maketitle
 
-% Begin Section{Alternating Data Automata}	
-\section{Alternating Data Automata}
+\begin{abstract}
+\end{abstract}
 
-	% Begin Subsection{Th(D)}
-	\subsection{$Th(\mathcal{D})$}
+\section{Introduction}
 
-	Given a possibly infinite data domain $\mathcal{D}$, we denote by $Th(\mathcal{D}) = \langle \mathcal{D}, f_1, f_2, ..., f_m \rangle$ the set of syntactically correct first-order formulae with function symbols $f_1, f_2, ..., f_m$.
-	% End Subsection{Th(D)}
+\section{Preliminaries}
 
-	% Begin Subsection{B^Q}
-	\subsection{$\mathcal{B}^{\mathcal{Q}}$}
+A \emph{signature} $\sig = (\ssorts{\sig},\sfuns{\sig})$ consists of a
+set $\ssorts{\sig}$ of \emph{sort symbols} and a set $\sfuns{\sig}$ of
+sorted \emph{function symbols}. To simplify the presentation, we
+assume w.l.o.g. that $\ssorts{\Sigma}$ consists of two distinct sorts
+$\Data$ and $\Bool$ and each function symbol $f \in \sfuns{\sig}$ has
+$\#(f) \geq 0$ arguments of sort $\Data$ and return value $\sigma(f)
+\in \ssorts{\sig}$. If $n=0$, $f$ is a \emph{constant} of sort
+$\sigma(f)$ and we consider the constants $\top,\bot \in \sfuns{\sig}$
+of sort $\Bool$, denoting the truth values true and false.
 
-	Denote by the symbol $\mathcal{B}$ the two-element Boolean algebra $\mathcal{B} = (\{ 0, 1 \}, \lor, \land, \lnot, 0, 1)$.\\
+We consider an infinite countable set of \emph{variables} $\vars$,
+where each $x \in \vars$ has an associated sort $\sigma(x)$. Terms are
+defined recursively: a term $t$ of sort $S$ is a variable $x \in
+\vars$ where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
+$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
+\emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
+between two terms of sort $\Data$. A \emph{literal} is an atom or its
+negation. A \emph{formula} is a combination of atoms using disjunction
+$\vee$, conjunction $\wedge$ and negation $\neg$.  A formula is in
+disjunctive normal form (DNF) if it is a disjunction of conjunctions
+of literals $\bigvee_{i=1}^N \bigwedge_{j=1}^{M_i} \ell_{ij}$.
 
-	Let $\mathcal{Q}$ be a set, then $\mathcal{B}^{\mathcal{Q}}$ is the set of all mappings from $\mathcal{Q}$ to $\mathcal{B}$.
-	% End Subsection{B^Q}
+An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
+\item the sort $\Data$ into a non-empty set $\Data^\I$, 
+\item the $\Bool$ sort into the set $\Bool^\I$ containing the boolean
+  values, and
+\item each function symbol $f$ into a total function $f^\I :
+  (\Data^\I)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
+  $\sigma(f)^I$ when $\#(f)=0$.
+\end{inparaenum}
 
-	% Begin Subsection{Definition}
-	\subsection{Definition}
-
-	An $Alternating\ Data\ Automaton$ (ADA) is a tuple $\mathcal{A} = \langle \mathcal{D}, \mathcal{X}, \Sigma, \mathcal{Q}, \bm{i}, \mathcal{F}, \bm{g} \rangle$ where \cite{HoFL}:
-	\begin{itemize}
-		\item $\mathcal{D}$ is an initial data domain;
-		\item $\mathcal{X} = \{ x_1, x_2, ..., x_n \}$ is a set of variables;
-		\item $\Sigma$ is a finite alphabet of input events;
-		\item $\mathcal{Q}$ is a finite set of states;
-		\item $\bm{i} \in \mathcal{Q}$ is an initial state;
-		\item $\mathcal{F} \subseteq \mathcal{Q}$ is a set of final states;
-		\item $\bm{g}$ is a function of $\mathcal{Q}$ into the set of all functions of $\Sigma$ into functions $\mathcal{D}^{\mathcal{X}} \times \mathcal{D}^{\mathcal{X}} \to ( \mathcal{B}^{\mathcal{Q}} \to \mathcal{B} )$.\\
-	\end{itemize}
-
-	We define $\bm{f} \in \mathcal{B}^{\mathcal{Q}}$ by the condition: $\bm{f}(q) = 1$ iff $q \in \mathcal{F}$.
-	% End Subsection{Definition}
-
-	% Begin Subsection{Function g}
-	\subsection{Function $\bm{g}$}
-
-	For each $q \in \mathcal{Q}$, $a \in \Sigma$, $v \in \mathcal{D}^{\mathcal{X}}$, $v' \in \mathcal{D}^{\mathcal{X}}$, $u \in \mathcal{B}^{\mathcal{Q}}$, $\bm{g}(q)(a)(v, v')(u)$ is a Boolean combination of $u(q_t)$ and $\phi_t(v, v')$ where $q_t \in \mathcal{Q}$ is the successor of $q$ with symbol $a$ and $\phi_t \in Th(\mathcal{D})$.\\
-
-	In the cases where $v'$ is not used, we write $\bm{g}(q)(a)(v)(u)$ instead of $\bm{g}(q)(a)(v, v')(u)$.\\
-
-	For example, we define an ADA $\mathcal{A} = \langle \mathcal{D}_{\mathcal{A}}, \mathcal{X}_{\mathcal{A}}, \Sigma_{\mathcal{A}}, \mathcal{Q}_{\mathcal{A}}, \bm{i}_{\mathcal{A}}, \mathcal{F}_{\mathcal{A}}, \bm{g}_{\mathcal{A}} \rangle$, where:
-	\begin{itemize}
-		\item $\mathcal{D}_{\mathcal{A}} = \mathcal{N}$;
-		\item $\mathcal{X}_{\mathcal{A}} = \{ x \}$;
-		\item $\Sigma_{\mathcal{A}} = \{ a, b \}$;
-		\item $\mathcal{Q}_{\mathcal{A}} = \{ q_0, q_1, q_2, q_3 \}$;
-		\item $\bm{i}_{\mathcal{A}} = q_0$;
-		\item $\mathcal{F}_{\mathcal{A}} = \{ q_0 \}$;
-		\item $\bm{g}_{\mathcal{A}}$ is given by:
-		\begin{tabular}{|c|p{5cm}|l|}
-		\hline  & $a$ & $b$\\
-		\hline $q_0$ & $u(q_1) \land (x > 3 \land x' = x + 1) \lor u(q_2) \land (x < 3 \land x' = x + 1) \lor u(q_3) \land (x < 1 \land x' = x + 1)$ & $0$\\
-		\hline $q_1$ & 0 & $u(q_0) \land x < 5$\\
-		\hline $q_2$ & 0 & $u(q_0) \land x = 3$\\
-		\hline $q_3$ & 0 & $u(q_1) \land x = 0$\\
-		\hline
-		\end{tabular}
-	\end{itemize}
-
-	\begin{align*}
-		&\qquad\ \bm{g}(q_0)(a)(2, 3)(\bm{f})\\
-		&= \quad \bm{f}(q_1) \land (2 > 3 \land 3 = 2 + 1) \lor \bm{f}(q_2) \land (2 < 3 \land 3 = 2 + 1) \lor \bm{f}(q_3) \land (2 < 1 \land 3 = 2 + 1)\\
-		&= \quad 0 \land 0 \lor 0 \land 1 \lor 0 \land 0\\
-		&= \quad 0 \lor 0 \lor 0\\
-		&= \quad 0 \\
-		\\
-		&\qquad\ \bm{g}(q_1)(b)(4)(\bm{f})\\
-		&=  \quad \bm{f}(q_0) \land 4 < 5\\
-		&=  \quad 1 \land 1\\
-		&=  \quad 1
-	\end{align*}
-	% End Subsection{Function g}
-
-	% Begin Subsection{Extension of Function g}
-	\subsection{Extension of Function $\bm{g}$}
-
-	Now we extend $\bm{g}$ to $\mathcal{Q} \to ((\mathcal{D}^{\mathcal{X}} \times \Sigma)^* \times \mathcal{B}^{\mathcal{Q}} \to \mathcal{B})$.
-Giving $q \in \mathcal{Q}$, $w \in (\mathcal{D}^{\mathcal{X}} \times \Sigma)^*$, $u \in \mathcal{B}^{\mathcal{Q}}$, we have:
-	\begin{itemize}
-		\item If $w = \lambda$, then $\bm{g}(q)(w, u) = u(q)$;
-		\item If $w = \langle v, a \rangle$ with $v \in \mathcal{D}^{\mathcal{X}}$ and $a \in \Sigma$, then $\bm{g}(q)(w, u) = \bm{g}(q)(a)(v)(u)$;
-		\item If $w = \langle v, a \rangle \langle v', b \rangle w'$ with $v, v' \in \mathcal{D}^{\mathcal{X}}$ and $a, b \in \Sigma$ and $w' \in (\mathcal{D}^{\mathcal{X}} \times \Sigma)^*$, then $\bm{g}(q)(w, u) = \bm{g}(q)(a)(v, v')(u)_{[u(q_t) / \bm{g}(q_t)(\langle v', b \rangle w', u)]}$ where $q_t \in \mathcal{Q}$ is the successor of $q$ with symbol $a$.
-	\end{itemize}
-	% End Subsection{Extension of Function g}
-
-	% Begin Subsection{Acceptance of a Word}
-	\subsection{Acceptance of a Word}
-
-	Let $\mathcal{A} = \langle \mathcal{D}, \mathcal{X}, \Sigma, \mathcal{Q}, \bm{i}, \mathcal{F}, \bm{g} \rangle$	be an ADA, a word $w \in (\mathcal{D}^{\mathcal{X}} \times \Sigma)^*$ is accepted by $\mathcal{A}$ iff $\bm{g}(\bm{i})(w, \bm{f})$.\\
-
-	Let's take the example in last page, and we try the following words:
-	\begin{align*}
-		&\qquad\ \bm{g}(q_0)(\langle v_0, a \rangle \langle v_1, b \rangle, \bm{f})\\
-		&= \quad \bm{g}(q_1)(\langle v_1, b \rangle, \bm{f}) \land (v_0 > 3 \land v_1 = v_0 + 1)\\
-			&\quad \lor\ \bm{g}(q_2)(\langle v_1, b \rangle, \bm{f}) \land (v_0 < 3 \land v_1 = v_0 + 1)\\
-			&\quad \lor\ \bm{g}(q_3)(\langle v_1, b \rangle, \bm{f}) \land (v_0 < 1 \land v_1 = v_0 + 1)\\
-		&= \quad \bm{f}(q_0) \land v_1 < 5 \land v_0 > 3 \land v_1 = v_0 + 1\\
-			&\quad \lor\ \bm{f}(q_0) \land v_1 = 3 \land v_0 < 3 \land v_1 = v_0 + 1\\
-			&\quad \lor\ \bm{f}(q_1) \land v_1 = 0 \land v_0 < 1 \land v_1 = v_0 + 1\\
-		&= \quad 1 \land v_1 < 5 \land v_0 > 3 \land v_1 = v_0 + 1\\
-			&\quad \lor\ 1 \land v_1 = 3 \land v_0 < 3 \land v_1 = v_0 + 1\\
-			&\quad \lor\ 0 \land v_1 = 0 \land v_0 < 1 \land v_1 = v_0 + 1\\
-		&= \quad v_1 < 5 \land v_0 > 3 \land v_1 = v_0 + 1 \lor v_1 = 3 \land v_0 < 3 \land v_1 = v_0 + 1\\
-		&= \quad v_1 = 3 \land v_0 < 3 \land v_1 = v_0 + 1\\
-		&= \quad v_0 = 2 \land v_1 = 3
-	\end{align*}
-
-	Therefore, the automaton $\mathcal{A}$ accepts the word $\langle 2, a \rangle \langle 3, b \rangle$.
-	% End Subsection{Acceptance of a Word}
-
-	% Begin Subsection{Language}
-	\subsection{Language}
-
-	The language accepted by an alternating automaton $\mathcal{A} = \langle \mathcal{D}, \mathcal{X}, \Sigma, \mathcal{Q}, \bm{i}, \mathcal{F}, \bm{g} \rangle$ is the set $\mathcal{L}(\mathcal{A}) = \{ w \in (\mathcal{D}^{\mathcal{X}} \times \Sigma)^*\ |\ \bm{g}(\bm{i})(w, \bm{f}) = 1 \}$.
-	% End Subsection{Language}
-
-	% Begin Subsection{Complementation}
-	\subsection{Complementation}
-
-	Let $\mathcal{A} = \langle \mathcal{D}, \mathcal{X}, \Sigma, \mathcal{Q}, \bm{i}, \mathcal{F}, \bm{g} \rangle$ be an ADA, we can now construct another ADA $\mathcal{\overline{A}} = \langle \mathcal{D}', \mathcal{X}', \Sigma', \mathcal{Q}', \bm{i}', \mathcal{F}', \bm{g}' \rangle$ such that $\mathcal{L}(\mathcal{\overline{A}}) = \overline{\mathcal{L}(\mathcal{A})}$:
-	\begin{itemize}
-		\item $\mathcal{D}' = \mathcal{D}$;
-		\item $\mathcal{X}' = \mathcal{X}$;
-		\item $\Sigma' = \Sigma$;
-		\item $\mathcal{Q}' = \mathcal{Q}$;
-		\item $\bm{i}' = \bm{i}$;
-		\item $\mathcal{F}' = \mathcal{Q} - \mathcal{F}$ therefore for each state $q \in \mathcal{Q}$, $\bm{f}'(q) = 1$ iff $q \in \mathcal{F}'$;
-		\item For each $q \in \mathcal{Q}$, $a \in \Sigma$, $v \in \mathcal{D}^{\mathcal{X}}$, $v' \in \mathcal{D}^{\mathcal{X}}$, $u \in \mathcal{B}^{\mathcal{Q}}$, $u' \in \mathcal{B}^{\mathcal{Q}}$ and $u'(q) = \overline{u(q)}$, we have $\bm{g}'(q)(a)(v, v')(u') = \overline{\bm{g}(q)(a)(v, v')(u)_{[u(q_t) / \overline{u'(q_t)}]}}$ where $q_t \in \mathcal{Q}$ is the successor of $q$ with symbol $a$.
-		\end{itemize}
-
-		Hence, for the extension of function $\bm{g}'$, we have:
-		\begin{itemize}
-			\item If $w = \lambda$, then:\\
-			$\bm{g}'(q)(w, u') = u'(q)$;
-			\item If $w = \langle v, a \rangle$ with $v \in \mathcal{D}^{\mathcal{X}}$ and $a \in \Sigma$, then:\\
-			$\bm{g}'(q)(w, u') = \bm{g}'(q)(a)(v)(u') = \overline{\bm{g}(q)(a)(v)(u)_{[u(q_t) / \overline{u'(q_t)}]}}$ where $q_t \in \mathcal{Q}$ is a successor of $q$ with symbol $a$;
-			\item If $w = \langle v, a \rangle \langle v', b \rangle w'$ with $v \in \mathcal{D}^{\mathcal{X}}$, $v' \in \mathcal{D}^{\mathcal{X}}$, $a \in \Sigma$, $b \in \Sigma$ and $w' \in (\mathcal{D}^{\mathcal{X}} \times \Sigma)^*$, then:\\
-			$\bm{g}'(q)(w, u') = \bm{g}'(q)(a)(v, v')(u')_{[u'(q_t) / \bm{g}'(q_t)(\langle v', b \rangle w', u')]} = \overline{\bm{g}(q)(a)(v, v')(u)_{[u(q_t) / \overline{\bm{g}'(q_t)(\langle v', b \rangle w', u')}]}}$ where $q_t \in \mathcal{Q}$ is the successor of $q$ with symbol $a$.
-		\end{itemize}
-
-		Let's still take the example in previous pages, now we construct the $\mathcal{\overline{A}} = \langle \mathcal{D}', \mathcal{X}', \Sigma', \mathcal{Q}', \bm{i}', \mathcal{F}', \bm{g}' \rangle$ such that $\mathcal{L}(\mathcal{\overline{A}}) = \overline{\mathcal{L}(\mathcal{A})}$:
-		\begin{itemize}
-			\item $\mathcal{D}' = \mathcal{N}$;
-			\item $\mathcal{X}' = \{ x \}$;
-			\item $\Sigma' = \{ a, b \}$;
-			\item $\mathcal{Q}' = \{ q_0, q_1, q_2, q_3 \}$;
-			\item $\bm{i}' = q_0$;
-			\item $\mathcal{F}' = \{ q_1, q_2, q_3 \}$;
-			\item $\bm{g}'$ is given by:
-			\begin{tabular}{|c|p{5cm}|l|}
-				\hline  & $a$ & $b$\\
-				\hline $q_0$ & $(u(q_1) \lor (x \leq 3 \lor x' \neq x + 1)) \land (u(q_2) \lor (x \geq 3 \lor x' \neq x + 1)) \land (u(q_3) \lor (x \geq 1 \lor x' \neq x + 1))$ & $1$\\
-				\hline $q_1$ & 1 & $u(q_0) \lor x \geq 5$\\
-				\hline $q_2$ & 1 & $u(q_0) \lor x \neq 3$\\
-				\hline $q_3$ & 1 & $u(q_1) \lor x \neq 0$\\
-				\hline
-			\end{tabular}
-		\end{itemize}
-
-		\begin{align*}
-			&\qquad\ \bm{g}'(q_0)(\langle v_0, a \rangle \langle v_1, b \rangle, \bm{f}')\\
-			&= \quad (\bm{g}'(q_1)(\langle v_1, b \rangle, \bm{f}') \lor (v_0 \leq 3 \lor v_1 \neq v_0 + 1))\\
-				&\quad \land\ (\bm{g}'(q_2)(\langle v_1, b \rangle, \bm{f}') \lor (v_0 \geq 3 \lor v_1 \neq v_0 + 1))\\
-				&\quad \land\ (\bm{g}'(q_3)(\langle v_1, b \rangle, \bm{f}') \lor (v_0 \geq 1 \lor v_1 \neq v_0 + 1))\\
-			&= \quad (\bm{f}'(q_0) \lor v_1 \geq 5 \lor v_0 \leq 3 \lor v_1 \neq v_0 + 1)\\
-				&\quad \land\ (\bm{f}'(q_0) \lor v_1 \neq 3 \lor v_0 \geq 3 \lor v_1 \neq v_0 + 1)\\
-				&\quad \land\ (\bm{f}'(q_1) \lor v_1 \neq 0 \lor v_0 \geq 1 \lor v_1 \neq v_0 + 1)\\
-			&= \quad (v_1 \geq 5 \lor v_0 \leq 3 \lor v_1 \neq v_0 + 1) \land (v_1 \neq 3 \lor v_0 \geq 3 \lor v_1 \neq v_0 + 1)\\
-			&= \quad v_1 \geq 5 \lor v_0 \leq 3 \land v_1 \neq 3 \lor v_0 = 3 \lor v_1 \neq v_0 + 1\\
-			&= \quad v_0 \neq 2 \lor v_1 \neq 3
-		\end{align*}
-
-		Therefore, the automaton $\mathcal{\overline{A}}$ does not accept the word $\langle 2, a \rangle \langle 3, b \rangle$.\\
-
-		$\bm{g}'(q_0)(\lambda, \bm{f}') = \bm{f}'(q_0) = 0$\\
-
-		Therefore, the automaton $\mathcal{\overline{A}}$ accepts the empty word.
-
-	% End Subsection{Complementation}
-
-% End Section{Alternating Data Automata}
-
-% References
-\bibliographystyle{plain}
-\bibliography{references}
-
-% Appendix
-\appendix
-
-	% Begin Section{Length of a Word}
-	\section{Length of a Word}
-
-	The length of a word $w$ is defined as below:\\
-	$length(w) = k$ iff $w \in (\mathcal{D}^{\mathcal{X}} \times \Sigma)^k$\\
-
-	Hence, $length(\lambda) = 0$.
-	% End Section{Length of a Word}
-
-	% Begin Section{Proof of the Correctness of Complementation}
-	\section{Proof of the Correctness of Complementation}
-
-	It is same to prove $\bm{g}(\bm{i})(w, \bm{f}) = \overline{\bm{g}'(\bm{i})(w, \bm{f}')}$ with $w \in (\mathcal{D}^{\mathcal{X}} \times \Sigma)^*$:
-	\begin{itemize}
-		\item If $w = \lambda$, then for each $q \in \mathcal{Q}$:
-		\begin{align*}
-			&\qquad\ \bm{g}(q)(w, \bm{f})\\
-			&= \quad \bm{f}(q)\\
-			&= \quad q \in \mathcal{F}\\
-			&= \quad \overline{q \in \mathcal{F}'}\\
-			&= \quad \overline{\bm{f}'(q)}\\
-			&= \quad \overline{\bm{g}'(q)(w, \bm{f}')}
-		\end{align*}
-		\item If $w = \langle v, a \rangle$ with $v \in \mathcal{D}^{\mathcal{X}}$ and $a \in \Sigma$, then for each $q \in \mathcal{Q}$:
-		\begin{align*}
-			&\qquad\ \bm{g}(q)(w, \bm{f})\\
-			&= \quad \bm{g}(q)(a)(v)(\bm{f})\\
-			&= \quad \bm{g}(q)(a)(v)(\bm{f})_{[\bm{f}(q_t) / \overline{\bm{f}'(q_t)}]} \quad \diamondsuit\\
-			&= \quad \overline{\overline{\bm{g}(q)(a)(v)(\bm{f})_{[\bm{f}(q_t) / \overline{\bm{f}'(q_t)}]}}}\\
-			&= \quad \overline{\bm{g}'(q)(a)(v)(\bm{f}')}\\
-			&= \quad \overline{\bm{g}'(q)(w, \bm{f}')}
-		\end{align*}
-		$\diamondsuit$ $q_t \in \mathcal{Q}$ is the successor of $q$ with symbol $a$.\\
-	\end{itemize}
-
-	Now, let's $\bm{suppose}$ that for each $k$-length non-empty($k \geq 1$) word $w' = \langle v', a' \rangle w''$ with $v' \in \mathcal{D}^{\mathcal{X}}$ and $a' \in \Sigma$ and $w'' \in (\mathcal{D}^{\mathcal{X}} \times \Sigma)^{k-1}$, we always have $\bm{g}(q)(w', \bm{f}) = \overline{\bm{g}'(q)(w', \bm{f}')}$, then for any $(k+1)$-length word $w = \langle v, a \rangle w' = \langle v, a \rangle \langle v', a' \rangle w''$ with $v \in \mathcal{D}^{\mathcal{X}}$ and $a \in \Sigma$, we can have:
-	\begin{align*}
-		&\qquad\ \bm{g}(q)(w, \bm{f})\\
-		&= \quad \bm{g}(q)(\langle v, a \rangle \langle v', a' \rangle w'', \bm{f})\\
-		&= \quad \bm{g}(q)(a)(v, v')(\bm{f})_{[\bm{f}(q_t) / \bm{g}(q_t)(\langle v', a' \rangle w'', \bm{f})]} \quad \clubsuit\\
-		&= \quad \bm{g}(q)(a)(v, v')(\bm{f})_{[\bm{f}(q_t) / \bm{g}(q_t)(w', \bm{f})]}\\
-		&= \quad \bm{g}(q)(a)(v, v')(\bm{f})_{[\bm{f}(q_t) / \overline{\bm{g}'(q_t)(w', \bm{f}')}]}\\
-		&= \quad \overline{\overline{\bm{g}(q)(a)(v, v')(\bm{f})_{[\bm{f}(q_t) / \overline{\bm{g}'(q_t)(w', \bm{f}')}]}}}\\
-		&= \quad \overline{\bm{g}'(q)(a)(v, v')(\bm{f}')_{[\bm{f}'(q_t) / \bm{g}'(q_t)(w', \bm{f}')]}}\\
-		&= \quad \overline{\bm{g}'(q)(w, \bm{f}')}
-	\end{align*}
-	$\clubsuit$ $q_t \in \mathcal{Q}$ is the successor of $q$ with symbol $a$.\\
-
-	We already have $\bm{g}(q)(w, \bm{f}) = \overline{\bm{g}'(q)(w, \bm{f}')}$ when the length is 1. Therefore, we can have it for all the length $k \geq 1$.\\
-
-	Hence, $\bm{g}(\bm{i})(w, \bm{f}) = \overline{\bm{g}'(\bm{i})(w, \bm{f}')}$ with $w \in (\mathcal{D}^{\mathcal{X}} \times \Sigma)^*$.
-	% End Section{Proof of the Correctness of Complementation}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \end{document}
-% ====================
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-05-01 draft.tex
first version

@@
 \in \ssorts{\sig}$. If $n=0$, $f$ is a \emph{constant} of sort
-$\sigma(f)$ and we consider the constants $\top,\bot \in \sfuns{\sig}$
-of sort $\Bool$, denoting the truth values true and false.
+$\sigma(f)$. We consider a constant $\mathbf{0}$ of sort $\Data$ and
+$\top,\bot$ of sort $\Bool$, denoting the truth values true and false.
 
@@
 disjunctive normal form (DNF) if it is a disjunction of conjunctions
-of literals $\bigvee_{i=1}^N \bigwedge_{j=1}^{M_i} \ell_{ij}$.
+of literals $\bigvee_{i=1}^N \bigwedge_{j=1}^{M_i} \ell_{ij}$. We
+denote by $\fv{S}{\phi}$ the set of variables of sort $S$ in $\phi$
+and write $\fv{}{\phi}$ for $\bigcup_{S \in \ssorts{\sig}}
+\fv{S}{\phi}$.
 
@@
 \item each function symbol $f$ into a total function $f^\I :
-  (\Data^\I)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
+  \left(\Data^\I\right)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
   $\sigma(f)^I$ when $\#(f)=0$.
+\end{inparaenum}
+Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
+variable $x \in \vars$ into an element $\nu(x) \in \sigma(x)^\I$. For
+a term $t$, we denote by $t^\I_\nu$ the value obtained by replacing
+each function symbol $f$ by its interpretation $f^\I$ and each
+variable $x$ by its valuation $\nu(x)$. For a formula $\phi$, we write
+$\I,\nu \models \phi$ if the formula obtained by replacing each term
+$t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.
+A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
+there exists a valuation $\nu$ such that $\I,\nu \models \phi$. Given
+formulae $\varphi$ and $\psi$, we say that \emph{$\phi$ entails
+  $\psi$}, denoted $\phi \models^\I \psi$ iff $\I,\nu \models \varphi$
+implies $\I,\nu \models \psi$, for each valuation $\nu$. We omit
+mentioning the interpretation when it is clear from the context.
+
+\section{Alternating Data Automata}
+
+In the rest of this section we fix an interpretation $\I$ and a finite
+alphabet $\Sigma$ of \emph{input events}. Given a finite set $\vec{x}
+\subset \vars$ of variables of sort $\Data$, a \emph{data word} is a
+finite sequence of pairs $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$
+where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n : \vec{x}
+\rightarrow \Data^\I$ are valuations. We denote by $\vec{x} \mapsto
+\Data^\I$ the set of valuations of the variables $\vec{x}$ and by
+$\left(\Sigma \times (\vec{x} \mapsto \Data^\I)\right)^*$ the set of
+data words over $\vec{x}$.
+
+Given two sets of variables $B,X \subset \vars$, we denote by
+$\Form^+(B,X)$ the set of formulae $\phi$ such that $\fv{\Bool}{\phi}
+\subseteq B$, $\fv{\Data}{\phi} \subseteq X$ and each variable from
+$B$ occurs under an even number of negations.
+
+An \emph{alternating data automaton} (ADA) is a tuple $\A =
+\tuple{\vec{x},Q,I,F,\Delta}$, where: \begin{compactitem}
+%
+\item $\vec{x} \subset \vars$ is a finite set of variables of sort
+  $\Data$,
+%
+\item $Q$ is a finite set of variables of sort $\Bool$, also called
+  \emph{states},
+%
+\item $I,F \subseteq Q$ are sets of \emph{initial} and \emph{final}
+  states, and
+%
+\item $\Delta : Q \times \Sigma \rightarrow
+  \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
+    function}, where $\overline{\vec{x}}$ is the set $\{\overline{x}
+  \mid x \in \vec{x}\}$.
+\end{compactitem}
+Intuitively, $\overline{\vec{x}}$ track the previous and $\vec{x}$ the
+current values of the variables of $\A$. Formally, let $\vec{x}_i =
+\set{x_i \mid x \in \vec{x}}$, $i\geq0$, be a time-stamped set of
+variables. For an input even $a \in \Sigma$ and a formula
+$\phi(\vec{x}_0,\ldots,\vec{x}_i)$ we write $\Delta(\phi,a)$ for the
+formula obtained from $\phi$ by simultaneously replacing each state $q
+\in \fv{\Bool}{\phi}$ by the formula
+$\Delta(q,a)[\vec{x}_i/\overline{\vec{x}},\vec{x}_{i+1}/\vec{x}]$, for
+all $i\geq0$.
+
+Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$, the
+\emph{run} of $\A$ over $w$ is the sequence of formulae
+$\phi_0,\phi_1,\ldots,\phi_n$, where $\phi_0 \equiv \bigwedge_{q \in
+  I} q$ and, for all $i > 0$, we have $\phi_i \equiv
+\Delta(\phi_{i-1},a_i)$. We slightly abuse notation and write
+$\Delta(\phi_0,a_1,\ldots,a_n)$ for
+$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ in the following. $\A$ 
+\emph{accepts} $w$ iff $\I,\nu \models \phi_n$, where $\nu$ is the
+valuation that maps:\begin{inparaenum}[(1)] 
+\item each $x \in \vec{x}_0$ to $\mathbf{0}$, 
+%
+\item each $x \in \vec{x}_i$ to $\nu_i(x)$, for all $i \in [1,n]$, 
+%
+\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$, and 
+%
+\item each $q \in \fv{\Bool}{\phi_n} \setminus F$ to $\bot$.
 \end{inparaenum}

### 2017-05-02 draft.tex
boolean closures

--- 2017-05-01 draft.tex
@@
 $\#(f) \geq 0$ arguments of sort $\Data$ and return value $\sigma(f)
-\in \ssorts{\sig}$. If $n=0$, $f$ is a \emph{constant} of sort
-$\sigma(f)$ and we consider the constants $\top,\bot \in \sfuns{\sig}$
-of sort $\Bool$, denoting the truth values true and false.
+\in \ssorts{\sig}$. If $\#(f)=0$ then $f$ is a \emph{constant}. We
+consider the constants $\mathbf{0}$ of sort $\Data$ and $\top,\bot$ of
+sort $\Bool$, denoting the truth values true and false.
 
@@
 where each $x \in \vars$ has an associated sort $\sigma(x)$. Terms are
-defined recursively: a term $t$ of sort $S$ is a variable $x \in
-\vars$ where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
+defined recursively: a term $t$ of sort $\sigma(t)=S$ is a variable $x
+\in \vars$ where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
 $t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
@@
 negation. A \emph{formula} is a combination of atoms using disjunction
-$\vee$, conjunction $\wedge$ and negation $\neg$.  A formula is in
-disjunctive normal form (DNF) if it is a disjunction of conjunctions
-of literals $\bigvee_{i=1}^N \bigwedge_{j=1}^{M_i} \ell_{ij}$.
+$\vee$, conjunction $\wedge$ and negation $\neg$. We denote by
+$\fv{S}{\phi}$ the set of variables of sort $S$ in $\phi$ and write
+$\fv{}{\phi}$ for $\bigcup_{S \in \set{\Data,\Bool}}
+\fv{S}{\phi}$. For a variable $x \in \fv{}{\phi}$ and a term $t$ such
+that $\sigma(t) = \sigma(x)$, we denote by $\phi[t/x]$ the result of
+replacing each occurrence of $x$ by $t$. For indexed sets
+$\vec{t}=\set{t_1,\ldots,t_n}$ and $\vec{x}=\set{x_1,\ldots,x_n}$, we
+write $\phi[\vec{t}/\vec{x}]$ for the formula obtained by
+simultaneously replacing $x_i$ with $t_i$ in $\phi$, for all
+$i=1,\ldots,n$. A formula is in positive normal form (PNF) if negation
+occurs only within literals and in disjunctive normal form (DNF) if it
+is a disjunction of conjunctions of literals $\bigvee_{i=1}^N
+\bigwedge_{j=1}^{M_i} \ell_{ij}$.
 
@@
 \item the sort $\Data$ into a non-empty set $\Data^\I$, 
-\item the $\Bool$ sort into the set $\Bool^\I$ containing the boolean
+%
+\item the sort $\Bool$ into the set $\Bool^\I$ containing the boolean
   values, and
+%
 \item each function symbol $f$ into a total function $f^\I :
-  (\Data^\I)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
+  \left(\Data^\I\right)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
   $\sigma(f)^I$ when $\#(f)=0$.
 \end{inparaenum}
+Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
+variable $x \in \vars$ into an element $\nu(x) \in \sigma(x)^\I$. For
+a term $t$, we denote by $t^\I_\nu$ the value obtained by replacing
+each function symbol $f$ by its interpretation $f^\I$ and each
+variable $x$ by its valuation $\nu(x)$. For a formula $\phi$, we write
+$\I,\nu \models \phi$ if the formula obtained by replacing each term
+$t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.
+A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
+there exists a valuation $\nu$ such that $\I,\nu \models \phi$. Given
+formulae $\varphi$ and $\psi$, we say that \emph{$\phi$ entails
+  $\psi$}, denoted $\phi \models^\I \psi$ iff $\I,\nu \models \varphi$
+implies $\I,\nu \models \psi$, for each valuation $\nu$, and $\phi
+\eqiff^\I \psi$ iff $\phi \models^\I \psi$ and $\psi \models^\I \phi$.
+We omit mentioning the interpretation $\I$ when it is clear from the
+context.
+
+\section{Alternating Data Automata}
+
+In the rest of this section we fix an interpretation $\I$ and a finite
+alphabet $\Sigma$ of \emph{input events}. Given a finite set $\vec{x}
+\subset \vars$ of variables of sort $\Data$, let $\vec{x} \mapsto
+\Data^\I$ be the set of valuations of the variables $\vec{x}$ and
+$\Sigma[\vec{x}] = \Sigma \times (\vec{x} \mapsto \Data^\I)$ be the
+set of \emph{data symbols}. A \emph{data word} (or, simply, a word) is
+a finite sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data
+symbols, where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n :
+\vec{x} \rightarrow \Data^\I$ are valuations. We denote by
+$\Sigma[\vec{x}]^*$ the set of data words over $\vec{x}$. This
+definition extends the classical notion of words from a finite
+alphabet with a possibly infinite alphabet $\Sigma[\vec{x}]$. Clearly,
+when $\Data^\I$ is sufficiently large or infinite, we can map the
+elements of $\Sigma$ into designated elements of $\Data^\I$ and use a
+special variable to encode the input events. However, keeping $\Sigma$
+explicit in the following simplifies several technical points below,
+without cluttering the presentation.
+
+Given sets of variables $\vec{b},\vec{x} \subset \vars$ of sort
+$\Bool$ and $\Data$, respectively, we denote by
+$\Form^+(\vec{b},\vec{x})$ the set of formulae $\phi$ such that
+$\fv{\Bool}{\phi} \subseteq \vec{b}$, $\fv{\Data}{\phi} \subseteq
+\vec{x}$ and each variable from $\vec{b}$ occurs under an even number
+of negations.
+
+An \emph{alternating data automaton} (or, simply, and automaton) is a
+tuple $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$: \begin{compactitem}
+%
+\item $\vec{x} \subset \vars$ is a finite set of variables of sort
+  $\Data$,
+%
+\item $Q$ is a finite set of variables of sort $\Bool$, also called
+  \emph{states},
+%
+\item $\iota \in \Form^+(Q,\emptyset)$ is a formula describing the
+  \emph{initial configurations},
+%
+\item $F \subseteq Q$ is a set of \emph{final} states, and
+%
+\item $\Delta : Q \times \Sigma \rightarrow
+  \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
+    function}, where $\overline{\vec{x}}$ is the set $\{\overline{x}
+  \mid x \in \vec{x}\}$.
+\end{compactitem}
+Intuitively, $\overline{\vec{x}}$ track the previous and $\vec{x}$ the
+current values of the variables of $\A$. Formally, let $\vec{x}_k =
+\set{x_k \mid x \in \vec{x}}$, $k\geq0$, be a time-stamped set of
+variables. For an input even $a \in \Sigma$ and a formula $\phi$ we
+write $\Delta^k(\phi,a)$ for the formula obtained from $\phi$ by
+simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
+formula
+$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
+all $k\geq0$.
+
+Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$, the
+\emph{run} of $\A$ over $w$ is the sequence of formulae
+$\phi_0,\phi_1,\ldots,\phi_n$, where $\phi_0 \equiv \iota$ and, for
+all $i > 0$, we have $\phi_k \equiv \Delta^k(\phi_{k-1},a_k)$. We
+slightly abuse notation and write $\Delta(\iota,a_1,\ldots,a_n)$ for
+the formula $\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ in the
+following. $\A$ \emph{accepts} $w$ iff $\I,\nu \models \phi_n$, where
+$\nu$ is the valuation that maps:\begin{inparaenum}[(1)]
+\item each $x \in \vec{x}_0$ to $\mathbf{0}$, 
+%
+\item each $x \in \vec{x}_k$ to $\nu_k(x)$, for all $k\in[1,n]$, 
+%
+\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$, and 
+%
+\item each $q \in \fv{\Bool}{\phi_n} \setminus F$ to $\bot$.
+\end{inparaenum}
+The language of $\A$, denoted as $L(\A)$, is the set of words accepted
+by $\A$.
+
+In this section we tackle the following problems: \begin{compactitem}
+\item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
+  with the same set of variables $\vec{x}$, does there exist automata
+  $\A_\cup$, $\A_\cap$ and $\overline{\A}$ such that $L(\A_\cup) =
+  \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and $L(\overline{\A_1})
+  = \Sigma[\vec{x}]^* \setminus L(\A_1)$ ?
+%
+\item \emph{emptiness}: given an automaton $\A$, is it the case that
+  $L(\A) = \emptyset$ ?
+\end{compactitem}
+
+It is well known that other problems, such as \emph{universality}
+(given an automaton $\A$ with variables $\vec{x}$, does $L(\A) =
+\Sigma[\vec{x}]^*$) and \emph{inclusion} (given automata $\A_1$ and
+$\A_2$ with the same set of variables, does $L(\A_1) \subseteq
+L(\A_2)$ ?) can be reduced to the above problems. Observe furthermore
+that we do not consider cases in which the sets of variables in the
+two automata differ. An interesting problem in this case is: given
+automata $\A_1$ and $\A_2$, with variables $\vec{x}_1$ and
+$\vec{x}_2$, respectively, such that $\vec{x}_1 \subseteq \vec{x}_2$,
+does $L(\A_1) \subseteq \proj{L(\A_2)}{\vec{x}_1}$, where
+$\proj{L(\A_2)}{\vec{x}_1}$ is the projection of the set of words
+$L(\A_2)$ onto the variables $\vec{x}_1$? This problem is considered
+as future work.
+
+\subsection{Boolean Closure}
+
+For a formula $\phi \in \Form^+(Q,\vec{x})$, with no negated
+occurrences of the boolean variables in $Q$, we define the
+formula $\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on
+the structure of $\phi$:
+\[\begin{array}{lclclclc}
+\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
+\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} \\
+\overline{\phi} & \equiv & \phi & \text{ if $\phi \in Q$}\hspace*{1cm} &
+\overline{\phi} & \equiv & \neg\phi & \text{if $\phi \not\in Q$ is an atom}
+\end{array}\]
+If $\phi$ does not contain variables from $Q$, then $\overline{\phi}$
+and $\neg\phi$ are equivalent, and in general:
+
+\begin{proposition}\label{prop:overline}
+  Given a formula $\phi \in \Form^+(Q,\vec{x})$ and a valuation $\nu$
+  mapping each $q \in Q$ a value $\nu(q) \in \Bool^\I$ and each $x \in
+  \vec{x}$ to a value $\nu(x) \in \Data^\I$, let $\nu'$ be the
+  valuation that assigns each $q \in Q$ the value $\neg\nu(q)$ and
+  each $x \in \vec{x}$ the value $\nu(x)$. Then we have 
+  $\I,\nu \models \phi \iff \I,\nu' \not\models \overline{\phi}$. 
+\end{proposition}
+\proof{Immediate, by induction on the structure of $\phi$. \qed}
+
+In the following let $\A_1 = \tuple{\vec{x},Q_1,\iota_1,F_1,\Delta_1}$
+and $\A_2 = \tuple{\vec{x},Q_2,\iota_2,F_2,\Delta_2}$, where
+w.l.o.g. we assume that $Q_1 \cap Q_2 = \emptyset$. We define $\A_\cup
+= \tuple{\vec{x},Q_1\cup Q_2,\iota_1 \vee \iota_2,F_1\cup
+  F_2,\Delta_1\cup\Delta_2}$, $\A_\cap = \tuple{\vec{x},Q_1\cup
+  Q_2,\iota_1 \wedge \iota_2, F_1\cup F_2, \Delta_1\cup\Delta_2}$ and
+$\overline{\A_1} = \langle \vec{x},Q_1,\iota_1,Q_1\setminus
+F_1,\overline{\Delta_1} \rangle$, where $\overline{\Delta_1}(q,a)
+\equiv \overline{\Delta_1(q,a)}$, for all $q \in Q_1$ and $a \in
+\Sigma$. The following lemma shows the correctness of these
+definitions:
+
+\begin{lemma}\label{lemma:closure}
+  For any automata $\A_1 = \tuple{\vec{x},Q_1,\iota_1,F_1,\Delta_1}$
+  and $\A_2 = \tuple{\vec{x},Q_2,\iota_2,F_2,\Delta_2}$ such that $Q_1
+  \cap Q_2 = \emptyset$, we have $L(\A_\cup) = L(\A_1) \cup L(\A_2)$,
+  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$ and $L(\overline{\A_1}) =
+  \Sigma[\vec{x}]^* \setminus L(\A_1)$.
+\end{lemma}
+\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
+  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
+  (a_n,\nu_n) \in L(\A_\cup)$ be a word, where $n=0$ corresponds to
+  the empty word. We prove by induction on $n\geq0$ that
+  $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n) \eqiff
+  \Delta(\iota_1,a_1\ldots a_n) \vee \Delta(\iota_2,a_1\ldots
+  a_n)$. The case $n=0$ follows from the definition of the initial
+  configuration of $\A_\cup$. For the inductive step $n>0$,
+  $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$ is obtained from
+  $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by replacing each
+  variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee \iota_2,a_1\ldots
+    a_{n-1})}$ with
+  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
+  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
+  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
+  \iota_2,a_1\ldots a_{n-1}) \eqiff \Delta(\iota_1,a_1\ldots a_{n-1})
+  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain: 
+  \[\begin{array}{rcl}
+  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iff & 
+  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee
+  \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) \\
+  & \iff & \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots
+  a_{n}) \enspace.
+  \end{array}\] 
+  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
+  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
+  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
+  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
+  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
+  \subseteq Q$ and thus $\overline{\iota_1} = \iota_1$. For the case
+  $n>0$, we compute:
+  \[\begin{array}{rcll}
+  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \eqiff & \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \text{ by induction hypothesis} \\
+  & \eqiff & \overline{\Delta(\iota_1,a_1\ldots,a_n)} & \text{ by the definition of $\overline{\phi}$ .}
+  \end{array}\]
+  Let $\nu,\nu' : Q \cup \bigcup_{i=0}^n\vec{x}_i$ be two valuations
+  such that: \begin{compactitem}
+  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
+    %
+  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
+    %
+  \item $\nu(x) = \nu'(x) = \mathbf{0}$, for each $x \in \vec{x}_0$, 
+    %
+  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
+    \in [1,n]$.
+  \end{compactitem}
+  By Proposition \ref{prop:overline}, we have $\I,\nu \models
+  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
+  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
+  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus $w \in L(\A_1) \iff
+  w \not\in L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) =
+  \Sigma[\vec{x}]^* \setminus L(\A_1)$. \qed}
 

### 2017-05-02 draft.tex
ok

@@
 
+\section{Predicate Abstraction for Emptiness}
+
+Unlike the boolean closure properties, showed to be effectively
+decidable (Lemma \ref{lemma:closure}), the emptiness problem is
+undecidable, even in very simple cases. For instance, if $\Data^\I$ is
+the set of positive integers, an ADA can simulate an Alternating
+Vector Addition System with States (AVASS) \cite{LazicSchmitz14} using
+only atoms $x = \overline{x} + k$, for $k \in \nat$ and $x \teq 0$,
+with the classical interpretation of function symbols on
+integers. Since reachability of a configuration $(q,\nu)$, where $q$
+is a control state and $\nu$ a valuation of the variables, is
+undecidable, the emptiness problem is undecidable for ADA, in general.
+
+We thus turn to semi-algorithms for this problem and develop methods
+that are supposed to be both: \begin{compactenum}
+\item \emph{sound}: given an automaton $\A$, if $L(\A) \neq\emptyset$,
+  the procedure will terminate and return a word $w \in L(\A)$), and
+%
+\item \emph{complete relative to termination}: if the procedure terminates
+  without returning such a word, then $L(\A) = \emptyset$. 
+\end{compactenum}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\bibliographystyle{abbrv} \bibliography{refs}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-05-02 draft.tex
ok

@@
 is a disjunction of conjunctions of literals $\bigvee_{i=1}^N
-\bigwedge_{j=1}^{M_i} \ell_{ij}$.
+\bigwedge_{j=1}^{M_i} \ell_{ij}$. The size of a formula $\phi$,
+denoted $\len{\phi}$ is the number of symbols occuring in it.
 
@@
 $\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
-all $k\geq0$.
+all $k\geq0$. The size of $\A$ is defined as $\len{\A} = \len{\iota} +
+\sum_{(q,a) \in Q\times\Sigma}\len{\Delta(q,a)}$.
 
@@
 
-For a formula $\phi \in \Form^+(Q,\vec{x})$, with no negated
-occurrences of the boolean variables in $Q$, we define the
-formula $\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on
-the structure of $\phi$:
+Given a set $Q$ of boolean variables and a set $\vec{x}$ of variables
+of sort $\Data$, for a formula $\phi \in \Form^+(Q,\vec{x})$, with no
+negated occurrences of the boolean variables, we define the formula
+$\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on the structure
+of $\phi$:
 \[\begin{array}{lclclclc}
@@
 \end{array}\]
-If $\phi$ does not contain variables from $Q$, then $\overline{\phi}$
-and $\neg\phi$ are equivalent, and in general:
+It is easy to check that $\len{\overline{\phi}} = \len{\phi}$, for
+every formula $\phi \in \Form^+(Q,\vec{x})$. If $\phi$ does not
+contain boolean variables, then $\overline{\phi}$ and $\neg\phi$ are
+equivalent, and in general:
 
@@
   \[\begin{array}{rcll}
-  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \eqiff & \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \text{ by induction hypothesis} \\
+  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \eqiff & 
+  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \text{ by induction hypothesis} \\
   & \eqiff & \overline{\Delta(\iota_1,a_1\ldots,a_n)} & \text{ by the definition of $\overline{\phi}$ .}
@@
 
+It is easy to see that $\len{\A_\cup} = \len{\A_\cap} = \len{\A_1} +
+\len{\A_2}$ and $\len{\overline{\A}} = \len{\A}$, thus the automata
+for the boolean operations, including complementation, can be
+implemented in linear time. This is not very surprising, since linear
+time intersection and complementation are possible for alternating
+automata over finite alphabets \cite{ChandraKozenStockmeyer81}.
+
 \section{Predicate Abstraction for Emptiness}
@@
 
-We thus turn to semi-algorithms for this problem and develop methods
-that are supposed to be both: \begin{compactenum}
+Consequently, we give up on the guarantee for termination and develop
+semi-algorithmic methods that are supposed to be
+both: \begin{compactenum}
 \item \emph{sound}: given an automaton $\A$, if $L(\A) \neq\emptyset$,
@@
 \end{compactenum}
+
+Let us fix an automaton $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ whose
+finite input alphabet is $\Sigma$, for the rest of this section. Given
+a finite sequence $u = a_1\ldots a_n \in \Sigma^*$ of input events and
+$\phi \in \Form^+(Q,\vec{x})$, we define $\Post(\phi,u) \equiv \exists
+\vec{x}_0 \ldots \exists \vec{x}_{n-1} ~.~
+\Delta(\phi,u)[\vec{x}/\vec{x}_n]$.
 

### 2017-05-03 draft.tex
ok

@@
 negation. A \emph{formula} is a combination of atoms using disjunction
-$\vee$, conjunction $\wedge$ and negation $\neg$. We denote by
-$\fv{S}{\phi}$ the set of variables of sort $S$ in $\phi$ and write
-$\fv{}{\phi}$ for $\bigcup_{S \in \set{\Data,\Bool}}
+$\vee$, conjunction $\wedge$ and negation $\neg$ and we write $\phi
+\rightarrow \psi$ for $\neg\phi \vee \psi$.
+
+We denote by $\fv{S}{\phi}$ the set of variables of sort $S$ in $\phi$
+and write $\fv{}{\phi}$ for $\bigcup_{S \in \set{\Data,\Bool}}
 \fv{S}{\phi}$. For a variable $x \in \fv{}{\phi}$ and a term $t$ such
@@
 %
-\item the sort $\Bool$ into the set $\Bool^\I$ containing the boolean
-  values, and
+\item the sort $\Bool$ into the set $\booli = \set{\true,\false}$, where
+  $\top^\I = \true$, $\bot^\I = \false$, and
 %
@@
 \vec{x} \rightarrow \Data^\I$ are valuations. We denote by
-$\Sigma[\vec{x}]^*$ the set of data words over $\vec{x}$. This
-definition extends the classical notion of words from a finite
-alphabet with a possibly infinite alphabet $\Sigma[\vec{x}]$. Clearly,
-when $\Data^\I$ is sufficiently large or infinite, we can map the
-elements of $\Sigma$ into designated elements of $\Data^\I$ and use a
-special variable to encode the input events. However, keeping $\Sigma$
-explicit in the following simplifies several technical points below,
-without cluttering the presentation.
+$\varepsilon$ the empty sequence, by $\Sigma^*$ the set of finite
+sequences of input events and by $\Sigma[\vec{x}]^*$ the set of data
+words over $\vec{x}$. This definition extends the classical notion of
+words from a finite alphabet with a possibly infinite alphabet
+$\Sigma[\vec{x}]$. Clearly, when $\Data^\I$ is sufficiently large or
+infinite, we can map the elements of $\Sigma$ into designated elements
+of $\Data^\I$ and use a special variable to encode the input
+events. However, keeping $\Sigma$ explicit in the following simplifies
+several technical points below, without cluttering the presentation.
 
@@
 
-In this section we tackle the following problems: \begin{compactitem}
+In this section we tackle the following problems: \begin{compactenum}
 \item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
@@
   $L(\A) = \emptyset$ ?
-\end{compactitem}
+\end{compactenum}
 
@@
   Given a formula $\phi \in \Form^+(Q,\vec{x})$ and a valuation $\nu$
-  mapping each $q \in Q$ a value $\nu(q) \in \Bool^\I$ and each $x \in
+  mapping each $q \in Q$ a value $\nu(q) \in \booli$ and each $x \in
   \vec{x}$ to a value $\nu(x) \in \Data^\I$, let $\nu'$ be the
@@
 
-\section{Predicate Abstraction for Emptiness}
+\section{Antichains for Emptiness}
 
@@
 Let us fix an automaton $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ whose
-finite input alphabet is $\Sigma$, for the rest of this section. Given
-a finite sequence $u = a_1\ldots a_n \in \Sigma^*$ of input events and
-$\phi \in \Form^+(Q,\vec{x})$, we define $\Post(\phi,u) \equiv \exists
-\vec{x}_0 \ldots \exists \vec{x}_{n-1} ~.~
-\Delta(\phi,u)[\vec{x}/\vec{x}_n]$.
+(finite) input event alphabet is $\Sigma$, for the rest of this
+section. Given a formula $\phi \in \Form^+(Q,\vec{x})$ and an input
+event $a \in \Sigma$, we define $\Post{\A}(\phi,a) \equiv \exists
+\overline{\vec{x}} ~.~ \Delta(\phi[\overline{\vec{x}}/\vec{x}], a) \in
+\Form^+(Q,\vec{x})$. We generalize $\Post{\A}$ to finite sequences of
+input events: $\Post{\A}(\phi,\varepsilon) \equiv \phi$ and
+$\Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a)$, for any $u
+\in \Sigma^*$. Then the emptiness problem for $\A$ becomes: does there
+exist a sequence $u \in \Sigma^*$ such that the formula $\Accept{\A}(u)
+\equiv \Post{\A}(\iota,u) \wedge \bigwedge_{q \in F} \top \rightarrow
+q \wedge \bigwedge_{q \in Q \setminus F} q \rightarrow \bot$ is
+satisfiable ? A na\"ive semi-algorithm then enumerates all finite
+sequences and checks for the satisfiability of $\Accept{\A}(u)$ for
+each $u \in \Sigma^*$.
+
+Since no boolean variable from $Q$ occurs under negation in $\phi$, it
+is easy to prove the following monotonicity property: given two
+formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
+then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
+\Sigma^*$. This suggest an improvement of the above semi-algorithm,
+that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
+$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
+antichain\footnote{Given a partial order $(D,\preceq)$ an antichain is
+  a set $A \subseteq D$ such that $a \not\preceq b$ for any $a,b \in
+  A$.} w.r.t. the entailment partial order. This is because, for any
+$u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models \Post{\A}(\iota,v)$
+and $\Accept{\A}(uw)$ is satisfiable, then $\Accept{\A}(vw)$ is
+satisfiable as well, and there is no need to store
+$\Post{\A}(\iota,u)$, since the non-emptiness of $\A$ can be proved
+using $v$ alone. However, in practice, the semi-algorithm using
+antichains diverges in a large number of cases, because, in many
+interpretations, there exist infinite antichains\footnote{For
+  instance, when $\Data^\I = \nat$, an infinite antichain is $q \wedge x \teq
+  0, q \wedge x \teq 1, \ldots$} $\set{\Post{\A}(\phi,u) \mid u \in U}$, where
+$U$ is an infinite subset of $\Sigma^*$.
+
+A \emph{safety invariant} for $\A$ is a function $I : (Q \mapsto
+\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
+boolean valuation $\beta : Q \rightarrow \booli$, the following
+hold: \begin{compactitem}
+\item if $\beta \models \iota$ then $\nu \in I(\beta)$ iff
+  $\nu(x)=\mathbf{0}$ for all $x \in \vec{x}$,
+%
+\item for each $u \in \Sigma^*$, if $\beta \cup \nu \models
+  \Post{\A}(\iota,u)$ then $\nu \in I(\beta)$, and
+%
+\item 
+\end{compactitem}
 

### 2017-05-03 draft.tex
ok

@@
 sequences of input events and by $\Sigma[\vec{x}]^*$ the set of data
-words over $\vec{x}$. This definition extends the classical notion of
-words from a finite alphabet with a possibly infinite alphabet
-$\Sigma[\vec{x}]$. Clearly, when $\Data^\I$ is sufficiently large or
-infinite, we can map the elements of $\Sigma$ into designated elements
-of $\Data^\I$ and use a special variable to encode the input
-events. However, keeping $\Sigma$ explicit in the following simplifies
-several technical points below, without cluttering the presentation.
+words over $\vec{x}$.
+
+This definition extends the classical notion of words from a finite
+alphabet with a possibly infinite alphabet $\Sigma[\vec{x}]$. Clearly,
+when $\Data^\I$ is sufficiently large or infinite, we can map the
+elements of $\Sigma$ into designated elements of $\Data^\I$ and use a
+special variable to encode the input events. However, keeping $\Sigma$
+explicit in the following simplifies several technical points below,
+without cluttering the presentation.
 
@@
 \len{\A_2}$ and $\len{\overline{\A}} = \len{\A}$, thus the automata
-for the boolean operations, including complementation, can be
-implemented in linear time. This is not very surprising, since linear
-time intersection and complementation are possible for alternating
+for the boolean operations, including complementation, can be built in
+linear time. This is not very surprising, since linear time
+intersection and complementation are also possible for alternating
 automata over finite alphabets \cite{ChandraKozenStockmeyer81}.
 
-\section{Antichains for Emptiness}
+\section{Antichains and Abstraction Refinement for Emptiness}
 
@@
 $\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
-antichain\footnote{Given a partial order $(D,\preceq)$ an antichain is
-  a set $A \subseteq D$ such that $a \not\preceq b$ for any $a,b \in
-  A$.} w.r.t. the entailment partial order. This is because, for any
-$u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models \Post{\A}(\iota,v)$
-and $\Accept{\A}(uw)$ is satisfiable, then $\Accept{\A}(vw)$ is
-satisfiable as well, and there is no need to store
-$\Post{\A}(\iota,u)$, since the non-emptiness of $\A$ can be proved
-using $v$ alone. However, in practice, the semi-algorithm using
+\emph{antichain}\footnote{Given a partial order $(D,\preceq)$ an
+  antichain is a set $A \subseteq D$ such that $a \not\preceq b$ for
+  any $a,b \in A$.} w.r.t. the entailment partial order. This is
+because, for any $u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models
+\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable, then
+$\Accept{\A}(vw)$ is satisfiable as well, and there is no need to
+store $\Post{\A}(\iota,u)$, since the non-emptiness of $\A$ can be
+proved using $v$ alone. However, in practice, the semi-algorithm using
 antichains diverges in a large number of cases, because, in many
 interpretations, there exist infinite antichains\footnote{For
-  instance, when $\Data^\I = \nat$, an infinite antichain is $q \wedge x \teq
-  0, q \wedge x \teq 1, \ldots$} $\set{\Post{\A}(\phi,u) \mid u \in U}$, where
-$U$ is an infinite subset of $\Sigma^*$.
+  instance, when $\Data^\I = \nat$, an infinite antichain is $q \wedge
+  x \teq 0, q \wedge x \teq 1, \ldots$} $\set{\Post{\A}(\phi,u_i) \mid
+  u_1,u_2, \ldots \in \Sigma^*}$.
 
-A \emph{safety invariant} for $\A$ is a function $I : (Q \mapsto
+A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
 \booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
-boolean valuation $\beta : Q \rightarrow \booli$, the following
-hold: \begin{compactitem}
-\item if $\beta \models \iota$ then $\nu \in I(\beta)$ iff
-  $\nu(x)=\mathbf{0}$ for all $x \in \vec{x}$,
+boolean valuation $\beta : Q \rightarrow \booli$, every valuation
+$\nu : \vec{x} \mapsto \Data^\I$ of the data
+variables and every finite sequence $u \in \Sigma^*$ of input events: \begin{compactitem}
+%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
+%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
 %
-\item for each $u \in \Sigma^*$, if $\beta \cup \nu \models
-  \Post{\A}(\iota,u)$ then $\nu \in I(\beta)$, and
+\item $\beta \cup \nu \models \Post{\A}(\iota,u) \Rightarrow \nu \in
+  \Inv(\beta)$, and
 %
-\item 
+\item $\nu \in \Inv(\beta) \Rightarrow \beta \cup \nu \not\models
+  \Accept{\A}(u)$.
 \end{compactitem}
+Intuitively, a safety invariant maps every boolean valuation into a
+set of data valuations, that contains the initial configuration (in
+which all data variables are assumed to be $\mathbf{0}\in\Data^\I$),
+over-approximates the set of reachable valuations and excludes those
+valuations that satisfying the acceptance condition.
+
+\begin{lemma}\label{lemma:safety-invariant}
+  For any automaton $\A$, we have $L(\A)
+= \emptyset$ iff $\A$ has a safety invariant. 
+\end{lemma}
+\proof{ Let $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ in the following.
+  ``$\Leftarrow$'' This direction is trivial. ``$\Rightarrow$'' We
+  define a function $\Inv : (Q \mapsto \booli) \rightarrow 2^{\vec{x}
+    \mapsto \Data^\I}$ as: for each $\beta : Q \rightarrow \booli$,
+  let $\Inv(\beta) = \{\nu : \vec{x} \rightarrow \Data^\I \mid \exists
+  u \in \Sigma^* ~.~ \beta\cup\nu \models
+  \Post{\A}(\iota,u)\}$. Checking that $\Inv$ is indeed a safety
+  invariant is straightforward. \qed}
+
+Turning back to the issue of divergence of language emptiness
+semi-algorithms in the case $L(\A) = \emptyset$, we can observe that
+an enumeration of input sequences $u_1,u_2,\ldots \in \Sigma^*$ can
+stop at step $k$ as soon as $\bigvee_{i=1}^k \Post(\iota,u_i)$ defines
+a safety invariant for $\A$. Although this condition can be
+effectively checked using a decision procedure for the quantifier-free
+first order theory $\theory(\sig,\I)$, of signature $\sig$ and
+interpretation $\I$, there is no guarantee that this check will ever
+succeed.
 

### 2017-05-04 draft.tex
ok

@@
 sorted \emph{function symbols}. To simplify the presentation, we
-assume w.l.o.g. that $\ssorts{\Sigma}$ consists of two distinct sorts
+assume w.l.o.g. that $\ssorts{\sig}$ consists of two distinct sorts
 $\Data$ and $\Bool$ and each function symbol $f \in \sfuns{\sig}$ has
@@
 consider the constants $\mathbf{0}$ of sort $\Data$ and $\top,\bot$ of
-sort $\Bool$, denoting the truth values true and false.
+sort $\Bool$.
 
@@
 \item each function symbol $f$ into a total function $f^\I :
-  \left(\Data^\I\right)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
+  (\Data^\I)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
   $\sigma(f)^I$ when $\#(f)=0$.
@@
 A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
-there exists a valuation $\nu$ such that $\I,\nu \models \phi$. Given
-formulae $\varphi$ and $\psi$, we say that \emph{$\phi$ entails
+there exists a valuation $\nu$ such that $\I,\nu \models \phi$. The
+\emph{theory} $\theory(\sig,\I)$ is the set of satisfiable formulae
+written in the signature $\sig$, with the interpretation $\I$.
+
+Given formulae $\varphi$ and $\psi$, we say that \emph{$\phi$ entails
   $\psi$}, denoted $\phi \models^\I \psi$ iff $\I,\nu \models \varphi$
@@
 
-This definition extends the classical notion of words from a finite
-alphabet with a possibly infinite alphabet $\Sigma[\vec{x}]$. Clearly,
-when $\Data^\I$ is sufficiently large or infinite, we can map the
-elements of $\Sigma$ into designated elements of $\Data^\I$ and use a
-special variable to encode the input events. However, keeping $\Sigma$
-explicit in the following simplifies several technical points below,
-without cluttering the presentation.
+This definition generalizes the classical notion of words from a
+finite alphabet to using a possibly infinite alphabet
+$\Sigma[\vec{x}]$. Clearly, when $\Data^\I$ is sufficiently large or
+infinite, we can map the elements of $\Sigma$ into designated elements
+of $\Data^\I$ and use a special variable to encode the input
+events. However, keeping $\Sigma$ explicit in the following simplifies
+several technical points below, without cluttering the presentation.
 
@@
 \item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
-  with the same set of variables $\vec{x}$, does there exist automata
+  with the same set of variables $\vec{x}$, do there exist automata
   $\A_\cup$, $\A_\cap$ and $\overline{\A}$ such that $L(\A_\cup) =
-  \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and $L(\overline{\A_1})
-  = \Sigma[\vec{x}]^* \setminus L(\A_1)$ ?
+  \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and
+  $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$ ?
 %
@@
 of $\phi$:
-\[\begin{array}{lclclclc}
+\[\begin{array}{lclclclclcl}
 \overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
-\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} \\
-\overline{\phi} & \equiv & \phi & \text{ if $\phi \in Q$}\hspace*{1cm} &
-\overline{\phi} & \equiv & \neg\phi & \text{if $\phi \not\in Q$ is an atom}
+\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} && 
+\overline{\neg\phi} & \equiv & \neg \overline{\phi} \text{ if $\phi$ is not an atom} \\
+\overline{\phi} & \equiv & \phi \text{  if $\phi \in Q$}\hspace*{1cm} &&
+\overline{\phi} & \equiv & \neg\phi \text{  if $\phi \not\in Q$ is an atom}
 \end{array}\]
@@
   for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
-  (a_n,\nu_n) \in L(\A_\cup)$ be a word, where $n=0$ corresponds to
-  the empty word. We prove by induction on $n\geq0$ that
-  $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n) \eqiff
-  \Delta(\iota_1,a_1\ldots a_n) \vee \Delta(\iota_2,a_1\ldots
-  a_n)$. The case $n=0$ follows from the definition of the initial
-  configuration of $\A_\cup$. For the inductive step $n>0$,
-  $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$ is obtained from
-  $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by replacing each
-  variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee \iota_2,a_1\ldots
-    a_{n-1})}$ with
+  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
+  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
+  \iota_2,a_1\ldots a_n) \eqiff \Delta(\iota_1,a_1\ldots a_n) \vee
+  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
+  definition of the initial configuration of $\A_\cup$. For the
+  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
+  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
+  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
+    \iota_2,a_1\ldots a_{n-1})}$ with
   $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
@@
 
-\section{Antichains and Abstraction Refinement for Emptiness}
+\section{Antichains and Interpolants for Emptiness}
 
@@
 with the classical interpretation of function symbols on
-integers. Since reachability of a configuration $(q,\nu)$, where $q$
-is a control state and $\nu$ a valuation of the variables, is
-undecidable, the emptiness problem is undecidable for ADA, in general.
+integers. Since reachability of a control state is undecidable for
+AVASS, the emptiness problem is undecidable for ADA.
 
-Consequently, we give up on the guarantee for termination and develop
+Consequently, we give up on the termination guarantee and develop
 semi-algorithmic methods that are supposed to be
@@
 section. Given a formula $\phi \in \Form^+(Q,\vec{x})$ and an input
-event $a \in \Sigma$, we define $\Post{\A}(\phi,a) \equiv \exists
-\overline{\vec{x}} ~.~ \Delta(\phi[\overline{\vec{x}}/\vec{x}], a) \in
-\Form^+(Q,\vec{x})$. We generalize $\Post{\A}$ to finite sequences of
-input events: $\Post{\A}(\phi,\varepsilon) \equiv \phi$ and
+event $a \in \Sigma$, we define the \emph{post-image} function
+$\Post{\A}(\phi,a) \equiv \exists \overline{\vec{x}} ~.~
+\Delta(\phi[\overline{\vec{x}}/\vec{x}], a)$, mapping each formula in
+$\Form^+(Q,\vec{x})$ to a formula, also in $\Form^+(Q,\vec{x})$, that
+defines the effect of firing a transition with input event $a$.
+
+We generalize the post-image function to finite sequences of input
+events, as follows: $\Post{\A}(\phi,\varepsilon) \equiv \phi$ and
 $\Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a)$, for any $u
 \in \Sigma^*$. Then the emptiness problem for $\A$ becomes: does there
-exist a sequence $u \in \Sigma^*$ such that the formula $\Accept{\A}(u)
-\equiv \Post{\A}(\iota,u) \wedge \bigwedge_{q \in F} \top \rightarrow
-q \wedge \bigwedge_{q \in Q \setminus F} q \rightarrow \bot$ is
-satisfiable ? A na\"ive semi-algorithm then enumerates all finite
-sequences and checks for the satisfiability of $\Accept{\A}(u)$ for
-each $u \in \Sigma^*$.
+exist $u \in \Sigma^*$ such that the formula $\Accept{\A}(u) \equiv
+\Post{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q
+\rightarrow \bot)$ is satisfiable ? A na\"ive semi-algorithm
+enumerates all finite sequences and checks the satisfiability of
+$\Accept{\A}(u)$ for each $u \in \Sigma^*$, using a decision procedure
+for the theory $\theory(\sig,\I)$.
 
@@
 \Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable, then
-$\Accept{\A}(vw)$ is satisfiable as well, and there is no need to
-store $\Post{\A}(\iota,u)$, since the non-emptiness of $\A$ can be
-proved using $v$ alone. However, in practice, the semi-algorithm using
-antichains diverges in a large number of cases, because, in many
-interpretations, there exist infinite antichains\footnote{For
-  instance, when $\Data^\I = \nat$, an infinite antichain is $q \wedge
-  x \teq 0, q \wedge x \teq 1, \ldots$} $\set{\Post{\A}(\phi,u_i) \mid
-  u_1,u_2, \ldots \in \Sigma^*}$.
+$\Post{\A}(\iota,uw) \models \Post{\A}(\iota,vw)$, thus
+$\Accept{\A}(vw)$ is satisfiable as well, and there is no need for
+$u$, since the non-emptiness of $\A$ can be proved using $v$
+alone. However, in practice, the antichain based semi-algorithm
+diverges in a large number of cases, since, in many interpretations,
+there exist infinite antichains\footnote{For instance, when $\Data^\I
+  = \nat$, an infinite antichain is $q \wedge x \teq 0, q \wedge x
+  \teq 1, \ldots$}.
 
@@
 $\nu : \vec{x} \mapsto \Data^\I$ of the data
-variables and every finite sequence $u \in \Sigma^*$ of input events: \begin{compactitem}
+variables and every finite sequence $u \in \Sigma^*$ of input events: \begin{compactenum}
 %% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
@@
 %
-\item $\beta \cup \nu \models \Post{\A}(\iota,u) \Rightarrow \nu \in
-  \Inv(\beta)$, and
+\item\label{it1:inv} $\beta \cup \nu \models \Post{\A}(\iota,u)
+  \Rightarrow \nu \in \Inv(\beta)$, and
 %
-\item $\nu \in \Inv(\beta) \Rightarrow \beta \cup \nu \not\models
-  \Accept{\A}(u)$.
-\end{compactitem}
-Intuitively, a safety invariant maps every boolean valuation into a
-set of data valuations, that contains the initial configuration (in
-which all data variables are assumed to be $\mathbf{0}\in\Data^\I$),
-over-approximates the set of reachable valuations and excludes those
-valuations that satisfying the acceptance condition.
+\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \beta \cup \nu
+  \not\models \Accept{\A}(u)$.
+\end{compactenum}
+If $\Inv$ satisfies only the first point above, we call it an
+invariant. Intuitively, a safety invariant maps every boolean
+valuation into a set of data valuations, that contains the initial
+configuration (in which all data variables are assumed to be
+$\mathbf{0}\in\Data^\I$), over-approximates the set of reachable
+valuations and excludes those valuations that satisfying the
+acceptance condition.
 
@@
 an enumeration of input sequences $u_1,u_2,\ldots \in \Sigma^*$ can
-stop at step $k$ as soon as $\bigvee_{i=1}^k \Post(\iota,u_i)$ defines
-a safety invariant for $\A$. Although this condition can be
-effectively checked using a decision procedure for the quantifier-free
-first order theory $\theory(\sig,\I)$, of signature $\sig$ and
-interpretation $\I$, there is no guarantee that this check will ever
-succeed.
+stop at step $k$ as soon as $\bigvee_{i=1}^k \Post{\A}(\iota,u_i)$
+defines\footnote{ A formula $\phi(Q,\vec{x})$ defines $\Inv$ iff for
+  all $\beta: Q \rightarrow \booli$ and $\nu : \vec{x} \rightarrow
+  \Data^\I$: $\beta\cup\nu \models \phi \iff \nu \in \Inv(\beta)$. } a
+safety invariant for $\A$. Although this condition can be effectively
+checked using a decision procedure for the theory $\theory(\sig,\I)$,
+there is no guarantee that this check will ever succeed.
+
+The solution we adopt is the use of \emph{abstraction} to ensure
+termination of invariant computations. However, it is worth pointing
+out from the start that abstraction will allow us to build invariants
+which are not necessarily safety invariants. To converge towards
+meeting the latter condition, we shall resort to \emph{counterexample
+  guided abstraction refinement} (CEGAR).
+
+Formally, let $\preds$ be a finite set of formulae with variables from
+$Q \cup \vec{x}$, such that $\bot \in \preds$ and the variables in $Q$
+can occur negated in some $\phi \in \preds$. We assume that $\preds$
+is fixed for now, and refer to these formulae as \emph{predicates}.
+Given a formula $\phi$, we denote by $\abs{\phi} \equiv \bigwedge
+\set{\pi \in \preds \mid \phi \models \pi}$ the abstraction of $\phi$
+w.r.t. the predicates in $\preds$. The abstract version of the
+post-image and acceptance formulae are defined as follows:
+{\footnotesize\[
+\abs{\Post{\A}}(\phi,\varepsilon) \equiv \phi 
+\hspace*{2mm}
+\abs{\Post{\A}}(\phi,ua) \equiv \abs{(\Post{\A}(\abs{\Post{\A}}(\phi,u),a))} 
+\hspace*{2mm}
+\abs{\Accept{\A}}(u) \equiv \abs{\Post{\A}}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q \rightarrow \bot)
+\]}
+\begin{lemma}\label{lemma:abstract-invariant}
+  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
+  $k>0$ such that $\bigvee_{i=1}^k \abs{\Post{\A}}(\iota,\mu(i))$ defines
+  an invariant for $\A$. 
+\end{lemma}
+\proof{ It is sufficient to show that there exists $k>0$ such that for
+  all $u \in \Sigma^*$ there exists $i \in [1,k]$ such that
+  $\Post{\A}(\iota,u) \models \abs{\Post{\A}}(\iota,\mu(i))$. We have
+  $\Post{\A}(\iota,u) \models \abs{\Post{\A}}(\iota,u)$ for all $u \in
+  \Sigma^*$. But since $\preds$ is a finite set, also the set
+  $\{\abs{\Post{\A}}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
+  there exists $k > 0$ such that, for all $u \in \Sigma^*$ there
+  exists $i \in [1,k]$ such that $\abs{\Post{\A}}(\iota,u) \eqiff
+  \abs{\Post{\A}}(\iota,\mu(i))$, which concludes the proof. \qed}
+
+We are left with fulfilling the second point from the definition of a
+safety invariant. To this end, suppose that, for a given set $\preds$
+of predicates, the invariant $\abs{\Inv}$, defined by the previous
+lemma, meets point (\ref{it1:inv}) but not point (\ref{it2:inv}) from
+the definition of invariants, where $\Post{\A}$ and $\Accept{\A}$
+replace $\abs{\Post{\A}}$ and $\abs{\Accept{\A}}$, respectively. In
+other words, there exists a finite sequence $u \in \Sigma^*$ such that
+$\nu \in \abs{\Inv}(\beta)$ and $\beta\cup\nu \models
+\abs{\Accept{\A}}(u)$, for some boolean $\beta : Q \rightarrow \booli$
+and data $\nu : \vec{x} \rightarrow \Data^\I$ valuations. Such
+a $u\in\Sigma^*$ is called a \emph{counterexample}.
+
+Once a counterexample $u$ is discovered, there are two
+possibilities. Either\begin{inparaenum}[(i)]
+\item $\Accept{\A}(u)$ is satisfiable, in which case $L(\A) \neq \emptyset$, or
+\item $\Accept{\A}(u)$ is unsatisfiable, in which case $u$ is called
+  \emph{spurious}.
+\end{inparaenum}
+In the first case, our semi-algorithm stops and returns a witness for
+non-emptiness, obtained from the satisfying valuation of
+$\Accept{\A}(u)$ and in the second case, we must strenghten the
+invariant by excluding from $\abs{\Inv}$ all pairs $(\beta,\nu)$ such
+that $\beta\cup\nu \models \abs{\Accept{\A}}(u)$. This strenghtening
+is carried out by adding to $\preds$ predicates sufficient to exclude
+the spurious counterexample.
+
+In general, given an unsatisfiable conjunction $\pi \equiv
+\phi_1(X_0,X_1) \wedge \phi_2(X_1,X_2) \wedge \ldots \wedge
+\phi_n(X_{n-1},X_n)$ of time-stamped variables $X_i = \set{x_i \mid x
+  \in X}$ of any sort, an \emph{interpolant} for $\pi$ is a tuple
+$\tuple{I_0(X), I_1(X), \ldots, I_n(X)}$ such
+that:\begin{inparaenum}[(i)]
+\item $I_0 \equiv \top$,  
+%
+\item $I_{i-1}[X_{i-1}/X] \wedge \phi_i(X_{i-1},X_i) \models
+  I_i[X_i/X]$, for all $i \in [1,n]$, and
+%
+\item $I_n \equiv \bot$. 
+\end{inparaenum}
+A classical method for abstraction refinement is to add the elements
+of the interpolant obtained from a proof of spuriousness to the set of
+predicates. This guarantees progress, meaning that the particular
+spurious counterexample, from which the interpolant was generated,
+will never be revisited in the future. In many practical test cases,
+this progress property eventually yields a safety invariant.
 

### 2017-05-05 draft.tex
ok

@@
 between two terms of sort $\Data$. A \emph{literal} is an atom or its
-negation. A \emph{formula} is a combination of atoms using disjunction
-$\vee$, conjunction $\wedge$ and negation $\neg$ and we write $\phi
-\rightarrow \psi$ for $\neg\phi \vee \psi$.
+negation. A \emph{formula} is an existentially quantified combination
+of atoms using disjunction $\vee$, conjunction $\wedge$ and negation
+$\neg$ and we write $\phi \rightarrow \psi$ for $\neg\phi \vee \psi$.
 
-We denote by $\fv{S}{\phi}$ the set of variables of sort $S$ in $\phi$
-and write $\fv{}{\phi}$ for $\bigcup_{S \in \set{\Data,\Bool}}
+We denote by $\fv{S}{\phi}$ the set of free (i.e.\ not under the scope
+of an existential quantifier) variables of sort $S$ in $\phi$ and
+write $\fv{}{\phi}$ for $\bigcup_{S \in \set{\Data,\Bool}}
 \fv{S}{\phi}$. For a variable $x \in \fv{}{\phi}$ and a term $t$ such
-that $\sigma(t) = \sigma(x)$, we denote by $\phi[t/x]$ the result of
+that $\sigma(t) = \sigma(x)$, let $\phi[t/x]$ be the result of
 replacing each occurrence of $x$ by $t$. For indexed sets
@@
 simultaneously replacing $x_i$ with $t_i$ in $\phi$, for all
-$i=1,\ldots,n$. A formula is in positive normal form (PNF) if negation
-occurs only within literals and in disjunctive normal form (DNF) if it
-is a disjunction of conjunctions of literals $\bigvee_{i=1}^N
-\bigwedge_{j=1}^{M_i} \ell_{ij}$. The size of a formula $\phi$,
-denoted $\len{\phi}$ is the number of symbols occuring in it.
+$i=1,\ldots,n$.
+%
+%% A formula is in positive normal form (PNF) if negation occurs only
+%% within literals and in disjunctive normal form (DNF) if it is a
+%% disjunction of conjunctions of literals $\bigvee_{i=1}^N
+%% \bigwedge_{j=1}^{M_i} \ell_{ij}$.
+%
+The size of a formula $\phi$, denoted $\len{\phi}$ is the number of
+symbols occuring in it.
 
@@
 \end{compactitem}
-Intuitively, $\overline{\vec{x}}$ track the previous and $\vec{x}$ the
-current values of the variables of $\A$. Formally, let $\vec{x}_k =
-\set{x_k \mid x \in \vec{x}}$, $k\geq0$, be a time-stamped set of
-variables. For an input even $a \in \Sigma$ and a formula $\phi$ we
-write $\Delta^k(\phi,a)$ for the formula obtained from $\phi$ by
-simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
-formula
+Intuitively, in each formula describing a transition rule,
+$\overline{\vec{x}}$ track the previous and $\vec{x}$ the current
+values of the variables of $\A$. Observe further, that the initial
+values of the variables are left unconstrained, because the initial
+configuration does not contain free data variables. The size of $\A$
+is defined as $\len{\A} = \len{\iota} + \sum_{(q,a) \in
+  Q\times\Sigma}\len{\Delta(q,a)}$.
+
+Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, $k\geq0$, be
+a time-stamped set of variables. For an input even $a \in \Sigma$ and
+a formula $\phi$ we write $\Delta^k(\phi,a)$ for the formula obtained
+from $\phi$ by simultaneously replacing each state $q \in
+\fv{\Bool}{\phi}$ by the formula
 $\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
-all $k\geq0$. The size of $\A$ is defined as $\len{\A} = \len{\iota} +
-\sum_{(q,a) \in Q\times\Sigma}\len{\Delta(q,a)}$.
-
-Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$, the
-\emph{run} of $\A$ over $w$ is the sequence of formulae
-$\phi_0,\phi_1,\ldots,\phi_n$, where $\phi_0 \equiv \iota$ and, for
-all $i > 0$, we have $\phi_k \equiv \Delta^k(\phi_{k-1},a_k)$. We
-slightly abuse notation and write $\Delta(\iota,a_1,\ldots,a_n)$ for
-the formula $\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ in the
-following. $\A$ \emph{accepts} $w$ iff $\I,\nu \models \phi_n$, where
-$\nu$ is the valuation that maps:\begin{inparaenum}[(1)]
-\item each $x \in \vec{x}_0$ to $\mathbf{0}$, 
+all $k\geq0$. Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
+(a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
+formulae $\phi_0,\phi_1,\ldots,\phi_n$, where $\phi_0 \equiv \iota$
+and, for all $k > 0$, we have $\phi_k \equiv
+\Delta^k(\phi_{k-1},a_k)$. We slightly abuse notation and write
+$\Delta(\iota,a_1,\ldots,a_n)$ for the formula
+$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ in the following. We say that
+$\A$ \emph{accepts} $w$ iff $\I,\nu \models \phi_n$, where $\nu$ is
+some valuation that maps:\begin{inparaenum}[(1)]
 %
@@
 \end{inparaenum}
-The language of $\A$, denoted as $L(\A)$, is the set of words accepted
-by $\A$.
+The language of $\A$ is the set $L(\A)$ of words accepted by $\A$.
 
@@
   \end{array}\]
-  Let $\nu,\nu' : Q \cup \bigcup_{i=0}^n\vec{x}_i$ be two valuations
-  such that: \begin{compactitem}
+  Let $\nu,\nu' : (Q \cup \bigcup_{i=0}^n\vec{x}_i) \rightarrow (\booli
+  \cup \Data^\I)$ be two valuations such that: \begin{compactitem}
   \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
@@
     %
-  \item $\nu(x) = \nu'(x) = \mathbf{0}$, for each $x \in \vec{x}_0$, 
+  \item $\nu(x) = \nu'(x)$, for each $x \in \vec{x}_0$, 
     %
@@
 Vector Addition System with States (AVASS) \cite{LazicSchmitz14} using
-only atoms $x = \overline{x} + k$, for $k \in \nat$ and $x \teq 0$,
-with the classical interpretation of function symbols on
+only atoms $x \geq 0$ and $x = \overline{x} + k$, for $k \in \zed$,
+with the classical interpretation of the function symbols on
 integers. Since reachability of a control state is undecidable for
@@
 
-Consequently, we give up on the termination guarantee and develop
+Consequently, we give up on the guarantee for termination and develop
 semi-algorithmic methods that are supposed to be
@@
 $\Post{\A}(\phi,a) \equiv \exists \overline{\vec{x}} ~.~
-\Delta(\phi[\overline{\vec{x}}/\vec{x}], a)$, mapping each formula in
-$\Form^+(Q,\vec{x})$ to a formula, also in $\Form^+(Q,\vec{x})$, that
-defines the effect of firing a transition with input event $a$.
+\Delta(\phi[\overline{\vec{x}}/\vec{x}], a) \in \Form^+(Q,\vec{x})$,
+mapping each formula in $\Form^+(Q,\vec{x})$ to a formula defining the
+effect of reading $a \in \Sigma$.
 
@@
 diverges in a large number of cases, since, in many interpretations,
-there exist infinite antichains\footnote{For instance, when $\Data^\I
-  = \nat$, an infinite antichain is $q \wedge x \teq 0, q \wedge x
-  \teq 1, \ldots$}.
+there exist infinite antichains of formulae\footnote{For instance,
+  when $\Data^\I = \nat$, an infinite antichain is $q \wedge x \teq 0,
+  q \wedge x \teq 1, \ldots$}.
 
@@
 \end{compactenum}
-If $\Inv$ satisfies only the first point above, we call it an
-invariant. Intuitively, a safety invariant maps every boolean
+If $\Inv$ satisfies only the first point above, we call it simply an
+\emph{invariant}. Intuitively, a safety invariant maps every boolean
 valuation into a set of data valuations, that contains the initial
-configuration (in which all data variables are assumed to be
-$\mathbf{0}\in\Data^\I$), over-approximates the set of reachable
-valuations and excludes those valuations that satisfying the
-acceptance condition.
+configuration $\iota$, in which the data variables are unconstrained
+and over-approximates the set of reachable valuations (\ref{it1:inv}),
+while excluding those valuations that satisfying the acceptance
+condition (\ref{it2:inv}).
 
@@
   ``$\Leftarrow$'' This direction is trivial. ``$\Rightarrow$'' We
-  define a function $\Inv : (Q \mapsto \booli) \rightarrow 2^{\vec{x}
-    \mapsto \Data^\I}$ as: for each $\beta : Q \rightarrow \booli$,
+  define $\Inv : (Q \mapsto \booli) \rightarrow 2^{\vec{x} \mapsto
+    \Data^\I}$ as follows. For each $\beta : Q \rightarrow \booli$,
   let $\Inv(\beta) = \{\nu : \vec{x} \rightarrow \Data^\I \mid \exists
   u \in \Sigma^* ~.~ \beta\cup\nu \models
-  \Post{\A}(\iota,u)\}$. Checking that $\Inv$ is indeed a safety
-  invariant is straightforward. \qed}
+  \Post{\A}(\iota,u)\}$. Checking that $\Inv$ is a safety invariant is
+  straightforward. \qed}
 
@@
 w.r.t. the predicates in $\preds$. The abstract version of the
-post-image and acceptance formulae are defined as follows:
-{\footnotesize\[
-\abs{\Post{\A}}(\phi,\varepsilon) \equiv \phi 
+post-image and acceptance condition are defined as follows:
+{\footnotesize\[ \abs{\Post{\A}}(\phi,\varepsilon) \equiv \phi
 \hspace*{2mm}
@@
 \hspace*{2mm}
-\abs{\Accept{\A}}(u) \equiv \abs{\Post{\A}}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q \rightarrow \bot)
-\]}
+\abs{\Accept{\A}}(u) \equiv \abs{\Post{\A}}(\iota,u) \wedge\!\!\!\bigwedge_{q \in Q \setminus F} 
+(q \rightarrow \bot)\]}
 \begin{lemma}\label{lemma:abstract-invariant}
@@
 \end{lemma}
-\proof{ It is sufficient to show that there exists $k>0$ such that for
-  all $u \in \Sigma^*$ there exists $i \in [1,k]$ such that
+\proof{ It is sufficient to show that there exists $k\geq0$ such that for
+  all $u \in \Sigma^*$ there exists $i \in [0,k]$ such that
   $\Post{\A}(\iota,u) \models \abs{\Post{\A}}(\iota,\mu(i))$. We have
@@
   $\{\abs{\Post{\A}}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
-  there exists $k > 0$ such that, for all $u \in \Sigma^*$ there
-  exists $i \in [1,k]$ such that $\abs{\Post{\A}}(\iota,u) \eqiff
+  there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
+  exists $i \in [0,k]$ such that $\abs{\Post{\A}}(\iota,u) \eqiff
   \abs{\Post{\A}}(\iota,\mu(i))$, which concludes the proof. \qed}
@@
 that $\beta\cup\nu \models \abs{\Accept{\A}}(u)$. This strenghtening
-is carried out by adding to $\preds$ predicates sufficient to exclude
-the spurious counterexample.
+is carried out by adding to $\preds$ several predicates that are
+sufficient to exclude the spurious counterexample.
 
@@
 \phi_n(X_{n-1},X_n)$ of time-stamped variables $X_i = \set{x_i \mid x
-  \in X}$ of any sort, an \emph{interpolant} for $\pi$ is a tuple
-$\tuple{I_0(X), I_1(X), \ldots, I_n(X)}$ such
-that:\begin{inparaenum}[(i)]
+  \in X}$ of any sort, a solution of the \emph{interpolation problem}
+$\pi$, simply called an \emph{interpolant}, is a tuple $\tuple{I_0(X),
+  I_1(X), \ldots, I_n(X)}$ such that:\begin{inparaenum}[(i)]
 \item $I_0 \equiv \top$,  
@@
 
+Given a non-empty spurious counterexample $u = a_1\ldots a_n$, where
+$n>0$, we consider the following interpolation problem: \[\pi(u)
+\equiv \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0 \cup
+\vec{x}_1) \wedge \ldots \wedge \theta_n(Q_{n-1} \cup
+Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n)\] where
+$Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped sets of
+boolean variables corresponding to the set $Q$ of states of $\A$. The
+first conjunct $\theta_0(Q_0) \equiv \iota[Q_0/Q]$ is the initial
+configuration of $\A$, with every $q \in \fv{\Bool}{\iota}$ replaced
+by $q_0$. The definition of $\theta_k$, $k\in[1,n]$ uses
+\emph{replacement sets} $R_\ell \in Q_\ell$, $\ell\in [0,n]$, which
+are defined inductively below: \begin{compactitem}
+\item $R_0 = \fv{\Bool}{\iota(Q_0)}$, 
+%
+\item for each $\ell\in[1,n]$: $\theta_\ell \equiv
+  \bigwedge_{q_{\ell-1}\in R_{\ell-1}} (q_{\ell-1} \rightarrow
+  \Delta(q,a_\ell)[Q_\ell/Q,\vec{x}_{\ell-1}/\overline{\vec{x}},\vec{x}_\ell/\vec{x}])$
+  and $R_\ell = \fv{\Bool}{\theta_\ell} \cap Q_\ell$.
+\end{compactitem}
+Finally, we set $\theta_{n+1}(Q_n) \equiv \bigwedge_{q \in Q \setminus
+  F} (q_n \rightarrow \bot)$. The intuition is that $\theta_0, \ldots,
+\theta_n$ simulates the run of $\A$ over $u$ and $\theta_{n+1}$ is the
+part of the acceptance condition which forces the last remaining
+non-final states to be false.
+
+We recall that a run of $\A$ over $u$ is a sequence $\phi_0(Q),
+\phi_1(Q,\vec{x}_0\cup\vec{x}_1), \ldots,
+\phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)$, where $\phi_0$ is the
+initial configuration $\iota$ and for each $k>0$, $\phi_k$ is obtained
+from $\phi_{k-1}$ by replacing each state $q \in
+\fv{\Bool}{\phi_{k-1}}$ by the formula $\Delta(q,a_k)$ given by the
+transition function of $\A$. The main reason behind the above
+definition of the interpolation problem is that we would like to
+obtain an interpolant $\tuple{\top,I_1(Q), I_2(Q,\vec{x}), \ldots,
+  I_{n}(Q,\vec{x}),\bot}$ whose formulae \emph{combine states with
+  the data constraints that must hold locally}, whenever the control
+reaches a certain boolean configuration. This association of states
+with data valuations is tantamount to defining efficient
+semi-algorithms, based on lazy abstraction \cite{HJMS02}. Observe
+that, because the states are replaced with transition formulae when
+moving one step in a run, the formulae thus produced lose track of the
+control history and are not directly suitable to producing
+interpolants that can be applied locally. Moreover, the abstraction
+defined by the interpolants generated in this way can
+\emph{over-approximate the control structure} of an automaton, in
+addition to the sets of data values encountered throughout its runs.
+
+The correctness of this interpolation-based abstraction refinement
+setup is captured by the progress property below, which guarantees
+that adding the formulae of an interpolant for $\pi(u)$ to the set
+$\preds$ of predicates suffices to exclude $u$ from future searches.
+
+\begin{lemma}\label{lemma:progress}
+  Given a non-empty sequence of input events $u \in \Sigma^*$ such
+  that $\Accept{\A}(u)$ is unsatisfiable, then the following
+  hold: \begin{compactenum}
+  \item\label{it1:progress} $\pi(u)$ is unsatisfiable, and
+    %
+  \item\label{it2:progress} if $\tuple{\top,I_1,\dots,I_n,\bot}$ is an
+    interpolant for $\pi(u)$ and $I_i \in \preds$, for all
+    $i\in[1,n]$, then $\abs{\Accept{\A}}(u)$ is unsatisfiable.
+  \end{compactenum}
+\end{lemma}
+\proof{
+
+\qed}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-05-05 draft.tex
ok

@@
 post-image and acceptance condition are defined as follows:
-{\footnotesize\[ \abs{\Post{\A}}(\phi,\varepsilon) \equiv \phi
+{\footnotesize\[ \AbsPost{\A}(\phi,\varepsilon) \equiv \phi
 \hspace*{2mm}
-\abs{\Post{\A}}(\phi,ua) \equiv \abs{(\Post{\A}(\abs{\Post{\A}}(\phi,u),a))} 
+\AbsPost{\A}(\phi,ua) \equiv \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} 
 \hspace*{2mm}
-\abs{\Accept{\A}}(u) \equiv \abs{\Post{\A}}(\iota,u) \wedge\!\!\!\bigwedge_{q \in Q \setminus F} 
+\AbsAccept{\A}(u) \equiv \AbsPost{\A}(\iota,u) \wedge\!\!\!\bigwedge_{q \in Q \setminus F} 
 (q \rightarrow \bot)\]}
@@
   For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
-  $k>0$ such that $\bigvee_{i=1}^k \abs{\Post{\A}}(\iota,\mu(i))$ defines
+  $k>0$ such that $\bigvee_{i=1}^k \AbsPost{\A}(\iota,\mu(i))$ defines
   an invariant for $\A$. 
@@
   all $u \in \Sigma^*$ there exists $i \in [0,k]$ such that
-  $\Post{\A}(\iota,u) \models \abs{\Post{\A}}(\iota,\mu(i))$. We have
-  $\Post{\A}(\iota,u) \models \abs{\Post{\A}}(\iota,u)$ for all $u \in
+  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,\mu(i))$. We have
+  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)$ for all $u \in
   \Sigma^*$. But since $\preds$ is a finite set, also the set
-  $\{\abs{\Post{\A}}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
+  $\{\AbsPost{\A}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
   there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
-  exists $i \in [0,k]$ such that $\abs{\Post{\A}}(\iota,u) \eqiff
-  \abs{\Post{\A}}(\iota,\mu(i))$, which concludes the proof. \qed}
+  exists $i \in [0,k]$ such that $\AbsPost{\A}(\iota,u) \eqiff
+  \AbsPost{\A}(\iota,\mu(i))$, which concludes the proof. \qed}
 
@@
 the definition of invariants, where $\Post{\A}$ and $\Accept{\A}$
-replace $\abs{\Post{\A}}$ and $\abs{\Accept{\A}}$, respectively. In
+replace $\AbsPost{\A}$ and $\AbsAccept{\A}$, respectively. In
 other words, there exists a finite sequence $u \in \Sigma^*$ such that
 $\nu \in \abs{\Inv}(\beta)$ and $\beta\cup\nu \models
-\abs{\Accept{\A}}(u)$, for some boolean $\beta : Q \rightarrow \booli$
+\AbsAccept{\A}(u)$, for some boolean $\beta : Q \rightarrow \booli$
 and data $\nu : \vec{x} \rightarrow \Data^\I$ valuations. Such
@@
 invariant by excluding from $\abs{\Inv}$ all pairs $(\beta,\nu)$ such
-that $\beta\cup\nu \models \abs{\Accept{\A}}(u)$. This strenghtening
+that $\beta\cup\nu \models \AbsAccept{\A}(u)$. This strenghtening
 is carried out by adding to $\preds$ several predicates that are
@@
     interpolant for $\pi(u)$ and $I_i \in \preds$, for all
-    $i\in[1,n]$, then $\abs{\Accept{\A}}(u)$ is unsatisfiable.
+    $i\in[1,n]$, then $\AbsAccept{\A}(u)$ is unsatisfiable.
   \end{compactenum}

### 2017-05-07 draft.tex
added predicate abstraction algo

@@
 Vector Addition System with States (AVASS) \cite{LazicSchmitz14} using
-only atoms $x \geq 0$ and $x = \overline{x} + k$, for $k \in \zed$,
+only atoms $x \geq k$ and $x = \overline{x} + k$, for $k \in \zed$,
 with the classical interpretation of the function symbols on
@@
 mapping each formula in $\Form^+(Q,\vec{x})$ to a formula defining the
-effect of reading $a \in \Sigma$.
+effect of reading the event $a$.
 
@@
 \Post{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q
-\rightarrow \bot)$ is satisfiable ? A na\"ive semi-algorithm
-enumerates all finite sequences and checks the satisfiability of
-$\Accept{\A}(u)$ for each $u \in \Sigma^*$, using a decision procedure
-for the theory $\theory(\sig,\I)$.
+\rightarrow \bot)$ is satisfiable ?  Observe that, since we ask a
+satisfiability query, the final states of $\A$ need not be
+constrained\footnote{ Since each state occurs positively in
+  $\Accept{\A}(u)$, this formula has a model iff it has a model with
+  every $q \in F$ set to true.}. A na\"ive semi-algorithm enumerates
+all finite sequences and checks the satisfiability of $\Accept{\A}(u)$
+for each $u \in \Sigma^*$, using a decision procedure for the theory
+$\theory(\sig,\I)$.
 
@@
 $u$, since the non-emptiness of $\A$ can be proved using $v$
-alone. However, in practice, the antichain based semi-algorithm
+alone. However, in practice, the antichain-based semi-algorithm
 diverges in a large number of cases, since, in many interpretations,
-there exist infinite antichains of formulae\footnote{For instance,
-  when $\Data^\I = \nat$, an infinite antichain is $q \wedge x \teq 0,
-  q \wedge x \teq 1, \ldots$}.
+there exist infinite antichains, e.g.\ $q \wedge x \teq 0, q \wedge x
+\teq 1, \ldots$ for $\Data^\I = \nat$.
 
@@
 \end{compactenum}
-If $\Inv$ satisfies only the first point above, we call it simply an
+A formula $\phi(Q,\vec{x})$ \emph{defines} $\Inv$ iff for all $\beta:
+Q \rightarrow \booli$ and $\nu : \vec{x} \rightarrow \Data^\I$,
+$\beta\cup\nu \models \phi \iff \nu \in \Inv(\beta)$. If $\Inv$
+satisfies only the first point above, we call it simply an
 \emph{invariant}. Intuitively, a safety invariant maps every boolean
 valuation into a set of data valuations, that contains the initial
-configuration $\iota$, in which the data variables are unconstrained
-and over-approximates the set of reachable valuations (\ref{it1:inv}),
-while excluding those valuations that satisfying the acceptance
-condition (\ref{it2:inv}).
+configuration $\iota = \Post{\A}(\iota,\varepsilon)$, in which the
+data variables are unconstrained, and over-approximates the set of
+reachable valuations (\ref{it1:inv}), while excluding those valuations
+which satisfy the acceptance condition (\ref{it2:inv}).
 
@@
 stop at step $k$ as soon as $\bigvee_{i=1}^k \Post{\A}(\iota,u_i)$
-defines\footnote{ A formula $\phi(Q,\vec{x})$ defines $\Inv$ iff for
-  all $\beta: Q \rightarrow \booli$ and $\nu : \vec{x} \rightarrow
-  \Data^\I$: $\beta\cup\nu \models \phi \iff \nu \in \Inv(\beta)$. } a
-safety invariant for $\A$. Although this condition can be effectively
-checked using a decision procedure for the theory $\theory(\sig,\I)$,
-there is no guarantee that this check will ever succeed.
+defines a safety invariant for $\A$. Although this condition can be
+effectively checked using a decision procedure for the theory
+$\theory(\sig,\I)$, there is no guarantee that this check will ever
+succeed.
 
@@
 out from the start that abstraction will allow us to build invariants
-which are not necessarily safety invariants. To converge towards
-meeting the latter condition, we shall resort to \emph{counterexample
-  guided abstraction refinement} (CEGAR).
+which are not necessarily safety invariants. To meet the latter
+condition, we resort to \emph{counterexample guided abstraction
+  refinement} (CEGAR).
 
@@
 spurious counterexample, from which the interpolant was generated,
-will never be revisited in the future. In many practical test cases,
-this progress property eventually yields a safety invariant.
+will never be revisited in the future. Though not always, in many
+practical test cases this progress property eventually yields a safety
+invariant.
 
 Given a non-empty spurious counterexample $u = a_1\ldots a_n$, where
-$n>0$, we consider the following interpolation problem: \[\pi(u)
-\equiv \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0 \cup
-\vec{x}_1) \wedge \ldots \wedge \theta_n(Q_{n-1} \cup
-Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n)\] where
-$Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped sets of
-boolean variables corresponding to the set $Q$ of states of $\A$. The
-first conjunct $\theta_0(Q_0) \equiv \iota[Q_0/Q]$ is the initial
-configuration of $\A$, with every $q \in \fv{\Bool}{\iota}$ replaced
-by $q_0$. The definition of $\theta_k$, $k\in[1,n]$ uses
-\emph{replacement sets} $R_\ell \in Q_\ell$, $\ell\in [0,n]$, which
-are defined inductively below: \begin{compactitem}
-\item $R_0 = \fv{\Bool}{\iota(Q_0)}$, 
+$n>0$, we consider the following interpolation problem: 
+\begin{equation}\label{eq:interpolation-problem}
+\pi(u) \equiv \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0
+\cup \vec{x}_1) \wedge \ldots \wedge \theta_n(Q_{n-1} \cup
+Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n)
+\end{equation} 
+where $Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped
+sets of boolean variables corresponding to the set $Q$ of states of
+$\A$. The first conjunct $\theta_0(Q_0) \equiv \iota[Q_0/Q]$ is the
+initial configuration of $\A$, with every $q \in \fv{\Bool}{\iota}$
+replaced by $q_0$. The definition of $\theta_k$, $k\in[1,n]$ uses
+\emph{replacement sets} $R_\ell \subseteq Q_\ell$, $\ell\in [0,n]$,
+which are defined inductively below: \begin{compactitem}
+\item $R_0 = \fv{\Bool}{\theta_0}$, 
 %
@@
   F} (q_n \rightarrow \bot)$. The intuition is that $\theta_0, \ldots,
-\theta_n$ simulates the run of $\A$ over $u$ and $\theta_{n+1}$ is the
-part of the acceptance condition which forces the last remaining
-non-final states to be false.
+\theta_n$ simulate the run of $\A$ over $u$ and $\theta_{n+1}$ is the
+acceptance condition, which forces the last remaining non-final states
+to be false.
 
@@
 \fv{\Bool}{\phi_{k-1}}$ by the formula $\Delta(q,a_k)$ given by the
-transition function of $\A$. The main reason behind the above
-definition of the interpolation problem is that we would like to
-obtain an interpolant $\tuple{\top,I_1(Q), I_2(Q,\vec{x}), \ldots,
-  I_{n}(Q,\vec{x}),\bot}$ whose formulae \emph{combine states with
-  the data constraints that must hold locally}, whenever the control
+transition function of $\A$. The main idea behind the above definition
+of the interpolation problem is that we would like to obtain an
+interpolant $\tuple{\top,I_1(Q), I_2(Q,\vec{x}), \ldots,
+  I_{n}(Q,\vec{x}),\bot}$ whose formulae \emph{combine states with the
+  data constraints that must hold locally}, whenever the control
 reaches a certain boolean configuration. This association of states
@@
 control history and are not directly suitable to producing
-interpolants that can be applied locally. Moreover, the abstraction
-defined by the interpolants generated in this way can
+interpolants that can be applied locally. Furthermore, the abstraction
+defined by the interpolants generated in this way can also
 \emph{over-approximate the control structure} of an automaton, in
@@
 that adding the formulae of an interpolant for $\pi(u)$ to the set
-$\preds$ of predicates suffices to exclude $u$ from future searches.
+$\preds$ of predicates suffices to exclude the spurious counterexample
+$u$ from future searches.
 
@@
     %
-  \item\label{it2:progress} if $\tuple{\top,I_1,\dots,I_n,\bot}$ is an
-    interpolant for $\pi(u)$ and $I_i \in \preds$, for all
-    $i\in[1,n]$, then $\AbsAccept{\A}(u)$ is unsatisfiable.
+  \item\label{it2:progress} if $\tuple{\top,I_1,\dots,I_{n+1},\bot}$
+    is an interpolant for $\pi(u)$ and $\set{I_i \mid i \in[1,n+1]}
+    \subseteq \preds$ then $\AbsAccept{\A}(u)$ is unsatisfiable.
   \end{compactenum}
 \end{lemma}
-\proof{
+\proof{ Let $u = a_1\ldots a_n$ and  $\pi(u) \equiv \theta_0(Q_0) \wedge
+  \theta_1(Q_0 \cup Q_1,\vec{x}_0 \cup \vec{x}_1) \wedge \ldots \wedge
+  \theta_n(Q_{n-1} \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge
+  \theta_{n+1}(Q_n)$ in the following.
 
-\qed}
+  (\ref{it1:progress}) Assume that $\pi(u)$ is satisfiable and let
+  $\beta_0,\ldots,\beta_n$ and $\nu_0,\ldots,\nu_n$ be boolean and
+  data valuations for $Q_0,\ldots,Q_n$ and
+  $\vec{x}_0,\ldots,\vec{x}_n$, respectively, such that
+  $\bigcup_{i=0}^n \beta_i \cup \bigcup_{i=0}^n \nu_i \models
+  \pi(u)$. We build a model for $\Accept{\A}(u) \equiv
+  \Post{\A}(\iota,u) \wedge \theta_{n+1}[Q/Q_n]$. Let
+  $\phi_0,\phi_1,\ldots,\phi_n$ be the run of $\A$ over $u$. We show
+  that $\phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)$ has a model, by
+  induction on $n\geq0$. For the base case $n=0$, $\phi_0 \equiv \iota
+  \equiv \theta_0[Q/Q_0]$ has the model $\beta(q) = \beta_0(q_0)$, for
+  all $q \in Q$. For the induction step $n > 0$, we assume that
+  $\phi_{n-1}(Q,\vec{x}_0\cup\ldots\vec{x}_{n-1})$ has the model
+  $\beta \cup \nu_0 \cup \ldots \nu_{n-1}$, where $\beta(q) =
+  \beta_{n-1}(q_{n-1})$, for all $q \in Q$. By the definition of
+  $\pi(u)$, we have: \[\theta_n \equiv \bigwedge_{q_{n-1} \in
+    \fv{\Bool}{\theta_{n-1}} \cap Q_{n-1}} \left(q_{n-1} \rightarrow
+  \Delta(q,a_n)[Q_n/Q,\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]\right)\]
+
+  But since $\beta_{n-1} \cup \beta_n, \nu_{n-1} \cup \nu_n \models
+  \theta_n(Q_{n-1}\cup Q_n, \vec{x}_{n-1} \cup \vec{x}_n)$, we have
+  that $\beta_n,\vec{x}_{n-1} \cup \vec{x}_n \models
+  \Delta(q,a_n)[Q_n/Q,\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
+  for every $q \in Q$ such that $\beta_{n-1}(q_{n-1}) = \true$. Then
+  we have $\beta,\vec{x}_0\cup\ldots\cup\vec{x}_n \models \phi_n$
+  where $\beta(q) = \beta_n(q_n)$, for all $q \in Q$ and $\phi_n$ is
+  obtained from $\phi_{n-1}$ by replacing each boolean variable $q$ by
+  $\Delta(a,q)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$. Observe
+  that, because each boolean variable occurs positively in
+  $\phi_{n-1}$, the truth value of
+  $\Delta(a,q)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$
+  under valuation $\beta\cup\nu_{n-1}\cup\nu_n$ is not important for
+  $q \in Q$ such that $\beta(q) = \false$. We have thus obtained a
+  model for $\phi_n$ and thus for $\Post{\A}(u)$. Since moreover,
+  $\theta_{n+1}[Q/Q_n] = \true$ in this model, we obtain a model for
+  $\Accept{\A}(u)$. Thus $\pi(u)$ is satisfiable implies that
+  $\Accept{\A}(u)$ is satisfiable, which proves the statement of point
+  (\ref{it1:progress}).
+
+  (\ref{it2:progress}) If $\tuple{\top,I_1,\dots,I_{n+1},\bot}$ is an
+    interpolant for $\pi(u)$, the following entailments hold: \begin{compactitem}
+    \item $\theta_0 \models I_1[Q_0/Q]$, 
+      %
+    \item $I_k[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k+1}[Q_k/Q,\vec{x}_k/\vec{x}]$, for all $k\in[1,n]$. 
+      %
+    \item $I_{n+1}[Q_{n+1}/Q] \wedge \theta_{n+1} \models \bot$. 
+    \end{compactitem}
+    We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n+1}$
+    by induction on $n \geq 0$. This is sufficient to conclude because
+    $\AbsAccept{\A}(a_1\ldots a_n) \equiv \AbsPost{\A}(\iota,a_1\ldots
+    a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n+1} \wedge
+    \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we
+    have $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv
+    \theta_0[Q/Q_0] \models I_1$. For the induction step $n>0$, we
+    compute:
+    \[\begin{array}{rcll}
+    \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] & \equiv & \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] & \text{ def. of $\AbsPost{\A}$} \\
+    & \models & \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n & \text{ ($\Asterisk$)} \\
+    & \models & \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n & \text{ by ind. hyp.} \\ 
+    & \models & I_n[Q_n/Q] & \hfill\text{\qed}
+    \end{array}\]
+    We prove the final point ($\Asterisk$). If
+    $\beta_n\cup\nu_{n-1}\cup\nu_n \models
+    \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q]$,
+    for some valuations $\beta_{n-1} : Q_{n-1} \rightarrow \booli$ and
+    $\nu_i : \vec{x}_i \rightarrow \Data^\I$, for $i=n-1,n$, then we
+    build a valuation $\beta_{n-1}$ such that $\beta_{n-1}\cup\beta_n
+    \cup \nu_{n-1}\cup\nu_n \models \AbsPost{\A}(\iota,a_1\ldots
+    a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$, as follows. For each
+    occurrence of a subformula $\Delta(q,a_n)$ in
+    $\AbsPost{\A}(\iota,a_1\ldots a_{n-1})$ we set
+    $\beta_{n-1}(q_{n-1}) = \true$ if $\beta_n\cup\nu_{n-1}\cup\nu_n
+    \models \Delta(q,a_n)[Q_n/Q]$ and $\beta_{n-1}(q_{n-1}) = \false$,
+    otherwise. Since there are no negated occurrences of such
+    subformulae, the definition of $\beta_{n-1}$ is consistent, and
+    the check $\beta_{n-1}\cup\beta_n \cup \nu_{n-1}\cup\nu_n \models
+    \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$
+    is straightforward. \qed}
+
+We have now all the ingredients to describe the first emptiness
+checking semi-algorithm for alternating data
+automata. Algorithm\footnote{Although termination is not guaranteed,
+  we call it an algorithm for conciseness.}  \ref{alg:predabs} builds
+an \emph{abstract reachability tree} (ART) whose edges are labeled
+with input events and each node $n$ is labeled with the formula
+$\AbsPost{\A}(\iota,u)$, where $u \in \Sigma^*$ is the event trace
+labeling the path from the root of the tree to $n$. It uses a covering
+relation $\lhd$ between nodes in order to ensure that the set of
+formulae labeling the nodes in the ART forms an antichain. Any
+spurious counterexample $u$ is eliminated by computing the
+interpolants corresponding to the least unfeasible suffix $v$ of $u$
+and recomputing the subtree of the ART rooted in the initial node of
+$v$.
+
+\begin{algorithm}[t!]
+\begin{algorithmic}[0]
+\State {\bf input}: an ADA $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$
+over the alphabet $\Sigma$ of input events
+
+\State {\bf output}: $\true$ if $L(\A)=\emptyset$ and a data word $w
+\in L(\A)$ otherwise
+\end{algorithmic}
+
+\begin{algorithmic}[1]
+  \State $\preds \leftarrow \set{\bot}$, $\worklist \leftarrow
+  \tuple{\rootNode}$, $\Art = \tuple{N,E,\Lambda,\lhd}$, where $N \leftarrow
+  \emptyset$, $E \leftarrow \emptyset$, $\Lambda \leftarrow
+  \set{(\rootNode,\iota)}$ and $\lhd \leftarrow \emptyset$
+
+  \While {$\worklist \neq \emptyset$}
+
+  \State dequeue $n$ from $\worklist$
+
+  \State $N \leftarrow N \cup \set{n}$
+
+  \If{$\Lambda(n) \wedge \bigwedge_{q \in Q \setminus F} (q
+    \rightarrow \bot)$ is satisfiable} \Comment{counterexample found}
+
+  \State let $u = a_1\ldots a_k$ be the label of the path from
+  $\rootNode$ to $n$ in $\Art$
+
+  \If{$\Accept{\A}(u)$ is satisfiable} \Comment{counterexample is
+    real}
+
+  \State let $\beta \cup \nu_1 \cup \ldots \cup \nu_k \models
+  \Accept{\A}(u)$, where $\beta : Q \rightarrow \booli$,
+  $\nu_1,\ldots,\nu_k : \vec{x} \rightarrow \Data^\I$
+
+  \State {\bf return} $w = (a_1,\nu_1) \ldots (a_k,\nu_k)$
+  \Comment{$w \in L(\A)$ by construction}
+
+  \Else \Comment{spurious counterexample}
+
+  \State let $p \leftarrow \Call{FindPivot}{u,\Art}$
+
+  \State let $v$ be the label of the path from $p$ to $n$ in $\Art$
+
+  \State $\preds \leftarrow \preds \cup \set{I_1,\ldots,I_\ell}$,
+  where $\tuple{\top,I_1,\ldots,I_\ell,\bot}$ is the interpolant for
+  $\overline{\pi}(v)$
+
+  \State let $\SubArt = \tuple{N',E',\Lambda',\lhd'}$ be the subtree
+  rooted at $p$ in $\Art$
+
+  \For{$(m,q) \in \lhd$ such that $q \in N'$}
+
+  \State remove $m$ from $N$ and enqueue $m$ into $\worklist$
+
+  \State remove $\SubArt$ from $\Art$
+
+  \State enqueue $p$ into $\worklist$
+
+  \EndFor
+
+  \EndIf
+
+  \Else
+
+  \For{$a \in \Sigma$}
+
+  \State let $\phi \leftarrow \AbsPost{\A}(\Lambda(n),a)$
+
+  \If{exist $m \in N$ such that $\phi \models \Lambda(m)$}
+
+  \State $\lhd \leftarrow \lhd \cup \set{(n,m)}$ \Comment{$n$ covers $m$}
+
+  \Else
+
+  \State let $s \leftarrow \Call{NewNode}{}$ and $\Lambda \leftarrow
+  \Lambda \cup \set{(s,\phi)}$
+
+  \State $R \leftarrow \set{m \in \worklist \mid \Lambda(m) \models
+    \phi}$ \Comment{worklist nodes covered by $s$}
+
+  \For{$r \in R$}
+
+  \For{$m \in N$ such that $(m,r) \in E$}
+
+  \State let $\lhd \leftarrow \lhd \cup \set{(m,s)}$ \Comment{redirect covered children from $R$ into $s$}
+
+  \EndFor
+
+  \For{$(m,r) \in \lhd$}
+
+  \State let $\lhd \leftarrow \lhd \cup \set{(m,s)}$ \Comment{redirect covered nodes from $R$ into $s$}
+
+  \EndFor
+
+  \EndFor
+
+  \State remove $R$ from $\Art$
+
+  \State enqueue $s$ into $\worklist$ \Comment{recompute the subtree rooted at $s$}
+
+  \EndIf
+
+  \EndFor
+
+  \EndIf
+
+  \EndWhile
+
+  \State {\bf return} $\true$
+\end{algorithmic}
+\caption{Lazy Predicate Abstraction for Checking Emptiness of ADA}
+\label{alg:predabs}
+\end{algorithm}
+
+Formally, let $\Art = \tuple{N,E,\Lambda,\lhd}$ be a tree, where $N$
+is a set of nodes, $E \subseteq N \times \Sigma \times N$ is a set of
+edges, $\Lambda : N \rightarrow \Form^+(Q,\vec{x})$ is a labeling of
+the nodes with formulae, and $\lhd \subseteq N \times N$ is a set of
+covering edges. The root of the tree is labeled with the initial
+configuration $\iota$ and each node $n \in N$ corresponds to a unique
+path from the root to $n$, labeled by a sequence $u$ of input
+events. The least unfeasible suffix of $u$ is the smallest sequence $v
+= a_1 \ldots a_k$, such that $u = wv$, for some $w \in \Sigma^*$ and
+the formula (\ref{eq:pivot}) below is unsatisfiable. The \emph{pivot}
+of $u$ is the node $p$ corresponding to the start of $v$ in $u$.
+\begin{equation}\label{eq:pivot}
+  \overline{\pi}(a_1\ldots a_k) \equiv \Lambda(p)[Q_0/Q] \wedge
+  \theta_1(Q_0 \cup Q_1,\vec{x}_0 \cup \vec{x}_1) \wedge \ldots \wedge
+  \theta_{k+1}(Q_k)
+\end{equation}
+where $\theta_1,\ldots,\theta_{k+1}$ are defined as in
+(\ref{eq:interpolation-problem}), for $\theta_0 \equiv
+\Lambda(p)[Q_0/Q]$.
+
+With these considerations, Algorithm \ref{alg:predabs} uses a worklist
+iteration to build a ART. We keep newly expanded nodes of $\Art$ in a
+use a queue $\worklist$ to ensure a breadth-first exploration
+strategy, which guarantees that the shortest counterexamples are
+explored first. When the search encounters a counterexample $u$, it is
+checked for spuriousness. If the counterexample is real, the procedure
+returns a data word $w \in L(\A)$, which interleaves the input events
+of $u$ with the data valuations from the model of $\Accept{\A}(u)$
+(since $u$ is feasible, clearly $\Accept{\A}(u)$ is
+satisfiable). Otherwise, if $u$ is spurious, we compute its pivot $p$,
+add the interpolants for the least unfeasible suffix of $u$ to the set
+of predicates $\preds$, remove and recompute the subtree of $\Art$
+rooted at $p$.
+
+Termination of Algorithm \ref{alg:predabs} depends on the ability of a
+given interpolating decision procedure for the combined boolean and
+data theory $\theory(\sig,\I)$ to provide interpolants that yield a
+safety invariant, provided that $L(\A) = \emptyset$. In this case, we
+use the covering relation $\lhd$ to ensure that, when a newly
+generated node is covered by a node in $N$, it is not added to the
+worklist, thus cutting the current branch of the search. Formally, for
+any two nodes $n,m \in N$, we have $n \lhd m$ iff
+$\AbsPost{\A}(\Lambda(n),a) \models \Lambda(m)$ for some $a \in
+\Sigma$, in other words, if $n$ has a successor whose label entails
+the label of $m$. The soundness and relative completness of Algorithm
+\ref{alg:predabs} are proved by the following theorem:
+
+\begin{theorem}\label{thm:predabs}
+  Given an automaton $\A$, such that $L(\A) \neq \emptyset$, Algorithm
+  \ref{alg:predabs} terminates and returns a word $w \in L(\A)$. If
+  Algorithm \ref{alg:predabs} terminates reporting $\true$, then
+  $L(\A) = \emptyset$.
+\end{theorem}
 

### 2017-05-07 draft.tex
switched to IEEE Trans

@@
-\documentclass{llncs}
+% \documentclass{llncs}
+\documentclass[10pt,conference,letterpaper,twocolumn]{IEEEtran}
 
@@
 
-\author{Radu Iosif \and Xiao Xu}
-
-\institute{Verimag/CNRS/Universit\'e de Grenoble Alpes}
+\author{Radu Iosif and Xiao Xu \\
+Verimag/CNRS/Universit\'e de Grenoble Alpes\\
+Email: \{Radu.Iosif,Xiao.Xu\}@univ-grenoble-alpes.fr}
 
@@
 of $\phi$:
-\[\begin{array}{lclclclclcl}
+\[\begin{array}{lclclcl}
 \overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
-\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} && 
-\overline{\neg\phi} & \equiv & \neg \overline{\phi} \text{ if $\phi$ is not an atom} \\
-\overline{\phi} & \equiv & \phi \text{  if $\phi \in Q$}\hspace*{1cm} &&
-\overline{\phi} & \equiv & \neg\phi \text{  if $\phi \not\in Q$ is an atom}
+\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} \\ 
+\overline{\neg\phi} & \equiv & \neg \overline{\phi} \text{ if $\phi$ not atom} &&
+\overline{\phi} & \equiv & \phi \text{ if $\phi \in Q$} \\
+\overline{\phi} & \equiv & \neg\phi \text{ if $\phi \not\in Q$ atom}
 \end{array}\]
@@
 \begin{lemma}\label{lemma:closure}
-  For any automata $\A_1 = \tuple{\vec{x},Q_1,\iota_1,F_1,\Delta_1}$
-  and $\A_2 = \tuple{\vec{x},Q_2,\iota_2,F_2,\Delta_2}$ such that $Q_1
-  \cap Q_2 = \emptyset$, we have $L(\A_\cup) = L(\A_1) \cup L(\A_2)$,
-  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$ and $L(\overline{\A_1}) =
-  \Sigma[\vec{x}]^* \setminus L(\A_1)$.
+  Given automata $\A_i = \tuple{\vec{x},Q_i,\iota_i,F_i,\Delta_i}$,
+  for $i=1,2$, such that $Q_1 \cap Q_2 = \emptyset$, we have
+  $L(\A_\cup) = L(\A_1) \cup L(\A_2)$, $L(\A_\cap) = L(\A_1) \cap
+  L(\A_2)$ and $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus
+  L(\A_1)$.
 \end{lemma}
@@
   \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
-  \[\begin{array}{rcl}
-  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iff & 
-  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee
-  \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) \\
-  & \iff & \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots
-  a_{n}) \enspace.
+  \[\begin{array}{ll}
+  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \eqiff \\
+  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) & \eqiff \\
+  \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots a_{n}) \enspace.
   \end{array}\] 
@@
   $n>0$, we compute:
-  \[\begin{array}{rcll}
-  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \eqiff & 
-  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \text{ by induction hypothesis} \\
-  & \eqiff & \overline{\Delta(\iota_1,a_1\ldots,a_n)} & \text{ by the definition of $\overline{\phi}$ .}
+  \[\begin{array}{lcl}
+  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \eqiff & \text{ by induction hypothesis} \\
+  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \eqiff & \text{ by the definition of $\overline{\phi}$} \\
+  \overline{\Delta(\iota_1,a_1\ldots,a_n)}
   \end{array}\]
   Let $\nu,\nu' : (Q \cup \bigcup_{i=0}^n\vec{x}_i) \rightarrow (\booli
-  \cup \Data^\I)$ be two valuations such that: \begin{compactitem}
+  \cup \Data^\I)$ be valuations such that: \begin{compactitem}
   \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
@@
 for the boolean operations, including complementation, can be built in
-linear time. This is not very surprising, since linear time
-intersection and complementation are also possible for alternating
-automata over finite alphabets \cite{ChandraKozenStockmeyer81}.
+linear time. This is not surprising, as linear time intersection and
+complementation are also possible for alternating automata over finite
+alphabets \cite{ChandraKozenStockmeyer81}.
 
@@
 Consequently, we give up on the guarantee for termination and develop
-semi-algorithmic methods that are supposed to be
-both: \begin{compactenum}
+semi-algorithms that are supposed to be: \begin{compactenum}
 \item \emph{sound}: given an automaton $\A$, if $L(\A) \neq\emptyset$,
@@
 
-The solution we adopt is the use of \emph{abstraction} to ensure
-termination of invariant computations. However, it is worth pointing
-out from the start that abstraction will allow us to build invariants
-which are not necessarily safety invariants. To meet the latter
-condition, we resort to \emph{counterexample guided abstraction
-  refinement} (CEGAR).
+The solution we adopt is the use of abstraction to ensure termination
+of invariant computations. However, it is worth pointing out from the
+start that abstraction will allow us to build invariants which are not
+necessarily safety invariants. To meet the latter condition, we resort
+to counterexample guided abstraction refinement (CEGAR).
 
@@
 post-image and acceptance condition are defined as follows:
-{\footnotesize\[ \AbsPost{\A}(\phi,\varepsilon) \equiv \phi
-\hspace*{2mm}
-\AbsPost{\A}(\phi,ua) \equiv \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} 
-\hspace*{2mm}
-\AbsAccept{\A}(u) \equiv \AbsPost{\A}(\iota,u) \wedge\!\!\!\bigwedge_{q \in Q \setminus F} 
-(q \rightarrow \bot)\]}
+\[\begin{array}{llcl}
+\AbsPost{\A}(\phi,\varepsilon) \equiv \phi & 
+\AbsPost{\A}(\phi,ua) & \equiv & \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} \\
+&
+\AbsAccept{\A}(u) & \equiv & \AbsPost{\A}(\iota,u) \wedge\!\!\!\bigwedge_{q \in Q \setminus F} 
+(q \rightarrow \bot)
+\end{array}\]
 \begin{lemma}\label{lemma:abstract-invariant}
@@
 $n>0$, we consider the following interpolation problem: 
-\begin{equation}\label{eq:interpolation-problem}
-\pi(u) \equiv \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0
-\cup \vec{x}_1) \wedge \ldots \wedge \theta_n(Q_{n-1} \cup
-Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n)
-\end{equation} 
+\begin{eqnarray}\label{eq:interpolation-problem}
+\pi(u) & \equiv & \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0
+\cup \vec{x}_1) \wedge \ldots \\ 
+&& \wedge~ \theta_n(Q_{n-1} \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n) \nonumber
+\end{eqnarray}
 where $Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped
@@
 %
-\item for each $\ell\in[1,n]$: $\theta_\ell \equiv
-  \bigwedge_{q_{\ell-1}\in R_{\ell-1}} (q_{\ell-1} \rightarrow
+\item $\theta_\ell \equiv \bigwedge_{q_{\ell-1}\in R_{\ell-1}}
+  (q_{\ell-1} \rightarrow
   \Delta(q,a_\ell)[Q_\ell/Q,\vec{x}_{\ell-1}/\overline{\vec{x}},\vec{x}_\ell/\vec{x}])$
-  and $R_\ell = \fv{\Bool}{\theta_\ell} \cap Q_\ell$.
+  and $R_\ell = \fv{\Bool}{\theta_\ell} \cap Q_\ell$, for each
+  $\ell\in[1,n]$.
 \end{compactitem}
@@
   \beta_{n-1}(q_{n-1})$, for all $q \in Q$. By the definition of
-  $\pi(u)$, we have: \[\theta_n \equiv \bigwedge_{q_{n-1} \in
+  $\pi(u)$: \[\theta_n \equiv \bigwedge_{q_{n-1} \in
     \fv{\Bool}{\theta_{n-1}} \cap Q_{n-1}} \left(q_{n-1} \rightarrow
@@
       %
-    \item $I_k[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k+1}[Q_k/Q,\vec{x}_k/\vec{x}]$, for all $k\in[1,n]$. 
+    \item $I_k[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k+1}[Q_k/Q,\vec{x}_k/\vec{x}]$, $\forall k\in[1,n]$. 
       %
@@
     compute:
-    \[\begin{array}{rcll}
-    \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] & \equiv & \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] & \text{ def. of $\AbsPost{\A}$} \\
-    & \models & \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n & \text{ ($\Asterisk$)} \\
-    & \models & \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n & \text{ by ind. hyp.} \\ 
-    & \models & I_n[Q_n/Q] & \hfill\text{\qed}
+    \[\begin{array}{lcl}
+    \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] & \equiv & \text{ def. of $\AbsPost{\A}$} \\
+    \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] & \models \\
+    \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n & \models & \text{ ($\Asterisk$)} \\
+    \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n & \models & \text{ ind. hyp.} \\ 
+    I_n[Q_n/Q] & \hfill\text{\qed}
     \end{array}\]
@@
 \begin{algorithm}[t!]
-\begin{algorithmic}[0]
+{\scriptsize\begin{algorithmic}[0]
 \State {\bf input}: an ADA $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$
@@
 \in L(\A)$ otherwise
-\end{algorithmic}
+\end{algorithmic}}
 
-\begin{algorithmic}[1]
+{\scriptsize\begin{algorithmic}[1]
   \State $\preds \leftarrow \set{\bot}$, $\worklist \leftarrow
@@
 
-  \State let $s \leftarrow \Call{NewNode}{}$ and $\Lambda \leftarrow
-  \Lambda \cup \set{(s,\phi)}$
+  \State let $s$ be a fresh node and $\Lambda \leftarrow \Lambda \cup
+  \set{(s,\phi)}$
 
@@
   \State {\bf return} $\true$
-\end{algorithmic}
-\caption{Lazy Predicate Abstraction for Checking Emptiness of ADA}
+\end{algorithmic}}
+\caption{Lazy Predicate Abstraction for ADA Emptiness}
 \label{alg:predabs}
@@
 \begin{equation}\label{eq:pivot}
-  \overline{\pi}(a_1\ldots a_k) \equiv \Lambda(p)[Q_0/Q] \wedge
+  \overline{\pi}(v) \equiv \Lambda(p)[Q_0/Q] \wedge
   \theta_1(Q_0 \cup Q_1,\vec{x}_0 \cup \vec{x}_1) \wedge \ldots \wedge

### 2017-05-08 draft.tex
proof of Thm. 1 added

@@
 %
-The size of a formula $\phi$, denoted $\len{\phi}$ is the number of
-symbols occuring in it.
+The size of a formula $\phi$, denoted as $\len{\phi}$, is the number
+of symbols occuring in it.
 
@@
 A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
-there exists a valuation $\nu$ such that $\I,\nu \models \phi$. The
-\emph{theory} $\theory(\sig,\I)$ is the set of satisfiable formulae
-written in the signature $\sig$, with the interpretation $\I$.
+there exists a valuation $\nu$ such that $\I,\nu \models \phi$, and
+\emph{valid} if $\I,\nu \models \phi$ for all valuations $\nu$.  The
+\emph{theory} $\theory(\sig,\I)$ is the set of valid formulae written
+in the signature $\sig$, with the interpretation $\I$.
 
@@
 This definition generalizes the classical notion of words from a
-finite alphabet to using a possibly infinite alphabet
+finite alphabet to the possibly infinite alphabet
 $\Sigma[\vec{x}]$. Clearly, when $\Data^\I$ is sufficiently large or
@@
 
-An \emph{alternating data automaton} (or, simply, and automaton) is a
-tuple $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$: \begin{compactitem}
+An \emph{alternating data automaton} (ADA) or, simply, an automaton,
+is a tuple $\A =
+\tuple{\vec{x},Q,\iota,F,\Delta}$, where: \begin{compactitem}
 %
@@
 %
-\item $Q$ is a finite set of variables of sort $\Bool$, also called
+\item $Q$ is a finite set of variables of sort $\Bool$, called
   \emph{states},
 %
-\item $\iota \in \Form^+(Q,\emptyset)$ is a formula describing the
-  \emph{initial configurations},
+\item $\iota \in \Form^+(Q,\emptyset)$ is positive boolean combination
+  of states defining the \emph{initial configurations},
 %
-\item $F \subseteq Q$ is a set of \emph{final} states, and
+\item $F \subseteq Q$ is a set of \emph{final states}, and
 %
@@
 \end{compactitem}
-Intuitively, in each formula describing a transition rule,
-$\overline{\vec{x}}$ track the previous and $\vec{x}$ the current
-values of the variables of $\A$. Observe further, that the initial
-values of the variables are left unconstrained, because the initial
-configuration does not contain free data variables. The size of $\A$
-is defined as $\len{\A} = \len{\iota} + \sum_{(q,a) \in
-  Q\times\Sigma}\len{\Delta(q,a)}$.
+Intuitively, in each formula $\Delta(q,a)$ describing a transition
+rule, the variables $\overline{\vec{x}}$ track the previous and
+$\vec{x}$ the current values of the variables of $\A$. Observe
+further, that the initial values of the variables are left
+unconstrained, as the initial configuration does not contain free
+data variables. The size of $\A$ is defined as $\len{\A} = \len{\iota}
++ \sum_{(q,a) \in Q\times\Sigma}\len{\Delta(q,a)}$.
 
 Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, $k\geq0$, be
-a time-stamped set of variables. For an input even $a \in \Sigma$ and
-a formula $\phi$ we write $\Delta^k(\phi,a)$ for the formula obtained
+a set of time-stamped variables. For an input event $a \in \Sigma$ and
+a formula $\phi$, we write $\Delta^k(\phi,a)$ for the formula obtained
 from $\phi$ by simultaneously replacing each state $q \in
@@
 and, for all $k > 0$, we have $\phi_k \equiv
-\Delta^k(\phi_{k-1},a_k)$. We slightly abuse notation and write
+\Delta^k(\phi_{k-1},a_k)$. Next, we slightly abuse notation and write
 $\Delta(\iota,a_1,\ldots,a_n)$ for the formula
-$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ in the following. We say that
-$\A$ \emph{accepts} $w$ iff $\I,\nu \models \phi_n$, where $\nu$ is
-some valuation that maps:\begin{inparaenum}[(1)]
+$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ above. We say that $\A$
+\emph{accepts} $w$ iff $\I,\nu \models \Delta(\iota,a_1,\ldots,a_n)$,
+where $\nu$ is some valuation that maps:\begin{inparaenum}[(1)]
 %
@@
 %
-\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$, and 
+\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$ and 
 %
@@
 
-In this section we tackle the following problems: \begin{compactenum}
+In this paper we tackle the following problems: \begin{compactenum}
 \item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
   with the same set of variables $\vec{x}$, do there exist automata
-  $\A_\cup$, $\A_\cap$ and $\overline{\A}$ such that $L(\A_\cup) =
+  $\A_\cup$, $\A_\cap$ and $\overline{\A_1}$ such that $L(\A_\cup) =
   \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and
@@
 It is well known that other problems, such as \emph{universality}
-(given an automaton $\A$ with variables $\vec{x}$, does $L(\A) =
-\Sigma[\vec{x}]^*$) and \emph{inclusion} (given automata $\A_1$ and
+(given automaton $\A$ with variables $\vec{x}$, does $L(\A) =
+\Sigma[\vec{x}]^*$?) and \emph{inclusion} (given automata $\A_1$ and
 $\A_2$ with the same set of variables, does $L(\A_1) \subseteq
-L(\A_2)$ ?) can be reduced to the above problems. Observe furthermore
+L(\A_2)$?) can be reduced to the above problems. Observe furthermore
 that we do not consider cases in which the sets of variables in the
-two automata differ. An interesting problem in this case is: given
+two automata differ. An interesting problem in this case would be: given
 automata $\A_1$ and $\A_2$, with variables $\vec{x}_1$ and
@@
   Given a formula $\phi \in \Form^+(Q,\vec{x})$ and a valuation $\nu$
-  mapping each $q \in Q$ a value $\nu(q) \in \booli$ and each $x \in
+  mapping each $q \in Q$ to a value $\nu(q) \in \booli$ and each $x \in
   \vec{x}$ to a value $\nu(x) \in \Data^\I$, let $\nu'$ be the
@@
   each $x \in \vec{x}$ the value $\nu(x)$. Then we have 
-  $\I,\nu \models \phi \iff \I,\nu' \not\models \overline{\phi}$. 
+  $\I,\nu \models \phi$ if and only if $\I,\nu' \not\models \overline{\phi}$. 
 \end{proposition}
@@
 
-In the following let $\A_1 = \tuple{\vec{x},Q_1,\iota_1,F_1,\Delta_1}$
-and $\A_2 = \tuple{\vec{x},Q_2,\iota_2,F_2,\Delta_2}$, where
-w.l.o.g. we assume that $Q_1 \cap Q_2 = \emptyset$. We define $\A_\cup
-= \tuple{\vec{x},Q_1\cup Q_2,\iota_1 \vee \iota_2,F_1\cup
-  F_2,\Delta_1\cup\Delta_2}$, $\A_\cap = \tuple{\vec{x},Q_1\cup
-  Q_2,\iota_1 \wedge \iota_2, F_1\cup F_2, \Delta_1\cup\Delta_2}$ and
-$\overline{\A_1} = \langle \vec{x},Q_1,\iota_1,Q_1\setminus
-F_1,\overline{\Delta_1} \rangle$, where $\overline{\Delta_1}(q,a)
-\equiv \overline{\Delta_1(q,a)}$, for all $q \in Q_1$ and $a \in
-\Sigma$. The following lemma shows the correctness of these
-definitions:
+In the following let $\A_i =
+\tuple{\vec{x},Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, where
+w.l.o.g. we assume that $Q_1 \cap Q_2 = \emptyset$. We define: 
+\[\begin{array}{rcl}
+\A_\cup & = & \tuple{\vec{x},Q_1\cup Q_2,\iota_1 \vee \iota_2,F_1\cup F_2,\Delta_1\cup\Delta_2} \\ 
+\A_\cap & = & \tuple{\vec{x},Q_1\cup Q_2,\iota_1 \wedge \iota_2, F_1\cup F_2, \Delta_1\cup\Delta_2} \\ 
+\overline{\A_1} & = & \langle \vec{x},Q_1,\iota_1,Q_1\setminus F_1,\overline{\Delta_1} \rangle
+\end{array}\] 
+where $\overline{\Delta_1}(q,a) \equiv \overline{\Delta_1(q,a)}$, for
+all $q \in Q_1$ and $a \in \Sigma$. The following lemma shows the
+correctness of the above definitions:
 
@@
 integers. Since reachability of a control state is undecidable for
-AVASS, the emptiness problem is undecidable for ADA.
+AVASS \cite{LINCOLN92}, the emptiness problem is undecidable for ADA.
 
@@
 \item \emph{sound}: given an automaton $\A$, if $L(\A) \neq\emptyset$,
-  the procedure will terminate and return a word $w \in L(\A)$), and
+  the procedure will terminate and return a word $w \in L(\A)$, and
 %
@@
 We generalize the post-image function to finite sequences of input
-events, as follows: $\Post{\A}(\phi,\varepsilon) \equiv \phi$ and
-$\Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a)$, for any $u
-\in \Sigma^*$. Then the emptiness problem for $\A$ becomes: does there
-exist $u \in \Sigma^*$ such that the formula $\Accept{\A}(u) \equiv
-\Post{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q
-\rightarrow \bot)$ is satisfiable ?  Observe that, since we ask a
-satisfiability query, the final states of $\A$ need not be
-constrained\footnote{ Since each state occurs positively in
-  $\Accept{\A}(u)$, this formula has a model iff it has a model with
-  every $q \in F$ set to true.}. A na\"ive semi-algorithm enumerates
-all finite sequences and checks the satisfiability of $\Accept{\A}(u)$
-for each $u \in \Sigma^*$, using a decision procedure for the theory
-$\theory(\sig,\I)$.
+events, as follows, for any $u \in \Sigma^*$:
+\[\begin{array}{l}
+\Post{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm} \Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a) \\
+\Accept{\A}(u) \equiv \Post{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q \rightarrow \bot)
+\end{array}\]
+Then the emptiness problem for $\A$ becomes: does there exist $u \in
+\Sigma^*$ such that the formula $\Accept{\A}(u)$ is satisfiable?
+Observe that, since we ask a satisfiability query, the final states of
+$\A$ need not be constrained\footnote{ Since each state occurs
+  positively in $\Accept{\A}(u)$, this formula has a model iff it has
+  a model with every $q \in F$ set to true.}. A na\"ive semi-algorithm
+enumerates all finite sequences and checks the satisfiability of
+$\Accept{\A}(u)$ for each $u \in \Sigma^*$, using a decision procedure
+for the theory $\theory(\sig,\I)$.
 
@@
 $u$, since the non-emptiness of $\A$ can be proved using $v$
-alone. However, in practice, the antichain-based semi-algorithm
-diverges in a large number of cases, since, in many interpretations,
-there exist infinite antichains, e.g.\ $q \wedge x \teq 0, q \wedge x
-\teq 1, \ldots$ for $\Data^\I = \nat$.
+alone. However, the antichain-based semi-algorithm diverges in a large
+number of cases, since infinite antichains exist in many
+interpretations, e.g.\ $q \wedge x \teq 0,~ q \wedge x \teq 1, \ldots$
+for $\Data^\I = \nat$.
 
@@
 \booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
-boolean valuation $\beta : Q \rightarrow \booli$, every valuation
-$\nu : \vec{x} \mapsto \Data^\I$ of the data
-variables and every finite sequence $u \in \Sigma^*$ of input events: \begin{compactenum}
+boolean valuation $\beta : Q \rightarrow \booli$, every valuation $\nu
+: \vec{x} \mapsto \Data^\I$ of the data variables and every finite
+sequence $u \in \Sigma^*$ of input events, the following
+hold: \begin{compactenum}
 %% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
@@
 %
-\item\label{it1:inv} $\beta \cup \nu \models \Post{\A}(\iota,u)
+\item\label{it1:inv} $\I,\beta \cup \nu \models \Post{\A}(\iota,u)
   \Rightarrow \nu \in \Inv(\beta)$, and
 %
-\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \beta \cup \nu
-  \not\models \Accept{\A}(u)$.
+\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \I,\beta \cup
+  \nu \not\models \Accept{\A}(u)$.
 \end{compactenum}
@@
 Q \rightarrow \booli$ and $\nu : \vec{x} \rightarrow \Data^\I$,
-$\beta\cup\nu \models \phi \iff \nu \in \Inv(\beta)$. If $\Inv$
-satisfies only the first point above, we call it simply an
-\emph{invariant}. Intuitively, a safety invariant maps every boolean
-valuation into a set of data valuations, that contains the initial
-configuration $\iota = \Post{\A}(\iota,\varepsilon)$, in which the
-data variables are unconstrained, and over-approximates the set of
-reachable valuations (\ref{it1:inv}), while excluding those valuations
-which satisfy the acceptance condition (\ref{it2:inv}).
+$\I,\beta\cup\nu \models \phi$ if and only if $\nu \in
+\Inv(\beta)$. If $\Inv$ satisfies only the first point above, we call
+it an \emph{invariant}. Intuitively, a safety invariant maps every
+boolean valuation into a set of data valuations, that contains the
+initial configuration $\iota = \Post{\A}(\iota,\varepsilon)$, whose
+data variables are unconstrained, over-approximates the set of
+reachable valuations (\ref{it1:inv}) and excludes the valuations
+satisfying the acceptance condition (\ref{it2:inv}).
 
 \begin{lemma}\label{lemma:safety-invariant}
-  For any automaton $\A$, we have $L(\A)
-= \emptyset$ iff $\A$ has a safety invariant. 
+  For any automaton $\A$, we have $L(\A) = \emptyset$ if and only if
+  $\A$ has a safety invariant.
 \end{lemma}
@@
 
-The solution we adopt is the use of abstraction to ensure termination
-of invariant computations. However, it is worth pointing out from the
+The solution we adopt is abstraction to ensure the termination of
+invariant computations. However, it is worth pointing out from the
 start that abstraction will allow us to build invariants which are not
@@
 $Q \cup \vec{x}$, such that $\bot \in \preds$ and the variables in $Q$
-can occur negated in some $\phi \in \preds$. We assume that $\preds$
-is fixed for now, and refer to these formulae as \emph{predicates}.
-Given a formula $\phi$, we denote by $\abs{\phi} \equiv \bigwedge
-\set{\pi \in \preds \mid \phi \models \pi}$ the abstraction of $\phi$
-w.r.t. the predicates in $\preds$. The abstract version of the
-post-image and acceptance condition are defined as follows:
-\[\begin{array}{llcl}
-\AbsPost{\A}(\phi,\varepsilon) \equiv \phi & 
-\AbsPost{\A}(\phi,ua) & \equiv & \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} \\
-&
-\AbsAccept{\A}(u) & \equiv & \AbsPost{\A}(\iota,u) \wedge\!\!\!\bigwedge_{q \in Q \setminus F} 
+can now occur negated within the formulae $\phi \in \preds$. We fix
+$\preds$ from now on and refer to these formulae as
+\emph{predicates}. Given a formula $\phi$, we denote by $\abs{\phi}
+\equiv \bigwedge \set{\pi \in \preds \mid \phi \models \pi}$ the
+abstraction of $\phi$ w.r.t. the predicates in $\preds$. The abstract
+version of the post-image and acceptance condition are defined as
+follows, for any $u \in \Sigma^*$:
+\[\begin{array}{l}
+\AbsPost{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm}
+\AbsPost{\A}(\phi,ua) \equiv \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} \\
+\AbsAccept{\A}(u) \equiv \AbsPost{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} 
 (q \rightarrow \bot)
@@
 other words, there exists a finite sequence $u \in \Sigma^*$ such that
-$\nu \in \abs{\Inv}(\beta)$ and $\beta\cup\nu \models
+$\nu \in \abs{\Inv}(\beta)$ and $\I,\beta\cup\nu \models
 \AbsAccept{\A}(u)$, for some boolean $\beta : Q \rightarrow \booli$
@@
 invariant by excluding from $\abs{\Inv}$ all pairs $(\beta,\nu)$ such
-that $\beta\cup\nu \models \AbsAccept{\A}(u)$. This strenghtening
+that $\I,\beta\cup\nu \models \AbsAccept{\A}(u)$. This strenghtening
 is carried out by adding to $\preds$ several predicates that are
@@
 which are defined inductively below: \begin{compactitem}
-\item $R_0 = \fv{\Bool}{\theta_0}$, 
+\item $R_0 = \fv{\Bool}{\theta_0}$ and 
 %
@@
 \fv{\Bool}{\phi_{k-1}}$ by the formula $\Delta(q,a_k)$ given by the
-transition function of $\A$. The main idea behind the above definition
-of the interpolation problem is that we would like to obtain an
-interpolant $\tuple{\top,I_1(Q), I_2(Q,\vec{x}), \ldots,
-  I_{n}(Q,\vec{x}),\bot}$ whose formulae \emph{combine states with the
-  data constraints that must hold locally}, whenever the control
-reaches a certain boolean configuration. This association of states
-with data valuations is tantamount to defining efficient
-semi-algorithms, based on lazy abstraction \cite{HJMS02}. Observe
-that, because the states are replaced with transition formulae when
-moving one step in a run, the formulae thus produced lose track of the
-control history and are not directly suitable to producing
-interpolants that can be applied locally. Furthermore, the abstraction
-defined by the interpolants generated in this way can also
-\emph{over-approximate the control structure} of an automaton, in
-addition to the sets of data values encountered throughout its runs.
+transition function of $\A$. Observe that, because the states are
+replaced with transition formulae when moving one step in a run, the
+formulae thus produced lose track of the control history and are not
+directly suitable to producing interpolants that relates states and
+data.
+
+The main idea behind the above definition of the interpolation problem
+is that we would like to obtain an interpolant $\tuple{\top,I_1(Q),
+  I_2(Q,\vec{x}), \ldots, I_{n}(Q,\vec{x}),\bot}$ whose formulae
+\emph{combine states with the data constraints that must hold
+  locally}, whenever the control reaches a certain boolean
+configuration. This association of states with data valuations is
+tantamount to defining efficient semi-algorithms, based on lazy
+abstraction \cite{HJMS02}. Furthermore, the abstraction defined by the
+interpolants generated in this way can also \emph{over-approximate the
+  control structure} of an automaton, in addition to the sets of data
+values encountered throughout its runs.
 
@@
 \begin{lemma}\label{lemma:progress}
-  Given a non-empty sequence of input events $u \in \Sigma^*$ such
-  that $\Accept{\A}(u)$ is unsatisfiable, then the following
+  For any non-empty sequence $u = a_1\ldots a_n \in \Sigma^*$
+  such that $\Accept{\A}(u)$ is unsatisfiable, the following
   hold: \begin{compactenum}
@@
 \end{lemma}
-\proof{ Let $u = a_1\ldots a_n$ and  $\pi(u) \equiv \theta_0(Q_0) \wedge
-  \theta_1(Q_0 \cup Q_1,\vec{x}_0 \cup \vec{x}_1) \wedge \ldots \wedge
-  \theta_n(Q_{n-1} \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge
-  \theta_{n+1}(Q_n)$ in the following.
+\proof{ Let $\pi(u) \equiv \theta_0(Q_0) \wedge \theta_1(Q_0 \cup
+  Q_1,\vec{x}_0 \cup \vec{x}_1) \wedge \ldots \wedge \theta_n(Q_{n-1}
+  \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n)$ in
+  the following.
 
@@
   $\vec{x}_0,\ldots,\vec{x}_n$, respectively, such that
-  $\bigcup_{i=0}^n \beta_i \cup \bigcup_{i=0}^n \nu_i \models
+  $\I,\bigcup_{i=0}^n \beta_i \cup \bigcup_{i=0}^n \nu_i \models
   \pi(u)$. We build a model for $\Accept{\A}(u) \equiv
@@
 
-  But since $\beta_{n-1} \cup \beta_n, \nu_{n-1} \cup \nu_n \models
+  But since $\I,\beta_{n-1} \cup \beta_n, \nu_{n-1} \cup \nu_n \models
   \theta_n(Q_{n-1}\cup Q_n, \vec{x}_{n-1} \cup \vec{x}_n)$, we have
-  that $\beta_n,\vec{x}_{n-1} \cup \vec{x}_n \models
+  that $\I,\beta_n,\nu_{n-1} \cup \nu_n \models
   \Delta(q,a_n)[Q_n/Q,\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
   for every $q \in Q$ such that $\beta_{n-1}(q_{n-1}) = \true$. Then
-  we have $\beta,\vec{x}_0\cup\ldots\cup\vec{x}_n \models \phi_n$
+  we have $\I,\beta\cup\bigcup_{i=0}^n\nu_i \models \phi_n$
   where $\beta(q) = \beta_n(q_n)$, for all $q \in Q$ and $\phi_n$ is
@@
   $\Delta(a,q)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$
-  under valuation $\beta\cup\nu_{n-1}\cup\nu_n$ is not important for
-  $q \in Q$ such that $\beta(q) = \false$. We have thus obtained a
+  under the valuation $\beta\cup\nu_{n-1}\cup\nu_n$ is not important for
+  any $q \in Q$ such that $\beta(q) = \false$. We have obtained a
   model for $\phi_n$ and thus for $\Post{\A}(u)$. Since moreover,
@@
     compute:
-    \[\begin{array}{lcl}
-    \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] & \equiv & \text{ def. of $\AbsPost{\A}$} \\
-    \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] & \models \\
-    \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n & \models & \text{ ($\Asterisk$)} \\
-    \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n & \models & \text{ ind. hyp.} \\ 
-    I_n[Q_n/Q] & \hfill\text{\qed}
+    \[\begin{array}{l}
+    \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] \equiv \text{ (by definition of $\AbsPost{\A}$)} \\
+    \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] \models \text{ ($\Asterisk$)} \\
+    \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n \models \text{ (ind. hyp.) } \\
+    \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q]  \hfill\text{\qed}
     \end{array}\]
-    We prove the final point ($\Asterisk$). If
-    $\beta_n\cup\nu_{n-1}\cup\nu_n \models
-    \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q]$,
-    for some valuations $\beta_{n-1} : Q_{n-1} \rightarrow \booli$ and
+   Let us prove the final point ($\Asterisk$). If
+   $\I,\beta_n\cup\nu_{n-1}\cup\nu_n \models
+   \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_{n}/Q]$,
+   for some valuations $\beta_{n} : Q_{n-1} \rightarrow \booli$ and
    $\nu_i : \vec{x}_i \rightarrow \Data^\I$, for $i=n-1,n$, then we
-    build a valuation $\beta_{n-1}$ such that $\beta_{n-1}\cup\beta_n
-    \cup \nu_{n-1}\cup\nu_n \models \AbsPost{\A}(\iota,a_1\ldots
-    a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$, as follows. For each
-    occurrence of a subformula $\Delta(q,a_n)$ in
+   build a valuation $\beta_{n-1}$ such that
+   $\I,\beta_{n-1}\cup\beta_n \cup \nu_{n-1}\cup\nu_n \models
+   \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$,
+   as follows. For each occurrence of a subformula $\Delta(q,a_n)$ in
    $\AbsPost{\A}(\iota,a_1\ldots a_{n-1})$ we set
-    $\beta_{n-1}(q_{n-1}) = \true$ if $\beta_n\cup\nu_{n-1}\cup\nu_n
+   $\beta_{n-1}(q_{n-1}) = \true$ if $\I,\beta_n\cup\nu_{n-1}\cup\nu_n
    \models \Delta(q,a_n)[Q_n/Q]$ and $\beta_{n-1}(q_{n-1}) = \false$,
    otherwise. Since there are no negated occurrences of such
-    subformulae, the definition of $\beta_{n-1}$ is consistent, and
-    the check $\beta_{n-1}\cup\beta_n \cup \nu_{n-1}\cup\nu_n \models
+   subformulae, the definition of $\beta_{n-1}$ is consistent, and the
+   check $\I,\beta_{n-1}\cup\beta_n \cup \nu_{n-1}\cup\nu_n \models
    \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$
@@
 checking semi-algorithm for alternating data
-automata. Algorithm\footnote{Although termination is not guaranteed,
-  we call it an algorithm for conciseness.}  \ref{alg:predabs} builds
+automata. Algorithm\footnote{Though termination is not guaranteed,
+  we call it algorithm for conciseness.}  \ref{alg:predabs} builds
 an \emph{abstract reachability tree} (ART) whose edges are labeled
@@
   \State $\preds \leftarrow \set{\bot}$, $\worklist \leftarrow
-  \tuple{\rootNode}$, $\Art = \tuple{N,E,\Lambda,\lhd}$, where $N \leftarrow
+  \tuple{\rootNode}$, $\Art = \tuple{N,E,\Lambda,\lhd}$, $N \leftarrow
   \emptyset$, $E \leftarrow \emptyset$, $\Lambda \leftarrow
-  \set{(\rootNode,\iota)}$ and $\lhd \leftarrow \emptyset$
+  \set{(\rootNode,\iota)}$, $\lhd \leftarrow \emptyset$
 
   \While {$\worklist \neq \emptyset$}
+  \label{ln:while}
 
@@
 
-  \State let $\beta \cup \nu_1 \cup \ldots \cup \nu_k \models
-  \Accept{\A}(u)$, where $\beta : Q \rightarrow \booli$,
-  $\nu_1,\ldots,\nu_k : \vec{x} \rightarrow \Data^\I$
+  \State get model $\beta : Q \rightarrow \booli$, $\{\nu_i :
+  \vec{x}_i \rightarrow \Data^\I \mid i=0\ldots k\}$ of $\Accept{\A}(u)$
 
   \State {\bf return} $w = (a_1,\nu_1) \ldots (a_k,\nu_k)$
+  \label{ln:cex}
   \Comment{$w \in L(\A)$ by construction}
@@
 
-  \State let $p \leftarrow \Call{FindPivot}{u,\Art}$
+  \State $p \leftarrow \Call{FindPivot}{u,\Art}$
 
@@
   \State enqueue $p$ into $\worklist$
+  \label{ln:refine}
+  \Comment{recompute the subtree rooted at $p$}
 
@@
 
-  \Else
+  \Else \label{ln:else}
 
-  \For{$a \in \Sigma$}
+  \For{$a \in \Sigma$}\label{ln:forall-events}
 
-  \State let $\phi \leftarrow \AbsPost{\A}(\Lambda(n),a)$
+  \State $\phi \leftarrow \AbsPost{\A}(\Lambda(n),a)$
 
-  \If{exist $m \in N$ such that $\phi \models \Lambda(m)$}
+  \If{exist $m \in N$ such that $\phi \models \Lambda(m)$} \label{ln:covered}
 
-  \State $\lhd \leftarrow \lhd \cup \set{(n,m)}$ \Comment{$n$ covers $m$}
+  \State $\lhd \leftarrow \lhd \cup \set{(n,m)}$ \Comment{$m$ covers $n$}
+  \label{ln:direct-cover}
 
@@
 
-  \State let $s$ be a fresh node and $\Lambda \leftarrow \Lambda \cup
-  \set{(s,\phi)}$
+  \State let $s$ be a fresh node
+
+  \State $E \leftarrow E \cup \set{(n,a,s)}$
+
+  \State $\Lambda \leftarrow \Lambda \cup \set{(s,\phi)}$
 
@@
 
-  \For{$m \in N$ such that $(m,r) \in E$}
+  \For{$m \in N$ such that $(m,b,r) \in E$, $b \in \Sigma$}
 
-  \State let $\lhd \leftarrow \lhd \cup \set{(m,s)}$ \Comment{redirect covered children from $R$ into $s$}
+  \State $\lhd \leftarrow \lhd \cup \set{(m,s)}$ 
+  \label{ln:child-cover}
+  \Comment{redirect covered children from $R$ into $s$}
 
@@
 
-  \State let $\lhd \leftarrow \lhd \cup \set{(m,s)}$ \Comment{redirect covered nodes from $R$ into $s$}
+  \State $\lhd \leftarrow \lhd \cup \set{(m,s)}$ 
+  \label{ln:indirect-cover}
+  \Comment{redirect covered nodes from $R$ into $s$}
 
@@
 
-  \State enqueue $s$ into $\worklist$ \Comment{recompute the subtree rooted at $s$}
+  \State enqueue $s$ into $\worklist$ 
+  \label{ln:expand}
 
@@
   \State {\bf return} $\true$
+  \label{ln:true}
 \end{algorithmic}}
@@
 
-Formally, let $\Art = \tuple{N,E,\Lambda,\lhd}$ be a tree, where $N$
-is a set of nodes, $E \subseteq N \times \Sigma \times N$ is a set of
-edges, $\Lambda : N \rightarrow \Form^+(Q,\vec{x})$ is a labeling of
-the nodes with formulae, and $\lhd \subseteq N \times N$ is a set of
-covering edges. The root of the tree is labeled with the initial
+Formally, an ART $\Art = \tuple{N,E,\Lambda,\lhd}$ is a tree, where
+$N$ is a set of nodes, $E \subseteq N \times \Sigma \times N$ is a set
+of edges, $\Lambda : N \rightarrow \Form^+(Q,\vec{x})$ is a labeling
+of the nodes with formulae and $\lhd \subseteq N \times N$ is a set
+of covering edges. The root of the tree is labeled with the initial
 configuration $\iota$ and each node $n \in N$ corresponds to a unique
@@
 With these considerations, Algorithm \ref{alg:predabs} uses a worklist
-iteration to build a ART. We keep newly expanded nodes of $\Art$ in a
+iteration to build an ART. We keep newly expanded nodes of $\Art$ in a
 use a queue $\worklist$ to ensure a breadth-first exploration
@@
 use the covering relation $\lhd$ to ensure that, when a newly
-generated node is covered by a node in $N$, it is not added to the
-worklist, thus cutting the current branch of the search. Formally, for
-any two nodes $n,m \in N$, we have $n \lhd m$ iff
+generated node is covered by a node already in $N$, it is not added to
+the worklist, thus cutting the current branch of the search. Formally,
+for any two nodes $n,m \in N$, we have $n \lhd m$ iff
 $\AbsPost{\A}(\Lambda(n),a) \models \Lambda(m)$ for some $a \in
@@
 \end{theorem}
+\proof{To prove the first point, assume w.l.o.g. that $\Sigma =
+  \set{a_1,\ldots,a_{\card{\Sigma}}}$ is a total ordering of the input event
+  alphabet of $\A$ and that the loop at line \ref{ln:forall-events}
+  considers the events in this order. If $L(\A) \neq \emptyset$, let
+  $(a_{i_1},\nu_1) \ldots (a_{i_k},\nu_k) \in L(\A)$ be a word, such
+  that $a_{i_1} \ldots a_{i_k}$ is the lexicographically minimal
+  feasible counterexample for $\A$.  Since Algorithm \ref{alg:predabs}
+  explores $\Sigma^*$ in this lexicographical order, it must encounter
+  $a_{i_1} \ldots a_{i_k}$ and return a word $(a_{i_1},\nu'_1) \ldots
+  (a_{i_k},\nu'_k) \in L(\A)$ at line \ref{ln:cex}.
+
+  For the second point, we prove the following invariant: each time
+  Algorithm \ref{alg:predabs} reaches line \ref{ln:while}, the set $W$
+  of nodes in $\worklist$ contains all the frontier nodes in $\Art =
+  \tuple{N,E,\Lambda,\lhd}$ which are not covered by some node in $N$,
+  namely that $W = \set{n \in N \mid \forall m \in N \forall a \in
+    \Sigma ~.~ (n,a,m) \not\in E ~\wedge~ (n,m) \not\in
+    \lhd}$. Initially, this is the case because $W = \set{\rootNode}$
+  and $E = \lhd = \emptyset$. If the invariant holds previously, at
+  line \ref{ln:while}, it will hold again after line \ref{ln:refine}
+  is executed, because, when the subtree rooted at the pivot $p$ is
+  removed, $p$ becomes a member of the set of uncovered frontier
+  nodes, and is added to $W$ at line \ref{ln:refine}. The last case
+  considers that the invariant holds at line \ref{ln:while} and the
+  control follows the else branch at line \ref{ln:else}. In this case,
+  the newly created frontier node $s$ is added to $W$ only if it is
+  not covered by an existing node in $N$ (line \ref{ln:covered}).
+
+  We use this invariant in the proof as follows. If Algorithm
+  \ref{alg:predabs} returns at line \ref{ln:true}, then it must be
+  that $W = \emptyset$. Because of the invariant, each node in $N$ is
+  either covered by another node in $N$, or all its successors are in
+  $N$. We prove next that $\bigvee_{n \in N} \Lambda(n)$ defines a
+  safety invariant, which allows to conclude that $L(\A) = \emptyset$,
+  by Lemma \ref{lemma:safety-invariant}. To this end, it is sufficient
+  to prove that, for any $u \in \Sigma^*$, there exists some node $n
+  \in N$ such that $\Post{\A}(\iota,u) \models \Lambda(n)$. Let $u \in
+  \Sigma^*$ be an arbitrary sequence. If $u$ labels the path from
+  $\rootNode$ to some $n \in N$, we have $\Post{\A}(\iota,u) \models
+  \AbsPost{\A}(\iota,u) \models \Lambda(n)$ and we are
+  done. Otherwise, let $v$ be the (possibly empty) prefix of $u$ which
+  labels the path from $\rootNode$ to some $n \in N$, which is covered
+  by another $m \in N$, where $(n,a,m) \in E$, that is $u = vav'$, for
+  some $a \in \Sigma$ and $v'\in\Sigma^*$. Moreover, we have
+  $\Post{\A}(\iota,va) \models \AbsPost{\A}(\iota,va) \models
+  \Lambda(m)$, by the construction of the set $\lhd$ of covering edges
+  --- lines \ref{ln:direct-cover}, \ref{ln:child-cover},
+  \ref{ln:indirect-cover}. Continuing this argument recursively from
+  $m$, since $\len{v'} < \len{u}$, we shall eventually discover a node
+  $p$ such that $\Post{\A}(\iota,u) \models \Lambda(p)$. This
+  concludes the proof. \qed}
 

### 2017-05-08 draft.tex
abstract

@@
 \begin{abstract}
+Alternating automata have been widely used to model and verify systems
+that handle data from finite domains, such as communication protocols
+or hardware. The main advantage of the alternating model of
+computation is that complementation is possible in linear time, thus
+allowing to concisely encode trace inclusion problems that occur often
+in verification. In this paper we consider alternating automata over
+infinite alphabets, whose transition rules are formulae in a combined
+theory of booleans and some infinite data domain, that relate past and
+current values of the data variables. The data theory is not fixed,
+but rather it is a parameter of the class. We show that union,
+intersection and complementation are possible in linear time in this
+model and, though the emptiness problem is undecidable, we provide two
+efficient semi-algorithms, inspired by two state-of-the-art
+abstraction refinement model checking methods: lazy predicate
+abstraction and the IMPACT algorithm. We have implemented both methods
+and report the results of an experimental comparison.
 \end{abstract}

### 2017-05-09 draft.tex
ok

@@
 abstraction refinement model checking methods: lazy predicate
-abstraction and the IMPACT algorithm. We have implemented both methods
-and report the results of an experimental comparison.
+abstraction \cite{HJMS02} and the \impact~ semi-algorithm
+\cite{mcmillan06}. We have implemented both methods and report the
+results of an experimental comparison.
 \end{abstract}
@@
 \in \ssorts{\sig}$. If $\#(f)=0$ then $f$ is a \emph{constant}. We
-consider the constants $\mathbf{0}$ of sort $\Data$ and $\top,\bot$ of
-sort $\Bool$.
+consider the constants $\top$ and $\bot$ of sort $\Bool$.
 
@@
 $\Sigma[\vec{x}] = \Sigma \times (\vec{x} \mapsto \Data^\I)$ be the
-set of \emph{data symbols}. A \emph{data word} (or, simply, a word) is
+set of \emph{data symbols}. A \emph{data word} (word in the sequel) is
 a finite sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data
@@
 
-An \emph{alternating data automaton} (ADA) or, simply, an automaton,
-is a tuple $\A =
-\tuple{\vec{x},Q,\iota,F,\Delta}$, where: \begin{compactitem}
+An \emph{alternating data automaton} (ADA or automaton in the sequel)
+is a tuple $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$,
+where: \begin{compactitem}
 %
@@
 
-Consequently, we give up on the guarantee for termination and develop
-semi-algorithms that are supposed to be: \begin{compactenum}
-\item \emph{sound}: given an automaton $\A$, if $L(\A) \neq\emptyset$,
-  the procedure will terminate and return a word $w \in L(\A)$, and
-%
-\item \emph{complete relative to termination}: if the procedure terminates
-  without returning such a word, then $L(\A) = \emptyset$. 
+Consequently, we give up on the guarantee for termination and build
+semi-algorithms that meet the requirements below: \begin{compactenum}[(i)]
+\item given an automaton $\A$, if $L(\A) \neq\emptyset$, the procedure
+  will terminate and return a word $w \in L(\A)$, and
+%
+\item if the procedure terminates without returning such a word, then
+  $L(\A) = \emptyset$.
 \end{compactenum}
@@
 invariant computations. However, it is worth pointing out from the
-start that abstraction will allow us to build invariants which are not
-necessarily safety invariants. To meet the latter condition, we resort
-to counterexample guided abstraction refinement (CEGAR).
+start that abstraction alone will only allow us to build invariants
+that are not necessarily safety invariants. To meet the latter
+condition, we resort to counterexample guided abstraction refinement
+(CEGAR).
 
@@
 \end{inparaenum}
+
+\paragraph{\em Remark} 
+We deviate from the standard notion of \emph{generalized interpolants}
+\cite{mcmillan06}, which requires $\fv{}{I_i} \subseteq \fv{}{\phi_i}
+\cap \fv{}{\phi_{i+1}}$, for each $i \in [1,n-1]$, and assume further
+that the variables have been already renamed by removing the
+time-stamps. \qed
+
 A classical method for abstraction refinement is to add the elements
@@
 {\scriptsize\begin{algorithmic}[1]
-  \State $\preds \leftarrow \set{\bot}$, $\worklist \leftarrow
-  \tuple{\rootNode}$, $\Art = \tuple{N,E,\Lambda,\lhd}$, $N \leftarrow
-  \emptyset$, $E \leftarrow \emptyset$, $\Lambda \leftarrow
-  \set{(\rootNode,\iota)}$, $\lhd \leftarrow \emptyset$
+
+  \State let $\Art = \tuple{N,E,\Lambda,\lhd}$ be an ART
+
+  \State initially $N = E = \lhd = \emptyset$, $\Lambda =
+  \set{(\rootNode,\iota)}$, $\preds = \set{\bot}$, $\worklist =
+  \tuple{\rootNode}$,
 
@@
   \If{$\Lambda(n) \wedge \bigwedge_{q \in Q \setminus F} (q
-    \rightarrow \bot)$ is satisfiable} \Comment{counterexample found}
+    \rightarrow \bot)$ is satisfiable} \Comment{counterexample candidate}
 
@@
   \State $p \leftarrow \Call{FindPivot}{u,\Art}$
+  \label{ln:pivot}
 
@@
   \State $\preds \leftarrow \preds \cup \set{I_1,\ldots,I_\ell}$,
-  where $\tuple{\top,I_1,\ldots,I_\ell,\bot}$ is the interpolant for
+  where $\tuple{\top,I_1,\ldots,I_\ell,\bot}$ is an interpolant for
   $\overline{\pi}(v)$
@@
   \For{$a \in \Sigma$}\label{ln:forall-events}
+  \Comment{expand $n$}
 
@@
 of edges, $\Lambda : N \rightarrow \Form^+(Q,\vec{x})$ is a labeling
-of the nodes with formulae and $\lhd \subseteq N \times N$ is a set
-of covering edges. The root of the tree is labeled with the initial
-configuration $\iota$ and each node $n \in N$ corresponds to a unique
-path from the root to $n$, labeled by a sequence $u$ of input
-events. The least unfeasible suffix of $u$ is the smallest sequence $v
-= a_1 \ldots a_k$, such that $u = wv$, for some $w \in \Sigma^*$ and
-the formula (\ref{eq:pivot}) below is unsatisfiable. The \emph{pivot}
-of $u$ is the node $p$ corresponding to the start of $v$ in $u$.
+of the nodes with formulae and $\lhd \subseteq N \times N$ is a set of
+covering edges. We require that $(N,E)$ is a directed tree, rooted at
+$\rootNode \in N$, where $\Lambda(\rootNode) = \iota$. Then each node
+$n \in N$ corresponds to a unique path from the root to $n$, labeled
+by a sequence $u$ of input events. The least unfeasible suffix of $u$
+is the smallest sequence $v = a_1 \ldots a_k$, such that $u = wv$, for
+some $w \in \Sigma^*$ and the formula (\ref{eq:pivot}) below is
+unsatisfiable. The \emph{pivot} of $u$ is the node $p$ corresponding
+to the start of $v$ in $u$.
 \begin{equation}\label{eq:pivot}
@@
 (since $u$ is feasible, clearly $\Accept{\A}(u)$ is
-satisfiable). Otherwise, if $u$ is spurious, we compute its pivot $p$,
-add the interpolants for the least unfeasible suffix of $u$ to the set
-of predicates $\preds$, remove and recompute the subtree of $\Art$
-rooted at $p$.
+satisfiable). Otherwise, if $u$ is spurious, we compute its pivot $p$
+(line \ref{ln:pivot}), add the interpolants for the least unfeasible
+suffix of $u$ to the set of predicates $\preds$, remove and recompute
+the subtree of $\Art$ rooted at $p$.
 
@@
 \Sigma$, in other words, if $n$ has a successor whose label entails
-the label of $m$. The soundness and relative completness of Algorithm
-\ref{alg:predabs} are proved by the following theorem:
+the label of $m$. The correctness criteria of Algorithm
+\ref{alg:predabs} are showed in the following theorem:
 
@@
 
+\section{Checking ADA Emptiness with \impact}
+
+As pointed out by a number of authors (e.g.\ \cite{mcmillan06}) the
+bottleneck of predicate abstraction is the high cost of existential
+quantifier elimination required to compute the $\Post{\A}(\phi,a)$
+function --- we recall that $\Post{\A}(\phi,a)$ is defined by the
+formula $\overline{\vec{x}} ~.~
+\Delta(\phi[\overline{\vec{x}}/\vec{x}],a)$. The main idea of the
+\impact~ procedure is that interpolation can be used instead
+quantifier elimination, because an interpolant over-approximates the
+post-images along the counterexample path in a way that is sufficient
+to prove its spuriousness.
+
+With this in mind, Algorithm \ref{alg:impact} builds an ART for the
+given automaton and uses interpolants, instead of abstract
+post-images, to label its nodes. To speedup the computation we use two
+additional labeling functions for bookkeeping. Formally, the ART
+structure used by Algorithm \ref{alg:impact} is defined as $\Art =
+\tuple{N,E,\Lambda,R,T,\lhd}$, where $N,E,\Lambda$ and $\lhd$ are
+defined as for Algorithm \ref{alg:predabs} and,
+moreover: \begin{compactitem}
+\item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
+  sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$, 
+%
+\item $T : E \rightarrow \bigcup_{i=0}^\infty
+  \Form(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is labeling of edges
+  with time-stamped formulae, in which $Q_i$ may occur under negation,
+  and
+%
+\item for each edge $e = (n,a,m) \in E$, if $n$ lies at
+  depth\footnote{We consider that the root node has zero depth.}
+  $k\geq0$ in the tree $(N,E)$, then we have:
+  \[\begin{array}{rcl}
+  T(e) & = & \Theta(n,k) \\
+  R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)} \text{ where} \\
+  \Theta(n,k) & \equiv & \bigwedge_{q \in R(n)} (q_k \rightarrow
+  \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}])
+  \end{array}\]
+\end{compactitem}
+For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is
+$(n,a,m)$. The idea behind this additional bookkeeping is that, given
+a sequence $u = a_1\ldots a_k \in \Sigma^*$ that labels the path
+from the root to a node $n$ in $\Art$, the conjunction: 
+\[\mu(n) \equiv \Lambda(\rootNode) \wedge T(\rootNode,a_1,n_1) \wedge \ldots
+T(n_{k-1},a_k,n_k) \wedge \bigwedge_{q \in R(n_k) \setminus F}(q
+\rightarrow \bot)\] is equivalent to the interpolation problem
+$\pi(u)$. We assume furthermore a total alphabetical order $\prec$ on
+$\Sigma$ and lift it to the total lexicographical order $\prec^*$ on
+$\Sigma^*$. We write $n \prec^* m$ if the sequence labeling the path
+from $\rootNode$ to $n$ is strictly smaller, in the lexicographical
+order, than the sequence labeling the path from $\rootNode$ to $m$. A
+node $n \in N$ is \emph{covered by $m$} if $(n,m) \in \lhd$ and simply
+covered when $m$ is not important. A node $n$ is \emph{closed} if for
+all nodes $m \in N$ such that $m \prec^* n$, either $\Lambda(n)
+\not\models \Lambda(m)$ or $(n,m) \in \lhd$.
+
+\begin{algorithm}[t!]
+{\scriptsize\begin{algorithmic}[0]
+  \State {\bf input}: an ADA $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$
+  over the alphabet $\Sigma$ of input events
+
+  \State {\bf output}: $\true$ if $L(\A)=\emptyset$ and a data word $w
+  \in L(\A)$ otherwise
+\end{algorithmic}}
+
+{\scriptsize\begin{algorithmic}[1]
+
+  \State let $\Art = \tuple{N,E,\Lambda,R,T,\lhd}$ be an ART
+
+  \State initially $N = E = T = \lhd = \emptyset$, $\Lambda =
+  \set{(\rootNode,\iota)}$, $R = \fv{\Bool}{\iota[Q_0/Q]}$,
+  $\worklist=\set{\rootNode}$
+
+  \While{$\worklist \neq \emptyset$}
+
+  \State dequeue $n$ from $\worklist$
+
+  \State $N \leftarrow N \cup \set{n}$
+
+  \For{$m \in N$ such that $m \prec^* n$}
+  \Comment{close $n$}
+
+  \If{$\Lambda(n) \models \Lambda(m)$}
+
+  \State $\lhd \leftarrow \lhd \setminus \set{ (p,q) \in \lhd \mid
+    \text{$q$ is a successor of $n$}}$
+
+  \State $\lhd \leftarrow \lhd \cup \set{(n,m)}$
+
+  \EndIf
+
+  \EndFor
+
+  \If{$n$ is not covered}
+
+  \If{$R(n) \cap F \neq \emptyset$} 
+  \Comment{counterexample candidate}
+
+  \If{$\mu(n)$ is satisfiable} 
+  \Comment{counterexample is real}
+
+  \State get model $\beta : Q_i \rightarrow \booli$, $\{\nu_i :
+  \vec{x}_i \rightarrow \Data^\I \mid i=0\ldots k\}$ of $\mu(n)$
+
+  \State {\bf return} $w = (a_1,\nu_1) \ldots (a_k,\nu_k)$
+  \Comment{$w \in L(\A)$ by construction}
+
+  \Else \Comment{spurious counterexample}
+
+  \State let $\rootNode,n_1,\ldots,n_k$ be the sequence of nodes from $\rootNode$ to $n$
+
+  \State let $\tuple{\top,I_1,\ldots,I_k,I_{k+1}\equiv\bot}$ be an interpolant for $\mu(n)$
+
+  \For{$i=1,\ldots,k+1$}
+
+  \If{$\Lambda(n_i) \not\models I_i$}
+
+  \State $\lhd \leftarrow \lhd \setminus \set{(m,n_i) \in \lhd \mid m \in N}$
+  \Comment{uncover $n_i$}
+
+  \State $\Lambda(n_i) \leftarrow \Lambda(n_i) \wedge I_i$
+  \Comment{strenghten the label of $n_i$}
+
+  \EndIf
+
+  \EndFor
+
+  \EndIf
+
+  \For{$a \in \Sigma$} 
+  \Comment{expand $n$}
+
+  \State let $s$ be a fresh node and $e \leftarrow (n,a,s)$
+
+  \State $E \leftarrow E \cup \set{e}$
+
+  \State $\Lambda \leftarrow \Lambda \cup \set{(s,\top)}$
+
+  \State $T \leftarrow T \cup \set{(e, \Theta(n,k))}$
+
+  \State $R \leftarrow R \cup \{(s,\bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)})\}$
+
+  \State enqueue $s$ into $\worklist$
+
+  \EndFor
+
+  \EndIf
+
+  \EndIf
+
+  \EndWhile  
+\end{algorithmic}}
+\caption{\impact~ for ADA Emptiness}
+\label{alg:impact}
+\end{algorithm}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-05-10 draft.tex
ok

@@
 $\Bool$ and $\Data$, respectively, we denote by
-$\Form^+(\vec{b},\vec{x})$ the set of formulae $\phi$ such that
-$\fv{\Bool}{\phi} \subseteq \vec{b}$, $\fv{\Data}{\phi} \subseteq
-\vec{x}$ and each variable from $\vec{b}$ occurs under an even number
-of negations.
+$\Form(\vec{b},\vec{x})$ the set of formulae $\phi$ such that
+$\fv{\Bool}{\phi} \subseteq \vec{b}$ and $\fv{\Data}{\phi} \subseteq
+\vec{x}$. By $\Form^+(\vec{b},\vec{x})$ we denote the set of formulae
+from $\Form(\vec{b},\vec{x})$ in which each boolean variable occurs
+under an even number of negations.
 
@@
 
+\paragraph{\em Remark}
+While computing the concrete post-image $\Post{\A}$ relies on an
+expensive quantifier elimination, it is common belief that this should
+hold also for $\AbsPost{\A}$. However, the choice of conjoined
+predicates that form $\AbsPost{\A}$ requires only a quantifier-free
+satisfiability query: $\Post{\A}(\phi,a) \models \pi \iff \exists
+\overline{\vec{x}} ~.~ \Delta(\phi[\overline{\vec{x}}/\vec{x}],a)
+\models \pi \iff \Delta(\phi[\overline{\vec{x}}/\vec{x}],a) \wedge
+\neg\pi$ unsatisfiable. \qed
+
 We are left with fulfilling the second point from the definition of a
@@
 
-As pointed out by a number of authors (e.g.\ \cite{mcmillan06}) the
-bottleneck of predicate abstraction is the high cost of existential
-quantifier elimination required to compute the $\Post{\A}(\phi,a)$
-function --- we recall that $\Post{\A}(\phi,a)$ is defined by the
-formula $\overline{\vec{x}} ~.~
-\Delta(\phi[\overline{\vec{x}}/\vec{x}],a)$. The main idea of the
-\impact~ procedure is that interpolation can be used instead
-quantifier elimination, because an interpolant over-approximates the
-post-images along the counterexample path in a way that is sufficient
-to prove its spuriousness.
-
-With this in mind, Algorithm \ref{alg:impact} builds an ART for the
-given automaton and uses interpolants, instead of abstract
-post-images, to label its nodes. To speedup the computation we use two
+As pointed out by a number of authors, the bottleneck of predicate
+abstraction is the high cost of reconstructing parts of the ART,
+subsequent to the refinement of the set of predicates. The main idea
+of the \impact~ procedure \cite{mcmillan06} is that this can be
+avoided and the refinement (strenghtening of the node labels of the
+ART) can be performed in-place. This refinement step requires an
+update of the covering relation, because a node that used to cover
+another node might not cover it after the strenghtening of its label.
+
+The ART built by Algorithm \ref{alg:impact} corresponds to an
+unfolding of the input automaton by enumerating the input event
+sequences in $\Sigma^*$. Each time a new node is created, its label is
+initially $\top$. This label is subject to subsequent strenghtening,
+when a spurious counterexample is encountered. To speedup the
+computation of interpolants for spurious counterexamples, we use two
 additional labeling functions for bookkeeping. Formally, the ART
@@
   \Form(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is labeling of edges
-  with time-stamped formulae, in which $Q_i$ may occur under negation,
-  and
+  with time-stamped formulae, and
 %
 \item for each edge $e = (n,a,m) \in E$, if $n$ lies at
-  depth\footnote{We consider that the root node has zero depth.}
-  $k\geq0$ in the tree $(N,E)$, then we have:
+  depth\footnote{We consider that the root node has zero depth.}  $k$
+  in the tree $(N,E)$, then we have:
   \[\begin{array}{rcl}
@@
 For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is
-$(n,a,m)$. The idea behind this additional bookkeeping is that, given
-a sequence $u = a_1\ldots a_k \in \Sigma^*$ that labels the path
-from the root to a node $n$ in $\Art$, the conjunction: 
+$(n,a,m)$. This additional bookkeeping is used as follows. Given a
+sequence $u = a_1\ldots a_k \in \Sigma^*$ that labels the path from
+the root to a node $n$ in $\Art$, the conjunction:
 \[\mu(n) \equiv \Lambda(\rootNode) \wedge T(\rootNode,a_1,n_1) \wedge \ldots
 T(n_{k-1},a_k,n_k) \wedge \bigwedge_{q \in R(n_k) \setminus F}(q
-\rightarrow \bot)\] is equivalent to the interpolation problem
-$\pi(u)$. We assume furthermore a total alphabetical order $\prec$ on
-$\Sigma$ and lift it to the total lexicographical order $\prec^*$ on
-$\Sigma^*$. We write $n \prec^* m$ if the sequence labeling the path
-from $\rootNode$ to $n$ is strictly smaller, in the lexicographical
-order, than the sequence labeling the path from $\rootNode$ to $m$. A
-node $n \in N$ is \emph{covered by $m$} if $(n,m) \in \lhd$ and simply
-covered when $m$ is not important. A node $n$ is \emph{closed} if for
-all nodes $m \in N$ such that $m \prec^* n$, either $\Lambda(n)
-\not\models \Lambda(m)$ or $(n,m) \in \lhd$.
+\rightarrow \bot)\] defines the interpolation problem $\pi(u)$.
+
+We consider a total alphabetical order $\prec$ on $\Sigma$ and lift it
+to the total lexicographical order $\prec^*$ on $\Sigma^*$. We write
+$n \prec^* m$ if the sequence labeling the path from $\rootNode$ to
+$n$ is strictly smaller, in the lexicographical order, than the
+sequence labeling the path from $\rootNode$ to $m$. A node $n \in N$
+is \emph{covered by $m$} if $(n,m) \in \lhd$. We say that $n$ is just
+\emph{covered} when $m$ is not important. A node $n$ is \emph{closed}
+if for all nodes $m \in N$ such that $n \prec^* m$, either $\Lambda(n)
+\not\models \Lambda(m)$ or $(n,m) \in \lhd$ holds. To efficiently
+check that a node is closed, one may keep all the nodes from the ART
+in a list, in preorder.
 
@@
 
-  \For{$m \in N$ such that $m \prec^* n$}
+  \For{$m \in N$ such that $n \prec^* m$ and $n$ is not an ancestor of $m$}
   \Comment{close $n$}
@@
 
+  \State {\bf goto} \ref{ln:not-covered}
+
   \EndIf
@@
   \If{$n$ is not covered}
+  \label{ln:not-covered}
 
@@
 
-  \State let $\rootNode,n_1,\ldots,n_k$ be the sequence of nodes from $\rootNode$ to $n$
+  \State let $\rootNode,n_1,\ldots,n_k,n$ be the sequence of nodes from $\rootNode$ to $n$
 
@@
 
-  \State let $s$ be a fresh node and $e \leftarrow (n,a,s)$
+  \State let $s$ be a fresh node and $e = (n,a,s)$ be a new edge
 
@@
 
+The execution of Algorithm \ref{alg:impact} consists of three
+phases\footnote{These phases correspond to the \textsc{Close},
+  \textsc{Refine} and \textsc{Expand} procedures from the original
+  description of \impact~\cite{mcmillan06}.}: \emph{close},
+\emph{refine} and \emph{expand}.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-05-10 draft.tex
ok

@@
 
-  \For{$m \in N$ such that $n \prec^* m$ and $n$ is not an ancestor of $m$}
-  \Comment{close $n$}
-
-  \If{$\Lambda(n) \models \Lambda(m)$}
-
-  \State $\lhd \leftarrow \lhd \setminus \set{ (p,q) \in \lhd \mid
-    \text{$q$ is a successor of $n$}}$
-
-  \State $\lhd \leftarrow \lhd \cup \set{(n,m)}$
-
-  \State {\bf goto} \ref{ln:not-covered}
-
-  \EndIf
-
-  \EndFor
+  \State $\Call{Close}{n}$
+  \label{ln:close-worklist}
 
@@
 
-  \State let $\rootNode,n_1,\ldots,n_k,n$ be the sequence of nodes from $\rootNode$ to $n$
+  \State let $\rootNode,n_1,\ldots,n_k,n_{k+1}=n$ be the sequence of nodes from $\rootNode$ to $n$
 
@@
 
+  \State $\Call{Close}{n_i}$
+
   \EndIf
@@
   \EndWhile  
+\end{algorithmic}
+
+\begin{algorithmic}[1]
+\Function{Close}{$x$}
+
+\For{$y \in N$ such that $x \prec^* y$ and $x$ is not an ancestor of $y$}
+
+\If{$\Lambda(x) \models \Lambda(y)$}
+
+\State $\lhd \leftarrow \left(\lhd \setminus \set{ (p,q) \in \lhd \mid
+  \text{$q$ is a successor of $x$}}\right) \cup \set{(x,y)}$
+
+\State {\bf return}
+
+\EndIf
+
+\EndFor
+
+\EndFunction
 \end{algorithmic}}
@@
   description of \impact~\cite{mcmillan06}.}: \emph{close},
-\emph{refine} and \emph{expand}.
+\emph{refine} and \emph{expand}. When a node $n$ is removed from the
+worklist, it is first closed (line \ref{ln:close-worklist}) by finding
+a lexicographically bigger node $m$ such that $\Lambda(n) \models
+\Lambda(m)$, if one exists, and adding $(n,m)$ to $\lhd$. As in
+Algorithm \ref{alg:predabs}, the intuition is that all successors of
+$n$ are covered by the successors of $m$, thus for any feasible
+counterexample that can be discovered from $n$ there exists a feasible
+counterexample stemming in $m$, and we do not need to process $n$
+further. If $n$ has not been covered in the close phase, we check if
+it is the endpoint of a counterexample candidate, which can only be
+the case
 

### 2017-05-10 draft.tex
ok

@@
   \State let $u = a_1\ldots a_k$ be the label of the path from
-  $\rootNode$ to $n$ in $\Art$
+  $\rootNode$ to $n$
 
   \If{$\Accept{\A}(u)$ is satisfiable} \Comment{counterexample is
-    real}
+    feasible}
 
@@
 explored first. When the search encounters a counterexample $u$, it is
-checked for spuriousness. If the counterexample is real, the procedure
+checked for spuriousness. If the counterexample is feasible, the procedure
 returns a data word $w \in L(\A)$, which interleaves the input events
@@
   \If{$n$ is not covered}
-  \label{ln:not-covered}
 
   \If{$R(n) \cap F \neq \emptyset$} 
+  \label{ln:impact-cex}
   \Comment{counterexample candidate}
@@
   \If{$\mu(n)$ is satisfiable} 
-  \Comment{counterexample is real}
+  \Comment{counterexample is feasible}
 
@@
 
+  \State let $a_1 \ldots a_k$ be the label of the path from
+  $\rootNode$ to $n$
+
   \State {\bf return} $w = (a_1,\nu_1) \ldots (a_k,\nu_k)$
@@
   \State let $\rootNode,n_1,\ldots,n_k,n_{k+1}=n$ be the sequence of nodes from $\rootNode$ to $n$
+  \label{ln:refine-begin}
 
@@
   \Comment{strenghten the label of $n_i$}
+  \label{ln:close-refine}
 
   \State $\Call{Close}{n_i}$
+  \label{ln:refine-end}
 
@@
   \EndFor
+
+  \EndIf
 
@@
 
-  \EndIf
-
   \EndWhile  
@@
   \text{$q$ is a successor of $x$}}\right) \cup \set{(x,y)}$
+\label{ln:close-uncover}
 
@@
 counterexample stemming in $m$, and we do not need to process $n$
-further. If $n$ has not been covered in the close phase, we check if
-it is the endpoint of a counterexample candidate, which can only be
-the case
+further.
+
+If $n$ has not been covered in the close phase, we check if it is the
+endpoint of a counterexample candidate, which can only be the case
+when the replacement set $R(n)$ contains at least one final state
+(line \ref{ln:cex}). Observe that, if $u$ is the label of the path
+from $\rootNode$ to $n$, then $\Accept{\A}(u)$ is satisfiable only if
+it contains a (positive occurrence of) a final state. The
+counterexample is feasible if and only if $\mu(n)$ is satisfiable, in
+which case a model of $\mu(n)$ is obtained and a word $w \in L(\A)$
+returned. Otherwise, the counterexample is spurious and the procedure
+enters the refinement phase (lines \ref{ln:refine-begin} -
+\ref{ln:refine-end}). The interpolant for $\mu(n)$ is used to
+strenghten the labels of all the ancestors of $n$, by conjoining the
+formulae of the interpolant to the existing labels. In this process,
+some node might become eligible for coverage, therefore we attempt to
+close each ancestor of $n$ that is impacted by the refinement (line
+\ref{ln:close-refine}). Observe that, in this case the call to
+$\Call{Close}{}$ must uncover each node which is covered by a
+successor of $n$ (line \ref{ln:close-uncover} in the $\Call{Close}{}$
+procedure). This is required because, due to the over-approximation of
+the sets of reachable configurations, the covering relation is not
+transitive, as explained in \cite{mcmillan06}.
+
+Finally, we expand $n$ (lines \ref{ln:expand-begin} -
+\ref{ln:expand-end}) by creating a new node for each successor via an
+input event $a \in \Sigma$ and inserting it into the worklist. The
+label of the
 

### 2017-05-10 draft.tex
ok

@@
 $n$ is strictly smaller, in the lexicographical order, than the
-sequence labeling the path from $\rootNode$ to $m$. A node $n \in N$
-is \emph{covered by $m$} if $(n,m) \in \lhd$. We say that $n$ is just
-\emph{covered} when $m$ is not important. A node $n$ is \emph{closed}
-if for all nodes $m \in N$ such that $n \prec^* m$, either $\Lambda(n)
-\not\models \Lambda(m)$ or $(n,m) \in \lhd$ holds. To efficiently
-check that a node is closed, one may keep all the nodes from the ART
-in a list, in preorder.
+sequence labeling the path from $\rootNode$ to $m$.
+
+A node $n \in N$ is \emph{covered} if it has an ancestor $m$ such that
+$(m,p) \in \lhd$, for some $p \in N$. A node $n$ is \emph{closed} if
+it is covered, or $\Lambda(n) \not\models \Lambda(m)$ for all $m \in
+N$ such that $n \prec^* m$.
 
@@
   \Comment{expand $n$}
+  \label{ln:expand-begin}
 
@@
   \State enqueue $s$ into $\worklist$
+  \label{ln:expand-end}
 
@@
 Finally, we expand $n$ (lines \ref{ln:expand-begin} -
-\ref{ln:expand-end}) by creating a new node for each successor via an
-input event $a \in \Sigma$ and inserting it into the worklist. The
-label of the
+\ref{ln:expand-end}) by creating a new node for each successor $s$ via
+an input event $a \in \Sigma$ and inserting it into the worklist. The
+label of the new node is $\top$, its replacement set $R(s)$ is
+computed based on $R(n)$ and the label of the edge $(n,a,s)$ is
+$\Theta(n,k)$, where $k$ is the level of $n$ in $\Art$.
 

### 2017-05-10 draft.tex
ok

@@
 
+  \State let $b$ be $\false$
+
   \For{$i=1,\ldots,k+1$}
@@
   \Comment{strenghten the label of $n_i$}
-  \label{ln:close-refine}
-
-  \State $\Call{Close}{n_i}$
+
+  \If{$\neg b$}
+
+  \State $b \leftarrow \Call{Close}{n_i}$
   \label{ln:refine-end}
+
+  \EndIf
 
@@
 \begin{algorithmic}[1]
-\Function{Close}{$x$}
+\Function{Close}{$x$} {\bf returns} $\Bool$
 
@@
 
-\State {\bf return}
+\State {\bf return} $\true$
 
@@
 \EndFor
+
+\State {\bf return} $\false$
 
@@
 \Lambda(m)$, if one exists, and adding $(n,m)$ to $\lhd$. As in
-Algorithm \ref{alg:predabs}, the intuition is that all successors of
-$n$ are covered by the successors of $m$, thus for any feasible
+Algorithm \ref{alg:predabs}, the intuition is that for any feasible
 counterexample that can be discovered from $n$ there exists a feasible
-counterexample stemming in $m$, and we do not need to process $n$
+counterexample stemming in $m$, thus we do not need to process $n$
 further.
 
-If $n$ has not been covered in the close phase, we check if it is the
-endpoint of a counterexample candidate, which can only be the case
-when the replacement set $R(n)$ contains at least one final state
-(line \ref{ln:cex}). Observe that, if $u$ is the label of the path
-from $\rootNode$ to $n$, then $\Accept{\A}(u)$ is satisfiable only if
-it contains a (positive occurrence of) a final state. The
-counterexample is feasible if and only if $\mu(n)$ is satisfiable, in
-which case a model of $\mu(n)$ is obtained and a word $w \in L(\A)$
-returned. Otherwise, the counterexample is spurious and the procedure
-enters the refinement phase (lines \ref{ln:refine-begin} -
+If $n$ has not been previously covered at line
+\ref{ln:close-worklist}, we check if it is the endpoint of a
+counterexample candidate, which can only be the case when the
+replacement set $R(n)$ contains at least one final state (line
+\ref{ln:cex}). Observe that, if $u$ is the label of the path from
+$\rootNode$ to $n$, then $\Accept{\A}(u)$ is satisfiable only if it
+contains a (positive occurrence of) a final state. The counterexample
+is moreover feasible iff $\mu(n)$ is satisfiable, in which case a model of
+$\mu(n)$ is obtained and a word $w \in L(\A)$ returned.
+
+Otherwise, the counterexample is spurious and the procedure enters the
+refinement phase (lines \ref{ln:refine-begin} -
 \ref{ln:refine-end}). The interpolant for $\mu(n)$ is used to
@@
 close each ancestor of $n$ that is impacted by the refinement (line
-\ref{ln:close-refine}). Observe that, in this case the call to
+\ref{ln:refine-end}). Observe that, in this case the call to
 $\Call{Close}{}$ must uncover each node which is covered by a

### 2017-05-10 draft.tex
ok

@@
     \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n \models \text{ (ind. hyp.) } \\
-    \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q]  \hfill\text{\qed}
+    \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q] 
     \end{array}\]
@@
 
-  \State let $\Art = \tuple{N,E,\Lambda,\lhd}$ be an ART
+  \State let $\Art = \tuple{N,E,\Lambda,\lhd}$ be an ART with root
+  node $\rootNode$
 
@@
   Algorithm \ref{alg:predabs} reaches line \ref{ln:while}, the set $W$
-  of nodes in $\worklist$ contains all the frontier nodes in $\Art =
-  \tuple{N,E,\Lambda,\lhd}$ which are not covered by some node in $N$,
-  namely that $W = \set{n \in N \mid \forall m \in N \forall a \in
-    \Sigma ~.~ (n,a,m) \not\in E ~\wedge~ (n,m) \not\in
-    \lhd}$. Initially, this is the case because $W = \set{\rootNode}$
-  and $E = \lhd = \emptyset$. If the invariant holds previously, at
-  line \ref{ln:while}, it will hold again after line \ref{ln:refine}
-  is executed, because, when the subtree rooted at the pivot $p$ is
+  of nodes in $\worklist$ contains all the frontier nodes in the ART
+  $\tuple{N\cup W,E,\Lambda,\lhd}$ which are not covered by some node
+  in $N$, namely that:
+  \begin{equation}\label{inv:predabs}
+  W = \set{n \mid \forall m \in N~ \forall a \in \Sigma ~.~ (n,a,m)
+    \not\in E ~\wedge~ (n,m) \not\in \lhd}
+  \end{equation}
+  Initially, this is the case because $W = \set{\rootNode}$ and $N = E
+  = \lhd = \emptyset$. If the invariant holds previously, at line
+  \ref{ln:while}, it will hold again after line \ref{ln:refine} is
+  executed, because, when the subtree rooted at the pivot $p$ is
   removed, $p$ becomes a member of the set of uncovered frontier
@@
 \[\mu(n) \equiv \Lambda(\rootNode) \wedge T(\rootNode,a_1,n_1) \wedge \ldots
-T(n_{k-1},a_k,n_k) \wedge \bigwedge_{q \in R(n_k) \setminus F}(q
+T(n_{k-1},a_k,n_k) \wedge \bigwedge_{q \in R(n_k) \setminus F}(q_k
 \rightarrow \bot)\] defines the interpolation problem $\pi(u)$.
@@
 
-A node $n \in N$ is \emph{covered} if it has an ancestor $m$ such that
-$(m,p) \in \lhd$, for some $p \in N$. A node $n$ is \emph{closed} if
-it is covered, or $\Lambda(n) \not\models \Lambda(m)$ for all $m \in
-N$ such that $n \prec^* m$.
+A node $n \in N$ is \emph{covered} if $(n,p) \in \lhd$ or it has an
+ancestor $m$ such that $(m,p) \in \lhd$, for some $p \in N$. A node
+$n$ is \emph{closed} if it is covered, or $\Lambda(n) \not\models
+\Lambda(m)$ for all $m \in N$ such that $n \prec^* m$.
 
@@
 
-  \State let $\Art = \tuple{N,E,\Lambda,R,T,\lhd}$ be an ART
+  \State let $\Art = \tuple{N,E,\Lambda,R,T,\lhd}$ be an ART with root
+  node $\rootNode$
 
@@
   \While{$\worklist \neq \emptyset$}
+  \label{ln:impact-while}
 
@@
   \If{$n$ is not covered}
+
+  \State let $(\rootNode,a_1,n_1),(n_1,a_2,n_2), \ldots,
+  (n_{k-1},a_k,n)$ be the path from $\rootNode$ to $n$
 
@@
 
-  \State let $a_1 \ldots a_k$ be the label of the path from
-  $\rootNode$ to $n$
-
   \State {\bf return} $w = (a_1,\nu_1) \ldots (a_k,\nu_k)$
@@
 
-  \State let $\rootNode,n_1,\ldots,n_k,n_{k+1}=n$ be the sequence of nodes from $\rootNode$ to $n$
+  \State let $\tuple{\top,I_1,\ldots,I_k \equiv\bot}$ be an interpolant for $\mu(n)$
   \label{ln:refine-begin}
 
-  \State let $\tuple{\top,I_1,\ldots,I_k,I_{k+1}\equiv\bot}$ be an interpolant for $\mu(n)$
-
   \State let $b$ be $\false$
 
-  \For{$i=1,\ldots,k+1$}
+  \For{$i=1,\ldots,k$}
 
@@
   \State $\lhd \leftarrow \lhd \setminus \set{(m,n_i) \in \lhd \mid m \in N}$
-  \Comment{uncover $n_i$}
 
@@
   \State $E \leftarrow E \cup \set{e}$
+  \label{ln:edge-insert}
 
@@
 \State $\lhd \leftarrow \left(\lhd \setminus \set{ (p,q) \in \lhd \mid
-  \text{$q$ is a successor of $x$}}\right) \cup \set{(x,y)}$
+  \text{$q$ is $x$ or a successor of $x$}}\right) \cup \set{(x,y)}$
 \label{ln:close-uncover}
@@
 a lexicographically bigger node $m$ such that $\Lambda(n) \models
-\Lambda(m)$, if one exists, and adding $(n,m)$ to $\lhd$. As in
-Algorithm \ref{alg:predabs}, the intuition is that for any feasible
+\Lambda(m)$, if one exists, and adding the pair $(n,m)$ to $\lhd$. As
+in Algorithm \ref{alg:predabs}, the intuition is that for any feasible
 counterexample that can be discovered from $n$ there exists a feasible
@@
 
-If $n$ has not been previously covered at line
-\ref{ln:close-worklist}, we check if it is the endpoint of a
+If $n$ is not covered, we check if it is the endpoint of a
 counterexample candidate, which can only be the case when the
@@
 contains a (positive occurrence of) a final state. The counterexample
-is moreover feasible iff $\mu(n)$ is satisfiable, in which case a model of
-$\mu(n)$ is obtained and a word $w \in L(\A)$ returned.
+is moreover feasible iff $\mu(n)$ is satisfiable, in which case a
+model of $\mu(n)$ is obtained and a word $w \in L(\A)$ returned.
 
@@
 strenghten the labels of all the ancestors of $n$, by conjoining the
-formulae of the interpolant to the existing labels. In this process,
-some node might become eligible for coverage, therefore we attempt to
-close each ancestor of $n$ that is impacted by the refinement (line
-\ref{ln:refine-end}). Observe that, in this case the call to
-$\Call{Close}{}$ must uncover each node which is covered by a
+formulae of the interpolant to the existing labels. Then $\Lambda(n)$
+becomes equivalent to $\bot$.
+
+In this process, the nodes on the path between $\rootNode$ and $n$,
+including $n$, might become eligible for coverage, therefore we
+attempt to close each ancestor of $n$ that is impacted by the
+refinement (line \ref{ln:refine-end}). Observe that, in this case the
+call to $\Call{Close}{}$ must uncover each node which is covered by a
 successor of $n$ (line \ref{ln:close-uncover} in the $\Call{Close}{}$
@@
 the sets of reachable configurations, the covering relation is not
-transitive, as explained in \cite{mcmillan06}.
+transitive, as explained in \cite{mcmillan06}. If $\Call{Close}{}$
+adds a covering edge $(n_i,m)$ to $\lhd$, it does not have to be
+called for the successors of $n_i$ on this path, which is handled via
+the boolean flag $b$.
 
@@
 
+The correctness of Algorithm \ref{alg:impact} is proved in the
+following:
+
+\begin{theorem}\label{thm:impact}
+  Given an automaton $\A$, such that $L(\A) \neq \emptyset$, Algorithm
+  \ref{alg:impact} terminates and returns a word $w \in L(\A)$. If
+  Algorithm \ref{alg:impact} terminates reporting $\true$, then
+  $L(\A) = \emptyset$.
+\end{theorem}
+\proof{ The first point is proved along the lines of the first point
+  of Theorem \ref{thm:predabs}. In particular, it is easy to see that
+  each call to $\Call{Close}{}$ terminates, because $N$ is finite at
+  each moment during the execution.
+
+  For the second point, we prove the following invariant: each time
+  Algorithm \ref{alg:impact} reaches the line \ref{ln:impact-while},
+  the set of nodes $W$ in $\worklist$ contains exactly those nodes in
+  the ART $\tuple{N \cup W,E,\Lambda,R,T,\lhd}$ which are
+  both\begin{inparaenum}[(i)]
+  \item uncovered, and such that
+  \item for some $a \in \Sigma$, the edge $(n,a,s)$ is missing from $E$:
+  \end{inparaenum}
+  \begin{equation}\label{inv:impact}
+  W = \set{n \mid n \text{ is uncovered and } \exists a \in \Sigma~
+    \forall s \in N ~.~ (n,a,s) \not\in E}
+  \end{equation}
+  Initially, $W = \set{\rootNode}$ and $N=E=\lhd=\emptyset$, thus the
+  invariant holds trivially. Suppose that the invariant holds at some
+  point, when reaching line \ref{ln:impact-while} and some node $n$
+  was removed from $W$ and inserted into $N$. We distinguish two
+  cases, either: \begin{itemize}
+  \item $n$ is covered at line \ref{ln:close-worklist}, in which case
+    $W$ remains unchanged, or
+  %
+  \item $n$ is not covered, in which case for each $a \in \Sigma$ a
+    new node $s$ is inserted into $W$ (line \ref{ln:expand-end}) and
+    the edge $(n,a,s)$ is inserted into $E$ (line
+    \ref{ln:edge-insert}). None of these new nodes are covered and
+    none of their successors inserted into $E$.
+  \end{itemize}
+  Then the condition (\ref{inv:impact}) holds next time the line
+  \ref{ln:impact-while} is reached, thus proving the invariant.  \qed}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-05-10 draft.tex
ok

@@
   \begin{equation}\label{inv:predabs}
-  W = \set{n \mid \forall m \in N~ \forall a \in \Sigma ~.~ (n,a,m)
+  W = \set{n \not\in N \mid \forall m \in N~ \forall a \in \Sigma ~.~ (n,a,m)
     \not\in E ~\wedge~ (n,m) \not\in \lhd}
@@
   \EndWhile
+
+  \State {\bf return} $\true$
 \end{algorithmic}
@@
   \begin{equation}\label{inv:impact}
-  W = \set{n \mid n \text{ is uncovered and } \exists a \in \Sigma~
-    \forall s \in N ~.~ (n,a,s) \not\in E}
+  W = \set{n \not\in N \mid n \text{ uncovered, } \exists a \in
+    \Sigma~ \forall s \in N ~.~ (n,a,s) \not\in E}
   \end{equation}
   Initially, $W = \set{\rootNode}$ and $N=E=\lhd=\emptyset$, thus the
-  invariant holds trivially. Suppose that the invariant holds at some
-  point, when reaching line \ref{ln:impact-while} and some node $n$
-  was removed from $W$ and inserted into $N$. We distinguish two
-  cases, either: \begin{itemize}
+  invariant holds trivially. Suppose that (\ref{inv:impact}) holds at
+  when reaching line \ref{ln:impact-while} and some node $n$ was
+  removed from $W$ and inserted into $N$. We distinguish two cases,
+  either: \begin{itemize}
   \item $n$ is covered at line \ref{ln:close-worklist}, in which case
@@
   \end{itemize}
-  Then the condition (\ref{inv:impact}) holds next time the line
-  \ref{ln:impact-while} is reached, thus proving the invariant.  \qed}
+  Then the condition (\ref{inv:impact}) holds next time line
+  \ref{ln:impact-while} is reached, thus it is invariant. If Algorithm
+  \ref{alg:impact} returns $\true$ then $W = \emptyset$ and, by
+  (\ref{inv:impact}), for each node in $n \in N$ one of the following
+  hold: \begin{compactitem}
+  \item $n$ is covered, or
+  \item for each $a \in \Sigma$ there exists $s \in N$ such that
+    $(n,a,s) \in E$. 
+  \end{compactitem}
+  We prove that, in this case, $\bigvee_{n \in N} \Lambda(n)$ defines
+  a safety invariant and conclude that $L(\A) = \emptyset$, by Lemma
+  \ref{lemma:safety-invariant}. To show that $\bigvee_{n \in N}
+  \Lambda(n)$ defines an invariant, let $u \in \Sigma^*$ be an
+  arbitrary sequence. \qed}
 

### 2017-05-10 draft.tex
ok

@@
 
-  \State let $\Art = \tuple{N,E,\Lambda,\lhd}$ be an ART with root
-  node $\rootNode$
+  \State let $\Art = \tuple{N,E,\rootNode,\Lambda,\lhd}$ be an ART
 
@@
 
-  \State let $\SubArt = \tuple{N',E',\Lambda',\lhd'}$ be the subtree
-  rooted at $p$ in $\Art$
+  \State let $\SubArt = \tuple{N',E',p,\Lambda',\lhd'}$ be the subtree
+  of $\Art$ rooted at $p$
 
@@
 
-Formally, an ART $\Art = \tuple{N,E,\Lambda,\lhd}$ is a tree, where
-$N$ is a set of nodes, $E \subseteq N \times \Sigma \times N$ is a set
-of edges, $\Lambda : N \rightarrow \Form^+(Q,\vec{x})$ is a labeling
-of the nodes with formulae and $\lhd \subseteq N \times N$ is a set of
-covering edges. We require that $(N,E)$ is a directed tree, rooted at
-$\rootNode \in N$, where $\Lambda(\rootNode) = \iota$. Then each node
-$n \in N$ corresponds to a unique path from the root to $n$, labeled
-by a sequence $u$ of input events. The least unfeasible suffix of $u$
-is the smallest sequence $v = a_1 \ldots a_k$, such that $u = wv$, for
-some $w \in \Sigma^*$ and the formula (\ref{eq:pivot}) below is
-unsatisfiable. The \emph{pivot} of $u$ is the node $p$ corresponding
-to the start of $v$ in $u$.
+Formally, an ART $\Art = \tuple{N,E,\rootNode,\Lambda,\lhd}$ is a
+tree, where $N$ is a set of nodes, $E \subseteq N \times \Sigma \times
+N$ is a set of edges, $\rootNode \in N$ is the root of the directed
+tree $(N,E)$, $\Lambda : N \rightarrow \Form^+(Q,\vec{x})$ is a
+labeling of the nodes with formulae, such that $\Lambda(\rootNode) =
+\iota$ and $\lhd \subseteq N \times N$ is a set of covering
+edges. Then each node $n \in N$ corresponds to a unique path from the
+root to $n$, labeled by a sequence $u$ of input events. The least
+unfeasible suffix of $u$ is the smallest sequence $v = a_1 \ldots
+a_k$, such that $u = wv$, for some $w \in \Sigma^*$ and the formula
+(\ref{eq:pivot}) below is unsatisfiable. The \emph{pivot} of $u$ is
+the node $p$ corresponding to the start of $v$ in $u$.
 \begin{equation}\label{eq:pivot}
@@
   of nodes in $\worklist$ contains all the frontier nodes in the ART
-  $\tuple{N\cup W,E,\Lambda,\lhd}$ which are not covered by some node
+  $\tuple{N\cup W,E,\rootNode,\Lambda,\lhd}$ which are not covered by some node
   in $N$, namely that:
@@
 structure used by Algorithm \ref{alg:impact} is defined as $\Art =
-\tuple{N,E,\Lambda,R,T,\lhd}$, where $N,E,\Lambda$ and $\lhd$ are
-defined as for Algorithm \ref{alg:predabs} and,
+\tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$, where $N,E,\Lambda$ and
+$\lhd$ are defined as for Algorithm \ref{alg:predabs} and,
 moreover: \begin{compactitem}
@@
   \[\begin{array}{rcl}
-  T(e) & = & \Theta(n,k) \\
+  T(e) & = & \Theta(n,a,k) \\
   R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)} \text{ where} \\
-  \Theta(n,k) & \equiv & \bigwedge_{q \in R(n)} (q_k \rightarrow
+  \Theta(n,a,k) & \equiv & \bigwedge_{q \in R(n)} (q_k \rightarrow
   \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}])
@@
 
-  \State let $\Art = \tuple{N,E,\Lambda,R,T,\lhd}$ be an ART with root
-  node $\rootNode$
+  \State let $\Art = \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$ be an ART
 
@@
 
-  \State $T \leftarrow T \cup \set{(e, \Theta(n,k))}$
+  \State $T \leftarrow T \cup \set{(e, \Theta(n,a,k))}$
 
@@
 Otherwise, the counterexample is spurious and the procedure enters the
-refinement phase (lines \ref{ln:refine-begin} -
-\ref{ln:refine-end}). The interpolant for $\mu(n)$ is used to
-strenghten the labels of all the ancestors of $n$, by conjoining the
-formulae of the interpolant to the existing labels. Then $\Lambda(n)$
-becomes equivalent to $\bot$.
+refinement phase (lines
+\ref{ln:refine-begin}-\ref{ln:refine-end}). The interpolant for
+$\mu(n)$ is used to strenghten the labels of all the ancestors of $n$,
+by conjoining the formulae of the interpolant to the existing
+labels. Then $\Lambda(n)$ becomes equivalent to $\bot$.
 
@@
 
-Finally, we expand $n$ (lines \ref{ln:expand-begin} -
-\ref{ln:expand-end}) by creating a new node for each successor $s$ via
-an input event $a \in \Sigma$ and inserting it into the worklist. The
-label of the new node is $\top$, its replacement set $R(s)$ is
-computed based on $R(n)$ and the label of the edge $(n,a,s)$ is
-$\Theta(n,k)$, where $k$ is the level of $n$ in $\Art$.
+Finally, we expand $n$ (lines
+\ref{ln:expand-begin}-\ref{ln:expand-end}) by creating a new node for
+each successor $s$ via the event $a \in \Sigma$ and inserting it
+into the worklist. The label of the new node is $\top$, its
+replacement set $R(s)$ is computed based on $R(n)$ and the label of
+the edge $(n,a,s)$ is $\Theta(n,a,k)$, where $k$ is the depth of $n$
+in $\Art$.
+
+The following lemma states that Algorithm \ref{alg:impact} builds an
+over-approximation of the concrete runs of $\A$.
+
+\begin{lemma}\label{lemma:impact-interpol}
+  Given an ART $\Art = \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$ built
+  by Algorithm \ref{alg:impact} and a node $n \in N$, for all $(n,a,m)
+  \in E$, we have $\Post{\A}(\Lambda(n),a) \models \Lambda(m)$.
+\end{lemma}
+\proof{ We distinguish two cases. First, if $(n,a,m)$ occurs on a path
+  in $\Art$ that has never been refined, then $\Lambda(m)=\top$ and
+  the entailment holds trivially. Otherwise, let $(n_0,a_1,n_1),
+  \ldots, (n_{k-1},a_k,n_k)$ be a path, where $n_0 = \rootNode$ and
+  $(n,a,m) = (n_{i-1},a_i,n_i)$, for some $i \in [1,k]$ and, moreover,
+  $a_1\ldots a_k$ was found, at some point, to be a spurious
+  counterexample. Let $\tuple{\top,I_1,\ldots,I_k\equiv\bot}$ be an
+  interpolant for $\mu(n_k) \equiv \Lambda(\rootNode) \wedge
+  \bigwedge_{i=1}^k \Theta(n_{i-1},a,i-1) \wedge \bigwedge_{q \in R(n_k)}
+  (q_k \rightarrow \bot)$. 
+\qed}
 
 The correctness of Algorithm \ref{alg:impact} is proved in the
-following:
+following.
 
@@
   a safety invariant and conclude that $L(\A) = \emptyset$, by Lemma
-  \ref{lemma:safety-invariant}. To show that $\bigvee_{n \in N}
-  \Lambda(n)$ defines an invariant, let $u \in \Sigma^*$ be an
-  arbitrary sequence. \qed}
+  \ref{lemma:safety-invariant}. To this end, let $u \in \Sigma^*$ be
+  an arbitrary sequence and let $v_1$ be the largest prefix of $u$
+  that labels a path from $\rootNode$ to some node $n_1 \in N$. 
+\qed}
 

### 2017-05-11 draft.tex
ok

@@
 implies $\I,\nu \models \psi$, for each valuation $\nu$, and $\phi
-\eqiff^\I \psi$ iff $\phi \models^\I \psi$ and $\psi \models^\I \phi$.
+\iffeq^\I \psi$ iff $\phi \models^\I \psi$ and $\psi \models^\I \phi$.
 We omit mentioning the interpretation $\I$ when it is clear from the
@@
 a set of time-stamped variables. For an input event $a \in \Sigma$ and
-a formula $\phi$, we write $\Delta^k(\phi,a)$ for the formula obtained
-from $\phi$ by simultaneously replacing each state $q \in
-\fv{\Bool}{\phi}$ by the formula
+a formula $\phi$, we write $\Delta(\phi)$ and, respectively,
+$\Delta^k(\phi,a)$, for the formula obtained from $\phi$ by
+simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
+formula $\Delta(q,a)$, respectively
 $\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
@@
   word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
-  \iota_2,a_1\ldots a_n) \eqiff \Delta(\iota_1,a_1\ldots a_n) \vee
+  \iota_2,a_1\ldots a_n) \iffeq \Delta(\iota_1,a_1\ldots a_n) \vee
   \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
@@
   a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
-  \iota_2,a_1\ldots a_{n-1}) \eqiff \Delta(\iota_1,a_1\ldots a_{n-1})
+  \iota_2,a_1\ldots a_{n-1}) \iffeq \Delta(\iota_1,a_1\ldots a_{n-1})
   \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
   \[\begin{array}{ll}
-  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \eqiff \\
-  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) & \eqiff \\
+  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
+  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
   \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots a_{n}) \enspace.
@@
   \[\begin{array}{lcl}
-  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \eqiff & \text{ by induction hypothesis} \\
-  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \eqiff & \text{ by the definition of $\overline{\phi}$} \\
+  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \iffeq & \text{ by induction hypothesis} \\
+  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \iffeq & \text{ by the definition of $\overline{\phi}$} \\
   \overline{\Delta(\iota_1,a_1\ldots,a_n)}
@@
 \end{compactenum}
-A formula $\phi(Q,\vec{x})$ \emph{defines} $\Inv$ iff for all $\beta:
-Q \rightarrow \booli$ and $\nu : \vec{x} \rightarrow \Data^\I$,
-$\I,\beta\cup\nu \models \phi$ if and only if $\nu \in
-\Inv(\beta)$. If $\Inv$ satisfies only the first point above, we call
-it an \emph{invariant}. Intuitively, a safety invariant maps every
-boolean valuation into a set of data valuations, that contains the
-initial configuration $\iota = \Post{\A}(\iota,\varepsilon)$, whose
-data variables are unconstrained, over-approximates the set of
-reachable valuations (\ref{it1:inv}) and excludes the valuations
-satisfying the acceptance condition (\ref{it2:inv}).
+If $\Inv$ satisfies only the first point above, we call it an
+\emph{invariant}. Intuitively, a safety invariant maps every boolean
+valuation into a set of data valuations, that contains the initial
+configuration $\iota = \Post{\A}(\iota,\varepsilon)$, whose data
+variables are unconstrained, over-approximates the set of reachable
+valuations (\ref{it1:inv}) and excludes the valuations satisfying the
+acceptance condition (\ref{it2:inv}).  A formula $\phi(Q,\vec{x})$
+\emph{defines} $\Inv$ iff for all $\beta: Q \rightarrow \booli$ and
+$\nu : \vec{x} \rightarrow \Data^\I$, we have $\I,\beta\cup\nu \models
+\phi$ iff $\nu \in \Inv(\beta)$.
 
@@
   there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
-  exists $i \in [0,k]$ such that $\AbsPost{\A}(\iota,u) \eqiff
+  exists $i \in [0,k]$ such that $\AbsPost{\A}(\iota,u) \iffeq
   \AbsPost{\A}(\iota,\mu(i))$, which concludes the proof. \qed}
@@
 
+\begin{proposition}\label{prop:asterisk}
+  Given a formula $\phi \in \Form^+(Q,\vec{x})$ and $a \in \Sigma$, we
+  have \(\Delta(\phi,a) \iffeq \exists Q' ~.~ \phi[Q'/Q] \wedge
+  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))\).
+\end{proposition}
+\proof{``$\Rightarrow$'' If $\I,\beta\cup\overline\nu\cup\nu \models
+  \Delta^n(\phi,a)$, for some valuations $\beta : Q \rightarrow
+  \booli$ and $\overline\nu : \overline{\vec{x}} \rightarrow
+  \Data^\I$, $\nu : \vec{x} \rightarrow \Data^\I$, then we build a
+  valuation $\beta' : Q' \rightarrow \booli$ such that
+  $\I,\beta'\cup\beta\cup\overline\nu\cup\nu \models \phi[Q'/Q] \wedge
+  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))$. For each
+  occurrence of a formula $\Delta(q,a)$ in $\Delta(\phi,a)$ we set
+  $\beta'(q') = \true$ if $\I,\beta\cup\overline\nu\cup\nu \models
+  \Delta(q,a)$ and $\beta'(q') = \false$, otherwise. Since there are
+  no negated occurrences of such subformulae, the definition of
+  $\beta'$ is consistent, and the check
+  $\I,\beta'\cup\beta\cup\overline\nu\cup\nu \models \phi[Q'/Q] \wedge
+  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))$ is
+  immediate. $''\Leftarrow''$ This direction is an easy check. \qed}
+
 We recall that a run of $\A$ over $u$ is a sequence $\phi_0(Q),
@@
 
-  (\ref{it1:progress}) Assume that $\pi(u)$ is satisfiable and let
-  $\beta_0,\ldots,\beta_n$ and $\nu_0,\ldots,\nu_n$ be boolean and
-  data valuations for $Q_0,\ldots,Q_n$ and
-  $\vec{x}_0,\ldots,\vec{x}_n$, respectively, such that
-  $\I,\bigcup_{i=0}^n \beta_i \cup \bigcup_{i=0}^n \nu_i \models
-  \pi(u)$. We build a model for $\Accept{\A}(u) \equiv
-  \Post{\A}(\iota,u) \wedge \theta_{n+1}[Q/Q_n]$. Let
-  $\phi_0,\phi_1,\ldots,\phi_n$ be the run of $\A$ over $u$. We show
-  that $\phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)$ has a model, by
-  induction on $n\geq0$. For the base case $n=0$, $\phi_0 \equiv \iota
-  \equiv \theta_0[Q/Q_0]$ has the model $\beta(q) = \beta_0(q_0)$, for
-  all $q \in Q$. For the induction step $n > 0$, we assume that
-  $\phi_{n-1}(Q,\vec{x}_0\cup\ldots\vec{x}_{n-1})$ has the model
-  $\beta \cup \nu_0 \cup \ldots \nu_{n-1}$, where $\beta(q) =
-  \beta_{n-1}(q_{n-1})$, for all $q \in Q$. By the definition of
-  $\pi(u)$: \[\theta_n \equiv \bigwedge_{q_{n-1} \in
-    \fv{\Bool}{\theta_{n-1}} \cap Q_{n-1}} \left(q_{n-1} \rightarrow
-  \Delta(q,a_n)[Q_n/Q,\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]\right)\]
-
-  But since $\I,\beta_{n-1} \cup \beta_n, \nu_{n-1} \cup \nu_n \models
-  \theta_n(Q_{n-1}\cup Q_n, \vec{x}_{n-1} \cup \vec{x}_n)$, we have
-  that $\I,\beta_n,\nu_{n-1} \cup \nu_n \models
-  \Delta(q,a_n)[Q_n/Q,\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
-  for every $q \in Q$ such that $\beta_{n-1}(q_{n-1}) = \true$. Then
-  we have $\I,\beta\cup\bigcup_{i=0}^n\nu_i \models \phi_n$
-  where $\beta(q) = \beta_n(q_n)$, for all $q \in Q$ and $\phi_n$ is
-  obtained from $\phi_{n-1}$ by replacing each boolean variable $q$ by
-  $\Delta(a,q)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$. Observe
-  that, because each boolean variable occurs positively in
-  $\phi_{n-1}$, the truth value of
-  $\Delta(a,q)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$
-  under the valuation $\beta\cup\nu_{n-1}\cup\nu_n$ is not important for
-  any $q \in Q$ such that $\beta(q) = \false$. We have obtained a
-  model for $\phi_n$ and thus for $\Post{\A}(u)$. Since moreover,
-  $\theta_{n+1}[Q/Q_n] = \true$ in this model, we obtain a model for
-  $\Accept{\A}(u)$. Thus $\pi(u)$ is satisfiable implies that
-  $\Accept{\A}(u)$ is satisfiable, which proves the statement of point
-  (\ref{it1:progress}).
+  (\ref{it1:progress}) We apply Proposition \ref{prop:asterisk}
+  recursively and get: \[\Post{\A}(\iota,u)[Q_n/Q,\vec{x}_n/\vec{x}]
+  \iff \exists Q_0 \ldots \exists Q_{n-1} \exists \vec{x}_0 \ldots
+  \exists \vec{x}_{n-1} ~.~ \bigwedge_{i=0}^n \theta_i\] Assuming that
+  $\pi(u) \equiv \bigwedge_{i=0}^{n+1} \theta_i$ is satisfiable, we
+  obtain a model for $\Accept{\A}(u) \equiv \Post{\A}(\iota,u) \wedge
+  \theta_{n+1}[Q/Q_n]$.
 
   (\ref{it2:progress}) If $\tuple{\top,I_1,\dots,I_{n+1},\bot}$ is an
-    interpolant for $\pi(u)$, the following entailments hold: \begin{compactitem}
+  interpolant for $\pi(u)$, the following entailments
+  hold: \begin{compactitem}
   \item $\theta_0 \models I_1[Q_0/Q]$,
@@
   \end{compactitem}
-    We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n+1}$
-    by induction on $n \geq 0$. This is sufficient to conclude because
+  We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n+1}$ by
+  induction on $n \geq 0$. This is sufficient to conclude because
   $\AbsAccept{\A}(a_1\ldots a_n) \equiv \AbsPost{\A}(\iota,a_1\ldots
   a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n+1} \wedge
-    \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we
-    have $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv
-    \theta_0[Q/Q_0] \models I_1$. For the induction step $n>0$, we
-    compute:
+  \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we have
+  $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv \theta_0[Q/Q_0]
+  \models I_1$. For the induction step $n>0$, we compute:
   \[\begin{array}{l}
   \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] \equiv \text{ (by definition of $\AbsPost{\A}$)} \\
-    \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] \models \text{ ($\Asterisk$)} \\
+  \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] \models \text{ (by Prop. \ref{prop:asterisk})} \\
   \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n \models \text{ (ind. hyp.) } \\
-    \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q] 
-    \end{array}\]
-   Let us prove the final point ($\Asterisk$). If
-   $\I,\beta_n\cup\nu_{n-1}\cup\nu_n \models
-   \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_{n}/Q]$,
-   for some valuations $\beta_{n} : Q_{n-1} \rightarrow \booli$ and
-   $\nu_i : \vec{x}_i \rightarrow \Data^\I$, for $i=n-1,n$, then we
-   build a valuation $\beta_{n-1}$ such that
-   $\I,\beta_{n-1}\cup\beta_n \cup \nu_{n-1}\cup\nu_n \models
-   \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$,
-   as follows. For each occurrence of a subformula $\Delta(q,a_n)$ in
-   $\AbsPost{\A}(\iota,a_1\ldots a_{n-1})$ we set
-   $\beta_{n-1}(q_{n-1}) = \true$ if $\I,\beta_n\cup\nu_{n-1}\cup\nu_n
-   \models \Delta(q,a_n)[Q_n/Q]$ and $\beta_{n-1}(q_{n-1}) = \false$,
-   otherwise. Since there are no negated occurrences of such
-   subformulae, the definition of $\beta_{n-1}$ is consistent, and the
-   check $\I,\beta_{n-1}\cup\beta_n \cup \nu_{n-1}\cup\nu_n \models
-   \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$
-   is straightforward. \qed}
-
+  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q] \hfill\text{\qed}
+  \end{array}\]}
 We have now all the ingredients to describe the first emptiness
 checking semi-algorithm for alternating data
-automata. Algorithm\footnote{Though termination is not guaranteed,
-  we call it algorithm for conciseness.}  \ref{alg:predabs} builds
-an \emph{abstract reachability tree} (ART) whose edges are labeled
-with input events and each node $n$ is labeled with the formula
+automata. Algorithm\footnote{Though termination is not guaranteed, we
+  call it algorithm for conciseness.}  \ref{alg:predabs} builds an
+\emph{abstract reachability tree} (ART) whose edges are labeled with
+input events and each node $n$ is labeled with the formula
 $\AbsPost{\A}(\iota,u)$, where $u \in \Sigma^*$ is the event trace
@@
   in the tree $(N,E)$, then we have:
+\end{compactitem}
   \[\begin{array}{rcl}
   T(e) & = & \Theta(n,a,k) \\
-  R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)} \text{ where} \\
+  R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)} \text{, where} \\
   \Theta(n,a,k) & \equiv & \bigwedge_{q \in R(n)} (q_k \rightarrow
@@
   \end{array}\]
-\end{compactitem}
 For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is
@@
 T(n_{k-1},a_k,n_k) \wedge \bigwedge_{q \in R(n_k) \setminus F}(q_k
-\rightarrow \bot)\] defines the interpolation problem $\pi(u)$.
+\rightarrow \bot)\] is equivalent to the interpolation problem
+$\pi(u)$ and can be directly computed by a traversal of the path from
+$\rootNode$ to $n$. In addition, we require that an interpolant
+$\tuple{\top,I_0,\ldots,I_k,\bot}$ for the insatisfiable interpolation
+problem $\mu(n)$ does not have negative occurrences of states,
+i.e.\ $I_i \in \Form^+(Q,\vec{x})$, for all $i \in [0,k]$. Such an
+interpolant can always be built, as showed below:
+
+\begin{proposition}\label{prop:positive-interpol}
+  For any $u = a_1 \ldots a_n \in \Sigma^*$, if $\pi(u)$ is
+  unsatisfiable then one can build an interpolant
+  $\tuple{\top,I_0,\ldots,I_n,\bot}$ for $\pi(u)$, such that $I_i
+  \in \Form^+(Q,\vec{x})$, for all $i \in [0,n]$.
+\end{proposition}
+\proof{ Let $\tuple{\top,I_0,\ldots,I_{n},\bot}$ be an interpolant,
+  and $i \in [0,n]$ be the first index for which $I_i \not\in
+  \Form^+(Q,\vec{x})$. By Proposition \ref{prop:asterisk}, we have
+  that $\Delta^{i-1}(I_{i-1},a_i) \iffeq \exists Q_{i-1} ~.~
+  I_{i-1}[Q_{i-1}/Q] \wedge \theta_i$. Since $I_{i-1} \in
+  \Form^+(Q,\vec{x})$, by the choice of $i$, we have that
+  $\Delta^{i-1}(I_{i-1},a_i) \in \Form^+(Q,\vec{x})$ as well. We
+  replace $I_i$ with $\Delta^{i-1}(I_{i-1},a_i)$. The result is still
+  an interpolant, because: \begin{compactitem}
+  \item $I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \models
+    \Delta^{i-1}(I_{i-1},a_i)$, by Proposition \ref{prop:asterisk}, and
+    %
+  \item $\Delta^{i-1}(I_{i-1},a_i) \wedge \theta_{i+1} \models \exists
+    Q_{i-1} ~.~ I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \wedge \theta_{i+1}
+    \models I_i[Q_i/Q] \wedge \theta_{i+1} \models I_{i+1}[Q_{i+1}/Q]$.
+  \end{compactitem}
+  We proceed with these replacements until there are no more formulae
+  $I_i \not\in \Form^+(Q,\vec{x})$ left. \qed}
 
@@
 
-  \State $\Call{Close}{n}$
-  \label{ln:close-worklist}
-
-  \If{$n$ is not covered}
-
   \State let $(\rootNode,a_1,n_1),(n_1,a_2,n_2), \ldots,
@@
 
-  \If{$R(n) \cap F \neq \emptyset$} 
-  \label{ln:impact-cex}
-  \Comment{counterexample candidate}
-
   \If{$\mu(n)$ is satisfiable} 
@@
 
-  \State let $\tuple{\top,I_1,\ldots,I_k \equiv\bot}$ be an interpolant for $\mu(n)$
+  \State let $\tuple{\top,I_0,\ldots,I_k,\bot}$ be an
+  interpolant for $\mu(n)$ with $I_0,\ldots,I_k \in
+  \Form^+(Q,\vec{x})$
   \label{ln:refine-begin}
@@
 
-  \For{$i=1,\ldots,k$}
+  \For{$i=0,\ldots,k$}
 
@@
 
-  \EndIf
-
-  \EndIf
+  \EndIf % $\neg b$
+
+  \EndIf % $\Lambda(n_i) \not\models I_i$
 
@@
 
-  \EndIf
-
-  \EndIf
+  \EndIf % $\mu(n)$ is satisfiable
+
+  \If{$n$ is not covered}
 
@@
 
-  \EndIf
+  \EndIf % $n$ is not covered
 
@@
 
-\For{$y \in N$ such that $x \prec^* y$ and $x$ is not an ancestor of $y$}
+\For{$y \in N$ such that $y \prec^* x$}
 
@@
   description of \impact~\cite{mcmillan06}.}: \emph{close},
-\emph{refine} and \emph{expand}. When a node $n$ is removed from the
-worklist, it is first closed (line \ref{ln:close-worklist}) by finding
-a lexicographically bigger node $m$ such that $\Lambda(n) \models
-\Lambda(m)$, if one exists, and adding the pair $(n,m)$ to $\lhd$. As
-in Algorithm \ref{alg:predabs}, the intuition is that for any feasible
-counterexample that can be discovered from $n$ there exists a feasible
-counterexample stemming in $m$, thus we do not need to process $n$
-further.
-
-If $n$ is not covered, we check if it is the endpoint of a
-counterexample candidate, which can only be the case when the
-replacement set $R(n)$ contains at least one final state (line
-\ref{ln:cex}). Observe that, if $u$ is the label of the path from
-$\rootNode$ to $n$, then $\Accept{\A}(u)$ is satisfiable only if it
-contains a (positive occurrence of) a final state. The counterexample
-is moreover feasible iff $\mu(n)$ is satisfiable, in which case a
-model of $\mu(n)$ is obtained and a word $w \in L(\A)$ returned.
-
-Otherwise, the counterexample is spurious and the procedure enters the
-refinement phase (lines
+\emph{refine} and \emph{expand}. 
+%
+%% When a node $n$ is removed from the worklist, it is first closed (line
+%% \ref{ln:close-worklist}) by finding a lexicographically bigger node
+%% $m$ such that $\Lambda(n) \models \Lambda(m)$, if one exists, and
+%% adding the pair $(n,m)$ to $\lhd$. As in Algorithm \ref{alg:predabs},
+%% the intuition is that for any feasible counterexample that can be
+%% discovered from $n$ there exists a feasible counterexample stemming in
+%% $m$, thus we do not need to process $n$ further.
+%
+%% If $n$ is not covered, we check if it is the endpoint of a
+%% counterexample candidate, which can only be the case when the
+%% replacement set $R(n)$ contains at least one final state (line
+%% \ref{ln:cex}). Observe that, if $u$ is the label of the path from
+%% $\rootNode$ to $n$, then $\Accept{\A}(u)$ is satisfiable only if it
+%% contains a (positive occurrence of) a final state. The counterexample
+%% is moreover feasible iff $\mu(n)$ is satisfiable, in which case a
+%% model of $\mu(n)$ is obtained and a word $w \in L(\A)$ returned.
+%
+Let $u$ be the label of the path from the root node $\rootNode$ to
+$n$. If $\mu(n)$ is satisfiable, then $u$ is a feasible
+counterexample, in which case a model of $\mu(n)$ is obtained and a
+word $w \in L(\A)$ returned. Otherwise, $u$ is a spurious
+counterexample and the procedure enters the refinement phase (lines
 \ref{ln:refine-begin}-\ref{ln:refine-end}). The interpolant for
@@
 by conjoining the formulae of the interpolant to the existing
-labels. Then $\Lambda(n)$ becomes equivalent to $\bot$.
+labels. For technical reasons, related to proving the correctness of
+Algorithm \ref{alg:impact}, we assume that all formulae from the
+interpolant for $\mu(n)$ have only positive occurrences of the
+states\footnote{According to Proposition \ref{prop:positive-interpol},
+  building such an interpolant is always possible} in $Q$.
 
@@
 
-Finally, we expand $n$ (lines
+Finally, if $n$ is still uncovered (i.e.\ it has not been previously
+covered during the refinement phase) we expand $n$ (lines
 \ref{ln:expand-begin}-\ref{ln:expand-end}) by creating a new node for
-each successor $s$ via the event $a \in \Sigma$ and inserting it
-into the worklist. The label of the new node is $\top$, its
-replacement set $R(s)$ is computed based on $R(n)$ and the label of
-the edge $(n,a,s)$ is $\Theta(n,a,k)$, where $k$ is the depth of $n$
-in $\Art$.
-
-The following lemma states that Algorithm \ref{alg:impact} builds an
-over-approximation of the concrete runs of $\A$.
-
-\begin{lemma}\label{lemma:impact-interpol}
+each successor $s$ via the event $a \in \Sigma$ and inserting it into
+the worklist. The label of the new node is $\top$, its replacement set
+$R(s)$ is computed based on $R(n)$ and the label of the edge $(n,a,s)$
+is $\Theta(n,a,k)$, where $k$ is the depth of $n$ in $\Art$.
+
+The following lemma states that the labeling of the nodes in an ART
+built by Algorithm \ref{alg:impact} is consistent with the semantics
+of the transition function of $\A$.
+
+\begin{lemma}\label{lemma:well-labeled}
   Given an ART $\Art = \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$ built
-  by Algorithm \ref{alg:impact} and a node $n \in N$, for all $(n,a,m)
-  \in E$, we have $\Post{\A}(\Lambda(n),a) \models \Lambda(m)$.
+  by Algorithm \ref{alg:impact}, $\Post{\A}(\Lambda(n),a) \models
+  \Lambda(m)$, for all $(n,a,m) \in E$.
 \end{lemma}
@@
   in $\Art$ that has never been refined, then $\Lambda(m)=\top$ and
-  the entailment holds trivially. Otherwise, let $(n_0,a_1,n_1),
-  \ldots, (n_{k-1},a_k,n_k)$ be a path, where $n_0 = \rootNode$ and
-  $(n,a,m) = (n_{i-1},a_i,n_i)$, for some $i \in [1,k]$ and, moreover,
-  $a_1\ldots a_k$ was found, at some point, to be a spurious
-  counterexample. Let $\tuple{\top,I_1,\ldots,I_k\equiv\bot}$ be an
+  the entailment holds trivially. Otherwise, let $\Omega$ be the set
+  of paths $\omega = (n_0,a_1,n_1), \ldots, (n_{k-1},a_k,n_k)$, where
+  $n_0 = \rootNode$ and $(n,a,m) = (n_{i-1},a_i,n_i)$, for some $i \in
+  [1,k]$ and, moreover, $a_1\ldots a_k$ was found, at some point, to
+  be a spurious counterexample. Let
+  $\langle\top,I^\omega_1,\ldots,I^\omega_k\equiv\bot\rangle$ be an
   interpolant for $\mu(n_k) \equiv \Lambda(\rootNode) \wedge
-  \bigwedge_{i=1}^k \Theta(n_{i-1},a,i-1) \wedge \bigwedge_{q \in R(n_k)}
-  (q_k \rightarrow \bot)$. 
-\qed}
+  \bigwedge_{i=1}^k \Theta(n_{i-1},a,i-1) \wedge \bigwedge_{q \in
+    R(n_k)} (q_k \rightarrow \bot)$, such that $I^\omega_i \in
+  \Form^+(Q,\vec{x})$, for all $i \in [1,k]$. According to Proposition
+  \ref{prop:positive-interpol}, it is always possible to build such an
+  interpolant, when $\mu(n_k)$ is unsatisfiable. By Proposition
+  \ref{prop:asterisk}, we obtain $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q]
+  \iffeq \exists Q_{i-1} ~.~
+  I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge \theta_i$
+  and, since $I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge
+  \theta_i \models I^\omega_i[Q_i/Q,\vec{x}_i/\vec{x}]$, we obtain
+  that $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q] \models
+  I^\omega_i[Q_i/Q,\vec{x}_i/\vec{x}]$. Since $\Lambda(n_{i-1}) =
+  \bigwedge_{\omega \in \Omega} I^\omega_{i-1}$ and $\Lambda(n_{i}) =
+  \bigwedge_{\omega \in \Omega} I^\omega_{i}$, we obtain
+  $\Post{\A}(\Lambda(n_{i-1}),a_i) \models \Lambda(n_i)$. \qed}
 
@@
   each call to $\Call{Close}{}$ terminates, because $N$ is finite at
-  each moment during the execution.
-
-  For the second point, we prove the following invariant: each time
-  Algorithm \ref{alg:impact} reaches the line \ref{ln:impact-while},
-  the set of nodes $W$ in $\worklist$ contains exactly those nodes in
-  the ART $\tuple{N \cup W,E,\Lambda,R,T,\lhd}$ which are
-  both\begin{inparaenum}[(i)]
-  \item uncovered, and such that
-  \item for some $a \in \Sigma$, the edge $(n,a,s)$ is missing from $E$:
-  \end{inparaenum}
+  each moment during the execution. For the second point, we prove
+  that each time Algorithm \ref{alg:impact} reaches the line
+  \ref{ln:impact-while}, we have:
   \begin{equation}\label{inv:impact}
@@
   \end{equation}
-  Initially, $W = \set{\rootNode}$ and $N=E=\lhd=\emptyset$, thus the
-  invariant holds trivially. Suppose that (\ref{inv:impact}) holds at
-  when reaching line \ref{ln:impact-while} and some node $n$ was
-  removed from $W$ and inserted into $N$. We distinguish two cases,
-  either: \begin{itemize}
-  \item $n$ is covered at line \ref{ln:close-worklist}, in which case
-    $W$ remains unchanged, or
-  %
-  \item $n$ is not covered, in which case for each $a \in \Sigma$ a
-    new node $s$ is inserted into $W$ (line \ref{ln:expand-end}) and
-    the edge $(n,a,s)$ is inserted into $E$ (line
-    \ref{ln:edge-insert}). None of these new nodes are covered and
-    none of their successors inserted into $E$.
+  Initially, $W = \set{\rootNode}$ and $N=E=\lhd=\emptyset$, thus
+  (\ref{inv:impact}) holds trivially. Suppose that (\ref{inv:impact})
+  holds at when reaching line \ref{ln:impact-while} and some node $n$
+  was removed from $W$ and inserted into $N$. We distinguish two
+  cases, either: \begin{itemize}
+  \item $n$ is covered, in which case $W$ becomes $W \setminus
+    \set{n}$, or
+  %
+  \item $n$ is not covered, in which case $W$ becomes $(W \setminus
+    \set{n}) \cup S$, where $S = \set{s \not\in N \mid (n,a,s) \in E,
+      a \in \Sigma}$ is the set of fresh successors of $n$. But then
+    no node $s \in S$ is covered and has successors in $E$.
   \end{itemize}
@@
   a safety invariant and conclude that $L(\A) = \emptyset$, by Lemma
-  \ref{lemma:safety-invariant}. To this end, let $u \in \Sigma^*$ be
-  an arbitrary sequence and let $v_1$ be the largest prefix of $u$
-  that labels a path from $\rootNode$ to some node $n_1 \in N$. 
+  \ref{lemma:safety-invariant}. To this end, let $u = a_1 \ldots a_k
+  \in \Sigma^*$ be an arbitrary sequence and let $v_1$ be the largest
+  prefix of $u$ that labels a path from $\rootNode$ to some node $n_1
+  \in N$. If $v_1 = u$ we are done. Otherwise, by the choice of $v_1$,
+  it must be the case that a successor of $n_1$ is missing from
+  $(N,E)$, thus $n_1$ must be covered. Let $n'_1$ be the closest
+  ancestor of $n_1$ such that $(n'_1,n''_1) \in \lhd$, for some $n''_1
+  \in N$, and let $v'_1$ be the prefix of $v_1$ leading to $n'_1$. By
+  the construction of $\lhd$ (line \ref{ln:close-uncover} in function
+  $\Call{Close}{}$), we have $\Lambda(n'_1) \models
+  \Lambda(n''_1)$. Applying Lemma \ref{lemma:well-labeled} inductively
+  on $v'_1$, we obtain that $\Post{\A}(\iota,v'_1) \models
+  \Lambda(n'_1)$, thus $\Post{\A}(\iota,v'_1) \models \Lambda(n''_1)$.
+  Continuing inductively from $n''_1$, we exhibit a sequence of
+  strings $v'_1,\ldots,v'_\ell \in \Sigma^*$ and nodes $\rootNode =
+  m_0,m_1,\ldots,m_\ell$ such that, for all $i \in
+  [1,\ell]$: \begin{compactitem}
+  \item $v'_i$ labels the path between $m_{i-1}$ and $m_i$ in $(N,E)$,    
+    %
+  \item $\Post{\A}(\iota,v'_1\ldots v'_i) \models \Lambda(m_i)$.
+  \end{compactitem}
+  Moreover, we have $u = v'_1 \ldots v'_\ell$, thus $\Post{\A}(\iota,u)
+  \models \Lambda(m_\ell)$ and we are done showing that $\bigvee_{n\in
+    N} \Lambda(n)$ is an invariant. To prove that it is, moreover, a
+  safety invariant, 
 \qed}

### 2017-05-12 draft.tex
ok

@@
 sorted \emph{function symbols}. To simplify the presentation, we
-assume w.l.o.g. that $\ssorts{\sig}$ consists of two distinct sorts
-$\Data$ and $\Bool$ and each function symbol $f \in \sfuns{\sig}$ has
-$\#(f) \geq 0$ arguments of sort $\Data$ and return value $\sigma(f)
-\in \ssorts{\sig}$. If $\#(f)=0$ then $f$ is a \emph{constant}. We
-consider the constants $\top$ and $\bot$ of sort $\Bool$.
+assume w.l.o.g. that $\ssorts{\sig} = \set{\Data,\Bool}$\footnote{The
+  generalization to more than two sorts is without difficulty, but
+  would unnecessarily clutter the technical presentation.} and each
+function symbol $f \in \sfuns{\sig}$ has $\#(f) \geq 0$ arguments of
+sort $\Data$ and return value $\sigma(f) \in \ssorts{\sig}$. If
+$\#(f)=0$ then $f$ is a \emph{constant}. We consider the constants
+$\top$ and $\bot$ of sort $\Bool$.
 
@@
 
-We denote by $\fv{S}{\phi}$ the set of free (i.e.\ not under the scope
-of an existential quantifier) variables of sort $S$ in $\phi$ and
-write $\fv{}{\phi}$ for $\bigcup_{S \in \set{\Data,\Bool}}
-\fv{S}{\phi}$. For a variable $x \in \fv{}{\phi}$ and a term $t$ such
-that $\sigma(t) = \sigma(x)$, let $\phi[t/x]$ be the result of
+We denote by $\fv{\sigma}{\phi}$ the set of free (not under the scope
+of an existential quantifier) variables of sort $\sigma$ in $\phi$ and
+write $\fv{}{\phi}$ for $\bigcup_{\sigma \in \ssorts{\sig}}
+\fv{\sigma}{\phi}$. For a variable $x \in \fv{}{\phi}$ and a term $t$
+such that $\sigma(t) = \sigma(x)$, let $\phi[t/x]$ be the result of
 replacing each occurrence of $x$ by $t$. For indexed sets
@@
 simultaneously replacing $x_i$ with $t_i$ in $\phi$, for all
-$i=1,\ldots,n$.
-%
-%% A formula is in positive normal form (PNF) if negation occurs only
-%% within literals and in disjunctive normal form (DNF) if it is a
-%% disjunction of conjunctions of literals $\bigvee_{i=1}^N
-%% \bigwedge_{j=1}^{M_i} \ell_{ij}$.
-%
-The size of a formula $\phi$, denoted as $\len{\phi}$, is the number
-of symbols occuring in it.
+$i\in[1,n]$. The size $\len{\phi}$ is the number of symbols occuring
+in $\phi$.
 
@@
 \emph{theory} $\theory(\sig,\I)$ is the set of valid formulae written
-in the signature $\sig$, with the interpretation $\I$.
+in the signature $\sig$, with the interpretation $\I$. A decision
+procedure for $\theory(\sig,\I)$ is an algorithm that takes a formula
+$\phi$ in the signature $\sig$ and returns yes iff $\phi \in
+\theory(\sig,\I)$.
 
@@
 a set of time-stamped variables. For an input event $a \in \Sigma$ and
-a formula $\phi$, we write $\Delta(\phi)$ and, respectively,
-$\Delta^k(\phi,a)$, for the formula obtained from $\phi$ by
+a formula $\phi$, we write $\Delta(\phi,a)$ (respectively,
+$\Delta^k(\phi,a)$) for the formula obtained from $\phi$ by
 simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
-formula $\Delta(q,a)$, respectively
+formula $\Delta(q,a)$ (respectively
 $\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
-all $k\geq0$. Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
+all $k\geq0$). Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
 (a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
 formulae $\phi_0,\phi_1,\ldots,\phi_n$, where $\phi_0 \equiv \iota$
-and, for all $k > 0$, we have $\phi_k \equiv
+and, for all $k\in[1,n]$, we have $\phi_k \equiv
 \Delta^k(\phi_{k-1},a_k)$. Next, we slightly abuse notation and write
@@
 \end{inparaenum}
-The language of $\A$ is the set $L(\A)$ of words accepted by $\A$.
-
+The language of $\A$ is the set $L(\A)$ of words from
+$\Sigma[\vec{x}]^*$ accepted by $\A$.
+%
 In this paper we tackle the following problems: \begin{compactenum}
@@
   $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
-  \subseteq Q$ and thus $\overline{\iota_1} = \iota_1$. For the case
+  \subseteq Q$ and thus $\overline{\iota_1} \equiv \iota_1$. For the case
   $n>0$, we compute:
   \[\begin{array}{lcl}
-  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \iffeq & \text{ by induction hypothesis} \\
-  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \iffeq & \text{ by the definition of $\overline{\phi}$} \\
+  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \iffeq & \text{ (by ind. hyp.)} \\
+  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \iffeq & \text{ (by the def. of $\overline{\phi}$)} \\
   \overline{\Delta(\iota_1,a_1\ldots,a_n)}
@@
   \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
-  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus $w \in L(\A_1) \iff
-  w \not\in L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) =
-  \Sigma[\vec{x}]^* \setminus L(\A_1)$. \qed}
+  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus for all $w \in
+  \Sigma[\vec{x}]^*$, we have $w \in L(\A_1) \iff w \not\in
+  L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) = \Sigma[\vec{x}]^*
+  \setminus L(\A_1)$. \qed}
 
@@
 because, for any $u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models
-\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable, then
-$\Post{\A}(\iota,uw) \models \Post{\A}(\iota,vw)$, thus
-$\Accept{\A}(vw)$ is satisfiable as well, and there is no need for
-$u$, since the non-emptiness of $\A$ can be proved using $v$
-alone. However, the antichain-based semi-algorithm diverges in a large
-number of cases, since infinite antichains exist in many
-interpretations, e.g.\ $q \wedge x \teq 0,~ q \wedge x \teq 1, \ldots$
-for $\Data^\I = \nat$.
+\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable for some $w
+\in \Sigma^*$, then $\Post{\A}(\iota,uw) \models \Post{\A}(\iota,vw)$,
+thus $\Accept{\A}(vw)$ is satisfiable as well, and there is no need
+for $u$, since the non-emptiness of $\A$ can be proved using $v$
+alone. However, even with this optimization, the enumeration of
+sequences from $\Sigma^*$ diverges in many real cases, because
+infinite antichains exist in many interpretations, e.g.\ $q \wedge x
+\teq 0,~ q \wedge x \teq 1, \ldots$ for $\Data^\I = \nat$.
 
@@
 variables are unconstrained, over-approximates the set of reachable
-valuations (\ref{it1:inv}) and excludes the valuations satisfying the
-acceptance condition (\ref{it2:inv}).  A formula $\phi(Q,\vec{x})$
+valuations (point \ref{it1:inv}) and excludes the valuations satisfying the
+acceptance condition (point \ref{it2:inv}).  A formula $\phi(Q,\vec{x})$
 \emph{defines} $\Inv$ iff for all $\beta: Q \rightarrow \booli$ and
@@
 
-The solution we adopt is abstraction to ensure the termination of
-invariant computations. However, it is worth pointing out from the
-start that abstraction alone will only allow us to build invariants
-that are not necessarily safety invariants. To meet the latter
-condition, we resort to counterexample guided abstraction refinement
-(CEGAR).
-
-Formally, let $\preds$ be a finite set of formulae with variables from
-$Q \cup \vec{x}$, such that $\bot \in \preds$ and the variables in $Q$
-can now occur negated within the formulae $\phi \in \preds$. We fix
-$\preds$ from now on and refer to these formulae as
-\emph{predicates}. Given a formula $\phi$, we denote by $\abs{\phi}
-\equiv \bigwedge \set{\pi \in \preds \mid \phi \models \pi}$ the
-abstraction of $\phi$ w.r.t. the predicates in $\preds$. The abstract
-version of the post-image and acceptance condition are defined as
-follows, for any $u \in \Sigma^*$:
+The solution we adopt in the sequel is abstraction to ensure the
+termination of invariant computations. However, it is worth pointing
+out from the start that abstraction alone will only allow us to build
+invariants that are not necessarily safety invariants. To meet the
+latter condition, we resort to counterexample guided abstraction
+refinement (CEGAR).
+
+Formally, we fix a set of formulae $\preds \subseteq
+\Form(Q,\vec{x})$, such that $\bot \in \preds$ and refer to these
+formulae as \emph{predicates}. Given a formula $\phi$, we denote by
+$\abs{\phi} \equiv \bigwedge \set{\pi \in \preds \mid \phi \models
+  \pi}$ the abstraction of $\phi$ w.r.t. the predicates in
+$\preds$. The abstract versions of the post-image and acceptance
+condition are defined as follows, for any $u \in \Sigma^*$:
 \[\begin{array}{l}
@@
   For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
-  $k>0$ such that $\bigvee_{i=1}^k \AbsPost{\A}(\iota,\mu(i))$ defines
+  $k>0$ such that $\bigvee_{i=0}^k \AbsPost{\A}(\iota,\mu(i))$ defines
   an invariant for $\A$. 
@@
 
-We are left with fulfilling the second point from the definition of a
-safety invariant. To this end, suppose that, for a given set $\preds$
-of predicates, the invariant $\abs{\Inv}$, defined by the previous
-lemma, meets point (\ref{it1:inv}) but not point (\ref{it2:inv}) from
-the definition of invariants, where $\Post{\A}$ and $\Accept{\A}$
-replace $\AbsPost{\A}$ and $\AbsAccept{\A}$, respectively. In
-other words, there exists a finite sequence $u \in \Sigma^*$ such that
-$\nu \in \abs{\Inv}(\beta)$ and $\I,\beta\cup\nu \models
-\AbsAccept{\A}(u)$, for some boolean $\beta : Q \rightarrow \booli$
-and data $\nu : \vec{x} \rightarrow \Data^\I$ valuations. Such
-a $u\in\Sigma^*$ is called a \emph{counterexample}.
+We are left with fulfilling point (\ref{it2:inv}) from the definition
+of a safety invariant. To this end, suppose that, for a given set
+$\preds$ of predicates, the invariant $\abs{\Inv}$, defined by the
+previous lemma, meets point (\ref{it1:inv}) but not point
+(\ref{it2:inv}), where $\Post{\A}$ and $\Accept{\A}$ replace
+$\AbsPost{\A}$ and $\AbsAccept{\A}$, respectively. In other words,
+there exists a finite sequence $u \in \Sigma^*$ such that $\nu \in
+\abs{\Inv}(\beta)$ and $\I,\beta\cup\nu \models \AbsAccept{\A}(u)$,
+for some boolean $\beta : Q \rightarrow \booli$ and data $\nu :
+\vec{x} \rightarrow \Data^\I$ valuations. Such a $u\in\Sigma^*$ is
+called a \emph{counterexample}.
 
@@
 possibilities. Either\begin{inparaenum}[(i)]
-\item $\Accept{\A}(u)$ is satisfiable, in which case $L(\A) \neq \emptyset$, or
-\item $\Accept{\A}(u)$ is unsatisfiable, in which case $u$ is called
+\item $\Accept{\A}(u)$ is satisfiable, in which case $u$ is
+  \emph{feasible} and $L(\A) \neq \emptyset$, or
+%
+\item $\Accept{\A}(u)$ is unsatisfiable, in which case $u$ is
   \emph{spurious}.
@@
 
-In general, given an unsatisfiable conjunction $\pi \equiv
+In general, given an unsatisfiable conjunction $\Phi \equiv
 \phi_1(X_0,X_1) \wedge \phi_2(X_1,X_2) \wedge \ldots \wedge
@@
   \in X}$ of any sort, a solution of the \emph{interpolation problem}
-$\pi$, simply called an \emph{interpolant}, is a tuple $\tuple{I_0(X),
+$\Phi$, simply called an \emph{interpolant}, is a tuple $\tuple{I_0(X),
   I_1(X), \ldots, I_n(X)}$ such that:\begin{inparaenum}[(i)]
@@
 \end{inparaenum}
+In the following, we shall assume the existence of an interpolating
+decision procedure for $\theory(\sig,\I)$.
 
@@
 \begin{eqnarray}\label{eq:interpolation-problem}
-\pi(u) & \equiv & \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0
+\Theta(u) & \equiv & \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0
 \cup \vec{x}_1) \wedge \ldots \\ 
@@
   F} (q_n \rightarrow \bot)$. The intuition is that $\theta_0, \ldots,
-\theta_n$ simulate the run of $\A$ over $u$ and $\theta_{n+1}$ is the
+\theta_n$ encode the run of $\A$ over $u$ and $\theta_{n+1}$ is the
 acceptance condition, which forces the last remaining non-final states
@@
 \proof{``$\Rightarrow$'' If $\I,\beta\cup\overline\nu\cup\nu \models
-  \Delta^n(\phi,a)$, for some valuations $\beta : Q \rightarrow
-  \booli$ and $\overline\nu : \overline{\vec{x}} \rightarrow
-  \Data^\I$, $\nu : \vec{x} \rightarrow \Data^\I$, then we build a
-  valuation $\beta' : Q' \rightarrow \booli$ such that
+  \Delta(\phi,a)$, for some valuations $\beta : Q \rightarrow \booli$
+  and $\overline\nu : \overline{\vec{x}} \rightarrow \Data^\I$, $\nu :
+  \vec{x} \rightarrow \Data^\I$, then we build a valuation $\beta' :
+  Q' \rightarrow \booli$ such that
   $\I,\beta'\cup\beta\cup\overline\nu\cup\nu \models \phi[Q'/Q] \wedge
@@
 
+Moreover, we require that an interpolant
+$\tuple{\top,I_0,\ldots,I_n,\bot}$ for the interpolation problem
+$\Theta(u)$ does not have negative occurrences of states, i.e.\ $I_i
+\in \Form^+(Q,\vec{x})$, for all $i \in [0,n]$. Such an interpolant
+can always be built, as showed below:
+
+\begin{proposition}\label{prop:positive-interpol}
+  For any $u = a_1 \ldots a_n \in \Sigma^*$, if $\pi(u)$ is
+  unsatisfiable then one can build an interpolant
+  $\tuple{\top,I_0,\ldots,I_n,\bot}$ for $\pi(u)$, such that $I_i \in
+  \Form^+(Q,\vec{x})$, for all $i \in [0,n]$.
+\end{proposition}
+\proof{ Let $\tuple{\top,I_0,\ldots,I_{n},\bot}$ be an interpolant,
+  and $i \in [0,n]$ be the first index for which $I_i \not\in
+  \Form^+(Q,\vec{x})$. By Proposition \ref{prop:asterisk}, we have
+  that $\Delta^{i-1}(I_{i-1},a_i) \iffeq \exists Q_{i-1} ~.~
+  I_{i-1}[Q_{i-1}/Q] \wedge \theta_i$. Since $I_{i-1} \in
+  \Form^+(Q,\vec{x})$, by the choice of $i$, we have that
+  $\Delta^{i-1}(I_{i-1},a_i) \in \Form^+(Q,\vec{x})$ as well. We
+  replace $I_i$ with $\Delta^{i-1}(I_{i-1},a_i)$. The result is still
+  an interpolant, because: \begin{compactitem}
+  \item $I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \models
+    \Delta^{i-1}(I_{i-1},a_i)$, by Proposition \ref{prop:asterisk}, and
+    %
+  \item $\Delta^{i-1}(I_{i-1},a_i) \wedge \theta_{i+1} \models \exists
+    Q_{i-1} ~.~ I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \wedge \theta_{i+1}
+    \models I_i[Q_i/Q] \wedge \theta_{i+1} \models I_{i+1}[Q_{i+1}/Q]$.
+  \end{compactitem}
+  We proceed with these replacements until there are no more formulae
+  $I_i \not\in \Form^+(Q,\vec{x})$ left. \qed}
+
 We recall that a run of $\A$ over $u$ is a sequence $\phi_0(Q),
@@
 from $\phi_{k-1}$ by replacing each state $q \in
-\fv{\Bool}{\phi_{k-1}}$ by the formula $\Delta(q,a_k)$ given by the
-transition function of $\A$. Observe that, because the states are
-replaced with transition formulae when moving one step in a run, the
-formulae thus produced lose track of the control history and are not
-directly suitable to producing interpolants that relates states and
-data.
+\fv{\Bool}{\phi_{k-1}}$ by the formula
+$\Delta(q,a_k)[\vec{x}_{k-1}/\overline{\vec{x}},\vec{x}_k/\vec{x}]$,
+given by the transition function of $\A$. Observe that, because the
+states are replaced with transition formulae when moving one step in a
+run, the formulae thus produced lose track of the control history and
+are not directly suitable to producing interpolants that relates
+states and data.
 
 The main idea behind the above definition of the interpolation problem
-is that we would like to obtain an interpolant $\tuple{\top,I_1(Q),
+is that we would like to obtain an interpolant $\tuple{\top,I_0(Q),
   I_2(Q,\vec{x}), \ldots, I_{n}(Q,\vec{x}),\bot}$ whose formulae
@@
   hold: \begin{compactenum}
-  \item\label{it1:progress} $\pi(u)$ is unsatisfiable, and
-    %
-  \item\label{it2:progress} if $\tuple{\top,I_1,\dots,I_{n+1},\bot}$
-    is an interpolant for $\pi(u)$ and $\set{I_i \mid i \in[1,n+1]}
-    \subseteq \preds$ then $\AbsAccept{\A}(u)$ is unsatisfiable.
+  \item\label{it1:progress} $\Theta(u)$ is unsatisfiable, and
+    %
+  \item\label{it2:progress} if $\tuple{\top,I_0,\dots,I_{n},\bot}$ is
+    an interpolant for $\Theta(u)$ such that $\set{I_i \mid i
+      \in[0,n]} \subseteq \preds$ then $\AbsAccept{\A}(u)$ is
+    unsatisfiable.
   \end{compactenum}
 \end{lemma}
-\proof{ Let $\pi(u) \equiv \theta_0(Q_0) \wedge \theta_1(Q_0 \cup
+\proof{ Let $\Theta(u) \equiv \theta_0(Q_0) \wedge \theta_1(Q_0 \cup
   Q_1,\vec{x}_0 \cup \vec{x}_1) \wedge \ldots \wedge \theta_n(Q_{n-1}
@@
   \exists \vec{x}_{n-1} ~.~ \bigwedge_{i=0}^n \theta_i\] Assuming that
-  $\pi(u) \equiv \bigwedge_{i=0}^{n+1} \theta_i$ is satisfiable, we
-  obtain a model for $\Accept{\A}(u) \equiv \Post{\A}(\iota,u) \wedge
-  \theta_{n+1}[Q/Q_n]$.
-
-  (\ref{it2:progress}) If $\tuple{\top,I_1,\dots,I_{n+1},\bot}$ is an
+  $\Theta(u)$ is satisfiable, we obtain a model for $\Accept{\A}(u)
+  \equiv \Post{\A}(\iota,u) \wedge \theta_{n+1}[Q/Q_n]$.
+
+  (\ref{it2:progress}) If $\tuple{\top,I_0,\dots,I_n,\bot}$ is an
   interpolant for $\pi(u)$, the following entailments
   hold: \begin{compactitem}
-  \item $\theta_0 \models I_1[Q_0/Q]$,
-    %
-  \item $I_k[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k+1}[Q_k/Q,\vec{x}_k/\vec{x}]$, $\forall k\in[1,n]$. 
-    %
-  \item $I_{n+1}[Q_{n+1}/Q] \wedge \theta_{n+1} \models \bot$. 
+  \item $\theta_0 \models I_0[Q_0/Q]$,
+    %
+  \item $I_{k-1}[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k}[Q_k/Q,\vec{x}_k/\vec{x}]$, $\forall k\in[1,n]$. 
+    %
+  \item $I_{n}[Q_{n}/Q] \wedge \theta_{n+1} \models \bot$. 
   \end{compactitem}
-  We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n+1}$ by
+  We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n}$ by
   induction on $n \geq 0$. This is sufficient to conclude because
   $\AbsAccept{\A}(a_1\ldots a_n) \equiv \AbsPost{\A}(\iota,a_1\ldots
-  a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n+1} \wedge
+  a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n} \wedge
   \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we have
   $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv \theta_0[Q/Q_0]
-  \models I_1$. For the induction step $n>0$, we compute:
+  \models I_0$. For the induction step $n>0$, we compute:
   \[\begin{array}{l}
-  \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] \equiv \text{ (by definition of $\AbsPost{\A}$)} \\
+  \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] \equiv \text{ (by def. of $\AbsPost{\A}$)} \\
   \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] \models \text{ (by Prop. \ref{prop:asterisk})} \\
@@
 We have now all the ingredients to describe the first emptiness
-checking semi-algorithm for alternating data
+semi-algorithm for alternating data
 automata. Algorithm\footnote{Though termination is not guaranteed, we
   call it algorithm for conciseness.} \ref{alg:predabs} builds an
-\emph{abstract reachability tree} (ART) whose edges are labeled with
-input events and each node $n$ is labeled with the formula
-$\AbsPost{\A}(\iota,u)$, where $u \in \Sigma^*$ is the event trace
-labeling the path from the root of the tree to $n$. It uses a covering
-relation $\lhd$ between nodes in order to ensure that the set of
-formulae labeling the nodes in the ART forms an antichain. Any
-spurious counterexample $u$ is eliminated by computing the
-interpolants corresponding to the least unfeasible suffix $v$ of $u$
-and recomputing the subtree of the ART rooted in the initial node of
-$v$.
+\emph{abstract reachability tree} (ART) whose nodes are labeled with
+formulae over-approximating the concrete sets of configurations, which
+uses a covering relation between nodes in order to ensure that the set
+of formulae labeling the nodes in the ART forms an antichain. Any
+spurious counterexample is eliminated by computing an interpolant and
+adding its formulae to the set of predicates.
 
@@
     feasible}
+  \label{ln:feasible}
 
@@
 
-  \State $\preds \leftarrow \preds \cup \set{I_1,\ldots,I_\ell}$,
-  where $\tuple{\top,I_1,\ldots,I_\ell,\bot}$ is an interpolant for
-  $\overline{\pi}(v)$
+  \State $\preds \leftarrow \preds \cup \set{I_0,\ldots,I_\ell}$,
+  where $\tuple{\top,I_0,\ldots,I_\ell,\bot}$ is an interpolant for
+  $\pivot{v}$
 
@@
 
-Formally, an ART $\Art = \tuple{N,E,\rootNode,\Lambda,\lhd}$ is a
-tree, where $N$ is a set of nodes, $E \subseteq N \times \Sigma \times
-N$ is a set of edges, $\rootNode \in N$ is the root of the directed
-tree $(N,E)$, $\Lambda : N \rightarrow \Form^+(Q,\vec{x})$ is a
-labeling of the nodes with formulae, such that $\Lambda(\rootNode) =
-\iota$ and $\lhd \subseteq N \times N$ is a set of covering
-edges. Then each node $n \in N$ corresponds to a unique path from the
-root to $n$, labeled by a sequence $u$ of input events. The least
-unfeasible suffix of $u$ is the smallest sequence $v = a_1 \ldots
-a_k$, such that $u = wv$, for some $w \in \Sigma^*$ and the formula
-(\ref{eq:pivot}) below is unsatisfiable. The \emph{pivot} of $u$ is
-the node $p$ corresponding to the start of $v$ in $u$.
+Formally, an ART is tuple $\Art =
+\tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$, where: \begin{compactitem}
+\item $N$ is a set of nodes, 
+%
+\item $E \subseteq N \times \Sigma \times
+N$ is a set of edges, 
+%
+\item $\rootNode \in N$ is the root of the directed
+tree $(N,E)$, 
+%
+\item $\Lambda : N \rightarrow \Form(Q,\vec{x})$ is a labeling of the
+  nodes with formulae, such that $\Lambda(\rootNode) = \iota$,
+%
+\item $R : N \rightarrow 2^Q$ is a labeling of nodes with
+\emph{replacement sets}, such that $R(\rootNode) = \fv{\Bool}{\iota}$,
+%
+\item $T : E \rightarrow \bigcup_{i=0}^\infty
+\Form(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is labeling of edges with
+time-stamped formulae, and 
+%
+\item $\lhd \subseteq N \times N$ is a set of
+\emph{covering edges}. 
+\end{compactitem}
+For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is $(n,a,m)$.
+The replacement sets and time-stamped edge labels are defined below,
+for each edge $e = (n,a,m) \in E$, if $n$ lies at depth\footnote{We
+  consider that the root node has zero depth.} $k$ in the tree
+$(N,E)$ then we have:
+\[\begin{array}{rcl}
+T(e) & = & \bigwedge_{q \in R(n)} (q_k \rightarrow \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]) \\
+R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)}
+\end{array}\]
+Each node $n \in N$ corresponds to a unique path from the root to $n$,
+labeled by a sequence $\lambda(n) \in \Sigma^*$ of input events. The
+least unfeasible suffix of $\lambda(n)$ is the smallest sequence $v =
+a_1 \ldots a_k$, such that $\lambda(n) = wv$, for some $w \in
+\Sigma^*$ and the formula (\ref{eq:pivot}) below is unsatisfiable. The
+\emph{pivot} of $n$ is the node $p$ corresponding to the start of $v$
+in $\lambda(n)$.
 \begin{equation}\label{eq:pivot}
-  \overline{\pi}(v) \equiv \Lambda(p)[Q_0/Q] \wedge
+  \pivot{v} \equiv \Lambda(p)[Q_0/Q] \wedge
   \theta_1(Q_0 \cup Q_1,\vec{x}_0 \cup \vec{x}_1) \wedge \ldots \wedge
@@
 iteration to build an ART. We keep newly expanded nodes of $\Art$ in a
-use a queue $\worklist$ to ensure a breadth-first exploration
+queue $\worklist$, thus implementing a breadth-first exploration
 strategy, which guarantees that the shortest counterexamples are
-explored first. When the search encounters a counterexample $u$, it is
-checked for spuriousness. If the counterexample is feasible, the procedure
-returns a data word $w \in L(\A)$, which interleaves the input events
-of $u$ with the data valuations from the model of $\Accept{\A}(u)$
-(since $u$ is feasible, clearly $\Accept{\A}(u)$ is
-satisfiable). Otherwise, if $u$ is spurious, we compute its pivot $p$
-(line \ref{ln:pivot}), add the interpolants for the least unfeasible
-suffix of $u$ to the set of predicates $\preds$, remove and recompute
-the subtree of $\Art$ rooted at $p$.
+explored first. When the search encounters a counterexample candidate
+$u$, it is checked for spuriousness. If the counterexample is
+feasible, the procedure returns a data word $w \in L(\A)$, which
+interleaves the input events of $u$ with the data valuations from the
+model of $\Accept{\A}(u)$ (since $u$ is feasible, clearly
+$\Accept{\A}(u)$ is satisfiable). Otherwise, if $u$ is spurious, we
+compute its pivot $p$ (line \ref{ln:pivot}), add the interpolants for
+the least unfeasible suffix of $u$ to the set of predicates $\preds$,
+remove and recompute the subtree of $\Art$ rooted at $p$.
 
@@
 data theory $\theory(\sig,\I)$ to provide interpolants that yield a
-safety invariant, provided that $L(\A) = \emptyset$. In this case, we
+safety invariant, whenever $L(\A) = \emptyset$. In this case, we
 use the covering relation $\lhd$ to ensure that, when a newly
@@
 \end{theorem}
-\proof{To prove the first point, assume w.l.o.g. that $\Sigma =
-  \set{a_1,\ldots,a_{\card{\Sigma}}}$ is a total ordering of the input event
-  alphabet of $\A$ and that the loop at line \ref{ln:forall-events}
-  considers the events in this order. If $L(\A) \neq \emptyset$, let
-  $(a_{i_1},\nu_1) \ldots (a_{i_k},\nu_k) \in L(\A)$ be a word, such
-  that $a_{i_1} \ldots a_{i_k}$ is the lexicographically minimal
-  feasible counterexample for $\A$.  Since Algorithm \ref{alg:predabs}
-  explores $\Sigma^*$ in this lexicographical order, it must encounter
-  $a_{i_1} \ldots a_{i_k}$ and return a word $(a_{i_1},\nu'_1) \ldots
-  (a_{i_k},\nu'_k) \in L(\A)$ at line \ref{ln:cex}.
-
-  For the second point, we prove the following invariant: each time
-  Algorithm \ref{alg:predabs} reaches line \ref{ln:while}, the set $W$
-  of nodes in $\worklist$ contains all the frontier nodes in the ART
-  $\tuple{N\cup W,E,\rootNode,\Lambda,\lhd}$ which are not covered by some node
-  in $N$, namely that:
+\proof{We prove the following invariant: each time Algorithm
+  \ref{alg:predabs} reaches line \ref{ln:while}, the set $W$ of nodes
+  in $\worklist$ contains all the frontier nodes in the ART
+  $\tuple{N\cup W,E,\rootNode,\Lambda,\lhd}$ which are not covered by
+  some node in $N$, namely that:
   \begin{equation}\label{inv:predabs}
@@
 
-  We use this invariant in the proof as follows. If Algorithm
-  \ref{alg:predabs} returns at line \ref{ln:true}, then it must be
-  that $W = \emptyset$. Because of the invariant, each node in $N$ is
-  either covered by another node in $N$, or all its successors are in
-  $N$. We prove next that $\bigvee_{n \in N} \Lambda(n)$ defines a
-  safety invariant, which allows to conclude that $L(\A) = \emptyset$,
-  by Lemma \ref{lemma:safety-invariant}. To this end, it is sufficient
-  to prove that, for any $u \in \Sigma^*$, there exists some node $n
-  \in N$ such that $\Post{\A}(\iota,u) \models \Lambda(n)$. Let $u \in
-  \Sigma^*$ be an arbitrary sequence. If $u$ labels the path from
-  $\rootNode$ to some $n \in N$, we have $\Post{\A}(\iota,u) \models
-  \AbsPost{\A}(\iota,u) \models \Lambda(n)$ and we are
-  done. Otherwise, let $v$ be the (possibly empty) prefix of $u$ which
-  labels the path from $\rootNode$ to some $n \in N$, which is covered
-  by another $m \in N$, where $(n,a,m) \in E$, that is $u = vav'$, for
-  some $a \in \Sigma$ and $v'\in\Sigma^*$. Moreover, we have
-  $\Post{\A}(\iota,va) \models \AbsPost{\A}(\iota,va) \models
-  \Lambda(m)$, by the construction of the set $\lhd$ of covering edges
-  --- lines \ref{ln:direct-cover}, \ref{ln:child-cover},
+  Next we prove that, if Algorithm \ref{alg:predabs} returns $\true$,
+  then $\bigvee_{n \in N} \Lambda(n)$ defines a safety
+  invariant. Suppose that Algorithm \ref{alg:predabs} returns at line
+  \ref{ln:true}. Then it must be that $W = \emptyset$. Because
+  (\ref{inv:predabs}) is invariant, each node in $N$ is either covered
+  by another node in $N$, or all its successors are in $N$. We prove
+  first that $\bigvee_{n \in N} \Lambda(n)$ is an invariant: for any
+  $u \in \Sigma^*$, there exists some node $n \in N$ such that
+  $\Post{\A}(\iota,u) \models \Lambda(n)$. Let $u \in \Sigma^*$ be an
+  arbitrary sequence. If $u$ labels the path from $\rootNode$ to some
+  $n \in N$, we have $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)
+  \models \Lambda(n)$ and we are done. Otherwise, let $v$ be the
+  (possibly empty) prefix of $u$ which labels the path from
+  $\rootNode$ to some $n \in N$, which is covered by another $m \in
+  N$, where $(n,a,m) \in E$, that is $u = vav'$, for some $a \in
+  \Sigma$ and $v'\in\Sigma^*$. Moreover, we have $\Post{\A}(\iota,va)
+  \models \AbsPost{\A}(\iota,va) \models \Lambda(m)$, by the
+  construction of the set $\lhd$ of covering edges --- lines
+  \ref{ln:direct-cover}, \ref{ln:child-cover},
   \ref{ln:indirect-cover}. Continuing this argument recursively from
   $m$, since $\len{v'} < \len{u}$, we shall eventually discover a node
-  $p$ such that $\Post{\A}(\iota,u) \models \Lambda(p)$. This
-  concludes the proof. \qed}
+  $p$ such that $\Post{\A}(\iota,u) \models \Lambda(p)$.
+
+  To prove that $\bigvee_{n \in N} \Lambda(n)$ is, moreover, a safety
+  invariant, suppose, by contradiction, that there exists $u \in
+  \Sigma^*$ such that $\Accept{\A}(u)$ is satisfiable. By the previous
+  point, there exists a node $p \in N$ such that $\Post{\A}(\iota,u)
+  \models \Lambda(p)$. But then we have $\Accept{\A}(\iota,u) \models
+  \Accept{\A}(\iota,\lambda(p))$, thus $\Accept{\A}(\iota,\lambda(p))$
+  is satisfiable as well. However, this cannot be the case, because
+  $p$ has been processed at line \ref{ln:feasible} and Algorithm
+  \ref{alg:predabs} would have returned a counterexample,
+  contradicting the assumption that it returns $\true$. This concludes
+  the proof that $\bigvee_{n\in N} \Lambda(n)$ is a safety invariant,
+  thus $L(\A) = \emptyset$, by Lemma \ref{lemma:safety-invariant}.
+
+  We have then proved the second point of the statement. For the first
+  point, assume that $L(\A) \neq 0$ and let $w = (a_1,\nu_1) \ldots
+  (a_k,\nu_k) \in L(\A)$ be a word. By the above, Algorithm
+  \ref{alg:predabs} cannot return $\true$. Suppose, by contradiction
+  that it does not terminate. Since the sequences from $\Sigma^*$ are
+  explored in breadth-first order, every sequence of length $k$ is
+  eventually explored, which leads to the discovery of $w$ at line
+  \ref{ln:feasible}. Then Algorithm \ref{alg:predabs} terminates
+  returning $w \in L(\A)$. \qed}
 
@@
 
-The ART built by Algorithm \ref{alg:impact} corresponds to an
-unfolding of the input automaton by enumerating the input event
-sequences in $\Sigma^*$. Each time a new node is created, its label is
-initially $\top$. This label is subject to subsequent strenghtening,
-when a spurious counterexample is encountered. To speedup the
-computation of interpolants for spurious counterexamples, we use two
-additional labeling functions for bookkeeping. Formally, the ART
-structure used by Algorithm \ref{alg:impact} is defined as $\Art =
-\tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$, where $N,E,\Lambda$ and
-$\lhd$ are defined as for Algorithm \ref{alg:predabs} and,
-moreover: \begin{compactitem}
-\item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
-  sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$, 
-%
-\item $T : E \rightarrow \bigcup_{i=0}^\infty
-  \Form(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is labeling of edges
-  with time-stamped formulae, and
-%
-\item for each edge $e = (n,a,m) \in E$, if $n$ lies at
-  depth\footnote{We consider that the root node has zero depth.}  $k$
-  in the tree $(N,E)$, then we have:
-\end{compactitem}
-  \[\begin{array}{rcl}
-  T(e) & = & \Theta(n,a,k) \\
-  R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)} \text{, where} \\
-  \Theta(n,a,k) & \equiv & \bigwedge_{q \in R(n)} (q_k \rightarrow
-  \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}])
-  \end{array}\]
-For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is
-$(n,a,m)$. This additional bookkeeping is used as follows. Given a
-sequence $u = a_1\ldots a_k \in \Sigma^*$ that labels the path from
-the root to a node $n$ in $\Art$, the conjunction:
-\[\mu(n) \equiv \Lambda(\rootNode) \wedge T(\rootNode,a_1,n_1) \wedge \ldots
-T(n_{k-1},a_k,n_k) \wedge \bigwedge_{q \in R(n_k) \setminus F}(q_k
-\rightarrow \bot)\] is equivalent to the interpolation problem
-$\pi(u)$ and can be directly computed by a traversal of the path from
-$\rootNode$ to $n$. In addition, we require that an interpolant
-$\tuple{\top,I_0,\ldots,I_k,\bot}$ for the insatisfiable interpolation
-problem $\mu(n)$ does not have negative occurrences of states,
-i.e.\ $I_i \in \Form^+(Q,\vec{x})$, for all $i \in [0,k]$. Such an
-interpolant can always be built, as showed below:
-
-\begin{proposition}\label{prop:positive-interpol}
-  For any $u = a_1 \ldots a_n \in \Sigma^*$, if $\pi(u)$ is
-  unsatisfiable then one can build an interpolant
-  $\tuple{\top,I_0,\ldots,I_n,\bot}$ for $\pi(u)$, such that $I_i
-  \in \Form^+(Q,\vec{x})$, for all $i \in [0,n]$.
-\end{proposition}
-\proof{ Let $\tuple{\top,I_0,\ldots,I_{n},\bot}$ be an interpolant,
-  and $i \in [0,n]$ be the first index for which $I_i \not\in
-  \Form^+(Q,\vec{x})$. By Proposition \ref{prop:asterisk}, we have
-  that $\Delta^{i-1}(I_{i-1},a_i) \iffeq \exists Q_{i-1} ~.~
-  I_{i-1}[Q_{i-1}/Q] \wedge \theta_i$. Since $I_{i-1} \in
-  \Form^+(Q,\vec{x})$, by the choice of $i$, we have that
-  $\Delta^{i-1}(I_{i-1},a_i) \in \Form^+(Q,\vec{x})$ as well. We
-  replace $I_i$ with $\Delta^{i-1}(I_{i-1},a_i)$. The result is still
-  an interpolant, because: \begin{compactitem}
-  \item $I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \models
-    \Delta^{i-1}(I_{i-1},a_i)$, by Proposition \ref{prop:asterisk}, and
-    %
-  \item $\Delta^{i-1}(I_{i-1},a_i) \wedge \theta_{i+1} \models \exists
-    Q_{i-1} ~.~ I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \wedge \theta_{i+1}
-    \models I_i[Q_i/Q] \wedge \theta_{i+1} \models I_{i+1}[Q_{i+1}/Q]$.
-  \end{compactitem}
-  We proceed with these replacements until there are no more formulae
-  $I_i \not\in \Form^+(Q,\vec{x})$ left. \qed}
+%% The ART built by Algorithm \ref{alg:impact} corresponds to an
+%% unfolding of the input automaton by enumerating the input event
+%% sequences in $\Sigma^*$. Each time a new node is created, its label is
+%% initially $\top$. This label is subject to subsequent strenghtening,
+%% when a spurious counterexample is encountered. To speedup the
+%% computation of interpolants for spurious counterexamples, we use two
+%% additional labeling functions for bookkeeping. Formally, the ART
+%% structure used by Algorithm \ref{alg:impact} is defined as $\Art =
+%% \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$, where $N,E,\Lambda$ and
+%% $\lhd$ are defined as for Algorithm \ref{alg:predabs} and,
+%% moreover: \begin{compactitem}
+%% \item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
+%%   sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$, 
+%% %
+%% \item $T : E \rightarrow \bigcup_{i=0}^\infty
+%%   \Form(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is labeling of edges
+%%   with time-stamped formulae, and
+%% %
+%% \item for each edge $e = (n,a,m) \in E$, if $n$ lies at
+%%   depth\footnote{We consider that the root node has zero depth.}  $k$
+%%   in the tree $(N,E)$, then we have:
+%% \end{compactitem}
+%%   \[\begin{array}{rcl}
+%%   T(e) & = & \Theta(n,a,k) \\
+%%   R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)} \text{, where} \\
+%%   \Theta(n,a,k) & \equiv & \bigwedge_{q \in R(n)} (q_k \rightarrow
+%%   \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}])
+%%   \end{array}\]
+%% For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is
+%% $(n,a,m)$. This additional bookkeeping is used as follows. Given a
+%% sequence $u = a_1\ldots a_k \in \Sigma^*$ that labels the path from
+%% the root to a node $n$ in $\Art$, the conjunction:
+%% \[\mu(n) \equiv \Lambda(\rootNode) \wedge T(\rootNode,a_1,n_1) \wedge \ldots
+%% T(n_{k-1},a_k,n_k) \wedge \bigwedge_{q \in R(n_k) \setminus F}(q_k
+%% \rightarrow \bot)\] is equivalent to the interpolation problem
+%% $\pi(u)$ and can be directly computed by a traversal of the path from
+%% $\rootNode$ to $n$.
 
@@
   \State dequeue $n$ from $\worklist$ 
+  \label{ln:impact-dequeue}
 
@@
 
-  \If{$\mu(n)$ is satisfiable} 
+  \If{$\Phi(a_1 \ldots a_k)$ is satisfiable} 
   \Comment{counterexample is feasible}
@@
   \State {\bf return} $w = (a_1,\nu_1) \ldots (a_k,\nu_k)$
+  \label{ln:impact-nonempty}
   \Comment{$w \in L(\A)$ by construction}
@@
 
-  \State let $\tuple{\top,I_0,\ldots,I_k,\bot}$ be an
-  interpolant for $\mu(n)$ with $I_0,\ldots,I_k \in
-  \Form^+(Q,\vec{x})$
+  \State let $\tuple{\top,I_0,\ldots,I_k,\bot}$ be an interpolant for
+  $\Phi(a_1 \ldots a_k)$
   \label{ln:refine-begin}
@@
   description of \impact~\cite{mcmillan06}.}: \emph{close},
-\emph{refine} and \emph{expand}. 
+\emph{refine} and \emph{expand}. Let $n$ be a node removed from the
+worklist at line \ref{ln:impact-dequeue}. If $\Phi(\lambda(n))$ is
+satisfiable, the counterexample $\lambda(n)$ is feasible, in which
+case a model of $\Phi(\lambda(n))$ is obtained and a word $w \in
+L(\A)$ is returned. 
 %
@@
 %
-Let $u$ be the label of the path from the root node $\rootNode$ to
-$n$. If $\mu(n)$ is satisfiable, then $u$ is a feasible
-counterexample, in which case a model of $\mu(n)$ is obtained and a
-word $w \in L(\A)$ returned. Otherwise, $u$ is a spurious
-counterexample and the procedure enters the refinement phase (lines
+Otherwise, $\lambda(n)$ is a spurious counterexample and the procedure
+enters the refinement phase (lines
 \ref{ln:refine-begin}-\ref{ln:refine-end}). The interpolant for
-$\mu(n)$ is used to strenghten the labels of all the ancestors of $n$,
-by conjoining the formulae of the interpolant to the existing
-labels. For technical reasons, related to proving the correctness of
-Algorithm \ref{alg:impact}, we assume that all formulae from the
-interpolant for $\mu(n)$ have only positive occurrences of the
-states\footnote{According to Proposition \ref{prop:positive-interpol},
-  building such an interpolant is always possible} in $Q$.
+$\Phi(\lambda(n))$ is used to strenghten the labels of all the
+ancestors of $n$, by conjoining the formulae of the interpolant to the
+existing labels.
 
@@
 
-Finally, if $n$ is still uncovered (i.e.\ it has not been previously
-covered during the refinement phase) we expand $n$ (lines
+Finally, if $n$ is still uncovered (it has not been previously covered
+during the refinement phase) we expand $n$ (lines
 \ref{ln:expand-begin}-\ref{ln:expand-end}) by creating a new node for
 each successor $s$ via the event $a \in \Sigma$ and inserting it into
-the worklist. The label of the new node is $\top$, its replacement set
-$R(s)$ is computed based on $R(n)$ and the label of the edge $(n,a,s)$
-is $\Theta(n,a,k)$, where $k$ is the depth of $n$ in $\Art$.
+the worklist.
 
@@
   be a spurious counterexample. Let
-  $\langle\top,I^\omega_1,\ldots,I^\omega_k\equiv\bot\rangle$ be an
-  interpolant for $\mu(n_k) \equiv \Lambda(\rootNode) \wedge
-  \bigwedge_{i=1}^k \Theta(n_{i-1},a,i-1) \wedge \bigwedge_{q \in
-    R(n_k)} (q_k \rightarrow \bot)$, such that $I^\omega_i \in
-  \Form^+(Q,\vec{x})$, for all $i \in [1,k]$. According to Proposition
-  \ref{prop:positive-interpol}, it is always possible to build such an
-  interpolant, when $\mu(n_k)$ is unsatisfiable. By Proposition
-  \ref{prop:asterisk}, we obtain $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q]
-  \iffeq \exists Q_{i-1} ~.~
+  $\langle\top,I^\omega_0,\ldots,I^\omega_k\equiv\bot\rangle$ be an
+  interpolant for $\Phi(a_1 \ldots a_k) \equiv \Lambda(\rootNode)
+  \wedge \bigwedge_{i=1}^k \theta_i \wedge \bigwedge_{q \in R(n_k)}
+  (q_k \rightarrow \bot)$, such that $I^\omega_i \in
+  \Form^+(Q,\vec{x})$, for all $i \in [0,k]$. According to Proposition
+  \ref{prop:positive-interpol}, it is possible to build such an
+  interpolant, when $\Phi(a_1 \ldots a_k)$ is unsatisfiable. By
+  Proposition \ref{prop:asterisk}, we obtain
+  $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q] \iffeq \exists Q_{i-1} ~.~
   I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge \theta_i$
@@
 \end{theorem}
-\proof{ The first point is proved along the lines of the first point
-  of Theorem \ref{thm:predabs}. In particular, it is easy to see that
-  each call to $\Call{Close}{}$ terminates, because $N$ is finite at
-  each moment during the execution. For the second point, we prove
-  that each time Algorithm \ref{alg:impact} reaches the line
-  \ref{ln:impact-while}, we have:
+\proof{ We prove first that, each time Algorithm \ref{alg:impact}
+  reaches the line \ref{ln:impact-while}, we have:
   \begin{equation}\label{inv:impact}
@@
   Then the condition (\ref{inv:impact}) holds next time line
-  \ref{ln:impact-while} is reached, thus it is invariant. If Algorithm
-  \ref{alg:impact} returns $\true$ then $W = \emptyset$ and, by
-  (\ref{inv:impact}), for each node in $n \in N$ one of the following
-  hold: \begin{compactitem}
+  \ref{ln:impact-while} is reached, thus it is invariant.
+
+  Suppose first that Algorithm \ref{alg:impact} returns $\true$, thus
+  $W = \emptyset$ and, by (\ref{inv:impact}), for each node in $n \in
+  N$ one of the following hold: \begin{compactitem}
   \item $n$ is covered, or
@@
   \end{compactitem}
-  Moreover, we have $u = v'_1 \ldots v'_\ell$, thus $\Post{\A}(\iota,u)
-  \models \Lambda(m_\ell)$ and we are done showing that $\bigvee_{n\in
-    N} \Lambda(n)$ is an invariant. To prove that it is, moreover, a
-  safety invariant, 
-\qed}
+  Moreover, we have $u = v'_1 \ldots v'_\ell$, thus
+  $\Post{\A}(\iota,u) \models \Lambda(m_\ell)$ and we are done showing
+  that $\bigvee_{n\in N} \Lambda(n)$ is an invariant.
+
+  To prove that $\bigvee_{n \in N} \Lambda(n)$ is, moreover, a safety
+  invariant, suppose that $\Accept{\A}(u)$ is satisfiable, for some $u
+  \in \Sigma^*$ and let $n \in N$ be a node such that
+  $\Post{\A}(\iota,u) \models \Lambda(n)$. By the previous point, such
+  a node must exist. But then $\Accept{\A}(u) \models
+  \Accept{\A}(\lambda(n))$, thus $\Accept{\A}(\lambda(n))$ is
+  satisfiable, and Algorithm \ref{alg:impact} returns at line
+  \ref{ln:impact-nonempty}, upon encountering $\lambda(n)$. But this
+  contradicts the assumption that Algorithm \ref{alg:impact} returns
+  $\true$, hence we have proved that $\bigvee_{n \in N} \Lambda(n)$ is
+  a safety invariant, and $L(\A) = \emptyset$ follows, by Lemma
+  \ref{lemma:safety-invariant}. We have then proved the second point
+  of the statement.
+
+  To prove the first point, assume that $L(\A) \neq \emptyset$. By the
+  previous point, Algorithm \ref{alg:impact} does not return
+  $\true$. Suppose, by contradiction, that it does not terminate and
+  conclude using the breadth-first exploration argument from the proof
+  of Theorem \ref{thm:predabs}. \qed}
 

### 2017-05-13 draft.tex
ok

@@
 is that we would like to obtain an interpolant $\tuple{\top,I_0(Q),
-  I_2(Q,\vec{x}), \ldots, I_{n}(Q,\vec{x}),\bot}$ whose formulae
+  I_1(Q,\vec{x}), \ldots, I_{n}(Q,\vec{x}),\bot}$ whose formulae
 \emph{combine states with the data constraints that must hold
@@
   $\true$. Suppose, by contradiction, that it does not terminate and
-  conclude using the breadth-first exploration argument from the proof
-  of Theorem \ref{thm:predabs}. \qed}
+  conclude using the breadth-first argument from the proof of Theorem
+  \ref{thm:predabs}. \qed}
 

### 2017-05-14 draft.tex
intro+related

@@
 \section{Introduction}
+
+The language inclusion problem is recognized as being central to
+verification of hardware, communication protocols and software
+systems. A property is a specification of the correct executions of a
+system, given as a set $\mathcal{P}$ of executions, and the
+verification problem asks if the set $\mathcal{S}$ of executions of
+the system under consideration is contained within $\mathcal{P}$.
+
+This problem is at the core of widespread verification techniques,
+such as automata-theoretic model checking \cite{VARDI94}, where
+systems are specified as finite-state automata and properties defined
+using Linear Temporal Logic \cite{Pnueli77}. However the bottleneck of
+this and other related verification techniques is the intractability
+of language inclusion (PSPACE-complete for finite-state automata on
+finite words over finite alphabets).
+
+Alternation \cite{ChandraKozenStockmeyer81} was introduced as a
+generalization of nondeterminism, introducing universal, in addition
+to existential transitions. For automata over finite alphabets, the
+language inclusion problem can be encoded as the emptiness problem of
+an alternating automaton of linear size. Moreover, efficient
+exploration techniques based on antichains are shown to perform well
+for alternating automata over finite alphabets \cite{DeWulf08}.
+
+Using finite alphabets for the specification of properties and models
+is however very restrictive, when dealing with real-life computer
+systems, mostly because of the following reasons. On one hand,
+programs handle data from very large domains, that can be assumed to
+be infinite (64-bit integers, floating point numbers, strings of
+characters, etc.) and their correctness must be specified in terms of
+the data values. On the other hand, systems must respond to strict
+deadlines, which requires temporal specifications as timed
+languages \cite{AlurDill94}.
+
+Although being convenient specification tools, automata over infinite
+alphabets lack the decidability properties ensured by finite
+alphabets. In general, when considering infinite data as part of the
+input alphabet, language inclusion is undecidable and, even
+complementation becomes impossible, for instance, for timed automata
+\cite{AlurDill94} or finite-memory register automata
+\cite{Kaminski94}. In some cases, one can recover theoretical
+decidability, by restricting the number of variables (clocks) in timed
+automata to one \cite{OuaknineWorrell04}, or forbidding relations
+between current and past/future values, as with symbolic automata
+\cite{symbTransd:POPL12}. In such cases, the emptiness problem for the
+alternating versions becomes decidable \cite{Lasota05,DAntoniKW16}.
+
+In this paper, we present a new model of alternating automata over
+infinite alphabets consisting of pairs $(a,\nu)$ where $a$ is an input
+event from a finite set and $\nu$ is a valuation of a finite set
+$\vec{x}$ of variables that range over an infinite domain. We assume
+that, at all times, the successive values taken by the variables in
+$\vec{x}$ are an observable part of the language, in other words,
+there are no hidden variables in our model. The transition rules are
+specified by a set of formulae, in a combined first-order theory of
+boolean control states and data, that relate past with present values
+of the variables. We do not fix the data theory a~priori, but rather
+consider it to be a parameter of the class.
+
+A run over an input word $(a_1,\nu_1) \ldots (a_n,\nu_n)$ is a
+sequence $\phi_0(\vec{x}_0) \Rightarrow \phi_1(\vec{x}_0,\vec{x}_1)
+\Rightarrow \ldots \Rightarrow \phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ of
+rewritings of the initial formula by substituting boolean states with
+time-stamped transition rules. The word is accepted if the final
+formula $\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ holds, when all
+time-stamped variables $\vec{x}_1,\ldots,\vec{x}_n$ are substituted by
+their values in $\nu_1,\ldots,\nu_n$, all non-final states replaced by
+false and all final states by true.
+
+The boolean operations of union, intersection and complement can be
+implemented in linear time in this model, thus matching the complexity
+of performing these operations in the finite-alphabet case. The price
+to be paid is that emptiness becomes undecidable, for which reason we
+provide two efficient semi-algorithms for emptiness, based on lazy
+predicate abstraction \cite{blast} and the \impact~method
+\cite{mcmillan06}. These algorithms are proven to terminate and return
+a word from the language, if the language is not empty, but
+termination is not guaranteed when the language is empty. Finally, we
+prove termination in the empty case, when the transition
+formulae\begin{inparaenum}[(i)]
+\item contain only constants of the data sort and boolean functions of
+  any arity, or they
+%
+\item constrain only the current but not the past values on the input,
+  as in the case of symbolic automata \cite{symbTransd:POPL12,DAntoniKW16}.
+\end{inparaenum}
+
+We have implemented the boolean operations and emptiness checking
+semi-algorithms in a prototype tool and carried out experiments with
+examples taken from array logics \cite{cav09}, timed automata
+\cite{henzinger:RealTimeSystems}, communication protocols \cite{abp}
+and hardware verification \cite{smrcka}.
+
+\paragraph{Related Work}
+Data languages and automata have been defined previously, in a
+classical nondeterministic setting. For instance, Kaminski and Francez
+\cite{Kaminski94} consider languages, over an infinite alphabet of
+data, recognized by automata with a finite number of registers, that
+store the input data and compare it using equality. Just as the timed
+languages recognized by timed automata \cite{AlurDill94}, these
+languages, called quasi-regular, are not closed under complement, but
+their emptiness is decidable. The impossibility of complementation
+here is caused by the use of hidden variables, which we do not
+allow. Even though closed under complement, emptiness is undecidable
+in our case, mainly because counting (incrementing and comparing to a
+constant) data values is allowed, in many data theories.
+
+Other approaches to defining notions of data automata
+\cite{Bojanczyk:2011:TLD,BouyerPetitTherien03} are different from ours
+mainly because these automata are controlled by equivalence relations
+of finite index, whereas in our case, the transitions are defined by
+unrestricted formulae in the first-order theory of the data domain. As
+in \cite{Kaminski94,AlurDill94}, the emptiness problem is decidable
+for these classes.
+
+Another related model is that of predicate automata
+\cite{Farzan15,Farzan16}, which recognize languages over integer data
+by labeling the words with conjunctions of uninterpreted
+predicates. We intend to explore further the connection with our model
+of alternating data automata, in order to apply our method to the
+verification of parallel programs.
+
+The model presented in this paper stems from the language inclusion
+problem considered in \cite{Tacas16}. There we provide a
+semi-algorithm for inclusion of data languages, based on an
+exponential determinization procedure and an abstraction refinement
+loop using lazy predicate abstraction \cite{blast}. In this work we
+consider the full model of alternation and rely entirely on the
+ability of SMT solvers to produce interpolants in the combined theory
+of booleans and data. Since determinisation is not needed and
+complementation is possible in linear time, the bulk of the work is
+shipped to the decision procedure.
+
+The emptiness check is implemented by adapting similar semi-algorithms
+for nondeterministic infinite-state programs to the alternating model
+of computation. In particular, we considered the state-of-the-art
+\impact~ procedure \cite{mcmillan06} that is shown to outperform lazy
+predicate abstraction in the nondeterministic case, and generalized it
+to cope with alternation. More recent approaches for interpolant-based
+abstraction refinement \cite{McMillan14,Hoder12} target Horn systems,
+used to encode recursive and concurrent programs
+\cite{Grebenshchikov12}. However, the emptiness of alternating word
+automata cannot be directly encoded using Horn clauses, because all
+the branches of the computation synchronize on the same input, which
+does not seem to be easily encoded using a finite number of local
+(equality) constraints. We believe that the lazy annotation techniques
+for Horn clauses are suited for branching computations, which we
+intend to consider in a future tree automata setting.
 

### 2017-05-14 draft.tex
added examples

@@
 \usepackage{paralist}
-\usepackage{graphics}
 \usepackage{stmaryrd}
@@
 
+\begin{figure}[htb]
+\centerline{\includegraphics[scale=0.8]{figures/Example.pdf}}
+\caption{Alternating Data Automaton Example}
+\label{fig:ex}
+\end{figure}
+
+\paragraph{\em Example}
+Figure \ref{fig:ex} depicts an alternating data automaton, with input
+alphabet $\Sigma = \set{a,b}$, variables $\vec{x} = \set{x,y}$, states
+$Q = \set{q_0,q_1,q_2,q_3,q_4}$, initial configuration $q_0$, final
+states $F=\set{q_3,q_4}$ and transitions:
+\[\begin{array}{rcl}
+\Delta(q_0,a) & \equiv & q_1 \wedge q_2 \wedge x=0 \wedge y=0 \\
+\Delta(q_1,a) & \equiv & q_1 \wedge q_3 \wedge x=\overline{y}+1 \wedge y=\overline{x}+1 \\
+\Delta(q_1,b) & \equiv & q_3 \wedge \overline{x} \geq \overline{y} \\
+\Delta(q_2,a) & \equiv & q_2 \wedge x > \overline{x} \wedge y > \overline{y} \\
+\Delta(q_2,b) & \equiv & q_4 \wedge \overline{x}>\overline{y}
+\end{array}\]
+The missing transition rules are assumed to be $\false$. Rules
+$\Delta(q_0,a)$ and $\Delta(q_1,a)$ are universal and there are no
+existential nondeterministic rules. Rules $\Delta(q_1,a)$ and
+$\Delta(q_2,b)$ compare past ($\overline{x},\overline{y}$) with present
+($x,y$) values, $\Delta(q_0,a)$ constrains the present and
+$\Delta(q_1,b)$, $\Delta(q_2,b)$ the past values, respectively. \qed
+
 Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, $k\geq0$, be
@@
 $\Sigma[\vec{x}]^*$ accepted by $\A$.
-%
+
+\paragraph{\em Example} 
+The following sequence is a non-accepting run of the ADA from Figure
+\ref{fig:ex} on the word
+$(a,\tuple{0,0}),(a,\tuple{1,1}),(b,\tuple{2,1})$: 
+\[\begin{array}{l}
+q_0 \stackrel{(a,\tuple{0,0})}{\Longrightarrow} 
+q_1 \wedge q_2 \wedge x_1=0 \wedge y_1=0 \stackrel{(a,\tuple{1,1})}{\Longrightarrow} \\ 
+(q_1 \wedge q_2 \wedge x_2=y_1+1 \wedge y_2=x_1+1) ~\wedge \\[-2mm]
+(q_2 \wedge x_2>x_1 \wedge y_2>y_1) \wedge x_1=0 \wedge y_1=0 \stackrel{(b,\tuple{2,1})}{\Longrightarrow} \\
+(q_3 \wedge x_2\geq y_2) \wedge (q_4 \wedge x_2>y_2) \wedge x_2=y_1+1 \wedge y_2=x_1+1 \\ 
+\wedge~ (q_4 \wedge x_2>y_2) \wedge x_2>x_1 \wedge y_2>y_1 \wedge x_1=0 \wedge y_1=0 
+\rightarrow \bot~ \hfill\text{\qed}
+\end{array}\]
+
 In this paper we tackle the following problems: \begin{compactenum}
@@
 
+\begin{figure}[htb]
+\begin{tabular}{cc}
+\includegraphics[scale=0.5]{figures/PA1.pdf} & \includegraphics[scale=0.5]{figures/PA2.pdf} \\
+\footnotesize{(a)} & \footnotesize{(b)} \\
+\\
+\includegraphics[scale=0.5]{figures/PA3.pdf} & \includegraphics[scale=0.5]{figures/PA4.pdf} \\
+\footnotesize{(c)} & \footnotesize{(d)}
+\end{tabular}
+\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex} using Algorithm \ref{alg:predabs}}
+\label{fig:pa}
+\end{figure}
+
+\paragraph{\em Example}
+Consider the automaton given in Figure \ref{fig:ex}. First, Algorithm
+\ref{alg:predabs} fires the sequence $a$, and since there are no other
+formulae than $\bot$ in $\preds$, the successor of $\iota \equiv q_0$
+is $\top$ (a). The pivot of $a$ is the root of the ART and the
+spuriousness check results in the predicates $\set{q_0,q_1}$, which
+are added to $\preds$ and the $\top$ node is deleted. The next time
+$a$ is fired, it creates a node labeled $q_1$ and the next sequence
+$aa$ creates another node $q_1$ which is covered by the first,
+depicted with a dashed arrow (b). The third sequence is $ab$, which
+results in a new node $\top$ and triggers a spuriousness check. The
+new predicate is $x\leq0\wedge q_2 \wedge y\geq0$ and the pivot is
+again the root. Then the entire ART is rebuild with the new predicates
+and the sequence $aab$ results in an uncovered node $\top$ (c). The
+new pivot is the endpoint of $a$ and the new predicates are $q_1\wedge
+q_2$ and $y>x-1 \wedge q_2$. Finally, the ART is rebuilt from the
+pivot node and the result is a tree in which all nodes are covered,
+thus proving the emptiness of the automaton. \qed
+
 Termination of Algorithm \ref{alg:predabs} depends on the ability of a

### 2017-05-14 draft.tex
ok

@@
 formulae than $\bot$ in $\preds$, the successor of $\iota \equiv q_0$
-is $\top$ (a). The pivot of $a$ is the root of the ART and the
-spuriousness check results in the predicates $\set{q_0,q_1}$, which
-are added to $\preds$ and the $\top$ node is deleted. The next time
-$a$ is fired, it creates a node labeled $q_1$ and the next sequence
-$aa$ creates another node $q_1$ which is covered by the first,
-depicted with a dashed arrow (b). The third sequence is $ab$, which
-results in a new node $\top$ and triggers a spuriousness check. The
-new predicate is $x\leq0\wedge q_2 \wedge y\geq0$ and the pivot is
-again the root. Then the entire ART is rebuild with the new predicates
-and the sequence $aab$ results in an uncovered node $\top$ (c). The
-new pivot is the endpoint of $a$ and the new predicates are $q_1\wedge
-q_2$ and $y>x-1 \wedge q_2$. Finally, the ART is rebuilt from the
-pivot node and the result is a tree in which all nodes are covered,
-thus proving the emptiness of the automaton. \qed
+is $\top$, in Figure \ref{fig:pa} (a). The pivot of $a$ is the root of
+the ART and the spuriousness check yields the predicates $q_0$ and
+$q_1$, which are added to $\preds$ and the $\top$ node is removed. The
+next time $a$ is fired, it creates a node labeled $q_1$ and the
+sequence $aa$ creates another node $q_1$, which is covered by the
+first, depicted with a dashed arrow, in Figure \ref{fig:pa} (b). The
+third sequence is $ab$, which results in a new uncovered node $\top$
+and triggers a spuriousness check. The new predicate obtained from
+this check is $x\leq0\wedge q_2 \wedge y\geq0$ and the pivot is again
+the root. Then the entire ART is rebuild with the new predicates and
+the sequence $aab$ results in an uncovered node $\top$, in Figure
+\ref{fig:pa} (c). The new pivot is the endpoint of $a$ and the new
+predicates are $q_1\wedge q_2$ and $y>x-1 \wedge q_2$. Finally, the
+ART is rebuilt from the pivot node and the result is a tree in which
+all nodes are covered, thus proving the emptiness of the automaton, in
+Figure \ref{fig:pa} (d). \qed
 
@@
 
+\begin{figure}[htb]
+\begin{tabular}{cc}
+\includegraphics[scale=0.42]{figures/IM1.pdf} & \includegraphics[scale=0.42]{figures/IM2.pdf} \\
+\footnotesize{(a)} & \footnotesize{(b)} \\
+\\
+\includegraphics[scale=0.42]{figures/IM3.pdf} & \includegraphics[scale=0.42]{figures/IM4.pdf} \\
+\footnotesize{(c)} & \footnotesize{(d)}
+\end{tabular}
+\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex} using Algorithm \ref{alg:impact}}
+\label{fig:im}
+\end{figure}
+
+\paragraph{\em Example}
+We show the execution of Algorithm \ref{alg:impact} on the automaton
+from Figure \ref{fig:ex}.  Initially, the procedure fires the sequence
+$a$, whose endpoint it labeled with $\top$, in Figure \ref{fig:im}
+(a). Since this node is uncovered, it checks the spuriousness of the
+counterexample $a$ and refines the label of the node to $q_1$. Since
+the node is still uncovered, two successors, labeled with $\top$ are
+computed, corresponding to the sequences $aa$ and $ab$, in Figure
+\ref{fig:im} (b). The spuriousness check for $aa$ yields the
+interpolant $\tuple{q_0,x\leq0\wedge q_2 \wedge y\geq0}$ which
+strenghtens the label of the endpoint of $a$ from $q_1$ to $q_1\wedge
+x\leq0\wedge q_2\wedge y\geq0$. The sequence $ab$ is also found to be
+spurious, which changes the label of its endpoint from $\top$ to
+$\bot$, and also covers it. Since the endpoint of $aa$ is not covered,
+it is expanded in Figure \ref{fig:im} (c). Both sequences $aaa$ and
+$aab$ are found to be spurious, and the enpoint of $aab$, whose label
+has changed from $\top$ to $\bot$, is now covered. In the process, the
+label of $aa$ has also changed from $q_1$ to $q_1 \wedge y>x-1 \wedge
+q_2$, due to the strenghtening with the interpolant from $aab$.
+Finally, the only uncovered node is expanded to $aaaa$ and $aaab$,
+both found to be spurious, in Figure \ref{fig:im} (d). The refinement
+of $aaab$ causes the label of $aaa$ to change from $q_1$ to $q_1
+\wedge y>x-1 \wedge q_2$ and this node is now covered by $aa$. Since
+its successors are also covered, there are no uncovered nodes and the
+procedure returns $\true$. \qed
+
 The following lemma states that the labeling of the nodes in an ART

### 2017-05-14 draft.tex
appendix

@@
 predicate abstraction \cite{blast} and the \impact~method
-\cite{mcmillan06}. These algorithms are proven to terminate and return
-a word from the language, if the language is not empty, but
-termination is not guaranteed when the language is empty. Finally, we
-prove termination in the empty case, when the transition
-formulae\begin{inparaenum}[(i)]
-\item contain only constants of the data sort and boolean functions of
-  any arity, or they
-%
-\item constrain only the current but not the past values on the input,
-  as in the case of symbolic automata \cite{symbTransd:POPL12,DAntoniKW16}.
-\end{inparaenum}
+\cite{mcmillan06}.  These algorithms are proven to terminate and
+return a word from the language, if the language is not empty, but
+termination is not guaranteed when the language is empty.
+
+%% Finally, we prove termination in the empty case, when the transition
+%% formulae\begin{inparaenum}[(i)]
+%% \item contain only constants of the data sort and boolean functions of
+%%   any arity, or they
+%% %
+%% \item constrain only the current but not the past values on the input,
+%%   as in the case of symbolic automata \cite{symbTransd:POPL12,DAntoniKW16}.
+%% \end{inparaenum}
 
@@
 It is easy to check that $\len{\overline{\phi}} = \len{\phi}$, for
-every formula $\phi \in \Form^+(Q,\vec{x})$. If $\phi$ does not
-contain boolean variables, then $\overline{\phi}$ and $\neg\phi$ are
-equivalent, and in general:
-
-\begin{proposition}\label{prop:overline}
-  Given a formula $\phi \in \Form^+(Q,\vec{x})$ and a valuation $\nu$
-  mapping each $q \in Q$ to a value $\nu(q) \in \booli$ and each $x \in
-  \vec{x}$ to a value $\nu(x) \in \Data^\I$, let $\nu'$ be the
-  valuation that assigns each $q \in Q$ the value $\neg\nu(q)$ and
-  each $x \in \vec{x}$ the value $\nu(x)$. Then we have 
-  $\I,\nu \models \phi$ if and only if $\I,\nu' \not\models \overline{\phi}$. 
-\end{proposition}
-\proof{Immediate, by induction on the structure of $\phi$. \qed}
+every formula $\phi \in \Form^+(Q,\vec{x})$.
 
@@
 \end{lemma}
-\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
-  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
-  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
-  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
-  \iota_2,a_1\ldots a_n) \iffeq \Delta(\iota_1,a_1\ldots a_n) \vee
-  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
-  definition of the initial configuration of $\A_\cup$. For the
-  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
-  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
-  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
-    \iota_2,a_1\ldots a_{n-1})}$ with
-  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
-  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
-  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
-  \iota_2,a_1\ldots a_{n-1}) \iffeq \Delta(\iota_1,a_1\ldots a_{n-1})
-  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
-  \[\begin{array}{ll}
-  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
-  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
-  \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots a_{n}) \enspace.
-  \end{array}\] 
-  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
-  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
-  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
-  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
-  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
-  \subseteq Q$ and thus $\overline{\iota_1} \equiv \iota_1$. For the case
-  $n>0$, we compute:
-  \[\begin{array}{lcl}
-  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \iffeq & \text{ (by ind. hyp.)} \\
-  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \iffeq & \text{ (by the def. of $\overline{\phi}$)} \\
-  \overline{\Delta(\iota_1,a_1\ldots,a_n)}
-  \end{array}\]
-  Let $\nu,\nu' : (Q \cup \bigcup_{i=0}^n\vec{x}_i) \rightarrow (\booli
-  \cup \Data^\I)$ be valuations such that: \begin{compactitem}
-  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
-    %
-  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
-    %
-  \item $\nu(x) = \nu'(x)$, for each $x \in \vec{x}_0$, 
-    %
-  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
-    \in [1,n]$.
-  \end{compactitem}
-  By Proposition \ref{prop:overline}, we have $\I,\nu \models
-  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
-  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
-  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus for all $w \in
-  \Sigma[\vec{x}]^*$, we have $w \in L(\A_1) \iff w \not\in
-  L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) = \Sigma[\vec{x}]^*
-  \setminus L(\A_1)$. \qed}
 
@@
 \end{lemma}
-\proof{ Let $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ in the following.
-  ``$\Leftarrow$'' This direction is trivial. ``$\Rightarrow$'' We
-  define $\Inv : (Q \mapsto \booli) \rightarrow 2^{\vec{x} \mapsto
-    \Data^\I}$ as follows. For each $\beta : Q \rightarrow \booli$,
-  let $\Inv(\beta) = \{\nu : \vec{x} \rightarrow \Data^\I \mid \exists
-  u \in \Sigma^* ~.~ \beta\cup\nu \models
-  \Post{\A}(\iota,u)\}$. Checking that $\Inv$ is a safety invariant is
-  straightforward. \qed}
 
@@
 \end{array}\]
+
 \begin{lemma}\label{lemma:abstract-invariant}
@@
 \end{lemma}
-\proof{ It is sufficient to show that there exists $k\geq0$ such that for
-  all $u \in \Sigma^*$ there exists $i \in [0,k]$ such that
-  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,\mu(i))$. We have
-  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)$ for all $u \in
-  \Sigma^*$. But since $\preds$ is a finite set, also the set
-  $\{\AbsPost{\A}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
-  there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
-  exists $i \in [0,k]$ such that $\AbsPost{\A}(\iota,u) \iffeq
-  \AbsPost{\A}(\iota,\mu(i))$, which concludes the proof. \qed}
 
@@
 
-\begin{proposition}\label{prop:asterisk}
-  Given a formula $\phi \in \Form^+(Q,\vec{x})$ and $a \in \Sigma$, we
-  have \(\Delta(\phi,a) \iffeq \exists Q' ~.~ \phi[Q'/Q] \wedge
-  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))\).
-\end{proposition}
-\proof{``$\Rightarrow$'' If $\I,\beta\cup\overline\nu\cup\nu \models
-  \Delta(\phi,a)$, for some valuations $\beta : Q \rightarrow \booli$
-  and $\overline\nu : \overline{\vec{x}} \rightarrow \Data^\I$, $\nu :
-  \vec{x} \rightarrow \Data^\I$, then we build a valuation $\beta' :
-  Q' \rightarrow \booli$ such that
-  $\I,\beta'\cup\beta\cup\overline\nu\cup\nu \models \phi[Q'/Q] \wedge
-  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))$. For each
-  occurrence of a formula $\Delta(q,a)$ in $\Delta(\phi,a)$ we set
-  $\beta'(q') = \true$ if $\I,\beta\cup\overline\nu\cup\nu \models
-  \Delta(q,a)$ and $\beta'(q') = \false$, otherwise. Since there are
-  no negated occurrences of such subformulae, the definition of
-  $\beta'$ is consistent, and the check
-  $\I,\beta'\cup\beta\cup\overline\nu\cup\nu \models \phi[Q'/Q] \wedge
-  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))$ is
-  immediate. $''\Leftarrow''$ This direction is an easy check. \qed}
-
 Moreover, we require that an interpolant
@@
 \end{proposition}
-\proof{ Let $\tuple{\top,I_0,\ldots,I_{n},\bot}$ be an interpolant,
-  and $i \in [0,n]$ be the first index for which $I_i \not\in
-  \Form^+(Q,\vec{x})$. By Proposition \ref{prop:asterisk}, we have
-  that $\Delta^{i-1}(I_{i-1},a_i) \iffeq \exists Q_{i-1} ~.~
-  I_{i-1}[Q_{i-1}/Q] \wedge \theta_i$. Since $I_{i-1} \in
-  \Form^+(Q,\vec{x})$, by the choice of $i$, we have that
-  $\Delta^{i-1}(I_{i-1},a_i) \in \Form^+(Q,\vec{x})$ as well. We
-  replace $I_i$ with $\Delta^{i-1}(I_{i-1},a_i)$. The result is still
-  an interpolant, because: \begin{compactitem}
-  \item $I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \models
-    \Delta^{i-1}(I_{i-1},a_i)$, by Proposition \ref{prop:asterisk}, and
-    %
-  \item $\Delta^{i-1}(I_{i-1},a_i) \wedge \theta_{i+1} \models \exists
-    Q_{i-1} ~.~ I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \wedge \theta_{i+1}
-    \models I_i[Q_i/Q] \wedge \theta_{i+1} \models I_{i+1}[Q_{i+1}/Q]$.
-  \end{compactitem}
-  We proceed with these replacements until there are no more formulae
-  $I_i \not\in \Form^+(Q,\vec{x})$ left. \qed}
 
@@
 \end{lemma}
-\proof{ Let $\Theta(u) \equiv \theta_0(Q_0) \wedge \theta_1(Q_0 \cup
-  Q_1,\vec{x}_0 \cup \vec{x}_1) \wedge \ldots \wedge \theta_n(Q_{n-1}
-  \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n)$ in
-  the following.
-
-  (\ref{it1:progress}) We apply Proposition \ref{prop:asterisk}
-  recursively and get: \[\Post{\A}(\iota,u)[Q_n/Q,\vec{x}_n/\vec{x}]
-  \iff \exists Q_0 \ldots \exists Q_{n-1} \exists \vec{x}_0 \ldots
-  \exists \vec{x}_{n-1} ~.~ \bigwedge_{i=0}^n \theta_i\] Assuming that
-  $\Theta(u)$ is satisfiable, we obtain a model for $\Accept{\A}(u)
-  \equiv \Post{\A}(\iota,u) \wedge \theta_{n+1}[Q/Q_n]$.
-
-  (\ref{it2:progress}) If $\tuple{\top,I_0,\dots,I_n,\bot}$ is an
-  interpolant for $\pi(u)$, the following entailments
-  hold: \begin{compactitem}
-  \item $\theta_0 \models I_0[Q_0/Q]$,
-    %
-  \item $I_{k-1}[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k}[Q_k/Q,\vec{x}_k/\vec{x}]$, $\forall k\in[1,n]$. 
-    %
-  \item $I_{n}[Q_{n}/Q] \wedge \theta_{n+1} \models \bot$. 
-  \end{compactitem}
-  We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n}$ by
-  induction on $n \geq 0$. This is sufficient to conclude because
-  $\AbsAccept{\A}(a_1\ldots a_n) \equiv \AbsPost{\A}(\iota,a_1\ldots
-  a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n} \wedge
-  \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we have
-  $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv \theta_0[Q/Q_0]
-  \models I_0$. For the induction step $n>0$, we compute:
-  \[\begin{array}{l}
-  \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] \equiv \text{ (by def. of $\AbsPost{\A}$)} \\
-  \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] \models \text{ (by Prop. \ref{prop:asterisk})} \\
-  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n \models \text{ (ind. hyp.) } \\
-  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q] \hfill\text{\qed}
-  \end{array}\]}
+
 We have now all the ingredients to describe the first emptiness
@@
 \end{theorem}
-\proof{We prove the following invariant: each time Algorithm
-  \ref{alg:predabs} reaches line \ref{ln:while}, the set $W$ of nodes
-  in $\worklist$ contains all the frontier nodes in the ART
-  $\tuple{N\cup W,E,\rootNode,\Lambda,\lhd}$ which are not covered by
-  some node in $N$, namely that:
-  \begin{equation}\label{inv:predabs}
-  W = \set{n \not\in N \mid \forall m \in N~ \forall a \in \Sigma ~.~ (n,a,m)
-    \not\in E ~\wedge~ (n,m) \not\in \lhd}
-  \end{equation}
-  Initially, this is the case because $W = \set{\rootNode}$ and $N = E
-  = \lhd = \emptyset$. If the invariant holds previously, at line
-  \ref{ln:while}, it will hold again after line \ref{ln:refine} is
-  executed, because, when the subtree rooted at the pivot $p$ is
-  removed, $p$ becomes a member of the set of uncovered frontier
-  nodes, and is added to $W$ at line \ref{ln:refine}. The last case
-  considers that the invariant holds at line \ref{ln:while} and the
-  control follows the else branch at line \ref{ln:else}. In this case,
-  the newly created frontier node $s$ is added to $W$ only if it is
-  not covered by an existing node in $N$ (line \ref{ln:covered}).
-
-  Next we prove that, if Algorithm \ref{alg:predabs} returns $\true$,
-  then $\bigvee_{n \in N} \Lambda(n)$ defines a safety
-  invariant. Suppose that Algorithm \ref{alg:predabs} returns at line
-  \ref{ln:true}. Then it must be that $W = \emptyset$. Because
-  (\ref{inv:predabs}) is invariant, each node in $N$ is either covered
-  by another node in $N$, or all its successors are in $N$. We prove
-  first that $\bigvee_{n \in N} \Lambda(n)$ is an invariant: for any
-  $u \in \Sigma^*$, there exists some node $n \in N$ such that
-  $\Post{\A}(\iota,u) \models \Lambda(n)$. Let $u \in \Sigma^*$ be an
-  arbitrary sequence. If $u$ labels the path from $\rootNode$ to some
-  $n \in N$, we have $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)
-  \models \Lambda(n)$ and we are done. Otherwise, let $v$ be the
-  (possibly empty) prefix of $u$ which labels the path from
-  $\rootNode$ to some $n \in N$, which is covered by another $m \in
-  N$, where $(n,a,m) \in E$, that is $u = vav'$, for some $a \in
-  \Sigma$ and $v'\in\Sigma^*$. Moreover, we have $\Post{\A}(\iota,va)
-  \models \AbsPost{\A}(\iota,va) \models \Lambda(m)$, by the
-  construction of the set $\lhd$ of covering edges --- lines
-  \ref{ln:direct-cover}, \ref{ln:child-cover},
-  \ref{ln:indirect-cover}. Continuing this argument recursively from
-  $m$, since $\len{v'} < \len{u}$, we shall eventually discover a node
-  $p$ such that $\Post{\A}(\iota,u) \models \Lambda(p)$.
-
-  To prove that $\bigvee_{n \in N} \Lambda(n)$ is, moreover, a safety
-  invariant, suppose, by contradiction, that there exists $u \in
-  \Sigma^*$ such that $\Accept{\A}(u)$ is satisfiable. By the previous
-  point, there exists a node $p \in N$ such that $\Post{\A}(\iota,u)
-  \models \Lambda(p)$. But then we have $\Accept{\A}(\iota,u) \models
-  \Accept{\A}(\iota,\lambda(p))$, thus $\Accept{\A}(\iota,\lambda(p))$
-  is satisfiable as well. However, this cannot be the case, because
-  $p$ has been processed at line \ref{ln:feasible} and Algorithm
-  \ref{alg:predabs} would have returned a counterexample,
-  contradicting the assumption that it returns $\true$. This concludes
-  the proof that $\bigvee_{n\in N} \Lambda(n)$ is a safety invariant,
-  thus $L(\A) = \emptyset$, by Lemma \ref{lemma:safety-invariant}.
-
-  We have then proved the second point of the statement. For the first
-  point, assume that $L(\A) \neq 0$ and let $w = (a_1,\nu_1) \ldots
-  (a_k,\nu_k) \in L(\A)$ be a word. By the above, Algorithm
-  \ref{alg:predabs} cannot return $\true$. Suppose, by contradiction
-  that it does not terminate. Since the sequences from $\Sigma^*$ are
-  explored in breadth-first order, every sequence of length $k$ is
-  eventually explored, which leads to the discovery of $w$ at line
-  \ref{ln:feasible}. Then Algorithm \ref{alg:predabs} terminates
-  returning $w \in L(\A)$. \qed}
 
@@
 \end{lemma}
+
+The correctness of Algorithm \ref{alg:impact} is proved in the
+following.
+
+\begin{theorem}\label{thm:impact}
+  Given an automaton $\A$, such that $L(\A) \neq \emptyset$, Algorithm
+  \ref{alg:impact} terminates and returns a word $w \in L(\A)$. If
+  Algorithm \ref{alg:impact} terminates reporting $\true$, then
+  $L(\A) = \emptyset$.
+\end{theorem}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\bibliographystyle{abbrv} \bibliography{refs}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\appendix
+
+\section{Proofs}
+
+\subsection{Proof of Lemma \ref{lemma:closure}}
+
+\begin{proposition}\label{prop:overline}
+  Given a formula $\phi \in \Form^+(Q,\vec{x})$ and a valuation $\nu$
+  mapping each $q \in Q$ to a value $\nu(q) \in \booli$ and each $x \in
+  \vec{x}$ to a value $\nu(x) \in \Data^\I$, let $\nu'$ be the
+  valuation that assigns each $q \in Q$ the value $\neg\nu(q)$ and
+  each $x \in \vec{x}$ the value $\nu(x)$. Then we have 
+  $\I,\nu \models \phi$ if and only if $\I,\nu' \not\models \overline{\phi}$. 
+\end{proposition}
+\proof{Immediate, by induction on the structure of $\phi$. \qed}
+
+\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
+  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
+  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
+  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
+  \iota_2,a_1\ldots a_n) \iffeq \Delta(\iota_1,a_1\ldots a_n) \vee
+  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
+  definition of the initial configuration of $\A_\cup$. For the
+  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
+  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
+  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
+    \iota_2,a_1\ldots a_{n-1})}$ with
+  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
+  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
+  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
+  \iota_2,a_1\ldots a_{n-1}) \iffeq \Delta(\iota_1,a_1\ldots a_{n-1})
+  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
+  \[\begin{array}{ll}
+  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
+  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
+  \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots a_{n}) \enspace.
+  \end{array}\] 
+  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
+  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
+  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
+  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
+  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
+  \subseteq Q$ and thus $\overline{\iota_1} \equiv \iota_1$. For the case
+  $n>0$, we compute:
+  \[\begin{array}{lcl}
+  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \iffeq & \text{ (by ind. hyp.)} \\
+  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \iffeq & \text{ (by the def. of $\overline{\phi}$)} \\
+  \overline{\Delta(\iota_1,a_1\ldots,a_n)}
+  \end{array}\]
+  Let $\nu,\nu' : (Q \cup \bigcup_{i=0}^n\vec{x}_i) \rightarrow (\booli
+  \cup \Data^\I)$ be valuations such that: \begin{compactitem}
+  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
+    %
+  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
+    %
+  \item $\nu(x) = \nu'(x)$, for each $x \in \vec{x}_0$, 
+    %
+  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
+    \in [1,n]$.
+  \end{compactitem}
+  By Proposition \ref{prop:overline}, we have $\I,\nu \models
+  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
+  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
+  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus for all $w \in
+  \Sigma[\vec{x}]^*$, we have $w \in L(\A_1) \iff w \not\in
+  L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) = \Sigma[\vec{x}]^*
+  \setminus L(\A_1)$. \qed}
+
+\subsection{Proof of Lemma \ref{lemma:safety-invariant}}
+
+\proof{ Let $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ in the following.
+  ``$\Leftarrow$'' This direction is trivial. ``$\Rightarrow$'' We
+  define $\Inv : (Q \mapsto \booli) \rightarrow 2^{\vec{x} \mapsto
+    \Data^\I}$ as follows. For each $\beta : Q \rightarrow \booli$,
+  let $\Inv(\beta) = \{\nu : \vec{x} \rightarrow \Data^\I \mid \exists
+  u \in \Sigma^* ~.~ \beta\cup\nu \models
+  \Post{\A}(\iota,u)\}$. Checking that $\Inv$ is a safety invariant is
+  straightforward. \qed}
+
+\subsection{Proof of Lemma \ref{lemma:abstract-invariant}}
+
+\proof{ It is sufficient to show that there exists $k\geq0$ such that for
+  all $u \in \Sigma^*$ there exists $i \in [0,k]$ such that
+  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,\mu(i))$. We have
+  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)$ for all $u \in
+  \Sigma^*$. But since $\preds$ is a finite set, also the set
+  $\{\AbsPost{\A}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
+  there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
+  exists $i \in [0,k]$ such that $\AbsPost{\A}(\iota,u) \iffeq
+  \AbsPost{\A}(\iota,\mu(i))$, which concludes the proof. \qed}
+
+\subsection{Proof of Proposition \ref{prop:positive-interpol}}
+
+\proof{ Let $\tuple{\top,I_0,\ldots,I_{n},\bot}$ be an interpolant,
+  and $i \in [0,n]$ be the first index for which $I_i \not\in
+  \Form^+(Q,\vec{x})$. By Proposition \ref{prop:asterisk}, we have
+  that $\Delta^{i-1}(I_{i-1},a_i) \iffeq \exists Q_{i-1} ~.~
+  I_{i-1}[Q_{i-1}/Q] \wedge \theta_i$. Since $I_{i-1} \in
+  \Form^+(Q,\vec{x})$, by the choice of $i$, we have that
+  $\Delta^{i-1}(I_{i-1},a_i) \in \Form^+(Q,\vec{x})$ as well. We
+  replace $I_i$ with $\Delta^{i-1}(I_{i-1},a_i)$. The result is still
+  an interpolant, because: \begin{compactitem}
+  \item $I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \models
+    \Delta^{i-1}(I_{i-1},a_i)$, by Proposition \ref{prop:asterisk}, and
+    %
+  \item $\Delta^{i-1}(I_{i-1},a_i) \wedge \theta_{i+1} \models \exists
+    Q_{i-1} ~.~ I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \wedge \theta_{i+1}
+    \models I_i[Q_i/Q] \wedge \theta_{i+1} \models I_{i+1}[Q_{i+1}/Q]$.
+  \end{compactitem}
+  We proceed with these replacements until there are no more formulae
+  $I_i \not\in \Form^+(Q,\vec{x})$ left. \qed}
+
+\subsection{Proof of Lemma \ref{lemma:progress}}
+
+\begin{proposition}\label{prop:asterisk}
+  Given a formula $\phi \in \Form^+(Q,\vec{x})$ and $a \in \Sigma$, we
+  have \(\Delta(\phi,a) \iffeq \exists Q' ~.~ \phi[Q'/Q] \wedge
+  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))\).
+\end{proposition}
+\proof{``$\Rightarrow$'' If $\I,\beta\cup\overline\nu\cup\nu \models
+  \Delta(\phi,a)$, for some valuations $\beta : Q \rightarrow \booli$
+  and $\overline\nu : \overline{\vec{x}} \rightarrow \Data^\I$, $\nu :
+  \vec{x} \rightarrow \Data^\I$, then we build a valuation $\beta' :
+  Q' \rightarrow \booli$ such that
+  $\I,\beta'\cup\beta\cup\overline\nu\cup\nu \models \phi[Q'/Q] \wedge
+  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))$. For each
+  occurrence of a formula $\Delta(q,a)$ in $\Delta(\phi,a)$ we set
+  $\beta'(q') = \true$ if $\I,\beta\cup\overline\nu\cup\nu \models
+  \Delta(q,a)$ and $\beta'(q') = \false$, otherwise. Since there are
+  no negated occurrences of such subformulae, the definition of
+  $\beta'$ is consistent, and the check
+  $\I,\beta'\cup\beta\cup\overline\nu\cup\nu \models \phi[Q'/Q] \wedge
+  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))$ is
+  immediate. $''\Leftarrow''$ This direction is an easy check. \qed}
+
+\proof{ Let $\Theta(u) \equiv \theta_0(Q_0) \wedge \theta_1(Q_0 \cup
+  Q_1,\vec{x}_0 \cup \vec{x}_1) \wedge \ldots \wedge \theta_n(Q_{n-1}
+  \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n)$ in
+  the following.
+
+  (\ref{it1:progress}) We apply Proposition \ref{prop:asterisk}
+  recursively and get: \[\Post{\A}(\iota,u)[Q_n/Q,\vec{x}_n/\vec{x}]
+  \iff \exists Q_0 \ldots \exists Q_{n-1} \exists \vec{x}_0 \ldots
+  \exists \vec{x}_{n-1} ~.~ \bigwedge_{i=0}^n \theta_i\] Assuming that
+  $\Theta(u)$ is satisfiable, we obtain a model for $\Accept{\A}(u)
+  \equiv \Post{\A}(\iota,u) \wedge \theta_{n+1}[Q/Q_n]$.
+
+  (\ref{it2:progress}) If $\tuple{\top,I_0,\dots,I_n,\bot}$ is an
+  interpolant for $\pi(u)$, the following entailments
+  hold: \begin{compactitem}
+  \item $\theta_0 \models I_0[Q_0/Q]$,
+    %
+  \item $I_{k-1}[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k}[Q_k/Q,\vec{x}_k/\vec{x}]$, $\forall k\in[1,n]$. 
+    %
+  \item $I_{n}[Q_{n}/Q] \wedge \theta_{n+1} \models \bot$. 
+  \end{compactitem}
+  We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n}$ by
+  induction on $n \geq 0$. This is sufficient to conclude because
+  $\AbsAccept{\A}(a_1\ldots a_n) \equiv \AbsPost{\A}(\iota,a_1\ldots
+  a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n} \wedge
+  \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we have
+  $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv \theta_0[Q/Q_0]
+  \models I_0$. For the induction step $n>0$, we compute:
+  \[\begin{array}{l}
+  \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] \equiv \text{ (by def. of $\AbsPost{\A}$)} \\
+  \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] \models \text{ (by Prop. \ref{prop:asterisk})} \\
+  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n \models \text{ (ind. hyp.) } \\
+  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q]
+  \end{array}\]
+\qed}
+
+\subsection{Proof of Theorem \ref{thm:predabs}}
+
+\proof{We prove the following invariant: each time Algorithm
+  \ref{alg:predabs} reaches line \ref{ln:while}, the set $W$ of nodes
+  in $\worklist$ contains all the frontier nodes in the ART
+  $\tuple{N\cup W,E,\rootNode,\Lambda,\lhd}$ which are not covered by
+  some node in $N$, namely that:
+  \begin{equation}\label{inv:predabs}
+  W = \set{n \not\in N \mid \forall m \in N~ \forall a \in \Sigma ~.~ (n,a,m)
+    \not\in E ~\wedge~ (n,m) \not\in \lhd}
+  \end{equation}
+  Initially, this is the case because $W = \set{\rootNode}$ and $N = E
+  = \lhd = \emptyset$. If the invariant holds previously, at line
+  \ref{ln:while}, it will hold again after line \ref{ln:refine} is
+  executed, because, when the subtree rooted at the pivot $p$ is
+  removed, $p$ becomes a member of the set of uncovered frontier
+  nodes, and is added to $W$ at line \ref{ln:refine}. The last case
+  considers that the invariant holds at line \ref{ln:while} and the
+  control follows the else branch at line \ref{ln:else}. In this case,
+  the newly created frontier node $s$ is added to $W$ only if it is
+  not covered by an existing node in $N$ (line \ref{ln:covered}).
+
+  Next we prove that, if Algorithm \ref{alg:predabs} returns $\true$,
+  then $\bigvee_{n \in N} \Lambda(n)$ defines a safety
+  invariant. Suppose that Algorithm \ref{alg:predabs} returns at line
+  \ref{ln:true}. Then it must be that $W = \emptyset$. Because
+  (\ref{inv:predabs}) is invariant, each node in $N$ is either covered
+  by another node in $N$, or all its successors are in $N$. We prove
+  first that $\bigvee_{n \in N} \Lambda(n)$ is an invariant: for any
+  $u \in \Sigma^*$, there exists some node $n \in N$ such that
+  $\Post{\A}(\iota,u) \models \Lambda(n)$. Let $u \in \Sigma^*$ be an
+  arbitrary sequence. If $u$ labels the path from $\rootNode$ to some
+  $n \in N$, we have $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)
+  \models \Lambda(n)$ and we are done. Otherwise, let $v$ be the
+  (possibly empty) prefix of $u$ which labels the path from
+  $\rootNode$ to some $n \in N$, which is covered by another $m \in
+  N$, where $(n,a,m) \in E$, that is $u = vav'$, for some $a \in
+  \Sigma$ and $v'\in\Sigma^*$. Moreover, we have $\Post{\A}(\iota,va)
+  \models \AbsPost{\A}(\iota,va) \models \Lambda(m)$, by the
+  construction of the set $\lhd$ of covering edges --- lines
+  \ref{ln:direct-cover}, \ref{ln:child-cover},
+  \ref{ln:indirect-cover}. Continuing this argument recursively from
+  $m$, since $\len{v'} < \len{u}$, we shall eventually discover a node
+  $p$ such that $\Post{\A}(\iota,u) \models \Lambda(p)$.
+
+  To prove that $\bigvee_{n \in N} \Lambda(n)$ is, moreover, a safety
+  invariant, suppose, by contradiction, that there exists $u \in
+  \Sigma^*$ such that $\Accept{\A}(u)$ is satisfiable. By the previous
+  point, there exists a node $p \in N$ such that $\Post{\A}(\iota,u)
+  \models \Lambda(p)$. But then we have $\Accept{\A}(\iota,u) \models
+  \Accept{\A}(\iota,\lambda(p))$, thus $\Accept{\A}(\iota,\lambda(p))$
+  is satisfiable as well. However, this cannot be the case, because
+  $p$ has been processed at line \ref{ln:feasible} and Algorithm
+  \ref{alg:predabs} would have returned a counterexample,
+  contradicting the assumption that it returns $\true$. This concludes
+  the proof that $\bigvee_{n\in N} \Lambda(n)$ is a safety invariant,
+  thus $L(\A) = \emptyset$, by Lemma \ref{lemma:safety-invariant}.
+
+  We have then proved the second point of the statement. For the first
+  point, assume that $L(\A) \neq 0$ and let $w = (a_1,\nu_1) \ldots
+  (a_k,\nu_k) \in L(\A)$ be a word. By the above, Algorithm
+  \ref{alg:predabs} cannot return $\true$. Suppose, by contradiction
+  that it does not terminate. Since the sequences from $\Sigma^*$ are
+  explored in breadth-first order, every sequence of length $k$ is
+  eventually explored, which leads to the discovery of $w$ at line
+  \ref{ln:feasible}. Then Algorithm \ref{alg:predabs} terminates
+  returning $w \in L(\A)$. \qed}
+
+\subsection{Proof of Lemma \ref{lemma:well-labeled}}
+
 \proof{ We distinguish two cases. First, if $(n,a,m)$ occurs on a path
@@
 
-The correctness of Algorithm \ref{alg:impact} is proved in the
-following.
-
-\begin{theorem}\label{thm:impact}
-  Given an automaton $\A$, such that $L(\A) \neq \emptyset$, Algorithm
-  \ref{alg:impact} terminates and returns a word $w \in L(\A)$. If
-  Algorithm \ref{alg:impact} terminates reporting $\true$, then
-  $L(\A) = \emptyset$.
-\end{theorem}
+\subsection{Proof of Theorem \ref{thm:impact}}
+
 \proof{ We prove first that, each time Algorithm \ref{alg:impact}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{abbrv} \bibliography{refs}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \end{document}

### 2017-05-14 draft.tex
ok

@@
 \begin{tabular}{cc}
-\includegraphics[scale=0.5]{figures/PA1.pdf} & \includegraphics[scale=0.5]{figures/PA2.pdf} \\
+\includegraphics[scale=0.52]{figures/PA1.pdf} & \includegraphics[scale=0.52]{figures/PA2.pdf} \\
 \footnotesize{(a)} & \footnotesize{(b)} \\
 \\
-\includegraphics[scale=0.5]{figures/PA3.pdf} & \includegraphics[scale=0.5]{figures/PA4.pdf} \\
+\includegraphics[scale=0.52]{figures/PA3.pdf} & \includegraphics[scale=0.52]{figures/PA4.pdf} \\
 \footnotesize{(c)} & \footnotesize{(d)}

### 2017-05-14 draft.tex
ok

@@
 \cite{mcmillan06}.  These algorithms are proven to terminate and
-return a word from the language, if the language is not empty, but
+return a word from the language of the automaton, if one exists, but
 termination is not guaranteed when the language is empty.
@@
 We have implemented the boolean operations and emptiness checking
-semi-algorithms in a prototype tool and carried out experiments with
-examples taken from array logics \cite{cav09}, timed automata
+semi-algorithms and carried out experiments with examples taken from
+array logics \cite{cav09}, timed automata
 \cite{henzinger:RealTimeSystems}, communication protocols \cite{abp}
@@
 here is caused by the use of hidden variables, which we do not
-allow. Even though closed under complement, emptiness is undecidable
-in our case, mainly because counting (incrementing and comparing to a
-constant) data values is allowed, in many data theories.
-
-Other approaches to defining notions of data automata
-\cite{Bojanczyk:2011:TLD,BouyerPetitTherien03} are different from ours
-mainly because these automata are controlled by equivalence relations
-of finite index, whereas in our case, the transitions are defined by
-unrestricted formulae in the first-order theory of the data domain. As
-in \cite{Kaminski94,AlurDill94}, the emptiness problem is decidable
-for these classes.
-
-Another related model is that of predicate automata
-\cite{Farzan15,Farzan16}, which recognize languages over integer data
-by labeling the words with conjunctions of uninterpreted
-predicates. We intend to explore further the connection with our model
-of alternating data automata, in order to apply our method to the
-verification of parallel programs.
+allow. Emptiness is however undecidable in our case, mainly because
+counting (incrementing and comparing to a constant) data values is
+allowed, in many data theories.
+
+%% Other approaches to defining classes of data automata
+%% \cite{Bojanczyk:2011:TLD,BouyerPetitTherien03} are different from ours
+%% mainly because these automata are controlled by equivalence relations
+%% of finite index, whereas in our case, the transitions are defined by
+%% unrestricted formulae in the first-order theory of the data domain. As
+%% in \cite{Kaminski94,AlurDill94}, the emptiness problem is decidable
+%% for these classes.
+
+Another related model is that of predicate automata \cite{Farzan15},
+which recognize languages over integer data by labeling the words with
+conjunctions of uninterpreted predicates. We intend to explore further
+the connection with our model of alternating data automata, in order
+to apply our method to the verification of parallel programs.
 
@@
 
-The emptiness check is implemented by adapting similar semi-algorithms
-for nondeterministic infinite-state programs to the alternating model
-of computation. In particular, we considered the state-of-the-art
-\impact~ procedure \cite{mcmillan06} that is shown to outperform lazy
-predicate abstraction in the nondeterministic case, and generalized it
-to cope with alternation. More recent approaches for interpolant-based
-abstraction refinement \cite{McMillan14,Hoder12} target Horn systems,
-used to encode recursive and concurrent programs
+The emptiness check for alternating data automata adapts similar
+semi-algorithms for nondeterministic infinite-state programs to the
+alternating model of computation. In particular, we considered the
+state-of-the-art \impact~ procedure \cite{mcmillan06} that is shown to
+outperform lazy predicate abstraction \cite{blast} in the
+nondeterministic case, and generalized it to cope with
+alternation. More recent approaches for interpolant-based abstraction
+refinement target Horn systems \cite{McMillan14,Hoder12}, used to
+encode recursive and concurrent programs
 \cite{Grebenshchikov12}. However, the emptiness of alternating word
@@
 the branches of the computation synchronize on the same input, which
-does not seem to be easily encoded using a finite number of local
-(equality) constraints. We believe that the lazy annotation techniques
-for Horn clauses are suited for branching computations, which we
-intend to consider in a future tree automata setting.
+cannot be encoded by a finite number of local (equality)
+constraints. We believe that the lazy annotation techniques for Horn
+clauses are suited for branching computations, which we intend to
+consider in a future tree automata setting.
 
@@
 We consider an infinite countable set of \emph{variables} $\vars$,
-where each $x \in \vars$ has an associated sort $\sigma(x)$. Terms are
-defined recursively: a term $t$ of sort $\sigma(t)=S$ is a variable $x
-\in \vars$ where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
+where each $x \in \vars$ has an associated sort $\sigma(x)$. A
+\emph{term} $t$ of sort $\sigma(t)=S$ is a variable $x \in \vars$
+where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
 $t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
 \emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
-between two terms of sort $\Data$. A \emph{literal} is an atom or its
-negation. A \emph{formula} is an existentially quantified combination
-of atoms using disjunction $\vee$, conjunction $\wedge$ and negation
-$\neg$ and we write $\phi \rightarrow \psi$ for $\neg\phi \vee \psi$.
-
-We denote by $\fv{\sigma}{\phi}$ the set of free (not under the scope
-of an existential quantifier) variables of sort $\sigma$ in $\phi$ and
-write $\fv{}{\phi}$ for $\bigcup_{\sigma \in \ssorts{\sig}}
-\fv{\sigma}{\phi}$. For a variable $x \in \fv{}{\phi}$ and a term $t$
-such that $\sigma(t) = \sigma(x)$, let $\phi[t/x]$ be the result of
-replacing each occurrence of $x$ by $t$. For indexed sets
-$\vec{t}=\set{t_1,\ldots,t_n}$ and $\vec{x}=\set{x_1,\ldots,x_n}$, we
-write $\phi[\vec{t}/\vec{x}]$ for the formula obtained by
-simultaneously replacing $x_i$ with $t_i$ in $\phi$, for all
-$i\in[1,n]$. The size $\len{\phi}$ is the number of symbols occuring
-in $\phi$.
+between two terms of sort $\Data$. A \emph{formula} is an
+existentially quantified combination of atoms using disjunction
+$\vee$, conjunction $\wedge$ and negation $\neg$ and we write $\phi
+\rightarrow \psi$ for $\neg\phi \vee \psi$.
+
+We denote by $\fv{\sigma}{\phi}$ the set of free variables of sort
+$\sigma$ in $\phi$ and write $\fv{}{\phi}$ for $\bigcup_{\sigma \in
+  \ssorts{\sig}} \fv{\sigma}{\phi}$. For a variable $x \in
+\fv{}{\phi}$ and a term $t$ such that $\sigma(t) = \sigma(x)$, let
+$\phi[t/x]$ be the result of replacing each occurrence of $x$ by
+$t$. For indexed sets $\vec{t}=\set{t_1,\ldots,t_n}$ and
+$\vec{x}=\set{x_1,\ldots,x_n}$, we write $\phi[\vec{t}/\vec{x}]$ for
+the formula obtained by simultaneously replacing $x_i$ with $t_i$ in
+$\phi$, for all $i\in[1,n]$. The size $\len{\phi}$ is the number of
+symbols occuring in $\phi$.
 
@@
 $t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.
+
 A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
@@
   \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
-    function}, where $\overline{\vec{x}}$ is the set $\{\overline{x}
-  \mid x \in \vec{x}\}$.
+    function},
 \end{compactitem}
-Intuitively, in each formula $\Delta(q,a)$ describing a transition
-rule, the variables $\overline{\vec{x}}$ track the previous and
-$\vec{x}$ the current values of the variables of $\A$. Observe
-further, that the initial values of the variables are left
-unconstrained, as the initial configuration does not contain free
-data variables. The size of $\A$ is defined as $\len{\A} = \len{\iota}
-+ \sum_{(q,a) \in Q\times\Sigma}\len{\Delta(q,a)}$.
+where $\overline{\vec{x}}=\{\overline{x} \mid x \in \vec{x}\}$. In
+each formula $\Delta(q,a)$ describing a transition rule, the variables
+$\overline{\vec{x}}$ track the previous and $\vec{x}$ the current
+values of the variables of $\A$. Observe that the initial values of
+the variables are left unconstrained, as the initial configuration
+does not contain free data variables. The size of $\A$ is defined as
+$\len{\A} = \len{\iota} + \sum_{(q,a) \in
+  Q\times\Sigma}\len{\Delta(q,a)}$.
 
@@
 \paragraph{\em Example}
-Figure \ref{fig:ex} depicts an alternating data automaton, with input
-alphabet $\Sigma = \set{a,b}$, variables $\vec{x} = \set{x,y}$, states
-$Q = \set{q_0,q_1,q_2,q_3,q_4}$, initial configuration $q_0$, final
-states $F=\set{q_3,q_4}$ and transitions:
+Figure \ref{fig:ex} depicts an ADA with input alphabet $\Sigma =
+\set{a,b}$, variables $\vec{x} = \set{x,y}$, states $Q =
+\set{q_0,q_1,q_2,q_3,q_4}$, initial configuration $q_0$, final states
+$F=\set{q_3,q_4}$ and transitions:
 \[\begin{array}{rcl}
@@
 \end{array}\]
-The missing transition rules are assumed to be $\false$. Rules
-$\Delta(q_0,a)$ and $\Delta(q_1,a)$ are universal and there are no
-existential nondeterministic rules. Rules $\Delta(q_1,a)$ and
-$\Delta(q_2,b)$ compare past ($\overline{x},\overline{y}$) with present
-($x,y$) values, $\Delta(q_0,a)$ constrains the present and
+The missing rules, such as $\Delta(q_0,b)$, are assumed to be
+$\bot$. Rules $\Delta(q_0,a)$ and $\Delta(q_1,a)$ are universal and
+there are no existential nondeterministic rules. Rules $\Delta(q_1,a)$
+and $\Delta(q_2,a)$ compare past ($\overline{x},\overline{y}$) with
+present ($x,y$) values, $\Delta(q_0,a)$ constrains the present and
 $\Delta(q_1,b)$, $\Delta(q_2,b)$ the past values, respectively. \qed
@@
 a set of time-stamped variables. For an input event $a \in \Sigma$ and
-a formula $\phi$, we write $\Delta(\phi,a)$ (respectively,
+a formula $\phi$, we write $\Delta(\phi,a)$ (respectively
 $\Delta^k(\phi,a)$) for the formula obtained from $\phi$ by
@@
 (a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
-formulae $\phi_0,\phi_1,\ldots,\phi_n$, where $\phi_0 \equiv \iota$
-and, for all $k\in[1,n]$, we have $\phi_k \equiv
-\Delta^k(\phi_{k-1},a_k)$. Next, we slightly abuse notation and write
-$\Delta(\iota,a_1,\ldots,a_n)$ for the formula
+formulae: \[\phi_0(Q) \Rightarrow \phi_1(Q,\vec{x}_0 \cup \vec{x}_1)
+\Rightarrow \ldots \Rightarrow \phi_n(Q,\vec{x}_0 \cup \ldots
+\vec{x}_n)\] where $\phi_0 \equiv \iota$ and, for all $k\in[1,n]$, we
+have $\phi_k \equiv \Delta^k(\phi_{k-1},a_k)$. Next, we slightly abuse
+notation and write $\Delta(\iota,a_1,\ldots,a_n)$ for the formula
 $\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ above. We say that $\A$
@@
 \end{array}\]
-It is easy to check that $\len{\overline{\phi}} = \len{\phi}$, for
-every formula $\phi \in \Form^+(Q,\vec{x})$.
+We have $\len{\overline{\phi}} = \len{\phi}$, for every formula $\phi
+\in \Form^+(Q,\vec{x})$.
 
@@
 for the boolean operations, including complementation, can be built in
-linear time. This is not surprising, as linear time intersection and
-complementation are also possible for alternating automata over finite
-alphabets \cite{ChandraKozenStockmeyer81}.
+linear time. This matches the linear-time bounds for intersection and
+complementation of alternating automata over finite alphabets
+\cite{ChandraKozenStockmeyer81}.
 
@@
 We generalize the post-image function to finite sequences of input
-events, as follows, for any $u \in \Sigma^*$:
+events, as follows:
 \[\begin{array}{l}
 \Post{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm} \Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a) \\
-\Accept{\A}(u) \equiv \Post{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q \rightarrow \bot)
+\Accept{\A}(u) \equiv \Post{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q \rightarrow \bot)\text{, for any $u \in \Sigma^*$}
 \end{array}\]
@@
 $\preds$. The abstract versions of the post-image and acceptance
-condition are defined as follows, for any $u \in \Sigma^*$:
+condition are defined as follows:
 \[\begin{array}{l}
@@
 \AbsAccept{\A}(u) \equiv \AbsPost{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} 
-(q \rightarrow \bot)
+(q \rightarrow \bot)\text{, for any $u \in \Sigma^*$}
 \end{array}\]
@@
   $k>0$ such that $\bigvee_{i=0}^k \AbsPost{\A}(\iota,\mu(i))$ defines
-  an invariant for $\A$. 
+  an invariant $\abs{\Inv}$ for $\A$. 
 \end{lemma}
@@
 initial configuration of $\A$, with every $q \in \fv{\Bool}{\iota}$
-replaced by $q_0$. The definition of $\theta_k$, $k\in[1,n]$ uses
-\emph{replacement sets} $R_\ell \subseteq Q_\ell$, $\ell\in [0,n]$,
-which are defined inductively below: \begin{compactitem}
-\item $R_0 = \fv{\Bool}{\theta_0}$ and 
+replaced by $q_0$. The definition of $\theta_k$, for all $k\in[1,n]$,
+uses \emph{replacement sets} $R_\ell \subseteq Q_\ell$, $\ell\in
+[0,n]$, which are defined inductively below: \begin{compactitem}
+\item $R_0 = \fv{\Bool}{\theta_0}$,
 %
@@
   $\ell\in[1,n]$.
+%
+\item $\theta_{n+1}(Q_n) \equiv \bigwedge_{q \in Q \setminus F} (q_n
+  \rightarrow \bot)$.
 \end{compactitem}
-Finally, we set $\theta_{n+1}(Q_n) \equiv \bigwedge_{q \in Q \setminus
-  F} (q_n \rightarrow \bot)$. The intuition is that $\theta_0, \ldots,
-\theta_n$ encode the run of $\A$ over $u$ and $\theta_{n+1}$ is the
-acceptance condition, which forces the last remaining non-final states
-to be false.
+The intuition is that $R_0,\ldots,R_n$ are the sets of states replaced,
+$\theta_0, \ldots, \theta_n$ are the transition rules fired on the run
+of $\A$ over $u$ and $\theta_{n+1}$ is the acceptance condition, which
+forces the last remaining non-final states to be false.
 
@@
 \begin{proposition}\label{prop:positive-interpol}
-  For any $u = a_1 \ldots a_n \in \Sigma^*$, if $\pi(u)$ is
-  unsatisfiable then one can build an interpolant
-  $\tuple{\top,I_0,\ldots,I_n,\bot}$ for $\pi(u)$, such that $I_i \in
-  \Form^+(Q,\vec{x})$, for all $i \in [0,n]$.
+  If $\Theta(a_1 \ldots a_n)$ is unsatisfiable then one can build an
+  interpolant $\tuple{\top,I_0,\ldots,I_n,\bot}$ for $\Theta(a_1\ldots
+  a_n)$, such that $I_i \in \Form^+(Q,\vec{x})$, for all $i \in
+  [0,n]$.
 \end{proposition}
 
-We recall that a run of $\A$ over $u$ is a sequence $\phi_0(Q),
-\phi_1(Q,\vec{x}_0\cup\vec{x}_1), \ldots,
-\phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)$, where $\phi_0$ is the
-initial configuration $\iota$ and for each $k>0$, $\phi_k$ is obtained
-from $\phi_{k-1}$ by replacing each state $q \in
+We recall that a run of $\A$ over $u$ is a sequence: \[\phi_0(Q)
+\Rightarrow \phi_1(Q,\vec{x}_0\cup\vec{x}_1) \Rightarrow \ldots
+\Rightarrow \phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)\] where
+$\phi_0$ is the initial configuration $\iota$ and for each $k>0$,
+$\phi_k$ is obtained from $\phi_{k-1}$ by replacing each state $q \in
 \fv{\Bool}{\phi_{k-1}}$ by the formula
@@
 run, the formulae thus produced lose track of the control history and
-are not directly suitable to producing interpolants that relates
-states and data.
+are not directly suitable to producing interpolants that relate states
+and data.
 
@@
 setup is captured by the progress property below, which guarantees
-that adding the formulae of an interpolant for $\pi(u)$ to the set
+that adding the formulae of an interpolant for $\Theta(u)$ to the set
 $\preds$ of predicates suffices to exclude the spurious counterexample
@@
 We have now all the ingredients to describe the first emptiness
-semi-algorithm for alternating data
+checking semi-algorithm for alternating data
 automata. Algorithm\footnote{Though termination is not guaranteed, we
@@
 \emph{abstract reachability tree} (ART) whose nodes are labeled with
-formulae over-approximating the concrete sets of configurations, which
-uses a covering relation between nodes in order to ensure that the set
-of formulae labeling the nodes in the ART forms an antichain. Any
+formulae over-approximating the concrete sets of configurations, and a
+covering relation between nodes in order to ensure that the set of
+formulae labeling the nodes in the ART forms an antichain. Any
 spurious counterexample is eliminated by computing an interpolant and
@@
 %
-\item $R : N \rightarrow 2^Q$ is a labeling of nodes with
-\emph{replacement sets}, such that $R(\rootNode) = \fv{\Bool}{\iota}$,
+\item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
+  sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$,
 %
 \item $T : E \rightarrow \bigcup_{i=0}^\infty
-\Form(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is labeling of edges with
-time-stamped formulae, and 
+  \Form(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is a labeling of edges
+  with time-stamped formulae, and
 %
@@
 \end{compactitem}
-For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is $(n,a,m)$.
-The replacement sets and time-stamped edge labels are defined below,
-for each edge $e = (n,a,m) \in E$, if $n$ lies at depth\footnote{We
-  consider that the root node has zero depth.} $k$ in the tree
-$(N,E)$ then we have:
-\[\begin{array}{rcl}
-T(e) & = & \bigwedge_{q \in R(n)} (q_k \rightarrow \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]) \\
-R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)}
-\end{array}\]
+%% For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is $(n,a,m)$.
+%% The replacement sets and time-stamped edge labels are defined below,
+%% for each edge $e = (n,a,m) \in E$, if $n$ lies at depth\footnote{We
+%%   consider that the root node has zero depth.} $k$ in the tree
+%% $(N,E)$ then we have:
+%% \[\begin{array}{rcl}
+%% T(e) & = & \bigwedge_{q \in R(n)} (q_k \rightarrow \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]) \\
+%% R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)}
+%% \end{array}\]
+
 Each node $n \in N$ corresponds to a unique path from the root to $n$,
 labeled by a sequence $\lambda(n) \in \Sigma^*$ of input events. The
-least unfeasible suffix of $\lambda(n)$ is the smallest sequence $v =
+least infeasible suffix of $\lambda(n)$ is the smallest sequence $v =
 a_1 \ldots a_k$, such that $\lambda(n) = wv$, for some $w \in
@@
 $q_1$, which are added to $\preds$ and the $\top$ node is removed. The
-next time $a$ is fired, it creates a node labeled $q_1$ and the
-sequence $aa$ creates another node $q_1$, which is covered by the
+next time $a$ is fired, it creates a node labeled $q_1$ and the next
+sequence $aa$ creates a successor node $q_1$, which is covered by the
 first, depicted with a dashed arrow, in Figure \ref{fig:pa} (b). The
@@
 this check is $x\leq0\wedge q_2 \wedge y\geq0$ and the pivot is again
-the root. Then the entire ART is rebuild with the new predicates and
-the sequence $aab$ results in an uncovered node $\top$, in Figure
+the root. Then the ART is rebuilt with the new predicates and
+the sequence $aab$ yields an uncovered node $\top$, in Figure
 \ref{fig:pa} (c). The new pivot is the endpoint of $a$ and the new
@@
 the label of $m$. The correctness criteria of Algorithm
-\ref{alg:predabs} are showed in the following theorem:
+\ref{alg:predabs} are stated by the following theorem:
 
@@
 $n$ is \emph{closed} if it is covered, or $\Lambda(n) \not\models
-\Lambda(m)$ for all $m \in N$ such that $n \prec^* m$.
+\Lambda(m)$ for all $m \in N$ such that $m \prec^* n$.
 

### 2017-05-15 draft.tex
ok

@@
 complementation is possible in linear time, the bulk of the work is
-shipped to the decision procedure.
+carried out by the decision procedure.
 
@@
 \emph{theory} $\theory(\sig,\I)$ is the set of valid formulae written
-in the signature $\sig$, with the interpretation $\I$. A decision
-procedure for $\theory(\sig,\I)$ is an algorithm that takes a formula
+in the signature $\sig$, with the interpretation $\I$. A \emph{decision
+procedure} for $\theory(\sig,\I)$ is an algorithm that takes a formula
 $\phi$ in the signature $\sig$ and returns yes iff $\phi \in
@@
 %
-\item $Q$ is a finite set of variables of sort $\Bool$, called
-  \emph{states},
-%
-\item $\iota \in \Form^+(Q,\emptyset)$ is positive boolean combination
-  of states defining the \emph{initial configurations},
+\item $Q \subset \vars$ is a finite set of variables of sort $\Bool$
+  (\emph{states}),
+%
+\item $\iota \in \Form^+(Q,\emptyset)$ is the \emph{initial
+  configuration},
 %
@@
 \begin{figure}[htb]
+\vspace*{-\baselineskip}
 \centerline{\includegraphics[scale=0.8]{figures/Example.pdf}}
+\vspace*{-\baselineskip}
 \caption{Alternating Data Automaton Example}
@@
 \[\begin{array}{rcl}
-\Delta(q_0,a) & \equiv & q_1 \wedge q_2 \wedge x=0 \wedge y=0 \\
-\Delta(q_1,a) & \equiv & q_1 \wedge q_3 \wedge x=\overline{y}+1 \wedge y=\overline{x}+1 \\
+\Delta(q_0,a) & \equiv & q_1 \wedge q_2 \wedge x\teq0 \wedge y\teq0 \\
+\Delta(q_1,a) & \equiv & q_1 \wedge q_3 \wedge x\teq\overline{y}+1 \wedge y\teq\overline{x}+1 \\
 \Delta(q_1,b) & \equiv & q_3 \wedge \overline{x} \geq \overline{y} \\
@@
 
-Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, $k\geq0$, be
-a set of time-stamped variables. For an input event $a \in \Sigma$ and
-a formula $\phi$, we write $\Delta(\phi,a)$ (respectively
-$\Delta^k(\phi,a)$) for the formula obtained from $\phi$ by
-simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
+Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, for any
+$k\geq0$, be a set of time-stamped variables. For an input event $a
+\in \Sigma$ and a formula $\phi$, we write $\Delta(\phi,a)$
+(respectively $\Delta^k(\phi,a)$) for the formula obtained from $\phi$
+by simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
 formula $\Delta(q,a)$ (respectively
 $\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
-all $k\geq0$). Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
+$k\geq0$). Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
 (a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
 formulae: \[\phi_0(Q) \Rightarrow \phi_1(Q,\vec{x}_0 \cup \vec{x}_1)
-\Rightarrow \ldots \Rightarrow \phi_n(Q,\vec{x}_0 \cup \ldots
+\Rightarrow \ldots \Rightarrow \phi_n(Q,\vec{x}_0 \cup \ldots \cup
 \vec{x}_n)\] where $\phi_0 \equiv \iota$ and, for all $k\in[1,n]$, we
@@
 \emph{accepts} $w$ iff $\I,\nu \models \Delta(\iota,a_1,\ldots,a_n)$,
-where $\nu$ is some valuation that maps:\begin{inparaenum}[(1)]
+for some valuation $\nu$ that maps:\begin{inparaenum}[(1)]
 %
@@
 \ref{fig:ex} on the word
-$(a,\tuple{0,0}),(a,\tuple{1,1}),(b,\tuple{2,1})$: 
+$(a,\tuple{0,0}),(a,\tuple{1,1}),(b,\tuple{2,1})$, where $\Data^\I =
+\zed$ and the function symbols have standard arithmetic
+interpretation:
 \[\begin{array}{l}
 q_0 \stackrel{(a,\tuple{0,0})}{\Longrightarrow} 
-q_1 \wedge q_2 \wedge x_1=0 \wedge y_1=0 \stackrel{(a,\tuple{1,1})}{\Longrightarrow} \\ 
-(q_1 \wedge q_2 \wedge x_2=y_1+1 \wedge y_2=x_1+1) ~\wedge \\[-2mm]
-(q_2 \wedge x_2>x_1 \wedge y_2>y_1) \wedge x_1=0 \wedge y_1=0 \stackrel{(b,\tuple{2,1})}{\Longrightarrow} \\
-(q_3 \wedge x_2\geq y_2) \wedge (q_4 \wedge x_2>y_2) \wedge x_2=y_1+1 \wedge y_2=x_1+1 \\ 
-\wedge~ (q_4 \wedge x_2>y_2) \wedge x_2>x_1 \wedge y_2>y_1 \wedge x_1=0 \wedge y_1=0 
-\rightarrow \bot~ \hfill\text{\qed}
+q_1 \wedge q_2 \wedge x_1\teq0 \wedge y_1\teq0 \stackrel{(a,\tuple{1,1})}{\Longrightarrow} \\ 
+\underbrace{q_1 \wedge q_2 \wedge x_2\teq y_1+1 \wedge y_2\teq x_1+1}_{q_1} ~\wedge~
+\underbrace{q_2 \wedge x_2>x_1 \wedge y_2>y_1}_{q_2} \\ 
+\wedge~ x_1\teq0 \wedge y_1\teq0 \stackrel{(b,\tuple{2,1})}{\Longrightarrow} 
+\underbrace{q_3 \wedge x_2\geq y_2}_{q_1} \wedge \underbrace{q_4 \wedge x_2>y_2}_{q_2} \wedge x_2\teq y_1+1 \\ 
+\wedge~ y_2\teq x_1+1 \wedge \underbrace{q_4 \wedge x_2>y_2}_{q_2} \wedge x_2>x_1 \wedge y_2>y_1 \wedge x_1\teq0 \wedge y_1\teq0 ~\hfill\text{\qed}
 \end{array}\]
@@
 %
-\item \emph{emptiness}: given an automaton $\A$, is it the case that
-  $L(\A) = \emptyset$ ?
+\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
+  \emptyset$ ?
 \end{compactenum}
@@
 valuation into a set of data valuations, that contains the initial
-configuration $\iota = \Post{\A}(\iota,\varepsilon)$, whose data
+configuration $\iota \equiv \Post{\A}(\iota,\varepsilon)$, whose data
 variables are unconstrained, over-approximates the set of reachable
@@
 acceptance condition (point \ref{it2:inv}).  A formula $\phi(Q,\vec{x})$
-\emph{defines} $\Inv$ iff for all $\beta: Q \rightarrow \booli$ and
+is said to \emph{define} $\Inv$ iff for all $\beta: Q \rightarrow \booli$ and
 $\nu : \vec{x} \rightarrow \Data^\I$, we have $\I,\beta\cup\nu \models
@@
 expensive quantifier elimination, it is common belief that this should
-hold also for $\AbsPost{\A}$. However, the choice of conjoined
+hold also for $\AbsPost{\A}$. However, chosing the conjoined
 predicates that form $\AbsPost{\A}$ requires only a quantifier-free
-satisfiability query: $\Post{\A}(\phi,a) \models \pi \iff \exists
+satisfiability query: $\Post{\A}(\phi,a) \models \pi$ iff $\exists
 \overline{\vec{x}} ~.~ \Delta(\phi[\overline{\vec{x}}/\vec{x}],a)
-\models \pi \iff \Delta(\phi[\overline{\vec{x}}/\vec{x}],a) \wedge
-\neg\pi$ unsatisfiable. \qed
+\models \pi$ iff $\Delta(\phi[\overline{\vec{x}}/\vec{x}],a) \wedge
+\neg\pi$ is unsatisfiable. \qed
 
@@
 \paragraph{\em Remark} 
-We deviate from the standard notion of \emph{generalized interpolants}
-\cite{mcmillan06}, which requires $\fv{}{I_i} \subseteq \fv{}{\phi_i}
-\cap \fv{}{\phi_{i+1}}$, for each $i \in [1,n-1]$, and assume further
-that the variables have been already renamed by removing the
-time-stamps. \qed
+We deviate w.l.o.g. from the standard notion of \emph{generalized
+  interpolants} \cite{mcmillan06}, which requires $\fv{}{I_i}
+\subseteq \fv{}{\phi_i} \cap \fv{}{\phi_{i+1}}$, for each $i \in
+          [1,n-1]$, and assume further that the variables have been
+          renamed by removing the time stamps. \qed
 
@@
 \end{compactitem}
-The intuition is that $R_0,\ldots,R_n$ are the sets of states replaced,
-$\theta_0, \ldots, \theta_n$ are the transition rules fired on the run
-of $\A$ over $u$ and $\theta_{n+1}$ is the acceptance condition, which
-forces the last remaining non-final states to be false.
+The intuition is that $R_0,\ldots,R_n$ are the sets of states
+replaced, $\theta_0, \ldots, \theta_n$ are the sets of transition
+rules fired on the run of $\A$ over $u$ and $\theta_{n+1}$ is the
+acceptance condition, which forces the last remaining non-final states
+to be false.
 
@@
 states are replaced with transition formulae when moving one step in a
-run, the formulae thus produced lose track of the control history and
-are not directly suitable to producing interpolants that relate states
-and data.
+run, these formulae lose track of the control history and are not
+suitable for producing interpolants that relate states and data.
 
@@
 \begin{lemma}\label{lemma:progress}
-  For any non-empty sequence $u = a_1\ldots a_n \in \Sigma^*$
-  such that $\Accept{\A}(u)$ is unsatisfiable, the following
+  For any sequence $u = a_1\ldots a_n \in \Sigma^*$, if
+  $\Accept{\A}(u)$ is unsatisfiable, the following
   hold: \begin{compactenum}
@@
 \end{lemma}
+
+\section{Lazy Predicate Abstraction for ADA Emptiness}
 
@@
 spurious counterexample is eliminated by computing an interpolant and
-adding its formulae to the set of predicates.
+adding its formulae to the set of predicates (cf. Lemma \ref{lemma:progress}).
 
@@
 
-  \If{$\Lambda(n) \wedge \bigwedge_{q \in Q \setminus F} (q
-    \rightarrow \bot)$ is satisfiable} \Comment{counterexample candidate}
-
-  \State let $u = a_1\ldots a_k$ be the label of the path from
+  \State let $\lambda(n) = a_1\ldots a_k$ be the label of the path from
   $\rootNode$ to $n$
 
-  \If{$\Accept{\A}(u)$ is satisfiable} \Comment{counterexample is
-    feasible}
+  \If{$\AbsPost{\A}(\lambda(n))$ is satisfiable} \Comment{counterexample candidate}
+
+  \If{$\Accept{\A}(u)$ is satisfiable} \Comment{feasible counterexample}
   \label{ln:feasible}
 
-  \State get model $\beta : Q \rightarrow \booli$, $\{\nu_i :
-  \vec{x}_i \rightarrow \Data^\I \mid i=0\ldots k\}$ of $\Accept{\A}(u)$
+  \State get model $(\beta,\nu_1,\ldots,\nu_k)$ of
+  $\Accept{\A}(\lambda(n))$
 
@@
 
-  \State $p \leftarrow \Call{FindPivot}{u,\Art}$
+  \State $p \leftarrow \Call{FindPivot}{\lambda(n),\Art}$
   \label{ln:pivot}
 
-  \State let $v$ be the label of the path from $p$ to $n$ in $\Art$
+  \State $v \leftarrow \Call{LeastInfeasibleSuffix}{\lambda(n),\Art}$
 
@@
 \item $T : E \rightarrow \bigcup_{i=0}^\infty
-  \Form(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is a labeling of edges
+  \Form^+(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is a labeling of edges
   with time-stamped formulae, and
@@
 labeled by a sequence $\lambda(n) \in \Sigma^*$ of input events. The
-least infeasible suffix of $\lambda(n)$ is the smallest sequence $v =
-a_1 \ldots a_k$, such that $\lambda(n) = wv$, for some $w \in
-\Sigma^*$ and the formula (\ref{eq:pivot}) below is unsatisfiable. The
-\emph{pivot} of $n$ is the node $p$ corresponding to the start of $v$
-in $\lambda(n)$.
+\emph{least infeasible suffix} of $\lambda(n)$ is the smallest
+sequence $v = a_1 \ldots a_k$, such that $\lambda(n) = wv$, for some
+$w \in \Sigma^*$ and the following formula is unsatisfiable:
 \begin{equation}\label{eq:pivot}
@@
 where $\theta_1,\ldots,\theta_{k+1}$ are defined as in
-(\ref{eq:interpolation-problem}), for $\theta_0 \equiv
-\Lambda(p)[Q_0/Q]$.
+(\ref{eq:interpolation-problem}) and $\theta_0 \equiv
+\Lambda(p)[Q_0/Q]$. The \emph{pivot} of $n$ is the node $p$
+corresponding to the start of the least infeasible suffix.  We assume
+the existence of two functions $\Call{FindPivot}{u,\Art}$ and
+$\Call{LeastInfeasibleSuffix}{u,\Art}$ that return the pivot and least
+infeasbile suffix of a sequence $u \in \Sigma^*$ in an ART $\Art$,
+without detailing their implementation.
 
@@
 
+Termination of Algorithm \ref{alg:predabs} depends on the ability of a
+given interpolating decision procedure for the combined boolean and
+data theory $\theory(\sig,\I)$ to provide interpolants that yield a
+safety invariant, whenever $L(\A) = \emptyset$. In this case, we use
+the covering relation $\lhd$ to ensure that, when a newly generated
+node is covered by a node already in $N$, it is not added to the
+worklist, thus cutting the current branch of the search.
+
+Formally, for any two nodes $n,m \in N$, we have $n \lhd m$ iff
+$\AbsPost{\A}(\Lambda(n),a) \models \Lambda(m)$ for some $a \in
+\Sigma$, in other words, if $n$ has a successor whose label entails
+the label of $m$.
+
 \paragraph{\em Example}
@@
 formulae than $\bot$ in $\preds$, the successor of $\iota \equiv q_0$
-is $\top$, in Figure \ref{fig:pa} (a). The pivot of $a$ is the root of
-the ART and the spuriousness check yields the predicates $q_0$ and
-$q_1$, which are added to $\preds$ and the $\top$ node is removed. The
-next time $a$ is fired, it creates a node labeled $q_1$ and the next
-sequence $aa$ creates a successor node $q_1$, which is covered by the
-first, depicted with a dashed arrow, in Figure \ref{fig:pa} (b). The
-third sequence is $ab$, which results in a new uncovered node $\top$
-and triggers a spuriousness check. The new predicate obtained from
-this check is $x\leq0\wedge q_2 \wedge y\geq0$ and the pivot is again
-the root. Then the ART is rebuilt with the new predicates and
-the sequence $aab$ yields an uncovered node $\top$, in Figure
-\ref{fig:pa} (c). The new pivot is the endpoint of $a$ and the new
-predicates are $q_1\wedge q_2$ and $y>x-1 \wedge q_2$. Finally, the
-ART is rebuilt from the pivot node and the result is a tree in which
-all nodes are covered, thus proving the emptiness of the automaton, in
-Figure \ref{fig:pa} (d). \qed
-
-Termination of Algorithm \ref{alg:predabs} depends on the ability of a
-given interpolating decision procedure for the combined boolean and
-data theory $\theory(\sig,\I)$ to provide interpolants that yield a
-safety invariant, whenever $L(\A) = \emptyset$. In this case, we
-use the covering relation $\lhd$ to ensure that, when a newly
-generated node is covered by a node already in $N$, it is not added to
-the worklist, thus cutting the current branch of the search. Formally,
-for any two nodes $n,m \in N$, we have $n \lhd m$ iff
-$\AbsPost{\A}(\Lambda(n),a) \models \Lambda(m)$ for some $a \in
-\Sigma$, in other words, if $n$ has a successor whose label entails
-the label of $m$. The correctness criteria of Algorithm
-\ref{alg:predabs} are stated by the following theorem:
+is $\top$, in Figure \ref{fig:pa} (a). The spuriousness check for $a$
+yields the root of the ART as pivot and the interpolant
+$\tuple{q_0,q_1}$, which is added to the set $\preds$. Then the $\top$
+node is removed and the next time $a$ is fired, it creates a node
+labeled $q_1$. The second sequence $aa$ creates a successor node
+$q_1$, which is covered by the first, depicted with a dashed arrow, in
+Figure \ref{fig:pa} (b). The third sequence is $ab$, which results in
+a new uncovered node $\top$ and triggers a spuriousness check. The new
+predicate obtained from this check is $x\leq0\wedge q_2 \wedge y\geq0$
+and the pivot is again the root. Then the entire ART is rebuilt with
+the new predicates and the fourth sequence $aab$ yields an uncovered
+node $\top$, in Figure \ref{fig:pa} (c). The new pivot is the endpoint
+of $a$ and the newly added predicates are $q_1\wedge q_2$ and $y>x-1
+\wedge q_2$. Finally, the ART is rebuilt from the pivot node and
+finally all nodes are covered, thus proving the emptiness of the
+automaton, in Figure \ref{fig:pa} (d). \qed
+
+The correctness of Algorithm \ref{alg:predabs} is proved below:
 
@@
 We consider a total alphabetical order $\prec$ on $\Sigma$ and lift it
-to the total lexicographical order $\prec^*$ on $\Sigma^*$. We write
-$n \prec^* m$ if the sequence labeling the path from $\rootNode$ to
-$n$ is strictly smaller, in the lexicographical order, than the
-sequence labeling the path from $\rootNode$ to $m$.
-
-A node $n \in N$ is \emph{covered} if $(n,p) \in \lhd$ or it has an
-ancestor $m$ such that $(m,p) \in \lhd$, for some $p \in N$. A node
-$n$ is \emph{closed} if it is covered, or $\Lambda(n) \not\models
-\Lambda(m)$ for all $m \in N$ such that $m \prec^* n$.
+to the total lexicographical order $\prec^*$ on $\Sigma^*$. A node $n
+\in N$ is \emph{covered} if $(n,p) \in \lhd$ or it has an ancestor $m$
+such that $(m,p) \in \lhd$, for some $p \in N$. A node $n$ is
+\emph{closed} if it is covered, or $\Lambda(n) \not\models \Lambda(m)$
+for all $m \in N$ such that $\lambda(m) \prec^* \lambda(n)$. Observe
+that we use the coverage here with a different meaning than in
+Algorithm \ref{alg:predabs}.
 
@@
 
-  \If{$\Phi(a_1 \ldots a_k)$ is satisfiable} 
+  \If{$\Accept{\A}(a_1\ldots a_k)$ is satisfiable} 
   \Comment{counterexample is feasible}
 
-  \State get model $\beta : Q_i \rightarrow \booli$, $\{\nu_i :
-  \vec{x}_i \rightarrow \Data^\I \mid i=0\ldots k\}$ of $\mu(n)$
+  \State get model $(\beta,\nu_1,\ldots,\nu_k)$ of
+  $\Accept{\A}(\lambda(n))$
 
@@
   \State let $\tuple{\top,I_0,\ldots,I_k,\bot}$ be an interpolant for
-  $\Phi(a_1 \ldots a_k)$
+  $\Theta(a_1 \ldots a_k)$
   \label{ln:refine-begin}
 
-  \State let $b$ be $\false$
+  \State $b \leftarrow \false$
 
@@
 
-  \State $T \leftarrow T \cup \set{(e, \Theta(n,a,k))}$
+  \State $T \leftarrow T \cup \set{(e, \theta_k)}$
 
@@
 
-\For{$y \in N$ such that $y \prec^* x$}
+\For{$y \in N$ such that $\lambda(y) \prec^* \lambda(x)$}
 
@@
 \emph{refine} and \emph{expand}. Let $n$ be a node removed from the
-worklist at line \ref{ln:impact-dequeue}. If $\Phi(\lambda(n))$ is
-satisfiable, the counterexample $\lambda(n)$ is feasible, in which
-case a model of $\Phi(\lambda(n))$ is obtained and a word $w \in
-L(\A)$ is returned. 
-%
-%% When a node $n$ is removed from the worklist, it is first closed (line
-%% \ref{ln:close-worklist}) by finding a lexicographically bigger node
-%% $m$ such that $\Lambda(n) \models \Lambda(m)$, if one exists, and
-%% adding the pair $(n,m)$ to $\lhd$. As in Algorithm \ref{alg:predabs},
-%% the intuition is that for any feasible counterexample that can be
-%% discovered from $n$ there exists a feasible counterexample stemming in
-%% $m$, thus we do not need to process $n$ further.
-%
-%% If $n$ is not covered, we check if it is the endpoint of a
-%% counterexample candidate, which can only be the case when the
-%% replacement set $R(n)$ contains at least one final state (line
-%% \ref{ln:cex}). Observe that, if $u$ is the label of the path from
-%% $\rootNode$ to $n$, then $\Accept{\A}(u)$ is satisfiable only if it
-%% contains a (positive occurrence of) a final state. The counterexample
-%% is moreover feasible iff $\mu(n)$ is satisfiable, in which case a
-%% model of $\mu(n)$ is obtained and a word $w \in L(\A)$ returned.
-%
-Otherwise, $\lambda(n)$ is a spurious counterexample and the procedure
-enters the refinement phase (lines
+worklist at line \ref{ln:impact-dequeue}. If $\Accept{\A}(\lambda(n))$
+is satisfiable, the counterexample $\lambda(n)$ is feasible, in which
+case a model of $\Accept{\A}(\lambda(n))$ is obtained and a word $w
+\in L(\A)$ is returned. Otherwise, $\lambda(n)$ is a spurious
+counterexample and the procedure enters the refinement phase (lines
 \ref{ln:refine-begin}-\ref{ln:refine-end}). The interpolant for
-$\Phi(\lambda(n))$ is used to strenghten the labels of all the
-ancestors of $n$, by conjoining the formulae of the interpolant to the
-existing labels.
+$\Theta(\lambda(n))$ (cf. formula \ref{eq:interpolation-problem}) is
+used to strenghten the labels of all the ancestors of $n$, by
+conjoining the formulae of the interpolant to the existing labels.
 
@@
 call to $\Call{Close}{}$ must uncover each node which is covered by a
-successor of $n$ (line \ref{ln:close-uncover} in the $\Call{Close}{}$
-procedure). This is required because, due to the over-approximation of
+successor of $n$ (line \ref{ln:close-uncover} of the $\Call{Close}{}$
+function). This is required because, due to the over-approximation of
 the sets of reachable configurations, the covering relation is not
@@
 \ref{ln:expand-begin}-\ref{ln:expand-end}) by creating a new node for
-each successor $s$ via the event $a \in \Sigma$ and inserting it into
-the worklist.
-
-\begin{figure}[htb]
+each successor $s$ via the input event $a \in \Sigma$ and inserting it
+into the worklist.
+
+\begin{figure}[t!]
 \begin{tabular}{cc}
@@
 from Figure \ref{fig:ex}.  Initially, the procedure fires the sequence
-$a$, whose endpoint it labeled with $\top$, in Figure \ref{fig:im}
-(a). Since this node is uncovered, it checks the spuriousness of the
-counterexample $a$ and refines the label of the node to $q_1$. Since
+$a$, whose endpoint is labeled with $\top$, in Figure \ref{fig:im}
+(a). Since this node is uncovered, we check the spuriousness of the
+counterexample $a$ and refine the label of the node to $q_1$. Since
 the node is still uncovered, two successors, labeled with $\top$ are
@@
 spurious, which changes the label of its endpoint from $\top$ to
-$\bot$, and also covers it. Since the endpoint of $aa$ is not covered,
-it is expanded in Figure \ref{fig:im} (c). Both sequences $aaa$ and
-$aab$ are found to be spurious, and the enpoint of $aab$, whose label
-has changed from $\top$ to $\bot$, is now covered. In the process, the
-label of $aa$ has also changed from $q_1$ to $q_1 \wedge y>x-1 \wedge
-q_2$, due to the strenghtening with the interpolant from $aab$.
-Finally, the only uncovered node is expanded to $aaaa$ and $aaab$,
-both found to be spurious, in Figure \ref{fig:im} (d). The refinement
-of $aaab$ causes the label of $aaa$ to change from $q_1$ to $q_1
-\wedge y>x-1 \wedge q_2$ and this node is now covered by $aa$. Since
-its successors are also covered, there are no uncovered nodes and the
-procedure returns $\true$. \qed
+$\bot$, and also covers it (depicted with a dashed edge). Since the
+endpoint of $aa$ is not covered, it is expanded to $aaa$ and $aab$, in
+Figure \ref{fig:im} (c). Both sequences $aaa$ and $aab$ are found to
+be spurious, and the enpoint of $aab$, whose label has changed from
+$\top$ to $\bot$, is now covered. In the process, the label of $aa$
+has also changed from $q_1$ to $q_1 \wedge y>x-1 \wedge q_2$, due to
+the strenghtening with the interpolant from $aab$.  Finally, the only
+uncovered node $aaa$ is expanded to $aaaa$ and $aaab$, both found to be
+spurious, in Figure \ref{fig:im} (d). The refinement of $aaab$ causes
+the label of $aaa$ to change from $q_1$ to $q_1 \wedge y>x-1 \wedge
+q_2$ and this node is now covered by $aa$. Since its successors are
+also covered, there are no uncovered nodes and the procedure returns
+$\true$. \qed
 
@@
 The correctness of Algorithm \ref{alg:impact} is proved in the
-following.
+following:
 
@@
 \end{theorem}
+
+\section{Experimental Evaluation}
 

### 2017-05-15 draft.tex
ok

@@
 provide two efficient semi-algorithms for emptiness, based on lazy
-predicate abstraction \cite{blast} and the \impact~method
+predicate abstraction \cite{HJMS02} and the \impact~method
 \cite{mcmillan06}.  These algorithms are proven to terminate and
@@
 exponential determinization procedure and an abstraction refinement
-loop using lazy predicate abstraction \cite{blast}. In this work we
+loop using lazy predicate abstraction \cite{HJMS02}. In this work we
 consider the full model of alternation and rely entirely on the
@@
 state-of-the-art \impact~ procedure \cite{mcmillan06} that is shown to
-outperform lazy predicate abstraction \cite{blast} in the
+outperform lazy predicate abstraction \cite{HJMS02} in the
 nondeterministic case, and generalized it to cope with
@@
 the set of positive integers, an ADA can simulate an Alternating
-Vector Addition System with States (AVASS) \cite{LazicSchmitz14} using
-only atoms $x \geq k$ and $x = \overline{x} + k$, for $k \in \zed$,
-with the classical interpretation of the function symbols on
-integers. Since reachability of a control state is undecidable for
-AVASS \cite{LINCOLN92}, the emptiness problem is undecidable for ADA.
+Vector Addition System with States (AVASS) using only atoms $x \geq k$
+and $x = \overline{x} + k$, for $k \in \zed$, with the classical
+interpretation of the function symbols on integers. Since reachability
+of a control state is undecidable for AVASS \cite{LINCOLN92}, the
+emptiness problem is undecidable for ADA.
 
@@
 for all $m \in N$ such that $\lambda(m) \prec^* \lambda(n)$. Observe
-that we use the coverage here with a different meaning than in
-Algorithm \ref{alg:predabs}.
+that we use the coverage relation $\lhd$ here with a different meaning
+than in Algorithm \ref{alg:predabs}.
 
@@
 
+\vspace*{\baselineskip}
 \proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
@@
   \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \iffeq & \text{ (by the def. of $\overline{\phi}$)} \\
-  \overline{\Delta(\iota_1,a_1\ldots,a_n)}
+  \overline{\Delta(\iota_1,a_1\ldots,a_n)} \enspace.
   \end{array}\]
@@
   \Sigma[\vec{x}]^*$, we have $w \in L(\A_1) \iff w \not\in
-  L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) = \Sigma[\vec{x}]^*
-  \setminus L(\A_1)$. \qed}
+  L(\overline{\A_1})$. \qed}
 
@@
 \subsection{Proof of Proposition \ref{prop:positive-interpol}}
-
-\proof{ Let $\tuple{\top,I_0,\ldots,I_{n},\bot}$ be an interpolant,
-  and $i \in [0,n]$ be the first index for which $I_i \not\in
-  \Form^+(Q,\vec{x})$. By Proposition \ref{prop:asterisk}, we have
-  that $\Delta^{i-1}(I_{i-1},a_i) \iffeq \exists Q_{i-1} ~.~
-  I_{i-1}[Q_{i-1}/Q] \wedge \theta_i$. Since $I_{i-1} \in
-  \Form^+(Q,\vec{x})$, by the choice of $i$, we have that
-  $\Delta^{i-1}(I_{i-1},a_i) \in \Form^+(Q,\vec{x})$ as well. We
-  replace $I_i$ with $\Delta^{i-1}(I_{i-1},a_i)$. The result is still
-  an interpolant, because: \begin{compactitem}
-  \item $I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \models
-    \Delta^{i-1}(I_{i-1},a_i)$, by Proposition \ref{prop:asterisk}, and
-    %
-  \item $\Delta^{i-1}(I_{i-1},a_i) \wedge \theta_{i+1} \models \exists
-    Q_{i-1} ~.~ I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \wedge \theta_{i+1}
-    \models I_i[Q_i/Q] \wedge \theta_{i+1} \models I_{i+1}[Q_{i+1}/Q]$.
-  \end{compactitem}
-  We proceed with these replacements until there are no more formulae
-  $I_i \not\in \Form^+(Q,\vec{x})$ left. \qed}
-
-\subsection{Proof of Lemma \ref{lemma:progress}}
 
@@
 
+\vspace*{\baselineskip} \proof{ Let
+  $\tuple{\top,I_0,\ldots,I_{n},\bot}$ be an interpolant, and $i \in
+  [0,n]$ be the first index for which $I_i \not\in
+  \Form^+(Q,\vec{x})$. If $i=0$, we replace $I_0$ with $\iota \in
+  \Form^+(Q,\vec{x})$ and verify that the result is still an
+  interpolant. If $i>0$, by Proposition \ref{prop:asterisk}, we have
+  that $\Delta^{i-1}(I_{i-1},a_i) \iffeq \exists Q_{i-1} ~.~
+  I_{i-1}[Q_{i-1}/Q] \wedge \theta_i$. Since $I_{i-1} \in
+  \Form^+(Q,\vec{x})$, by the choice of $i$, we have that
+  $\Delta^{i-1}(I_{i-1},a_i) \in \Form^+(Q,\vec{x})$ as well. We
+  replace $I_i$ with $\Delta^{i-1}(I_{i-1},a_i)$. The result is still
+  an interpolant, because: \begin{compactitem}
+  \item $I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \models
+    \Delta^{i-1}(I_{i-1},a_i)$, by Proposition \ref{prop:asterisk}, and
+    %
+  \item $\Delta^{i-1}(I_{i-1},a_i) \wedge \theta_{i+1} \models \exists
+    Q_{i-1} ~.~ I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \wedge \theta_{i+1}
+    \models I_i[Q_i/Q] \wedge \theta_{i+1} \models I_{i+1}[Q_{i+1}/Q]$.
+  \end{compactitem}
+  We proceed with these replacements until there are no more formulae
+  $I_i \not\in \Form^+(Q,\vec{x})$ left. \qed}
+
+\subsection{Proof of Lemma \ref{lemma:progress}}
+
 \proof{ Let $\Theta(u) \equiv \theta_0(Q_0) \wedge \theta_1(Q_0 \cup
@@
   (\ref{it2:progress}) If $\tuple{\top,I_0,\dots,I_n,\bot}$ is an
-  interpolant for $\pi(u)$, the following entailments
+  interpolant for $\Theta(u)$, the following entailments
   hold: \begin{compactitem}
@@
   \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n \models \text{ (ind. hyp.) } \\
-  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q]
-  \end{array}\]
-\qed}
+  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q] \hfill\text{\qed}
+  \end{array}\]}
 
@@
   \begin{equation}\label{inv:predabs}
-  W = \set{n \not\in N \mid \forall m \in N~ \forall a \in \Sigma ~.~ (n,a,m)
+  W = \set{n \mid \forall m \in N~ \forall a \in \Sigma ~.~ (n,a,m)
     \not\in E ~\wedge~ (n,m) \not\in \lhd}
   \end{equation}
-  Initially, this is the case because $W = \set{\rootNode}$ and $N = E
-  = \lhd = \emptyset$. If the invariant holds previously, at line
+  Initially, this is the case because $W = \set{\rootNode}$ and $E =
+  \lhd = \emptyset$. If the invariant holds previously, at line
   \ref{ln:while}, it will hold again after line \ref{ln:refine} is
@@
   removed, $p$ becomes a member of the set of uncovered frontier
-  nodes, and is added to $W$ at line \ref{ln:refine}. The last case
-  considers that the invariant holds at line \ref{ln:while} and the
-  control follows the else branch at line \ref{ln:else}. In this case,
-  the newly created frontier node $s$ is added to $W$ only if it is
-  not covered by an existing node in $N$ (line \ref{ln:covered}).
+  nodes, and is added to $W$ at line \ref{ln:refine}. Otherwise, the
+  invariant holds at line \ref{ln:while} and the control follows the
+  else branch at line \ref{ln:else}. In this case, the newly created
+  frontier node $s$ is added to $W$ only if it is not covered by an
+  existing node in $N$ (line \ref{ln:covered}).
 
@@
   the proof that $\bigvee_{n\in N} \Lambda(n)$ is a safety invariant,
-  thus $L(\A) = \emptyset$, by Lemma \ref{lemma:safety-invariant}.
-
-  We have then proved the second point of the statement. For the first
-  point, assume that $L(\A) \neq 0$ and let $w = (a_1,\nu_1) \ldots
-  (a_k,\nu_k) \in L(\A)$ be a word. By the above, Algorithm
-  \ref{alg:predabs} cannot return $\true$. Suppose, by contradiction
-  that it does not terminate. Since the sequences from $\Sigma^*$ are
-  explored in breadth-first order, every sequence of length $k$ is
-  eventually explored, which leads to the discovery of $w$ at line
-  \ref{ln:feasible}. Then Algorithm \ref{alg:predabs} terminates
-  returning $w \in L(\A)$. \qed}
+  thus $L(\A) = \emptyset$, by Lemma \ref{lemma:safety-invariant}. We
+  have then proved the second point of the statement.
+
+  For the first point, assume that $L(\A) \neq 0$ and let $w =
+  (a_1,\nu_1) \ldots (a_k,\nu_k) \in L(\A)$ be a word. By the above,
+  Algorithm \ref{alg:predabs} cannot return $\true$. Suppose, by
+  contradiction that it does not terminate. Since the sequences from
+  $\Sigma^*$ are explored in breadth-first order, every sequence of
+  length $k$ is eventually explored, which leads to the discovery of
+  $w$ at line \ref{ln:feasible}. Then Algorithm \ref{alg:predabs}
+  terminates returning $w \in L(\A)$. \qed}
 
@@
   be a spurious counterexample. Let
-  $\langle\top,I^\omega_0,\ldots,I^\omega_k\equiv\bot\rangle$ be an
+  $\langle\top,I^\omega_0,\ldots,I^\omega_k,\bot\rangle$ be an
   interpolant for $\Phi(a_1 \ldots a_k) \equiv \Lambda(\rootNode)
@@
   \begin{equation}\label{inv:impact}
-  W = \set{n \not\in N \mid n \text{ uncovered, } \exists a \in
-    \Sigma~ \forall s \in N ~.~ (n,a,s) \not\in E}
+  W = \set{n \mid n \text{ uncovered, } \exists a \in \Sigma~ \forall
+    s \in N ~.~ (n,a,s) \not\in E}
   \end{equation}
-  Initially, $W = \set{\rootNode}$ and $N=E=\lhd=\emptyset$, thus
+  Initially, $W = \set{\rootNode}$ and $E=\lhd=\emptyset$, thus
   (\ref{inv:impact}) holds trivially. Suppose that (\ref{inv:impact})
@@
   \item $n$ is covered, in which case $W$ becomes $W \setminus
-    \set{n}$, or
+    \set{n}$ and (\ref{inv:impact}) holds, or
   %
@@
       a \in \Sigma}$ is the set of fresh successors of $n$. But then
-    no node $s \in S$ is covered and has successors in $E$.
+    no node $s \in S$ is covered and has successors in $E$, thus
+    (\ref{inv:impact}) holds.
   \end{itemize}
@@
   it must be the case that a successor of $n_1$ is missing from
-  $(N,E)$, thus $n_1$ must be covered. Let $n'_1$ be the closest
-  ancestor of $n_1$ such that $(n'_1,n''_1) \in \lhd$, for some $n''_1
-  \in N$, and let $v'_1$ be the prefix of $v_1$ leading to $n'_1$. By
-  the construction of $\lhd$ (line \ref{ln:close-uncover} in function
+  $(N,E)$, thus $n_1$ must be covered, by (\ref{inv:impact}) and the
+  fact that $W=\emptyset$. Let $n'_1$ be the closest ancestor of $n_1$
+  such that $(n'_1,n''_1) \in \lhd$, for some $n''_1 \in N$, and let
+  $v'_1$ be the prefix of $v_1$ leading to $n'_1$. By the construction
+  of $\lhd$ (line \ref{ln:close-uncover} in function
   $\Call{Close}{}$), we have $\Lambda(n'_1) \models

### 2017-05-15 draft.tex
ok

@@
 \author{Radu Iosif and Xiao Xu \\
-Verimag/CNRS/Universit\'e de Grenoble Alpes\\
+CNRS, Verimag, Universit\'e de Grenoble Alpes \\
 Email: \{Radu.Iosif,Xiao.Xu\}@univ-grenoble-alpes.fr}
@@
 
-\paragraph{\em Remark}
-While computing the concrete post-image $\Post{\A}$ relies on an
-expensive quantifier elimination, it is common belief that this should
-hold also for $\AbsPost{\A}$. However, chosing the conjoined
-predicates that form $\AbsPost{\A}$ requires only a quantifier-free
-satisfiability query: $\Post{\A}(\phi,a) \models \pi$ iff $\exists
-\overline{\vec{x}} ~.~ \Delta(\phi[\overline{\vec{x}}/\vec{x}],a)
-\models \pi$ iff $\Delta(\phi[\overline{\vec{x}}/\vec{x}],a) \wedge
-\neg\pi$ is unsatisfiable. \qed
+%% \paragraph{\em Remark}
+%% While computing the concrete post-image $\Post{\A}$ relies on an
+%% expensive quantifier elimination, it is common belief that this should
+%% hold also for $\AbsPost{\A}$. However, chosing the conjoined
+%% predicates that form $\AbsPost{\A}$ requires only a quantifier-free
+%% satisfiability query: $\Post{\A}(\phi,a) \models \pi$ iff $\exists
+%% \overline{\vec{x}} ~.~ \Delta(\phi[\overline{\vec{x}}/\vec{x}],a)
+%% \models \pi$ iff $\Delta(\phi[\overline{\vec{x}}/\vec{x}],a) \wedge
+%% \neg\pi$ is unsatisfiable. \qed
 
@@
 
-\paragraph{\em Remark} 
-We deviate w.l.o.g. from the standard notion of \emph{generalized
-  interpolants} \cite{mcmillan06}, which requires $\fv{}{I_i}
-\subseteq \fv{}{\phi_i} \cap \fv{}{\phi_{i+1}}$, for each $i \in
-          [1,n-1]$, and assume further that the variables have been
-          renamed by removing the time stamps. \qed
+%% \paragraph{\em Remark} 
+%% We deviate w.l.o.g. from the standard notion of \emph{generalized
+%%   interpolants} \cite{mcmillan06}, which requires $\fv{}{I_i}
+%% \subseteq \fv{}{\phi_i} \cap \fv{}{\phi_{i+1}}$, for each $i \in
+%%           [1,n-1]$, and assume further that the variables have been
+%%           renamed by removing the time stamps. \qed
 
@@
 
-\begin{figure}[htb]
-\begin{tabular}{cc}
-\includegraphics[scale=0.52]{figures/PA1.pdf} & \includegraphics[scale=0.52]{figures/PA2.pdf} \\
-\footnotesize{(a)} & \footnotesize{(b)} \\
-\\
-\includegraphics[scale=0.52]{figures/PA3.pdf} & \includegraphics[scale=0.52]{figures/PA4.pdf} \\
-\footnotesize{(c)} & \footnotesize{(d)}
-\end{tabular}
-\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex} using Algorithm \ref{alg:predabs}}
-\label{fig:pa}
-\end{figure}
-
 Termination of Algorithm \ref{alg:predabs} depends on the ability of a
@@
 the label of $m$.
+
+\begin{figure}[t!]
+\begin{tabular}{cc}
+\includegraphics[scale=0.5]{figures/PA1.pdf} & \includegraphics[scale=0.5]{figures/PA2.pdf} \\[-2mm]
+\tiny{(a)} & \tiny{(b)} \\
+\\
+\includegraphics[scale=0.5]{figures/PA3.pdf} & \includegraphics[scale=0.5]{figures/PA4.pdf} \\[-2mm]
+\tiny{(c)} & \tiny{(d)}
+\end{tabular}
+\vspace*{-\baselineskip}
+\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex} by Algorithm \ref{alg:predabs}}
+\label{fig:pa}
+\end{figure}
 
@@
 
-%% The ART built by Algorithm \ref{alg:impact} corresponds to an
-%% unfolding of the input automaton by enumerating the input event
-%% sequences in $\Sigma^*$. Each time a new node is created, its label is
-%% initially $\top$. This label is subject to subsequent strenghtening,
-%% when a spurious counterexample is encountered. To speedup the
-%% computation of interpolants for spurious counterexamples, we use two
-%% additional labeling functions for bookkeeping. Formally, the ART
-%% structure used by Algorithm \ref{alg:impact} is defined as $\Art =
-%% \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$, where $N,E,\Lambda$ and
-%% $\lhd$ are defined as for Algorithm \ref{alg:predabs} and,
-%% moreover: \begin{compactitem}
-%% \item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
-%%   sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$, 
-%% %
-%% \item $T : E \rightarrow \bigcup_{i=0}^\infty
-%%   \Form(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is labeling of edges
-%%   with time-stamped formulae, and
-%% %
-%% \item for each edge $e = (n,a,m) \in E$, if $n$ lies at
-%%   depth\footnote{We consider that the root node has zero depth.}  $k$
-%%   in the tree $(N,E)$, then we have:
-%% \end{compactitem}
-%%   \[\begin{array}{rcl}
-%%   T(e) & = & \Theta(n,a,k) \\
-%%   R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)} \text{, where} \\
-%%   \Theta(n,a,k) & \equiv & \bigwedge_{q \in R(n)} (q_k \rightarrow
-%%   \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}])
-%%   \end{array}\]
-%% For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is
-%% $(n,a,m)$. This additional bookkeeping is used as follows. Given a
-%% sequence $u = a_1\ldots a_k \in \Sigma^*$ that labels the path from
-%% the root to a node $n$ in $\Art$, the conjunction:
-%% \[\mu(n) \equiv \Lambda(\rootNode) \wedge T(\rootNode,a_1,n_1) \wedge \ldots
-%% T(n_{k-1},a_k,n_k) \wedge \bigwedge_{q \in R(n_k) \setminus F}(q_k
-%% \rightarrow \bot)\] is equivalent to the interpolation problem
-%% $\pi(u)$ and can be directly computed by a traversal of the path from
-%% $\rootNode$ to $n$.
-
 We consider a total alphabetical order $\prec$ on $\Sigma$ and lift it
@@
 
-\begin{figure}[t!]
-\begin{tabular}{cc}
-\includegraphics[scale=0.42]{figures/IM1.pdf} & \includegraphics[scale=0.42]{figures/IM2.pdf} \\
-\footnotesize{(a)} & \footnotesize{(b)} \\
-\\
-\includegraphics[scale=0.42]{figures/IM3.pdf} & \includegraphics[scale=0.42]{figures/IM4.pdf} \\
-\footnotesize{(c)} & \footnotesize{(d)}
+\begin{figure}[htb]
+\vspace*{-\baselineskip}
+\begin{tabular}{c}
+\includegraphics[scale=0.65]{figures/IM1.pdf} \\[-2mm]
+\tiny{(a)} \\
+\includegraphics[scale=0.65]{figures/IM2.pdf} \\[-2mm]
+\tiny{(b)} \\
+\includegraphics[scale=0.65]{figures/IM3.pdf} \\[-2mm]
+\tiny{(c)}
 \end{tabular}
-\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex} using Algorithm \ref{alg:impact}}
-\label{fig:im}
+\vspace*{-0.5\baselineskip}
+\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex}
+  by Algorithm \ref{alg:impact} (1/2)}
+\label{fig:im1}
+\end{figure}
+
+\begin{figure}[htb]
+\centerline{\includegraphics[scale=0.65]{figures/IM4.pdf}}
+\vspace*{-\baselineskip}
+\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex}
+  by Algorithm \ref{alg:impact} (2/2)}
+\label{fig:im2}
 \end{figure}
@@
 from Figure \ref{fig:ex}.  Initially, the procedure fires the sequence
-$a$, whose endpoint is labeled with $\top$, in Figure \ref{fig:im}
+$a$, whose endpoint is labeled with $\top$, in Figure \ref{fig:im1}
 (a). Since this node is uncovered, we check the spuriousness of the
@@
 computed, corresponding to the sequences $aa$ and $ab$, in Figure
-\ref{fig:im} (b). The spuriousness check for $aa$ yields the
+\ref{fig:im1} (b). The spuriousness check for $aa$ yields the
 interpolant $\tuple{q_0,x\leq0\wedge q_2 \wedge y\geq0}$ which
@@
 endpoint of $aa$ is not covered, it is expanded to $aaa$ and $aab$, in
-Figure \ref{fig:im} (c). Both sequences $aaa$ and $aab$ are found to
+Figure \ref{fig:im1} (c). Both sequences $aaa$ and $aab$ are found to
 be spurious, and the enpoint of $aab$, whose label has changed from
@@
 uncovered node $aaa$ is expanded to $aaaa$ and $aaab$, both found to be
-spurious, in Figure \ref{fig:im} (d). The refinement of $aaab$ causes
+spurious, in Figure \ref{fig:im2}. The refinement of $aaab$ causes
 the label of $aaa$ to change from $q_1$ to $q_1 \wedge y>x-1 \wedge
@@
 $\true$. \qed
-
-The following lemma states that the labeling of the nodes in an ART
-built by Algorithm \ref{alg:impact} is consistent with the semantics
-of the transition function of $\A$.
-
-\begin{lemma}\label{lemma:well-labeled}
-  Given an ART $\Art = \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$ built
-  by Algorithm \ref{alg:impact}, $\Post{\A}(\Lambda(n),a) \models
-  \Lambda(m)$, for all $(n,a,m) \in E$.
-\end{lemma}
 
@@
 \section{Experimental Evaluation}
+
+xxx
 
@@
 
-\subsection{Proof of Lemma \ref{lemma:well-labeled}}
-
+\subsection{Proof of Theorem \ref{thm:impact}}
+
+\begin{lemma}\label{lemma:well-labeled}
+  Given an ART $\Art = \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$ built
+  by Algorithm \ref{alg:impact}, $\Post{\A}(\Lambda(n),a) \models
+  \Lambda(m)$, for all $(n,a,m) \in E$.
+\end{lemma}
 \proof{ We distinguish two cases. First, if $(n,a,m)$ occurs on a path
@@
 
-\subsection{Proof of Theorem \ref{thm:impact}}
-
+\vspace*{\baselineskip}
 \proof{ We prove first that, each time Algorithm \ref{alg:impact}

### 2017-05-15 draft.tex
ok

@@
 
-xxx
+We have implemented both Algorithm \ref{alg:predabs} and
+\ref{alg:impact} in a prototype tool\footnote{\url{XXXX}} that uses
+the Z3 SMT solver\footnote{\url{https://github.com/Z3Prover/z3}} to
+check satisfiability queries and generate interpolants in the theory
+of linear integer arithmetic variables (LIA). We compared both
+algorithms with a previous implementation of trace inclusion
+procedure, called
+\textsc{Includer}\footnote{\url{http://www.fit.vutbr.cz/research/groups/verifit/tools/includer/}},
+that uses on-the-fly determinisation and lazy predicate abstraction
+with interpolant-based refinement \cite{Tacas16}.
+
+The results of the experiments are given in Table
+\ref{tab:experiments}. We applied the tool on several array logic
+entailments, which occur as verification conditions for imperative
+programs with arrays \cite{cav09} (Array shift, Array rotation 1+2,
+Array split) available online \cite{ntslib}. Then, we applied it on
+proving safety properties of hardware circuits (HW Counter 1+2)
+\cite{smrcka}. Finally, we considered two timed communication
+protocols, consisting of systems that are asynchronous compositions of
+timed automata, whom correctness specification are given by timed
+automata monitors: two versions (correct and faulty) of the timed
+Alternating Bit Protocol \cite{abp} and a~controller of a railroad
+crossing \cite{henzinger:RealTimeSystems}. All results were obtained
+on an XXXX CPU @ YYYY GHz machine with ZZZZ GB of RAM. Execution times
+are given in seconds.
+
+\begin{table}[htb]
+\begin{center}
+{\fontsize{8}{9}\selectfont
+\begin{tabular}{||l|c|c|c|c|c||}
+\hline
+Example & $\len{\A}$ & $L(\A)=\emptyset$ ? & Algorithm \ref{alg:predabs} & Algorithm \ref{alg:impact} & \textsc{Includer} \\
+\hline 
+\hline
+\end{tabular}
+}
+\caption{}\label{tab:experiments}
+\end{center}
+\end{table}
 
@@
 \appendix
-
-\section{Proofs}
 

### 2017-05-15 draft.tex
ok

@@
 We have implemented both Algorithm \ref{alg:predabs} and
-\ref{alg:impact} in a prototype tool\footnote{\url{XXXX}} that uses
+\ref{alg:impact} in a prototype tool\footnote{\url{IMPLEMENTATION URL}} that uses
 the Z3 SMT solver\footnote{\url{https://github.com/Z3Prover/z3}} to

### 2017-05-15 draft.tex
ok

--- 2017-05-15 draft.tex
@@
 We have implemented both Algorithm \ref{alg:predabs} and
-\ref{alg:impact} in a prototype tool\footnote{\url{XXXX}} that uses
+\ref{alg:impact} in a prototype
+tool\footnote{\url{https://github.com/cathiec/AltImpact}} that uses
 the Z3 SMT solver\footnote{\url{https://github.com/Z3Prover/z3}} to
@@
 crossing \cite{henzinger:RealTimeSystems}. All results were obtained
-on an XXXX CPU @ YYYY GHz machine with ZZZZ GB of RAM. Execution times
-are given in seconds.
+on an Intel(R) Core(TM) i7-4650U CPU @ 1.70GHz with 8GB of
+RAM. Execution times are given in seconds.
 

### 2017-05-15 draft.tex
ok

@@
 The execution of Algorithm \ref{alg:impact} consists of three
-phases\footnote{These phases correspond to the \textsc{Close},
-  \textsc{Refine} and \textsc{Expand} procedures from the original
-  description of \impact~\cite{mcmillan06}.}: \emph{close},
+phases\footnote{Corresponding to the \textsc{Close}, \textsc{Refine}
+  and \textsc{Expand} in \cite{mcmillan06}.}: \emph{close},
 \emph{refine} and \emph{expand}. Let $n$ be a node removed from the
@@
 \begin{tabular}{c}
-\includegraphics[scale=0.65]{figures/IM1.pdf} \\[-2mm]
+\includegraphics[scale=0.65]{figures/IM1.pdf} \\
 \tiny{(a)} \\
@@
 \label{fig:im1}
+\vspace*{-\baselineskip}
 \end{figure}
@@
 
-The correctness of Algorithm \ref{alg:impact} is proved in the
-following:
-
 \begin{theorem}\label{thm:impact}
@@
 We have implemented both Algorithm \ref{alg:predabs} and
-\ref{alg:impact} in a prototype
-tool\footnote{\url{https://github.com/cathiec/AltImpact}} that uses
-the Z3 SMT solver\footnote{\url{https://github.com/Z3Prover/z3}} to
-check satisfiability queries and generate interpolants in the theory
-of linear integer arithmetic variables (LIA). We compared both
-algorithms with a previous implementation of trace inclusion
+\ref{alg:impact} in a prototype tool\footnote{The implementation is
+  available at \url{https://github.com/cathiec/AltImpact}} that uses
+the Z3 SMT solver\footnote{\url{https://github.com/Z3Prover/z3}} for
+the satisfiability queries and interpolant generation, in the theory
+of linear integer arithmetic (LIA) combined with booleans. We compared
+both algorithms with a previous implementation of a trace inclusion
 procedure, called
@@
 that uses on-the-fly determinisation and lazy predicate abstraction
-with interpolant-based refinement \cite{Tacas16}.
+with interpolant-based refinement \cite{Tacas16} in the LIA theory,
+without booleans.
 
 The results of the experiments are given in Table
-\ref{tab:experiments}. We applied the tool on several array logic
-entailments, which occur as verification conditions for imperative
-programs with arrays \cite{cav09} (Array shift, Array rotation 1+2,
-Array split) available online \cite{ntslib}. Then, we applied it on
-proving safety properties of hardware circuits (HW Counter 1+2)
-\cite{smrcka}. Finally, we considered two timed communication
-protocols, consisting of systems that are asynchronous compositions of
-timed automata, whom correctness specification are given by timed
-automata monitors: two versions (correct and faulty) of the timed
-Alternating Bit Protocol \cite{abp} and a~controller of a railroad
-crossing \cite{henzinger:RealTimeSystems}. All results were obtained
-on an Intel(R) Core(TM) i7-4650U CPU @ 1.70GHz with 8GB of
-RAM. Execution times are given in seconds.
+\ref{tab:experiments}. We applied the tool first to several array
+logic entailments, which occur as verification conditions for
+imperative programs with arrays \cite{cav09} (array\_shift,
+array\_simple, array\_rotation 1+2) available online
+\cite{ntslib}. Next, we applied it on proving safety properties of
+hardware circuits (hw1+2) \cite{smrcka}. Finally, we considered two
+timed communication protocols, consisting of systems that are
+asynchronous compositions of timed automata, whom correctness
+specifications are given by timed automata monitors: a timed version of
+the Alternating Bit Protocol (abp) \cite{abp} and a~controller of a
+railroad crossing (train) \cite{henzinger:RealTimeSystems}. All
+results were obtained on an Intel(R) Core(TM) i7-4650U CPU @ 1.70GHz
+with 8GB of RAM. The automata sizes are in bytes and the execution
+times are in seconds.
 
@@
 \hline
+simple1 & 309 & no & 0.641 & 0.061 & 0.021\\
+\hline
+simple2 & 504 & yes & 0.667 & 0.179 & 0.031\\
+\hline
+simple3 & 214 & yes & 0.825 & 0.150 & 0.051\\
+\hline
+array\_shift & 874 & yes & 2.138 & 0.209 & 0.060\\
+\hline
+array\_simple & 3440 & yes & timeout & 16.435 & 5.496\\
+\hline
+array\_rotation1 & 1834 & yes & 6.827 & 0.865 & 0.114\\
+\hline
+array\_rotation2 & 15182 & yes & timeout & timeout & 23.056\\
+\hline
+abp & 6909 & no & 8.292 & 1.410 & 1.715\\
+\hline
+train & 1823 & yes & 16.989 & 2.738 & 0.308\\
+\hline
+hw1 & 322 & yes & 1.566 & 0.298 & 0.159\\
+\hline
+hw2 & 674 & yes & 22.914 & 0.552 & 0.419\\
 \hline
@@
 \end{table}
+
+As in the non-deterministic case \cite{mcmillan06}, \impact~
+outperforms lazy predicate abstraction for checking emptiness by at
+least one order of magnitude. However, both our implementations are
+slower than \textsc{Includer}, on average (except for the abp
+example). The reason for this is currently under investigation, one
+possible bottleneck being the hardness of the combined (LIA+booleans)
+interpolation problems, as opposed to converting the entire formula
+into DNF, eliminating the boolean variables and using interpolation in
+the pure LIA theory.
 

### 2017-05-16 predicate-automata.tex
ok

--- 2017-05-15 draft.tex
@@
-% \documentclass{llncs}
-\documentclass[10pt,conference,letterpaper,twocolumn]{IEEEtran}
+\documentclass{llncs}
 
@@
 \usepackage{pslatex}
-\usepackage{epsfig}
+% \usepackage{epsfig}
 \usepackage{wrapfig}
@@
 
-\title{The Impact of Alternation}
+\title{From Predicate Automata to Alternating Data Automata}
 
-\author{Radu Iosif and Xiao Xu \\
-CNRS, Verimag, Universit\'e de Grenoble Alpes \\
-Email: \{Radu.Iosif,Xiao.Xu\}@univ-grenoble-alpes.fr}
+\author{Radu Iosif}
+\institute{CNRS, Verimag \\
+Radu.Iosif@univ-grenoble-alpes.fr}
 
@@
 
-\begin{abstract}
-Alternating automata have been widely used to model and verify systems
-that handle data from finite domains, such as communication protocols
-or hardware. The main advantage of the alternating model of
-computation is that complementation is possible in linear time, thus
-allowing to concisely encode trace inclusion problems that occur often
-in verification. In this paper we consider alternating automata over
-infinite alphabets, whose transition rules are formulae in a combined
-theory of booleans and some infinite data domain, that relate past and
-current values of the data variables. The data theory is not fixed,
-but rather it is a parameter of the class. We show that union,
-intersection and complementation are possible in linear time in this
-model and, though the emptiness problem is undecidable, we provide two
-efficient semi-algorithms, inspired by two state-of-the-art
-abstraction refinement model checking methods: lazy predicate
-abstraction \cite{HJMS02} and the \impact~ semi-algorithm
-\cite{mcmillan06}. We have implemented both methods and report the
-results of an experimental comparison.
-\end{abstract}
+\section{Definitions}
 
-\section{Introduction}
+Let $Q$ be a finite set of predicates, where for each $q \in Q$, we
+denote by $\#(q)\geq0$ its arity. We denote by $Q_0$ the subset of $Q$
+consisting of predicates of arity zero. By $\Sigma$ we denote a
+finite set of symbols, called alphabet.
 
-The language inclusion problem is recognized as being central to
-verification of hardware, communication protocols and software
-systems. A property is a specification of the correct executions of a
-system, given as a set $\mathcal{P}$ of executions, and the
-verification problem asks if the set $\mathcal{S}$ of executions of
-the system under consideration is contained within $\mathcal{P}$.
+We consider an infinite countable set $\vars$ of first-order
+variables, ranging over $\nat$, and we associate each predicate $q \in
+Q$ a tuple of arguments $\vec{x}_q = \tuple{x_{q,1}, \ldots,
+  x_{q,\#(q)}}$ such that, for any two $p,q \in Q$, $\vec{x}_p$ and
+$\vec{x}_q$ have no elements in common, when $p \neq q$.
 
-This problem is at the core of widespread verification techniques,
-such as automata-theoretic model checking \cite{VARDI94}, where
-systems are specified as finite-state automata and properties defined
-using Linear Temporal Logic \cite{Pnueli77}. However the bottleneck of
-this and other related verification techniques is the intractability
-of language inclusion (PSPACE-complete for finite-state automata on
-finite words over finite alphabets).
-
-Alternation \cite{ChandraKozenStockmeyer81} was introduced as a
-generalization of nondeterminism, introducing universal, in addition
-to existential transitions. For automata over finite alphabets, the
-language inclusion problem can be encoded as the emptiness problem of
-an alternating automaton of linear size. Moreover, efficient
-exploration techniques based on antichains are shown to perform well
-for alternating automata over finite alphabets \cite{DeWulf08}.
-
-Using finite alphabets for the specification of properties and models
-is however very restrictive, when dealing with real-life computer
-systems, mostly because of the following reasons. On one hand,
-programs handle data from very large domains, that can be assumed to
-be infinite (64-bit integers, floating point numbers, strings of
-characters, etc.) and their correctness must be specified in terms of
-the data values. On the other hand, systems must respond to strict
-deadlines, which requires temporal specifications as timed
-languages \cite{AlurDill94}.
-
-Although being convenient specification tools, automata over infinite
-alphabets lack the decidability properties ensured by finite
-alphabets. In general, when considering infinite data as part of the
-input alphabet, language inclusion is undecidable and, even
-complementation becomes impossible, for instance, for timed automata
-\cite{AlurDill94} or finite-memory register automata
-\cite{Kaminski94}. In some cases, one can recover theoretical
-decidability, by restricting the number of variables (clocks) in timed
-automata to one \cite{OuaknineWorrell04}, or forbidding relations
-between current and past/future values, as with symbolic automata
-\cite{symbTransd:POPL12}. In such cases, the emptiness problem for the
-alternating versions becomes decidable \cite{Lasota05,DAntoniKW16}.
-
-In this paper, we present a new model of alternating automata over
-infinite alphabets consisting of pairs $(a,\nu)$ where $a$ is an input
-event from a finite set and $\nu$ is a valuation of a finite set
-$\vec{x}$ of variables that range over an infinite domain. We assume
-that, at all times, the successive values taken by the variables in
-$\vec{x}$ are an observable part of the language, in other words,
-there are no hidden variables in our model. The transition rules are
-specified by a set of formulae, in a combined first-order theory of
-boolean control states and data, that relate past with present values
-of the variables. We do not fix the data theory a~priori, but rather
-consider it to be a parameter of the class.
-
-A run over an input word $(a_1,\nu_1) \ldots (a_n,\nu_n)$ is a
-sequence $\phi_0(\vec{x}_0) \Rightarrow \phi_1(\vec{x}_0,\vec{x}_1)
-\Rightarrow \ldots \Rightarrow \phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ of
-rewritings of the initial formula by substituting boolean states with
-time-stamped transition rules. The word is accepted if the final
-formula $\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ holds, when all
-time-stamped variables $\vec{x}_1,\ldots,\vec{x}_n$ are substituted by
-their values in $\nu_1,\ldots,\nu_n$, all non-final states replaced by
-false and all final states by true.
-
-The boolean operations of union, intersection and complement can be
-implemented in linear time in this model, thus matching the complexity
-of performing these operations in the finite-alphabet case. The price
-to be paid is that emptiness becomes undecidable, for which reason we
-provide two efficient semi-algorithms for emptiness, based on lazy
-predicate abstraction \cite{HJMS02} and the \impact~method
-\cite{mcmillan06}.  These algorithms are proven to terminate and
-return a word from the language of the automaton, if one exists, but
-termination is not guaranteed when the language is empty.
-
-%% Finally, we prove termination in the empty case, when the transition
-%% formulae\begin{inparaenum}[(i)]
-%% \item contain only constants of the data sort and boolean functions of
-%%   any arity, or they
-%% %
-%% \item constrain only the current but not the past values on the input,
-%%   as in the case of symbolic automata \cite{symbTransd:POPL12,DAntoniKW16}.
-%% \end{inparaenum}
-
-We have implemented the boolean operations and emptiness checking
-semi-algorithms and carried out experiments with examples taken from
-array logics \cite{cav09}, timed automata
-\cite{henzinger:RealTimeSystems}, communication protocols \cite{abp}
-and hardware verification \cite{smrcka}.
-
-\paragraph{Related Work}
-Data languages and automata have been defined previously, in a
-classical nondeterministic setting. For instance, Kaminski and Francez
-\cite{Kaminski94} consider languages, over an infinite alphabet of
-data, recognized by automata with a finite number of registers, that
-store the input data and compare it using equality. Just as the timed
-languages recognized by timed automata \cite{AlurDill94}, these
-languages, called quasi-regular, are not closed under complement, but
-their emptiness is decidable. The impossibility of complementation
-here is caused by the use of hidden variables, which we do not
-allow. Emptiness is however undecidable in our case, mainly because
-counting (incrementing and comparing to a constant) data values is
-allowed, in many data theories.
-
-%% Other approaches to defining classes of data automata
-%% \cite{Bojanczyk:2011:TLD,BouyerPetitTherien03} are different from ours
-%% mainly because these automata are controlled by equivalence relations
-%% of finite index, whereas in our case, the transitions are defined by
-%% unrestricted formulae in the first-order theory of the data domain. As
-%% in \cite{Kaminski94,AlurDill94}, the emptiness problem is decidable
-%% for these classes.
-
-Another related model is that of predicate automata \cite{Farzan15},
-which recognize languages over integer data by labeling the words with
-conjunctions of uninterpreted predicates. We intend to explore further
-the connection with our model of alternating data automata, in order
-to apply our method to the verification of parallel programs.
-
-The model presented in this paper stems from the language inclusion
-problem considered in \cite{Tacas16}. There we provide a
-semi-algorithm for inclusion of data languages, based on an
-exponential determinization procedure and an abstraction refinement
-loop using lazy predicate abstraction \cite{HJMS02}. In this work we
-consider the full model of alternation and rely entirely on the
-ability of SMT solvers to produce interpolants in the combined theory
-of booleans and data. Since determinisation is not needed and
-complementation is possible in linear time, the bulk of the work is
-carried out by the decision procedure.
-
-The emptiness check for alternating data automata adapts similar
-semi-algorithms for nondeterministic infinite-state programs to the
-alternating model of computation. In particular, we considered the
-state-of-the-art \impact~ procedure \cite{mcmillan06} that is shown to
-outperform lazy predicate abstraction \cite{HJMS02} in the
-nondeterministic case, and generalized it to cope with
-alternation. More recent approaches for interpolant-based abstraction
-refinement target Horn systems \cite{McMillan14,Hoder12}, used to
-encode recursive and concurrent programs
-\cite{Grebenshchikov12}. However, the emptiness of alternating word
-automata cannot be directly encoded using Horn clauses, because all
-the branches of the computation synchronize on the same input, which
-cannot be encoded by a finite number of local (equality)
-constraints. We believe that the lazy annotation techniques for Horn
-clauses are suited for branching computations, which we intend to
-consider in a future tree automata setting.
-
-\section{Preliminaries}
-
-A \emph{signature} $\sig = (\ssorts{\sig},\sfuns{\sig})$ consists of a
-set $\ssorts{\sig}$ of \emph{sort symbols} and a set $\sfuns{\sig}$ of
-sorted \emph{function symbols}. To simplify the presentation, we
-assume w.l.o.g. that $\ssorts{\sig} = \set{\Data,\Bool}$\footnote{The
-  generalization to more than two sorts is without difficulty, but
-  would unnecessarily clutter the technical presentation.} and each
-function symbol $f \in \sfuns{\sig}$ has $\#(f) \geq 0$ arguments of
-sort $\Data$ and return value $\sigma(f) \in \ssorts{\sig}$. If
-$\#(f)=0$ then $f$ is a \emph{constant}. We consider the constants
-$\top$ and $\bot$ of sort $\Bool$.
-
-We consider an infinite countable set of \emph{variables} $\vars$,
-where each $x \in \vars$ has an associated sort $\sigma(x)$. A
-\emph{term} $t$ of sort $\sigma(t)=S$ is a variable $x \in \vars$
-where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
-$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
-\emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
-between two terms of sort $\Data$. A \emph{formula} is an
-existentially quantified combination of atoms using disjunction
-$\vee$, conjunction $\wedge$ and negation $\neg$ and we write $\phi
-\rightarrow \psi$ for $\neg\phi \vee \psi$.
-
-We denote by $\fv{\sigma}{\phi}$ the set of free variables of sort
-$\sigma$ in $\phi$ and write $\fv{}{\phi}$ for $\bigcup_{\sigma \in
-  \ssorts{\sig}} \fv{\sigma}{\phi}$. For a variable $x \in
-\fv{}{\phi}$ and a term $t$ such that $\sigma(t) = \sigma(x)$, let
-$\phi[t/x]$ be the result of replacing each occurrence of $x$ by
-$t$. For indexed sets $\vec{t}=\set{t_1,\ldots,t_n}$ and
-$\vec{x}=\set{x_1,\ldots,x_n}$, we write $\phi[\vec{t}/\vec{x}]$ for
-the formula obtained by simultaneously replacing $x_i$ with $t_i$ in
-$\phi$, for all $i\in[1,n]$. The size $\len{\phi}$ is the number of
-symbols occuring in $\phi$.
-
-An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
-\item the sort $\Data$ into a non-empty set $\Data^\I$, 
-%
-\item the sort $\Bool$ into the set $\booli = \set{\true,\false}$, where
-  $\top^\I = \true$, $\bot^\I = \false$, and
-%
-\item each function symbol $f$ into a total function $f^\I :
-  (\Data^\I)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
-  $\sigma(f)^I$ when $\#(f)=0$.
-\end{inparaenum}
-Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
-variable $x \in \vars$ into an element $\nu(x) \in \sigma(x)^\I$. For
-a term $t$, we denote by $t^\I_\nu$ the value obtained by replacing
-each function symbol $f$ by its interpretation $f^\I$ and each
-variable $x$ by its valuation $\nu(x)$. For a formula $\phi$, we write
-$\I,\nu \models \phi$ if the formula obtained by replacing each term
-$t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.
-
-A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
-there exists a valuation $\nu$ such that $\I,\nu \models \phi$, and
-\emph{valid} if $\I,\nu \models \phi$ for all valuations $\nu$.  The
-\emph{theory} $\theory(\sig,\I)$ is the set of valid formulae written
-in the signature $\sig$, with the interpretation $\I$. A \emph{decision
-procedure} for $\theory(\sig,\I)$ is an algorithm that takes a formula
-$\phi$ in the signature $\sig$ and returns yes iff $\phi \in
-\theory(\sig,\I)$.
-
-Given formulae $\varphi$ and $\psi$, we say that \emph{$\phi$ entails
-  $\psi$}, denoted $\phi \models^\I \psi$ iff $\I,\nu \models \varphi$
-implies $\I,\nu \models \psi$, for each valuation $\nu$, and $\phi
-\iffeq^\I \psi$ iff $\phi \models^\I \psi$ and $\psi \models^\I \phi$.
-We omit mentioning the interpretation $\I$ when it is clear from the
-context.
-
-\section{Alternating Data Automata}
-
-In the rest of this section we fix an interpretation $\I$ and a finite
-alphabet $\Sigma$ of \emph{input events}. Given a finite set $\vec{x}
-\subset \vars$ of variables of sort $\Data$, let $\vec{x} \mapsto
-\Data^\I$ be the set of valuations of the variables $\vec{x}$ and
-$\Sigma[\vec{x}] = \Sigma \times (\vec{x} \mapsto \Data^\I)$ be the
-set of \emph{data symbols}. A \emph{data word} (word in the sequel) is
-a finite sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data
-symbols, where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n :
-\vec{x} \rightarrow \Data^\I$ are valuations. We denote by
-$\varepsilon$ the empty sequence, by $\Sigma^*$ the set of finite
-sequences of input events and by $\Sigma[\vec{x}]^*$ the set of data
-words over $\vec{x}$.
-
-This definition generalizes the classical notion of words from a
-finite alphabet to the possibly infinite alphabet
-$\Sigma[\vec{x}]$. Clearly, when $\Data^\I$ is sufficiently large or
-infinite, we can map the elements of $\Sigma$ into designated elements
-of $\Data^\I$ and use a special variable to encode the input
-events. However, keeping $\Sigma$ explicit in the following simplifies
-several technical points below, without cluttering the presentation.
-
-Given sets of variables $\vec{b},\vec{x} \subset \vars$ of sort
-$\Bool$ and $\Data$, respectively, we denote by
-$\Form(\vec{b},\vec{x})$ the set of formulae $\phi$ such that
-$\fv{\Bool}{\phi} \subseteq \vec{b}$ and $\fv{\Data}{\phi} \subseteq
-\vec{x}$. By $\Form^+(\vec{b},\vec{x})$ we denote the set of formulae
-from $\Form(\vec{b},\vec{x})$ in which each boolean variable occurs
-under an even number of negations.
-
-An \emph{alternating data automaton} (ADA or automaton in the sequel)
-is a tuple $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$,
-where: \begin{compactitem}
-%
-\item $\vec{x} \subset \vars$ is a finite set of variables of sort
-  $\Data$,
-%
-\item $Q \subset \vars$ is a finite set of variables of sort $\Bool$
-  (\emph{states}),
-%
-\item $\iota \in \Form^+(Q,\emptyset)$ is the \emph{initial
-  configuration},
-%
-\item $F \subseteq Q$ is a set of \emph{final states}, and
-%
-\item $\Delta : Q \times \Sigma \rightarrow
-  \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
-    function},
-\end{compactitem}
-where $\overline{\vec{x}}=\{\overline{x} \mid x \in \vec{x}\}$. In
-each formula $\Delta(q,a)$ describing a transition rule, the variables
-$\overline{\vec{x}}$ track the previous and $\vec{x}$ the current
-values of the variables of $\A$. Observe that the initial values of
-the variables are left unconstrained, as the initial configuration
-does not contain free data variables. The size of $\A$ is defined as
-$\len{\A} = \len{\iota} + \sum_{(q,a) \in
-  Q\times\Sigma}\len{\Delta(q,a)}$.
-
-\begin{figure}[htb]
-\vspace*{-\baselineskip}
-\centerline{\includegraphics[scale=0.8]{figures/Example.pdf}}
-\vspace*{-\baselineskip}
-\caption{Alternating Data Automaton Example}
-\label{fig:ex}
-\end{figure}
-
-\paragraph{\em Example}
-Figure \ref{fig:ex} depicts an ADA with input alphabet $\Sigma =
-\set{a,b}$, variables $\vec{x} = \set{x,y}$, states $Q =
-\set{q_0,q_1,q_2,q_3,q_4}$, initial configuration $q_0$, final states
-$F=\set{q_3,q_4}$ and transitions:
-\[\begin{array}{rcl}
-\Delta(q_0,a) & \equiv & q_1 \wedge q_2 \wedge x\teq0 \wedge y\teq0 \\
-\Delta(q_1,a) & \equiv & q_1 \wedge q_3 \wedge x\teq\overline{y}+1 \wedge y\teq\overline{x}+1 \\
-\Delta(q_1,b) & \equiv & q_3 \wedge \overline{x} \geq \overline{y} \\
-\Delta(q_2,a) & \equiv & q_2 \wedge x > \overline{x} \wedge y > \overline{y} \\
-\Delta(q_2,b) & \equiv & q_4 \wedge \overline{x}>\overline{y}
-\end{array}\]
-The missing rules, such as $\Delta(q_0,b)$, are assumed to be
-$\bot$. Rules $\Delta(q_0,a)$ and $\Delta(q_1,a)$ are universal and
-there are no existential nondeterministic rules. Rules $\Delta(q_1,a)$
-and $\Delta(q_2,a)$ compare past ($\overline{x},\overline{y}$) with
-present ($x,y$) values, $\Delta(q_0,a)$ constrains the present and
-$\Delta(q_1,b)$, $\Delta(q_2,b)$ the past values, respectively. \qed
-
-Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, for any
-$k\geq0$, be a set of time-stamped variables. For an input event $a
-\in \Sigma$ and a formula $\phi$, we write $\Delta(\phi,a)$
-(respectively $\Delta^k(\phi,a)$) for the formula obtained from $\phi$
-by simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
-formula $\Delta(q,a)$ (respectively
-$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
-$k\geq0$). Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
-(a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
-formulae: \[\phi_0(Q) \Rightarrow \phi_1(Q,\vec{x}_0 \cup \vec{x}_1)
-\Rightarrow \ldots \Rightarrow \phi_n(Q,\vec{x}_0 \cup \ldots \cup
-\vec{x}_n)\] where $\phi_0 \equiv \iota$ and, for all $k\in[1,n]$, we
-have $\phi_k \equiv \Delta^k(\phi_{k-1},a_k)$. Next, we slightly abuse
-notation and write $\Delta(\iota,a_1,\ldots,a_n)$ for the formula
-$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ above. We say that $\A$
-\emph{accepts} $w$ iff $\I,\nu \models \Delta(\iota,a_1,\ldots,a_n)$,
-for some valuation $\nu$ that maps:\begin{inparaenum}[(1)]
-%
-\item each $x \in \vec{x}_k$ to $\nu_k(x)$, for all $k\in[1,n]$, 
-%
-\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$ and 
-%
-\item each $q \in \fv{\Bool}{\phi_n} \setminus F$ to $\bot$.
-\end{inparaenum}
-The language of $\A$ is the set $L(\A)$ of words from
-$\Sigma[\vec{x}]^*$ accepted by $\A$.
-
-\paragraph{\em Example} 
-The following sequence is a non-accepting run of the ADA from Figure
-\ref{fig:ex} on the word
-$(a,\tuple{0,0}),(a,\tuple{1,1}),(b,\tuple{2,1})$, where $\Data^\I =
-\zed$ and the function symbols have standard arithmetic
-interpretation:
-\[\begin{array}{l}
-q_0 \stackrel{(a,\tuple{0,0})}{\Longrightarrow} 
-q_1 \wedge q_2 \wedge x_1\teq0 \wedge y_1\teq0 \stackrel{(a,\tuple{1,1})}{\Longrightarrow} \\ 
-\underbrace{q_1 \wedge q_2 \wedge x_2\teq y_1+1 \wedge y_2\teq x_1+1}_{q_1} ~\wedge~
-\underbrace{q_2 \wedge x_2>x_1 \wedge y_2>y_1}_{q_2} \\ 
-\wedge~ x_1\teq0 \wedge y_1\teq0 \stackrel{(b,\tuple{2,1})}{\Longrightarrow} 
-\underbrace{q_3 \wedge x_2\geq y_2}_{q_1} \wedge \underbrace{q_4 \wedge x_2>y_2}_{q_2} \wedge x_2\teq y_1+1 \\ 
-\wedge~ y_2\teq x_1+1 \wedge \underbrace{q_4 \wedge x_2>y_2}_{q_2} \wedge x_2>x_1 \wedge y_2>y_1 \wedge x_1\teq0 \wedge y_1\teq0 ~\hfill\text{\qed}
-\end{array}\]
-
-In this paper we tackle the following problems: \begin{compactenum}
-\item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
-  with the same set of variables $\vec{x}$, do there exist automata
-  $\A_\cup$, $\A_\cap$ and $\overline{\A_1}$ such that $L(\A_\cup) =
-  \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and
-  $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$ ?
-%
-\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
-  \emptyset$ ?
-\end{compactenum}
-
-It is well known that other problems, such as \emph{universality}
-(given automaton $\A$ with variables $\vec{x}$, does $L(\A) =
-\Sigma[\vec{x}]^*$?) and \emph{inclusion} (given automata $\A_1$ and
-$\A_2$ with the same set of variables, does $L(\A_1) \subseteq
-L(\A_2)$?) can be reduced to the above problems. Observe furthermore
-that we do not consider cases in which the sets of variables in the
-two automata differ. An interesting problem in this case would be: given
-automata $\A_1$ and $\A_2$, with variables $\vec{x}_1$ and
-$\vec{x}_2$, respectively, such that $\vec{x}_1 \subseteq \vec{x}_2$,
-does $L(\A_1) \subseteq \proj{L(\A_2)}{\vec{x}_1}$, where
-$\proj{L(\A_2)}{\vec{x}_1}$ is the projection of the set of words
-$L(\A_2)$ onto the variables $\vec{x}_1$? This problem is considered
-as future work.
-
-\subsection{Boolean Closure}
-
-Given a set $Q$ of boolean variables and a set $\vec{x}$ of variables
-of sort $\Data$, for a formula $\phi \in \Form^+(Q,\vec{x})$, with no
-negated occurrences of the boolean variables, we define the formula
-$\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on the structure
-of $\phi$:
-\[\begin{array}{lclclcl}
-\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
-\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} \\ 
-\overline{\neg\phi} & \equiv & \neg \overline{\phi} \text{ if $\phi$ not atom} &&
-\overline{\phi} & \equiv & \phi \text{ if $\phi \in Q$} \\
-\overline{\phi} & \equiv & \neg\phi \text{ if $\phi \not\in Q$ atom}
-\end{array}\]
-We have $\len{\overline{\phi}} = \len{\phi}$, for every formula $\phi
-\in \Form^+(Q,\vec{x})$.
-
-In the following let $\A_i =
-\tuple{\vec{x},Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, where
-w.l.o.g. we assume that $Q_1 \cap Q_2 = \emptyset$. We define:
-\[\begin{array}{rcl}
-\A_\cup & = & \tuple{\vec{x},Q_1\cup Q_2,\iota_1 \vee \iota_2,F_1\cup F_2,\Delta_1\cup\Delta_2} \\ 
-\A_\cap & = & \tuple{\vec{x},Q_1\cup Q_2,\iota_1 \wedge \iota_2, F_1\cup F_2, \Delta_1\cup\Delta_2} \\ 
-\overline{\A_1} & = & \langle \vec{x},Q_1,\iota_1,Q_1\setminus F_1,\overline{\Delta_1} \rangle
-\end{array}\] 
-where $\overline{\Delta_1}(q,a) \equiv \overline{\Delta_1(q,a)}$, for
-all $q \in Q_1$ and $a \in \Sigma$. The following lemma shows the
-correctness of the above definitions:
-
-\begin{lemma}\label{lemma:closure}
-  Given automata $\A_i = \tuple{\vec{x},Q_i,\iota_i,F_i,\Delta_i}$,
-  for $i=1,2$, such that $Q_1 \cap Q_2 = \emptyset$, we have
-  $L(\A_\cup) = L(\A_1) \cup L(\A_2)$, $L(\A_\cap) = L(\A_1) \cap
-  L(\A_2)$ and $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus
-  L(\A_1)$.
-\end{lemma}
-
-It is easy to see that $\len{\A_\cup} = \len{\A_\cap} = \len{\A_1} +
-\len{\A_2}$ and $\len{\overline{\A}} = \len{\A}$, thus the automata
-for the boolean operations, including complementation, can be built in
-linear time. This matches the linear-time bounds for intersection and
-complementation of alternating automata over finite alphabets
-\cite{ChandraKozenStockmeyer81}.
-
-\section{Antichains and Interpolants for Emptiness}
-
-Unlike the boolean closure properties, showed to be effectively
-decidable (Lemma \ref{lemma:closure}), the emptiness problem is
-undecidable, even in very simple cases. For instance, if $\Data^\I$ is
-the set of positive integers, an ADA can simulate an Alternating
-Vector Addition System with States (AVASS) using only atoms $x \geq k$
-and $x = \overline{x} + k$, for $k \in \zed$, with the classical
-interpretation of the function symbols on integers. Since reachability
-of a control state is undecidable for AVASS \cite{LINCOLN92}, the
-emptiness problem is undecidable for ADA.
-
-Consequently, we give up on the guarantee for termination and build
-semi-algorithms that meet the requirements below: \begin{compactenum}[(i)]
-\item given an automaton $\A$, if $L(\A) \neq\emptyset$, the procedure
-  will terminate and return a word $w \in L(\A)$, and
-%
-\item if the procedure terminates without returning such a word, then
-  $L(\A) = \emptyset$.
-\end{compactenum}
-
-Let us fix an automaton $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ whose
-(finite) input event alphabet is $\Sigma$, for the rest of this
-section. Given a formula $\phi \in \Form^+(Q,\vec{x})$ and an input
-event $a \in \Sigma$, we define the \emph{post-image} function
-$\Post{\A}(\phi,a) \equiv \exists \overline{\vec{x}} ~.~
-\Delta(\phi[\overline{\vec{x}}/\vec{x}], a) \in \Form^+(Q,\vec{x})$,
-mapping each formula in $\Form^+(Q,\vec{x})$ to a formula defining the
-effect of reading the event $a$.
-
-We generalize the post-image function to finite sequences of input
-events, as follows:
-\[\begin{array}{l}
-\Post{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm} \Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a) \\
-\Accept{\A}(u) \equiv \Post{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q \rightarrow \bot)\text{, for any $u \in \Sigma^*$}
-\end{array}\]
-Then the emptiness problem for $\A$ becomes: does there exist $u \in
-\Sigma^*$ such that the formula $\Accept{\A}(u)$ is satisfiable?
-Observe that, since we ask a satisfiability query, the final states of
-$\A$ need not be constrained\footnote{ Since each state occurs
-  positively in $\Accept{\A}(u)$, this formula has a model iff it has
-  a model with every $q \in F$ set to true.}. A na\"ive semi-algorithm
-enumerates all finite sequences and checks the satisfiability of
-$\Accept{\A}(u)$ for each $u \in \Sigma^*$, using a decision procedure
-for the theory $\theory(\sig,\I)$.
-
-Since no boolean variable from $Q$ occurs under negation in $\phi$, it
-is easy to prove the following monotonicity property: given two
-formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
-then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
-\Sigma^*$. This suggest an improvement of the above semi-algorithm,
-that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
-$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
-\emph{antichain}\footnote{Given a partial order $(D,\preceq)$ an
-  antichain is a set $A \subseteq D$ such that $a \not\preceq b$ for
-  any $a,b \in A$.} w.r.t. the entailment partial order. This is
-because, for any $u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models
-\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable for some $w
-\in \Sigma^*$, then $\Post{\A}(\iota,uw) \models \Post{\A}(\iota,vw)$,
-thus $\Accept{\A}(vw)$ is satisfiable as well, and there is no need
-for $u$, since the non-emptiness of $\A$ can be proved using $v$
-alone. However, even with this optimization, the enumeration of
-sequences from $\Sigma^*$ diverges in many real cases, because
-infinite antichains exist in many interpretations, e.g.\ $q \wedge x
-\teq 0,~ q \wedge x \teq 1, \ldots$ for $\Data^\I = \nat$.
-
-A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
-\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
-boolean valuation $\beta : Q \rightarrow \booli$, every valuation $\nu
-: \vec{x} \mapsto \Data^\I$ of the data variables and every finite
-sequence $u \in \Sigma^*$ of input events, the following
-hold: \begin{compactenum}
-%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
-%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
-%
-\item\label{it1:inv} $\I,\beta \cup \nu \models \Post{\A}(\iota,u)
-  \Rightarrow \nu \in \Inv(\beta)$, and
-%
-\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \I,\beta \cup
-  \nu \not\models \Accept{\A}(u)$.
-\end{compactenum}
-If $\Inv$ satisfies only the first point above, we call it an
-\emph{invariant}. Intuitively, a safety invariant maps every boolean
-valuation into a set of data valuations, that contains the initial
-configuration $\iota \equiv \Post{\A}(\iota,\varepsilon)$, whose data
-variables are unconstrained, over-approximates the set of reachable
-valuations (point \ref{it1:inv}) and excludes the valuations satisfying the
-acceptance condition (point \ref{it2:inv}).  A formula $\phi(Q,\vec{x})$
-is said to \emph{define} $\Inv$ iff for all $\beta: Q \rightarrow \booli$ and
-$\nu : \vec{x} \rightarrow \Data^\I$, we have $\I,\beta\cup\nu \models
-\phi$ iff $\nu \in \Inv(\beta)$.
-
-\begin{lemma}\label{lemma:safety-invariant}
-  For any automaton $\A$, we have $L(\A) = \emptyset$ if and only if
-  $\A$ has a safety invariant.
-\end{lemma}
-
-Turning back to the issue of divergence of language emptiness
-semi-algorithms in the case $L(\A) = \emptyset$, we can observe that
-an enumeration of input sequences $u_1,u_2,\ldots \in \Sigma^*$ can
-stop at step $k$ as soon as $\bigvee_{i=1}^k \Post{\A}(\iota,u_i)$
-defines a safety invariant for $\A$. Although this condition can be
-effectively checked using a decision procedure for the theory
-$\theory(\sig,\I)$, there is no guarantee that this check will ever
-succeed.
-
-The solution we adopt in the sequel is abstraction to ensure the
-termination of invariant computations. However, it is worth pointing
-out from the start that abstraction alone will only allow us to build
-invariants that are not necessarily safety invariants. To meet the
-latter condition, we resort to counterexample guided abstraction
-refinement (CEGAR).
-
-Formally, we fix a set of formulae $\preds \subseteq
-\Form(Q,\vec{x})$, such that $\bot \in \preds$ and refer to these
-formulae as \emph{predicates}. Given a formula $\phi$, we denote by
-$\abs{\phi} \equiv \bigwedge \set{\pi \in \preds \mid \phi \models
-  \pi}$ the abstraction of $\phi$ w.r.t. the predicates in
-$\preds$. The abstract versions of the post-image and acceptance
-condition are defined as follows:
-\[\begin{array}{l}
-\AbsPost{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm}
-\AbsPost{\A}(\phi,ua) \equiv \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} \\
-\AbsAccept{\A}(u) \equiv \AbsPost{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} 
-(q \rightarrow \bot)\text{, for any $u \in \Sigma^*$}
-\end{array}\]
-
-\begin{lemma}\label{lemma:abstract-invariant}
-  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
-  $k>0$ such that $\bigvee_{i=0}^k \AbsPost{\A}(\iota,\mu(i))$ defines
-  an invariant $\abs{\Inv}$ for $\A$. 
-\end{lemma}
-
-%% \paragraph{\em Remark}
-%% While computing the concrete post-image $\Post{\A}$ relies on an
-%% expensive quantifier elimination, it is common belief that this should
-%% hold also for $\AbsPost{\A}$. However, chosing the conjoined
-%% predicates that form $\AbsPost{\A}$ requires only a quantifier-free
-%% satisfiability query: $\Post{\A}(\phi,a) \models \pi$ iff $\exists
-%% \overline{\vec{x}} ~.~ \Delta(\phi[\overline{\vec{x}}/\vec{x}],a)
-%% \models \pi$ iff $\Delta(\phi[\overline{\vec{x}}/\vec{x}],a) \wedge
-%% \neg\pi$ is unsatisfiable. \qed
-
-We are left with fulfilling point (\ref{it2:inv}) from the definition
-of a safety invariant. To this end, suppose that, for a given set
-$\preds$ of predicates, the invariant $\abs{\Inv}$, defined by the
-previous lemma, meets point (\ref{it1:inv}) but not point
-(\ref{it2:inv}), where $\Post{\A}$ and $\Accept{\A}$ replace
-$\AbsPost{\A}$ and $\AbsAccept{\A}$, respectively. In other words,
-there exists a finite sequence $u \in \Sigma^*$ such that $\nu \in
-\abs{\Inv}(\beta)$ and $\I,\beta\cup\nu \models \AbsAccept{\A}(u)$,
-for some boolean $\beta : Q \rightarrow \booli$ and data $\nu :
-\vec{x} \rightarrow \Data^\I$ valuations. Such a $u\in\Sigma^*$ is
-called a \emph{counterexample}.
-
-Once a counterexample $u$ is discovered, there are two
-possibilities. Either\begin{inparaenum}[(i)]
-\item $\Accept{\A}(u)$ is satisfiable, in which case $u$ is
-  \emph{feasible} and $L(\A) \neq \emptyset$, or
-%
-\item $\Accept{\A}(u)$ is unsatisfiable, in which case $u$ is
-  \emph{spurious}.
-\end{inparaenum}
-In the first case, our semi-algorithm stops and returns a witness for
-non-emptiness, obtained from the satisfying valuation of
-$\Accept{\A}(u)$ and in the second case, we must strenghten the
-invariant by excluding from $\abs{\Inv}$ all pairs $(\beta,\nu)$ such
-that $\I,\beta\cup\nu \models \AbsAccept{\A}(u)$. This strenghtening
-is carried out by adding to $\preds$ several predicates that are
-sufficient to exclude the spurious counterexample.
-
-In general, given an unsatisfiable conjunction $\Phi \equiv
-\phi_1(X_0,X_1) \wedge \phi_2(X_1,X_2) \wedge \ldots \wedge
-\phi_n(X_{n-1},X_n)$ of time-stamped variables $X_i = \set{x_i \mid x
-  \in X}$ of any sort, a solution of the \emph{interpolation problem}
-$\Phi$, simply called an \emph{interpolant}, is a tuple $\tuple{I_0(X),
-  I_1(X), \ldots, I_n(X)}$ such that:\begin{inparaenum}[(i)]
-\item $I_0 \equiv \top$,  
-%
-\item $I_{i-1}[X_{i-1}/X] \wedge \phi_i(X_{i-1},X_i) \models
-  I_i[X_i/X]$, for all $i \in [1,n]$, and
-%
-\item $I_n \equiv \bot$. 
-\end{inparaenum}
-In the following, we shall assume the existence of an interpolating
-decision procedure for $\theory(\sig,\I)$.
-
-%% \paragraph{\em Remark} 
-%% We deviate w.l.o.g. from the standard notion of \emph{generalized
-%%   interpolants} \cite{mcmillan06}, which requires $\fv{}{I_i}
-%% \subseteq \fv{}{\phi_i} \cap \fv{}{\phi_{i+1}}$, for each $i \in
-%%           [1,n-1]$, and assume further that the variables have been
-%%           renamed by removing the time stamps. \qed
-
-A classical method for abstraction refinement is to add the elements
-of the interpolant obtained from a proof of spuriousness to the set of
-predicates. This guarantees progress, meaning that the particular
-spurious counterexample, from which the interpolant was generated,
-will never be revisited in the future. Though not always, in many
-practical test cases this progress property eventually yields a safety
-invariant.
-
-Given a non-empty spurious counterexample $u = a_1\ldots a_n$, where
-$n>0$, we consider the following interpolation problem: 
-\begin{eqnarray}\label{eq:interpolation-problem}
-\Theta(u) & \equiv & \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0
-\cup \vec{x}_1) \wedge \ldots \\ 
-&& \wedge~ \theta_n(Q_{n-1} \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n) \nonumber
-\end{eqnarray}
-where $Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped
-sets of boolean variables corresponding to the set $Q$ of states of
-$\A$. The first conjunct $\theta_0(Q_0) \equiv \iota[Q_0/Q]$ is the
-initial configuration of $\A$, with every $q \in \fv{\Bool}{\iota}$
-replaced by $q_0$. The definition of $\theta_k$, for all $k\in[1,n]$,
-uses \emph{replacement sets} $R_\ell \subseteq Q_\ell$, $\ell\in
-[0,n]$, which are defined inductively below: \begin{compactitem}
-\item $R_0 = \fv{\Bool}{\theta_0}$,
-%
-\item $\theta_\ell \equiv \bigwedge_{q_{\ell-1}\in R_{\ell-1}}
-  (q_{\ell-1} \rightarrow
-  \Delta(q,a_\ell)[Q_\ell/Q,\vec{x}_{\ell-1}/\overline{\vec{x}},\vec{x}_\ell/\vec{x}])$
-  and $R_\ell = \fv{\Bool}{\theta_\ell} \cap Q_\ell$, for each
-  $\ell\in[1,n]$.
-%
-\item $\theta_{n+1}(Q_n) \equiv \bigwedge_{q \in Q \setminus F} (q_n
-  \rightarrow \bot)$.
-\end{compactitem}
-The intuition is that $R_0,\ldots,R_n$ are the sets of states
-replaced, $\theta_0, \ldots, \theta_n$ are the sets of transition
-rules fired on the run of $\A$ over $u$ and $\theta_{n+1}$ is the
-acceptance condition, which forces the last remaining non-final states
-to be false.
-
-Moreover, we require that an interpolant
-$\tuple{\top,I_0,\ldots,I_n,\bot}$ for the interpolation problem
-$\Theta(u)$ does not have negative occurrences of states, i.e.\ $I_i
-\in \Form^+(Q,\vec{x})$, for all $i \in [0,n]$. Such an interpolant
-can always be built, as showed below:
-
-\begin{proposition}\label{prop:positive-interpol}
-  If $\Theta(a_1 \ldots a_n)$ is unsatisfiable then one can build an
-  interpolant $\tuple{\top,I_0,\ldots,I_n,\bot}$ for $\Theta(a_1\ldots
-  a_n)$, such that $I_i \in \Form^+(Q,\vec{x})$, for all $i \in
-  [0,n]$.
-\end{proposition}
-
-We recall that a run of $\A$ over $u$ is a sequence: \[\phi_0(Q)
-\Rightarrow \phi_1(Q,\vec{x}_0\cup\vec{x}_1) \Rightarrow \ldots
-\Rightarrow \phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)\] where
-$\phi_0$ is the initial configuration $\iota$ and for each $k>0$,
-$\phi_k$ is obtained from $\phi_{k-1}$ by replacing each state $q \in
-\fv{\Bool}{\phi_{k-1}}$ by the formula
-$\Delta(q,a_k)[\vec{x}_{k-1}/\overline{\vec{x}},\vec{x}_k/\vec{x}]$,
-given by the transition function of $\A$. Observe that, because the
-states are replaced with transition formulae when moving one step in a
-run, these formulae lose track of the control history and are not
-suitable for producing interpolants that relate states and data.
-
-The main idea behind the above definition of the interpolation problem
-is that we would like to obtain an interpolant $\tuple{\top,I_0(Q),
-  I_1(Q,\vec{x}), \ldots, I_{n}(Q,\vec{x}),\bot}$ whose formulae
-\emph{combine states with the data constraints that must hold
-  locally}, whenever the control reaches a certain boolean
-configuration. This association of states with data valuations is
-tantamount to defining efficient semi-algorithms, based on lazy
-abstraction \cite{HJMS02}. Furthermore, the abstraction defined by the
-interpolants generated in this way can also \emph{over-approximate the
-  control structure} of an automaton, in addition to the sets of data
-values encountered throughout its runs.
-
-The correctness of this interpolation-based abstraction refinement
-setup is captured by the progress property below, which guarantees
-that adding the formulae of an interpolant for $\Theta(u)$ to the set
-$\preds$ of predicates suffices to exclude the spurious counterexample
-$u$ from future searches.
-
-\begin{lemma}\label{lemma:progress}
-  For any sequence $u = a_1\ldots a_n \in \Sigma^*$, if
-  $\Accept{\A}(u)$ is unsatisfiable, the following
-  hold: \begin{compactenum}
-  \item\label{it1:progress} $\Theta(u)$ is unsatisfiable, and
-    %
-  \item\label{it2:progress} if $\tuple{\top,I_0,\dots,I_{n},\bot}$ is
-    an interpolant for $\Theta(u)$ such that $\set{I_i \mid i
-      \in[0,n]} \subseteq \preds$ then $\AbsAccept{\A}(u)$ is
-    unsatisfiable.
-  \end{compactenum}
-\end{lemma}
-
-\section{Lazy Predicate Abstraction for ADA Emptiness}
-
-We have now all the ingredients to describe the first emptiness
-checking semi-algorithm for alternating data
-automata. Algorithm\footnote{Though termination is not guaranteed, we
-  call it algorithm for conciseness.} \ref{alg:predabs} builds an
-\emph{abstract reachability tree} (ART) whose nodes are labeled with
-formulae over-approximating the concrete sets of configurations, and a
-covering relation between nodes in order to ensure that the set of
-formulae labeling the nodes in the ART forms an antichain. Any
-spurious counterexample is eliminated by computing an interpolant and
-adding its formulae to the set of predicates (cf. Lemma \ref{lemma:progress}).
-
-\begin{algorithm}[t!]
-{\scriptsize\begin{algorithmic}[0]
-\State {\bf input}: an ADA $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$
-over the alphabet $\Sigma$ of input events
-
-\State {\bf output}: $\true$ if $L(\A)=\emptyset$ and a data word $w
-\in L(\A)$ otherwise
-\end{algorithmic}}
-
-{\scriptsize\begin{algorithmic}[1]
-
-  \State let $\Art = \tuple{N,E,\rootNode,\Lambda,\lhd}$ be an ART
-
-  \State initially $N = E = \lhd = \emptyset$, $\Lambda =
-  \set{(\rootNode,\iota)}$, $\preds = \set{\bot}$, $\worklist =
-  \tuple{\rootNode}$,
-
-  \While {$\worklist \neq \emptyset$}
-  \label{ln:while}
-
-  \State dequeue $n$ from $\worklist$
-
-  \State $N \leftarrow N \cup \set{n}$
-
-  \State let $\lambda(n) = a_1\ldots a_k$ be the label of the path from
-  $\rootNode$ to $n$
-
-  \If{$\AbsPost{\A}(\lambda(n))$ is satisfiable} \Comment{counterexample candidate}
-
-  \If{$\Accept{\A}(u)$ is satisfiable} \Comment{feasible counterexample}
-  \label{ln:feasible}
-
-  \State get model $(\beta,\nu_1,\ldots,\nu_k)$ of
-  $\Accept{\A}(\lambda(n))$
-
-  \State {\bf return} $w = (a_1,\nu_1) \ldots (a_k,\nu_k)$
-  \label{ln:cex}
-  \Comment{$w \in L(\A)$ by construction}
-
-  \Else \Comment{spurious counterexample}
-
-  \State $p \leftarrow \Call{FindPivot}{\lambda(n),\Art}$
-  \label{ln:pivot}
-
-  \State $v \leftarrow \Call{LeastInfeasibleSuffix}{\lambda(n),\Art}$
-
-  \State $\preds \leftarrow \preds \cup \set{I_0,\ldots,I_\ell}$,
-  where $\tuple{\top,I_0,\ldots,I_\ell,\bot}$ is an interpolant for
-  $\pivot{v}$
-
-  \State let $\SubArt = \tuple{N',E',p,\Lambda',\lhd'}$ be the subtree
-  of $\Art$ rooted at $p$
-
-  \For{$(m,q) \in \lhd$ such that $q \in N'$}
-
-  \State remove $m$ from $N$ and enqueue $m$ into $\worklist$
-
-  \State remove $\SubArt$ from $\Art$
-
-  \State enqueue $p$ into $\worklist$
-  \label{ln:refine}
-  \Comment{recompute the subtree rooted at $p$}
-
-  \EndFor
-
-  \EndIf
-
-  \Else \label{ln:else}
-
-  \For{$a \in \Sigma$}\label{ln:forall-events}
-  \Comment{expand $n$}
-
-  \State $\phi \leftarrow \AbsPost{\A}(\Lambda(n),a)$
-
-  \If{exist $m \in N$ such that $\phi \models \Lambda(m)$} \label{ln:covered}
-
-  \State $\lhd \leftarrow \lhd \cup \set{(n,m)}$ \Comment{$m$ covers $n$}
-  \label{ln:direct-cover}
-
-  \Else
-
-  \State let $s$ be a fresh node
-
-  \State $E \leftarrow E \cup \set{(n,a,s)}$
-
-  \State $\Lambda \leftarrow \Lambda \cup \set{(s,\phi)}$
-
-  \State $R \leftarrow \set{m \in \worklist \mid \Lambda(m) \models
-    \phi}$ \Comment{worklist nodes covered by $s$}
-
-  \For{$r \in R$}
-
-  \For{$m \in N$ such that $(m,b,r) \in E$, $b \in \Sigma$}
-
-  \State $\lhd \leftarrow \lhd \cup \set{(m,s)}$ 
-  \label{ln:child-cover}
-  \Comment{redirect covered children from $R$ into $s$}
-
-  \EndFor
-
-  \For{$(m,r) \in \lhd$}
-
-  \State $\lhd \leftarrow \lhd \cup \set{(m,s)}$ 
-  \label{ln:indirect-cover}
-  \Comment{redirect covered nodes from $R$ into $s$}
-
-  \EndFor
-
-  \EndFor
-
-  \State remove $R$ from $\Art$
-
-  \State enqueue $s$ into $\worklist$ 
-  \label{ln:expand}
-
-  \EndIf
-
-  \EndFor
-
-  \EndIf
-
-  \EndWhile
-
-  \State {\bf return} $\true$
-  \label{ln:true}
-\end{algorithmic}}
-\caption{Lazy Predicate Abstraction for ADA Emptiness}
-\label{alg:predabs}
-\end{algorithm}
-
-Formally, an ART is tuple $\Art =
-\tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$, where: \begin{compactitem}
-\item $N$ is a set of nodes, 
-%
-\item $E \subseteq N \times \Sigma \times
-N$ is a set of edges, 
-%
-\item $\rootNode \in N$ is the root of the directed
-tree $(N,E)$, 
-%
-\item $\Lambda : N \rightarrow \Form(Q,\vec{x})$ is a labeling of the
-  nodes with formulae, such that $\Lambda(\rootNode) = \iota$,
-%
-\item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
-  sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$,
-%
-\item $T : E \rightarrow \bigcup_{i=0}^\infty
-  \Form^+(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is a labeling of edges
-  with time-stamped formulae, and
-%
-\item $\lhd \subseteq N \times N$ is a set of
-\emph{covering edges}. 
-\end{compactitem}
-%% For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is $(n,a,m)$.
-%% The replacement sets and time-stamped edge labels are defined below,
-%% for each edge $e = (n,a,m) \in E$, if $n$ lies at depth\footnote{We
-%%   consider that the root node has zero depth.} $k$ in the tree
-%% $(N,E)$ then we have:
-%% \[\begin{array}{rcl}
-%% T(e) & = & \bigwedge_{q \in R(n)} (q_k \rightarrow \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]) \\
-%% R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)}
-%% \end{array}\]
-
-Each node $n \in N$ corresponds to a unique path from the root to $n$,
-labeled by a sequence $\lambda(n) \in \Sigma^*$ of input events. The
-\emph{least infeasible suffix} of $\lambda(n)$ is the smallest
-sequence $v = a_1 \ldots a_k$, such that $\lambda(n) = wv$, for some
-$w \in \Sigma^*$ and the following formula is unsatisfiable:
-\begin{equation}\label{eq:pivot}
-  \pivot{v} \equiv \Lambda(p)[Q_0/Q] \wedge
-  \theta_1(Q_0 \cup Q_1,\vec{x}_0 \cup \vec{x}_1) \wedge \ldots \wedge
-  \theta_{k+1}(Q_k)
-\end{equation}
-where $\theta_1,\ldots,\theta_{k+1}$ are defined as in
-(\ref{eq:interpolation-problem}) and $\theta_0 \equiv
-\Lambda(p)[Q_0/Q]$. The \emph{pivot} of $n$ is the node $p$
-corresponding to the start of the least infeasible suffix.  We assume
-the existence of two functions $\Call{FindPivot}{u,\Art}$ and
-$\Call{LeastInfeasibleSuffix}{u,\Art}$ that return the pivot and least
-infeasbile suffix of a sequence $u \in \Sigma^*$ in an ART $\Art$,
-without detailing their implementation.
-
-With these considerations, Algorithm \ref{alg:predabs} uses a worklist
-iteration to build an ART. We keep newly expanded nodes of $\Art$ in a
-queue $\worklist$, thus implementing a breadth-first exploration
-strategy, which guarantees that the shortest counterexamples are
-explored first. When the search encounters a counterexample candidate
-$u$, it is checked for spuriousness. If the counterexample is
-feasible, the procedure returns a data word $w \in L(\A)$, which
-interleaves the input events of $u$ with the data valuations from the
-model of $\Accept{\A}(u)$ (since $u$ is feasible, clearly
-$\Accept{\A}(u)$ is satisfiable). Otherwise, if $u$ is spurious, we
-compute its pivot $p$ (line \ref{ln:pivot}), add the interpolants for
-the least unfeasible suffix of $u$ to the set of predicates $\preds$,
-remove and recompute the subtree of $\Art$ rooted at $p$.
-
-Termination of Algorithm \ref{alg:predabs} depends on the ability of a
-given interpolating decision procedure for the combined boolean and
-data theory $\theory(\sig,\I)$ to provide interpolants that yield a
-safety invariant, whenever $L(\A) = \emptyset$. In this case, we use
-the covering relation $\lhd$ to ensure that, when a newly generated
-node is covered by a node already in $N$, it is not added to the
-worklist, thus cutting the current branch of the search.
-
-Formally, for any two nodes $n,m \in N$, we have $n \lhd m$ iff
-$\AbsPost{\A}(\Lambda(n),a) \models \Lambda(m)$ for some $a \in
-\Sigma$, in other words, if $n$ has a successor whose label entails
-the label of $m$.
-
-\begin{figure}[t!]
-\begin{tabular}{cc}
-\includegraphics[scale=0.5]{figures/PA1.pdf} & \includegraphics[scale=0.5]{figures/PA2.pdf} \\[-2mm]
-\tiny{(a)} & \tiny{(b)} \\
-\\
-\includegraphics[scale=0.5]{figures/PA3.pdf} & \includegraphics[scale=0.5]{figures/PA4.pdf} \\[-2mm]
-\tiny{(c)} & \tiny{(d)}
-\end{tabular}
-\vspace*{-\baselineskip}
-\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex} by Algorithm \ref{alg:predabs}}
-\label{fig:pa}
-\end{figure}
-
-\paragraph{\em Example}
-Consider the automaton given in Figure \ref{fig:ex}. First, Algorithm
-\ref{alg:predabs} fires the sequence $a$, and since there are no other
-formulae than $\bot$ in $\preds$, the successor of $\iota \equiv q_0$
-is $\top$, in Figure \ref{fig:pa} (a). The spuriousness check for $a$
-yields the root of the ART as pivot and the interpolant
-$\tuple{q_0,q_1}$, which is added to the set $\preds$. Then the $\top$
-node is removed and the next time $a$ is fired, it creates a node
-labeled $q_1$. The second sequence $aa$ creates a successor node
-$q_1$, which is covered by the first, depicted with a dashed arrow, in
-Figure \ref{fig:pa} (b). The third sequence is $ab$, which results in
-a new uncovered node $\top$ and triggers a spuriousness check. The new
-predicate obtained from this check is $x\leq0\wedge q_2 \wedge y\geq0$
-and the pivot is again the root. Then the entire ART is rebuilt with
-the new predicates and the fourth sequence $aab$ yields an uncovered
-node $\top$, in Figure \ref{fig:pa} (c). The new pivot is the endpoint
-of $a$ and the newly added predicates are $q_1\wedge q_2$ and $y>x-1
-\wedge q_2$. Finally, the ART is rebuilt from the pivot node and
-finally all nodes are covered, thus proving the emptiness of the
-automaton, in Figure \ref{fig:pa} (d). \qed
-
-The correctness of Algorithm \ref{alg:predabs} is proved below:
-
-\begin{theorem}\label{thm:predabs}
-  Given an automaton $\A$, such that $L(\A) \neq \emptyset$, Algorithm
-  \ref{alg:predabs} terminates and returns a word $w \in L(\A)$. If
-  Algorithm \ref{alg:predabs} terminates reporting $\true$, then
-  $L(\A) = \emptyset$.
-\end{theorem}
-
-\section{Checking ADA Emptiness with \impact}
-
-As pointed out by a number of authors, the bottleneck of predicate
-abstraction is the high cost of reconstructing parts of the ART,
-subsequent to the refinement of the set of predicates. The main idea
-of the \impact~ procedure \cite{mcmillan06} is that this can be
-avoided and the refinement (strenghtening of the node labels of the
-ART) can be performed in-place. This refinement step requires an
-update of the covering relation, because a node that used to cover
-another node might not cover it after the strenghtening of its label.
-
-We consider a total alphabetical order $\prec$ on $\Sigma$ and lift it
-to the total lexicographical order $\prec^*$ on $\Sigma^*$. A node $n
-\in N$ is \emph{covered} if $(n,p) \in \lhd$ or it has an ancestor $m$
-such that $(m,p) \in \lhd$, for some $p \in N$. A node $n$ is
-\emph{closed} if it is covered, or $\Lambda(n) \not\models \Lambda(m)$
-for all $m \in N$ such that $\lambda(m) \prec^* \lambda(n)$. Observe
-that we use the coverage relation $\lhd$ here with a different meaning
-than in Algorithm \ref{alg:predabs}.
-
-\begin{algorithm}[t!]
-{\scriptsize\begin{algorithmic}[0]
-  \State {\bf input}: an ADA $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$
-  over the alphabet $\Sigma$ of input events
-
-  \State {\bf output}: $\true$ if $L(\A)=\emptyset$ and a data word $w
-  \in L(\A)$ otherwise
-\end{algorithmic}}
-
-{\scriptsize\begin{algorithmic}[1]
-
-  \State let $\Art = \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$ be an ART
-
-  \State initially $N = E = T = \lhd = \emptyset$, $\Lambda =
-  \set{(\rootNode,\iota)}$, $R = \fv{\Bool}{\iota[Q_0/Q]}$,
-  $\worklist=\set{\rootNode}$
-
-  \While{$\worklist \neq \emptyset$}
-  \label{ln:impact-while}
-
-  \State dequeue $n$ from $\worklist$ 
-  \label{ln:impact-dequeue}
-
-  \State $N \leftarrow N \cup \set{n}$
-
-  \State let $(\rootNode,a_1,n_1),(n_1,a_2,n_2), \ldots,
-  (n_{k-1},a_k,n)$ be the path from $\rootNode$ to $n$
-
-  \If{$\Accept{\A}(a_1\ldots a_k)$ is satisfiable} 
-  \Comment{counterexample is feasible}
-
-  \State get model $(\beta,\nu_1,\ldots,\nu_k)$ of
-  $\Accept{\A}(\lambda(n))$
-
-  \State {\bf return} $w = (a_1,\nu_1) \ldots (a_k,\nu_k)$
-  \label{ln:impact-nonempty}
-  \Comment{$w \in L(\A)$ by construction}
-
-  \Else \Comment{spurious counterexample}
-
-  \State let $\tuple{\top,I_0,\ldots,I_k,\bot}$ be an interpolant for
-  $\Theta(a_1 \ldots a_k)$
-  \label{ln:refine-begin}
-
-  \State $b \leftarrow \false$
-
-  \For{$i=0,\ldots,k$}
-
-  \If{$\Lambda(n_i) \not\models I_i$}
-
-  \State $\lhd \leftarrow \lhd \setminus \set{(m,n_i) \in \lhd \mid m \in N}$
-
-  \State $\Lambda(n_i) \leftarrow \Lambda(n_i) \wedge I_i$
-  \Comment{strenghten the label of $n_i$}
-
-  \If{$\neg b$}
-
-  \State $b \leftarrow \Call{Close}{n_i}$
-  \label{ln:refine-end}
-
-  \EndIf % $\neg b$
-
-  \EndIf % $\Lambda(n_i) \not\models I_i$
-
-  \EndFor
-
-  \EndIf % $\mu(n)$ is satisfiable
-
-  \If{$n$ is not covered}
-
-  \For{$a \in \Sigma$} 
-  \Comment{expand $n$}
-  \label{ln:expand-begin}
-
-  \State let $s$ be a fresh node and $e = (n,a,s)$ be a new edge
-
-  \State $E \leftarrow E \cup \set{e}$
-  \label{ln:edge-insert}
-
-  \State $\Lambda \leftarrow \Lambda \cup \set{(s,\top)}$
-
-  \State $T \leftarrow T \cup \set{(e, \theta_k)}$
-
-  \State $R \leftarrow R \cup \{(s,\bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)})\}$
-
-  \State enqueue $s$ into $\worklist$
-  \label{ln:expand-end}
-
-  \EndFor
-
-  \EndIf % $n$ is not covered
-
-  \EndWhile
-
-  \State {\bf return} $\true$
-\end{algorithmic}
-
-\begin{algorithmic}[1]
-\Function{Close}{$x$} {\bf returns} $\Bool$
-
-\For{$y \in N$ such that $\lambda(y) \prec^* \lambda(x)$}
-
-\If{$\Lambda(x) \models \Lambda(y)$}
-
-\State $\lhd \leftarrow \left(\lhd \setminus \set{ (p,q) \in \lhd \mid
-  \text{$q$ is $x$ or a successor of $x$}}\right) \cup \set{(x,y)}$
-\label{ln:close-uncover}
-
-\State {\bf return} $\true$
-
-\EndIf
-
-\EndFor
-
-\State {\bf return} $\false$
-
-\EndFunction
-\end{algorithmic}}
-\caption{\impact~ for ADA Emptiness}
-\label{alg:impact}
-\end{algorithm}
-
-The execution of Algorithm \ref{alg:impact} consists of three
-phases\footnote{These phases correspond to the \textsc{Close},
-  \textsc{Refine} and \textsc{Expand} procedures from the original
-  description of \impact~\cite{mcmillan06}.}: \emph{close},
-\emph{refine} and \emph{expand}. Let $n$ be a node removed from the
-worklist at line \ref{ln:impact-dequeue}. If $\Accept{\A}(\lambda(n))$
-is satisfiable, the counterexample $\lambda(n)$ is feasible, in which
-case a model of $\Accept{\A}(\lambda(n))$ is obtained and a word $w
-\in L(\A)$ is returned. Otherwise, $\lambda(n)$ is a spurious
-counterexample and the procedure enters the refinement phase (lines
-\ref{ln:refine-begin}-\ref{ln:refine-end}). The interpolant for
-$\Theta(\lambda(n))$ (cf. formula \ref{eq:interpolation-problem}) is
-used to strenghten the labels of all the ancestors of $n$, by
-conjoining the formulae of the interpolant to the existing labels.
-
-In this process, the nodes on the path between $\rootNode$ and $n$,
-including $n$, might become eligible for coverage, therefore we
-attempt to close each ancestor of $n$ that is impacted by the
-refinement (line \ref{ln:refine-end}). Observe that, in this case the
-call to $\Call{Close}{}$ must uncover each node which is covered by a
-successor of $n$ (line \ref{ln:close-uncover} of the $\Call{Close}{}$
-function). This is required because, due to the over-approximation of
-the sets of reachable configurations, the covering relation is not
-transitive, as explained in \cite{mcmillan06}. If $\Call{Close}{}$
-adds a covering edge $(n_i,m)$ to $\lhd$, it does not have to be
-called for the successors of $n_i$ on this path, which is handled via
-the boolean flag $b$.
-
-Finally, if $n$ is still uncovered (it has not been previously covered
-during the refinement phase) we expand $n$ (lines
-\ref{ln:expand-begin}-\ref{ln:expand-end}) by creating a new node for
-each successor $s$ via the input event $a \in \Sigma$ and inserting it
-into the worklist.
-
-\begin{figure}[htb]
-\vspace*{-\baselineskip}
-\begin{tabular}{c}
-\includegraphics[scale=0.65]{figures/IM1.pdf} \\[-2mm]
-\tiny{(a)} \\
-\includegraphics[scale=0.65]{figures/IM2.pdf} \\[-2mm]
-\tiny{(b)} \\
-\includegraphics[scale=0.65]{figures/IM3.pdf} \\[-2mm]
-\tiny{(c)}
-\end{tabular}
-\vspace*{-0.5\baselineskip}
-\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex}
-  by Algorithm \ref{alg:impact} (1/2)}
-\label{fig:im1}
-\end{figure}
-
-\begin{figure}[htb]
-\centerline{\includegraphics[scale=0.65]{figures/IM4.pdf}}
-\vspace*{-\baselineskip}
-\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex}
-  by Algorithm \ref{alg:impact} (2/2)}
-\label{fig:im2}
-\end{figure}
-
-\paragraph{\em Example}
-We show the execution of Algorithm \ref{alg:impact} on the automaton
-from Figure \ref{fig:ex}.  Initially, the procedure fires the sequence
-$a$, whose endpoint is labeled with $\top$, in Figure \ref{fig:im1}
-(a). Since this node is uncovered, we check the spuriousness of the
-counterexample $a$ and refine the label of the node to $q_1$. Since
-the node is still uncovered, two successors, labeled with $\top$ are
-computed, corresponding to the sequences $aa$ and $ab$, in Figure
-\ref{fig:im1} (b). The spuriousness check for $aa$ yields the
-interpolant $\tuple{q_0,x\leq0\wedge q_2 \wedge y\geq0}$ which
-strenghtens the label of the endpoint of $a$ from $q_1$ to $q_1\wedge
-x\leq0\wedge q_2\wedge y\geq0$. The sequence $ab$ is also found to be
-spurious, which changes the label of its endpoint from $\top$ to
-$\bot$, and also covers it (depicted with a dashed edge). Since the
-endpoint of $aa$ is not covered, it is expanded to $aaa$ and $aab$, in
-Figure \ref{fig:im1} (c). Both sequences $aaa$ and $aab$ are found to
-be spurious, and the enpoint of $aab$, whose label has changed from
-$\top$ to $\bot$, is now covered. In the process, the label of $aa$
-has also changed from $q_1$ to $q_1 \wedge y>x-1 \wedge q_2$, due to
-the strenghtening with the interpolant from $aab$.  Finally, the only
-uncovered node $aaa$ is expanded to $aaaa$ and $aaab$, both found to be
-spurious, in Figure \ref{fig:im2}. The refinement of $aaab$ causes
-the label of $aaa$ to change from $q_1$ to $q_1 \wedge y>x-1 \wedge
-q_2$ and this node is now covered by $aa$. Since its successors are
-also covered, there are no uncovered nodes and the procedure returns
-$\true$. \qed
-
-The correctness of Algorithm \ref{alg:impact} is proved in the
-following:
-
-\begin{theorem}\label{thm:impact}
-  Given an automaton $\A$, such that $L(\A) \neq \emptyset$, Algorithm
-  \ref{alg:impact} terminates and returns a word $w \in L(\A)$. If
-  Algorithm \ref{alg:impact} terminates reporting $\true$, then
-  $L(\A) = \emptyset$.
-\end{theorem}
-
-\section{Experimental Evaluation}
-
-We have implemented both Algorithm \ref{alg:predabs} and
-\ref{alg:impact} in a prototype tool\footnote{\url{IMPLEMENTATION URL}} that uses
-the Z3 SMT solver\footnote{\url{https://github.com/Z3Prover/z3}} to
-check satisfiability queries and generate interpolants in the theory
-of linear integer arithmetic variables (LIA). We compared both
-algorithms with a previous implementation of trace inclusion
-procedure, called
-\textsc{Includer}\footnote{\url{http://www.fit.vutbr.cz/research/groups/verifit/tools/includer/}},
-that uses on-the-fly determinisation and lazy predicate abstraction
-with interpolant-based refinement \cite{Tacas16}.
-
-The results of the experiments are given in Table
-\ref{tab:experiments}. We applied the tool on several array logic
-entailments, which occur as verification conditions for imperative
-programs with arrays \cite{cav09} (Array shift, Array rotation 1+2,
-Array split) available online \cite{ntslib}. Then, we applied it on
-proving safety properties of hardware circuits (HW Counter 1+2)
-\cite{smrcka}. Finally, we considered two timed communication
-protocols, consisting of systems that are asynchronous compositions of
-timed automata, whom correctness specification are given by timed
-automata monitors: two versions (correct and faulty) of the timed
-Alternating Bit Protocol \cite{abp} and a~controller of a railroad
-crossing \cite{henzinger:RealTimeSystems}. All results were obtained
-on an XXXX CPU @ YYYY GHz machine with ZZZZ GB of RAM. Execution times
-are given in seconds.
-
-\begin{table}[htb]
-\begin{center}
-{\fontsize{8}{9}\selectfont
-\begin{tabular}{||l|c|c|c|c|c||}
-\hline
-Example & $\len{\A}$ & $L(\A)=\emptyset$ ? & Algorithm \ref{alg:predabs} & Algorithm \ref{alg:impact} & \textsc{Includer} \\
-\hline 
-\hline
-\end{tabular}
-}
-\caption{}\label{tab:experiments}
-\end{center}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{abbrv} \bibliography{refs}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\appendix
-
-\subsection{Proof of Lemma \ref{lemma:closure}}
-
-\begin{proposition}\label{prop:overline}
-  Given a formula $\phi \in \Form^+(Q,\vec{x})$ and a valuation $\nu$
-  mapping each $q \in Q$ to a value $\nu(q) \in \booli$ and each $x \in
-  \vec{x}$ to a value $\nu(x) \in \Data^\I$, let $\nu'$ be the
-  valuation that assigns each $q \in Q$ the value $\neg\nu(q)$ and
-  each $x \in \vec{x}$ the value $\nu(x)$. Then we have 
-  $\I,\nu \models \phi$ if and only if $\I,\nu' \not\models \overline{\phi}$. 
-\end{proposition}
-\proof{Immediate, by induction on the structure of $\phi$. \qed}
-
-\vspace*{\baselineskip}
-\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
-  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
-  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
-  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
-  \iota_2,a_1\ldots a_n) \iffeq \Delta(\iota_1,a_1\ldots a_n) \vee
-  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
-  definition of the initial configuration of $\A_\cup$. For the
-  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
-  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
-  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
-    \iota_2,a_1\ldots a_{n-1})}$ with
-  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
-  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
-  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
-  \iota_2,a_1\ldots a_{n-1}) \iffeq \Delta(\iota_1,a_1\ldots a_{n-1})
-  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
-  \[\begin{array}{ll}
-  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
-  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
-  \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots a_{n}) \enspace.
-  \end{array}\] 
-  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
-  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
-  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
-  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
-  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
-  \subseteq Q$ and thus $\overline{\iota_1} \equiv \iota_1$. For the case
-  $n>0$, we compute:
-  \[\begin{array}{lcl}
-  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \iffeq & \text{ (by ind. hyp.)} \\
-  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \iffeq & \text{ (by the def. of $\overline{\phi}$)} \\
-  \overline{\Delta(\iota_1,a_1\ldots,a_n)} \enspace.
-  \end{array}\]
-  Let $\nu,\nu' : (Q \cup \bigcup_{i=0}^n\vec{x}_i) \rightarrow (\booli
-  \cup \Data^\I)$ be valuations such that: \begin{compactitem}
-  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
-    %
-  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
-    %
-  \item $\nu(x) = \nu'(x)$, for each $x \in \vec{x}_0$, 
-    %
-  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
-    \in [1,n]$.
-  \end{compactitem}
-  By Proposition \ref{prop:overline}, we have $\I,\nu \models
-  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
-  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
-  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus for all $w \in
-  \Sigma[\vec{x}]^*$, we have $w \in L(\A_1) \iff w \not\in
-  L(\overline{\A_1})$. \qed}
-
-\subsection{Proof of Lemma \ref{lemma:safety-invariant}}
-
-\proof{ Let $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ in the following.
-  ``$\Leftarrow$'' This direction is trivial. ``$\Rightarrow$'' We
-  define $\Inv : (Q \mapsto \booli) \rightarrow 2^{\vec{x} \mapsto
-    \Data^\I}$ as follows. For each $\beta : Q \rightarrow \booli$,
-  let $\Inv(\beta) = \{\nu : \vec{x} \rightarrow \Data^\I \mid \exists
-  u \in \Sigma^* ~.~ \beta\cup\nu \models
-  \Post{\A}(\iota,u)\}$. Checking that $\Inv$ is a safety invariant is
-  straightforward. \qed}
-
-\subsection{Proof of Lemma \ref{lemma:abstract-invariant}}
-
-\proof{ It is sufficient to show that there exists $k\geq0$ such that for
-  all $u \in \Sigma^*$ there exists $i \in [0,k]$ such that
-  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,\mu(i))$. We have
-  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)$ for all $u \in
-  \Sigma^*$. But since $\preds$ is a finite set, also the set
-  $\{\AbsPost{\A}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
-  there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
-  exists $i \in [0,k]$ such that $\AbsPost{\A}(\iota,u) \iffeq
-  \AbsPost{\A}(\iota,\mu(i))$, which concludes the proof. \qed}
-
-\subsection{Proof of Proposition \ref{prop:positive-interpol}}
-
-\begin{proposition}\label{prop:asterisk}
-  Given a formula $\phi \in \Form^+(Q,\vec{x})$ and $a \in \Sigma$, we
-  have \(\Delta(\phi,a) \iffeq \exists Q' ~.~ \phi[Q'/Q] \wedge
-  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))\).
-\end{proposition}
-\proof{``$\Rightarrow$'' If $\I,\beta\cup\overline\nu\cup\nu \models
-  \Delta(\phi,a)$, for some valuations $\beta : Q \rightarrow \booli$
-  and $\overline\nu : \overline{\vec{x}} \rightarrow \Data^\I$, $\nu :
-  \vec{x} \rightarrow \Data^\I$, then we build a valuation $\beta' :
-  Q' \rightarrow \booli$ such that
-  $\I,\beta'\cup\beta\cup\overline\nu\cup\nu \models \phi[Q'/Q] \wedge
-  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))$. For each
-  occurrence of a formula $\Delta(q,a)$ in $\Delta(\phi,a)$ we set
-  $\beta'(q') = \true$ if $\I,\beta\cup\overline\nu\cup\nu \models
-  \Delta(q,a)$ and $\beta'(q') = \false$, otherwise. Since there are
-  no negated occurrences of such subformulae, the definition of
-  $\beta'$ is consistent, and the check
-  $\I,\beta'\cup\beta\cup\overline\nu\cup\nu \models \phi[Q'/Q] \wedge
-  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))$ is
-  immediate. $''\Leftarrow''$ This direction is an easy check. \qed}
-
-\vspace*{\baselineskip} \proof{ Let
-  $\tuple{\top,I_0,\ldots,I_{n},\bot}$ be an interpolant, and $i \in
-  [0,n]$ be the first index for which $I_i \not\in
-  \Form^+(Q,\vec{x})$. If $i=0$, we replace $I_0$ with $\iota \in
-  \Form^+(Q,\vec{x})$ and verify that the result is still an
-  interpolant. If $i>0$, by Proposition \ref{prop:asterisk}, we have
-  that $\Delta^{i-1}(I_{i-1},a_i) \iffeq \exists Q_{i-1} ~.~
-  I_{i-1}[Q_{i-1}/Q] \wedge \theta_i$. Since $I_{i-1} \in
-  \Form^+(Q,\vec{x})$, by the choice of $i$, we have that
-  $\Delta^{i-1}(I_{i-1},a_i) \in \Form^+(Q,\vec{x})$ as well. We
-  replace $I_i$ with $\Delta^{i-1}(I_{i-1},a_i)$. The result is still
-  an interpolant, because: \begin{compactitem}
-  \item $I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \models
-    \Delta^{i-1}(I_{i-1},a_i)$, by Proposition \ref{prop:asterisk}, and
-    %
-  \item $\Delta^{i-1}(I_{i-1},a_i) \wedge \theta_{i+1} \models \exists
-    Q_{i-1} ~.~ I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \wedge \theta_{i+1}
-    \models I_i[Q_i/Q] \wedge \theta_{i+1} \models I_{i+1}[Q_{i+1}/Q]$.
-  \end{compactitem}
-  We proceed with these replacements until there are no more formulae
-  $I_i \not\in \Form^+(Q,\vec{x})$ left. \qed}
-
-\subsection{Proof of Lemma \ref{lemma:progress}}
-
-\proof{ Let $\Theta(u) \equiv \theta_0(Q_0) \wedge \theta_1(Q_0 \cup
-  Q_1,\vec{x}_0 \cup \vec{x}_1) \wedge \ldots \wedge \theta_n(Q_{n-1}
-  \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n)$ in
-  the following.
-
-  (\ref{it1:progress}) We apply Proposition \ref{prop:asterisk}
-  recursively and get: \[\Post{\A}(\iota,u)[Q_n/Q,\vec{x}_n/\vec{x}]
-  \iff \exists Q_0 \ldots \exists Q_{n-1} \exists \vec{x}_0 \ldots
-  \exists \vec{x}_{n-1} ~.~ \bigwedge_{i=0}^n \theta_i\] Assuming that
-  $\Theta(u)$ is satisfiable, we obtain a model for $\Accept{\A}(u)
-  \equiv \Post{\A}(\iota,u) \wedge \theta_{n+1}[Q/Q_n]$.
-
-  (\ref{it2:progress}) If $\tuple{\top,I_0,\dots,I_n,\bot}$ is an
-  interpolant for $\Theta(u)$, the following entailments
-  hold: \begin{compactitem}
-  \item $\theta_0 \models I_0[Q_0/Q]$,
-    %
-  \item $I_{k-1}[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k}[Q_k/Q,\vec{x}_k/\vec{x}]$, $\forall k\in[1,n]$. 
-    %
-  \item $I_{n}[Q_{n}/Q] \wedge \theta_{n+1} \models \bot$. 
-  \end{compactitem}
-  We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n}$ by
-  induction on $n \geq 0$. This is sufficient to conclude because
-  $\AbsAccept{\A}(a_1\ldots a_n) \equiv \AbsPost{\A}(\iota,a_1\ldots
-  a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n} \wedge
-  \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we have
-  $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv \theta_0[Q/Q_0]
-  \models I_0$. For the induction step $n>0$, we compute:
-  \[\begin{array}{l}
-  \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] \equiv \text{ (by def. of $\AbsPost{\A}$)} \\
-  \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] \models \text{ (by Prop. \ref{prop:asterisk})} \\
-  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n \models \text{ (ind. hyp.) } \\
-  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q] \hfill\text{\qed}
-  \end{array}\]}
-
-\subsection{Proof of Theorem \ref{thm:predabs}}
-
-\proof{We prove the following invariant: each time Algorithm
-  \ref{alg:predabs} reaches line \ref{ln:while}, the set $W$ of nodes
-  in $\worklist$ contains all the frontier nodes in the ART
-  $\tuple{N\cup W,E,\rootNode,\Lambda,\lhd}$ which are not covered by
-  some node in $N$, namely that:
-  \begin{equation}\label{inv:predabs}
-  W = \set{n \mid \forall m \in N~ \forall a \in \Sigma ~.~ (n,a,m)
-    \not\in E ~\wedge~ (n,m) \not\in \lhd}
-  \end{equation}
-  Initially, this is the case because $W = \set{\rootNode}$ and $E =
-  \lhd = \emptyset$. If the invariant holds previously, at line
-  \ref{ln:while}, it will hold again after line \ref{ln:refine} is
-  executed, because, when the subtree rooted at the pivot $p$ is
-  removed, $p$ becomes a member of the set of uncovered frontier
-  nodes, and is added to $W$ at line \ref{ln:refine}. Otherwise, the
-  invariant holds at line \ref{ln:while} and the control follows the
-  else branch at line \ref{ln:else}. In this case, the newly created
-  frontier node $s$ is added to $W$ only if it is not covered by an
-  existing node in $N$ (line \ref{ln:covered}).
-
-  Next we prove that, if Algorithm \ref{alg:predabs} returns $\true$,
-  then $\bigvee_{n \in N} \Lambda(n)$ defines a safety
-  invariant. Suppose that Algorithm \ref{alg:predabs} returns at line
-  \ref{ln:true}. Then it must be that $W = \emptyset$. Because
-  (\ref{inv:predabs}) is invariant, each node in $N$ is either covered
-  by another node in $N$, or all its successors are in $N$. We prove
-  first that $\bigvee_{n \in N} \Lambda(n)$ is an invariant: for any
-  $u \in \Sigma^*$, there exists some node $n \in N$ such that
-  $\Post{\A}(\iota,u) \models \Lambda(n)$. Let $u \in \Sigma^*$ be an
-  arbitrary sequence. If $u$ labels the path from $\rootNode$ to some
-  $n \in N$, we have $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)
-  \models \Lambda(n)$ and we are done. Otherwise, let $v$ be the
-  (possibly empty) prefix of $u$ which labels the path from
-  $\rootNode$ to some $n \in N$, which is covered by another $m \in
-  N$, where $(n,a,m) \in E$, that is $u = vav'$, for some $a \in
-  \Sigma$ and $v'\in\Sigma^*$. Moreover, we have $\Post{\A}(\iota,va)
-  \models \AbsPost{\A}(\iota,va) \models \Lambda(m)$, by the
-  construction of the set $\lhd$ of covering edges --- lines
-  \ref{ln:direct-cover}, \ref{ln:child-cover},
-  \ref{ln:indirect-cover}. Continuing this argument recursively from
-  $m$, since $\len{v'} < \len{u}$, we shall eventually discover a node
-  $p$ such that $\Post{\A}(\iota,u) \models \Lambda(p)$.
-
-  To prove that $\bigvee_{n \in N} \Lambda(n)$ is, moreover, a safety
-  invariant, suppose, by contradiction, that there exists $u \in
-  \Sigma^*$ such that $\Accept{\A}(u)$ is satisfiable. By the previous
-  point, there exists a node $p \in N$ such that $\Post{\A}(\iota,u)
-  \models \Lambda(p)$. But then we have $\Accept{\A}(\iota,u) \models
-  \Accept{\A}(\iota,\lambda(p))$, thus $\Accept{\A}(\iota,\lambda(p))$
-  is satisfiable as well. However, this cannot be the case, because
-  $p$ has been processed at line \ref{ln:feasible} and Algorithm
-  \ref{alg:predabs} would have returned a counterexample,
-  contradicting the assumption that it returns $\true$. This concludes
-  the proof that $\bigvee_{n\in N} \Lambda(n)$ is a safety invariant,
-  thus $L(\A) = \emptyset$, by Lemma \ref{lemma:safety-invariant}. We
-  have then proved the second point of the statement.
-
-  For the first point, assume that $L(\A) \neq 0$ and let $w =
-  (a_1,\nu_1) \ldots (a_k,\nu_k) \in L(\A)$ be a word. By the above,
-  Algorithm \ref{alg:predabs} cannot return $\true$. Suppose, by
-  contradiction that it does not terminate. Since the sequences from
-  $\Sigma^*$ are explored in breadth-first order, every sequence of
-  length $k$ is eventually explored, which leads to the discovery of
-  $w$ at line \ref{ln:feasible}. Then Algorithm \ref{alg:predabs}
-  terminates returning $w \in L(\A)$. \qed}
-
-\subsection{Proof of Theorem \ref{thm:impact}}
-
-\begin{lemma}\label{lemma:well-labeled}
-  Given an ART $\Art = \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$ built
-  by Algorithm \ref{alg:impact}, $\Post{\A}(\Lambda(n),a) \models
-  \Lambda(m)$, for all $(n,a,m) \in E$.
-\end{lemma}
-\proof{ We distinguish two cases. First, if $(n,a,m)$ occurs on a path
-  in $\Art$ that has never been refined, then $\Lambda(m)=\top$ and
-  the entailment holds trivially. Otherwise, let $\Omega$ be the set
-  of paths $\omega = (n_0,a_1,n_1), \ldots, (n_{k-1},a_k,n_k)$, where
-  $n_0 = \rootNode$ and $(n,a,m) = (n_{i-1},a_i,n_i)$, for some $i \in
-  [1,k]$ and, moreover, $a_1\ldots a_k$ was found, at some point, to
-  be a spurious counterexample. Let
-  $\langle\top,I^\omega_0,\ldots,I^\omega_k,\bot\rangle$ be an
-  interpolant for $\Phi(a_1 \ldots a_k) \equiv \Lambda(\rootNode)
-  \wedge \bigwedge_{i=1}^k \theta_i \wedge \bigwedge_{q \in R(n_k)}
-  (q_k \rightarrow \bot)$, such that $I^\omega_i \in
-  \Form^+(Q,\vec{x})$, for all $i \in [0,k]$. According to Proposition
-  \ref{prop:positive-interpol}, it is possible to build such an
-  interpolant, when $\Phi(a_1 \ldots a_k)$ is unsatisfiable. By
-  Proposition \ref{prop:asterisk}, we obtain
-  $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q] \iffeq \exists Q_{i-1} ~.~
-  I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge \theta_i$
-  and, since $I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge
-  \theta_i \models I^\omega_i[Q_i/Q,\vec{x}_i/\vec{x}]$, we obtain
-  that $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q] \models
-  I^\omega_i[Q_i/Q,\vec{x}_i/\vec{x}]$. Since $\Lambda(n_{i-1}) =
-  \bigwedge_{\omega \in \Omega} I^\omega_{i-1}$ and $\Lambda(n_{i}) =
-  \bigwedge_{\omega \in \Omega} I^\omega_{i}$, we obtain
-  $\Post{\A}(\Lambda(n_{i-1}),a_i) \models \Lambda(n_i)$. \qed}
-
-\vspace*{\baselineskip}
-\proof{ We prove first that, each time Algorithm \ref{alg:impact}
-  reaches the line \ref{ln:impact-while}, we have:
-  \begin{equation}\label{inv:impact}
-  W = \set{n \mid n \text{ uncovered, } \exists a \in \Sigma~ \forall
-    s \in N ~.~ (n,a,s) \not\in E}
-  \end{equation}
-  Initially, $W = \set{\rootNode}$ and $E=\lhd=\emptyset$, thus
-  (\ref{inv:impact}) holds trivially. Suppose that (\ref{inv:impact})
-  holds at when reaching line \ref{ln:impact-while} and some node $n$
-  was removed from $W$ and inserted into $N$. We distinguish two
-  cases, either: \begin{itemize}
-  \item $n$ is covered, in which case $W$ becomes $W \setminus
-    \set{n}$ and (\ref{inv:impact}) holds, or
-  %
-  \item $n$ is not covered, in which case $W$ becomes $(W \setminus
-    \set{n}) \cup S$, where $S = \set{s \not\in N \mid (n,a,s) \in E,
-      a \in \Sigma}$ is the set of fresh successors of $n$. But then
-    no node $s \in S$ is covered and has successors in $E$, thus
-    (\ref{inv:impact}) holds.
-  \end{itemize}
-  Then the condition (\ref{inv:impact}) holds next time line
-  \ref{ln:impact-while} is reached, thus it is invariant.
-
-  Suppose first that Algorithm \ref{alg:impact} returns $\true$, thus
-  $W = \emptyset$ and, by (\ref{inv:impact}), for each node in $n \in
-  N$ one of the following hold: \begin{compactitem}
-  \item $n$ is covered, or
-  \item for each $a \in \Sigma$ there exists $s \in N$ such that
-    $(n,a,s) \in E$. 
-  \end{compactitem}
-  We prove that, in this case, $\bigvee_{n \in N} \Lambda(n)$ defines
-  a safety invariant and conclude that $L(\A) = \emptyset$, by Lemma
-  \ref{lemma:safety-invariant}. To this end, let $u = a_1 \ldots a_k
-  \in \Sigma^*$ be an arbitrary sequence and let $v_1$ be the largest
-  prefix of $u$ that labels a path from $\rootNode$ to some node $n_1
-  \in N$. If $v_1 = u$ we are done. Otherwise, by the choice of $v_1$,
-  it must be the case that a successor of $n_1$ is missing from
-  $(N,E)$, thus $n_1$ must be covered, by (\ref{inv:impact}) and the
-  fact that $W=\emptyset$. Let $n'_1$ be the closest ancestor of $n_1$
-  such that $(n'_1,n''_1) \in \lhd$, for some $n''_1 \in N$, and let
-  $v'_1$ be the prefix of $v_1$ leading to $n'_1$. By the construction
-  of $\lhd$ (line \ref{ln:close-uncover} in function
-  $\Call{Close}{}$), we have $\Lambda(n'_1) \models
-  \Lambda(n''_1)$. Applying Lemma \ref{lemma:well-labeled} inductively
-  on $v'_1$, we obtain that $\Post{\A}(\iota,v'_1) \models
-  \Lambda(n'_1)$, thus $\Post{\A}(\iota,v'_1) \models \Lambda(n''_1)$.
-  Continuing inductively from $n''_1$, we exhibit a sequence of
-  strings $v'_1,\ldots,v'_\ell \in \Sigma^*$ and nodes $\rootNode =
-  m_0,m_1,\ldots,m_\ell$ such that, for all $i \in
-  [1,\ell]$: \begin{compactitem}
-  \item $v'_i$ labels the path between $m_{i-1}$ and $m_i$ in $(N,E)$,    
-    %
-  \item $\Post{\A}(\iota,v'_1\ldots v'_i) \models \Lambda(m_i)$.
-  \end{compactitem}
-  Moreover, we have $u = v'_1 \ldots v'_\ell$, thus
-  $\Post{\A}(\iota,u) \models \Lambda(m_\ell)$ and we are done showing
-  that $\bigvee_{n\in N} \Lambda(n)$ is an invariant.
-
-  To prove that $\bigvee_{n \in N} \Lambda(n)$ is, moreover, a safety
-  invariant, suppose that $\Accept{\A}(u)$ is satisfiable, for some $u
-  \in \Sigma^*$ and let $n \in N$ be a node such that
-  $\Post{\A}(\iota,u) \models \Lambda(n)$. By the previous point, such
-  a node must exist. But then $\Accept{\A}(u) \models
-  \Accept{\A}(\lambda(n))$, thus $\Accept{\A}(\lambda(n))$ is
-  satisfiable, and Algorithm \ref{alg:impact} returns at line
-  \ref{ln:impact-nonempty}, upon encountering $\lambda(n)$. But this
-  contradicts the assumption that Algorithm \ref{alg:impact} returns
-  $\true$, hence we have proved that $\bigvee_{n \in N} \Lambda(n)$ is
-  a safety invariant, and $L(\A) = \emptyset$ follows, by Lemma
-  \ref{lemma:safety-invariant}. We have then proved the second point
-  of the statement.
-
-  To prove the first point, assume that $L(\A) \neq \emptyset$. By the
-  previous point, Algorithm \ref{alg:impact} does not return
-  $\true$. Suppose, by contradiction, that it does not terminate and
-  conclude using the breadth-first argument from the proof of Theorem
-  \ref{thm:predabs}. \qed}
+\section{Translation of PA into ADA}
 

### 2017-05-17 predicate-automata.tex
ok

@@
 
-\include{commands}
+\include{commands-lncs}
 
@@
 denote by $\#(q)\geq0$ its arity. We denote by $Q_0$ the subset of $Q$
-consisting of predicates of arity zero. By $\Sigma$ we denote a
-finite set of symbols, called alphabet.
+consisting of predicates of arity zero. By $\Sigma$ we denote a finite
+set of symbols, disjoint from $Q$, called an alphabet.
 
 We consider an infinite countable set $\vars$ of first-order
-variables, ranging over $\nat$, and we associate each predicate $q \in
-Q$ a tuple of arguments $\vec{x}_q = \tuple{x_{q,1}, \ldots,
-  x_{q,\#(q)}}$ such that, for any two $p,q \in Q$, $\vec{x}_p$ and
-$\vec{x}_q$ have no elements in common, when $p \neq q$.
+variables, ranging over $\nat$. For a tuple of variables $\vec{x} =
+\tuple{x_1,\ldots,x_n}$, we define a one-to-one corresponding tuple
+$\overline{\vec{x}} = \tuple{\overline{x_1},\ldots,\vec{x_n}}$. We
+associate each predicate $q \in Q$ a tuple of arguments $\vec{x}_q =
+\tuple{x_{q,1}, \ldots, x_{q,\#(q)}}$ such that, for any two $p,q \in
+Q$, if $p \neq q$ then any two of $\vec{x}_p$, $\overline{\vec{x}}_p$,
+$\vec{x}_q$ and $\overline{\vec{x}}_q$ have no elements in common,
+respectively.
+
+An \emph{atom} is a predicate $q(\vec{x}_q)$,
+$q(\overline{\vec{x}}_q)$, $x \teq y$ or $x \not\teq y$, for $x,y \in
+\vars$ and a \emph{formula} is a positive boolean combination of
+atoms. We denote by $\Form(Q,\vec{x})$ the set of formulae with
+predicates from $Q$ and variables from $\vec{x}$.
+
+\begin{definition}[Predicate Automata]
+\end{definition}
 

### 2017-05-17 predicate-automata.tex
ok

--- 2017-05-16 predicate-automata.tex
@@
 Radu.Iosif@univ-grenoble-alpes.fr}
-
 \maketitle
@@
 denote by $\#(q)\geq0$ its arity. We denote by $Q_0$ the subset of $Q$
-consisting of predicates of arity zero. By $\Sigma$ we denote a
-finite set of symbols, called alphabet.
+consisting of predicates of arity zero. We consider an infinite
+countable set $\vars$ of first-order variables, ranging over
+$\nat$. For a tuple of variables $\vec{x} = \tuple{x_1,\ldots,x_n}$,
+we define a one-to-one corresponding tuple $\overline{\vec{x}} =
+\tuple{\overline{x_1},\ldots,\vec{x_n}}$. We associate each predicate
+$q \in Q$ a tuple of variables $\vec{x}_q = \tuple{x_{q,1}, \ldots,
+  x_{q,\#(q)}}$, called \emph{arguments}, such that, for any two $p,q
+\in Q$, if $p \neq q$ then any two of $\vec{x}_p$,
+$\overline{\vec{x}}_p$, $\vec{x}_q$ and $\overline{\vec{x}}_q$ have no
+elements in common, respectively.
 
-We consider an infinite countable set $\vars$ of first-order
-variables, ranging over $\nat$, and we associate each predicate $q \in
-Q$ a tuple of arguments $\vec{x}_q = \tuple{x_{q,1}, \ldots,
-  x_{q,\#(q)}}$ such that, for any two $p,q \in Q$, $\vec{x}_p$ and
-$\vec{x}_q$ have no elements in common, when $p \neq q$.
+An \emph{atom} is a predicate $q(\vec{x}_q)$,
+$q(\overline{\vec{x}}_q)$, $x \teq y$ or $x \not\teq y$, for $x,y \in
+\vars$ and a \emph{formula} is a positive boolean combination of
+atoms. We denote by $\Form(Q,\vec{x})$ the set of formulae with
+predicates from $Q$ and free variables from $\vec{x}$. A \emph{ground
+  formula} is a formula in which each predicate occurs as
+$q(n_1,\ldots,n_{\#(q)})$, for some $n_1,\ldots,n_{\#(q)} \in \nat$.
+
+\subsection{Predicate Automata}
+
+We denote by $\varepsilon$ the empty sequence and by $\Sigma^*$ the
+set of finite sequences of input symbols.
+
+\begin{definition}
+  A predicate automaton (PA) is a tuple $\A =
+  \tuple{\Sigma,Q,\iota,F,\Delta}$, where: \begin{compactitem}
+  \item $\Sigma$ is a finite input alphabet,
+  \item $Q$ is the set of predicates,
+  \item $\iota \in \Form(Q,\emptyset)$ is a ground formula describing
+    the initial configurations,
+  \item $F \subseteq Q$ is a set of accepting predicates,
+  \item $\Delta : Q \times \Sigma \rightarrow
+    \Form(Q,\vec{x}\cup\overline{\vec{x}})$ is a transition function.
+  \end{compactitem}
+\end{definition}
+We write transition rules as follows: 
+\[q(\vec{x}_q) \arrow{\alpha,x_0}{} \Delta(q,\alpha) 
+\text{, where $\fv{}{\Delta(q,\alpha)} \subseteq \bigcup \{\vec{x}_p
+  \mid p \text{ occurs in } \Delta(q,\alpha)\} \cup
+  \overline{\vec{x}}_q \cup \set{x_0}$}\] Intuitively, $\vec{x}_p$
+denotes the current values of the arguments $\vec{x}_p$ of $p$, where
+$p$ is any predicate that occurs in $\Delta(q,\alpha)$, $x_0$ is the
+input value associated with the symbol $\alpha$ read by the automaton
+and $\overline{\vec{x}}_q$ are the past values of the arguments of
+$q$. We denote by $\Delta(\phi,\alpha)$ the formula obtained from
+$\phi$ by replacing each occurrence of $q(\vec{x}_q)$ in $\phi$ by
+$\Delta(q,\alpha)$.
+
+\begin{example}
+For example, instead of \(q(i,j) \arrow{\alpha,k}{} i = k \wedge j
+\neq k \wedge p(k,j) \wedge q(i,k)\), we write: \[q(x_{q,1},x_{q,2})
+\arrow{\alpha,x_0}{} \overline{x}_{q,1} = x_0 \wedge
+\overline{x}_{q,2} \neq x_0 \wedge x_{p,1} = x_0 \wedge x_{p,2} =
+\overline{x}_{q,2} \wedge p(x_{p,1},x_{p,2}) \wedge x_{q,1} =
+\overline{x}_{q,1} \wedge x_{q,2} = x_0 \wedge q(x_{q,1},x_{q,2})\] We
+assume that this preprocessing is already done for PA. \hfill\qed
+\end{example}
+
+A \emph{configuration} of $\A$ is a ground formula. We abuse notation
+and blur the distinction between the terms denoting natural numbers
+and their interpretations. The transition relation is a subset of
+$\Form(Q,\emptyset) \times \Sigma \times \Form(Q,\emptyset)$ denoted
+as \(C \stackrel{\alpha,n}{\Longrightarrow} C'\),
+where: \begin{compactitem}
+\item $C = \phi\sigma$, for a formula $\phi \in \Form(Q,\vec{x})$ and
+  a substitution $\sigma : \vec{x} \rightarrow \nat$, and
+%
+\item $C' = \Delta(q,\alpha)\sigma'$, where $\sigma' :
+  \fv{}{\Delta(q,\alpha)} \rightarrow \nat$ is a substitution defined
+  as follows:
+  \[\sigma'(y) = \left\{\begin{array}{ll}
+  \sigma(x) & \text{if $y = \overline{x}$} \\
+  n & \text{if $y = x_0$} \\
+  \text{arbitrary $m \in \nat$} & \text{otherwise}
+  \end{array}\right.\]
+\end{compactitem}
+Observe that, because of the arbitrary choices of current values for
+the arguments, a configuration can have infinitely many outgoing
+transitions. However, since we identify configurations that are
+logically equivalent, a false configuration is a sink with only false
+successors, for any pair $(\alpha,n) \in \Sigma \times \nat$.
+
+A word $w = (\alpha_1,n_1) \ldots (\alpha_k,n_k) \in (\Sigma \times
+\nat)^*$ is \emph{accepted} by $\A$ iff there exists a sequence of
+configurations $C_k, \ldots, C_0$ such that $C_k = \iota$, $C_{i+1}
+\stackrel{\alpha_i,n_i}{\Longrightarrow} C_i$ for all $i \in [0,n-1]$
+and $C_0 \in \Form(F,\emptyset)$ contains only accepting
+predicates. The language of $\A$ is the set $L(\A)$ of accepted words.
+
+\subsection{Alternating Data Automata}
+
+Given a finite set $\vec{x} \subset \vars$ of variables, let $\vec{x}
+\mapsto \nat$ be the set of valuations of the variables $\vec{x}$ and
+$\Sigma[\vec{x}] = \Sigma \times (\vec{x} \mapsto \nat)$ be the set of
+\emph{data symbols}. A \emph{data word} (word in the sequel) is a
+finite sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data
+symbols, where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n :
+\vec{x} \rightarrow \nat$ are valuations. We denote by
+$\Sigma[\vec{x}]^*$ the set of data words over $\vec{x}$.
+
+\begin{definition}
+  An alternating data automaton (ADA) is a tuple $\A =
+  \tuple{\Sigma,\vec{x},Q_0,\iota,F,\Delta}$,
+  where: \begin{compactitem}
+    \item $\Sigma$ is a finite input alphabet,
+    \item $\vec{x}$ is a finite set of variables, ranging over $\nat$,
+    \item $Q_0$ is a set of predicates of arity zero, called \emph{states}
+    \item $\iota \in \Form(Q_0,\emptyset)$ is a boolean combination of states,
+    \item $F \subseteq Q_0$ is a set of final states,
+    \item $\Delta : Q_0 \times \Sigma \rightarrow
+      \Form(Q_0,\vec{x}\cup\overline{\vec{x}})$ is a transition function.
+  \end{compactitem}
+\end{definition}
+In each formula $\Delta(q,a)$ describing a transition rule, the
+variables $\overline{\vec{x}}$ track the previous and $\vec{x}$ the
+current values of the variables of $\A$. Observe that the initial
+values of the variables are left unconstrained, as the initial
+configuration does not contain free data variables.
+
+Let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, for any $k\geq0$, be a
+set of time-stamped variables. For an input symbol $\alpha \in \Sigma$
+and a formula $\phi \in \Form(Q_0,\vec{x}\cup\overline{\vec{x}})$, we
+write $\Delta^k(\phi,a)$ for the formula obtained from $\phi$ by
+simultaneously replacing each $q \in Q_0$ occurring in $\phi$ by the
+formula
+$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$,
+$k\geq0$. Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
+(a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
+formulae: \[\phi_0(Q) \Rightarrow \phi_1(Q,\vec{x}_0 \cup \vec{x}_1)
+\Rightarrow \ldots \Rightarrow \phi_n(Q,\vec{x}_0 \cup \ldots \cup
+\vec{x}_n)\] where $\phi_0 \equiv \iota$ and, for all $k\in[1,n]$, we
+have $\phi_k \equiv \Delta^k(\phi_{k-1},a_k)$. Next, we slightly abuse
+notation and write $\Delta(\iota,a_1,\ldots,a_n)$ for the formula
+$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ above. We say that $\A$
+\emph{accepts} $w$ iff $\I,\nu \models \Delta(\iota,a_1,\ldots,a_n)$,
+for some valuation $\nu$ that maps:\begin{compactitem}
+%
+\item each $x \in \vec{x}_k$ to $\nu_k(x)$, for all $k\in[1,n]$, 
+%
+\item each $q \in F$ to $\top$ and 
+%
+\item each $q \in Q_0 \setminus F$ to $\bot$.
+\end{compactitem}
+The language of $\A$ is the set $L(\A)$ of words from
+$\Sigma[\vec{x}]^*$ accepted by $\A$.
 
 \section{Translation of PA into ADA}
+
+Given a predicate automaton $\A_\PA = \tuple{\Sigma, Q_\PA, \iota_\PA,
+  F_\PA, \Delta_\PA}$, we define an alternating data automaton
+$\A_\ADA = \tuple{\Sigma, \vec{x}_\ADA, Q_\ADA, \iota_\ADA, F_\ADA,
+  \Delta_\ADA}$ as follows: \begin{compactitem}
+\item $\vec{x}_\ADA = \set{x \in \vars \mid x \in \fv{}{\Delta_\PA(q,\alpha)}, q \in Q_\PA, \alpha \in \Sigma}$, 
+%
+\item $Q_\ADA = \set{\overline{q} \mid q \in Q_\PA}$, where $\#(\overline{q}) = 0$, for all $\overline{q} \in Q_\ADA$,
+%
+\item $\iota_\ADA$ is obtained from $\iota_\PA$ by replacing each
+  occurrence of a ground term $q(n_1,\ldots,n_{\#(q)})$ from
+  $\iota_\PA$ with $\overline{q}$,
+%
+\item $F_\ADA = \set{\overline{q} \mid q \in F_\PA}$, and 
+%
+\item for each transition rule \(q(k_1,\ldots,k_{\#(q)}) \arrow{\alpha,x_0}{}
+  \Delta_\PA(q,\alpha)\) \begin{compactitem}
+  \item if \(k_1,\ldots,k_q\) is the tuple of variables $\vec{x}_q$
+    then \(\Delta_\ADA(q,\alpha)\) is obtained from
+    \(\Delta_\PA(q,\alpha)\) by replacing each occurrence of a
+    predicate $q(\vec{x}_q)$ with $\overline{q}$, and
+  %
+\item else, if \(k_1,\ldots,k_{\#(q)} \in \nat\), we add the
+  conjunction $\bigwedge_{i=1}^{\#(q)} \overline{x}_i = k_i$ to
+  \(\Delta_\ADA(q,\alpha)\).
+  \end{compactitem}
+  Nothing else is in $\Delta_\ADA$. 
+\end{compactitem}
+
+\begin{example}
+The PA transition rule: 
+\[q(x_{q,1},x_{q,2})
+\arrow{\alpha,x_0}{} \overline{x}_{q,1} = x_0 \wedge
+\overline{x}_{q,2} \neq x_0 \wedge x_{p,1} = x_0 \wedge x_{p,2} =
+\overline{x}_{q,2} \wedge p(x_{p,1},x_{p,2}) \wedge x_{q,1} =
+\overline{x}_{q,1} \wedge x_{q,2} = x_0 \wedge q(x_{q,1},x_{q,2})\]
+becomes the following ADA transition rule:
+\[\overline{q} \arrow{\alpha}{} \overline{x}_{q,1} = x_0 \wedge
+\overline{x}_{q,2} \neq x_0 \wedge x_{p,1} = x_0 \wedge x_{p,2} =
+\overline{x}_{q,2} \wedge \overline{p} \wedge x_{q,1} =
+\overline{x}_{q,1} \wedge x_{q,2} = x_0 \wedge \overline{q}\]
+\end{example}
+
+\begin{claim}
+  $L(\A_\PA) \neq \emptyset$ if and only if $L(\A_\ADA) \neq \emptyset$.
+\end{claim}
 

### 2017-05-22 quantifiers/draft.tex
added quantifiers

--- 2017-05-15 draft.tex
@@
-% \documentclass{llncs}
-\documentclass[10pt,conference,letterpaper,twocolumn]{IEEEtran}
+\documentclass{llncs}
+% \documentclass[10pt,conference,letterpaper,twocolumn]{IEEEtran}
 
@@
 
-\title{The Impact of Alternation}
+\title{First Order Alternation}
 
@@
 \begin{abstract}
-Alternating automata have been widely used to model and verify systems
-that handle data from finite domains, such as communication protocols
-or hardware. The main advantage of the alternating model of
-computation is that complementation is possible in linear time, thus
-allowing to concisely encode trace inclusion problems that occur often
-in verification. In this paper we consider alternating automata over
-infinite alphabets, whose transition rules are formulae in a combined
-theory of booleans and some infinite data domain, that relate past and
-current values of the data variables. The data theory is not fixed,
-but rather it is a parameter of the class. We show that union,
-intersection and complementation are possible in linear time in this
-model and, though the emptiness problem is undecidable, we provide two
-efficient semi-algorithms, inspired by two state-of-the-art
-abstraction refinement model checking methods: lazy predicate
-abstraction \cite{HJMS02} and the \impact~ semi-algorithm
-\cite{mcmillan06}. We have implemented both methods and report the
-results of an experimental comparison.
 \end{abstract}
-
-\section{Introduction}
-
-The language inclusion problem is recognized as being central to
-verification of hardware, communication protocols and software
-systems. A property is a specification of the correct executions of a
-system, given as a set $\mathcal{P}$ of executions, and the
-verification problem asks if the set $\mathcal{S}$ of executions of
-the system under consideration is contained within $\mathcal{P}$.
-
-This problem is at the core of widespread verification techniques,
-such as automata-theoretic model checking \cite{VARDI94}, where
-systems are specified as finite-state automata and properties defined
-using Linear Temporal Logic \cite{Pnueli77}. However the bottleneck of
-this and other related verification techniques is the intractability
-of language inclusion (PSPACE-complete for finite-state automata on
-finite words over finite alphabets).
-
-Alternation \cite{ChandraKozenStockmeyer81} was introduced as a
-generalization of nondeterminism, introducing universal, in addition
-to existential transitions. For automata over finite alphabets, the
-language inclusion problem can be encoded as the emptiness problem of
-an alternating automaton of linear size. Moreover, efficient
-exploration techniques based on antichains are shown to perform well
-for alternating automata over finite alphabets \cite{DeWulf08}.
-
-Using finite alphabets for the specification of properties and models
-is however very restrictive, when dealing with real-life computer
-systems, mostly because of the following reasons. On one hand,
-programs handle data from very large domains, that can be assumed to
-be infinite (64-bit integers, floating point numbers, strings of
-characters, etc.) and their correctness must be specified in terms of
-the data values. On the other hand, systems must respond to strict
-deadlines, which requires temporal specifications as timed
-languages \cite{AlurDill94}.
-
-Although being convenient specification tools, automata over infinite
-alphabets lack the decidability properties ensured by finite
-alphabets. In general, when considering infinite data as part of the
-input alphabet, language inclusion is undecidable and, even
-complementation becomes impossible, for instance, for timed automata
-\cite{AlurDill94} or finite-memory register automata
-\cite{Kaminski94}. In some cases, one can recover theoretical
-decidability, by restricting the number of variables (clocks) in timed
-automata to one \cite{OuaknineWorrell04}, or forbidding relations
-between current and past/future values, as with symbolic automata
-\cite{symbTransd:POPL12}. In such cases, the emptiness problem for the
-alternating versions becomes decidable \cite{Lasota05,DAntoniKW16}.
-
-In this paper, we present a new model of alternating automata over
-infinite alphabets consisting of pairs $(a,\nu)$ where $a$ is an input
-event from a finite set and $\nu$ is a valuation of a finite set
-$\vec{x}$ of variables that range over an infinite domain. We assume
-that, at all times, the successive values taken by the variables in
-$\vec{x}$ are an observable part of the language, in other words,
-there are no hidden variables in our model. The transition rules are
-specified by a set of formulae, in a combined first-order theory of
-boolean control states and data, that relate past with present values
-of the variables. We do not fix the data theory a~priori, but rather
-consider it to be a parameter of the class.
-
-A run over an input word $(a_1,\nu_1) \ldots (a_n,\nu_n)$ is a
-sequence $\phi_0(\vec{x}_0) \Rightarrow \phi_1(\vec{x}_0,\vec{x}_1)
-\Rightarrow \ldots \Rightarrow \phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ of
-rewritings of the initial formula by substituting boolean states with
-time-stamped transition rules. The word is accepted if the final
-formula $\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ holds, when all
-time-stamped variables $\vec{x}_1,\ldots,\vec{x}_n$ are substituted by
-their values in $\nu_1,\ldots,\nu_n$, all non-final states replaced by
-false and all final states by true.
-
-The boolean operations of union, intersection and complement can be
-implemented in linear time in this model, thus matching the complexity
-of performing these operations in the finite-alphabet case. The price
-to be paid is that emptiness becomes undecidable, for which reason we
-provide two efficient semi-algorithms for emptiness, based on lazy
-predicate abstraction \cite{HJMS02} and the \impact~method
-\cite{mcmillan06}.  These algorithms are proven to terminate and
-return a word from the language of the automaton, if one exists, but
-termination is not guaranteed when the language is empty.
-
-%% Finally, we prove termination in the empty case, when the transition
-%% formulae\begin{inparaenum}[(i)]
-%% \item contain only constants of the data sort and boolean functions of
-%%   any arity, or they
-%% %
-%% \item constrain only the current but not the past values on the input,
-%%   as in the case of symbolic automata \cite{symbTransd:POPL12,DAntoniKW16}.
-%% \end{inparaenum}
-
-We have implemented the boolean operations and emptiness checking
-semi-algorithms and carried out experiments with examples taken from
-array logics \cite{cav09}, timed automata
-\cite{henzinger:RealTimeSystems}, communication protocols \cite{abp}
-and hardware verification \cite{smrcka}.
-
-\paragraph{Related Work}
-Data languages and automata have been defined previously, in a
-classical nondeterministic setting. For instance, Kaminski and Francez
-\cite{Kaminski94} consider languages, over an infinite alphabet of
-data, recognized by automata with a finite number of registers, that
-store the input data and compare it using equality. Just as the timed
-languages recognized by timed automata \cite{AlurDill94}, these
-languages, called quasi-regular, are not closed under complement, but
-their emptiness is decidable. The impossibility of complementation
-here is caused by the use of hidden variables, which we do not
-allow. Emptiness is however undecidable in our case, mainly because
-counting (incrementing and comparing to a constant) data values is
-allowed, in many data theories.
-
-%% Other approaches to defining classes of data automata
-%% \cite{Bojanczyk:2011:TLD,BouyerPetitTherien03} are different from ours
-%% mainly because these automata are controlled by equivalence relations
-%% of finite index, whereas in our case, the transitions are defined by
-%% unrestricted formulae in the first-order theory of the data domain. As
-%% in \cite{Kaminski94,AlurDill94}, the emptiness problem is decidable
-%% for these classes.
-
-Another related model is that of predicate automata \cite{Farzan15},
-which recognize languages over integer data by labeling the words with
-conjunctions of uninterpreted predicates. We intend to explore further
-the connection with our model of alternating data automata, in order
-to apply our method to the verification of parallel programs.
-
-The model presented in this paper stems from the language inclusion
-problem considered in \cite{Tacas16}. There we provide a
-semi-algorithm for inclusion of data languages, based on an
-exponential determinization procedure and an abstraction refinement
-loop using lazy predicate abstraction \cite{HJMS02}. In this work we
-consider the full model of alternation and rely entirely on the
-ability of SMT solvers to produce interpolants in the combined theory
-of booleans and data. Since determinisation is not needed and
-complementation is possible in linear time, the bulk of the work is
-carried out by the decision procedure.
-
-The emptiness check for alternating data automata adapts similar
-semi-algorithms for nondeterministic infinite-state programs to the
-alternating model of computation. In particular, we considered the
-state-of-the-art \impact~ procedure \cite{mcmillan06} that is shown to
-outperform lazy predicate abstraction \cite{HJMS02} in the
-nondeterministic case, and generalized it to cope with
-alternation. More recent approaches for interpolant-based abstraction
-refinement target Horn systems \cite{McMillan14,Hoder12}, used to
-encode recursive and concurrent programs
-\cite{Grebenshchikov12}. However, the emptiness of alternating word
-automata cannot be directly encoded using Horn clauses, because all
-the branches of the computation synchronize on the same input, which
-cannot be encoded by a finite number of local (equality)
-constraints. We believe that the lazy annotation techniques for Horn
-clauses are suited for branching computations, which we intend to
-consider in a future tree automata setting.
 
@@
 sort $\Data$ and return value $\sigma(f) \in \ssorts{\sig}$. If
-$\#(f)=0$ then $f$ is a \emph{constant}. We consider the constants
-$\top$ and $\bot$ of sort $\Bool$.
+$\#(f)=0$ then $f$ is a \emph{constant}. Let
+$\top$ and $\bot$ be constants of sort $\Bool$.
 
@@
 \emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
-between two terms of sort $\Data$. A \emph{formula} is an
-existentially quantified combination of atoms using disjunction
-$\vee$, conjunction $\wedge$ and negation $\neg$ and we write $\phi
-\rightarrow \psi$ for $\neg\phi \vee \psi$.
+between two terms of sort $\Data$. A \emph{formula} is a quantified
+boolean combination of atoms, where the quantifiers are as $\exists x
+~.~ \phi$ and $\forall x ~.~ \phi$.
 
@@
 $\phi[t/x]$ be the result of replacing each occurrence of $x$ by
-$t$. For indexed sets $\vec{t}=\set{t_1,\ldots,t_n}$ and
-$\vec{x}=\set{x_1,\ldots,x_n}$, we write $\phi[\vec{t}/\vec{x}]$ for
+$t$. For indexed sets $T=\set{t_1,\ldots,t_n}$ and
+$X=\set{x_1,\ldots,x_n}$, we write $\phi[T/X]$ for
 the formula obtained by simultaneously replacing $x_i$ with $t_i$ in
@@
 $t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.
+The semantics of quantifiers is given as usual: $\I,\nu \models
+\exists x ~.~ \phi$ ($\forall x ~.~ \phi$) if and only if $\I,\nu[x
+  \leftarrow u] \models \phi$, for some (for all) value(s) $u \in
+\sigma(x)^\I$, where $\nu[x \leftarrow u]$ is the valuation that
+assigns $u$ to $x$ and $\nu(y)$ to all $y \neq x$.
 
@@
 \emph{valid} if $\I,\nu \models \phi$ for all valuations $\nu$.  The
-\emph{theory} $\theory(\sig,\I)$ is the set of valid formulae written
-in the signature $\sig$, with the interpretation $\I$. A \emph{decision
-procedure} for $\theory(\sig,\I)$ is an algorithm that takes a formula
-$\phi$ in the signature $\sig$ and returns yes iff $\phi \in
-\theory(\sig,\I)$.
+\emph{first-order theory} $\theory(\sig,\I)$ is the set of valid
+formulae written in the signature $\sig$, with the interpretation
+$\I$. A \emph{decision procedure} for $\theory(\sig,\I)$ is an
+algorithm that takes a formula $\phi$ in the signature $\sig$ and
+returns yes if and only if $\phi \in \theory(\sig,\I)$.
 
@@
 
-\section{Alternating Data Automata}
+\section{First Order Alternating Data Automata}
 
-In the rest of this section we fix an interpretation $\I$ and a finite
-alphabet $\Sigma$ of \emph{input events}. Given a finite set $\vec{x}
-\subset \vars$ of variables of sort $\Data$, let $\vec{x} \mapsto
-\Data^\I$ be the set of valuations of the variables $\vec{x}$ and
-$\Sigma[\vec{x}] = \Sigma \times (\vec{x} \mapsto \Data^\I)$ be the
-set of \emph{data symbols}. A \emph{data word} (word in the sequel) is
-a finite sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data
+We fix an interpretation $\I$ and a finite alphabet $\Sigma$ of
+\emph{input events}. Given a finite set $X \subset \vars$ of
+variables of sort $\Data$, let $X \mapsto \Data^\I$ be the set
+of valuations of the variables $X$ and $\Sigma[X] = \Sigma
+\times (X \mapsto \Data^\I)$ be the possibly infinite set of
+\emph{data symbols}. A \emph{data word} (word in the sequel) is a
+finite sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data
 symbols, where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n :
-\vec{x} \rightarrow \Data^\I$ are valuations. We denote by
+X \rightarrow \Data^\I$ are valuations. We denote by
 $\varepsilon$ the empty sequence, by $\Sigma^*$ the set of finite
-sequences of input events and by $\Sigma[\vec{x}]^*$ the set of data
-words over $\vec{x}$.
+sequences of input events and by $\Sigma[X]^*$ the set of data
+words over the variables $X$.
 
-This definition generalizes the classical notion of words from a
-finite alphabet to the possibly infinite alphabet
-$\Sigma[\vec{x}]$. Clearly, when $\Data^\I$ is sufficiently large or
-infinite, we can map the elements of $\Sigma$ into designated elements
-of $\Data^\I$ and use a special variable to encode the input
-events. However, keeping $\Sigma$ explicit in the following simplifies
-several technical points below, without cluttering the presentation.
+Given sets of variables $B,X \subset \vars$ of sort $\Bool$ and
+$\Data$, respectively, we denote by $\Form(B,X,Y)$ the set of formulae
+$\phi$ such that $\fv{\Bool}{\phi} = B$, $\fv{\Data}{\phi} = X$ and
+each variable from $Y$ is of sort $\Data$ and occurs under the scope
+of a quantifier. By $\Form^+(B,X,Y)$ we denote the subset of
+$\Form(B,X,Y)$ in which each boolean variable occurs under an even
+number of negations.
 
-Given sets of variables $\vec{b},\vec{x} \subset \vars$ of sort
-$\Bool$ and $\Data$, respectively, we denote by
-$\Form(\vec{b},\vec{x})$ the set of formulae $\phi$ such that
-$\fv{\Bool}{\phi} \subseteq \vec{b}$ and $\fv{\Data}{\phi} \subseteq
-\vec{x}$. By $\Form^+(\vec{b},\vec{x})$ we denote the set of formulae
-from $\Form(\vec{b},\vec{x})$ in which each boolean variable occurs
-under an even number of negations.
-
-An \emph{alternating data automaton} (ADA or automaton in the sequel)
-is a tuple $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$,
-where: \begin{compactitem}
+A \emph{first-order alternating data automaton} (fo-ADA or automaton
+in the sequel) is a tuple $\A =
+\tuple{G,L,Q,\iota,F,\Delta}$, where: \begin{compactitem}
 %
-\item $\vec{x} \subset \vars$ is a finite set of variables of sort
-  $\Data$,
+\item $G \subseteq \vars$ is a finite set of \emph{global
+  variables} of sort $\Data$, 
+%
+\item $L \subseteq \vars$ is a finite set of \emph{local
+  variables} of sort $\Data$,
 %
@@
 %
-\item $\iota \in \Form^+(Q,\emptyset)$ is the \emph{initial
+\item $\iota \in \Form^+(Q,\emptyset,\emptyset)$ is the \emph{initial
   configuration},
 %
-\item $F \subseteq Q$ is a set of \emph{final states}, and
-%
 \item $\Delta : Q \times \Sigma \rightarrow
-  \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
-    function},
+  \Form^+(Q,\overline{G}\cup G\cup\overline{L},L)$
+  is a \emph{transition function},
 \end{compactitem}
-where $\overline{\vec{x}}=\{\overline{x} \mid x \in \vec{x}\}$. In
-each formula $\Delta(q,a)$ describing a transition rule, the variables
-$\overline{\vec{x}}$ track the previous and $\vec{x}$ the current
-values of the variables of $\A$. Observe that the initial values of
-the variables are left unconstrained, as the initial configuration
-does not contain free data variables. The size of $\A$ is defined as
+where, for a set $X \subseteq \vars$, we denote $\overline{X} =
+\set{\overline{x} \mid x \in X}$. Intuitively, the variables
+$\overline{x}$ from a transition formula $\Delta(q,a)$ refer to past
+values, whereas the other variables refer to the current
+values. Observe that the initial values are unconstrained (since the
+initial configuration is a boolean combination of states without
+variables) and quantification is allowed only over the current local
+variables. In other words, we do not allow quantification over past
+values, which are beyond our control. The size of $\A$ is defined as
 $\len{\A} = \len{\iota} + \sum_{(q,a) \in
   Q\times\Sigma}\len{\Delta(q,a)}$.
-
-\begin{figure}[htb]
-\vspace*{-\baselineskip}
-\centerline{\includegraphics[scale=0.8]{figures/Example.pdf}}
-\vspace*{-\baselineskip}
-\caption{Alternating Data Automaton Example}
-\label{fig:ex}
-\end{figure}
-
-\paragraph{\em Example}
-Figure \ref{fig:ex} depicts an ADA with input alphabet $\Sigma =
-\set{a,b}$, variables $\vec{x} = \set{x,y}$, states $Q =
-\set{q_0,q_1,q_2,q_3,q_4}$, initial configuration $q_0$, final states
-$F=\set{q_3,q_4}$ and transitions:
-\[\begin{array}{rcl}
-\Delta(q_0,a) & \equiv & q_1 \wedge q_2 \wedge x\teq0 \wedge y\teq0 \\
-\Delta(q_1,a) & \equiv & q_1 \wedge q_3 \wedge x\teq\overline{y}+1 \wedge y\teq\overline{x}+1 \\
-\Delta(q_1,b) & \equiv & q_3 \wedge \overline{x} \geq \overline{y} \\
-\Delta(q_2,a) & \equiv & q_2 \wedge x > \overline{x} \wedge y > \overline{y} \\
-\Delta(q_2,b) & \equiv & q_4 \wedge \overline{x}>\overline{y}
-\end{array}\]
-The missing rules, such as $\Delta(q_0,b)$, are assumed to be
-$\bot$. Rules $\Delta(q_0,a)$ and $\Delta(q_1,a)$ are universal and
-there are no existential nondeterministic rules. Rules $\Delta(q_1,a)$
-and $\Delta(q_2,a)$ compare past ($\overline{x},\overline{y}$) with
-present ($x,y$) values, $\Delta(q_0,a)$ constrains the present and
-$\Delta(q_1,b)$, $\Delta(q_2,b)$ the past values, respectively. \qed
-
-Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, for any
-$k\geq0$, be a set of time-stamped variables. For an input event $a
-\in \Sigma$ and a formula $\phi$, we write $\Delta(\phi,a)$
-(respectively $\Delta^k(\phi,a)$) for the formula obtained from $\phi$
-by simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
-formula $\Delta(q,a)$ (respectively
-$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
-$k\geq0$). Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
-(a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
-formulae: \[\phi_0(Q) \Rightarrow \phi_1(Q,\vec{x}_0 \cup \vec{x}_1)
-\Rightarrow \ldots \Rightarrow \phi_n(Q,\vec{x}_0 \cup \ldots \cup
-\vec{x}_n)\] where $\phi_0 \equiv \iota$ and, for all $k\in[1,n]$, we
-have $\phi_k \equiv \Delta^k(\phi_{k-1},a_k)$. Next, we slightly abuse
-notation and write $\Delta(\iota,a_1,\ldots,a_n)$ for the formula
-$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ above. We say that $\A$
-\emph{accepts} $w$ iff $\I,\nu \models \Delta(\iota,a_1,\ldots,a_n)$,
-for some valuation $\nu$ that maps:\begin{inparaenum}[(1)]
-%
-\item each $x \in \vec{x}_k$ to $\nu_k(x)$, for all $k\in[1,n]$, 
-%
-\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$ and 
-%
-\item each $q \in \fv{\Bool}{\phi_n} \setminus F$ to $\bot$.
-\end{inparaenum}
-The language of $\A$ is the set $L(\A)$ of words from
-$\Sigma[\vec{x}]^*$ accepted by $\A$.
-
-\paragraph{\em Example} 
-The following sequence is a non-accepting run of the ADA from Figure
-\ref{fig:ex} on the word
-$(a,\tuple{0,0}),(a,\tuple{1,1}),(b,\tuple{2,1})$, where $\Data^\I =
-\zed$ and the function symbols have standard arithmetic
-interpretation:
-\[\begin{array}{l}
-q_0 \stackrel{(a,\tuple{0,0})}{\Longrightarrow} 
-q_1 \wedge q_2 \wedge x_1\teq0 \wedge y_1\teq0 \stackrel{(a,\tuple{1,1})}{\Longrightarrow} \\ 
-\underbrace{q_1 \wedge q_2 \wedge x_2\teq y_1+1 \wedge y_2\teq x_1+1}_{q_1} ~\wedge~
-\underbrace{q_2 \wedge x_2>x_1 \wedge y_2>y_1}_{q_2} \\ 
-\wedge~ x_1\teq0 \wedge y_1\teq0 \stackrel{(b,\tuple{2,1})}{\Longrightarrow} 
-\underbrace{q_3 \wedge x_2\geq y_2}_{q_1} \wedge \underbrace{q_4 \wedge x_2>y_2}_{q_2} \wedge x_2\teq y_1+1 \\ 
-\wedge~ y_2\teq x_1+1 \wedge \underbrace{q_4 \wedge x_2>y_2}_{q_2} \wedge x_2>x_1 \wedge y_2>y_1 \wedge x_1\teq0 \wedge y_1\teq0 ~\hfill\text{\qed}
-\end{array}\]
-
-In this paper we tackle the following problems: \begin{compactenum}
-\item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
-  with the same set of variables $\vec{x}$, do there exist automata
-  $\A_\cup$, $\A_\cap$ and $\overline{\A_1}$ such that $L(\A_\cup) =
-  \A_1 \cup \A_2$, $L(A_\cap) = \A_1 \cap \A_2$ and
-  $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$ ?
-%
-\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
-  \emptyset$ ?
-\end{compactenum}
-
-It is well known that other problems, such as \emph{universality}
-(given automaton $\A$ with variables $\vec{x}$, does $L(\A) =
-\Sigma[\vec{x}]^*$?) and \emph{inclusion} (given automata $\A_1$ and
-$\A_2$ with the same set of variables, does $L(\A_1) \subseteq
-L(\A_2)$?) can be reduced to the above problems. Observe furthermore
-that we do not consider cases in which the sets of variables in the
-two automata differ. An interesting problem in this case would be: given
-automata $\A_1$ and $\A_2$, with variables $\vec{x}_1$ and
-$\vec{x}_2$, respectively, such that $\vec{x}_1 \subseteq \vec{x}_2$,
-does $L(\A_1) \subseteq \proj{L(\A_2)}{\vec{x}_1}$, where
-$\proj{L(\A_2)}{\vec{x}_1}$ is the projection of the set of words
-$L(\A_2)$ onto the variables $\vec{x}_1$? This problem is considered
-as future work.
-
-\subsection{Boolean Closure}
-
-Given a set $Q$ of boolean variables and a set $\vec{x}$ of variables
-of sort $\Data$, for a formula $\phi \in \Form^+(Q,\vec{x})$, with no
-negated occurrences of the boolean variables, we define the formula
-$\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on the structure
-of $\phi$:
-\[\begin{array}{lclclcl}
-\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
-\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} \\ 
-\overline{\neg\phi} & \equiv & \neg \overline{\phi} \text{ if $\phi$ not atom} &&
-\overline{\phi} & \equiv & \phi \text{ if $\phi \in Q$} \\
-\overline{\phi} & \equiv & \neg\phi \text{ if $\phi \not\in Q$ atom}
-\end{array}\]
-We have $\len{\overline{\phi}} = \len{\phi}$, for every formula $\phi
-\in \Form^+(Q,\vec{x})$.
-
-In the following let $\A_i =
-\tuple{\vec{x},Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, where
-w.l.o.g. we assume that $Q_1 \cap Q_2 = \emptyset$. We define:
-\[\begin{array}{rcl}
-\A_\cup & = & \tuple{\vec{x},Q_1\cup Q_2,\iota_1 \vee \iota_2,F_1\cup F_2,\Delta_1\cup\Delta_2} \\ 
-\A_\cap & = & \tuple{\vec{x},Q_1\cup Q_2,\iota_1 \wedge \iota_2, F_1\cup F_2, \Delta_1\cup\Delta_2} \\ 
-\overline{\A_1} & = & \langle \vec{x},Q_1,\iota_1,Q_1\setminus F_1,\overline{\Delta_1} \rangle
-\end{array}\] 
-where $\overline{\Delta_1}(q,a) \equiv \overline{\Delta_1(q,a)}$, for
-all $q \in Q_1$ and $a \in \Sigma$. The following lemma shows the
-correctness of the above definitions:
-
-\begin{lemma}\label{lemma:closure}
-  Given automata $\A_i = \tuple{\vec{x},Q_i,\iota_i,F_i,\Delta_i}$,
-  for $i=1,2$, such that $Q_1 \cap Q_2 = \emptyset$, we have
-  $L(\A_\cup) = L(\A_1) \cup L(\A_2)$, $L(\A_\cap) = L(\A_1) \cap
-  L(\A_2)$ and $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus
-  L(\A_1)$.
-\end{lemma}
-
-It is easy to see that $\len{\A_\cup} = \len{\A_\cap} = \len{\A_1} +
-\len{\A_2}$ and $\len{\overline{\A}} = \len{\A}$, thus the automata
-for the boolean operations, including complementation, can be built in
-linear time. This matches the linear-time bounds for intersection and
-complementation of alternating automata over finite alphabets
-\cite{ChandraKozenStockmeyer81}.
-
-\section{Antichains and Interpolants for Emptiness}
-
-Unlike the boolean closure properties, showed to be effectively
-decidable (Lemma \ref{lemma:closure}), the emptiness problem is
-undecidable, even in very simple cases. For instance, if $\Data^\I$ is
-the set of positive integers, an ADA can simulate an Alternating
-Vector Addition System with States (AVASS) using only atoms $x \geq k$
-and $x = \overline{x} + k$, for $k \in \zed$, with the classical
-interpretation of the function symbols on integers. Since reachability
-of a control state is undecidable for AVASS \cite{LINCOLN92}, the
-emptiness problem is undecidable for ADA.
-
-Consequently, we give up on the guarantee for termination and build
-semi-algorithms that meet the requirements below: \begin{compactenum}[(i)]
-\item given an automaton $\A$, if $L(\A) \neq\emptyset$, the procedure
-  will terminate and return a word $w \in L(\A)$, and
-%
-\item if the procedure terminates without returning such a word, then
-  $L(\A) = \emptyset$.
-\end{compactenum}
-
-Let us fix an automaton $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ whose
-(finite) input event alphabet is $\Sigma$, for the rest of this
-section. Given a formula $\phi \in \Form^+(Q,\vec{x})$ and an input
-event $a \in \Sigma$, we define the \emph{post-image} function
-$\Post{\A}(\phi,a) \equiv \exists \overline{\vec{x}} ~.~
-\Delta(\phi[\overline{\vec{x}}/\vec{x}], a) \in \Form^+(Q,\vec{x})$,
-mapping each formula in $\Form^+(Q,\vec{x})$ to a formula defining the
-effect of reading the event $a$.
-
-We generalize the post-image function to finite sequences of input
-events, as follows:
-\[\begin{array}{l}
-\Post{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm} \Post{\A}(\phi,ua) \equiv \Post{\A}(\Post{\A}(\phi,u),a) \\
-\Accept{\A}(u) \equiv \Post{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} (q \rightarrow \bot)\text{, for any $u \in \Sigma^*$}
-\end{array}\]
-Then the emptiness problem for $\A$ becomes: does there exist $u \in
-\Sigma^*$ such that the formula $\Accept{\A}(u)$ is satisfiable?
-Observe that, since we ask a satisfiability query, the final states of
-$\A$ need not be constrained\footnote{ Since each state occurs
-  positively in $\Accept{\A}(u)$, this formula has a model iff it has
-  a model with every $q \in F$ set to true.}. A na\"ive semi-algorithm
-enumerates all finite sequences and checks the satisfiability of
-$\Accept{\A}(u)$ for each $u \in \Sigma^*$, using a decision procedure
-for the theory $\theory(\sig,\I)$.
-
-Since no boolean variable from $Q$ occurs under negation in $\phi$, it
-is easy to prove the following monotonicity property: given two
-formulae $\phi,\psi \in \Form^+(Q,\vec{x})$ if $\phi \models \psi$
-then $\Post{\A}(\phi,u) \models \Post{\A}(\psi,u)$, for any $u \in
-\Sigma^*$. This suggest an improvement of the above semi-algorithm,
-that enumerates and stores only a set $U \subseteq \Sigma^*$ for which
-$\set{\Post{\A}(\phi,u) \mid u \in U}$ forms an
-\emph{antichain}\footnote{Given a partial order $(D,\preceq)$ an
-  antichain is a set $A \subseteq D$ such that $a \not\preceq b$ for
-  any $a,b \in A$.} w.r.t. the entailment partial order. This is
-because, for any $u,v \in \Sigma^*$, if $\Post{\A}(\iota,u) \models
-\Post{\A}(\iota,v)$ and $\Accept{\A}(uw)$ is satisfiable for some $w
-\in \Sigma^*$, then $\Post{\A}(\iota,uw) \models \Post{\A}(\iota,vw)$,
-thus $\Accept{\A}(vw)$ is satisfiable as well, and there is no need
-for $u$, since the non-emptiness of $\A$ can be proved using $v$
-alone. However, even with this optimization, the enumeration of
-sequences from $\Sigma^*$ diverges in many real cases, because
-infinite antichains exist in many interpretations, e.g.\ $q \wedge x
-\teq 0,~ q \wedge x \teq 1, \ldots$ for $\Data^\I = \nat$.
-
-A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
-\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
-boolean valuation $\beta : Q \rightarrow \booli$, every valuation $\nu
-: \vec{x} \mapsto \Data^\I$ of the data variables and every finite
-sequence $u \in \Sigma^*$ of input events, the following
-hold: \begin{compactenum}
-%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
-%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
-%
-\item\label{it1:inv} $\I,\beta \cup \nu \models \Post{\A}(\iota,u)
-  \Rightarrow \nu \in \Inv(\beta)$, and
-%
-\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \I,\beta \cup
-  \nu \not\models \Accept{\A}(u)$.
-\end{compactenum}
-If $\Inv$ satisfies only the first point above, we call it an
-\emph{invariant}. Intuitively, a safety invariant maps every boolean
-valuation into a set of data valuations, that contains the initial
-configuration $\iota \equiv \Post{\A}(\iota,\varepsilon)$, whose data
-variables are unconstrained, over-approximates the set of reachable
-valuations (point \ref{it1:inv}) and excludes the valuations satisfying the
-acceptance condition (point \ref{it2:inv}).  A formula $\phi(Q,\vec{x})$
-is said to \emph{define} $\Inv$ iff for all $\beta: Q \rightarrow \booli$ and
-$\nu : \vec{x} \rightarrow \Data^\I$, we have $\I,\beta\cup\nu \models
-\phi$ iff $\nu \in \Inv(\beta)$.
-
-\begin{lemma}\label{lemma:safety-invariant}
-  For any automaton $\A$, we have $L(\A) = \emptyset$ if and only if
-  $\A$ has a safety invariant.
-\end{lemma}
-
-Turning back to the issue of divergence of language emptiness
-semi-algorithms in the case $L(\A) = \emptyset$, we can observe that
-an enumeration of input sequences $u_1,u_2,\ldots \in \Sigma^*$ can
-stop at step $k$ as soon as $\bigvee_{i=1}^k \Post{\A}(\iota,u_i)$
-defines a safety invariant for $\A$. Although this condition can be
-effectively checked using a decision procedure for the theory
-$\theory(\sig,\I)$, there is no guarantee that this check will ever
-succeed.
-
-The solution we adopt in the sequel is abstraction to ensure the
-termination of invariant computations. However, it is worth pointing
-out from the start that abstraction alone will only allow us to build
-invariants that are not necessarily safety invariants. To meet the
-latter condition, we resort to counterexample guided abstraction
-refinement (CEGAR).
-
-Formally, we fix a set of formulae $\preds \subseteq
-\Form(Q,\vec{x})$, such that $\bot \in \preds$ and refer to these
-formulae as \emph{predicates}. Given a formula $\phi$, we denote by
-$\abs{\phi} \equiv \bigwedge \set{\pi \in \preds \mid \phi \models
-  \pi}$ the abstraction of $\phi$ w.r.t. the predicates in
-$\preds$. The abstract versions of the post-image and acceptance
-condition are defined as follows:
-\[\begin{array}{l}
-\AbsPost{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm}
-\AbsPost{\A}(\phi,ua) \equiv \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} \\
-\AbsAccept{\A}(u) \equiv \AbsPost{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} 
-(q \rightarrow \bot)\text{, for any $u \in \Sigma^*$}
-\end{array}\]
-
-\begin{lemma}\label{lemma:abstract-invariant}
-  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
-  $k>0$ such that $\bigvee_{i=0}^k \AbsPost{\A}(\iota,\mu(i))$ defines
-  an invariant $\abs{\Inv}$ for $\A$. 
-\end{lemma}
-
-%% \paragraph{\em Remark}
-%% While computing the concrete post-image $\Post{\A}$ relies on an
-%% expensive quantifier elimination, it is common belief that this should
-%% hold also for $\AbsPost{\A}$. However, chosing the conjoined
-%% predicates that form $\AbsPost{\A}$ requires only a quantifier-free
-%% satisfiability query: $\Post{\A}(\phi,a) \models \pi$ iff $\exists
-%% \overline{\vec{x}} ~.~ \Delta(\phi[\overline{\vec{x}}/\vec{x}],a)
-%% \models \pi$ iff $\Delta(\phi[\overline{\vec{x}}/\vec{x}],a) \wedge
-%% \neg\pi$ is unsatisfiable. \qed
-
-We are left with fulfilling point (\ref{it2:inv}) from the definition
-of a safety invariant. To this end, suppose that, for a given set
-$\preds$ of predicates, the invariant $\abs{\Inv}$, defined by the
-previous lemma, meets point (\ref{it1:inv}) but not point
-(\ref{it2:inv}), where $\Post{\A}$ and $\Accept{\A}$ replace
-$\AbsPost{\A}$ and $\AbsAccept{\A}$, respectively. In other words,
-there exists a finite sequence $u \in \Sigma^*$ such that $\nu \in
-\abs{\Inv}(\beta)$ and $\I,\beta\cup\nu \models \AbsAccept{\A}(u)$,
-for some boolean $\beta : Q \rightarrow \booli$ and data $\nu :
-\vec{x} \rightarrow \Data^\I$ valuations. Such a $u\in\Sigma^*$ is
-called a \emph{counterexample}.
-
-Once a counterexample $u$ is discovered, there are two
-possibilities. Either\begin{inparaenum}[(i)]
-\item $\Accept{\A}(u)$ is satisfiable, in which case $u$ is
-  \emph{feasible} and $L(\A) \neq \emptyset$, or
-%
-\item $\Accept{\A}(u)$ is unsatisfiable, in which case $u$ is
-  \emph{spurious}.
-\end{inparaenum}
-In the first case, our semi-algorithm stops and returns a witness for
-non-emptiness, obtained from the satisfying valuation of
-$\Accept{\A}(u)$ and in the second case, we must strenghten the
-invariant by excluding from $\abs{\Inv}$ all pairs $(\beta,\nu)$ such
-that $\I,\beta\cup\nu \models \AbsAccept{\A}(u)$. This strenghtening
-is carried out by adding to $\preds$ several predicates that are
-sufficient to exclude the spurious counterexample.
-
-In general, given an unsatisfiable conjunction $\Phi \equiv
-\phi_1(X_0,X_1) \wedge \phi_2(X_1,X_2) \wedge \ldots \wedge
-\phi_n(X_{n-1},X_n)$ of time-stamped variables $X_i = \set{x_i \mid x
-  \in X}$ of any sort, a solution of the \emph{interpolation problem}
-$\Phi$, simply called an \emph{interpolant}, is a tuple $\tuple{I_0(X),
-  I_1(X), \ldots, I_n(X)}$ such that:\begin{inparaenum}[(i)]
-\item $I_0 \equiv \top$,  
-%
-\item $I_{i-1}[X_{i-1}/X] \wedge \phi_i(X_{i-1},X_i) \models
-  I_i[X_i/X]$, for all $i \in [1,n]$, and
-%
-\item $I_n \equiv \bot$. 
-\end{inparaenum}
-In the following, we shall assume the existence of an interpolating
-decision procedure for $\theory(\sig,\I)$.
-
-%% \paragraph{\em Remark} 
-%% We deviate w.l.o.g. from the standard notion of \emph{generalized
-%%   interpolants} \cite{mcmillan06}, which requires $\fv{}{I_i}
-%% \subseteq \fv{}{\phi_i} \cap \fv{}{\phi_{i+1}}$, for each $i \in
-%%           [1,n-1]$, and assume further that the variables have been
-%%           renamed by removing the time stamps. \qed
-
-A classical method for abstraction refinement is to add the elements
-of the interpolant obtained from a proof of spuriousness to the set of
-predicates. This guarantees progress, meaning that the particular
-spurious counterexample, from which the interpolant was generated,
-will never be revisited in the future. Though not always, in many
-practical test cases this progress property eventually yields a safety
-invariant.
-
-Given a non-empty spurious counterexample $u = a_1\ldots a_n$, where
-$n>0$, we consider the following interpolation problem: 
-\begin{eqnarray}\label{eq:interpolation-problem}
-\Theta(u) & \equiv & \theta_0(Q_0) \wedge \theta_1(Q_0 \cup Q_1,\vec{x}_0
-\cup \vec{x}_1) \wedge \ldots \\ 
-&& \wedge~ \theta_n(Q_{n-1} \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n) \nonumber
-\end{eqnarray}
-where $Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped
-sets of boolean variables corresponding to the set $Q$ of states of
-$\A$. The first conjunct $\theta_0(Q_0) \equiv \iota[Q_0/Q]$ is the
-initial configuration of $\A$, with every $q \in \fv{\Bool}{\iota}$
-replaced by $q_0$. The definition of $\theta_k$, for all $k\in[1,n]$,
-uses \emph{replacement sets} $R_\ell \subseteq Q_\ell$, $\ell\in
-[0,n]$, which are defined inductively below: \begin{compactitem}
-\item $R_0 = \fv{\Bool}{\theta_0}$,
-%
-\item $\theta_\ell \equiv \bigwedge_{q_{\ell-1}\in R_{\ell-1}}
-  (q_{\ell-1} \rightarrow
-  \Delta(q,a_\ell)[Q_\ell/Q,\vec{x}_{\ell-1}/\overline{\vec{x}},\vec{x}_\ell/\vec{x}])$
-  and $R_\ell = \fv{\Bool}{\theta_\ell} \cap Q_\ell$, for each
-  $\ell\in[1,n]$.
-%
-\item $\theta_{n+1}(Q_n) \equiv \bigwedge_{q \in Q \setminus F} (q_n
-  \rightarrow \bot)$.
-\end{compactitem}
-The intuition is that $R_0,\ldots,R_n$ are the sets of states
-replaced, $\theta_0, \ldots, \theta_n$ are the sets of transition
-rules fired on the run of $\A$ over $u$ and $\theta_{n+1}$ is the
-acceptance condition, which forces the last remaining non-final states
-to be false.
-
-Moreover, we require that an interpolant
-$\tuple{\top,I_0,\ldots,I_n,\bot}$ for the interpolation problem
-$\Theta(u)$ does not have negative occurrences of states, i.e.\ $I_i
-\in \Form^+(Q,\vec{x})$, for all $i \in [0,n]$. Such an interpolant
-can always be built, as showed below:
-
-\begin{proposition}\label{prop:positive-interpol}
-  If $\Theta(a_1 \ldots a_n)$ is unsatisfiable then one can build an
-  interpolant $\tuple{\top,I_0,\ldots,I_n,\bot}$ for $\Theta(a_1\ldots
-  a_n)$, such that $I_i \in \Form^+(Q,\vec{x})$, for all $i \in
-  [0,n]$.
-\end{proposition}
-
-We recall that a run of $\A$ over $u$ is a sequence: \[\phi_0(Q)
-\Rightarrow \phi_1(Q,\vec{x}_0\cup\vec{x}_1) \Rightarrow \ldots
-\Rightarrow \phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)\] where
-$\phi_0$ is the initial configuration $\iota$ and for each $k>0$,
-$\phi_k$ is obtained from $\phi_{k-1}$ by replacing each state $q \in
-\fv{\Bool}{\phi_{k-1}}$ by the formula
-$\Delta(q,a_k)[\vec{x}_{k-1}/\overline{\vec{x}},\vec{x}_k/\vec{x}]$,
-given by the transition function of $\A$. Observe that, because the
-states are replaced with transition formulae when moving one step in a
-run, these formulae lose track of the control history and are not
-suitable for producing interpolants that relate states and data.
-
-The main idea behind the above definition of the interpolation problem
-is that we would like to obtain an interpolant $\tuple{\top,I_0(Q),
-  I_1(Q,\vec{x}), \ldots, I_{n}(Q,\vec{x}),\bot}$ whose formulae
-\emph{combine states with the data constraints that must hold
-  locally}, whenever the control reaches a certain boolean
-configuration. This association of states with data valuations is
-tantamount to defining efficient semi-algorithms, based on lazy
-abstraction \cite{HJMS02}. Furthermore, the abstraction defined by the
-interpolants generated in this way can also \emph{over-approximate the
-  control structure} of an automaton, in addition to the sets of data
-values encountered throughout its runs.
-
-The correctness of this interpolation-based abstraction refinement
-setup is captured by the progress property below, which guarantees
-that adding the formulae of an interpolant for $\Theta(u)$ to the set
-$\preds$ of predicates suffices to exclude the spurious counterexample
-$u$ from future searches.
-
-\begin{lemma}\label{lemma:progress}
-  For any sequence $u = a_1\ldots a_n \in \Sigma^*$, if
-  $\Accept{\A}(u)$ is unsatisfiable, the following
-  hold: \begin{compactenum}
-  \item\label{it1:progress} $\Theta(u)$ is unsatisfiable, and
-    %
-  \item\label{it2:progress} if $\tuple{\top,I_0,\dots,I_{n},\bot}$ is
-    an interpolant for $\Theta(u)$ such that $\set{I_i \mid i
-      \in[0,n]} \subseteq \preds$ then $\AbsAccept{\A}(u)$ is
-    unsatisfiable.
-  \end{compactenum}
-\end{lemma}
-
-\section{Lazy Predicate Abstraction for ADA Emptiness}
-
-We have now all the ingredients to describe the first emptiness
-checking semi-algorithm for alternating data
-automata. Algorithm\footnote{Though termination is not guaranteed, we
-  call it algorithm for conciseness.} \ref{alg:predabs} builds an
-\emph{abstract reachability tree} (ART) whose nodes are labeled with
-formulae over-approximating the concrete sets of configurations, and a
-covering relation between nodes in order to ensure that the set of
-formulae labeling the nodes in the ART forms an antichain. Any
-spurious counterexample is eliminated by computing an interpolant and
-adding its formulae to the set of predicates (cf. Lemma \ref{lemma:progress}).
-
-\begin{algorithm}[t!]
-{\scriptsize\begin{algorithmic}[0]
-\State {\bf input}: an ADA $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$
-over the alphabet $\Sigma$ of input events
-
-\State {\bf output}: $\true$ if $L(\A)=\emptyset$ and a data word $w
-\in L(\A)$ otherwise
-\end{algorithmic}}
-
-{\scriptsize\begin{algorithmic}[1]
-
-  \State let $\Art = \tuple{N,E,\rootNode,\Lambda,\lhd}$ be an ART
-
-  \State initially $N = E = \lhd = \emptyset$, $\Lambda =
-  \set{(\rootNode,\iota)}$, $\preds = \set{\bot}$, $\worklist =
-  \tuple{\rootNode}$,
-
-  \While {$\worklist \neq \emptyset$}
-  \label{ln:while}
-
-  \State dequeue $n$ from $\worklist$
-
-  \State $N \leftarrow N \cup \set{n}$
-
-  \State let $\lambda(n) = a_1\ldots a_k$ be the label of the path from
-  $\rootNode$ to $n$
-
-  \If{$\AbsPost{\A}(\lambda(n))$ is satisfiable} \Comment{counterexample candidate}
-
-  \If{$\Accept{\A}(u)$ is satisfiable} \Comment{feasible counterexample}
-  \label{ln:feasible}
-
-  \State get model $(\beta,\nu_1,\ldots,\nu_k)$ of
-  $\Accept{\A}(\lambda(n))$
-
-  \State {\bf return} $w = (a_1,\nu_1) \ldots (a_k,\nu_k)$
-  \label{ln:cex}
-  \Comment{$w \in L(\A)$ by construction}
-
-  \Else \Comment{spurious counterexample}
-
-  \State $p \leftarrow \Call{FindPivot}{\lambda(n),\Art}$
-  \label{ln:pivot}
-
-  \State $v \leftarrow \Call{LeastInfeasibleSuffix}{\lambda(n),\Art}$
-
-  \State $\preds \leftarrow \preds \cup \set{I_0,\ldots,I_\ell}$,
-  where $\tuple{\top,I_0,\ldots,I_\ell,\bot}$ is an interpolant for
-  $\pivot{v}$
-
-  \State let $\SubArt = \tuple{N',E',p,\Lambda',\lhd'}$ be the subtree
-  of $\Art$ rooted at $p$
-
-  \For{$(m,q) \in \lhd$ such that $q \in N'$}
-
-  \State remove $m$ from $N$ and enqueue $m$ into $\worklist$
-
-  \State remove $\SubArt$ from $\Art$
-
-  \State enqueue $p$ into $\worklist$
-  \label{ln:refine}
-  \Comment{recompute the subtree rooted at $p$}
-
-  \EndFor
-
-  \EndIf
-
-  \Else \label{ln:else}
-
-  \For{$a \in \Sigma$}\label{ln:forall-events}
-  \Comment{expand $n$}
-
-  \State $\phi \leftarrow \AbsPost{\A}(\Lambda(n),a)$
-
-  \If{exist $m \in N$ such that $\phi \models \Lambda(m)$} \label{ln:covered}
-
-  \State $\lhd \leftarrow \lhd \cup \set{(n,m)}$ \Comment{$m$ covers $n$}
-  \label{ln:direct-cover}
-
-  \Else
-
-  \State let $s$ be a fresh node
-
-  \State $E \leftarrow E \cup \set{(n,a,s)}$
-
-  \State $\Lambda \leftarrow \Lambda \cup \set{(s,\phi)}$
-
-  \State $R \leftarrow \set{m \in \worklist \mid \Lambda(m) \models
-    \phi}$ \Comment{worklist nodes covered by $s$}
-
-  \For{$r \in R$}
-
-  \For{$m \in N$ such that $(m,b,r) \in E$, $b \in \Sigma$}
-
-  \State $\lhd \leftarrow \lhd \cup \set{(m,s)}$ 
-  \label{ln:child-cover}
-  \Comment{redirect covered children from $R$ into $s$}
-
-  \EndFor
-
-  \For{$(m,r) \in \lhd$}
-
-  \State $\lhd \leftarrow \lhd \cup \set{(m,s)}$ 
-  \label{ln:indirect-cover}
-  \Comment{redirect covered nodes from $R$ into $s$}
-
-  \EndFor
-
-  \EndFor
-
-  \State remove $R$ from $\Art$
-
-  \State enqueue $s$ into $\worklist$ 
-  \label{ln:expand}
-
-  \EndIf
-
-  \EndFor
-
-  \EndIf
-
-  \EndWhile
-
-  \State {\bf return} $\true$
-  \label{ln:true}
-\end{algorithmic}}
-\caption{Lazy Predicate Abstraction for ADA Emptiness}
-\label{alg:predabs}
-\end{algorithm}
-
-Formally, an ART is tuple $\Art =
-\tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$, where: \begin{compactitem}
-\item $N$ is a set of nodes, 
-%
-\item $E \subseteq N \times \Sigma \times
-N$ is a set of edges, 
-%
-\item $\rootNode \in N$ is the root of the directed
-tree $(N,E)$, 
-%
-\item $\Lambda : N \rightarrow \Form(Q,\vec{x})$ is a labeling of the
-  nodes with formulae, such that $\Lambda(\rootNode) = \iota$,
-%
-\item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
-  sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$,
-%
-\item $T : E \rightarrow \bigcup_{i=0}^\infty
-  \Form^+(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is a labeling of edges
-  with time-stamped formulae, and
-%
-\item $\lhd \subseteq N \times N$ is a set of
-\emph{covering edges}. 
-\end{compactitem}
-%% For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is $(n,a,m)$.
-%% The replacement sets and time-stamped edge labels are defined below,
-%% for each edge $e = (n,a,m) \in E$, if $n$ lies at depth\footnote{We
-%%   consider that the root node has zero depth.} $k$ in the tree
-%% $(N,E)$ then we have:
-%% \[\begin{array}{rcl}
-%% T(e) & = & \bigwedge_{q \in R(n)} (q_k \rightarrow \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]) \\
-%% R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)}
-%% \end{array}\]
-
-Each node $n \in N$ corresponds to a unique path from the root to $n$,
-labeled by a sequence $\lambda(n) \in \Sigma^*$ of input events. The
-\emph{least infeasible suffix} of $\lambda(n)$ is the smallest
-sequence $v = a_1 \ldots a_k$, such that $\lambda(n) = wv$, for some
-$w \in \Sigma^*$ and the following formula is unsatisfiable:
-\begin{equation}\label{eq:pivot}
-  \pivot{v} \equiv \Lambda(p)[Q_0/Q] \wedge
-  \theta_1(Q_0 \cup Q_1,\vec{x}_0 \cup \vec{x}_1) \wedge \ldots \wedge
-  \theta_{k+1}(Q_k)
-\end{equation}
-where $\theta_1,\ldots,\theta_{k+1}$ are defined as in
-(\ref{eq:interpolation-problem}) and $\theta_0 \equiv
-\Lambda(p)[Q_0/Q]$. The \emph{pivot} of $n$ is the node $p$
-corresponding to the start of the least infeasible suffix.  We assume
-the existence of two functions $\Call{FindPivot}{u,\Art}$ and
-$\Call{LeastInfeasibleSuffix}{u,\Art}$ that return the pivot and least
-infeasbile suffix of a sequence $u \in \Sigma^*$ in an ART $\Art$,
-without detailing their implementation.
-
-With these considerations, Algorithm \ref{alg:predabs} uses a worklist
-iteration to build an ART. We keep newly expanded nodes of $\Art$ in a
-queue $\worklist$, thus implementing a breadth-first exploration
-strategy, which guarantees that the shortest counterexamples are
-explored first. When the search encounters a counterexample candidate
-$u$, it is checked for spuriousness. If the counterexample is
-feasible, the procedure returns a data word $w \in L(\A)$, which
-interleaves the input events of $u$ with the data valuations from the
-model of $\Accept{\A}(u)$ (since $u$ is feasible, clearly
-$\Accept{\A}(u)$ is satisfiable). Otherwise, if $u$ is spurious, we
-compute its pivot $p$ (line \ref{ln:pivot}), add the interpolants for
-the least unfeasible suffix of $u$ to the set of predicates $\preds$,
-remove and recompute the subtree of $\Art$ rooted at $p$.
-
-Termination of Algorithm \ref{alg:predabs} depends on the ability of a
-given interpolating decision procedure for the combined boolean and
-data theory $\theory(\sig,\I)$ to provide interpolants that yield a
-safety invariant, whenever $L(\A) = \emptyset$. In this case, we use
-the covering relation $\lhd$ to ensure that, when a newly generated
-node is covered by a node already in $N$, it is not added to the
-worklist, thus cutting the current branch of the search.
-
-Formally, for any two nodes $n,m \in N$, we have $n \lhd m$ iff
-$\AbsPost{\A}(\Lambda(n),a) \models \Lambda(m)$ for some $a \in
-\Sigma$, in other words, if $n$ has a successor whose label entails
-the label of $m$.
-
-\begin{figure}[t!]
-\begin{tabular}{cc}
-\includegraphics[scale=0.5]{figures/PA1.pdf} & \includegraphics[scale=0.5]{figures/PA2.pdf} \\[-2mm]
-\tiny{(a)} & \tiny{(b)} \\
-\\
-\includegraphics[scale=0.5]{figures/PA3.pdf} & \includegraphics[scale=0.5]{figures/PA4.pdf} \\[-2mm]
-\tiny{(c)} & \tiny{(d)}
-\end{tabular}
-\vspace*{-\baselineskip}
-\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex} by Algorithm \ref{alg:predabs}}
-\label{fig:pa}
-\end{figure}
-
-\paragraph{\em Example}
-Consider the automaton given in Figure \ref{fig:ex}. First, Algorithm
-\ref{alg:predabs} fires the sequence $a$, and since there are no other
-formulae than $\bot$ in $\preds$, the successor of $\iota \equiv q_0$
-is $\top$, in Figure \ref{fig:pa} (a). The spuriousness check for $a$
-yields the root of the ART as pivot and the interpolant
-$\tuple{q_0,q_1}$, which is added to the set $\preds$. Then the $\top$
-node is removed and the next time $a$ is fired, it creates a node
-labeled $q_1$. The second sequence $aa$ creates a successor node
-$q_1$, which is covered by the first, depicted with a dashed arrow, in
-Figure \ref{fig:pa} (b). The third sequence is $ab$, which results in
-a new uncovered node $\top$ and triggers a spuriousness check. The new
-predicate obtained from this check is $x\leq0\wedge q_2 \wedge y\geq0$
-and the pivot is again the root. Then the entire ART is rebuilt with
-the new predicates and the fourth sequence $aab$ yields an uncovered
-node $\top$, in Figure \ref{fig:pa} (c). The new pivot is the endpoint
-of $a$ and the newly added predicates are $q_1\wedge q_2$ and $y>x-1
-\wedge q_2$. Finally, the ART is rebuilt from the pivot node and
-finally all nodes are covered, thus proving the emptiness of the
-automaton, in Figure \ref{fig:pa} (d). \qed
-
-The correctness of Algorithm \ref{alg:predabs} is proved below:
-
-\begin{theorem}\label{thm:predabs}
-  Given an automaton $\A$, such that $L(\A) \neq \emptyset$, Algorithm
-  \ref{alg:predabs} terminates and returns a word $w \in L(\A)$. If
-  Algorithm \ref{alg:predabs} terminates reporting $\true$, then
-  $L(\A) = \emptyset$.
-\end{theorem}
-
-\section{Checking ADA Emptiness with \impact}
-
-As pointed out by a number of authors, the bottleneck of predicate
-abstraction is the high cost of reconstructing parts of the ART,
-subsequent to the refinement of the set of predicates. The main idea
-of the \impact~ procedure \cite{mcmillan06} is that this can be
-avoided and the refinement (strenghtening of the node labels of the
-ART) can be performed in-place. This refinement step requires an
-update of the covering relation, because a node that used to cover
-another node might not cover it after the strenghtening of its label.
-
-We consider a total alphabetical order $\prec$ on $\Sigma$ and lift it
-to the total lexicographical order $\prec^*$ on $\Sigma^*$. A node $n
-\in N$ is \emph{covered} if $(n,p) \in \lhd$ or it has an ancestor $m$
-such that $(m,p) \in \lhd$, for some $p \in N$. A node $n$ is
-\emph{closed} if it is covered, or $\Lambda(n) \not\models \Lambda(m)$
-for all $m \in N$ such that $\lambda(m) \prec^* \lambda(n)$. Observe
-that we use the coverage relation $\lhd$ here with a different meaning
-than in Algorithm \ref{alg:predabs}.
-
-\begin{algorithm}[t!]
-{\scriptsize\begin{algorithmic}[0]
-  \State {\bf input}: an ADA $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$
-  over the alphabet $\Sigma$ of input events
-
-  \State {\bf output}: $\true$ if $L(\A)=\emptyset$ and a data word $w
-  \in L(\A)$ otherwise
-\end{algorithmic}}
-
-{\scriptsize\begin{algorithmic}[1]
-
-  \State let $\Art = \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$ be an ART
-
-  \State initially $N = E = T = \lhd = \emptyset$, $\Lambda =
-  \set{(\rootNode,\iota)}$, $R = \fv{\Bool}{\iota[Q_0/Q]}$,
-  $\worklist=\set{\rootNode}$
-
-  \While{$\worklist \neq \emptyset$}
-  \label{ln:impact-while}
-
-  \State dequeue $n$ from $\worklist$ 
-  \label{ln:impact-dequeue}
-
-  \State $N \leftarrow N \cup \set{n}$
-
-  \State let $(\rootNode,a_1,n_1),(n_1,a_2,n_2), \ldots,
-  (n_{k-1},a_k,n)$ be the path from $\rootNode$ to $n$
-
-  \If{$\Accept{\A}(a_1\ldots a_k)$ is satisfiable} 
-  \Comment{counterexample is feasible}
-
-  \State get model $(\beta,\nu_1,\ldots,\nu_k)$ of
-  $\Accept{\A}(\lambda(n))$
-
-  \State {\bf return} $w = (a_1,\nu_1) \ldots (a_k,\nu_k)$
-  \label{ln:impact-nonempty}
-  \Comment{$w \in L(\A)$ by construction}
-
-  \Else \Comment{spurious counterexample}
-
-  \State let $\tuple{\top,I_0,\ldots,I_k,\bot}$ be an interpolant for
-  $\Theta(a_1 \ldots a_k)$
-  \label{ln:refine-begin}
-
-  \State $b \leftarrow \false$
-
-  \For{$i=0,\ldots,k$}
-
-  \If{$\Lambda(n_i) \not\models I_i$}
-
-  \State $\lhd \leftarrow \lhd \setminus \set{(m,n_i) \in \lhd \mid m \in N}$
-
-  \State $\Lambda(n_i) \leftarrow \Lambda(n_i) \wedge I_i$
-  \Comment{strenghten the label of $n_i$}
-
-  \If{$\neg b$}
-
-  \State $b \leftarrow \Call{Close}{n_i}$
-  \label{ln:refine-end}
-
-  \EndIf % $\neg b$
-
-  \EndIf % $\Lambda(n_i) \not\models I_i$
-
-  \EndFor
-
-  \EndIf % $\mu(n)$ is satisfiable
-
-  \If{$n$ is not covered}
-
-  \For{$a \in \Sigma$} 
-  \Comment{expand $n$}
-  \label{ln:expand-begin}
-
-  \State let $s$ be a fresh node and $e = (n,a,s)$ be a new edge
-
-  \State $E \leftarrow E \cup \set{e}$
-  \label{ln:edge-insert}
-
-  \State $\Lambda \leftarrow \Lambda \cup \set{(s,\top)}$
-
-  \State $T \leftarrow T \cup \set{(e, \theta_k)}$
-
-  \State $R \leftarrow R \cup \{(s,\bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)})\}$
-
-  \State enqueue $s$ into $\worklist$
-  \label{ln:expand-end}
-
-  \EndFor
-
-  \EndIf % $n$ is not covered
-
-  \EndWhile
-
-  \State {\bf return} $\true$
-\end{algorithmic}
-
-\begin{algorithmic}[1]
-\Function{Close}{$x$} {\bf returns} $\Bool$
-
-\For{$y \in N$ such that $\lambda(y) \prec^* \lambda(x)$}
-
-\If{$\Lambda(x) \models \Lambda(y)$}
-
-\State $\lhd \leftarrow \left(\lhd \setminus \set{ (p,q) \in \lhd \mid
-  \text{$q$ is $x$ or a successor of $x$}}\right) \cup \set{(x,y)}$
-\label{ln:close-uncover}
-
-\State {\bf return} $\true$
-
-\EndIf
-
-\EndFor
-
-\State {\bf return} $\false$
-
-\EndFunction
-\end{algorithmic}}
-\caption{\impact~ for ADA Emptiness}
-\label{alg:impact}
-\end{algorithm}
-
-The execution of Algorithm \ref{alg:impact} consists of three
-phases\footnote{These phases correspond to the \textsc{Close},
-  \textsc{Refine} and \textsc{Expand} procedures from the original
-  description of \impact~\cite{mcmillan06}.}: \emph{close},
-\emph{refine} and \emph{expand}. Let $n$ be a node removed from the
-worklist at line \ref{ln:impact-dequeue}. If $\Accept{\A}(\lambda(n))$
-is satisfiable, the counterexample $\lambda(n)$ is feasible, in which
-case a model of $\Accept{\A}(\lambda(n))$ is obtained and a word $w
-\in L(\A)$ is returned. Otherwise, $\lambda(n)$ is a spurious
-counterexample and the procedure enters the refinement phase (lines
-\ref{ln:refine-begin}-\ref{ln:refine-end}). The interpolant for
-$\Theta(\lambda(n))$ (cf. formula \ref{eq:interpolation-problem}) is
-used to strenghten the labels of all the ancestors of $n$, by
-conjoining the formulae of the interpolant to the existing labels.
-
-In this process, the nodes on the path between $\rootNode$ and $n$,
-including $n$, might become eligible for coverage, therefore we
-attempt to close each ancestor of $n$ that is impacted by the
-refinement (line \ref{ln:refine-end}). Observe that, in this case the
-call to $\Call{Close}{}$ must uncover each node which is covered by a
-successor of $n$ (line \ref{ln:close-uncover} of the $\Call{Close}{}$
-function). This is required because, due to the over-approximation of
-the sets of reachable configurations, the covering relation is not
-transitive, as explained in \cite{mcmillan06}. If $\Call{Close}{}$
-adds a covering edge $(n_i,m)$ to $\lhd$, it does not have to be
-called for the successors of $n_i$ on this path, which is handled via
-the boolean flag $b$.
-
-Finally, if $n$ is still uncovered (it has not been previously covered
-during the refinement phase) we expand $n$ (lines
-\ref{ln:expand-begin}-\ref{ln:expand-end}) by creating a new node for
-each successor $s$ via the input event $a \in \Sigma$ and inserting it
-into the worklist.
-
-\begin{figure}[htb]
-\vspace*{-\baselineskip}
-\begin{tabular}{c}
-\includegraphics[scale=0.65]{figures/IM1.pdf} \\[-2mm]
-\tiny{(a)} \\
-\includegraphics[scale=0.65]{figures/IM2.pdf} \\[-2mm]
-\tiny{(b)} \\
-\includegraphics[scale=0.65]{figures/IM3.pdf} \\[-2mm]
-\tiny{(c)}
-\end{tabular}
-\vspace*{-0.5\baselineskip}
-\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex}
-  by Algorithm \ref{alg:impact} (1/2)}
-\label{fig:im1}
-\end{figure}
-
-\begin{figure}[htb]
-\centerline{\includegraphics[scale=0.65]{figures/IM4.pdf}}
-\vspace*{-\baselineskip}
-\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex}
-  by Algorithm \ref{alg:impact} (2/2)}
-\label{fig:im2}
-\end{figure}
-
-\paragraph{\em Example}
-We show the execution of Algorithm \ref{alg:impact} on the automaton
-from Figure \ref{fig:ex}.  Initially, the procedure fires the sequence
-$a$, whose endpoint is labeled with $\top$, in Figure \ref{fig:im1}
-(a). Since this node is uncovered, we check the spuriousness of the
-counterexample $a$ and refine the label of the node to $q_1$. Since
-the node is still uncovered, two successors, labeled with $\top$ are
-computed, corresponding to the sequences $aa$ and $ab$, in Figure
-\ref{fig:im1} (b). The spuriousness check for $aa$ yields the
-interpolant $\tuple{q_0,x\leq0\wedge q_2 \wedge y\geq0}$ which
-strenghtens the label of the endpoint of $a$ from $q_1$ to $q_1\wedge
-x\leq0\wedge q_2\wedge y\geq0$. The sequence $ab$ is also found to be
-spurious, which changes the label of its endpoint from $\top$ to
-$\bot$, and also covers it (depicted with a dashed edge). Since the
-endpoint of $aa$ is not covered, it is expanded to $aaa$ and $aab$, in
-Figure \ref{fig:im1} (c). Both sequences $aaa$ and $aab$ are found to
-be spurious, and the enpoint of $aab$, whose label has changed from
-$\top$ to $\bot$, is now covered. In the process, the label of $aa$
-has also changed from $q_1$ to $q_1 \wedge y>x-1 \wedge q_2$, due to
-the strenghtening with the interpolant from $aab$.  Finally, the only
-uncovered node $aaa$ is expanded to $aaaa$ and $aaab$, both found to be
-spurious, in Figure \ref{fig:im2}. The refinement of $aaab$ causes
-the label of $aaa$ to change from $q_1$ to $q_1 \wedge y>x-1 \wedge
-q_2$ and this node is now covered by $aa$. Since its successors are
-also covered, there are no uncovered nodes and the procedure returns
-$\true$. \qed
-
-The correctness of Algorithm \ref{alg:impact} is proved in the
-following:
-
-\begin{theorem}\label{thm:impact}
-  Given an automaton $\A$, such that $L(\A) \neq \emptyset$, Algorithm
-  \ref{alg:impact} terminates and returns a word $w \in L(\A)$. If
-  Algorithm \ref{alg:impact} terminates reporting $\true$, then
-  $L(\A) = \emptyset$.
-\end{theorem}
-
-\section{Experimental Evaluation}
-
-We have implemented both Algorithm \ref{alg:predabs} and
-\ref{alg:impact} in a prototype tool\footnote{\url{IMPLEMENTATION URL}} that uses
-the Z3 SMT solver\footnote{\url{https://github.com/Z3Prover/z3}} to
-check satisfiability queries and generate interpolants in the theory
-of linear integer arithmetic variables (LIA). We compared both
-algorithms with a previous implementation of trace inclusion
-procedure, called
-\textsc{Includer}\footnote{\url{http://www.fit.vutbr.cz/research/groups/verifit/tools/includer/}},
-that uses on-the-fly determinisation and lazy predicate abstraction
-with interpolant-based refinement \cite{Tacas16}.
-
-The results of the experiments are given in Table
-\ref{tab:experiments}. We applied the tool on several array logic
-entailments, which occur as verification conditions for imperative
-programs with arrays \cite{cav09} (Array shift, Array rotation 1+2,
-Array split) available online \cite{ntslib}. Then, we applied it on
-proving safety properties of hardware circuits (HW Counter 1+2)
-\cite{smrcka}. Finally, we considered two timed communication
-protocols, consisting of systems that are asynchronous compositions of
-timed automata, whom correctness specification are given by timed
-automata monitors: two versions (correct and faulty) of the timed
-Alternating Bit Protocol \cite{abp} and a~controller of a railroad
-crossing \cite{henzinger:RealTimeSystems}. All results were obtained
-on an XXXX CPU @ YYYY GHz machine with ZZZZ GB of RAM. Execution times
-are given in seconds.
-
-\begin{table}[htb]
-\begin{center}
-{\fontsize{8}{9}\selectfont
-\begin{tabular}{||l|c|c|c|c|c||}
-\hline
-Example & $\len{\A}$ & $L(\A)=\emptyset$ ? & Algorithm \ref{alg:predabs} & Algorithm \ref{alg:impact} & \textsc{Includer} \\
-\hline 
-\hline
-\end{tabular}
-}
-\caption{}\label{tab:experiments}
-\end{center}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{abbrv} \bibliography{refs}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\appendix
-
-\subsection{Proof of Lemma \ref{lemma:closure}}
-
-\begin{proposition}\label{prop:overline}
-  Given a formula $\phi \in \Form^+(Q,\vec{x})$ and a valuation $\nu$
-  mapping each $q \in Q$ to a value $\nu(q) \in \booli$ and each $x \in
-  \vec{x}$ to a value $\nu(x) \in \Data^\I$, let $\nu'$ be the
-  valuation that assigns each $q \in Q$ the value $\neg\nu(q)$ and
-  each $x \in \vec{x}$ the value $\nu(x)$. Then we have 
-  $\I,\nu \models \phi$ if and only if $\I,\nu' \not\models \overline{\phi}$. 
-\end{proposition}
-\proof{Immediate, by induction on the structure of $\phi$. \qed}
-
-\vspace*{\baselineskip}
-\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
-  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
-  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
-  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
-  \iota_2,a_1\ldots a_n) \iffeq \Delta(\iota_1,a_1\ldots a_n) \vee
-  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
-  definition of the initial configuration of $\A_\cup$. For the
-  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
-  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
-  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
-    \iota_2,a_1\ldots a_{n-1})}$ with
-  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
-  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
-  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
-  \iota_2,a_1\ldots a_{n-1}) \iffeq \Delta(\iota_1,a_1\ldots a_{n-1})
-  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
-  \[\begin{array}{ll}
-  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
-  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
-  \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots a_{n}) \enspace.
-  \end{array}\] 
-  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
-  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
-  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
-  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
-  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
-  \subseteq Q$ and thus $\overline{\iota_1} \equiv \iota_1$. For the case
-  $n>0$, we compute:
-  \[\begin{array}{lcl}
-  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \iffeq & \text{ (by ind. hyp.)} \\
-  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \iffeq & \text{ (by the def. of $\overline{\phi}$)} \\
-  \overline{\Delta(\iota_1,a_1\ldots,a_n)} \enspace.
-  \end{array}\]
-  Let $\nu,\nu' : (Q \cup \bigcup_{i=0}^n\vec{x}_i) \rightarrow (\booli
-  \cup \Data^\I)$ be valuations such that: \begin{compactitem}
-  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
-    %
-  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
-    %
-  \item $\nu(x) = \nu'(x)$, for each $x \in \vec{x}_0$, 
-    %
-  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
-    \in [1,n]$.
-  \end{compactitem}
-  By Proposition \ref{prop:overline}, we have $\I,\nu \models
-  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
-  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
-  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus for all $w \in
-  \Sigma[\vec{x}]^*$, we have $w \in L(\A_1) \iff w \not\in
-  L(\overline{\A_1})$. \qed}
-
-\subsection{Proof of Lemma \ref{lemma:safety-invariant}}
-
-\proof{ Let $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ in the following.
-  ``$\Leftarrow$'' This direction is trivial. ``$\Rightarrow$'' We
-  define $\Inv : (Q \mapsto \booli) \rightarrow 2^{\vec{x} \mapsto
-    \Data^\I}$ as follows. For each $\beta : Q \rightarrow \booli$,
-  let $\Inv(\beta) = \{\nu : \vec{x} \rightarrow \Data^\I \mid \exists
-  u \in \Sigma^* ~.~ \beta\cup\nu \models
-  \Post{\A}(\iota,u)\}$. Checking that $\Inv$ is a safety invariant is
-  straightforward. \qed}
-
-\subsection{Proof of Lemma \ref{lemma:abstract-invariant}}
-
-\proof{ It is sufficient to show that there exists $k\geq0$ such that for
-  all $u \in \Sigma^*$ there exists $i \in [0,k]$ such that
-  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,\mu(i))$. We have
-  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)$ for all $u \in
-  \Sigma^*$. But since $\preds$ is a finite set, also the set
-  $\{\AbsPost{\A}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
-  there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
-  exists $i \in [0,k]$ such that $\AbsPost{\A}(\iota,u) \iffeq
-  \AbsPost{\A}(\iota,\mu(i))$, which concludes the proof. \qed}
-
-\subsection{Proof of Proposition \ref{prop:positive-interpol}}
-
-\begin{proposition}\label{prop:asterisk}
-  Given a formula $\phi \in \Form^+(Q,\vec{x})$ and $a \in \Sigma$, we
-  have \(\Delta(\phi,a) \iffeq \exists Q' ~.~ \phi[Q'/Q] \wedge
-  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))\).
-\end{proposition}
-\proof{``$\Rightarrow$'' If $\I,\beta\cup\overline\nu\cup\nu \models
-  \Delta(\phi,a)$, for some valuations $\beta : Q \rightarrow \booli$
-  and $\overline\nu : \overline{\vec{x}} \rightarrow \Data^\I$, $\nu :
-  \vec{x} \rightarrow \Data^\I$, then we build a valuation $\beta' :
-  Q' \rightarrow \booli$ such that
-  $\I,\beta'\cup\beta\cup\overline\nu\cup\nu \models \phi[Q'/Q] \wedge
-  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))$. For each
-  occurrence of a formula $\Delta(q,a)$ in $\Delta(\phi,a)$ we set
-  $\beta'(q') = \true$ if $\I,\beta\cup\overline\nu\cup\nu \models
-  \Delta(q,a)$ and $\beta'(q') = \false$, otherwise. Since there are
-  no negated occurrences of such subformulae, the definition of
-  $\beta'$ is consistent, and the check
-  $\I,\beta'\cup\beta\cup\overline\nu\cup\nu \models \phi[Q'/Q] \wedge
-  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))$ is
-  immediate. $''\Leftarrow''$ This direction is an easy check. \qed}
-
-\vspace*{\baselineskip} \proof{ Let
-  $\tuple{\top,I_0,\ldots,I_{n},\bot}$ be an interpolant, and $i \in
-  [0,n]$ be the first index for which $I_i \not\in
-  \Form^+(Q,\vec{x})$. If $i=0$, we replace $I_0$ with $\iota \in
-  \Form^+(Q,\vec{x})$ and verify that the result is still an
-  interpolant. If $i>0$, by Proposition \ref{prop:asterisk}, we have
-  that $\Delta^{i-1}(I_{i-1},a_i) \iffeq \exists Q_{i-1} ~.~
-  I_{i-1}[Q_{i-1}/Q] \wedge \theta_i$. Since $I_{i-1} \in
-  \Form^+(Q,\vec{x})$, by the choice of $i$, we have that
-  $\Delta^{i-1}(I_{i-1},a_i) \in \Form^+(Q,\vec{x})$ as well. We
-  replace $I_i$ with $\Delta^{i-1}(I_{i-1},a_i)$. The result is still
-  an interpolant, because: \begin{compactitem}
-  \item $I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \models
-    \Delta^{i-1}(I_{i-1},a_i)$, by Proposition \ref{prop:asterisk}, and
-    %
-  \item $\Delta^{i-1}(I_{i-1},a_i) \wedge \theta_{i+1} \models \exists
-    Q_{i-1} ~.~ I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \wedge \theta_{i+1}
-    \models I_i[Q_i/Q] \wedge \theta_{i+1} \models I_{i+1}[Q_{i+1}/Q]$.
-  \end{compactitem}
-  We proceed with these replacements until there are no more formulae
-  $I_i \not\in \Form^+(Q,\vec{x})$ left. \qed}
-
-\subsection{Proof of Lemma \ref{lemma:progress}}
-
-\proof{ Let $\Theta(u) \equiv \theta_0(Q_0) \wedge \theta_1(Q_0 \cup
-  Q_1,\vec{x}_0 \cup \vec{x}_1) \wedge \ldots \wedge \theta_n(Q_{n-1}
-  \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n)$ in
-  the following.
-
-  (\ref{it1:progress}) We apply Proposition \ref{prop:asterisk}
-  recursively and get: \[\Post{\A}(\iota,u)[Q_n/Q,\vec{x}_n/\vec{x}]
-  \iff \exists Q_0 \ldots \exists Q_{n-1} \exists \vec{x}_0 \ldots
-  \exists \vec{x}_{n-1} ~.~ \bigwedge_{i=0}^n \theta_i\] Assuming that
-  $\Theta(u)$ is satisfiable, we obtain a model for $\Accept{\A}(u)
-  \equiv \Post{\A}(\iota,u) \wedge \theta_{n+1}[Q/Q_n]$.
-
-  (\ref{it2:progress}) If $\tuple{\top,I_0,\dots,I_n,\bot}$ is an
-  interpolant for $\Theta(u)$, the following entailments
-  hold: \begin{compactitem}
-  \item $\theta_0 \models I_0[Q_0/Q]$,
-    %
-  \item $I_{k-1}[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k}[Q_k/Q,\vec{x}_k/\vec{x}]$, $\forall k\in[1,n]$. 
-    %
-  \item $I_{n}[Q_{n}/Q] \wedge \theta_{n+1} \models \bot$. 
-  \end{compactitem}
-  We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n}$ by
-  induction on $n \geq 0$. This is sufficient to conclude because
-  $\AbsAccept{\A}(a_1\ldots a_n) \equiv \AbsPost{\A}(\iota,a_1\ldots
-  a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n} \wedge
-  \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we have
-  $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv \theta_0[Q/Q_0]
-  \models I_0$. For the induction step $n>0$, we compute:
-  \[\begin{array}{l}
-  \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] \equiv \text{ (by def. of $\AbsPost{\A}$)} \\
-  \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] \models \text{ (by Prop. \ref{prop:asterisk})} \\
-  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n \models \text{ (ind. hyp.) } \\
-  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q] \hfill\text{\qed}
-  \end{array}\]}
-
-\subsection{Proof of Theorem \ref{thm:predabs}}
-
-\proof{We prove the following invariant: each time Algorithm
-  \ref{alg:predabs} reaches line \ref{ln:while}, the set $W$ of nodes
-  in $\worklist$ contains all the frontier nodes in the ART
-  $\tuple{N\cup W,E,\rootNode,\Lambda,\lhd}$ which are not covered by
-  some node in $N$, namely that:
-  \begin{equation}\label{inv:predabs}
-  W = \set{n \mid \forall m \in N~ \forall a \in \Sigma ~.~ (n,a,m)
-    \not\in E ~\wedge~ (n,m) \not\in \lhd}
-  \end{equation}
-  Initially, this is the case because $W = \set{\rootNode}$ and $E =
-  \lhd = \emptyset$. If the invariant holds previously, at line
-  \ref{ln:while}, it will hold again after line \ref{ln:refine} is
-  executed, because, when the subtree rooted at the pivot $p$ is
-  removed, $p$ becomes a member of the set of uncovered frontier
-  nodes, and is added to $W$ at line \ref{ln:refine}. Otherwise, the
-  invariant holds at line \ref{ln:while} and the control follows the
-  else branch at line \ref{ln:else}. In this case, the newly created
-  frontier node $s$ is added to $W$ only if it is not covered by an
-  existing node in $N$ (line \ref{ln:covered}).
-
-  Next we prove that, if Algorithm \ref{alg:predabs} returns $\true$,
-  then $\bigvee_{n \in N} \Lambda(n)$ defines a safety
-  invariant. Suppose that Algorithm \ref{alg:predabs} returns at line
-  \ref{ln:true}. Then it must be that $W = \emptyset$. Because
-  (\ref{inv:predabs}) is invariant, each node in $N$ is either covered
-  by another node in $N$, or all its successors are in $N$. We prove
-  first that $\bigvee_{n \in N} \Lambda(n)$ is an invariant: for any
-  $u \in \Sigma^*$, there exists some node $n \in N$ such that
-  $\Post{\A}(\iota,u) \models \Lambda(n)$. Let $u \in \Sigma^*$ be an
-  arbitrary sequence. If $u$ labels the path from $\rootNode$ to some
-  $n \in N$, we have $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)
-  \models \Lambda(n)$ and we are done. Otherwise, let $v$ be the
-  (possibly empty) prefix of $u$ which labels the path from
-  $\rootNode$ to some $n \in N$, which is covered by another $m \in
-  N$, where $(n,a,m) \in E$, that is $u = vav'$, for some $a \in
-  \Sigma$ and $v'\in\Sigma^*$. Moreover, we have $\Post{\A}(\iota,va)
-  \models \AbsPost{\A}(\iota,va) \models \Lambda(m)$, by the
-  construction of the set $\lhd$ of covering edges --- lines
-  \ref{ln:direct-cover}, \ref{ln:child-cover},
-  \ref{ln:indirect-cover}. Continuing this argument recursively from
-  $m$, since $\len{v'} < \len{u}$, we shall eventually discover a node
-  $p$ such that $\Post{\A}(\iota,u) \models \Lambda(p)$.
-
-  To prove that $\bigvee_{n \in N} \Lambda(n)$ is, moreover, a safety
-  invariant, suppose, by contradiction, that there exists $u \in
-  \Sigma^*$ such that $\Accept{\A}(u)$ is satisfiable. By the previous
-  point, there exists a node $p \in N$ such that $\Post{\A}(\iota,u)
-  \models \Lambda(p)$. But then we have $\Accept{\A}(\iota,u) \models
-  \Accept{\A}(\iota,\lambda(p))$, thus $\Accept{\A}(\iota,\lambda(p))$
-  is satisfiable as well. However, this cannot be the case, because
-  $p$ has been processed at line \ref{ln:feasible} and Algorithm
-  \ref{alg:predabs} would have returned a counterexample,
-  contradicting the assumption that it returns $\true$. This concludes
-  the proof that $\bigvee_{n\in N} \Lambda(n)$ is a safety invariant,
-  thus $L(\A) = \emptyset$, by Lemma \ref{lemma:safety-invariant}. We
-  have then proved the second point of the statement.
-
-  For the first point, assume that $L(\A) \neq 0$ and let $w =
-  (a_1,\nu_1) \ldots (a_k,\nu_k) \in L(\A)$ be a word. By the above,
-  Algorithm \ref{alg:predabs} cannot return $\true$. Suppose, by
-  contradiction that it does not terminate. Since the sequences from
-  $\Sigma^*$ are explored in breadth-first order, every sequence of
-  length $k$ is eventually explored, which leads to the discovery of
-  $w$ at line \ref{ln:feasible}. Then Algorithm \ref{alg:predabs}
-  terminates returning $w \in L(\A)$. \qed}
-
-\subsection{Proof of Theorem \ref{thm:impact}}
-
-\begin{lemma}\label{lemma:well-labeled}
-  Given an ART $\Art = \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$ built
-  by Algorithm \ref{alg:impact}, $\Post{\A}(\Lambda(n),a) \models
-  \Lambda(m)$, for all $(n,a,m) \in E$.
-\end{lemma}
-\proof{ We distinguish two cases. First, if $(n,a,m)$ occurs on a path
-  in $\Art$ that has never been refined, then $\Lambda(m)=\top$ and
-  the entailment holds trivially. Otherwise, let $\Omega$ be the set
-  of paths $\omega = (n_0,a_1,n_1), \ldots, (n_{k-1},a_k,n_k)$, where
-  $n_0 = \rootNode$ and $(n,a,m) = (n_{i-1},a_i,n_i)$, for some $i \in
-  [1,k]$ and, moreover, $a_1\ldots a_k$ was found, at some point, to
-  be a spurious counterexample. Let
-  $\langle\top,I^\omega_0,\ldots,I^\omega_k,\bot\rangle$ be an
-  interpolant for $\Phi(a_1 \ldots a_k) \equiv \Lambda(\rootNode)
-  \wedge \bigwedge_{i=1}^k \theta_i \wedge \bigwedge_{q \in R(n_k)}
-  (q_k \rightarrow \bot)$, such that $I^\omega_i \in
-  \Form^+(Q,\vec{x})$, for all $i \in [0,k]$. According to Proposition
-  \ref{prop:positive-interpol}, it is possible to build such an
-  interpolant, when $\Phi(a_1 \ldots a_k)$ is unsatisfiable. By
-  Proposition \ref{prop:asterisk}, we obtain
-  $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q] \iffeq \exists Q_{i-1} ~.~
-  I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge \theta_i$
-  and, since $I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge
-  \theta_i \models I^\omega_i[Q_i/Q,\vec{x}_i/\vec{x}]$, we obtain
-  that $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q] \models
-  I^\omega_i[Q_i/Q,\vec{x}_i/\vec{x}]$. Since $\Lambda(n_{i-1}) =
-  \bigwedge_{\omega \in \Omega} I^\omega_{i-1}$ and $\Lambda(n_{i}) =
-  \bigwedge_{\omega \in \Omega} I^\omega_{i}$, we obtain
-  $\Post{\A}(\Lambda(n_{i-1}),a_i) \models \Lambda(n_i)$. \qed}
-
-\vspace*{\baselineskip}
-\proof{ We prove first that, each time Algorithm \ref{alg:impact}
-  reaches the line \ref{ln:impact-while}, we have:
-  \begin{equation}\label{inv:impact}
-  W = \set{n \mid n \text{ uncovered, } \exists a \in \Sigma~ \forall
-    s \in N ~.~ (n,a,s) \not\in E}
-  \end{equation}
-  Initially, $W = \set{\rootNode}$ and $E=\lhd=\emptyset$, thus
-  (\ref{inv:impact}) holds trivially. Suppose that (\ref{inv:impact})
-  holds at when reaching line \ref{ln:impact-while} and some node $n$
-  was removed from $W$ and inserted into $N$. We distinguish two
-  cases, either: \begin{itemize}
-  \item $n$ is covered, in which case $W$ becomes $W \setminus
-    \set{n}$ and (\ref{inv:impact}) holds, or
-  %
-  \item $n$ is not covered, in which case $W$ becomes $(W \setminus
-    \set{n}) \cup S$, where $S = \set{s \not\in N \mid (n,a,s) \in E,
-      a \in \Sigma}$ is the set of fresh successors of $n$. But then
-    no node $s \in S$ is covered and has successors in $E$, thus
-    (\ref{inv:impact}) holds.
-  \end{itemize}
-  Then the condition (\ref{inv:impact}) holds next time line
-  \ref{ln:impact-while} is reached, thus it is invariant.
-
-  Suppose first that Algorithm \ref{alg:impact} returns $\true$, thus
-  $W = \emptyset$ and, by (\ref{inv:impact}), for each node in $n \in
-  N$ one of the following hold: \begin{compactitem}
-  \item $n$ is covered, or
-  \item for each $a \in \Sigma$ there exists $s \in N$ such that
-    $(n,a,s) \in E$. 
-  \end{compactitem}
-  We prove that, in this case, $\bigvee_{n \in N} \Lambda(n)$ defines
-  a safety invariant and conclude that $L(\A) = \emptyset$, by Lemma
-  \ref{lemma:safety-invariant}. To this end, let $u = a_1 \ldots a_k
-  \in \Sigma^*$ be an arbitrary sequence and let $v_1$ be the largest
-  prefix of $u$ that labels a path from $\rootNode$ to some node $n_1
-  \in N$. If $v_1 = u$ we are done. Otherwise, by the choice of $v_1$,
-  it must be the case that a successor of $n_1$ is missing from
-  $(N,E)$, thus $n_1$ must be covered, by (\ref{inv:impact}) and the
-  fact that $W=\emptyset$. Let $n'_1$ be the closest ancestor of $n_1$
-  such that $(n'_1,n''_1) \in \lhd$, for some $n''_1 \in N$, and let
-  $v'_1$ be the prefix of $v_1$ leading to $n'_1$. By the construction
-  of $\lhd$ (line \ref{ln:close-uncover} in function
-  $\Call{Close}{}$), we have $\Lambda(n'_1) \models
-  \Lambda(n''_1)$. Applying Lemma \ref{lemma:well-labeled} inductively
-  on $v'_1$, we obtain that $\Post{\A}(\iota,v'_1) \models
-  \Lambda(n'_1)$, thus $\Post{\A}(\iota,v'_1) \models \Lambda(n''_1)$.
-  Continuing inductively from $n''_1$, we exhibit a sequence of
-  strings $v'_1,\ldots,v'_\ell \in \Sigma^*$ and nodes $\rootNode =
-  m_0,m_1,\ldots,m_\ell$ such that, for all $i \in
-  [1,\ell]$: \begin{compactitem}
-  \item $v'_i$ labels the path between $m_{i-1}$ and $m_i$ in $(N,E)$,    
-    %
-  \item $\Post{\A}(\iota,v'_1\ldots v'_i) \models \Lambda(m_i)$.
-  \end{compactitem}
-  Moreover, we have $u = v'_1 \ldots v'_\ell$, thus
-  $\Post{\A}(\iota,u) \models \Lambda(m_\ell)$ and we are done showing
-  that $\bigvee_{n\in N} \Lambda(n)$ is an invariant.
-
-  To prove that $\bigvee_{n \in N} \Lambda(n)$ is, moreover, a safety
-  invariant, suppose that $\Accept{\A}(u)$ is satisfiable, for some $u
-  \in \Sigma^*$ and let $n \in N$ be a node such that
-  $\Post{\A}(\iota,u) \models \Lambda(n)$. By the previous point, such
-  a node must exist. But then $\Accept{\A}(u) \models
-  \Accept{\A}(\lambda(n))$, thus $\Accept{\A}(\lambda(n))$ is
-  satisfiable, and Algorithm \ref{alg:impact} returns at line
-  \ref{ln:impact-nonempty}, upon encountering $\lambda(n)$. But this
-  contradicts the assumption that Algorithm \ref{alg:impact} returns
-  $\true$, hence we have proved that $\bigvee_{n \in N} \Lambda(n)$ is
-  a safety invariant, and $L(\A) = \emptyset$ follows, by Lemma
-  \ref{lemma:safety-invariant}. We have then proved the second point
-  of the statement.
-
-  To prove the first point, assume that $L(\A) \neq \emptyset$. By the
-  previous point, Algorithm \ref{alg:impact} does not return
-  $\true$. Suppose, by contradiction, that it does not terminate and
-  conclude using the breadth-first argument from the proof of Theorem
-  \ref{thm:predabs}. \qed}
 

### 2017-05-22 quantifiers/draft.tex
ok

@@
 $\Form(B,X,Y)$ in which each boolean variable occurs under an even
-number of negations.
+number of negations. We denote by $\phi(B,X,Y)$ any formula from
+$\Form(B,X,Y)$.
 
@@
 
+In order to define the semantics of automata, we consider time-stamped
+sets of variables $\stamp{X}{n} = \{\stamp{x}{n} \mid x \in X\}$ for
+any $n\geq0$. For an input event $a \in \Sigma$, some constant
+$n\geq0$ and a formula $\phi$, we write $\stamp{\Delta}{n}(\phi,a)$
+for the formula obtained from $\phi$ by simultaneously replacing each
+state $q \in \fv{\Bool}{\phi}$ with
+$\Delta(\phi,a)[\stamp{G}{n}/\overline{G}, \stamp{G}{n+1}/G,
+  \stamp{L}{n}/\overline{L}, \stamp{L}{n+1}/L]$.
+
+Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in
+\Sigma[G]^*$ the \emph{run} of $\A$ over $w$ is the sequence of
+formulae: \[\phi_0 \Rightarrow \phi_1(Q,\stamp{G}{0} \cup \stamp{G}{1},
+\stamp{L}{0} \cup \stamp{L}{1}) \Rightarrow \ldots \Rightarrow
+\phi_n(Q,\stamp{G}{0} \cup \ldots \cup \stamp{G}{n}, \stamp{L}{0} \cup
+\ldots \cup \stamp{L}{n})\]
+where: \begin{compactitem}
+\item $\phi_0 \equiv \iota$, 
+%
+\item $\phi_1 \equiv \exists \stamp{L}{0} ~.~ \stamp{\Delta}{0}(\iota,a_1)$, and
+%
+\item $\phi_{i+1} \equiv \stamp{\Delta}{i}(\phi_i,a_{i+1})$, for all $i \in [1,n-1]$.
+\end{compactitem}
+We say that $\A$ \emph{accepts} $w$ if $\I,\nu \models
+\phi_n(Q,\stamp{G}{0} \cup \ldots \cup \stamp{G}{n}, \stamp{L}{0} \cup
+\ldots \cup \stamp{L}{n})$, where $\nu$ is a valuation such
+that: \begin{compactitem}
+\item $\nu(\stamp{g}{i}) = \nu_i(g)$ for each $g \in G$, 
+%
+\item $\nu(q) = \true$ for each $q \in F$, and
+%
+\item $\nu(q) = \false$ for each $q \in Q \setminus F$.
+\end{compactitem}
+The language of $\A$ is the set $L(\A)$ of words from $\Sigma[G]^*$
+accepted by $\A$. In the following, we ask two
+questions: \begin{compactenum}
+\item given automata $\A_i = \tuple{G,L_i,Q_i,\iota_i,F_i,\Delta_i}$,
+  with $L_1 \cap L_2 = \emptyset$ and $Q_1 \cap Q_2 = \emptyset$, do
+  there exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such
+  that $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
+  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[G]^* \setminus L(\A_1)$ ?
+%
+\item given an automaton $\A$, is $L(\A) = \emptyset$ ? 
+\end{compactenum}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-05-22 quantifiers/draft.tex
ok

@@
 sort $\Data$ and return value $\sigma(f) \in \ssorts{\sig}$. If
-$\#(f)=0$ then $f$ is a \emph{constant}. Let
-$\top$ and $\bot$ be constants of sort $\Bool$.
+$\#(f)=0$ then $f$ is a \emph{constant}. Let $\top$ and $\bot$ be
+constants of sort $\Bool$ and $\mathbf{0}$ be a constant of sort
+$\Data$.
 
@@
 where, for a set $X \subseteq \vars$, we denote $\overline{X} =
-\set{\overline{x} \mid x \in X}$. Intuitively, the variables
-$\overline{x}$ from a transition formula $\Delta(q,a)$ refer to past
-values, whereas the other variables refer to the current
-values. Observe that the initial values are unconstrained (since the
-initial configuration is a boolean combination of states without
-variables) and quantification is allowed only over the current local
-variables. In other words, we do not allow quantification over past
-values, which are beyond our control. The size of $\A$ is defined as
+\set{\overline{x} \mid x \in X}$. The size of $\A$ is defined as
 $\len{\A} = \len{\iota} + \sum_{(q,a) \in
   Q\times\Sigma}\len{\Delta(q,a)}$.
+
+Intuitively, the variables $\overline{x}$ from a transition formula
+$\Delta(q,a)$ refer to past values, whereas the other variables refer
+to the current values. Observe that the initial values of the global
+variables are unconstrained (since the initial configuration is a
+boolean combination of states without variables) and quantification is
+allowed only over the current local variables. In other words, we do
+not quantify over global variables which are visible throughout the
+execution and also, we do not quantify over over past values, which
+are beyond our control.
 
@@
 \Sigma[G]^*$ the \emph{run} of $\A$ over $w$ is the sequence of
-formulae: \[\phi_0 \Rightarrow \phi_1(Q,\stamp{G}{0} \cup \stamp{G}{1},
-\stamp{L}{0} \cup \stamp{L}{1}) \Rightarrow \ldots \Rightarrow
-\phi_n(Q,\stamp{G}{0} \cup \ldots \cup \stamp{G}{n}, \stamp{L}{0} \cup
-\ldots \cup \stamp{L}{n})\]
-where: \begin{compactitem}
-\item $\phi_0 \equiv \iota$, 
-%
-\item $\phi_1 \equiv \exists \stamp{L}{0} ~.~ \stamp{\Delta}{0}(\iota,a_1)$, and
+formulae: \[\iota \stackrel{(a_1,\nu_1)}{\Longrightarrow}
+\phi_1(Q,\stamp{G}{1},
+\stamp{L}{1}) \stackrel{(a_2,\nu_2)}{\Longrightarrow} \ldots
+\stackrel{(a_n,\nu_n)}{\Longrightarrow} \phi_n(Q,\stamp{G}{1} \cup
+\ldots \cup \stamp{G}{n}, \stamp{L}{1} \cup \ldots \cup
+\stamp{L}{n})\] where: \begin{compactitem}
+\item $\phi_1 \equiv \stamp{\Delta}{0}(\iota,a_1)[\mathbf{0}/\stamp{G}{0},\mathbf{0}/\stamp{L}{0}]$, and
 %
@@
 We say that $\A$ \emph{accepts} $w$ if $\I,\nu \models
-\phi_n(Q,\stamp{G}{0} \cup \ldots \cup \stamp{G}{n}, \stamp{L}{0} \cup
+\phi_n(Q,\stamp{G}{1} \cup \ldots \cup \stamp{G}{n}, \stamp{L}{1} \cup
 \ldots \cup \stamp{L}{n})$, where $\nu$ is a valuation such
 that: \begin{compactitem}
-\item $\nu(\stamp{g}{i}) = \nu_i(g)$ for each $g \in G$, 
+\item $\nu(\stamp{g}{i}) = \nu_i(g)$ for each $g \in G$ and each $i \in [1,n]$, 
 %
@@
 \item given automata $\A_i = \tuple{G,L_i,Q_i,\iota_i,F_i,\Delta_i}$,
-  with $L_1 \cap L_2 = \emptyset$ and $Q_1 \cap Q_2 = \emptyset$, do
-  there exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such
-  that $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
-  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[G]^* \setminus L(\A_1)$ ?
+  for $i=1,2$, with $L_1 \cap L_2 = \emptyset$ and $Q_1 \cap Q_2 =
+  \emptyset$, do there exist automata $\A_\cap$, $\A_\cup$ and
+  $\overline{\A}_1$ such that $L(\A_\cap) = L(\A_1) \cap L(\A_2)$,
+  $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ and $L(\overline{\A}_1) =
+  \Sigma[G]^* \setminus L(\A_1)$ ?
 %

### 2017-07-14 symbolic_nfa_universality/symbolic_subset_construction.tex
adding symbolic nfa universality (without abstraction)

\documentclass{llncs}

\usepackage[usenames,dvipsnames]{color}

\let\iint\undefined 
\let\iiint\undefined 
\let\iiiint\undefined 
\let\idotsint\undefined

\usepackage{amsmath} 
\usepackage{amsxtra} 
\usepackage{amssymb}
\usepackage{mathabx}
\usepackage{mathtools}
\usepackage{textcomp}

\usepackage{latexsym}
\usepackage{pslatex}
\usepackage{epsfig}
\usepackage{wrapfig}
%\usepackage{graphics}
\usepackage{enumerate}
\usepackage{cancel}

\usepackage{paralist}

\usepackage{ listings }

\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage[noend]{algpseudocode}
\let\Asterisk\undefined

\pagestyle{empty}

\include{commands}

\newcommand{\Xobs}{\vec{x}}
\newcommand{\Xinv}{\vec{y}}
\newcommand{\Xinvz}{\vec{z}}

\newcommand{\vx}{\vec{x}}
\newcommand{\vy}{\vec{y}}
\newcommand{\vz}{\vec{z}}
\newcommand{\dom}{\mathcal{D}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\title{Trace Inclusion of Infinite State Systems with Local Information}
\title{Predicate Abstraction for Trace Inclusion}

\author{Luk\'{a}\v{s} Hol\'{i}k\inst{1} \and Radu Iosif\inst{2} \and Adam Rogalewicz\inst{1} \and
Tom\'{a}\v{s}~Vojnar\inst{1}}

\institute{
  FIT, Brno University of Technology, IT4Innovations Centre of Excellence, 
  Czech Republic
  \and 
  University Grenoble Alpes, CNRS, VERIMAG, Grenoble, France
}

 \maketitle

\begin{abstract}\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Treansition Syatems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Let $\nat$ denote the set of non-negative integers including zero. For
any $k,\ell \in \nat$, $k \leq \ell$, we write $[k,\ell]$ for the set
$\set{k,k+1,\ldots,\ell}$. We write $\false$ and $\true$ for the
boolean constants \emph{false} and \emph{true}, respectively. Given a
possibly infinite data domain $\mathcal{D}$, we denote by
$\thd=\tuple{\mathcal{D}, p_1,\ldots,p_n, f_1,\ldots,f_m}$ the set of
syntactically correct first-order formulae with predicate
symbols $p_1,\ldots,p_n$ and function symbols $f_1,\ldots,f_m$. A variable $x$
is said to be \emph{free} in a~formula $\phi$, denoted as $\phi(x)$,
iff it does not occur under the scope of a~quantifier.

Let $\vec{x} = \set{x_1,\ldots,x_n}$ be a finite set of variables. A
\emph{valuation} $\nu : \vec{x} \rightarrow \mathcal{D}$ is an
assignment of the variables in $\vec{x}$ with values from
$\mathcal{D}$. We denote by $\mathcal{D}^\vec{x}$ the set of such
valuations. For a formula $\phi(\vec{x})$, we denote by $\nu
\modelsthd \phi$ the fact that substituting each variable $x \in
\vec{x}$ by $\nu(x)$ yields a valid formula in the theory $\thd$. In
this case, $\nu$ is said to be a~\emph{model} of $\phi$. A formula is
said to be {\em satisfiable} iff it has a model. For a~formula
$\phi(\vec{x}, \vec{x'})$ where $\vec{x'} = \set{x' ~|~ x \in
  \vec{x}}$ and two valuations $\nu,\nu' \in \mathcal{D}^{\vec{x}}$,
we denote by $(\nu,\nu') \modelsthd \phi$ the fact that the formula
obtained from $\phi$ by substituting each $x$ with $\nu(x)$ and each
$x'$ with $\nu'(x')$ is valid in $\thd$.

\newcommand{\initial}{I}
\newcommand{\accepting}{F}
\newcommand{\transition}{\tau}
\paragraph{Transition system.} 
Let $\mathcal{D}$ be an infinite date domain equipped with a first order theory $\thd$.
%
Let $\Xobs$ be the set of observable and $\Xinv$ the set of invisible variables.
%
A transition system (TS) is a triple formula $T = (\initial(\Xobs,\Xinv),\transition(\Xobs,\Xinv,\Xobs', \Xinv'),\accepting(\Xobs,\Xinv))$ where $\initial$ is a fomrula describing initial valuations, $\transition(\Xobs,\Xinv,\Xobs', \Xinv')$ describes the transition relation, and $\accepting(\Xobs,\Xinv)$ describes accepting valuations.
%
We say that a valuation $\nu'$ is 
a~\emph{successor} of $\nu$ if and only if and $(\nu,\nu') \modelsthd
\transition$. We denote the successor relation by $\nu
\arrow{\transition}{} \nu'$, and we omit writing $\transition$ 
when no confusion may arise. We write 
$\mathit{succ}(\nu)$ to denote $\nu'$.

A \emph{run} of $\transition$ is a sequence
of valuations $\pi : \nu_0 \arrow{\transition}{}  \ldots \arrow{\transition}{} \nu_n$ where $\nu_0\models\initial$. It is accepting if also $\nu_n\models\accepting$.
Each run corresponds to a \emph{trace}, which is a finite sequence
$w(\pi)=\nu_0\proj_{\Xobs}, \ldots,
\nu_n\proj_{\Xobs}$.
%
The \emph{language} of $A$ is the set of traces 
$\mathcal{L}(T) = \{w(\pi)\mid \pi \text{ is a run of } T\}$.

\paragraph{Trace Inclusion.} Let 
$\transition({\Xobs},{\Xinv},{\Xobs}',{\Xinv}')$ 
and
$\mu({\Xobs},{\Xinvz},{\Xobs}',{\Xinvz}')$ be two TS. The \emph{trace
  inclusion problem} asks whether $\mathcal{L}(\mathcal{A})
\subseteq \mathcal{L}(B)$.

\newcommand{\viz}{x}
\newcommand{\inviz}{y}
\newcommand{\domain}{\mathcal {D}}
\newcommand{\dvalue}{d}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Subset Construction and Universality}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For simplicity, let us have one visible variable $x$ and one invisible variable $y$.
A trace is then just a sequence $d_0\cdots d_k$ of elements of the domain.

This is a formula describing a relation between the current state after $k$ steps (that is, the values of $\viz$ and $\inviz$) and the global trace leading to it (that is, the values of $\viz_0,\ldots,\viz_k$):
$$
M_k(\viz_0,\ldots,\viz_k,\inviz,\viz) \equiv \exists \inviz_0\cdots \inviz_{k}:  I[x,y/x_0,y_0] \land \bigwedge_{i=1}^k \transition[\viz,\inviz,\viz',\inviz'/\viz_{i-1},\inviz_{i-1},\viz_i,\inviz_i] \land \viz = \viz_k \land \inviz = \inviz_k 
$$
%We note that in our automata, the state is determined by the last transition taken. 
We note that in the formula, the current state is the $k$th value of the variables, which is copied as into the unindexed variables $x,y$.  
%our automata, the state is determined by the last transition taken, and in the formu.

Every global trace gives a macrostate (subset of states of the system) to which the normal subset construction would arrive via that trace.
That is, for every $k$-tuple (trace) $w = \dvalue_0\cdots\dvalue_k \in \domain^k$, the formula 
$$M_{w}(\viz,\inviz) \equiv \exists \viz_0\cdots \viz_{k}: M_k \land \bigwedge_{i=0}^k{\viz_i = \dvalue_i}$$ 
or equivalently
$$M_{w}(\viz,\inviz) \equiv M_k [\viz_0,\ldots,\viz_k/\dvalue_0,\ldots,\dvalue_k] $$ 
is one macrostate.

The relation between macrostates and the traces up-to $k$ steps by which they are reached is captured by the formula  
$$\bigvee_{i=0}^k M_k$$

That all macrostates reached by traces of the lenght $k+1$ have already been reached by shorter traces should be expressed by the formula
$$
\mathit{fixpoint}_k = \forall \viz_0'\cdots\viz_{k+1}'\exists \viz_0\cdots\viz_{k} :
M_{k+1}[\viz_0,\ldots,\viz_{k+1}/\viz'_0,\ldots,\viz'_{k+1}]
\iff 
 \bigvee_{i=0}^k M_k 
$$

So the symbolic version of the standart subset construction would generate the sequence $M_0,M_1,\ldots$ and terminate with $\bigvee_{i=0}^k M_k$ when $\mathit{fixpoint}_k$ holds.    
Of course it does not have to terminate and we will try to use predicate abstraction to make it terminate more often.

A universality test would ask whether every reachable macrostate contains an accepting valuation: 
$$
\forall \viz_0\cdots\viz_k \exists x y:  \bigvee_{i=0}^k M_k \land \accepting
$$

For the purpose of the universality checking,
the fixpoint condition can be weakened in the spirit of antichains:
$$
\mathit{fixpoint}'_k = \forall \viz_0'\cdots\viz_{k+1}'\exists \viz_0\cdots\viz_{k} :
M_{k+1}[\viz_0,\ldots,\viz_{k+1}/\viz'_0,\ldots,\viz'_{k+1}]
\Leftarrow
\bigvee_{i=0}^k M_k 
$$
That is, for every newly generated macroste, there already has been a smaller one.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{splncs03}
\bibliography{ref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-07-17 symbolic_nfa_universality/symbolic_subset_construction.tex
some excercise on emptiness

@@
-\documentclass{llncs}
+\documentclass[10pt]{llncs}
+%\documentclass[acmsmall]{acmart}
 
-\usepackage[usenames,dvipsnames]{color}
+%\usepackage[usenames,dvipsnames]{color}
 
-\let\iint\undefined 
-\let\iiint\undefined 
-\let\iiiint\undefined 
-\let\idotsint\undefined
+%\let\iint\undefined 
+%\let\iiint\undefined 
+%\let\iiiint\undefined 
+%\let\idotsint\undefined
 
@@
 \usepackage{amssymb}
-\usepackage{mathabx}
+%\usepackage{mathabx}
 \usepackage{mathtools}
 \usepackage{textcomp}
+\usepackage{xifthen}
 
 \usepackage{latexsym}
-\usepackage{pslatex}
-\usepackage{epsfig}
-\usepackage{wrapfig}
+%\usepackage{pslatex}
+%\usepackage{epsfig}
+%\usepackage{wrapfig}
 %\usepackage{graphics}
 \usepackage{enumerate}
-\usepackage{cancel}
+%\usepackage{cancel}
 
@@
 
-\usepackage{ listings }
+%\usepackage{ listings }
 
-\usepackage{algorithm}
-\usepackage{algorithmicx}
-\usepackage[noend]{algpseudocode}
-\let\Asterisk\undefined
+%\usepackage{algorithm}
+%\usepackage{algorithmicx}
+%\usepackage[noend]{algpseudocode}
+%\let\Asterisk\undefined
+
+\usepackage{xparse}
+\usepackage{pdfsync}
 
@@
 
-\include{commands}
+%needed macros from the coopy-pasted preliminaries
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newcommand{\nat}{{\mathbb N}}
+%\newcommand{\set}[1]{\left\{ #1 \right\}}
+\newcommand{\true}{\mathtt{true}}
+\newcommand{\false}{\mathtt{false}}
+\newcommand{\tuple}[1]{\left\langle #1 \right\rangle}
+\newcommand{\theo}{\mbox{Th}}
+\newcommand{\thd}{\theo(\mathcal{D})}
+\newcommand{\modelsthd}{\models_{\mbox{\tiny Th}(\scriptscriptstyle\mathcal{D})}}
+\def\proj{\mathbin{\downarrow}}
 
-\newcommand{\Xobs}{\vec{x}}
-\newcommand{\Xinv}{\vec{y}}
-\newcommand{\Xinvz}{\vec{z}}
+\DeclareDocumentCommand\vect{ m o o }{%
+    {\bar #1%
+        \IfNoValueF {#2} {%
+			\IfNoValueT	{#3} {_{(#2)}}%
+			\IfNoValueF {#3} {^{(#2,#3)}}%
+		}%
+    }%
+}
 
-\newcommand{\vx}{\vec{x}}
-\newcommand{\vy}{\vec{y}}
-\newcommand{\vz}{\vec{z}}
-\newcommand{\dom}{\mathcal{D}}
+\newcommand{\set}{\vect}
+%\DeclareDocumentCommand\set{ g g }{%
+%    {%	
+%		\IfNoValueT {#2} {\{\bar #1\}}%
+%       	\IfNoValueF {#2} {\{\bar #1_{(#2)}\}}%
+%    }%
+%}
+
+\newcommand{\subst}[3]{#1[#2/#3]}
+\newcommand{\Nonempty}{\mathit{Nonemp}}
+\newcommand{\Universal}[1]{\mathit{Univ_{#1}}}
+%\newcommand{\Empty}{\mathit{Emp}}
+\newcommand{\Language}[1]{\mathit{Lang_{#1}}}
+\newcommand{\Path}[1]{\mathit{Path_{#1}}}
+\newcommand{\Run}[1]{\mathit{Run_{#1}}}
+\newcommand{\Prefix}[1]{\mathit{Pref_{#1}}}
+\newcommand{\Suffix}[1]{\mathit{Suff_{#1}}}
+\newcommand{\States}[1]{\mathit{States_{#1}}}
+
+\newcommand{\arrow}[2]{\xrightarrow{{\scriptstyle #1}}_{{\scriptstyle #2}}}
+\def\projobs{\mathbin{\downarrow_{\Xobs}}}
+
+\newcommand{\viz}{x}
+\newcommand{\inviz}{y}
+\newcommand{\domain}{\mathcal {D}}
+\newcommand{\lang}{\mathcal {L}}
+\newcommand{\dvalue}{d}
+
+\newcommand{\Xobs}{\vect{x}}
+\newcommand{\Xinv}{\vect{y}}
+\newcommand{\Xinvz}{\vect{z}}
+
+\newcommand{\vx}{\vect{x}}
+\newcommand{\vy}{\vect{y}}
+\newcommand{\vz}{\vect{z}}
+
+\newcommand{\initial}{I}
+\newcommand{\final}{F}
+\newcommand{\accepting}{F}
+\newcommand{\transition}{\tau}
+
+\newcommand{\fdef}{\mathrel{:=}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\include{commands}
 
@@
 
-%\title{Trace Inclusion of Infinite State Systems with Local Information}
 \title{Predicate Abstraction for Trace Inclusion}
@@
  \maketitle
+% \titlepage
 
-\begin{abstract}\end{abstract}
+\begin{abstract}
+A sketch of a subset construction for transition systems with hidden variables.
+\end{abstract}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
+%\section{Introduction}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Treansition Syatems}
+\section{Transition Systems}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-Let $\nat$ denote the set of non-negative integers including zero. For
-any $k,\ell \in \nat$, $k \leq \ell$, we write $[k,\ell]$ for the set
-$\set{k,k+1,\ldots,\ell}$. We write $\false$ and $\true$ for the
-boolean constants \emph{false} and \emph{true}, respectively. Given a
-possibly infinite data domain $\mathcal{D}$, we denote by
-$\thd=\tuple{\mathcal{D}, p_1,\ldots,p_n, f_1,\ldots,f_m}$ the set of
-syntactically correct first-order formulae with predicate
-symbols $p_1,\ldots,p_n$ and function symbols $f_1,\ldots,f_m$. A variable $x$
-is said to be \emph{free} in a~formula $\phi$, denoted as $\phi(x)$,
+Vectors $x_1,\ldots,x_k,k\in\nat$ of indexed variables are denoted by $\vect{x}$ and we use $\vect{x}[k]$ to denote the particular vector $x_1,\ldots,x_k$.
+%
+We will sometimes abuse the notation and write $\vect{x}$ to interchangeably denote 
+the vector and also the set $\{x_1,\ldots,x_k\}$ of variables appearing in it.
+%
+The primed variant $x_1',\ldots,x_k'$ of the vector $\vect{x}$ is denoted by $\vect{x}'$.
+%
+Let $\nat$ denote the set of non-negative integers including zero. 
+%
+%For
+%any $k,\ell \in \nat$, $k \leq \ell$, we write $[k,\ell]$ for the set
+%$\set{k,k+1,\ldots,\ell}$. 
+We use write $\false$ and $\true$ for the
+boolean constants \emph{false} and \emph{true}, respectively. 
+%
+Let $\theo$ denote the set of all syntactically correct formulae of some first order theory. 
+%
+%Given a
+%(possibly infinite) data domain $\mathcal{D}$, 
+%we denote by $\theo$  =\tuple{\mathcal{D}, p_1,\ldots,p_n, f_1,\ldots,f_m}$ the set of
+%syntactically correct first-order formulae with predicate
+%symbols $p_1,\ldots,p_n$ and function symbols $f_1,\ldots,f_m$. 
+%
+A variable $x$
+is said to be \emph{free} in a~formula $\phi\in\theo$
 iff it does not occur under the scope of a~quantifier.
+We write $\phi(\vect{x})$ to denote that the variables of $\vect{x}$ are free in $\phi$.
+%
+Given a vector $\vect \omega = \vect \omega[k]$ of formulae in $\theo$, 
+we denote by $\subst{\phi}{\vect x}{\vect \omega}$ the formula that arises from $\phi$ by substituting all occurrences of $x_i,1\leq i \leq k$, by the formula $\omega_i$.  
+%
+%
+%We use $\vect{x}$ to denote a vector $x_1,\ldots,x_k$ of variables,
+%and we write $\vect{x}{k}$ if we wish to also indicate the $k$. 
+%
+%We will write $\set{x}{k}$ or $\set{x}$ to denote the set $\{x_1,\ldots,x_k\}$.
+%%
+%We use $\vect{x}{k}$ to denote the set $\{x_1,\ldots,x_k\}$ of variables.
+%from which we omit the subscript $(k)$ if $k$ is not relevant. 
+%
+%We use $\vect{x}$ to denote a set $\{x_1,\ldots,x_k\}$ of variables,
+%and we write $\vect{x}{k}$ if we wish to indicate the $k$. 
+%
+%or alternatively $\vect{x}{k}$ if we wish to indicate $k$.
+%
+%We write $\vect{x}$ to denote a vector $x_1,\ldots,x_k$ of variables,
+%or alternatively $\vect{x}{k}$ if we wish to indicate its length $k$.
+%
+%We will sometimes abuse notation and write $\{\vect{x}{n}\}$ to denote the set $\{x_1,\ldots,x_n\}$.
+%
+%We will use the notation $\vect{x}{k}$ interchangeably for vectors and sets $\{x_1,\ldots,x_k\}$.
 
-Let $\vec{x} = \set{x_1,\ldots,x_n}$ be a finite set of variables. A
-\emph{valuation} $\nu : \vec{x} \rightarrow \mathcal{D}$ is an
-assignment of the variables in $\vec{x}$ with values from
-$\mathcal{D}$. We denote by $\mathcal{D}^\vec{x}$ the set of such
-valuations. For a formula $\phi(\vec{x})$, we denote by $\nu
-\modelsthd \phi$ the fact that substituting each variable $x \in
-\vec{x}$ by $\nu(x)$ yields a valid formula in the theory $\thd$. In
-this case, $\nu$ is said to be a~\emph{model} of $\phi$. A formula is
-said to be {\em satisfiable} iff it has a model. For a~formula
-$\phi(\vec{x}, \vec{x'})$ where $\vec{x'} = \set{x' ~|~ x \in
-  \vec{x}}$ and two valuations $\nu,\nu' \in \mathcal{D}^{\vec{x}}$,
-we denote by $(\nu,\nu') \modelsthd \phi$ the fact that the formula
-obtained from $\phi$ by substituting each $x$ with $\nu(x)$ and each
-$x'$ with $\nu'(x')$ is valid in $\thd$.
+Let $\vect{x} = \vect{x}[k]$ and
+%
+let $\thd$ be an interpretation of the first order theory with a possibly infinite domain $\domain$. 
+%
+A \emph{valuation} $\nu : \set{x} \rightarrow \domain$ is an
+assignment of the variables in $\vect{x}$ with values from
+$\domain$. We denote by $\domain^{\set{x}}$ the set of such
+valuations. 
+%
+For a formula $\phi(\vect{x})\in\theo$, we denote by $\nu
+\modelsthd \phi$ that $\nu$ satisfies $\phi$ under the interpretation $\thd$. 
+%the fact that substituting each variable $x \in
+%\vect{x}$ by $\nu(x)$ yields a valid formula in the theory $\thd$. 
+%
+%For a formula $\phi(\vect{x})$, we denote by $\nu
+%\modelsthd \phi$ the fact that substituting each variable $x \in
+%\vect{x}$ by $\nu(x)$ yields a valid formula in the theory $\thd$. 
+%
+%In
+%this case, $\nu$ is said to be a~\emph{model} of $\phi$. A formula is
+%said to be {\em satisfiable} iff it has a model. 
+%
+%For a~formula
+%$\phi(\vect{x}, \vect{x}')$ where $\vect{x}' = \set{x' ~|~ x \in \vect{x}}$ and two valuations $\nu,\nu' \in \domain^{\vect{x}}$,
+%we denote by $(\nu,\nu') \modelsthd \phi$ the fact that the formula
+%obtained from $\phi$ by substituting each $x$ with $\nu(x)$ and each
+%$x'$ with $\nu'(x')$ is valid in $\thd$.
+%
 
-\newcommand{\initial}{I}
-\newcommand{\accepting}{F}
-\newcommand{\transition}{\tau}
 \paragraph{Transition system.} 
-Let $\mathcal{D}$ be an infinite date domain equipped with a first order theory $\thd$.
+%Let $\domain$ be an infinite data domain equipped with a first order theory $\thd$.
 %
-Let $\Xobs$ be the set of observable and $\Xinv$ the set of invisible variables.
+Let $\vect{x} = \vect{x}[k]$ be a vector of visible and $\vect{y}=\vect{y}[\ell]$ a vector of invisible variables.
 %
-A transition system (TS) is a triple formula $T = (\initial(\Xobs,\Xinv),\transition(\Xobs,\Xinv,\Xobs', \Xinv'),\accepting(\Xobs,\Xinv))$ where $\initial$ is a fomrula describing initial valuations, $\transition(\Xobs,\Xinv,\Xobs', \Xinv')$ describes the transition relation, and $\accepting(\Xobs,\Xinv)$ describes accepting valuations.
+A transition system is a triple $T = (\initial(\Xobs,\Xinv),\transition(\Xobs,\Xinv,\Xobs', \Xinv'),\accepting(\Xobs,\Xinv))$ where $\initial$ is the initial formula, 
+$\transition(\Xobs,\Xinv,\Xobs', \Xinv')$ is the transition formula, 
+and $\accepting(\Xobs,\Xinv)$ is the final formula.
 %
 We say that a valuation $\nu'$ is 
-a~\emph{successor} of $\nu$ if and only if and $(\nu,\nu') \modelsthd
-\transition$. We denote the successor relation by $\nu
-\arrow{\transition}{} \nu'$, and we omit writing $\transition$ 
-when no confusion may arise. We write 
-$\mathit{succ}(\nu)$ to denote $\nu'$.
+a~\emph{successor} of $\nu$ if and only if and $(\nu\cup\nu') \modelsthd
+\transition$, denoted by $\nu
+\arrow{\transition}{} \nu'$, from where we omit the $\transition$ 
+when no confusion may arise. 
+%We write 
+%$\mathit{succ}(\nu)$ to denote $\nu'$.
 
 A \emph{run} of $\transition$ is a sequence
-of valuations $\pi : \nu_0 \arrow{\transition}{}  \ldots \arrow{\transition}{} \nu_n$ where $\nu_0\models\initial$. It is accepting if also $\nu_n\models\accepting$.
+of valuations $\pi : \nu_0 \arrow{\transition}{}  \cdots \arrow{\transition}{} \nu_n$ with $\nu_0\models\initial$. 
+It is accepting if also $\nu_n\models\accepting$.
 Each run corresponds to a \emph{trace}, which is a finite sequence
 $w(\pi)=\nu_0\proj_{\Xobs}, \ldots,
-\nu_n\proj_{\Xobs}$.
+\nu_n\proj_{\Xobs}$, where $\nu_i\proj_{\Xobs}{}$ denotes the restriction of $\nu_i$ to $\Xobs$.
 %
-The \emph{language} of $A$ is the set of traces 
-$\mathcal{L}(T) = \{w(\pi)\mid \pi \text{ is a run of } T\}$.
+The \emph{language} of $T$ is the set of accepted traces 
+$\lang(T) = \{w(\pi)\mid \pi \text{ is an accepting run of } T\}$.
 
-\paragraph{Trace Inclusion.} Let 
-$\transition({\Xobs},{\Xinv},{\Xobs}',{\Xinv}')$ 
-and
-$\mu({\Xobs},{\Xinvz},{\Xobs}',{\Xinvz}')$ be two TS. The \emph{trace
-  inclusion problem} asks whether $\mathcal{L}(\mathcal{A})
-\subseteq \mathcal{L}(B)$.
+%\paragraph{Trace Inclusion.} Let 
+%$\transition({\Xobs},{\Xinv},{\Xobs}',{\Xinv}')$ 
+%and
+%$\mu({\Xobs},{\Xinvz},{\Xobs}',{\Xinvz}')$ be two TS. The \emph{trace
+%  inclusion problem} asks whether $\mathcal{L}(\mathcal{A})
+%\subseteq \mathcal{L}(B)$.
 
-\newcommand{\viz}{x}
-\newcommand{\inviz}{y}
-\newcommand{\domain}{\mathcal {D}}
-\newcommand{\dvalue}{d}
+\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Subset Construction and Universality}
+\section{Emptiness and Universality of Transition Systems}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-For simplicity, let us have one visible variable $x$ and one invisible variable $y$.
+For simplicity, let us assume wlog. a transition system $T$ with only one visible variable $x$ and only one invisible variable $y$.
 A trace is then just a sequence $d_0\cdots d_k$ of elements of the domain.
+%
+We will elaborate on the problems of emptiness and universality of $T$, that is, the problems of deciding whether $\lang(T) = \emptyset$ and  whether $\lang(T) = \domain^*$.
 
+For a formula $\phi(x,y,x',y')$ over variables $\vect{z}$ and their primed versions $\vect z'$, we write $\phi_i,i\geq 0$ to denote the formula $\subst{\phi}{x,y,x',y'}{x_i,y_i,x_{i+1},y_{i+1}}$ where the unprimed variables are indexed by $i$ and the primed by $i+1$.
+%
+
+All runs of the length $k\geq 0$ are described by the formula 
+$$
+\Run k(\vect x[k],\vect y[k]) \fdef
+\initial_0(x_0,y_0) \land \bigwedge_{i=1}^{k} \tau_{i-1}(x_{i-1},y_{i-1},x_{i},y_{i})
+$$
+The language of $T$ would correspond to the union over all $k\geq 0$ of the satisfying assignments of formulae $\Language k$  that describe traces of the accepting runs of the length $k$:  
+$$
+\Language k(\vect x[k]) \fdef  \exists \vect y[k]:\Run k(\vect x[k],\vect y[k]) \land \final(x_k,y_k) \ \ .
+$$
+%The correspondence is a bit weird since every trace in $w\in\lang(T)$ corresponds to all assignments $\nu:\vect x[\infty]\rightarrow \domain$ such that $w = \nu(x_1)\cdots \nu(x_{|w|})$. 
+%
+We have that $T$ is nonempty iff for some $k\geq 0$, $\Language k$ is satisfiable,
+ and that $T$ is universal iff $\Language k$ is valid for all $k\geq 0$.
+
+\paragraph{Emptiness}
+Emptiness can be solved by standard techniques of exploration of the space of all reachable assignments.
+As an excercise, let us ponder about what this corresponds to from the perspective of testing satisfiability of the infinite disjunction
+$$
+%\begin{multline*}
+\Nonempty \fdef
+\bigvee_{k\geq 0} \exists \vect x[k]:\Language k(\vect x[k])
+=
+%\bigvee_{k\geq 0} \exists \vect x[k]\,\exists \vect y[k]:\Run k(\vect x[k])
+% = 
+%\\ = 
+%\bigvee_{k\geq 0} \exists \vect x[k]\,\exists \vect y[k]:\initial_0(x_0,y_0) \land \bigwedge_{i=0}^{k-1} \tau_i(x_i,y_i,x_{i+1},y_{i+1}) \land \final_k(x_{k},y_k) 
+%\bigvee_{k\geq 0} \exists \vect x[k]\,\exists \vect y[k]:\Run k(\vect x[k])
+% = 
+%\\ = 
+\bigvee_{k\geq 0} \exists \vect x[k]\,\exists \vect y[k]:\Run k(\vect x[k]\,\vect y[k]) \land \final_k(x_{k},y_k)\ \ . 
+%\end{multline*}
+$$
+
+For any $0\geq k$, the $k$th disjunct is equivalent to 
+$$
+\exists xy: \subst{(\exists \vect x[k-1]\vect y[k-1]:\Run {k}(\vect x[k]\vect y[k]))}{x_k,y_k}{x,y}
+\land 
+\final(x,y)
+$$
+where the inner formula represents the set of states reached after $k$ steps.
+Let us denote it $\States k(x,y)$. $\Nonempty$ is therefore equivalent to
+$\exists xy:\bigvee_{k\geq 0} \States k(x,y) \land \final(x,y)$, that is, it is enough to compute a formula representing states reachable by runs of all lengths and test existence of an accepting one.
+%% 
+%% For any $0\geq k$, the $k$th disjunct is equivalent to 
+%% $$
+%% \exists x_{k}y_{k}: (\exists \vect x[k-1]\vect y[k-1]:\Run {k})
+%% \land 
+%% \final_k
+%% $$
+%% where the inner formula represents the set of states reached after $k$ steps.
+%% Let us denote it $\States k(x_k,y_k)$. $\Nonempty$ is therefore equivalent to
+%% $\bigvee_{k\geq 0} \exists \vect x[k]\,\exists\vect y[k]:\States k(x_k,y_k) \land \final_k(x_k,y_k)$, that is, it is enough to compute a formula representing states reachable by runs of all lengths and test existence of an accepting one.
+%% %
+%% To compute all reachable states, we use that 
+%% $$
+%% \States k \equiv \exists x_{k-1}y_{k-1}:\States {k-1} 
+%% \land 
+%% \transition_{k-1}
+%% \ \ .
+%% $$
+%% That is, reachable states at length $k$ can be computed from those reached at length $k-1$ by applying the transition relation.  
+%% %
+%% So it is enough to compute a formula that describes reached states that is stable under the transition relation:
+%% $$\States {\leq n} \fdef \bigvee_{k=0}^n \subst{\States k}{x_k,y_k}{x,y}$$ 
+%% such that 
+%% $$\States {\leq n}  \equiv \subst{(\exists x,y: \States {\leq n}\land \transition)}{x',y'}{x,y}\ \ .$$
+%
+To compute all reachable states, we use that 
+$$
+\States k \equiv \subst{(\exists xy:\States {k-1}(x,y)
+\land 
+\transition)}{x',y'}{x,y}
+\ \ .
+$$
+That is, reachable states at length $k$ can be computed from those reached at length $k-1$ by applying the transition relation.  
+%
+So it is enough to compute a formula that describes reached states that is stable under the transition relation:
+$$\States {\leq n} \fdef \bigvee_{k=0}^n \States k$$ 
+such that 
+$$\States {\leq n}  \equiv \subst{(\exists x,y: \States {\leq n}\land \transition)}{x',y'}{x,y}\ \ .$$
+
+%% PREFIX SUFFIX
+%% $$
+%% \exists x_jy_jx_{j+1}y_{j+1} :\Prefix j(x_j,y_j) 
+%% \land 
+%% \transition_j(x_j,y_j,x_{j+1},y_{j+1})
+%% \land 
+%% \Suffix {j+1}(x_{j+1},y_{j+1})
+%% $$
+%% where
+%% \begin{align*}
+%% \Prefix j(\vect x[j],\vect y[j]) &\fdef \exists\vect x[j-1]\,\exists \vect y[j-1]:\initial_0(x_0,y_0)\land\bigwedge_{i=0}^{j-1} \tau_i(x_i,y_i,x_{i+1},y_{i+1})
+%% \\
+%% \Suffix {j+1}(\vect x[j+1,k],\vect y[j+1,k]) &\fdef \exists\vect x[j+2,k]\,\exists \vect y[j+2,k]:\bigwedge_{i=j+1}^{k-1} \tau_i(x_i,y_i,x_{i+1},y_{i+1})\land\final_k(x_k,y_k)
+%% \end{align*}
+
+\paragraph{Universality (TODO)}
+We will sketch an incomplete universality check that performs a symbolic version of the standard subset construction.
+It does not use any abstraction or acceleration, so it is easy to come up with transition systems on which it does not terminate.
+Some kind of predicate abstraction needs to be added.
+universality of $T$ corresponds to the validity of each formula $\Universal k$ for $k\geq 0$ where
+$$
+\Universal k \fdef
+\forall \vect x[k]\, \exists \vect y[k]:\Run k(\vect x[k],\vect y[k]) \ \ .
+$$
+
+As normal subset construction, it is based on computing the set of all macrostates that $T$ can reach,
+where a macrostate is the set of all states  of $T$ (a state is an assignment $\nu$ to $x$ and $y$) that can be reached via a single trace.
+%
+
+\newpage
+\section{Old Stuff}
 This is a formula describing a relation between the current state after $k$ steps (that is, the values of $\viz$ and $\inviz$) and the global trace leading to it (that is, the values of $\viz_0,\ldots,\viz_k$):
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{splncs03}
-\bibliography{ref}
+%\bibliographystyle{splncs03}
+%\bibliography{ref}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-07-31 symbolic_nfa_universality/symbolic_subset_construction.tex
an excercise on emptiness and universality of transition systems

@@
-\documentclass[10pt]{llncs}
-%\documentclass[acmsmall]{acmart}
+%\documentclass{llncs}
+\documentclass[acmsmall]{acmart}
 
@@
 \usepackage{xifthen}
+%\usepackage{silence}
+%\WarningsOff
 
@@
 \newcommand{\nat}{{\mathbb N}}
+\newcommand{\domain}{\mathcal {D}}
 %\newcommand{\set}[1]{\left\{ #1 \right\}}
@@
 \newcommand{\tuple}[1]{\left\langle #1 \right\rangle}
-\newcommand{\theo}{\mbox{Th}}
+\newcommand{\theo}{\mathsf{Th}}
 \newcommand{\thd}{\theo(\mathcal{D})}
-\newcommand{\modelsthd}{\models_{\mbox{\tiny Th}(\scriptscriptstyle\mathcal{D})}}
+\newcommand{\modelsthd}{\models_{\thd}}
 \def\proj{\mathbin{\downarrow}}
@@
 \DeclareDocumentCommand\vect{ m o o }{%
-    {\bar #1%
+    {% 
+        \IfNoValueT {#2} {\bar #1} 
         \IfNoValueF {#2} {%
-			\IfNoValueT	{#3} {_{(#2)}}%
-			\IfNoValueF {#3} {^{(#2,#3)}}%
+			\IfNoValueT	{#3} {\bar #1_{#2}}%
+			%\IfNoValueF {#3} {^{(#2,#3)}}%
+			\IfNoValueF {#3} {\bar #1^{(#2,#3)}}%
 		}%
@@
 
-\newcommand{\subst}[3]{#1[#2/#3]}
+\newcommand{\subst}[3]{#1[#2/#3]}%substitute in #1 occurences of #2 by #3
 \newcommand{\Nonempty}{\mathit{Nonemp}}
-\newcommand{\Universal}[1]{\mathit{Univ_{#1}}}
+\newcommand{\Universal}{\mathit{Univ}}
+\newcommand{\Nonuniversal}{\mathit{Nonuniv}}
 %\newcommand{\Empty}{\mathit{Emp}}
@@
 \newcommand{\States}[1]{\mathit{States_{#1}}}
+\newcommand{\Configurations}[1]{\mathit{Confs_{#1}}}
+\newcommand{\Configuration}[1]{\mathit{Conf_{#1}}}
 
+%\newcommand{\arrow}[2]{\xrightarrow{_{_{#1}}}_{{\scriptstyle #2}}}
 \newcommand{\arrow}[2]{\xrightarrow{{\scriptstyle #1}}_{{\scriptstyle #2}}}
+%\newcommand{\arrow}[2]{\rightarrow_{{#1}}^{{\scriptstyle #2}}}
 \def\projobs{\mathbin{\downarrow_{\Xobs}}}
@@
 \newcommand{\inviz}{y}
-\newcommand{\domain}{\mathcal {D}}
 \newcommand{\lang}{\mathcal {L}}
-\newcommand{\dvalue}{d}
+\newcommand{\dvalue}{v}
+\newcommand{\dval}{\dvalue}
 
@@
 
-\title{Predicate Abstraction for Trace Inclusion}
+%\title{Predicate Abstraction for Trace Inclusion}
 
-\author{Luk\'{a}\v{s} Hol\'{i}k\inst{1} \and Radu Iosif\inst{2} \and Adam Rogalewicz\inst{1} \and
-Tom\'{a}\v{s}~Vojnar\inst{1}}
+%\author{Luk\'{a}\v{s} Hol\'{i}k\inst{1} \and Radu Iosif\inst{2} \and Adam Rogalewicz\inst{1} \and
+%Tom\'{a}\v{s}~Vojnar\inst{1}}
 
-\institute{
-  FIT, Brno University of Technology, IT4Innovations Centre of Excellence, 
-  Czech Republic
-  \and 
-  University Grenoble Alpes, CNRS, VERIMAG, Grenoble, France
-}
+%\institute{
+%  FIT, Brno University of Technology, IT4Innovations Centre of Excellence, 
+%  Czech Republic
+%  \and 
+%  University Grenoble Alpes, CNRS, VERIMAG, Grenoble, France
+%}
 
- \maketitle
+% \maketitle
 % \titlepage
@@
 %
-%For
-%any $k,\ell \in \nat$, $k \leq \ell$, we write $[k,\ell]$ for the set
-%$\set{k,k+1,\ldots,\ell}$. 
-We use write $\false$ and $\true$ for the
-boolean constants \emph{false} and \emph{true}, respectively. 
+We write $\false$ and $\true$ to denote the two boolean constants. 
 %
 Let $\theo$ denote the set of all syntactically correct formulae of some first order theory. 
-%
-%Given a
-%(possibly infinite) data domain $\mathcal{D}$, 
-%we denote by $\theo$  =\tuple{\mathcal{D}, p_1,\ldots,p_n, f_1,\ldots,f_m}$ the set of
-%syntactically correct first-order formulae with predicate
-%symbols $p_1,\ldots,p_n$ and function symbols $f_1,\ldots,f_m$. 
 %
@@
 %
+
+%Let $\vect{x} = \vect{x}[k]$ and
 %
-%We use $\vect{x}$ to denote a vector $x_1,\ldots,x_k$ of variables,
-%and we write $\vect{x}{k}$ if we wish to also indicate the $k$. 
-%
-%We will write $\set{x}{k}$ or $\set{x}$ to denote the set $\{x_1,\ldots,x_k\}$.
-%%
-%We use $\vect{x}{k}$ to denote the set $\{x_1,\ldots,x_k\}$ of variables.
-%from which we omit the subscript $(k)$ if $k$ is not relevant. 
-%
-%We use $\vect{x}$ to denote a set $\{x_1,\ldots,x_k\}$ of variables,
-%and we write $\vect{x}{k}$ if we wish to indicate the $k$. 
-%
-%or alternatively $\vect{x}{k}$ if we wish to indicate $k$.
-%
-%We write $\vect{x}$ to denote a vector $x_1,\ldots,x_k$ of variables,
-%or alternatively $\vect{x}{k}$ if we wish to indicate its length $k$.
-%
-%We will sometimes abuse notation and write $\{\vect{x}{n}\}$ to denote the set $\{x_1,\ldots,x_n\}$.
-%
-%We will use the notation $\vect{x}{k}$ interchangeably for vectors and sets $\{x_1,\ldots,x_k\}$.
-
-Let $\vect{x} = \vect{x}[k]$ and
-%
-let $\thd$ be an interpretation of the first order theory with a possibly infinite domain $\domain$. 
+Let $\thd$ be an interpretation of the first order theory $\theo$ with a possibly infinite domain $\domain$. 
 %
 A \emph{valuation} $\nu : \set{x} \rightarrow \domain$ is an
-assignment of the variables in $\vect{x}$ with values from
+assignment of the variables in a vector $\vect{x}$ with values from
 $\domain$. We denote by $\domain^{\set{x}}$ the set of such
@@
 \modelsthd \phi$ that $\nu$ satisfies $\phi$ under the interpretation $\thd$.
-%the fact that substituting each variable $x \in
-%\vect{x}$ by $\nu(x)$ yields a valid formula in the theory $\thd$. 
-%
-%For a formula $\phi(\vect{x})$, we denote by $\nu
-%\modelsthd \phi$ the fact that substituting each variable $x \in
-%\vect{x}$ by $\nu(x)$ yields a valid formula in the theory $\thd$. 
-%
-%In
-%this case, $\nu$ is said to be a~\emph{model} of $\phi$. A formula is
-%said to be {\em satisfiable} iff it has a model. 
-%
-%For a~formula
-%$\phi(\vect{x}, \vect{x}')$ where $\vect{x}' = \set{x' ~|~ x \in \vect{x}}$ and two valuations $\nu,\nu' \in \domain^{\vect{x}}$,
-%we denote by $(\nu,\nu') \modelsthd \phi$ the fact that the formula
-%obtained from $\phi$ by substituting each $x$ with $\nu(x)$ and each
-%$x'$ with $\nu'(x')$ is valid in $\thd$.
-%
 
 \paragraph{Transition system.} 
-%Let $\domain$ be an infinite data domain equipped with a first order theory $\thd$.
 %
-Let $\vect{x} = \vect{x}[k]$ be a vector of visible and $\vect{y}=\vect{y}[\ell]$ a vector of invisible variables.
+%Let $\vect{x} = \vect{x}[k]$ be a vector of visible and $\vect{y}=\vect{y}[\ell]$ a vector of invisible variables.
+Let $\vect{x}$ be a vector of visible and $\vect{y}$ a vector of invisible variables.
 %
@@
 \transition$, denoted by $\nu
-\arrow{\transition}{} \nu'$, from where we omit the $\transition$ 
-when no confusion may arise. 
-%We write 
-%$\mathit{succ}(\nu)$ to denote $\nu'$.
+\arrow{\transition}{} \nu'$. 
+%, from where we omit the $\transition$ when no confusion may arise.
 
@@
 
-%\paragraph{Trace Inclusion.} Let 
-%$\transition({\Xobs},{\Xinv},{\Xobs}',{\Xinv}')$ 
-%and
-%$\mu({\Xobs},{\Xinvz},{\Xobs}',{\Xinvz}')$ be two TS. The \emph{trace
-%  inclusion problem} asks whether $\mathcal{L}(\mathcal{A})
-%\subseteq \mathcal{L}(B)$.
-
-\newpage
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 For simplicity, let us assume wlog. a transition system $T$ with only one visible variable $x$ and only one invisible variable $y$.
-A trace is then just a sequence $d_0\cdots d_k$ of elements of the domain.
+A trace is then just a sequence $d_0,\ldots, d_k$ of elements of the domain.
 %
@@
 
-For a formula $\phi(x,y,x',y')$ over variables $\vect{z}$ and their primed versions $\vect z'$, we write $\phi_i,i\geq 0$ to denote the formula $\subst{\phi}{x,y,x',y'}{x_i,y_i,x_{i+1},y_{i+1}}$ where the unprimed variables are indexed by $i$ and the primed by $i+1$.
+For a formula $\phi(x,y,x',y')$ over variables $\vect{z}$ and their primed variants $\vect z'$, we write $\phi_i,i\geq 0$ to denote the formula $\subst{\phi}{x,y,x',y'}{x_i,y_i,x_{i+1},y_{i+1}}$ where the unprimed variables are indexed by $i$ and the primed by $i+1$.
 %
@@
 $$
-%The correspondence is a bit weird since every trace in $w\in\lang(T)$ corresponds to all assignments $\nu:\vect x[\infty]\rightarrow \domain$ such that $w = \nu(x_1)\cdots \nu(x_{|w|})$. 
 %
@@
 
-\paragraph{Emptiness}
-Emptiness can be solved by standard techniques of exploration of the space of all reachable assignments.
-As an excercise, let us ponder about what this corresponds to from the perspective of testing satisfiability of the infinite disjunction
+\subsection{Emptiness}
+As an exercise, before we elaborate on universality checking, we discuss language emptiness.  
+Emptiness corresponds to unsatisfiability of the formula
 $$
-%\begin{multline*}
 \Nonempty \fdef
@@
 =
-%\bigvee_{k\geq 0} \exists \vect x[k]\,\exists \vect y[k]:\Run k(\vect x[k])
-% = 
-%\\ = 
-%\bigvee_{k\geq 0} \exists \vect x[k]\,\exists \vect y[k]:\initial_0(x_0,y_0) \land \bigwedge_{i=0}^{k-1} \tau_i(x_i,y_i,x_{i+1},y_{i+1}) \land \final_k(x_{k},y_k) 
-%\bigvee_{k\geq 0} \exists \vect x[k]\,\exists \vect y[k]:\Run k(\vect x[k])
-% = 
-%\\ = 
-\bigvee_{k\geq 0} \exists \vect x[k]\,\exists \vect y[k]:\Run k(\vect x[k]\,\vect y[k]) \land \final_k(x_{k},y_k)\ \ . 
-%\end{multline*}
+\bigvee_{k\geq 0} \exists \vect x[k]\,\exists \vect y[k]:\Run k(\vect x[k],\vect y[k]) \land \final_k(x_{k},y_k)\ \ . 
 $$
-
+%
 For any $0\geq k$, the $k$th disjunct is equivalent to 
 $$
-\exists xy: \subst{(\exists \vect x[k-1]\vect y[k-1]:\Run {k}(\vect x[k]\vect y[k]))}{x_k,y_k}{x,y}
+\exists xy: \subst{(\exists \vect x[k-1]\vect y[k-1]:\Run {k}(\vect x[k],\vect y[k]))}{x_k,y_k}{x,y}
 \land 
@@
 Let us denote it $\States k(x,y)$. $\Nonempty$ is therefore equivalent to
-$\exists xy:\bigvee_{k\geq 0} \States k(x,y) \land \final(x,y)$, that is, it is enough to compute a formula representing states reachable by runs of all lengths and test existence of an accepting one.
-%% 
-%% For any $0\geq k$, the $k$th disjunct is equivalent to 
-%% $$
-%% \exists x_{k}y_{k}: (\exists \vect x[k-1]\vect y[k-1]:\Run {k})
-%% \land 
-%% \final_k
-%% $$
-%% where the inner formula represents the set of states reached after $k$ steps.
-%% Let us denote it $\States k(x_k,y_k)$. $\Nonempty$ is therefore equivalent to
-%% $\bigvee_{k\geq 0} \exists \vect x[k]\,\exists\vect y[k]:\States k(x_k,y_k) \land \final_k(x_k,y_k)$, that is, it is enough to compute a formula representing states reachable by runs of all lengths and test existence of an accepting one.
-%% %
-%% To compute all reachable states, we use that 
-%% $$
-%% \States k \equiv \exists x_{k-1}y_{k-1}:\States {k-1} 
-%% \land 
-%% \transition_{k-1}
-%% \ \ .
-%% $$
-%% That is, reachable states at length $k$ can be computed from those reached at length $k-1$ by applying the transition relation.  
-%% %
-%% So it is enough to compute a formula that describes reached states that is stable under the transition relation:
-%% $$\States {\leq n} \fdef \bigvee_{k=0}^n \subst{\States k}{x_k,y_k}{x,y}$$ 
-%% such that 
-%% $$\States {\leq n}  \equiv \subst{(\exists x,y: \States {\leq n}\land \transition)}{x',y'}{x,y}\ \ .$$
-%
+$\exists xy:\bigvee_{k\geq 0} \States k(x,y) \land \final(x,y)$, that is, it is enough to compute a formula representing states reachable at all distances and test existence of an accepting one.
 To compute all reachable states, we use that 
@@
 \land 
-\transition)}{x',y'}{x,y}
+\transition(x,y,x',y'))}{x',y'}{x,y}
 \ \ .
 $$
-That is, reachable states at length $k$ can be computed from those reached at length $k-1$ by applying the transition relation.  
+That is, reachable states at the distance $k$ can be computed from those reached at the distance $k-1$ by applying the transition relation.  
 %
@@
 such that 
-$$\States {\leq n}  \equiv \subst{(\exists x,y: \States {\leq n}\land \transition)}{x',y'}{x,y}\ \ .$$
-
-%% PREFIX SUFFIX
-%% $$
-%% \exists x_jy_jx_{j+1}y_{j+1} :\Prefix j(x_j,y_j) 
-%% \land 
-%% \transition_j(x_j,y_j,x_{j+1},y_{j+1})
-%% \land 
-%% \Suffix {j+1}(x_{j+1},y_{j+1})
-%% $$
-%% where
-%% \begin{align*}
-%% \Prefix j(\vect x[j],\vect y[j]) &\fdef \exists\vect x[j-1]\,\exists \vect y[j-1]:\initial_0(x_0,y_0)\land\bigwedge_{i=0}^{j-1} \tau_i(x_i,y_i,x_{i+1},y_{i+1})
-%% \\
-%% \Suffix {j+1}(\vect x[j+1,k],\vect y[j+1,k]) &\fdef \exists\vect x[j+2,k]\,\exists \vect y[j+2,k]:\bigwedge_{i=j+1}^{k-1} \tau_i(x_i,y_i,x_{i+1},y_{i+1})\land\final_k(x_k,y_k)
-%% \end{align*}
-
-\paragraph{Universality (TODO)}
-We will sketch an incomplete universality check that performs a symbolic version of the standard subset construction.
-It does not use any abstraction or acceleration, so it is easy to come up with transition systems on which it does not terminate.
-Some kind of predicate abstraction needs to be added.
-universality of $T$ corresponds to the validity of each formula $\Universal k$ for $k\geq 0$ where
-$$
-\Universal k \fdef
-\forall \vect x[k]\, \exists \vect y[k]:\Run k(\vect x[k],\vect y[k]) \ \ .
-$$
-
-As normal subset construction, it is based on computing the set of all macrostates that $T$ can reach,
-where a macrostate is the set of all states  of $T$ (a state is an assignment $\nu$ to $x$ and $y$) that can be reached via a single trace.
-%
+$$\States {\leq n}  \equiv \States {\leq n} \lor \subst{(\exists x,y: \States {n}\land \transition)}{x',y'}{x,y}\ \ .$$
 
 \newpage
-\section{Old Stuff}
-This is a formula describing a relation between the current state after $k$ steps (that is, the values of $\viz$ and $\inviz$) and the global trace leading to it (that is, the values of $\viz_0,\ldots,\viz_k$):
+\subsection{Universality}
+We will snow ketch a basic universality check. 
+As in the case of emptiness, it is incomplete due to the infinite data domain, but additionally also due to the need of working with unbounded histories. 
+%
+Universality of $T$ corresponds to the validity the formula
 $$
-M_k(\viz_0,\ldots,\viz_k,\inviz,\viz) \equiv \exists \inviz_0\cdots \inviz_{k}:  I[x,y/x_0,y_0] \land \bigwedge_{i=1}^k \transition[\viz,\inviz,\viz',\inviz'/\viz_{i-1},\inviz_{i-1},\viz_i,\inviz_i] \land \viz = \viz_k \land \inviz = \inviz_k
+\Universal \fdef
+\bigwedge_{k\geq 0} \forall \vect x[k]:\Language k(\vect x[k])
+=
+\bigwedge_{k\geq 0} \forall \vect x[k]\,\exists \vect y[k]:\Run k(\vect x[k],\vect y[k]) \land \final_k(x_{k},y_k) 
 $$
-%We note that in our automata, the state is determined by the last transition taken. 
-We note that in the formula, the current state is the $k$th value of the variables, which is copied as into the unindexed variables $x,y$.  
-%our automata, the state is determined by the last transition taken, and in the formu.
+(the difference from the formula $\Nonempty$ is the leading $\bigwedge$ and the universal quantification over the visible variables). 
+%
+Let 
+$$
+\Configurations k(\vect x[k],x,y) \fdef \exists \vect y[k]:\Run k \land x = x_k \land y = y_k  
+$$
+be a formula which associates a visible trace $\vect x[k]$ of the length $k$ with the resulting values of $x$ and $y$.
+Fixing a particular trace $\vect \dval [k]$ results in a so called \emph{configuration}, 
+a constraint over $x$ and $y$ with $x$ fixed at the last value $\dval_k$: 
+$$
+\subst{\Configurations k}{\vect x[k]}{\vect \dval[k]}(x,y)\ \ .
+$$
+One may hence understand $\Configurations k$ as a function $\lambda \vect x[k].\,\Configurations k$ from traces to configurations. 
+%
+%
+We may rewrite conjunction $\Universal$ as 
+$$
+\Universal 
+\equiv
+\bigwedge_{k\geq 0} \forall \vect x[k]\,\exists y:\Configurations k(\vect x[k],x,y) \land \final(x,y) 
+$$
+That is, every configuration must have $y$ to satisfy $\final$.
+Notice that the universality test does not require all but only the smallest (strongest) configurations.
 
-Every global trace gives a macrostate (subset of states of the system) to which the normal subset construction would arrive via that trace.
-That is, for every $k$-tuple (trace) $w = \dvalue_0\cdots\dvalue_k \in \domain^k$, the formula 
-$$M_{w}(\viz,\inviz) \equiv \exists \viz_0\cdots \viz_{k}: M_k \land \bigwedge_{i=0}^k{\viz_i = \dvalue_i}$$ 
-or equivalently
-$$M_{w}(\viz,\inviz) \equiv M_k [\viz_0,\ldots,\viz_k/\dvalue_0,\ldots,\dvalue_k] $$ 
-is one macrostate.
-
-The relation between macrostates and the traces up-to $k$ steps by which they are reached is captured by the formula  
-$$\bigvee_{i=0}^k M_k$$
-
-That all macrostates reached by traces of the lenght $k+1$ have already been reached by shorter traces should be expressed by the formula
+The task is therefore to compute and test all configurations $\subst{\Configurations {k}}{\vect x[k]}{\vect \dval[k]}$. 
+To compute a representation of all configurations, we use the fact that configurations at the distance $k+1$ can be obtained as successors of those at the distance $k$ as follows. 
+For every visible trace $\vect \dval [k+1]$,
 $$
-\mathit{fixpoint}_k = \forall \viz_0'\cdots\viz_{k+1}'\exists \viz_0\cdots\viz_{k} :
-M_{k+1}[\viz_0,\ldots,\viz_{k+1}/\viz'_0,\ldots,\viz'_{k+1}]
-\iff 
- \bigvee_{i=0}^k M_k 
+\subst{\Configurations {k+1}}{\vect x[k+1]}{\vect \dval[k+1]} \equiv 
+\subst{(\exists{xy}:\subst{\Configurations {k}}{\vect x[k]}{\vect \dval[k]} \land \tau)}{x',y'}{x,y}\land x = \dval_{k+1}\ \ .
 $$
-
-So the symbolic version of the standart subset construction would generate the sequence $M_0,M_1,\ldots$ and terminate with $\bigvee_{i=0}^k M_k$ when $\mathit{fixpoint}_k$ holds.    
-Of course it does not have to terminate and we will try to use predicate abstraction to make it terminate more often.
-
-A universality test would ask whether every reachable macrostate contains an accepting valuation: 
+The set of all configurations can be hence computed as the least fixpoint of the transition relation that includes the initial configurations.
+%
+Notice that the lowest configurations at the distance $k+1$ are successors of the lowest configurations at the distance $k+1$.
+%
+That is, all smallest configurations are found within the distance $k$, i.e.
 $$
-\forall \viz_0\cdots\viz_k \exists x y:  \bigvee_{i=0}^k M_k \land \accepting
+\bigwedge_{\ell\geq 0} \forall \vect x[\ell]:\Configurations {\ell} \rightarrow \bigvee_{m\leq k} \exists \vect x[m]: \Configurations {m}
 $$
-
-For the purpose of the universality checking,
-the fixpoint condition can be weakened in the spirit of antichains:
+%if and only if the distance $k+1$ does not hold any any new configurations, i.e 
+if and only if there are no new or smaller configurations at the distance $k+1$, i.e 
 $$
-\mathit{fixpoint}'_k = \forall \viz_0'\cdots\viz_{k+1}'\exists \viz_0\cdots\viz_{k} :
-M_{k+1}[\viz_0,\ldots,\viz_{k+1}/\viz'_0,\ldots,\viz'_{k+1}]
-\Leftarrow
-\bigvee_{i=0}^k M_k 
+\forall \vect x[k+1]:\Configurations {k+1} \rightarrow \bigvee_{\ell\leq k} \exists \vect x[\ell]: \Configurations {\ell}\ \ .
 $$
-That is, for every newly generated macroste, there already has been a smaller one.
 

### 2017-07-31 symbolic_nfa_universality/symbolic_subset_construction.tex
little fixes in emp univ trans sys

@@
 \subsection{Universality}
-We will snow ketch a basic universality check. 
+We will now sketch a basic universality check. 
 As in the case of emptiness, it is incomplete due to the infinite data domain, but additionally also due to the need of working with unbounded histories. 
@@
 be a formula which associates a visible trace $\vect x[k]$ of the length $k$ with the resulting values of $x$ and $y$.
-Fixing a particular trace $\vect \dval [k]$ results in a so called \emph{configuration}, 
+Fixing a particular trace $\vect \dval [k]\in\domain^k$ results in a so called \emph{configuration}, 
 a constraint over $x$ and $y$ with $x$ fixed at the last value $\dval_k$: 
@@
 $$
-That is, every configuration must have $y$ to satisfy $\final$.
-Notice that the universality test does not require all but only the smallest (strongest) configurations.
+That is, every configuration must posses $y$ to satisfy $\final$.
+Notice that the universality test does not require to test all but only the smallest (strongest) configurations.
 
-The task is therefore to compute and test all configurations $\subst{\Configurations {k}}{\vect x[k]}{\vect \dval[k]}$. 
-To compute a representation of all configurations, we use the fact that configurations at the distance $k+1$ can be obtained as successors of those at the distance $k$ as follows. 
+The task is therefore to compute and test all smallest configurations $\subst{\Configurations {k}}{\vect x[k]}{\vect \dval[k]}$. 
+To compute a representation of all smallest configurations, we use the fact that configurations at the distance $k+1$ can be obtained as successors of those at the distance $k$ as follows. 
 For every visible trace $\vect \dval [k+1]$,
@@
 %
-Notice that the lowest configurations at the distance $k+1$ are successors of the lowest configurations at the distance $k+1$.
+Notice that the smallest configurations at the distance $k+1$ are successors of the smallest configurations at the distance $k+1$.
 %
+
 That is, all smallest configurations are found within the distance $k$, i.e.

### 2017-07-31 symbolic_nfa_universality/symbolic_subset_construction.tex
some typos in trans sys uni emp

@@
 
-Vectors $x_1,\ldots,x_k,k\in\nat$ of indexed variables are denoted by $\vect{x}$ and we use $\vect{x}[k]$ to denote the particular vector $x_1,\ldots,x_k$.
+Vectors $x_1,x_2,\ldots,x_k,k\geq 0$ of indexed variables are denoted by $\vect{x}$ and we use $\vect{x}[k]$ to denote the particular vector $x_1,\ldots,x_k$.
 %
@@
 %
-Let $\nat$ denote the set of non-negative integers including zero. 
+%Let $\nat$ denote the set of non-negative integers including zero. 
 %
-We write $\false$ and $\true$ to denote the two boolean constants. 
+%We write $\false$ and $\true$ to denote the two boolean constants. 
 %
@@
 %
-
 %Let $\vect{x} = \vect{x}[k]$ and
@@
 We say that a valuation $\nu'$ is 
-a~\emph{successor} of $\nu$ if and only if and $(\nu\cup\nu') \modelsthd
-\transition$, denoted by $\nu
-\arrow{\transition}{} \nu'$. 
+a~\emph{successor} of $\nu$ if and only if $(\nu\cup\nu') \modelsthd
+\transition$. 
+%denoted by $\nu
+%\arrow{\transition}{} \nu'$. 
 %, from where we omit the $\transition$ when no confusion may arise. 
-
+%
+%
 A \emph{run} of $\transition$ is a sequence
-of valuations $\pi : \nu_0 \arrow{\transition}{}  \cdots \arrow{\transition}{} \nu_n$ with $\nu_0\models\initial$. 
-It is accepting if also $\nu_n\models\accepting$.
+of valuations $\pi : \nu_0, \ldots, \nu_n$ with $\nu_0\models\initial$ and $\nu_i$ a successor of $\nu_{i-1}$ for all $1\leq i \leq n$. 
+It is \emph{accepting} if also $\nu_n\models\accepting$.
 Each run corresponds to a \emph{trace}, which is a finite sequence
 $w(\pi)=\nu_0\proj_{\Xobs}, \ldots,
-\nu_n\proj_{\Xobs}$, where $\nu_i\proj_{\Xobs}{}$ denotes the restriction of $\nu_i$ to $\Xobs$.
+\nu_n\proj_{\Xobs}$, where $\nu_i\proj_{\Xobs}$ denotes the restriction of $\nu_i$ to $\Xobs$.
 %
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-For simplicity, let us assume wlog. a transition system $T$ with only one visible variable $x$ and only one invisible variable $y$.
+For simplicity, let us assume that the transition system $T$ has only one visible variable $x$ and only one invisible variable $y$ (this is without loss of generality, since a variable may hold an $n$-tuple).
 A trace is then just a sequence $d_0,\ldots, d_k$ of elements of the domain.
@@
 
+\newpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Emptiness}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 As an exercise, before we elaborate on universality checking, we discuss language emptiness.  
@@
 where the inner formula represents the set of states reached after $k$ steps.
-Let us denote it $\States k(x,y)$. $\Nonempty$ is therefore equivalent to
-$\exists xy:\bigvee_{k\geq 0} \States k(x,y) \land \final(x,y)$, that is, it is enough to compute a formula representing states reachable at all distances and test existence of an accepting one.
+Let us denote it 
+$$\States k(x,y)\fdef \subst{(\exists \vect x[k-1]\vect y[k-1]:\Run {k}(\vect x[k],\vect y[k]))}{x_k,y_k}{x,y}\ \ .$$ $\Nonempty$ is therefore equivalent to
+$\exists xy:\bigvee_{k\geq 0} \States k(x,y) \land \final(x,y)$.
+To test emptiness, it is enough to compute a formula representing states reachable at all distances and test existence of an accepting state.
 To compute all reachable states, we use that 
@@
 $$
-That is, reachable states at the distance $k$ can be computed from those reached at the distance $k-1$ by applying the transition relation.  
+That is, reachable states at the distance $k$ can be computed as the image wrt. the transition relation of those reached at the distance $k-1$.  
 %
-So it is enough to compute a formula that describes reached states that is stable under the transition relation:
-$$\States {\leq n} \fdef \bigvee_{k=0}^n \States k$$ 
+So a formula that describes states including the initial ones and stable under the transition relation describes all reachable states. E.g., the formula
+$$
+\States {\leq n} \fdef \bigvee_{k=0}^n \States k
+$$ 
 such that 
-$$\States {\leq n}  \equiv \States {\leq n} \lor \subst{(\exists x,y: \States {n}\land \transition)}{x',y'}{x,y}\ \ .$$
+$$
+%\States {\leq n}  \equiv \States {\leq n} \lor \subst{(\exists x,y: \States {n}\land \transition)}{x',y'}{x,y}\ \ .
+\States {\leq n}  \equiv \States {\leq n} \lor \States {n+1}\ \ .
+$$
 
 \newpage
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Universality}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 We will now sketch a basic universality check. 
@@
 %
-Universality of $T$ corresponds to the validity the formula
+Universality of $T$ corresponds to the validity of the formula
 $$
@@
 $$
-(the difference from the formula $\Nonempty$ is the leading $\bigwedge$ and the universal quantification over the visible variables). 
+(the difference from the formula $\Nonempty$ is the leading $\bigwedge$ and the universal quantification over the visible trace). 
 %
@@
 Fixing a particular trace $\vect \dval [k]\in\domain^k$ results in a so called \emph{configuration}, 
-a constraint over $x$ and $y$ with $x$ fixed at the last value $\dval_k$: 
+a constraint over $x$ and $y$ (with $x$ fixed to the last value $\dval_k$ of the trace): 
 $$
@@
 %
-We may rewrite conjunction $\Universal$ as 
+We may rewrite the conjunction $\Universal$ as 
 $$
@@
 $$
-That is, every configuration must posses $y$ to satisfy $\final$.
-Notice that the universality test does not require to test all but only the smallest (strongest) configurations.
-
+That is, every configuration must admit $y$ to satisfy $\final$.
+Notice that this does not require to test all, but only the smallest (strongest) configurations.
+%
 The task is therefore to compute and test all smallest configurations $\subst{\Configurations {k}}{\vect x[k]}{\vect \dval[k]}$. 
@@
 %
-
 That is, all smallest configurations are found within the distance $k$, i.e.
 $$
-\bigwedge_{\ell\geq 0} \forall \vect x[\ell]:\Configurations {\ell} \rightarrow \bigvee_{m\leq k} \exists \vect x[m]: \Configurations {m}
+\bigwedge_{m\geq 0} \forall \vect x[m]:\Configurations {m} \rightarrow \bigvee_{\ell\leq k} \exists \vect x[\ell]: \Configurations {\ell}
 $$
 %if and only if the distance $k+1$ does not hold any any new configurations, i.e 
-if and only if there are no new or smaller configurations at the distance $k+1$, i.e 
+if and only if the configurations found at the distance $k+1$ are only the same or larger, i.e.
 $$

### 2017-07-31 symbolic_nfa_universality/symbolic_subset_construction.tex
little thing

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-For simplicity, let us assume that the transition system $T$ has only one visible variable $x$ and only one invisible variable $y$ (this is without loss of generality, since a variable may hold an $n$-tuple).
+For simplicity, let us assume that the transition system $T$ has only one visible variable $x$ and only one invisible variable $y$ (this is without loss of generality since $\domain$ may contain tuples).
 A trace is then just a sequence $d_0,\ldots, d_k$ of elements of the domain.

### 2017-08-16 arxiv/draft.tex
added arxiv

--- 2017-05-15 draft.tex
@@
 \vspace*{-\baselineskip}
-\centerline{\includegraphics[scale=0.8]{figures/Example.pdf}}
+\centerline{\includegraphics[scale=0.8]{Example.pdf}}
 \vspace*{-\baselineskip}
@@
 \begin{tabular}{cc}
-\includegraphics[scale=0.5]{figures/PA1.pdf} & \includegraphics[scale=0.5]{figures/PA2.pdf} \\[-2mm]
+\includegraphics[scale=0.5]{PA1.pdf} & \includegraphics[scale=0.5]{PA2.pdf} \\[-2mm]
 \tiny{(a)} & \tiny{(b)} \\
 \\
-\includegraphics[scale=0.5]{figures/PA3.pdf} & \includegraphics[scale=0.5]{figures/PA4.pdf} \\[-2mm]
+\includegraphics[scale=0.5]{PA3.pdf} & \includegraphics[scale=0.5]{PA4.pdf} \\[-2mm]
 \tiny{(c)} & \tiny{(d)}
@@
 \begin{tabular}{c}
-\includegraphics[scale=0.65]{figures/IM1.pdf} \\
+\includegraphics[scale=0.65]{IM1.pdf} \\
 \tiny{(a)} \\
-\includegraphics[scale=0.65]{figures/IM2.pdf} \\[-2mm]
+\includegraphics[scale=0.65]{IM2.pdf} \\[-2mm]
 \tiny{(b)} \\
-\includegraphics[scale=0.65]{figures/IM3.pdf} \\[-2mm]
+\includegraphics[scale=0.65]{IM3.pdf} \\[-2mm]
 \tiny{(c)}
@@
 \begin{figure}[htb]
-\centerline{\includegraphics[scale=0.65]{figures/IM4.pdf}}
+\centerline{\includegraphics[scale=0.65]{IM4.pdf}}
 \vspace*{-\baselineskip}

### 2017-08-16 arxiv/draft.tex
added arxiv

@@
 \A_\cap & = & \tuple{\vec{x},Q_1\cup Q_2,\iota_1 \wedge \iota_2, F_1\cup F_2, \Delta_1\cup\Delta_2} \\ 
-\overline{\A_1} & = & \langle \vec{x},Q_1,\iota_1,Q_1\setminus F_1,\overline{\Delta_1} \rangle
+\overline{\A_1} & = & \langle \vec{x},Q_1,\overline{\iota_1},Q_1\setminus F_1,\overline{\Delta_1} \rangle
 \end{array}\] 
@@
   let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
-  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
-  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
-  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
-  \subseteq Q$ and thus $\overline{\iota_1} \equiv \iota_1$. For the case
-  $n>0$, we compute:
+  $\overline{\Delta}(\overline{\iota_1},a_1 \ldots a_n) =
+  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on $n\geq0$.
+
+  In the case $n=0$, we have $\overline{\Delta}(\overline{\iota_1},a_1
+  \ldots a_n) \equiv \overline{\iota_1}$. Then $\varepsilon$ is
+  accepted by $\A_1$ iff $\nu_0 \models \iota_1$, where $\nu_0(q) =
+  \top$ if $q \in F_1$ and $\nu_0(q) = \bot$, otherwise. But $\nu_0
+  \models \iota_1$ iff $\overline{\nu_0} \models \overline{\iota_1}$,
+  where $\overline{\nu_0}(q) = \top$ if $q \not\in F_1$ and
+  $\overline{\nu_0}(q) = \bot$, otherwise. Thus $\varepsilon$ is
+  accepted by $\A_1$ iff it is not accepted by $\overline{\A_1}$.
+
+  For the case $n>0$, we compute:
   \[\begin{array}{lcl}

### 2017-08-16 quantifiers/draft.tex
ok

--- 2017-05-22 quantifiers/draft.tex
@@
 
+For two integers $0 \leq i \leq j$, we denote by $[i,j]$ the set
+$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[1,i]$, where $[0]$ is
+the empty set.
+
 A \emph{signature} $\sig = (\ssorts{\sig},\sfuns{\sig})$ consists of a
@@
 function symbol $f \in \sfuns{\sig}$ has $\#(f) \geq 0$ arguments of
-sort $\Data$ and return value $\sigma(f) \in \ssorts{\sig}$. If
-$\#(f)=0$ then $f$ is a \emph{constant}. Let
-$\top$ and $\bot$ be constants of sort $\Bool$.
+sort $\Data$ and return value $f_\sigma \in \ssorts{\sig}$. If
+$\#(f)=0$ then $f$ is a \emph{constant}. Let $\top$ and $\bot$ be
+constants of sort $\Bool$ and $\mathbf{0}$ be a constant of sort
+$\Data$.
 
 We consider an infinite countable set of \emph{variables} $\vars$,
-where each $x \in \vars$ has an associated sort $\sigma(x)$. A
-\emph{term} $t$ of sort $\sigma(t)=S$ is a variable $x \in \vars$
-where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
-$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
-\emph{atom} is a term of sort $\Bool$ or an equality $t \teq s$
-between two terms of sort $\Data$. A \emph{formula} is a quantified
-boolean combination of atoms, where the quantifiers are as $\exists x
-~.~ \phi$ and $\forall x ~.~ \phi$.
+where each $x \in \vars$ has an associated sort $x_\sigma$. A
+\emph{term} $t$ of sort $t_\sigma = S$ is a variable $x \in \vars$
+where $x_\sigma=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
+$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $f_\sigma=S$. An
+\emph{atom} is a term $p(t_1,\ldots,t_{\#(p)})$ of sort $\Bool$, or an
+equality $t \teq s$ between two terms of sort $\Data$. A
+\emph{formula} is a possibly quantified boolean combination of atoms.
 
@@
   \ssorts{\sig}} \fv{\sigma}{\phi}$. For a variable $x \in
-\fv{}{\phi}$ and a term $t$ such that $\sigma(t) = \sigma(x)$, let
+\fv{}{\phi}$ and a term $t$ such that $t_\sigma = x_\sigma$, let
 $\phi[t/x]$ be the result of replacing each occurrence of $x$ by
 $t$. For indexed sets $T=\set{t_1,\ldots,t_n}$ and
-$X=\set{x_1,\ldots,x_n}$, we write $\phi[T/X]$ for
-the formula obtained by simultaneously replacing $x_i$ with $t_i$ in
-$\phi$, for all $i\in[1,n]$. The size $\len{\phi}$ is the number of
-symbols occuring in $\phi$.
+$X=\set{x_1,\ldots,x_n}$, we write $\phi[T/X]$ for the formula
+obtained by simultaneously replacing $x_i$ with $t_i$ in $\phi$, for
+all $i\in[n]$. The size $\len{\phi}$ is the number of symbols
+occuring in $\phi$.
 
@@
 \item each function symbol $f$ into a total function $f^\I :
-  (\Data^\I)^{\#(f)} \rightarrow \sigma(f)^I$, or an element of
-  $\sigma(f)^I$ when $\#(f)=0$.
+  (\Data^\I)^{\#(f)} \rightarrow f_\sigma^I$, or an element of
+  $f_\sigma^I$ when $\#(f)=0$.
 \end{inparaenum}
 Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
-variable $x \in \vars$ into an element $\nu(x) \in \sigma(x)^\I$. For
-a term $t$, we denote by $t^\I_\nu$ the value obtained by replacing
-each function symbol $f$ by its interpretation $f^\I$ and each
-variable $x$ by its valuation $\nu(x)$. For a formula $\phi$, we write
-$\I,\nu \models \phi$ if the formula obtained by replacing each term
-$t$ in $\phi$ by the value $t^\I_\nu$ is logically equivalent to true.
-The semantics of quantifiers is given as usual: $\I,\nu \models
-\exists x ~.~ \phi$ ($\forall x ~.~ \phi$) if and only if $\I,\nu[x
-  \leftarrow u] \models \phi$, for some (for all) value(s) $u \in
-\sigma(x)^\I$, where $\nu[x \leftarrow u]$ is the valuation that
-assigns $u$ to $x$ and $\nu(y)$ to all $y \neq x$.
+variable $x \in \vars$ into an element of $x_\sigma^\I$. For a term
+$t$, we denote by $t^\I_\nu$ the value obtained by replacing each
+function symbol $f$ by its interpretation $f^\I$ and each variable $x$
+by its valuation $\nu(x)$. For a formula $\phi$, we write $\I,\nu
+\models \phi$ if the formula obtained by replacing each term $t$ in
+$\phi$ by the value $t^\I_\nu$ is logically equivalent to true.  The
+semantics of quantifiers is given as usual: $\I,\nu \models \exists x
+~.~ \phi$ if and only if $\I,\nu[x \leftarrow u] \models \phi$, for
+some value $u \in x_\sigma^\I$, where $\nu[x \leftarrow u]$ is the
+valuation that assigns $u$ to $x$ and $\nu(y)$ to all $y \neq x$.
 
@@
 
-Given sets of variables $B,X \subset \vars$ of sort $\Bool$ and
-$\Data$, respectively, we denote by $\Form(B,X,Y)$ the set of formulae
-$\phi$ such that $\fv{\Bool}{\phi} = B$, $\fv{\Data}{\phi} = X$ and
-each variable from $Y$ is of sort $\Data$ and occurs under the scope
-of a quantifier. By $\Form^+(B,X,Y)$ we denote the subset of
-$\Form(B,X,Y)$ in which each boolean variable occurs under an even
-number of negations.
+Given sets of variables $B,X,Y \subset \vars$ of sorts $\Bool$,
+$\Data$ and $\Data$, respectively, we denote by $\Form\tuple{B;X;Y}$
+the set of formulae $\phi$ such that $\fv{\Bool}{\phi} = B$,
+$\fv{\Data}{\phi} = X$ and each variable from $Y$ occurs under the
+scope of a quantifier. By $\Form^+\tuple{B;X;Y}$ we denote the subset
+of $\Form\tuple{B;X;Y}$ in which each boolean variable from $B$ occurs
+under an even number of negations. We denote by $\phi\tuple{B;X;Y}$
+any formula in $\Form\tuple{B;X;Y}$.
 
@@
 %
-\item $\iota \in \Form^+(Q,\emptyset,\emptyset)$ is the \emph{initial
+\item $\iota \in \Form^+\tuple{Q;\emptyset;\emptyset}$ is the \emph{initial
   configuration},
@@
 \item $\Delta : Q \times \Sigma \rightarrow
-  \Form^+(Q,\overline{G}\cup G\cup\overline{L},L)$
-  is a \emph{transition function},
+  \Form^+\tuple{Q;\overline{G}\cup G\cup\overline{L};L}$ is a
+  \emph{transition function},
 \end{compactitem}
 where, for a set $X \subseteq \vars$, we denote $\overline{X} =
-\set{\overline{x} \mid x \in X}$. Intuitively, the variables
-$\overline{x}$ from a transition formula $\Delta(q,a)$ refer to past
-values, whereas the other variables refer to the current
-values. Observe that the initial values are unconstrained (since the
-initial configuration is a boolean combination of states without
-variables) and quantification is allowed only over the current local
-variables. In other words, we do not allow quantification over past
-values, which are beyond our control. The size of $\A$ is defined as
-$\len{\A} = \len{\iota} + \sum_{(q,a) \in
+\set{\overline{x} \mid x \in X}$. The \emph{size} of $\A$ is defined
+as $\len{\A} = \len{\iota} + \sum_{(q,a) \in
   Q\times\Sigma}\len{\Delta(q,a)}$.
+
+Intuitively, the variables $\overline{x}$ from a transition formula
+$\Delta(q,a)$ refer to past values, whereas the other variables refer
+to the current values. Observe that the initial values of the global
+variables are unconstrained (since the initial configuration is a
+boolean combination of states without variables) and quantification
+occurs only over the current copy of the local variables $L$. As a
+remark, the transition relation does not quantify over global
+variables $G \cup \overline{G}$, which must be visible throughout the
+execution. Furthermore, we assume that the past values of the local
+variables are coined and beyond our control, thus quantification over
+$\overline{L}$ is not allowed either.
+
+In order to define the semantics of automata, we consider time-stamped
+sets of variables $\stamp{X}{n} = \{\stamp{x}{n} \mid x \in X\}$ for
+any $n\geq0$. For an input event $a \in \Sigma$, some constant
+$n\geq0$ and a formula $\phi$, we write $\stamp{\Delta}{n}(\phi,a)$
+for the formula obtained from $\phi$ by simultaneously replacing each
+state $q \in \fv{\Bool}{\phi}$ with
+$\Delta(\phi,a)[\stamp{G}{n}/\overline{G}, \stamp{G}{n+1}/G,
+  \stamp{L}{n}/\overline{L}, \stamp{L}{n+1}/L]$.
+
+Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in
+\Sigma[G]^*$ the \emph{run} of $\A$ over $w$ is the sequence of
+formulae: \[\iota \stackrel{(a_1,\nu_1)}{\Longrightarrow} \phi_1
+\tuple{Q,\stamp{G}{1}, \stamp{L}{1}}
+\stackrel{(a_2,\nu_2)}{\Longrightarrow} \ldots
+\stackrel{(a_n,\nu_n)}{\Longrightarrow} \phi_n\tuple{Q,\stamp{G}{1}
+  \cup \ldots \cup \stamp{G}{n}, \stamp{L}{1} \cup \ldots \cup
+  \stamp{L}{n}}\] where $\phi_1 \equiv
+\stamp{\Delta}{0}(\iota,a_1)[\mathbf{0}/\stamp{G}{0},\mathbf{0}/\stamp{L}{0}]$
+and $\phi_{i+1} \equiv \stamp{\Delta}{i}(\phi_i,a_{i+1})$, for all $i
+\in [1,n-1]$. In other words, we assume that all data variables are
+initially $\mathbf{0}$.
+
+We say that $\A$ \emph{accepts} $w$ if there exists a valuation $\nu$
+such that $\nu(\stamp{x}{i}) = \nu_i(x)$ for each global variable $x
+\in G$ and each timestamp $i \in [n]$, $\nu(q) = \true$ for each $q
+\in F$, $\nu(q) = \false$ for each $q \in Q \setminus F$ and $\I,\nu
+\models \phi_n$. The \emph{language} of $\A$ is the set $L(\A)$ of
+words from $\Sigma[G]^*$ accepted by $\A$. In this paper, we ask the
+following two questions: \begin{compactenum}
+\item \emph{boolean closure}: given automata $\A_i =
+  \tuple{G,L_i,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, with $L_1 \cap
+  L_2 = \emptyset$ and $Q_1 \cap Q_2 = \emptyset$, do there exist
+  automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
+  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
+  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[G]^* \setminus L(\A_1)$ ?
+%
+\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
+  \emptyset$ ?
+\end{compactenum}
+
+\subsection{Boolean Closure}
+
+Given a set $B$ of boolean variables, two disjoint sets $X$ and $Y$ of
+data variables, for any formula $\phi \in \Form^+\tuple{B;X;Y}$,
+define $\overline{\phi} \in \Form^+\tuple{B;X;Y}$ recursively:
+\[\begin{array}{rclcrclcrcl}
+\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} & \hspace*{5mm} & 
+\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} & \hspace*{5mm} & 
+\overline{\phi} & \equiv & \neg\phi \text{ if $\phi \not\in B$ is an atom} \\ 
+\overline{\exists x ~.~ \phi_1} & \equiv & \forall x ~.~ \overline{\phi_1} && 
+\overline{\forall x ~.~ \phi_1} & \equiv & \exists x ~.~ \overline{\phi_1} && 
+\overline{\neg\phi} & \equiv & \neg \overline{\phi} \text{ if $\phi$ is not an atom} \\
+&& && && && \overline{\phi} & \equiv & \phi \text{ if $\phi \in B$}
+\end{array}\]
+Clearly $\phi$ and $\overline{\phi}$ have the same size. The relation
+between $\phi$ and $\overline{\phi}$ is formalized below:
+
+\begin{proposition}\label{prop:complement}
+  Given sets of variables $B,X,Y \subset \vars$ of sorts boolean, data
+  and data respectively, and a formula $\phi \in
+  \Form^+\tuple{B;X;Y}$, for any valuation $\nu$ we have $\nu \models
+  \phi$ if and only if $\overline{\nu} \models \neg\overline{\phi}$,
+  where $\overline{\nu}(x) = \nu(x)$ for all $x \in X$ and
+  $\overline{\nu}(b) = \neg\nu(b)$ for all $b \in B$.
+\end{proposition}
+\proof{By induction on the structure of $\phi$. \qed}
+
+Consider the automata $\A_i = \tuple{G,L_i,Q_i,\iota_i,F_i,\Delta_i}$,
+for $i=1,2$, where $L_1 \cap L_2 = \emptyset$ and $Q_1 \cap Q_2 =
+\emptyset$. We define the following automata: 
+\[\begin{array}{rcl}
+\A_\cup & = & \tuple{G,~ L_1 \cup L_2,~ Q_1 \cup Q_2,~ \iota_1 \vee \iota_2,~ F_1 \cup F_2,~ \Delta_1 \cup \Delta_2} \\
+\A_\cap & = & \tuple{G,~ L_1 \cup L_2,~ Q_1 \cup Q_2,~ \iota_1 \wedge \iota_2,~ F_1 \cup F_2,~ \Delta_1 \cup \Delta_2} \\
+\overline{A_1} & = & \tuple{G,~ L_1,~ Q_1,~ \overline{\iota_1},~ Q_1 \setminus F_1,~ \lambda q \in Q_1~ \lambda a \in \Sigma ~.~ \overline{\Delta_1(q,a)}}
+\end{array}\]
+
+\begin{lemma}\label{lemma:boolean-closure}
+  Given automata $\A_i = \tuple{G,L_i,Q_i,\iota_i,F_i,\Delta_i}$, for
+  $i=1,2$, where $L_1 \cap L_2 = \emptyset$ and $Q_1 \cap Q_2 =
+  \emptyset$, we have $L(\A_\cup) = L(\A_1) \cup L(\A_2)$, $L(A_\cap)
+  = L(\A_1) \cap L(\A_2)$ and $L(\overline{\A_1}) = \Sigma[G]*
+  \setminus L(\A_1)$. 
+\end{lemma}
+\proof{
+  Same as \cite[Lemma 1]{IosifX17}. 
+\qed}
+
+\section{Checking Emptiness with Interpolants and Skolem Witnesses}
+
+Since the emptiness problem is undecidable even for automata with only
+two global variables and no local variables, by reduction from the
+emptiness problem for 2-counter machines \cite{Minsky67}, we give a
+semi-algorithm with the following properties. Given an automaton $\A$
+as input, the procedure will\begin{inparaenum}[(1)] 
+\item return a word $w \in L(\A)$ if $L(\A)
+\neq \emptyset$, and 
+%
+\item if the procedure terminates without returning such a word, then
+  we can safely conclude that $L(\A) = \emptyset$. \end{inparaenum}
+Due to the undecidability of the emptiness problem, it is possible for
+the semi-algorithm to run forever in the case when the language of the
+automaton is empty.
+
+\begin{proposition}\label{prop:substitution}
+  Given a formula $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$, where $Q_1 \ldots
+  Q_n \in \set{\exists,\forall}$ and $\phi$ is a quantifier-free
+  formula with free variables $\fv{}{\phi} = \set{x_1,\ldots,x_n,q}$,
+  for any formula $\psi$, the following equivalence holds:
+  \[Q_1 x_1 \ldots Q_n x_n ~.~ \phi[\psi/q] \iff 
+  Q_1 x_1 \ldots Q_n x_n \exists q ~.~ \phi \wedge (q \rightarrow \psi)\]
+\end{proposition}
+\proof{
+
+\qed}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\bibliographystyle{abbrv} \bibliography{refs}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2017-08-16 quantifiers/draft.tex
ok

@@
   Q_n \in \set{\exists,\forall}$ and $\phi$ is a quantifier-free
-  formula with free variables $\fv{}{\phi} = \set{x_1,\ldots,x_n,q}$,
-  for any formula $\psi$, the following equivalence holds:
+  formula with free variables $\fv{}{\phi} = \set{x_1,\ldots,x_n,q}$
+  in which the boolean variable $q$ occurs positively, for any formula
+  $\psi$, the following equivalence holds:
   \[Q_1 x_1 \ldots Q_n x_n ~.~ \phi[\psi/q] \iff 
@@
 \end{proposition}
-\proof{
+\proof{ Let $\nu$ be an arbitrary valuation of the variables in
+  $\fv{}{\phi} \cup \fv{}{\psi}$. It is sufficient to prove that $\nu
+  \models \phi[\psi/q]$ if and only if $\nu \models \exists q ~.~ \phi
+  \wedge (q \rightarrow \psi)$.
 
+  ``$\Leftarrow$'' Suppose that $\nu \models \phi[\psi/q]$ and
+  distinguish the following cases: \begin{compactenum}
+  \item if $\nu \models \psi$, since $q$ occurs positively in $\phi$, we
+    obtain that $\nu \models \phi[\top/q]$, thus $\nu \models
+    \phi[\top/q] \wedge (\top \rightarrow \psi)$ and $\nu \models
+    \exists q ~.~ \phi \wedge (q \rightarrow \psi)$ follows.
+    %
+  \item if $\nu \not\models \psi$, since $q$ occurs positively in $\phi$, we
+    obtain that $\nu \models \phi[\bot/q]$, thus $\nu \models
+    \phi[\bot/q] \wedge (\bot \rightarrow \psi)$ and $\nu \models
+    \exists q ~.~ \phi \wedge (q \rightarrow \psi)$ follows.
+  \end{compactenum}
+
+  ``$\Rightarrow$'' Suppose that $\nu \models \exists q ~.~ \phi
+  \wedge (q \rightarrow \psi)$, or equivalently, $\nu \models
+  (\phi[\top/q] \wedge \psi) \vee \phi[\bot/q]$. If $\nu \models
+  \phi[\top/q] \wedge \psi$ then $\nu \models
+  \phi[\psi/q]$. Otherwise, if $\nu \models \phi[\bot/q]$, we
+  distinguish two cases: \begin{compactitem} 
+  \item if $\nu \not\models \psi$ then clearly $\nu \models
+    \phi[\psi/q]$.
+    %
+  \item else, if $\nu \models \psi$, since $q$ occurs positively in
+    $\phi$ and $\nu \models \phi[\bot/q]$ then also $\nu \models
+    \phi[\top/q]$, because $\phi$ is monotone in $q$ in the entailment
+    partial order. But then $\nu \models \phi[\psi/q]$.
+  \end{compactitem}
   \qed}

### 2017-09-12 rebuttal.txt
ok

We are grateful to all reviewers for their great work. The main points
of criticism are addressed as follows:

1. Why are our semi-algorithms performing slower than Includer [11] ?

First of all, Includer is specifically designed to solve language
inclusion problems between nondeterministic data automata, which are a
(significant) special case of alternation, whereas the semi-algorithms
presented in this paper can deal with the entire class of alternating
data automata, not just the ones resulting from inclusion problems. It
is common that procedures for more general problems are slower than
the ones for specific problems, although ideally, it is desirable to
have matching speed on the same class of problems.

As reviewer 2 points out, the generalized subset construction used by
Includer does not cause a huge blowup for the examples considered in
this paper and in [11]. Moreover, Includer uses the LIA interpolation
procedure of MathSAT5, which seems to be faster than the more generic
UFLIA interpolation of Z3 used here (UFLIA is needed in order to deal
with the boolean variables representing the states). In some cases, Z3
answers unsat almost immediately and then takes up to 5 seconds to
extract the interpolant. We believe that further experimenting with
different Z3 tactics and tuning the encoding of the interpolation
procedure, as well as considering a larger set of benchmarks than the
preliminary benchmarks used to test our first prototype will greatly
improve the performance of our tool.

2. What is the difference between the semi-algorithms in this paper
and the original ones of Blast [8] and Impact [15] ?

Both Blast and Impact are designed for checking emptiness (safety) of
nondeterministic data automata (programs), whereas our methods address
the emptiness problem for alternating data automata. A first
difference between our methods and [8,15] is a more sophisticated
Abstract Reachability Tree (ART) data structure (page 5, 2nd paragraph
of section 5), which requires labeling both nodes and edges with
formulae. However, on the positive side, our encoding of path formulae
(interpolation problems) allows the basic principles of [8,15] to be
carried over to the alternating setting (we represent path formulae as
conjunctions of implications and abstract sets of configurations as
combinations of boolean and data theory atoms), with some specific
differences listed below:

2.a) Difference with Blast [8]. Algorithm 1 (page 6) uses antichains
to prune the search space (lines 29-35). Using antichains is a
specific technique for classical finite-alphabet alternating
automata. We adapt this idea from [5] to optimize the predicate
abstraction emptiness checking of ADA.

2.b) Difference with Impact [15]. The original description of Impact
[15] is based on a nondeterministic application of the Cover, Refine
and Expand procedures, whereas Algorithm 2 (page 8) is a detailed
deterministic procedure that closely mirrors our implementation. In
particular, we chose to cover nodes lazily, only if their labels are
strenghtened by refinement (line 18).

3. Using the Lyndon interpolation theorem instead of post-images to
prove Proposition 1. This is a very interesting remark that we did not
consider initially and which allows to use positive interpolants
directly instead of post-images (quantifiers can be eliminated if the
data theory is LIA, but quantifier elimination has a non-negligeable
cost). If accepted, we are willing to change this point in the final
version, provided that we can find a published reference for the fact
that McMillan's Z3 interpolation procedure fits with Lyndon's theorem
(their CAV13, FMCAD11 or TCS05 papers on interpolation do not seem to
refer to it).

4. Several reviewers expressed concerns relative to the paper being
too theoretical and out of the scope of FMCAD. As authors of a paper
on formal verification techniques, we felt the need to define and
prove each point formally. However, the paper does not address issues
such as decidability or computational complexity of subclasses of ADA,
which could fit a more theoretical conference. Although interpolation
can be used as an argument for decidability (when suitable
restrictions are imposed, such as for symbolic alternating automata)
it is perhaps not a very good framework to derive complexity
bounds. In general, theoretical complexity proofs require different
techniques (e.g. reductions to other decidable problems whose
complexity is known) which are clearly out of the scope of the present
paper.

### 2017-10-19 tacas/draft.tex
ok

--- 2017-08-16 arxiv/draft.tex
@@
-% \documentclass{llncs}
-\documentclass[10pt,conference,letterpaper,twocolumn]{IEEEtran}
+\documentclass[10pt]{llncs}
+
+% \documentclass[10pt,conference,letterpaper,twocolumn]{IEEEtran}
 
@@
 
-\author{Radu Iosif and Xiao Xu \\
-CNRS, Verimag, Universit\'e de Grenoble Alpes \\
-Email: \{Radu.Iosif,Xiao.Xu\}@univ-grenoble-alpes.fr}
+\author{Radu Iosif \and Xiao Xu}
+\institute{CNRS, Verimag, Universit\'e de Grenoble Alpes \\
+\email{\{Radu.Iosif,Xiao.Xu\}@univ-grenoble-alpes.fr}
+}
 
@@
 \begin{figure}[htb]
-\vspace*{-\baselineskip}
+\begin{center}
+\begin{minipage}{6cm}
 \centerline{\includegraphics[scale=0.8]{Example.pdf}}
-\vspace*{-\baselineskip}
-\caption{Alternating Data Automaton Example}
-\label{fig:ex}
-\end{figure}
-
-\paragraph{\em Example}
-Figure \ref{fig:ex} depicts an ADA with input alphabet $\Sigma =
-\set{a,b}$, variables $\vec{x} = \set{x,y}$, states $Q =
-\set{q_0,q_1,q_2,q_3,q_4}$, initial configuration $q_0$, final states
-$F=\set{q_3,q_4}$ and transitions:
+\centerline{(a)}
+\end{minipage}
+\begin{minipage}{6cm}
 \[\begin{array}{rcl}
@@
 \Delta(q_2,a) & \equiv & q_2 \wedge x > \overline{x} \wedge y > \overline{y} \\
-\Delta(q_2,b) & \equiv & q_4 \wedge \overline{x}>\overline{y}
+\Delta(q_2,b) & \equiv & q_4 \wedge \overline{x}>\overline{y} \\\\
 \end{array}\]
-The missing rules, such as $\Delta(q_0,b)$, are assumed to be
+\centerline{(b)}
+\end{minipage}
+\end{center}
+\caption{Alternating Data Automaton Example}
+\label{fig:ex}
+\end{figure}
+
+\paragraph{\em Example}
+Figure \ref{fig:ex} (a) depicts an ADA with input alphabet $\Sigma =
+\set{a,b}$, variables $\vec{x} = \set{x,y}$, states $Q =
+\set{q_0,q_1,q_2,q_3,q_4}$, initial configuration $q_0$, final states
+$F=\set{q_3,q_4}$ and transitions given in Figure \ref{fig:ex} (b),
+where missing rules, such as $\Delta(q_0,b)$, are assumed to be
 $\bot$. Rules $\Delta(q_0,a)$ and $\Delta(q_1,a)$ are universal and
@@
 of $\phi$:
-\[\begin{array}{lclclcl}
+\[\begin{array}{rclcrcl}
 \overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} && 
@@
 interpretation of the function symbols on integers. Since reachability
-of a control state is undecidable for AVASS \cite{LINCOLN92}, the
-emptiness problem is undecidable for ADA.
+of a control state is undecidable for AVASS \cite{LINCOLN92}, 
+emptiness is undecidable for ADA.
 
@@
 mapping each formula in $\Form^+(Q,\vec{x})$ to a formula defining the
-effect of reading the event $a$.
-
-We generalize the post-image function to finite sequences of input
-events, as follows:
+effect of reading the event $a$. We generalize the post-image function
+to finite sequences of input events, as follows:
 \[\begin{array}{l}
@@
 
-In general, given an unsatisfiable conjunction $\Phi \equiv
-\phi_1(X_0,X_1) \wedge \phi_2(X_1,X_2) \wedge \ldots \wedge
-\phi_n(X_{n-1},X_n)$ of time-stamped variables $X_i = \set{x_i \mid x
-  \in X}$ of any sort, a solution of the \emph{interpolation problem}
-$\Phi$, simply called an \emph{interpolant}, is a tuple $\tuple{I_0(X),
-  I_1(X), \ldots, I_n(X)}$ such that:\begin{inparaenum}[(i)]
-\item $I_0 \equiv \top$,  
-%
-\item $I_{i-1}[X_{i-1}/X] \wedge \phi_i(X_{i-1},X_i) \models
-  I_i[X_i/X]$, for all $i \in [1,n]$, and
-%
-\item $I_n \equiv \bot$. 
-\end{inparaenum}
-In the following, we shall assume the existence of an interpolating
-decision procedure for $\theory(\sig,\I)$.
+Given an unsatisfiable conjunction of formulae $\psi_1 \wedge \ldots
+\wedge \psi_n$, an \emph{interpolant} is a tuple of formulae
+$\tuple{I_1,\ldots,I_{n-1},I_n}$ such that $I_n \equiv \bot$, $I_i
+\wedge \psi_i \models_\T I_{i+1}$ and $I_i$ contains only variables
+and function symbols that are common to $\psi_i$ and $\psi_{i+1}$, for
+all $i \in [n-1]$. Moreover, by Lyndon's Interpolation Theorem
+\cite{Lyndon59}, we can assume without loss of generality that every
+boolean variable with at least one positive (negative) occurrence in
+$I_i$ has at least one positive (negative) occurrence in both $\psi_i$
+and $\psi_{i+1}$. In the following, we shall assume the existence of
+an interpolating decision procedure for $\theory(\sig,\I)$ that meets
+the requirements of Lyndon's Interpolation Theorem.
 
@@
 to be false.
-
-Moreover, we require that an interpolant
-$\tuple{\top,I_0,\ldots,I_n,\bot}$ for the interpolation problem
-$\Theta(u)$ does not have negative occurrences of states, i.e.\ $I_i
-\in \Form^+(Q,\vec{x})$, for all $i \in [0,n]$. Such an interpolant
-can always be built, as showed below:
-
-\begin{proposition}\label{prop:positive-interpol}
-  If $\Theta(a_1 \ldots a_n)$ is unsatisfiable then one can build an
-  interpolant $\tuple{\top,I_0,\ldots,I_n,\bot}$ for $\Theta(a_1\ldots
-  a_n)$, such that $I_i \in \Form^+(Q,\vec{x})$, for all $i \in
-  [0,n]$.
-\end{proposition}
-
+%
+%% Moreover, we require that an interpolant
+%% $\tuple{\top,I_0,\ldots,I_n,\bot}$ for the interpolation problem
+%% $\Theta(u)$ does not have negative occurrences of states, i.e.\ $I_i
+%% \in \Form^+(Q,\vec{x})$, for all $i \in [0,n]$. Such an interpolant
+%% can always be built, as showed below:
+%
+%% \begin{proposition}\label{prop:positive-interpol}
+%%   If $\Theta(a_1 \ldots a_n)$ is unsatisfiable then one can build an
+%%   interpolant $\tuple{\top,I_0,\ldots,I_n,\bot}$ for $\Theta(a_1\ldots
+%%   a_n)$, such that $I_i \in \Form^+(Q,\vec{x})$, for all $i \in
+%%   [0,n]$.
+%% \end{proposition}
+%
 We recall that a run of $\A$ over $u$ is a sequence: \[\phi_0(Q)
@@
 spurious counterexample is eliminated by computing an interpolant and
-adding its formulae to the set of predicates (cf. Lemma \ref{lemma:progress}).
+adding its formulae to the set of predicates (cf. Lemma
+\ref{lemma:progress}). Formally, an ART is tuple $\Art =
+\tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$, where: \begin{compactitem}
+\item $N$ is a set of nodes, 
+%
+\item $E \subseteq N \times \Sigma \times
+N$ is a set of edges, 
+%
+\item $\rootNode \in N$ is the root of the directed
+tree $(N,E)$, 
+%
+\item $\Lambda : N \rightarrow \Form(Q,\vec{x})$ is a labeling of the
+  nodes with formulae, such that $\Lambda(\rootNode) = \iota$,
+%
+\item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
+  sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$,
+%
+\item $T : E \rightarrow \bigcup_{i=0}^\infty
+  \Form^+(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is a labeling of edges
+  with time-stamped formulae, and
+%
+\item $\lhd \subseteq N \times N$ is a set of
+\emph{covering edges}. 
+\end{compactitem}
 
@@
 
-Formally, an ART is tuple $\Art =
-\tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$, where: \begin{compactitem}
-\item $N$ is a set of nodes, 
-%
-\item $E \subseteq N \times \Sigma \times
-N$ is a set of edges, 
-%
-\item $\rootNode \in N$ is the root of the directed
-tree $(N,E)$, 
-%
-\item $\Lambda : N \rightarrow \Form(Q,\vec{x})$ is a labeling of the
-  nodes with formulae, such that $\Lambda(\rootNode) = \iota$,
-%
-\item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
-  sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$,
-%
-\item $T : E \rightarrow \bigcup_{i=0}^\infty
-  \Form^+(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is a labeling of edges
-  with time-stamped formulae, and
-%
-\item $\lhd \subseteq N \times N$ is a set of
-\emph{covering edges}. 
-\end{compactitem}
 %% For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is $(n,a,m)$.
@@
 model of $\Accept{\A}(u)$ (since $u$ is feasible, clearly
-$\Accept{\A}(u)$ is satisfiable). Otherwise, if $u$ is spurious, we
+$\Accept{\A}(u)$ is satisfiable). Otherwise, $u$ is spurious and we
 compute its pivot $p$ (line \ref{ln:pivot}), add the interpolants for
-the least unfeasible suffix of $u$ to the set of predicates $\preds$,
-remove and recompute the subtree of $\Art$ rooted at $p$.
+the least unfeasible suffix of $u$ to $\preds$, remove and recompute
+the subtree of $\Art$ rooted at $p$.
 
@@
 
-\begin{figure}[t!]
+\begin{figure}[htb]
+\begin{center}
 \begin{tabular}{cc}
@@
 \end{tabular}
+\end{center}
 \vspace*{-\baselineskip}
@@
 \begin{figure}[htb]
-\vspace*{-\baselineskip}
+\begin{center}
+\begin{minipage}{6cm}
 \begin{tabular}{c}
-\includegraphics[scale=0.65]{IM1.pdf} \\
+\includegraphics[scale=0.60]{IM1.pdf} \\
 \tiny{(a)} \\
-\includegraphics[scale=0.65]{IM2.pdf} \\[-2mm]
+\includegraphics[scale=0.60]{IM2.pdf} \\[-2mm]
 \tiny{(b)} \\
-\includegraphics[scale=0.65]{IM3.pdf} \\[-2mm]
+\includegraphics[scale=0.60]{IM3.pdf} \\[-2mm]
 \tiny{(c)}
 \end{tabular}
-\vspace*{-0.5\baselineskip}
-\caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex}
-  by Algorithm \ref{alg:impact} (1/2)}
-\label{fig:im1}
-\vspace*{-\baselineskip}
-\end{figure}
-
-\begin{figure}[htb]
-\centerline{\includegraphics[scale=0.65]{IM4.pdf}}
+\end{minipage}
+\begin{minipage}{6cm}
+\includegraphics[scale=0.60]{IM4.pdf}
+\centerline{\tiny{(d)}}
+\end{minipage}
+\end{center}
 \vspace*{-\baselineskip}
 \caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex}
-  by Algorithm \ref{alg:impact} (2/2)}
-\label{fig:im2}
+  by Algorithm \ref{alg:impact}}
+\label{fig:im}
 \end{figure}
@@
 from Figure \ref{fig:ex}.  Initially, the procedure fires the sequence
-$a$, whose endpoint is labeled with $\top$, in Figure \ref{fig:im1}
+$a$, whose endpoint is labeled with $\top$, in Figure \ref{fig:im}
 (a). Since this node is uncovered, we check the spuriousness of the
@@
 computed, corresponding to the sequences $aa$ and $ab$, in Figure
-\ref{fig:im1} (b). The spuriousness check for $aa$ yields the
+\ref{fig:im} (b). The spuriousness check for $aa$ yields the
 interpolant $\tuple{q_0,x\leq0\wedge q_2 \wedge y\geq0}$ which
@@
 endpoint of $aa$ is not covered, it is expanded to $aaa$ and $aab$, in
-Figure \ref{fig:im1} (c). Both sequences $aaa$ and $aab$ are found to
+Figure \ref{fig:im} (c). Both sequences $aaa$ and $aab$ are found to
 be spurious, and the enpoint of $aab$, whose label has changed from
@@
 uncovered node $aaa$ is expanded to $aaaa$ and $aaab$, both found to be
-spurious, in Figure \ref{fig:im2}. The refinement of $aaab$ causes
+spurious, in Figure \ref{fig:im} (d). The refinement of $aaab$ causes
 the label of $aaa$ to change from $q_1$ to $q_1 \wedge y>x-1 \wedge
@@
 
-\subsection{Proof of Lemma \ref{lemma:closure}}
+\subsection*{Proof of Lemma \ref{lemma:closure}}
 
@@
   let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
+  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
+  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
+  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
+  \subseteq Q$ and thus $\overline{\iota_1} \equiv \iota_1$. For the case
+  $n>0$, we compute:
   $\overline{\Delta}(\overline{\iota_1},a_1 \ldots a_n) =
@@
 
-\subsection{Proof of Lemma \ref{lemma:safety-invariant}}
+\subsection*{Proof of Lemma \ref{lemma:safety-invariant}}
 
@@
 
-\subsection{Proof of Lemma \ref{lemma:abstract-invariant}}
+\subsection*{Proof of Lemma \ref{lemma:abstract-invariant}}
 
@@
 
-\subsection{Proof of Proposition \ref{prop:positive-interpol}}
+%% \subsection*{Proof of Proposition \ref{prop:positive-interpol}}
+
+%% \vspace*{\baselineskip} \proof{ Let
+%%   $\tuple{\top,I_0,\ldots,I_{n},\bot}$ be an interpolant, and $i \in
+%%   [0,n]$ be the first index for which $I_i \not\in
+%%   \Form^+(Q,\vec{x})$. If $i=0$, we replace $I_0$ with $\iota \in
+%%   \Form^+(Q,\vec{x})$ and verify that the result is still an
+%%   interpolant. If $i>0$, by Proposition \ref{prop:asterisk}, we have
+%%   that $\Delta^{i-1}(I_{i-1},a_i) \iffeq \exists Q_{i-1} ~.~
+%%   I_{i-1}[Q_{i-1}/Q] \wedge \theta_i$. Since $I_{i-1} \in
+%%   \Form^+(Q,\vec{x})$, by the choice of $i$, we have that
+%%   $\Delta^{i-1}(I_{i-1},a_i) \in \Form^+(Q,\vec{x})$ as well. We
+%%   replace $I_i$ with $\Delta^{i-1}(I_{i-1},a_i)$. The result is still
+%%   an interpolant, because: \begin{compactitem}
+%%   \item $I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \models
+%%     \Delta^{i-1}(I_{i-1},a_i)$, by Proposition \ref{prop:asterisk}, and
+%%     %
+%%   \item $\Delta^{i-1}(I_{i-1},a_i) \wedge \theta_{i+1} \models \exists
+%%     Q_{i-1} ~.~ I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \wedge \theta_{i+1}
+%%     \models I_i[Q_i/Q] \wedge \theta_{i+1} \models I_{i+1}[Q_{i+1}/Q]$.
+%%   \end{compactitem}
+%%   We proceed with these replacements until there are no more formulae
+%%   $I_i \not\in \Form^+(Q,\vec{x})$ left. \qed}
+
+\subsection*{Proof of Lemma \ref{lemma:progress}}
 
@@
 
-\vspace*{\baselineskip} \proof{ Let
-  $\tuple{\top,I_0,\ldots,I_{n},\bot}$ be an interpolant, and $i \in
-  [0,n]$ be the first index for which $I_i \not\in
-  \Form^+(Q,\vec{x})$. If $i=0$, we replace $I_0$ with $\iota \in
-  \Form^+(Q,\vec{x})$ and verify that the result is still an
-  interpolant. If $i>0$, by Proposition \ref{prop:asterisk}, we have
-  that $\Delta^{i-1}(I_{i-1},a_i) \iffeq \exists Q_{i-1} ~.~
-  I_{i-1}[Q_{i-1}/Q] \wedge \theta_i$. Since $I_{i-1} \in
-  \Form^+(Q,\vec{x})$, by the choice of $i$, we have that
-  $\Delta^{i-1}(I_{i-1},a_i) \in \Form^+(Q,\vec{x})$ as well. We
-  replace $I_i$ with $\Delta^{i-1}(I_{i-1},a_i)$. The result is still
-  an interpolant, because: \begin{compactitem}
-  \item $I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \models
-    \Delta^{i-1}(I_{i-1},a_i)$, by Proposition \ref{prop:asterisk}, and
-    %
-  \item $\Delta^{i-1}(I_{i-1},a_i) \wedge \theta_{i+1} \models \exists
-    Q_{i-1} ~.~ I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \wedge \theta_{i+1}
-    \models I_i[Q_i/Q] \wedge \theta_{i+1} \models I_{i+1}[Q_{i+1}/Q]$.
-  \end{compactitem}
-  We proceed with these replacements until there are no more formulae
-  $I_i \not\in \Form^+(Q,\vec{x})$ left. \qed}
-
-\subsection{Proof of Lemma \ref{lemma:progress}}
-
 \proof{ Let $\Theta(u) \equiv \theta_0(Q_0) \wedge \theta_1(Q_0 \cup
@@
 
-\subsection{Proof of Theorem \ref{thm:predabs}}
+\subsection*{Proof of Theorem \ref{thm:predabs}}
 
@@
 
-\subsection{Proof of Theorem \ref{thm:impact}}
+\subsection*{Proof of Theorem \ref{thm:impact}}
 
@@
   (q_k \rightarrow \bot)$, such that $I^\omega_i \in
-  \Form^+(Q,\vec{x})$, for all $i \in [0,k]$. According to Proposition
-  \ref{prop:positive-interpol}, it is possible to build such an
-  interpolant, when $\Phi(a_1 \ldots a_k)$ is unsatisfiable. By
-  Proposition \ref{prop:asterisk}, we obtain
-  $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q] \iffeq \exists Q_{i-1} ~.~
+  \Form^+(Q,\vec{x})$, for all $i \in [0,k]$. According to Lyndon's
+  Interpolation Theorem, it is possible to build such an interpolant,
+  when $\Phi(a_1 \ldots a_k)$ is unsatisfiable. By Proposition
+  \ref{prop:asterisk}, we obtain $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q]
+  \iffeq \exists Q_{i-1} ~.~
   I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge \theta_i$

### 2017-10-20 quantifiers/draft.tex

@@
 \usepackage{paralist}
-\usepackage{stmaryrd}
+% \usepackage{stmaryrd}
 \usepackage{txfonts}
@@
 $\#(f)=0$ then $f$ is a \emph{constant}. Let $\top$ and $\bot$ be
-constants of sort $\Bool$ and $\mathbf{0}$ be a constant of sort
-$\Data$.
+constants of sort $\Bool$.
+
+%% and $\mathbf{0}$ be a constant of sort $\Data$.
 
@@
 equality $t \teq s$ between two terms of sort $\Data$. A
-\emph{formula} is a possibly quantified boolean combination of atoms.
+\emph{formula} is a possibly quantified boolean combination of atoms
+and a \emph{sentence} is a formula in which each variable occurs under
+the scope of a quantifier.
 
@@
 obtained by simultaneously replacing $x_i$ with $t_i$ in $\phi$, for
-all $i\in[n]$. The size $\len{\phi}$ is the number of symbols
-occuring in $\phi$.
+all $i\in[n]$. The size $\len{\phi}$ is the number of symbols occuring
+in $\phi$.
 
@@
 Given formulae $\varphi$ and $\psi$, we say that \emph{$\phi$ entails
-  $\psi$}, denoted $\phi \models^\I \psi$ iff $\I,\nu \models \varphi$
-implies $\I,\nu \models \psi$, for each valuation $\nu$, and $\phi
-\iffeq^\I \psi$ iff $\phi \models^\I \psi$ and $\psi \models^\I \phi$.
-We omit mentioning the interpretation $\I$ when it is clear from the
-context.
+  $\psi$ modulo $\theory(\sig,\I)$}, denoted $\phi
+\models_{\theory(\sig,\I)} \psi$ iff $\varphi \rightarrow \psi \in
+\theory(\sig,\I)$. We say that $\varphi$ and $\psi$ are
+\emph{$\theory(\sig,\I)$-equivalent}, denoted $\phi \iffeq_{\theory(\sig,\I)} \psi$ iff
+$\phi \models_{\theory(\sig,\I)} \psi$ and $\psi
+\models_{\theory(\sig,\I)} \phi$. The theory $\theory(\sig,\I)$ has
+\emph{quantifier elimination} iff for any formula $\exists x ~.~
+\phi(x)$, where $\phi$ is quantifier-free, there exists a quantifier
+free formula $\psi$ such that $\phi$ and $\psi$ are
+$\theory(\sig,\I)$-equivalent.
 
@@
 
-We fix an interpretation $\I$ and a finite alphabet $\Sigma$ of
-\emph{input events}. Given a finite set $X \subset \vars$ of
-variables of sort $\Data$, let $X \mapsto \Data^\I$ be the set
-of valuations of the variables $X$ and $\Sigma[X] = \Sigma
-\times (X \mapsto \Data^\I)$ be the possibly infinite set of
-\emph{data symbols}. A \emph{data word} (word in the sequel) is a
-finite sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data
-symbols, where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n :
-X \rightarrow \Data^\I$ are valuations. We denote by
-$\varepsilon$ the empty sequence, by $\Sigma^*$ the set of finite
-sequences of input events and by $\Sigma[X]^*$ the set of data
-words over the variables $X$.
+We fix a theory $\T = \theory(\sig,\I)$ and a finite alphabet $\Sigma$
+of \emph{input events}. Given a finite set $X \subset \vars$ of
+variables of sort $\Data$, let $X \mapsto \Data^\I$ be the set of
+valuations of the variables $X$ and $\Sigma[X] = \Sigma \times (X
+\mapsto \Data^\I)$ be the possibly infinite set of \emph{data
+  symbols}. A \emph{data word} (word in the sequel) is a finite
+sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data symbols,
+where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n : X
+\rightarrow \Data^\I$ are valuations. We denote by $\varepsilon$ the
+empty sequence, by $\Sigma^*$ the set of finite sequences of input
+events and by $\Sigma[X]^*$ the set of data words over the variables
+$X$.
 
 Given sets of variables $B,X,Y \subset \vars$ of sorts $\Bool$,
-$\Data$ and $\Data$, respectively, we denote by $\Form\tuple{B;X;Y}$
-the set of formulae $\phi$ such that $\fv{\Bool}{\phi} = B$,
-$\fv{\Data}{\phi} = X$ and each variable from $Y$ occurs under the
-scope of a quantifier. By $\Form^+\tuple{B;X;Y}$ we denote the subset
-of $\Form\tuple{B;X;Y}$ in which each boolean variable from $B$ occurs
-under an even number of negations. We denote by $\phi\tuple{B;X;Y}$
-any formula in $\Form\tuple{B;X;Y}$.
+$\Data$ and $\Data$, respectively, we denote by $\Form(B;X;Y)$ the set
+of formulae $\phi$ such that $\fv{\Bool}{\phi} = B$, $\fv{\Data}{\phi}
+= X$ and each variable from $Y$ occurs under the scope of a
+quantifier. By $\Form^+(B;X;Y)$ we denote the subset of $\Form(B;X;Y)$
+in which each boolean variable from $B$ occurs under an even number of
+negations. We denote by $\phi(B;X;Y)$ any formula in $\Form(B;X;Y)$.
 
-A \emph{first-order alternating data automaton} (fo-ADA or automaton
-in the sequel) is a tuple $\A =
-\tuple{G,L,Q,\iota,F,\Delta}$, where: \begin{compactitem}
+A \emph{first-order alternating data automaton} (automaton in the
+sequel) is a tuple $\A = \tuple{G,L,Q,\iota,F,\Delta}$,
+where: \begin{compactitem}
 %
@@
 %
-\item $\iota \in \Form^+\tuple{Q;\emptyset;\emptyset}$ is the \emph{initial
+\item $\iota \in \Form^+(Q;\emptyset;\emptyset)$ is the \emph{initial
   configuration},
@@
 \item $\Delta : Q \times \Sigma \rightarrow
-  \Form^+\tuple{Q;\overline{G}\cup G\cup\overline{L};L}$ is a
+  \Form^+(Q;G\cup\overline{G}\cup\overline{L};L)$ is a
   \emph{transition function},
@@
 where, for a set $X \subseteq \vars$, we denote $\overline{X} =
-\set{\overline{x} \mid x \in X}$. The \emph{size} of $\A$ is defined
-as $\len{\A} = \len{\iota} + \sum_{(q,a) \in
-  Q\times\Sigma}\len{\Delta(q,a)}$.
-
-Intuitively, the variables $\overline{x}$ from a transition formula
-$\Delta(q,a)$ refer to past values, whereas the other variables refer
-to the current values. Observe that the initial values of the global
-variables are unconstrained (since the initial configuration is a
-boolean combination of states without variables) and quantification
-occurs only over the current copy of the local variables $L$. As a
-remark, the transition relation does not quantify over global
-variables $G \cup \overline{G}$, which must be visible throughout the
+\set{\overline{x} \mid x \in X}$. Intuitively, the variables
+$\overline{x}$ from a transition formula $\Delta(q,a)$ refer to past
+values, whereas the other variables refer to the current
+values. Observe that the initial values of the global variables are
+unconstrained (since the initial configuration is a boolean
+combination of states without variables) and quantification occurs
+only over the current copy of the local variables $L$. As a remark,
+the transition relation does not quantify over global variables $G
+\cup \overline{G}$, which must be visible throughout the
 execution. Furthermore, we assume that the past values of the local
 variables are coined and beyond our control, thus quantification over
-$\overline{L}$ is not allowed either.
+$\overline{L}$ is not allowed either. Last, observe that $\Delta$ is a
+total function, defined\footnote{A partial transition relation $R$ can
+  be encoded by setting $\Delta(q,a) = \bot$ when $R(q,a)$ is
+  undefined.} for all pairs $(q,a) \in Q \times \Sigma$. The
+\emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
+\sum_{(q,a) \in Q\times\Sigma}\len{\Delta(q,a)}$.
 
-In order to define the semantics of automata, we consider time-stamped
-sets of variables $\stamp{X}{n} = \{\stamp{x}{n} \mid x \in X\}$ for
-any $n\geq0$. For an input event $a \in \Sigma$, some constant
-$n\geq0$ and a formula $\phi$, we write $\stamp{\Delta}{n}(\phi,a)$
-for the formula obtained from $\phi$ by simultaneously replacing each
-state $q \in \fv{\Bool}{\phi}$ with
-$\Delta(\phi,a)[\stamp{G}{n}/\overline{G}, \stamp{G}{n+1}/G,
-  \stamp{L}{n}/\overline{L}, \stamp{L}{n+1}/L]$.
-
-Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in
-\Sigma[G]^*$ the \emph{run} of $\A$ over $w$ is the sequence of
-formulae: \[\iota \stackrel{(a_1,\nu_1)}{\Longrightarrow} \phi_1
-\tuple{Q,\stamp{G}{1}, \stamp{L}{1}}
-\stackrel{(a_2,\nu_2)}{\Longrightarrow} \ldots
-\stackrel{(a_n,\nu_n)}{\Longrightarrow} \phi_n\tuple{Q,\stamp{G}{1}
-  \cup \ldots \cup \stamp{G}{n}, \stamp{L}{1} \cup \ldots \cup
-  \stamp{L}{n}}\] where $\phi_1 \equiv
-\stamp{\Delta}{0}(\iota,a_1)[\mathbf{0}/\stamp{G}{0},\mathbf{0}/\stamp{L}{0}]$
-and $\phi_{i+1} \equiv \stamp{\Delta}{i}(\phi_i,a_{i+1})$, for all $i
-\in [1,n-1]$. In other words, we assume that all data variables are
-initially $\mathbf{0}$.
-
-We say that $\A$ \emph{accepts} $w$ if there exists a valuation $\nu$
-such that $\nu(\stamp{x}{i}) = \nu_i(x)$ for each global variable $x
-\in G$ and each timestamp $i \in [n]$, $\nu(q) = \true$ for each $q
-\in F$, $\nu(q) = \false$ for each $q \in Q \setminus F$ and $\I,\nu
-\models \phi_n$. The \emph{language} of $\A$ is the set $L(\A)$ of
-words from $\Sigma[G]^*$ accepted by $\A$. In this paper, we ask the
-following two questions: \begin{compactenum}
+In order to define the semantics of automata, we consider the
+\emph{time-stamped} sets of variables $\stamp{X}{n} = \{\stamp{x}{n}
+\mid x \in X\}$ for any $n\in\zed$. For an input event $a \in \Sigma$,
+a positive integer $n\geq0$ and a formula $\varphi\in\Form^+(Q;G;L)$,
+we write $\stamp{\Delta}{n}(\varphi,a)$ for the formula obtained from
+$\varphi$ by simultaneously replacing each state $q \in
+\fv{\Bool}{\phi}$ with the formula
+$\Delta(q,a)[\stamp{G}{n}/\overline{G}, \stamp{G}{n+1}/G,
+  \stamp{L}{n}/\overline{L}, \stamp{L}{n+1}/L]$. For a sequence $a_1
+\ldots a_n \in \Sigma^*$, we consider the formulae \(\iota \equiv
+\phi_0, \phi_1, \ldots, \phi_n \), where $\phi_{i} \equiv
+\stamp{\Delta}{i-1}(\phi_{i-1},a_{i})$, for all $i \in [n]$. $\A$
+\emph{accepts} a word $w = (a_1,\nu_1) \ldots (a_n,\nu_n) \in
+\Sigma[G]^*$ iff there exists a valuation $\nu$ such that
+$\I,\nu \models \phi_n$ and: \begin{compactitem}
+%% \item $\nu(\stamp{x}{0}) = \mathbf{0}^\I$ for each variable $x \in G
+%%   \cup L$, i.e.\ all data variables are initially $\mathbf{0}$,
+%
+\item $\nu(\stamp{x}{i}) = \nu_i(x)$ for each global variable $x\in G$
+  and each time-stamp $i \in [n]$,
+%
+\item $\nu(q) = \true$ if and only if $q \in F$. 
+\end{compactitem}
+The \emph{language} of $\A$ is the set $L(\A)$ of words from
+$\Sigma[G]^*$ accepted by $\A$. In this paper, we ask the following
+questions: \begin{compactenum}
 \item \emph{boolean closure}: given automata $\A_i =
@@
 Given a set $B$ of boolean variables, two disjoint sets $X$ and $Y$ of
-data variables, for any formula $\phi \in \Form^+\tuple{B;X;Y}$,
-define $\overline{\phi} \in \Form^+\tuple{B;X;Y}$ recursively:
+data variables, for any formula $\phi \in \Form^+(B;X;Y)$,
+define $\overline{\phi} \in \Form^+(B;X;Y)$ recursively:
 \[\begin{array}{rclcrclcrcl}
@@
   and data respectively, and a formula $\phi \in
-  \Form^+\tuple{B;X;Y}$, for any valuation $\nu$ we have $\nu \models
+  \Form^+(B;X;Y)$, for any valuation $\nu$ we have $\nu \models
   \phi$ if and only if $\overline{\nu} \models \neg\overline{\phi}$,
@@
 emptiness problem for 2-counter machines \cite{Minsky67}, we give a
-semi-algorithm with the following properties. Given an automaton $\A$
-as input, the procedure will\begin{inparaenum}[(1)] 
-\item return a word $w \in L(\A)$ if $L(\A)
-\neq \emptyset$, and 
+semi-algorithm with the following properties. Given an automaton $\A =
+\tuple{G,L,Q,\iota,F,\Delta}$ as input, the procedure
+will\begin{inparaenum}[(1)]
+\item return a word $w \in L(\A)$ if $L(\A) \neq \emptyset$, and
 %
@@
   we can safely conclude that $L(\A) = \emptyset$. \end{inparaenum}
-Due to the undecidability of the emptiness problem, it is possible for
-the semi-algorithm to run forever in the case when the language of the
-automaton is empty.
+Due to the undecidability of the emptiness problem, there exist
+automata with empty language on which the procedure runs forever.
+
+The semi-algorithm described in this section reduces the language
+emptiness problem for $\A$ to a (possibly infinite) sequence of
+satisfiability problems. We generate these formulae systematically
+using a tree whose backbone is a finite set of input events and whose
+labels are quantifier-free formulae. Given two input sequences
+$\alpha,\beta \in \Sigma^*$, we say that $\beta$ is a prefix of
+$\alpha$, written $\beta \preceq \alpha$ if $\alpha = \beta \gamma$,
+for some sequence $\gamma \in \Sigma^*$. A set $S \subseteq \Sigma^*$
+is \emph{prefix-closed} if for each $\alpha \in S$, $\beta \preceq
+\alpha$ implies $\beta \in S$ and \emph{complete} if for each $\alpha
+\in S$ there exists $a \in \Sigma$ such that $\alpha a \in S$ if and
+only if $\alpha b \in S$ for all $b \in \Sigma$.
+
+For a set of variables $X \subseteq \vars$, let $\stamp{X}{-\infty} =
+\bigcup_{n\leq0} \stamp{X}{n}$ be a set of \emph{history} variables,
+where $\stamp{X}{0}$ track the current values, $\stamp{X}{-1}$ the
+immediate predecessors, etc. For a formula $\varphi$ and an integer $n
+\in \zed$, we denote by $\stamp{\varphi}{n}$ the formula in which each
+time-stamped free variable $\stamp{x}{i}$ is replaced with
+$\stamp{x}{i+n}$. Given formulae $\varphi,\psi \in
+\Form^+(Q;\stamp{X}{-\infty}; \emptyset)$ and a sequence $a_1 \ldots
+a_n \in \Sigma^*$, we write $\varphi \darrow{a_1 \ldots a_n}{\A} \psi$
+if there exist formulae $\phi_0, \ldots, \phi_n$ such
+that\begin{inparaenum}[(i)]
+\item $\varphi \equiv \phi_0$, 
+\item $\phi_i \equiv \stamp{\Delta}{i-1}(\phi_{i-1},a_i)$, for all $i
+  \in [n]$, and
+\item $\psi \equiv \stamp{\phi_n}{-n}$. \end{inparaenum} Observe that
+$\psi$ is uniquely determined by $\varphi$, because the result of the
+substitutions $\stamp{\Delta}{i-1}(\phi_{i-1},a_i)$ and
+$\stamp{\phi_n}{-n}$ is unique.
+
+\begin{proposition}\label{prop:monotonicity}
+  For all $\varphi,\psi \in \Form^+(Q;\stamp{X}{-\infty}; \emptyset)$
+  and $w \in \Sigma^*$, if $\varphi \models_\T \psi$ and $\varphi',
+  \psi'$ are the unique formulae such that $\varphi \darrow{w}{\A}
+  \varphi'$ and $\psi \darrow{w}{\A} \psi'$, then $\varphi' \models_\T
+  \psi'$.
+\end{proposition}
+\proof{By induction on the length of $w = a_1 \ldots a_n$, using the
+  fact that any state $q \in Q$ occurs positively in each formula from
+  the sequences $\varphi \equiv \varphi_0, \ldots, \varphi_n$ and
+  $\psi \equiv \psi_0, \ldots, \psi_n$, defining $\varphi
+  \darrow{w}{\A} \varphi' \equiv \stamp{\psi_n}{-n}$ and $\psi
+  \darrow{w}{\A} \stamp{\psi_n}{-n} \equiv \psi'$, respectively. \qed}
+
+\begin{definition}\label{def:unfolding}
+An \emph{unfolding} of $\A$ is a finite mapping $U : \Sigma^*
+\rightharpoonup_{\mathit{fin}} \Form^+(Q;\stamp{(G\cup L)}{-\infty};
+\emptyset)$ such that $\dom(U)$ is a prefix-closed complete set,
+$U(\varepsilon) = \iota$ and, for all $\alpha a \in \dom(U)$, we have
+$\stamp{\Delta}{0}(U(\alpha),a) \models_\T \stamp{U(\alpha a)}{1}$.
+Moreover, $U$ is \emph{safe} if for all $\alpha \in \dom(U)$, the
+formula $U(\alpha) \wedge \mathit{Final}$ is unsatisfiable, where
+$\mathit{Final} \equiv \bigwedge_{q \not\in F} q \rightarrow \bot$.
+%% $\mathit{Init} \equiv \bigwedge_{x \in \stamp{G}{0} \cup
+%% \stamp{L}{0}} x \teq \mathbf{0}$ and
+\end{definition}
+
+The elements of $\dom(U)$ are called \emph{nodes} and the
+prefix-maximal nodes \emph{leaves}. A node $\alpha \in \dom(U)$ is
+\emph{covered} by another node $\beta \in \dom(U)$ if there exists a
+prefix $\alpha' \preceq \alpha$ such that $U(\alpha') \models_\T
+U(\beta)$. A node is \emph{uncovered} if it is not covered by any
+other node. We say that $U$ is \emph{closed} when each leaf in
+$\dom(U)$ is covered by an uncovered node.
+
+\begin{lemma}\label{lemma:art}
+  If $\A$ has a nonempty safe closed unfolding, then $L(\A) =
+  \emptyset$.
+\end{lemma}
+\proof{ Let $U$ be a safe closed unfolding of $\A$ and suppose, for a
+  contradiction, that $w = (a_1,\nu_1) \ldots (a_n,\nu_n) \in
+  L(\A)$. Let $\alpha = a_1 \ldots a_n \in \Sigma^*$ and
+  $\phi_{\alpha}$ be the unique formula such that $\iota
+  \darrow{\alpha}{\A} \phi_{\alpha}$. Since $w \in L(\A)$, we have
+  that $\phi_{\alpha} \wedge \mathit{Final}$ is satisfiable. It is
+  sufficient to show that there exists a node $\beta \in \dom(U)$ such
+  that $\phi_{\alpha} \models_\T U(\beta)$, because then $U(\beta)
+  \wedge \mathit{Final}$ is satisfiable as well and $U$ is not safe,
+  contradiction.
+
+  If $\alpha \in \dom(U)$ then we obtain $\phi_{\alpha} \models_\T
+  U(\alpha)$ from the fact that $U$ is an unfolding, applying
+  Proposition \ref{prop:monotonicity} inductively on the length of
+  $\alpha$. Otherwise, let $\alpha' \in \dom(U)$ be the unique leaf
+  which is a prefix of $\alpha = \alpha'\gamma$, for a unique nonempty
+  sequence $\gamma$. Such a leaf must exist because $\dom(U) \neq
+  \emptyset$. Moreover, since $U$ is closed, $\alpha'$ must be covered
+  and let $\alpha_1$ be the prefix of $\alpha'$ such that $U(\alpha_1)
+  \models_\T U(\beta_1)$, for some uncovered node $\beta_1$. Let
+  $\phi_1$ be the unique formula such that $\iota
+  \darrow{\alpha_1}{\A} \phi_1$. Applying Proposition
+  \ref{prop:monotonicity} inductively, we obtain that $\phi_1
+  \models_\T U(\alpha_1)$, hence $\phi_1 \models_\T U(\beta_1)$.
+  Because $\beta_1$ is uncovered, it cannot be a leaf and let
+  $\gamma_1$ be the unique prefix of $\gamma$ such that
+  $\beta_1\gamma_1$ is a leaf. Since $U$ is closed, $\beta_1\gamma_1$
+  is covered and, because $\beta_1$ is uncovered, there exists a
+  nonempty sequence $\gamma'_1 \preceq \gamma_1$ such that
+  $U(\beta_1\gamma'_1) \models_\T U(\beta_2)$ for some uncovered node
+  $\beta_2 \in \dom(U)$. Let $\alpha_2 = \alpha_1\gamma'_1$ and
+  $\phi_2$ be the unique formula such that $\iota
+  \darrow{\alpha_2}{\A} \phi_2$. Since $\phi_1 \models_\T U(\beta_1)$
+  and $\phi_1 \darrow{\gamma'_1}{\A} \phi_2$, we obtain that $\phi_2
+  \models_\T U(\beta_2)$ by an inductive application of Proposition
+  \ref{prop:monotonicity}. Continuing this argument inductively, we
+  discover a strictly increasing sequence $\alpha_1 \prec \alpha_2
+  \prec \ldots$ of prefixes of $\alpha$, a sequence of formulae $\iota
+  \darrow{\alpha_1}{\A} \phi_1, \iota \darrow{\alpha_2}{\A} \phi_2,
+  \ldots$ and a sequence of nodes $\beta_1, \beta_2, \ldots \in
+  \dom(U)$ such that $\phi_i \models_\T U(\beta_i)$, for all $i =
+  1,2,\ldots$. Since $\alpha$ is finite, it must be the case that
+  $\phi_\alpha \models_\T U(\beta_n)$, for some node $\beta_n$ in the
+  sequence. \qed}
+
+Lazy annotation semi-algorithms \cite{McMillan06,McMillan14}
+essentially try to prove language emptiness by building safe closed
+unfoldings of automata. The labels of these unfoldings are
+interpolants obtained from the proof of unsatisfiability of the
+formulae that encode the acceptance conditions for the nodes
+(sequences of input events) in the unfolding. Given an unsatisfiable
+conjunction of formulae $\psi_1 \wedge \ldots \wedge \psi_n$, an
+\emph{interpolant} is a tuple of formulae
+$\tuple{I_1,\ldots,I_{n-1},I_n}$ such that $I_n \equiv \bot$, $I_i
+\wedge \psi_i \models_\T I_{i+1}$ and $I_i$ contains only variables
+and function symbols that are common to $\psi_i$ and $\psi_{i+1}$, for
+all $i \in [n-1]$. Moreover, by Lyndon's Interpolation Theorem
+\cite{Lyndon59}, we can assume without loss of generality that every
+boolean variable with at least one positive (negative) occurrence in
+$I_i$ has at least one positive (negative) occurrence in both $\psi_i$
+and $\psi_{i+1}$.
+
+The execution semantics of automata is defined in terms of successive
+rewritings of the boolean variables (states) by time-stamped
+transition formulae. Formally, given a sequence $a_1 \ldots a_n \in
+\Sigma^*$ of input events, we have a sequence of formulae $\Phi_1,
+\ldots, \Phi_n$, where $\iota \darrow{a_1 \ldots a_i}{\A} \Phi_i$, for
+all $i \in [n]$. Altough natural for defining acceptance of
+alternating automata, this semantics does not cope well with
+interpolation, because\begin{inparaenum}[(1)]
+\item rewriting states loses track of the history of the computation,
+  and
+%
+\item quantifiers may occur nested in each of the intermediate
+  formulae $\Phi_1,\ldots,\Phi_n$.
+\end{inparaenum}
+
+More precisely, each formula in the sequence defining acceptance is of
+the form: \[\Phi_i \equiv \mathbf{Q}_1 \stamp{\mathcal{L}}{1}_1 \ldots
+\mathbf{Q}_i \stamp{\mathcal{L}}{i}_i ~.~ \psi_i(Q,\bigcup_{j=0}^i
+\stamp{G}{j} \cup \stamp{\mathcal{L}}{j}_j)\] where $\mathcal{L}_1,
+\ldots, \mathcal{L}_i$ are sequences of local variables from $L$ and
+$\mathbf{Q} \mathcal{X}$ denotes the quantifier prefix $\mathcal{Q}_1
+x_1 \ldots \mathcal{Q}_k x_k$, with $\mathbf{Q} = \mathcal{Q}_1 \ldots
+\mathcal{Q}_k$ and $\mathcal{X} = x_1,\ldots,x_k$. We deal with the
+first issue above by explicitly encoding the computation history as a
+possibly quantified conjunction of implications of the form
+$\stamp{q}{i} \rightarrow \stamp{\Delta}{i}(q,a_i)[\stamp{q}{i+1}/q]$,
+for $i \in [0,n-1]$. To this end, we use the following fact:
 
 \begin{proposition}\label{prop:substitution}
-  Given a formula $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$, where $Q_1 \ldots
-  Q_n \in \set{\exists,\forall}$ and $\phi$ is a quantifier-free
-  formula with free variables $\fv{}{\phi} = \set{x_1,\ldots,x_n,q}$
-  in which the boolean variable $q$ occurs positively, for any formula
-  $\psi$, the following equivalence holds:
-  \[Q_1 x_1 \ldots Q_n x_n ~.~ \phi[\psi/q] \iff 
-  Q_1 x_1 \ldots Q_n x_n \exists q ~.~ \phi \wedge (q \rightarrow \psi)\]
+  Given a formula $\phi$ in which $q \in \fv{\Bool}{\phi}$ occurs
+  positively, for any formula $\psi$ and any valuation $\nu$ of
+  $\fv{}{\phi} \cup \fv{}{\psi}$, \(\nu \models \phi[\psi/q]\) iff
+  \(\nu \models \exists q ~.~ \phi \wedge (q \rightarrow \psi)\).
 \end{proposition}
-\proof{ Let $\nu$ be an arbitrary valuation of the variables in
-  $\fv{}{\phi} \cup \fv{}{\psi}$. It is sufficient to prove that $\nu
-  \models \phi[\psi/q]$ if and only if $\nu \models \exists q ~.~ \phi
-  \wedge (q \rightarrow \psi)$.
-
-  ``$\Leftarrow$'' Suppose that $\nu \models \phi[\psi/q]$ and
+\proof{ ``$\Leftarrow$'' Suppose that $\nu \models \phi[\psi/q]$ and
   distinguish the following cases: \begin{compactenum}
@@
     \phi[\top/q]$, because $\phi$ is monotone in $q$ in the entailment
-    partial order. But then $\nu \models \phi[\psi/q]$.
-  \end{compactitem}
-  \qed}
+    partial order. But then $\nu \models \phi[\psi/q]$. \qed
+  \end{compactitem}}
+Next, we consider the following formulae, logically equivalent to
+$\Phi_i$, for all $i \in [n]$:
+\[\begin{array}{rcl} 
+\Theta_i & \equiv & \mathbf{\exists} \stamp{Q}{0} \mathbf{Q}_1
+\stamp{\mathcal{L}}{1}_1 \mathbf{\exists} \stamp{Q}{1} \mathbf{Q}_2
+\stamp{\mathcal{L}}{2}_2 \mathbf{\exists} \stamp{Q}{2} \ldots
+\mathbf{Q}_i \stamp{\mathcal{L}}{i}_i \mathbf{\exists} \stamp{Q}{i} 
+~.~ (\top \rightarrow \stamp{\iota}{0}) ~\wedge~ \bigwedge_{j=0}^i \Psi_j \\
+&& \text{where } \Psi_j \equiv \bigwedge_{q \in Q} \left(\stamp{q}{j} \rightarrow
+\stamp{\Delta}{j}(q,a_j)[\stamp{q}{j+1}/q]\right)\enspace.
+\end{array}\] These formulae are
+still not proper interpolation problems, due to the nested quantifier
+prefix. Thus we use Skolem witnesses to reduce the (un)satisfiability
+of the above formulae to the problem of computing an interpolant for a
+quantifier-free formula.
+
+\begin{theorem}\label{thm:skolem}
+  Given $Q_1 x_1 \ldots Q_n x_n ~.~ \varphi$ a first order sentence, 
+  where $Q_1, \ldots, Q_n \in \set{\exists,\forall}$ and $\varphi$ is
+  quantifier-free, let: 
+  \[\eta_i \equiv \left\{\begin{array}{ll}
+  f_{x_i}(y_1,\ldots,y_{k_i}) & \text{if } Q_i \equiv \forall \\
+  x_i & \text{if } Q_i \equiv \exists
+  \end{array}\right.\]
+  where $f_{x_i}$ is a fresh function symbol and $\set{y_1, \ldots,
+    y_{k_i}} = \set{x_j \mid j < i,~ Q_j \equiv \exists}$. Then we
+  have \(Q_1 x_1 \ldots Q_n x_n ~.~ \varphi \models_\T
+  \varphi[\eta_1/x_1,\ldots,\eta_n/x_n]\).
+\end{theorem}
+\proof{See \cite[Theorem 2.1.8]{BorgerGraedelGurevich97} and
+  \cite[Lemma 2.1.9]{BorgerGraedelGurevich97}. \qed} We further
+transform $\Theta_i$ into a quantifier-free formula
+$\widehat{\Theta}_i \equiv (\top \rightarrow \stamp{\iota}{0}) \wedge
+\bigwedge_{j=0}^i \widehat{\Psi}_j$, where $\widehat{\Psi}_j$ are
+obtained from $\Psi_j$ by replacing each universally quantified
+variable $x_i$ by a term $f_{x_i}(y_1, \ldots, y_{k_i})$, whose free
+variables are among the existentially quantified variables $x_j$ that
+precede $x_i$ in the quantifier prefix. By Theorem \ref{thm:skolem},
+replacing a universally quantified variable by an uninterpreted Skolem
+function symbol yields an overapproximation of the original formula,
+thus proving unsatisfiability of $\widehat{\Theta}_i$ suffices to
+prove unsatisfiability of $\Theta_i$.
+
+Furthermore, a proof of unsatisfiability for $\widehat{\Theta}_i$
+yields an interpolant $\tuple{I_0, \ldots, I_i}$, where $\fv{}{I_j}
+\subseteq \bigcup_{\ell=0}^j \stamp{G}{\ell} \cup \stamp{L}{\ell}$,
+for all $j \in [0,i]$. That is, the interpolants obtained from the
+proof of unsatisfiability of $\widehat{\Theta}_i$ may refer to all
+values taken by the global and local variables from the beginning of
+the path up to the $i$-th step.
 

### 2017-10-20 tacas/draft.tex
ok

@@
 
-\title{The Impact of Alternation}
+\title{Abstraction Refinement for Emptiness Checking of Alternating
+  Data Automata}
 
@@
 the system under consideration is contained within $\mathcal{P}$.
-
 This problem is at the core of widespread verification techniques,
@@
 this and other related verification techniques is the intractability
-of language inclusion (PSPACE-complete for finite-state automata on
-finite words over finite alphabets).
+of language inclusion (PSPACE-complete for finite-state automata over
+finite alphabets).
 
@@
 \cite{AlurDill94} or finite-memory register automata
-\cite{Kaminski94}. In some cases, one can recover theoretical
-decidability, by restricting the number of variables (clocks) in timed
-automata to one \cite{OuaknineWorrell04}, or forbidding relations
-between current and past/future values, as with symbolic automata
-\cite{symbTransd:POPL12}. In such cases, the emptiness problem for the
-alternating versions becomes decidable \cite{Lasota05,DAntoniKW16}.
+\cite{Kaminski94}. One can recover theoretical decidability, by
+restricting the number of variables (clocks) in timed automata to one
+\cite{OuaknineWorrell04}, or forbidding relations between current and
+past/future values, as with symbolic automata
+\cite{symbTransd:POPL12}. In such cases, also the emptiness problem
+for the alternating versions becomes decidable
+\cite{Lasota05,DAntoniKW16}.
 
@@
 complementation is possible in linear time, the bulk of the work is
-carried out by the decision procedure.
+carried out by the solver.
 
@@
 sort $\Data$ and return value $\sigma(f) \in \ssorts{\sig}$. If
-$\#(f)=0$ then $f$ is a \emph{constant}. We consider the constants
-$\top$ and $\bot$ of sort $\Bool$.
-
-We consider an infinite countable set of \emph{variables} $\vars$,
-where each $x \in \vars$ has an associated sort $\sigma(x)$. A
-\emph{term} $t$ of sort $\sigma(t)=S$ is a variable $x \in \vars$
-where $\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
+$\#(f)=0$ then $f$ is a \emph{constant}. We consider constants $\top$
+and $\bot$ of sort $\Bool$.
+
+Let $\vars$ be an infinite countable set of \emph{variables}, where
+each $x \in \vars$ has an associated sort $\sigma(x)$. A \emph{term}
+$t$ of sort $\sigma(t)=S$ is a variable $x \in \vars$ where
+$\sigma(x)=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
 $t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $\sigma(f)=S$. An
@@
   \Form^+(Q,\overline{\vec{x}}\cup\vec{x})$ is a \emph{transition
-    function},
+    function}, where $\overline{\vec{x}}$ denotes $\{\overline{x} \mid
+  x \in \vec{x}\}$.
 \end{compactitem}
-where $\overline{\vec{x}}=\{\overline{x} \mid x \in \vec{x}\}$. In
-each formula $\Delta(q,a)$ describing a transition rule, the variables
-$\overline{\vec{x}}$ track the previous and $\vec{x}$ the current
-values of the variables of $\A$. Observe that the initial values of
-the variables are left unconstrained, as the initial configuration
-does not contain free data variables. The size of $\A$ is defined as
-$\len{\A} = \len{\iota} + \sum_{(q,a) \in
+In each formula $\Delta(q,a)$ describing a transition rule, the
+variables $\overline{\vec{x}}$ track the previous and $\vec{x}$ the
+current values of the variables of $\A$. Observe that the initial
+values of the variables are left unconstrained, as the initial
+configuration does not contain free data variables. The size of $\A$
+is defined as $\len{\A} = \len{\iota} + \sum_{(q,a) \in
   Q\times\Sigma}\len{\Delta(q,a)}$.

### 2017-10-20 tacas/draft.tex
ok

@@
 \ref{alg:impact} in a prototype tool\footnote{The implementation is
-  available at \url{https://github.com/cathiec/AltImpact}} that uses
-the Z3 SMT solver\footnote{\url{https://github.com/Z3Prover/z3}} for
-the satisfiability queries and interpolant generation, in the theory
-of linear integer arithmetic (LIA) combined with booleans. We compared
+  available at \url{https://github.com/cathiec/JAltImpact}} that uses
+the MathSAT5 SMT solver\footnote{\url{http://mathsat.fbk.eu/}} via the
+Java SMT
+interface\footnote{\url{https://github.com/sosy-lab/java-smt}} for the
+satisfiability queries and interpolant generation, in the theory of
+linear integer arithmetic (LIA) combined with booleans. We compared
 both algorithms with a previous implementation of a trace inclusion
@@
 asynchronous compositions of timed automata, whom correctness
-specifications are given by timed automata monitors: a timed version of
-the Alternating Bit Protocol (abp) \cite{abp} and a~controller of a
+specifications are given by timed automata monitors: a timed version
+of the Alternating Bit Protocol (abp) \cite{abp} and a~controller of a
 railroad crossing (train) \cite{henzinger:RealTimeSystems}. All
-results were obtained on an Intel(R) Core(TM) i7-4650U CPU @ 1.70GHz
-with 8GB of RAM. The automata sizes are in bytes and the execution
-times are in seconds.
+results were obtained on x86\_64 Linux Ubuntu virtual machine with 8GB
+of RAM. The automata sizes are in bytes and the execution times are in
+seconds.
 
@@
 least one order of magnitude. However, both our implementations are
-slower than \textsc{Includer}, on average (except for the abp
-example). The reason for this is currently under investigation, one
-possible bottleneck being the hardness of the combined (LIA+booleans)
-interpolation problems, as opposed to converting the entire formula
-into DNF, eliminating the boolean variables and using interpolation in
-the pure LIA theory.
+slower than \textsc{Includer}, on average. The reason for this is
+currently under investigation, one possible bottleneck being the
+hardness of the combined (LIA+booleans) interpolation problems, as
+opposed to converting the entire formula into DNF, eliminating the
+boolean variables and using interpolation in the pure LIA theory.
 

### 2017-10-21 tacas/draft.tex
experiment result
ok

--- 2017-10-20 tacas/draft.tex
@@
 
-Unlike the boolean closure properties, showed to be effectively
-decidable (Lemma \ref{lemma:closure}), the emptiness problem is
-undecidable, even in very simple cases. For instance, if $\Data^\I$ is
-the set of positive integers, an ADA can simulate an Alternating
-Vector Addition System with States (AVASS) using only atoms $x \geq k$
-and $x = \overline{x} + k$, for $k \in \zed$, with the classical
-interpretation of the function symbols on integers. Since reachability
-of a control state is undecidable for AVASS \cite{LINCOLN92}, 
-emptiness is undecidable for ADA.
+The emptiness problem for ADA is undecidable, even in very simple
+cases. For instance, if $\Data^\I$ is the set of positive integers, an
+ADA can simulate an Alternating Vector Addition System with States
+(AVASS) using only atoms $x \geq k$ and $x = \overline{x} + k$, for $k
+\in \zed$, with the classical interpretation of the function symbols
+on integers. Since reachability of a control state is undecidable for
+AVASS \cite{LINCOLN92}, ADA emptiness is undecidable.
 
@@
 \end{center}
-\vspace*{-\baselineskip}
 \caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex} by Algorithm \ref{alg:predabs}}
@@
   \If{$\Lambda(n_i) \not\models I_i$}
+  \label{ln:entailment1}
 
@@
 \If{$\Lambda(x) \models \Lambda(y)$}
+\label{ln:entailment2}
 
@@
 \end{center}
-\vspace*{-\baselineskip}
 \caption{Proving Emptiness of the Automaton from Fig. \ref{fig:ex}
@@
 
+The correctness of Algorithm \ref{alg:impact} is coined by the theorem
+below:
+
 \begin{theorem}\label{thm:impact}
@@
 \ref{alg:impact} in a prototype tool\footnote{The implementation is
-  available at \url{https://github.com/cathiec/AltImpact}} that uses
-the Z3 SMT solver\footnote{\url{https://github.com/Z3Prover/z3}} for
-the satisfiability queries and interpolant generation, in the theory
-of linear integer arithmetic (LIA) combined with booleans. We compared
-both algorithms with a previous implementation of a trace inclusion
-procedure, called
+  available at \url{https://github.com/cathiec/JAltImpact}} that uses
+the MathSAT5 SMT solver\footnote{\url{http://mathsat.fbk.eu/}} via the
+Java SMT
+interface\footnote{\url{https://github.com/sosy-lab/java-smt}} for the
+satisfiability queries and interpolant generation, in the theory of
+linear integer arithmetic with uninterpreted boolean functions
+(UFLIA). We compared both algorithms with a previous implementation of
+a trace inclusion procedure, called
 \textsc{Includer}\footnote{\url{http://www.fit.vutbr.cz/research/groups/verifit/tools/includer/}},
 that uses on-the-fly determinisation and lazy predicate abstraction
-with interpolant-based refinement \cite{Tacas16} in the LIA theory,
-without booleans.
+with interpolant-based refinement \cite{Tacas16} in the LIA theory.
 
@@
 imperative programs with arrays \cite{cav09} (array\_shift,
-array\_simple, array\_rotation 1+2) available online
+array\_simple, array\_rotation1+2) available online
 \cite{ntslib}. Next, we applied it on proving safety properties of
@@
 asynchronous compositions of timed automata, whom correctness
-specifications are given by timed automata monitors: a timed version of
-the Alternating Bit Protocol (abp) \cite{abp} and a~controller of a
+specifications are given by timed automata monitors: a timed version
+of the Alternating Bit Protocol (abp) \cite{abp} and a~controller of a
 railroad crossing (train) \cite{henzinger:RealTimeSystems}. All
-results were obtained on an Intel(R) Core(TM) i7-4650U CPU @ 1.70GHz
-with 8GB of RAM. The automata sizes are in bytes and the execution
-times are in seconds.
+results were obtained on x86\_64 Linux Ubuntu virtual machine with 8GB
+of RAM running on an Intel(R) Xeon(R) CPU E5-2683 v3 @ 2.00GHz. The
+automata sizes are given in bytes needed to store their ASCII
+description on file and the execution times are in seconds.
 
 \begin{table}[htb]
+\vspace*{-\baselineskip}
 \begin{center}
@@
 \hline
-Example & $\len{\A}$ & $L(\A)=\emptyset$ ? & Algorithm \ref{alg:predabs} & Algorithm \ref{alg:impact} & \textsc{Includer} \\
+Example & $\len{\A}$ (bytes) & $L(\A)=\emptyset$ ? & Algorithm \ref{alg:predabs} (sec) & Algorithm \ref{alg:impact} (sec) & \textsc{Includer} (sec) \\
 \hline
-simple1 & 309 & no & 0.641 & 0.061 & 0.021\\
+simple1 & 309 & no & 0.774 & 0.064 & 0.076\\
 \hline
-simple2 & 504 & yes & 0.667 & 0.179 & 0.031\\
+simple2 & 504 & yes & 0.867 & 0.070 & 0.070\\
 \hline
-simple3 & 214 & yes & 0.825 & 0.150 & 0.051\\
+simple3 & 214 & yes & 0.899 & 0.095 & 0.095\\
 \hline
-array\_shift & 874 & yes & 2.138 & 0.209 & 0.060\\
+array\_shift & 874 & yes & 2.889 & 0.126 & 0.078\\
 \hline
-array\_simple & 3440 & yes & timeout & 16.435 & 5.496\\
+array\_simple & 3440 & yes & timeout & 9.998 & 7.154\\
 \hline
-array\_rotation1 & 1834 & yes & 6.827 & 0.865 & 0.114\\
+array\_rotation1 & 1834 & yes & 7.227 & 0.331 & 0.229\\
 \hline
-array\_rotation2 & 15182 & yes & timeout & timeout & 23.056\\
+array\_rotation2 & 15182 & yes & timeout & timeout & 31.632\\
 \hline
-abp & 6909 & no & 8.292 & 1.410 & 1.715\\
+abp & 6909 & no & 9.492 & 0.631 & 2.288\\
 \hline
-train & 1823 & yes & 16.989 & 2.738 & 0.308\\
+train & 1823 & yes & 19.237 & 0.763 & 0.678\\
 \hline
-hw1 & 322 & yes & 1.566 & 0.298 & 0.159\\
+hw1 & 322 & yes & 1.861 & 0.163 & 0.172\\
 \hline
-hw2 & 674 & yes & 22.914 & 0.552 & 0.419\\
+hw2 & 674 & yes & 24.111 & 0.308 & 0.473\\
 \hline
@@
 \end{center}
+\vspace*{-2\baselineskip}
 \end{table}
 
-As in the non-deterministic case \cite{mcmillan06}, \impact~
-outperforms lazy predicate abstraction for checking emptiness by at
-least one order of magnitude. However, both our implementations are
-slower than \textsc{Includer}, on average (except for the abp
-example). The reason for this is currently under investigation, one
-possible bottleneck being the hardness of the combined (LIA+booleans)
-interpolation problems, as opposed to converting the entire formula
-into DNF, eliminating the boolean variables and using interpolation in
-the pure LIA theory.
+As in the case of non-alternating nondeterministic integer programs
+\cite{mcmillan06}, the alternating version of \impact~ (Algorithm
+\ref{alg:impact}) outperforms lazy predicate abstraction for checking
+emptiness by at least one order of magnitude. Moreover, \impact~ is
+comparable, on average, to the previous implementation of
+\textsc{Includer}, which uses also MathSAT5 via the C API. We believe
+the reason for which \textsc{Includer} outperforms \impact~ on some
+examples is the hardness of the UFLIA entailment checks used in
+Algorithm \ref{alg:impact} (line \ref{ln:entailment1} and line
+\ref{ln:entailment2} in the function \Call{Close}{}) as opposed to the
+pure LIA entailment checks used in \textsc{Includer}. According to our
+statistics, Algorithm \ref{alg:impact} spends more than 50\% of the
+time waiting for the SMT solver to finish answering entailment
+queries.
 

### 2018-02-05 tacas/draft.tex
ok

@@
 interpretation:
-\[\begin{array}{l}
+{\scriptsize\[\begin{array}{l}
 q_0 \stackrel{(a,\tuple{0,0})}{\Longrightarrow} 
-q_1 \wedge q_2 \wedge x_1\teq0 \wedge y_1\teq0 \stackrel{(a,\tuple{1,1})}{\Longrightarrow} \\ 
+q_1 \wedge q_2 \wedge x_1\teq0 \wedge y_1\teq0 \stackrel{(a,\tuple{1,1})}{\Longrightarrow} 
 \underbrace{q_1 \wedge q_2 \wedge x_2\teq y_1+1 \wedge y_2\teq x_1+1}_{q_1} ~\wedge~
-\underbrace{q_2 \wedge x_2>x_1 \wedge y_2>y_1}_{q_2} \\ 
-\wedge~ x_1\teq0 \wedge y_1\teq0 \stackrel{(b,\tuple{2,1})}{\Longrightarrow} 
-\underbrace{q_3 \wedge x_2\geq y_2}_{q_1} \wedge \underbrace{q_4 \wedge x_2>y_2}_{q_2} \wedge x_2\teq y_1+1 \\ 
-\wedge~ y_2\teq x_1+1 \wedge \underbrace{q_4 \wedge x_2>y_2}_{q_2} \wedge x_2>x_1 \wedge y_2>y_1 \wedge x_1\teq0 \wedge y_1\teq0 ~\hfill\text{\qed}
-\end{array}\]
+\underbrace{q_2 \wedge x_2>x_1 \wedge y_2>y_1}_{q_2} 
+\wedge~ x_1\teq0 \wedge y_1\teq0 \stackrel{(b,\tuple{2,1})}{\Longrightarrow} \\
+\underbrace{q_3 \wedge x_2\geq y_2}_{q_1} \wedge \underbrace{q_4 \wedge x_2>y_2}_{q_2} \wedge x_2\teq y_1+1 
+\wedge~ y_2\teq x_1+1 \wedge \underbrace{q_4 \wedge x_2>y_2}_{q_2} \wedge x_2>x_1 \wedge y_2>y_1 \wedge x_1\teq0 \wedge y_1\teq0 ~\hfill\text{\qed}\end{array}\]}
 

### 2018-02-22 tacas/draft.tex
final

@@
 infinite alphabets, whose transition rules are formulae in a combined
-theory of booleans and some infinite data domain, that relate past and
+theory of Booleans and some infinite data domain, that relate past and
 current values of the data variables. The data theory is not fixed,
@@
 specified by a set of formulae, in a combined first-order theory of
-boolean control states and data, that relate past with present values
+Boolean control states and data, that relate past with present values
 of the variables. We do not fix the data theory a~priori, but rather
@@
 \Rightarrow \ldots \Rightarrow \phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ of
-rewritings of the initial formula by substituting boolean states with
+rewritings of the initial formula by substituting Boolean states with
 time-stamped transition rules. The word is accepted if the final
@@
 
-The boolean operations of union, intersection and complement can be
+The Boolean operations of union, intersection and complement can be
 implemented in linear time in this model, thus matching the complexity
@@
 %% formulae\begin{inparaenum}[(i)]
-%% \item contain only constants of the data sort and boolean functions of
+%% \item contain only constants of the data sort and Boolean functions of
 %%   any arity, or they
@@
 
-We have implemented the boolean operations and emptiness checking
+We have implemented the Boolean operations and emptiness checking
 semi-algorithms and carried out experiments with examples taken from
@@
 ability of SMT solvers to produce interpolants in the combined theory
-of booleans and data. Since determinisation is not needed and
+of Booleans and data. Since determinisation is not needed and
 complementation is possible in linear time, the bulk of the work is
@@
 \vec{x}$. By $\Form^+(\vec{b},\vec{x})$ we denote the set of formulae
-from $\Form(\vec{b},\vec{x})$ in which each boolean variable occurs
+from $\Form(\vec{b},\vec{x})$ in which each Boolean variable occurs
 under an even number of negations.
@@
 \Delta(q_0,a) & \equiv & q_1 \wedge q_2 \wedge x\teq0 \wedge y\teq0 \\
-\Delta(q_1,a) & \equiv & q_1 \wedge q_3 \wedge x\teq\overline{y}+1 \wedge y\teq\overline{x}+1 \\
+\Delta(q_1,a) & \equiv & q_1 \wedge q_2 \wedge x\teq\overline{y}+1 \wedge y\teq\overline{x}+1 \\
 \Delta(q_1,b) & \equiv & q_3 \wedge \overline{x} \geq \overline{y} \\
@@
 In this paper we tackle the following problems: \begin{compactenum}
-\item \emph{boolean closure}: given automata $\A_1$ and $\A_2$, both
+\item \emph{Boolean closure}: given automata $\A_1$ and $\A_2$, both
   with the same set of variables $\vec{x}$, do there exist automata
@@
 
-Given a set $Q$ of boolean variables and a set $\vec{x}$ of variables
+Given a set $Q$ of Boolean variables and a set $\vec{x}$ of variables
 of sort $\Data$, for a formula $\phi \in \Form^+(Q,\vec{x})$, with no
-negated occurrences of the boolean variables, we define the formula
+negated occurrences of the Boolean variables, we define the formula
 $\overline{\phi} \in \Form^+(Q,\vec{x})$ recursively on the structure
@@
 \len{\A_2}$ and $\len{\overline{\A}} = \len{\A}$, thus the automata
-for the boolean operations, including complementation, can be built in
+for the Boolean operations, including complementation, can be built in
 linear time. This matches the linear-time bounds for intersection and
@@
 
-Since no boolean variable from $Q$ occurs under negation in $\phi$, it
+Since no Boolean variable from $Q$ occurs under negation in $\phi$, it
 is easy to prove the following monotonicity property: given two
@@
 \booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
-boolean valuation $\beta : Q \rightarrow \booli$, every valuation $\nu
+Boolean valuation $\beta : Q \rightarrow \booli$, every valuation $\nu
 : \vec{x} \mapsto \Data^\I$ of the data variables and every finite
@@
 If $\Inv$ satisfies only the first point above, we call it an
-\emph{invariant}. Intuitively, a safety invariant maps every boolean
+\emph{invariant}. Intuitively, a safety invariant maps every Boolean
 valuation into a set of data valuations, that contains the initial
@@
 \abs{\Inv}(\beta)$ and $\I,\beta\cup\nu \models \AbsAccept{\A}(u)$,
-for some boolean $\beta : Q \rightarrow \booli$ and data $\nu :
+for some Boolean $\beta : Q \rightarrow \booli$ and data $\nu :
 \vec{x} \rightarrow \Data^\I$ valuations. Such a $u\in\Sigma^*$ is
@@
 \cite{Lyndon59}, we can assume without loss of generality that every
-boolean variable with at least one positive (negative) occurrence in
+Boolean variable with at least one positive (negative) occurrence in
 $I_i$ has at least one positive (negative) occurrence in both $\psi_i$
@@
 where $Q_k = \set{q_k \mid q \in Q}$, $k \in [0,n]$ are time-stamped
-sets of boolean variables corresponding to the set $Q$ of states of
+sets of Boolean variables corresponding to the set $Q$ of states of
 $\A$. The first conjunct $\theta_0(Q_0) \equiv \iota[Q_0/Q]$ is the
@@
 \emph{combine states with the data constraints that must hold
-  locally}, whenever the control reaches a certain boolean
+  locally}, whenever the control reaches a certain Boolean
 configuration. This association of states with data valuations is
@@
 Termination of Algorithm \ref{alg:predabs} depends on the ability of a
-given interpolating decision procedure for the combined boolean and
+given interpolating decision procedure for the combined Boolean and
 data theory $\theory(\sig,\I)$ to provide interpolants that yield a
@@
 called for the successors of $n_i$ on this path, which is handled via
-the boolean flag $b$.
+the Boolean flag $b$.
 
@@
 satisfiability queries and interpolant generation, in the theory of
-linear integer arithmetic with uninterpreted boolean functions
+linear integer arithmetic with uninterpreted Boolean functions
 (UFLIA). We compared both algorithms with a previous implementation of

### 2018-02-23 tacas/final.tex
added final

--- 2018-02-22 tacas/draft.tex
@@
 
-\appendix
-
-\subsection*{Proof of Lemma \ref{lemma:closure}}
-
-\begin{proposition}\label{prop:overline}
-  Given a formula $\phi \in \Form^+(Q,\vec{x})$ and a valuation $\nu$
-  mapping each $q \in Q$ to a value $\nu(q) \in \booli$ and each $x \in
-  \vec{x}$ to a value $\nu(x) \in \Data^\I$, let $\nu'$ be the
-  valuation that assigns each $q \in Q$ the value $\neg\nu(q)$ and
-  each $x \in \vec{x}$ the value $\nu(x)$. Then we have 
-  $\I,\nu \models \phi$ if and only if $\I,\nu' \not\models \overline{\phi}$. 
-\end{proposition}
-\proof{Immediate, by induction on the structure of $\phi$. \qed}
-
-\vspace*{\baselineskip}
-\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
-  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
-  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
-  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
-  \iota_2,a_1\ldots a_n) \iffeq \Delta(\iota_1,a_1\ldots a_n) \vee
-  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
-  definition of the initial configuration of $\A_\cup$. For the
-  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
-  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
-  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
-    \iota_2,a_1\ldots a_{n-1})}$ with
-  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
-  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
-  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
-  \iota_2,a_1\ldots a_{n-1}) \iffeq \Delta(\iota_1,a_1\ldots a_{n-1})
-  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
-  \[\begin{array}{ll}
-  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
-  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
-  \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots a_{n}) \enspace.
-  \end{array}\] 
-  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
-  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
-  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
-  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
-  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
-  \subseteq Q$ and thus $\overline{\iota_1} \equiv \iota_1$. For the case
-  $n>0$, we compute:
-  $\overline{\Delta}(\overline{\iota_1},a_1 \ldots a_n) =
-  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on $n\geq0$.
-
-  In the case $n=0$, we have $\overline{\Delta}(\overline{\iota_1},a_1
-  \ldots a_n) \equiv \overline{\iota_1}$. Then $\varepsilon$ is
-  accepted by $\A_1$ iff $\nu_0 \models \iota_1$, where $\nu_0(q) =
-  \top$ if $q \in F_1$ and $\nu_0(q) = \bot$, otherwise. But $\nu_0
-  \models \iota_1$ iff $\overline{\nu_0} \models \overline{\iota_1}$,
-  where $\overline{\nu_0}(q) = \top$ if $q \not\in F_1$ and
-  $\overline{\nu_0}(q) = \bot$, otherwise. Thus $\varepsilon$ is
-  accepted by $\A_1$ iff it is not accepted by $\overline{\A_1}$.
-
-  For the case $n>0$, we compute:
-  \[\begin{array}{lcl}
-  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \iffeq & \text{ (by ind. hyp.)} \\
-  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \iffeq & \text{ (by the def. of $\overline{\phi}$)} \\
-  \overline{\Delta(\iota_1,a_1\ldots,a_n)} \enspace.
-  \end{array}\]
-  Let $\nu,\nu' : (Q \cup \bigcup_{i=0}^n\vec{x}_i) \rightarrow (\booli
-  \cup \Data^\I)$ be valuations such that: \begin{compactitem}
-  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
-    %
-  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
-    %
-  \item $\nu(x) = \nu'(x)$, for each $x \in \vec{x}_0$, 
-    %
-  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
-    \in [1,n]$.
-  \end{compactitem}
-  By Proposition \ref{prop:overline}, we have $\I,\nu \models
-  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
-  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
-  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus for all $w \in
-  \Sigma[\vec{x}]^*$, we have $w \in L(\A_1) \iff w \not\in
-  L(\overline{\A_1})$. \qed}
-
-\subsection*{Proof of Lemma \ref{lemma:safety-invariant}}
-
-\proof{ Let $\A = \tuple{\vec{x},Q,\iota,F,\Delta}$ in the following.
-  ``$\Leftarrow$'' This direction is trivial. ``$\Rightarrow$'' We
-  define $\Inv : (Q \mapsto \booli) \rightarrow 2^{\vec{x} \mapsto
-    \Data^\I}$ as follows. For each $\beta : Q \rightarrow \booli$,
-  let $\Inv(\beta) = \{\nu : \vec{x} \rightarrow \Data^\I \mid \exists
-  u \in \Sigma^* ~.~ \beta\cup\nu \models
-  \Post{\A}(\iota,u)\}$. Checking that $\Inv$ is a safety invariant is
-  straightforward. \qed}
-
-\subsection*{Proof of Lemma \ref{lemma:abstract-invariant}}
-
-\proof{ It is sufficient to show that there exists $k\geq0$ such that for
-  all $u \in \Sigma^*$ there exists $i \in [0,k]$ such that
-  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,\mu(i))$. We have
-  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)$ for all $u \in
-  \Sigma^*$. But since $\preds$ is a finite set, also the set
-  $\{\AbsPost{\A}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
-  there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
-  exists $i \in [0,k]$ such that $\AbsPost{\A}(\iota,u) \iffeq
-  \AbsPost{\A}(\iota,\mu(i))$, which concludes the proof. \qed}
-
-%% \subsection*{Proof of Proposition \ref{prop:positive-interpol}}
-
-%% \vspace*{\baselineskip} \proof{ Let
-%%   $\tuple{\top,I_0,\ldots,I_{n},\bot}$ be an interpolant, and $i \in
-%%   [0,n]$ be the first index for which $I_i \not\in
-%%   \Form^+(Q,\vec{x})$. If $i=0$, we replace $I_0$ with $\iota \in
-%%   \Form^+(Q,\vec{x})$ and verify that the result is still an
-%%   interpolant. If $i>0$, by Proposition \ref{prop:asterisk}, we have
-%%   that $\Delta^{i-1}(I_{i-1},a_i) \iffeq \exists Q_{i-1} ~.~
-%%   I_{i-1}[Q_{i-1}/Q] \wedge \theta_i$. Since $I_{i-1} \in
-%%   \Form^+(Q,\vec{x})$, by the choice of $i$, we have that
-%%   $\Delta^{i-1}(I_{i-1},a_i) \in \Form^+(Q,\vec{x})$ as well. We
-%%   replace $I_i$ with $\Delta^{i-1}(I_{i-1},a_i)$. The result is still
-%%   an interpolant, because: \begin{compactitem}
-%%   \item $I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \models
-%%     \Delta^{i-1}(I_{i-1},a_i)$, by Proposition \ref{prop:asterisk}, and
-%%     %
-%%   \item $\Delta^{i-1}(I_{i-1},a_i) \wedge \theta_{i+1} \models \exists
-%%     Q_{i-1} ~.~ I_{i-1}[Q_{i-1}/Q] \wedge \theta_i \wedge \theta_{i+1}
-%%     \models I_i[Q_i/Q] \wedge \theta_{i+1} \models I_{i+1}[Q_{i+1}/Q]$.
-%%   \end{compactitem}
-%%   We proceed with these replacements until there are no more formulae
-%%   $I_i \not\in \Form^+(Q,\vec{x})$ left. \qed}
-
-\subsection*{Proof of Lemma \ref{lemma:progress}}
-
-\begin{proposition}\label{prop:asterisk}
-  Given a formula $\phi \in \Form^+(Q,\vec{x})$ and $a \in \Sigma$, we
-  have \(\Delta(\phi,a) \iffeq \exists Q' ~.~ \phi[Q'/Q] \wedge
-  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))\).
-\end{proposition}
-\proof{``$\Rightarrow$'' If $\I,\beta\cup\overline\nu\cup\nu \models
-  \Delta(\phi,a)$, for some valuations $\beta : Q \rightarrow \booli$
-  and $\overline\nu : \overline{\vec{x}} \rightarrow \Data^\I$, $\nu :
-  \vec{x} \rightarrow \Data^\I$, then we build a valuation $\beta' :
-  Q' \rightarrow \booli$ such that
-  $\I,\beta'\cup\beta\cup\overline\nu\cup\nu \models \phi[Q'/Q] \wedge
-  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))$. For each
-  occurrence of a formula $\Delta(q,a)$ in $\Delta(\phi,a)$ we set
-  $\beta'(q') = \true$ if $\I,\beta\cup\overline\nu\cup\nu \models
-  \Delta(q,a)$ and $\beta'(q') = \false$, otherwise. Since there are
-  no negated occurrences of such subformulae, the definition of
-  $\beta'$ is consistent, and the check
-  $\I,\beta'\cup\beta\cup\overline\nu\cup\nu \models \phi[Q'/Q] \wedge
-  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))$ is
-  immediate. $''\Leftarrow''$ This direction is an easy check. \qed}
-
-\proof{ Let $\Theta(u) \equiv \theta_0(Q_0) \wedge \theta_1(Q_0 \cup
-  Q_1,\vec{x}_0 \cup \vec{x}_1) \wedge \ldots \wedge \theta_n(Q_{n-1}
-  \cup Q_n,\vec{x}_{n-1} \cup \vec{x}_n) \wedge \theta_{n+1}(Q_n)$ in
-  the following.
-
-  (\ref{it1:progress}) We apply Proposition \ref{prop:asterisk}
-  recursively and get: \[\Post{\A}(\iota,u)[Q_n/Q,\vec{x}_n/\vec{x}]
-  \iff \exists Q_0 \ldots \exists Q_{n-1} \exists \vec{x}_0 \ldots
-  \exists \vec{x}_{n-1} ~.~ \bigwedge_{i=0}^n \theta_i\] Assuming that
-  $\Theta(u)$ is satisfiable, we obtain a model for $\Accept{\A}(u)
-  \equiv \Post{\A}(\iota,u) \wedge \theta_{n+1}[Q/Q_n]$.
-
-  (\ref{it2:progress}) If $\tuple{\top,I_0,\dots,I_n,\bot}$ is an
-  interpolant for $\Theta(u)$, the following entailments
-  hold: \begin{compactitem}
-  \item $\theta_0 \models I_0[Q_0/Q]$,
-    %
-  \item $I_{k-1}[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k}[Q_k/Q,\vec{x}_k/\vec{x}]$, $\forall k\in[1,n]$. 
-    %
-  \item $I_{n}[Q_{n}/Q] \wedge \theta_{n+1} \models \bot$. 
-  \end{compactitem}
-  We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n}$ by
-  induction on $n \geq 0$. This is sufficient to conclude because
-  $\AbsAccept{\A}(a_1\ldots a_n) \equiv \AbsPost{\A}(\iota,a_1\ldots
-  a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n} \wedge
-  \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we have
-  $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv \theta_0[Q/Q_0]
-  \models I_0$. For the induction step $n>0$, we compute:
-  \[\begin{array}{l}
-  \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] \equiv \text{ (by def. of $\AbsPost{\A}$)} \\
-  \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] \models \text{ (by Prop. \ref{prop:asterisk})} \\
-  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n \models \text{ (ind. hyp.) } \\
-  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q] \hfill\text{\qed}
-  \end{array}\]}
-
-\subsection*{Proof of Theorem \ref{thm:predabs}}
-
-\proof{We prove the following invariant: each time Algorithm
-  \ref{alg:predabs} reaches line \ref{ln:while}, the set $W$ of nodes
-  in $\worklist$ contains all the frontier nodes in the ART
-  $\tuple{N\cup W,E,\rootNode,\Lambda,\lhd}$ which are not covered by
-  some node in $N$, namely that:
-  \begin{equation}\label{inv:predabs}
-  W = \set{n \mid \forall m \in N~ \forall a \in \Sigma ~.~ (n,a,m)
-    \not\in E ~\wedge~ (n,m) \not\in \lhd}
-  \end{equation}
-  Initially, this is the case because $W = \set{\rootNode}$ and $E =
-  \lhd = \emptyset$. If the invariant holds previously, at line
-  \ref{ln:while}, it will hold again after line \ref{ln:refine} is
-  executed, because, when the subtree rooted at the pivot $p$ is
-  removed, $p$ becomes a member of the set of uncovered frontier
-  nodes, and is added to $W$ at line \ref{ln:refine}. Otherwise, the
-  invariant holds at line \ref{ln:while} and the control follows the
-  else branch at line \ref{ln:else}. In this case, the newly created
-  frontier node $s$ is added to $W$ only if it is not covered by an
-  existing node in $N$ (line \ref{ln:covered}).
-
-  Next we prove that, if Algorithm \ref{alg:predabs} returns $\true$,
-  then $\bigvee_{n \in N} \Lambda(n)$ defines a safety
-  invariant. Suppose that Algorithm \ref{alg:predabs} returns at line
-  \ref{ln:true}. Then it must be that $W = \emptyset$. Because
-  (\ref{inv:predabs}) is invariant, each node in $N$ is either covered
-  by another node in $N$, or all its successors are in $N$. We prove
-  first that $\bigvee_{n \in N} \Lambda(n)$ is an invariant: for any
-  $u \in \Sigma^*$, there exists some node $n \in N$ such that
-  $\Post{\A}(\iota,u) \models \Lambda(n)$. Let $u \in \Sigma^*$ be an
-  arbitrary sequence. If $u$ labels the path from $\rootNode$ to some
-  $n \in N$, we have $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)
-  \models \Lambda(n)$ and we are done. Otherwise, let $v$ be the
-  (possibly empty) prefix of $u$ which labels the path from
-  $\rootNode$ to some $n \in N$, which is covered by another $m \in
-  N$, where $(n,a,m) \in E$, that is $u = vav'$, for some $a \in
-  \Sigma$ and $v'\in\Sigma^*$. Moreover, we have $\Post{\A}(\iota,va)
-  \models \AbsPost{\A}(\iota,va) \models \Lambda(m)$, by the
-  construction of the set $\lhd$ of covering edges --- lines
-  \ref{ln:direct-cover}, \ref{ln:child-cover},
-  \ref{ln:indirect-cover}. Continuing this argument recursively from
-  $m$, since $\len{v'} < \len{u}$, we shall eventually discover a node
-  $p$ such that $\Post{\A}(\iota,u) \models \Lambda(p)$.
-
-  To prove that $\bigvee_{n \in N} \Lambda(n)$ is, moreover, a safety
-  invariant, suppose, by contradiction, that there exists $u \in
-  \Sigma^*$ such that $\Accept{\A}(u)$ is satisfiable. By the previous
-  point, there exists a node $p \in N$ such that $\Post{\A}(\iota,u)
-  \models \Lambda(p)$. But then we have $\Accept{\A}(\iota,u) \models
-  \Accept{\A}(\iota,\lambda(p))$, thus $\Accept{\A}(\iota,\lambda(p))$
-  is satisfiable as well. However, this cannot be the case, because
-  $p$ has been processed at line \ref{ln:feasible} and Algorithm
-  \ref{alg:predabs} would have returned a counterexample,
-  contradicting the assumption that it returns $\true$. This concludes
-  the proof that $\bigvee_{n\in N} \Lambda(n)$ is a safety invariant,
-  thus $L(\A) = \emptyset$, by Lemma \ref{lemma:safety-invariant}. We
-  have then proved the second point of the statement.
-
-  For the first point, assume that $L(\A) \neq 0$ and let $w =
-  (a_1,\nu_1) \ldots (a_k,\nu_k) \in L(\A)$ be a word. By the above,
-  Algorithm \ref{alg:predabs} cannot return $\true$. Suppose, by
-  contradiction that it does not terminate. Since the sequences from
-  $\Sigma^*$ are explored in breadth-first order, every sequence of
-  length $k$ is eventually explored, which leads to the discovery of
-  $w$ at line \ref{ln:feasible}. Then Algorithm \ref{alg:predabs}
-  terminates returning $w \in L(\A)$. \qed}
-
-\subsection*{Proof of Theorem \ref{thm:impact}}
-
-\begin{lemma}\label{lemma:well-labeled}
-  Given an ART $\Art = \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$ built
-  by Algorithm \ref{alg:impact}, $\Post{\A}(\Lambda(n),a) \models
-  \Lambda(m)$, for all $(n,a,m) \in E$.
-\end{lemma}
-\proof{ We distinguish two cases. First, if $(n,a,m)$ occurs on a path
-  in $\Art$ that has never been refined, then $\Lambda(m)=\top$ and
-  the entailment holds trivially. Otherwise, let $\Omega$ be the set
-  of paths $\omega = (n_0,a_1,n_1), \ldots, (n_{k-1},a_k,n_k)$, where
-  $n_0 = \rootNode$ and $(n,a,m) = (n_{i-1},a_i,n_i)$, for some $i \in
-  [1,k]$ and, moreover, $a_1\ldots a_k$ was found, at some point, to
-  be a spurious counterexample. Let
-  $\langle\top,I^\omega_0,\ldots,I^\omega_k,\bot\rangle$ be an
-  interpolant for $\Phi(a_1 \ldots a_k) \equiv \Lambda(\rootNode)
-  \wedge \bigwedge_{i=1}^k \theta_i \wedge \bigwedge_{q \in R(n_k)}
-  (q_k \rightarrow \bot)$, such that $I^\omega_i \in
-  \Form^+(Q,\vec{x})$, for all $i \in [0,k]$. According to Lyndon's
-  Interpolation Theorem, it is possible to build such an interpolant,
-  when $\Phi(a_1 \ldots a_k)$ is unsatisfiable. By Proposition
-  \ref{prop:asterisk}, we obtain $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q]
-  \iffeq \exists Q_{i-1} ~.~
-  I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge \theta_i$
-  and, since $I^\omega_{i-1}[Q_{i-1}/Q,\vec{x}_{i-1}/\vec{x}] \wedge
-  \theta_i \models I^\omega_i[Q_i/Q,\vec{x}_i/\vec{x}]$, we obtain
-  that $\Delta^i(I^\omega_{i-1},a_i)[Q_i/Q] \models
-  I^\omega_i[Q_i/Q,\vec{x}_i/\vec{x}]$. Since $\Lambda(n_{i-1}) =
-  \bigwedge_{\omega \in \Omega} I^\omega_{i-1}$ and $\Lambda(n_{i}) =
-  \bigwedge_{\omega \in \Omega} I^\omega_{i}$, we obtain
-  $\Post{\A}(\Lambda(n_{i-1}),a_i) \models \Lambda(n_i)$. \qed}
-
-\vspace*{\baselineskip}
-\proof{ We prove first that, each time Algorithm \ref{alg:impact}
-  reaches the line \ref{ln:impact-while}, we have:
-  \begin{equation}\label{inv:impact}
-  W = \set{n \mid n \text{ uncovered, } \exists a \in \Sigma~ \forall
-    s \in N ~.~ (n,a,s) \not\in E}
-  \end{equation}
-  Initially, $W = \set{\rootNode}$ and $E=\lhd=\emptyset$, thus
-  (\ref{inv:impact}) holds trivially. Suppose that (\ref{inv:impact})
-  holds at when reaching line \ref{ln:impact-while} and some node $n$
-  was removed from $W$ and inserted into $N$. We distinguish two
-  cases, either: \begin{itemize}
-  \item $n$ is covered, in which case $W$ becomes $W \setminus
-    \set{n}$ and (\ref{inv:impact}) holds, or
-  %
-  \item $n$ is not covered, in which case $W$ becomes $(W \setminus
-    \set{n}) \cup S$, where $S = \set{s \not\in N \mid (n,a,s) \in E,
-      a \in \Sigma}$ is the set of fresh successors of $n$. But then
-    no node $s \in S$ is covered and has successors in $E$, thus
-    (\ref{inv:impact}) holds.
-  \end{itemize}
-  Then the condition (\ref{inv:impact}) holds next time line
-  \ref{ln:impact-while} is reached, thus it is invariant.
-
-  Suppose first that Algorithm \ref{alg:impact} returns $\true$, thus
-  $W = \emptyset$ and, by (\ref{inv:impact}), for each node in $n \in
-  N$ one of the following hold: \begin{compactitem}
-  \item $n$ is covered, or
-  \item for each $a \in \Sigma$ there exists $s \in N$ such that
-    $(n,a,s) \in E$. 
-  \end{compactitem}
-  We prove that, in this case, $\bigvee_{n \in N} \Lambda(n)$ defines
-  a safety invariant and conclude that $L(\A) = \emptyset$, by Lemma
-  \ref{lemma:safety-invariant}. To this end, let $u = a_1 \ldots a_k
-  \in \Sigma^*$ be an arbitrary sequence and let $v_1$ be the largest
-  prefix of $u$ that labels a path from $\rootNode$ to some node $n_1
-  \in N$. If $v_1 = u$ we are done. Otherwise, by the choice of $v_1$,
-  it must be the case that a successor of $n_1$ is missing from
-  $(N,E)$, thus $n_1$ must be covered, by (\ref{inv:impact}) and the
-  fact that $W=\emptyset$. Let $n'_1$ be the closest ancestor of $n_1$
-  such that $(n'_1,n''_1) \in \lhd$, for some $n''_1 \in N$, and let
-  $v'_1$ be the prefix of $v_1$ leading to $n'_1$. By the construction
-  of $\lhd$ (line \ref{ln:close-uncover} in function
-  $\Call{Close}{}$), we have $\Lambda(n'_1) \models
-  \Lambda(n''_1)$. Applying Lemma \ref{lemma:well-labeled} inductively
-  on $v'_1$, we obtain that $\Post{\A}(\iota,v'_1) \models
-  \Lambda(n'_1)$, thus $\Post{\A}(\iota,v'_1) \models \Lambda(n''_1)$.
-  Continuing inductively from $n''_1$, we exhibit a sequence of
-  strings $v'_1,\ldots,v'_\ell \in \Sigma^*$ and nodes $\rootNode =
-  m_0,m_1,\ldots,m_\ell$ such that, for all $i \in
-  [1,\ell]$: \begin{compactitem}
-  \item $v'_i$ labels the path between $m_{i-1}$ and $m_i$ in $(N,E)$,    
-    %
-  \item $\Post{\A}(\iota,v'_1\ldots v'_i) \models \Lambda(m_i)$.
-  \end{compactitem}
-  Moreover, we have $u = v'_1 \ldots v'_\ell$, thus
-  $\Post{\A}(\iota,u) \models \Lambda(m_\ell)$ and we are done showing
-  that $\bigvee_{n\in N} \Lambda(n)$ is an invariant.
-
-  To prove that $\bigvee_{n \in N} \Lambda(n)$ is, moreover, a safety
-  invariant, suppose that $\Accept{\A}(u)$ is satisfiable, for some $u
-  \in \Sigma^*$ and let $n \in N$ be a node such that
-  $\Post{\A}(\iota,u) \models \Lambda(n)$. By the previous point, such
-  a node must exist. But then $\Accept{\A}(u) \models
-  \Accept{\A}(\lambda(n))$, thus $\Accept{\A}(\lambda(n))$ is
-  satisfiable, and Algorithm \ref{alg:impact} returns at line
-  \ref{ln:impact-nonempty}, upon encountering $\lambda(n)$. But this
-  contradicts the assumption that Algorithm \ref{alg:impact} returns
-  $\true$, hence we have proved that $\bigvee_{n \in N} \Lambda(n)$ is
-  a safety invariant, and $L(\A) = \emptyset$ follows, by Lemma
-  \ref{lemma:safety-invariant}. We have then proved the second point
-  of the statement.
-
-  To prove the first point, assume that $L(\A) \neq \emptyset$. By the
-  previous point, Algorithm \ref{alg:impact} does not return
-  $\true$. Suppose, by contradiction, that it does not terminate and
-  conclude using the breadth-first argument from the proof of Theorem
-  \ref{thm:predabs}. \qed}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-05-25 quantifiers/draft-old.tex
ok

--- 2017-10-20 quantifiers/draft.tex
@@
-\documentclass{llncs}
-% \documentclass[10pt,conference,letterpaper,twocolumn]{IEEEtran}
-
-\usepackage[usenames,dvipsnames]{color}
-
-\usepackage{latexsym}
-\usepackage{amsxtra} 
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{pslatex}
-\usepackage{epsfig}
-\usepackage{wrapfig}
-\usepackage{paralist}
-% \usepackage{stmaryrd}
-\usepackage{txfonts}
-\usepackage{framed}
-\usepackage{makecell}
-\usepackage{url}
-\usepackage{tikz}
-\usetikzlibrary{automata,positioning, calc}
-\usepackage[inline,shortlabels]{enumitem}
-
-\usepackage{proof}
-
-\usepackage{algorithm}
-\usepackage{algorithmicx}
-\usepackage[noend]{algpseudocode}
-
-\usepackage[draft]{commenting}
-
-\pagestyle{plain}
-
-\include{commands-lncs}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\title{First Order Alternation}
-
-\author{Radu Iosif and Xiao Xu \\
-CNRS, Verimag, Universit\'e de Grenoble Alpes \\
-Email: \{Radu.Iosif,Xiao.Xu\}@univ-grenoble-alpes.fr}
-
-\maketitle
-
-\begin{abstract}
-\end{abstract}
-
-\section{Preliminaries}
-
-For two integers $0 \leq i \leq j$, we denote by $[i,j]$ the set
-$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[1,i]$, where $[0]$ is
-the empty set.
-
-A \emph{signature} $\sig = (\ssorts{\sig},\sfuns{\sig})$ consists of a
-set $\ssorts{\sig}$ of \emph{sort symbols} and a set $\sfuns{\sig}$ of
-sorted \emph{function symbols}. To simplify the presentation, we
-assume w.l.o.g. that $\ssorts{\sig} = \set{\Data,\Bool}$\footnote{The
-  generalization to more than two sorts is without difficulty, but
-  would unnecessarily clutter the technical presentation.} and each
-function symbol $f \in \sfuns{\sig}$ has $\#(f) \geq 0$ arguments of
-sort $\Data$ and return value $f_\sigma \in \ssorts{\sig}$. If
-$\#(f)=0$ then $f$ is a \emph{constant}. Let $\top$ and $\bot$ be
-constants of sort $\Bool$.
-
-%% and $\mathbf{0}$ be a constant of sort $\Data$.
-
-We consider an infinite countable set of \emph{variables} $\vars$,
-where each $x \in \vars$ has an associated sort $x_\sigma$. A
-\emph{term} $t$ of sort $t_\sigma = S$ is a variable $x \in \vars$
-where $x_\sigma=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
-$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $f_\sigma=S$. An
-\emph{atom} is a term $p(t_1,\ldots,t_{\#(p)})$ of sort $\Bool$, or an
-equality $t \teq s$ between two terms of sort $\Data$. A
-\emph{formula} is a possibly quantified boolean combination of atoms
-and a \emph{sentence} is a formula in which each variable occurs under
-the scope of a quantifier.
-
-We denote by $\fv{\sigma}{\phi}$ the set of free variables of sort
-$\sigma$ in $\phi$ and write $\fv{}{\phi}$ for $\bigcup_{\sigma \in
-  \ssorts{\sig}} \fv{\sigma}{\phi}$. For a variable $x \in
-\fv{}{\phi}$ and a term $t$ such that $t_\sigma = x_\sigma$, let
-$\phi[t/x]$ be the result of replacing each occurrence of $x$ by
-$t$. For indexed sets $T=\set{t_1,\ldots,t_n}$ and
-$X=\set{x_1,\ldots,x_n}$, we write $\phi[T/X]$ for the formula
-obtained by simultaneously replacing $x_i$ with $t_i$ in $\phi$, for
-all $i\in[n]$. The size $\len{\phi}$ is the number of symbols occuring
-in $\phi$.
-
-An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
-\item the sort $\Data$ into a non-empty set $\Data^\I$, 
-%
-\item the sort $\Bool$ into the set $\booli = \set{\true,\false}$, where
-  $\top^\I = \true$, $\bot^\I = \false$, and
-%
-\item each function symbol $f$ into a total function $f^\I :
-  (\Data^\I)^{\#(f)} \rightarrow f_\sigma^I$, or an element of
-  $f_\sigma^I$ when $\#(f)=0$.
-\end{inparaenum}
-Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
-variable $x \in \vars$ into an element of $x_\sigma^\I$. For a term
-$t$, we denote by $t^\I_\nu$ the value obtained by replacing each
-function symbol $f$ by its interpretation $f^\I$ and each variable $x$
-by its valuation $\nu(x)$. For a formula $\phi$, we write $\I,\nu
-\models \phi$ if the formula obtained by replacing each term $t$ in
-$\phi$ by the value $t^\I_\nu$ is logically equivalent to true.  The
-semantics of quantifiers is given as usual: $\I,\nu \models \exists x
-~.~ \phi$ if and only if $\I,\nu[x \leftarrow u] \models \phi$, for
-some value $u \in x_\sigma^\I$, where $\nu[x \leftarrow u]$ is the
-valuation that assigns $u$ to $x$ and $\nu(y)$ to all $y \neq x$.
-
-A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
-there exists a valuation $\nu$ such that $\I,\nu \models \phi$, and
-\emph{valid} if $\I,\nu \models \phi$ for all valuations $\nu$.  The
-\emph{first-order theory} $\theory(\sig,\I)$ is the set of valid
-formulae written in the signature $\sig$, with the interpretation
-$\I$. A \emph{decision procedure} for $\theory(\sig,\I)$ is an
-algorithm that takes a formula $\phi$ in the signature $\sig$ and
-returns yes if and only if $\phi \in \theory(\sig,\I)$.
-
-Given formulae $\varphi$ and $\psi$, we say that \emph{$\phi$ entails
-  $\psi$ modulo $\theory(\sig,\I)$}, denoted $\phi
-\models_{\theory(\sig,\I)} \psi$ iff $\varphi \rightarrow \psi \in
-\theory(\sig,\I)$. We say that $\varphi$ and $\psi$ are
-\emph{$\theory(\sig,\I)$-equivalent}, denoted $\phi \iffeq_{\theory(\sig,\I)} \psi$ iff
-$\phi \models_{\theory(\sig,\I)} \psi$ and $\psi
-\models_{\theory(\sig,\I)} \phi$. The theory $\theory(\sig,\I)$ has
-\emph{quantifier elimination} iff for any formula $\exists x ~.~
-\phi(x)$, where $\phi$ is quantifier-free, there exists a quantifier
-free formula $\psi$ such that $\phi$ and $\psi$ are
-$\theory(\sig,\I)$-equivalent.
-
-\section{First Order Alternating Data Automata}
-
-We fix a theory $\T = \theory(\sig,\I)$ and a finite alphabet $\Sigma$
-of \emph{input events}. Given a finite set $X \subset \vars$ of
-variables of sort $\Data$, let $X \mapsto \Data^\I$ be the set of
-valuations of the variables $X$ and $\Sigma[X] = \Sigma \times (X
-\mapsto \Data^\I)$ be the possibly infinite set of \emph{data
-  symbols}. A \emph{data word} (word in the sequel) is a finite
-sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data symbols,
-where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n : X
-\rightarrow \Data^\I$ are valuations. We denote by $\varepsilon$ the
-empty sequence, by $\Sigma^*$ the set of finite sequences of input
-events and by $\Sigma[X]^*$ the set of data words over the variables
-$X$.
-
 Given sets of variables $B,X,Y \subset \vars$ of sorts $\Bool$,
@@
 the path up to the $i$-th step.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{abbrv} \bibliography{refs}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\end{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-05-25 quantifiers/draft.tex
ok

--- 2017-08-16 quantifiers/draft.tex
@@
 
+\declareauthor{ri}{Radu}{blue}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\author{Radu Iosif and Xiao Xu \\
+\author{Radu Iosif and Xiao Xu}
+\institute{
   CNRS, Verimag, Universit\'e de Grenoble Alpes \\
-Email: \{Radu.Iosif,Xiao.Xu\}@univ-grenoble-alpes.fr}
+  Email: \{Radu.Iosif,Xiao.Xu\}@univ-grenoble-alpes.fr
+}
 
@@
 function symbol $f \in \sfuns{\sig}$ has $\#(f) \geq 0$ arguments of
-sort $\Data$ and return value $f_\sigma \in \ssorts{\sig}$. If
-$\#(f)=0$ then $f$ is a \emph{constant}. Let $\top$ and $\bot$ be
-constants of sort $\Bool$ and $\mathbf{0}$ be a constant of sort
-$\Data$.
+sort $\Data$ and return value of sort $\sort{f}$, which can be either
+$\Bool$ or $\Data$. If $\#(f)=0$ then $f$ is a \emph{constant} and we
+denote by $\top$ and $\bot$ the only constants of sort $\Bool$,
+representing the true and false values, respectively. We write $\preds
+= \set{f \in \sfuns{\sig} \mid \sort{f} = \Bool}$ for the set of
+boolean function symbols, also called \emph{predicates} in the
+following.
 
 We consider an infinite countable set of \emph{variables} $\vars$,
-where each $x \in \vars$ has an associated sort $x_\sigma$. A
-\emph{term} $t$ of sort $t_\sigma = S$ is a variable $x \in \vars$
-where $x_\sigma=S$, or $f(t_1,\ldots,t_{\#(f)})$ where
-$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and $f_\sigma=S$. An
-\emph{atom} is a term $p(t_1,\ldots,t_{\#(p)})$ of sort $\Bool$, or an
-equality $t \teq s$ between two terms of sort $\Data$. A
-\emph{formula} is a possibly quantified boolean combination of atoms.
+where each $x \in \vars$ has an associated sort, denoted $\sort{x}$. A
+\emph{term} $t$ of sort $\sort{t}$ is either a variable $x \in \vars$
+of sort $\sort{t}$, or $f(t_1,\ldots,t_{\#(f)})$ where
+$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and
+$\sort{f}=\sort{t}$. An \emph{atom} is a term
+$p(t_1,\ldots,t_{\#(p)})$ of sort $\Bool$, or an equality $t \teq s$
+between two terms of sort $\Data$. A \emph{formula} is a possibly
+quantified boolean combination of atoms and a \emph{sentence} is a
+formula in which each variable occurs under the scope of a quantifier.
+A formula is \emph{positive} if and only if each predicate occurs
+under an even number of negations.
 
@@
   \ssorts{\sig}} \fv{\sigma}{\phi}$. For a variable $x \in
-\fv{}{\phi}$ and a term $t$ such that $t_\sigma = x_\sigma$, let
+\fv{}{\phi}$ and a term $t$ such that $\sort{t} = \sort{x}$, let
 $\phi[t/x]$ be the result of replacing each occurrence of $x$ by
 $t$. For indexed sets $T=\set{t_1,\ldots,t_n}$ and
-$X=\set{x_1,\ldots,x_n}$, we write $\phi[T/X]$ for the formula
-obtained by simultaneously replacing $x_i$ with $t_i$ in $\phi$, for
-all $i\in[n]$. The size $\len{\phi}$ is the number of symbols
-occuring in $\phi$.
+$X=\set{x_1,\ldots,x_n}$, where $\sort{t_i} = \sort{x_i}$, we write
+$\phi[T/X]$ for the formula obtained by simultaneously replacing $x_i$
+with $t_i$ in $\phi$, for all $i\in[n]$. The size $\len{\phi}$ is the
+number of symbols occuring in $\phi$.
 
-An \emph{interpretation} $\I$ maps\begin{inparaenum}[(1)]
+An \emph{interpretation} $\I$ maps \begin{inparaenum}[(1)]
 \item the sort $\Data$ into a non-empty set $\Data^\I$, 
@@
 \item each function symbol $f$ into a total function $f^\I :
-  (\Data^\I)^{\#(f)} \rightarrow f_\sigma^I$, or an element of
-  $f_\sigma^I$ when $\#(f)=0$.
+  \left(\Data^\I\right)^{\#(f)} \rightarrow \sort{f}^I$, or an element
+  of $\sort{f}^I$ when $\#(f)=0$.
 \end{inparaenum}
@@
 ~.~ \phi$ if and only if $\I,\nu[x \leftarrow u] \models \phi$, for
-some value $u \in x_\sigma^\I$, where $\nu[x \leftarrow u]$ is the
+some value $u \in \sort{x}^\I$, where $\nu[x \leftarrow u]$ is the
 valuation that assigns $u$ to $x$ and $\nu(y)$ to all $y \neq x$.
@@
 Given formulae $\varphi$ and $\psi$, we say that \emph{$\phi$ entails
-  $\psi$}, denoted $\phi \models^\I \psi$ iff $\I,\nu \models \varphi$
-implies $\I,\nu \models \psi$, for each valuation $\nu$, and $\phi
-\iffeq^\I \psi$ iff $\phi \models^\I \psi$ and $\psi \models^\I \phi$.
-We omit mentioning the interpretation $\I$ when it is clear from the
-context.
+  $\psi$ modulo $\theory(\sig,\I)$}, denoted $\phi
+\models_{\theory(\sig,\I)} \psi$ iff $\varphi \rightarrow \psi \in
+\theory(\sig,\I)$. We say that $\varphi$ and $\psi$ are
+\emph{$\theory(\sig,\I)$-equivalent}, denoted $\phi
+\equiv_{\theory(\sig,\I)} \psi$ iff $\phi \models_{\theory(\sig,\I)}
+\psi$ and $\psi \models_{\theory(\sig,\I)} \phi$. The theory
+$\theory(\sig,\I)$ has \emph{quantifier elimination} iff for any
+formula $\exists x ~.~ \phi(x)$, where $\phi$ is quantifier-free,
+there exists a quantifier free formula $\psi$ such that $\phi$ and
+$\psi$ are $\theory(\sig,\I)$-equivalent.
 
@@
 
-We fix an interpretation $\I$ and a finite alphabet $\Sigma$ of
-\emph{input events}. Given a finite set $X \subset \vars$ of
-variables of sort $\Data$, let $X \mapsto \Data^\I$ be the set
-of valuations of the variables $X$ and $\Sigma[X] = \Sigma
-\times (X \mapsto \Data^\I)$ be the possibly infinite set of
-\emph{data symbols}. A \emph{data word} (word in the sequel) is a
-finite sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data
-symbols, where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n :
-X \rightarrow \Data^\I$ are valuations. We denote by
-$\varepsilon$ the empty sequence, by $\Sigma^*$ the set of finite
-sequences of input events and by $\Sigma[X]^*$ the set of data
+In the rest of the paper, we we fix a theory $\theory(\sig,\I)$ and
+omit mentioning it in the context of entailments and equivalences. Let
+$\Sigma$ be a finite alphabet $\Sigma$ of \emph{input events}. Given a
+finite set $X \subset \vars$ of variables of sort $\Data$, we denote
+by $X \mapsto \Data^\I$ the set of valuations of the variables $X$ and
+$\Sigma[X] = \Sigma \times (X \mapsto \Data^\I)$ be the possibly
+infinite set of \emph{data symbols}. A \emph{data word} (word in the
+sequel) is a finite sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots
+(a_n,\nu_n)$ of data symbols, where $a_1,\ldots,a_n \in \Sigma$ and
+$\nu_1,\ldots,\nu_n : X \rightarrow \Data^\I$ are valuations. We
+denote by $\varepsilon$ the empty sequence, by $\Sigma^*$ the set of
+finite sequences of input events and by $\Sigma[X]^*$ the set of data
 words over the variables $X$.
 
-Given sets of variables $B,X,Y \subset \vars$ of sorts $\Bool$,
-$\Data$ and $\Data$, respectively, we denote by $\Form\tuple{B;X;Y}$
-the set of formulae $\phi$ such that $\fv{\Bool}{\phi} = B$,
-$\fv{\Data}{\phi} = X$ and each variable from $Y$ occurs under the
-scope of a quantifier. By $\Form^+\tuple{B;X;Y}$ we denote the subset
-of $\Form\tuple{B;X;Y}$ in which each boolean variable from $B$ occurs
-under an even number of negations. We denote by $\phi\tuple{B;X;Y}$
-any formula in $\Form\tuple{B;X;Y}$.
+A \emph{first-order alternating data automaton} (automaton in the
+sequel) is a tuple $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, where
+$\Sigma$ is the finite event alphabet, $X$ is a
+finite set of variables, $Q \subseteq \preds$ is a finite set of
+\emph{state predicates}, $\iota$ is a sentence, $F \subseteq Q$ is a
+set of \emph{final states} and $\Delta$ is a set of \emph{transition
+  rules} of the form:
+\[q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\]
+where $q \in Q$, $a \in \Sigma$, $\psi$ is a positive formula such
+that $\fv{}{\phi} \subseteq X \cup \set{y_1,\ldots,y_{\#(q)}}$, $X
+\cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$ and each predicate that
+occurs in $\psi$ belongs also to $Q$. We stress the fact that each
+predicate occurs in $\psi$ under an even number of negations.
 
-A \emph{first-order alternating data automaton} (fo-ADA or automaton
-in the sequel) is a tuple $\A =
-\tuple{G,L,Q,\iota,F,\Delta}$, where: \begin{compactitem}
-%
-\item $G \subseteq \vars$ is a finite set of \emph{global
-  variables} of sort $\Data$, 
-%
-\item $L \subseteq \vars$ is a finite set of \emph{local
-  variables} of sort $\Data$,
-%
-\item $Q \subset \vars$ is a finite set of variables of sort $\Bool$
-  (\emph{states}),
-%
-\item $\iota \in \Form^+\tuple{Q;\emptyset;\emptyset}$ is the \emph{initial
-  configuration},
-%
-\item $\Delta : Q \times \Sigma \rightarrow
-  \Form^+\tuple{Q;\overline{G}\cup G\cup\overline{L};L}$ is a
-  \emph{transition function},
-\end{compactitem}
-where, for a set $X \subseteq \vars$, we denote $\overline{X} =
-\set{\overline{x} \mid x \in X}$. The \emph{size} of $\A$ is defined
-as $\len{\A} = \len{\iota} + \sum_{(q,a) \in
-  Q\times\Sigma}\len{\Delta(q,a)}$.
+Intuitively, $a$ is the input event and $X$ are the input data values
+that trigger the transition, whereas $q$ and $y_1,\ldots,y_{\#(q)}$
+are the current control state and data values pertaining to that
+state, respectively. Without loss of generality, we consider, for each
+predicate $q \in Q$ and each input event $a \in \Sigma$, at most one
+such rule (two or more rules can be joined using disjunction). The
+\emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
+\sum_{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta} \len{\psi}$.
 
-Intuitively, the variables $\overline{x}$ from a transition formula
-$\Delta(q,a)$ refer to past values, whereas the other variables refer
-to the current values. Observe that the initial values of the global
-variables are unconstrained (since the initial configuration is a
-boolean combination of states without variables) and quantification
-occurs only over the current copy of the local variables $L$. As a
-remark, the transition relation does not quantify over global
-variables $G \cup \overline{G}$, which must be visible throughout the
-execution. Furthermore, we assume that the past values of the local
-variables are coined and beyond our control, thus quantification over
-$\overline{L}$ is not allowed either.
+Given a predicate $q \in Q$ and a tuple of data values
+$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
+a \emph{configuration}\footnote{Note that a configuration is not a
+  logical term since data values cannot be written in logic.}.
+Because we consider an alternating model of computation, an automaton
+can find itself, at the same time, in more than one
+configuration. Typically, in the classical literature on alternating
+machines, this set is finite. In our case, due to the presence of
+first-order quantifiers ranging over an infinite data domain, we
+consider that an automaton can be also in an infinite set of
+configurations, all at once.
 
-In order to define the semantics of automata, we consider time-stamped
-sets of variables $\stamp{X}{n} = \{\stamp{x}{n} \mid x \in X\}$ for
-any $n\geq0$. For an input event $a \in \Sigma$, some constant
-$n\geq0$ and a formula $\phi$, we write $\stamp{\Delta}{n}(\phi,a)$
-for the formula obtained from $\phi$ by simultaneously replacing each
-state $q \in \fv{\Bool}{\phi}$ with
-$\Delta(\phi,a)[\stamp{G}{n}/\overline{G}, \stamp{G}{n+1}/G,
-  \stamp{L}{n}/\overline{L}, \stamp{L}{n+1}/L]$.
+To simplify the presentation, at this point, we shall give the
+execution semantics of automata in terms of symbolic rewriting rules.
+The connection between this symbolic semantics and the concrete
+semantics, expressed in terms of sets of configurations, will be given
+at a later time. Let $\stamp{X}{i} = \set{\stamp{x}{i} \mid x \in X}$
+be a set of \emph{time-stamped} variables. For a formula $\psi$ and $i
+\in \nat$, we define $\stamp{\psi}{i} = \psi[X/\stamp{X}{i}]$,
+i.e.\ the formula in which each variable $x \in X$, denoting an input
+data element, is replaced by $\stamp{x}{i}$. Further, given a formula
+$\phi$, let $\Accept{F}{\phi}$ be the formula obtained from $\phi$ by
+replacing each predicate atom $q(z_1,\ldots,z_{\#(q)})$ by $\top$ if
+$q \in F$, and by $\bot$, otherwise.
 
-Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in
-\Sigma[G]^*$ the \emph{run} of $\A$ over $w$ is the sequence of
-formulae: \[\iota \stackrel{(a_1,\nu_1)}{\Longrightarrow} \phi_1
-\tuple{Q,\stamp{G}{1}, \stamp{L}{1}}
-\stackrel{(a_2,\nu_2)}{\Longrightarrow} \ldots
-\stackrel{(a_n,\nu_n)}{\Longrightarrow} \phi_n\tuple{Q,\stamp{G}{1}
-  \cup \ldots \cup \stamp{G}{n}, \stamp{L}{1} \cup \ldots \cup
-  \stamp{L}{n}}\] where $\phi_1 \equiv
-\stamp{\Delta}{0}(\iota,a_1)[\mathbf{0}/\stamp{G}{0},\mathbf{0}/\stamp{L}{0}]$
-and $\phi_{i+1} \equiv \stamp{\Delta}{i}(\phi_i,a_{i+1})$, for all $i
-\in [1,n-1]$. In other words, we assume that all data variables are
-initially $\mathbf{0}$.
+For an input word $(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$,
+we define the sequence of formulae \(\phi_0 \Rightarrow \phi_1
+\Rightarrow \ldots \Rightarrow \phi_n\), where $\phi_0 = \iota$ and
+$\phi_{i+1}$ is obtained from $\phi_i$ by replacing each predicate
+atom $q(z_1,\ldots,z_{\#(q)})$ from $\phi_i$ by the formula
+$\stamp{\psi}{i+1}[z_1, \ldots, z_{\#(q)}/y_1,\ldots,y_{\#(q)}]$,
+where $q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta$ is
+the only transition rule that applies, or by $\bot$, if no transition
+rule applies. Observe that $\phi_n$ is uniquely defined by the event
+sequence $a_1 \ldots a_n \in \Sigma^*$. The word $(a_1,\nu_1) \ldots
+(a_n,\nu_n)$ is \emph{accepted} by $\A$ if and only if $\I,\nu \models
+\Accept{F}{\phi_n}$, where the valuation $\nu : \bigcup_{i=1}^n
+\stamp{X}{i} \rightarrow \Data^\I$ is defined as $\nu(\stamp{x}{i}) =
+\nu_i(x)$, for each $x \in X$ and $i \in [n]$.
 
-We say that $\A$ \emph{accepts} $w$ if there exists a valuation $\nu$
-such that $\nu(\stamp{x}{i}) = \nu_i(x)$ for each global variable $x
-\in G$ and each timestamp $i \in [n]$, $\nu(q) = \true$ for each $q
-\in F$, $\nu(q) = \false$ for each $q \in Q \setminus F$ and $\I,\nu
-\models \phi_n$. The \emph{language} of $\A$ is the set $L(\A)$ of
-words from $\Sigma[G]^*$ accepted by $\A$. In this paper, we ask the
-following two questions: \begin{compactenum}
-\item \emph{boolean closure}: given automata $\A_i =
-  \tuple{G,L_i,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, with $L_1 \cap
-  L_2 = \emptyset$ and $Q_1 \cap Q_2 = \emptyset$, do there exist
-  automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
+As usual, the \emph{language} of $\A$, denoted $\lang{\A}$, is the set
+of words accepted by $\A$. In this paper, we ask the following
+questions: \begin{compactenum}
+\item \emph{closure}: given automata $\A_i =
+  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
+  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
   $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
-  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[G]^* \setminus L(\A_1)$ ?
+  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
 %
 \item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
-  \emptyset$ ?
+  \emptyset$?
 \end{compactenum}
 
-\subsection{Boolean Closure}
+\subsection{Closure Properties}
 
-Given a set $B$ of boolean variables, two disjoint sets $X$ and $Y$ of
-data variables, for any formula $\phi \in \Form^+\tuple{B;X;Y}$,
-define $\overline{\phi} \in \Form^+\tuple{B;X;Y}$ recursively:
+Given a positive formula $\phi$, we define the \emph{dual} formula
+$\dual{\phi}$ recursively on the structure of $\phi$ as follows: 
 \[\begin{array}{rclcrclcrcl}
-\overline{\phi_1 \vee \phi_2} & \equiv & \overline{\phi_1} \wedge \overline{\phi_2} & \hspace*{5mm} & 
-\overline{\phi_1 \wedge \phi_2} & \equiv & \overline{\phi_1} \vee \overline{\phi_2} & \hspace*{5mm} & 
-\overline{\phi} & \equiv & \neg\phi \text{ if $\phi \not\in B$ is an atom} \\ 
-\overline{\exists x ~.~ \phi_1} & \equiv & \forall x ~.~ \overline{\phi_1} && 
-\overline{\forall x ~.~ \phi_1} & \equiv & \exists x ~.~ \overline{\phi_1} && 
-\overline{\neg\phi} & \equiv & \neg \overline{\phi} \text{ if $\phi$ is not an atom} \\
-&& && && && \overline{\phi} & \equiv & \phi \text{ if $\phi \in B$}
+\dual{(\phi_1 \vee \phi_2)} & = & \dual{\phi_1} \wedge \dual{\phi_2} && 
+\dual{(\phi_1 \wedge \phi_2)} & = & \dual{\phi_1} \wedge \dual{\phi_2} &&
+\dual{(t \teq s)} & = & \neg(t \teq s) \\
+\dual{(\exists x ~.~ \phi_1)} & = & \forall x ~.~ \dual{\phi_1} && 
+\dual{(\forall x ~.~ \phi_1)} & = & \exists x ~.~ \dual{\phi_1} && 
+\dual{(\neg(t \teq s))} & = & t \teq s \\
+\dual{(q(x_1,\ldots,x_{\#(q)}))} & = & q(x_1,\ldots,x_{\#(q)})
 \end{array}\]
-Clearly $\phi$ and $\overline{\phi}$ have the same size. The relation
-between $\phi$ and $\overline{\phi}$ is formalized below:
+Observe that, because predicate atoms do not occur negated in $\phi$,
+there is no need to define dualization for formulae of the form $\neg
+q(x_1,\ldots,x_{\#(q)})$. The dualization operation has the following
+property:
+\begin{lemma}\label{lemma:dual}
+  Given a formula $\phi$ and an automaton
+  $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$, we have the equivalence
+  $\Accept{F}{\phi} \equiv \neg(\dual{\Accept{Q\setminus F}{\phi}})$. 
+\end{lemma}
+\proof{ By straightforward induction on the structure of $\phi$. \qed}
 
-\begin{proposition}\label{prop:complement}
-  Given sets of variables $B,X,Y \subset \vars$ of sorts boolean, data
-  and data respectively, and a formula $\phi \in
-  \Form^+\tuple{B;X;Y}$, for any valuation $\nu$ we have $\nu \models
-  \phi$ if and only if $\overline{\nu} \models \neg\overline{\phi}$,
-  where $\overline{\nu}(x) = \nu(x)$ for all $x \in X$ and
-  $\overline{\nu}(b) = \neg\nu(b)$ for all $b \in B$.
-\end{proposition}
-\proof{By induction on the structure of $\phi$. \qed}
+The following theorem shows that first-order alternating data automata
+are closed under all boolean operations:
 
-Consider the automata $\A_i = \tuple{G,L_i,Q_i,\iota_i,F_i,\Delta_i}$,
-for $i=1,2$, where $L_1 \cap L_2 = \emptyset$ and $Q_1 \cap Q_2 =
-\emptyset$. We define the following automata: 
-\[\begin{array}{rcl}
-\A_\cup & = & \tuple{G,~ L_1 \cup L_2,~ Q_1 \cup Q_2,~ \iota_1 \vee \iota_2,~ F_1 \cup F_2,~ \Delta_1 \cup \Delta_2} \\
-\A_\cap & = & \tuple{G,~ L_1 \cup L_2,~ Q_1 \cup Q_2,~ \iota_1 \wedge \iota_2,~ F_1 \cup F_2,~ \Delta_1 \cup \Delta_2} \\
-\overline{A_1} & = & \tuple{G,~ L_1,~ Q_1,~ \overline{\iota_1},~ Q_1 \setminus F_1,~ \lambda q \in Q_1~ \lambda a \in \Sigma ~.~ \overline{\Delta_1(q,a)}}
-\end{array}\]
-
-\begin{lemma}\label{lemma:boolean-closure}
-  Given automata $\A_i = \tuple{G,L_i,Q_i,\iota_i,F_i,\Delta_i}$, for
-  $i=1,2$, where $L_1 \cap L_2 = \emptyset$ and $Q_1 \cap Q_2 =
-  \emptyset$, we have $L(\A_\cup) = L(\A_1) \cup L(\A_2)$, $L(A_\cap)
-  = L(\A_1) \cap L(\A_2)$ and $L(\overline{\A_1}) = \Sigma[G]*
-  \setminus L(\A_1)$. 
-\end{lemma}
-\proof{
-  Same as \cite[Lemma 1]{IosifX17}. 
-\qed}
-
-\section{Checking Emptiness with Interpolants and Skolem Witnesses}
-
-Since the emptiness problem is undecidable even for automata with only
-two global variables and no local variables, by reduction from the
-emptiness problem for 2-counter machines \cite{Minsky67}, we give a
-semi-algorithm with the following properties. Given an automaton $\A$
-as input, the procedure will\begin{inparaenum}[(1)] 
-\item return a word $w \in L(\A)$ if $L(\A)
-\neq \emptyset$, and 
+\begin{theorem}\label{thm:closure}
+  Given automata $\A_i = \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$,
+  for $i=1,2$, such that $Q_1 \cap Q_2 = \emptyset$, the following hold: 
+  \begin{compactitem}
+    \item\label{it1:closure} $\lang{\A_\cap} = \lang{\A_1} \cap \lang{\A_2}$, where
+      $\A_\cap = \tuple{\Sigma,X,Q_1 \cup Q_2, \iota_1 \wedge \iota_2,
+      F_1 \cup F_2, \Delta_1 \cup \Delta_2}$, 
       %
-\item if the procedure terminates without returning such a word, then
-  we can safely conclude that $L(\A) = \emptyset$. \end{inparaenum}
-Due to the undecidability of the emptiness problem, it is possible for
-the semi-algorithm to run forever in the case when the language of the
-automaton is empty.
-
-\begin{proposition}\label{prop:substitution}
-  Given a formula $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$, where $Q_1 \ldots
-  Q_n \in \set{\exists,\forall}$ and $\phi$ is a quantifier-free
-  formula with free variables $\fv{}{\phi} = \set{x_1,\ldots,x_n,q}$,
-  for any formula $\psi$, the following equivalence holds:
-  \[Q_1 x_1 \ldots Q_n x_n ~.~ \phi[\psi/q] \iff 
-  Q_1 x_1 \ldots Q_n x_n \exists q ~.~ \phi \wedge (q \rightarrow \psi)\]
-\end{proposition}
-\proof{
-
-\qed}
+    \item\label{it2:closure} $\lang{\overline{\A_i}} = \Sigma[X]^* \setminus
+      \lang{\A_i}$, where $\overline{\A_i} =
+      \tuple{\Sigma,X,Q_i,\dual{\iota},Q_i\setminus
+        F_i,\overline{\Delta_i}}$ and, for all
+      $i=1,2$: \[q(x_1,\ldots,x_{\#(q)}) \arrow{a(X)}{} \phi \in
+      \Delta_i \iff q(x_1,\ldots,x_{\#(q)}) \arrow{a(X)}{} \dual{\phi}
+      \in \overline{\Delta_i} \enspace.\]
+  \end{compactitem}
+  Moreover, $\len{\A_\cap} = \bigO{\len{\A_1}+\len{\A_2}}$ and
+  $\len{\overline{\A_i}} = \bigO{\len{\A_i}}$, for all $i=1,2$.
+\end{theorem}
+\proof{\qed}
 

### 2018-05-25 quantifiers/draft.tex
ok

@@
 
+\subsection{Configuration-based Semantics}
+
+We give a low-level operational semantics of automata, in terms of
+sets of configurations, with a conjunctive meaning. A (possibly
+infinite) set $C$ of configurations is also called a
+\emph{cube}. Intuitively, an automaton $\A$ is in a cube $C$ if it is
+simultaneously in all configurations $q(d_1, \ldots, d_{\#(q)})$ from
+$C$. A positive sentence $\phi$ containing at least one occurrence of
+a predicate, defines a (possibly infinite) set of cubes $\sem{\phi}$,
+defined next.
+
+A \emph{context} $q[x_1,\ldots,x_k] = q(\delta_1, \ldots,
+\delta_{\#(q)})$ is a configuration in which one or more variables may
+occur, i.e. $\delta_i \in \Data^\I \cup \set{x_1,\ldots,x_k}$, for all
+$i \in [\#(q)]$. For a set of data values $D \subseteq \Data^\I$, we
+let $q[x_1,\ldots,x_{i-1},D,x_{i+1}, \ldots, x_k] =
+\{q[x_1,\ldots,x_{i-1},d,x_{i+1},\ldots,x_k] \mid d \in D\}$ be a set
+of contexts. Observe that $q[D]$ is a cube, when $k=1$.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-06-05 quantifiers/draft.tex
ok

@@
 $\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[1,i]$, where $[0]$ is
-the empty set.
+the empty set. To simplify the presentation, we consider two sorts
+$\Data$ and $\Bool$, where $\Data$ ranges over an infinite domain and
+$\Bool$ has the usual interpretation. The $\Data$ sort is equipped
+with a finite set of function symbols $f : \Data^{\#(f)} \rightarrow
+\Data$, where $\#(f)\geq0$ denotes the number of arguments (arity) of
+$f$. When $\#(f)=0$, we say that $f$ is a constant, and denote by
+$\top$ and $\bot$ the only constants of sort $\Bool$, representing the
+true and false values, respectively.  For conciseness, in the
+following we blur the distinction between the sort and function
+symbols and their interpretations. A \emph{predicate} is a function
+symbol $p : \Data^{\#(p)} \rightarrow \Bool$, denoting a relation of
+arity $\#(p)$.
 
-A \emph{signature} $\sig = (\ssorts{\sig},\sfuns{\sig})$ consists of a
-set $\ssorts{\sig}$ of \emph{sort symbols} and a set $\sfuns{\sig}$ of
-sorted \emph{function symbols}. To simplify the presentation, we
-assume w.l.o.g. that $\ssorts{\sig} = \set{\Data,\Bool}$\footnote{The
-  generalization to more than two sorts is without difficulty, but
-  would unnecessarily clutter the technical presentation.} and each
-function symbol $f \in \sfuns{\sig}$ has $\#(f) \geq 0$ arguments of
-sort $\Data$ and return value of sort $\sort{f}$, which can be either
-$\Bool$ or $\Data$. If $\#(f)=0$ then $f$ is a \emph{constant} and we
-denote by $\top$ and $\bot$ the only constants of sort $\Bool$,
-representing the true and false values, respectively. We write $\preds
-= \set{f \in \sfuns{\sig} \mid \sort{f} = \Bool}$ for the set of
-boolean function symbols, also called \emph{predicates} in the
-following.
+Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
+variables, ranging over the interpretation of the $\Data$ sort. Terms
+are either constants of sort $\Data$, variables or function
+applications. The set of \emph{positive} formulae is defined by the
+syntax below: 
+\[\phi := t \teq s \mid t \not\teq s \mid p(t_1,\ldots,t_{\#(p)}) 
+\mid \phi_1 \wedge \phi_2 \mid \phi_1 \vee \phi_2 \mid \exists x ~.~
+\phi_1 \mid \forall x ~.~ \phi_1\] where $t,s,t_1,\ldots,t_{\#(p)}$
+denote terms. We denote by $\fv{}{\phi}$ the set of free variables in
+$\phi$. A \emph{sentence} is a formula $\phi$ in which each variable
+occurs under the scope of a quantifier, i.e. $\fv{}{\phi} =
+\emptyset$. The size $\len{\phi}$ is the number of symbols occuring in
+$\phi$.
 
-We consider an infinite countable set of \emph{variables} $\vars$,
-where each $x \in \vars$ has an associated sort, denoted $\sort{x}$. A
-\emph{term} $t$ of sort $\sort{t}$ is either a variable $x \in \vars$
-of sort $\sort{t}$, or $f(t_1,\ldots,t_{\#(f)})$ where
-$t_1,\ldots,t_{\#(f)}$ are terms of sort $\Data$ and
-$\sort{f}=\sort{t}$. An \emph{atom} is a term
-$p(t_1,\ldots,t_{\#(p)})$ of sort $\Bool$, or an equality $t \teq s$
-between two terms of sort $\Data$. A \emph{formula} is a possibly
-quantified boolean combination of atoms and a \emph{sentence} is a
-formula in which each variable occurs under the scope of a quantifier.
-A formula is \emph{positive} if and only if each predicate occurs
-under an even number of negations.
+%% For a variable $x \in \fv{}{\phi}$ and a term $t$, let $\phi[t/x]$ be
+%% the result of replacing each occurrence of $x$ by $t$. For indexed
+%% sets of terms $T=\set{t_1,\ldots,t_n}$ and variables
+%% $X=\set{x_1,\ldots,x_n}$, we write $\phi[T/X]$ for the formula
+%% obtained by simultaneously replacing each occurrence of $x_i$ with
+%% $t_i$ in $\phi$, for all $i\in[n]$.
 
-We denote by $\fv{\sigma}{\phi}$ the set of free variables of sort
-$\sigma$ in $\phi$ and write $\fv{}{\phi}$ for $\bigcup_{\sigma \in
-  \ssorts{\sig}} \fv{\sigma}{\phi}$. For a variable $x \in
-\fv{}{\phi}$ and a term $t$ such that $\sort{t} = \sort{x}$, let
-$\phi[t/x]$ be the result of replacing each occurrence of $x$ by
-$t$. For indexed sets $T=\set{t_1,\ldots,t_n}$ and
-$X=\set{x_1,\ldots,x_n}$, where $\sort{t_i} = \sort{x_i}$, we write
-$\phi[T/X]$ for the formula obtained by simultaneously replacing $x_i$
-with $t_i$ in $\phi$, for all $i\in[n]$. The size $\len{\phi}$ is the
-number of symbols occuring in $\phi$.
+An \emph{interpretation} $\I$ maps each predicate $p$ into a set $p^\I
+\subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element of $\Data$
+if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable $x$ into an
+element of $\Data$. Given a term $t$, we denote by $t^\nu$ the value
+obtained by replacing each variable $x$ by the value $\nu(x)$ and
+evaluating each function application. For a formula $\phi$, we define
+the forcing relation $\I,\nu \models \phi$ recursively on the
+structure of $\phi$, as follows:
+\[
+\begin{array}{rcl}
+\I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
+\I,\nu \models t \not\teq s & \iff & t^\nu \neq s^\nu \\
+\I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
+\I,\nu \models \phi_1 \vee \phi_2 & \iff & \I,\nu \models \phi_i \text{, for some } i=1,2 \\
+\I,\nu \models \forall x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for all } d \in \Data \\
+\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data \\
+\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I 
+\end{array}
+\]
+For a formula $\phi$, we define $\sem{\phi}_\nu \isdef \set{\I \mid
+  \I,\nu \models \phi}$, where $\nu(x)$ is defined for each $x \in
+\fv{}{\phi}$, and drop the subscript for sentences. An element of
+$\sem{\phi}$ is called a \emph{model} of $\phi$. A formula is
+satisfiable if $\sem{\phi}\neq\emptyset$.
 
-An \emph{interpretation} $\I$ maps \begin{inparaenum}[(1)]
-\item the sort $\Data$ into a non-empty set $\Data^\I$, 
-%
-\item the sort $\Bool$ into the set $\booli = \set{\true,\false}$, where
-  $\top^\I = \true$, $\bot^\I = \false$, and
-%
-\item each function symbol $f$ into a total function $f^\I :
-  \left(\Data^\I\right)^{\#(f)} \rightarrow \sort{f}^I$, or an element
-  of $\sort{f}^I$ when $\#(f)=0$.
-\end{inparaenum}
-Given an interpretation $\I$, a \emph{valuation} $\nu$ maps each
-variable $x \in \vars$ into an element of $x_\sigma^\I$. For a term
-$t$, we denote by $t^\I_\nu$ the value obtained by replacing each
-function symbol $f$ by its interpretation $f^\I$ and each variable $x$
-by its valuation $\nu(x)$. For a formula $\phi$, we write $\I,\nu
-\models \phi$ if the formula obtained by replacing each term $t$ in
-$\phi$ by the value $t^\I_\nu$ is logically equivalent to true.  The
-semantics of quantifiers is given as usual: $\I,\nu \models \exists x
-~.~ \phi$ if and only if $\I,\nu[x \leftarrow u] \models \phi$, for
-some value $u \in \sort{x}^\I$, where $\nu[x \leftarrow u]$ is the
-valuation that assigns $u$ to $x$ and $\nu(y)$ to all $y \neq x$.
+\section{First Order Alternating Automata}
 
-A formula $\phi$ is \emph{satisfiable} in the interpretation $\I$ if
-there exists a valuation $\nu$ such that $\I,\nu \models \phi$, and
-\emph{valid} if $\I,\nu \models \phi$ for all valuations $\nu$.  The
-\emph{first-order theory} $\theory(\sig,\I)$ is the set of valid
-formulae written in the signature $\sig$, with the interpretation
-$\I$. A \emph{decision procedure} for $\theory(\sig,\I)$ is an
-algorithm that takes a formula $\phi$ in the signature $\sig$ and
-returns yes if and only if $\phi \in \theory(\sig,\I)$.
+Let $\Sigma$ be a finite alphabet $\Sigma$ of \emph{input
+  events}. Given a finite set of variables $X$, we denote by $X
+\mapsto \Data^\I$ the set of valuations of the variables $X$ and
+$\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly infinite
+set of \emph{data symbols}. A \emph{data word} (word) is a finite
+sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data symbols,
+where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n : X
+\rightarrow \Data$ are valuations. We denote by $\varepsilon$ the
+empty sequence, by $\Sigma^*$ the set of finite sequences of input
+events and by $\Sigma[X]^*$ the set of data words over the variables
+$X$.
 
-Given formulae $\varphi$ and $\psi$, we say that \emph{$\phi$ entails
-  $\psi$ modulo $\theory(\sig,\I)$}, denoted $\phi
-\models_{\theory(\sig,\I)} \psi$ iff $\varphi \rightarrow \psi \in
-\theory(\sig,\I)$. We say that $\varphi$ and $\psi$ are
-\emph{$\theory(\sig,\I)$-equivalent}, denoted $\phi
-\equiv_{\theory(\sig,\I)} \psi$ iff $\phi \models_{\theory(\sig,\I)}
-\psi$ and $\psi \models_{\theory(\sig,\I)} \phi$. The theory
-$\theory(\sig,\I)$ has \emph{quantifier elimination} iff for any
-formula $\exists x ~.~ \phi(x)$, where $\phi$ is quantifier-free,
-there exists a quantifier free formula $\psi$ such that $\phi$ and
-$\psi$ are $\theory(\sig,\I)$-equivalent.
-
-\section{First Order Alternating Data Automata}
-
-In the rest of the paper, we we fix a theory $\theory(\sig,\I)$ and
-omit mentioning it in the context of entailments and equivalences. Let
-$\Sigma$ be a finite alphabet $\Sigma$ of \emph{input events}. Given a
-finite set $X \subset \vars$ of variables of sort $\Data$, we denote
-by $X \mapsto \Data^\I$ the set of valuations of the variables $X$ and
-$\Sigma[X] = \Sigma \times (X \mapsto \Data^\I)$ be the possibly
-infinite set of \emph{data symbols}. A \emph{data word} (word in the
-sequel) is a finite sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots
-(a_n,\nu_n)$ of data symbols, where $a_1,\ldots,a_n \in \Sigma$ and
-$\nu_1,\ldots,\nu_n : X \rightarrow \Data^\I$ are valuations. We
-denote by $\varepsilon$ the empty sequence, by $\Sigma^*$ the set of
-finite sequences of input events and by $\Sigma[X]^*$ the set of data
-words over the variables $X$.
-
-A \emph{first-order alternating data automaton} (automaton in the
-sequel) is a tuple $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, where
-$\Sigma$ is the finite event alphabet, $X$ is a
-finite set of variables, $Q \subseteq \preds$ is a finite set of
-\emph{state predicates}, $\iota$ is a sentence, $F \subseteq Q$ is a
-set of \emph{final states} and $\Delta$ is a set of \emph{transition
-  rules} of the form:
-\[q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\]
-where $q \in Q$, $a \in \Sigma$, $\psi$ is a positive formula such
-that $\fv{}{\phi} \subseteq X \cup \set{y_1,\ldots,y_{\#(q)}}$, $X
-\cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$ and each predicate that
-occurs in $\psi$ belongs also to $Q$. We stress the fact that each
-predicate occurs in $\psi$ under an even number of negations.
+A \emph{first-order alternating automaton} (automaton) is a tuple $\A
+= \tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is the finite
+event alphabet, $X$ is a finite set of variables, $Q$ is a finite set
+of predicates, $\iota$ is a satisfiable sentence, $F \subseteq Q$ is a
+set of \emph{final predicates} and $\Delta$ is a set of
+\emph{transition rules} of the form \(q(y_1,\ldots,y_{\#(q)})
+\arrow{a(X)}{} \psi\), where $q \in Q$ is a predicate, $a \in \Sigma$
+is an input event, $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$,
+$\psi$ is a positive formula such that $\fv{}{\phi} \subseteq X \cup
+\set{y_1,\ldots,y_{\#(q)}}$ and each predicate that occurs in $\psi$
+belongs also to $Q$. The \emph{size} of $\A$ is defined as $\len{\A} =
+\len{\iota} + \sum_{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}
+\len{\psi}$.
 
@@
 predicate $q \in Q$ and each input event $a \in \Sigma$, at most one
-such rule (two or more rules can be joined using disjunction). The
-\emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
-\sum_{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta} \len{\psi}$.
+such rule (two or more rules can be joined using disjunction).
 
@@
 
-To simplify the presentation, at this point, we shall give the
-execution semantics of automata in terms of symbolic rewriting rules.
-The connection between this symbolic semantics and the concrete
-semantics, expressed in terms of sets of configurations, will be given
-at a later time. Let $\stamp{X}{i} = \set{\stamp{x}{i} \mid x \in X}$
-be a set of \emph{time-stamped} variables. For a formula $\psi$ and $i
-\in \nat$, we define $\stamp{\psi}{i} = \psi[X/\stamp{X}{i}]$,
-i.e.\ the formula in which each variable $x \in X$, denoting an input
-data element, is replaced by $\stamp{x}{i}$. Further, given a formula
-$\phi$, let $\Accept{F}{\phi}$ be the formula obtained from $\phi$ by
-replacing each predicate atom $q(z_1,\ldots,z_{\#(q)})$ by $\top$ if
-$q \in F$, and by $\bot$, otherwise.
+To formalize the execution semantics of automata, we relate sets of
+configurations to models of first-order sentences, as follows. Given a
+formula $\phi$ and a valuation of its free variables $\nu$, each
+interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
+configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid
+  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a \emph{cube}. For a
+set $\mathcal{M}$ of interpretations, we define $\cube{\mathcal{M}}
+\isdef \set{\cube{\I} \mid \I \in \mathcal{M}}$.
 
-For an input word $(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$,
-we define the sequence of formulae \(\phi_0 \Rightarrow \phi_1
-\Rightarrow \ldots \Rightarrow \phi_n\), where $\phi_0 = \iota$ and
-$\phi_{i+1}$ is obtained from $\phi_i$ by replacing each predicate
-atom $q(z_1,\ldots,z_{\#(q)})$ from $\phi_i$ by the formula
-$\stamp{\psi}{i+1}[z_1, \ldots, z_{\#(q)}/y_1,\ldots,y_{\#(q)}]$,
-where $q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta$ is
-the only transition rule that applies, or by $\bot$, if no transition
-rule applies. Observe that $\phi_n$ is uniquely defined by the event
-sequence $a_1 \ldots a_n \in \Sigma^*$. The word $(a_1,\nu_1) \ldots
-(a_n,\nu_n)$ is \emph{accepted} by $\A$ if and only if $\I,\nu \models
-\Accept{F}{\phi_n}$, where the valuation $\nu : \bigcup_{i=1}^n
-\stamp{X}{i} \rightarrow \Data^\I$ is defined as $\nu(\stamp{x}{i}) =
-\nu_i(x)$, for each $x \in X$ and $i \in [n]$.
-
-As usual, the \emph{language} of $\A$, denoted $\lang{\A}$, is the set
-of words accepted by $\A$. In this paper, we ask the following
+Given a word $w=(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in
+\Sigma[X]^*$, an \emph{execution} of
+$\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is a (possibly
+infinite) forest $\F_w = \set{T_1,T_2,\ldots}$, where each $T_i$ is
+defined below:
+\begin{compactitem}
+\item $T_i$ is rooted in a configuration from some cube $C \in
+  \cube{\sem{\iota}}$ --- since $\iota$ is satisfiable, such cube
+  always exists,
+%
+\item all paths from the root to a leaf of $T_i$ have length $n$,
+%
+\item if $q(d_1,\ldots,d_{\#(q)})$ is a node on level $j \in [0,n-1]$
+  in $T_i$, then its children form a cube $C \in
+  \cube{\sem{\psi}_\eta}$, where \(q(y_1,\ldots,y_{\#(q)})
+  \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is a transition rule and
+  $\eta$ is a valuation that assigns each variable $y_k$ the value
+  $d_k$, for $k \in [1,\#(q)]$ and each variable $x \in X$ the value
+  $\nu_{j+1}(x)$ --- because $\fv{}{\psi} \subseteq
+  \set{y_1,\ldots,y_{\#(q)}} \cup X$, the values assigned by $\eta$ to
+  variables outside $\set{y_1,\ldots,y_{\#(q)}} \cup X$ are not
+  important for $\cube{\sem{\psi}_\eta}$.
+\end{compactitem}
+A forest $\F_w$ is said to be \emph{accepting} if and only if the
+frontier of each tree $T \in \F_w$ is labeled with configurations
+$q(d_1,\ldots,d_{\#(q)})$ such that $q \in F$ is a final predicate. In
+this case, we say that $\A$ accepts $w$ and denote by $\lang{\A}$ the
+set of words accepted by $\A$. In this paper, we answer the following
 questions: \begin{compactenum}
@@
 \end{compactenum}
+
+\subsection{Symbolic Execution}
+
+In several of the upcoming developments it is more convenient to work
+with a symbolic execution semantics, that describes a set of execution
+forests by a sequence of formulae obtained one from another using
+rewriting rules. Let $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an
+automaton and $\stamp{X}{i} = \set{\stamp{x}{i} \mid x \in X}$ be a
+set of \emph{time-stamped} variables. For a formula $\psi$ and $i \in
+\nat$, we define $\stamp{\psi}{i} = \psi[X/\stamp{X}{i}]$, i.e.\ the
+formula in which each variable $x \in X$, denoting an input data
+element, is replaced by $\stamp{x}{i}$. Further, given a formula
+$\phi$, let $\Accept{F}{\phi}$ be the formula obtained from $\phi$ by
+replacing each predicate atom $q(z_1,\ldots,z_{\#(q)})$ by $\top$ if
+$q \in F$, and by $\bot$, otherwise.
+
+For a sequence of input events $a_1 \ldots a_n \in \Sigma^*$, we
+consider the sequence of formulae \(\phi_0 \stackrel{a_1}{\Rightarrow}
+\phi_1 \stackrel{a_2}{\Rightarrow} \ldots \stackrel{a_n}{\Rightarrow}
+\phi_n\), where $\phi_0 = \iota$ and $\phi_{i+1}$ is obtained from
+$\phi_i$ by replacing each predicate atom $q(t_1,\ldots,t_{\#(q)})$
+from $\phi_i$ by the formula $\stamp{\psi}{i+1}[t_1, \ldots,
+  t_{\#(q)}/y_1,\ldots,y_{\#(q)}]$, where $q(y_1,\ldots,y_{\#(q)})
+\arrow{a_i(X)}{} \psi \in \Delta$ is the only transition rule that
+applies, or by $\bot$, if no transition rule applies. The following
+lemma formalizes the relation between symbolic execution and the
+low-level configuration-based semantics of automata:
+
+\begin{lemma}\label{lemma:symbolic-execution}
+  A word $(a_1,\nu_1) \ldots (a_n,\nu_n)$ is accepted by $\A =
+  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ if and only if $\I,\nu \models
+  \Accept{F}{\phi_n}$ for each interpretation $\I$, where the
+  valuation $\nu : \bigcup_{i=1}^n \stamp{X}{i} \rightarrow \Data$ is
+  defined as $\nu(\stamp{x}{i}) = \nu_i(x)$, for each $x \in X$ and $i
+  \in [n]$.
+\end{lemma}
+\proof{ ``$\Rightarrow$'' Assume that $w = (a_1,\nu_1) \ldots
+  (a_n,\nu_n)$ is accepted by $\A$ and let $\F_w$ be an execution
+  forest of $\A$ over $w$. First, it is easy to check that $\phi_i$ is
+  a positive formula, for each $i \in [n]$. Next, we define the
+  interpretation: \[\I^w_n(q) = \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
+    q(d_1,\ldots,d_{\#(q)}) \text{ occurs on the $n$-th level of }
+    F_w}\] and prove $\I^w_n,\nu \models \phi_n$, by induction on
+  $n\geq0$. For $n=0$, we have $\cube{\I^w_0} \in
+  \cube{\sem{\iota}}$ and $\I^w_0 \models \iota$ follows, by
+  definition. For the induction step, assume that $\I^w_{n-1},\mu
+  \models \phi_{n-1}$, where $\mu(\stamp{x}{i}) = \nu_i(x)$, for each
+  $x \in X$ and $i \in [n-1]$.
+
+  \noindent''$\Leftarrow$'' 
+\qed}
 

### 2018-06-08 quantifiers/draft.tex
ok

@@
 For two integers $0 \leq i \leq j$, we denote by $[i,j]$ the set
-$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[1,i]$, where $[0]$ is
-the empty set. To simplify the presentation, we consider two sorts
-$\Data$ and $\Bool$, where $\Data$ ranges over an infinite domain and
-$\Bool$ has the usual interpretation. The $\Data$ sort is equipped
-with a finite set of function symbols $f : \Data^{\#(f)} \rightarrow
-\Data$, where $\#(f)\geq0$ denotes the number of arguments (arity) of
-$f$. When $\#(f)=0$, we say that $f$ is a constant, and denote by
-$\top$ and $\bot$ the only constants of sort $\Bool$, representing the
-true and false values, respectively.  For conciseness, in the
-following we blur the distinction between the sort and function
-symbols and their interpretations. A \emph{predicate} is a function
-symbol $p : \Data^{\#(p)} \rightarrow \Bool$, denoting a relation of
-arity $\#(p)$.
+$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[0,i]$. To simplify the
+presentation, we consider two sorts $\Data$ and $\Bool$, where $\Data$
+ranges over an infinite domain and $\Bool$ has the usual
+interpretation. The $\Data$ sort is equipped with a finite set of
+function symbols $f : \Data^{\#(f)} \rightarrow \Data$, where
+$\#(f)\geq0$ denotes the number of arguments (arity) of $f$. When
+$\#(f)=0$, we say that $f$ is a constant, and denote by $\top$ and
+$\bot$ the only constants of sort $\Bool$, representing the true and
+false values, respectively. A \emph{predicate} is a function symbol
+$p : \Data^{\#(p)} \rightarrow \Bool$, denoting a relation of arity
+$\#(p)$ and we write $\preds$ for the set of predicates.
 
@@
 are either constants of sort $\Data$, variables or function
-applications. The set of \emph{positive} formulae is defined by the
-syntax below: 
+applications. The set of \emph{positive} first-order formulae is
+defined by the syntax below:
 \[\phi := t \teq s \mid t \not\teq s \mid p(t_1,\ldots,t_{\#(p)}) 
@@
 
+Interpretations are partially ordered by the pointwise subset order,
+defined as $\I_1 \subseteq \I_2$ if and only if $p^{\I_1} \subseteq
+p^{\I_2}$ for each predicate $p \in \preds$. Given a set $\mathcal{M}$
+of interpretations, a minimal element $\I \in \mathcal{M}$ is an
+interpretation such that for no other interpretation $\I' \in
+\mathcal{M} \setminus \set{\I}$ do we have $\I' \subseteq \I$. For a
+formula $\phi$ and a valuation $\nu$, we denote by $\minsem{\phi}_\nu$
+and $\minsem{\phi}$ the set of minimal interpretations from
+$\sem{\phi}_\nu$ and $\sem{\phi}$, respectively.
+
 \section{First Order Alternating Automata}
@@
 predicate $q \in Q$ and each input event $a \in \Sigma$, at most one
-such rule (two or more rules can be joined using disjunction).
+such rule --- two or more rules can be joined using disjunction.
+
+The execution semantics of of automata is given in close analogy with
+the case of boolean alternating automata, with transition rules of the
+form $q \arrow{a}{} \phi$, where $q$ is a boolean constant and $\phi$
+a positive boolean combination of such constants. For instance, $q_0
+\arrow{a}{} q_1 \wedge q_2 \vee q_3$ means that the automaton can
+choose to transition in either both $q_1$ and $q_2$ or in $q_3$
+alone. This intuition leads to saying that the steps of the automaton
+are defined by the minimal boolean models of the transition
+formulae. In this case, both $\set{q_1 \leftarrow \top, q_2 \leftarrow
+  \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow \bot, q_2
+  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, but not
+$\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
+which is a model but is not minimal.
 
@@
 a \emph{configuration}\footnote{Note that a configuration is not a
-  logical term since data values cannot be written in logic.}.
-Because we consider an alternating model of computation, an automaton
-can find itself, at the same time, in more than one
-configuration. Typically, in the classical literature on alternating
-machines, this set is finite. In our case, due to the presence of
-first-order quantifiers ranging over an infinite data domain, we
-consider that an automaton can be also in an infinite set of
-configurations, all at once.
-
-To formalize the execution semantics of automata, we relate sets of
+  logical term since data values cannot be written in logic.}. To
+formalize the execution semantics of automata, we relate sets of
 configurations to models of first-order sentences, as follows. Given a
@@
 set $\mathcal{M}$ of interpretations, we define $\cube{\mathcal{M}}
-\isdef \set{\cube{\I} \mid \I \in \mathcal{M}}$.
-
-Given a word $w=(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in
-\Sigma[X]^*$, an \emph{execution} of
-$\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is a (possibly
-infinite) forest $\F_w = \set{T_1,T_2,\ldots}$, where each $T_i$ is
-defined below:
+\isdef \set{\cube{\I} \mid \I \in \mathcal{M}}$. Given a word
+$w=(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in \Sigma[X]^*$, an
+\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is
+a (possibly infinite) forest $\F_w = \set{T_1,T_2,\ldots}$, where each
+$T_i$ is defined below:
 \begin{compactitem}
-\item $T_i$ is rooted in a configuration from some cube $C \in
-  \cube{\sem{\iota}}$ --- since $\iota$ is satisfiable, such cube
+\item $T_i$ is rooted in a configuration from some cube from
+  $\cube{\minsem{\iota}}$ --- because $\iota$ is satisfiable, such cube
   always exists,
 %
-\item all paths from the root to a leaf of $T_i$ have length $n$,
+\item all paths from the root to a leaf of $T_i$ have length $n$, and
 %
 \item if $q(d_1,\ldots,d_{\#(q)})$ is a node on level $j \in [0,n-1]$
-  in $T_i$, then its children form a cube $C \in
-  \cube{\sem{\psi}_\eta}$, where \(q(y_1,\ldots,y_{\#(q)})
+  in $T_i$, then its children form a cube from
+  $\cube{\minsem{\psi}_\eta}$, where \(q(y_1,\ldots,y_{\#(q)})
   \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is a transition rule and
@@
   $d_k$, for $k \in [1,\#(q)]$ and each variable $x \in X$ the value
-  $\nu_{j+1}(x)$ --- because $\fv{}{\psi} \subseteq
-  \set{y_1,\ldots,y_{\#(q)}} \cup X$, the values assigned by $\eta$ to
-  variables outside $\set{y_1,\ldots,y_{\#(q)}} \cup X$ are not
-  important for $\cube{\sem{\psi}_\eta}$.
+  $\nu_{j+1}(x)$ --- because $\fv{}{\psi} \subseteq X \cup
+  \set{y_1,\ldots,y_{\#(q)}}$, the values assigned by $\eta$ to
+  variables outside $X \cup \set{y_1,\ldots,y_{\#(q)}}$ have no impact
+  on $\cube{\minsem{\psi}_\eta}$.
 \end{compactitem}
@@
 \phi_1 \stackrel{a_2}{\Rightarrow} \ldots \stackrel{a_n}{\Rightarrow}
-\phi_n\), where $\phi_0 = \iota$ and $\phi_{i+1}$ is obtained from
-$\phi_i$ by replacing each predicate atom $q(t_1,\ldots,t_{\#(q)})$
-from $\phi_i$ by the formula $\stamp{\psi}{i+1}[t_1, \ldots,
-  t_{\#(q)}/y_1,\ldots,y_{\#(q)}]$, where $q(y_1,\ldots,y_{\#(q)})
-\arrow{a_i(X)}{} \psi \in \Delta$ is the only transition rule that
-applies, or by $\bot$, if no transition rule applies. The following
-lemma formalizes the relation between symbolic execution and the
-low-level configuration-based semantics of automata:
+\phi_n\), where $\phi_0 = \iota$ and, for each $i \in [n-1]$, the
+formula $\phi_{i+1}$ is obtained from $\phi_i$ by replacing each
+predicate atom $q(t_1,\ldots,t_{\#(q)})$ from $\phi_i$ by the formula
+$\stamp{\psi}{i+1}[t_1, \ldots, t_{\#(q)}/y_1,\ldots,y_{\#(q)}]$,
+where $q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta$ is
+the only transition rule that applies, or by $\bot$, if no transition
+rule applies. The following lemma formalizes the relation between the
+symbolic and the configuration-based semantics of automata:
 
@@
 
-\subsection{Configuration-based Semantics}
-
-We give a low-level operational semantics of automata, in terms of
-sets of configurations, with a conjunctive meaning. A (possibly
-infinite) set $C$ of configurations is also called a
-\emph{cube}. Intuitively, an automaton $\A$ is in a cube $C$ if it is
-simultaneously in all configurations $q(d_1, \ldots, d_{\#(q)})$ from
-$C$. A positive sentence $\phi$ containing at least one occurrence of
-a predicate, defines a (possibly infinite) set of cubes $\sem{\phi}$,
-defined next.
-
-A \emph{context} $q[x_1,\ldots,x_k] = q(\delta_1, \ldots,
-\delta_{\#(q)})$ is a configuration in which one or more variables may
-occur, i.e. $\delta_i \in \Data^\I \cup \set{x_1,\ldots,x_k}$, for all
-$i \in [\#(q)]$. For a set of data values $D \subseteq \Data^\I$, we
-let $q[x_1,\ldots,x_{i-1},D,x_{i+1}, \ldots, x_k] =
-\{q[x_1,\ldots,x_{i-1},d,x_{i+1},\ldots,x_k] \mid d \in D\}$ be a set
-of contexts. Observe that $q[D]$ is a cube, when $k=1$.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-06-08 quantifiers/draft.tex
ok

@@
 $\bot$ the only constants of sort $\Bool$, representing the true and
-false values, respectively. A \emph{predicate} is a function symbol
-$p : \Data^{\#(p)} \rightarrow \Bool$, denoting a relation of arity
-$\#(p)$ and we write $\preds$ for the set of predicates.
+false values, respectively. A \emph{predicate} is a function symbol $p
+: \Data^{\#(p)} \rightarrow \Bool$, denoting a relation of arity
+$\#(p)$ and we write $\preds$ for the set of predicates. In the
+following, we shall consider that the interpretation of all function
+symbols $f : \Data^{\#(f)} \rightarrow \Data$ that are not predicates
+is fixed and relative to the interpretation of the $\Data$ sort.
 
@@
 \I,\nu \models t \not\teq s & \iff & t^\nu \neq s^\nu \\
+\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I \\
 \I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
@@
 \I,\nu \models \forall x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for all } d \in \Data \\
-\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data \\
-\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I 
+\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data 
 \end{array}
@@
   \I,\nu \models \phi}$, where $\nu(x)$ is defined for each $x \in
-\fv{}{\phi}$, and drop the subscript for sentences. An element of
-$\sem{\phi}$ is called a \emph{model} of $\phi$. A formula is
-satisfiable if $\sem{\phi}\neq\emptyset$.
+\fv{}{\phi}$, and drop the subscript for sentences. A sentence $\phi$
+is \emph{satisfiable} if $\sem{\phi}\neq\emptyset$, an element of
+$\sem{\phi}$ being called a \emph{model} of $\phi$.
 
@@
 defined as $\I_1 \subseteq \I_2$ if and only if $p^{\I_1} \subseteq
-p^{\I_2}$ for each predicate $p \in \preds$. Given a set $\mathcal{M}$
-of interpretations, a minimal element $\I \in \mathcal{M}$ is an
+p^{\I_2}$ for each predicate $p \in \preds$. Given a set $\mathcal{S}$
+of interpretations, a minimal element $\I \in \mathcal{S}$ is an
 interpretation such that for no other interpretation $\I' \in
-\mathcal{M} \setminus \set{\I}$ do we have $\I' \subseteq \I$. For a
+\mathcal{S} \setminus \set{\I}$ do we have $\I' \subseteq \I$. For a
 formula $\phi$ and a valuation $\nu$, we denote by $\minsem{\phi}_\nu$
@@
 set of \emph{data symbols}. A \emph{data word} (word) is a finite
-sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data symbols,
-where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n : X
-\rightarrow \Data$ are valuations. We denote by $\varepsilon$ the
-empty sequence, by $\Sigma^*$ the set of finite sequences of input
-events and by $\Sigma[X]^*$ the set of data words over the variables
-$X$.
+sequence $w=(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data
+symbols, where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n : X
+\rightarrow \Data$ are valuations. Given a word $w$, we denote by
+$\event{w} \isdef a_1\ldots a_n$ its sequence of input events and by
+$\data{w}$ the valuation associating each time-stamped variable
+$\stamp{x}{i}$ the value $\nu_i(x)$, for all $x \in \vars$ and
+$i\in[1,n]$. We denote by $\varepsilon$ the empty sequence, by
+$\Sigma^*$ the set of finite sequences of input events and by
+$\Sigma[X]^*$ the set of data words over the variables $X$.
 
@@
 $\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
-which is a model but is not minimal.
+which is a model but is not minimal\footnote{To avoid the confusion of
+  considering potentially non-minimal models, the original definition
+  of alternating finite-state automata \cite{ChandraKozenStockmeyer81}
+  uses boolean valuations instead of formulae.}.
 
@@
   \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a \emph{cube}. For a
-set $\mathcal{M}$ of interpretations, we define $\cube{\mathcal{M}}
-\isdef \set{\cube{\I} \mid \I \in \mathcal{M}}$. Given a word
+set $\mathcal{S}$ of interpretations, we define $\cube{\mathcal{S}}
+\isdef \set{\cube{\I} \mid \I \in \mathcal{S}}$. Given a word
 $w=(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in \Sigma[X]^*$, an
@@
 a (possibly infinite) forest $\F_w = \set{T_1,T_2,\ldots}$, where each
-$T_i$ is defined below:
+tree $T_i$ is defined below:
 \begin{compactitem}
 \item $T_i$ is rooted in a configuration from some cube from
-  $\cube{\minsem{\iota}}$ --- because $\iota$ is satisfiable, such cube
+  $\cube{\minsem{\iota}}$ --- because $\iota$ is satisfiable, such a cube
   always exists,
@@
 %
-\item if $q(d_1,\ldots,d_{\#(q)})$ is a node on level $j \in [0,n-1]$
-  in $T_i$, then its children form a cube from
+\item if $q(d_1,\ldots,d_{\#(q)})$ is a node on the level $j \in
+  [n-1]$ in $T_i$, then its children form a cube from
   $\cube{\minsem{\psi}_\eta}$, where \(q(y_1,\ldots,y_{\#(q)})
@@
 \subsection{Symbolic Execution}
+
+In the upcoming developments it is more convenient to work with path
+formulae defining executions.  Let $\A =
+\tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the rest of
+this session. For any $i \in \nat$, we denote by $\stamp{Q}{i} =
+\set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} = \set{\stamp{x}{i}
+  \mid x \in X}$ the sets of time-stamped predicates and variables,
+respectively. For a formula $\psi$ and $i \in \nat$, we define
+$\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the
+formula in which all input variables and state predicates are replaced
+by their time-stamped counterparts.
+
+Given a sequence of input events $\alpha = a_1 \ldots a_n$, the
+\emph{path formula} of $\alpha$ in $\A$ is defined as follows: 
+\[\pathform{\A}{\alpha} \isdef \stamp{\iota}{0} \wedge 
+\bigwedge_{i=1}^n \bigwedge_{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{}
+  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
+\stamp{q}{i-1}(y_1,\ldots,y_{\#(q)}) \Longrightarrow \stamp{\psi}{i}\]
+Observe that path formulae are not positive, due to the occurrence of
+predicate atoms to the left of the implication as in
+$\stamp{q}{i-1}(y_1,\ldots,y_{\#(q)}) \Longrightarrow
+\stamp{\psi}{i}$. The intuition is that, whenever the automaton is in
+a configuration $q(d_1, \ldots, d_{\#(q)})$ at step $i-1$, it proceeds
+at step $i$, according to $\psi$.
+
+To formalize the relation between the low-level configuration-based
+execution semantics and the symbolic path formulae, consider a word
+$w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any execution forest $\F_w$ of
+$\A$ over $w$ is associated a unique interpretation defined as
+$\I_{[\F_w]}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}}
+  \mid q(d_1,\ldots,d_{\#(q)}) \text{ occurs on level $i$ in $\F_w$}}$
+for each $q \in \preds$ and $i \in [n]$.
+
+\begin{lemma}\label{lemma:path-formula}
+  For any word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have: 
+  \[\minsem{\pathform{\A}{w_\Sigma}}_{w_\Data} = \set{\I_{[\F_w]} \mid \F_w
+    \text{ is an execution of $\A$ over $w$}}\enspace.\]
+\end{lemma}
+
+\comment[ri]{I am here}
 

### 2018-06-27 quantifiers/draft.tex
ok

@@
 presentation, we consider two sorts $\Data$ and $\Bool$, where $\Data$
-ranges over an infinite domain and $\Bool$ has the usual
-interpretation. The $\Data$ sort is equipped with a finite set of
-function symbols $f : \Data^{\#(f)} \rightarrow \Data$, where
+is an infinite domain and $\Bool$ is the set of truth values $\top$
+(true) and $\bot$ (false). The $\Data$ sort is equipped with finitely
+many function symbols $f : \Data^{\#(f)} \rightarrow \Data$, where
 $\#(f)\geq0$ denotes the number of arguments (arity) of $f$. When
-$\#(f)=0$, we say that $f$ is a constant, and denote by $\top$ and
-$\bot$ the only constants of sort $\Bool$, representing the true and
-false values, respectively. A \emph{predicate} is a function symbol $p
-: \Data^{\#(p)} \rightarrow \Bool$, denoting a relation of arity
-$\#(p)$ and we write $\preds$ for the set of predicates. In the
-following, we shall consider that the interpretation of all function
-symbols $f : \Data^{\#(f)} \rightarrow \Data$ that are not predicates
-is fixed and relative to the interpretation of the $\Data$ sort.
+$\#(f)=0$, we say that $f$ is a constant. A \emph{predicate} is a
+function symbol $p : \Data^{\#(p)} \rightarrow \Bool$, denoting a
+relation of arity $\#(p)$ and we write $\preds$ for the set of
+predicates. In the following, we shall consider that the
+interpretation of all function symbols $f : \Data^{\#(f)} \rightarrow
+\Data$ that are not predicates is fixed and relative to the
+interpretation of the $\Data$ sort. For simplicity, we further blur
+the notational distinction between function symbols and their
+interpretations.
 
@@
 defined by the syntax below:
-\[\phi := t \teq s \mid t \not\teq s \mid p(t_1,\ldots,t_{\#(p)}) 
-\mid \phi_1 \wedge \phi_2 \mid \phi_1 \vee \phi_2 \mid \exists x ~.~
-\phi_1 \mid \forall x ~.~ \phi_1\] where $t,s,t_1,\ldots,t_{\#(p)}$
-denote terms. We denote by $\fv{}{\phi}$ the set of free variables in
-$\phi$. A \emph{sentence} is a formula $\phi$ in which each variable
-occurs under the scope of a quantifier, i.e. $\fv{}{\phi} =
-\emptyset$. The size $\len{\phi}$ is the number of symbols occuring in
-$\phi$.
+\[\phi := t \teq s \mid p(t_1,\ldots,t_{\#(p)}) 
+\mid \neg \phi_1 \mid \phi_1 \wedge \phi_2 \mid \exists x ~.~
+\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms. We write
+$\phi_1 \vee \phi_2$, $\phi_1 \Rightarrow \phi_2$ and $\forall x ~.~
+\phi_1$ for $\neg(\neg\phi_1 \wedge \neg\phi_2)$, $\neg\phi_1 \vee
+\phi_2$ and $\neg\exists x ~.~ \neg\phi_1$, respectively. We denote by
+$\fv{}{\phi}$ the set of free variables in $\phi$. A \emph{sentence}
+is a formula $\phi$ in which each variable occurs under the scope of a
+quantifier, i.e. $\fv{}{\phi} = \emptyset$. A formula is
+\emph{positive} if each predicate symbol occurs under an even number
+of negations. The size $\len{\phi}$ of a formula $\phi$ is the number
+of symbols needed to write it down.
 
@@
 \I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
-\I,\nu \models t \not\teq s & \iff & t^\nu \neq s^\nu \\
 \I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I \\
+\I,\nu \models \neg\phi_1 & \iff & \I,\nu \not\models \phi_1 \\
 \I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
-\I,\nu \models \phi_1 \vee \phi_2 & \iff & \I,\nu \models \phi_i \text{, for some } i=1,2 \\
-\I,\nu \models \forall x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for all } d \in \Data \\
 \I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data 
@@
   events}. Given a finite set of variables $X$, we denote by $X
-\mapsto \Data^\I$ the set of valuations of the variables $X$ and
+\mapsto \Data$ the set of valuations of the variables $X$ and
 $\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly infinite
-set of \emph{data symbols}. A \emph{data word} (word) is a finite
-sequence $w=(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data
-symbols, where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n : X
-\rightarrow \Data$ are valuations. Given a word $w$, we denote by
+set of \emph{data symbols} $(a,\nu)$, where $a$ is an input symbol and
+$\nu$ is a valuation. A \emph{data word} (simply called word in the
+following) is a finite sequence $w=(a_1,\nu_1)(a_2,\nu_2) \ldots
+(a_n,\nu_n)$ of data symbols. Given a word $w$, we denote by
 $\event{w} \isdef a_1\ldots a_n$ its sequence of input events and by
@@
 
-A \emph{first-order alternating automaton} (automaton) is a tuple $\A
-= \tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is the finite
-event alphabet, $X$ is a finite set of variables, $Q$ is a finite set
-of predicates, $\iota$ is a satisfiable sentence, $F \subseteq Q$ is a
-set of \emph{final predicates} and $\Delta$ is a set of
-\emph{transition rules} of the form \(q(y_1,\ldots,y_{\#(q)})
-\arrow{a(X)}{} \psi\), where $q \in Q$ is a predicate, $a \in \Sigma$
-is an input event, $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$,
-$\psi$ is a positive formula such that $\fv{}{\phi} \subseteq X \cup
-\set{y_1,\ldots,y_{\#(q)}}$ and each predicate that occurs in $\psi$
-belongs also to $Q$. The \emph{size} of $\A$ is defined as $\len{\A} =
-\len{\iota} + \sum_{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}
-\len{\psi}$.
+A \emph{first-order alternating automaton} (simply called automaton in
+the following) is a tuple $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$,
+where $\Sigma$ is the finite event alphabet, $X$ is a finite set of
+variables, $Q$ is a finite set of predicates, $\iota$ is a satisfiable
+sentence, $F \subseteq Q$ is a set of \emph{final predicates} and
+$\Delta$ is a set of \emph{transition rules} of the form
+\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
+predicate, $a \in \Sigma$ is an input event, $X \cap
+\set{y_1,\ldots,y_{\#(q)}} = \emptyset$, $\psi$ is a positive formula
+such that $\fv{}{\phi} \subseteq X \cup \set{y_1,\ldots,y_{\#(q)}}$
+and each predicate that occurs in $\psi$ belongs also to $Q$. The
+\emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
+\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}} \len{\psi}$.
 
-Intuitively, $a$ is the input event and $X$ are the input data values
-that trigger the transition, whereas $q$ and $y_1,\ldots,y_{\#(q)}$
-are the current control state and data values pertaining to that
-state, respectively. Without loss of generality, we consider, for each
-predicate $q \in Q$ and each input event $a \in \Sigma$, at most one
-such rule --- two or more rules can be joined using disjunction.
+The intuition of a transition rule \(q(y_1,\ldots,y_{\#(q)})
+\arrow{a(X)}{} \psi\) is the following: $a$ is the input event and $X$
+are the input data values that trigger the transition, whereas $q$ and
+$y_1,\ldots,y_{\#(q)}$ are the current control state and data values
+in that state, respectively. Without loss of generality, we consider,
+for each predicate $q \in Q$ and each input event $a \in \Sigma$, at
+most one such rule --- two or more rules can be joined using
+disjunction.
 
-The execution semantics of of automata is given in close analogy with
+The execution semantics of automata is given in close analogy with
 the case of boolean alternating automata, with transition rules of the
@@
 \emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is
-a (possibly infinite) forest $\F_w = \set{T_1,T_2,\ldots}$, where each
+a (possibly infinite) forest $\F = \set{T_1,T_2,\ldots}$, where each
 tree $T_i$ is defined below:
@@
 \item $T_i$ is rooted in a configuration from some cube from
-  $\cube{\minsem{\iota}}$ --- because $\iota$ is satisfiable, such a cube
-  always exists,
+  $\cube{\minsem{\iota}}$. % --- because $\iota$ is satisfiable, such
+                           % a cube always exists,
 %
@@
   $d_k$, for $k \in [1,\#(q)]$ and each variable $x \in X$ the value
-  $\nu_{j+1}(x)$ --- because $\fv{}{\psi} \subseteq X \cup
-  \set{y_1,\ldots,y_{\#(q)}}$, the values assigned by $\eta$ to
-  variables outside $X \cup \set{y_1,\ldots,y_{\#(q)}}$ have no impact
-  on $\cube{\minsem{\psi}_\eta}$.
+  $\nu_{j+1}(x)$. % --- because $\fv{}{\psi} \subseteq X \cup
+                  % \set{y_1,\ldots,y_{\#(q)}}$, the values assigned
+                  % by $\eta$ to variables outside $X \cup
+                  % \set{y_1,\ldots,y_{\#(q)}}$ have no impact on
+                  % $\cube{\minsem{\psi}_\eta}$.
 \end{compactitem}
-A forest $\F_w$ is said to be \emph{accepting} if and only if the
-frontier of each tree $T \in \F_w$ is labeled with configurations
+An execution $\F$ is said to be \emph{accepting} if and only if the
+frontier of each tree $T \in \F$ is labeled with configurations
 $q(d_1,\ldots,d_{\#(q)})$ such that $q \in F$ is a final predicate. In
@@
 
-In the upcoming developments it is more convenient to work with path
-formulae defining executions.  Let $\A =
+In the upcoming developments it is more convenient to work with
+logical formulae defining executions. Let $\A =
 \tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the rest of
@@
 formula in which all input variables and state predicates are replaced
-by their time-stamped counterparts.
-
-Given a sequence of input events $\alpha = a_1 \ldots a_n$, the
-\emph{path formula} of $\alpha$ in $\A$ is defined as follows: 
+by their time-stamped counterparts. Given a sequence of input events
+$\alpha = a_1 \ldots a_n$, the \emph{path formula} of $\alpha$ in $\A$
+is defined as follows:
 \[\pathform{\A}{\alpha} \isdef \stamp{\iota}{0} \wedge 
@@
 \stamp{q}{i-1}(y_1,\ldots,y_{\#(q)}) \Longrightarrow \stamp{\psi}{i}\]
-Observe that path formulae are not positive, due to the occurrence of
-predicate atoms to the left of the implication as in
-$\stamp{q}{i-1}(y_1,\ldots,y_{\#(q)}) \Longrightarrow
-\stamp{\psi}{i}$. The intuition is that, whenever the automaton is in
-a configuration $q(d_1, \ldots, d_{\#(q)})$ at step $i-1$, it proceeds
-at step $i$, according to $\psi$.
-
 To formalize the relation between the low-level configuration-based
 execution semantics and the symbolic path formulae, consider a word
-$w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any execution forest $\F_w$ of
-$\A$ over $w$ is associated a unique interpretation defined as
-$\I_{[\F_w]}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}}
-  \mid q(d_1,\ldots,d_{\#(q)}) \text{ occurs on level $i$ in $\F_w$}}$
-for each $q \in \preds$ and $i \in [n]$.
+$w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any execution forest $\F$ of $\A$
+over $w$ is associated an interpretation $\I_{\F}$ of the set of
+time-stamped predicates $\set{\stamp{q}{i} \mid q \in Q, i \in [n]}$,
+defined as follows:
+\[\I_{\F}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}}
+  \mid q(d_1,\ldots,d_{\#(q)}) \text{ occurs on level $i$ in
+    $\F$}}\enspace.\]
 
@@
   For any word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have: 
-  \[\minsem{\pathform{\A}{w_\Sigma}}_{w_\Data} = \set{\I_{[\F_w]} \mid \F_w
+  \[\minsem{\pathform{\A}{w_\Sigma}}_{w_\Data} = \set{\I_{\F} \mid \F
     \text{ is an execution of $\A$ over $w$}}\enspace.\]

### 2018-06-29 quantifiers/draft.tex
ok

@@
 are either constants of sort $\Data$, variables or function
-applications. The set of \emph{positive} first-order formulae is
-defined by the syntax below:
+applications. The set of first-order formulae is defined by the syntax
+below:
 \[\phi := t \teq s \mid p(t_1,\ldots,t_{\#(p)}) 
@@
 
-The execution semantics of automata is given in close analogy with
-the case of boolean alternating automata, with transition rules of the
+The execution semantics of automata is given in close analogy with the
+case of boolean alternating automata, with transition rules of the
 form $q \arrow{a}{} \phi$, where $q$ is a boolean constant and $\phi$
@@
   \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow \bot, q_2
-  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, but not
+  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, however
 $\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
-which is a model but is not minimal\footnote{To avoid the confusion of
+is a model but is not minimal\footnote{To avoid the confusion of
   considering potentially non-minimal models, the original definition
@@
 configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid
-  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a \emph{cube}. For a
-set $\mathcal{S}$ of interpretations, we define $\cube{\mathcal{S}}
-\isdef \set{\cube{\I} \mid \I \in \mathcal{S}}$. Given a word
-$w=(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in \Sigma[X]^*$, an
-\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is
-a (possibly infinite) forest $\F = \set{T_1,T_2,\ldots}$, where each
-tree $T_i$ is defined below:
+  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I, q \in Q}$, called a
+\emph{cube}. For a set $\mathcal{S}$ of interpretations, we define
+$\cube{\mathcal{S}} \isdef \set{\cube{\I} \mid \I \in
+  \mathcal{S}}$. Given a word $w=(a_1,\nu_1)(a_2,\nu_2) \ldots
+(a_n,\nu_n) \in \Sigma[X]^*$, an \emph{execution} of
+$\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is a (possibly
+infinite) forest $\F = \set{T_1,T_2,\ldots}$, where each tree $T_i$ is
+such that:
 \begin{compactitem}
-\item $T_i$ is rooted in a configuration from some cube from
+\item the roots of $T_1, T_2, \ldots$ form a cube from
   $\cube{\minsem{\iota}}$. % --- because $\iota$ is satisfiable, such
@@
 \end{lemma}
-
-\comment[ri]{I am here}
-
-In several of the upcoming developments it is more convenient to work
-with a symbolic execution semantics, that describes a set of execution
-forests by a sequence of formulae obtained one from another using
-rewriting rules. Let $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an
-automaton and $\stamp{X}{i} = \set{\stamp{x}{i} \mid x \in X}$ be a
-set of \emph{time-stamped} variables. For a formula $\psi$ and $i \in
-\nat$, we define $\stamp{\psi}{i} = \psi[X/\stamp{X}{i}]$, i.e.\ the
-formula in which each variable $x \in X$, denoting an input data
-element, is replaced by $\stamp{x}{i}$. Further, given a formula
-$\phi$, let $\Accept{F}{\phi}$ be the formula obtained from $\phi$ by
-replacing each predicate atom $q(z_1,\ldots,z_{\#(q)})$ by $\top$ if
-$q \in F$, and by $\bot$, otherwise.
-
-For a sequence of input events $a_1 \ldots a_n \in \Sigma^*$, we
-consider the sequence of formulae \(\phi_0 \stackrel{a_1}{\Rightarrow}
-\phi_1 \stackrel{a_2}{\Rightarrow} \ldots \stackrel{a_n}{\Rightarrow}
-\phi_n\), where $\phi_0 = \iota$ and, for each $i \in [n-1]$, the
-formula $\phi_{i+1}$ is obtained from $\phi_i$ by replacing each
-predicate atom $q(t_1,\ldots,t_{\#(q)})$ from $\phi_i$ by the formula
-$\stamp{\psi}{i+1}[t_1, \ldots, t_{\#(q)}/y_1,\ldots,y_{\#(q)}]$,
-where $q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta$ is
-the only transition rule that applies, or by $\bot$, if no transition
-rule applies. The following lemma formalizes the relation between the
-symbolic and the configuration-based semantics of automata:
-
-\begin{lemma}\label{lemma:symbolic-execution}
-  A word $(a_1,\nu_1) \ldots (a_n,\nu_n)$ is accepted by $\A =
-  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ if and only if $\I,\nu \models
-  \Accept{F}{\phi_n}$ for each interpretation $\I$, where the
-  valuation $\nu : \bigcup_{i=1}^n \stamp{X}{i} \rightarrow \Data$ is
-  defined as $\nu(\stamp{x}{i}) = \nu_i(x)$, for each $x \in X$ and $i
-  \in [n]$.
-\end{lemma}
-\proof{ ``$\Rightarrow$'' Assume that $w = (a_1,\nu_1) \ldots
-  (a_n,\nu_n)$ is accepted by $\A$ and let $\F_w$ be an execution
-  forest of $\A$ over $w$. First, it is easy to check that $\phi_i$ is
-  a positive formula, for each $i \in [n]$. Next, we define the
-  interpretation: \[\I^w_n(q) = \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
-    q(d_1,\ldots,d_{\#(q)}) \text{ occurs on the $n$-th level of }
-    F_w}\] and prove $\I^w_n,\nu \models \phi_n$, by induction on
-  $n\geq0$. For $n=0$, we have $\cube{\I^w_0} \in
-  \cube{\sem{\iota}}$ and $\I^w_0 \models \iota$ follows, by
-  definition. For the induction step, assume that $\I^w_{n-1},\mu
-  \models \phi_{n-1}$, where $\mu(\stamp{x}{i}) = \nu_i(x)$, for each
-  $x \in X$ and $i \in [n-1]$.
-
-  \noindent''$\Leftarrow$'' 
-\qed}
+\proof{ ``$\subseteq$'' Let $\I$ be a minimal interpretation such that
+  $\I,w_\Data \models \pathform{\A}{w_\Sigma}$. We show that there
+  exists an execution $\F$ of $\A$ over $w$ such that $\I = \I_\F$, by
+  induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
+  $\pathform{\A}{w_\Sigma} = \stamp{\iota}{0}$. Because $\iota$ is a
+  sentence, the valuation $w_\Data$ is not important in $\I,w_\Data
+  \models \stamp{\iota}{0}$ and, moreover, since $\I$ is minimal, we
+  have $\I \in \minsem{\stamp{\iota}{0}}$. We define the
+  interpretation $\J(q)=\I(\stamp{q}{0})$, for all $q \in Q$. Then
+  $\cube{\J}$ is an execution of $\A$ over $\epsilon$ and $\I =
+  \I_{\cube{\J}}$ is immediate. For the inductive case $n>0$, we
+  assume that $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\J$ be the
+  interpretation defined as $\I$ for all $\stamp{q}{i}$, with $q \in
+  Q$ and $i \in [n-1]$, and $\emptyset$ everywhere else. Then
+  $\J,u_\Data \models \pathform{\A}{u_\Sigma}$ and $\J$ is moreover
+  minimal. By the induction hypothesis, there exists an execution $\G$
+  of $\A$ over $u$, such that $\J = \I_\G$. We extend this to an
+  execution over $w$, by appending to each configuration
+  $q(d_1,\ldots,d_{\#(q)})$ on the frontier [($n-1$)-th level] of a
+  tree $T \in \G$ the leaves corresponding to \qed}
 

### 2018-06-30 quantifiers/draft.tex
ok

@@
 \declareauthor{ri}{Radu}{blue}
+\declareauthor{xx}{Xiao}{red}
 
@@
 \item the roots of $T_1, T_2, \ldots$ form a cube from
-  $\cube{\minsem{\iota}}$. % --- because $\iota$ is satisfiable, such
+  $\cube{\minsem{\iota}}$, % --- because $\iota$ is satisfiable, such
                            % a cube always exists,
 %
-\item all paths from the root to a leaf of $T_i$ have length $n$, and
+% \item all paths from the root to a leaf of $T_i$ have length $n$, and
 %
@@
   minimal. By the induction hypothesis, there exists an execution $\G$
-  of $\A$ over $u$, such that $\J = \I_\G$. We extend this to an
-  execution over $w$, by appending to each configuration
-  $q(d_1,\ldots,d_{\#(q)})$ on the frontier [($n-1$)-th level] of a
-  tree $T \in \G$ the leaves corresponding to \qed}
+  of $\A$ over $u$, such that $\J = \I_\G$. Consider a leaf of a tree
+  $T \in \G$, labeled with a configuration $q(d_1,\ldots,d_{\#(q)})$
+  and let $\forall y_1 \ldots \forall y_{\#(q)} ~.~
+  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{n}$ be
+  the subformula of $\pathform{\A}{w_\Sigma}$ corresponding to the
+  application(s) of the transition rule $q(y_1,\ldots,y_{\#(q)})
+  \arrow{a_n}{} \psi$ at the $(n-1)$-th step. Let $\nu$ be the
+  valuation that assigns each $y_i$ the value $d_i$, for all $i \in
+  [1,\#(q)]$ and behaves like $w_\Data$ outside of
+  $\set{y_1,\ldots,y_{\#(d)}}$. Because $I, w_\Data \models \forall
+  y_1 \ldots \forall y_{\#(q)} ~.~
+  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{n}$,
+  we have $\I \in \sem{\stamp{\psi}{n}}_{w_\Data}$ and let $\K$ be one
+  of the minimal interpretations such that $\K \subseteq \I$ and $\K
+  \in \sem{\stamp{\psi}{n}}_{w_\Data}$. It is not hard to see that
+  $\K$ exists and is unique, otherwise we could take the pointwise
+  intersection of two or more such interpretations. We define the
+  interpretation $\bar{\K}(q) = \bar{\K}(\stamp{q}{n})$ for all $q \in
+  Q$. We have that $\bar{\K} \in \minsem{\psi}_{\nu}$ --- if
+  $\bar{\K}$ was not minimal, $\K$ was minimal to start with,
+  contradiction. Then we extend the execution $\G$ by appending to
+  each node labeled with a configuration $q(d_1, \ldots, d_{\#(q)})$
+  the cube $\cube{\bar{\K}}$. Clearly the result is an execution of
+  $\A$ over $w$.
+
+  ``$\supseteq$'' \comment[ri]{todo} \qed}
 
@@
 there is no need to define dualization for formulae of the form $\neg
-q(x_1,\ldots,x_{\#(q)})$. The dualization operation has the following
-property:
-\begin{lemma}\label{lemma:dual}
-  Given a formula $\phi$ and an automaton
-  $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$, we have the equivalence
-  $\Accept{F}{\phi} \equiv \neg(\dual{\Accept{Q\setminus F}{\phi}})$. 
-\end{lemma}
-\proof{ By straightforward induction on the structure of $\phi$. \qed}
+q(x_1,\ldots,x_{\#(q)})$.
+
+%% The dualization operation has the following property:
+%% \begin{lemma}\label{lemma:dual}
+%%   Given a formula $\phi$ and an automaton
+%%   $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$, we have the equivalence
+%%   $\Accept{F}{\phi} \equiv \neg(\dual{\Accept{Q\setminus F}{\phi}})$. 
+%% \end{lemma}
+%% \proof{ By straightforward induction on the structure of $\phi$. \qed}
 

### 2018-07-09 quantifiers/draft.tex
ok

@@
 $\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the
-formula in which all input variables and state predicates are replaced
-by their time-stamped counterparts. Given a sequence of input events
-$\alpha = a_1 \ldots a_n$, the \emph{path formula} of $\alpha$ in $\A$
-is defined as follows:
-\[\pathform{\A}{\alpha} \isdef \stamp{\iota}{0} \wedge 
+formula in which all input variables and state predicates (and only
+those symbols) are replaced by their time-stamped counterparts. Given
+a sequence of input events $\alpha = a_1 \ldots a_n$, the \emph{path
+  formula} of $\alpha$ (in $\A$) is defined as follows:
+\[\pathform{\alpha} \isdef \stamp{\iota}{0} \wedge 
 \bigwedge_{i=1}^n \bigwedge_{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{}
@@
 \stamp{q}{i-1}(y_1,\ldots,y_{\#(q)}) \Longrightarrow \stamp{\psi}{i}\]
-To formalize the relation between the low-level configuration-based
-execution semantics and the symbolic path formulae, consider a word
-$w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any execution forest $\F$ of $\A$
-over $w$ is associated an interpretation $\I_{\F}$ of the set of
-time-stamped predicates $\set{\stamp{q}{i} \mid q \in Q, i \in [n]}$,
-defined as follows:
+The automaton $\A$, to which $\pathform{\alpha}$ refers, will always
+be clear from the context. To formalize the relation between the
+low-level configuration-based execution semantics and the symbolic
+path formulae, consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any
+execution forest $\F$ of $\A$ over $w$ is associated an interpretation
+$\I_{\F}$ of the set of time-stamped predicates $\set{\stamp{q}{i}
+  \mid q \in Q, i \in [n]}$, defined as follows:
 \[\I_{\F}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}}
@@
 \begin{lemma}\label{lemma:path-formula}
-  For any word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have: 
-  \[\minsem{\pathform{\A}{w_\Sigma}}_{w_\Data} = \set{\I_{\F} \mid \F
+  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
+  word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have:
+  \[\minsem{\pathform{w_\Sigma}}_{w_\Data} = \set{\I_{\F} \mid \F
     \text{ is an execution of $\A$ over $w$}}\enspace.\]
@@
 \proof{ ``$\subseteq$'' Let $\I$ be a minimal interpretation such that
-  $\I,w_\Data \models \pathform{\A}{w_\Sigma}$. We show that there
-  exists an execution $\F$ of $\A$ over $w$ such that $\I = \I_\F$, by
+  $\I,w_\Data \models \pathform{w_\Sigma}$. We show that there exists
+  an execution $\F$ of $\A$ over $w$ such that $\I = \I_\F$, by
   induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
-  $\pathform{\A}{w_\Sigma} = \stamp{\iota}{0}$. Because $\iota$ is a
+  $\pathform{w_\Sigma} = \stamp{\iota}{0}$. Because $\iota$ is a
   sentence, the valuation $w_\Data$ is not important in $\I,w_\Data
@@
   Q$ and $i \in [n-1]$, and $\emptyset$ everywhere else. Then
-  $\J,u_\Data \models \pathform{\A}{u_\Sigma}$ and $\J$ is moreover
+  $\J,u_\Data \models \pathform{u_\Sigma}$ and $\J$ is moreover
   minimal. By the induction hypothesis, there exists an execution $\G$
@@
   \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{n}$ be
-  the subformula of $\pathform{\A}{w_\Sigma}$ corresponding to the
+  the subformula of $\pathform{w_\Sigma}$ corresponding to the
   application(s) of the transition rule $q(y_1,\ldots,y_{\#(q)})
@@
   [1,\#(q)]$ and behaves like $w_\Data$ outside of
-  $\set{y_1,\ldots,y_{\#(d)}}$. Because $I, w_\Data \models \forall
+  $\set{y_1,\ldots,y_{\#(d)}}$. Because $\I, w_\Data \models \forall
   y_1 \ldots \forall y_{\#(q)} ~.~
   \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{n}$,
-  we have $\I \in \sem{\stamp{\psi}{n}}_{w_\Data}$ and let $\K$ be one
-  of the minimal interpretations such that $\K \subseteq \I$ and $\K
-  \in \sem{\stamp{\psi}{n}}_{w_\Data}$. It is not hard to see that
-  $\K$ exists and is unique, otherwise we could take the pointwise
-  intersection of two or more such interpretations. We define the
-  interpretation $\bar{\K}(q) = \bar{\K}(\stamp{q}{n})$ for all $q \in
-  Q$. We have that $\bar{\K} \in \minsem{\psi}_{\nu}$ --- if
-  $\bar{\K}$ was not minimal, $\K$ was minimal to start with,
+  we have $\I \in \sem{\stamp{\psi}{n}}_{\nu}$ and let $\K$ be one of
+  the minimal interpretations such that $\K \subseteq \I$ and $\K \in
+  \sem{\stamp{\psi}{n}}_{\nu}$. It is not hard to see that $\K$ exists
+  and is unique, otherwise we could take the pointwise intersection of
+  two or more such interpretations. We define the interpretation
+  $\overline{\K}(q) = \overline{\K}(\stamp{q}{n})$ for all $q \in
+  Q$. We have that $\overline{\K} \in \minsem{\psi}_{\nu}$ --- if
+  $\overline{\K}$ was not minimal, $\K$ was not minimal to start with,
   contradiction. Then we extend the execution $\G$ by appending to
   each node labeled with a configuration $q(d_1, \ldots, d_{\#(q)})$
-  the cube $\cube{\bar{\K}}$. Clearly the result is an execution of
-  $\A$ over $w$.
+  the cube $\cube{\overline{\K}}$. By repeating this step for all
+  leaves of a tree in $\G$, we obtain an execution of $\A$ over $w$.
 
-  ``$\supseteq$'' \comment[ri]{todo} \qed}
+  ``$\supseteq$'' Let $\F$ be an execution of $\A$ over $w$. We show
+  that $\I_{\F}$ is a minimal interpretation such that $\I_{\F},
+  w_\Data \models \pathform{w_\Sigma}$, by induction on $n \geq
+  0$. For $n=0$, $\F$ is a cube from $\cube{\minsem{\iota}}$, by
+  definition. Then $\I_{\F} \models \stamp{\iota}{0}$ and moreover, it
+  is a minimal such interpretation. For the inductive case $n > 0$,
+  let $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\G$ be the
+  restriction of $\F$ to $u$. Consequently, $\I_\G$ is the restriction
+  of $\I_\F$ to $\bigcup_{i=0}^{n-1} \stamp{Q}{i}$. By the inductive
+  hypothesis, $\I_{\G}$ is a minimal interpretation such that
+  \(\I_{\G}, u_{\Data} \models \pathform{u_\Sigma}\). Since
+  $\I_{\F}(\stamp{q}{n}) = \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
+    q(d_1,\ldots,d_{\#(q)}) \text{ occurs on the $n$-th level in
+      $\F$}}$, we have $\I_{\F}, w_\Data \models \varphi$, for each
+  subformula $\varphi = \forall y_1 \ldots \forall y_{\#(q)} ~.~
+  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{n}$ of
+  $\pathform{w_\Sigma}$, by the execution semantics of $\A$. This is
+  the case because the children of each node labeled with
+  $q(d_1,\ldots,d_{\#(q)})$ on the $(n-1)$-th level of $\F$ form a
+  cube from $\cube{\minsem{\psi}_\nu}$, where $\nu$ is a valuation
+  that assigns each $y_i$ the value $d_i$ and behaves like $w_\Data$,
+  otherwise. Now supppose, for a contradiction, that $\I_{\F}$ is not
+  minimal and let $\J \subsetneq \I_{\F}$ be an interpretation such
+  that $\J,w_\Data \models \pathform{w_\Sigma}$. First, we show that
+  the restriction $\J'$ of $\J$ to $\bigcup_{i=0}^{n-1} \stamp{Q}{i}$
+  must coincide with $\I_{\G}$. Assuming this is not the case,
+  i.e.\ $\J' \subsetneq \I_{\G}$, contradicts the minimality of
+  $\I_{\G}$. Then the only possibility is that $\J(\stamp{q}{n})
+  \subsetneq \I_{\F}(\stamp{q}{n})$, for some $q \in Q$. Let
+  $p_1(y_1,\ldots,y_{\#(p_1)}) \arrow{a_n}{} \psi_1, \ldots,
+  p_k(y_1,\ldots,y_{\#(p_k)}) \arrow{a_n}{} \psi_k$ be the set of
+  transition rules in which the predicate symbol $q$ occurs. Then it
+  must be the case that, for some node $p_i(d_1,\ldots,d_{\#(p_i)})$
+  on the $(n-1)$-th level of $\G$, the set of children does not form a
+  minimal cube from $\cube{\minsem{\stamp{\psi_i}{n}}}$, which
+  contradicts the execution semantics of $\A$. \qed}
 

### 2018-07-10 quantifiers/draft.tex
ok

@@
 variables, $Q$ is a finite set of predicates, $\iota$ is a satisfiable
-sentence, $F \subseteq Q$ is a set of \emph{final predicates} and
+sentence, $F \subseteq Q$ is the set of \emph{final} predicates and
 $\Delta$ is a set of \emph{transition rules} of the form
@@
 \emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
-\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}} \len{\psi}$.
+\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
+\len{\psi}$.
 
@@
 $\cube{\mathcal{S}} \isdef \set{\cube{\I} \mid \I \in
-  \mathcal{S}}$. Given a word $w=(a_1,\nu_1)(a_2,\nu_2) \ldots
-(a_n,\nu_n) \in \Sigma[X]^*$, an \emph{execution} of
+  \mathcal{S}}$. Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in
+\Sigma[X]^*$, an \emph{execution} of
 $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is a (possibly
@@
 \end{compactitem}
-An execution $\F$ is said to be \emph{accepting} if and only if the
-frontier of each tree $T \in \F$ is labeled with configurations
-$q(d_1,\ldots,d_{\#(q)})$ such that $q \in F$ is a final predicate. In
-this case, we say that $\A$ accepts $w$ and denote by $\lang{\A}$ the
-set of words accepted by $\A$. In this paper, we answer the following
-questions: \begin{compactenum}
+An execution $\F$ over $w$ is said to be \emph{accepting} if and only
+if \begin{inparaenum}[(i)]
+\item all paths in $\F$ have the same length $n$, and
+\item the frontier of each tree $T \in \F$ is labeled with \emph{final
+  configurations} $q(d_1,\ldots,d_{\#(q)})$, where $q \in F$ is a
+  final predicate.
+\end{inparaenum} In this case, we say that $\A$ accepts $w$ and
+denote by $\lang{\A}$ the set of words accepted by $\A$.
+
+In this paper, we answer the following questions: \begin{compactenum}
 \item \emph{closure}: given automata $\A_i =
@@
   Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
-  word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have:
-  \[\minsem{\pathform{w_\Sigma}}_{w_\Data} = \set{\I_{\F} \mid \F
-    \text{ is an execution of $\A$ over $w$}}\enspace.\]
+  word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have 
+  \(\minsem{\pathform{w_\Sigma}}_{w_\Data} = \set{\I_{\F} \mid \F
+    \text{ is an execution of $\A$ over $w$}}\). 
 \end{lemma}
@@
 
+Next, we give a logical characterization of acceptance, relative to a
+given sequence of input events.
+
+\begin{lemma}\label{lemma:acceptance}
+  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ and a
+  sequence of input events $\alpha = a_1 \ldots a_n \in \Sigma^*$, the
+  formula $\accform{\alpha} = \pathform{\alpha} \wedge \bigwedge_{q
+    \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
+  \stamp{q}{n}(y_1,\ldots,y_{\#(q)}) \Rightarrow \bot$ is satisfiable
+  if and only if $\A$ accepts a word $w$ such that $w_\Sigma =
+  \alpha$.
+\end{lemma}
+\proof{ ``$\Rightarrow$'' Let $\I$ be an interpretation and $\nu$ be a
+  valuation such that $\I,\nu \models \accform{\alpha}$. W.l.o.g. we
+  consider $\I$ to be a minimal interpretation with this property. Let
+  $w$ be the unique word such that $w_\Sigma = \alpha$ and $w_\Data =
+  \nu$. By Lemma \ref{lemma:path-formula}, $\A$ has an execution $\F$
+  over $w$ such that $\I = \I_{\F}$. To prove that $\F$ is accepting,
+  we show that \begin{inparaenum}[(i)]
+    \item\label{it1:acceptance} all paths in $\F$ have length $n$ and
+      that
+    \item\label{it2:acceptance} the frontier of $\F$ is labeled with
+      final configurations only. \end{inparaenum} First, assume that
+  (\ref{it1:acceptance}) there exists a path in $\F$ of length $0 \leq
+  m < n$. Then there exists a node on the $m$-th level, labeled with
+  some configuration $q(d_1,\ldots,q_{\#(q)})$, that has no
+  children. By the definition of the execution semantics of $\A$, we
+  have $\cube{\minsem{\psi}_{\eta}} = \emptyset$, where
+  $q(y_1,\ldots,y_{\#(q)}) \arrow{a_{m+1}(X)}{} \psi$ is the transition
+  rule of $\A$ that applies for $q$ and $a_{m+1}$ and $\eta$ stands
+  for $\nu[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow
+    d_{\#(q)}]$. Hence $\sem{\psi}_{\eta} = \emptyset$, and because
+  $\I,\nu \models \accform{\alpha}$, we obtain that $\I,\eta \models
+  q(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{m+1}$, thus
+  $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$. However, this
+  contradicts the fact that $\I=\I_{\F}$ and that
+  $q(d_1,\ldots,d_{\#(q)})$ labels a node of $\F$. Second, assume that
+  (\ref{it2:acceptance}), there exists a frontier node of $\F$ labeled
+  with a configuration $q(d_1,\ldots,d_{\#(q)})$ such that $q \in Q
+  \setminus F$. Since $\I,\nu \models \forall y_1 \ldots \forall
+  y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \Rightarrow \bot$, by a
+  similar reasoning as in the above case, we obtain that
+  $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$, contradiction.
+
+  ``$\Leftarrow$'' Let $\F$ be an accepting execution of $\A$ over
+  $w$, where $w_\Sigma = \alpha$. We prove that $\I_\F,w_\Data \models
+  \accform{\alpha}$. By Lemma \ref{lemma:path-formula}, we obtain
+  $\I_\F,w_\Data \models \pathform{\alpha}$. Since every path in $\F$
+  is of length $n$ and all nodes on the $n$-th level of $\F$ are
+  labeled by final configurations, we obtain that $\I_F,w_\Data
+  \models \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
+  y_{\#(q)} ~.~ \stamp{q}{n}(y_1,\ldots,y_{\#(q)}) \Rightarrow \bot$,
+  trivially.  \qed}
+
 \subsection{Closure Properties}
@@
 Given a positive formula $\phi$, we define the \emph{dual} formula
-$\dual{\phi}$ recursively on the structure of $\phi$ as follows: 
+$\dual{\phi}$ recursively as follows:
 \[\begin{array}{rclcrclcrcl}
@@
       %
-    \item\label{it2:closure} $\lang{\overline{\A_i}} = \Sigma[X]^* \setminus
-      \lang{\A_i}$, where $\overline{\A_i} =
+    \item\label{it2:closure} $\lang{\overline{\A_i}} = \Sigma[X]^*
+      \setminus \lang{\A_i}$, where $\overline{\A_i} =
       \tuple{\Sigma,X,Q_i,\dual{\iota},Q_i\setminus
-        F_i,\overline{\Delta_i}}$ and, for all
-      $i=1,2$: \[q(x_1,\ldots,x_{\#(q)}) \arrow{a(X)}{} \phi \in
-      \Delta_i \iff q(x_1,\ldots,x_{\#(q)}) \arrow{a(X)}{} \dual{\phi}
-      \in \overline{\Delta_i} \enspace.\]
+        F_i,\dual{\Delta}_i}$ and, for $i=1,2$: \[\dual{\Delta}_i =
+      \set{q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \dual{\psi} \mid
+        q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi \in
+        \Delta_i}\enspace.\]
   \end{compactitem}
@@
 \proof{\qed}
+
+\section{The Emptiness Problem}
+
+The problem of checking emptiness of a given automata is undecidable,
+for automata using predicates of arity two or more, even if the theory
+in which the (right-hand sides of the) transition rules are written
+consists of only equalities and disequalities between variables, with
+no quantifiers \cite{FarzanKincaidPodelski15}. Thus, we resort to an
+abstraction-refinement loop that looks for an accepted word, by firing
+longer and longer sequences of input events.
+
+Since, for a countable data domain $\Data$, the set of finite words
+$\Sigma[X]^*$ over a finite set of variables $X$ is countable, the
+emptiness problem is recursively enumerable in this case. As a
+consequence of Lemma \ref{lemma:acceptance}, if the given automaton is
+not empty, a semi-algorithm based on a systematic exploration of the
+sequences of input events will eventually discover an accepted word,
+provided that the first-order data theory is decidable.
+
+However, if the automaton is empty, the enumeration of input event
+sequences will fail to terminate, in general. The classical way to
+fight this divergence problem is to consider a \emph{coverage} partial
+order, which is a subset of the following relation:
+
+\begin{definition}\label{def:coverage}
+  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ and
+  $S_1, S_2 \subseteq \bigcup_{q \in Q} \set{q(d_1,\ldots,d_{\#(q)})
+    \mid d_1,\ldots,d_{\#(q)} \in \Data}$, we have $S_1 \sqsubseteq
+  S_2$ if and only if 
+  \begin{equation}\label{eq:po}
+    \text{
+      \begin{minipage}{0.7\textwidth}
+        $\A$ accepts a word $w$ when started in a set of cubes $C_1
+        \in S_1$ only if $\A$ accepts $w$ when started in a set of
+        cubes $C_2 \in S_2$.
+    \end{minipage}} 
+  \end{equation}
+\end{definition}
+Since, in general the $\sqsubseteq$ relation is not definable using
+available decidable (SMT) theories, we consider the following stronger
+relation: $S_1 \supsetforex S_2$ if and only if, for each set of cubes
+$C_1 \in S_1$, there exists a set of cubes $C_2 \in S_2$ such that
+$C_1 \supseteq C_2$.
+
+\begin{lemma}\label{lemma:coverage}
+  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
+  sets of cubes $S_1, S_2 \subseteq \bigcup_{q \in Q}
+  \set{q(d_1,\ldots,d_{\#(q)}) \mid d_1,\ldots,d_{\#(q)} \in \Data}$
+  such that $S_1 \supsetforex S_2$, we have $S_1 \sqsubseteq S_2$.
+\end{lemma}
+\proof{  
+\qed}
 

### 2018-07-17 quantifiers/draft.tex
ok

@@
 execution forest $\F$ of $\A$ over $w$ is associated an interpretation
-$\I_{\F}$ of the set of time-stamped predicates $\set{\stamp{q}{i}
-  \mid q \in Q, i \in [n]}$, defined as follows:
-\[\I_{\F}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}}
-  \mid q(d_1,\ldots,d_{\#(q)}) \text{ occurs on level $i$ in
-    $\F$}}\enspace.\]
+$\I_{\F}$ of the set of time-stamped predicates $\stamp{Q}{\leq n}
+\isdef \set{\stamp{q}{i} \mid q \in Q, i \in [n]}$, defined as: 
+\[\I_{\F}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
+  q(d_1,\ldots,d_{\#(q)}) \text{ occurs on level $i$ in $\F$}}\]
 
@@
   restriction of $\F$ to $u$. Consequently, $\I_\G$ is the restriction
-  of $\I_\F$ to $\bigcup_{i=0}^{n-1} \stamp{Q}{i}$. By the inductive
-  hypothesis, $\I_{\G}$ is a minimal interpretation such that
-  \(\I_{\G}, u_{\Data} \models \pathform{u_\Sigma}\). Since
-  $\I_{\F}(\stamp{q}{n}) = \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
-    q(d_1,\ldots,d_{\#(q)}) \text{ occurs on the $n$-th level in
-      $\F$}}$, we have $\I_{\F}, w_\Data \models \varphi$, for each
-  subformula $\varphi = \forall y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{n}$ of
-  $\pathform{w_\Sigma}$, by the execution semantics of $\A$. This is
-  the case because the children of each node labeled with
-  $q(d_1,\ldots,d_{\#(q)})$ on the $(n-1)$-th level of $\F$ form a
-  cube from $\cube{\minsem{\psi}_\nu}$, where $\nu$ is a valuation
-  that assigns each $y_i$ the value $d_i$ and behaves like $w_\Data$,
+  of $\I_\F$ to $\stamp{Q}{\leq n-1}$. By the inductive hypothesis,
+  $\I_{\G}$ is a minimal interpretation such that \(\I_{\G}, u_{\Data}
+  \models \pathform{u_\Sigma}\). Since $\I_{\F}(\stamp{q}{n}) =
+  \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid q(d_1,\ldots,d_{\#(q)})
+    \text{ occurs on the $n$-th level in $\F$}}$, we have $\I_{\F},
+  w_\Data \models \varphi$, for each subformula $\varphi = \forall y_1
+  \ldots \forall y_{\#(q)} ~.~ \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)})
+  \Rightarrow \stamp{\psi}{n}$ of $\pathform{w_\Sigma}$, by the
+  execution semantics of $\A$. This is the case because the children
+  of each node labeled with $q(d_1,\ldots,d_{\#(q)})$ on the
+  $(n-1)$-th level of $\F$ form a cube from
+  $\cube{\minsem{\psi}_\nu}$, where $\nu$ is a valuation that assigns
+  each $y_i$ the value $d_i$ and behaves like $w_\Data$,
   otherwise. Now supppose, for a contradiction, that $\I_{\F}$ is not
@@
   p_k(y_1,\ldots,y_{\#(p_k)}) \arrow{a_n}{} \psi_k$ be the set of
-  transition rules in which the predicate symbol $q$ occurs. Then it
-  must be the case that, for some node $p_i(d_1,\ldots,d_{\#(p_i)})$
-  on the $(n-1)$-th level of $\G$, the set of children does not form a
+  transition rules in which the predicate symbol $q$ occurs on the
+  right-hand side. Then it must be the case that, for some node on the
+  $(n-1)$-th level of $\G$, labeled with a configuration
+  $p_i(d_1,\ldots,d_{\#(p_i)})$, the set of children does not form a
   minimal cube from $\cube{\minsem{\stamp{\psi_i}{n}}}$, which
@@
 The problem of checking emptiness of a given automata is undecidable,
-for automata using predicates of arity two or more, even if the theory
-in which the (right-hand sides of the) transition rules are written
-consists of only equalities and disequalities between variables, with
-no quantifiers \cite{FarzanKincaidPodelski15}. Thus, we resort to an
-abstraction-refinement loop that looks for an accepted word, by firing
-longer and longer sequences of input events.
+for automata using predicates of arity two or more, even only when the
+theory in which the (right-hand sides of the) transition rules are
+written consists of only equalities and disequalities between
+variables, with no quantifiers \cite{FarzanKincaidPodelski15}.
+Consequently, we cannot tackle the emptiness problem via a complete
+decision procedure and resort to an abstraction-refinement loop that
+tries to prove non-emptiess by discovering a word that it accepted,
+firing longer and longer sequences of input events.
 
-Since, for a countable data domain $\Data$, the set of finite words
-$\Sigma[X]^*$ over a finite set of variables $X$ is countable, the
-emptiness problem is recursively enumerable in this case. As a
-consequence of Lemma \ref{lemma:acceptance}, if the given automaton is
-not empty, a semi-algorithm based on a systematic exploration of the
-sequences of input events will eventually discover an accepted word,
-provided that the first-order data theory is decidable.
+Assuming that the data domain is countable, the set of finite words
+$\Sigma[X]^*$ is countable and the emptiness problem is recursively
+enumerable. Moreover, by Lemma \ref{lemma:acceptance}, a systematic
+enumeration of sequences of input events $\alpha \in \Sigma^*$ will
+eventually discover an accepted word, provided that all path formulae
+$\pathform{\alpha}$ belong to a decidable first-order theory.
 

### 2018-07-24 quantifiers/draft.tex
ok

@@
 $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is a (possibly
-infinite) forest $\F = \set{T_1,T_2,\ldots}$, where each tree $T_i$ is
+infinite) forest $\T = \set{T_1,T_2,\ldots}$, where each tree $T_i$ is
 such that:
@@
 \end{compactitem}
-An execution $\F$ over $w$ is said to be \emph{accepting} if and only
+An execution $\T$ over $w$ is said to be \emph{accepting} if and only
 if \begin{inparaenum}[(i)]
-\item all paths in $\F$ have the same length $n$, and
-\item the frontier of each tree $T \in \F$ is labeled with \emph{final
+\item all paths in $\T$ have the same length $n$, and
+\item the frontier of each tree $T \in \T$ is labeled with \emph{final
   configurations} $q(d_1,\ldots,d_{\#(q)})$, where $q \in F$ is a
@@
 \end{inparaenum} In this case, we say that $\A$ accepts $w$ and
-denote by $\lang{\A}$ the set of words accepted by $\A$.
-
-In this paper, we answer the following questions: \begin{compactenum}
-\item \emph{closure}: given automata $\A_i =
+denote by $\lang{\A}$ the set of words accepted by $\A$. We address
+the following questions: \begin{compactenum}
+\item \emph{boolean closure}: given automata $\A_i =
   \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
@@
   \mid x \in X}$ the sets of time-stamped predicates and variables,
-respectively. For a formula $\psi$ and $i \in \nat$, we define
-$\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the
-formula in which all input variables and state predicates (and only
-those symbols) are replaced by their time-stamped counterparts. Given
-a sequence of input events $\alpha = a_1 \ldots a_n$, the \emph{path
-  formula} of $\alpha$ (in $\A$) is defined as follows:
+respectively. As a shorthand, we write $\stamp{Q}{\leq n}$ (resp.
+$\stamp{X}{\leq n}$) for the set $\set{\stamp{q}{i} \mid q \in Q, i
+  \in [n]}$ (resp. $\set{\stamp{x}{i} \mid x \in X, i \in [n]}$). For
+a formula $\psi$ and $i \in \nat$, we define $\stamp{\psi}{i} \isdef
+\psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the formula in which all input
+variables and state predicates (and only those symbols) are replaced
+by their time-stamped counterparts. For a sequence of input events
+$\alpha = a_1 \ldots a_n$, the \emph{path formula} of $\alpha$ (in
+$\A$) is defined as:
 \[\pathform{\alpha} \isdef \stamp{\iota}{0} \wedge 
@@
 path formulae, consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any
-execution forest $\F$ of $\A$ over $w$ is associated an interpretation
-$\I_{\F}$ of the set of time-stamped predicates $\stamp{Q}{\leq n}
-\isdef \set{\stamp{q}{i} \mid q \in Q, i \in [n]}$, defined as: 
-\[\I_{\F}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
-  q(d_1,\ldots,d_{\#(q)}) \text{ occurs on level $i$ in $\F$}}\]
+execution forest $\T$ of $\A$ over $w$ is associated an interpretation
+$\I_{\T}$ of the set of time-stamped predicates $\stamp{Q}{\leq n}$,
+defined as:
+\[\I_{\T}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
+  q(d_1,\ldots,d_{\#(q)}) \text{ occurs on level $i$ in $\T$}}\]
 
@@
   word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have 
-  \(\minsem{\pathform{w_\Sigma}}_{w_\Data} = \set{\I_{\F} \mid \F
+  \(\minsem{\pathform{w_\Sigma}}_{w_\Data} = \set{\I_{\T} \mid \T
     \text{ is an execution of $\A$ over $w$}}\). 
@@
   $\I,w_\Data \models \pathform{w_\Sigma}$. We show that there exists
-  an execution $\F$ of $\A$ over $w$ such that $\I = \I_\F$, by
+  an execution $\T$ of $\A$ over $w$ such that $\I = \I_\T$, by
   induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
@@
 
-  ``$\supseteq$'' Let $\F$ be an execution of $\A$ over $w$. We show
-  that $\I_{\F}$ is a minimal interpretation such that $\I_{\F},
+  ``$\supseteq$'' Let $\T$ be an execution of $\A$ over $w$. We show
+  that $\I_{\T}$ is a minimal interpretation such that $\I_{\T},
   w_\Data \models \pathform{w_\Sigma}$, by induction on $n \geq
-  0$. For $n=0$, $\F$ is a cube from $\cube{\minsem{\iota}}$, by
-  definition. Then $\I_{\F} \models \stamp{\iota}{0}$ and moreover, it
+  0$. For $n=0$, $\T$ is a cube from $\cube{\minsem{\iota}}$, by
+  definition. Then $\I_{\T} \models \stamp{\iota}{0}$ and moreover, it
   is a minimal such interpretation. For the inductive case $n > 0$,
   let $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\G$ be the
-  restriction of $\F$ to $u$. Consequently, $\I_\G$ is the restriction
-  of $\I_\F$ to $\stamp{Q}{\leq n-1}$. By the inductive hypothesis,
+  restriction of $\T$ to $u$. Consequently, $\I_\G$ is the restriction
+  of $\I_\T$ to $\stamp{Q}{\leq n-1}$. By the inductive hypothesis,
   $\I_{\G}$ is a minimal interpretation such that \(\I_{\G}, u_{\Data}
-  \models \pathform{u_\Sigma}\). Since $\I_{\F}(\stamp{q}{n}) =
+  \models \pathform{u_\Sigma}\). Since $\I_{\T}(\stamp{q}{n}) =
   \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid q(d_1,\ldots,d_{\#(q)})
-    \text{ occurs on the $n$-th level in $\F$}}$, we have $\I_{\F},
+    \text{ occurs on the $n$-th level in $\T$}}$, we have $\I_{\T},
   w_\Data \models \varphi$, for each subformula $\varphi = \forall y_1
@@
   of each node labeled with $q(d_1,\ldots,d_{\#(q)})$ on the
-  $(n-1)$-th level of $\F$ form a cube from
+  $(n-1)$-th level of $\T$ form a cube from
   $\cube{\minsem{\psi}_\nu}$, where $\nu$ is a valuation that assigns
   each $y_i$ the value $d_i$ and behaves like $w_\Data$,
-  otherwise. Now supppose, for a contradiction, that $\I_{\F}$ is not
-  minimal and let $\J \subsetneq \I_{\F}$ be an interpretation such
+  otherwise. Now supppose, for a contradiction, that $\I_{\T}$ is not
+  minimal and let $\J \subsetneq \I_{\T}$ be an interpretation such
   that $\J,w_\Data \models \pathform{w_\Sigma}$. First, we show that
@@
   $\I_{\G}$. Then the only possibility is that $\J(\stamp{q}{n})
-  \subsetneq \I_{\F}(\stamp{q}{n})$, for some $q \in Q$. Let
+  \subsetneq \I_{\T}(\stamp{q}{n})$, for some $q \in Q$. Let
   $p_1(y_1,\ldots,y_{\#(p_1)}) \arrow{a_n}{} \psi_1, \ldots,
@@
 Next, we give a logical characterization of acceptance, relative to a
-given sequence of input events.
+given sequence of input events $\alpha \in \Sigma^*$. To this end, we
+extend the path formula $\pathform{\alpha}$ with a characterization of
+the set of final states of $\A$:
+
+\[\accform{\alpha} \isdef \pathform{\alpha} \wedge \bigwedge_{q
+  \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
+\stamp{q}{n}(y_1,\ldots,y_{\#(q)}) \Rightarrow \bot\]
 
 \begin{lemma}\label{lemma:acceptance}
-  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ and a
-  sequence of input events $\alpha = a_1 \ldots a_n \in \Sigma^*$, the
-  formula $\accform{\alpha} = \pathform{\alpha} \wedge \bigwedge_{q
-    \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{n}(y_1,\ldots,y_{\#(q)}) \Rightarrow \bot$ is satisfiable
-  if and only if $\A$ accepts a word $w$ such that $w_\Sigma =
-  \alpha$.
+  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for
+  every word $w \in \Sigma[X]^*$, the following are
+  equivalent: \begin{compactenum}
+  \item\label{it1:lemma:acceptance} there exists an interpretation
+    $\I$ such that $\I,w_\Data \models \accform{w_\Sigma}$, 
+    %
+  \item\label{it2:lemma:acceptance} $w \in \lang{\A}$. 
+  \end{compactenum}
 \end{lemma}
-\proof{ ``$\Rightarrow$'' Let $\I$ be an interpretation and $\nu$ be a
-  valuation such that $\I,\nu \models \accform{\alpha}$. W.l.o.g. we
-  consider $\I$ to be a minimal interpretation with this property. Let
-  $w$ be the unique word such that $w_\Sigma = \alpha$ and $w_\Data =
-  \nu$. By Lemma \ref{lemma:path-formula}, $\A$ has an execution $\F$
-  over $w$ such that $\I = \I_{\F}$. To prove that $\F$ is accepting,
-  we show that \begin{inparaenum}[(i)]
-    \item\label{it1:acceptance} all paths in $\F$ have length $n$ and
+\proof{ ``$(\ref{it1:lemma:acceptance}) \Rightarrow
+  (\ref{it2:lemma:acceptance})$'' Let $\I$ be an interpretation such
+  that $\I,w_\Data \models \accform{w_\Sigma}$. By Lemma
+  \ref{lemma:path-formula}, $\A$ has an execution $\T$ over $w$ such
+  that $\I = \I_{\T}$. To prove that $\T$ is accepting, we show
+  that \begin{inparaenum}[(i)]
+    \item\label{it1:acceptance} all paths in $\T$ have length $n$ and
       that
-    \item\label{it2:acceptance} the frontier of $\F$ is labeled with
+    \item\label{it2:acceptance} the frontier of $\T$ is labeled with
       final configurations only. \end{inparaenum} First, assume that
-  (\ref{it1:acceptance}) there exists a path in $\F$ of length $0 \leq
+  (\ref{it1:acceptance}) there exists a path in $\T$ of length $0 \leq
   m < n$. Then there exists a node on the $m$-th level, labeled with
@@
   $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$. However, this
-  contradicts the fact that $\I=\I_{\F}$ and that
-  $q(d_1,\ldots,d_{\#(q)})$ labels a node of $\F$. Second, assume that
-  (\ref{it2:acceptance}), there exists a frontier node of $\F$ labeled
+  contradicts the fact that $\I=\I_{\T}$ and that
+  $q(d_1,\ldots,d_{\#(q)})$ labels a node of $\T$. Second, assume that
+  (\ref{it2:acceptance}), there exists a frontier node of $\T$ labeled
   with a configuration $q(d_1,\ldots,d_{\#(q)})$ such that $q \in Q
@@
 
-  ``$\Leftarrow$'' Let $\F$ be an accepting execution of $\A$ over
-  $w$, where $w_\Sigma = \alpha$. We prove that $\I_\F,w_\Data \models
-  \accform{\alpha}$. By Lemma \ref{lemma:path-formula}, we obtain
-  $\I_\F,w_\Data \models \pathform{\alpha}$. Since every path in $\F$
-  is of length $n$ and all nodes on the $n$-th level of $\F$ are
-  labeled by final configurations, we obtain that $\I_F,w_\Data
+  ``$(\ref{it2:lemma:acceptance}) \Rightarrow
+  (\ref{it1:lemma:acceptance})$'' Let $\T$ be an accepting execution
+  of $\A$ over $w$. We prove that $\I_\T,w_\Data \models
+  \accform{w_\Sigma}$. By Lemma \ref{lemma:path-formula}, we obtain
+  $\I_\T,w_\Data \models \pathform{w_\Sigma}$. Since every path in
+  $\T$ is of length $n$ and all nodes on the $n$-th level of $\T$ are
+  labeled by final configurations, we obtain that $\I_\T,w_\Data
   \models \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
@@
   trivially.  \qed}
+
+In the following, it will be useful to define the set of data
+valuations of a word by a formula without predicates. Given a sequence
+of input events $\alpha = a_1 \ldots a_n$, let $\substform{\alpha}$ be
+the formula obtained from $\pathform{\alpha}$, by successively
+replacing in $\accform{\alpha}$: \begin{compactitem}
+\item each occurrence of a predicate
+  $\stamp{q}{i}(t_1,\ldots,t_{\#(q)})$ by the formula
+  $\stamp{\psi}{i+1}[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$, where
+  $q(y_1,\ldots,y_{\#(q)}) \arrow{a_{i+1}(X)}{} \psi \in \Delta$ is the
+  unique transition rule for $q$ and $a_i$, for all $i=0,\ldots,n-1$,
+%
+\item each occurrence of a predicate $\stamp{q}{n}$, such that $q \in
+  Q \setminus F$, by $\top$, and
+%
+\item each occurrence of a predicate $\stamp{q}{n}$, such that $q \in
+  F$, by $\bot$.
+\end{compactitem}
+Since there are no occurrences of predicates in $\substform{\alpha}$,
+for each valuation $\nu : \stamp{X}{\leq n} \rightarrow \Data$, there
+exists an interpretation $\I$ such that $\I,\nu \models
+\substform{\alpha}$ if and only if $\J,\nu \models \substform{\alpha}$, 
+for every interpretation $\J$. Thus we omit $\I$ and simply write $\nu
+\models \substform{\alpha}$ in this case.
+
+\begin{proposition}\label{prop:subst}
+  Let $\phi$ be a positive formula and let $q(t_1,\ldots,t_{\#(q)})$
+  be the only predicate atom occurring within $\phi$. Then, for every
+  valuation $\nu$ of $\fv{}{\phi}$ and every formula $\psi$ with no
+  predicate occurrences, the following are
+  equivalent: \begin{compactenum}
+  \item\label{it1:prop:subst} $\nu \models \phi[\psi[t_1/y_1, \ldots,
+      t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$, 
+    %
+  \item\label{it2:prop:subst} $\I,\nu \models \phi \wedge \forall y_1
+    \ldots \forall y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow
+    \psi$, for some interpretation $\I$.
+  \end{compactenum}
+\end{proposition}
+\proof{ We give the proof in the case $\phi$ is quantifier-free. The
+  generalization to a non-empty quantifier prefix is straightforward.
+  ``$(\ref{it1:prop:subst}) \Rightarrow (\ref{it2:prop:subst})$'' We
+  distinguish two cases:
+  \begin{compactitem}
+  \item if $\nu \models \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$
+    then let $\I(q) = \set{\tuple{t_1^\nu,\ldots,t_{\#(q)}^\nu}}$ and
+    $\emptyset$ everywhere else. We have $\I,\nu \models
+    q(t_1,\ldots,t_{\#(q)})$ and, since $\phi$ is positive, replacing
+    $\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ with
+    $q(t_1,\ldots,t_{\#(q)})$ in $\phi$ does not change its
+    satisfiability under $\I,\nu$, hence $\I,\nu \models
+    \phi$. Moreover, $\nu \models \psi[t_1/y_1, \ldots,
+      t_{\#(q)}/y_{\#(q)}]$ thus $\I,\nu \models \forall y_1 \ldots
+    \forall y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$.
+    %
+  \item else $\nu \not\models \psi[t_1/y_1, \ldots,
+    t_{\#(q)}/y_{\#(q)}]$ and let $\I$ be the interpretation assigning
+    each predicate symbol with $\emptyset$. Then replacing
+    $\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ with
+    $q(t_1,\ldots,t_{\#(q)})$ in $\phi$ does not change its
+    satisfiability under $\I,\nu$. Moreover, since $\I$ maps $q$ to
+    $\emptyset$, we have $\I,\nu \models \forall y_1 \ldots \forall
+    y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$, trivially.
+  \end{compactitem}
+
+  ``$(\ref{it2:prop:subst}) \Rightarrow (\ref{it1:prop:subst})$'' We
+  distinguish two cases: \begin{compactitem}
+  \item if $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \in \I(q)$, since
+    $\I,\nu \models \forall y_1 \ldots \forall y_{\#(q)} ~.~
+    q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$, we have $\nu \models
+    \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$. Since $\phi$ is
+    positive, replacing $q(t_1,\ldots,t_{\#(q)})$ by $\psi$ in $\phi$
+    does not change its satisfiability under $\nu$, hence $\nu \models
+    \phi[\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]/q(t_1, \ldots,
+      t_{\#(q)})]$.
+    %
+  \item else $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \not\in \I(q)$,
+    that is $\I,\nu \not\models q(t_1,\ldots,t_{\#(q)})$. Then $\nu
+    \models \phi[\bot/q(t_1,\ldots,t_{\#(q)})]$ and we obtain $\nu
+    \models \phi[\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]/q(t_1,
+      \ldots, t_{\#(q)})]$, because $\phi$ is positive, thus
+    $\phi[\bot/q(t_1,\ldots,t_{\#(q)})] \models \phi[\psi[t_1/y_1,
+        \ldots, t_{\#(q)}/y_{\#(q)}]/q(t_1, \ldots, t_{\#(q)})]$. \qed
+  \end{compactitem}}
+
+\begin{lemma}\label{lemma:substitution}
+  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for
+  every word $w \in \Sigma[X]^*$, we have $w_\Data \models
+  \substform{w_\Sigma}$ if and only if $w \in \lang{\A}$.
+\end{lemma}
+\proof{ By Lemma \ref{lemma:acceptance}, $w \in \lang{\A}$ if and only
+  if $\I,w_\Data \models \accform{w_\Sigma}$, for some interpretation
+  $\I$. By induction on the number of predicate atoms in
+  $\accform{w_\Sigma}$, using Proposition \ref{prop:subst} in the
+  inductive step, we prove that, $\I,w_\Data \models
+  \accform{w_\Sigma}$ if and only if $w_\Data \models
+  \substform{w_\Sigma}$. \qed}
 
@@
 there is no need to define dualization for formulae of the form $\neg
-q(x_1,\ldots,x_{\#(q)})$.
+q(x_1,\ldots,x_{\#(q)})$. The following theorem shows closure of
+automata under all boolean operations:
 
@@
 
-The following theorem shows that first-order alternating data automata
-are closed under all boolean operations:
-
 \begin{theorem}\label{thm:closure}
@@
   for $i=1,2$, such that $Q_1 \cap Q_2 = \emptyset$, the following hold: 
-  \begin{compactitem}
+  \begin{compactenum}
     \item\label{it1:closure} $\lang{\A_\cap} = \lang{\A_1} \cap \lang{\A_2}$, where
@@
         \Delta_i}\enspace.\]
-  \end{compactitem}
+  \end{compactenum}
   Moreover, $\len{\A_\cap} = \bigO{\len{\A_1}+\len{\A_2}}$ and

### 2018-07-27 quantifiers/draft.tex
ok

@@
 \phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms. We write
-$\phi_1 \vee \phi_2$, $\phi_1 \Rightarrow \phi_2$ and $\forall x ~.~
+$\phi_1 \vee \phi_2$, $\phi_1 \rightarrow \phi_2$ and $\forall x ~.~
 \phi_1$ for $\neg(\neg\phi_1 \wedge \neg\phi_2)$, $\neg\phi_1 \vee
@@
 \fv{}{\phi}$, and drop the subscript for sentences. A sentence $\phi$
-is \emph{satisfiable} if $\sem{\phi}\neq\emptyset$, an element of
-$\sem{\phi}$ being called a \emph{model} of $\phi$.
+is \emph{satisfiable} (\emph{unsatisfiable}) if
+$\sem{\phi}\neq\emptyset$ [$\sem{\phi}=\emptyset$]. An element of
+$\sem{\phi}$ is called a \emph{model} of $\phi$. A formula $\phi$ is
+\emph{valid} if $\I,\nu \models \phi$ for every interpretation $\I$
+and every valuation $\nu$.
 
@@
   \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-\stamp{q}{i-1}(y_1,\ldots,y_{\#(q)}) \Longrightarrow \stamp{\psi}{i}\]
+\stamp{q}{i-1}(y_1,\ldots,y_{\#(q)}) \rightarrow \stamp{\psi}{i}\]
 The automaton $\A$, to which $\pathform{\alpha}$ refers, will always
@@
   and let $\forall y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{n}$ be
+  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \rightarrow \stamp{\psi}{n}$ be
   the subformula of $\pathform{w_\Sigma}$ corresponding to the
@@
   y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{n}$,
+  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \rightarrow \stamp{\psi}{n}$,
   we have $\I \in \sem{\stamp{\psi}{n}}_{\nu}$ and let $\K$ be one of
@@
   \ldots \forall y_{\#(q)} ~.~ \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)})
-  \Rightarrow \stamp{\psi}{n}$ of $\pathform{w_\Sigma}$, by the
+  \rightarrow \stamp{\psi}{n}$ of $\pathform{w_\Sigma}$, by the
   execution semantics of $\A$. This is the case because the children
@@
   \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-\stamp{q}{n}(y_1,\ldots,y_{\#(q)}) \Rightarrow \bot\]
+\stamp{q}{n}(y_1,\ldots,y_{\#(q)}) \rightarrow \bot\]
 
@@
   $\I,\nu \models \accform{\alpha}$, we obtain that $\I,\eta \models
-  q(y_1,\ldots,y_{\#(q)}) \Rightarrow \stamp{\psi}{m+1}$, thus
+  q(y_1,\ldots,y_{\#(q)}) \rightarrow \stamp{\psi}{m+1}$, thus
   $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$. However, this
@@
   \setminus F$. Since $\I,\nu \models \forall y_1 \ldots \forall
-  y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \Rightarrow \bot$, by a
+  y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow \bot$, by a
   similar reasoning as in the above case, we obtain that
@@
   \models \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
-  y_{\#(q)} ~.~ \stamp{q}{n}(y_1,\ldots,y_{\#(q)}) \Rightarrow \bot$,
+  y_{\#(q)} ~.~ \stamp{q}{n}(y_1,\ldots,y_{\#(q)}) \rightarrow \bot$,
   trivially.  \qed}
@@
 the formula obtained from $\pathform{\alpha}$, by successively
-replacing in $\accform{\alpha}$: \begin{compactitem}
-\item each occurrence of a predicate
+replacing in $\accform{\alpha}$: \begin{compactenum}[(a)]
+\item\label{it1:substform} each occurrence of a predicate
   $\stamp{q}{i}(t_1,\ldots,t_{\#(q)})$ by the formula
   $\stamp{\psi}{i+1}[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$, where
-  $q(y_1,\ldots,y_{\#(q)}) \arrow{a_{i+1}(X)}{} \psi \in \Delta$ is the
-  unique transition rule for $q$ and $a_i$, for all $i=0,\ldots,n-1$,
+  $q(y_1,\ldots,y_{\#(q)}) \arrow{a_{i+1}(X)}{} \psi \in \Delta$ is
+  the unique transition rule for $q$ and $a_i$, for all
+  $i=0,\ldots,n-1$,
 %
-\item each occurrence of a predicate $\stamp{q}{n}$, such that $q \in
-  Q \setminus F$, by $\top$, and
+\item\label{it2:substform} each occurrence of a predicate
+  $\stamp{q}{n}$, such that $q \in Q \setminus F$, by $\top$, and
 %
-\item each occurrence of a predicate $\stamp{q}{n}$, such that $q \in
-  F$, by $\bot$.
-\end{compactitem}
+\item\label{it3:substform} each occurrence of a predicate
+  $\stamp{q}{n}$, such that $q \in F$, by $\bot$.
+\end{compactenum}
 Since there are no occurrences of predicates in $\substform{\alpha}$,
@@
   \end{compactenum}
+  where $y_1,\ldots,y_{\#(q)} \not\in \fv{}{\phi}$. 
 \end{proposition}
@@
   \begin{compactenum}
-    \item\label{it1:closure} $\lang{\A_\cap} = \lang{\A_1} \cap \lang{\A_2}$, where
+    \item\label{it1:thm:closure} $\lang{\A_\cap} = \lang{\A_1} \cap \lang{\A_2}$, where
       $\A_\cap = \tuple{\Sigma,X,Q_1 \cup Q_2, \iota_1 \wedge \iota_2,
@@
       %
-    \item\label{it2:closure} $\lang{\overline{\A_i}} = \Sigma[X]^*
+    \item\label{it2:thm:closure} $\lang{\overline{\A_i}} = \Sigma[X]^*
       \setminus \lang{\A_i}$, where $\overline{\A_i} =
@@
 \end{theorem}
-\proof{\qed}
+\proof{ (\ref{it1:thm:closure}) ``$\subseteq$'' Let $w \in
+  \lang{\A_\cap}$ be a word and $\T$ be an execution of $\A_\cap$ over
+  $w$. Since $Q_1 \cap Q_2 = \emptyset$, it is possible to partition
+  $\T$ into $\T_1$ and $\T_2$ such that the roots of $\T_i$ form a
+  cube from $\cube{\minsem{\iota_i}}$, for all $i=1,2$. Because
+  $\Delta_1 \cap \Delta_2 = \emptyset$, by induction on
+  $\len{w}\geq0$, one shows that $\T_i$ is an execution of $\A_i$ over
+  $w$, for all $i=1,2$. Finally, because $\T$ is accepting, we obtain
+  that $\T_1$ and $\T_2$ are accepting, respectively, hence $w \in
+  \lang{\A_1} \cap \lang{\A_2}$. ``$\supseteq$'' Let $w \in
+  \lang{\A_1} \cap \lang{\A_2}$ and let $\T_i$ an accepting execution
+  of $\A_i$ over $w$, for all $i=1,2$. We show that $\T_1 \cup \T_2$
+  is an execution of $\A_\cap$ over $w$, by induction on
+  $\len{w}\geq0$. For the base case $\len{w}=0$, we have $\T_i \in
+  \cube{\minsem{\iota_i}}$ for all $i=1,2$ and since $Q_1 \cap Q_2 =
+  \emptyset$, we have $\T_1 \cup \T_2 \in \cube{\minsem{\iota_1 \wedge
+      \iota_2}}$. The induction step follows as a consequence of the
+  fact that $\Delta_1 \cup \Delta_2$ is the set of transition rules of
+  $\A_\cap$. Finally, since both $\T_1$ and $\T_2$ are accepting,
+  $\T_1 \cup \T_2$ is accepting as well. Moreover, we have: 
+  \[
+  \len{\A_\cap} = \len{\iota_1 \wedge \iota_2} + \sum_{q(\vec{y})
+    \arrow{a(X)}{} \psi \in \Delta_1 \cup \Delta_2} \len{\psi} =
+  1 + \len{\iota_1} + \len{\iota_2} + \sum_{q(\vec{y})
+    \arrow{a(X)}{} \psi \in \Delta_1} \len{\psi} + \sum_{q(\vec{y})
+    \arrow{a(X)}{} \psi \in \Delta_2} \len{\psi} \enspace. 
+  \]
+
+  \noindent(\ref{it2:thm:closure}) Let $w \in \Sigma[X]^*$ be a
+  word. We denote by $\accformA{\A_1}{w_\Sigma}$ and
+  $\substformA{\A_1}{w_\Sigma}$
+  [resp. $\accformA{\overline{\A}_1}{w_\Sigma}$ and
+    $\substformA{\overline{\A}_1}{w_\Sigma}$] the formulae
+  $\accform{w_\Sigma}$ and $\substform{w_\Sigma}$ for $\A_1$ and
+  $\overline{\A}_1$, respectively. It is enough to show that
+  $\substformA{\overline{\A}_1}{w_\Sigma} =
+  \neg\substformA{\A_1}{w_\Sigma}$ and apply Lemma
+  \ref{lemma:substitution} to prove that $w \in \lang{\A_1} \iff w
+  \not\in \lang{\overline{\A}_1}$. Since the choice of $w$ was
+  arbitrary, this proves $\lang{\overline{\A}_1} = \Sigma[X]^*
+  \setminus \lang{\A_1}$. By induction on the number of predicate
+  atoms in $\accformA{\A_1}{w_\Sigma}$ that are replaced during the
+  generation of $\substformA{\A_1}{w_\Sigma}$. The proof relies on the
+  following fact:
+
+  \begin{fact}\label{fact:dual-neg}
+    Let $\phi$ be a positive formula and let $q(t_1,\ldots,t_{\#(q)})$
+    be the only occurrence of a predicate symbol within $\phi$. Then,
+    every formula $\phi$ with no predicate occurrences:
+    $\neg\phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]
+    \equiv
+    \dual{\phi}[\neg\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$.
+  \end{fact}
+  \proof{By induction on the structure of $\phi$. \qed}}
 
@@
 The problem of checking emptiness of a given automata is undecidable,
-for automata using predicates of arity two or more, even only when the
-theory in which the (right-hand sides of the) transition rules are
-written consists of only equalities and disequalities between
-variables, with no quantifiers \cite{FarzanKincaidPodelski15}.
-Consequently, we cannot tackle the emptiness problem via a complete
-decision procedure and resort to an abstraction-refinement loop that
-tries to prove non-emptiess by discovering a word that it accepted,
-firing longer and longer sequences of input events.
+even for automata with predicates of arity two whose transition rules
+use only equalities and disequalities, with no quantifiers
+\cite{Farzan15}. Since there is no decision procedure even in the most
+simple case, we use an abstraction-refinement semi-algorithm based on
+\emph{lazy annotation} \cite{McMillan06,McMillan14}. In a nutshell, a
+lazy annotation procedure systematically explores the set of execution
+traces of a program in search for a counterexample. When a spurious
+counterexample is found, the search backtracks and the program path is
+annotated with a set of learned facts, that excludes the path from
+future searches. In general, these facts are interpolants extracted
+from the proof of spuriousness of the given (finite) path.
+
+Typically, in a lazy annotation procedure that checks the
+unreachability of a given control state in a program (or, more
+generally, the infeasability of a set of non-linear Horn clauses
+\cite{BjornerGurfinkelMcMillanRybalchenko15}), the interpolants used
+to annotate the unfolded control structure are local assertions about
+the values of the program variables. These assertions define
+overapproximations of the set of configurations reachable at a given
+control state.
+
+When considering an alternating model of computation, computing local
+interpolants involving a single control state is not possible anymore,
+because of the fact the several branches of the computation
+synchronize on the same sequence of input values. Consider, for
+instance, an automaton with the following transition rules:
+\[\begin{array}{rclcrcl}
+q_0(y) & \arrow{a(x)}{} & q_1(y+x) \wedge q_2(y-x) & \hspace*{2mm} & 
+q_1(y) \arrow{a(x)}{} y+x > 0 \\
+q_1(y) & \arrow{a(x)}{} & q_1(y+x) && q_2(y) \arrow{a(x)}{} y-x > 0 \\
+q_2(y) & \arrow{a(x)}{} & q_2(y-x)
+\end{array}\]
+When started in an initial configuration $q_0(0)$, on an input word
+$(a,k_1) \ldots (a,k_n)$, the automaton executes as follows: 
+\[q_0(0) \arrow{(a,k_1)}{} q_1(k_1)
+\wedge q_2(-k_1) \ldots \arrow{(a,k_{n-1})}{} q_1(\sum_{i=1}^{n-1}
+k_i) \wedge q_2(-\sum_{i=1}^{n-1} k_i) \arrow{(a,k_n)}{}
+\sum_{i=1}^{n} k_i > 0 \wedge \sum_{i=1}^{n} k_i < 0\] A fairly good
+overapproximation of the set of cubes generated after the initial
+configuration is given by the formula $\exists x_1 \exists x_2 ~.~
+q_1(x_1) \wedge q_2(x_2) \wedge x_1+x_2 \teq 0$. Observe that a
+formula using one occurrence of a predicate would give a too rough
+overapproximation of this set, unable to prove the emptiness of the
+automaton.
+
+The rest of this section is concerned with the building blocks of a
+lazy annotation semi-algorithm for checking emptiness of automata. In
+particular, we address the following problems: \begin{compactenum}
+%
+\item generation of interpolants that prove the infeasability of a
+  given path formula, provided that the satisfiability of the
+  first-order theory thereof is decidable, in the presence of
+  predicate symbols.
+%
+\item generation of non-local interpolants using history variables,
+  obtained by Skolemization of the universal quantifiers that occur
+  within transition rules.
+%
+\item computing the coverage between two node labels in the unfolding
+  of an automaton, when history variables occur. 
+\end{compactenum} 
+Based on these results, we show how a version of the classical
+\textsc{Impact} \cite{McMillan06} semi-algorithm can be used to
+(dis-)prove the emptiness of first-order alternating automata.
+
+\subsection{Lyndon Interpolants}
+
+Let $\phi$ and $\psi$ be first-order formulae, such that $\phi \wedge
+\psi$ is unsatisfiable. The \emph{vocabulary} of $\phi$, denoted as
+$\voc{\phi}$, is the set of function symbols that occur in $\phi$.  By
+$\pset{+}{\phi}$ [$\pset{-}{\phi}$] we denote the set of predicate
+symbols that occur in $\phi$ under an even (odd) number of
+negations. A \emph{Lyndon interpolant} \cite{Lyndon59} for $\phi$ and
+$\psi$ is a first-order formula $I$ such that: \begin{compactitem}
+\item $\phi \rightarrow I$ is valid,  
+\item $I \wedge \psi$ is unsatisfiable, 
+\item $\voc{I} \subseteq \voc{\phi} \cap \voc{\psi}$, 
+\item $\pset{+}{I} \subseteq \pset{+}{\phi} \cap \pset{+}{\psi}$ and
+  $\pset{-}{I} \subseteq \pset{-}{\phi} \cap \pset{-}{\psi}$.
+\end{compactitem}
+The existence of Lyndon interpolants for first-order logic was proved
+by a non-constructive model-theoretic argument \cite{Lyndon59}. For
+now, we silently assume that the satisfiability problem for the theory
+in which $\phi$ and $\psi$ are written is decidable and, moreover,
+that a decision procedure generating Lyndon interpolants exists, for
+this theory. This is a rather unrealistic assumption, given that
+classical decidable theories, such as Presburger arithmetic, become
+undecidable in the presence of (monadic) predicate symbols
+\cite{Halpern91}.
+
+Let $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the
+rest of this section. Given a non-empty sequence of input events
+$\alpha = a_1 \ldots a_n \in \Sigma^*$, a \emph{local interpolant} for
+$\alpha$ is a sequence $I_0,I_1,\ldots,I_n$ of first-order formulae
+such that: \begin{compactitem}
+\item $\iota \rightarrow I_0$ is valid, 
+%
+\item $\left[I_k \wedge \left(\bigwedge_{q(\vec{y}_q) \arrow{a_i(X)}{} \psi}
+  \forall \vec{y}_q ~.~ \stamp{q}{k}(\vec{y}_q) \rightarrow
+  \nextf{\psi}\right)\right] \rightarrow \nextf{I}_{k+1}$ is valid, for all $k \in [n-1]$,
+%
+\item $I_n \wedge \left(\bigwedge_{q \in Q \setminus F} \forall \vec{y}_q
+  ~.~ q(\vec{y}_q) \rightarrow \bot\right)$ is unsatisfiable,
+%
+\item $\voc{I_k} = \pset{+}{I_k}$ and $\pset{-}{I_k} = \emptyset$. 
+\end{compactitem}
+
+\begin{lemma}\label{lemma:lyndon-existence}
+
+\end{lemma}
+
+\subsection{Skolem Witnesses}
+
+\subsection{Lazy Annotation}
 

### 2018-07-29 quantifiers/draft.tex
ok

@@
 variables and state predicates (and only those symbols) are replaced
-by their time-stamped counterparts. For a sequence of input events
-$\alpha = a_1 \ldots a_n$, the \emph{path formula} of $\alpha$ (in
-$\A$) is defined as:
-\[\pathform{\alpha} \isdef \stamp{\iota}{0} \wedge 
-\bigwedge_{i=1}^n \bigwedge_{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{}
+by their time-stamped counterparts. As a shorthand, we shall write
+$q(\vec{y})$ for $q(y_1,\ldots,y_{\#(q)})$, when no confusion arises.
+
+Given a sequence of input events $\alpha = a_1 \ldots a_n$, the
+\emph{path formula} of $\alpha$ (in $\A$) is defined as follows:
+\begin{equation}\label{eq:pathform}
+\pathform{\alpha} \isdef \stamp{\iota}{0} \wedge 
+\bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{}
   \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-\stamp{q}{i-1}(y_1,\ldots,y_{\#(q)}) \rightarrow \stamp{\psi}{i}\]
+\stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i}
+\end{equation}
 The automaton $\A$, to which $\pathform{\alpha}$ refers, will always
@@
   and let $\forall y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \rightarrow \stamp{\psi}{n}$ be
+  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$ be
   the subformula of $\pathform{w_\Sigma}$ corresponding to the
-  application(s) of the transition rule $q(y_1,\ldots,y_{\#(q)})
+  application(s) of the transition rule $q(\vec{y})
   \arrow{a_n}{} \psi$ at the $(n-1)$-th step. Let $\nu$ be the
@@
   y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)}) \rightarrow \stamp{\psi}{n}$,
+  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$,
   we have $\I \in \sem{\stamp{\psi}{n}}_{\nu}$ and let $\K$ be one of
@@
   w_\Data \models \varphi$, for each subformula $\varphi = \forall y_1
-  \ldots \forall y_{\#(q)} ~.~ \stamp{q}{n-1}(y_1,\ldots,y_{\#(q)})
+  \ldots \forall y_{\#(q)} ~.~ \stamp{q}{n-1}(\vec{y})
   \rightarrow \stamp{\psi}{n}$ of $\pathform{w_\Sigma}$, by the
@@
 the set of final states of $\A$: 
-
-\[\accform{\alpha} \isdef \pathform{\alpha} \wedge \bigwedge_{q
+\begin{equation}\label{eq:accform}
+\accform{\alpha} \isdef \pathform{\alpha} \wedge \bigwedge_{q
   \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-\stamp{q}{n}(y_1,\ldots,y_{\#(q)}) \rightarrow \bot\]
+\stamp{q}{n}(\vec{y}) \rightarrow \bot
+\end{equation}
 
@@
   have $\cube{\minsem{\psi}_{\eta}} = \emptyset$, where
-  $q(y_1,\ldots,y_{\#(q)}) \arrow{a_{m+1}(X)}{} \psi$ is the transition
+  $q(\vec{y}) \arrow{a_{m+1}(X)}{} \psi$ is the transition
   rule of $\A$ that applies for $q$ and $a_{m+1}$ and $\eta$ stands
@@
   $\I,\nu \models \accform{\alpha}$, we obtain that $\I,\eta \models
-  q(y_1,\ldots,y_{\#(q)}) \rightarrow \stamp{\psi}{m+1}$, thus
+  q(\vec{y}) \rightarrow \stamp{\psi}{m+1}$, thus
   $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$. However, this
@@
   \setminus F$. Since $\I,\nu \models \forall y_1 \ldots \forall
-  y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow \bot$, by a
+  y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot$, by a
   similar reasoning as in the above case, we obtain that
@@
   \models \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
-  y_{\#(q)} ~.~ \stamp{q}{n}(y_1,\ldots,y_{\#(q)}) \rightarrow \bot$,
+  y_{\#(q)} ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot$,
   trivially.  \qed}
 
-In the following, it will be useful to define the set of data
-valuations of a word by a formula without predicates. Given a sequence
-of input events $\alpha = a_1 \ldots a_n$, let $\substform{\alpha}$ be
-the formula obtained from $\pathform{\alpha}$, by successively
-replacing in $\accform{\alpha}$: \begin{compactenum}[(a)]
+In the following, it will be useful, in several places, to define the
+data sequence of an accepted word using the predicate-free formula
+defined below:
+
+\begin{definition}\label{def:substform}
+Given a sequence of input events $\alpha = a_1 \ldots a_n$, let
+$\substform{\alpha}$ be the formula obtained by successively replacing
+in $\accform{\alpha}$: \begin{compactenum}[(a)]
 \item\label{it1:substform} each occurrence of a predicate
@@
   $\stamp{\psi}{i+1}[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$, where
-  $q(y_1,\ldots,y_{\#(q)}) \arrow{a_{i+1}(X)}{} \psi \in \Delta$ is
-  the unique transition rule for $q$ and $a_i$, for all
-  $i=0,\ldots,n-1$,
+  $q(\vec{y}) \arrow{a_{i+1}(X)}{} \psi \in \Delta$ is the transition
+  rule for $q$ and $a_i$, for all $i=0,\ldots,n-1$,
 %
@@
 \end{compactenum}
+\end{definition}
 Since there are no occurrences of predicates in $\substform{\alpha}$,
@@
 exists an interpretation $\I$ such that $\I,\nu \models
-\substform{\alpha}$ if and only if $\J,\nu \models \substform{\alpha}$, 
-for every interpretation $\J$. Thus we omit $\I$ and simply write $\nu
-\models \substform{\alpha}$ in this case.
+\substform{\alpha}$ if and only if $\J,\nu \models
+\substform{\alpha}$, for every interpretation $\J$. Thus we omit $\I$
+and simply write $\nu \models \substform{\alpha}$ in this case.
 
@@
   \item\label{it2:prop:subst} $\I,\nu \models \phi \wedge \forall y_1
-    \ldots \forall y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow
+    \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow
     \psi$, for some interpretation $\I$.
@@
       t_{\#(q)}/y_{\#(q)}]$ thus $\I,\nu \models \forall y_1 \ldots
-    \forall y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$.
+    \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \psi$.
     %
@@
     $\emptyset$, we have $\I,\nu \models \forall y_1 \ldots \forall
-    y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$, trivially.
+    y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \psi$, trivially.
   \end{compactitem}
@@
     $\I,\nu \models \forall y_1 \ldots \forall y_{\#(q)} ~.~
-    q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$, we have $\nu \models
+    q(\vec{y}) \rightarrow \psi$, we have $\nu \models
     \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$. Since $\phi$ is
@@
         F_i,\dual{\Delta}_i}$ and, for $i=1,2$: \[\dual{\Delta}_i =
-      \set{q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \dual{\psi} \mid
-        q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi \in
+      \set{q(\vec{y}) \arrow{a(X)}{} \dual{\psi} \mid
+        q(\vec{y}) \arrow{a(X)}{} \psi \in
         \Delta_i}\enspace.\]
@@
 use only equalities and disequalities, with no quantifiers
-\cite{Farzan15}. Since there is no decision procedure even in the most
-simple case, we use an abstraction-refinement semi-algorithm based on
-\emph{lazy annotation} \cite{McMillan06,McMillan14}. In a nutshell, a
-lazy annotation procedure systematically explores the set of execution
-traces of a program in search for a counterexample. When a spurious
-counterexample is found, the search backtracks and the program path is
-annotated with a set of learned facts, that excludes the path from
-future searches. In general, these facts are interpolants extracted
-from the proof of spuriousness of the given (finite) path.
+\cite{Farzan15}. Since even such simple classes of alternating
+automata have no general decision procedure, we use an
+abstraction-refinement semi-algorithm based on \emph{lazy annotation}
+\cite{McMillan06,McMillan14}. In a nutshell, a lazy annotation
+procedure systematically explores the set of execution traces of a
+program in search for a counterexample. When a spurious counterexample
+is found, the search backtracks and the program path is annotated with
+a set of learned facts, that excludes the path from future
+searches. These facts are interpolants extracted from the proof of
+spuriousness of the given (finite) path.
 
-Typically, in a lazy annotation procedure that checks the
-unreachability of a given control state in a program (or, more
-generally, the infeasability of a set of non-linear Horn clauses
-\cite{BjornerGurfinkelMcMillanRybalchenko15}), the interpolants used
-to annotate the unfolded control structure are local assertions about
-the values of the program variables. These assertions define
-overapproximations of the set of configurations reachable at a given
-control state.
+Typically, when checking the unreachability of a set of program
+configurations (or, more generally, the infeasability of a set of Horn
+clauses \cite{BjornerGurfinkelMcMillanRybalchenko15}), the
+interpolants used to annotate the unfolded control structure are local
+assertions about the values of the program variables in a given
+control state. These assertions define inductive overapproximations of
+the set of configurations reachable at that control state, that
+suffice to prove emptiness or, equivalently, a safety property of the
+system.
 
-When considering an alternating model of computation, computing local
-interpolants involving a single control state is not possible anymore,
-because of the fact the several branches of the computation
-synchronize on the same sequence of input values. Consider, for
-instance, an automaton with the following transition rules:
+When considering an alternating model of computation, local
+interpolants that involve a single control state are no longer enough,
+because of the synchronization of several branches of the computation
+on the same sequence of input values. Consider, for instance, an
+automaton with the following transition rules and final state $q_f$:
 \[\begin{array}{rclcrcl}
 q_0(y) & \arrow{a(x)}{} & q_1(y+x) \wedge q_2(y-x) & \hspace*{2mm} & 
-q_1(y) \arrow{a(x)}{} y+x > 0 \\
-q_1(y) & \arrow{a(x)}{} & q_1(y+x) && q_2(y) \arrow{a(x)}{} y-x > 0 \\
+q_1(y) \arrow{a(x)}{} y+x > 0 \wedge q_f \\
+q_1(y) & \arrow{a(x)}{} & q_1(y+x) && q_2(y) \arrow{a(x)}{} y-x > 0 \wedge q_f \\
 q_2(y) & \arrow{a(x)}{} & q_2(y-x)
 \end{array}\]
-When started in an initial configuration $q_0(0)$, on an input word
-$(a,k_1) \ldots (a,k_n)$, the automaton executes as follows: 
-\[q_0(0) \arrow{(a,k_1)}{} q_1(k_1)
-\wedge q_2(-k_1) \ldots \arrow{(a,k_{n-1})}{} q_1(\sum_{i=1}^{n-1}
-k_i) \wedge q_2(-\sum_{i=1}^{n-1} k_i) \arrow{(a,k_n)}{}
-\sum_{i=1}^{n} k_i > 0 \wedge \sum_{i=1}^{n} k_i < 0\] A fairly good
-overapproximation of the set of cubes generated after the initial
-configuration is given by the formula $\exists x_1 \exists x_2 ~.~
-q_1(x_1) \wedge q_2(x_2) \wedge x_1+x_2 \teq 0$. Observe that a
-formula using one occurrence of a predicate would give a too rough
-overapproximation of this set, unable to prove the emptiness of the
-automaton.
+Started in an initial configuration $q_0(0)$ with an input word
+$(a,\nu_1) \ldots (a,\nu_{n-1}) (b,\nu_n)$, such that $\nu_i(x)=k_i$,
+the automaton executes as follows:
+\[\begin{array}{l}
+q_0(0) \arrow{(a,\nu_1)}{} q_1(k_1)
+\wedge q_2(-k_1) \ldots \arrow{(a,\nu_{n-1})}{} 
+q_1(\sum_{i=1}^{n-1} k_i) \wedge q_2(-\sum_{i=1}^{n-1} k_i) \arrow{(a,\nu_n)}{} \\
+\sum_{i=1}^{n} k_i > 0 \wedge \sum_{i=1}^{n} k_i < 0 \wedge q_f \equiv \bot
+\end{array}\] 
+An overapproximation of the set of cubes generated after one or more
+steps is defined by the formula: $\exists x_1 \exists x_2 ~.~ q_1(x_1)
+\wedge q_2(x_2) \wedge x_1+x_2 \teq 0$. Observe that a formula using
+one occurrence of a predicate would give a too rough overapproximation
+of this set, unable to prove the emptiness of the automaton.
 
-The rest of this section is concerned with the building blocks of a
-lazy annotation semi-algorithm for checking emptiness of automata. In
-particular, we address the following problems: \begin{compactenum}
+The solution we adopt is a generalization of the classical approach of
+extracting the interpolants from the proof of unsatisfiability of a
+time-stamped path formula. Given a sequence of input events $\alpha$,
+we consider the path formula $\accform{\alpha}$, constrained by the
+fact that all non-final predicates $\stamp{q}{\len{\alpha}}$ must be
+empty sets, defined as in (\ref{eq:accform}). If this formula is
+unsatisfiable, we show the existence of a sequence of (possibly
+quantified) positive interpolants that overapproximate the set of
+cubes generated at each step during an execution over a word $w$,
+whose even sequence is $w_\Sigma = \alpha$. Moreover, these
+interpolants are local, in the sense that each formula uses only the
+values of the variables and the state predicates stamped with its
+index in the sequence.
+
+Unlike with the classical path formulae for integer programs
+\cite{McMillan06} and Horn clauses
+\cite{BjornerGurfinkelMcMillanRybalchenko15}, that are expressed using
+linear or Presburger arithmetic, the existence of local interpolants
+for theories using first-order predicates relies on a non-constructive
+argument, the Lyndon Interpolation Theorem \cite{Lyndon59}. Since
+usual decidable theories, such as Presburger arithmetic, become
+undecidable in the presence of monadic predicate symbols
+\cite{Halpern91}, we cannot hope to find an automated interpolant
+generation method, in general.
+
+But here is where the predicate-free path formulae (Definition
+\ref{def:substform}) come into play. For a given event sequence
+$\alpha$, the automaton $\A$ accepts a word $w$ such that $w_\Sigma =
+\alpha$ if and only if $\substform{\alpha}$ is satisfiable (Lemma
+\ref{lemma:substitution}). Assuming further that the equality atoms in
+the transition rules of $\A$ are written in the language of a
+decidable first-order theory, such as Presburger arithmetic, Lemma
+\ref{lemma:substitution} gives us an effective way of checking
+emptiness of $\A$ relative to a given event sequence. However, this
+method does not cope well with lazy annotation, because there is no
+way to extract, from the unsatisfiability proof of
+$\substform{\alpha}$, the interpolants needed to annotate
+$\alpha$. This is because \begin{inparaenum}[(i)]
+\item the formula $\substform{\alpha}$, obtained by repeated
+  substitutions (Definition \ref{def:substform}) loses track of the
+  execution semantics of $\A$, and 
+%
+\item quantifiers may occur nested in $\substform{\alpha}$, making it
+  difficult to write $\substform{\alpha}$ as a proper interpolation
+  problem.
+\end{inparaenum}
+
+The solution we adopt in this paper relies on an additional assumption
+related to the decidability of the predicate-free theory in which
+$\substform{\alpha}$ formulae are written, namely that the theory has
+an effective \emph{witness-providing quantifier elimination}
+procedure. More precisely, we assume that, for each formula $\exists x
+~.~ \phi(x)$, there exists an effectively computable term $t$, in
+which $x$ does not occur, such that $\exists x ~.~ \phi$ and
+$\phi[t/x]$ are equi-satisfiable. These terms, called \emph{witnesses}
+in the following, are actual interpretations of the Skolem function
+symbols from the following folklore theorem:
+
+\begin{theorem}\label{thm:skolem}
+  Given $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$ a first order sentence, 
+  where $Q_1, \ldots, Q_n \in \set{\exists,\forall}$ and $\phi$ is
+  quantifier-free, let: 
+  \[\eta_i \equiv \left\{\begin{array}{ll}
+  f_i(y_1,\ldots,y_{k_i}) & \text{if } Q_i \equiv \forall \\
+  x_i & \text{if } Q_i \equiv \exists
+  \end{array}\right.\]
+  where $f_i$ is a fresh function symbol and $\set{y_1, \ldots,
+    y_{k_i}} = \set{x_j \mid j < i,~ Q_j \equiv \exists}$. Then \(Q_1
+  x_1 \ldots Q_n x_n ~.~ \phi \rightarrow
+  \phi[\eta_1/x_1,\ldots,\eta_n/x_n]\) is valid.
+\end{theorem}
+\proof{See \cite[Theorem 2.1.8]{BorgerGraedelGurevich97} and
+  \cite[Lemma 2.1.9]{BorgerGraedelGurevich97}. \qed}
+Observe that our assumption on the existence of an effective
+witness-providing quantifier elimination procedure is stronger than
+the above theorem, as we require that \begin{inparaenum}[(i)] 
+\item the $f_i$ functions be definable in the language of the theory,
+  and 
+\item $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$ and
+  $\phi[\eta_1/x_1,\ldots,\eta_n/x_n]$ are equi-satisfiable. 
+\end{inparaenum}
+Among the theories that support witness-providing quantifier
+elimination, one mentions \comment[ri]{LIA, LRA, Presburger, Tarski?,
+  BAPA?}
+
+Turning back to the problem of generating interpolants from the proof
+of unsatisfiability of a path formula $\accform{\alpha}$, assume that
+the equivalent predicate-free formula $\substform{\alpha} = \exists
+x_1 \forall y_1 \ldots \exists x_n \forall y_n ~.~ \psi$ is
+unsatisfiable and let $t_1(x_1)$, $\ldots, t_n(x_1,\ldots,x_n)$ be
+witnesses such that $\psi[t_1/y_1,\ldots,t_n/y_n] \equiv \bot$. Then
+we show that the quantifier-free formula (with predicates) obtained
+from $\accform{\alpha}$ by (roughly) replacing the universal
+quantifiers with $t_1, \ldots, t_n$ in $\accform{\alpha}$ is also
+unsatisfiable.  Moreover, this new formula is layered in the same way
+as $\accform{\alpha}$, namely as a conjunction of the form:
+\[\stamp{\iota}{0} \wedge \bigwedge_{i=1}^n \bigwedge_{q(\vec{y})
+  \arrow{a_i(X)}{} \psi \in \Delta} \stamp{q}{i-1}(\vec{t}_{q,i-1})
+\rightarrow \stamp{\psi}{i} \wedge \bigwedge_{q \in Q \setminus F}
+\stamp{q}{n}(\vec{t}_{q,n}) \rightarrow \bot\]
+
+\comment[ri]{i am here}
+
+However, the existence of a decision procedure for the theory in which
+$\accform{\alpha}$ is written is not guaranteed, as it was usually the
+case of classical path formulae for integer programs \cite{McMillan06}
+and Horn clauses \cite{BjornerGurfinkelMcMillanRybalchenko15} using
+linear or Presburger arithmetic. The problem is mainly that decidable
+theories, such as Presburger arithmetic become undecidable in the
+presence of monadic predicate symbols \cite{Halpern91}. A way around
+this problem is to check whether a given event sequence $\alpha$ may
+
+The rest of this section is concerned with defining the building
+blocks of a lazy annotation semi-algorithm for checking emptiness of
+automata.
+
+In particular, we address the following problems: \begin{compactenum}
 %
 \item generation of interpolants that prove the infeasability of a
-  given path formula, provided that the satisfiability of the
-  first-order theory thereof is decidable, in the presence of
-  predicate symbols.
+  given path formula, assuming the decidability of the first-order
+  theory in which the path formulae are written. 
 %

### 2018-09-09 tacasPPTandSources/src/main/java/ADA.java
source&pdf

@@ -1,0 +1,712 @@
+/*
+	JAltImpact
+    Copyright (C) 2017  Xiao XU & Radu IOSIF
+
+	This file is part of JAltImpact.
+
+    JAltImpact is free software: you can redistribute it and/or modify
+    it under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    This program is distributed in the hope that it will be useful,
+    but WITHOUT ANY WARRANTY; without even the implied warranty of
+    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+    GNU General Public License for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with this program.  If not, see <https://www.gnu.org/licenses/>.
+    
+    If you have any questions, please contact Xiao XU <xiao.xu.cathiec@gmail.com>.
+*/
+
+import java.io.BufferedReader;
+import java.io.File;
+import java.io.FileReader;
+import java.io.FileWriter;
+import java.io.IOException;
+import java.util.ArrayList;
+import java.util.Collection;
+import java.util.HashMap;
+import java.util.HashSet;
+import java.util.Iterator;
+import java.util.List;
+import java.util.Map;
+import java.util.Map.Entry;
+import java.util.Set;
+
+import org.sosy_lab.common.ShutdownManager;
+import org.sosy_lab.common.configuration.Configuration;
+import org.sosy_lab.common.configuration.InvalidConfigurationException;
+import org.sosy_lab.common.log.LogManager;
+import org.sosy_lab.java_smt.SolverContextFactory;
+import org.sosy_lab.java_smt.SolverContextFactory.Solvers;
+import org.sosy_lab.java_smt.api.BooleanFormula;
+import org.sosy_lab.java_smt.api.BooleanFormulaManager;
+import org.sosy_lab.java_smt.api.Formula;
+import org.sosy_lab.java_smt.api.FormulaManager;
+import org.sosy_lab.java_smt.api.FormulaType;
+import org.sosy_lab.java_smt.api.IntegerFormulaManager;
+import org.sosy_lab.java_smt.api.InterpolatingProverEnvironment;
+import org.sosy_lab.java_smt.api.ProverEnvironment;
[…]

### 2018-09-09 tacasPPTandSources/src/main/java/CheckResult.java

@@ -1,0 +1,49 @@
+/*
+	JAltImpact
+    Copyright (C) 2017  Xiao XU & Radu IOSIF
+
+	This file is part of JAltImpact.
+
+    JAltImpact is free software: you can redistribute it and/or modify
+    it under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    This program is distributed in the hope that it will be useful,
+    but WITHOUT ANY WARRANTY; without even the implied warranty of
+    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+    GNU General Public License for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with this program.  If not, see <https://www.gnu.org/licenses/>.
+    
+    If you have any questions, please contact Xiao XU <xiao.xu.cathiec@gmail.com>.
+*/
+
+import java.util.ArrayList;
+
+import org.sosy_lab.java_smt.api.BooleanFormula;
+import org.sosy_lab.java_smt.api.Model;
+
+public class CheckResult {
+
+	Boolean value;
+	Model model;
+	ArrayList<BooleanFormula> interpolants;
+	
+	public CheckResult() {
+		value = null;
+		model = null;
+		interpolants = new ArrayList<BooleanFormula>();
+	}
+	
+	public String toString() {
+		if(!value) {
+			return "$ Non-Accepting @ Interpolants :\n$ " + interpolants.toString();
+		}
+		else {
+			return "$ Accepting @ Model :\n" + ( model == null ? "[Empty Model]" : model.toString());
+		}
+	}
+	
+}

### 2018-09-09 tacasPPTandSources/src/main/java/PA.java

@@ -1,0 +1,869 @@
+/*
+	JAltImpact
+    Copyright (C) 2017  Xiao XU & Radu IOSIF
+
+	This file is part of JAltImpact.
+
+    JAltImpact is free software: you can redistribute it and/or modify
+    it under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    This program is distributed in the hope that it will be useful,
+    but WITHOUT ANY WARRANTY; without even the implied warranty of
+    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+    GNU General Public License for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with this program.  If not, see <https://www.gnu.org/licenses/>.
+    
+    If you have any questions, please contact Xiao XU <xiao.xu.cathiec@gmail.com>.
+*/
+
+import java.io.BufferedReader;
+import java.io.File;
+import java.io.FileReader;
+import java.io.IOException;
+import java.util.ArrayList;
+import java.util.Collection;
+import java.util.HashMap;
+import java.util.HashSet;
+import java.util.Iterator;
+import java.util.List;
+import java.util.Map;
+import java.util.Map.Entry;
+import java.util.Set;
+
+import org.sosy_lab.common.ShutdownManager;
+import org.sosy_lab.common.configuration.Configuration;
+import org.sosy_lab.common.configuration.InvalidConfigurationException;
+import org.sosy_lab.common.log.LogManager;
+import org.sosy_lab.java_smt.SolverContextFactory;
+import org.sosy_lab.java_smt.SolverContextFactory.Solvers;
+import org.sosy_lab.java_smt.api.BooleanFormula;
+import org.sosy_lab.java_smt.api.BooleanFormulaManager;
+import org.sosy_lab.java_smt.api.Formula;
+import org.sosy_lab.java_smt.api.FormulaManager;
+import org.sosy_lab.java_smt.api.FormulaType;
+import org.sosy_lab.java_smt.api.IntegerFormulaManager;
+import org.sosy_lab.java_smt.api.InterpolatingProverEnvironment;
+import org.sosy_lab.java_smt.api.ProverEnvironment;
+import org.sosy_lab.java_smt.api.SolverContext;
[…]

### 2018-09-09 tacasPPTandSources/src/main/java/empty.java

@@ -1,0 +1,134 @@
+/*
+	JAltImpact
+    Copyright (C) 2017  Xiao XU & Radu IOSIF
+
+	This file is part of JAltImpact.
+
+    JAltImpact is free software: you can redistribute it and/or modify
+    it under the terms of the GNU General Public License as published by
+    the Free Software Foundation, either version 3 of the License, or
+    (at your option) any later version.
+
+    This program is distributed in the hope that it will be useful,
+    but WITHOUT ANY WARRANTY; without even the implied warranty of
+    MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+    GNU General Public License for more details.
+
+    You should have received a copy of the GNU General Public License
+    along with this program.  If not, see <https://www.gnu.org/licenses/>.
+    
+    If you have any questions, please contact Xiao XU <xiao.xu.cathiec@gmail.com>.
+*/
+
+import java.io.IOException;
+
+import org.sosy_lab.common.NativeLibraries;
+import org.sosy_lab.common.ShutdownManager;
+import org.sosy_lab.common.configuration.Configuration;
+import org.sosy_lab.common.configuration.InvalidConfigurationException;
+import org.sosy_lab.common.log.BasicLogManager;
+import org.sosy_lab.common.log.LogManager;
+import org.sosy_lab.java_smt.SolverContextFactory.Solvers;
+import org.sosy_lab.java_smt.api.SolverException;
+
+public class empty {
+	
+  public static void main(String[] args)
+		  throws
+		  InvalidConfigurationException,
+		  SolverException,
+		  InterruptedException,
+		  IOException
+  {  
+	  Configuration config = Configuration.defaultConfiguration();
+	  LogManager logger = BasicLogManager.create(config);
+	  ShutdownManager shutdown = ShutdownManager.create();
+    
+	  //System.out.println(NativeLibraries.getNativeLibraryPath());
+	  
+	  Solvers solver = Solvers.MATHSAT5;
+	  String inputFile = null;
+	  int backStep = 1000;
+	  int mode = 1;
+	  
+	  if(args.length == 0) {
+		  System.out.println("# Error : No input file.");
+		  return;
+	  }
+	  else if(args.length >= 1) {
+		  inputFile = args[0];
+		  if(args.length >= 2) {
[…]

### 2018-09-11 quantifiers/draft.tex
ok

@@
 For two integers $0 \leq i \leq j$, we denote by $[i,j]$ the set
-$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[0,i]$. To simplify the
-presentation, we consider two sorts $\Data$ and $\Bool$, where $\Data$
-is an infinite domain and $\Bool$ is the set of truth values $\top$
-(true) and $\bot$ (false). The $\Data$ sort is equipped with finitely
-many function symbols $f : \Data^{\#(f)} \rightarrow \Data$, where
-$\#(f)\geq0$ denotes the number of arguments (arity) of $f$. When
-$\#(f)=0$, we say that $f$ is a constant. A \emph{predicate} is a
-function symbol $p : \Data^{\#(p)} \rightarrow \Bool$, denoting a
+$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[0,i]$. We consider two
+sorts $\Data$ and $\Bool$, where $\Data$ is an infinite domain and
+$\Bool = \set{\top,\bot}$ is the set of boolean values true ($\top$)
+and false ($\bot$), respectively. The $\Data$ sort is equipped with
+finitely many function symbols $f : \Data^{\#(f)} \rightarrow \Data$,
+where $\#(f)\geq0$ denotes the number of arguments (arity) of
+$f$. When $\#(f)=0$, we say that $f$ is a constant. A \emph{predicate}
+is a function symbol $p : \Data^{\#(p)} \rightarrow \Bool$, denoting a
 relation of arity $\#(p)$ and we write $\preds$ for the set of
-predicates. In the following, we shall consider that the
-interpretation of all function symbols $f : \Data^{\#(f)} \rightarrow
-\Data$ that are not predicates is fixed and relative to the
-interpretation of the $\Data$ sort. For simplicity, we further blur
-the notational distinction between function symbols and their
+predicates.
+
+In the following, we shall consider that the interpretation of all
+function symbols $f : \Data^{\#(f)} \rightarrow \Data$ that are not
+predicates is fixed by the interpretation of the $\Data$ sort,
+e.g.\ if $\Data$ is the set of integers $\zed$, the function symbols
+are zero, the successor function and the arithmetic operations of
+addition and multiplication. For simplicity, we further blur the
+notational distinction between function symbols and their
 interpretations.
@@
 Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
-variables, ranging over the interpretation of the $\Data$ sort. Terms
-are either constants of sort $\Data$, variables or function
-applications. The set of first-order formulae is defined by the syntax
-below:
+variables, ranging over $\Data$. Terms are either constants of sort
+$\Data$, variables or function applications $f(t_1,\ldots,t_{\#(f)})$,
+where $t_1,\ldots,t_{\#(f)}$ are terms. The set of first-order
+formulae is defined by the syntax below:
 \[\phi := t \teq s \mid p(t_1,\ldots,t_{\#(p)}) 
@@
 \]
-For a formula $\phi$, we define $\sem{\phi}_\nu \isdef \set{\I \mid
-  \I,\nu \models \phi}$, where $\nu(x)$ is defined for each $x \in
-\fv{}{\phi}$, and drop the subscript for sentences. A sentence $\phi$
+where $\nu[x\leftarrow d]$ is the valuation which assigns $d$ to $x$
+and behaves like $\nu$ elsewhere. For a formula $\phi$ and a valuation
+$\nu$, we define $\sem{\phi}_\nu \isdef \set{\I \mid \I,\nu \models
+  \phi}$ and drop the $\nu$ subscript for sentences. A sentence $\phi$
 is \emph{satisfiable} (\emph{unsatisfiable}) if
-$\sem{\phi}\neq\emptyset$ [$\sem{\phi}=\emptyset$]. An element of
+$\sem{\phi}\neq\emptyset$ ($\sem{\phi}=\emptyset$). An element of
 $\sem{\phi}$ is called a \emph{model} of $\phi$. A formula $\phi$ is
@@
 Let $\Sigma$ be a finite alphabet $\Sigma$ of \emph{input
-  events}. Given a finite set of variables $X$, we denote by $X
+  events}. Given a finite set of variables $X \subseteq \vars$, we denote by $X
 \mapsto \Data$ the set of valuations of the variables $X$ and
@@
 
-A \emph{first-order alternating automaton} (simply called automaton in
-the following) is a tuple $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$,
-where $\Sigma$ is the finite event alphabet, $X$ is a finite set of
-variables, $Q$ is a finite set of predicates, $\iota$ is a satisfiable
-sentence, $F \subseteq Q$ is the set of \emph{final} predicates and
-$\Delta$ is a set of \emph{transition rules} of the form
-\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
-predicate, $a \in \Sigma$ is an input event, $X \cap
-\set{y_1,\ldots,y_{\#(q)}} = \emptyset$, $\psi$ is a positive formula
-such that $\fv{}{\phi} \subseteq X \cup \set{y_1,\ldots,y_{\#(q)}}$
-and each predicate that occurs in $\psi$ belongs also to $Q$. The
-\emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
+A \emph{first-order alternating automaton} (automaton) is a tuple $\A
+= \tuple{\Sigma,X,Q,\iota,F,\Delta}$, where: \begin{compactitem}
+\item $\Sigma$ is the finite set of input events, 
+\item $X$ is a finite set of input variables, 
+\item $Q$ is a finite set of predicates denoting control states, 
+\item $\iota$ is a sentence defining initial configurations,
+\item $F \subseteq Q$ is the set of predicates denoting final states, and 
+\item $\Delta$ is a set of \emph{transition rules} of the form
+  \(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is
+  a predicate, $a \in \Sigma$ is an input event, $X \cap
+  \set{y_1,\ldots,y_{\#(q)}} = \emptyset$, $\psi$ is a positive
+  formula such that $\fv{}{\phi} \subseteq X \cup
+  \set{y_1,\ldots,y_{\#(q)}}$ and each predicate that occurs (under an
+  even number of negations) in $\psi$ belongs also to $Q$. The
+  quantifiers occurring in $\psi$ are referred to as \emph{transition
+    quantifiers}.
+\end{compactitem}
+The \emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
 \sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
@@
 interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
-configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid
-  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I, q \in Q}$, called a
+configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid q
+  \in Q,~ \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a
 \emph{cube}. For a set $\mathcal{S}$ of interpretations, we define
 $\cube{\mathcal{S}} \isdef \set{\cube{\I} \mid \I \in
-  \mathcal{S}}$. Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in
-\Sigma[X]^*$, an \emph{execution} of
-$\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is a (possibly
-infinite) forest $\T = \set{T_1,T_2,\ldots}$, where each tree $T_i$ is
-such that:
+  \mathcal{S}}$.
+
+Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$, an
+\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is
+a (possibly infinite) forest $\T = \set{T_1,T_2,\ldots}$, where each
+$T_i$ is a tree labeled with configurations, such that:
 \begin{compactitem}
-\item the roots of $T_1, T_2, \ldots$ form a cube from
-  $\cube{\minsem{\iota}}$, % --- because $\iota$ is satisfiable, such
-                           % a cube always exists,
+\item the labels of the roots of $T_1, T_2, \ldots$ form a cube from
+  $\cube{\minsem{\iota}}$,
 %
-% \item all paths from the root to a leaf of $T_i$ have length $n$, and
-%
-\item if $q(d_1,\ldots,d_{\#(q)})$ is a node on the level $j \in
-  [n-1]$ in $T_i$, then its children form a cube from
-  $\cube{\minsem{\psi}_\eta}$, where \(q(y_1,\ldots,y_{\#(q)})
-  \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is a transition rule and
-  $\eta$ is a valuation that assigns each variable $y_k$ the value
-  $d_k$, for $k \in [1,\#(q)]$ and each variable $x \in X$ the value
-  $\nu_{j+1}(x)$. % --- because $\fv{}{\psi} \subseteq X \cup
-                  % \set{y_1,\ldots,y_{\#(q)}}$, the values assigned
-                  % by $\eta$ to variables outside $X \cup
-                  % \set{y_1,\ldots,y_{\#(q)}}$ have no impact on
-                  % $\cube{\minsem{\psi}_\eta}$.
+\item if $q(d_1,\ldots,d_{\#(q)})$ labels a node on the level $j \in
+  [n-1]$ in $T_i$, then the labels of its children form a cube from
+  $\cube{\minsem{\psi}_{\nu_{j+1}[y_1 \leftarrow d_1,\ldots,y_{\#(q)}
+        \leftarrow d_{\#(q)}]}}$, where \(q(y_1,\ldots,y_{\#(q)})
+  \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is a transition rule.
 \end{compactitem}
@@
 \item the frontier of each tree $T \in \T$ is labeled with \emph{final
-  configurations} $q(d_1,\ldots,d_{\#(q)})$, where $q \in F$ is a
-  final predicate.
+  configurations} $q(d_1,\ldots,d_{\#(q)})$, where $q \in F$.
 \end{inparaenum} In this case, we say that $\A$ accepts $w$ and
-denote by $\lang{\A}$ the set of words accepted by $\A$. We address
-the following questions: \begin{compactenum}
+denote by $\lang{\A}$ the set of words accepted by $\A$.
+
+In this paper, we address the following questions: \begin{compactenum}
 \item \emph{boolean closure}: given automata $\A_i =
@@
 
-In the upcoming developments it is more convenient to work with
-logical formulae defining executions. Let $\A =
-\tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the rest of
-this session. For any $i \in \nat$, we denote by $\stamp{Q}{i} =
-\set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} = \set{\stamp{x}{i}
-  \mid x \in X}$ the sets of time-stamped predicates and variables,
-respectively. As a shorthand, we write $\stamp{Q}{\leq n}$ (resp.
-$\stamp{X}{\leq n}$) for the set $\set{\stamp{q}{i} \mid q \in Q, i
-  \in [n]}$ (resp. $\set{\stamp{x}{i} \mid x \in X, i \in [n]}$). For
-a formula $\psi$ and $i \in \nat$, we define $\stamp{\psi}{i} \isdef
-\psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the formula in which all input
-variables and state predicates (and only those symbols) are replaced
-by their time-stamped counterparts. As a shorthand, we shall write
-$q(\vec{y})$ for $q(y_1,\ldots,y_{\#(q)})$, when no confusion arises.
+In the upcoming developments it is sometimes more convenient to work
+with logical formulae defining executions of automata, than with
+low-level execution forests. For this reason, we first introduce
+\emph{path formulae} $\pathform{\alpha}$, which are formulae defining
+the executions of an automaton, over words that share a given sequence
+$\alpha$ of input events.  Second, we restrict a path formula
+$\pathform{\alpha}$ to an \emph{acceptance formula}
+$\accform{\alpha}$, which defines only accepting executions over words
+that share a given input sequence. Otherwise stated,
+$\accform{\alpha}$ is satisfiable if and only if the automaton accepts
+a word $w$ such that $w_\Sigma = \alpha$.
+
+Let $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the
+rest of this section. For any $i \in \nat$, we denote by $\stamp{Q}{i}
+= \set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} =
+\set{\stamp{x}{i} \mid x \in X}$ the sets of time-stamped predicates
+and variables, respectively. As a shorthand, we write $\stamp{Q}{\leq
+  n}$ (resp.  $\stamp{X}{\leq n}$) for the set $\set{\stamp{q}{i} \mid
+  q \in Q, i \in [n]}$ (resp. $\set{\stamp{x}{i} \mid x \in X, i \in
+  [n]}$). For a formula $\psi$ and $i \in \nat$, we define
+$\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the
+formula in which all input variables and state predicates (and only
+those symbols) are replaced by their time-stamped counterparts. As a
+shorthand, we shall write $q(\vec{y})$ for $q(y_1,\ldots,y_{\#(q)})$,
+when no confusion arises.
 
 Given a sequence of input events $\alpha = a_1 \ldots a_n$, the
-\emph{path formula} of $\alpha$ (in $\A$) is defined as follows:
+\emph{path formula} of $\alpha$ is defined as:
 \begin{equation}\label{eq:pathform}
@@
 \[\I_{\T}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
-  q(d_1,\ldots,d_{\#(q)}) \text{ occurs on level $i$ in $\T$}}\]
+  q(d_1,\ldots,d_{\#(q)}) \text{ labels a node on level $i$ in $\T$}}\]
 
@@
   and let $\forall y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$ be
-  the subformula of $\pathform{w_\Sigma}$ corresponding to the
-  application(s) of the transition rule $q(\vec{y})
-  \arrow{a_n}{} \psi$ at the $(n-1)$-th step. Let $\nu$ be the
-  valuation that assigns each $y_i$ the value $d_i$, for all $i \in
-  [1,\#(q)]$ and behaves like $w_\Data$ outside of
-  $\set{y_1,\ldots,y_{\#(d)}}$. Because $\I, w_\Data \models \forall
-  y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$,
-  we have $\I \in \sem{\stamp{\psi}{n}}_{\nu}$ and let $\K$ be one of
-  the minimal interpretations such that $\K \subseteq \I$ and $\K \in
+  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$ be the
+  subformula of $\pathform{w_\Sigma}$ corresponding to the
+  application(s) of the transition rule $q(\vec{y}) \arrow{a_n}{}
+  \psi$ at the $(n-1)$-th step. Let $\nu = w_\Data[y_1 \leftarrow
+    d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$. Because $\I, w_\Data
+  \models \forall y_1 \ldots \forall y_{\#(q)} ~.~
+  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$, we have $\I
+  \in \sem{\stamp{\psi}{n}}_{\nu}$ and let $\K$ be one of the minimal
+  interpretations such that $\K \subseteq \I$ and $\K \in
   \sem{\stamp{\psi}{n}}_{\nu}$. It is not hard to see that $\K$ exists
@@
   \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid q(d_1,\ldots,d_{\#(q)})
-    \text{ occurs on the $n$-th level in $\T$}}$, we have $\I_{\T},
+    \text{ labels a node on the $n$-th level in $\T$}}$, we have $\I_{\T},
   w_\Data \models \varphi$, for each subformula $\varphi = \forall y_1
@@
 \end{equation}
+The top-level universal quantifiers from a subformula $\forall y_1
+\ldots \forall y_{\#(q)} ~.~ \stamp{q}{i}(\vec{y}) \rightarrow \psi$
+of $\accform{\alpha}$ will be referred to as \emph{path quantifiers},
+in the following. Without loss of generality, path quantifiers are
+assumed to be distinct from the transition quantifiers that occur
+within a formula $\psi$ of a transition rule $q(y_1,\ldots,y_{\#(q)})
+\arrow{a(X)}{} \psi$ of $\A$.
+
+Observe that the acceptance formula $\accform{\A}$ is false in every
+interpretation of the predicates that assigns a non-empty set to a
+non-final predicate occurring on the last level in the execution
+forest. The relation between the words accepted by $\A$ and the
+acceptance formula above, is formally captured by the following lemma:
 
@@
   $q(\vec{y}) \arrow{a_{m+1}(X)}{} \psi$ is the transition
-  rule of $\A$ that applies for $q$ and $a_{m+1}$ and $\eta$ stands
-  for $\nu[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow
+  rule of $\A$ that applies for $q$ and $a_{m+1}$ and $\eta=
+  w_\Data[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow
     d_{\#(q)}]$. Hence $\sem{\psi}_{\eta} = \emptyset$, and because
-  $\I,\nu \models \accform{\alpha}$, we obtain that $\I,\eta \models
+  $\I,w_\Data \models \accform{\alpha}$, we obtain that $\I,\eta \models
   q(\vec{y}) \rightarrow \stamp{\psi}{m+1}$, thus
@@
   with a configuration $q(d_1,\ldots,d_{\#(q)})$ such that $q \in Q
-  \setminus F$. Since $\I,\nu \models \forall y_1 \ldots \forall
+  \setminus F$. Since $\I,w_\Data \models \forall y_1 \ldots \forall
   y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot$, by a
@@
 
-In the following, it will be useful, in several places, to define the
-data sequence of an accepted word using the predicate-free formula
-defined below:
+As an immediate consequence, one can decide whether $\A$ accepts some
+word $w$ with a given input sequence $w_\Sigma=\alpha$, by checking
+whether $\accform{\alpha}$ is satisfiable. However, unlike
+non-alternating infinite-state models of computation, such as counter
+automata (nondeterministic programs with integer variables), the
+satisfiability query for an acceptance (path) formula falls outside of
+known decidable theories, supported by standard SMT solvers. There are
+basically two reasons for this, namely \begin{inparaenum}[(i)]
+\item the presence of predicate symbols, and
+\item non-trivial alternation of quantifiers.
+\end{inparaenum}
+To understand this point, consider for example, the quantified theory
+of Presburger arithmetic. Adding even only one monadic predicate
+symbol to it yields undecidability in the presence of non-trivial
+quantifier alternation \cite{Halpern91}, however the quantifier-free
+fragment with predicate symbols is decidable\footnote{Linear integer
+  arithmetic with uninterpreted function symbols is decidable, by a
+  Nelson-Oppen style theory combination argument.}.
+
+To tackle this problem, we start from the observation that acceptance
+formulae have a particular form that allows the elimination of path
+quantifiers and of predicates, by a couple of equivalence-preserving
+transformations. The result will be a formula with no predicate
+symbols, whose only quantifiers are the transition quantifiers
+introduced by the transition rules of the automaton. We shall further
+assume (\S\ref{sec:emptiness}) that the first-order theory of the data
+sort $\Data$ has quantifier elimination, which allows to effectively
+decide these queries.
+
+For the time being, let us formally define the elimination of
+transition quantifiers and predicates, respectively. Given a sequence
+of input events $\alpha = a_1 \ldots a_n$, consider the acceptance
+formula: 
+\[\accform{\alpha} = \stamp{\iota}{0} \wedge 
+\bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in
+  \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
+\stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i} \wedge
+\bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
+~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot\]
+
+\begin{definition}\label{def:quantform}
+  Let $\quantformA{0}{\alpha}, \ldots, \quantformA{n}{\alpha}$ be the
+  following sequence of formulae: \begin{compactitem}
+  \item $\quantformA{0}{\alpha} \isdef \stamp{\iota}{0}$, 
+  \item $\quantformA{i}{\alpha} \isdef \quantformA{i-1}{\alpha} \wedge
+    \bigwedge \set{\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow
+      \stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}] \mid
+      \stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \text{ occurs in }
+      \quantformA{i-1}{\alpha}}$, for all $i \in [1,n]$.    
+  \end{compactitem}
+  We write $\quantform{\alpha}$ instead of $\quantformA{n}{\alpha}$,
+  when no confusion arises.
+\end{definition}
+Observe that $\quantform{\alpha}$ contains no path quantifiers, as
+required. On the other hand, the scope of zero or more transition
+quantifiers might extend outside the formula given by the transition
+rule, as shown by the following example:
+
+\begin{example}\label{ex:path-quant-elim}
+
+\end{example}
 
@@
 \subsection{Closure Properties}
+\label{sec:closure}
 
@@
 \section{The Emptiness Problem}
+\label{sec:emptiness}
 

### 2018-09-14 quantifiers/draft.tex
ok

@@
 \usepackage{tikz}
+% \usepackage{hyperref}
 \usetikzlibrary{automata,positioning, calc}
@@
   \set{y_1,\ldots,y_{\#(q)}}$ and each predicate that occurs (under an
-  even number of negations) in $\psi$ belongs also to $Q$. The
-  quantifiers occurring in $\psi$ are referred to as \emph{transition
-    quantifiers}.
+  even number of negations) in $\psi$ belongs also to $Q$. 
 \end{compactitem}
-The \emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
+The quantifiers occurring in the right-hand side formula of a
+transition rule are referred to as \emph{transition quantifiers}. The
+\emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
 \sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
@@
 $\accform{\alpha}$ is satisfiable if and only if the automaton accepts
-a word $w$ such that $w_\Sigma = \alpha$.
+a word $w$ such that $\event{w} = \alpha$.
 
@@
 \[\I_{\T}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
-  q(d_1,\ldots,d_{\#(q)}) \text{ labels a node on level $i$ in $\T$}}\]
+  q(d_1,\ldots,d_{\#(q)}) \text{ labels a node on level $i$ in $\T$}} 
+,~\forall q\in Q ~\forall i \in [n]\]
 
@@
   word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have 
-  \(\minsem{\pathform{w_\Sigma}}_{w_\Data} = \set{\I_{\T} \mid \T
+  \(\minsem{\pathform{\event{w}}}_{\data{w}} = \set{\I_{\T} \mid \T
     \text{ is an execution of $\A$ over $w$}}\). 
@@
 \proof{ ``$\subseteq$'' Let $\I$ be a minimal interpretation such that
-  $\I,w_\Data \models \pathform{w_\Sigma}$. We show that there exists
+  $\I,\data{w} \models \pathform{\event{w}}$. We show that there exists
   an execution $\T$ of $\A$ over $w$ such that $\I = \I_\T$, by
   induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
-  $\pathform{w_\Sigma} = \stamp{\iota}{0}$. Because $\iota$ is a
-  sentence, the valuation $w_\Data$ is not important in $\I,w_\Data
+  $\pathform{\event{w}} = \stamp{\iota}{0}$. Because $\iota$ is a
+  sentence, the valuation $\data{w}$ is not important in $\I,\data{w}
   \models \stamp{\iota}{0}$ and, moreover, since $\I$ is minimal, we
@@
   \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$ be the
-  subformula of $\pathform{w_\Sigma}$ corresponding to the
+  subformula of $\pathform{\event{w}}$ corresponding to the
   application(s) of the transition rule $q(\vec{y}) \arrow{a_n}{}
-  \psi$ at the $(n-1)$-th step. Let $\nu = w_\Data[y_1 \leftarrow
-    d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$. Because $\I, w_\Data
+  \psi$ at the $(n-1)$-th step. Let $\nu = \data{w}[y_1 \leftarrow
+    d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$. Because $\I, \data{w}
   \models \forall y_1 \ldots \forall y_{\#(q)} ~.~
@@
   that $\I_{\T}$ is a minimal interpretation such that $\I_{\T},
-  w_\Data \models \pathform{w_\Sigma}$, by induction on $n \geq
+  \data{w} \models \pathform{\event{w}}$, by induction on $n \geq
   0$. For $n=0$, $\T$ is a cube from $\cube{\minsem{\iota}}$, by
@@
     \text{ labels a node on the $n$-th level in $\T$}}$, we have $\I_{\T},
-  w_\Data \models \varphi$, for each subformula $\varphi = \forall y_1
+  \data{w} \models \varphi$, for each subformula $\varphi = \forall y_1
   \ldots \forall y_{\#(q)} ~.~ \stamp{q}{n-1}(\vec{y})
-  \rightarrow \stamp{\psi}{n}$ of $\pathform{w_\Sigma}$, by the
+  \rightarrow \stamp{\psi}{n}$ of $\pathform{\event{w}}$, by the
   execution semantics of $\A$. This is the case because the children
@@
   $\cube{\minsem{\psi}_\nu}$, where $\nu$ is a valuation that assigns
-  each $y_i$ the value $d_i$ and behaves like $w_\Data$,
+  each $y_i$ the value $d_i$ and behaves like $\data{w}$,
   otherwise. Now supppose, for a contradiction, that $\I_{\T}$ is not
   minimal and let $\J \subsetneq \I_{\T}$ be an interpretation such
-  that $\J,w_\Data \models \pathform{w_\Sigma}$. First, we show that
+  that $\J,\data{w} \models \pathform{\event{w}}$. First, we show that
   the restriction $\J'$ of $\J$ to $\bigcup_{i=0}^{n-1} \stamp{Q}{i}$
@@
 given sequence of input events $\alpha \in \Sigma^*$. To this end, we
-extend the path formula $\pathform{\alpha}$ with a characterization of
-the set of final states of $\A$: 
+constrain the path formula $\pathform{\alpha}$ by requiring that only
+final states of $\A$ occur on the last level of the execution.  The
+result is the \emph{acceptance formula} for $\alpha$:
 \begin{equation}\label{eq:accform}
@@
 of $\accform{\alpha}$ will be referred to as \emph{path quantifiers},
-in the following. Without loss of generality, path quantifiers are
-assumed to be distinct from the transition quantifiers that occur
-within a formula $\psi$ of a transition rule $q(y_1,\ldots,y_{\#(q)})
-\arrow{a(X)}{} \psi$ of $\A$.
-
-Observe that the acceptance formula $\accform{\A}$ is false in every
-interpretation of the predicates that assigns a non-empty set to a
-non-final predicate occurring on the last level in the execution
-forest. The relation between the words accepted by $\A$ and the
-acceptance formula above, is formally captured by the following lemma:
+in the following. Notice that path quantifiers are distinct from the
+transition quantifiers that occur within a formula $\psi$ of a
+transition rule $q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi$ of $\A$.
+
+The acceptance formula $\accform{\A}$ is false in every interpretation
+of the predicates that assigns a non-empty set to a non-final
+predicate occurring on the last level in the execution forest. The
+relation between the words accepted by $\A$ and the acceptance formula
+above, is formally captured by the following lemma:
 
@@
   \item\label{it1:lemma:acceptance} there exists an interpretation
-    $\I$ such that $\I,w_\Data \models \accform{w_\Sigma}$, 
+    $\I$ such that $\I,\data{w} \models \accform{\event{w}}$, 
     %
@@
   (\ref{it2:lemma:acceptance})$'' Let $\I$ be an interpretation such
-  that $\I,w_\Data \models \accform{w_\Sigma}$. By Lemma
+  that $\I,\data{w} \models \accform{\event{w}}$. By Lemma
   \ref{lemma:path-formula}, $\A$ has an execution $\T$ over $w$ such
@@
   rule of $\A$ that applies for $q$ and $a_{m+1}$ and $\eta=
-  w_\Data[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow
+  \data{w}[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow
     d_{\#(q)}]$. Hence $\sem{\psi}_{\eta} = \emptyset$, and because
-  $\I,w_\Data \models \accform{\alpha}$, we obtain that $\I,\eta \models
+  $\I,\data{w} \models \accform{\alpha}$, we obtain that $\I,\eta \models
   q(\vec{y}) \rightarrow \stamp{\psi}{m+1}$, thus
@@
   with a configuration $q(d_1,\ldots,d_{\#(q)})$ such that $q \in Q
-  \setminus F$. Since $\I,w_\Data \models \forall y_1 \ldots \forall
+  \setminus F$. Since $\I,\data{w} \models \forall y_1 \ldots \forall
   y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot$, by a
@@
   (\ref{it1:lemma:acceptance})$'' Let $\T$ be an accepting execution
-  of $\A$ over $w$. We prove that $\I_\T,w_\Data \models
-  \accform{w_\Sigma}$. By Lemma \ref{lemma:path-formula}, we obtain
-  $\I_\T,w_\Data \models \pathform{w_\Sigma}$. Since every path in
+  of $\A$ over $w$. We prove that $\I_\T,\data{w} \models
+  \accform{\event{w}}$. By Lemma \ref{lemma:path-formula}, we obtain
+  $\I_\T,\data{w} \models \pathform{\event{w}}$. Since every path in
   $\T$ is of length $n$ and all nodes on the $n$-th level of $\T$ are
-  labeled by final configurations, we obtain that $\I_\T,w_\Data
+  labeled by final configurations, we obtain that $\I_\T,\data{w}
   \models \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
@@
 As an immediate consequence, one can decide whether $\A$ accepts some
-word $w$ with a given input sequence $w_\Sigma=\alpha$, by checking
+word $w$ with a given input sequence $\event{w}=\alpha$, by checking
 whether $\accform{\alpha}$ is satisfiable. However, unlike
@@
 \item the presence of predicate symbols, and
-\item non-trivial alternation of quantifiers.
+\item the non-trivial alternation of quantifiers.
 \end{inparaenum}
-To understand this point, consider for example, the quantified theory
-of Presburger arithmetic. Adding even only one monadic predicate
-symbol to it yields undecidability in the presence of non-trivial
-quantifier alternation \cite{Halpern91}, however the quantifier-free
-fragment with predicate symbols is decidable\footnote{Linear integer
-  arithmetic with uninterpreted function symbols is decidable, by a
-  Nelson-Oppen style theory combination argument.}.
+To understand this point, consider for example, the decidable theory
+of Presburger arithmetic \cite{Presburger29}. Adding even only one
+monadic predicate symbol to it yields undecidability in the presence
+of non-trivial quantifier alternation \cite{Halpern91}. However the
+quantifier-free fragment of Presburger arithmetic extended with
+predicate symbols can be shown to be decidable, using a Nelson-Oppen
+style congruence closure argument \cite{NelsonOppen80}.
 
 To tackle this problem, we start from the observation that acceptance
-formulae have a particular form that allows the elimination of path
-quantifiers and of predicates, by a couple of equivalence-preserving
-transformations. The result will be a formula with no predicate
-symbols, whose only quantifiers are the transition quantifiers
-introduced by the transition rules of the automaton. We shall further
+formulae have a particular form, which allows the elimination of path
+quantifiers and of predicates, by a couple of
+satisfiability-preserving transformations. The result of applying
+these transformations is a formula with no predicate symbols, whose
+only quantifiers are those introduced by the transition rules of the
+automaton, referred to as transition quantifiers. We shall further
 assume (\S\ref{sec:emptiness}) that the first-order theory of the data
 sort $\Data$ has quantifier elimination, which allows to effectively
-decide these queries.
+decide the satisfiability of such formulae.
 
 For the time being, let us formally define the elimination of
-transition quantifiers and predicates, respectively. Given a sequence
-of input events $\alpha = a_1 \ldots a_n$, consider the acceptance
-formula: 
-\[\accform{\alpha} = \stamp{\iota}{0} \wedge 
-\bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in
-  \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-\stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i} \wedge
-\bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
-~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot\]
+transition quantifiers and predicates, respectively. Consider a given
+sequence of input events $\alpha = a_1 \ldots a_n$ and denote by
+$\alpha_i$ the prefix $a_1 \ldots a_i$ of $\alpha$, for $i \in [n]$,
+where $\alpha_0=\epsilon$.
+
+%% \[\accform{\alpha} = \stamp{\iota}{0} \wedge 
+%% \bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in
+%%   \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
+%% \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i} \wedge
+%% \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
+%% ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot\]
 
 \begin{definition}\label{def:quantform}
-  Let $\quantformA{0}{\alpha}, \ldots, \quantformA{n}{\alpha}$ be the
-  following sequence of formulae: \begin{compactitem}
-  \item $\quantformA{0}{\alpha} \isdef \stamp{\iota}{0}$, 
-  \item $\quantformA{i}{\alpha} \isdef \quantformA{i-1}{\alpha} \wedge
-    \bigwedge \set{\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow
-      \stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}] \mid
-      \stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \text{ occurs in }
-      \quantformA{i-1}{\alpha}}$, for all $i \in [1,n]$.    
-  \end{compactitem}
-  We write $\quantform{\alpha}$ instead of $\quantformA{n}{\alpha}$,
-  when no confusion arises.
+  Let $\quantformn{\alpha_0}, \ldots, \quantformn{\alpha_n}$ be the
+  sequence of formulae defined by $\quantformn{\alpha_0} \isdef
+  \stamp{\iota}{0}$ and, for all $i \in [1,n]$:
+  \[\quantformn{\alpha_i} \isdef \quantformn{\alpha_{i-1}} \wedge
+  \bigwedge_{\begin{array}{l}
+      \scriptstyle{\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_{i-1}}} \\
+      \scriptstyle{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta} 
+  \end{array}} \stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow 
+    \stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]
+      \]  We write
+    $\quantform{\alpha}$ for the prenex normal form of the formula: 
+      \[\quantformn{\alpha_n} \wedge \bigwedge_{\begin{array}{l}
+          \scriptstyle{\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_n}} \\
+          \scriptstyle{q \in Q \setminus F}
+      \end{array}} \stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot \enspace.\]
 \end{definition}
 Observe that $\quantform{\alpha}$ contains no path quantifiers, as
-required. On the other hand, the scope of zero or more transition
-quantifiers might extend outside the formula given by the transition
-rule, as shown by the following example:
-
-\begin{example}\label{ex:path-quant-elim}
-
+required. On the other hand, the scope of the transition quantifiers
+in $\quantform{\alpha}$ exceeds the right-hand side formulae from the
+transition rules, as shown by the following example.
+
+\begin{example}\label{ex:quant-elim}
+  Consider the automaton $\A = \tuple{\set{a_1,a_2}, \set{x}, \set{q},
+    \iota, \set{q}, \Delta}$, where:
+  \[\begin{array}{rcl}
+  \iota & = & \exists z ~.~ q(z) \wedge z \geq 0 \\
+  \Delta & = & \set{q(y) \arrow{a_1(x)}{} \forall z ~.~ z \leq x+y \rightarrow q(y-z+x),~
+    q(y) \arrow{a_2(x)}{} \exists z ~.~ z \geq x-y \wedge q(y+z+x)}
+  \end{array}\]
+  For the input event sequence $\alpha = a_1a_2$, the path and acceptance formulae are:
+  \[\begin{array}{rcl}
+  \pathform{\alpha} = \accform{\alpha} & = & \exists z ~.~ \stamp{q}{0}(z) \wedge z \geq 0 \\
+  && \forall y ~.~ \stamp{q}{0}(y) \rightarrow (\forall z ~.~ z\leq\stamp{x}{1}+y \rightarrow \stamp{q}{1}(y-z+\stamp{x}{1})) \\
+  && \forall y ~.~ \stamp{q}{1}(y) \rightarrow \exists z ~.~ z\geq\stamp{x}{2}-y \wedge \stamp{q}{2}(y+z+\stamp{x}{2})
+  \end{array}\]
+  The result of eliminating the path quantifiers, in prenex normal form, is shown below:
+  \[\begin{array}{rcl}
+  \quantform{\alpha} & = & \exists z_1\forall z_2\exists z_3 ~.~ \stamp{q}{0}(z_1) \wedge z_1 \geq 0 \\
+  && \wedge~ [\stamp{q}{0}(z_1) \rightarrow (z_2 \leq \stamp{x}{1}+z_1 \rightarrow \stamp{q}{1}(z_1-z_2+\stamp{x}{1}))] \\
+  && \wedge~ [\stamp{q}{1}(z_1-z_2+\stamp{x}{1}) \rightarrow z_3 \geq \stamp{x}{2} -z_1+z_2-\stamp{x}{1} \wedge 
+    \stamp{q}{2}(z_1-z_2+\stamp{x}{1}+z_3+\stamp{x}{2})] \enspace\enspace\hfill\blacksquare
+  \end{array}\]
 \end{example}
 
+The next lemma establishes a formal relation between the
+satisfiability of an acceptance formula $\accform{\alpha}$ and that of
+the formula $\quantform{\alpha}$, obtained by eliminating the path
+quantifiers from $\accform{\alpha}$.
+
+\begin{lemma}\label{lemma:quant}
+  For any input event sequence $\alpha=a_1\ldots a_n$ and each
+  valuation $\nu : \stamp{X}{\leq n} \rightarrow \Data$, the following
+  hold: \begin{compactenum}
+  \item\label{it1:quant} for all interpretations $\I$, if $\I,\nu
+    \models \accform{\alpha}$ then $\I,\nu \models
+    \quantform{\alpha}$. 
+    %
+  \item\label{it2:quant} if there exists an interpretation $\I$ such
+    that $\I,\nu \models \quantform{\alpha}$ then there exists an
+    interpretation $\J \subseteq \I$ such that $\J,\nu \models
+    \accform{\alpha}$. 
+  \end{compactenum}
+\end{lemma}
+\proof{(\ref{it1:quant}) Trivial, since every subformula
+  $q(t_1,\ldots,t_{\#(q)}) \rightarrow
+  \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ of $\quantform{\alpha}$ is
+  entailed by a subformula $\forall y_1 \ldots \forall y_{\#(q)} ~.~
+  q(y_1, \ldots, y_{\#(q)}) \rightarrow \psi$ of $\accform{\alpha}$.
+
+  \noindent(\ref{it2:quant}) By repeated applications of the following fact:
+
+  \begin{fact}
+    Given formulae $\phi$ and $\psi$, such that no predicate atom with
+    predicate symbol $q$ occurs in $\psi(y_1,\ldots,y_{\#(q)})$, for
+    each valuation $\nu$, if there exists an interpretation $\I$ such
+    that $\I,\nu \models \phi \wedge
+    \bigwedge_{q(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \phi}
+    q(t_1,\ldots,t_{\#(q)}) \rightarrow
+    \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ then there exists a
+    valuation $\J$ such that $\J(q) \subseteq \I(q)$ and $\J(q') =
+    \I(q')$ for all $q' \in Q \setminus \set{q}$ and $\J,\nu \models
+    \phi \wedge \forall y_1 \ldots \forall y_{\#(q)} ~.~
+    q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$.
+  \end{fact}
+  \proof{Assume w.l.o.g. that $\phi$ is quantifier free. The proof can
+    be easily generalized to the case $\phi$ has quantifiers. Let
+    $\J(q) = \set{\langle t_1^\nu, \ldots, t_{\#(q)}^\nu \rangle \in
+      \I(q) \mid q(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \phi}$ and
+    $\J(q') = \I(q')$ for all $q' \in Q \setminus \set{q}$. Since
+    $\I,\nu \models \phi$, we obtain that also $\J,\nu \models \phi$
+    because the tuples of values in $\I(q) \setminus \J(q)$ are not
+    interpretations of terms that occur within subformulae
+    $q(t_1,\ldots,t_{\#(q)})$ of $\phi$. Moreover,
+    $\bigwedge_{q(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \phi}
+    q(t_1,\ldots,t_{\#(q)}) \rightarrow
+    \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ and $\forall y_1 \ldots
+    \forall y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$
+    are equivalent under $\J$, thus $\J,\nu \models \forall y_{\#(q)}
+    ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$, as required. \qed}
+
+  \noindent This concludes the proof. \qed}
+
+We proceed with the elimination of predicate atoms from
+$\quantform{\alpha}$, defined below.
+
 \begin{definition}\label{def:substform}
-Given a sequence of input events $\alpha = a_1 \ldots a_n$, let
-$\substform{\alpha}$ be the formula obtained by successively replacing
-in $\accform{\alpha}$: \begin{compactenum}[(a)]
-\item\label{it1:substform} each occurrence of a predicate
-  $\stamp{q}{i}(t_1,\ldots,t_{\#(q)})$ by the formula
-  $\stamp{\psi}{i+1}[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$, where
-  $q(\vec{y}) \arrow{a_{i+1}(X)}{} \psi \in \Delta$ is the transition
-  rule for $q$ and $a_i$, for all $i=0,\ldots,n-1$,
-%
-\item\label{it2:substform} each occurrence of a predicate
-  $\stamp{q}{n}$, such that $q \in Q \setminus F$, by $\top$, and
-%
-\item\label{it3:substform} each occurrence of a predicate
-  $\stamp{q}{n}$, such that $q \in F$, by $\bot$.
-\end{compactenum}
+  Let $\substformn{\alpha_0}, \ldots, \substformn{\alpha_n}$ be the
+  sequence of formulae defined by $\substformn{\alpha_0} \isdef
+  \stamp{\iota}{0}$ and, for all $i \in [1,n]$,
+  $\substformn{\alpha_i}$ is obtained by replacing each occurrence of
+  a predicate atom $\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)})$ in
+  $\substformn{\alpha_{i-1}}$ with the formula
+  $\stamp{\psi}{i}[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$, where
+  $q(\vec{y}) \arrow{a_{i}(X)}{} \psi \in \Delta$. We write
+  $\substform{\alpha}$ for the formula obtained by replacing, in
+  $\substformn{\alpha}$, each occurrence of a predicate
+  $\stamp{q}{n}$, such that $q \in Q \setminus F$ (resp. $q \in F$),
+  by $\bot$ (resp. $\top$).
 \end{definition}
-Since there are no occurrences of predicates in $\substform{\alpha}$,
-for each valuation $\nu : \stamp{X}{\leq n} \rightarrow \Data$, there
+
+\begin{example}[Contd. from Example \ref{ex:quant-elim}]\label{ex:pred-elim}
+  The result of the elimination of predicate atoms from the acceptance
+  formula in Example \ref{ex:quant-elim} is shown below:
+  \[\exists z_1 \forall z_2 \exists z_3 ~.~ (z_2 \leq \stamp{x}{1}+z_1 
+  \rightarrow z_3 \geq \stamp{x}{2} -z_1+z_2-\stamp{x}{1}) \wedge z_1
+  \geq 0\] Since this formula is satisfiable, by Lemma
+  \ref{lemma:quant-pred-acceptance} below, there exists a word $w$
+  with input event sequence $\event{w} = a_1a_2$ that is accepted by the
+  automaton $\A$ from Example
+  \ref{ex:quant-elim}. \hfill$\blacksquare$
+\end{example}
+
+At this point, we prove the formal relation between the satisfiability
+of the formulae $\quantform{\alpha}$ and $\substform{\alpha}$. Since
+there are no occurrences of predicates in $\substform{\alpha}$, for
+each valuation $\nu : \stamp{X}{\leq n} \rightarrow \Data$, there
 exists an interpretation $\I$ such that $\I,\nu \models
 \substform{\alpha}$ if and only if $\J,\nu \models
-\substform{\alpha}$, for every interpretation $\J$. Thus we omit $\I$
-and simply write $\nu \models \substform{\alpha}$ in this case.
-
-\begin{proposition}\label{prop:subst}
-  Let $\phi$ be a positive formula and let $q(t_1,\ldots,t_{\#(q)})$
-  be the only predicate atom occurring within $\phi$. Then, for every
-  valuation $\nu$ of $\fv{}{\phi}$ and every formula $\psi$ with no
-  predicate occurrences, the following are
-  equivalent: \begin{compactenum}
-  \item\label{it1:prop:subst} $\nu \models \phi[\psi[t_1/y_1, \ldots,
-      t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$, 
-    %
-  \item\label{it2:prop:subst} $\I,\nu \models \phi \wedge \forall y_1
-    \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow
-    \psi$, for some interpretation $\I$.
-  \end{compactenum}
-  where $y_1,\ldots,y_{\#(q)} \not\in \fv{}{\phi}$. 
-\end{proposition}
-\proof{ We give the proof in the case $\phi$ is quantifier-free. The
-  generalization to a non-empty quantifier prefix is straightforward.
-  ``$(\ref{it1:prop:subst}) \Rightarrow (\ref{it2:prop:subst})$'' We
-  distinguish two cases:
-  \begin{compactitem}
-  \item if $\nu \models \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$
-    then let $\I(q) = \set{\tuple{t_1^\nu,\ldots,t_{\#(q)}^\nu}}$ and
-    $\emptyset$ everywhere else. We have $\I,\nu \models
-    q(t_1,\ldots,t_{\#(q)})$ and, since $\phi$ is positive, replacing
-    $\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ with
-    $q(t_1,\ldots,t_{\#(q)})$ in $\phi$ does not change its
-    satisfiability under $\I,\nu$, hence $\I,\nu \models
-    \phi$. Moreover, $\nu \models \psi[t_1/y_1, \ldots,
-      t_{\#(q)}/y_{\#(q)}]$ thus $\I,\nu \models \forall y_1 \ldots
-    \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \psi$.
-    %
-  \item else $\nu \not\models \psi[t_1/y_1, \ldots,
-    t_{\#(q)}/y_{\#(q)}]$ and let $\I$ be the interpretation assigning
-    each predicate symbol with $\emptyset$. Then replacing
-    $\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ with
-    $q(t_1,\ldots,t_{\#(q)})$ in $\phi$ does not change its
-    satisfiability under $\I,\nu$. Moreover, since $\I$ maps $q$ to
-    $\emptyset$, we have $\I,\nu \models \forall y_1 \ldots \forall
-    y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \psi$, trivially.
+\substform{\alpha}$, for every interpretation $\J$. In this case we
+omit $\I$ and simply write $\nu \models \substform{\alpha}$.
+
+\begin{lemma}\label{lemma:subst}
+  For any input event sequence $\alpha=a_1\ldots a_n$ and each
+  valuation $\nu : \stamp{X}{\leq n} \rightarrow \Data$, there exists
+  a valuation $\I$ such that $\I,\nu \models \quantform{\alpha}$ if
+  and only if $\nu \models \substform{\alpha}$. 
+\end{lemma}
+\proof{ By induction on $n\geq0$. The base case $n=0$ is trivial,
+  since $\quantform{A} = \substform{A} = \stamp{\iota}{0}$. For the
+  induction step, we rely on the following fact:
+
+  \begin{fact}
+    Given formulae $\phi$ and $\psi$, such that $\phi$ is positive
+    $q(t_1,\ldots,t_{\#(q)})$ is the only one occurrence of the
+    predicate symbol $q$ in $\phi$ and no predicate atom with
+    predicate symbol $q$ occurs in $\psi(y_1,\ldots,y_{\#(q)})$, for
+    each interpretation $\I$ and each valuation $\nu$, we have:
+    \[\begin{array}{l} 
+    \I,\nu \models \phi \wedge q(t_1,\ldots,t_{\#(q)}) \rightarrow
+    \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}] \iff \\ \nu \models
+    \phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]\enspace.
+    \end{array}\]
+  \end{fact}
+  \proof{ We assume w.l.o.g. that $\phi$ is quantifier-free. The proof
+    can be easily generalized to the case $\phi$ has quantifiers.
+
+    \noindent''$\Rightarrow$'' 
+    We distinguish two cases: \begin{compactitem}
+    \item if $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \in \I(q)$ then
+      $\I,\nu \models \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$. Since
+      $\phi$ is positive, replacing $q(t_1,\ldots,t_{\#(q)})$ with
+      $\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ does not change the
+      truth value of $\phi$ under $\nu$, thus $\nu \models
+      \phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$.
+      %
+    \item else, $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \not\in
+      \I(q)$, thus $\nu \models
+      \phi[\bot/q(t_1,\ldots,t_{\#(q)})]$. Since $\phi$ is positive
+      and $\bot$ entails $\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$,
+      we obtain $\nu \models
+      \phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$
+      by monotonicity. 
     \end{compactitem}
 
-  ``$(\ref{it2:prop:subst}) \Rightarrow (\ref{it1:prop:subst})$'' We
+    \noindent``$\Leftarrow$'' Let $\I$ be any interpretation such that
+    $\I(q) = \set{\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \mid \nu
+      \models \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]}$. We
     distinguish two cases: \begin{compactitem}
-  \item if $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \in \I(q)$, since
-    $\I,\nu \models \forall y_1 \ldots \forall y_{\#(q)} ~.~
-    q(\vec{y}) \rightarrow \psi$, we have $\nu \models
-    \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$. Since $\phi$ is
-    positive, replacing $q(t_1,\ldots,t_{\#(q)})$ by $\psi$ in $\phi$
-    does not change its satisfiability under $\nu$, hence $\nu \models
-    \phi[\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]/q(t_1, \ldots,
-      t_{\#(q)})]$.
-    %
-  \item else $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \not\in \I(q)$,
-    that is $\I,\nu \not\models q(t_1,\ldots,t_{\#(q)})$. Then $\nu
-    \models \phi[\bot/q(t_1,\ldots,t_{\#(q)})]$ and we obtain $\nu
-    \models \phi[\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]/q(t_1,
-      \ldots, t_{\#(q)})]$, because $\phi$ is positive, thus
-    $\phi[\bot/q(t_1,\ldots,t_{\#(q)})] \models \phi[\psi[t_1/y_1,
-        \ldots, t_{\#(q)}/y_{\#(q)}]/q(t_1, \ldots, t_{\#(q)})]$. \qed
+      \item if $\I(q) \not \emptyset$ then $\I,\nu \models
+        q(t_1,\ldots,t_{\#(q)})$ and $\nu \models \psi[t_1/y_1,
+          \ldots, t_{\#(q)}/y_{\#(q)}]$. Thus replacing $\psi[t_1/y_1,
+          \ldots, t_{\#(q)}/y_{\#(q)}]$ by $q(t_1,\ldots,t_{\#(q)})$
+        does not change the truth value of $\phi$ under $\I$ and
+        $\nu$, and we obtain $\I,\nu \models \phi$. Moreover, $\I,\nu
+        \models \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ implies
+        $\I,\nu \models q(t_1,\ldots,t_{\#(q)}) \rightarrow
+        \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$.
+        %
+      \item else $\I(q) = \emptyset$, hence $\nu \not\models
+        \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$, thus $\nu \models
+        \phi[\bot/q(t_1,\ldots,t_{\#(q)})]$. Because $\phi$ is
+        positive, we obtain $\I,\nu \models \phi$ by monotonicity. But
+        $\I,\nu \models q(t_1,\ldots,t_{\#(q)}) \rightarrow
+        \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ trivially, because
+        $\I,\nu \not\models q(t_1,\ldots,t_{\#(q)})$.\qed
     \end{compactitem}}
-
-\begin{lemma}\label{lemma:substitution}
+This concludes the proof. \qed}
+
+%% \begin{proposition}\label{prop:subst}
+%%   Let $\phi$ be a positive formula and let $q(t_1,\ldots,t_{\#(q)})$
+%%   be the only predicate atom occurring within $\phi$. Then, for every
+%%   valuation $\nu$ of $\fv{}{\phi}$ and every formula $\psi$ with no
+%%   predicate occurrences, the following are
+%%   equivalent: \begin{compactenum}
+%%   \item\label{it1:prop:subst} $\nu \models \phi[\psi[t_1/y_1, \ldots,
+%%       t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$, 
+%%     %
+%%   \item\label{it2:prop:subst} $\I,\nu \models \phi \wedge \forall y_1
+%%     \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow
+%%     \psi$, for some interpretation $\I$.
+%%   \end{compactenum}
+%%   where $y_1,\ldots,y_{\#(q)} \not\in \fv{}{\phi}$. 
+%% \end{proposition}
+%% \proof{ We give the proof in the case $\phi$ is quantifier-free. The
+%%   generalization to a non-empty quantifier prefix is straightforward.
+%%   ``$(\ref{it1:prop:subst}) \Rightarrow (\ref{it2:prop:subst})$'' We
+%%   distinguish two cases:
+%%   \begin{compactitem}
+%%   \item if $\nu \models \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$
+%%     then let $\I(q) = \set{\tuple{t_1^\nu,\ldots,t_{\#(q)}^\nu}}$ and
+%%     $\emptyset$ everywhere else. We have $\I,\nu \models
+%%     q(t_1,\ldots,t_{\#(q)})$ and, since $\phi$ is positive, replacing
+%%     $\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ with
+%%     $q(t_1,\ldots,t_{\#(q)})$ in $\phi$ does not change its
+%%     satisfiability under $\I,\nu$, hence $\I,\nu \models
+%%     \phi$. Moreover, $\nu \models \psi[t_1/y_1, \ldots,
+%%       t_{\#(q)}/y_{\#(q)}]$ thus $\I,\nu \models \forall y_1 \ldots
+%%     \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \psi$.
+%%     %
+%%   \item else $\nu \not\models \psi[t_1/y_1, \ldots,
+%%     t_{\#(q)}/y_{\#(q)}]$ and let $\I$ be the interpretation assigning
+%%     each predicate symbol with $\emptyset$. Then replacing
+%%     $\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ with
+%%     $q(t_1,\ldots,t_{\#(q)})$ in $\phi$ does not change its
+%%     satisfiability under $\I,\nu$. Moreover, since $\I$ maps $q$ to
+%%     $\emptyset$, we have $\I,\nu \models \forall y_1 \ldots \forall
+%%     y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \psi$, trivially.
+%%   \end{compactitem}
+
+%%   ``$(\ref{it2:prop:subst}) \Rightarrow (\ref{it1:prop:subst})$'' We
+%%   distinguish two cases: \begin{compactitem}
+%%   \item if $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \in \I(q)$, since
+%%     $\I,\nu \models \forall y_1 \ldots \forall y_{\#(q)} ~.~
+%%     q(\vec{y}) \rightarrow \psi$, we have $\nu \models
+%%     \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$. Since $\phi$ is
+%%     positive, replacing $q(t_1,\ldots,t_{\#(q)})$ by $\psi$ in $\phi$
+%%     does not change its satisfiability under $\nu$, hence $\nu \models
+%%     \phi[\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]/q(t_1, \ldots,
+%%       t_{\#(q)})]$.
+%%     %
+%%   \item else $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \not\in \I(q)$,
+%%     that is $\I,\nu \not\models q(t_1,\ldots,t_{\#(q)})$. Then $\nu
+%%     \models \phi[\bot/q(t_1,\ldots,t_{\#(q)})]$ and we obtain $\nu
+%%     \models \phi[\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]/q(t_1,
+%%       \ldots, t_{\#(q)})]$, because $\phi$ is positive, thus
+%%     $\phi[\bot/q(t_1,\ldots,t_{\#(q)})] \models \phi[\psi[t_1/y_1,
+%%         \ldots, t_{\#(q)}/y_{\#(q)}]/q(t_1, \ldots, t_{\#(q)})]$. \qed
+%%   \end{compactitem}}
+
+Finally, we define the acceptance of a word with a given input event
+sequence by means of a formula in which no predicate atom occurs. As
+previously discussed, several decidable theories, such as Presburger
+arithmetic, become undecidable if predicate atoms are added to
+them. Therefore, the result below makes a step towards deciding
+whether the automaton accepts a word with a given input sequence, by
+reducing this problem to the satisfiability of a quantified formula
+without predicates.
+
+\begin{lemma}\label{lemma:quant-pred-acceptance}
   Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for
-  every word $w \in \Sigma[X]^*$, we have $w_\Data \models
-  \substform{w_\Sigma}$ if and only if $w \in \lang{\A}$.
+  every word $w \in \Sigma[X]^*$, we have $\data{w} \models
+  \substform{\event{w}}$ if and only if $w \in \lang{\A}$.
 \end{lemma}
 \proof{ By Lemma \ref{lemma:acceptance}, $w \in \lang{\A}$ if and only
-  if $\I,w_\Data \models \accform{w_\Sigma}$, for some interpretation
-  $\I$. By induction on the number of predicate atoms in
-  $\accform{w_\Sigma}$, using Proposition \ref{prop:subst} in the
-  inductive step, we prove that, $\I,w_\Data \models
-  \accform{w_\Sigma}$ if and only if $w_\Data \models
-  \substform{w_\Sigma}$. \qed}
+  if $\I,\data{w} \models \accform{\event{w}}$, for some
+  interpretation $\I$. By Lemma \ref{lemma:quant}, there exists an
+  interpretation $\I$ such that $\I,\data{w} \models
+  \accform{\event{w}}$ if and only if there exists an interpretation
+  $\J$ such that $\J,\nu \models \quantform{\event{w}}$. By Lemma
+  \ref{lemma:subst}, there exists an interpretation $\J$ such that
+  $\J,\nu \models \quantform{\event{w}}$ if and only if $\nu \models
+  \substform{\event{w}}$. \qed}
 
@@
 \dual{(\phi_1 \vee \phi_2)} & = & \dual{\phi_1} \wedge \dual{\phi_2} && 
-\dual{(\phi_1 \wedge \phi_2)} & = & \dual{\phi_1} \wedge \dual{\phi_2} &&
+\dual{(\phi_1 \wedge \phi_2)} & = & \dual{\phi_1} \vee \dual{\phi_2} &&
 \dual{(t \teq s)} & = & \neg(t \teq s) \\
@@
   \noindent(\ref{it2:thm:closure}) Let $w \in \Sigma[X]^*$ be a
-  word. We denote by $\accformA{\A_1}{w_\Sigma}$ and
-  $\substformA{\A_1}{w_\Sigma}$
-  [resp. $\accformA{\overline{\A}_1}{w_\Sigma}$ and
-    $\substformA{\overline{\A}_1}{w_\Sigma}$] the formulae
-  $\accform{w_\Sigma}$ and $\substform{w_\Sigma}$ for $\A_1$ and
+  word. We denote by $\accformA{\A_1}{\event{w}}$ and
+  $\substformA{\A_1}{\event{w}}$
+  [resp. $\accformA{\overline{\A}_1}{\event{w}}$ and
+    $\substformA{\overline{\A}_1}{\event{w}}$] the formulae
+  $\accform{\event{w}}$ and $\substform{\event{w}}$ for $\A_1$ and
   $\overline{\A}_1$, respectively. It is enough to show that
-  $\substformA{\overline{\A}_1}{w_\Sigma} =
-  \neg\substformA{\A_1}{w_\Sigma}$ and apply Lemma
-  \ref{lemma:substitution} to prove that $w \in \lang{\A_1} \iff w
-  \not\in \lang{\overline{\A}_1}$. Since the choice of $w$ was
+  $\substformA{\overline{\A}_1}{\event{w}} =
+  \neg\substformA{\A_1}{\event{w}}$ and apply Lemma
+  \ref{lemma:quant-pred-acceptance} to prove that $w \in \lang{\A_1}
+  \iff w \not\in \lang{\overline{\A}_1}$. Since the choice of $w$ was
   arbitrary, this proves $\lang{\overline{\A}_1} = \Sigma[X]^*
   \setminus \lang{\A_1}$. By induction on the number of predicate
-  atoms in $\accformA{\A_1}{w_\Sigma}$ that are replaced during the
-  generation of $\substformA{\A_1}{w_\Sigma}$. The proof relies on the
-  following fact:
+  atoms in $\accformA{\A_1}{\event{w}}$ that are replaced during the
+  generation of $\substformA{\A_1}{\event{w}}$. The proof relies on
+  the following fact:
 
@@
 The problem of checking emptiness of a given automata is undecidable,
-even for automata with predicates of arity two whose transition rules
-use only equalities and disequalities, with no quantifiers
+even for automata with predicates of arity two, whose transition rules
+use only equalities and disequalities, with no transition quantifiers
 \cite{Farzan15}. Since even such simple classes of alternating
@@
 clauses \cite{BjornerGurfinkelMcMillanRybalchenko15}), the
-interpolants used to annotate the unfolded control structure are local
+interpolants used to annotate the unfolded control structure are
 assertions about the values of the program variables in a given
-control state. These assertions define inductive overapproximations of
-the set of configurations reachable at that control state, that
-suffice to prove emptiness or, equivalently, a safety property of the
-system.
-
-When considering an alternating model of computation, local
-interpolants that involve a single control state are no longer enough,
-because of the synchronization of several branches of the computation
-on the same sequence of input values. Consider, for instance, an
-automaton with the following transition rules and final state $q_f$:
+control state $q$, at a certain step $i \geq 0$ of a given
+execution. As we discuss next, it is useful to distinguish between
+\begin{inparaenum}[(i)]
+\item locality of interpolants w.r.t. a given control state and
+%
+\item locality w.r.t. a given time stamp.
+\end{inparaenum} 
+In logical terms, \emph{control-local} interpolants are defined by
+formulae involving a single predicate symbol, whereas
+\emph{time-local} interpolants involve only predicates $\stamp{q}{i}$
+and variables $\stamp{x}{i}$, for a single $i \geq 0$.
+
+\paragraph{Remark}
+When considering an alternating model of computation, control-local
+interpolants are not enough to prove emptiness, because of the
+synchronization of several branches of the computation on the same
+sequence of input values. Consider, for instance, an automaton with
+the following transition rules and final state $q_f$:
 \[\begin{array}{rclcrcl}
@@
 one occurrence of a predicate would give a too rough overapproximation
-of this set, unable to prove the emptiness of the automaton.
+of this set, unable to prove the emptiness of the
+automaton. \hfill$\blacksquare$
+
+First, let us give the formal definition of the class of interpolants
+we shall work with. Given a formula $\phi$, the \emph{vocabulary} of
+$\phi$ is the set of uninterpreted function symbols, predicate symbols
+$q \in \stamp{Q}{i}$ and variables $x \in \stamp{X}{i}$, occurring in
+$\phi$, for some $i\geq0$. Observe that quantified variables and the
+interpreted function symbols of the data theory\footnote{E.g.,\ the
+  arithmetic operators of addition and multiplication, when $\Data$ is
+  the set of integers.} do not belong to the vocabulary of a
+formula. By $\pset{+}{\phi}$ [$\pset{-}{\phi}$] we denote the set of
+predicate symbols that occur in $\phi$ under an even [odd] number of
+negations.
+
+Given formulae $\phi$ and $\psi$ such that $\phi \wedge \psi$ is
+unsatisfiable, a \emph{Lyndon interpolant} \cite{Lyndon59} is a
+first-order formula $I$ such that: \begin{compactitem}
+\item the formula $\phi \rightarrow I$ is valid,
+\item the formula $I \wedge \psi$ is unsatisfiable, 
+\item $\voc{I} \subseteq \voc{\phi} \cap \voc{\psi}$, 
+\item $\pset{+}{I} \subseteq \pset{+}{\phi} \cap \pset{+}{\psi}$ and
+  $\pset{-}{I} \subseteq \pset{-}{\phi} \cap \pset{-}{\psi}$.
+\end{compactitem}
+The existence of Lyndon interpolants for first-order logic was proved
+by a non-constructive model-theoretic argument \cite{Lyndon59}. In the
+following, we shall consider only theories of the data domain $\Data$,
+for which there exists a decision procedure for satisfiability, able
+to produce Lyndon interpolants. For instance, interpolating decision
+procedures exist for linear real and integer arithmetic, with
+uninterpreted function symbols (UFLRA and UFLIA)
+\cite{RybalchenkoSofronieStokkermans}. In the rest of this section, we
+fix an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$.
+
+\begin{definition}\label{def:generalized-lyndon-interpolant}
+  Given a non-empty sequence of input events $\alpha = a_1 \ldots a_n
+  \in \Sigma^*$, a \emph{generalized Lyndon interpolant} is a sequence
+  $I_0,I_1,\ldots,I_n$ of formulae such that, for all $k =
+  0,\ldots,n-1$: \begin{compactenum}
+    \item $\voc{I_k} = \pset{+}{I_k}$ and $\pset{-}{I_k} = \emptyset$, 
+      %
+    \item $\iota \models I_0$ and 
+      \(I_k \wedge \left(\bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi}
+      \forall \vec{y} ~.~ q(\vec{y}) \rightarrow
+      \nextf{\psi}\right) \models \nextf{I}_{k+1}\), 
+      %
+    \item the formula $I_n \wedge \bigwedge_{q \in Q \setminus F}
+      \forall \vec{y} ~.~ q(\vec{y}) \rightarrow \bot$ is
+      unsatisfiable.
+  \end{compactenum}
+\end{definition}
+
+\comment[ri]{I am here}
 
@@
 time-stamped path formula. Given a sequence of input events $\alpha$,
-we consider the path formula $\accform{\alpha}$, constrained by the
-fact that all non-final predicates $\stamp{q}{\len{\alpha}}$ must be
-empty sets, defined as in (\ref{eq:accform}). If this formula is
-unsatisfiable, we show the existence of a sequence of (possibly
+we consider the acceptance formula $\accform{\alpha}$. If this formula
+is unsatisfiable, we show the existence of a sequence of (possibly
 quantified) positive interpolants that overapproximate the set of
 cubes generated at each step during an execution over a word $w$,
-whose even sequence is $w_\Sigma = \alpha$. Moreover, these
+whose even sequence is $\event{w} = \alpha$. Moreover, these
 interpolants are local, in the sense that each formula uses only the
@@
 \ref{def:substform}) come into play. For a given event sequence
-$\alpha$, the automaton $\A$ accepts a word $w$ such that $w_\Sigma =
+$\alpha$, the automaton $\A$ accepts a word $w$ such that $\event{w} =
 \alpha$ if and only if $\substform{\alpha}$ is satisfiable (Lemma
@@
 
-\comment[ri]{i am here}
-
 However, the existence of a decision procedure for the theory in which

### 2018-09-14 quantifiers/draft.tex
ok

@@
 
+\begin{definition}\label{def:lyndon-interpolant}
 Given formulae $\phi$ and $\psi$ such that $\phi \wedge \psi$ is
 unsatisfiable, a \emph{Lyndon interpolant} \cite{Lyndon59} is a
-first-order formula $I$ such that: \begin{compactitem}
-\item the formula $\phi \rightarrow I$ is valid,
-\item the formula $I \wedge \psi$ is unsatisfiable, 
-\item $\voc{I} \subseteq \voc{\phi} \cap \voc{\psi}$, 
-\item $\pset{+}{I} \subseteq \pset{+}{\phi} \cap \pset{+}{\psi}$ and
+formula $I$ such that $\phi \models I$, the formula $I \wedge \psi$ is
+unsatisfiable, $\voc{I} \subseteq \voc{\phi} \cap \voc{\psi}$,
+$\pset{+}{I} \subseteq \pset{+}{\phi} \cap \pset{+}{\psi}$ and
 $\pset{-}{I} \subseteq \pset{-}{\phi} \cap \pset{-}{\psi}$.
-\end{compactitem}
+\end{definition}
 The existence of Lyndon interpolants for first-order logic was proved
@@
 for which there exists a decision procedure for satisfiability, able
-to produce Lyndon interpolants. For instance, interpolating decision
-procedures exist for linear real and integer arithmetic, with
+to produce Lyndon interpolants. For instance, such interpolating
+decision procedures exist for linear real and integer arithmetic, with
 uninterpreted function symbols (UFLRA and UFLIA)

### 2018-09-18 quantifiers/draft.tex
ok

--- 2018-09-14 quantifiers/draft.tex
@@
 \phi_2$ and $\neg\exists x ~.~ \neg\phi_1$, respectively. We denote by
-$\fv{}{\phi}$ the set of free variables in $\phi$. A \emph{sentence}
-is a formula $\phi$ in which each variable occurs under the scope of a
-quantifier, i.e. $\fv{}{\phi} = \emptyset$. A formula is
-\emph{positive} if each predicate symbol occurs under an even number
-of negations. The size $\len{\phi}$ of a formula $\phi$ is the number
-of symbols needed to write it down.
+$\fv{}{\phi}$ the set of free variables in $\phi$. The size
+$\len{\phi}$ of a formula $\phi$ is the number of symbols needed to
+write it down.
+
+A \emph{sentence} is a formula $\phi$ in which each variable occurs
+under the scope of a quantifier, i.e. $\fv{}{\phi} = \emptyset$.  A
+formula is \emph{positive} if each predicate symbol occurs under an
+even number of negations. A formula is in \emph{prenex form} if it is
+of the form $\varphi = Q_1x_1 \ldots Q_nx_n ~.~ \phi$, where $\phi$
+has no quantifiers. In this case we call $\phi$ the \emph{matrix} of
+$\varphi$. Every first-order formula can be written in prenex form, by
+renaming each quantified variable to a unique name and moving the
+quantifiers upfront.
 
@@
 \emph{valid} if $\I,\nu \models \phi$ for every interpretation $\I$
-and every valuation $\nu$.
+and every valuation $\nu$. For two formulae $\phi$ and $\psi$ we write
+$\phi \models \psi$ for $\sem{\phi} \subseteq \sem{\psi}$, in which
+case we say that $\phi$ \emph{entails} $\psi$.
 
@@
 
-%% \[\accform{\alpha} = \stamp{\iota}{0} \wedge 
-%% \bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in
-%%   \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-%% \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i} \wedge
-%% \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
-%% ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot\]
-
 \begin{definition}\label{def:quantform}
@@
 \begin{example}\label{ex:quant-elim}
-  Consider the automaton $\A = \tuple{\set{a_1,a_2}, \set{x}, \set{q},
-    \iota, \set{q}, \Delta}$, where:
+  Consider the automaton $\A = \tuple{\set{a_1,a_2}, \set{x},
+    \set{q,q_f}, \iota, \set{q_f}, \Delta}$, where:
   \[\begin{array}{rcl}
-  \iota & = & \exists z ~.~ q(z) \wedge z \geq 0 \\
-  \Delta & = & \set{q(y) \arrow{a_1(x)}{} \forall z ~.~ z \leq x+y \rightarrow q(y-z+x),~
-    q(y) \arrow{a_2(x)}{} \exists z ~.~ z \geq x-y \wedge q(y+z+x)}
+  \iota & = & \exists z ~.~ z \geq 0 \wedge q(z) \\
+  \Delta & = & \set{
+    q(y) \arrow{a_1(x)}{} x\geq 0 \wedge \forall z ~.~ z \leq y \rightarrow q(x+z),~
+    q(y) \arrow{a_2(x)}{} y<0 \wedge q_f(x+y)
+  }
   \end{array}\]
-  For the input event sequence $\alpha = a_1a_2$, the path and acceptance formulae are:
+  For the input event sequence $\alpha = a_1a_2$, the acceptance
+  formula is:
   \[\begin{array}{rcl}
-  \pathform{\alpha} = \accform{\alpha} & = & \exists z ~.~ \stamp{q}{0}(z) \wedge z \geq 0 \\
-  && \forall y ~.~ \stamp{q}{0}(y) \rightarrow (\forall z ~.~ z\leq\stamp{x}{1}+y \rightarrow \stamp{q}{1}(y-z+\stamp{x}{1})) \\
-  && \forall y ~.~ \stamp{q}{1}(y) \rightarrow \exists z ~.~ z\geq\stamp{x}{2}-y \wedge \stamp{q}{2}(y+z+\stamp{x}{2})
+  \accform{\alpha} & = & \exists z ~.~ z \geq 0 \wedge \stamp{q}{0}(z) ~\wedge \\
+  && \forall y ~.~ \stamp{q}{0}(y) \rightarrow [\stamp{x}{1}\geq0 \wedge \forall z ~.~ z\geq y 
+    \rightarrow \stamp{q}{1}(\stamp{x}{1}+z)] ~\wedge \\
+  && \forall y ~.~ \stamp{q}{1}(y) \rightarrow [y<0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+y)]
   \end{array}\]
@@
   \[\begin{array}{rcl}
-  \quantform{\alpha} & = & \exists z_1\forall z_2\exists z_3 ~.~ \stamp{q}{0}(z_1) \wedge z_1 \geq 0 \\
-  && \wedge~ [\stamp{q}{0}(z_1) \rightarrow (z_2 \leq \stamp{x}{1}+z_1 \rightarrow \stamp{q}{1}(z_1-z_2+\stamp{x}{1}))] \\
-  && \wedge~ [\stamp{q}{1}(z_1-z_2+\stamp{x}{1}) \rightarrow z_3 \geq \stamp{x}{2} -z_1+z_2-\stamp{x}{1} \wedge 
-    \stamp{q}{2}(z_1-z_2+\stamp{x}{1}+z_3+\stamp{x}{2})] \enspace\enspace\hfill\blacksquare
+  \quantform{\alpha} & = & \exists z_1\forall z_2 ~.~ z_1 \geq 0 \wedge \stamp{q}{0}(z_1) ~\wedge \\
+  && [\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge (z_2 \geq z_1 \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_2))] ~\wedge \\
+  && [\stamp{q}{1}(\stamp{x}{1}+z_2) \rightarrow \stamp{x}{1}+z_2 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_2)] 
+  \enspace\enspace\hfill\blacksquare
   \end{array}\]
@@
   formula in Example \ref{ex:quant-elim} is shown below:
-  \[\exists z_1 \forall z_2 \exists z_3 ~.~ (z_2 \leq \stamp{x}{1}+z_1 
-  \rightarrow z_3 \geq \stamp{x}{2} -z_1+z_2-\stamp{x}{1}) \wedge z_1
-  \geq 0\] Since this formula is satisfiable, by Lemma
-  \ref{lemma:quant-pred-acceptance} below, there exists a word $w$
-  with input event sequence $\event{w} = a_1a_2$ that is accepted by the
-  automaton $\A$ from Example
+  \[\substform{\alpha} = 
+  \exists z_1 \forall z_2 ~.~ z_1 \geq 0 \wedge 
+          [\stamp{x}{1} \geq 0 \wedge (z_2 \geq z_1 \rightarrow
+            \stamp{x}{1}+z_2 < 0)]\] Since this formula is
+          unsatisfiable, by Lemma \ref{lemma:quant-pred-acceptance}
+          below, no word $w$ with input event sequence $\event{w} =
+          a_1a_2$ is accepted by the automaton $\A$ from Example
           \ref{ex:quant-elim}. \hfill$\blacksquare$
@@
 
-  \begin{fact}
+  \begin{fact}\label{fact:subst}
     Given formulae $\phi$ and $\psi$, such that $\phi$ is positive
@@
 \cite{Farzan15}. Since even such simple classes of alternating
-automata have no general decision procedure, we use an
+automata have no general decision procedure for emptiness, we use an
 abstraction-refinement semi-algorithm based on \emph{lazy annotation}
@@
 searches. These facts are interpolants extracted from the proof of
-spuriousness of the given (finite) path.
+spuriousness of the given (finite) path. The semi-algorithm uses
+moreover a coverage relation to ensure that a spurious counterexample
+is never explore twice, thus making progress towards either finding a
+real counterexample or a proof of emptiness. In the rest of this
+section, we discuss the generation of interpolants
+(\S\ref{sec:interpolants}) and the coverage (\S\ref{sec:coverage})
+used by our emptiness checking semi-algorithm (\S\ref{sec:impact}).
+
+\subsection{Interpolants}\label{sec:interpolants}
 
 Typically, when checking the unreachability of a set of program
-configurations (or, more generally, the infeasability of a set of Horn
-clauses \cite{BjornerGurfinkelMcMillanRybalchenko15}), the
-interpolants used to annotate the unfolded control structure are
-assertions about the values of the program variables in a given
-control state $q$, at a certain step $i \geq 0$ of a given
-execution. As we discuss next, it is useful to distinguish between
+configurations or, more generally, the infeasability of a set of Horn
+clauses \cite{BjornerGurfinkelMcMillanRybalchenko15}, the interpolants
+used to annotate the unfolded control structure are assertions about
+the values of the program variables in a given control state, at a
+certain step of a given execution. As we discuss next, it is useful to
+distinguish between
 \begin{inparaenum}[(i)]
-\item locality of interpolants w.r.t. a given control state and
-%
-\item locality w.r.t. a given time stamp.
+\item locality of interpolants w.r.t. a given control state (control
+  locality) and
+%
+\item locality w.r.t. a given time stamp (time locality).
 \end{inparaenum} 
@@
 When considering an alternating model of computation, control-local
-interpolants are not enough to prove emptiness, because of the
+interpolants are not always enough to prove emptiness, because of the
 synchronization of several branches of the computation on the same
@@
 Started in an initial configuration $q_0(0)$ with an input word
-$(a,\nu_1) \ldots (a,\nu_{n-1}) (b,\nu_n)$, such that $\nu_i(x)=k_i$,
+$(a,\nu_1) \ldots (a,\nu_{n-1}) (a,\nu_n)$, such that $\nu_i(x)=k_i$,
 the automaton executes as follows:
@@
 steps is defined by the formula: $\exists x_1 \exists x_2 ~.~ q_1(x_1)
-\wedge q_2(x_2) \wedge x_1+x_2 \teq 0$. Observe that a formula using
-one occurrence of a predicate would give a too rough overapproximation
-of this set, unable to prove the emptiness of the
+\wedge q_2(x_2) \wedge x_1+x_2 \teq 0$. Observe that a control-local
+formula using one occurrence of a predicate would give a too rough
+overapproximation of this set, unable to prove the emptiness of the
 automaton. \hfill$\blacksquare$
@@
 we shall work with. Given a formula $\phi$, the \emph{vocabulary} of
-$\phi$ is the set of uninterpreted function symbols, predicate symbols
-$q \in \stamp{Q}{i}$ and variables $x \in \stamp{X}{i}$, occurring in
-$\phi$, for some $i\geq0$. Observe that quantified variables and the
-interpreted function symbols of the data theory\footnote{E.g.,\ the
-  arithmetic operators of addition and multiplication, when $\Data$ is
-  the set of integers.} do not belong to the vocabulary of a
-formula. By $\pset{+}{\phi}$ [$\pset{-}{\phi}$] we denote the set of
-predicate symbols that occur in $\phi$ under an even [odd] number of
-negations.
-
+$\phi$, denoted $\voc{\phi}$ is the set of predicate symbols $q \in
+\stamp{Q}{i}$ and variables $x \in \stamp{X}{i}$, occurring in $\phi$,
+for some $i\geq0$. For a term $t$, its vocabulary $\voc{t}$ is the set
+of variables that occur in $t$. Observe that quantified variables and
+the interpreted function symbols of the data
+theory\footnote{E.g.,\ the arithmetic operators of addition and
+  multiplication, when $\Data$ is the set of integers.}  do not belong
+to the vocabulary of a formula. By $\pset{+}{\phi}$ [$\pset{-}{\phi}$]
+we denote the set of predicate symbols that occur in $\phi$ under an
+even [odd] number of negations.
+
+\begin{definition}[\cite{Lyndon59}]\label{def:lyndon-interpolant}
 Given formulae $\phi$ and $\psi$ such that $\phi \wedge \psi$ is
-unsatisfiable, a \emph{Lyndon interpolant} \cite{Lyndon59} is a
-first-order formula $I$ such that: \begin{compactitem}
-\item the formula $\phi \rightarrow I$ is valid,
-\item the formula $I \wedge \psi$ is unsatisfiable, 
-\item $\voc{I} \subseteq \voc{\phi} \cap \voc{\psi}$, 
-\item $\pset{+}{I} \subseteq \pset{+}{\phi} \cap \pset{+}{\psi}$ and
-  $\pset{-}{I} \subseteq \pset{-}{\phi} \cap \pset{-}{\psi}$.
-\end{compactitem}
-The existence of Lyndon interpolants for first-order logic was proved
-by a non-constructive model-theoretic argument \cite{Lyndon59}. In the
-following, we shall consider only theories of the data domain $\Data$,
-for which there exists a decision procedure for satisfiability, able
-to produce Lyndon interpolants. For instance, interpolating decision
-procedures exist for linear real and integer arithmetic, with
-uninterpreted function symbols (UFLRA and UFLIA)
-\cite{RybalchenkoSofronieStokkermans}. In the rest of this section, we
-fix an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$.
+unsatisfiable, a \emph{Lyndon interpolant} is a formula $I$ such that
+$\phi \models I$, the formula $I \wedge \psi$ is unsatisfiable,
+$\voc{I} \subseteq \voc{\phi} \cap \voc{\psi}$, $\pset{+}{I} \subseteq
+\pset{+}{\phi} \cap \pset{+}{\psi}$ and $\pset{-}{I} \subseteq
+\pset{-}{\phi} \cap \pset{-}{\psi}$.
+\end{definition}
+The existence of Lyndon interpolants for first-order logic, with no
+predefined interpretation of function symbols\footnote{A model
+  specifies the universe and the interpretation of function symbols.},
+was proved by a non-constructive model-theoretic argument
+\cite{Lyndon59}. In the following, we shall consider only those
+theories of the data domain $\Data$, for which there exists a decision
+procedure for satisfiability, able to produce Lyndon interpolants. For
+instance, such interpolating decision procedures exist for linear real
+and integer arithmetic, with uninterpreted function symbols (UFLRA and
+UFLIA) \cite{RybalchenkoSofronieStokkermans}. In the rest of this
+section, let us fix an automaton $\A =
+\tuple{\Sigma,X,Q,\iota,F,\Delta}$.
 
@@
   Given a non-empty sequence of input events $\alpha = a_1 \ldots a_n
-  \in \Sigma^*$, a \emph{generalized Lyndon interpolant} is a sequence
-  $I_0,I_1,\ldots,I_n$ of formulae such that, for all $k =
-  0,\ldots,n-1$: \begin{compactenum}
-    \item $\voc{I_k} = \pset{+}{I_k}$ and $\pset{-}{I_k} = \emptyset$, 
-      %
-    \item $\iota \models I_0$ and 
-      \(I_k \wedge \left(\bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi}
-      \forall \vec{y} ~.~ q(\vec{y}) \rightarrow
-      \nextf{\psi}\right) \models \nextf{I}_{k+1}\), 
-      %
-    \item the formula $I_n \wedge \bigwedge_{q \in Q \setminus F}
-      \forall \vec{y} ~.~ q(\vec{y}) \rightarrow \bot$ is
-      unsatisfiable.
+  \in \Sigma^*$, a \emph{generalized Lyndon interpolant (GLI)} is a
+  sequence $(I_0,\ldots,I_n)$ of formulae such that, for all $k \in
+  [n-1]$: \begin{compactenum}
+    \item $\pset{-}{I_k} = \emptyset$,
+      %
+    \item $\stamp{\iota}{0} \models I_0$ and \(I_k \wedge
+      \Big(\bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in \Delta}
+      \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
+      \rightarrow \stamp{\psi}{k+1}\Big) \models I_{k+1}\),
+      %
+    \item $I_n \wedge \bigwedge_{q \in Q \setminus F} \forall y_1
+      \ldots \forall y_{\#(q)} ~.~ q(\vec{y})$ is unsatisfiable.
   \end{compactenum}
+  Moreover, the GLI is \emph{local} iff $\voc{I_k} \cap \stamp{Q}{\leq n}
+  \subseteq \stamp{Q}{k}$ and $\voc{I_k} \cap \stamp{X}{\leq n}
+  \subseteq \stamp{X}{k}$, for all $k \in [n]$.
 \end{definition}
-
-\comment[ri]{I am here}
-
-The solution we adopt is a generalization of the classical approach of
-extracting the interpolants from the proof of unsatisfiability of a
-time-stamped path formula. Given a sequence of input events $\alpha$,
-we consider the acceptance formula $\accform{\alpha}$. If this formula
-is unsatisfiable, we show the existence of a sequence of (possibly
-quantified) positive interpolants that overapproximate the set of
-cubes generated at each step during an execution over a word $w$,
-whose even sequence is $\event{w} = \alpha$. Moreover, these
-interpolants are local, in the sense that each formula uses only the
-values of the variables and the state predicates stamped with its
-index in the sequence.
-
-Unlike with the classical path formulae for integer programs
-\cite{McMillan06} and Horn clauses
-\cite{BjornerGurfinkelMcMillanRybalchenko15}, that are expressed using
-linear or Presburger arithmetic, the existence of local interpolants
-for theories using first-order predicates relies on a non-constructive
-argument, the Lyndon Interpolation Theorem \cite{Lyndon59}. Since
-usual decidable theories, such as Presburger arithmetic, become
-undecidable in the presence of monadic predicate symbols
-\cite{Halpern91}, we cannot hope to find an automated interpolant
-generation method, in general.
-
-But here is where the predicate-free path formulae (Definition
+The following proposition states the existence of local GLI for the
+theories in which Lyndon's Interpolation Theorem holds.
+
+\begin{proposition}\label{prop:local-gli}
+  If there exists a Lyndon interpolant for any two formulae $\phi$ and
+  $\psi$, such that $\phi \wedge \psi$ is unsatisfiable, then any
+  sequence of input events $\alpha = a_1 \ldots a_n \in \Sigma^*$,
+  such that $\accform{\alpha}$ is unsatisfiable, has a local GLI
+  $(I_0,\ldots,I_n)$.
+\end{proposition}
+\proof{ By definition, $\accform{\alpha}$ is the formula:
+  \[\stamp{\iota}{0} \wedge 
+ \bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in
+   \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
+ \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i} \wedge
+ \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
+ ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot\]
+ We define the formulae: 
+ \[\begin{array}{rcl}
+ \varphi_i & \isdef & \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
+ \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i},~ \text{for all}~ i \in [1,n] \\
+ \psi & \isdef & \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
+ ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot
+ \end{array}\]
+ Observe that $\voc{\stamp{\iota}{0}} \subseteq \stamp{Q}{0}$,
+ $\voc{\varphi_i} \subseteq
+ \stamp{Q}{i-1}\cup\stamp{Q}{i}\cup\stamp{X}{i}$, for all $i \in
+       [1,n]$, and $\voc{\psi} \subseteq \stamp{Q}{n}$.  We apply
+       Lyndon's Interpolation Theorem for the formulae
+       $\stamp{\iota}{0}$ and $\bigwedge_{i=1}^n \varphi_i \wedge
+       \psi$ and obtain a formula $I_0$, such that $\stamp{\iota}{0}
+       \models I_0$, $I_0 \wedge \bigwedge_{i=1}^n \varphi_i \wedge
+       \psi$ is unsatisfiable, $\voc{I_0} \subseteq
+       \voc{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n \voc{\varphi_i}
+       \cup \voc{\psi}) \subseteq \stamp{Q}{0}$ and $\pset{-}{I_0}
+       \subseteq \pset{-}{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n
+       \pset{-}{\varphi} \cup \pset{-}{\psi}) = \emptyset$. Repeating
+       the reasoning for the formulae $I_0 \wedge \varphi_1$ and
+       $\bigwedge_{i=2}^n \varphi_i \wedge \psi$, we obtain $I_1$,
+       such that $I_0 \wedge \varphi_1 \models I_1$, $I_1 \wedge
+       \bigwedge_{i=2}^n \varphi_i \wedge \psi$ is unsatisfiable,
+       $\voc{I_1} \subseteq (\voc{I_0} \cup \voc{\varphi_1}) \cap
+       (\bigcup_{i=2}^n \voc{\varphi_i} \cup \voc{\psi}) \subseteq
+       \stamp{Q}{1}$ and $\pset{-}{I_1} \subseteq (\pset{-}{I_0} \cup
+       \pset{-}{\varphi_1}) \cap (\bigcup_{i=2}^n \pset{-}{\varphi_i}
+       \cup \pset{-}{\psi}) = \emptyset$. Continuing in this way, we
+       obtain formulae $I_0, I_1, \ldots, I_{n}$ as required. \qed}
+
+The main problem with the local GLI construction described in the
+proof of Proposition \ref{prop:local-gli} is that the existence of
+Lyndon interpolants (Definition \ref{def:lyndon-interpolant}) is
+guaranteed in principle, but the proof is non-constructive. Building
+an interpolant for an unsatisfiable conjunction of formulae $\phi
+\wedge \psi$ is typically the job of the decision procedure that
+proves the unsatisfiability and, in general, there is no such
+procedure, when $\phi$ and $\psi$ contain predicates and have
+non-trivial quantifier alternation.
+
+This is where the predicate-free path formulae (Definition
 \ref{def:substform}) come into play. For a given event sequence
 $\alpha$, the automaton $\A$ accepts a word $w$ such that $\event{w} =
-\alpha$ if and only if $\substform{\alpha}$ is satisfiable (Lemma
-\ref{lemma:substitution}). Assuming further that the equality atoms in
-the transition rules of $\A$ are written in the language of a
-decidable first-order theory, such as Presburger arithmetic, Lemma
-\ref{lemma:substitution} gives us an effective way of checking
-emptiness of $\A$ relative to a given event sequence. However, this
-method does not cope well with lazy annotation, because there is no
-way to extract, from the unsatisfiability proof of
-$\substform{\alpha}$, the interpolants needed to annotate
-$\alpha$. This is because \begin{inparaenum}[(i)]
-\item the formula $\substform{\alpha}$, obtained by repeated
-  substitutions (Definition \ref{def:substform}) loses track of the
-  execution semantics of $\A$, and 
-%
-\item quantifiers may occur nested in $\substform{\alpha}$, making it
-  difficult to write $\substform{\alpha}$ as a proper interpolation
-  problem.
+\alpha$ if and only if $\substform{\alpha}$ is satisfiable. Assuming
+further that the equality atoms in the transition rules of $\A$ are
+written in the language of a decidable first-order theory, such as
+Presburger arithmetic, Lemma \ref{lemma:quant-pred-acceptance} gives
+us an effective way of checking emptiness of $\A$, relative to a given
+event sequence. However, this method does not cope well with lazy
+annotation, because there is no way to extract, from the
+unsatisfiability proof of $\substform{\alpha}$, the interpolants
+needed to annotate $\alpha$. This is because \begin{inparaenum}[(i)]
+\item\label{it1:gli} the formula $\substform{\alpha}$, obtained by
+  repeated substitutions (Definition \ref{def:substform}) loses track
+  of the steps of the execution, and
+%
+\item\label{it2:gli} quantifiers that occur nested in
+  $\substform{\alpha}$ make it difficult to write $\substform{\alpha}$
+  as an unsatisfiable conjunction of formulae from which interpolants
+  are extracted (Definition \ref{def:lyndon-interpolant}).
 \end{inparaenum}
 
-The solution we adopt in this paper relies on an additional assumption
-related to the decidability of the predicate-free theory in which
-$\substform{\alpha}$ formulae are written, namely that the theory has
-an effective \emph{witness-providing quantifier elimination}
-procedure. More precisely, we assume that, for each formula $\exists x
-~.~ \phi(x)$, there exists an effectively computable term $t$, in
-which $x$ does not occur, such that $\exists x ~.~ \phi$ and
-$\phi[t/x]$ are equi-satisfiable. These terms, called \emph{witnesses}
-in the following, are actual interpretations of the Skolem function
-symbols from the following folklore theorem:
+The solution we adopt for the first issue (\ref{it1:gli}) consists in
+partially recovering the time-stamped structure of the acceptance
+formula $\accform{\alpha}$ using the formula $\quantform{\alpha}$, in
+which only transition quantifiers occur. The second issue
+(\ref{it2:gli}) is solved under the additional assuption that the
+theory of the data domain $\Data$ has \emph{witness-producing
+  quantifier elimination}. More precisely, we assume that, for each
+formula $\exists x ~.~ \phi(x)$, there exists an effectively
+computable term $\tau$, in which $x$ does not occur, such that
+$\exists x ~.~ \phi$ and $\phi[\tau/x]$ are equisatisfiable. These
+terms, called \emph{witness terms} in the following, are actual
+definitions of the Skolem function symbols from the following folklore
+theorem:
 
@@
   \[\eta_i \equiv \left\{\begin{array}{ll}
-  f_i(y_1,\ldots,y_{k_i}) & \text{if } Q_i \equiv \forall \\
-  x_i & \text{if } Q_i \equiv \exists
+  f_i(y_1,\ldots,y_{k_i}) & \text{if } Q_i = \forall \\
+  x_i & \text{if } Q_i = \exists
   \end{array}\right.\]
   where $f_i$ is a fresh function symbol and $\set{y_1, \ldots,
-    y_{k_i}} = \set{x_j \mid j < i,~ Q_j \equiv \exists}$. Then \(Q_1
-  x_1 \ldots Q_n x_n ~.~ \phi \rightarrow
-  \phi[\eta_1/x_1,\ldots,\eta_n/x_n]\) is valid.
+    y_{k_i}} = \set{x_j \mid j < i,~ Q_j = \exists}$. Then the entailment \(Q_1
+  x_1 \ldots Q_n x_n ~.~ \phi \models
+  \phi[\eta_1/x_1,\ldots,\eta_n/x_n]\) holds. 
 \end{theorem}
@@
   \cite[Lemma 2.1.9]{BorgerGraedelGurevich97}. \qed}
-Observe that our assumption on the existence of an effective
-witness-providing quantifier elimination procedure is stronger than
-the above theorem, as we require that \begin{inparaenum}[(i)] 
-\item the $f_i$ functions be definable in the language of the theory,
-  and 
-\item $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$ and
-  $\phi[\eta_1/x_1,\ldots,\eta_n/x_n]$ are equi-satisfiable. 
-\end{inparaenum}
-Among the theories that support witness-providing quantifier
-elimination, one mentions \comment[ri]{LIA, LRA, Presburger, Tarski?,
-  BAPA?}
-
-Turning back to the problem of generating interpolants from the proof
-of unsatisfiability of a path formula $\accform{\alpha}$, assume that
-the equivalent predicate-free formula $\substform{\alpha} = \exists
-x_1 \forall y_1 \ldots \exists x_n \forall y_n ~.~ \psi$ is
-unsatisfiable and let $t_1(x_1)$, $\ldots, t_n(x_1,\ldots,x_n)$ be
-witnesses such that $\psi[t_1/y_1,\ldots,t_n/y_n] \equiv \bot$. Then
-we show that the quantifier-free formula (with predicates) obtained
-from $\accform{\alpha}$ by (roughly) replacing the universal
-quantifiers with $t_1, \ldots, t_n$ in $\accform{\alpha}$ is also
-unsatisfiable.  Moreover, this new formula is layered in the same way
-as $\accform{\alpha}$, namely as a conjunction of the form:
-\[\stamp{\iota}{0} \wedge \bigwedge_{i=1}^n \bigwedge_{q(\vec{y})
-  \arrow{a_i(X)}{} \psi \in \Delta} \stamp{q}{i-1}(\vec{t}_{q,i-1})
-\rightarrow \stamp{\psi}{i} \wedge \bigwedge_{q \in Q \setminus F}
-\stamp{q}{n}(\vec{t}_{q,n}) \rightarrow \bot\]
-
-However, the existence of a decision procedure for the theory in which
-$\accform{\alpha}$ is written is not guaranteed, as it was usually the
-case of classical path formulae for integer programs \cite{McMillan06}
-and Horn clauses \cite{BjornerGurfinkelMcMillanRybalchenko15} using
-linear or Presburger arithmetic. The problem is mainly that decidable
-theories, such as Presburger arithmetic become undecidable in the
-presence of monadic predicate symbols \cite{Halpern91}. A way around
-this problem is to check whether a given event sequence $\alpha$ may
-
-The rest of this section is concerned with defining the building
-blocks of a lazy annotation semi-algorithm for checking emptiness of
-automata.
-
-In particular, we address the following problems: \begin{compactenum}
-%
-\item generation of interpolants that prove the infeasability of a
-  given path formula, assuming the decidability of the first-order
-  theory in which the path formulae are written. 
-%
-\item generation of non-local interpolants using history variables,
-  obtained by Skolemization of the universal quantifiers that occur
-  within transition rules.
-%
-\item computing the coverage between two node labels in the unfolding
-  of an automaton, when history variables occur. 
+
+Examples of witness-producing quantifier elimination procedures can be
+found in the literature for e.g. linear integer (real) arithmetic
+(LIA,LRA), Presburger arithmetic and boolean algebra of sets and
+Presburger cardinality constraints (BAPA)
+\cite{KuncakMayerPiskacSuter12}. Furthermore, we describe a cubic-time
+witness generation for difference bounds constraints, a simple subset
+of linear integer (real) arithmetic (\S\ref{sec:applications}).
+
+Under the assumption that witness terms can be effectively built, let
+us describe the generation of a non-local GLI for a given input event
+sequence $\alpha = a_1 \ldots a_n$. First, we generate successively
+the acceptance formula $\accform{\alpha}$ and its equisatisfiable
+forms $\quantform{\alpha} = Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$
+and $\substform{\alpha} = Q_1x_1 \ldots Q_mx_m ~.~ \overline{\Phi}$,
+both written in prenex form, with matrices $\widehat{\Phi}$ and
+$\overline{\Phi}$, respectively. W.l.o.g. we consider a monotonic
+mapping $\xi : [1,m] \rightarrow [n]$ assigning each transition
+quantifier from $\substform{\alpha}$ its corresponding time-stamp from
+$\accform{\alpha}$. Because we assumed that the first-order theory of
+$\Data$ has quantifier elimination, the satisfiability problem for
+$\substform{\alpha}$ is decidable. If $\substform{\alpha}$ is
+satisfiable, we build a counterexample for emptiness $w$ such that
+$\event{w}=\alpha$ and $\data{w}$ is a satisfying assignment for
+$\substform{\alpha}$. Otherwise, $\substform{\alpha}$ is unsatisfiable
+and there exist witness terms $\tau_{i_1} \ldots \tau_{i_\ell}$, such
+that $\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq\xi(i_j)}$ (Theorem
+\ref{thm:skolem}) and the formula $\overline{\Phi}[\tau_{i_1}/x_{i_1},
+  \ldots, \tau_{i_\ell}/x_{i_\ell}]$, obtained from the matrix of
+$\substform{\alpha}$, by replacing the universally quantified
+variables with witness terms, is unsatisfiable. Then it turns out that
+the formula $\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
+  \tau_{i_\ell}/x_{i_\ell}]$, obtained analogously from the matrix of
+$\quantform{\alpha}$, is unsatisfiable as well (Lemma
+\ref{lemma:subst-quant}).
+
+Because this latter formula is structured as a conjunction of formulae
+$\stamp{\iota}{0} \wedge \phi_1 \ldots \wedge \phi_n \wedge \psi$,
+where $\voc{\phi_k} \subseteq \stamp{Q}{k-1} \cup \stamp{Q}{k} \cup
+\stamp{X}{\leq k}$ for all $k \in [1,n]$ and $\voc{\psi} \subseteq
+\stamp{Q}{n}$, it is now possible to use an existing interpolation
+procedure for the quantifier-free theory of $\Data$, extended with
+uninterpreted function symbols, to compute a sequence of non-local GLI
+$(I_0, \ldots, I_n)$ such that $\voc{I_k} \cap \stamp{Q}{\leq n}
+\subseteq \stamp{Q}{k}$ and $\voc{I_k} \cap \stamp{X}{\leq n}
+\subseteq \stamp{X}{\leq k}$, for all $k \in [n]$. Since each formula
+$I_k$ of the GLI may refer to the current predicates $\stamp{Q}{k}$
+and to the past and current input variables $\stamp{X}{\leq k}$, we
+further give up on the locality of interpolants and work with
+non-local GLI that are effectively computable by a combination of
+quantifier elimination with interpolation for quantifier-free data
+theories with uninterpreted functions.
+
+\begin{example}[Contd. from Examples \ref{ex:quant-elim} and \ref{ex:pred-elim}]\label{ex:gli}
+The formula $\substform{\alpha}$ (Example \ref{ex:pred-elim}) is
+unsatisfiable and let $\tau_2 = z_1$ be the witness term for the
+universally quantified variable $z_2$. Replacing $z_2$ with $\tau_2$
+in the matrix of $\quantform{\alpha}$ (Example \ref{ex:quant-elim})
+yields the unsatisfiable conjunction: 
+\[\begin{array}{l}
+z_1 \geq 0 \wedge \stamp{q}{0}(z_1) ~\wedge \\
+\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge (z_2 \geq z_1 \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_2)) ~\wedge \\
+\stamp{q}{1}(\stamp{x}{1}+z_2) \rightarrow \stamp{x}{1}+z_2 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_2)
+\end{array}\]
+A non-local GLI for the above is $(\stamp{q}{0}(z_1) \wedge
+z_1\geq 0,~ \stamp{x}{1} \geq 0 \wedge \stamp{q}{1}(\stamp{x}{1}+z_1)
+\wedge z_1\geq0,~\bot)$. \hfill$\blacksquare$
+\end{example}
+
+The following lemma gives the formal proof of correctness for this
+construction.
+
+\begin{lemma}\label{lemma:subst-quant}
+  Given a non-empty input event sequence $\alpha = a_1 \ldots a_n \in
+  \Sigma^*$, such that $\accform{\alpha}$ is unsatisfiable, let
+  $Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$ be a prenex form of
+  $\quantform{\alpha}$, where $Q_ix_i$ are transition quantifiers and
+  $\xi : [1,m] \rightarrow [n]$ is a monotonic function mapping each
+  quantified variable to the time-stamp of the step where it occurs in
+  $\accform{\alpha}$. Then the following hold: \begin{compactenum}
+  \item\label{it1:subst-quant} there exist witness terms $\tau_{i_1},
+    \ldots, \tau_{i_\ell}$, where $1 \leq i_1 < \ldots < i_\ell \leq
+    m$ and $\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)}$ for
+    all $j \in [1,\ell]$, such that $\widehat{\Phi}[\tau_{i_1}/x_{i_1},
+      \ldots, \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable, and
+    %
+  \item\label{it2:subst-quant} there exists a GLI $(I_0, \ldots, I_n)$
+    for $\alpha$, such that $\voc{I_k} \cap \stamp{Q}{\leq n}
+    \subseteq \stamp{Q}{k}$ and $\voc{I_k} \cap \stamp{X}{\leq n}
+    \subseteq \stamp{X}{\leq k}$, for all $k \in [n]$.
   \end{compactenum}
-Based on these results, we show how a version of the classical
-\textsc{Impact} \cite{McMillan06} semi-algorithm can be used to
-(dis-)prove the emptiness of first-order alternating automata.
-
-\subsection{Lyndon Interpolants}
-
-Let $\phi$ and $\psi$ be first-order formulae, such that $\phi \wedge
-\psi$ is unsatisfiable. The \emph{vocabulary} of $\phi$, denoted as
-$\voc{\phi}$, is the set of function symbols that occur in $\phi$.  By
-$\pset{+}{\phi}$ [$\pset{-}{\phi}$] we denote the set of predicate
-symbols that occur in $\phi$ under an even (odd) number of
-negations. A \emph{Lyndon interpolant} \cite{Lyndon59} for $\phi$ and
-$\psi$ is a first-order formula $I$ such that: \begin{compactitem}
-\item $\phi \rightarrow I$ is valid,  
-\item $I \wedge \psi$ is unsatisfiable, 
-\item $\voc{I} \subseteq \voc{\phi} \cap \voc{\psi}$, 
-\item $\pset{+}{I} \subseteq \pset{+}{\phi} \cap \pset{+}{\psi}$ and
-  $\pset{-}{I} \subseteq \pset{-}{\phi} \cap \pset{-}{\psi}$.
+\end{lemma}
+\proof{(\ref{it1:subst-quant}) If $\accform{\alpha}$ is unsatisfiable,
+  by Lemmas \ref{lemma:quant} and \ref{lemma:subst}, we obtain that,
+  successively $\quantform{\alpha}$ and $\substform{\alpha}$ are
+  unsatisfiable. Let $Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$ and
+  $Q_1x_1 \ldots Q_mx_m ~.~ \overline{\Phi}$ be prenex forms for
+  $\quantform{\alpha}$ and $\substform{\alpha}$,
+  respectively. Further, let $\xi : [1,m] \rightarrow [n]$ be the
+  monotone function mapping of each quantified variable to time-stamp
+  where it was introduced. Then there exist witness terms $\tau_{i_1},
+  \ldots, \tau_{i_\ell}$, where $\set{i_1,\ldots,i_\ell}=\set{i \in
+    [1,m] \mid Q_i=\forall}$, such that $\voc{\tau_{i_j}} \subseteq
+  \stamp{X}{\xi(i_j)}$, for all $j \in [1,\ell]$ and
+  $\overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
+    \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.
+
+  Let $\widehat{\Phi}_0$, \ldots, $\widehat{\Phi}_n$ be the sequence
+  of quantifier-free formulae, defined as follows: \begin{compactitem}
+  \item $\widehat{\Phi}_0$ is the matrix of the prenex form of
+    $\stamp{\iota}{0}$,
+    %
+  \item for all $i=1,\ldots,n$, let $\widehat{\Phi}_i$ be a prenex
+    form matrix of the formula:
+    \[\widehat{\Phi}_i \isdef \widehat{\Phi}_{i-1} \wedge
+     \underbrace{\bigwedge_{\begin{array}{l}
+         \scriptstyle{\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \widehat{\Phi}_{i-1}} \\
+        \scriptstyle{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta} 
+     \end{array}} \stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow 
+     \stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]}_{\isdef \phi_i}\]  
   \end{compactitem}
-The existence of Lyndon interpolants for first-order logic was proved
-by a non-constructive model-theoretic argument \cite{Lyndon59}. For
-now, we silently assume that the satisfiability problem for the theory
-in which $\phi$ and $\psi$ are written is decidable and, moreover,
-that a decision procedure generating Lyndon interpolants exists, for
-this theory. This is a rather unrealistic assumption, given that
-classical decidable theories, such as Presburger arithmetic, become
-undecidable in the presence of (monadic) predicate symbols
-\cite{Halpern91}.
-
-Let $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the
-rest of this section. Given a non-empty sequence of input events
-$\alpha = a_1 \ldots a_n \in \Sigma^*$, a \emph{local interpolant} for
-$\alpha$ is a sequence $I_0,I_1,\ldots,I_n$ of first-order formulae
-such that: \begin{compactitem}
-\item $\iota \rightarrow I_0$ is valid, 
-%
-\item $\left[I_k \wedge \left(\bigwedge_{q(\vec{y}_q) \arrow{a_i(X)}{} \psi}
-  \forall \vec{y}_q ~.~ \stamp{q}{k}(\vec{y}_q) \rightarrow
-  \nextf{\psi}\right)\right] \rightarrow \nextf{I}_{k+1}$ is valid, for all $k \in [n-1]$,
-%
-\item $I_n \wedge \left(\bigwedge_{q \in Q \setminus F} \forall \vec{y}_q
-  ~.~ q(\vec{y}_q) \rightarrow \bot\right)$ is unsatisfiable,
-%
-\item $\voc{I_k} = \pset{+}{I_k}$ and $\pset{-}{I_k} = \emptyset$. 
+  It is easy to see that $\widehat{\Phi}$ is a prenex form matrix of the formula:   
+  \[\widehat{\Phi}_n \wedge \underbrace{\bigwedge_{\begin{array}{l}
+        \scriptstyle{\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \widehat{\Phi}_n} \\
+        \scriptstyle{q \in Q \setminus F}
+    \end{array}} \stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot}_{\isdef \psi}\]
+  Applying the equivalence from Fact \ref{fact:subst} in the proof of
+  Lemma \ref{lemma:subst}, we obtain a sequence of quantifier-free
+  formulae $\overline{\Phi}_0, \ldots, \overline{\Phi}_n$ such that
+  $\overline{\Phi}_i \equiv \widehat{\Phi}_i$, for all $i \in [n]$ and
+  $\overline{\Phi}$ is obtained from $\overline{\Phi}_n$ by replacing
+  each occurrence of a predicate atom $q(t_1,\ldots,t_{\#(q)})$ in
+  $\overline{\Phi}_n$ by $\bot$ if $q \in Q\setminus F$ and by $\top$
+  if $q \in F$. Clearly $\overline{\Phi} \equiv \widehat{\Phi}$, thus
+  $\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
+    \tau_{i_\ell}/x_{i_\ell}] \equiv
+  \overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
+    \tau_{i_\ell}/x_{i_\ell}] \equiv \bot$.
+
+  \noindent(\ref{it2:subst-quant}) With the notation introduced at
+  point (\ref{it1:subst-quant}), we have $\widehat{\Phi} =
+  \widehat{\Phi}_0 \wedge \bigwedge_{i=1}^n \phi_i \wedge
+  \psi$. Consider the sequence of witness terms $\tau_{i_1}, \ldots,
+  \tau_{i_\ell}$, such that $\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq
+    \xi(i_j)}$, for all $j \in [1,\ell]$, whose existence is proved by
+  point (\ref{it1:subst-quant}). By the definition of $\xi$, we obtain
+  the following: \begin{compactitem}
+  \item $\voc{\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots, \tau_{i_\ell}/x_{i_\ell}]} \subseteq 
+    \stamp{Q}{0} \cup \stamp{X}{0}$, 
+    %
+  \item $\voc{\phi_i[\tau_{i_1}/x_{i_1}, \ldots, \tau_{i_\ell}/x_{i_\ell}]} \subseteq 
+    \stamp{Q}{i-1} \cup \stamp{Q}{i} \cup \stamp{X}{\leq i}$, for all $i \in [1,n]$,
+    %
+  \item $\voc{\psi[\tau_{i_1}/x_{i_1}, \ldots, \tau_{i_\ell}/x_{i_\ell}]} \subseteq 
+    \stamp{Q}{n} \cup \stamp{X}{\leq n}$. 
   \end{compactitem}
-
-\begin{lemma}\label{lemma:lyndon-existence}
-
-\end{lemma}
-
-\subsection{Skolem Witnesses}
-
-\subsection{Lazy Annotation}
+  By repeatedly applying Lyndon's Interpolation Theorem, we obtain a
+  sequence of formulae $(I_0, \ldots, I_n)$ such that: \begin{compactitem}
+    \item $\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
+      \tau_{i_\ell}/x_{i_\ell}] \models I_0$ and $\voc{I_0} \subseteq
+      \stamp{Q}{0} \cup \stamp{X}{0}$,
+      %
+    \item $I_{k-1} \wedge \phi_i[\tau_{i_1}/x_{i_1}, \ldots,
+      \tau_{i_\ell}/x_{i_\ell}] \models I_k$ and $\voc{I_k} \subseteq
+      \stamp{Q}{k} \cup \stamp{X}{\leq k}$, for all $k \in [1,n]$,
+      %
+    \item $I_n \wedge \psi[\tau_{i_1}/x_{i_1}, \ldots,
+      \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.
+  \end{compactitem}
+  To show that $(I_0,\ldots,I_n)$ is a GLI for $a_1\ldots a_n$, it is
+  sufficient to notice that \[\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
+    \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
+  \stamp{q}{k}(\vec{y}) \rightarrow \stamp{\psi}{k+1} \models \phi_k\]
+  for all $k \in [1,n]$. Consequently, we obtain: \begin{compactitem}
+    \item $\stamp{\iota}{0} \models \widehat{\Phi}_0 \models I_0$, by
+      Theorem \ref{thm:skolem},
+      %
+    \item $I_{k-1} \wedge \Big(\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
+      \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
+      \stamp{q}{k}(\vec{y}) \rightarrow \stamp{\psi}{k+1} \models
+      \phi_k\Big) \models I_k$, and 
+      %
+    \item $I_n \wedge \bigwedge_{q \in Q \setminus F} \forall y_1
+      \ldots \forall y_{\#(q)} ~.~ q(\vec{y})$ is
+      unsatisfiable,
+  \end{compactitem}
+  as required by Definition
+  \ref{def:generalized-lyndon-interpolant}. \qed}
+
+\comment[ri]{*** I am here ***}
+
+\subsection{Coverage}
 
@@
 
+\subsection{Lazy Annotation}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-09-20 quantifiers/draft.tex
ok

@@
 formula is \emph{positive} if each predicate symbol occurs under an
-even number of negations. A formula is in \emph{prenex form} if it is
-of the form $\varphi = Q_1x_1 \ldots Q_nx_n ~.~ \phi$, where $\phi$
-has no quantifiers. In this case we call $\phi$ the \emph{matrix} of
-$\varphi$. Every first-order formula can be written in prenex form, by
-renaming each quantified variable to a unique name and moving the
-quantifiers upfront.
+even number of negations and we denote by $\posforms(Q,X)$ the set of
+positive formulae with predicates from the set $Q \subseteq \preds$
+and free variables from the set $X \subseteq \vars$.
+
+A formula is in \emph{prenex form} if it is of the form $\varphi =
+Q_1x_1 \ldots Q_nx_n ~.~ \phi$, where $\phi$ has no quantifiers. In
+this case we call $\phi$ the \emph{matrix} of $\varphi$. Every
+first-order formula can be written in prenex form, by renaming each
+quantified variable to a unique name and moving the quantifiers
+upfront.
 
@@
 \item $Q$ is a finite set of predicates denoting control states, 
-\item $\iota$ is a sentence defining initial configurations,
+\item $\iota \in \posforms(Q,\emptyset)$ is a sentence defining initial configurations,
 \item $F \subseteq Q$ is the set of predicates denoting final states, and 
@@
   \(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is
-  a predicate, $a \in \Sigma$ is an input event, $X \cap
-  \set{y_1,\ldots,y_{\#(q)}} = \emptyset$, $\psi$ is a positive
-  formula such that $\fv{}{\phi} \subseteq X \cup
-  \set{y_1,\ldots,y_{\#(q)}}$ and each predicate that occurs (under an
-  even number of negations) in $\psi$ belongs also to $Q$. 
+  a predicate, $a \in \Sigma$ is an input event and $\psi \in
+  \posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive
+  formula, where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$.
 \end{compactitem}
@@
 
-Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$, an
-\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is
+\begin{definition}\label{def:execution}
+Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$ and a cube $c$, an
+\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$, starting with $c$, is
 a (possibly infinite) forest $\T = \set{T_1,T_2,\ldots}$, where each
 $T_i$ is a tree labeled with configurations, such that:
-\begin{compactitem}
-\item the labels of the roots of $T_1, T_2, \ldots$ form a cube from
-  $\cube{\minsem{\iota}}$,
-%
-\item if $q(d_1,\ldots,d_{\#(q)})$ labels a node on the level $j \in
+\begin{compactenum}
+\item\label{it1:execution} $c = \set{T(\epsilon) \mid T \in \T}$ is the set of
+  configurations labeling the roots of $T_1,T_2,\ldots$ and
+%
+\item\label{it2:execution} if $q(d_1,\ldots,d_{\#(q)})$ labels a node on the level $j \in
   [n-1]$ in $T_i$, then the labels of its children form a cube from
-  $\cube{\minsem{\psi}_{\nu_{j+1}[y_1 \leftarrow d_1,\ldots,y_{\#(q)}
-        \leftarrow d_{\#(q)}]}}$, where \(q(y_1,\ldots,y_{\#(q)})
-  \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is a transition rule.
+  $\cube{\minsem{\psi}_{\eta}}$, where $\eta = \nu_{j+1}[y_1
+    \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$ and
+  \(q(y_1,\ldots,y_{\#(q)}) \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is
+  a transition rule of $\A$.
+\end{compactenum}
+\end{definition}
+
+\begin{definition}\label{def:accepting}
+An execution $\T$ over $w$, starting with $c$, is \emph{accepting} if
+and only if \begin{compactitem}
+\item\label{it1:accepting} all paths in $\T$ have the same length $n$,
+  and
+  %
+\item\label{it2:accepting} the frontier of each tree $T \in \T$ is
+  labeled with \emph{final configurations} $q(d_1,\ldots,d_{\#(q)})$,
+  where $q \in F$.
 \end{compactitem} 
-An execution $\T$ over $w$ is said to be \emph{accepting} if and only
-if \begin{inparaenum}[(i)]
-\item all paths in $\T$ have the same length $n$, and
-\item the frontier of each tree $T \in \T$ is labeled with \emph{final
-  configurations} $q(d_1,\ldots,d_{\#(q)})$, where $q \in F$.
-\end{inparaenum} In this case, we say that $\A$ accepts $w$ and
-denote by $\lang{\A}$ the set of words accepted by $\A$.
-
+If $\A$ has an accepting execution over $w$ starting with a cube $c
+\in \cube{\minsem{\iota}}$, then $\A$ \emph{accepts} $w$ and let
+$\lang{\A}$ be the set of words accepted by $\A$.
+\end{definition}
 In this paper, we address the following questions: \begin{compactenum}
@@
 searches. These facts are interpolants extracted from the proof of
-spuriousness of the given (finite) path. The semi-algorithm uses
-moreover a coverage relation to ensure that a spurious counterexample
-is never explore twice, thus making progress towards either finding a
-real counterexample or a proof of emptiness. In the rest of this
-section, we discuss the generation of interpolants
-(\S\ref{sec:interpolants}) and the coverage (\S\ref{sec:coverage})
-used by our emptiness checking semi-algorithm (\S\ref{sec:impact}).
-
-\subsection{Interpolants}\label{sec:interpolants}
+spuriousness of the given (finite) path. The emptiness checking
+semi-algorithm uses moreover a coverage relation to ensure that a
+spurious counterexample is never explore twice, thus making progress
+towards either finding a real counterexample or a proof of emptiness.
+
+Our implementation is a version of the \impact~ lazy annotation
+semi-algorithm \cite{McMillan06}. The detailed pseudocode has been
+given for a simpler model of boolean alternating automata extended
+with variables, in \cite{IosifXu18} and is reported, for
+self-completeness, in Appendix \ref{app:pseudocode}. For the time
+being, we shall content ourselves of a high-level definition of the
+data structure used to prove emptiness of an automaton.
+
+Given a finite input event alphabet $\Sigma$, for two sequences
+$\alpha, \beta \in \Sigma^*$, we say that $\alpha$ is a prefix of
+$\beta$, written $\alpha \prefix \beta$, if $\alpha=\beta\gamma$ for
+some sequence $\gamma\in\Sigma^*$. A set $S$ of sequences
+is: \begin{compactitem}
+\item \emph{prefix-closed} if for each $\alpha \in S$, if $\beta \prefix
+\alpha$ then $\beta \in S$, and
+\item \emph{complete} if for each $\alpha \in S$, there exists $a \in
+  \Sigma$ such that $\alpha a \in S$ if and only if $\alpha b \in S$
+  for all $b \in \Sigma$. 
+\end{compactitem}
+Observe that a prefix-closed set is the backbone of a tree whose edges
+are labeled with input events. If the set is, moreover, complete, then
+every node of the tree has either zero successors, in which case it is
+called a \emph{leaf}, or it has a successor edge labeled with $a$ for
+each input event $a \in \Sigma$.
+
+\begin{definition}\label{def:unfolding}
+  An \emph{unfolding} of an automaton $\A =
+  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ is a finite partial mapping $U :
+  \Sigma^* \rightharpoonup_{\mathit{fin}} \posforms(Q,X)$, such
+  that: \begin{compactenum}
+  \item\label{it1:unfolding} $\dom(U)$ is a finite prefix-closed
+    complete set,
+    %
+  \item\label{it2:unfolding} $U(\epsilon) = \iota$, and
+    %
+  \item\label{it3:unfolding} for each sequence $\alpha a \in \dom(U)$,
+    such that $\alpha \in \Sigma^*$ and $a \in \Sigma$:
+    \[\stamp{U(\alpha)}{0} \wedge
+    \bigwedge_{q(\vec{y}) \arrow{a(X)}{} \psi} \forall y_1 \ldots
+    \forall y_{\#q} ~.~ \stamp{q}{0}(\vec{y}) \arrow{a(X)}{}
+    \stamp{\psi}{1} \models \stamp{U(\alpha a)}{1}\]
+  \end{compactenum}
+  Moreover, $U$ is \emph{safe} if for each $\alpha \in \dom(U)$, the
+  formula $U(\alpha) \wedge \bigwedge_{q \in Q \setminus F} \forall
+  y_1 \ldots \forall_{y_{\#(q)}} ~.~ q(\vec{y}) \rightarrow \bot$ is
+  unsatisfiable. 
+\end{definition}
+
+Lazy annotation semi-algorithms \cite{McMillan06,McMillan14}
+essentially build unfoldings of automata trying to discover
+counterexamples for emptiness. If the automaton $\A$ in question is
+non-empty, a systematic enumeration of the input event
+sequences\footnote{For instance, using breadth-first search.} from
+$\Sigma^*$ will suffice to discover a word $w \in \lang{\A}$, provided
+that the first-order theory of the data domain $\Data$ is decidable
+(Lemma \ref{lemma:acceptance}). However, if $\lang{\A} = \emptyset$,
+the enumeration of input event sequences may, in principle, run
+forever. The typical way of fighting this divergence problem is to
+define a \emph{coverage} relation between the nodes of the unfolding
+tree.
+
+\begin{definition}\label{def:coverage}
+  Given an unfolding $U$ of an automaton $\A =
+  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ a node $\alpha \in \dom(U)$ is
+  \emph{covered} by another node $\beta \in \dom(U)$, denoted $\alpha
+  \cover \beta$, if and only if there exists a node $\alpha' \prefix
+  \alpha$ such that, for all words $w \in \Sigma[X]^*$ and all cubes
+  $c \in \cube{\sem{U(\alpha')}}$, if $\A$ accepts $w$ started in $c$
+  then there exists a cube $d \in \cube{\sem{U(\beta)}}$ such that
+  $\A$ accepts $w$ started in $d$. Moreover, $U$ is \emph{closed} if
+  and only if every leaf from $\dom(U)$ is covered.
+\end{definition}
+
+A lazy annotation semi-algorithm can stop and report emptiness if it
+succeeds in building a closed and safe unfolding of the automaton. For
+instance, the \impact~ semi-algorithm \cite{McMillan06}
+
+The higher complexity of the first-order alternating automata model
+versus the automata considered in \cite{IosifXu18}, which uses boolean
+constants for control states and whose transition rules have no
+quantifiers, manifests itself within the method used to generate the
+labels of the unfolding, which is discussed in the rest of this
+section.
+
+\subsection{Interpolant Generation}\label{sec:interpolants}
 
@@
 the values of the program variables in a given control state, at a
-certain step of a given execution. As we discuss next, it is useful to
-distinguish between
+certain step of a given execution. It is useful to distinguish between
 \begin{inparaenum}[(i)]
@@
     %
-  \item\label{it2:subst-quant} there exists a GLI $(I_0, \ldots, I_n)$
-    for $\alpha$, such that $\voc{I_k} \cap \stamp{Q}{\leq n}
-    \subseteq \stamp{Q}{k}$ and $\voc{I_k} \cap \stamp{X}{\leq n}
-    \subseteq \stamp{X}{\leq k}$, for all $k \in [n]$.
+  \item\label{it2:subst-quant} one can effectively build a GLI $(I_0,
+    \ldots, I_n)$ for $\alpha$, such that $\voc{I_k} \cap
+    \stamp{Q}{\leq n} \subseteq \stamp{Q}{k}$ and $\voc{I_k} \cap
+    \stamp{X}{\leq n} \subseteq \stamp{X}{\leq k}$, for all $k \in
+          [n]$.
   \end{compactenum}

### 2018-09-20 quantifiers/draft.tex
ok

@@
 
-A lazy annotation semi-algorithm can stop and report emptiness if it
-succeeds in building a closed and safe unfolding of the automaton. For
-instance, the \impact~ semi-algorithm \cite{McMillan06}
-
-The higher complexity of the first-order alternating automata model
-versus the automata considered in \cite{IosifXu18}, which uses boolean
-constants for control states and whose transition rules have no
-quantifiers, manifests itself within the method used to generate the
-labels of the unfolding, which is discussed in the rest of this
-section.
+A lazy annotation semi-algorithm will stop and report emptiness
+provided that it succeeds in building a closed and safe unfolding of
+the automaton. For instance, the \impact~ semi-algorithm
+\cite{McMillan06} upon which our procedure is based, continuosly
+expands the domain of the unfolding, while partially recomputing its
+labeling and the coverage, and stops when a safe complete unfolding
+has been found. The following theorem proves the soundness of this
+method:
+
+\begin{theorem}\label{thm:soundness}
+  Given an automaton $\A$, if $\A$ has a nonempty safe closed
+  unfolding then $\lang{A} = \emptyset$. 
+\end{theorem}
+\proof{
+
+\qed}
+
+As mentioned above, we check emptiness of first-order alternating
+automata using the same method previously used to check emptiness of a
+simpler model of alternating automata, which uses boolean constants
+for control states and whose transition rules have no quantifiers
+\cite{IosifXu18}. The higher complexity of the automata model
+considered here, manifests itself within the interpolant generation
+procedure, used to label the unfoldings. We shall discuss generation
+of interpolants in the remainder of this section.
 

### 2018-09-20 quantifiers/draft.tex
ok

@@
 
-\comment[ri]{*** I am here ***}
-
-\subsection{Coverage}
-
-Assuming that the data domain is countable, the set of finite words
-$\Sigma[X]^*$ is countable and the emptiness problem is recursively
-enumerable. Moreover, by Lemma \ref{lemma:acceptance}, a systematic
-enumeration of sequences of input events $\alpha \in \Sigma^*$ will
-eventually discover an accepted word, provided that all path formulae
-$\pathform{\alpha}$ belong to a decidable first-order theory.
-
-However, if the automaton is empty, the enumeration of input event
-sequences will fail to terminate, in general. The classical way to
-fight this divergence problem is to consider a \emph{coverage} partial
-order, which is a subset of the following relation:
-
-\begin{definition}\label{def:coverage}
-  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ and
-  $S_1, S_2 \subseteq \bigcup_{q \in Q} \set{q(d_1,\ldots,d_{\#(q)})
-    \mid d_1,\ldots,d_{\#(q)} \in \Data}$, we have $S_1 \sqsubseteq
-  S_2$ if and only if 
-  \begin{equation}\label{eq:po}
-    \text{
-      \begin{minipage}{0.7\textwidth}
-        $\A$ accepts a word $w$ when started in a set of cubes $C_1
-        \in S_1$ only if $\A$ accepts $w$ when started in a set of
-        cubes $C_2 \in S_2$.
-    \end{minipage}} 
-  \end{equation}
-\end{definition}
-Since, in general the $\sqsubseteq$ relation is not definable using
-available decidable (SMT) theories, we consider the following stronger
-relation: $S_1 \supsetforex S_2$ if and only if, for each set of cubes
-$C_1 \in S_1$, there exists a set of cubes $C_2 \in S_2$ such that
-$C_1 \supseteq C_2$.
-
-\begin{lemma}\label{lemma:coverage}
-  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
-  sets of cubes $S_1, S_2 \subseteq \bigcup_{q \in Q}
-  \set{q(d_1,\ldots,d_{\#(q)}) \mid d_1,\ldots,d_{\#(q)} \in \Data}$
-  such that $S_1 \supsetforex S_2$, we have $S_1 \sqsubseteq S_2$.
-\end{lemma}
-\proof{  
-\qed}
-
-\subsection{Lazy Annotation}
+%% \subsection{Coverage}
+
+%% Assuming that the data domain is countable, the set of finite words
+%% $\Sigma[X]^*$ is countable and the emptiness problem is recursively
+%% enumerable. Moreover, by Lemma \ref{lemma:acceptance}, a systematic
+%% enumeration of sequences of input events $\alpha \in \Sigma^*$ will
+%% eventually discover an accepted word, provided that all path formulae
+%% $\pathform{\alpha}$ belong to a decidable first-order theory.
+
+%% However, if the automaton is empty, the enumeration of input event
+%% sequences will fail to terminate, in general. The classical way to
+%% fight this divergence problem is to consider a \emph{coverage} partial
+%% order, which is a subset of the following relation:
+
+%% \begin{definition}\label{def:coverage}
+%%   Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ and
+%%   $S_1, S_2 \subseteq \bigcup_{q \in Q} \set{q(d_1,\ldots,d_{\#(q)})
+%%     \mid d_1,\ldots,d_{\#(q)} \in \Data}$, we have $S_1 \sqsubseteq
+%%   S_2$ if and only if 
+%%   \begin{equation}\label{eq:po}
+%%     \text{
+%%       \begin{minipage}{0.7\textwidth}
+%%         $\A$ accepts a word $w$ when started in a set of cubes $C_1
+%%         \in S_1$ only if $\A$ accepts $w$ when started in a set of
+%%         cubes $C_2 \in S_2$.
+%%     \end{minipage}} 
+%%   \end{equation}
+%% \end{definition}
+%% Since, in general the $\sqsubseteq$ relation is not definable using
+%% available decidable (SMT) theories, we consider the following stronger
+%% relation: $S_1 \supsetforex S_2$ if and only if, for each set of cubes
+%% $C_1 \in S_1$, there exists a set of cubes $C_2 \in S_2$ such that
+%% $C_1 \supseteq C_2$.
+
+%% \begin{lemma}\label{lemma:coverage}
+%%   Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
+%%   sets of cubes $S_1, S_2 \subseteq \bigcup_{q \in Q}
+%%   \set{q(d_1,\ldots,d_{\#(q)}) \mid d_1,\ldots,d_{\#(q)} \in \Data}$
+%%   such that $S_1 \supsetforex S_2$, we have $S_1 \sqsubseteq S_2$.
+%% \end{lemma}
+%% \proof{  
+%% \qed}
 

### 2018-09-21 quantifiers/draft.tex
ok

@@
 abstraction-refinement semi-algorithm based on \emph{lazy annotation}
-\cite{McMillan06,McMillan14}. In a nutshell, a lazy annotation
-procedure systematically explores the set of execution traces of a
-program in search for a counterexample. When a spurious counterexample
-is found, the search backtracks and the program path is annotated with
-a set of learned facts, that excludes the path from future
-searches. These facts are interpolants extracted from the proof of
-spuriousness of the given (finite) path. The emptiness checking
-semi-algorithm uses moreover a coverage relation to ensure that a
-spurious counterexample is never explore twice, thus making progress
-towards either finding a real counterexample or a proof of emptiness.
-
-Our implementation is a version of the \impact~ lazy annotation
-semi-algorithm \cite{McMillan06}. The detailed pseudocode has been
-given for a simpler model of boolean alternating automata extended
-with variables, in \cite{IosifXu18} and is reported, for
-self-completeness, in Appendix \ref{app:pseudocode}. For the time
-being, we shall content ourselves of a high-level definition of the
-data structure used to prove emptiness of an automaton.
+\cite{McMillan06,McMillan14}.
+
+In a nutshell, a lazy annotation procedure systematically explores the
+set of execution paths (in our case, sequences of input events) in
+search of an accepting execution. Each path has a corresponding path
+formula that defines all words accepted along that path. If the
+current path formula is satisfiable, the automaton accepts a word and
+is not empty. Otherwise, the path is \emph{spurious}. When a spurious
+path is encountered, the search backtracks and the path is annotated
+with a set of learned facts, that marks this path as spurious in all
+future searches. The semi-algorithm uses moreover a coverage relation
+between paths, that optimize the search (the continuations of already
+covered paths are never explored) and sometimes provide a sound
+termination argument, when the automaton is empty.
+
+We check emptiness of first-order alternating automata using a version
+of the \impact~ lazy annotation semi-algorithm \cite{McMillan06}. The
+detailed pseudocode has been given for a simpler model of boolean
+alternating automata extended with variables \cite{IosifXu18}. To save
+space, we do not present the details of this algorithm (given in
+Appendix \ref{app:pseudocode}) and shall content ourselves of several
+high-level definitions.
 
@@
   \tuple{\Sigma,X,Q,\iota,F,\Delta}$ is a finite partial mapping $U :
-  \Sigma^* \rightharpoonup_{\mathit{fin}} \posforms(Q,X)$, such
-  that: \begin{compactenum}
+  \Sigma^* \rightharpoonup_{\mathit{fin}} \posforms(Q,\emptyset)$,
+  such that: \begin{compactenum}
   \item\label{it1:unfolding} $\dom(U)$ is a finite prefix-closed
@@
     \bigwedge_{q(\vec{y}) \arrow{a(X)}{} \psi} \forall y_1 \ldots
-    \forall y_{\#q} ~.~ \stamp{q}{0}(\vec{y}) \arrow{a(X)}{}
+    \forall y_{\#q} ~.~ \stamp{q}{0}(\vec{y}) \rightarrow
     \stamp{\psi}{1} \models \stamp{U(\alpha a)}{1}\]
@@
 
-Lazy annotation semi-algorithms \cite{McMillan06,McMillan14}
-essentially build unfoldings of automata trying to discover
-counterexamples for emptiness. If the automaton $\A$ in question is
-non-empty, a systematic enumeration of the input event
-sequences\footnote{For instance, using breadth-first search.} from
-$\Sigma^*$ will suffice to discover a word $w \in \lang{\A}$, provided
-that the first-order theory of the data domain $\Data$ is decidable
-(Lemma \ref{lemma:acceptance}). However, if $\lang{\A} = \emptyset$,
-the enumeration of input event sequences may, in principle, run
+Lazy annotation semi-algorithms \cite{McMillan06,McMillan14} build
+unfoldings of automata trying to discover counterexamples for
+emptiness. If the automaton $\A$ in question is non-empty, a
+systematic enumeration of the input event sequences\footnote{For
+  instance, using breadth-first search.} from $\Sigma^*$ will suffice
+to discover a word $w \in \lang{\A}$, provided that the first-order
+theory of the data domain $\Data$ is decidable (Lemma
+\ref{lemma:acceptance}). However, if $\lang{\A} = \emptyset$, the
+enumeration of input event sequences may, in principle, run
 forever. The typical way of fighting this divergence problem is to
@@
 provided that it succeeds in building a closed and safe unfolding of
-the automaton. For instance, the \impact~ semi-algorithm
-\cite{McMillan06} upon which our procedure is based, continuosly
-expands the domain of the unfolding, while partially recomputing its
-labeling and the coverage, and stops when a safe complete unfolding
-has been found. The following theorem proves the soundness of this
-method:
+the automaton. Notice that, by Definition \ref{def:coverage}, for any
+three nodes of an unfolding $U$, say $\alpha,\beta,\gamma \in
+\dom(U)$, if $\alpha \prec \beta$ and $\alpha \cover \gamma$, then
+$\beta \cover \gamma$ as well. As we show next (Theorem
+\ref{thm:soundness}), there is no need to expand covered nodes,
+because, intuitively, there is a word $w \in \lang{\A}$ such that
+$\alpha \preceq \event{w}$ and $\alpha \cover \gamma$ only if there
+exists another word $u \in \lang{\A}$ such that $\gamma \preceq
+\event{u}$.
+
+An unfolding node $\alpha \in \dom(U)$ is said to be \emph{spurious}
+if and only if $\accform{\alpha}$ is unsatisfiable. In this case, we
+change (refine) the labels $U(\beta)$ of (some of the) prefixes $\beta
+\prefix \alpha$, including $\alpha$ itself, such that $U(\alpha)$
+becomes $\bot$, thus indicating that there is no real execution of the
+automaton along that input event sequence. As a result of the change
+of labels, if a node $\gamma \prefix \alpha$ used to cover another
+node from $\dom(U)$, it might not cover it with the new
+label. Therefore, the coverage relation has to be recomputed after
+each refinement of the labeling. For a detailed presentation of the
+emptiness procedure, we refer the reader to \cite{IosifXu18} or
+Appendix \ref{app:pseudocode}.
+
+The semi-algorithm stops when (and if) a safe complete unfolding has
+been found. The following theorem proves the soundness of this
+termination argument:
 
@@
 \end{theorem}
-\proof{
+\proof{ Let $U$ be a safe and complete unfolding of $\A$, such that
+  $\dom(U) \neq \emptyset$. Suppose, by contradiction, that there
+  exists a word $w \in \lang{A}$ and let $\alpha \isdef
+  \event{w}$.
+
+  Since $w \in \lang{A}$, by Lemma \ref{lemma:acceptance}, that there
+  exists an interpretation $\I$ such that $\I,\data{w} \models
+  \accform{\alpha}$. If $\alpha \in \dom(U)$ then one can show, by
+  induction on the length $n\geq0$ of $w$, that $\pathform{\alpha}
+  \models \stamp{U(\alpha)}{n}$, thus $\I,\data{w} \models
+  \stamp{U(\alpha)}{n}$. But $\I,\data{w} \models \bigwedge_{q \in Q
+    \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
+  \stamp{q}{n}(\vec{y}) \rightarrow \bot$, hence $\stamp{U(\alpha)}{n}
+  \wedge \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
+  y_{\#(q)} ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot$. By renaming
+  $\stamp{q}{n}$ with $q$ in the previous formula, we obtain that
+  $U(\alpha) \wedge \forall y_1 \ldots \forall y_{\#(q)} ~.~
+  q(\vec{y}) \rightarrow \bot$ is satisfiable, thus $U$ is not safe,
+  contradiction.
 
@@
 considered here, manifests itself within the interpolant generation
-procedure, used to label the unfoldings. We shall discuss generation
-of interpolants in the remainder of this section.
-
-\subsection{Interpolant Generation}\label{sec:interpolants}
+procedure, used to refine the labeling of the unfolding. We discuss
+generation of interpolants in the next section.
+
+\section{Interpolant Generation}\label{sec:interpolants}
 

### 2018-09-25 quantifiers/draft.tex
ok

@@
 
-The problem of checking emptiness of a given automata is undecidable,
+The problem of checking emptiness of a given automaton is undecidable,
 even for automata with predicates of arity two, whose transition rules
-use only equalities and disequalities, with no transition quantifiers
-\cite{Farzan15}. Since even such simple classes of alternating
-automata have no general decision procedure for emptiness, we use an
-abstraction-refinement semi-algorithm based on \emph{lazy annotation}
-\cite{McMillan06,McMillan14}.
+use only equalities and disequalities, having no transition
+quantifiers \cite{Farzan15}. Since even such simple classes of
+alternating automata have no general decision procedure for emptiness,
+we use an abstraction-refinement semi-algorithm based on \emph{lazy
+  annotation} \cite{McMillan06,McMillan14}.
 
@@
 current path formula is satisfiable, the automaton accepts a word and
-is not empty. Otherwise, the path is \emph{spurious}. When a spurious
-path is encountered, the search backtracks and the path is annotated
-with a set of learned facts, that marks this path as spurious in all
-future searches. The semi-algorithm uses moreover a coverage relation
-between paths, that optimize the search (the continuations of already
-covered paths are never explored) and sometimes provide a sound
-termination argument, when the automaton is empty.
+is not empty. Otherwise, the path is said to be \emph{spurious}. When
+a spurious path is encountered, the search backtracks and the path is
+annotated with a set of learned facts, that marks this path as
+spurious in all future searches. The semi-algorithm uses moreover a
+coverage relation between paths, that optimize the search (the
+continuations of already covered paths are never explored) and
+sometimes provide a sound termination argument, when the automaton is
+empty.
 
@@
 \ref{thm:soundness}), there is no need to expand covered nodes,
-because, intuitively, there is a word $w \in \lang{\A}$ such that
+because, intuitively, there exists a word $w \in \lang{\A}$ such that
 $\alpha \preceq \event{w}$ and $\alpha \cover \gamma$ only if there
 exists another word $u \in \lang{\A}$ such that $\gamma \preceq
-\event{u}$.
+\event{u}$. Hence, exploring only those input event sequences that are
+continuations of $\gamma$ (and ignoring those of $\alpha$) suffices in
+order to find a counterexample for emptiness, if one exists.
 
@@
 if and only if $\accform{\alpha}$ is unsatisfiable. In this case, we
-change (refine) the labels $U(\beta)$ of (some of the) prefixes $\beta
-\prefix \alpha$, including $\alpha$ itself, such that $U(\alpha)$
-becomes $\bot$, thus indicating that there is no real execution of the
-automaton along that input event sequence. As a result of the change
-of labels, if a node $\gamma \prefix \alpha$ used to cover another
-node from $\dom(U)$, it might not cover it with the new
-label. Therefore, the coverage relation has to be recomputed after
-each refinement of the labeling. For a detailed presentation of the
-emptiness procedure, we refer the reader to \cite{IosifXu18} or
-Appendix \ref{app:pseudocode}.
+change (refine) the labels of (some of the) prefixes of $\alpha$ (and
+that of $\alpha$), such that $U(\alpha)$ becomes $\bot$, thus
+indicating that there is no real execution of the automaton along that
+input event sequence. As a result of the change of labels, if a node
+$\gamma \prefix \alpha$ used to cover another node from $\dom(U)$, it
+might not cover it with the new label. Therefore, the coverage
+relation has to be recomputed after each refinement of the
+labeling. For a detailed presentation of the emptiness procedure, we
+refer the reader to \cite{IosifXu18} (or Appendix
+\ref{app:pseudocode}).
 
@@
 \begin{theorem}\label{thm:soundness}
-  Given an automaton $\A$, if $\A$ has a nonempty safe closed
-  unfolding then $\lang{A} = \emptyset$. 
+  If an automaton $\A$ has a nonempty safe closed unfolding then
+  $\lang{A} = \emptyset$.
 \end{theorem}
-\proof{ Let $U$ be a safe and complete unfolding of $\A$, such that
+\proof{ 
+  Let $U$ be a safe and complete unfolding of $\A$, such that
   $\dom(U) \neq \emptyset$. Suppose, by contradiction, that there
@@
 
-  Since $w \in \lang{A}$, by Lemma \ref{lemma:acceptance}, that there
+  Since $w \in \lang{A}$, by Lemma \ref{lemma:acceptance}, there
   exists an interpretation $\I$ such that $\I,\data{w} \models
-  \accform{\alpha}$. If $\alpha \in \dom(U)$ then one can show, by
-  induction on the length $n\geq0$ of $w$, that $\pathform{\alpha}
-  \models \stamp{U(\alpha)}{n}$, thus $\I,\data{w} \models
-  \stamp{U(\alpha)}{n}$. But $\I,\data{w} \models \bigwedge_{q \in Q
+  \accform{\alpha}$. Assume first that $\alpha \in \dom(U)$.  In this
+  case, one can show, by induction on the length $n\geq0$ of $w$, that
+  $\pathform{\alpha} \models \stamp{U(\alpha)}{n}$, thus $\I,\data{w}
+  \models \stamp{U(\alpha)}{n}$. Since $\I,\data{w} \models
+  \accform{\alpha}$, we have $\I,\data{w} \models \bigwedge_{q \in Q
     \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
@@
   y_{\#(q)} ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot$. By renaming
-  $\stamp{q}{n}$ with $q$ in the previous formula, we obtain that
+  $\stamp{q}{n}$ with $q$ in the previous formula, we obtain
   $U(\alpha) \wedge \forall y_1 \ldots \forall y_{\#(q)} ~.~
@@
 
-\qed}
+  We proceed thus under the assumption that $\alpha \not\in
+  \dom(U)$. Since $\dom(U)$ is a nonempty prefix-closed set, there
+  exists a strict prefix $\alpha'$ of $\alpha$ that is a leaf of
+  $\dom(U)$. Since $U$ is closed, $\alpha'$ is closed and let
+  $\alpha_1 \prefix \alpha' \prefix \alpha$ such that $\alpha_1 \cover
+  \beta_1$, for some node $\beta_1 \in \dom(U)$. Let $\gamma_1$ be the
+  unique sequence such that $\alpha_1\gamma_1 = \alpha$. By Definition
+  \ref{def:coverage}, since $\alpha_1 \cover \beta_1$ and $\event{w} =
+  \alpha_1\gamma_1 \in \lang{\A}$, there exists a word $w_1$ and a
+  cube $c_1 \in \cube{\sem{U(\beta_1)}}$ such that $\event{w_1} =
+  \gamma_1$ and $\A$ accepts $w_1$ starting with $c_1$. \qed}
 

### 2018-09-26 quantifiers/draft.tex
ok

@@
   $\A$ accepts $w$ started in $d$. Moreover, $U$ is \emph{closed} if
-  and only if every leaf from $\dom(U)$ is covered.
+  and only if every leaf from $\dom(U)$ is covered by an uncovered
+  node.
 \end{definition}
@@
   exists a strict prefix $\alpha'$ of $\alpha$ that is a leaf of
-  $\dom(U)$. Since $U$ is closed, $\alpha'$ is closed and let
-  $\alpha_1 \prefix \alpha' \prefix \alpha$ such that $\alpha_1 \cover
-  \beta_1$, for some node $\beta_1 \in \dom(U)$. Let $\gamma_1$ be the
-  unique sequence such that $\alpha_1\gamma_1 = \alpha$. By Definition
-  \ref{def:coverage}, since $\alpha_1 \cover \beta_1$ and $\event{w} =
-  \alpha_1\gamma_1 \in \lang{\A}$, there exists a word $w_1$ and a
-  cube $c_1 \in \cube{\sem{U(\beta_1)}}$ such that $\event{w_1} =
-  \gamma_1$ and $\A$ accepts $w_1$ starting with $c_1$. \qed}
+  $\dom(U)$. Since $U$ is closed, the leaf $\alpha'$ must be covered
+  and let $\alpha_1 \prefix \alpha' \prefix \alpha$ be a node such
+  that, for some uncovered node $\beta_1 \in \dom(U)$, for every word
+  $w'$ and every cube $c \in \cube{\sem{U(\alpha_1)}}$ such that $\A$
+  accepts $w'$ starting with $c$, there exists a cube $d \in
+  \cube{\sem{U(\beta_1)}}$ such that $\A$ accepts $w'$ starting with
+  $d$. Let $\gamma_1$ be the unique sequence such that
+  $\alpha_1\gamma_1 = \alpha$. By Definition \ref{def:coverage}, since
+  $\alpha_1 \cover \beta_1$ and $\event{w} = \alpha_1\gamma_1 \in
+  \lang{\A}$, there exists a word $w_1$ and a cube $c_1 \in
+  \cube{\sem{U(\beta_1)}}$ such that $\event{w_1} = \gamma_1$ and $\A$
+  accepts $w_1$ starting with $c_1$. If $\beta_1\gamma_1 \in \dom(U)$,
+  we obtain a contradiction by a similar argument as above. Hence
+  $\beta_1\gamma_1 \not\in \dom(U)$ and there exists a leaf of
+  $\dom(U)$ which is also a prefix of $\beta_1\gamma_1$. Since $U$ is
+  closed, this leaf is covered by an uncovered node $\beta_2 \in
+  \dom(U)$ and let $\alpha_2 \in \dom(U)$ be the minimal (in the
+  prefix partial order) node such that $\beta_1 \prefix \alpha_2
+  \prefix \beta_1\gamma_1$ and $\alpha_2 \cover \beta_2$. Let
+  $\gamma_2$ be the unique sequence such that $\alpha_2\gamma_2 =
+  \beta_1\gamma_1$. Since $\beta_1$ is uncovered, we have $\beta_1
+  \neq \alpha_2$ and thus $\len{\gamma_1} > \len{\gamma_2}$. By
+  repeating the above reasoning for $\alpha_2$, $\beta_2$ and
+  $\gamma_2$, we obtain an infinite sequence $\len{\gamma_1} >
+  \len{\gamma_2} > \ldots$, which is again a contradiction. \qed}
 

### 2018-09-26 quantifiers/draft.tex
ok

--- 2018-09-25 quantifiers/draft.tex
@@
 search of an accepting execution. Each path has a corresponding path
-formula that defines all words accepted along that path. If the
-current path formula is satisfiable, the automaton accepts a word and
-is not empty. Otherwise, the path is said to be \emph{spurious}. When
-a spurious path is encountered, the search backtracks and the path is
-annotated with a set of learned facts, that marks this path as
-spurious in all future searches. The semi-algorithm uses moreover a
-coverage relation between paths, that optimize the search (the
-continuations of already covered paths are never explored) and
-sometimes provide a sound termination argument, when the automaton is
-empty.
+formula that defines all words accepted along that path. If the path
+formula is satisfiable, the automaton accepts a word. Otherwise, the
+path is said to be \emph{spurious}. When a spurious path is
+encountered, the search backtracks and the path is annotated with a
+set of learned facts, that marks this path as infeasible. The
+semi-algorithm uses moreover a coverage relation between paths,
+ensuring that the continuations of already covered paths are never
+explored. Sometimes this coverage relation provides a sound
+termination argument, when the automaton is empty.
 
 We check emptiness of first-order alternating automata using a version
-of the \impact~ lazy annotation semi-algorithm \cite{McMillan06}. The
-detailed pseudocode has been given for a simpler model of boolean
-alternating automata extended with variables \cite{IosifXu18}. To save
-space, we do not present the details of this algorithm (given in
-Appendix \ref{app:pseudocode}) and shall content ourselves of several
+of the \impact~ lazy annotation semi-algorithm \cite{McMillan06}. An
+analogous procedure is given in \cite{IosifXu18}, for a simpler model
+of alternating automata, that uses only predicates or arity zero
+(booleans) and no transition quantifiers. For simplicity, we do not
+present the details of this algorithm (see Appendix
+\ref{app:pseudocode}) and shall content ourselves of several
 high-level definitions.

### 2018-10-04 quantifiers/draft.tex
ok

@@
 the values of the program variables in a given control state, at a
-certain step of a given execution. It is useful to distinguish between
+certain step of a given execution. However, in an alternating model of
+computation, it is useful to distinguish between
 \begin{inparaenum}[(i)]
@@
 section, let us fix an automaton $\A =
-\tuple{\Sigma,X,Q,\iota,F,\Delta}$.
+\tuple{\Sigma,X,Q,\iota,F,\Delta}$. Due to the above observation, none
+of the interpolants considered will be control-local and we shall use
+the term \emph{local} to denote time-local interpolants, with no free
+variables.
 
@@
   \end{compactenum}
-  Moreover, the GLI is \emph{local} iff $\voc{I_k} \cap \stamp{Q}{\leq n}
-  \subseteq \stamp{Q}{k}$ and $\voc{I_k} \cap \stamp{X}{\leq n}
-  \subseteq \stamp{X}{k}$, for all $k \in [n]$.
+  Moreover, the GLI is \emph{local} iff $\voc{I_k} \cap \stamp{Q}{\leq
+    n} \subseteq \stamp{Q}{k}$ and $\voc{I_k} \cap \stamp{X}{\leq n} =
+  \emptyset$, for all $k \in [n]$.
 \end{definition}

### 2018-10-04 quantifiers/draft.tex
ok

@@
   \end{compactenum}
-  Moreover, the GLI is \emph{local} iff $\voc{I_k} \cap \stamp{Q}{\leq
-    n} \subseteq \stamp{Q}{k}$ and $\voc{I_k} \cap \stamp{X}{\leq n} =
-  \emptyset$, for all $k \in [n]$.
+  Moreover, the GLI is \emph{local} if and only if $\voc{I_k}
+  \subseteq \stamp{Q}{k}$, for all $k \in [n]$.
 \end{definition}
@@
 procedure, when $\phi$ and $\psi$ contain predicates and have
-non-trivial quantifier alternation.
+non-trivial quantifier alternation. In this case, some provers use
+instantiation heuristics for the universal quantifiers that are
+sufficient for proving unsatisfiability, however these heuristics are
+not always suitable for interpolant generation. Consequently, from now
+on, we assume the existence of an effective Lyndon interpolation
+procedure only for decidable theories, such as the quantifier-free
+linear (integer) arithmetic with uninterpreted functions (QFLIA,
+QFLRA, etc.).
 
@@
 Presburger cardinality constraints (BAPA)
-\cite{KuncakMayerPiskacSuter12}. Furthermore, we describe a cubic-time
-witness generation for difference bounds constraints, a simple subset
-of linear integer (real) arithmetic (\S\ref{sec:applications}).
+\cite{KuncakMayerPiskacSuter12}.
+
+%% Furthermore, we describe a cubic-time witness generation for
+%% difference bounds constraints, a simple subset of linear integer
+%% (real) arithmetic (\S\ref{sec:applications}).
 
@@
 both written in prenex form, with matrices $\widehat{\Phi}$ and
-$\overline{\Phi}$, respectively. W.l.o.g. we consider a monotonic
-mapping $\xi : [1,m] \rightarrow [n]$ assigning each transition
-quantifier from $\substform{\alpha}$ its corresponding time-stamp from
-$\accform{\alpha}$. Because we assumed that the first-order theory of
-$\Data$ has quantifier elimination, the satisfiability problem for
-$\substform{\alpha}$ is decidable. If $\substform{\alpha}$ is
-satisfiable, we build a counterexample for emptiness $w$ such that
-$\event{w}=\alpha$ and $\data{w}$ is a satisfying assignment for
-$\substform{\alpha}$. Otherwise, $\substform{\alpha}$ is unsatisfiable
-and there exist witness terms $\tau_{i_1} \ldots \tau_{i_\ell}$, such
-that $\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq\xi(i_j)}$ (Theorem
-\ref{thm:skolem}) and the formula $\overline{\Phi}[\tau_{i_1}/x_{i_1},
-  \ldots, \tau_{i_\ell}/x_{i_\ell}]$, obtained from the matrix of
-$\substform{\alpha}$, by replacing the universally quantified
-variables with witness terms, is unsatisfiable. Then it turns out that
-the formula $\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
+$\overline{\Phi}$, respectively. Because we assumed that the
+first-order theory of $\Data$ has quantifier elimination, the
+satisfiability problem for $\substform{\alpha}$ is decidable. If
+$\substform{\alpha}$ is satisfiable, we build a counterexample for
+emptiness $w$ such that $\event{w}=\alpha$ and $\data{w}$ is a
+satisfying assignment for $\substform{\alpha}$. Otherwise,
+$\substform{\alpha}$ is unsatisfiable and there exist witness terms
+$\tau_{i_1} \ldots \tau_{i_\ell}$, where $\set{i_1, \ldots, i_\ell} =
+\set{j \in [1,m] \mid Q_j = \forall}$, such that
+$\overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
+  \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable (Theorem
+\ref{thm:skolem}). Then it turns out that the formula
+$\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
   \tau_{i_\ell}/x_{i_\ell}]$, obtained analogously from the matrix of
 $\quantform{\alpha}$, is unsatisfiable as well (Lemma
-\ref{lemma:subst-quant}).
-
-Because this latter formula is structured as a conjunction of formulae
-$\stamp{\iota}{0} \wedge \phi_1 \ldots \wedge \phi_n \wedge \psi$,
-where $\voc{\phi_k} \subseteq \stamp{Q}{k-1} \cup \stamp{Q}{k} \cup
-\stamp{X}{\leq k}$ for all $k \in [1,n]$ and $\voc{\psi} \subseteq
-\stamp{Q}{n}$, it is now possible to use an existing interpolation
-procedure for the quantifier-free theory of $\Data$, extended with
-uninterpreted function symbols, to compute a sequence of non-local GLI
-$(I_0, \ldots, I_n)$ such that $\voc{I_k} \cap \stamp{Q}{\leq n}
-\subseteq \stamp{Q}{k}$ and $\voc{I_k} \cap \stamp{X}{\leq n}
-\subseteq \stamp{X}{\leq k}$, for all $k \in [n]$. Since each formula
-$I_k$ of the GLI may refer to the current predicates $\stamp{Q}{k}$
-and to the past and current input variables $\stamp{X}{\leq k}$, we
-further give up on the locality of interpolants and work with
-non-local GLI that are effectively computable by a combination of
-quantifier elimination with interpolation for quantifier-free data
-theories with uninterpreted functions.
+\ref{lemma:subst-quant}). Because this latter formula is structured as
+a conjunction of formulae $\stamp{\iota}{0} \wedge \phi_1 \ldots
+\wedge \phi_n \wedge \psi$, where $\voc{\phi_k} \cap \stamp{Q}{\leq n}
+\subseteq \stamp{Q}{k-1} \cup \stamp{Q}{k}$ and $\voc{\psi} \cap
+\stamp{Q}{\leq n} \subseteq \stamp{Q}{n}$, it is now possible to use
+an existing interpolation procedure for the quantifier-free theory of
+$\Data$, extended with uninterpreted function symbols, to compute a
+sequence of non-local GLI $(I_0, \ldots, I_n)$ such that $\voc{I_k}
+\cap \stamp{Q}{\leq n} \subseteq \stamp{Q}{k}$, for all $k \in [n]$.
+
+%% Since each formula $I_k$ of the GLI may refer to the current
+%% predicates $\stamp{Q}{k}$ and to the past and current input variables
+%% $\stamp{X}{\leq k}$, we further give up on the locality of
+%% interpolants and work with non-local GLI that are effectively
+%% computable by a combination of quantifier elimination with
+%% interpolation for quantifier-free data theories with uninterpreted
+%% functions.
 
@@
   $Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$ be a prenex form of
-  $\quantform{\alpha}$, where $Q_ix_i$ are transition quantifiers and
-  $\xi : [1,m] \rightarrow [n]$ is a monotonic function mapping each
-  quantified variable to the time-stamp of the step where it occurs in
-  $\accform{\alpha}$. Then the following hold: \begin{compactenum}
-  \item\label{it1:subst-quant} there exist witness terms $\tau_{i_1},
-    \ldots, \tau_{i_\ell}$, where $1 \leq i_1 < \ldots < i_\ell \leq
-    m$ and $\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)}$ for
-    all $j \in [1,\ell]$, such that $\widehat{\Phi}[\tau_{i_1}/x_{i_1},
-      \ldots, \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable, and
+  $\quantform{\alpha}$ and let $\xi : [1,m] \rightarrow [n]$ be a
+  monotonic function mapping each transition quantifier to the minimal
+  index from the sequence $\quantformn{\alpha_0}, \ldots,
+  \quantformn{\alpha_n}$ where it occurs. Then the following
+  hold: \begin{compactenum}
+  \item\label{it1:subst-quant} one can effectively build witness terms
+    $\tau_{i_1}, \ldots, \tau_{i_\ell}$, where $\set{i_1, \ldots,
+    i_\ell} = \set{j \in [1,m] \mid Q_j = \forall}$ and
+    $\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup \set{x_k
+    \mid k < i_j, Q_k = \exists}$, for all $j \in [1,\ell]$, such that
+    $\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
+    \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable, and
     %
   \item\label{it2:subst-quant} one can effectively build a GLI $(I_0,
-    \ldots, I_n)$ for $\alpha$, such that $\voc{I_k} \cap
-    \stamp{Q}{\leq n} \subseteq \stamp{Q}{k}$ and $\voc{I_k} \cap
-    \stamp{X}{\leq n} \subseteq \stamp{X}{\leq k}$, for all $k \in
-          [n]$.
+    \ldots, I_n)$ for $\alpha$, such that $\voc{I_k} \subseteq
+    \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
+      \max(\xi^{-1}(k)),~ Q_j = \exists}$, for all $k \in [n]$.
   \end{compactenum}
@@
   $Q_1x_1 \ldots Q_mx_m ~.~ \overline{\Phi}$ be prenex forms for
-  $\quantform{\alpha}$ and $\substform{\alpha}$,
-  respectively. Further, let $\xi : [1,m] \rightarrow [n]$ be the
-  monotone function mapping of each quantified variable to time-stamp
-  where it was introduced. Then there exist witness terms $\tau_{i_1},
-  \ldots, \tau_{i_\ell}$, where $\set{i_1,\ldots,i_\ell}=\set{i \in
-    [1,m] \mid Q_i=\forall}$, such that $\voc{\tau_{i_j}} \subseteq
-  \stamp{X}{\xi(i_j)}$, for all $j \in [1,\ell]$ and
-  $\overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
+  $\quantform{\alpha}$ and $\substform{\alpha}$, respectively. Since
+  we assumed that the first-order theory of the data domain has
+  witness-producing quantifier elimination, using Theorem
+  \ref{thm:skolem} one can effectively build witness terms
+  $\tau_{i_1}, \ldots, \tau_{i_\ell}$, where
+  $\set{i_1,\ldots,i_\ell}=\set{i \in [1,m] \mid Q_i=\forall}$, such
+  that: \begin{compactitem}
+  \item $\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup
+    \set{x_k \mid k < i_j, Q_k = \exists}$, for all $j \in [1,\ell]$ and
+    %
+  \item $\overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
     \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.
-
+  \end{compactitem}
   Let $\widehat{\Phi}_0$, \ldots, $\widehat{\Phi}_n$ be the sequence
   of quantifier-free formulae, defined as follows: \begin{compactitem}
-  \item $\widehat{\Phi}_0$ is the matrix of the prenex form of
+  \item $\widehat{\Phi}_0$ is the matrix of some prenex form of
     $\stamp{\iota}{0}$,
     %
-  \item for all $i=1,\ldots,n$, let $\widehat{\Phi}_i$ be a prenex
-    form matrix of the formula:
+  \item for all $i=1,\ldots,n$, let $\widehat{\Phi}_i$ be the matrix
+    of some prenex form of the formula:
     \[\widehat{\Phi}_i \isdef \widehat{\Phi}_{i-1} \wedge
@@
   \end{compactitem}
-  It is easy to see that $\widehat{\Phi}$ is a prenex form matrix of the formula:   
+  It is easy to see that $\widehat{\Phi}$ is the matrix of some prenex
+  form of the formula:
   \[\widehat{\Phi}_n \wedge \underbrace{\bigwedge_{\begin{array}{l}
@@
   \psi$. Consider the sequence of witness terms $\tau_{i_1}, \ldots,
-  \tau_{i_\ell}$, such that $\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq
-    \xi(i_j)}$, for all $j \in [1,\ell]$, whose existence is proved by
-  point (\ref{it1:subst-quant}). By the definition of $\xi$, we obtain
-  the following: \begin{compactitem}
-  \item $\voc{\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots, \tau_{i_\ell}/x_{i_\ell}]} \subseteq 
-    \stamp{Q}{0} \cup \stamp{X}{0}$, 
-    %
-  \item $\voc{\phi_i[\tau_{i_1}/x_{i_1}, \ldots, \tau_{i_\ell}/x_{i_\ell}]} \subseteq 
-    \stamp{Q}{i-1} \cup \stamp{Q}{i} \cup \stamp{X}{\leq i}$, for all $i \in [1,n]$,
-    %
-  \item $\voc{\psi[\tau_{i_1}/x_{i_1}, \ldots, \tau_{i_\ell}/x_{i_\ell}]} \subseteq 
-    \stamp{Q}{n} \cup \stamp{X}{\leq n}$. 
+  \tau_{i_\ell}$, whose existence is proved by point
+  (\ref{it1:subst-quant}). Because $\voc{\tau_{i_j}} \subseteq
+  \stamp{X}{\leq \xi(i_j)} \cup \set{x_k \mid k < i_j, Q_k =
+    \exists}$, for all $j \in [1,\ell]$, and moreover $\xi$ is
+  monotonic, we obtain: \begin{compactitem}
+  \item $\voc{\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
+      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{0} \cup
+    \stamp{X}{0} \cup \set{x_j \mid j < \max(\xi^{-1}(0)), Q_j =
+      \exists}$,
+    %
+  \item $\voc{\phi_i[\tau_{i_1}/x_{i_1}, \ldots,
+      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{i-1} \cup
+    \stamp{Q}{i} \cup \stamp{X}{\leq i} \cup \set{x_j \mid j <
+      \max(\xi^{-1}(i)), Q_j = \exists}$, for all $i \in [1,n]$,
+    %
+  \item $\voc{\psi[\tau_{i_1}/x_{i_1}, \ldots,
+      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{n} \cup
+    \stamp{X}{\leq n} \cup \set{x_j \mid j \in [1,m], Q_j = \exists}$.
   \end{compactitem}
@@
       \tau_{i_\ell}/x_{i_\ell}] \models I_0$ and $\voc{I_0} \subseteq
-      \stamp{Q}{0} \cup \stamp{X}{0}$,
+      \stamp{Q}{0} \cup \stamp{X}{0} \cup \set{x_j \mid j <
+        \max(\xi^{-1}(0)), Q_j = \exists}$,
       %
@@
       \tau_{i_\ell}/x_{i_\ell}] \models I_k$ and $\voc{I_k} \subseteq
-      \stamp{Q}{k} \cup \stamp{X}{\leq k}$, for all $k \in [1,n]$,
+      \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
+        \max(\xi^{-1}(k)), Q_j = \exists}$, for all $k \in [1,n]$,
       %
@@
       \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-      \stamp{q}{k}(\vec{y}) \rightarrow \stamp{\psi}{k+1} \models
-      \phi_k\Big) \models I_k$, and 
-      %
-    \item $I_n \wedge \bigwedge_{q \in Q \setminus F} \forall y_1
-      \ldots \forall y_{\#(q)} ~.~ q(\vec{y})$ is
-      unsatisfiable,
+      \stamp{q}{k}(\vec{y}) \rightarrow \stamp{\psi}{k+1}\Big) \models
+      I_{k-1} \wedge \phi_k \models I_k$, and 
+      %
+    \item $I_n \wedge \Big(\bigwedge_{q \in Q \setminus F} \forall y_1
+      \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot\Big)
+      \models I_n \wedge \psi \models \bot$,     
   \end{compactitem}
@@
   \ref{def:generalized-lyndon-interpolant}. \qed}
+
+In conclusion, under two assumptions about the first-order theory of
+the data domain, namely \begin{inparaenum}[(i)]
+\item witness-producing quantifier elimination, and
+\item Lyndon interpolation for the quantifier-free fragment with
+  uninterpreted functions,
+\end{inparaenum}
+we developped a rather generic method that produces generalized Lyndon
+interpolants for unfeasible input event sequences. Moreover, each
+formula $I_k$ in the interpolant refers only to the current predicate
+symbols $\stamp{Q}{I_k}$, the current and past input variables
+$\stamp{X}{\leq k}$ and the existentially quantified transition
+variables introduced at the previous steps $\set{x_j \mid j <
+  \max(\xi^{-1}(k)), Q_j = \exists}$. The remaining question is how to
+use such non-local interpolants for the coverage between the nodes of
+the unfolding of an automaton (Definition \ref{def:coverage}).
+
+\subsection{Coverage with Non-local Interpolants}
 

### 2018-10-05 quantifiers/draft.tex
ok

@@
   [n-1]$: \begin{compactenum}
-    \item $\pset{-}{I_k} = \emptyset$,
-      %
-    \item $\stamp{\iota}{0} \models I_0$ and \(I_k \wedge
-      \Big(\bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in \Delta}
-      \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
-      \rightarrow \stamp{\psi}{k+1}\Big) \models I_{k+1}\),
-      %
-    \item $I_n \wedge \bigwedge_{q \in Q \setminus F} \forall y_1
-      \ldots \forall y_{\#(q)} ~.~ q(\vec{y})$ is unsatisfiable.
+    \item\label{it1:generalized-lyndon-interpolant} $\pset{-}{I_k} =
+      \emptyset$,
+      %
+    \item\label{it2:generalized-lyndon-interpolant} $\stamp{\iota}{0}
+      \models I_0$ and \(I_k \wedge \Big(\bigwedge_{q(\vec{y})
+        \arrow{a_i(X)}{} \psi \in \Delta} \forall y_1 \ldots \forall
+      y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y}) \rightarrow
+      \stamp{\psi}{k+1}\Big) \models I_{k+1}\),
+      %
+    \item\label{it3:generalized-lyndon-interpolant} $I_n \wedge
+      \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
+      y_{\#(q)} ~.~ q(\vec{y})$ is unsatisfiable.
   \end{compactenum}
@@
 yields the unsatisfiable conjunction: 
-\[\begin{array}{l}
-z_1 \geq 0 \wedge \stamp{q}{0}(z_1) ~\wedge \\
-\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge (z_2 \geq z_1 \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_2)) ~\wedge \\
-\stamp{q}{1}(\stamp{x}{1}+z_2) \rightarrow \stamp{x}{1}+z_2 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_2)
+\[\begin{array}{c}
+z_1 \geq 0 \wedge \stamp{q}{0}(z_1)  \\ \wedge \\
+\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge (z_1 \geq z_1 \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_1)) \\ \wedge \\
+\stamp{q}{1}(\stamp{x}{1}+z_1) \rightarrow \stamp{x}{1}+z_1 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_1)
 \end{array}\]
@@
 In conclusion, under two assumptions about the first-order theory of
-the data domain, namely \begin{inparaenum}[(i)]
+the data domain, namely\begin{inparaenum}[(i)]
 \item witness-producing quantifier elimination, and
@@
   \max(\xi^{-1}(k)), Q_j = \exists}$. The remaining question is how to
-use such non-local interpolants for the coverage between the nodes of
-the unfolding of an automaton (Definition \ref{def:coverage}).
-
-\subsection{Coverage with Non-local Interpolants}
+use such non-local interpolants to label the unfolding of an automaton
+(Definition \ref{def:unfolding}) and to compute the coverage between
+nodes of the unfolding (Definition \ref{def:coverage}).
+
+\subsection{Unfolding with Non-local Interpolants}
+
+As required by Definition \ref{def:unfolding}, the unfolding $U$ of an
+automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ is labeled by
+formulae $U(\alpha) \in \posforms(Q,\emptyset)$, with no free symbols,
+other than predicate symbols, such that the labeling is compatible
+with the transition relation of the automaton (point
+\ref{it3:unfolding} of Definition \ref{def:unfolding}).
+
+Suppose that a given input event sequence $\alpha \in \dom(U)$ is
+labeled with a sequence of formulae $U(\alpha_0) = \iota, U(\alpha_1),
+\ldots U(\alpha_n) = U(\alpha)$ in a way that fulfills the requirement
+from Definition \ref{def:unfolding} (\ref{it3:unfolding}) and,
+moreover, it has a non-local GLI $(I_0, \ldots, I_n)$, such that
+$\voc{I_k} \subseteq \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup
+\set{x_{i_1},\ldots,x_{i_k}}$, where $x_{i_1},\ldots,x_{i_n}$ are the
+renamed variables occurring in the rules of $\Delta$ under the scope
+of existential transition quantifiers (formally these are the
+existentially quantified variables from the prenex normal form of
+$\quantform{\alpha_k}$). The interpolant is used in refining the
+labeling of the sequence as follows, for all $k \in [n]$: \begin{compactitem}
+\item let $J_k \in \posforms(Q,\emptyset)$ be the formula obtained
+  from $I_k$ by\begin{inparaenum}[(i)]
+    \item replacing each time-stamped predicate symbol $\stamp{q}{k}$ by $q$, and
+      %
+    \item existentially quantifying each variable $y \in
+      \stamp{X}{\leq k} \cup \set{x_{i_1},\ldots,x_{i_k}}$.
+  \end{inparaenum}
+%
+\item define the new labeling as $U'(\alpha_k) = U(\alpha_k) \wedge J_k$. 
+\end{compactitem}
+To see that the new set of formulae $U'(\alpha_0), \ldots,
+U'(\alpha_n)$ complies with Definition \ref{def:unfolding}, it is
+enough to notice the following facts: \begin{compactitem}
+\item $U'(\alpha_0) \equiv \iota$, because\footnote{By point
+  \ref{it2:generalized-lyndon-interpolant} of Definition
+  \ref{def:generalized-lyndon-interpolant}.} $\stamp{\iota}{0} \models
+  I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) = U(\alpha_0)
+  \wedge J_0 \equiv \iota \wedge J_0$.
+%
+\item 
+\end{compactitem}
 

### 2018-10-05 quantifiers/draft.tex
ok

@@
   \cover \beta$, if and only if there exists a node $\alpha' \prefix
-  \alpha$ such that, for all words $w \in \Sigma[X]^*$ and all cubes
-  $c \in \cube{\sem{U(\alpha')}}$, if $\A$ accepts $w$ started in $c$
-  then there exists a cube $d \in \cube{\sem{U(\beta)}}$ such that
-  $\A$ accepts $w$ started in $d$. Moreover, $U$ is \emph{closed} if
-  and only if every leaf from $\dom(U)$ is covered by an uncovered
-  node.
+  \alpha$ such that $U(\alpha') \models U(\beta)$. Moreover, $U$ is
+  \emph{closed} if and only if every leaf from $\dom(U)$ is covered by
+  an uncovered node.
 \end{definition}
+
+%% , for all words $w \in \Sigma[X]^*$ and all cubes $c \in
+%% \cube{\sem{U(\alpha')}}$, if $\A$ accepts $w$ started in $c$ then
+%% there exists a cube $d \in \cube{\sem{U(\beta)}}$ such that $\A$
+%% accepts $w$ started in $d$.
 
@@
 relation has to be recomputed after each refinement of the
-labeling. For a detailed presentation of the emptiness procedure, we
-refer the reader to \cite{IosifXu18} (or Appendix
+labeling. The semi-algorithm stops when (and if) a safe complete
+unfolding has been found. For a detailed presentation of the emptiness
+procedure, we refer the reader to \cite{IosifXu18} (or Appendix
 \ref{app:pseudocode}).
-
-The semi-algorithm stops when (and if) a safe complete unfolding has
-been found. The following theorem proves the soundness of this
-termination argument:
 
@@
 \end{theorem}
-\proof{ 
-  Let $U$ be a safe and complete unfolding of $\A$, such that
+\proof{ Let $U$ be a safe and complete unfolding of $\A$, such that
   $\dom(U) \neq \emptyset$. Suppose, by contradiction, that there
   exists a word $w \in \lang{A}$ and let $\alpha \isdef
-  \event{w}$.
-
-  Since $w \in \lang{A}$, by Lemma \ref{lemma:acceptance}, there
-  exists an interpretation $\I$ such that $\I,\data{w} \models
+  \event{w}$. Since $w \in \lang{A}$, by Lemma \ref{lemma:acceptance},
+  there exists an interpretation $\I$ such that $\I,\data{w} \models
   \accform{\alpha}$. Assume first that $\alpha \in \dom(U)$.  In this
@@
   and let $\alpha_1 \prefix \alpha' \prefix \alpha$ be a node such
-  that, for some uncovered node $\beta_1 \in \dom(U)$, for every word
-  $w'$ and every cube $c \in \cube{\sem{U(\alpha_1)}}$ such that $\A$
-  accepts $w'$ starting with $c$, there exists a cube $d \in
-  \cube{\sem{U(\beta_1)}}$ such that $\A$ accepts $w'$ starting with
-  $d$. Let $\gamma_1$ be the unique sequence such that
-  $\alpha_1\gamma_1 = \alpha$. By Definition \ref{def:coverage}, since
-  $\alpha_1 \cover \beta_1$ and $\event{w} = \alpha_1\gamma_1 \in
-  \lang{\A}$, there exists a word $w_1$ and a cube $c_1 \in
-  \cube{\sem{U(\beta_1)}}$ such that $\event{w_1} = \gamma_1$ and $\A$
-  accepts $w_1$ starting with $c_1$. If $\beta_1\gamma_1 \in \dom(U)$,
-  we obtain a contradiction by a similar argument as above. Hence
-  $\beta_1\gamma_1 \not\in \dom(U)$ and there exists a leaf of
-  $\dom(U)$ which is also a prefix of $\beta_1\gamma_1$. Since $U$ is
-  closed, this leaf is covered by an uncovered node $\beta_2 \in
-  \dom(U)$ and let $\alpha_2 \in \dom(U)$ be the minimal (in the
-  prefix partial order) node such that $\beta_1 \prefix \alpha_2
-  \prefix \beta_1\gamma_1$ and $\alpha_2 \cover \beta_2$. Let
-  $\gamma_2$ be the unique sequence such that $\alpha_2\gamma_2 =
-  \beta_1\gamma_1$. Since $\beta_1$ is uncovered, we have $\beta_1
-  \neq \alpha_2$ and thus $\len{\gamma_1} > \len{\gamma_2}$. By
-  repeating the above reasoning for $\alpha_2$, $\beta_2$ and
-  $\gamma_2$, we obtain an infinite sequence $\len{\gamma_1} >
-  \len{\gamma_2} > \ldots$, which is again a contradiction. \qed}
+  that $U(\alpha_1) \models U(\beta_1)$, for some uncovered node
+  $\beta_1 \in \dom(U)$.
+  %% for every word $w'$ and every cube $c \in
+  %% \cube{\sem{U(\alpha_1)}}$ such that $\A$ accepts $w'$ starting
+  %% with $c$, there exists a cube $d \in \cube{\sem{U(\beta_1)}}$
+  %% such that $\A$ accepts $w'$ starting with $d$.
+  Let $\gamma_1$ be the unique sequence such that $\alpha_1\gamma_1 =
+  \alpha$. By Definition \ref{def:coverage}, since $\alpha_1 \cover
+  \beta_1$ and $\event{w} = \alpha_1\gamma_1 \in \lang{\A}$, there
+  exists a word $w_1$ and a cube $c_1 \in \cube{\sem{U(\alpha_1)}}
+  \subseteq \cube{\sem{U(\beta_1)}}$, such that $\event{w_1} =
+  \gamma_1$ and $\A$ accepts $w_1$ starting with $c_1$. If
+  $\beta_1\gamma_1 \in \dom(U)$, we obtain a contradiction by a
+  similar argument as above. Hence $\beta_1\gamma_1 \not\in \dom(U)$
+  and there exists a leaf of $\dom(U)$ which is also a prefix of
+  $\beta_1\gamma_1$. Since $U$ is closed, this leaf is covered by an
+  uncovered node $\beta_2 \in \dom(U)$ and let $\alpha_2 \in \dom(U)$
+  be the minimal (in the prefix partial order) node such that $\beta_1
+  \prefix \alpha_2 \prefix \beta_1\gamma_1$ and $\alpha_2 \cover
+  \beta_2$. Let $\gamma_2$ be the unique sequence such that
+  $\alpha_2\gamma_2 = \beta_1\gamma_1$. Since $\beta_1$ is uncovered,
+  we have $\beta_1 \neq \alpha_2$ and thus $\len{\gamma_1} >
+  \len{\gamma_2}$. By repeating the above reasoning for $\alpha_2$,
+  $\beta_2$ and $\gamma_2$, we obtain an infinite sequence
+  $\len{\gamma_1} > \len{\gamma_2} > \ldots$, which is again a
+  contradiction. \qed}
 
@@
       \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-      \stamp{q}{k}(\vec{y}) \rightarrow \stamp{\psi}{k+1}\Big) \models
+      \stamp{q}{k-1}(\vec{y}) \rightarrow \stamp{\psi}{k}\Big) \models
       I_{k-1} \wedge \phi_k \models I_k$, and 
@@
 \end{compactitem}
-To see that the new set of formulae $U'(\alpha_0), \ldots,
-U'(\alpha_n)$ complies with Definition \ref{def:unfolding}, it is
-enough to notice the following facts: \begin{compactitem}
+
+Observe that, by Lemma \ref{lemma:subst-quant}
+(\ref{it2:subst-quant}), the set of free variables of a GLI formula
+$I_k$ consists of\begin{inparaenum}[(i)]
+\item variables $\stamp{X}{\leq k}$ keeping track of data
+values seen in the input at some earlier moment in time, or 
+%
+\item variables $\set{x_{i_1},\ldots,x_{i_k}}$ that track past choices
+  made within the transition rules. 
+\end{inparaenum}
+From the present point of view, it is not important when exactly in
+the past a certain input has been read or when a choice has been made,
+but rather the (symbolic) value and its associated data
+constraints. Intuitively, existential quantification of these
+variables does the job of ignoring the particular moments in the past
+when they have been encountered. Formally, to see that the new set of
+formulae $U'(\alpha_0), \ldots, U'(\alpha_n)$ complies with Definition
+\ref{def:unfolding}, it is enough to notice the following
+facts: \begin{compactitem}
 \item $U'(\alpha_0) \equiv \iota$, because\footnote{By point
@@
 %
-\item 
+\item by Definition \ref{def:generalized-lyndon-interpolant}
+  (\ref{it3:generalized-lyndon-interpolant}) we have, for all $k \in
+  [n-1]$:
+  \[I_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
+  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
+  \leftarrow \stamp{\psi}{k+1} \models I_{k+1}\] By slightly abusing
+  notation, let $\stamp{I}{n}_\ell$ be the formula in which every
+  predicate symbol $\stamp{q}{\ell}$ has been replaced with
+  $\stamp{q}{n}$. Then the following entailment holds: 
+  \[\stamp{I}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
+  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
+  \leftarrow \stamp{\psi}{1} \models \stamp{I}{1}_{k+1}\]
+  Because $J_k$ is obtained by existentially quantifying the free variables of $I_k$, we also obtain: 
+  \[\stamp{J}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
+  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
+  \leftarrow \stamp{\psi}{1} \models \stamp{J}{1}_{k+1}\] Since $U$
+  satisfies the labeling condition of Definition \ref{def:unfolding}
+  (\ref{it3:unfolding}) and $U'(\alpha_k) = U(\alpha_k) \wedge J_k$,
+  we obtain, as required: 
+  \[\stamp{U'(\alpha_k)}{0} \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
+  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
+  \leftarrow \stamp{\psi}{1} \models
+  \stamp{U'(\alpha_{k+1})}{1}\enspace.\]
 \end{compactitem}
 
-%% \subsection{Coverage}
-
-%% Assuming that the data domain is countable, the set of finite words
-%% $\Sigma[X]^*$ is countable and the emptiness problem is recursively
-%% enumerable. Moreover, by Lemma \ref{lemma:acceptance}, a systematic
-%% enumeration of sequences of input events $\alpha \in \Sigma^*$ will
-%% eventually discover an accepted word, provided that all path formulae
-%% $\pathform{\alpha}$ belong to a decidable first-order theory.
-
-%% However, if the automaton is empty, the enumeration of input event
-%% sequences will fail to terminate, in general. The classical way to
-%% fight this divergence problem is to consider a \emph{coverage} partial
-%% order, which is a subset of the following relation:
-
-%% \begin{definition}\label{def:coverage}
-%%   Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ and
-%%   $S_1, S_2 \subseteq \bigcup_{q \in Q} \set{q(d_1,\ldots,d_{\#(q)})
-%%     \mid d_1,\ldots,d_{\#(q)} \in \Data}$, we have $S_1 \sqsubseteq
-%%   S_2$ if and only if 
-%%   \begin{equation}\label{eq:po}
-%%     \text{
-%%       \begin{minipage}{0.7\textwidth}
-%%         $\A$ accepts a word $w$ when started in a set of cubes $C_1
-%%         \in S_1$ only if $\A$ accepts $w$ when started in a set of
-%%         cubes $C_2 \in S_2$.
-%%     \end{minipage}} 
-%%   \end{equation}
-%% \end{definition}
-%% Since, in general the $\sqsubseteq$ relation is not definable using
-%% available decidable (SMT) theories, we consider the following stronger
-%% relation: $S_1 \supsetforex S_2$ if and only if, for each set of cubes
-%% $C_1 \in S_1$, there exists a set of cubes $C_2 \in S_2$ such that
-%% $C_1 \supseteq C_2$.
-
-%% \begin{lemma}\label{lemma:coverage}
-%%   Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
-%%   sets of cubes $S_1, S_2 \subseteq \bigcup_{q \in Q}
-%%   \set{q(d_1,\ldots,d_{\#(q)}) \mid d_1,\ldots,d_{\#(q)} \in \Data}$
-%%   such that $S_1 \supsetforex S_2$, we have $S_1 \sqsubseteq S_2$.
-%% \end{lemma}
-%% \proof{  
-%% \qed}
+The last ingredient of the lazy annotation semi-algorithm based on
+unfolding refinement with coverage consist in the implementation of
+the coverage check, when the unfolding of an automaton is labeled with
+conjunctions of existentially quantified formulae with predicate
+symbols, obtained from interpolation. By Definition
+\ref{def:coverage}, checking whether a given node $\alpha \in \dom(U)$
+is covered amounts to finding a prefix $\alpha' \prefix \alpha$ and a
+node $\beta \in \dom(U)$ such that $U(\alpha') \models U(\beta)$, or
+equivalently, the formula $U(\alpha') \wedge \neg U(\beta)$ is
+unsatisfiable. However, the latter formula, in prenex form, has
+quantifier prefix in the language $\exists^*\forall^*$ and, as
+previously mentioned, the satisfiability problem for such formulae
+becomes undecidable when the data theory subsumes Presburger
+arithmetic \cite{Halpern91}.
+
+Nevertheless, if we require just a yes/no answer (i.e.\ not an
+interpolant) recently developped quantifier instantiation heuristics,
+such as \cite{ReynoldsKK17}, perform rather well in answering a large
+number of queries in this class. Observe, moreover, that coverage does
+not need to rely on a complete decision procedure. If the prover fails
+in answering the above satisfiability query, then the semi-algorithm
+assumes the node is not covered and continues exploring its
+successors. Failure to compute complete coverage may lead to
+divergence (non-termination) and failure to prove emptiness, but the
+soundness of the procedure is preserved.
 

### 2018-10-05 quantifiers/draft.tex
ok

@@
 Nevertheless, if we require just a yes/no answer (i.e.\ not an
-interpolant) recently developped quantifier instantiation heuristics,
-such as \cite{ReynoldsKK17}, perform rather well in answering a large
-number of queries in this class. Observe, moreover, that coverage does
-not need to rely on a complete decision procedure. If the prover fails
-in answering the above satisfiability query, then the semi-algorithm
+interpolant) recently developped quantifier instantiation heuristics
+\cite{ReynoldsKK17} perform rather well in answering a large number of
+queries in this class. Observe, moreover, that coverage does not need
+to rely on a complete decision procedure. If the prover fails in
+answering the above satisfiability query, then the semi-algorithm
 assumes the node is not covered and continues exploring its
@@
 divergence (non-termination) and failure to prove emptiness, but the
-soundness of the procedure is preserved.
+soundness of the procedure is not affected.
 

### 2018-10-12 quantifiers/draft.tex
ok

@@
 Let $\Sigma$ be a finite alphabet $\Sigma$ of \emph{input
-  events}. Given a finite set of variables $X \subseteq \vars$, we denote by $X
-\mapsto \Data$ the set of valuations of the variables $X$ and
-$\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly infinite
-set of \emph{data symbols} $(a,\nu)$, where $a$ is an input symbol and
-$\nu$ is a valuation. A \emph{data word} (simply called word in the
-following) is a finite sequence $w=(a_1,\nu_1)(a_2,\nu_2) \ldots
-(a_n,\nu_n)$ of data symbols. Given a word $w$, we denote by
+  events}. Given a finite set of variables $X \subseteq \vars$, we
+denote by $X \mapsto \Data$ the set of valuations of the variables $X$
+and $\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly
+infinite set of \emph{data symbols} $(a,\nu)$, where $a$ is an input
+symbol and $\nu$ is a valuation. A \emph{data word} (simply called
+word in the following) is a finite sequence $w=(a_1,\nu_1)(a_2,\nu_2)
+\ldots (a_n,\nu_n)$ of data symbols. Given a word $w$, we denote by
 $\event{w} \isdef a_1\ldots a_n$ its sequence of input events and by
@@
 $\Sigma^*$ the set of finite sequences of input events and by
-$\Sigma[X]^*$ the set of data words over the variables $X$.
-
-A \emph{first-order alternating automaton} (automaton) is a tuple $\A
-= \tuple{\Sigma,X,Q,\iota,F,\Delta}$, where: \begin{compactitem}
-\item $\Sigma$ is the finite set of input events, 
+$\Sigma[X]^*$ the set of data words over the variables $X$. A
+\emph{first-order alternating automaton} is a tuple $\A =
+\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where: \begin{compactitem}
+\item $\Sigma$ is a finite set of input events, 
 \item $X$ is a finite set of input variables, 
@@
 other than predicate symbols, such that the labeling is compatible
-with the transition relation of the automaton (point
-\ref{it3:unfolding} of Definition \ref{def:unfolding}).
-
-Suppose that a given input event sequence $\alpha \in \dom(U)$ is
-labeled with a sequence of formulae $U(\alpha_0) = \iota, U(\alpha_1),
-\ldots U(\alpha_n) = U(\alpha)$ in a way that fulfills the requirement
-from Definition \ref{def:unfolding} (\ref{it3:unfolding}) and,
-moreover, it has a non-local GLI $(I_0, \ldots, I_n)$, such that
-$\voc{I_k} \subseteq \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup
+with the transition relation of the automaton, according to the point
+(\ref{it3:unfolding}) of Definition \ref{def:unfolding}. Suppose that
+a given input event sequence $\alpha \in \dom(U)$ is labeled with a
+sequence of formulae $U(\alpha_0) = \iota, U(\alpha_1), \ldots
+U(\alpha_n) = U(\alpha)$ in a way that fulfills the requirement from
+Definition \ref{def:unfolding} (\ref{it3:unfolding}) and, moreover, it
+has a non-local GLI $(I_0, \ldots, I_n)$, such that $\voc{I_k}
+\subseteq \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup
 \set{x_{i_1},\ldots,x_{i_k}}$, where $x_{i_1},\ldots,x_{i_n}$ are the
 renamed variables occurring in the rules of $\Delta$ under the scope
-of existential transition quantifiers (formally these are the
+of existential transition quantifiers\footnote{These are the
   existentially quantified variables from the prenex normal form of
-$\quantform{\alpha_k}$). The interpolant is used in refining the
-labeling of the sequence as follows, for all $k \in [n]$: \begin{compactitem}
+  $\quantform{\alpha_k}$.}. The interpolant is used to refine the
+labeling of the sequence as follows, for all $k \in
+[n]$: \begin{compactitem}
 \item let $J_k \in \posforms(Q,\emptyset)$ be the formula obtained
-  from $I_k$ by\begin{inparaenum}[(i)]
+  from $I_k$ by \begin{inparaenum}[(i)]
     \item replacing each time-stamped predicate symbol $\stamp{q}{k}$ by $q$, and
@@
     \item existentially quantifying each variable $y \in
-      \stamp{X}{\leq k} \cup \set{x_{i_1},\ldots,x_{i_k}}$.
+      \stamp{X}{\leq k} \cup \set{x_{i_1},\ldots,x_{i_k}}$,
   \end{inparaenum}
@@
 (\ref{it2:subst-quant}), the set of free variables of a GLI formula
-$I_k$ consists of\begin{inparaenum}[(i)]
+$I_k$ consists of \begin{inparaenum}[(i)]
 \item variables $\stamp{X}{\leq k}$ keeping track of data
@@
 \end{inparaenum}
-From the present point of view, it is not important when exactly in
-the past a certain input has been read or when a choice has been made,
-but rather the (symbolic) value and its associated data
-constraints. Intuitively, existential quantification of these
-variables does the job of ignoring the particular moments in the past
-when they have been encountered. Formally, to see that the new set of
+Basically, it is not important when exactly in the past a certain
+input has been read or when a choice has been made, as only the value
+of the variable determines the future behavior. Intuitively,
+existential quantification of these variables does the job of ignoring
+when in the past these values have been seen. Formally, the new set of
 formulae $U'(\alpha_0), \ldots, U'(\alpha_n)$ complies with Definition
-\ref{def:unfolding}, it is enough to notice the following
-facts: \begin{compactitem}
-\item $U'(\alpha_0) \equiv \iota$, because\footnote{By point
+\ref{def:unfolding}, because: \begin{compactitem}
+\item $U'(\alpha_0) \equiv \iota$, since, by point
   \ref{it2:generalized-lyndon-interpolant} of Definition
-  \ref{def:generalized-lyndon-interpolant}.} $\stamp{\iota}{0} \models
-  I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) = U(\alpha_0)
-  \wedge J_0 \equiv \iota \wedge J_0$.
+  \ref{def:generalized-lyndon-interpolant}, we have $\stamp{\iota}{0}
+  \models I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) =
+  U(\alpha_0) \wedge J_0 \equiv \iota \wedge J_0 \equiv \iota$, and
 %
@@
   \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
-  \leftarrow \stamp{\psi}{k+1} \models I_{k+1}\] By slightly abusing
-  notation, let $\stamp{I}{n}_\ell$ be the formula in which every
-  predicate symbol $\stamp{q}{\ell}$ has been replaced with
-  $\stamp{q}{n}$. Then the following entailment holds: 
+  \rightarrow \stamp{\psi}{k+1} \models I_{k+1}\] Then the following
+  entailment holds:
   \[\stamp{I}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
   \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
-  \leftarrow \stamp{\psi}{1} \models \stamp{I}{1}_{k+1}\]
+  \rightarrow \stamp{\psi}{1} \models \stamp{I}{1}_{k+1}\]
   Because $J_k$ is obtained by existentially quantifying the free variables of $I_k$, we also obtain: 
@@
   \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
-  \leftarrow \stamp{\psi}{1} \models \stamp{J}{1}_{k+1}\] Since $U$
+  \rightarrow \stamp{\psi}{1} \models \stamp{J}{1}_{k+1}\] Since $U$
   satisfies the labeling condition of Definition \ref{def:unfolding}
@@
   \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
-  \leftarrow \stamp{\psi}{1} \models
+  \rightarrow \stamp{\psi}{1} \models
   \stamp{U'(\alpha_{k+1})}{1}\enspace.\]
@@
 successors. Failure to compute complete coverage may lead to
-divergence (non-termination) and failure to prove emptiness, but the
-soundness of the procedure is not affected.
+divergence (non-termination) and failure to prove emptiness, but does
+not affect the soundness of the semi-algorithm (real counterexamples
+will still be found).
+
+\section{Applications}
+\label{sec:applications}
+
+The main application of first-order alternating automata is checking
+inclusion between various classes of automata extended with variables
+ranging over infinite domains that recognize languages over infinite
+alphabets. The most widely known such classes are \emph{timed
+  automata} \cite{AlurDill94} and \emph{finite-memory (register)
+  automata} \cite{KaminskiFrancez94}. In both cases, complementation
+is not possible inside the class and inclusion is undecidable. Our
+contribution is providing a systematic semi-algorithm for these
+decision problems. In addition, we can extend our previous
+\emph{generic register automata} \cite{HolikIosifRogalewiczVojnar18}
+inclusion checking framework, by allowing monitor (right-hand side)
+automata to have local variables, that are not visible in the
+language\footnote{We conjecture that hidden computations using
+  invisible variables (clocks) are the reason for which timed and
+  register automata cannot be complemented within their class.}.
+
+Another application is checking safety (mutual exclusion, absence of
+deadlocks, etc.) and liveness (termination, lack of starvation, etc.)
+properties of parameterized concurrent programs, consisting of an
+unbounded number of replicated threads that communicate via a fixed
+set of global variables (locks, counters, etc.). The verification of
+parametric programs has been reduced to checking the emptiness of a
+(possibly infinite) sequence of first-order alternating automata,
+called \emph{predicate automata} \cite{Farzan15,Farzan16}, encoding
+the inclusion of the set of traces of a parametric concurrent program
+into increasingly general \emph{proof spaces}, obtained by
+generalization of counterexamples. The program and the proof spaces
+are first-order alternating automata over the infinite alphabet of
+pairs consisting of program statements and thread identifiers.
+
+\subsection{Timed Automata}
+\label{sec:ta}
+
+The standard definition of a finite \emph{timed word} is a sequence of
+pairs $(a_1,\tau_1), \ldots, (a_n,\tau_n) \in (\Sigma \times
+\real)^*$, where $\real$ is the set of real numbers, such that $0 \leq
+\tau_i < \tau_{i+1}$, for all $i \in [1,n-1]$. Intuitively, $\tau_i$
+is the moment in time where the input event $a_i$ occurs. Given a set
+$C$ of \emph{clocks}, the set $\Phi(C)$ of \emph{clock constraints} is
+defined inductively as the set of formulae $x \leq c$, $x \geq c$,
+$\neg\delta$, $\delta_1 \wedge \delta_2$, where $x \in C$, $c \in
+\rat$ is a rational constant and $\delta,\delta_1,\delta_2 \in
+\Phi(X)$.
+
+A \emph{timed automaton} is a tuple $T = \tuple{\Sigma,S,S_0,F,C,E}$,
+where: $\Sigma$ is a finite set of input events, $S$ is a finite set
+of states, $S_0, F \subseteq S$ are sets of initial and final states,
+respectively, $C$ is a finite set of clocks and $E \subseteq S \times
+\Sigma \times S \times 2^C \times \Phi(C)$ is the set of transitions
+$(s,a,s',\lambda,\delta)$ from state $s$ to state $s'$ with symbol
+$a$, $\lambda$ is the set of clocks to be reset and $\delta$ is a
+clock constraint. A run of $T$ over a timed word $w = (a_1,\tau_1)
+\ldots (a_n,\tau_n)$ is a sequence $(s_0,\gamma_0) \ldots
+(s_n,\gamma_n)$, where $s_i \in S$, $\gamma_i : C \rightarrow \real$
+are clock valuations, for all $i \in [n]$, and: \begin{compactitem}
+\item $s_0 \in S_0$ and $\gamma_0(x)$ for all $x \in C$, 
+%
+\item for all $i \in [n]$, there exists a transition
+  $(s_i,a_i,s_{i+1},\lambda_i,\delta_i) \in E$ such that $\gamma_i +
+  \tau_{i+1} - \tau_i \models \delta_i$, and for all $x \in C$,
+  $\gamma_{i+1}(x) = 0$ if $x \in \lambda_i$ and $\gamma_{i+1}(x) = \gamma_i(x)
+  + \tau_{i+1}-\tau_i$, otherwise. Here $\tau_0 \isdef 0$ and $\gamma_i +
+  \tau_{i+1} - \tau_i$ is the valuation mapping each $x \in C$ to
+  $\gamma_i(x) + \tau_{i+1} - \tau_i$.
+\end{compactitem}
+The run is accepting iff $s_n \in F$, in which case $T$ accepts
+$w$. As usual, we denote by $\lang{T}$ the set of finite words
+accepted by $T$. It is well-known \cite{AlurDill94} that, in general,
+there is no timed automaton accepting the complement language
+$(\Sigma\times\real)^* \setminus \lang{T}$ and, moreover, the language
+inclusion problem is undecidable.
+
+Given a timed automaton $T = \tuple{\Sigma,S,S_0,F,C,E}$, we define a
+first-order alternating automaton $\A_T =
+\tuple{\Sigma,\set{t},Q_T,\iota_T,F_T,\Delta_T}$, with a single input
+variable $t$, ranging over $\real$, such that each timed word $w =
+(a_1,\tau_1) \ldots (a_n,\tau_n)$ corresponds to a unique data word
+$d(w) = (a_1,\nu_1) \ldots (a_n,\nu_n)$ such that $\nu_i(t) = \tau_i$
+for all $i \in [1,n]$ and $\lang{\A_T} = \set{d(w) \mid
+  w\in\lang{T}}$. The only difficulty here is capturing the fact that
+all the clocks of $T$ evolve at the same pace, which is easily done
+using a technique from \cite{Fribourg98}, that consists in replacing
+each clock $x_i$ of $T$ by a variable $y_i$ who tracks the difference
+between the values of $t$ and $x_i$, instead of the value of $x_i$
+directly.
+
+Formally, if $C = \set{x_1,\ldots,x_k}$ and $S =
+\set{s_1,\ldots,s_m}$, we define $Q_T \isdef \set{q_1, \ldots, q_m}$ ,
+where $\#(q_i) = k+1$ for all $i \in [1,m]$, $\iota_T \isdef
+\bigvee_{s_i \in S_0} q_i(0,\ldots,0)$, $F_T \isdef \set{q_i \mid s_i
+  \in F}$ and, for each transition $(s_i,a,s_j,\lambda,\delta) \in E$,
+$\Delta_T$ contains the rule:
+\[\begin{array}{rcl}
+q_i(y_1,\ldots,y_k,z) \arrow{a(t)}{} \exists y'_1 \ldots \exists y'_k
+~.~ t > z \wedge \delta(z-y_1,\ldots,z-y_k) & \wedge & 
+ \bigwedge_{x_i \in \lambda} y'_i = z \wedge \bigwedge_{x_i \not\in \lambda} y'_i = y_i \\ 
+& \wedge & q_j(y'_1,\ldots,y'_k,t)
+\end{array}\]
+Moreover, nothing else is in $\Delta_T$. We establish the following
+connection between a timed automaton and its corresponding first-order
+alternating automaton.
+
+\begin{proposition}\label{prop:timed-alternating}
+  Given a timed automaton $T = \tuple{\Sigma,S,S_0,F,C,E}$, the
+  first-order alternating automaton $\A_T =
+  \tuple{\Sigma,\set{t},Q_T,\iota_T,F_T,\Delta_T}$ recognizes the
+  language $\lang{\A_T} = \set{d(w) \mid w \in \lang{T}}$. 
+\end{proposition}
+\proof{ ``$\subseteq$'' Let $w = (a_1,\nu_1) \ldots (a_n,\nu_n) \in
+  \lang{\A_T}$ be a data word. We show the existence of a timed word
+  $(a_1,\tau_1) \ldots (a_n,\tau_n) \in \lang{T}$ such that
+  $\nu_i(t)=\tau_i$, for all $i \in [1,n]$, by induction on
+  $n\geq0$. In fact we shall prove the following stronger
+  statements: \begin{compactenum}
+  \item each execution of $\A_T$ over $w$ starting with a cube $c \in
+    \cube{\minsem{\iota_T}}$ is a linear tree, in which each node has
+    at most one child.
+    %
+  \item for each execution $q_{i_0}(d^0_1,\ldots,d^0_k,\tau_0) \ldots
+    q_{i_n}(d^n_1,\ldots,d^n_k,\tau_n)$ of $\A_T$, $T$ has an
+    execution $(s_{i_0}, \gamma_0) \ldots (s_{i_n}, \gamma_n)$ over
+    the timed word $(a_1, \tau_1) \ldots (a_n, \tau_n)$, such that,
+    for all $i \in [1,n]$ and all $\ell \in [1,k]$, we have
+    $\gamma_i(x_\ell) = \tau_{i-1} - d^i_\ell$.
+  \end{compactenum}
+  The first point above is by inspection of $\iota_T = \bigvee_{s_i
+    \in S_0} q_i(0,\ldots,0)$ and of the rules from
+  $\Delta_T$. Indeed, each minimal model of $\iota_T$ corresponds to a
+  cube $q(0,\ldots,0)$ and each rule has exactly one predicate atom on
+  its right-hand side, thus each node of the execution will have at
+  most one successor. The second point is by induction on $n\geq0$.
+
+  ``$\supseteq$'' Let $w = (a_1,\tau_1) \ldots (a_n,\tau_n) \in
+  \lang{T}$ be a timed word. By induction on $n\geq0$, we show that
+  for each run $(s_{i_0}, \gamma_0) \ldots (s_{i_n}, \gamma_n)$ of $T$
+  over $w$, $\A_T$ has a linear execution
+  $q_{i_0}(d^0_1,\ldots,d^0_k,\tau_0) \ldots
+  q_{i_n}(d^n_1,\ldots,d^n_k,\tau_n)$ such that, for all $i \in [1,n]$
+  and all $\ell \in [1,k]$, we have $\gamma_i(x_\ell) = \tau_{i-1} -
+  d^i_\ell$. \qed}
+
+An easy consequence is that the timed language inclusion problem
+``given timed automata $T_1$ and $T_2$, does $\lang{T_1} \subseteq
+\lang{T_2}$ ?'' is reduced in polynomial time to the emptiness problem
+$\lang{\A_{T_1}} \cap \lang{\overline{\A_{T_2}}} = \emptyset$, for
+which (\S\ref{sec:emptiness}) provides a semi-algorithm. Observe
+further that the transition quantifiers from the encoding of timed
+automata into first-order alternating automata can be easily
+eliminated by substitution as in $\exists x ~.~ x = t \wedge \phi
+\equiv \phi[t/x]$.
+
+\subsection{Register Automata}
+\label{sec:ra}
+
+Finite-memory automata, most commonly referred to as register automata
+\cite{KaminskiFrancez94} are among the first attempts at lifting the
+finite alphabet restriction of classical Rabin-Scott automata. In a
+nutshell, a register automaton is a finite-state automaton equipped
+with a finite set of registers $x_1,\ldots,x_r$ able to copy input
+values and compare them with later the input. Consequently, basic
+results from classical automata theory, such as the pumping lemma or
+the closure under complement do not hold in this model and inclusion
+of languages recognized by register automata is undecidable
+\cite{NevenSchwentickVianu03}.
+
+Let $\Sigma$ be an infinite alphabet, $\#$ be a symbol not in $\Sigma$
+and $r>0$ be an integer constant. An \emph{assignment} is a word
+$\vec{v} = v_1 \ldots v_r$ such that if $v_i = v_j$ and $i \neq j$
+then $v_i = \#$, for all $i,j \in [1,r]$. We write $[\vec{v}]$ for the
+set $\set{v_i \mid i \in [1,r]}$. A \emph{finite-memory (register)
+  automaton} is a tuple $R = \tuple{S,q_0,\vec{u}, \rho, \mu, F}$,
+where $S$ is a finite set of states, $q_0 \in S$ is the initial state,
+$\vec{u} = u_1 \ldots u_r$ is the initial assignment, $\rho : S
+\rightarrow [1,r]$ is the reassignment partial function, $\mu
+\subseteq S \times [1,r] \times S$ is the transition relation and $F
+\subseteq S$ is the set of final states. A run of $A$ over an input
+word $a_1 \ldots a_n \in \Sigma^*$ is a sequence $(s_0,\vec{v}_0)
+\ldots (s_n,\vec{v}_n)$ such that $\vec{v}_0 = \vec{u}$ and, for all
+$i \in [1,n]$, exactly one of the following holds: \begin{compactitem}
+\item if there exists $k\in[1,r]$ such that $a_i=(\vec{v}_{i-1})_k$
+  then $\vec{v}_i = \vec{v}_{i-1}$ and $(s_{i-1},k,s_i) \in \mu$, 
+%
+\item otherwise $a_i\not\in[\vec{v}_{i-1}]$, $\rho(s_{i-1})$ is
+  defined, $(\vec{v}_i)_{\rho(s_{i-1})} = a_i$, for each $k \in
+  [1,r] \setminus \set{\rho(s_{i-1})}$, we have $(\vec{v}_i)_k =
+  (\vec{v}_{i-1})_k$ and $(s_{i-1},\rho(s_{i-1}),s_i) \in \mu$.
+\end{compactitem}
+Intuitively, if the input symbol is already stored in some register,
+the automaton moves to the next state if, moreover, the transition
+relation allows it, otherwise it copies the input to the register
+indicated by the reassignment, erasing its the previous value, and
+moves according to the transition relation.
+
+The translation of register automata to first-order alternating
+automata is quite natural, because registers can be encoded as
+arguments of predicate atoms. Formally, given a register automaton $R
+= \tuple{S,s_0,\vec{u},\rho,\mu,F}$, such that $S =
+\set{s_0,\ldots,s_m}$, we define the alternating automaton $\A_R =
+\tuple{\set{\bullet},\set{x},Q_R,\iota_R,F_R,\Delta_R}$, where
+$\bullet \not\in \Sigma$, $Q_R \isdef \set{q_0,\ldots,q_m}$ and
+$\#(q_i) = r$ for all $i \in [m]$, $\iota_R \isdef q_0(\vec{u})$, $F_R
+\isdef \set{q_i \mid s_i \in F}$ and, for each transition $(s_i, k,
+s_j) \in \mu$, $\Delta_T$ contains the rule:
+\[q_i(y_1,\ldots,y_r) \arrow{a(x)}{} y_k=x \wedge q_j(y_1,\ldots,y_r) ~\vee~
+\bigwedge_{i=1}^r x \neq y_i \wedge q_j(y_1,\ldots,y_{k-1},x,y_{k+1},
+\ldots, y_r)\] Moreover, nothing else is in $\Delta_R$. The connection
+between register automata and first-order alternating automata is
+stated below.
+
+\begin{proposition}\label{prop:register-alternating}
+  Given a register automaton $R = \tuple{S,s_0,\vec{u},\rho,\mu,F}$
+  over an infinite alphabet $\Sigma$, the first-order alternating
+  automaton $\A_R = \tuple{\set{\bullet},Q_R,\iota_R,F_R,\Delta_R}$
+  recognizes the languge $\lang{\A_R} = \set{ (\bullet,a_1) \ldots
+    (\bullet,a_n) \mid a_1 \ldots a_n \in \lang{R} }$. 
+\end{proposition}
+\proof{ ``$\subseteq$'' Let $w = (\bullet,a_1) \ldots (\bullet,a_n)
+  \in \lang{\A_R}$. First, it is easy to show that each execution of
+  $\A_R$, that starts in some cube $c \in \cube{\minsem{\iota_R}}$, is
+  a linear tree with labels $q_0(\vec{v}_0), \ldots, q_n(\vec{v}_0)$
+  such that $\vec{v}_0 = \vec{u}$. Second by induction on $n\geq0$, we
+  prove that $\A_R$ has a run as above over $w$ only if $R$ has a run
+  $(q_0,\vec{v}_0), \ldots, (q_n,\vec{v}_n)$ over $a_1 \ldots
+  a_n$. ``$\supseteq$'' Let $w = a_1 \ldots a_n \in \lang{R}$ and
+  $q_0(\vec{v}_0), \ldots, q_n(\vec{v}_0)$ be a run of $R$ over $w$,
+  such that $\vec{v}_0 = \vec{u}$. By induction on $n\geq0$, we can
+  build an execution of $\A_R$ over $(\bullet,a_1) \ldots
+  (\bullet,a_n)$ that is a linear tree with labels $q_0(\vec{v}_0),
+  \ldots, q_n(\vec{v}_n)$. \qed}
+
+Consequently, the language inclusion problem ``given register automata
+$R_1$ and $R_2$, does $\lang{R_1} \subseteq \lang{R_2}$?'' is reduced
+in polynomial time to emptiness problem $\lang{\A_{R_1}} \cap
+\lang{\overline{\A_{R_2}}} = \emptyset$, for which
+(\S\ref{sec:emptiness}) provides a semi-algorithm. Notice further that
+the encoding of register automata uses no transition quantifiers.
+
+\subsection{Predicate Automata}
+\label{sec:pa}
+
+The model of \emph{predicate automata} \cite{Farzan15,Farzan16} has
+emerged recently as a tool for checking safety and liveness properties
+of parameterized concurrent programs, in which there is an unbounded
+number of replicated threads that communicate via global
+variables. Predicate automata recognize finite sequences of actions
+that are pairs $(\sigma,i)$, where $\sigma$ is from a finite set
+$\Sigma$ of program statements and $i \in \nat$ ranges over an
+unbounded set of thread identifiers. To avoid clutter, we shall view a
+pair $(\sigma,i)$ as a data symbol $(\sigma,\nu)$ where $\nu(x)=i$,
+for a designated input variable $x$.
+
+Since thread identifiers can only be compared for equality, the data
+theory of predicate automata is the first-order theory of
+equality. Moreover, transition quantifiers are only needed for
+checking termination and, generally, liveness properties
+\cite{Farzan16}. Thus, syntactically speaking, predicate automata are
+a subset of first-order alternating automata.
+
+However, the execution semantics of predicate automata differs from
+that of first-order automata with respect to the following detail:
+initial configurations and successors of predicate automata are
+defined using the entire sets of models of the initial sentence and
+transition rules, not just the minimal ones, as in our case. Formally,
+a run of a predicate automaton $P =
+\tuple{\Sigma,\set{x},Q,\iota,F,\Delta}$ over a word $(a_1,\nu_1)
+\ldots$ $(a_n,\nu_n)$ is a sequence of interpretations $\I_0, \ldots,
+\I_n$ such that $\I_0 \in \sem{\iota}$ and for each $i \in [1,n]$,
+each $q \in Q$ and each tuple $\tuple{d_1,\ldots,d_{\#(q)}} \in
+\I_{i-1}(q)$, we have $\I_i \in \sem{\psi}_\mu$, for each rule
+$q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(x)}{} \psi \in \Delta$, where
+$\mu=\nu_i[y_1\leftarrow d_1, \ldots, y_{\#(q)} \leftarrow
+  d_{\#(q)}]$. The run is accepting if and only if $\I(q)=\emptyset$
+for all $q \in Q \setminus F$.
+
+In fact, as shown next, this more simple execution semantics is
+equivalent, from the language point of view, with the semantics given
+by Definitions \ref{def:execution} and \ref{def:accepting}. However,
+the semantics of first-order alternating automata based on minimal
+models is important for its relation to the classical textbook
+semantics of boolean alternating automata
+\cite{ChandraKozenStockmeyer81}.
+
+\begin{proposition}
+  Given a predicate automaton $P =
+  \tuple{\Sigma,\set{x},Q,\iota,F,\Delta}$, let $\A_P$ be the
+  first-order alternating automaton that has the same description as
+  $P$. Then $\lang{P} = \lang{\A_P}$. 
+\end{proposition}
+\proof{
+
+\qed}
+
+As before, this result enables using the semi-algorithm of
+(\S\ref{sec:emptiness}) for checking emptiness of predicate automata.
+We point out that, although quantifier-free predicate automata with
+predicates of arity one are decidable for emptiness \cite{Farzan15},
+currently there is no method for checking emptiness of predicate
+automata with \begin{inparaenum}[(i)]
+\item predicates of arity greater than one, or 
+\item transition quantifiers. 
+\end{inparaenum}
 

### 2018-10-16 quantifiers/draft.tex
ok

@@
 $\Sigma^*$ the set of finite sequences of input events and by
-$\Sigma[X]^*$ the set of data words over the variables $X$. A
-\emph{first-order alternating automaton} is a tuple $\A =
-\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where: \begin{compactitem}
-\item $\Sigma$ is a finite set of input events, 
-\item $X$ is a finite set of input variables, 
-\item $Q$ is a finite set of predicates denoting control states, 
-\item $\iota \in \posforms(Q,\emptyset)$ is a sentence defining initial configurations,
-\item $F \subseteq Q$ is the set of predicates denoting final states, and 
-\item $\Delta$ is a set of \emph{transition rules} of the form
-  \(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is
-  a predicate, $a \in \Sigma$ is an input event and $\psi \in
-  \posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive
-  formula, where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$.
-\end{compactitem}
-The quantifiers occurring in the right-hand side formula of a
-transition rule are referred to as \emph{transition quantifiers}. The
-\emph{size} of $\A$ is defined as $\len{\A} = \len{\iota} +
+$\Sigma[X]^*$ the set of data words over the variables $X$.
+
+Formally, a \emph{first-order alternating automaton} is a tuple $\A =
+\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is a finite set of
+input events, $X$ is a finite set of input variables, $Q$ is a finite
+set of predicates denoting control states, $\iota \in
+\posforms(Q,\emptyset)$ is a sentence defining initial configurations,
+$F \subseteq Q$ is the set of predicates denoting final states, and
+$\Delta$ is a set of \emph{transition rules} of the form
+\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
+predicate, $a \in \Sigma$ is an input event and $\psi \in
+\posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive formula,
+where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$. The quantifiers
+occurring in the right-hand side formula of a transition rule are
+referred to as \emph{transition quantifiers}. The \emph{size} of $\A$
+is defined as $\len{\A} = \len{\iota} +
 \sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
@@
 for each predicate $q \in Q$ and each input event $a \in \Sigma$, at
-most one such rule --- two or more rules can be joined using
+most one such rule, as two or more rules can be joined using
 disjunction.
@@
 $\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
-is a model but is not minimal\footnote{To avoid the confusion of
-  considering potentially non-minimal models, the original definition
-  of alternating finite-state automata \cite{ChandraKozenStockmeyer81}
-  uses boolean valuations instead of formulae.}.
+is a model but is not minimal. The original definition of alternating
+finite-state automata \cite{ChandraKozenStockmeyer81} works around
+this problem by considering boolean valuations (models) instead of
+formulae.
 
@@
 Typically, when checking the unreachability of a set of program
-configurations or, more generally, the infeasability of a set of Horn
-clauses \cite{BjornerGurfinkelMcMillanRybalchenko15}, the interpolants
-used to annotate the unfolded control structure are assertions about
-the values of the program variables in a given control state, at a
-certain step of a given execution. However, in an alternating model of
-computation, it is useful to distinguish between
+configurations \cite{mcmillan06}, the interpolants used to annotate
+the unfolded control structure are assertions about the values of the
+program variables in a given control state, at a certain step of a
+given execution. However, in an alternating model of computation, it
+is useful to distinguish between
 \begin{inparaenum}[(i)]
@@
 the automaton executes as follows:
-\[\begin{array}{l}
-q_0(0) \arrow{(a,\nu_1)}{} q_1(k_1)
-\wedge q_2(-k_1) \ldots \arrow{(a,\nu_{n-1})}{} 
-q_1(\sum_{i=1}^{n-1} k_i) \wedge q_2(-\sum_{i=1}^{n-1} k_i) \arrow{(a,\nu_n)}{} \\
-\sum_{i=1}^{n} k_i > 0 \wedge \sum_{i=1}^{n} k_i < 0 \wedge q_f \equiv \bot
+\[\begin{array}{c}
+q_0(0) \arrow{(a,\nu_1)}{} \set{q_1(k_1),q_2(-k_1)} \ldots \arrow{(a,\nu_{n-1})}{} 
+\set{q_1(\sum_{i=1}^{n-1} k_i),q_2(-\sum_{i=1}^{n-1} k_i)}
+\arrow{(a,\nu_n)}{} \emptyset
 \end{array}\]
@@
 \end{definition}
-The existence of Lyndon interpolants for first-order logic, with no
-predefined interpretation of function symbols\footnote{A model
-  specifies the universe and the interpretation of function symbols.},
-was proved by a non-constructive model-theoretic argument
-\cite{Lyndon59}. In the following, we shall consider only those
-theories of the data domain $\Data$, for which there exists a decision
-procedure for satisfiability, able to produce Lyndon interpolants. For
-instance, such interpolating decision procedures exist for linear real
-and integer arithmetic, with uninterpreted function symbols (UFLRA and
-UFLIA) \cite{RybalchenkoSofronieStokkermans}. In the rest of this
-section, let us fix an automaton $\A =
+%% The existence of Lyndon interpolants for first-order logic, with no
+%% predefined interpretation of function symbols\footnote{A model
+%%   specifies the universe and the interpretation of function symbols.},
+%% was proved by a non-constructive model-theoretic argument
+%% \cite{Lyndon59}. In the following, we shall consider only those
+%% theories of the data domain $\Data$, for which there exists a decision
+%% procedure for satisfiability, able to produce Lyndon interpolants. For
+%% instance, such interpolating decision procedures exist for linear real
+%% and integer arithmetic, with uninterpreted function symbols (UFLRA and
+%% UFLIA).
+
+In the rest of this section, let us fix an automaton $\A =
 \tuple{\Sigma,X,Q,\iota,F,\Delta}$. Due to the above observation, none
@@
 procedure only for decidable theories, such as the quantifier-free
-linear (integer) arithmetic with uninterpreted functions (QFLIA,
-QFLRA, etc.).
+linear (integer) arithmetic with uninterpreted functions (UFLIA,
+UFLRA, etc.) \cite{RybalchenkoSofronieStokkermans}.
 
@@
 
-The following lemma gives the formal proof of correctness for this
-construction.
+A function $\xi : \nat \rightarrow \nat$ is [strictly]
+\emph{monotonic} iff for each $n < m$ we have $\xi(n) \leq \xi(m)$
+     [$\xi(n) < \xi(m)$] and \emph{finite-range} iff for each $n \in
+     \nat$ the set $\set{m \mid \xi(m)=n}$ is finite. If $\xi$ is
+     finite-range, we denote by $\xi^{-1}(n) \in \nat$ the maximal
+     value $m$ such that $\xi(m)=n$. It is easy to see that $\xi^{-1}$
+     is strictly monotonic if $\xi$ is monotonic and finite-range. The
+     lemma below gives the proof of correctness for the previous
+     construction of non-local GLI.
 
@@
   \item\label{it1:subst-quant} one can effectively build witness terms
-    $\tau_{i_1}, \ldots, \tau_{i_\ell}$, where $\set{i_1, \ldots,
-    i_\ell} = \set{j \in [1,m] \mid Q_j = \forall}$ and
-    $\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup \set{x_k
-    \mid k < i_j, Q_k = \exists}$, for all $j \in [1,\ell]$, such that
-    $\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
+    $\tau_{i_1}, \ldots, \tau_{i_\ell}$, where 
+    \[\begin{array}{l}   
+    \set{i_1, \ldots,i_\ell} = \set{j \in [1,m] \mid Q_j = \forall} \text{ and} \\
+    \voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup \set{x_k
+    \mid k < i_j, Q_k = \exists},~ \forall j \in [1,\ell]
+    \end{array}\]
+    such that $\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
       \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable, and
@@
   \item\label{it2:subst-quant} one can effectively build a GLI $(I_0,
-    \ldots, I_n)$ for $\alpha$, such that $\voc{I_k} \subseteq
+    \ldots, I_n)$ for $\alpha$, such that \[\voc{I_k} \subseteq
     \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
-      \max(\xi^{-1}(k)),~ Q_j = \exists}$, for all $k \in [n]$.
+      \xi^{-1}(k),~ Q_j = \exists},~ \forall k \in [n]\enspace.\]
   \end{compactenum}
@@
   $\tau_{i_1}, \ldots, \tau_{i_\ell}$, where
-  $\set{i_1,\ldots,i_\ell}=\set{i \in [1,m] \mid Q_i=\forall}$, such
-  that: \begin{compactitem}
+  $\set{i_1,\ldots,i_\ell}=\set{i \in [1,m] \mid Q_i=\forall}$
+  and: \begin{compactitem}
   \item $\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup
@@
   \item for all $i=1,\ldots,n$, let $\widehat{\Phi}_i$ be the matrix
-    of some prenex form of the formula:
+    of some prenex form of:
     \[\widehat{\Phi}_i \isdef \widehat{\Phi}_{i-1} \wedge
@@
   It is easy to see that $\widehat{\Phi}$ is the matrix of some prenex
-  form of the formula:
+  form of:
   \[\widehat{\Phi}_n \wedge \underbrace{\bigwedge_{\begin{array}{l}
@@
   \stamp{X}{\leq \xi(i_j)} \cup \set{x_k \mid k < i_j, Q_k =
-    \exists}$, for all $j \in [1,\ell]$, and moreover $\xi$ is
-  monotonic, we obtain: \begin{compactitem}
+    \exists}$, for all $j \in [1,\ell]$, and moreover $\xi^{-1}$ is
+  strictly monotonic, we obtain: \begin{compactitem}
   \item $\voc{\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
       \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{0} \cup
-    \stamp{X}{0} \cup \set{x_j \mid j < \max(\xi^{-1}(0)), Q_j =
-      \exists}$,
+    \stamp{X}{0} \cup \set{x_j \mid j < \xi^{-1}(0), Q_j = \exists}$,
     %

### 2018-10-18 quantifiers/draft.tex
ok

@@
      \nat$ the set $\set{m \mid \xi(m)=n}$ is finite. If $\xi$ is
-     finite-range, we denote by $\xi^{-1}(n) \in \nat$ the maximal
-     value $m$ such that $\xi(m)=n$. It is easy to see that $\xi^{-1}$
-     is strictly monotonic if $\xi$ is monotonic and finite-range. The
-     lemma below gives the proof of correctness for the previous
-     construction of non-local GLI.
+     finite-range, we denote by $\xi_{\max}^{-1}(n) \in \nat$ the
+     maximal value $m$ such that $\xi(m)=n$. It is easy to see that
+     $\xi_{\max}^{-1}$ is strictly monotonic if $\xi$ is monotonic and
+     finite-range. The lemma below gives the proof of correctness for
+     the previous construction of non-local GLI.
 
@@
       \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{0} \cup
-    \stamp{X}{0} \cup \set{x_j \mid j < \xi^{-1}(0), Q_j = \exists}$,
+    \stamp{X}{0} \cup \set{x_j \mid j < \xi_{\max}^{-1}(0), Q_j = \exists}$,
     %
@@
     \stamp{Q}{i} \cup \stamp{X}{\leq i} \cup \set{x_j \mid j <
-      \max(\xi^{-1}(i)), Q_j = \exists}$, for all $i \in [1,n]$,
+      \xi_{\max}^{-1}(i), Q_j = \exists}$, for all $i \in [1,n]$,
     %
@@
       \stamp{Q}{0} \cup \stamp{X}{0} \cup \set{x_j \mid j <
-        \max(\xi^{-1}(0)), Q_j = \exists}$,
+        \xi_{\max}^{-1}(0), Q_j = \exists}$,
       %
@@
       \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
-        \max(\xi^{-1}(k)), Q_j = \exists}$, for all $k \in [1,n]$,
+       \xi_{\max}^{-1}(k), Q_j = \exists}$, for all $k \in [1,n]$,
       %
@@
 In conclusion, under two assumptions about the first-order theory of
-the data domain, namely\begin{inparaenum}[(i)]
+the data domain, namely the \begin{inparaenum}[(i)]
 \item witness-producing quantifier elimination, and
@@
 variables introduced at the previous steps $\set{x_j \mid j <
-  \max(\xi^{-1}(k)), Q_j = \exists}$. The remaining question is how to
-use such non-local interpolants to label the unfolding of an automaton
-(Definition \ref{def:unfolding}) and to compute the coverage between
-nodes of the unfolding (Definition \ref{def:coverage}).
+  \xi_{\max}^{-1}(k), Q_j = \exists}$. The remaining question is how
+to use such non-local interpolants to label the unfolding of an
+automaton (Definition \ref{def:unfolding}) and to compute the coverage
+between nodes of the unfolding (Definition \ref{def:coverage}).
 
@@
 with the transition relation of the automaton, according to the point
-(\ref{it3:unfolding}) of Definition \ref{def:unfolding}. Suppose that
-a given input event sequence $\alpha \in \dom(U)$ is labeled with a
-sequence of formulae $U(\alpha_0) = \iota, U(\alpha_1), \ldots
-U(\alpha_n) = U(\alpha)$ in a way that fulfills the requirement from
-Definition \ref{def:unfolding} (\ref{it3:unfolding}) and, moreover, it
-has a non-local GLI $(I_0, \ldots, I_n)$, such that $\voc{I_k}
-\subseteq \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup
-\set{x_{i_1},\ldots,x_{i_k}}$, where $x_{i_1},\ldots,x_{i_n}$ are the
-renamed variables occurring in the rules of $\Delta$ under the scope
-of existential transition quantifiers\footnote{These are the
-  existentially quantified variables from the prenex normal form of
-  $\quantform{\alpha_k}$.}. The interpolant is used to refine the
-labeling of the sequence as follows, for all $k \in
-[n]$: \begin{compactitem}
-\item let $J_k \in \posforms(Q,\emptyset)$ be the formula obtained
-  from $I_k$ by \begin{inparaenum}[(i)]
-    \item replacing each time-stamped predicate symbol $\stamp{q}{k}$ by $q$, and
-      %
-    \item existentially quantifying each variable $y \in
-      \stamp{X}{\leq k} \cup \set{x_{i_1},\ldots,x_{i_k}}$,
-  \end{inparaenum}
-%
-\item define the new labeling as $U'(\alpha_k) = U(\alpha_k) \wedge J_k$. 
+(\ref{it3:unfolding}) of Definition \ref{def:unfolding}. The following
+lemma describes the refinement of the labeling of an input sequence
+$\alpha$ of length $n$ by a non-local GLI $(I_0, \ldots, I_n)$, such
+that $\voc{I_k} \subseteq \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup
+\vec{x}_k$, where $\vec{x}_k$ are the existentially quantified
+variables from the prenex normal form of $\quantform{\alpha_k}$.
+
+\begin{lemma}\label{lemma:refinement}
+  Let $U$ be an unfolding of an automaton $\A =
+  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ such that $\alpha = a_1\ldots a_n
+  \in \dom(U)$ and $(I_0, \ldots, I_n)$ be a GLI for $\alpha$. The
+  mapping $U' : \dom(U) \rightarrow \posforms(Q,\emptyset)$ defined as: 
+  \begin{compactitem}
+  \item $U'(\alpha_k) = U(\alpha_k) \wedge J_k$, for all $k \in [n]$,
+    where $J_k$ is the formula obtained from $I_k$ by replacing each
+    time-stamped predicate symbol $\stamp{q}{k}$ by $q$ and
+    existentially quantifying each free variable in $I_k$,
+  %
+\item $U'(\beta) = U(\beta)$ if $\beta \in \dom(U)$ and $\beta
+  \not\prefix \alpha$,
   \end{compactitem}
-
-Observe that, by Lemma \ref{lemma:subst-quant}
-(\ref{it2:subst-quant}), the set of free variables of a GLI formula
-$I_k$ consists of \begin{inparaenum}[(i)]
-\item variables $\stamp{X}{\leq k}$ keeping track of data
-values seen in the input at some earlier moment in time, or 
-%
-\item variables $\set{x_{i_1},\ldots,x_{i_k}}$ that track past choices
-  made within the transition rules. 
-\end{inparaenum}
-Basically, it is not important when exactly in the past a certain
-input has been read or when a choice has been made, as only the value
-of the variable determines the future behavior. Intuitively,
-existential quantification of these variables does the job of ignoring
-when in the past these values have been seen. Formally, the new set of
-formulae $U'(\alpha_0), \ldots, U'(\alpha_n)$ complies with Definition
-\ref{def:unfolding}, because: \begin{compactitem}
+  is an unfolding of $\A$. 
+\end{lemma}
+\proof{
+  The new set of formulae $U'(\alpha_0), \ldots, U'(\alpha_n)$ complies
+  with Definition \ref{def:unfolding}, because: \begin{compactitem}
   \item $U'(\alpha_0) \equiv \iota$, since, by point
@@
     \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
-  \rightarrow \stamp{\psi}{k+1} \models I_{k+1}\] Then the following
-  entailment holds:
-  \[\stamp{I}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
+    \rightarrow \stamp{\psi}{k+1} \models I_{k+1}\] We write
+    $\overstamp{I}{j}_k$ for the formula in which each predicate
+    symbol $\stamp{q}{k}$ is replaced by $\stamp{q}{j}$. Then the
+    following entailment holds:
+    \[\overstamp{I}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
     \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
-  \rightarrow \stamp{\psi}{1} \models \stamp{I}{1}_{k+1}\]
-  Because $J_k$ is obtained by existentially quantifying the free variables of $I_k$, we also obtain: 
+    \rightarrow \stamp{\psi}{1} \models \overstamp{I}{1}_{k+1}\]
+    Because $J_k$ is obtained by removing the time stamps from the predicate symbols and  
+    existentially quantifying all the free variables of $I_k$, we also
+    obtain, applying Fact \ref{fact:ex-entail} below:
     \[\stamp{J}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
@@
 
+  \begin{fact}\label{fact:ex-entail}
+    Given formulae $\phi(\vec{x}, \vec{y})$ and $\psi(\vec{x})$ such
+    that $\phi(\vec{x}, \vec{y}) \models \psi(\vec{x})$, we also have
+    $\exists \vec{x} ~.~ \phi(\vec{x},\vec{y}) \models \exists \vec{x}
+    ~.~ \psi(\vec{x})$. 
+  \end{fact}
+  \proof{ For each choice of a valuation for the existentially
+    quantified variables on the left-hand side, we chose the same
+    valuation for the variables on the right-hand side. \qed}\qed}
+
+Observe that, by Lemma \ref{lemma:subst-quant}
+(\ref{it2:subst-quant}), the set of free variables of a GLI formula
+$I_k$ consists of \begin{inparaenum}[(i)]
+\item variables $\stamp{X}{\leq k}$ keeping track of data values seen
+  in the input at some earlier moment in time, and
+%
+\item variables that track past choices made within the transition
+  rules.
+\end{inparaenum}
+Basically, it is not important when exactly in the past a certain
+input has been read or when a choice has been made, as only the value
+of the variable determines the future behavior. Intuitively,
+existential quantification of these variables does the job of ignoring
+when in the past these values have been seen.
+
 The last ingredient of the lazy annotation semi-algorithm based on
-unfolding refinement with coverage consist in the implementation of
-the coverage check, when the unfolding of an automaton is labeled with
-conjunctions of existentially quantified formulae with predicate
-symbols, obtained from interpolation. By Definition
-\ref{def:coverage}, checking whether a given node $\alpha \in \dom(U)$
-is covered amounts to finding a prefix $\alpha' \prefix \alpha$ and a
-node $\beta \in \dom(U)$ such that $U(\alpha') \models U(\beta)$, or
-equivalently, the formula $U(\alpha') \wedge \neg U(\beta)$ is
-unsatisfiable. However, the latter formula, in prenex form, has
-quantifier prefix in the language $\exists^*\forall^*$ and, as
-previously mentioned, the satisfiability problem for such formulae
-becomes undecidable when the data theory subsumes Presburger
-arithmetic \cite{Halpern91}.
+unfoldings consist in the implementation of the coverage check, when
+the unfolding of an automaton is labeled with conjunctions of
+existentially quantified formulae with predicate symbols, obtained
+from interpolation. By Definition \ref{def:coverage}, checking whether
+a given node $\alpha \in \dom(U)$ is covered amounts to finding a
+prefix $\alpha' \prefix \alpha$ and a node $\beta \in \dom(U)$ such
+that $U(\alpha') \models U(\beta)$, or equivalently, the formula
+$U(\alpha') \wedge \neg U(\beta)$ is unsatisfiable. However, the
+latter formula, in prenex form, has quantifier prefix in the language
+$\exists^*\forall^*$ and, as previously mentioned, the satisfiability
+problem for such formulae becomes undecidable when the data theory
+subsumes Presburger arithmetic \cite{Halpern91}.
 
@@
 answering the above satisfiability query, then the semi-algorithm
-assumes the node is not covered and continues exploring its
+assumes that the node is not covered and continues exploring its
 successors. Failure to compute complete coverage may lead to
-divergence (non-termination) and failure to prove emptiness, but does
-not affect the soundness of the semi-algorithm (real counterexamples
-will still be found).
+divergence (non-termination) and ultimately, to failure to prove
+emptiness, but does not affect the soundness of the semi-algorithm
+(real counterexamples will still be found).
 
@@
 contribution is providing a systematic semi-algorithm for these
-decision problems. In addition, we can extend our previous
-\emph{generic register automata} \cite{HolikIosifRogalewiczVojnar18}
-inclusion checking framework, by allowing monitor (right-hand side)
-automata to have local variables, that are not visible in the
-language\footnote{We conjecture that hidden computations using
-  invisible variables (clocks) are the reason for which timed and
-  register automata cannot be complemented within their class.}.
+decision problems. In addition, the method described in
+\S\ref{sec:emptiness} can extend our previous \emph{generic register
+  automata} \cite{HolikIosifRogalewiczVojnar18} inclusion checking
+framework, by allowing monitor (right-hand side) automata to have
+local variables, that are not visible in the language\footnote{We
+  conjecture that hidden computations using invisible variables
+  (clocks) are the reason for which timed and register automata cannot
+  be complemented within their class.}.
 
@@
 are clock valuations, for all $i \in [n]$, and: \begin{compactitem}
-\item $s_0 \in S_0$ and $\gamma_0(x)$ for all $x \in C$, 
+\item $s_0 \in S_0$ and $\gamma_0(x)=0$ for all $x \in C$, 
 %
@@
 $w$. As usual, we denote by $\lang{T}$ the set of finite words
-accepted by $T$. It is well-known \cite{AlurDill94} that, in general,
-there is no timed automaton accepting the complement language
-$(\Sigma\times\real)^* \setminus \lang{T}$ and, moreover, the language
-inclusion problem is undecidable.
+accepted by $T$. It is well-known that, in general, there is no timed
+automaton accepting the complement language $(\Sigma\times\real)^*
+\setminus \lang{T}$ and, moreover, the language inclusion problem is
+undecidable \cite{AlurDill94}.
 
@@
 all the clocks of $T$ evolve at the same pace, which is easily done
-using a technique from \cite{Fribourg98}, that consists in replacing
-each clock $x_i$ of $T$ by a variable $y_i$ who tracks the difference
-between the values of $t$ and $x_i$, instead of the value of $x_i$
-directly.
+using a technique from \cite{Fribourg98}, which replaces
+each clock $x_i$ of $T$ by a variable $y_i$ tracking the difference
+between the values of $t$ and $x_i$.
 
@@
 $\Delta_T$ contains the rule:
-\[\begin{array}{rcl}
-q_i(y_1,\ldots,y_k,z) \arrow{a(t)}{} \exists y'_1 \ldots \exists y'_k
-~.~ t > z \wedge \delta(z-y_1,\ldots,z-y_k) & \wedge & 
- \bigwedge_{x_i \in \lambda} y'_i = z \wedge \bigwedge_{x_i \not\in \lambda} y'_i = y_i \\ 
-& \wedge & q_j(y'_1,\ldots,y'_k,t)
-\end{array}\]
-Moreover, nothing else is in $\Delta_T$. We establish the following
-connection between a timed automaton and its corresponding first-order
-alternating automaton.
+\[q_i(y_1,\ldots,y_k,z) \arrow{a(t)}{} t > z \wedge \delta(z-y_1,\ldots,z-y_k) \wedge q_j(y'_1,\ldots,y'_k,t)\]
+where $y'_i$ stands for $z$ if $x_i \in \lambda$ and for $y_i$, 
+otherwise.  Moreover, nothing else is in $\Delta_T$. We establish the
+following connection between a timed automaton and its corresponding
+first-order alternating automaton.
 
@@
 $\lang{\A_{T_1}} \cap \lang{\overline{\A_{T_2}}} = \emptyset$, for
-which (\S\ref{sec:emptiness}) provides a semi-algorithm. Observe
-further that the transition quantifiers from the encoding of timed
-automata into first-order alternating automata can be easily
-eliminated by substitution as in $\exists x ~.~ x = t \wedge \phi
-\equiv \phi[t/x]$.
+which (\S\ref{sec:emptiness}) provides a semi-algorithm. Observe,
+moreover, that no transition quantifiers are needed to encode timed
+automata as first-order alternating automata.
 
@@
 with a finite set of registers $x_1,\ldots,x_r$ able to copy input
-values and compare them with later the input. Consequently, basic
+values and compare them with subsequent input. Consequently, basic
 results from classical automata theory, such as the pumping lemma or
-the closure under complement do not hold in this model and inclusion
-of languages recognized by register automata is undecidable
+the closure under complement do not hold in this model and, moreover,
+inclusion of languages recognized by register automata is undecidable
 \cite{NevenSchwentickVianu03}.
@@
 Let $\Sigma$ be an infinite alphabet, $\#$ be a symbol not in $\Sigma$
-and $r>0$ be an integer constant. An \emph{assignment} is a word
-$\vec{v} = v_1 \ldots v_r$ such that if $v_i = v_j$ and $i \neq j$
-then $v_i = \#$, for all $i,j \in [1,r]$. We write $[\vec{v}]$ for the
-set $\set{v_i \mid i \in [1,r]}$. A \emph{finite-memory (register)
-  automaton} is a tuple $R = \tuple{S,q_0,\vec{u}, \rho, \mu, F}$,
-where $S$ is a finite set of states, $q_0 \in S$ is the initial state,
-$\vec{u} = u_1 \ldots u_r$ is the initial assignment, $\rho : S
-\rightarrow [1,r]$ is the reassignment partial function, $\mu
-\subseteq S \times [1,r] \times S$ is the transition relation and $F
-\subseteq S$ is the set of final states. A run of $A$ over an input
-word $a_1 \ldots a_n \in \Sigma^*$ is a sequence $(s_0,\vec{v}_0)
-\ldots (s_n,\vec{v}_n)$ such that $\vec{v}_0 = \vec{u}$ and, for all
-$i \in [1,n]$, exactly one of the following holds: \begin{compactitem}
+and $r>0$ be an integer constant, denoting the number of registers. An
+\emph{assignment} is a word $\vec{v} = v_1 \ldots v_r$ such that if
+$v_i = v_j$ and $i \neq j$ then $v_i = \#$, for all $i,j \in
+     [1,r]$. We write $[\vec{v}]$ for the set $\set{v_i \mid i \in
+       [1,r]}$ of values in the assignment $\vec{v}$. A
+     \emph{finite-memory (register) automaton} is a tuple $R =
+     \tuple{S,q_0,\vec{u}, \rho, \mu, F}$, where $S$ is a finite set
+     of states, $q_0 \in S$ is the initial state, $\vec{u} = u_1
+     \ldots u_r$ is the initial assignment, $\rho : S \rightarrow
+            [1,r]$ is the reassignment partial function, $\mu
+            \subseteq S \times [1,r] \times S$ is the transition
+            relation and $F \subseteq S$ is the set of final states. A
+            run of $A$ over an input word $a_1 \ldots a_n \in
+            \Sigma^*$ is a sequence $(s_0,\vec{v}_0) \ldots
+            (s_n,\vec{v}_n)$ such that $\vec{v}_0 = \vec{u}$ and, for
+            all $i \in [1,n]$, exactly one of the following
+            holds: \begin{compactitem}
 \item if there exists $k\in[1,r]$ such that $a_i=(\vec{v}_{i-1})_k$
@@
 \set{s_0,\ldots,s_m}$, we define the alternating automaton $\A_R =
-\tuple{\set{\bullet},\set{x},Q_R,\iota_R,F_R,\Delta_R}$, where
-$\bullet \not\in \Sigma$, $Q_R \isdef \set{q_0,\ldots,q_m}$ and
+\tuple{\set{\alpha},\set{x},Q_R,\iota_R,F_R,\Delta_R}$, where
+$\alpha \not\in \Sigma$, $Q_R \isdef \set{q_0,\ldots,q_m}$ and
 $\#(q_i) = r$ for all $i \in [m]$, $\iota_R \isdef q_0(\vec{u})$, $F_R
@@
 s_j) \in \mu$, $\Delta_T$ contains the rule:
-\[q_i(y_1,\ldots,y_r) \arrow{a(x)}{} y_k=x \wedge q_j(y_1,\ldots,y_r) ~\vee~
+\[q_i(y_1,\ldots,y_r) \arrow{\alpha(x)}{} y_k=x \wedge q_j(y_1,\ldots,y_r) ~\vee~
 \bigwedge_{i=1}^r x \neq y_i \wedge q_j(y_1,\ldots,y_{k-1},x,y_{k+1},
@@
   over an infinite alphabet $\Sigma$, the first-order alternating
-  automaton $\A_R = \tuple{\set{\bullet},Q_R,\iota_R,F_R,\Delta_R}$
-  recognizes the languge $\lang{\A_R} = \set{ (\bullet,a_1) \ldots
-    (\bullet,a_n) \mid a_1 \ldots a_n \in \lang{R} }$. 
+  automaton $\A_R = \tuple{\set{\alpha},Q_R,\iota_R,F_R,\Delta_R}$
+  recognizes the languge $\lang{\A_R} = \set{ (\alpha,a_1) \ldots
+    (\alpha,a_n) \mid a_1 \ldots a_n \in \lang{R} }$. 
 \end{proposition}
-\proof{ ``$\subseteq$'' Let $w = (\bullet,a_1) \ldots (\bullet,a_n)
+\proof{ ``$\subseteq$'' Let $w = (\alpha,a_1) \ldots (\alpha,a_n)
   \in \lang{\A_R}$. First, it is easy to show that each execution of
@@
   such that $\vec{v}_0 = \vec{u}$. By induction on $n\geq0$, we can
-  build an execution of $\A_R$ over $(\bullet,a_1) \ldots
-  (\bullet,a_n)$ that is a linear tree with labels $q_0(\vec{v}_0),
+  build an execution of $\A_R$ over $(\alpha,a_1) \ldots
+  (\alpha,a_n)$ that is a linear tree with labels $q_0(\vec{v}_0),
   \ldots, q_n(\vec{v}_n)$. \qed}
@@
 (\S\ref{sec:emptiness}) provides a semi-algorithm. Notice further that
-the encoding of register automata uses no transition quantifiers.
+the encoding of register automata as first-order alternating automata
+uses no transition quantifiers.
 
@@
 checking termination and, generally, liveness properties
-\cite{Farzan16}. Thus, syntactically speaking, predicate automata are
-a subset of first-order alternating automata.
+\cite{Farzan16}.
 
@@
 defined using the entire sets of models of the initial sentence and
-transition rules, not just the minimal ones, as in our case. Formally,
-a run of a predicate automaton $P =
+transition rules, not just the minimal ones, as in our case.
+
+Formally, a run of a predicate automaton $P =
 \tuple{\Sigma,\set{x},Q,\iota,F,\Delta}$ over a word $(a_1,\nu_1)
@@
 equivalent, from the language point of view, with the semantics given
-by Definitions \ref{def:execution} and \ref{def:accepting}. However,
-the semantics of first-order alternating automata based on minimal
-models is important for its relation to the classical textbook
+by Definitions \ref{def:execution} and \ref{def:accepting}.  We
+believe that the semantics of first-order alternating automata based
+on minimal models is important for its relation to the textbook
 semantics of boolean alternating automata
@@
 \proof{
-
 \qed}
-
-As before, this result enables using the semi-algorithm of
-(\S\ref{sec:emptiness}) for checking emptiness of predicate automata.
+As before, this result enables using the semi-algorithm from
+\S\ref{sec:emptiness} for checking emptiness of predicate automata.
 We point out that, although quantifier-free predicate automata with
@@
 currently there is no method for checking emptiness of predicate
-automata with \begin{inparaenum}[(i)]
-\item predicates of arity greater than one, or 
-\item transition quantifiers. 
-\end{inparaenum}
+automata with predicates of arity greater than one, other than the
+explicit enumeration of cubes. Moreover, no method for dealing with
+emptiness in the presence of transition quantifiers is known to exist.
 

### 2018-10-19 quantifiers/draft.tex
ok

@@
 \end{abstract}
+
+\section{Introduction}
+
+Many results in formal language theory rely on the assumption that
+languages are defined over finite alphabets. In practice, this
+assumption is not realistic when attempting to model real-time
+hardware and communication protocols or even simple programs, whose
+input and observable output requires taking into account data values,
+ranging over very large domains, better viewed as infinite
+mathematical abstractions.
+
+Alternating automata are a generalization of nondeterministic automata
+with universal transitions, that create several copies of the
+automaton, which synchronize on the same input word. Alternating
+automata are appealing for verification because they allow encoding of
+problems such as temporal logic model checking in linear time, as
+opposed to the exponential time required by nondeterministic automata
+\cite{Vardi95}.
+
+%% However, checking the emptiness of an alternating automaton over a
+%% finite alphabet is \pspace-complete, instead of
+%% \nlogspace-complete, which is the case for nondeterministic automata.
+%% In practice, this complexity can be dealt with using efficient
+%% antichain algorithms \cite{Ganty09}.
+
+A finite-alphabet alternating automaton is typically described by
+transitions $q \arrow{a}{} \phi$, where $q$ is a state, $a$ is an
+input symbol and $\phi$ is a positive boolean combinations of states,
+viewed as propositional variables. Here we introduce a generalized
+alternating automata model in which states are predicate symbols
+$q(y_1,\ldots,y_k)$, the input has associated data variables
+$x_1,\ldots,x_n$, ranging over an infinite domain and transitions are
+of the form $q(y_1,\ldots,y_k) \arrow{a(x_1,\ldots,x_n)}{} \phi$,
+where $\phi$ is any formula in the first-order theory of the data
+domain, in which each state predicate occurs under an even number of
+negations. In this model, the arguments of a predicate atom
+$q(y_1,\ldots,y_k)$ track the values of the \emph{internal variables}
+associated with the state. Together with the input values
+$x_1,\ldots,x_n$, these values are used to compute the successor
+states and are invisible in the input sequence.
+
+Previous attempts to generalized classical Rabin-Scott automata to
+infinite alphabets, such as timed automata \cite{AlurDill94} and
+finite-memory (register) automata \cite{KaminskiFrancez94} face the
+\emph{complement closure} problem: there exist automata for which the
+complement language cannot be recognized by an automaton in the same
+class. This excludes the possibility of encoding a language inclusion
+problem $\lang{A} \subseteq \lang{B}$ as the emptiness of an automaton
+recognizing the language $\lang{A} \cap \clang{B}$, where $\clang{B}$
+denotes the complement of $\lang{B}$. However, when all internal
+variables are made visible in the input, determinization and
+complementation are possible, for data theories closed under negation,
+such as integer or real arithmetics \cite{IosifRV16}.
+
+The solution we adopt is a tight coupling of internal variables to
+control states. As we show, this allows for linear-time
+complementation just as in the case of boolean alternating
+automata. Complementation is, moreover, possible when the transition
+formulae contain first-order quantifiers, generating
+infinitely-branching executions.
+
+The price to be paid for this expressivity is that emptiness of
+first-order alternating automata is undecidable, even for the simplest
+data theory of equality \cite{Farzan15}. We work around this problem
+by developing an abstraction refinement lazy annotation semi-algorithm
+in the spirit of the \impact~procedure for checking safety of
+nondeterministic integer programs \cite{McMillan06}. However, checking
+emptiness of first-order alternating automata by lazy annotation with
+interpolants faces two problems: \begin{compactenum}
+\item Quantified transition rules are problematic to decide if a given
+  trace is spurious, mainly because adding uninterpreted predicate
+  symbols to decidable first-order theories, such as Presburger
+  arithmetic, results in undecidability \cite{Halpern91}. We deal with
+  this problem by assuming that the first-order data theory, without
+  uninterpreted predicate symbols, has an effective quantifier
+  elimination procedure, that instantiates quantifers with witness
+  terms.
+%
+\item The interpolants used to prove the spuriousness of an input
+  sequence are not \emph{local} anymore, as they may refer to input
+  values encountered in the past. However, the future executions are
+  oblivious to when these values have been seen in the past and depend
+  only on the data constraints between the values. We use this
+  observation to define a labeling of nodes visited by the lazy
+  annotation semi-algorithm as conjunctions of existentially
+  quantified interpolants that combine predicate symbols with data
+  constraints.
+\end{compactenum}
+
+The main contribution of this paper is an effective emptiness checking
+semi-algorithm for first-order alternating automata. Further, we
+describe applications of this method to several undecidable but
+recursively enumerable problems, for which no effective
+semi-algorithms exist, to date: inclusion between recognizable timed
+languages \cite{AlurDill94} and between languages recognized by
+finite-memory automata \cite{KaminskiFrancez94}, as well as emptiness
+of predicate automata, a model of first-order alternating automata
+used to check safety and liveness properties of parameterized
+concurrent programs \cite{Farzan15,Farzan16}.
+
+\paragraph{\bf Related Work} 
+The first-order alternating automata model presented in this paper
+stems from our previous work on boolean alternating automata extended
+with variables ranging over infinite data \cite{IosifXu18}. There we
+considered that states are propositional variables, as in the
+classical textbook alternating automata model, and that all the
+variables of the automaton are reflected in the input. The model in
+this paper overcomes this latter restriction by allowing for internal
+variables, whose variables are not visible in the language. This
+solves an older language inclusion problem $\bigcap_{i=1}^n\lang{A_i}
+\subseteq \lang{B}$, between finite-state automata with data variables
+\cite{IosifRV16}. There, we assumed that all variables of the observer
+automaton $B$ must be declared in the automata $A_1, \ldots, A_n$ that
+model the concurrent components of the system under check.
 
@@
 Typically, when checking the unreachability of a set of program
-configurations \cite{mcmillan06}, the interpolants used to annotate
+configurations \cite{McMillan06}, the interpolants used to annotate
 the unfolded control structure are assertions about the values of the
@@
 \S\ref{sec:emptiness} can extend our previous \emph{generic register
-  automata} \cite{HolikIosifRogalewiczVojnar18} inclusion checking
-framework, by allowing monitor (right-hand side) automata to have
-local variables, that are not visible in the language\footnote{We
-  conjecture that hidden computations using invisible variables
-  (clocks) are the reason for which timed and register automata cannot
-  be complemented within their class.}.
+  automata} \cite{IosifRV16} inclusion checking framework, by allowing
+monitor (right-hand side) automata to have local variables, that are
+not visible in the language.
 
@@
 each $q \in Q$ and each tuple $\tuple{d_1,\ldots,d_{\#(q)}} \in
-\I_{i-1}(q)$, we have $\I_i \in \sem{\psi}_\mu$, for each rule
+\I_{i-1}(q)$, we have $\I_i \in \sem{\psi}_\nu$, for each rule
 $q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(x)}{} \psi \in \Delta$, where
-$\mu=\nu_i[y_1\leftarrow d_1, \ldots, y_{\#(q)} \leftarrow
+$\nu=\nu_i[y_1\leftarrow d_1, \ldots, y_{\#(q)} \leftarrow
   d_{\#(q)}]$. The run is accepting if and only if $\I(q)=\emptyset$
@@
 \end{proposition}
-\proof{
-\qed}
-As before, this result enables using the semi-algorithm from
-\S\ref{sec:emptiness} for checking emptiness of predicate automata.
-We point out that, although quantifier-free predicate automata with
-predicates of arity one are decidable for emptiness \cite{Farzan15},
-currently there is no method for checking emptiness of predicate
-automata with predicates of arity greater than one, other than the
-explicit enumeration of cubes. Moreover, no method for dealing with
-emptiness in the presence of transition quantifiers is known to exist.
+\proof{ ``$\subseteq$'' Let $w = (a_1,\nu_1) \ldots (a_n,\nu_n) \in
+  \lang{P}$ be a word and $\I_0, \ldots, \I_n$ be an accepting
+  execution of $P$ over $w$. Let $\stamp{\I}{i}_j$ be the
+  interpretation that associates each predicate $\stamp{q}{i}$ the set
+  $\I_j(q)$, for $i,j \in [n]$. Then one builds, by induction on
+  $n\geq0$, an execution $\T$ of $\A_P$ such that $\I_\T \subseteq
+  \bigcup_{i=0}^n \stamp{\I}{i}_i$, where $\I_\T$ is the unique
+  interpretation associated with $\T$. Since $\I_0, \ldots, \I_n$ is
+  accepting, we have $\stamp{\I}{n}_n(\stamp{q}{n}) = \emptyset$, for
+  all $q \in Q \setminus F$ and hence $\I_\T(\stamp{q}{n}) =
+  \emptyset$, for all $q \in Q \setminus F$ and, consequently $w \in
+  \lang{\A_P}$. ``$\supseteq$'' Let $w = (a_1,\nu_1) \ldots
+  (a_n,\nu_n) \in \lang{\A_P}$ be a word and $\T$ be an accepting
+  execution of $\A_P$ over $w$. We define the sequence of
+  interpretations $\I_0, \ldots, \I_n$ as $\I_i(q) =
+  \I_\T(\stamp{q}{i})$, for each $i \in [n]$ and each $q \in Q$. By
+  induction on $n\geq0$ one shows that $\I_0, \ldots, \I_n$ is an
+  execution $P$. Moreover, since $\T$ is accepting, we have $\I_n(q) =
+  \I_\T(\stamp{q}{n}) = \emptyset$, for each $q \in Q \setminus F$,
+  thus $w \in \lang{P}$. \qed}
+
+As before, this result enables using the
+semi-algorithm from \S\ref{sec:emptiness} for checking emptiness of
+predicate automata.  We point out that, although quantifier-free
+predicate automata with predicates of arity one are decidable for
+emptiness \cite{Farzan15}, currently there is no method for checking
+emptiness of predicate automata with predicates of arity greater than
+one, other than the explicit enumeration of cubes. Moreover, no method
+for dealing with emptiness in the presence of transition quantifiers
+is known to exist.
 

### 2018-10-22 quantifiers/draft.tex
ok

@@
 \begin{abstract}
+We introduce first order alternating automata, a generalization of
+boolean alternating automata, in which transition rules are described
+by multisorted first order formulae, with states and internal
+variables given by uninterpreted predicate terms. The model is closed
+under union, intersection and complement, and its emptiness problem is
+undecidable, even for the simplest data theory of equality. To cope
+with this limitation, we develop an abstraction refinement
+semi-algorithm based on lazy annotation of the symbolic execution
+paths with interpolants, obtained by applying (i) quantifier
+elimination with witness term generation and (ii) Lyndon interpolation
+in the quantifier-free data theory with uninterpreted predicate
+symbols. This provides a method for checking inclusion of timed and
+finite-memory register automata, and emptiness of quantified predicate
+automata, previously used in the verification of parameterized
+concurrent programs, composed of replicated threads, with a
+shared-memory communication model.
 \end{abstract}
@@
 opposed to the exponential time required by nondeterministic automata
-\cite{Vardi95}.
-
-%% However, checking the emptiness of an alternating automaton over a
-%% finite alphabet is \pspace-complete, instead of
-%% \nlogspace-complete, which is the case for nondeterministic automata.
-%% In practice, this complexity can be dealt with using efficient
-%% antichain algorithms \cite{Ganty09}.
-
-A finite-alphabet alternating automaton is typically described by
-transitions $q \arrow{a}{} \phi$, where $q$ is a state, $a$ is an
-input symbol and $\phi$ is a positive boolean combinations of states,
-viewed as propositional variables. Here we introduce a generalized
-alternating automata model in which states are predicate symbols
-$q(y_1,\ldots,y_k)$, the input has associated data variables
-$x_1,\ldots,x_n$, ranging over an infinite domain and transitions are
-of the form $q(y_1,\ldots,y_k) \arrow{a(x_1,\ldots,x_n)}{} \phi$,
-where $\phi$ is any formula in the first-order theory of the data
-domain, in which each state predicate occurs under an even number of
-negations. In this model, the arguments of a predicate atom
-$q(y_1,\ldots,y_k)$ track the values of the \emph{internal variables}
-associated with the state. Together with the input values
-$x_1,\ldots,x_n$, these values are used to compute the successor
-states and are invisible in the input sequence.
-
-Previous attempts to generalized classical Rabin-Scott automata to
+\cite{Vardi95}. A finite-alphabet alternating automaton is typically
+described by a set of transition rules $q \arrow{a}{} \phi$, where $q$
+is a state, $a$ is an input symbol and $\phi$ is a positive boolean
+combinations of states, viewed as propositional variables.
+
+Here we introduce a generalized alternating automata model in which
+states are predicate symbols $q(y_1,\ldots,y_k)$, the input has
+associated data variables $x_1,\ldots,x_n$, ranging over an infinite
+domain and transitions are of the form $q(y_1,\ldots,y_k)
+\arrow{a(x_1,\ldots,x_n)}{} \phi$, where $\phi$ is any formula in the
+first-order theory of the data domain, in which each state predicate
+occurs under an even number of negations. In this model, the arguments
+of a predicate atom $q(y_1,\ldots,y_k)$ track the values of the
+\emph{internal variables} associated with the state. Together with the
+input values $x_1,\ldots,x_n$, these values are used to compute the
+successor states and are invisible in the input sequence.
+
+Previous attempts to generalize classical Rabin-Scott automata to
 infinite alphabets, such as timed automata \cite{AlurDill94} and
@@
 recognizing the language $\lang{A} \cap \clang{B}$, where $\clang{B}$
-denotes the complement of $\lang{B}$. However, when all internal
-variables are made visible in the input, determinization and
-complementation are possible, for data theories closed under negation,
-such as integer or real arithmetics \cite{IosifRV16}.
-
-The solution we adopt is a tight coupling of internal variables to
-control states. As we show, this allows for linear-time
-complementation just as in the case of boolean alternating
-automata. Complementation is, moreover, possible when the transition
-formulae contain first-order quantifiers, generating
-infinitely-branching executions.
-
-The price to be paid for this expressivity is that emptiness of
-first-order alternating automata is undecidable, even for the simplest
-data theory of equality \cite{Farzan15}. We work around this problem
-by developing an abstraction refinement lazy annotation semi-algorithm
-in the spirit of the \impact~procedure for checking safety of
-nondeterministic integer programs \cite{McMillan06}. However, checking
-emptiness of first-order alternating automata by lazy annotation with
-interpolants faces two problems: \begin{compactenum}
-\item Quantified transition rules are problematic to decide if a given
-  trace is spurious, mainly because adding uninterpreted predicate
-  symbols to decidable first-order theories, such as Presburger
-  arithmetic, results in undecidability \cite{Halpern91}. We deal with
-  this problem by assuming that the first-order data theory, without
-  uninterpreted predicate symbols, has an effective quantifier
-  elimination procedure, that instantiates quantifers with witness
-  terms.
-%
-\item The interpolants used to prove the spuriousness of an input
-  sequence are not \emph{local} anymore, as they may refer to input
-  values encountered in the past. However, the future executions are
-  oblivious to when these values have been seen in the past and depend
-  only on the data constraints between the values. We use this
-  observation to define a labeling of nodes visited by the lazy
-  annotation semi-algorithm as conjunctions of existentially
-  quantified interpolants that combine predicate symbols with data
-  constraints.
+denotes the complement of $\lang{B}$.
+
+The solution we adopt here is a tight coupling of internal variables
+to control states, using uninterpreted predicate symbols. As we show,
+this allows for linear-time complementation just as in the case of
+boolean alternating automata. Complementation is, moreover, possible
+when the transition formulae contain first-order quantifiers,
+generating infinitely-branching execution trees. The price to be paid
+for this expressivity is that emptiness of first-order alternating
+automata is undecidable, even for the simplest data theory of equality
+\cite{Farzan15}.
+
+The main contribution of this paper is an effective emptiness checking
+semi-algorithm for first-order alternating automata, in the spirit of
+the \impact~procedure, originally developed for checking safety of
+nondeterministic integer programs \cite{McMillan06}. Essentially,
+checking emptiness of first-order alternating automata by lazy
+annotation with interpolants faces two problems: \begin{compactenum}
+\item Quantified transition rules make it hard, or even impossible, to
+  decide if a given symbolic trace is spurious. This is mainly because
+  adding uninterpreted predicate symbols to decidable first-order
+  theories, such as Presburger arithmetic, results in undecidability
+  \cite{Halpern91}. To deal with this problem, we assume that the
+  first order data theory, without uninterpreted predicate symbols,
+  has a quantifier elimination procedure, that instantiates quantifers
+  with effectively computable \emph{witness terms}.
+%
+\item The interpolants that prove the spuriousness of a symbolic path
+  are not \emph{local}, as they may refer to input values encountered
+  in the past. However, the future executions are oblivious to
+  \emph{when} these values have been seen in the past and depend only
+  on the data constraints between the values. We use this fact to
+  define a labeling of nodes, visited by the lazy annotation
+  procedure, with conjunctions of existentially quantified
+  interpolants combining predicate atoms with data constraints.
 \end{compactenum}
 
-The main contribution of this paper is an effective emptiness checking
-semi-algorithm for first-order alternating automata. Further, we
-describe applications of this method to several undecidable but
-recursively enumerable problems, for which no effective
-semi-algorithms exist, to date: inclusion between recognizable timed
+We describe applications of this semi-algorithmic method to several
+undecidable but recursively enumerable problems, for which no
+effective method exist, to date: inclusion between recognizable timed
 languages \cite{AlurDill94} and between languages recognized by
@@
 \paragraph{\bf Related Work} 
-The first-order alternating automata model presented in this paper
+The first order alternating automata model presented in this paper
 stems from our previous work on boolean alternating automata extended
 with variables ranging over infinite data \cite{IosifXu18}. There we
-considered that states are propositional variables, as in the
-classical textbook alternating automata model, and that all the
-variables of the automaton are reflected in the input. The model in
-this paper overcomes this latter restriction by allowing for internal
-variables, whose variables are not visible in the language. This
-solves an older language inclusion problem $\bigcap_{i=1}^n\lang{A_i}
-\subseteq \lang{B}$, between finite-state automata with data variables
+considered states to be propositional variables, as in the classical
+textbook alternating automata model, and all variables of the
+automaton to be observable in the input. The model in this paper
+overcomes this latter restriction by allowing for internal variables,
+whose variables are not visible in the language.
+
+This solves an older language inclusion problem
+$\bigcap_{i=1}^n\lang{A_i} \subseteq \lang{B}$, between finite-state
+automata with data variables, whose languages are alternating
+sequences of input events and variable valuations
 \cite{IosifRV16}. There, we assumed that all variables of the observer
@@
 model the concurrent components of the system under check.
+
+The work probably closest to the one reported here concerns the model
+of predicate automata (PA) \cite{Farzan15,Farzan16,KincaidPhD},
+applied to the verification of parameterized concurrent programs with
+shared memory. In this model, the alphabet consists of pairs of
+program statements and thread identifiers, thus being infinite because
+the number of threads is unbounded. Because thread identifiers can
+only be compared for (dis-)equality, the data theory in PA is the
+theory of equality. Even with this simplification, the emptiness
+problem is undecidable when either the predicates have arity greater
+than one \cite{Farzan15} or quantified transition rules
+\cite{KincaidPhD}. Checking emptiness of quantifier free PA is
+possible semi-algorithmically, by explicitly enumerating reachable
+configurations and checking coverage by looking for permuations of
+argument values. However, no semi-algorithm is given for quantified
+PA.  Dealing with quantified transition rules is one of the
+contributions of the work reported in this paper.
 
@@
 $\Data$, variables or function applications $f(t_1,\ldots,t_{\#(f)})$,
-where $t_1,\ldots,t_{\#(f)}$ are terms. The set of first-order
+where $t_1,\ldots,t_{\#(f)}$ are terms. The set of first order
 formulae is defined by the syntax below:
@@
 this case we call $\phi$ the \emph{matrix} of $\varphi$. Every
-first-order formula can be written in prenex form, by renaming each
+first order formula can be written in prenex form, by renaming each
 quantified variable to a unique name and moving the quantifiers
@@
 
-Formally, a \emph{first-order alternating automaton} is a tuple $\A =
+Formally, a \emph{first order alternating automaton} is a tuple $\A =
 \tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is a finite set of
@@
 formalize the execution semantics of automata, we relate sets of
-configurations to models of first-order sentences, as follows. Given a
+configurations to models of first order sentences, as follows. Given a
 formula $\phi$ and a valuation of its free variables $\nu$, each
@@
 automaton, referred to as transition quantifiers. We shall further
-assume (\S\ref{sec:emptiness}) that the first-order theory of the data
+assume (\S\ref{sec:emptiness}) that the first order theory of the data
 sort $\Data$ has quantifier elimination, which allows to effectively
@@
 
-We check emptiness of first-order alternating automata using a version
+We check emptiness of first order alternating automata using a version
 of the \impact~ lazy annotation semi-algorithm \cite{McMillan06}. An
@@
   instance, using breadth-first search.} from $\Sigma^*$ will suffice
-to discover a word $w \in \lang{\A}$, provided that the first-order
+to discover a word $w \in \lang{\A}$, provided that the first order
 theory of the data domain $\Data$ is decidable (Lemma
@@
 
-As mentioned above, we check emptiness of first-order alternating
+As mentioned above, we check emptiness of first order alternating
 automata using the same method previously used to check emptiness of a
@@
 \end{definition}
-%% The existence of Lyndon interpolants for first-order logic, with no
+%% The existence of Lyndon interpolants for first order logic, with no
 %% predefined interpretation of function symbols\footnote{A model
@@
 further that the equality atoms in the transition rules of $\A$ are
-written in the language of a decidable first-order theory, such as
+written in the language of a decidable first order theory, such as
 Presburger arithmetic, Lemma \ref{lemma:quant-pred-acceptance} gives
@@
 $\overline{\Phi}$, respectively. Because we assumed that the
-first-order theory of $\Data$ has quantifier elimination, the
+first order theory of $\Data$ has quantifier elimination, the
 satisfiability problem for $\substform{\alpha}$ is decidable. If
@@
   $\quantform{\alpha}$ and $\substform{\alpha}$, respectively. Since
-  we assumed that the first-order theory of the data domain has
+  we assumed that the first order theory of the data domain has
   witness-producing quantifier elimination, using Theorem
@@
 
-In conclusion, under two assumptions about the first-order theory of
+In conclusion, under two assumptions about the first order theory of
 the data domain, namely the \begin{inparaenum}[(i)]
@@
 
-The main application of first-order alternating automata is checking
+The main application of first order alternating automata is checking
 inclusion between various classes of automata extended with variables
@@
 parametric programs has been reduced to checking the emptiness of a
-(possibly infinite) sequence of first-order alternating automata,
+(possibly infinite) sequence of first order alternating automata,
 called \emph{predicate automata} \cite{Farzan15,Farzan16}, encoding
@@
 generalization of counterexamples. The program and the proof spaces
-are first-order alternating automata over the infinite alphabet of
+are first order alternating automata over the infinite alphabet of
 pairs consisting of program statements and thread identifiers.
@@
 Given a timed automaton $T = \tuple{\Sigma,S,S_0,F,C,E}$, we define a
-first-order alternating automaton $\A_T =
+first order alternating automaton $\A_T =
 \tuple{\Sigma,\set{t},Q_T,\iota_T,F_T,\Delta_T}$, with a single input
@@
 following connection between a timed automaton and its corresponding
-first-order alternating automaton.
+first order alternating automaton.
 
@@
   Given a timed automaton $T = \tuple{\Sigma,S,S_0,F,C,E}$, the
-  first-order alternating automaton $\A_T =
+  first order alternating automaton $\A_T =
   \tuple{\Sigma,\set{t},Q_T,\iota_T,F_T,\Delta_T}$ recognizes the
@@
 moreover, that no transition quantifiers are needed to encode timed
-automata as first-order alternating automata.
+automata as first order alternating automata.
 
@@
 
-The translation of register automata to first-order alternating
+The translation of register automata to first order alternating
 automata is quite natural, because registers can be encoded as
@@
 \ldots, y_r)\] Moreover, nothing else is in $\Delta_R$. The connection
-between register automata and first-order alternating automata is
+between register automata and first order alternating automata is
 stated below.
@@
   Given a register automaton $R = \tuple{S,s_0,\vec{u},\rho,\mu,F}$
-  over an infinite alphabet $\Sigma$, the first-order alternating
+  over an infinite alphabet $\Sigma$, the first order alternating
   automaton $\A_R = \tuple{\set{\alpha},Q_R,\iota_R,F_R,\Delta_R}$
@@
 (\S\ref{sec:emptiness}) provides a semi-algorithm. Notice further that
-the encoding of register automata as first-order alternating automata
+the encoding of register automata as first order alternating automata
 uses no transition quantifiers.
@@
 Since thread identifiers can only be compared for equality, the data
-theory of predicate automata is the first-order theory of
+theory of predicate automata is the first order theory of
 equality. Moreover, transition quantifiers are only needed for
@@
 However, the execution semantics of predicate automata differs from
-that of first-order automata with respect to the following detail:
+that of first order automata with respect to the following detail:
 initial configurations and successors of predicate automata are
@@
 by Definitions \ref{def:execution} and \ref{def:accepting}.  We
-believe that the semantics of first-order alternating automata based
+believe that the semantics of first order alternating automata based
 on minimal models is important for its relation to the textbook
@@
   \tuple{\Sigma,\set{x},Q,\iota,F,\Delta}$, let $\A_P$ be the
-  first-order alternating automaton that has the same description as
+  first order alternating automaton that has the same description as
   $P$. Then $\lang{P} = \lang{\A_P}$. 

### 2018-11-06 quantifiers/abstract.tex
ok

--- 2018-10-18 quantifiers/draft.tex
@@
-\documentclass{llncs}
-% \documentclass[10pt,conference,letterpaper,twocolumn]{IEEEtran}
-
-\usepackage[usenames,dvipsnames]{color}
-
-\usepackage{latexsym}
-\usepackage{amsxtra} 
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{pslatex}
-\usepackage{epsfig}
-\usepackage{wrapfig}
-\usepackage{paralist}
-% \usepackage{stmaryrd}
-\usepackage{txfonts}
-\usepackage{framed}
-\usepackage{makecell}
-\usepackage{url}
-\usepackage{tikz}
-% \usepackage{hyperref}
-\usetikzlibrary{automata,positioning, calc}
-\usepackage[inline,shortlabels]{enumitem}
-
-\usepackage{proof}
-
-\usepackage{algorithm}
-\usepackage{algorithmicx}
-\usepackage[noend]{algpseudocode}
-
-\usepackage[draft]{commenting}
-
-\pagestyle{plain}
-
-\include{commands-lncs}
-
-\declareauthor{ri}{Radu}{blue}
-\declareauthor{xx}{Xiao}{red}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\title{First Order Alternation}
-
-\author{Radu Iosif and Xiao Xu}
-\institute{
-  CNRS, Verimag, Universit\'e de Grenoble Alpes \\
-  Email: \{Radu.Iosif,Xiao.Xu\}@univ-grenoble-alpes.fr
-}
-
-\maketitle
-
 \begin{abstract}
+We introduce first order alternating automata, a generalization of
+boolean alternating automata, in which transition rules are described
+by multisorted first order formulae, with states and internal
+variables given by uninterpreted predicate terms. The model is closed
+under union, intersection and complement, and its emptiness problem is
+undecidable, even for the simplest data theory of equality. To cope
+with this limitation, we develop an abstraction refinement
+semi-algorithm based on lazy annotation of the symbolic execution
+paths with interpolants, obtained by applying (i) quantifier
+elimination with witness term generation and (ii) Lyndon interpolation
+in the quantifier-free data theory with uninterpreted predicate
+symbols. This provides a method for checking inclusion of timed and
+finite-memory register automata, and emptiness of quantified predicate
+automata, previously used in the verification of parameterized
+concurrent programs, composed of replicated threads, with a
+shared-memory communication model.
 \end{abstract}
-
-\section{Preliminaries}
-
-For two integers $0 \leq i \leq j$, we denote by $[i,j]$ the set
-$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[0,i]$. We consider two
-sorts $\Data$ and $\Bool$, where $\Data$ is an infinite domain and
-$\Bool = \set{\top,\bot}$ is the set of boolean values true ($\top$)
-and false ($\bot$), respectively. The $\Data$ sort is equipped with
-finitely many function symbols $f : \Data^{\#(f)} \rightarrow \Data$,
-where $\#(f)\geq0$ denotes the number of arguments (arity) of
-$f$. When $\#(f)=0$, we say that $f$ is a constant. A \emph{predicate}
-is a function symbol $p : \Data^{\#(p)} \rightarrow \Bool$, denoting a
-relation of arity $\#(p)$ and we write $\preds$ for the set of
-predicates.
-
-In the following, we shall consider that the interpretation of all
-function symbols $f : \Data^{\#(f)} \rightarrow \Data$ that are not
-predicates is fixed by the interpretation of the $\Data$ sort,
-e.g.\ if $\Data$ is the set of integers $\zed$, the function symbols
-are zero, the successor function and the arithmetic operations of
-addition and multiplication. For simplicity, we further blur the
-notational distinction between function symbols and their
-interpretations.
-
-Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
-variables, ranging over $\Data$. Terms are either constants of sort
-$\Data$, variables or function applications $f(t_1,\ldots,t_{\#(f)})$,
-where $t_1,\ldots,t_{\#(f)}$ are terms. The set of first-order
-formulae is defined by the syntax below:
-\[\phi := t \teq s \mid p(t_1,\ldots,t_{\#(p)}) 
-\mid \neg \phi_1 \mid \phi_1 \wedge \phi_2 \mid \exists x ~.~
-\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms. We write
-$\phi_1 \vee \phi_2$, $\phi_1 \rightarrow \phi_2$ and $\forall x ~.~
-\phi_1$ for $\neg(\neg\phi_1 \wedge \neg\phi_2)$, $\neg\phi_1 \vee
-\phi_2$ and $\neg\exists x ~.~ \neg\phi_1$, respectively. We denote by
-$\fv{}{\phi}$ the set of free variables in $\phi$. The size
-$\len{\phi}$ of a formula $\phi$ is the number of symbols needed to
-write it down.
-
-A \emph{sentence} is a formula $\phi$ in which each variable occurs
-under the scope of a quantifier, i.e. $\fv{}{\phi} = \emptyset$.  A
-formula is \emph{positive} if each predicate symbol occurs under an
-even number of negations and we denote by $\posforms(Q,X)$ the set of
-positive formulae with predicates from the set $Q \subseteq \preds$
-and free variables from the set $X \subseteq \vars$.
-
-A formula is in \emph{prenex form} if it is of the form $\varphi =
-Q_1x_1 \ldots Q_nx_n ~.~ \phi$, where $\phi$ has no quantifiers. In
-this case we call $\phi$ the \emph{matrix} of $\varphi$. Every
-first-order formula can be written in prenex form, by renaming each
-quantified variable to a unique name and moving the quantifiers
-upfront.
-
-%% For a variable $x \in \fv{}{\phi}$ and a term $t$, let $\phi[t/x]$ be
-%% the result of replacing each occurrence of $x$ by $t$. For indexed
-%% sets of terms $T=\set{t_1,\ldots,t_n}$ and variables
-%% $X=\set{x_1,\ldots,x_n}$, we write $\phi[T/X]$ for the formula
-%% obtained by simultaneously replacing each occurrence of $x_i$ with
-%% $t_i$ in $\phi$, for all $i\in[n]$.
-
-An \emph{interpretation} $\I$ maps each predicate $p$ into a set $p^\I
-\subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element of $\Data$
-if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable $x$ into an
-element of $\Data$. Given a term $t$, we denote by $t^\nu$ the value
-obtained by replacing each variable $x$ by the value $\nu(x)$ and
-evaluating each function application. For a formula $\phi$, we define
-the forcing relation $\I,\nu \models \phi$ recursively on the
-structure of $\phi$, as follows:
-\[
-\begin{array}{rcl}
-\I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
-\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I \\
-\I,\nu \models \neg\phi_1 & \iff & \I,\nu \not\models \phi_1 \\
-\I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
-\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data 
-\end{array}
-\]
-where $\nu[x\leftarrow d]$ is the valuation which assigns $d$ to $x$
-and behaves like $\nu$ elsewhere. For a formula $\phi$ and a valuation
-$\nu$, we define $\sem{\phi}_\nu \isdef \set{\I \mid \I,\nu \models
-  \phi}$ and drop the $\nu$ subscript for sentences. A sentence $\phi$
-is \emph{satisfiable} (\emph{unsatisfiable}) if
-$\sem{\phi}\neq\emptyset$ ($\sem{\phi}=\emptyset$). An element of
-$\sem{\phi}$ is called a \emph{model} of $\phi$. A formula $\phi$ is
-\emph{valid} if $\I,\nu \models \phi$ for every interpretation $\I$
-and every valuation $\nu$. For two formulae $\phi$ and $\psi$ we write
-$\phi \models \psi$ for $\sem{\phi} \subseteq \sem{\psi}$, in which
-case we say that $\phi$ \emph{entails} $\psi$.
-
-Interpretations are partially ordered by the pointwise subset order,
-defined as $\I_1 \subseteq \I_2$ if and only if $p^{\I_1} \subseteq
-p^{\I_2}$ for each predicate $p \in \preds$. Given a set $\mathcal{S}$
-of interpretations, a minimal element $\I \in \mathcal{S}$ is an
-interpretation such that for no other interpretation $\I' \in
-\mathcal{S} \setminus \set{\I}$ do we have $\I' \subseteq \I$. For a
-formula $\phi$ and a valuation $\nu$, we denote by $\minsem{\phi}_\nu$
-and $\minsem{\phi}$ the set of minimal interpretations from
-$\sem{\phi}_\nu$ and $\sem{\phi}$, respectively.
-
-\section{First Order Alternating Automata}
-
-Let $\Sigma$ be a finite alphabet $\Sigma$ of \emph{input
-  events}. Given a finite set of variables $X \subseteq \vars$, we
-denote by $X \mapsto \Data$ the set of valuations of the variables $X$
-and $\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly
-infinite set of \emph{data symbols} $(a,\nu)$, where $a$ is an input
-symbol and $\nu$ is a valuation. A \emph{data word} (simply called
-word in the following) is a finite sequence $w=(a_1,\nu_1)(a_2,\nu_2)
-\ldots (a_n,\nu_n)$ of data symbols. Given a word $w$, we denote by
-$\event{w} \isdef a_1\ldots a_n$ its sequence of input events and by
-$\data{w}$ the valuation associating each time-stamped variable
-$\stamp{x}{i}$ the value $\nu_i(x)$, for all $x \in \vars$ and
-$i\in[1,n]$. We denote by $\varepsilon$ the empty sequence, by
-$\Sigma^*$ the set of finite sequences of input events and by
-$\Sigma[X]^*$ the set of data words over the variables $X$.
-
-Formally, a \emph{first-order alternating automaton} is a tuple $\A =
-\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is a finite set of
-input events, $X$ is a finite set of input variables, $Q$ is a finite
-set of predicates denoting control states, $\iota \in
-\posforms(Q,\emptyset)$ is a sentence defining initial configurations,
-$F \subseteq Q$ is the set of predicates denoting final states, and
-$\Delta$ is a set of \emph{transition rules} of the form
-\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
-predicate, $a \in \Sigma$ is an input event and $\psi \in
-\posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive formula,
-where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$. The quantifiers
-occurring in the right-hand side formula of a transition rule are
-referred to as \emph{transition quantifiers}. The \emph{size} of $\A$
-is defined as $\len{\A} = \len{\iota} +
-\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
-\len{\psi}$.
-
-The intuition of a transition rule \(q(y_1,\ldots,y_{\#(q)})
-\arrow{a(X)}{} \psi\) is the following: $a$ is the input event and $X$
-are the input data values that trigger the transition, whereas $q$ and
-$y_1,\ldots,y_{\#(q)}$ are the current control state and data values
-in that state, respectively. Without loss of generality, we consider,
-for each predicate $q \in Q$ and each input event $a \in \Sigma$, at
-most one such rule, as two or more rules can be joined using
-disjunction.
-
-The execution semantics of automata is given in close analogy with the
-case of boolean alternating automata, with transition rules of the
-form $q \arrow{a}{} \phi$, where $q$ is a boolean constant and $\phi$
-a positive boolean combination of such constants. For instance, $q_0
-\arrow{a}{} q_1 \wedge q_2 \vee q_3$ means that the automaton can
-choose to transition in either both $q_1$ and $q_2$ or in $q_3$
-alone. This intuition leads to saying that the steps of the automaton
-are defined by the minimal boolean models of the transition
-formulae. In this case, both $\set{q_1 \leftarrow \top, q_2 \leftarrow
-  \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow \bot, q_2
-  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, however
-$\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
-is a model but is not minimal. The original definition of alternating
-finite-state automata \cite{ChandraKozenStockmeyer81} works around
-this problem by considering boolean valuations (models) instead of
-formulae.
-
-Given a predicate $q \in Q$ and a tuple of data values
-$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
-a \emph{configuration}\footnote{Note that a configuration is not a
-  logical term since data values cannot be written in logic.}. To
-formalize the execution semantics of automata, we relate sets of
-configurations to models of first-order sentences, as follows. Given a
-formula $\phi$ and a valuation of its free variables $\nu$, each
-interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
-configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid q
-  \in Q,~ \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a
-\emph{cube}. For a set $\mathcal{S}$ of interpretations, we define
-$\cube{\mathcal{S}} \isdef \set{\cube{\I} \mid \I \in
-  \mathcal{S}}$.
-
-\begin{definition}\label{def:execution}
-Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$ and a cube $c$, an
-\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$, starting with $c$, is
-a (possibly infinite) forest $\T = \set{T_1,T_2,\ldots}$, where each
-$T_i$ is a tree labeled with configurations, such that:
-\begin{compactenum}
-\item\label{it1:execution} $c = \set{T(\epsilon) \mid T \in \T}$ is the set of
-  configurations labeling the roots of $T_1,T_2,\ldots$ and
-%
-\item\label{it2:execution} if $q(d_1,\ldots,d_{\#(q)})$ labels a node on the level $j \in
-  [n-1]$ in $T_i$, then the labels of its children form a cube from
-  $\cube{\minsem{\psi}_{\eta}}$, where $\eta = \nu_{j+1}[y_1
-    \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$ and
-  \(q(y_1,\ldots,y_{\#(q)}) \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is
-  a transition rule of $\A$.
-\end{compactenum}
-\end{definition}
-
-\begin{definition}\label{def:accepting}
-An execution $\T$ over $w$, starting with $c$, is \emph{accepting} if
-and only if \begin{compactitem}
-\item\label{it1:accepting} all paths in $\T$ have the same length $n$,
-  and
-  %
-\item\label{it2:accepting} the frontier of each tree $T \in \T$ is
-  labeled with \emph{final configurations} $q(d_1,\ldots,d_{\#(q)})$,
-  where $q \in F$.
-\end{compactitem} 
-If $\A$ has an accepting execution over $w$ starting with a cube $c
-\in \cube{\minsem{\iota}}$, then $\A$ \emph{accepts} $w$ and let
-$\lang{\A}$ be the set of words accepted by $\A$.
-\end{definition}
-In this paper, we address the following questions: \begin{compactenum}
-\item \emph{boolean closure}: given automata $\A_i =
-  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
-  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
-  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
-  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
-%
-\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
-  \emptyset$?
-\end{compactenum}
-
-\subsection{Symbolic Execution}
-
-In the upcoming developments it is sometimes more convenient to work
-with logical formulae defining executions of automata, than with
-low-level execution forests. For this reason, we first introduce
-\emph{path formulae} $\pathform{\alpha}$, which are formulae defining
-the executions of an automaton, over words that share a given sequence
-$\alpha$ of input events.  Second, we restrict a path formula
-$\pathform{\alpha}$ to an \emph{acceptance formula}
-$\accform{\alpha}$, which defines only accepting executions over words
-that share a given input sequence. Otherwise stated,
-$\accform{\alpha}$ is satisfiable if and only if the automaton accepts
-a word $w$ such that $\event{w} = \alpha$.
-
-Let $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the
-rest of this section. For any $i \in \nat$, we denote by $\stamp{Q}{i}
-= \set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} =
-\set{\stamp{x}{i} \mid x \in X}$ the sets of time-stamped predicates
-and variables, respectively. As a shorthand, we write $\stamp{Q}{\leq
-  n}$ (resp.  $\stamp{X}{\leq n}$) for the set $\set{\stamp{q}{i} \mid
-  q \in Q, i \in [n]}$ (resp. $\set{\stamp{x}{i} \mid x \in X, i \in
-  [n]}$). For a formula $\psi$ and $i \in \nat$, we define
-$\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the
-formula in which all input variables and state predicates (and only
-those symbols) are replaced by their time-stamped counterparts. As a
-shorthand, we shall write $q(\vec{y})$ for $q(y_1,\ldots,y_{\#(q)})$,
-when no confusion arises.
-
-Given a sequence of input events $\alpha = a_1 \ldots a_n$, the
-\emph{path formula} of $\alpha$ is defined as:
-\begin{equation}\label{eq:pathform}
-\pathform{\alpha} \isdef \stamp{\iota}{0} \wedge 
-\bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{}
-  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-\stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i}
-\end{equation}
-The automaton $\A$, to which $\pathform{\alpha}$ refers, will always
-be clear from the context. To formalize the relation between the
-low-level configuration-based execution semantics and the symbolic
-path formulae, consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any
-execution forest $\T$ of $\A$ over $w$ is associated an interpretation
-$\I_{\T}$ of the set of time-stamped predicates $\stamp{Q}{\leq n}$,
-defined as:
-\[\I_{\T}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
-  q(d_1,\ldots,d_{\#(q)}) \text{ labels a node on level $i$ in $\T$}} 
-,~\forall q\in Q ~\forall i \in [n]\]
-
-\begin{lemma}\label{lemma:path-formula}
-  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
-  word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have 
-  \(\minsem{\pathform{\event{w}}}_{\data{w}} = \set{\I_{\T} \mid \T
-    \text{ is an execution of $\A$ over $w$}}\). 
-\end{lemma}
-\proof{ ``$\subseteq$'' Let $\I$ be a minimal interpretation such that
-  $\I,\data{w} \models \pathform{\event{w}}$. We show that there exists
-  an execution $\T$ of $\A$ over $w$ such that $\I = \I_\T$, by
-  induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
-  $\pathform{\event{w}} = \stamp{\iota}{0}$. Because $\iota$ is a
-  sentence, the valuation $\data{w}$ is not important in $\I,\data{w}
-  \models \stamp{\iota}{0}$ and, moreover, since $\I$ is minimal, we
-  have $\I \in \minsem{\stamp{\iota}{0}}$. We define the
-  interpretation $\J(q)=\I(\stamp{q}{0})$, for all $q \in Q$. Then
-  $\cube{\J}$ is an execution of $\A$ over $\epsilon$ and $\I =
-  \I_{\cube{\J}}$ is immediate. For the inductive case $n>0$, we
-  assume that $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\J$ be the
-  interpretation defined as $\I$ for all $\stamp{q}{i}$, with $q \in
-  Q$ and $i \in [n-1]$, and $\emptyset$ everywhere else. Then
-  $\J,u_\Data \models \pathform{u_\Sigma}$ and $\J$ is moreover
-  minimal. By the induction hypothesis, there exists an execution $\G$
-  of $\A$ over $u$, such that $\J = \I_\G$. Consider a leaf of a tree
-  $T \in \G$, labeled with a configuration $q(d_1,\ldots,d_{\#(q)})$
-  and let $\forall y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$ be the
-  subformula of $\pathform{\event{w}}$ corresponding to the
-  application(s) of the transition rule $q(\vec{y}) \arrow{a_n}{}
-  \psi$ at the $(n-1)$-th step. Let $\nu = \data{w}[y_1 \leftarrow
-    d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$. Because $\I, \data{w}
-  \models \forall y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$, we have $\I
-  \in \sem{\stamp{\psi}{n}}_{\nu}$ and let $\K$ be one of the minimal
-  interpretations such that $\K \subseteq \I$ and $\K \in
-  \sem{\stamp{\psi}{n}}_{\nu}$. It is not hard to see that $\K$ exists
-  and is unique, otherwise we could take the pointwise intersection of
-  two or more such interpretations. We define the interpretation
-  $\overline{\K}(q) = \overline{\K}(\stamp{q}{n})$ for all $q \in
-  Q$. We have that $\overline{\K} \in \minsem{\psi}_{\nu}$ --- if
-  $\overline{\K}$ was not minimal, $\K$ was not minimal to start with,
-  contradiction. Then we extend the execution $\G$ by appending to
-  each node labeled with a configuration $q(d_1, \ldots, d_{\#(q)})$
-  the cube $\cube{\overline{\K}}$. By repeating this step for all
-  leaves of a tree in $\G$, we obtain an execution of $\A$ over $w$.
-
-  ``$\supseteq$'' Let $\T$ be an execution of $\A$ over $w$. We show
-  that $\I_{\T}$ is a minimal interpretation such that $\I_{\T},
-  \data{w} \models \pathform{\event{w}}$, by induction on $n \geq
-  0$. For $n=0$, $\T$ is a cube from $\cube{\minsem{\iota}}$, by
-  definition. Then $\I_{\T} \models \stamp{\iota}{0}$ and moreover, it
-  is a minimal such interpretation. For the inductive case $n > 0$,
-  let $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\G$ be the
-  restriction of $\T$ to $u$. Consequently, $\I_\G$ is the restriction
-  of $\I_\T$ to $\stamp{Q}{\leq n-1}$. By the inductive hypothesis,
-  $\I_{\G}$ is a minimal interpretation such that \(\I_{\G}, u_{\Data}
-  \models \pathform{u_\Sigma}\). Since $\I_{\T}(\stamp{q}{n}) =
-  \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid q(d_1,\ldots,d_{\#(q)})
-    \text{ labels a node on the $n$-th level in $\T$}}$, we have $\I_{\T},
-  \data{w} \models \varphi$, for each subformula $\varphi = \forall y_1
-  \ldots \forall y_{\#(q)} ~.~ \stamp{q}{n-1}(\vec{y})
-  \rightarrow \stamp{\psi}{n}$ of $\pathform{\event{w}}$, by the
-  execution semantics of $\A$. This is the case because the children
-  of each node labeled with $q(d_1,\ldots,d_{\#(q)})$ on the
-  $(n-1)$-th level of $\T$ form a cube from
-  $\cube{\minsem{\psi}_\nu}$, where $\nu$ is a valuation that assigns
-  each $y_i$ the value $d_i$ and behaves like $\data{w}$,
-  otherwise. Now supppose, for a contradiction, that $\I_{\T}$ is not
-  minimal and let $\J \subsetneq \I_{\T}$ be an interpretation such
-  that $\J,\data{w} \models \pathform{\event{w}}$. First, we show that
-  the restriction $\J'$ of $\J$ to $\bigcup_{i=0}^{n-1} \stamp{Q}{i}$
-  must coincide with $\I_{\G}$. Assuming this is not the case,
-  i.e.\ $\J' \subsetneq \I_{\G}$, contradicts the minimality of
-  $\I_{\G}$. Then the only possibility is that $\J(\stamp{q}{n})
-  \subsetneq \I_{\T}(\stamp{q}{n})$, for some $q \in Q$. Let
-  $p_1(y_1,\ldots,y_{\#(p_1)}) \arrow{a_n}{} \psi_1, \ldots,
-  p_k(y_1,\ldots,y_{\#(p_k)}) \arrow{a_n}{} \psi_k$ be the set of
-  transition rules in which the predicate symbol $q$ occurs on the
-  right-hand side. Then it must be the case that, for some node on the
-  $(n-1)$-th level of $\G$, labeled with a configuration
-  $p_i(d_1,\ldots,d_{\#(p_i)})$, the set of children does not form a
-  minimal cube from $\cube{\minsem{\stamp{\psi_i}{n}}}$, which
-  contradicts the execution semantics of $\A$. \qed}
-
-Next, we give a logical characterization of acceptance, relative to a
-given sequence of input events $\alpha \in \Sigma^*$. To this end, we
-constrain the path formula $\pathform{\alpha}$ by requiring that only
-final states of $\A$ occur on the last level of the execution.  The
-result is the \emph{acceptance formula} for $\alpha$:
-\begin{equation}\label{eq:accform}
-\accform{\alpha} \isdef \pathform{\alpha} \wedge \bigwedge_{q
-  \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-\stamp{q}{n}(\vec{y}) \rightarrow \bot
-\end{equation}
-The top-level universal quantifiers from a subformula $\forall y_1
-\ldots \forall y_{\#(q)} ~.~ \stamp{q}{i}(\vec{y}) \rightarrow \psi$
-of $\accform{\alpha}$ will be referred to as \emph{path quantifiers},
-in the following. Notice that path quantifiers are distinct from the
-transition quantifiers that occur within a formula $\psi$ of a
-transition rule $q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi$ of $\A$.
-
-The acceptance formula $\accform{\A}$ is false in every interpretation
-of the predicates that assigns a non-empty set to a non-final
-predicate occurring on the last level in the execution forest. The
-relation between the words accepted by $\A$ and the acceptance formula
-above, is formally captured by the following lemma:
-
-\begin{lemma}\label{lemma:acceptance}
-  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for
-  every word $w \in \Sigma[X]^*$, the following are
-  equivalent: \begin{compactenum}
-  \item\label{it1:lemma:acceptance} there exists an interpretation
-    $\I$ such that $\I,\data{w} \models \accform{\event{w}}$, 
-    %
-  \item\label{it2:lemma:acceptance} $w \in \lang{\A}$. 
-  \end{compactenum}
-\end{lemma}
-\proof{ ``$(\ref{it1:lemma:acceptance}) \Rightarrow
-  (\ref{it2:lemma:acceptance})$'' Let $\I$ be an interpretation such
-  that $\I,\data{w} \models \accform{\event{w}}$. By Lemma
-  \ref{lemma:path-formula}, $\A$ has an execution $\T$ over $w$ such
-  that $\I = \I_{\T}$. To prove that $\T$ is accepting, we show
-  that \begin{inparaenum}[(i)]
-    \item\label{it1:acceptance} all paths in $\T$ have length $n$ and
-      that
-    \item\label{it2:acceptance} the frontier of $\T$ is labeled with
-      final configurations only. \end{inparaenum} First, assume that
-  (\ref{it1:acceptance}) there exists a path in $\T$ of length $0 \leq
-  m < n$. Then there exists a node on the $m$-th level, labeled with
-  some configuration $q(d_1,\ldots,q_{\#(q)})$, that has no
-  children. By the definition of the execution semantics of $\A$, we
-  have $\cube{\minsem{\psi}_{\eta}} = \emptyset$, where
-  $q(\vec{y}) \arrow{a_{m+1}(X)}{} \psi$ is the transition
-  rule of $\A$ that applies for $q$ and $a_{m+1}$ and $\eta=
-  \data{w}[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow
-    d_{\#(q)}]$. Hence $\sem{\psi}_{\eta} = \emptyset$, and because
-  $\I,\data{w} \models \accform{\alpha}$, we obtain that $\I,\eta \models
-  q(\vec{y}) \rightarrow \stamp{\psi}{m+1}$, thus
-  $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$. However, this
-  contradicts the fact that $\I=\I_{\T}$ and that
-  $q(d_1,\ldots,d_{\#(q)})$ labels a node of $\T$. Second, assume that
-  (\ref{it2:acceptance}), there exists a frontier node of $\T$ labeled
-  with a configuration $q(d_1,\ldots,d_{\#(q)})$ such that $q \in Q
-  \setminus F$. Since $\I,\data{w} \models \forall y_1 \ldots \forall
-  y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot$, by a
-  similar reasoning as in the above case, we obtain that
-  $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$, contradiction.
-
-  ``$(\ref{it2:lemma:acceptance}) \Rightarrow
-  (\ref{it1:lemma:acceptance})$'' Let $\T$ be an accepting execution
-  of $\A$ over $w$. We prove that $\I_\T,\data{w} \models
-  \accform{\event{w}}$. By Lemma \ref{lemma:path-formula}, we obtain
-  $\I_\T,\data{w} \models \pathform{\event{w}}$. Since every path in
-  $\T$ is of length $n$ and all nodes on the $n$-th level of $\T$ are
-  labeled by final configurations, we obtain that $\I_\T,\data{w}
-  \models \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
-  y_{\#(q)} ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot$,
-  trivially.  \qed}
-
-As an immediate consequence, one can decide whether $\A$ accepts some
-word $w$ with a given input sequence $\event{w}=\alpha$, by checking
-whether $\accform{\alpha}$ is satisfiable. However, unlike
-non-alternating infinite-state models of computation, such as counter
-automata (nondeterministic programs with integer variables), the
-satisfiability query for an acceptance (path) formula falls outside of
-known decidable theories, supported by standard SMT solvers. There are
-basically two reasons for this, namely \begin{inparaenum}[(i)]
-\item the presence of predicate symbols, and
-\item the non-trivial alternation of quantifiers.
-\end{inparaenum}
-To understand this point, consider for example, the decidable theory
-of Presburger arithmetic \cite{Presburger29}. Adding even only one
-monadic predicate symbol to it yields undecidability in the presence
-of non-trivial quantifier alternation \cite{Halpern91}. However the
-quantifier-free fragment of Presburger arithmetic extended with
-predicate symbols can be shown to be decidable, using a Nelson-Oppen
-style congruence closure argument \cite{NelsonOppen80}.
-
-To tackle this problem, we start from the observation that acceptance
-formulae have a particular form, which allows the elimination of path
-quantifiers and of predicates, by a couple of
-satisfiability-preserving transformations. The result of applying
-these transformations is a formula with no predicate symbols, whose
-only quantifiers are those introduced by the transition rules of the
-automaton, referred to as transition quantifiers. We shall further
-assume (\S\ref{sec:emptiness}) that the first-order theory of the data
-sort $\Data$ has quantifier elimination, which allows to effectively
-decide the satisfiability of such formulae.
-
-For the time being, let us formally define the elimination of
-transition quantifiers and predicates, respectively. Consider a given
-sequence of input events $\alpha = a_1 \ldots a_n$ and denote by
-$\alpha_i$ the prefix $a_1 \ldots a_i$ of $\alpha$, for $i \in [n]$,
-where $\alpha_0=\epsilon$.
-
-\begin{definition}\label{def:quantform}
-  Let $\quantformn{\alpha_0}, \ldots, \quantformn{\alpha_n}$ be the
-  sequence of formulae defined by $\quantformn{\alpha_0} \isdef
-  \stamp{\iota}{0}$ and, for all $i \in [1,n]$:
-  \[\quantformn{\alpha_i} \isdef \quantformn{\alpha_{i-1}} \wedge
-  \bigwedge_{\begin{array}{l}
-      \scriptstyle{\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_{i-1}}} \\
-      \scriptstyle{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta} 
-  \end{array}} \stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow 
-    \stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]
-      \]  We write
-    $\quantform{\alpha}$ for the prenex normal form of the formula: 
-      \[\quantformn{\alpha_n} \wedge \bigwedge_{\begin{array}{l}
-          \scriptstyle{\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_n}} \\
-          \scriptstyle{q \in Q \setminus F}
-      \end{array}} \stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot \enspace.\]
-\end{definition}
-Observe that $\quantform{\alpha}$ contains no path quantifiers, as
-required. On the other hand, the scope of the transition quantifiers
-in $\quantform{\alpha}$ exceeds the right-hand side formulae from the
-transition rules, as shown by the following example.
-
-\begin{example}\label{ex:quant-elim}
-  Consider the automaton $\A = \tuple{\set{a_1,a_2}, \set{x},
-    \set{q,q_f}, \iota, \set{q_f}, \Delta}$, where:
-  \[\begin{array}{rcl}
-  \iota & = & \exists z ~.~ z \geq 0 \wedge q(z) \\
-  \Delta & = & \set{
-    q(y) \arrow{a_1(x)}{} x\geq 0 \wedge \forall z ~.~ z \leq y \rightarrow q(x+z),~
-    q(y) \arrow{a_2(x)}{} y<0 \wedge q_f(x+y)
-  }
-  \end{array}\]
-  For the input event sequence $\alpha = a_1a_2$, the acceptance
-  formula is:
-  \[\begin{array}{rcl}
-  \accform{\alpha} & = & \exists z ~.~ z \geq 0 \wedge \stamp{q}{0}(z) ~\wedge \\
-  && \forall y ~.~ \stamp{q}{0}(y) \rightarrow [\stamp{x}{1}\geq0 \wedge \forall z ~.~ z\geq y 
-    \rightarrow \stamp{q}{1}(\stamp{x}{1}+z)] ~\wedge \\
-  && \forall y ~.~ \stamp{q}{1}(y) \rightarrow [y<0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+y)]
-  \end{array}\]
-  The result of eliminating the path quantifiers, in prenex normal form, is shown below:
-  \[\begin{array}{rcl}
-  \quantform{\alpha} & = & \exists z_1\forall z_2 ~.~ z_1 \geq 0 \wedge \stamp{q}{0}(z_1) ~\wedge \\
-  && [\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge (z_2 \geq z_1 \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_2))] ~\wedge \\
-  && [\stamp{q}{1}(\stamp{x}{1}+z_2) \rightarrow \stamp{x}{1}+z_2 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_2)] 
-  \enspace\enspace\hfill\blacksquare
-  \end{array}\]
-\end{example}
-
-The next lemma establishes a formal relation between the
-satisfiability of an acceptance formula $\accform{\alpha}$ and that of
-the formula $\quantform{\alpha}$, obtained by eliminating the path
-quantifiers from $\accform{\alpha}$.
-
-\begin{lemma}\label{lemma:quant}
-  For any input event sequence $\alpha=a_1\ldots a_n$ and each
-  valuation $\nu : \stamp{X}{\leq n} \rightarrow \Data$, the following
-  hold: \begin{compactenum}
-  \item\label{it1:quant} for all interpretations $\I$, if $\I,\nu
-    \models \accform{\alpha}$ then $\I,\nu \models
-    \quantform{\alpha}$. 
-    %
-  \item\label{it2:quant} if there exists an interpretation $\I$ such
-    that $\I,\nu \models \quantform{\alpha}$ then there exists an
-    interpretation $\J \subseteq \I$ such that $\J,\nu \models
-    \accform{\alpha}$. 
-  \end{compactenum}
-\end{lemma}
-\proof{(\ref{it1:quant}) Trivial, since every subformula
-  $q(t_1,\ldots,t_{\#(q)}) \rightarrow
-  \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ of $\quantform{\alpha}$ is
-  entailed by a subformula $\forall y_1 \ldots \forall y_{\#(q)} ~.~
-  q(y_1, \ldots, y_{\#(q)}) \rightarrow \psi$ of $\accform{\alpha}$.
-
-  \noindent(\ref{it2:quant}) By repeated applications of the following fact:
-
-  \begin{fact}
-    Given formulae $\phi$ and $\psi$, such that no predicate atom with
-    predicate symbol $q$ occurs in $\psi(y_1,\ldots,y_{\#(q)})$, for
-    each valuation $\nu$, if there exists an interpretation $\I$ such
-    that $\I,\nu \models \phi \wedge
-    \bigwedge_{q(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \phi}
-    q(t_1,\ldots,t_{\#(q)}) \rightarrow
-    \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ then there exists a
-    valuation $\J$ such that $\J(q) \subseteq \I(q)$ and $\J(q') =
-    \I(q')$ for all $q' \in Q \setminus \set{q}$ and $\J,\nu \models
-    \phi \wedge \forall y_1 \ldots \forall y_{\#(q)} ~.~
-    q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$.
-  \end{fact}
-  \proof{Assume w.l.o.g. that $\phi$ is quantifier free. The proof can
-    be easily generalized to the case $\phi$ has quantifiers. Let
-    $\J(q) = \set{\langle t_1^\nu, \ldots, t_{\#(q)}^\nu \rangle \in
-      \I(q) \mid q(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \phi}$ and
-    $\J(q') = \I(q')$ for all $q' \in Q \setminus \set{q}$. Since
-    $\I,\nu \models \phi$, we obtain that also $\J,\nu \models \phi$
-    because the tuples of values in $\I(q) \setminus \J(q)$ are not
-    interpretations of terms that occur within subformulae
-    $q(t_1,\ldots,t_{\#(q)})$ of $\phi$. Moreover,
-    $\bigwedge_{q(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \phi}
-    q(t_1,\ldots,t_{\#(q)}) \rightarrow
-    \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ and $\forall y_1 \ldots
-    \forall y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$
-    are equivalent under $\J$, thus $\J,\nu \models \forall y_{\#(q)}
-    ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$, as required. \qed}
-
-  \noindent This concludes the proof. \qed}
-
-We proceed with the elimination of predicate atoms from
-$\quantform{\alpha}$, defined below.
-
-\begin{definition}\label{def:substform}
-  Let $\substformn{\alpha_0}, \ldots, \substformn{\alpha_n}$ be the
-  sequence of formulae defined by $\substformn{\alpha_0} \isdef
-  \stamp{\iota}{0}$ and, for all $i \in [1,n]$,
-  $\substformn{\alpha_i}$ is obtained by replacing each occurrence of
-  a predicate atom $\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)})$ in
-  $\substformn{\alpha_{i-1}}$ with the formula
-  $\stamp{\psi}{i}[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$, where
-  $q(\vec{y}) \arrow{a_{i}(X)}{} \psi \in \Delta$. We write
-  $\substform{\alpha}$ for the formula obtained by replacing, in
-  $\substformn{\alpha}$, each occurrence of a predicate
-  $\stamp{q}{n}$, such that $q \in Q \setminus F$ (resp. $q \in F$),
-  by $\bot$ (resp. $\top$).
-\end{definition}
-
-\begin{example}[Contd. from Example \ref{ex:quant-elim}]\label{ex:pred-elim}
-  The result of the elimination of predicate atoms from the acceptance
-  formula in Example \ref{ex:quant-elim} is shown below:
-  \[\substform{\alpha} = 
-  \exists z_1 \forall z_2 ~.~ z_1 \geq 0 \wedge 
-          [\stamp{x}{1} \geq 0 \wedge (z_2 \geq z_1 \rightarrow
-            \stamp{x}{1}+z_2 < 0)]\] Since this formula is
-          unsatisfiable, by Lemma \ref{lemma:quant-pred-acceptance}
-          below, no word $w$ with input event sequence $\event{w} =
-          a_1a_2$ is accepted by the automaton $\A$ from Example
-          \ref{ex:quant-elim}. \hfill$\blacksquare$
-\end{example}
-
-At this point, we prove the formal relation between the satisfiability
-of the formulae $\quantform{\alpha}$ and $\substform{\alpha}$. Since
-there are no occurrences of predicates in $\substform{\alpha}$, for
-each valuation $\nu : \stamp{X}{\leq n} \rightarrow \Data$, there
-exists an interpretation $\I$ such that $\I,\nu \models
-\substform{\alpha}$ if and only if $\J,\nu \models
-\substform{\alpha}$, for every interpretation $\J$. In this case we
-omit $\I$ and simply write $\nu \models \substform{\alpha}$.
-
-\begin{lemma}\label{lemma:subst}
-  For any input event sequence $\alpha=a_1\ldots a_n$ and each
-  valuation $\nu : \stamp{X}{\leq n} \rightarrow \Data$, there exists
-  a valuation $\I$ such that $\I,\nu \models \quantform{\alpha}$ if
-  and only if $\nu \models \substform{\alpha}$. 
-\end{lemma}
-\proof{ By induction on $n\geq0$. The base case $n=0$ is trivial,
-  since $\quantform{A} = \substform{A} = \stamp{\iota}{0}$. For the
-  induction step, we rely on the following fact:
-
-  \begin{fact}\label{fact:subst}
-    Given formulae $\phi$ and $\psi$, such that $\phi$ is positive
-    $q(t_1,\ldots,t_{\#(q)})$ is the only one occurrence of the
-    predicate symbol $q$ in $\phi$ and no predicate atom with
-    predicate symbol $q$ occurs in $\psi(y_1,\ldots,y_{\#(q)})$, for
-    each interpretation $\I$ and each valuation $\nu$, we have:
-    \[\begin{array}{l} 
-    \I,\nu \models \phi \wedge q(t_1,\ldots,t_{\#(q)}) \rightarrow
-    \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}] \iff \\ \nu \models
-    \phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]\enspace.
-    \end{array}\]
-  \end{fact}
-  \proof{ We assume w.l.o.g. that $\phi$ is quantifier-free. The proof
-    can be easily generalized to the case $\phi$ has quantifiers.
-
-    \noindent''$\Rightarrow$'' 
-    We distinguish two cases: \begin{compactitem}
-    \item if $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \in \I(q)$ then
-      $\I,\nu \models \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$. Since
-      $\phi$ is positive, replacing $q(t_1,\ldots,t_{\#(q)})$ with
-      $\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ does not change the
-      truth value of $\phi$ under $\nu$, thus $\nu \models
-      \phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$.
-      %
-    \item else, $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \not\in
-      \I(q)$, thus $\nu \models
-      \phi[\bot/q(t_1,\ldots,t_{\#(q)})]$. Since $\phi$ is positive
-      and $\bot$ entails $\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$,
-      we obtain $\nu \models
-      \phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$
-      by monotonicity. 
-    \end{compactitem}
-
-    \noindent``$\Leftarrow$'' Let $\I$ be any interpretation such that
-    $\I(q) = \set{\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \mid \nu
-      \models \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]}$. We
-    distinguish two cases: \begin{compactitem}
-      \item if $\I(q) \not \emptyset$ then $\I,\nu \models
-        q(t_1,\ldots,t_{\#(q)})$ and $\nu \models \psi[t_1/y_1,
-          \ldots, t_{\#(q)}/y_{\#(q)}]$. Thus replacing $\psi[t_1/y_1,
-          \ldots, t_{\#(q)}/y_{\#(q)}]$ by $q(t_1,\ldots,t_{\#(q)})$
-        does not change the truth value of $\phi$ under $\I$ and
-        $\nu$, and we obtain $\I,\nu \models \phi$. Moreover, $\I,\nu
-        \models \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ implies
-        $\I,\nu \models q(t_1,\ldots,t_{\#(q)}) \rightarrow
-        \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$.
-        %
-      \item else $\I(q) = \emptyset$, hence $\nu \not\models
-        \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$, thus $\nu \models
-        \phi[\bot/q(t_1,\ldots,t_{\#(q)})]$. Because $\phi$ is
-        positive, we obtain $\I,\nu \models \phi$ by monotonicity. But
-        $\I,\nu \models q(t_1,\ldots,t_{\#(q)}) \rightarrow
-        \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ trivially, because
-        $\I,\nu \not\models q(t_1,\ldots,t_{\#(q)})$.\qed
-    \end{compactitem}}
-This concludes the proof. \qed}
-
-%% \begin{proposition}\label{prop:subst}
-%%   Let $\phi$ be a positive formula and let $q(t_1,\ldots,t_{\#(q)})$
-%%   be the only predicate atom occurring within $\phi$. Then, for every
-%%   valuation $\nu$ of $\fv{}{\phi}$ and every formula $\psi$ with no
-%%   predicate occurrences, the following are
-%%   equivalent: \begin{compactenum}
-%%   \item\label{it1:prop:subst} $\nu \models \phi[\psi[t_1/y_1, \ldots,
-%%       t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$, 
-%%     %
-%%   \item\label{it2:prop:subst} $\I,\nu \models \phi \wedge \forall y_1
-%%     \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow
-%%     \psi$, for some interpretation $\I$.
-%%   \end{compactenum}
-%%   where $y_1,\ldots,y_{\#(q)} \not\in \fv{}{\phi}$. 
-%% \end{proposition}
-%% \proof{ We give the proof in the case $\phi$ is quantifier-free. The
-%%   generalization to a non-empty quantifier prefix is straightforward.
-%%   ``$(\ref{it1:prop:subst}) \Rightarrow (\ref{it2:prop:subst})$'' We
-%%   distinguish two cases:
-%%   \begin{compactitem}
-%%   \item if $\nu \models \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$
-%%     then let $\I(q) = \set{\tuple{t_1^\nu,\ldots,t_{\#(q)}^\nu}}$ and
-%%     $\emptyset$ everywhere else. We have $\I,\nu \models
-%%     q(t_1,\ldots,t_{\#(q)})$ and, since $\phi$ is positive, replacing
-%%     $\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ with
-%%     $q(t_1,\ldots,t_{\#(q)})$ in $\phi$ does not change its
-%%     satisfiability under $\I,\nu$, hence $\I,\nu \models
-%%     \phi$. Moreover, $\nu \models \psi[t_1/y_1, \ldots,
-%%       t_{\#(q)}/y_{\#(q)}]$ thus $\I,\nu \models \forall y_1 \ldots
-%%     \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \psi$.
-%%     %
-%%   \item else $\nu \not\models \psi[t_1/y_1, \ldots,
-%%     t_{\#(q)}/y_{\#(q)}]$ and let $\I$ be the interpretation assigning
-%%     each predicate symbol with $\emptyset$. Then replacing
-%%     $\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ with
-%%     $q(t_1,\ldots,t_{\#(q)})$ in $\phi$ does not change its
-%%     satisfiability under $\I,\nu$. Moreover, since $\I$ maps $q$ to
-%%     $\emptyset$, we have $\I,\nu \models \forall y_1 \ldots \forall
-%%     y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \psi$, trivially.
-%%   \end{compactitem}
-
-%%   ``$(\ref{it2:prop:subst}) \Rightarrow (\ref{it1:prop:subst})$'' We
-%%   distinguish two cases: \begin{compactitem}
-%%   \item if $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \in \I(q)$, since
-%%     $\I,\nu \models \forall y_1 \ldots \forall y_{\#(q)} ~.~
-%%     q(\vec{y}) \rightarrow \psi$, we have $\nu \models
-%%     \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$. Since $\phi$ is
-%%     positive, replacing $q(t_1,\ldots,t_{\#(q)})$ by $\psi$ in $\phi$
-%%     does not change its satisfiability under $\nu$, hence $\nu \models
-%%     \phi[\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]/q(t_1, \ldots,
-%%       t_{\#(q)})]$.
-%%     %
-%%   \item else $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \not\in \I(q)$,
-%%     that is $\I,\nu \not\models q(t_1,\ldots,t_{\#(q)})$. Then $\nu
-%%     \models \phi[\bot/q(t_1,\ldots,t_{\#(q)})]$ and we obtain $\nu
-%%     \models \phi[\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]/q(t_1,
-%%       \ldots, t_{\#(q)})]$, because $\phi$ is positive, thus
-%%     $\phi[\bot/q(t_1,\ldots,t_{\#(q)})] \models \phi[\psi[t_1/y_1,
-%%         \ldots, t_{\#(q)}/y_{\#(q)}]/q(t_1, \ldots, t_{\#(q)})]$. \qed
-%%   \end{compactitem}}
-
-Finally, we define the acceptance of a word with a given input event
-sequence by means of a formula in which no predicate atom occurs. As
-previously discussed, several decidable theories, such as Presburger
-arithmetic, become undecidable if predicate atoms are added to
-them. Therefore, the result below makes a step towards deciding
-whether the automaton accepts a word with a given input sequence, by
-reducing this problem to the satisfiability of a quantified formula
-without predicates.
-
-\begin{lemma}\label{lemma:quant-pred-acceptance}
-  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for
-  every word $w \in \Sigma[X]^*$, we have $\data{w} \models
-  \substform{\event{w}}$ if and only if $w \in \lang{\A}$.
-\end{lemma}
-\proof{ By Lemma \ref{lemma:acceptance}, $w \in \lang{\A}$ if and only
-  if $\I,\data{w} \models \accform{\event{w}}$, for some
-  interpretation $\I$. By Lemma \ref{lemma:quant}, there exists an
-  interpretation $\I$ such that $\I,\data{w} \models
-  \accform{\event{w}}$ if and only if there exists an interpretation
-  $\J$ such that $\J,\nu \models \quantform{\event{w}}$. By Lemma
-  \ref{lemma:subst}, there exists an interpretation $\J$ such that
-  $\J,\nu \models \quantform{\event{w}}$ if and only if $\nu \models
-  \substform{\event{w}}$. \qed}
-
-\subsection{Closure Properties}
-\label{sec:closure}
-
-Given a positive formula $\phi$, we define the \emph{dual} formula
-$\dual{\phi}$ recursively as follows:
-\[\begin{array}{rclcrclcrcl}
-\dual{(\phi_1 \vee \phi_2)} & = & \dual{\phi_1} \wedge \dual{\phi_2} && 
-\dual{(\phi_1 \wedge \phi_2)} & = & \dual{\phi_1} \vee \dual{\phi_2} &&
-\dual{(t \teq s)} & = & \neg(t \teq s) \\
-\dual{(\exists x ~.~ \phi_1)} & = & \forall x ~.~ \dual{\phi_1} && 
-\dual{(\forall x ~.~ \phi_1)} & = & \exists x ~.~ \dual{\phi_1} && 
-\dual{(\neg(t \teq s))} & = & t \teq s \\
-\dual{(q(x_1,\ldots,x_{\#(q)}))} & = & q(x_1,\ldots,x_{\#(q)})
-\end{array}\]
-Observe that, because predicate atoms do not occur negated in $\phi$,
-there is no need to define dualization for formulae of the form $\neg
-q(x_1,\ldots,x_{\#(q)})$. The following theorem shows closure of
-automata under all boolean operations:
-
-%% The dualization operation has the following property:
-%% \begin{lemma}\label{lemma:dual}
-%%   Given a formula $\phi$ and an automaton
-%%   $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$, we have the equivalence
-%%   $\Accept{F}{\phi} \equiv \neg(\dual{\Accept{Q\setminus F}{\phi}})$. 
-%% \end{lemma}
-%% \proof{ By straightforward induction on the structure of $\phi$. \qed}
-
-\begin{theorem}\label{thm:closure}
-  Given automata $\A_i = \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$,
-  for $i=1,2$, such that $Q_1 \cap Q_2 = \emptyset$, the following hold: 
-  \begin{compactenum}
-    \item\label{it1:thm:closure} $\lang{\A_\cap} = \lang{\A_1} \cap \lang{\A_2}$, where
-      $\A_\cap = \tuple{\Sigma,X,Q_1 \cup Q_2, \iota_1 \wedge \iota_2,
-      F_1 \cup F_2, \Delta_1 \cup \Delta_2}$, 
-      %
-    \item\label{it2:thm:closure} $\lang{\overline{\A_i}} = \Sigma[X]^*
-      \setminus \lang{\A_i}$, where $\overline{\A_i} =
-      \tuple{\Sigma,X,Q_i,\dual{\iota},Q_i\setminus
-        F_i,\dual{\Delta}_i}$ and, for $i=1,2$: \[\dual{\Delta}_i =
-      \set{q(\vec{y}) \arrow{a(X)}{} \dual{\psi} \mid
-        q(\vec{y}) \arrow{a(X)}{} \psi \in
-        \Delta_i}\enspace.\]
-  \end{compactenum}
-  Moreover, $\len{\A_\cap} = \bigO{\len{\A_1}+\len{\A_2}}$ and
-  $\len{\overline{\A_i}} = \bigO{\len{\A_i}}$, for all $i=1,2$.
-\end{theorem}
-\proof{ (\ref{it1:thm:closure}) ``$\subseteq$'' Let $w \in
-  \lang{\A_\cap}$ be a word and $\T$ be an execution of $\A_\cap$ over
-  $w$. Since $Q_1 \cap Q_2 = \emptyset$, it is possible to partition
-  $\T$ into $\T_1$ and $\T_2$ such that the roots of $\T_i$ form a
-  cube from $\cube{\minsem{\iota_i}}$, for all $i=1,2$. Because
-  $\Delta_1 \cap \Delta_2 = \emptyset$, by induction on
-  $\len{w}\geq0$, one shows that $\T_i$ is an execution of $\A_i$ over
-  $w$, for all $i=1,2$. Finally, because $\T$ is accepting, we obtain
-  that $\T_1$ and $\T_2$ are accepting, respectively, hence $w \in
-  \lang{\A_1} \cap \lang{\A_2}$. ``$\supseteq$'' Let $w \in
-  \lang{\A_1} \cap \lang{\A_2}$ and let $\T_i$ an accepting execution
-  of $\A_i$ over $w$, for all $i=1,2$. We show that $\T_1 \cup \T_2$
-  is an execution of $\A_\cap$ over $w$, by induction on
-  $\len{w}\geq0$. For the base case $\len{w}=0$, we have $\T_i \in
-  \cube{\minsem{\iota_i}}$ for all $i=1,2$ and since $Q_1 \cap Q_2 =
-  \emptyset$, we have $\T_1 \cup \T_2 \in \cube{\minsem{\iota_1 \wedge
-      \iota_2}}$. The induction step follows as a consequence of the
-  fact that $\Delta_1 \cup \Delta_2$ is the set of transition rules of
-  $\A_\cap$. Finally, since both $\T_1$ and $\T_2$ are accepting,
-  $\T_1 \cup \T_2$ is accepting as well. Moreover, we have: 
-  \[
-  \len{\A_\cap} = \len{\iota_1 \wedge \iota_2} + \sum_{q(\vec{y})
-    \arrow{a(X)}{} \psi \in \Delta_1 \cup \Delta_2} \len{\psi} =
-  1 + \len{\iota_1} + \len{\iota_2} + \sum_{q(\vec{y})
-    \arrow{a(X)}{} \psi \in \Delta_1} \len{\psi} + \sum_{q(\vec{y})
-    \arrow{a(X)}{} \psi \in \Delta_2} \len{\psi} \enspace. 
-  \]
-
-  \noindent(\ref{it2:thm:closure}) Let $w \in \Sigma[X]^*$ be a
-  word. We denote by $\accformA{\A_1}{\event{w}}$ and
-  $\substformA{\A_1}{\event{w}}$
-  [resp. $\accformA{\overline{\A}_1}{\event{w}}$ and
-    $\substformA{\overline{\A}_1}{\event{w}}$] the formulae
-  $\accform{\event{w}}$ and $\substform{\event{w}}$ for $\A_1$ and
-  $\overline{\A}_1$, respectively. It is enough to show that
-  $\substformA{\overline{\A}_1}{\event{w}} =
-  \neg\substformA{\A_1}{\event{w}}$ and apply Lemma
-  \ref{lemma:quant-pred-acceptance} to prove that $w \in \lang{\A_1}
-  \iff w \not\in \lang{\overline{\A}_1}$. Since the choice of $w$ was
-  arbitrary, this proves $\lang{\overline{\A}_1} = \Sigma[X]^*
-  \setminus \lang{\A_1}$. By induction on the number of predicate
-  atoms in $\accformA{\A_1}{\event{w}}$ that are replaced during the
-  generation of $\substformA{\A_1}{\event{w}}$. The proof relies on
-  the following fact:
-
-  \begin{fact}\label{fact:dual-neg}
-    Let $\phi$ be a positive formula and let $q(t_1,\ldots,t_{\#(q)})$
-    be the only occurrence of a predicate symbol within $\phi$. Then,
-    every formula $\phi$ with no predicate occurrences:
-    $\neg\phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]
-    \equiv
-    \dual{\phi}[\neg\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$.
-  \end{fact}
-  \proof{By induction on the structure of $\phi$. \qed}}
-
-\section{The Emptiness Problem}
-\label{sec:emptiness}
-
-The problem of checking emptiness of a given automaton is undecidable,
-even for automata with predicates of arity two, whose transition rules
-use only equalities and disequalities, having no transition
-quantifiers \cite{Farzan15}. Since even such simple classes of
-alternating automata have no general decision procedure for emptiness,
-we use an abstraction-refinement semi-algorithm based on \emph{lazy
-  annotation} \cite{McMillan06,McMillan14}.
-
-In a nutshell, a lazy annotation procedure systematically explores the
-set of execution paths (in our case, sequences of input events) in
-search of an accepting execution. Each path has a corresponding path
-formula that defines all words accepted along that path. If the path
-formula is satisfiable, the automaton accepts a word. Otherwise, the
-path is said to be \emph{spurious}. When a spurious path is
-encountered, the search backtracks and the path is annotated with a
-set of learned facts, that marks this path as infeasible. The
-semi-algorithm uses moreover a coverage relation between paths,
-ensuring that the continuations of already covered paths are never
-explored. Sometimes this coverage relation provides a sound
-termination argument, when the automaton is empty.
-
-We check emptiness of first-order alternating automata using a version
-of the \impact~ lazy annotation semi-algorithm \cite{McMillan06}. An
-analogous procedure is given in \cite{IosifXu18}, for a simpler model
-of alternating automata, that uses only predicates or arity zero
-(booleans) and no transition quantifiers. For simplicity, we do not
-present the details of this algorithm (see Appendix
-\ref{app:pseudocode}) and shall content ourselves of several
-high-level definitions.
-
-Given a finite input event alphabet $\Sigma$, for two sequences
-$\alpha, \beta \in \Sigma^*$, we say that $\alpha$ is a prefix of
-$\beta$, written $\alpha \prefix \beta$, if $\alpha=\beta\gamma$ for
-some sequence $\gamma\in\Sigma^*$. A set $S$ of sequences
-is: \begin{compactitem}
-\item \emph{prefix-closed} if for each $\alpha \in S$, if $\beta \prefix
-\alpha$ then $\beta \in S$, and
-\item \emph{complete} if for each $\alpha \in S$, there exists $a \in
-  \Sigma$ such that $\alpha a \in S$ if and only if $\alpha b \in S$
-  for all $b \in \Sigma$. 
-\end{compactitem}
-Observe that a prefix-closed set is the backbone of a tree whose edges
-are labeled with input events. If the set is, moreover, complete, then
-every node of the tree has either zero successors, in which case it is
-called a \emph{leaf}, or it has a successor edge labeled with $a$ for
-each input event $a \in \Sigma$.
-
-\begin{definition}\label{def:unfolding}
-  An \emph{unfolding} of an automaton $\A =
-  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ is a finite partial mapping $U :
-  \Sigma^* \rightharpoonup_{\mathit{fin}} \posforms(Q,\emptyset)$,
-  such that: \begin{compactenum}
-  \item\label{it1:unfolding} $\dom(U)$ is a finite prefix-closed
-    complete set,
-    %
-  \item\label{it2:unfolding} $U(\epsilon) = \iota$, and
-    %
-  \item\label{it3:unfolding} for each sequence $\alpha a \in \dom(U)$,
-    such that $\alpha \in \Sigma^*$ and $a \in \Sigma$:
-    \[\stamp{U(\alpha)}{0} \wedge
-    \bigwedge_{q(\vec{y}) \arrow{a(X)}{} \psi} \forall y_1 \ldots
-    \forall y_{\#q} ~.~ \stamp{q}{0}(\vec{y}) \rightarrow
-    \stamp{\psi}{1} \models \stamp{U(\alpha a)}{1}\]
-  \end{compactenum}
-  Moreover, $U$ is \emph{safe} if for each $\alpha \in \dom(U)$, the
-  formula $U(\alpha) \wedge \bigwedge_{q \in Q \setminus F} \forall
-  y_1 \ldots \forall_{y_{\#(q)}} ~.~ q(\vec{y}) \rightarrow \bot$ is
-  unsatisfiable. 
-\end{definition}
-
-Lazy annotation semi-algorithms \cite{McMillan06,McMillan14} build
-unfoldings of automata trying to discover counterexamples for
-emptiness. If the automaton $\A$ in question is non-empty, a
-systematic enumeration of the input event sequences\footnote{For
-  instance, using breadth-first search.} from $\Sigma^*$ will suffice
-to discover a word $w \in \lang{\A}$, provided that the first-order
-theory of the data domain $\Data$ is decidable (Lemma
-\ref{lemma:acceptance}). However, if $\lang{\A} = \emptyset$, the
-enumeration of input event sequences may, in principle, run
-forever. The typical way of fighting this divergence problem is to
-define a \emph{coverage} relation between the nodes of the unfolding
-tree.
-
-\begin{definition}\label{def:coverage}
-  Given an unfolding $U$ of an automaton $\A =
-  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ a node $\alpha \in \dom(U)$ is
-  \emph{covered} by another node $\beta \in \dom(U)$, denoted $\alpha
-  \cover \beta$, if and only if there exists a node $\alpha' \prefix
-  \alpha$ such that $U(\alpha') \models U(\beta)$. Moreover, $U$ is
-  \emph{closed} if and only if every leaf from $\dom(U)$ is covered by
-  an uncovered node.
-\end{definition}
-
-%% , for all words $w \in \Sigma[X]^*$ and all cubes $c \in
-%% \cube{\sem{U(\alpha')}}$, if $\A$ accepts $w$ started in $c$ then
-%% there exists a cube $d \in \cube{\sem{U(\beta)}}$ such that $\A$
-%% accepts $w$ started in $d$.
-
-A lazy annotation semi-algorithm will stop and report emptiness
-provided that it succeeds in building a closed and safe unfolding of
-the automaton. Notice that, by Definition \ref{def:coverage}, for any
-three nodes of an unfolding $U$, say $\alpha,\beta,\gamma \in
-\dom(U)$, if $\alpha \prec \beta$ and $\alpha \cover \gamma$, then
-$\beta \cover \gamma$ as well. As we show next (Theorem
-\ref{thm:soundness}), there is no need to expand covered nodes,
-because, intuitively, there exists a word $w \in \lang{\A}$ such that
-$\alpha \preceq \event{w}$ and $\alpha \cover \gamma$ only if there
-exists another word $u \in \lang{\A}$ such that $\gamma \preceq
-\event{u}$. Hence, exploring only those input event sequences that are
-continuations of $\gamma$ (and ignoring those of $\alpha$) suffices in
-order to find a counterexample for emptiness, if one exists.
-
-An unfolding node $\alpha \in \dom(U)$ is said to be \emph{spurious}
-if and only if $\accform{\alpha}$ is unsatisfiable. In this case, we
-change (refine) the labels of (some of the) prefixes of $\alpha$ (and
-that of $\alpha$), such that $U(\alpha)$ becomes $\bot$, thus
-indicating that there is no real execution of the automaton along that
-input event sequence. As a result of the change of labels, if a node
-$\gamma \prefix \alpha$ used to cover another node from $\dom(U)$, it
-might not cover it with the new label. Therefore, the coverage
-relation has to be recomputed after each refinement of the
-labeling. The semi-algorithm stops when (and if) a safe complete
-unfolding has been found. For a detailed presentation of the emptiness
-procedure, we refer the reader to \cite{IosifXu18} (or Appendix
-\ref{app:pseudocode}).
-
-\begin{theorem}\label{thm:soundness}
-  If an automaton $\A$ has a nonempty safe closed unfolding then
-  $\lang{A} = \emptyset$.
-\end{theorem}
-\proof{ Let $U$ be a safe and complete unfolding of $\A$, such that
-  $\dom(U) \neq \emptyset$. Suppose, by contradiction, that there
-  exists a word $w \in \lang{A}$ and let $\alpha \isdef
-  \event{w}$. Since $w \in \lang{A}$, by Lemma \ref{lemma:acceptance},
-  there exists an interpretation $\I$ such that $\I,\data{w} \models
-  \accform{\alpha}$. Assume first that $\alpha \in \dom(U)$.  In this
-  case, one can show, by induction on the length $n\geq0$ of $w$, that
-  $\pathform{\alpha} \models \stamp{U(\alpha)}{n}$, thus $\I,\data{w}
-  \models \stamp{U(\alpha)}{n}$. Since $\I,\data{w} \models
-  \accform{\alpha}$, we have $\I,\data{w} \models \bigwedge_{q \in Q
-    \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{n}(\vec{y}) \rightarrow \bot$, hence $\stamp{U(\alpha)}{n}
-  \wedge \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
-  y_{\#(q)} ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot$. By renaming
-  $\stamp{q}{n}$ with $q$ in the previous formula, we obtain
-  $U(\alpha) \wedge \forall y_1 \ldots \forall y_{\#(q)} ~.~
-  q(\vec{y}) \rightarrow \bot$ is satisfiable, thus $U$ is not safe,
-  contradiction.
-
-  We proceed thus under the assumption that $\alpha \not\in
-  \dom(U)$. Since $\dom(U)$ is a nonempty prefix-closed set, there
-  exists a strict prefix $\alpha'$ of $\alpha$ that is a leaf of
-  $\dom(U)$. Since $U$ is closed, the leaf $\alpha'$ must be covered
-  and let $\alpha_1 \prefix \alpha' \prefix \alpha$ be a node such
-  that $U(\alpha_1) \models U(\beta_1)$, for some uncovered node
-  $\beta_1 \in \dom(U)$.
-  %% for every word $w'$ and every cube $c \in
-  %% \cube{\sem{U(\alpha_1)}}$ such that $\A$ accepts $w'$ starting
-  %% with $c$, there exists a cube $d \in \cube{\sem{U(\beta_1)}}$
-  %% such that $\A$ accepts $w'$ starting with $d$.
-  Let $\gamma_1$ be the unique sequence such that $\alpha_1\gamma_1 =
-  \alpha$. By Definition \ref{def:coverage}, since $\alpha_1 \cover
-  \beta_1$ and $\event{w} = \alpha_1\gamma_1 \in \lang{\A}$, there
-  exists a word $w_1$ and a cube $c_1 \in \cube{\sem{U(\alpha_1)}}
-  \subseteq \cube{\sem{U(\beta_1)}}$, such that $\event{w_1} =
-  \gamma_1$ and $\A$ accepts $w_1$ starting with $c_1$. If
-  $\beta_1\gamma_1 \in \dom(U)$, we obtain a contradiction by a
-  similar argument as above. Hence $\beta_1\gamma_1 \not\in \dom(U)$
-  and there exists a leaf of $\dom(U)$ which is also a prefix of
-  $\beta_1\gamma_1$. Since $U$ is closed, this leaf is covered by an
-  uncovered node $\beta_2 \in \dom(U)$ and let $\alpha_2 \in \dom(U)$
-  be the minimal (in the prefix partial order) node such that $\beta_1
-  \prefix \alpha_2 \prefix \beta_1\gamma_1$ and $\alpha_2 \cover
-  \beta_2$. Let $\gamma_2$ be the unique sequence such that
-  $\alpha_2\gamma_2 = \beta_1\gamma_1$. Since $\beta_1$ is uncovered,
-  we have $\beta_1 \neq \alpha_2$ and thus $\len{\gamma_1} >
-  \len{\gamma_2}$. By repeating the above reasoning for $\alpha_2$,
-  $\beta_2$ and $\gamma_2$, we obtain an infinite sequence
-  $\len{\gamma_1} > \len{\gamma_2} > \ldots$, which is again a
-  contradiction. \qed}
-
-As mentioned above, we check emptiness of first-order alternating
-automata using the same method previously used to check emptiness of a
-simpler model of alternating automata, which uses boolean constants
-for control states and whose transition rules have no quantifiers
-\cite{IosifXu18}. The higher complexity of the automata model
-considered here, manifests itself within the interpolant generation
-procedure, used to refine the labeling of the unfolding. We discuss
-generation of interpolants in the next section.
-
-\section{Interpolant Generation}\label{sec:interpolants}
-
-Typically, when checking the unreachability of a set of program
-configurations \cite{mcmillan06}, the interpolants used to annotate
-the unfolded control structure are assertions about the values of the
-program variables in a given control state, at a certain step of a
-given execution. However, in an alternating model of computation, it
-is useful to distinguish between
-\begin{inparaenum}[(i)]
-\item locality of interpolants w.r.t. a given control state (control
-  locality) and
-%
-\item locality w.r.t. a given time stamp (time locality).
-\end{inparaenum} 
-In logical terms, \emph{control-local} interpolants are defined by
-formulae involving a single predicate symbol, whereas
-\emph{time-local} interpolants involve only predicates $\stamp{q}{i}$
-and variables $\stamp{x}{i}$, for a single $i \geq 0$.
-
-\paragraph{Remark}
-When considering an alternating model of computation, control-local
-interpolants are not always enough to prove emptiness, because of the
-synchronization of several branches of the computation on the same
-sequence of input values. Consider, for instance, an automaton with
-the following transition rules and final state $q_f$:
-\[\begin{array}{rclcrcl}
-q_0(y) & \arrow{a(x)}{} & q_1(y+x) \wedge q_2(y-x) & \hspace*{2mm} & 
-q_1(y) \arrow{a(x)}{} y+x > 0 \wedge q_f \\
-q_1(y) & \arrow{a(x)}{} & q_1(y+x) && q_2(y) \arrow{a(x)}{} y-x > 0 \wedge q_f \\
-q_2(y) & \arrow{a(x)}{} & q_2(y-x)
-\end{array}\]
-Started in an initial configuration $q_0(0)$ with an input word
-$(a,\nu_1) \ldots (a,\nu_{n-1}) (a,\nu_n)$, such that $\nu_i(x)=k_i$,
-the automaton executes as follows:
-\[\begin{array}{c}
-q_0(0) \arrow{(a,\nu_1)}{} \set{q_1(k_1),q_2(-k_1)} \ldots \arrow{(a,\nu_{n-1})}{} 
-\set{q_1(\sum_{i=1}^{n-1} k_i),q_2(-\sum_{i=1}^{n-1} k_i)}
-\arrow{(a,\nu_n)}{} \emptyset
-\end{array}\]
-An overapproximation of the set of cubes generated after one or more
-steps is defined by the formula: $\exists x_1 \exists x_2 ~.~ q_1(x_1)
-\wedge q_2(x_2) \wedge x_1+x_2 \teq 0$. Observe that a control-local
-formula using one occurrence of a predicate would give a too rough
-overapproximation of this set, unable to prove the emptiness of the
-automaton. \hfill$\blacksquare$
-
-First, let us give the formal definition of the class of interpolants
-we shall work with. Given a formula $\phi$, the \emph{vocabulary} of
-$\phi$, denoted $\voc{\phi}$ is the set of predicate symbols $q \in
-\stamp{Q}{i}$ and variables $x \in \stamp{X}{i}$, occurring in $\phi$,
-for some $i\geq0$. For a term $t$, its vocabulary $\voc{t}$ is the set
-of variables that occur in $t$. Observe that quantified variables and
-the interpreted function symbols of the data
-theory\footnote{E.g.,\ the arithmetic operators of addition and
-  multiplication, when $\Data$ is the set of integers.}  do not belong
-to the vocabulary of a formula. By $\pset{+}{\phi}$ [$\pset{-}{\phi}$]
-we denote the set of predicate symbols that occur in $\phi$ under an
-even [odd] number of negations.
-
-\begin{definition}[\cite{Lyndon59}]\label{def:lyndon-interpolant}
-Given formulae $\phi$ and $\psi$ such that $\phi \wedge \psi$ is
-unsatisfiable, a \emph{Lyndon interpolant} is a formula $I$ such that
-$\phi \models I$, the formula $I \wedge \psi$ is unsatisfiable,
-$\voc{I} \subseteq \voc{\phi} \cap \voc{\psi}$, $\pset{+}{I} \subseteq
-\pset{+}{\phi} \cap \pset{+}{\psi}$ and $\pset{-}{I} \subseteq
-\pset{-}{\phi} \cap \pset{-}{\psi}$.
-\end{definition}
-%% The existence of Lyndon interpolants for first-order logic, with no
-%% predefined interpretation of function symbols\footnote{A model
-%%   specifies the universe and the interpretation of function symbols.},
-%% was proved by a non-constructive model-theoretic argument
-%% \cite{Lyndon59}. In the following, we shall consider only those
-%% theories of the data domain $\Data$, for which there exists a decision
-%% procedure for satisfiability, able to produce Lyndon interpolants. For
-%% instance, such interpolating decision procedures exist for linear real
-%% and integer arithmetic, with uninterpreted function symbols (UFLRA and
-%% UFLIA).
-
-In the rest of this section, let us fix an automaton $\A =
-\tuple{\Sigma,X,Q,\iota,F,\Delta}$. Due to the above observation, none
-of the interpolants considered will be control-local and we shall use
-the term \emph{local} to denote time-local interpolants, with no free
-variables.
-
-\begin{definition}\label{def:generalized-lyndon-interpolant}
-  Given a non-empty sequence of input events $\alpha = a_1 \ldots a_n
-  \in \Sigma^*$, a \emph{generalized Lyndon interpolant (GLI)} is a
-  sequence $(I_0,\ldots,I_n)$ of formulae such that, for all $k \in
-  [n-1]$: \begin{compactenum}
-    \item\label{it1:generalized-lyndon-interpolant} $\pset{-}{I_k} =
-      \emptyset$,
-      %
-    \item\label{it2:generalized-lyndon-interpolant} $\stamp{\iota}{0}
-      \models I_0$ and \(I_k \wedge \Big(\bigwedge_{q(\vec{y})
-        \arrow{a_i(X)}{} \psi \in \Delta} \forall y_1 \ldots \forall
-      y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y}) \rightarrow
-      \stamp{\psi}{k+1}\Big) \models I_{k+1}\),
-      %
-    \item\label{it3:generalized-lyndon-interpolant} $I_n \wedge
-      \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
-      y_{\#(q)} ~.~ q(\vec{y})$ is unsatisfiable.
-  \end{compactenum}
-  Moreover, the GLI is \emph{local} if and only if $\voc{I_k}
-  \subseteq \stamp{Q}{k}$, for all $k \in [n]$.
-\end{definition}
-The following proposition states the existence of local GLI for the
-theories in which Lyndon's Interpolation Theorem holds.
-
-\begin{proposition}\label{prop:local-gli}
-  If there exists a Lyndon interpolant for any two formulae $\phi$ and
-  $\psi$, such that $\phi \wedge \psi$ is unsatisfiable, then any
-  sequence of input events $\alpha = a_1 \ldots a_n \in \Sigma^*$,
-  such that $\accform{\alpha}$ is unsatisfiable, has a local GLI
-  $(I_0,\ldots,I_n)$.
-\end{proposition}
-\proof{ By definition, $\accform{\alpha}$ is the formula:
-  \[\stamp{\iota}{0} \wedge 
- \bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in
-   \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
- \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i} \wedge
- \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
- ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot\]
- We define the formulae: 
- \[\begin{array}{rcl}
- \varphi_i & \isdef & \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
- \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i},~ \text{for all}~ i \in [1,n] \\
- \psi & \isdef & \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
- ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot
- \end{array}\]
- Observe that $\voc{\stamp{\iota}{0}} \subseteq \stamp{Q}{0}$,
- $\voc{\varphi_i} \subseteq
- \stamp{Q}{i-1}\cup\stamp{Q}{i}\cup\stamp{X}{i}$, for all $i \in
-       [1,n]$, and $\voc{\psi} \subseteq \stamp{Q}{n}$.  We apply
-       Lyndon's Interpolation Theorem for the formulae
-       $\stamp{\iota}{0}$ and $\bigwedge_{i=1}^n \varphi_i \wedge
-       \psi$ and obtain a formula $I_0$, such that $\stamp{\iota}{0}
-       \models I_0$, $I_0 \wedge \bigwedge_{i=1}^n \varphi_i \wedge
-       \psi$ is unsatisfiable, $\voc{I_0} \subseteq
-       \voc{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n \voc{\varphi_i}
-       \cup \voc{\psi}) \subseteq \stamp{Q}{0}$ and $\pset{-}{I_0}
-       \subseteq \pset{-}{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n
-       \pset{-}{\varphi} \cup \pset{-}{\psi}) = \emptyset$. Repeating
-       the reasoning for the formulae $I_0 \wedge \varphi_1$ and
-       $\bigwedge_{i=2}^n \varphi_i \wedge \psi$, we obtain $I_1$,
-       such that $I_0 \wedge \varphi_1 \models I_1$, $I_1 \wedge
-       \bigwedge_{i=2}^n \varphi_i \wedge \psi$ is unsatisfiable,
-       $\voc{I_1} \subseteq (\voc{I_0} \cup \voc{\varphi_1}) \cap
-       (\bigcup_{i=2}^n \voc{\varphi_i} \cup \voc{\psi}) \subseteq
-       \stamp{Q}{1}$ and $\pset{-}{I_1} \subseteq (\pset{-}{I_0} \cup
-       \pset{-}{\varphi_1}) \cap (\bigcup_{i=2}^n \pset{-}{\varphi_i}
-       \cup \pset{-}{\psi}) = \emptyset$. Continuing in this way, we
-       obtain formulae $I_0, I_1, \ldots, I_{n}$ as required. \qed}
-
-The main problem with the local GLI construction described in the
-proof of Proposition \ref{prop:local-gli} is that the existence of
-Lyndon interpolants (Definition \ref{def:lyndon-interpolant}) is
-guaranteed in principle, but the proof is non-constructive. Building
-an interpolant for an unsatisfiable conjunction of formulae $\phi
-\wedge \psi$ is typically the job of the decision procedure that
-proves the unsatisfiability and, in general, there is no such
-procedure, when $\phi$ and $\psi$ contain predicates and have
-non-trivial quantifier alternation. In this case, some provers use
-instantiation heuristics for the universal quantifiers that are
-sufficient for proving unsatisfiability, however these heuristics are
-not always suitable for interpolant generation. Consequently, from now
-on, we assume the existence of an effective Lyndon interpolation
-procedure only for decidable theories, such as the quantifier-free
-linear (integer) arithmetic with uninterpreted functions (UFLIA,
-UFLRA, etc.) \cite{RybalchenkoSofronieStokkermans}.
-
-This is where the predicate-free path formulae (Definition
-\ref{def:substform}) come into play. For a given event sequence
-$\alpha$, the automaton $\A$ accepts a word $w$ such that $\event{w} =
-\alpha$ if and only if $\substform{\alpha}$ is satisfiable. Assuming
-further that the equality atoms in the transition rules of $\A$ are
-written in the language of a decidable first-order theory, such as
-Presburger arithmetic, Lemma \ref{lemma:quant-pred-acceptance} gives
-us an effective way of checking emptiness of $\A$, relative to a given
-event sequence. However, this method does not cope well with lazy
-annotation, because there is no way to extract, from the
-unsatisfiability proof of $\substform{\alpha}$, the interpolants
-needed to annotate $\alpha$. This is because \begin{inparaenum}[(i)]
-\item\label{it1:gli} the formula $\substform{\alpha}$, obtained by
-  repeated substitutions (Definition \ref{def:substform}) loses track
-  of the steps of the execution, and
-%
-\item\label{it2:gli} quantifiers that occur nested in
-  $\substform{\alpha}$ make it difficult to write $\substform{\alpha}$
-  as an unsatisfiable conjunction of formulae from which interpolants
-  are extracted (Definition \ref{def:lyndon-interpolant}).
-\end{inparaenum}
-
-The solution we adopt for the first issue (\ref{it1:gli}) consists in
-partially recovering the time-stamped structure of the acceptance
-formula $\accform{\alpha}$ using the formula $\quantform{\alpha}$, in
-which only transition quantifiers occur. The second issue
-(\ref{it2:gli}) is solved under the additional assuption that the
-theory of the data domain $\Data$ has \emph{witness-producing
-  quantifier elimination}. More precisely, we assume that, for each
-formula $\exists x ~.~ \phi(x)$, there exists an effectively
-computable term $\tau$, in which $x$ does not occur, such that
-$\exists x ~.~ \phi$ and $\phi[\tau/x]$ are equisatisfiable. These
-terms, called \emph{witness terms} in the following, are actual
-definitions of the Skolem function symbols from the following folklore
-theorem:
-
-\begin{theorem}\label{thm:skolem}
-  Given $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$ a first order sentence, 
-  where $Q_1, \ldots, Q_n \in \set{\exists,\forall}$ and $\phi$ is
-  quantifier-free, let: 
-  \[\eta_i \equiv \left\{\begin{array}{ll}
-  f_i(y_1,\ldots,y_{k_i}) & \text{if } Q_i = \forall \\
-  x_i & \text{if } Q_i = \exists
-  \end{array}\right.\]
-  where $f_i$ is a fresh function symbol and $\set{y_1, \ldots,
-    y_{k_i}} = \set{x_j \mid j < i,~ Q_j = \exists}$. Then the entailment \(Q_1
-  x_1 \ldots Q_n x_n ~.~ \phi \models
-  \phi[\eta_1/x_1,\ldots,\eta_n/x_n]\) holds. 
-\end{theorem}
-\proof{See \cite[Theorem 2.1.8]{BorgerGraedelGurevich97} and
-  \cite[Lemma 2.1.9]{BorgerGraedelGurevich97}. \qed}
-
-Examples of witness-producing quantifier elimination procedures can be
-found in the literature for e.g. linear integer (real) arithmetic
-(LIA,LRA), Presburger arithmetic and boolean algebra of sets and
-Presburger cardinality constraints (BAPA)
-\cite{KuncakMayerPiskacSuter12}.
-
-%% Furthermore, we describe a cubic-time witness generation for
-%% difference bounds constraints, a simple subset of linear integer
-%% (real) arithmetic (\S\ref{sec:applications}).
-
-Under the assumption that witness terms can be effectively built, let
-us describe the generation of a non-local GLI for a given input event
-sequence $\alpha = a_1 \ldots a_n$. First, we generate successively
-the acceptance formula $\accform{\alpha}$ and its equisatisfiable
-forms $\quantform{\alpha} = Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$
-and $\substform{\alpha} = Q_1x_1 \ldots Q_mx_m ~.~ \overline{\Phi}$,
-both written in prenex form, with matrices $\widehat{\Phi}$ and
-$\overline{\Phi}$, respectively. Because we assumed that the
-first-order theory of $\Data$ has quantifier elimination, the
-satisfiability problem for $\substform{\alpha}$ is decidable. If
-$\substform{\alpha}$ is satisfiable, we build a counterexample for
-emptiness $w$ such that $\event{w}=\alpha$ and $\data{w}$ is a
-satisfying assignment for $\substform{\alpha}$. Otherwise,
-$\substform{\alpha}$ is unsatisfiable and there exist witness terms
-$\tau_{i_1} \ldots \tau_{i_\ell}$, where $\set{i_1, \ldots, i_\ell} =
-\set{j \in [1,m] \mid Q_j = \forall}$, such that
-$\overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
-  \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable (Theorem
-\ref{thm:skolem}). Then it turns out that the formula
-$\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
-  \tau_{i_\ell}/x_{i_\ell}]$, obtained analogously from the matrix of
-$\quantform{\alpha}$, is unsatisfiable as well (Lemma
-\ref{lemma:subst-quant}). Because this latter formula is structured as
-a conjunction of formulae $\stamp{\iota}{0} \wedge \phi_1 \ldots
-\wedge \phi_n \wedge \psi$, where $\voc{\phi_k} \cap \stamp{Q}{\leq n}
-\subseteq \stamp{Q}{k-1} \cup \stamp{Q}{k}$ and $\voc{\psi} \cap
-\stamp{Q}{\leq n} \subseteq \stamp{Q}{n}$, it is now possible to use
-an existing interpolation procedure for the quantifier-free theory of
-$\Data$, extended with uninterpreted function symbols, to compute a
-sequence of non-local GLI $(I_0, \ldots, I_n)$ such that $\voc{I_k}
-\cap \stamp{Q}{\leq n} \subseteq \stamp{Q}{k}$, for all $k \in [n]$.
-
-%% Since each formula $I_k$ of the GLI may refer to the current
-%% predicates $\stamp{Q}{k}$ and to the past and current input variables
-%% $\stamp{X}{\leq k}$, we further give up on the locality of
-%% interpolants and work with non-local GLI that are effectively
-%% computable by a combination of quantifier elimination with
-%% interpolation for quantifier-free data theories with uninterpreted
-%% functions.
-
-\begin{example}[Contd. from Examples \ref{ex:quant-elim} and \ref{ex:pred-elim}]\label{ex:gli}
-The formula $\substform{\alpha}$ (Example \ref{ex:pred-elim}) is
-unsatisfiable and let $\tau_2 = z_1$ be the witness term for the
-universally quantified variable $z_2$. Replacing $z_2$ with $\tau_2$
-in the matrix of $\quantform{\alpha}$ (Example \ref{ex:quant-elim})
-yields the unsatisfiable conjunction: 
-\[\begin{array}{c}
-z_1 \geq 0 \wedge \stamp{q}{0}(z_1)  \\ \wedge \\
-\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge (z_1 \geq z_1 \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_1)) \\ \wedge \\
-\stamp{q}{1}(\stamp{x}{1}+z_1) \rightarrow \stamp{x}{1}+z_1 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_1)
-\end{array}\]
-A non-local GLI for the above is $(\stamp{q}{0}(z_1) \wedge
-z_1\geq 0,~ \stamp{x}{1} \geq 0 \wedge \stamp{q}{1}(\stamp{x}{1}+z_1)
-\wedge z_1\geq0,~\bot)$. \hfill$\blacksquare$
-\end{example}
-
-A function $\xi : \nat \rightarrow \nat$ is [strictly]
-\emph{monotonic} iff for each $n < m$ we have $\xi(n) \leq \xi(m)$
-     [$\xi(n) < \xi(m)$] and \emph{finite-range} iff for each $n \in
-     \nat$ the set $\set{m \mid \xi(m)=n}$ is finite. If $\xi$ is
-     finite-range, we denote by $\xi_{\max}^{-1}(n) \in \nat$ the
-     maximal value $m$ such that $\xi(m)=n$. It is easy to see that
-     $\xi_{\max}^{-1}$ is strictly monotonic if $\xi$ is monotonic and
-     finite-range. The lemma below gives the proof of correctness for
-     the previous construction of non-local GLI.
-
-\begin{lemma}\label{lemma:subst-quant}
-  Given a non-empty input event sequence $\alpha = a_1 \ldots a_n \in
-  \Sigma^*$, such that $\accform{\alpha}$ is unsatisfiable, let
-  $Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$ be a prenex form of
-  $\quantform{\alpha}$ and let $\xi : [1,m] \rightarrow [n]$ be a
-  monotonic function mapping each transition quantifier to the minimal
-  index from the sequence $\quantformn{\alpha_0}, \ldots,
-  \quantformn{\alpha_n}$ where it occurs. Then the following
-  hold: \begin{compactenum}
-  \item\label{it1:subst-quant} one can effectively build witness terms
-    $\tau_{i_1}, \ldots, \tau_{i_\ell}$, where 
-    \[\begin{array}{l}   
-    \set{i_1, \ldots,i_\ell} = \set{j \in [1,m] \mid Q_j = \forall} \text{ and} \\
-    \voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup \set{x_k
-    \mid k < i_j, Q_k = \exists},~ \forall j \in [1,\ell]
-    \end{array}\]
-    such that $\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable, and
-    %
-  \item\label{it2:subst-quant} one can effectively build a GLI $(I_0,
-    \ldots, I_n)$ for $\alpha$, such that \[\voc{I_k} \subseteq
-    \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
-      \xi^{-1}(k),~ Q_j = \exists},~ \forall k \in [n]\enspace.\]
-  \end{compactenum}
-\end{lemma}
-\proof{(\ref{it1:subst-quant}) If $\accform{\alpha}$ is unsatisfiable,
-  by Lemmas \ref{lemma:quant} and \ref{lemma:subst}, we obtain that,
-  successively $\quantform{\alpha}$ and $\substform{\alpha}$ are
-  unsatisfiable. Let $Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$ and
-  $Q_1x_1 \ldots Q_mx_m ~.~ \overline{\Phi}$ be prenex forms for
-  $\quantform{\alpha}$ and $\substform{\alpha}$, respectively. Since
-  we assumed that the first-order theory of the data domain has
-  witness-producing quantifier elimination, using Theorem
-  \ref{thm:skolem} one can effectively build witness terms
-  $\tau_{i_1}, \ldots, \tau_{i_\ell}$, where
-  $\set{i_1,\ldots,i_\ell}=\set{i \in [1,m] \mid Q_i=\forall}$
-  and: \begin{compactitem}
-  \item $\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup
-    \set{x_k \mid k < i_j, Q_k = \exists}$, for all $j \in [1,\ell]$ and
-    %
-  \item $\overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
-    \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.
-  \end{compactitem}
-  Let $\widehat{\Phi}_0$, \ldots, $\widehat{\Phi}_n$ be the sequence
-  of quantifier-free formulae, defined as follows: \begin{compactitem}
-  \item $\widehat{\Phi}_0$ is the matrix of some prenex form of
-    $\stamp{\iota}{0}$,
-    %
-  \item for all $i=1,\ldots,n$, let $\widehat{\Phi}_i$ be the matrix
-    of some prenex form of:
-    \[\widehat{\Phi}_i \isdef \widehat{\Phi}_{i-1} \wedge
-     \underbrace{\bigwedge_{\begin{array}{l}
-         \scriptstyle{\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \widehat{\Phi}_{i-1}} \\
-        \scriptstyle{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta} 
-     \end{array}} \stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow 
-     \stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]}_{\isdef \phi_i}\]  
-  \end{compactitem}
-  It is easy to see that $\widehat{\Phi}$ is the matrix of some prenex
-  form of:
-  \[\widehat{\Phi}_n \wedge \underbrace{\bigwedge_{\begin{array}{l}
-        \scriptstyle{\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \widehat{\Phi}_n} \\
-        \scriptstyle{q \in Q \setminus F}
-    \end{array}} \stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot}_{\isdef \psi}\]
-  Applying the equivalence from Fact \ref{fact:subst} in the proof of
-  Lemma \ref{lemma:subst}, we obtain a sequence of quantifier-free
-  formulae $\overline{\Phi}_0, \ldots, \overline{\Phi}_n$ such that
-  $\overline{\Phi}_i \equiv \widehat{\Phi}_i$, for all $i \in [n]$ and
-  $\overline{\Phi}$ is obtained from $\overline{\Phi}_n$ by replacing
-  each occurrence of a predicate atom $q(t_1,\ldots,t_{\#(q)})$ in
-  $\overline{\Phi}_n$ by $\bot$ if $q \in Q\setminus F$ and by $\top$
-  if $q \in F$. Clearly $\overline{\Phi} \equiv \widehat{\Phi}$, thus
-  $\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
-    \tau_{i_\ell}/x_{i_\ell}] \equiv
-  \overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
-    \tau_{i_\ell}/x_{i_\ell}] \equiv \bot$.
-
-  \noindent(\ref{it2:subst-quant}) With the notation introduced at
-  point (\ref{it1:subst-quant}), we have $\widehat{\Phi} =
-  \widehat{\Phi}_0 \wedge \bigwedge_{i=1}^n \phi_i \wedge
-  \psi$. Consider the sequence of witness terms $\tau_{i_1}, \ldots,
-  \tau_{i_\ell}$, whose existence is proved by point
-  (\ref{it1:subst-quant}). Because $\voc{\tau_{i_j}} \subseteq
-  \stamp{X}{\leq \xi(i_j)} \cup \set{x_k \mid k < i_j, Q_k =
-    \exists}$, for all $j \in [1,\ell]$, and moreover $\xi^{-1}$ is
-  strictly monotonic, we obtain: \begin{compactitem}
-  \item $\voc{\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{0} \cup
-    \stamp{X}{0} \cup \set{x_j \mid j < \xi_{\max}^{-1}(0), Q_j = \exists}$,
-    %
-  \item $\voc{\phi_i[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{i-1} \cup
-    \stamp{Q}{i} \cup \stamp{X}{\leq i} \cup \set{x_j \mid j <
-      \xi_{\max}^{-1}(i), Q_j = \exists}$, for all $i \in [1,n]$,
-    %
-  \item $\voc{\psi[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{n} \cup
-    \stamp{X}{\leq n} \cup \set{x_j \mid j \in [1,m], Q_j = \exists}$.
-  \end{compactitem}
-  By repeatedly applying Lyndon's Interpolation Theorem, we obtain a
-  sequence of formulae $(I_0, \ldots, I_n)$ such that: \begin{compactitem}
-    \item $\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}] \models I_0$ and $\voc{I_0} \subseteq
-      \stamp{Q}{0} \cup \stamp{X}{0} \cup \set{x_j \mid j <
-        \xi_{\max}^{-1}(0), Q_j = \exists}$,
-      %
-    \item $I_{k-1} \wedge \phi_i[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}] \models I_k$ and $\voc{I_k} \subseteq
-      \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
-       \xi_{\max}^{-1}(k), Q_j = \exists}$, for all $k \in [1,n]$,
-      %
-    \item $I_n \wedge \psi[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.
-  \end{compactitem}
-  To show that $(I_0,\ldots,I_n)$ is a GLI for $a_1\ldots a_n$, it is
-  sufficient to notice that \[\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
-    \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{k}(\vec{y}) \rightarrow \stamp{\psi}{k+1} \models \phi_k\]
-  for all $k \in [1,n]$. Consequently, we obtain: \begin{compactitem}
-    \item $\stamp{\iota}{0} \models \widehat{\Phi}_0 \models I_0$, by
-      Theorem \ref{thm:skolem},
-      %
-    \item $I_{k-1} \wedge \Big(\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
-      \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-      \stamp{q}{k-1}(\vec{y}) \rightarrow \stamp{\psi}{k}\Big) \models
-      I_{k-1} \wedge \phi_k \models I_k$, and 
-      %
-    \item $I_n \wedge \Big(\bigwedge_{q \in Q \setminus F} \forall y_1
-      \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot\Big)
-      \models I_n \wedge \psi \models \bot$,     
-  \end{compactitem}
-  as required by Definition
-  \ref{def:generalized-lyndon-interpolant}. \qed}
-
-In conclusion, under two assumptions about the first-order theory of
-the data domain, namely the \begin{inparaenum}[(i)]
-\item witness-producing quantifier elimination, and
-\item Lyndon interpolation for the quantifier-free fragment with
-  uninterpreted functions,
-\end{inparaenum}
-we developped a rather generic method that produces generalized Lyndon
-interpolants for unfeasible input event sequences. Moreover, each
-formula $I_k$ in the interpolant refers only to the current predicate
-symbols $\stamp{Q}{I_k}$, the current and past input variables
-$\stamp{X}{\leq k}$ and the existentially quantified transition
-variables introduced at the previous steps $\set{x_j \mid j <
-  \xi_{\max}^{-1}(k), Q_j = \exists}$. The remaining question is how
-to use such non-local interpolants to label the unfolding of an
-automaton (Definition \ref{def:unfolding}) and to compute the coverage
-between nodes of the unfolding (Definition \ref{def:coverage}).
-
-\subsection{Unfolding with Non-local Interpolants}
-
-As required by Definition \ref{def:unfolding}, the unfolding $U$ of an
-automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ is labeled by
-formulae $U(\alpha) \in \posforms(Q,\emptyset)$, with no free symbols,
-other than predicate symbols, such that the labeling is compatible
-with the transition relation of the automaton, according to the point
-(\ref{it3:unfolding}) of Definition \ref{def:unfolding}. The following
-lemma describes the refinement of the labeling of an input sequence
-$\alpha$ of length $n$ by a non-local GLI $(I_0, \ldots, I_n)$, such
-that $\voc{I_k} \subseteq \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup
-\vec{x}_k$, where $\vec{x}_k$ are the existentially quantified
-variables from the prenex normal form of $\quantform{\alpha_k}$.
-
-\begin{lemma}\label{lemma:refinement}
-  Let $U$ be an unfolding of an automaton $\A =
-  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ such that $\alpha = a_1\ldots a_n
-  \in \dom(U)$ and $(I_0, \ldots, I_n)$ be a GLI for $\alpha$. The
-  mapping $U' : \dom(U) \rightarrow \posforms(Q,\emptyset)$ defined as: 
-  \begin{compactitem}
-  \item $U'(\alpha_k) = U(\alpha_k) \wedge J_k$, for all $k \in [n]$,
-    where $J_k$ is the formula obtained from $I_k$ by replacing each
-    time-stamped predicate symbol $\stamp{q}{k}$ by $q$ and
-    existentially quantifying each free variable in $I_k$,
-  %
-\item $U'(\beta) = U(\beta)$ if $\beta \in \dom(U)$ and $\beta
-  \not\prefix \alpha$,
-  \end{compactitem}
-  is an unfolding of $\A$. 
-\end{lemma}
-\proof{
-  The new set of formulae $U'(\alpha_0), \ldots, U'(\alpha_n)$ complies
-  with Definition \ref{def:unfolding}, because: \begin{compactitem}
-  \item $U'(\alpha_0) \equiv \iota$, since, by point
-    \ref{it2:generalized-lyndon-interpolant} of Definition
-    \ref{def:generalized-lyndon-interpolant}, we have $\stamp{\iota}{0}
-    \models I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) =
-    U(\alpha_0) \wedge J_0 \equiv \iota \wedge J_0 \equiv \iota$, and
-    %
-  \item by Definition \ref{def:generalized-lyndon-interpolant}
-    (\ref{it3:generalized-lyndon-interpolant}) we have, for all $k \in
-    [n-1]$:
-    \[I_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
-    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
-    \rightarrow \stamp{\psi}{k+1} \models I_{k+1}\] We write
-    $\overstamp{I}{j}_k$ for the formula in which each predicate
-    symbol $\stamp{q}{k}$ is replaced by $\stamp{q}{j}$. Then the
-    following entailment holds:
-    \[\overstamp{I}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
-    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
-    \rightarrow \stamp{\psi}{1} \models \overstamp{I}{1}_{k+1}\]
-    Because $J_k$ is obtained by removing the time stamps from the predicate symbols and  
-    existentially quantifying all the free variables of $I_k$, we also
-    obtain, applying Fact \ref{fact:ex-entail} below:
-    \[\stamp{J}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
-    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
-    \rightarrow \stamp{\psi}{1} \models \stamp{J}{1}_{k+1}\] Since $U$
-    satisfies the labeling condition of Definition \ref{def:unfolding}
-    (\ref{it3:unfolding}) and $U'(\alpha_k) = U(\alpha_k) \wedge J_k$,
-    we obtain, as required: 
-    \[\stamp{U'(\alpha_k)}{0} \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
-    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
-    \rightarrow \stamp{\psi}{1} \models
-    \stamp{U'(\alpha_{k+1})}{1}\enspace.\]
-  \end{compactitem}
-
-  \begin{fact}\label{fact:ex-entail}
-    Given formulae $\phi(\vec{x}, \vec{y})$ and $\psi(\vec{x})$ such
-    that $\phi(\vec{x}, \vec{y}) \models \psi(\vec{x})$, we also have
-    $\exists \vec{x} ~.~ \phi(\vec{x},\vec{y}) \models \exists \vec{x}
-    ~.~ \psi(\vec{x})$. 
-  \end{fact}
-  \proof{ For each choice of a valuation for the existentially
-    quantified variables on the left-hand side, we chose the same
-    valuation for the variables on the right-hand side. \qed}\qed}
-
-Observe that, by Lemma \ref{lemma:subst-quant}
-(\ref{it2:subst-quant}), the set of free variables of a GLI formula
-$I_k$ consists of \begin{inparaenum}[(i)]
-\item variables $\stamp{X}{\leq k}$ keeping track of data values seen
-  in the input at some earlier moment in time, and
-%
-\item variables that track past choices made within the transition
-  rules.
-\end{inparaenum}
-Basically, it is not important when exactly in the past a certain
-input has been read or when a choice has been made, as only the value
-of the variable determines the future behavior. Intuitively,
-existential quantification of these variables does the job of ignoring
-when in the past these values have been seen.
-
-The last ingredient of the lazy annotation semi-algorithm based on
-unfoldings consist in the implementation of the coverage check, when
-the unfolding of an automaton is labeled with conjunctions of
-existentially quantified formulae with predicate symbols, obtained
-from interpolation. By Definition \ref{def:coverage}, checking whether
-a given node $\alpha \in \dom(U)$ is covered amounts to finding a
-prefix $\alpha' \prefix \alpha$ and a node $\beta \in \dom(U)$ such
-that $U(\alpha') \models U(\beta)$, or equivalently, the formula
-$U(\alpha') \wedge \neg U(\beta)$ is unsatisfiable. However, the
-latter formula, in prenex form, has quantifier prefix in the language
-$\exists^*\forall^*$ and, as previously mentioned, the satisfiability
-problem for such formulae becomes undecidable when the data theory
-subsumes Presburger arithmetic \cite{Halpern91}.
-
-Nevertheless, if we require just a yes/no answer (i.e.\ not an
-interpolant) recently developped quantifier instantiation heuristics
-\cite{ReynoldsKK17} perform rather well in answering a large number of
-queries in this class. Observe, moreover, that coverage does not need
-to rely on a complete decision procedure. If the prover fails in
-answering the above satisfiability query, then the semi-algorithm
-assumes that the node is not covered and continues exploring its
-successors. Failure to compute complete coverage may lead to
-divergence (non-termination) and ultimately, to failure to prove
-emptiness, but does not affect the soundness of the semi-algorithm
-(real counterexamples will still be found).
-
-\section{Applications}
-\label{sec:applications}
-
-The main application of first-order alternating automata is checking
-inclusion between various classes of automata extended with variables
-ranging over infinite domains that recognize languages over infinite
-alphabets. The most widely known such classes are \emph{timed
-  automata} \cite{AlurDill94} and \emph{finite-memory (register)
-  automata} \cite{KaminskiFrancez94}. In both cases, complementation
-is not possible inside the class and inclusion is undecidable. Our
-contribution is providing a systematic semi-algorithm for these
-decision problems. In addition, the method described in
-\S\ref{sec:emptiness} can extend our previous \emph{generic register
-  automata} \cite{HolikIosifRogalewiczVojnar18} inclusion checking
-framework, by allowing monitor (right-hand side) automata to have
-local variables, that are not visible in the language\footnote{We
-  conjecture that hidden computations using invisible variables
-  (clocks) are the reason for which timed and register automata cannot
-  be complemented within their class.}.
-
-Another application is checking safety (mutual exclusion, absence of
-deadlocks, etc.) and liveness (termination, lack of starvation, etc.)
-properties of parameterized concurrent programs, consisting of an
-unbounded number of replicated threads that communicate via a fixed
-set of global variables (locks, counters, etc.). The verification of
-parametric programs has been reduced to checking the emptiness of a
-(possibly infinite) sequence of first-order alternating automata,
-called \emph{predicate automata} \cite{Farzan15,Farzan16}, encoding
-the inclusion of the set of traces of a parametric concurrent program
-into increasingly general \emph{proof spaces}, obtained by
-generalization of counterexamples. The program and the proof spaces
-are first-order alternating automata over the infinite alphabet of
-pairs consisting of program statements and thread identifiers.
-
-\subsection{Timed Automata}
-\label{sec:ta}
-
-The standard definition of a finite \emph{timed word} is a sequence of
-pairs $(a_1,\tau_1), \ldots, (a_n,\tau_n) \in (\Sigma \times
-\real)^*$, where $\real$ is the set of real numbers, such that $0 \leq
-\tau_i < \tau_{i+1}$, for all $i \in [1,n-1]$. Intuitively, $\tau_i$
-is the moment in time where the input event $a_i$ occurs. Given a set
-$C$ of \emph{clocks}, the set $\Phi(C)$ of \emph{clock constraints} is
-defined inductively as the set of formulae $x \leq c$, $x \geq c$,
-$\neg\delta$, $\delta_1 \wedge \delta_2$, where $x \in C$, $c \in
-\rat$ is a rational constant and $\delta,\delta_1,\delta_2 \in
-\Phi(X)$.
-
-A \emph{timed automaton} is a tuple $T = \tuple{\Sigma,S,S_0,F,C,E}$,
-where: $\Sigma$ is a finite set of input events, $S$ is a finite set
-of states, $S_0, F \subseteq S$ are sets of initial and final states,
-respectively, $C$ is a finite set of clocks and $E \subseteq S \times
-\Sigma \times S \times 2^C \times \Phi(C)$ is the set of transitions
-$(s,a,s',\lambda,\delta)$ from state $s$ to state $s'$ with symbol
-$a$, $\lambda$ is the set of clocks to be reset and $\delta$ is a
-clock constraint. A run of $T$ over a timed word $w = (a_1,\tau_1)
-\ldots (a_n,\tau_n)$ is a sequence $(s_0,\gamma_0) \ldots
-(s_n,\gamma_n)$, where $s_i \in S$, $\gamma_i : C \rightarrow \real$
-are clock valuations, for all $i \in [n]$, and: \begin{compactitem}
-\item $s_0 \in S_0$ and $\gamma_0(x)=0$ for all $x \in C$, 
-%
-\item for all $i \in [n]$, there exists a transition
-  $(s_i,a_i,s_{i+1},\lambda_i,\delta_i) \in E$ such that $\gamma_i +
-  \tau_{i+1} - \tau_i \models \delta_i$, and for all $x \in C$,
-  $\gamma_{i+1}(x) = 0$ if $x \in \lambda_i$ and $\gamma_{i+1}(x) = \gamma_i(x)
-  + \tau_{i+1}-\tau_i$, otherwise. Here $\tau_0 \isdef 0$ and $\gamma_i +
-  \tau_{i+1} - \tau_i$ is the valuation mapping each $x \in C$ to
-  $\gamma_i(x) + \tau_{i+1} - \tau_i$.
-\end{compactitem}
-The run is accepting iff $s_n \in F$, in which case $T$ accepts
-$w$. As usual, we denote by $\lang{T}$ the set of finite words
-accepted by $T$. It is well-known that, in general, there is no timed
-automaton accepting the complement language $(\Sigma\times\real)^*
-\setminus \lang{T}$ and, moreover, the language inclusion problem is
-undecidable \cite{AlurDill94}.
-
-Given a timed automaton $T = \tuple{\Sigma,S,S_0,F,C,E}$, we define a
-first-order alternating automaton $\A_T =
-\tuple{\Sigma,\set{t},Q_T,\iota_T,F_T,\Delta_T}$, with a single input
-variable $t$, ranging over $\real$, such that each timed word $w =
-(a_1,\tau_1) \ldots (a_n,\tau_n)$ corresponds to a unique data word
-$d(w) = (a_1,\nu_1) \ldots (a_n,\nu_n)$ such that $\nu_i(t) = \tau_i$
-for all $i \in [1,n]$ and $\lang{\A_T} = \set{d(w) \mid
-  w\in\lang{T}}$. The only difficulty here is capturing the fact that
-all the clocks of $T$ evolve at the same pace, which is easily done
-using a technique from \cite{Fribourg98}, which replaces
-each clock $x_i$ of $T$ by a variable $y_i$ tracking the difference
-between the values of $t$ and $x_i$.
-
-Formally, if $C = \set{x_1,\ldots,x_k}$ and $S =
-\set{s_1,\ldots,s_m}$, we define $Q_T \isdef \set{q_1, \ldots, q_m}$ ,
-where $\#(q_i) = k+1$ for all $i \in [1,m]$, $\iota_T \isdef
-\bigvee_{s_i \in S_0} q_i(0,\ldots,0)$, $F_T \isdef \set{q_i \mid s_i
-  \in F}$ and, for each transition $(s_i,a,s_j,\lambda,\delta) \in E$,
-$\Delta_T$ contains the rule:
-\[q_i(y_1,\ldots,y_k,z) \arrow{a(t)}{} t > z \wedge \delta(z-y_1,\ldots,z-y_k) \wedge q_j(y'_1,\ldots,y'_k,t)\]
-where $y'_i$ stands for $z$ if $x_i \in \lambda$ and for $y_i$, 
-otherwise.  Moreover, nothing else is in $\Delta_T$. We establish the
-following connection between a timed automaton and its corresponding
-first-order alternating automaton.
-
-\begin{proposition}\label{prop:timed-alternating}
-  Given a timed automaton $T = \tuple{\Sigma,S,S_0,F,C,E}$, the
-  first-order alternating automaton $\A_T =
-  \tuple{\Sigma,\set{t},Q_T,\iota_T,F_T,\Delta_T}$ recognizes the
-  language $\lang{\A_T} = \set{d(w) \mid w \in \lang{T}}$. 
-\end{proposition}
-\proof{ ``$\subseteq$'' Let $w = (a_1,\nu_1) \ldots (a_n,\nu_n) \in
-  \lang{\A_T}$ be a data word. We show the existence of a timed word
-  $(a_1,\tau_1) \ldots (a_n,\tau_n) \in \lang{T}$ such that
-  $\nu_i(t)=\tau_i$, for all $i \in [1,n]$, by induction on
-  $n\geq0$. In fact we shall prove the following stronger
-  statements: \begin{compactenum}
-  \item each execution of $\A_T$ over $w$ starting with a cube $c \in
-    \cube{\minsem{\iota_T}}$ is a linear tree, in which each node has
-    at most one child.
-    %
-  \item for each execution $q_{i_0}(d^0_1,\ldots,d^0_k,\tau_0) \ldots
-    q_{i_n}(d^n_1,\ldots,d^n_k,\tau_n)$ of $\A_T$, $T$ has an
-    execution $(s_{i_0}, \gamma_0) \ldots (s_{i_n}, \gamma_n)$ over
-    the timed word $(a_1, \tau_1) \ldots (a_n, \tau_n)$, such that,
-    for all $i \in [1,n]$ and all $\ell \in [1,k]$, we have
-    $\gamma_i(x_\ell) = \tau_{i-1} - d^i_\ell$.
-  \end{compactenum}
-  The first point above is by inspection of $\iota_T = \bigvee_{s_i
-    \in S_0} q_i(0,\ldots,0)$ and of the rules from
-  $\Delta_T$. Indeed, each minimal model of $\iota_T$ corresponds to a
-  cube $q(0,\ldots,0)$ and each rule has exactly one predicate atom on
-  its right-hand side, thus each node of the execution will have at
-  most one successor. The second point is by induction on $n\geq0$.
-
-  ``$\supseteq$'' Let $w = (a_1,\tau_1) \ldots (a_n,\tau_n) \in
-  \lang{T}$ be a timed word. By induction on $n\geq0$, we show that
-  for each run $(s_{i_0}, \gamma_0) \ldots (s_{i_n}, \gamma_n)$ of $T$
-  over $w$, $\A_T$ has a linear execution
-  $q_{i_0}(d^0_1,\ldots,d^0_k,\tau_0) \ldots
-  q_{i_n}(d^n_1,\ldots,d^n_k,\tau_n)$ such that, for all $i \in [1,n]$
-  and all $\ell \in [1,k]$, we have $\gamma_i(x_\ell) = \tau_{i-1} -
-  d^i_\ell$. \qed}
-
-An easy consequence is that the timed language inclusion problem
-``given timed automata $T_1$ and $T_2$, does $\lang{T_1} \subseteq
-\lang{T_2}$ ?'' is reduced in polynomial time to the emptiness problem
-$\lang{\A_{T_1}} \cap \lang{\overline{\A_{T_2}}} = \emptyset$, for
-which (\S\ref{sec:emptiness}) provides a semi-algorithm. Observe,
-moreover, that no transition quantifiers are needed to encode timed
-automata as first-order alternating automata.
-
-\subsection{Register Automata}
-\label{sec:ra}
-
-Finite-memory automata, most commonly referred to as register automata
-\cite{KaminskiFrancez94} are among the first attempts at lifting the
-finite alphabet restriction of classical Rabin-Scott automata. In a
-nutshell, a register automaton is a finite-state automaton equipped
-with a finite set of registers $x_1,\ldots,x_r$ able to copy input
-values and compare them with subsequent input. Consequently, basic
-results from classical automata theory, such as the pumping lemma or
-the closure under complement do not hold in this model and, moreover,
-inclusion of languages recognized by register automata is undecidable
-\cite{NevenSchwentickVianu03}.
-
-Let $\Sigma$ be an infinite alphabet, $\#$ be a symbol not in $\Sigma$
-and $r>0$ be an integer constant, denoting the number of registers. An
-\emph{assignment} is a word $\vec{v} = v_1 \ldots v_r$ such that if
-$v_i = v_j$ and $i \neq j$ then $v_i = \#$, for all $i,j \in
-     [1,r]$. We write $[\vec{v}]$ for the set $\set{v_i \mid i \in
-       [1,r]}$ of values in the assignment $\vec{v}$. A
-     \emph{finite-memory (register) automaton} is a tuple $R =
-     \tuple{S,q_0,\vec{u}, \rho, \mu, F}$, where $S$ is a finite set
-     of states, $q_0 \in S$ is the initial state, $\vec{u} = u_1
-     \ldots u_r$ is the initial assignment, $\rho : S \rightarrow
-            [1,r]$ is the reassignment partial function, $\mu
-            \subseteq S \times [1,r] \times S$ is the transition
-            relation and $F \subseteq S$ is the set of final states. A
-            run of $A$ over an input word $a_1 \ldots a_n \in
-            \Sigma^*$ is a sequence $(s_0,\vec{v}_0) \ldots
-            (s_n,\vec{v}_n)$ such that $\vec{v}_0 = \vec{u}$ and, for
-            all $i \in [1,n]$, exactly one of the following
-            holds: \begin{compactitem}
-\item if there exists $k\in[1,r]$ such that $a_i=(\vec{v}_{i-1})_k$
-  then $\vec{v}_i = \vec{v}_{i-1}$ and $(s_{i-1},k,s_i) \in \mu$, 
-%
-\item otherwise $a_i\not\in[\vec{v}_{i-1}]$, $\rho(s_{i-1})$ is
-  defined, $(\vec{v}_i)_{\rho(s_{i-1})} = a_i$, for each $k \in
-  [1,r] \setminus \set{\rho(s_{i-1})}$, we have $(\vec{v}_i)_k =
-  (\vec{v}_{i-1})_k$ and $(s_{i-1},\rho(s_{i-1}),s_i) \in \mu$.
-\end{compactitem}
-Intuitively, if the input symbol is already stored in some register,
-the automaton moves to the next state if, moreover, the transition
-relation allows it, otherwise it copies the input to the register
-indicated by the reassignment, erasing its the previous value, and
-moves according to the transition relation.
-
-The translation of register automata to first-order alternating
-automata is quite natural, because registers can be encoded as
-arguments of predicate atoms. Formally, given a register automaton $R
-= \tuple{S,s_0,\vec{u},\rho,\mu,F}$, such that $S =
-\set{s_0,\ldots,s_m}$, we define the alternating automaton $\A_R =
-\tuple{\set{\alpha},\set{x},Q_R,\iota_R,F_R,\Delta_R}$, where
-$\alpha \not\in \Sigma$, $Q_R \isdef \set{q_0,\ldots,q_m}$ and
-$\#(q_i) = r$ for all $i \in [m]$, $\iota_R \isdef q_0(\vec{u})$, $F_R
-\isdef \set{q_i \mid s_i \in F}$ and, for each transition $(s_i, k,
-s_j) \in \mu$, $\Delta_T$ contains the rule:
-\[q_i(y_1,\ldots,y_r) \arrow{\alpha(x)}{} y_k=x \wedge q_j(y_1,\ldots,y_r) ~\vee~
-\bigwedge_{i=1}^r x \neq y_i \wedge q_j(y_1,\ldots,y_{k-1},x,y_{k+1},
-\ldots, y_r)\] Moreover, nothing else is in $\Delta_R$. The connection
-between register automata and first-order alternating automata is
-stated below.
-
-\begin{proposition}\label{prop:register-alternating}
-  Given a register automaton $R = \tuple{S,s_0,\vec{u},\rho,\mu,F}$
-  over an infinite alphabet $\Sigma$, the first-order alternating
-  automaton $\A_R = \tuple{\set{\alpha},Q_R,\iota_R,F_R,\Delta_R}$
-  recognizes the languge $\lang{\A_R} = \set{ (\alpha,a_1) \ldots
-    (\alpha,a_n) \mid a_1 \ldots a_n \in \lang{R} }$. 
-\end{proposition}
-\proof{ ``$\subseteq$'' Let $w = (\alpha,a_1) \ldots (\alpha,a_n)
-  \in \lang{\A_R}$. First, it is easy to show that each execution of
-  $\A_R$, that starts in some cube $c \in \cube{\minsem{\iota_R}}$, is
-  a linear tree with labels $q_0(\vec{v}_0), \ldots, q_n(\vec{v}_0)$
-  such that $\vec{v}_0 = \vec{u}$. Second by induction on $n\geq0$, we
-  prove that $\A_R$ has a run as above over $w$ only if $R$ has a run
-  $(q_0,\vec{v}_0), \ldots, (q_n,\vec{v}_n)$ over $a_1 \ldots
-  a_n$. ``$\supseteq$'' Let $w = a_1 \ldots a_n \in \lang{R}$ and
-  $q_0(\vec{v}_0), \ldots, q_n(\vec{v}_0)$ be a run of $R$ over $w$,
-  such that $\vec{v}_0 = \vec{u}$. By induction on $n\geq0$, we can
-  build an execution of $\A_R$ over $(\alpha,a_1) \ldots
-  (\alpha,a_n)$ that is a linear tree with labels $q_0(\vec{v}_0),
-  \ldots, q_n(\vec{v}_n)$. \qed}
-
-Consequently, the language inclusion problem ``given register automata
-$R_1$ and $R_2$, does $\lang{R_1} \subseteq \lang{R_2}$?'' is reduced
-in polynomial time to emptiness problem $\lang{\A_{R_1}} \cap
-\lang{\overline{\A_{R_2}}} = \emptyset$, for which
-(\S\ref{sec:emptiness}) provides a semi-algorithm. Notice further that
-the encoding of register automata as first-order alternating automata
-uses no transition quantifiers.
-
-\subsection{Predicate Automata}
-\label{sec:pa}
-
-The model of \emph{predicate automata} \cite{Farzan15,Farzan16} has
-emerged recently as a tool for checking safety and liveness properties
-of parameterized concurrent programs, in which there is an unbounded
-number of replicated threads that communicate via global
-variables. Predicate automata recognize finite sequences of actions
-that are pairs $(\sigma,i)$, where $\sigma$ is from a finite set
-$\Sigma$ of program statements and $i \in \nat$ ranges over an
-unbounded set of thread identifiers. To avoid clutter, we shall view a
-pair $(\sigma,i)$ as a data symbol $(\sigma,\nu)$ where $\nu(x)=i$,
-for a designated input variable $x$.
-
-Since thread identifiers can only be compared for equality, the data
-theory of predicate automata is the first-order theory of
-equality. Moreover, transition quantifiers are only needed for
-checking termination and, generally, liveness properties
-\cite{Farzan16}.
-
-However, the execution semantics of predicate automata differs from
-that of first-order automata with respect to the following detail:
-initial configurations and successors of predicate automata are
-defined using the entire sets of models of the initial sentence and
-transition rules, not just the minimal ones, as in our case.
-
-Formally, a run of a predicate automaton $P =
-\tuple{\Sigma,\set{x},Q,\iota,F,\Delta}$ over a word $(a_1,\nu_1)
-\ldots$ $(a_n,\nu_n)$ is a sequence of interpretations $\I_0, \ldots,
-\I_n$ such that $\I_0 \in \sem{\iota}$ and for each $i \in [1,n]$,
-each $q \in Q$ and each tuple $\tuple{d_1,\ldots,d_{\#(q)}} \in
-\I_{i-1}(q)$, we have $\I_i \in \sem{\psi}_\mu$, for each rule
-$q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(x)}{} \psi \in \Delta$, where
-$\mu=\nu_i[y_1\leftarrow d_1, \ldots, y_{\#(q)} \leftarrow
-  d_{\#(q)}]$. The run is accepting if and only if $\I(q)=\emptyset$
-for all $q \in Q \setminus F$.
-
-In fact, as shown next, this more simple execution semantics is
-equivalent, from the language point of view, with the semantics given
-by Definitions \ref{def:execution} and \ref{def:accepting}.  We
-believe that the semantics of first-order alternating automata based
-on minimal models is important for its relation to the textbook
-semantics of boolean alternating automata
-\cite{ChandraKozenStockmeyer81}.
-
-\begin{proposition}
-  Given a predicate automaton $P =
-  \tuple{\Sigma,\set{x},Q,\iota,F,\Delta}$, let $\A_P$ be the
-  first-order alternating automaton that has the same description as
-  $P$. Then $\lang{P} = \lang{\A_P}$. 
-\end{proposition}
-\proof{
-\qed}
-As before, this result enables using the semi-algorithm from
-\S\ref{sec:emptiness} for checking emptiness of predicate automata.
-We point out that, although quantifier-free predicate automata with
-predicates of arity one are decidable for emptiness \cite{Farzan15},
-currently there is no method for checking emptiness of predicate
-automata with predicates of arity greater than one, other than the
-explicit enumeration of cubes. Moreover, no method for dealing with
-emptiness in the presence of transition quantifiers is known to exist.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{abbrv} \bibliography{refs}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\end{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-11-06 quantifiers/body.tex
ok

--- 2018-10-22 quantifiers/draft.tex
@@
-\documentclass{llncs}
-% \documentclass[10pt,conference,letterpaper,twocolumn]{IEEEtran}
-
-\usepackage[usenames,dvipsnames]{color}
-
-\usepackage{latexsym}
-\usepackage{amsxtra} 
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{pslatex}
-\usepackage{epsfig}
-\usepackage{wrapfig}
-\usepackage{paralist}
-% \usepackage{stmaryrd}
-\usepackage{txfonts}
-\usepackage{framed}
-\usepackage{makecell}
-\usepackage{url}
-\usepackage{tikz}
-% \usepackage{hyperref}
-\usetikzlibrary{automata,positioning, calc}
-\usepackage[inline,shortlabels]{enumitem}
-
-\usepackage{proof}
-
-\usepackage{algorithm}
-\usepackage{algorithmicx}
-\usepackage[noend]{algpseudocode}
-
-\usepackage[draft]{commenting}
-
-\pagestyle{plain}
-
-\include{commands-lncs}
-
-\declareauthor{ri}{Radu}{blue}
-\declareauthor{xx}{Xiao}{red}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\title{First Order Alternation}
-
-\author{Radu Iosif and Xiao Xu}
-\institute{
-  CNRS, Verimag, Universit\'e de Grenoble Alpes \\
-  Email: \{Radu.Iosif,Xiao.Xu\}@univ-grenoble-alpes.fr
-}
-
-\maketitle
-
-\begin{abstract}
-We introduce first order alternating automata, a generalization of
-boolean alternating automata, in which transition rules are described
-by multisorted first order formulae, with states and internal
-variables given by uninterpreted predicate terms. The model is closed
-under union, intersection and complement, and its emptiness problem is
-undecidable, even for the simplest data theory of equality. To cope
-with this limitation, we develop an abstraction refinement
-semi-algorithm based on lazy annotation of the symbolic execution
-paths with interpolants, obtained by applying (i) quantifier
-elimination with witness term generation and (ii) Lyndon interpolation
-in the quantifier-free data theory with uninterpreted predicate
-symbols. This provides a method for checking inclusion of timed and
-finite-memory register automata, and emptiness of quantified predicate
-automata, previously used in the verification of parameterized
-concurrent programs, composed of replicated threads, with a
-shared-memory communication model.
-\end{abstract}
-
 \section{Introduction}
@@
 the forcing relation $\I,\nu \models \phi$ recursively on the
-structure of $\phi$, as follows:
+structure of $\phi$, as usual. 
+\ifLongVersion
 \[
@@
 where $\nu[x\leftarrow d]$ is the valuation which assigns $d$ to $x$
-and behaves like $\nu$ elsewhere. For a formula $\phi$ and a valuation
-$\nu$, we define $\sem{\phi}_\nu \isdef \set{\I \mid \I,\nu \models
-  \phi}$ and drop the $\nu$ subscript for sentences. A sentence $\phi$
-is \emph{satisfiable} (\emph{unsatisfiable}) if
-$\sem{\phi}\neq\emptyset$ ($\sem{\phi}=\emptyset$). An element of
-$\sem{\phi}$ is called a \emph{model} of $\phi$. A formula $\phi$ is
-\emph{valid} if $\I,\nu \models \phi$ for every interpretation $\I$
-and every valuation $\nu$. For two formulae $\phi$ and $\psi$ we write
-$\phi \models \psi$ for $\sem{\phi} \subseteq \sem{\psi}$, in which
-case we say that $\phi$ \emph{entails} $\psi$.
+and behaves like $\nu$ elsewhere.  
+\fi 
+For a formula $\phi$ and a valuation $\nu$, we define $\sem{\phi}_\nu
+\isdef \set{\I \mid \I,\nu \models \phi}$ and drop the $\nu$ subscript
+for sentences. A sentence $\phi$ is \emph{satisfiable}
+(\emph{unsatisfiable}) if $\sem{\phi}\neq\emptyset$
+($\sem{\phi}=\emptyset$). An element of $\sem{\phi}$ is called a
+\emph{model} of $\phi$. A formula $\phi$ is \emph{valid} if $\I,\nu
+\models \phi$ for every interpretation $\I$ and every valuation
+$\nu$. For two formulae $\phi$ and $\psi$ we write $\phi \models \psi$
+for $\sem{\phi} \subseteq \sem{\psi}$, in which case we say that
+$\phi$ \emph{entails} $\psi$.
 
@@
   \stamp{\iota}{0}$ and, for all $i \in [1,n]$:
+  \vspace*{-.5\baselineskip}
   \[\quantformn{\alpha_i} \isdef \quantformn{\alpha_{i-1}} \wedge
@@
     \stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]
-      \]  We write
+      \]  
+      \vspace*{-.5\baselineskip}
+      We write
     $\quantform{\alpha}$ for the prenex normal form of the formula: 
@@
 (booleans) and no transition quantifiers. For simplicity, we do not
-present the details of this algorithm (see Appendix
-\ref{app:pseudocode}) and shall content ourselves of several
-high-level definitions.
+present the details of this algorithm and shall content ourselves of
+several high-level definitions.
 
@@
 unfolding has been found. For a detailed presentation of the emptiness
-procedure, we refer the reader to \cite{IosifXu18} (or Appendix
-\ref{app:pseudocode}).
+procedure, we refer the reader to \cite{IosifXu18}.
 
@@
 the following transition rules and final state $q_f$:
-\[\begin{array}{rclcrcl}
+\[\begin{array}{rclcrclcrcl}
 q_0(y) & \arrow{a(x)}{} & q_1(y+x) \wedge q_2(y-x) & \hspace*{2mm} & 
-q_1(y) \arrow{a(x)}{} y+x > 0 \wedge q_f \\
-q_1(y) & \arrow{a(x)}{} & q_1(y+x) && q_2(y) \arrow{a(x)}{} y-x > 0 \wedge q_f \\
+q_1(y) & \arrow{a(x)}{} & y+x > 0 \wedge q_f & \hspace*{2mm} & 
+q_1(y) & \arrow{a(x)}{} & q_1(y+x) \\ 
+q_2(y) & \arrow{a(x)}{} & y-x > 0 \wedge q_f && 
 q_2(y) & \arrow{a(x)}{} & q_2(y-x)
@@
   where $Q_1, \ldots, Q_n \in \set{\exists,\forall}$ and $\phi$ is
-  quantifier-free, let: 
-  \[\eta_i \equiv \left\{\begin{array}{ll}
-  f_i(y_1,\ldots,y_{k_i}) & \text{if } Q_i = \forall \\
-  x_i & \text{if } Q_i = \exists
-  \end{array}\right.\]
-  where $f_i$ is a fresh function symbol and $\set{y_1, \ldots,
-    y_{k_i}} = \set{x_j \mid j < i,~ Q_j = \exists}$. Then the entailment \(Q_1
-  x_1 \ldots Q_n x_n ~.~ \phi \models
-  \phi[\eta_1/x_1,\ldots,\eta_n/x_n]\) holds. 
+  quantifier-free, let $\eta_i \isdef f_i(y_1,\ldots,y_{k_i})$ if $Q_i
+  = \forall$ and $\eta_i \isdef x_i$ if $Q_i = \exists$, where $f_i$
+  is a fresh function symbol and $\set{y_1, \ldots, y_{k_i}} =
+  \set{x_j \mid j < i,~ Q_j = \exists}$. Then the entailment \(Q_1 x_1
+  \ldots Q_n x_n ~.~ \phi \models \phi[\eta_1/x_1,\ldots,\eta_n/x_n]\)
+  holds.
 \end{theorem}
@@
 yields the unsatisfiable conjunction: 
+\vspace*{-\baselineskip}
 \[\begin{array}{c}
-z_1 \geq 0 \wedge \stamp{q}{0}(z_1)  \\ \wedge \\
-\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge (z_1 \geq z_1 \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_1)) \\ \wedge \\
+z_1 \geq 0 \wedge \stamp{q}{0}(z_1)  ~\wedge~
+\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge (z_1 \geq z_1 \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_1)) ~\wedge~ \\
 \stamp{q}{1}(\stamp{x}{1}+z_1) \rightarrow \stamp{x}{1}+z_1 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_1)
@@
   \item\label{it1:subst-quant} one can effectively build witness terms
-    $\tau_{i_1}, \ldots, \tau_{i_\ell}$, where 
-    \[\begin{array}{l}   
-    \set{i_1, \ldots,i_\ell} = \set{j \in [1,m] \mid Q_j = \forall} \text{ and} \\
-    \voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup \set{x_k
-    \mid k < i_j, Q_k = \exists},~ \forall j \in [1,\ell]
-    \end{array}\]
+    $\tau_{i_1}, \ldots, \tau_{i_\ell}$, where \(\set{i_1,
+    \ldots,i_\ell} = \set{j \in [1,m] \mid Q_j = \forall}\) and
+    \(\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup
+    \set{x_k \mid k < i_j, Q_k = \exists},~ \forall j \in [1,\ell]\)
     such that $\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
@@
   \item\label{it2:subst-quant} one can effectively build a GLI $(I_0,
-    \ldots, I_n)$ for $\alpha$, such that \[\voc{I_k} \subseteq
+    \ldots, I_n)$ for $\alpha$, such that \(\voc{I_k} \subseteq
     \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
-      \xi^{-1}(k),~ Q_j = \exists},~ \forall k \in [n]\enspace.\]
+      \xi^{-1}(k),~ Q_j = \exists},~ \forall k \in [n]\).
   \end{compactenum}
@@
 
+\ifLongVersion
 \section{Applications}
@@
 is known to exist.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{abbrv} \bibliography{refs}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\end{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\fi

### 2018-11-06 quantifiers/draft.tex
ok

--- 2018-10-18 quantifiers/draft.tex
@@
 
+\renewcommand{\baselinestretch}{0.99}
+
 \declareauthor{ri}{Radu}{blue}
@@
 
-\begin{abstract}
-\end{abstract}
-
-\section{Preliminaries}
-
-For two integers $0 \leq i \leq j$, we denote by $[i,j]$ the set
-$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[0,i]$. We consider two
-sorts $\Data$ and $\Bool$, where $\Data$ is an infinite domain and
-$\Bool = \set{\top,\bot}$ is the set of boolean values true ($\top$)
-and false ($\bot$), respectively. The $\Data$ sort is equipped with
-finitely many function symbols $f : \Data^{\#(f)} \rightarrow \Data$,
-where $\#(f)\geq0$ denotes the number of arguments (arity) of
-$f$. When $\#(f)=0$, we say that $f$ is a constant. A \emph{predicate}
-is a function symbol $p : \Data^{\#(p)} \rightarrow \Bool$, denoting a
-relation of arity $\#(p)$ and we write $\preds$ for the set of
-predicates.
-
-In the following, we shall consider that the interpretation of all
-function symbols $f : \Data^{\#(f)} \rightarrow \Data$ that are not
-predicates is fixed by the interpretation of the $\Data$ sort,
-e.g.\ if $\Data$ is the set of integers $\zed$, the function symbols
-are zero, the successor function and the arithmetic operations of
-addition and multiplication. For simplicity, we further blur the
-notational distinction between function symbols and their
-interpretations.
-
-Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
-variables, ranging over $\Data$. Terms are either constants of sort
-$\Data$, variables or function applications $f(t_1,\ldots,t_{\#(f)})$,
-where $t_1,\ldots,t_{\#(f)}$ are terms. The set of first-order
-formulae is defined by the syntax below:
-\[\phi := t \teq s \mid p(t_1,\ldots,t_{\#(p)}) 
-\mid \neg \phi_1 \mid \phi_1 \wedge \phi_2 \mid \exists x ~.~
-\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms. We write
-$\phi_1 \vee \phi_2$, $\phi_1 \rightarrow \phi_2$ and $\forall x ~.~
-\phi_1$ for $\neg(\neg\phi_1 \wedge \neg\phi_2)$, $\neg\phi_1 \vee
-\phi_2$ and $\neg\exists x ~.~ \neg\phi_1$, respectively. We denote by
-$\fv{}{\phi}$ the set of free variables in $\phi$. The size
-$\len{\phi}$ of a formula $\phi$ is the number of symbols needed to
-write it down.
-
-A \emph{sentence} is a formula $\phi$ in which each variable occurs
-under the scope of a quantifier, i.e. $\fv{}{\phi} = \emptyset$.  A
-formula is \emph{positive} if each predicate symbol occurs under an
-even number of negations and we denote by $\posforms(Q,X)$ the set of
-positive formulae with predicates from the set $Q \subseteq \preds$
-and free variables from the set $X \subseteq \vars$.
-
-A formula is in \emph{prenex form} if it is of the form $\varphi =
-Q_1x_1 \ldots Q_nx_n ~.~ \phi$, where $\phi$ has no quantifiers. In
-this case we call $\phi$ the \emph{matrix} of $\varphi$. Every
-first-order formula can be written in prenex form, by renaming each
-quantified variable to a unique name and moving the quantifiers
-upfront.
-
-%% For a variable $x \in \fv{}{\phi}$ and a term $t$, let $\phi[t/x]$ be
-%% the result of replacing each occurrence of $x$ by $t$. For indexed
-%% sets of terms $T=\set{t_1,\ldots,t_n}$ and variables
-%% $X=\set{x_1,\ldots,x_n}$, we write $\phi[T/X]$ for the formula
-%% obtained by simultaneously replacing each occurrence of $x_i$ with
-%% $t_i$ in $\phi$, for all $i\in[n]$.
-
-An \emph{interpretation} $\I$ maps each predicate $p$ into a set $p^\I
-\subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element of $\Data$
-if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable $x$ into an
-element of $\Data$. Given a term $t$, we denote by $t^\nu$ the value
-obtained by replacing each variable $x$ by the value $\nu(x)$ and
-evaluating each function application. For a formula $\phi$, we define
-the forcing relation $\I,\nu \models \phi$ recursively on the
-structure of $\phi$, as follows:
-\[
-\begin{array}{rcl}
-\I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
-\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I \\
-\I,\nu \models \neg\phi_1 & \iff & \I,\nu \not\models \phi_1 \\
-\I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
-\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data 
-\end{array}
-\]
-where $\nu[x\leftarrow d]$ is the valuation which assigns $d$ to $x$
-and behaves like $\nu$ elsewhere. For a formula $\phi$ and a valuation
-$\nu$, we define $\sem{\phi}_\nu \isdef \set{\I \mid \I,\nu \models
-  \phi}$ and drop the $\nu$ subscript for sentences. A sentence $\phi$
-is \emph{satisfiable} (\emph{unsatisfiable}) if
-$\sem{\phi}\neq\emptyset$ ($\sem{\phi}=\emptyset$). An element of
-$\sem{\phi}$ is called a \emph{model} of $\phi$. A formula $\phi$ is
-\emph{valid} if $\I,\nu \models \phi$ for every interpretation $\I$
-and every valuation $\nu$. For two formulae $\phi$ and $\psi$ we write
-$\phi \models \psi$ for $\sem{\phi} \subseteq \sem{\psi}$, in which
-case we say that $\phi$ \emph{entails} $\psi$.
-
-Interpretations are partially ordered by the pointwise subset order,
-defined as $\I_1 \subseteq \I_2$ if and only if $p^{\I_1} \subseteq
-p^{\I_2}$ for each predicate $p \in \preds$. Given a set $\mathcal{S}$
-of interpretations, a minimal element $\I \in \mathcal{S}$ is an
-interpretation such that for no other interpretation $\I' \in
-\mathcal{S} \setminus \set{\I}$ do we have $\I' \subseteq \I$. For a
-formula $\phi$ and a valuation $\nu$, we denote by $\minsem{\phi}_\nu$
-and $\minsem{\phi}$ the set of minimal interpretations from
-$\sem{\phi}_\nu$ and $\sem{\phi}$, respectively.
-
-\section{First Order Alternating Automata}
-
-Let $\Sigma$ be a finite alphabet $\Sigma$ of \emph{input
-  events}. Given a finite set of variables $X \subseteq \vars$, we
-denote by $X \mapsto \Data$ the set of valuations of the variables $X$
-and $\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly
-infinite set of \emph{data symbols} $(a,\nu)$, where $a$ is an input
-symbol and $\nu$ is a valuation. A \emph{data word} (simply called
-word in the following) is a finite sequence $w=(a_1,\nu_1)(a_2,\nu_2)
-\ldots (a_n,\nu_n)$ of data symbols. Given a word $w$, we denote by
-$\event{w} \isdef a_1\ldots a_n$ its sequence of input events and by
-$\data{w}$ the valuation associating each time-stamped variable
-$\stamp{x}{i}$ the value $\nu_i(x)$, for all $x \in \vars$ and
-$i\in[1,n]$. We denote by $\varepsilon$ the empty sequence, by
-$\Sigma^*$ the set of finite sequences of input events and by
-$\Sigma[X]^*$ the set of data words over the variables $X$.
-
-Formally, a \emph{first-order alternating automaton} is a tuple $\A =
-\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is a finite set of
-input events, $X$ is a finite set of input variables, $Q$ is a finite
-set of predicates denoting control states, $\iota \in
-\posforms(Q,\emptyset)$ is a sentence defining initial configurations,
-$F \subseteq Q$ is the set of predicates denoting final states, and
-$\Delta$ is a set of \emph{transition rules} of the form
-\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
-predicate, $a \in \Sigma$ is an input event and $\psi \in
-\posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive formula,
-where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$. The quantifiers
-occurring in the right-hand side formula of a transition rule are
-referred to as \emph{transition quantifiers}. The \emph{size} of $\A$
-is defined as $\len{\A} = \len{\iota} +
-\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
-\len{\psi}$.
-
-The intuition of a transition rule \(q(y_1,\ldots,y_{\#(q)})
-\arrow{a(X)}{} \psi\) is the following: $a$ is the input event and $X$
-are the input data values that trigger the transition, whereas $q$ and
-$y_1,\ldots,y_{\#(q)}$ are the current control state and data values
-in that state, respectively. Without loss of generality, we consider,
-for each predicate $q \in Q$ and each input event $a \in \Sigma$, at
-most one such rule, as two or more rules can be joined using
-disjunction.
-
-The execution semantics of automata is given in close analogy with the
-case of boolean alternating automata, with transition rules of the
-form $q \arrow{a}{} \phi$, where $q$ is a boolean constant and $\phi$
-a positive boolean combination of such constants. For instance, $q_0
-\arrow{a}{} q_1 \wedge q_2 \vee q_3$ means that the automaton can
-choose to transition in either both $q_1$ and $q_2$ or in $q_3$
-alone. This intuition leads to saying that the steps of the automaton
-are defined by the minimal boolean models of the transition
-formulae. In this case, both $\set{q_1 \leftarrow \top, q_2 \leftarrow
-  \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow \bot, q_2
-  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, however
-$\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
-is a model but is not minimal. The original definition of alternating
-finite-state automata \cite{ChandraKozenStockmeyer81} works around
-this problem by considering boolean valuations (models) instead of
-formulae.
-
-Given a predicate $q \in Q$ and a tuple of data values
-$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
-a \emph{configuration}\footnote{Note that a configuration is not a
-  logical term since data values cannot be written in logic.}. To
-formalize the execution semantics of automata, we relate sets of
-configurations to models of first-order sentences, as follows. Given a
-formula $\phi$ and a valuation of its free variables $\nu$, each
-interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
-configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid q
-  \in Q,~ \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a
-\emph{cube}. For a set $\mathcal{S}$ of interpretations, we define
-$\cube{\mathcal{S}} \isdef \set{\cube{\I} \mid \I \in
-  \mathcal{S}}$.
-
-\begin{definition}\label{def:execution}
-Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$ and a cube $c$, an
-\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$, starting with $c$, is
-a (possibly infinite) forest $\T = \set{T_1,T_2,\ldots}$, where each
-$T_i$ is a tree labeled with configurations, such that:
-\begin{compactenum}
-\item\label{it1:execution} $c = \set{T(\epsilon) \mid T \in \T}$ is the set of
-  configurations labeling the roots of $T_1,T_2,\ldots$ and
-%
-\item\label{it2:execution} if $q(d_1,\ldots,d_{\#(q)})$ labels a node on the level $j \in
-  [n-1]$ in $T_i$, then the labels of its children form a cube from
-  $\cube{\minsem{\psi}_{\eta}}$, where $\eta = \nu_{j+1}[y_1
-    \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$ and
-  \(q(y_1,\ldots,y_{\#(q)}) \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is
-  a transition rule of $\A$.
-\end{compactenum}
-\end{definition}
-
-\begin{definition}\label{def:accepting}
-An execution $\T$ over $w$, starting with $c$, is \emph{accepting} if
-and only if \begin{compactitem}
-\item\label{it1:accepting} all paths in $\T$ have the same length $n$,
-  and
-  %
-\item\label{it2:accepting} the frontier of each tree $T \in \T$ is
-  labeled with \emph{final configurations} $q(d_1,\ldots,d_{\#(q)})$,
-  where $q \in F$.
-\end{compactitem} 
-If $\A$ has an accepting execution over $w$ starting with a cube $c
-\in \cube{\minsem{\iota}}$, then $\A$ \emph{accepts} $w$ and let
-$\lang{\A}$ be the set of words accepted by $\A$.
-\end{definition}
-In this paper, we address the following questions: \begin{compactenum}
-\item \emph{boolean closure}: given automata $\A_i =
-  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
-  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
-  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
-  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
-%
-\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
-  \emptyset$?
-\end{compactenum}
-
-\subsection{Symbolic Execution}
-
-In the upcoming developments it is sometimes more convenient to work
-with logical formulae defining executions of automata, than with
-low-level execution forests. For this reason, we first introduce
-\emph{path formulae} $\pathform{\alpha}$, which are formulae defining
-the executions of an automaton, over words that share a given sequence
-$\alpha$ of input events.  Second, we restrict a path formula
-$\pathform{\alpha}$ to an \emph{acceptance formula}
-$\accform{\alpha}$, which defines only accepting executions over words
-that share a given input sequence. Otherwise stated,
-$\accform{\alpha}$ is satisfiable if and only if the automaton accepts
-a word $w$ such that $\event{w} = \alpha$.
-
-Let $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the
-rest of this section. For any $i \in \nat$, we denote by $\stamp{Q}{i}
-= \set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} =
-\set{\stamp{x}{i} \mid x \in X}$ the sets of time-stamped predicates
-and variables, respectively. As a shorthand, we write $\stamp{Q}{\leq
-  n}$ (resp.  $\stamp{X}{\leq n}$) for the set $\set{\stamp{q}{i} \mid
-  q \in Q, i \in [n]}$ (resp. $\set{\stamp{x}{i} \mid x \in X, i \in
-  [n]}$). For a formula $\psi$ and $i \in \nat$, we define
-$\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the
-formula in which all input variables and state predicates (and only
-those symbols) are replaced by their time-stamped counterparts. As a
-shorthand, we shall write $q(\vec{y})$ for $q(y_1,\ldots,y_{\#(q)})$,
-when no confusion arises.
-
-Given a sequence of input events $\alpha = a_1 \ldots a_n$, the
-\emph{path formula} of $\alpha$ is defined as:
-\begin{equation}\label{eq:pathform}
-\pathform{\alpha} \isdef \stamp{\iota}{0} \wedge 
-\bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{}
-  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-\stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i}
-\end{equation}
-The automaton $\A$, to which $\pathform{\alpha}$ refers, will always
-be clear from the context. To formalize the relation between the
-low-level configuration-based execution semantics and the symbolic
-path formulae, consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any
-execution forest $\T$ of $\A$ over $w$ is associated an interpretation
-$\I_{\T}$ of the set of time-stamped predicates $\stamp{Q}{\leq n}$,
-defined as:
-\[\I_{\T}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
-  q(d_1,\ldots,d_{\#(q)}) \text{ labels a node on level $i$ in $\T$}} 
-,~\forall q\in Q ~\forall i \in [n]\]
-
-\begin{lemma}\label{lemma:path-formula}
-  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for any
-  word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have 
-  \(\minsem{\pathform{\event{w}}}_{\data{w}} = \set{\I_{\T} \mid \T
-    \text{ is an execution of $\A$ over $w$}}\). 
-\end{lemma}
-\proof{ ``$\subseteq$'' Let $\I$ be a minimal interpretation such that
-  $\I,\data{w} \models \pathform{\event{w}}$. We show that there exists
-  an execution $\T$ of $\A$ over $w$ such that $\I = \I_\T$, by
-  induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
-  $\pathform{\event{w}} = \stamp{\iota}{0}$. Because $\iota$ is a
-  sentence, the valuation $\data{w}$ is not important in $\I,\data{w}
-  \models \stamp{\iota}{0}$ and, moreover, since $\I$ is minimal, we
-  have $\I \in \minsem{\stamp{\iota}{0}}$. We define the
-  interpretation $\J(q)=\I(\stamp{q}{0})$, for all $q \in Q$. Then
-  $\cube{\J}$ is an execution of $\A$ over $\epsilon$ and $\I =
-  \I_{\cube{\J}}$ is immediate. For the inductive case $n>0$, we
-  assume that $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\J$ be the
-  interpretation defined as $\I$ for all $\stamp{q}{i}$, with $q \in
-  Q$ and $i \in [n-1]$, and $\emptyset$ everywhere else. Then
-  $\J,u_\Data \models \pathform{u_\Sigma}$ and $\J$ is moreover
-  minimal. By the induction hypothesis, there exists an execution $\G$
-  of $\A$ over $u$, such that $\J = \I_\G$. Consider a leaf of a tree
-  $T \in \G$, labeled with a configuration $q(d_1,\ldots,d_{\#(q)})$
-  and let $\forall y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$ be the
-  subformula of $\pathform{\event{w}}$ corresponding to the
-  application(s) of the transition rule $q(\vec{y}) \arrow{a_n}{}
-  \psi$ at the $(n-1)$-th step. Let $\nu = \data{w}[y_1 \leftarrow
-    d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$. Because $\I, \data{w}
-  \models \forall y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$, we have $\I
-  \in \sem{\stamp{\psi}{n}}_{\nu}$ and let $\K$ be one of the minimal
-  interpretations such that $\K \subseteq \I$ and $\K \in
-  \sem{\stamp{\psi}{n}}_{\nu}$. It is not hard to see that $\K$ exists
-  and is unique, otherwise we could take the pointwise intersection of
-  two or more such interpretations. We define the interpretation
-  $\overline{\K}(q) = \overline{\K}(\stamp{q}{n})$ for all $q \in
-  Q$. We have that $\overline{\K} \in \minsem{\psi}_{\nu}$ --- if
-  $\overline{\K}$ was not minimal, $\K$ was not minimal to start with,
-  contradiction. Then we extend the execution $\G$ by appending to
-  each node labeled with a configuration $q(d_1, \ldots, d_{\#(q)})$
-  the cube $\cube{\overline{\K}}$. By repeating this step for all
-  leaves of a tree in $\G$, we obtain an execution of $\A$ over $w$.
-
-  ``$\supseteq$'' Let $\T$ be an execution of $\A$ over $w$. We show
-  that $\I_{\T}$ is a minimal interpretation such that $\I_{\T},
-  \data{w} \models \pathform{\event{w}}$, by induction on $n \geq
-  0$. For $n=0$, $\T$ is a cube from $\cube{\minsem{\iota}}$, by
-  definition. Then $\I_{\T} \models \stamp{\iota}{0}$ and moreover, it
-  is a minimal such interpretation. For the inductive case $n > 0$,
-  let $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\G$ be the
-  restriction of $\T$ to $u$. Consequently, $\I_\G$ is the restriction
-  of $\I_\T$ to $\stamp{Q}{\leq n-1}$. By the inductive hypothesis,
-  $\I_{\G}$ is a minimal interpretation such that \(\I_{\G}, u_{\Data}
-  \models \pathform{u_\Sigma}\). Since $\I_{\T}(\stamp{q}{n}) =
-  \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid q(d_1,\ldots,d_{\#(q)})
-    \text{ labels a node on the $n$-th level in $\T$}}$, we have $\I_{\T},
-  \data{w} \models \varphi$, for each subformula $\varphi = \forall y_1
-  \ldots \forall y_{\#(q)} ~.~ \stamp{q}{n-1}(\vec{y})
-  \rightarrow \stamp{\psi}{n}$ of $\pathform{\event{w}}$, by the
-  execution semantics of $\A$. This is the case because the children
-  of each node labeled with $q(d_1,\ldots,d_{\#(q)})$ on the
-  $(n-1)$-th level of $\T$ form a cube from
-  $\cube{\minsem{\psi}_\nu}$, where $\nu$ is a valuation that assigns
-  each $y_i$ the value $d_i$ and behaves like $\data{w}$,
-  otherwise. Now supppose, for a contradiction, that $\I_{\T}$ is not
-  minimal and let $\J \subsetneq \I_{\T}$ be an interpretation such
-  that $\J,\data{w} \models \pathform{\event{w}}$. First, we show that
-  the restriction $\J'$ of $\J$ to $\bigcup_{i=0}^{n-1} \stamp{Q}{i}$
-  must coincide with $\I_{\G}$. Assuming this is not the case,
-  i.e.\ $\J' \subsetneq \I_{\G}$, contradicts the minimality of
-  $\I_{\G}$. Then the only possibility is that $\J(\stamp{q}{n})
-  \subsetneq \I_{\T}(\stamp{q}{n})$, for some $q \in Q$. Let
-  $p_1(y_1,\ldots,y_{\#(p_1)}) \arrow{a_n}{} \psi_1, \ldots,
-  p_k(y_1,\ldots,y_{\#(p_k)}) \arrow{a_n}{} \psi_k$ be the set of
-  transition rules in which the predicate symbol $q$ occurs on the
-  right-hand side. Then it must be the case that, for some node on the
-  $(n-1)$-th level of $\G$, labeled with a configuration
-  $p_i(d_1,\ldots,d_{\#(p_i)})$, the set of children does not form a
-  minimal cube from $\cube{\minsem{\stamp{\psi_i}{n}}}$, which
-  contradicts the execution semantics of $\A$. \qed}
-
-Next, we give a logical characterization of acceptance, relative to a
-given sequence of input events $\alpha \in \Sigma^*$. To this end, we
-constrain the path formula $\pathform{\alpha}$ by requiring that only
-final states of $\A$ occur on the last level of the execution.  The
-result is the \emph{acceptance formula} for $\alpha$:
-\begin{equation}\label{eq:accform}
-\accform{\alpha} \isdef \pathform{\alpha} \wedge \bigwedge_{q
-  \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-\stamp{q}{n}(\vec{y}) \rightarrow \bot
-\end{equation}
-The top-level universal quantifiers from a subformula $\forall y_1
-\ldots \forall y_{\#(q)} ~.~ \stamp{q}{i}(\vec{y}) \rightarrow \psi$
-of $\accform{\alpha}$ will be referred to as \emph{path quantifiers},
-in the following. Notice that path quantifiers are distinct from the
-transition quantifiers that occur within a formula $\psi$ of a
-transition rule $q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi$ of $\A$.
-
-The acceptance formula $\accform{\A}$ is false in every interpretation
-of the predicates that assigns a non-empty set to a non-final
-predicate occurring on the last level in the execution forest. The
-relation between the words accepted by $\A$ and the acceptance formula
-above, is formally captured by the following lemma:
-
-\begin{lemma}\label{lemma:acceptance}
-  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for
-  every word $w \in \Sigma[X]^*$, the following are
-  equivalent: \begin{compactenum}
-  \item\label{it1:lemma:acceptance} there exists an interpretation
-    $\I$ such that $\I,\data{w} \models \accform{\event{w}}$, 
-    %
-  \item\label{it2:lemma:acceptance} $w \in \lang{\A}$. 
-  \end{compactenum}
-\end{lemma}
-\proof{ ``$(\ref{it1:lemma:acceptance}) \Rightarrow
-  (\ref{it2:lemma:acceptance})$'' Let $\I$ be an interpretation such
-  that $\I,\data{w} \models \accform{\event{w}}$. By Lemma
-  \ref{lemma:path-formula}, $\A$ has an execution $\T$ over $w$ such
-  that $\I = \I_{\T}$. To prove that $\T$ is accepting, we show
-  that \begin{inparaenum}[(i)]
-    \item\label{it1:acceptance} all paths in $\T$ have length $n$ and
-      that
-    \item\label{it2:acceptance} the frontier of $\T$ is labeled with
-      final configurations only. \end{inparaenum} First, assume that
-  (\ref{it1:acceptance}) there exists a path in $\T$ of length $0 \leq
-  m < n$. Then there exists a node on the $m$-th level, labeled with
-  some configuration $q(d_1,\ldots,q_{\#(q)})$, that has no
-  children. By the definition of the execution semantics of $\A$, we
-  have $\cube{\minsem{\psi}_{\eta}} = \emptyset$, where
-  $q(\vec{y}) \arrow{a_{m+1}(X)}{} \psi$ is the transition
-  rule of $\A$ that applies for $q$ and $a_{m+1}$ and $\eta=
-  \data{w}[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow
-    d_{\#(q)}]$. Hence $\sem{\psi}_{\eta} = \emptyset$, and because
-  $\I,\data{w} \models \accform{\alpha}$, we obtain that $\I,\eta \models
-  q(\vec{y}) \rightarrow \stamp{\psi}{m+1}$, thus
-  $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$. However, this
-  contradicts the fact that $\I=\I_{\T}$ and that
-  $q(d_1,\ldots,d_{\#(q)})$ labels a node of $\T$. Second, assume that
-  (\ref{it2:acceptance}), there exists a frontier node of $\T$ labeled
-  with a configuration $q(d_1,\ldots,d_{\#(q)})$ such that $q \in Q
-  \setminus F$. Since $\I,\data{w} \models \forall y_1 \ldots \forall
-  y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot$, by a
-  similar reasoning as in the above case, we obtain that
-  $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$, contradiction.
-
-  ``$(\ref{it2:lemma:acceptance}) \Rightarrow
-  (\ref{it1:lemma:acceptance})$'' Let $\T$ be an accepting execution
-  of $\A$ over $w$. We prove that $\I_\T,\data{w} \models
-  \accform{\event{w}}$. By Lemma \ref{lemma:path-formula}, we obtain
-  $\I_\T,\data{w} \models \pathform{\event{w}}$. Since every path in
-  $\T$ is of length $n$ and all nodes on the $n$-th level of $\T$ are
-  labeled by final configurations, we obtain that $\I_\T,\data{w}
-  \models \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
-  y_{\#(q)} ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot$,
-  trivially.  \qed}
-
-As an immediate consequence, one can decide whether $\A$ accepts some
-word $w$ with a given input sequence $\event{w}=\alpha$, by checking
-whether $\accform{\alpha}$ is satisfiable. However, unlike
-non-alternating infinite-state models of computation, such as counter
-automata (nondeterministic programs with integer variables), the
-satisfiability query for an acceptance (path) formula falls outside of
-known decidable theories, supported by standard SMT solvers. There are
-basically two reasons for this, namely \begin{inparaenum}[(i)]
-\item the presence of predicate symbols, and
-\item the non-trivial alternation of quantifiers.
-\end{inparaenum}
-To understand this point, consider for example, the decidable theory
-of Presburger arithmetic \cite{Presburger29}. Adding even only one
-monadic predicate symbol to it yields undecidability in the presence
-of non-trivial quantifier alternation \cite{Halpern91}. However the
-quantifier-free fragment of Presburger arithmetic extended with
-predicate symbols can be shown to be decidable, using a Nelson-Oppen
-style congruence closure argument \cite{NelsonOppen80}.
-
-To tackle this problem, we start from the observation that acceptance
-formulae have a particular form, which allows the elimination of path
-quantifiers and of predicates, by a couple of
-satisfiability-preserving transformations. The result of applying
-these transformations is a formula with no predicate symbols, whose
-only quantifiers are those introduced by the transition rules of the
-automaton, referred to as transition quantifiers. We shall further
-assume (\S\ref{sec:emptiness}) that the first-order theory of the data
-sort $\Data$ has quantifier elimination, which allows to effectively
-decide the satisfiability of such formulae.
-
-For the time being, let us formally define the elimination of
-transition quantifiers and predicates, respectively. Consider a given
-sequence of input events $\alpha = a_1 \ldots a_n$ and denote by
-$\alpha_i$ the prefix $a_1 \ldots a_i$ of $\alpha$, for $i \in [n]$,
-where $\alpha_0=\epsilon$.
-
-\begin{definition}\label{def:quantform}
-  Let $\quantformn{\alpha_0}, \ldots, \quantformn{\alpha_n}$ be the
-  sequence of formulae defined by $\quantformn{\alpha_0} \isdef
-  \stamp{\iota}{0}$ and, for all $i \in [1,n]$:
-  \[\quantformn{\alpha_i} \isdef \quantformn{\alpha_{i-1}} \wedge
-  \bigwedge_{\begin{array}{l}
-      \scriptstyle{\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_{i-1}}} \\
-      \scriptstyle{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta} 
-  \end{array}} \stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow 
-    \stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]
-      \]  We write
-    $\quantform{\alpha}$ for the prenex normal form of the formula: 
-      \[\quantformn{\alpha_n} \wedge \bigwedge_{\begin{array}{l}
-          \scriptstyle{\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_n}} \\
-          \scriptstyle{q \in Q \setminus F}
-      \end{array}} \stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot \enspace.\]
-\end{definition}
-Observe that $\quantform{\alpha}$ contains no path quantifiers, as
-required. On the other hand, the scope of the transition quantifiers
-in $\quantform{\alpha}$ exceeds the right-hand side formulae from the
-transition rules, as shown by the following example.
-
-\begin{example}\label{ex:quant-elim}
-  Consider the automaton $\A = \tuple{\set{a_1,a_2}, \set{x},
-    \set{q,q_f}, \iota, \set{q_f}, \Delta}$, where:
-  \[\begin{array}{rcl}
-  \iota & = & \exists z ~.~ z \geq 0 \wedge q(z) \\
-  \Delta & = & \set{
-    q(y) \arrow{a_1(x)}{} x\geq 0 \wedge \forall z ~.~ z \leq y \rightarrow q(x+z),~
-    q(y) \arrow{a_2(x)}{} y<0 \wedge q_f(x+y)
-  }
-  \end{array}\]
-  For the input event sequence $\alpha = a_1a_2$, the acceptance
-  formula is:
-  \[\begin{array}{rcl}
-  \accform{\alpha} & = & \exists z ~.~ z \geq 0 \wedge \stamp{q}{0}(z) ~\wedge \\
-  && \forall y ~.~ \stamp{q}{0}(y) \rightarrow [\stamp{x}{1}\geq0 \wedge \forall z ~.~ z\geq y 
-    \rightarrow \stamp{q}{1}(\stamp{x}{1}+z)] ~\wedge \\
-  && \forall y ~.~ \stamp{q}{1}(y) \rightarrow [y<0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+y)]
-  \end{array}\]
-  The result of eliminating the path quantifiers, in prenex normal form, is shown below:
-  \[\begin{array}{rcl}
-  \quantform{\alpha} & = & \exists z_1\forall z_2 ~.~ z_1 \geq 0 \wedge \stamp{q}{0}(z_1) ~\wedge \\
-  && [\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge (z_2 \geq z_1 \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_2))] ~\wedge \\
-  && [\stamp{q}{1}(\stamp{x}{1}+z_2) \rightarrow \stamp{x}{1}+z_2 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_2)] 
-  \enspace\enspace\hfill\blacksquare
-  \end{array}\]
-\end{example}
-
-The next lemma establishes a formal relation between the
-satisfiability of an acceptance formula $\accform{\alpha}$ and that of
-the formula $\quantform{\alpha}$, obtained by eliminating the path
-quantifiers from $\accform{\alpha}$.
-
-\begin{lemma}\label{lemma:quant}
-  For any input event sequence $\alpha=a_1\ldots a_n$ and each
-  valuation $\nu : \stamp{X}{\leq n} \rightarrow \Data$, the following
-  hold: \begin{compactenum}
-  \item\label{it1:quant} for all interpretations $\I$, if $\I,\nu
-    \models \accform{\alpha}$ then $\I,\nu \models
-    \quantform{\alpha}$. 
-    %
-  \item\label{it2:quant} if there exists an interpretation $\I$ such
-    that $\I,\nu \models \quantform{\alpha}$ then there exists an
-    interpretation $\J \subseteq \I$ such that $\J,\nu \models
-    \accform{\alpha}$. 
-  \end{compactenum}
-\end{lemma}
-\proof{(\ref{it1:quant}) Trivial, since every subformula
-  $q(t_1,\ldots,t_{\#(q)}) \rightarrow
-  \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ of $\quantform{\alpha}$ is
-  entailed by a subformula $\forall y_1 \ldots \forall y_{\#(q)} ~.~
-  q(y_1, \ldots, y_{\#(q)}) \rightarrow \psi$ of $\accform{\alpha}$.
-
-  \noindent(\ref{it2:quant}) By repeated applications of the following fact:
-
-  \begin{fact}
-    Given formulae $\phi$ and $\psi$, such that no predicate atom with
-    predicate symbol $q$ occurs in $\psi(y_1,\ldots,y_{\#(q)})$, for
-    each valuation $\nu$, if there exists an interpretation $\I$ such
-    that $\I,\nu \models \phi \wedge
-    \bigwedge_{q(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \phi}
-    q(t_1,\ldots,t_{\#(q)}) \rightarrow
-    \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ then there exists a
-    valuation $\J$ such that $\J(q) \subseteq \I(q)$ and $\J(q') =
-    \I(q')$ for all $q' \in Q \setminus \set{q}$ and $\J,\nu \models
-    \phi \wedge \forall y_1 \ldots \forall y_{\#(q)} ~.~
-    q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$.
-  \end{fact}
-  \proof{Assume w.l.o.g. that $\phi$ is quantifier free. The proof can
-    be easily generalized to the case $\phi$ has quantifiers. Let
-    $\J(q) = \set{\langle t_1^\nu, \ldots, t_{\#(q)}^\nu \rangle \in
-      \I(q) \mid q(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \phi}$ and
-    $\J(q') = \I(q')$ for all $q' \in Q \setminus \set{q}$. Since
-    $\I,\nu \models \phi$, we obtain that also $\J,\nu \models \phi$
-    because the tuples of values in $\I(q) \setminus \J(q)$ are not
-    interpretations of terms that occur within subformulae
-    $q(t_1,\ldots,t_{\#(q)})$ of $\phi$. Moreover,
-    $\bigwedge_{q(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \phi}
-    q(t_1,\ldots,t_{\#(q)}) \rightarrow
-    \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ and $\forall y_1 \ldots
-    \forall y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$
-    are equivalent under $\J$, thus $\J,\nu \models \forall y_{\#(q)}
-    ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$, as required. \qed}
-
-  \noindent This concludes the proof. \qed}
-
-We proceed with the elimination of predicate atoms from
-$\quantform{\alpha}$, defined below.
-
-\begin{definition}\label{def:substform}
-  Let $\substformn{\alpha_0}, \ldots, \substformn{\alpha_n}$ be the
-  sequence of formulae defined by $\substformn{\alpha_0} \isdef
-  \stamp{\iota}{0}$ and, for all $i \in [1,n]$,
-  $\substformn{\alpha_i}$ is obtained by replacing each occurrence of
-  a predicate atom $\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)})$ in
-  $\substformn{\alpha_{i-1}}$ with the formula
-  $\stamp{\psi}{i}[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$, where
-  $q(\vec{y}) \arrow{a_{i}(X)}{} \psi \in \Delta$. We write
-  $\substform{\alpha}$ for the formula obtained by replacing, in
-  $\substformn{\alpha}$, each occurrence of a predicate
-  $\stamp{q}{n}$, such that $q \in Q \setminus F$ (resp. $q \in F$),
-  by $\bot$ (resp. $\top$).
-\end{definition}
-
-\begin{example}[Contd. from Example \ref{ex:quant-elim}]\label{ex:pred-elim}
-  The result of the elimination of predicate atoms from the acceptance
-  formula in Example \ref{ex:quant-elim} is shown below:
-  \[\substform{\alpha} = 
-  \exists z_1 \forall z_2 ~.~ z_1 \geq 0 \wedge 
-          [\stamp{x}{1} \geq 0 \wedge (z_2 \geq z_1 \rightarrow
-            \stamp{x}{1}+z_2 < 0)]\] Since this formula is
-          unsatisfiable, by Lemma \ref{lemma:quant-pred-acceptance}
-          below, no word $w$ with input event sequence $\event{w} =
-          a_1a_2$ is accepted by the automaton $\A$ from Example
-          \ref{ex:quant-elim}. \hfill$\blacksquare$
-\end{example}
-
-At this point, we prove the formal relation between the satisfiability
-of the formulae $\quantform{\alpha}$ and $\substform{\alpha}$. Since
-there are no occurrences of predicates in $\substform{\alpha}$, for
-each valuation $\nu : \stamp{X}{\leq n} \rightarrow \Data$, there
-exists an interpretation $\I$ such that $\I,\nu \models
-\substform{\alpha}$ if and only if $\J,\nu \models
-\substform{\alpha}$, for every interpretation $\J$. In this case we
-omit $\I$ and simply write $\nu \models \substform{\alpha}$.
-
-\begin{lemma}\label{lemma:subst}
-  For any input event sequence $\alpha=a_1\ldots a_n$ and each
-  valuation $\nu : \stamp{X}{\leq n} \rightarrow \Data$, there exists
-  a valuation $\I$ such that $\I,\nu \models \quantform{\alpha}$ if
-  and only if $\nu \models \substform{\alpha}$. 
-\end{lemma}
-\proof{ By induction on $n\geq0$. The base case $n=0$ is trivial,
-  since $\quantform{A} = \substform{A} = \stamp{\iota}{0}$. For the
-  induction step, we rely on the following fact:
-
-  \begin{fact}\label{fact:subst}
-    Given formulae $\phi$ and $\psi$, such that $\phi$ is positive
-    $q(t_1,\ldots,t_{\#(q)})$ is the only one occurrence of the
-    predicate symbol $q$ in $\phi$ and no predicate atom with
-    predicate symbol $q$ occurs in $\psi(y_1,\ldots,y_{\#(q)})$, for
-    each interpretation $\I$ and each valuation $\nu$, we have:
-    \[\begin{array}{l} 
-    \I,\nu \models \phi \wedge q(t_1,\ldots,t_{\#(q)}) \rightarrow
-    \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}] \iff \\ \nu \models
-    \phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]\enspace.
-    \end{array}\]
-  \end{fact}
-  \proof{ We assume w.l.o.g. that $\phi$ is quantifier-free. The proof
-    can be easily generalized to the case $\phi$ has quantifiers.
-
-    \noindent''$\Rightarrow$'' 
-    We distinguish two cases: \begin{compactitem}
-    \item if $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \in \I(q)$ then
-      $\I,\nu \models \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$. Since
-      $\phi$ is positive, replacing $q(t_1,\ldots,t_{\#(q)})$ with
-      $\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ does not change the
-      truth value of $\phi$ under $\nu$, thus $\nu \models
-      \phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$.
-      %
-    \item else, $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \not\in
-      \I(q)$, thus $\nu \models
-      \phi[\bot/q(t_1,\ldots,t_{\#(q)})]$. Since $\phi$ is positive
-      and $\bot$ entails $\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$,
-      we obtain $\nu \models
-      \phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$
-      by monotonicity. 
-    \end{compactitem}
-
-    \noindent``$\Leftarrow$'' Let $\I$ be any interpretation such that
-    $\I(q) = \set{\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \mid \nu
-      \models \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]}$. We
-    distinguish two cases: \begin{compactitem}
-      \item if $\I(q) \not \emptyset$ then $\I,\nu \models
-        q(t_1,\ldots,t_{\#(q)})$ and $\nu \models \psi[t_1/y_1,
-          \ldots, t_{\#(q)}/y_{\#(q)}]$. Thus replacing $\psi[t_1/y_1,
-          \ldots, t_{\#(q)}/y_{\#(q)}]$ by $q(t_1,\ldots,t_{\#(q)})$
-        does not change the truth value of $\phi$ under $\I$ and
-        $\nu$, and we obtain $\I,\nu \models \phi$. Moreover, $\I,\nu
-        \models \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ implies
-        $\I,\nu \models q(t_1,\ldots,t_{\#(q)}) \rightarrow
-        \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$.
-        %
-      \item else $\I(q) = \emptyset$, hence $\nu \not\models
-        \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$, thus $\nu \models
-        \phi[\bot/q(t_1,\ldots,t_{\#(q)})]$. Because $\phi$ is
-        positive, we obtain $\I,\nu \models \phi$ by monotonicity. But
-        $\I,\nu \models q(t_1,\ldots,t_{\#(q)}) \rightarrow
-        \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ trivially, because
-        $\I,\nu \not\models q(t_1,\ldots,t_{\#(q)})$.\qed
-    \end{compactitem}}
-This concludes the proof. \qed}
-
-%% \begin{proposition}\label{prop:subst}
-%%   Let $\phi$ be a positive formula and let $q(t_1,\ldots,t_{\#(q)})$
-%%   be the only predicate atom occurring within $\phi$. Then, for every
-%%   valuation $\nu$ of $\fv{}{\phi}$ and every formula $\psi$ with no
-%%   predicate occurrences, the following are
-%%   equivalent: \begin{compactenum}
-%%   \item\label{it1:prop:subst} $\nu \models \phi[\psi[t_1/y_1, \ldots,
-%%       t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$, 
-%%     %
-%%   \item\label{it2:prop:subst} $\I,\nu \models \phi \wedge \forall y_1
-%%     \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow
-%%     \psi$, for some interpretation $\I$.
-%%   \end{compactenum}
-%%   where $y_1,\ldots,y_{\#(q)} \not\in \fv{}{\phi}$. 
-%% \end{proposition}
-%% \proof{ We give the proof in the case $\phi$ is quantifier-free. The
-%%   generalization to a non-empty quantifier prefix is straightforward.
-%%   ``$(\ref{it1:prop:subst}) \Rightarrow (\ref{it2:prop:subst})$'' We
-%%   distinguish two cases:
-%%   \begin{compactitem}
-%%   \item if $\nu \models \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$
-%%     then let $\I(q) = \set{\tuple{t_1^\nu,\ldots,t_{\#(q)}^\nu}}$ and
-%%     $\emptyset$ everywhere else. We have $\I,\nu \models
-%%     q(t_1,\ldots,t_{\#(q)})$ and, since $\phi$ is positive, replacing
-%%     $\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ with
-%%     $q(t_1,\ldots,t_{\#(q)})$ in $\phi$ does not change its
-%%     satisfiability under $\I,\nu$, hence $\I,\nu \models
-%%     \phi$. Moreover, $\nu \models \psi[t_1/y_1, \ldots,
-%%       t_{\#(q)}/y_{\#(q)}]$ thus $\I,\nu \models \forall y_1 \ldots
-%%     \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \psi$.
-%%     %
-%%   \item else $\nu \not\models \psi[t_1/y_1, \ldots,
-%%     t_{\#(q)}/y_{\#(q)}]$ and let $\I$ be the interpretation assigning
-%%     each predicate symbol with $\emptyset$. Then replacing
-%%     $\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ with
-%%     $q(t_1,\ldots,t_{\#(q)})$ in $\phi$ does not change its
-%%     satisfiability under $\I,\nu$. Moreover, since $\I$ maps $q$ to
-%%     $\emptyset$, we have $\I,\nu \models \forall y_1 \ldots \forall
-%%     y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \psi$, trivially.
-%%   \end{compactitem}
-
-%%   ``$(\ref{it2:prop:subst}) \Rightarrow (\ref{it1:prop:subst})$'' We
-%%   distinguish two cases: \begin{compactitem}
-%%   \item if $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \in \I(q)$, since
-%%     $\I,\nu \models \forall y_1 \ldots \forall y_{\#(q)} ~.~
-%%     q(\vec{y}) \rightarrow \psi$, we have $\nu \models
-%%     \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$. Since $\phi$ is
-%%     positive, replacing $q(t_1,\ldots,t_{\#(q)})$ by $\psi$ in $\phi$
-%%     does not change its satisfiability under $\nu$, hence $\nu \models
-%%     \phi[\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]/q(t_1, \ldots,
-%%       t_{\#(q)})]$.
-%%     %
-%%   \item else $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \not\in \I(q)$,
-%%     that is $\I,\nu \not\models q(t_1,\ldots,t_{\#(q)})$. Then $\nu
-%%     \models \phi[\bot/q(t_1,\ldots,t_{\#(q)})]$ and we obtain $\nu
-%%     \models \phi[\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]/q(t_1,
-%%       \ldots, t_{\#(q)})]$, because $\phi$ is positive, thus
-%%     $\phi[\bot/q(t_1,\ldots,t_{\#(q)})] \models \phi[\psi[t_1/y_1,
-%%         \ldots, t_{\#(q)}/y_{\#(q)}]/q(t_1, \ldots, t_{\#(q)})]$. \qed
-%%   \end{compactitem}}
-
-Finally, we define the acceptance of a word with a given input event
-sequence by means of a formula in which no predicate atom occurs. As
-previously discussed, several decidable theories, such as Presburger
-arithmetic, become undecidable if predicate atoms are added to
-them. Therefore, the result below makes a step towards deciding
-whether the automaton accepts a word with a given input sequence, by
-reducing this problem to the satisfiability of a quantified formula
-without predicates.
-
-\begin{lemma}\label{lemma:quant-pred-acceptance}
-  Given an automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$, for
-  every word $w \in \Sigma[X]^*$, we have $\data{w} \models
-  \substform{\event{w}}$ if and only if $w \in \lang{\A}$.
-\end{lemma}
-\proof{ By Lemma \ref{lemma:acceptance}, $w \in \lang{\A}$ if and only
-  if $\I,\data{w} \models \accform{\event{w}}$, for some
-  interpretation $\I$. By Lemma \ref{lemma:quant}, there exists an
-  interpretation $\I$ such that $\I,\data{w} \models
-  \accform{\event{w}}$ if and only if there exists an interpretation
-  $\J$ such that $\J,\nu \models \quantform{\event{w}}$. By Lemma
-  \ref{lemma:subst}, there exists an interpretation $\J$ such that
-  $\J,\nu \models \quantform{\event{w}}$ if and only if $\nu \models
-  \substform{\event{w}}$. \qed}
-
-\subsection{Closure Properties}
-\label{sec:closure}
-
-Given a positive formula $\phi$, we define the \emph{dual} formula
-$\dual{\phi}$ recursively as follows:
-\[\begin{array}{rclcrclcrcl}
-\dual{(\phi_1 \vee \phi_2)} & = & \dual{\phi_1} \wedge \dual{\phi_2} && 
-\dual{(\phi_1 \wedge \phi_2)} & = & \dual{\phi_1} \vee \dual{\phi_2} &&
-\dual{(t \teq s)} & = & \neg(t \teq s) \\
-\dual{(\exists x ~.~ \phi_1)} & = & \forall x ~.~ \dual{\phi_1} && 
-\dual{(\forall x ~.~ \phi_1)} & = & \exists x ~.~ \dual{\phi_1} && 
-\dual{(\neg(t \teq s))} & = & t \teq s \\
-\dual{(q(x_1,\ldots,x_{\#(q)}))} & = & q(x_1,\ldots,x_{\#(q)})
-\end{array}\]
-Observe that, because predicate atoms do not occur negated in $\phi$,
-there is no need to define dualization for formulae of the form $\neg
-q(x_1,\ldots,x_{\#(q)})$. The following theorem shows closure of
-automata under all boolean operations:
-
-%% The dualization operation has the following property:
-%% \begin{lemma}\label{lemma:dual}
-%%   Given a formula $\phi$ and an automaton
-%%   $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$, we have the equivalence
-%%   $\Accept{F}{\phi} \equiv \neg(\dual{\Accept{Q\setminus F}{\phi}})$. 
-%% \end{lemma}
-%% \proof{ By straightforward induction on the structure of $\phi$. \qed}
-
-\begin{theorem}\label{thm:closure}
-  Given automata $\A_i = \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$,
-  for $i=1,2$, such that $Q_1 \cap Q_2 = \emptyset$, the following hold: 
-  \begin{compactenum}
-    \item\label{it1:thm:closure} $\lang{\A_\cap} = \lang{\A_1} \cap \lang{\A_2}$, where
-      $\A_\cap = \tuple{\Sigma,X,Q_1 \cup Q_2, \iota_1 \wedge \iota_2,
-      F_1 \cup F_2, \Delta_1 \cup \Delta_2}$, 
-      %
-    \item\label{it2:thm:closure} $\lang{\overline{\A_i}} = \Sigma[X]^*
-      \setminus \lang{\A_i}$, where $\overline{\A_i} =
-      \tuple{\Sigma,X,Q_i,\dual{\iota},Q_i\setminus
-        F_i,\dual{\Delta}_i}$ and, for $i=1,2$: \[\dual{\Delta}_i =
-      \set{q(\vec{y}) \arrow{a(X)}{} \dual{\psi} \mid
-        q(\vec{y}) \arrow{a(X)}{} \psi \in
-        \Delta_i}\enspace.\]
-  \end{compactenum}
-  Moreover, $\len{\A_\cap} = \bigO{\len{\A_1}+\len{\A_2}}$ and
-  $\len{\overline{\A_i}} = \bigO{\len{\A_i}}$, for all $i=1,2$.
-\end{theorem}
-\proof{ (\ref{it1:thm:closure}) ``$\subseteq$'' Let $w \in
-  \lang{\A_\cap}$ be a word and $\T$ be an execution of $\A_\cap$ over
-  $w$. Since $Q_1 \cap Q_2 = \emptyset$, it is possible to partition
-  $\T$ into $\T_1$ and $\T_2$ such that the roots of $\T_i$ form a
-  cube from $\cube{\minsem{\iota_i}}$, for all $i=1,2$. Because
-  $\Delta_1 \cap \Delta_2 = \emptyset$, by induction on
-  $\len{w}\geq0$, one shows that $\T_i$ is an execution of $\A_i$ over
-  $w$, for all $i=1,2$. Finally, because $\T$ is accepting, we obtain
-  that $\T_1$ and $\T_2$ are accepting, respectively, hence $w \in
-  \lang{\A_1} \cap \lang{\A_2}$. ``$\supseteq$'' Let $w \in
-  \lang{\A_1} \cap \lang{\A_2}$ and let $\T_i$ an accepting execution
-  of $\A_i$ over $w$, for all $i=1,2$. We show that $\T_1 \cup \T_2$
-  is an execution of $\A_\cap$ over $w$, by induction on
-  $\len{w}\geq0$. For the base case $\len{w}=0$, we have $\T_i \in
-  \cube{\minsem{\iota_i}}$ for all $i=1,2$ and since $Q_1 \cap Q_2 =
-  \emptyset$, we have $\T_1 \cup \T_2 \in \cube{\minsem{\iota_1 \wedge
-      \iota_2}}$. The induction step follows as a consequence of the
-  fact that $\Delta_1 \cup \Delta_2$ is the set of transition rules of
-  $\A_\cap$. Finally, since both $\T_1$ and $\T_2$ are accepting,
-  $\T_1 \cup \T_2$ is accepting as well. Moreover, we have: 
-  \[
-  \len{\A_\cap} = \len{\iota_1 \wedge \iota_2} + \sum_{q(\vec{y})
-    \arrow{a(X)}{} \psi \in \Delta_1 \cup \Delta_2} \len{\psi} =
-  1 + \len{\iota_1} + \len{\iota_2} + \sum_{q(\vec{y})
-    \arrow{a(X)}{} \psi \in \Delta_1} \len{\psi} + \sum_{q(\vec{y})
-    \arrow{a(X)}{} \psi \in \Delta_2} \len{\psi} \enspace. 
-  \]
-
-  \noindent(\ref{it2:thm:closure}) Let $w \in \Sigma[X]^*$ be a
-  word. We denote by $\accformA{\A_1}{\event{w}}$ and
-  $\substformA{\A_1}{\event{w}}$
-  [resp. $\accformA{\overline{\A}_1}{\event{w}}$ and
-    $\substformA{\overline{\A}_1}{\event{w}}$] the formulae
-  $\accform{\event{w}}$ and $\substform{\event{w}}$ for $\A_1$ and
-  $\overline{\A}_1$, respectively. It is enough to show that
-  $\substformA{\overline{\A}_1}{\event{w}} =
-  \neg\substformA{\A_1}{\event{w}}$ and apply Lemma
-  \ref{lemma:quant-pred-acceptance} to prove that $w \in \lang{\A_1}
-  \iff w \not\in \lang{\overline{\A}_1}$. Since the choice of $w$ was
-  arbitrary, this proves $\lang{\overline{\A}_1} = \Sigma[X]^*
-  \setminus \lang{\A_1}$. By induction on the number of predicate
-  atoms in $\accformA{\A_1}{\event{w}}$ that are replaced during the
-  generation of $\substformA{\A_1}{\event{w}}$. The proof relies on
-  the following fact:
-
-  \begin{fact}\label{fact:dual-neg}
-    Let $\phi$ be a positive formula and let $q(t_1,\ldots,t_{\#(q)})$
-    be the only occurrence of a predicate symbol within $\phi$. Then,
-    every formula $\phi$ with no predicate occurrences:
-    $\neg\phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]
-    \equiv
-    \dual{\phi}[\neg\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$.
-  \end{fact}
-  \proof{By induction on the structure of $\phi$. \qed}}
-
-\section{The Emptiness Problem}
-\label{sec:emptiness}
-
-The problem of checking emptiness of a given automaton is undecidable,
-even for automata with predicates of arity two, whose transition rules
-use only equalities and disequalities, having no transition
-quantifiers \cite{Farzan15}. Since even such simple classes of
-alternating automata have no general decision procedure for emptiness,
-we use an abstraction-refinement semi-algorithm based on \emph{lazy
-  annotation} \cite{McMillan06,McMillan14}.
-
-In a nutshell, a lazy annotation procedure systematically explores the
-set of execution paths (in our case, sequences of input events) in
-search of an accepting execution. Each path has a corresponding path
-formula that defines all words accepted along that path. If the path
-formula is satisfiable, the automaton accepts a word. Otherwise, the
-path is said to be \emph{spurious}. When a spurious path is
-encountered, the search backtracks and the path is annotated with a
-set of learned facts, that marks this path as infeasible. The
-semi-algorithm uses moreover a coverage relation between paths,
-ensuring that the continuations of already covered paths are never
-explored. Sometimes this coverage relation provides a sound
-termination argument, when the automaton is empty.
-
-We check emptiness of first-order alternating automata using a version
-of the \impact~ lazy annotation semi-algorithm \cite{McMillan06}. An
-analogous procedure is given in \cite{IosifXu18}, for a simpler model
-of alternating automata, that uses only predicates or arity zero
-(booleans) and no transition quantifiers. For simplicity, we do not
-present the details of this algorithm (see Appendix
-\ref{app:pseudocode}) and shall content ourselves of several
-high-level definitions.
-
-Given a finite input event alphabet $\Sigma$, for two sequences
-$\alpha, \beta \in \Sigma^*$, we say that $\alpha$ is a prefix of
-$\beta$, written $\alpha \prefix \beta$, if $\alpha=\beta\gamma$ for
-some sequence $\gamma\in\Sigma^*$. A set $S$ of sequences
-is: \begin{compactitem}
-\item \emph{prefix-closed} if for each $\alpha \in S$, if $\beta \prefix
-\alpha$ then $\beta \in S$, and
-\item \emph{complete} if for each $\alpha \in S$, there exists $a \in
-  \Sigma$ such that $\alpha a \in S$ if and only if $\alpha b \in S$
-  for all $b \in \Sigma$. 
-\end{compactitem}
-Observe that a prefix-closed set is the backbone of a tree whose edges
-are labeled with input events. If the set is, moreover, complete, then
-every node of the tree has either zero successors, in which case it is
-called a \emph{leaf}, or it has a successor edge labeled with $a$ for
-each input event $a \in \Sigma$.
-
-\begin{definition}\label{def:unfolding}
-  An \emph{unfolding} of an automaton $\A =
-  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ is a finite partial mapping $U :
-  \Sigma^* \rightharpoonup_{\mathit{fin}} \posforms(Q,\emptyset)$,
-  such that: \begin{compactenum}
-  \item\label{it1:unfolding} $\dom(U)$ is a finite prefix-closed
-    complete set,
-    %
-  \item\label{it2:unfolding} $U(\epsilon) = \iota$, and
-    %
-  \item\label{it3:unfolding} for each sequence $\alpha a \in \dom(U)$,
-    such that $\alpha \in \Sigma^*$ and $a \in \Sigma$:
-    \[\stamp{U(\alpha)}{0} \wedge
-    \bigwedge_{q(\vec{y}) \arrow{a(X)}{} \psi} \forall y_1 \ldots
-    \forall y_{\#q} ~.~ \stamp{q}{0}(\vec{y}) \rightarrow
-    \stamp{\psi}{1} \models \stamp{U(\alpha a)}{1}\]
-  \end{compactenum}
-  Moreover, $U$ is \emph{safe} if for each $\alpha \in \dom(U)$, the
-  formula $U(\alpha) \wedge \bigwedge_{q \in Q \setminus F} \forall
-  y_1 \ldots \forall_{y_{\#(q)}} ~.~ q(\vec{y}) \rightarrow \bot$ is
-  unsatisfiable. 
-\end{definition}
-
-Lazy annotation semi-algorithms \cite{McMillan06,McMillan14} build
-unfoldings of automata trying to discover counterexamples for
-emptiness. If the automaton $\A$ in question is non-empty, a
-systematic enumeration of the input event sequences\footnote{For
-  instance, using breadth-first search.} from $\Sigma^*$ will suffice
-to discover a word $w \in \lang{\A}$, provided that the first-order
-theory of the data domain $\Data$ is decidable (Lemma
-\ref{lemma:acceptance}). However, if $\lang{\A} = \emptyset$, the
-enumeration of input event sequences may, in principle, run
-forever. The typical way of fighting this divergence problem is to
-define a \emph{coverage} relation between the nodes of the unfolding
-tree.
-
-\begin{definition}\label{def:coverage}
-  Given an unfolding $U$ of an automaton $\A =
-  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ a node $\alpha \in \dom(U)$ is
-  \emph{covered} by another node $\beta \in \dom(U)$, denoted $\alpha
-  \cover \beta$, if and only if there exists a node $\alpha' \prefix
-  \alpha$ such that $U(\alpha') \models U(\beta)$. Moreover, $U$ is
-  \emph{closed} if and only if every leaf from $\dom(U)$ is covered by
-  an uncovered node.
-\end{definition}
-
-%% , for all words $w \in \Sigma[X]^*$ and all cubes $c \in
-%% \cube{\sem{U(\alpha')}}$, if $\A$ accepts $w$ started in $c$ then
-%% there exists a cube $d \in \cube{\sem{U(\beta)}}$ such that $\A$
-%% accepts $w$ started in $d$.
-
-A lazy annotation semi-algorithm will stop and report emptiness
-provided that it succeeds in building a closed and safe unfolding of
-the automaton. Notice that, by Definition \ref{def:coverage}, for any
-three nodes of an unfolding $U$, say $\alpha,\beta,\gamma \in
-\dom(U)$, if $\alpha \prec \beta$ and $\alpha \cover \gamma$, then
-$\beta \cover \gamma$ as well. As we show next (Theorem
-\ref{thm:soundness}), there is no need to expand covered nodes,
-because, intuitively, there exists a word $w \in \lang{\A}$ such that
-$\alpha \preceq \event{w}$ and $\alpha \cover \gamma$ only if there
-exists another word $u \in \lang{\A}$ such that $\gamma \preceq
-\event{u}$. Hence, exploring only those input event sequences that are
-continuations of $\gamma$ (and ignoring those of $\alpha$) suffices in
-order to find a counterexample for emptiness, if one exists.
-
-An unfolding node $\alpha \in \dom(U)$ is said to be \emph{spurious}
-if and only if $\accform{\alpha}$ is unsatisfiable. In this case, we
-change (refine) the labels of (some of the) prefixes of $\alpha$ (and
-that of $\alpha$), such that $U(\alpha)$ becomes $\bot$, thus
-indicating that there is no real execution of the automaton along that
-input event sequence. As a result of the change of labels, if a node
-$\gamma \prefix \alpha$ used to cover another node from $\dom(U)$, it
-might not cover it with the new label. Therefore, the coverage
-relation has to be recomputed after each refinement of the
-labeling. The semi-algorithm stops when (and if) a safe complete
-unfolding has been found. For a detailed presentation of the emptiness
-procedure, we refer the reader to \cite{IosifXu18} (or Appendix
-\ref{app:pseudocode}).
-
-\begin{theorem}\label{thm:soundness}
-  If an automaton $\A$ has a nonempty safe closed unfolding then
-  $\lang{A} = \emptyset$.
-\end{theorem}
-\proof{ Let $U$ be a safe and complete unfolding of $\A$, such that
-  $\dom(U) \neq \emptyset$. Suppose, by contradiction, that there
-  exists a word $w \in \lang{A}$ and let $\alpha \isdef
-  \event{w}$. Since $w \in \lang{A}$, by Lemma \ref{lemma:acceptance},
-  there exists an interpretation $\I$ such that $\I,\data{w} \models
-  \accform{\alpha}$. Assume first that $\alpha \in \dom(U)$.  In this
-  case, one can show, by induction on the length $n\geq0$ of $w$, that
-  $\pathform{\alpha} \models \stamp{U(\alpha)}{n}$, thus $\I,\data{w}
-  \models \stamp{U(\alpha)}{n}$. Since $\I,\data{w} \models
-  \accform{\alpha}$, we have $\I,\data{w} \models \bigwedge_{q \in Q
-    \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{n}(\vec{y}) \rightarrow \bot$, hence $\stamp{U(\alpha)}{n}
-  \wedge \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
-  y_{\#(q)} ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot$. By renaming
-  $\stamp{q}{n}$ with $q$ in the previous formula, we obtain
-  $U(\alpha) \wedge \forall y_1 \ldots \forall y_{\#(q)} ~.~
-  q(\vec{y}) \rightarrow \bot$ is satisfiable, thus $U$ is not safe,
-  contradiction.
-
-  We proceed thus under the assumption that $\alpha \not\in
-  \dom(U)$. Since $\dom(U)$ is a nonempty prefix-closed set, there
-  exists a strict prefix $\alpha'$ of $\alpha$ that is a leaf of
-  $\dom(U)$. Since $U$ is closed, the leaf $\alpha'$ must be covered
-  and let $\alpha_1 \prefix \alpha' \prefix \alpha$ be a node such
-  that $U(\alpha_1) \models U(\beta_1)$, for some uncovered node
-  $\beta_1 \in \dom(U)$.
-  %% for every word $w'$ and every cube $c \in
-  %% \cube{\sem{U(\alpha_1)}}$ such that $\A$ accepts $w'$ starting
-  %% with $c$, there exists a cube $d \in \cube{\sem{U(\beta_1)}}$
-  %% such that $\A$ accepts $w'$ starting with $d$.
-  Let $\gamma_1$ be the unique sequence such that $\alpha_1\gamma_1 =
-  \alpha$. By Definition \ref{def:coverage}, since $\alpha_1 \cover
-  \beta_1$ and $\event{w} = \alpha_1\gamma_1 \in \lang{\A}$, there
-  exists a word $w_1$ and a cube $c_1 \in \cube{\sem{U(\alpha_1)}}
-  \subseteq \cube{\sem{U(\beta_1)}}$, such that $\event{w_1} =
-  \gamma_1$ and $\A$ accepts $w_1$ starting with $c_1$. If
-  $\beta_1\gamma_1 \in \dom(U)$, we obtain a contradiction by a
-  similar argument as above. Hence $\beta_1\gamma_1 \not\in \dom(U)$
-  and there exists a leaf of $\dom(U)$ which is also a prefix of
-  $\beta_1\gamma_1$. Since $U$ is closed, this leaf is covered by an
-  uncovered node $\beta_2 \in \dom(U)$ and let $\alpha_2 \in \dom(U)$
-  be the minimal (in the prefix partial order) node such that $\beta_1
-  \prefix \alpha_2 \prefix \beta_1\gamma_1$ and $\alpha_2 \cover
-  \beta_2$. Let $\gamma_2$ be the unique sequence such that
-  $\alpha_2\gamma_2 = \beta_1\gamma_1$. Since $\beta_1$ is uncovered,
-  we have $\beta_1 \neq \alpha_2$ and thus $\len{\gamma_1} >
-  \len{\gamma_2}$. By repeating the above reasoning for $\alpha_2$,
-  $\beta_2$ and $\gamma_2$, we obtain an infinite sequence
-  $\len{\gamma_1} > \len{\gamma_2} > \ldots$, which is again a
-  contradiction. \qed}
-
-As mentioned above, we check emptiness of first-order alternating
-automata using the same method previously used to check emptiness of a
-simpler model of alternating automata, which uses boolean constants
-for control states and whose transition rules have no quantifiers
-\cite{IosifXu18}. The higher complexity of the automata model
-considered here, manifests itself within the interpolant generation
-procedure, used to refine the labeling of the unfolding. We discuss
-generation of interpolants in the next section.
-
-\section{Interpolant Generation}\label{sec:interpolants}
-
-Typically, when checking the unreachability of a set of program
-configurations \cite{mcmillan06}, the interpolants used to annotate
-the unfolded control structure are assertions about the values of the
-program variables in a given control state, at a certain step of a
-given execution. However, in an alternating model of computation, it
-is useful to distinguish between
-\begin{inparaenum}[(i)]
-\item locality of interpolants w.r.t. a given control state (control
-  locality) and
-%
-\item locality w.r.t. a given time stamp (time locality).
-\end{inparaenum} 
-In logical terms, \emph{control-local} interpolants are defined by
-formulae involving a single predicate symbol, whereas
-\emph{time-local} interpolants involve only predicates $\stamp{q}{i}$
-and variables $\stamp{x}{i}$, for a single $i \geq 0$.
-
-\paragraph{Remark}
-When considering an alternating model of computation, control-local
-interpolants are not always enough to prove emptiness, because of the
-synchronization of several branches of the computation on the same
-sequence of input values. Consider, for instance, an automaton with
-the following transition rules and final state $q_f$:
-\[\begin{array}{rclcrcl}
-q_0(y) & \arrow{a(x)}{} & q_1(y+x) \wedge q_2(y-x) & \hspace*{2mm} & 
-q_1(y) \arrow{a(x)}{} y+x > 0 \wedge q_f \\
-q_1(y) & \arrow{a(x)}{} & q_1(y+x) && q_2(y) \arrow{a(x)}{} y-x > 0 \wedge q_f \\
-q_2(y) & \arrow{a(x)}{} & q_2(y-x)
-\end{array}\]
-Started in an initial configuration $q_0(0)$ with an input word
-$(a,\nu_1) \ldots (a,\nu_{n-1}) (a,\nu_n)$, such that $\nu_i(x)=k_i$,
-the automaton executes as follows:
-\[\begin{array}{c}
-q_0(0) \arrow{(a,\nu_1)}{} \set{q_1(k_1),q_2(-k_1)} \ldots \arrow{(a,\nu_{n-1})}{} 
-\set{q_1(\sum_{i=1}^{n-1} k_i),q_2(-\sum_{i=1}^{n-1} k_i)}
-\arrow{(a,\nu_n)}{} \emptyset
-\end{array}\]
-An overapproximation of the set of cubes generated after one or more
-steps is defined by the formula: $\exists x_1 \exists x_2 ~.~ q_1(x_1)
-\wedge q_2(x_2) \wedge x_1+x_2 \teq 0$. Observe that a control-local
-formula using one occurrence of a predicate would give a too rough
-overapproximation of this set, unable to prove the emptiness of the
-automaton. \hfill$\blacksquare$
-
-First, let us give the formal definition of the class of interpolants
-we shall work with. Given a formula $\phi$, the \emph{vocabulary} of
-$\phi$, denoted $\voc{\phi}$ is the set of predicate symbols $q \in
-\stamp{Q}{i}$ and variables $x \in \stamp{X}{i}$, occurring in $\phi$,
-for some $i\geq0$. For a term $t$, its vocabulary $\voc{t}$ is the set
-of variables that occur in $t$. Observe that quantified variables and
-the interpreted function symbols of the data
-theory\footnote{E.g.,\ the arithmetic operators of addition and
-  multiplication, when $\Data$ is the set of integers.}  do not belong
-to the vocabulary of a formula. By $\pset{+}{\phi}$ [$\pset{-}{\phi}$]
-we denote the set of predicate symbols that occur in $\phi$ under an
-even [odd] number of negations.
-
-\begin{definition}[\cite{Lyndon59}]\label{def:lyndon-interpolant}
-Given formulae $\phi$ and $\psi$ such that $\phi \wedge \psi$ is
-unsatisfiable, a \emph{Lyndon interpolant} is a formula $I$ such that
-$\phi \models I$, the formula $I \wedge \psi$ is unsatisfiable,
-$\voc{I} \subseteq \voc{\phi} \cap \voc{\psi}$, $\pset{+}{I} \subseteq
-\pset{+}{\phi} \cap \pset{+}{\psi}$ and $\pset{-}{I} \subseteq
-\pset{-}{\phi} \cap \pset{-}{\psi}$.
-\end{definition}
-%% The existence of Lyndon interpolants for first-order logic, with no
-%% predefined interpretation of function symbols\footnote{A model
-%%   specifies the universe and the interpretation of function symbols.},
-%% was proved by a non-constructive model-theoretic argument
-%% \cite{Lyndon59}. In the following, we shall consider only those
-%% theories of the data domain $\Data$, for which there exists a decision
-%% procedure for satisfiability, able to produce Lyndon interpolants. For
-%% instance, such interpolating decision procedures exist for linear real
-%% and integer arithmetic, with uninterpreted function symbols (UFLRA and
-%% UFLIA).
-
-In the rest of this section, let us fix an automaton $\A =
-\tuple{\Sigma,X,Q,\iota,F,\Delta}$. Due to the above observation, none
-of the interpolants considered will be control-local and we shall use
-the term \emph{local} to denote time-local interpolants, with no free
-variables.
-
-\begin{definition}\label{def:generalized-lyndon-interpolant}
-  Given a non-empty sequence of input events $\alpha = a_1 \ldots a_n
-  \in \Sigma^*$, a \emph{generalized Lyndon interpolant (GLI)} is a
-  sequence $(I_0,\ldots,I_n)$ of formulae such that, for all $k \in
-  [n-1]$: \begin{compactenum}
-    \item\label{it1:generalized-lyndon-interpolant} $\pset{-}{I_k} =
-      \emptyset$,
-      %
-    \item\label{it2:generalized-lyndon-interpolant} $\stamp{\iota}{0}
-      \models I_0$ and \(I_k \wedge \Big(\bigwedge_{q(\vec{y})
-        \arrow{a_i(X)}{} \psi \in \Delta} \forall y_1 \ldots \forall
-      y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y}) \rightarrow
-      \stamp{\psi}{k+1}\Big) \models I_{k+1}\),
-      %
-    \item\label{it3:generalized-lyndon-interpolant} $I_n \wedge
-      \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
-      y_{\#(q)} ~.~ q(\vec{y})$ is unsatisfiable.
-  \end{compactenum}
-  Moreover, the GLI is \emph{local} if and only if $\voc{I_k}
-  \subseteq \stamp{Q}{k}$, for all $k \in [n]$.
-\end{definition}
-The following proposition states the existence of local GLI for the
-theories in which Lyndon's Interpolation Theorem holds.
-
-\begin{proposition}\label{prop:local-gli}
-  If there exists a Lyndon interpolant for any two formulae $\phi$ and
-  $\psi$, such that $\phi \wedge \psi$ is unsatisfiable, then any
-  sequence of input events $\alpha = a_1 \ldots a_n \in \Sigma^*$,
-  such that $\accform{\alpha}$ is unsatisfiable, has a local GLI
-  $(I_0,\ldots,I_n)$.
-\end{proposition}
-\proof{ By definition, $\accform{\alpha}$ is the formula:
-  \[\stamp{\iota}{0} \wedge 
- \bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in
-   \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
- \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i} \wedge
- \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
- ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot\]
- We define the formulae: 
- \[\begin{array}{rcl}
- \varphi_i & \isdef & \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
- \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i},~ \text{for all}~ i \in [1,n] \\
- \psi & \isdef & \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
- ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot
- \end{array}\]
- Observe that $\voc{\stamp{\iota}{0}} \subseteq \stamp{Q}{0}$,
- $\voc{\varphi_i} \subseteq
- \stamp{Q}{i-1}\cup\stamp{Q}{i}\cup\stamp{X}{i}$, for all $i \in
-       [1,n]$, and $\voc{\psi} \subseteq \stamp{Q}{n}$.  We apply
-       Lyndon's Interpolation Theorem for the formulae
-       $\stamp{\iota}{0}$ and $\bigwedge_{i=1}^n \varphi_i \wedge
-       \psi$ and obtain a formula $I_0$, such that $\stamp{\iota}{0}
-       \models I_0$, $I_0 \wedge \bigwedge_{i=1}^n \varphi_i \wedge
-       \psi$ is unsatisfiable, $\voc{I_0} \subseteq
-       \voc{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n \voc{\varphi_i}
-       \cup \voc{\psi}) \subseteq \stamp{Q}{0}$ and $\pset{-}{I_0}
-       \subseteq \pset{-}{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n
-       \pset{-}{\varphi} \cup \pset{-}{\psi}) = \emptyset$. Repeating
-       the reasoning for the formulae $I_0 \wedge \varphi_1$ and
-       $\bigwedge_{i=2}^n \varphi_i \wedge \psi$, we obtain $I_1$,
-       such that $I_0 \wedge \varphi_1 \models I_1$, $I_1 \wedge
-       \bigwedge_{i=2}^n \varphi_i \wedge \psi$ is unsatisfiable,
-       $\voc{I_1} \subseteq (\voc{I_0} \cup \voc{\varphi_1}) \cap
-       (\bigcup_{i=2}^n \voc{\varphi_i} \cup \voc{\psi}) \subseteq
-       \stamp{Q}{1}$ and $\pset{-}{I_1} \subseteq (\pset{-}{I_0} \cup
-       \pset{-}{\varphi_1}) \cap (\bigcup_{i=2}^n \pset{-}{\varphi_i}
-       \cup \pset{-}{\psi}) = \emptyset$. Continuing in this way, we
-       obtain formulae $I_0, I_1, \ldots, I_{n}$ as required. \qed}
-
-The main problem with the local GLI construction described in the
-proof of Proposition \ref{prop:local-gli} is that the existence of
-Lyndon interpolants (Definition \ref{def:lyndon-interpolant}) is
-guaranteed in principle, but the proof is non-constructive. Building
-an interpolant for an unsatisfiable conjunction of formulae $\phi
-\wedge \psi$ is typically the job of the decision procedure that
-proves the unsatisfiability and, in general, there is no such
-procedure, when $\phi$ and $\psi$ contain predicates and have
-non-trivial quantifier alternation. In this case, some provers use
-instantiation heuristics for the universal quantifiers that are
-sufficient for proving unsatisfiability, however these heuristics are
-not always suitable for interpolant generation. Consequently, from now
-on, we assume the existence of an effective Lyndon interpolation
-procedure only for decidable theories, such as the quantifier-free
-linear (integer) arithmetic with uninterpreted functions (UFLIA,
-UFLRA, etc.) \cite{RybalchenkoSofronieStokkermans}.
-
-This is where the predicate-free path formulae (Definition
-\ref{def:substform}) come into play. For a given event sequence
-$\alpha$, the automaton $\A$ accepts a word $w$ such that $\event{w} =
-\alpha$ if and only if $\substform{\alpha}$ is satisfiable. Assuming
-further that the equality atoms in the transition rules of $\A$ are
-written in the language of a decidable first-order theory, such as
-Presburger arithmetic, Lemma \ref{lemma:quant-pred-acceptance} gives
-us an effective way of checking emptiness of $\A$, relative to a given
-event sequence. However, this method does not cope well with lazy
-annotation, because there is no way to extract, from the
-unsatisfiability proof of $\substform{\alpha}$, the interpolants
-needed to annotate $\alpha$. This is because \begin{inparaenum}[(i)]
-\item\label{it1:gli} the formula $\substform{\alpha}$, obtained by
-  repeated substitutions (Definition \ref{def:substform}) loses track
-  of the steps of the execution, and
-%
-\item\label{it2:gli} quantifiers that occur nested in
-  $\substform{\alpha}$ make it difficult to write $\substform{\alpha}$
-  as an unsatisfiable conjunction of formulae from which interpolants
-  are extracted (Definition \ref{def:lyndon-interpolant}).
-\end{inparaenum}
-
-The solution we adopt for the first issue (\ref{it1:gli}) consists in
-partially recovering the time-stamped structure of the acceptance
-formula $\accform{\alpha}$ using the formula $\quantform{\alpha}$, in
-which only transition quantifiers occur. The second issue
-(\ref{it2:gli}) is solved under the additional assuption that the
-theory of the data domain $\Data$ has \emph{witness-producing
-  quantifier elimination}. More precisely, we assume that, for each
-formula $\exists x ~.~ \phi(x)$, there exists an effectively
-computable term $\tau$, in which $x$ does not occur, such that
-$\exists x ~.~ \phi$ and $\phi[\tau/x]$ are equisatisfiable. These
-terms, called \emph{witness terms} in the following, are actual
-definitions of the Skolem function symbols from the following folklore
-theorem:
-
-\begin{theorem}\label{thm:skolem}
-  Given $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$ a first order sentence, 
-  where $Q_1, \ldots, Q_n \in \set{\exists,\forall}$ and $\phi$ is
-  quantifier-free, let: 
-  \[\eta_i \equiv \left\{\begin{array}{ll}
-  f_i(y_1,\ldots,y_{k_i}) & \text{if } Q_i = \forall \\
-  x_i & \text{if } Q_i = \exists
-  \end{array}\right.\]
-  where $f_i$ is a fresh function symbol and $\set{y_1, \ldots,
-    y_{k_i}} = \set{x_j \mid j < i,~ Q_j = \exists}$. Then the entailment \(Q_1
-  x_1 \ldots Q_n x_n ~.~ \phi \models
-  \phi[\eta_1/x_1,\ldots,\eta_n/x_n]\) holds. 
-\end{theorem}
-\proof{See \cite[Theorem 2.1.8]{BorgerGraedelGurevich97} and
-  \cite[Lemma 2.1.9]{BorgerGraedelGurevich97}. \qed}
-
-Examples of witness-producing quantifier elimination procedures can be
-found in the literature for e.g. linear integer (real) arithmetic
-(LIA,LRA), Presburger arithmetic and boolean algebra of sets and
-Presburger cardinality constraints (BAPA)
-\cite{KuncakMayerPiskacSuter12}.
-
-%% Furthermore, we describe a cubic-time witness generation for
-%% difference bounds constraints, a simple subset of linear integer
-%% (real) arithmetic (\S\ref{sec:applications}).
-
-Under the assumption that witness terms can be effectively built, let
-us describe the generation of a non-local GLI for a given input event
-sequence $\alpha = a_1 \ldots a_n$. First, we generate successively
-the acceptance formula $\accform{\alpha}$ and its equisatisfiable
-forms $\quantform{\alpha} = Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$
-and $\substform{\alpha} = Q_1x_1 \ldots Q_mx_m ~.~ \overline{\Phi}$,
-both written in prenex form, with matrices $\widehat{\Phi}$ and
-$\overline{\Phi}$, respectively. Because we assumed that the
-first-order theory of $\Data$ has quantifier elimination, the
-satisfiability problem for $\substform{\alpha}$ is decidable. If
-$\substform{\alpha}$ is satisfiable, we build a counterexample for
-emptiness $w$ such that $\event{w}=\alpha$ and $\data{w}$ is a
-satisfying assignment for $\substform{\alpha}$. Otherwise,
-$\substform{\alpha}$ is unsatisfiable and there exist witness terms
-$\tau_{i_1} \ldots \tau_{i_\ell}$, where $\set{i_1, \ldots, i_\ell} =
-\set{j \in [1,m] \mid Q_j = \forall}$, such that
-$\overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
-  \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable (Theorem
-\ref{thm:skolem}). Then it turns out that the formula
-$\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
-  \tau_{i_\ell}/x_{i_\ell}]$, obtained analogously from the matrix of
-$\quantform{\alpha}$, is unsatisfiable as well (Lemma
-\ref{lemma:subst-quant}). Because this latter formula is structured as
-a conjunction of formulae $\stamp{\iota}{0} \wedge \phi_1 \ldots
-\wedge \phi_n \wedge \psi$, where $\voc{\phi_k} \cap \stamp{Q}{\leq n}
-\subseteq \stamp{Q}{k-1} \cup \stamp{Q}{k}$ and $\voc{\psi} \cap
-\stamp{Q}{\leq n} \subseteq \stamp{Q}{n}$, it is now possible to use
-an existing interpolation procedure for the quantifier-free theory of
-$\Data$, extended with uninterpreted function symbols, to compute a
-sequence of non-local GLI $(I_0, \ldots, I_n)$ such that $\voc{I_k}
-\cap \stamp{Q}{\leq n} \subseteq \stamp{Q}{k}$, for all $k \in [n]$.
-
-%% Since each formula $I_k$ of the GLI may refer to the current
-%% predicates $\stamp{Q}{k}$ and to the past and current input variables
-%% $\stamp{X}{\leq k}$, we further give up on the locality of
-%% interpolants and work with non-local GLI that are effectively
-%% computable by a combination of quantifier elimination with
-%% interpolation for quantifier-free data theories with uninterpreted
-%% functions.
-
-\begin{example}[Contd. from Examples \ref{ex:quant-elim} and \ref{ex:pred-elim}]\label{ex:gli}
-The formula $\substform{\alpha}$ (Example \ref{ex:pred-elim}) is
-unsatisfiable and let $\tau_2 = z_1$ be the witness term for the
-universally quantified variable $z_2$. Replacing $z_2$ with $\tau_2$
-in the matrix of $\quantform{\alpha}$ (Example \ref{ex:quant-elim})
-yields the unsatisfiable conjunction: 
-\[\begin{array}{c}
-z_1 \geq 0 \wedge \stamp{q}{0}(z_1)  \\ \wedge \\
-\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge (z_1 \geq z_1 \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_1)) \\ \wedge \\
-\stamp{q}{1}(\stamp{x}{1}+z_1) \rightarrow \stamp{x}{1}+z_1 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_1)
-\end{array}\]
-A non-local GLI for the above is $(\stamp{q}{0}(z_1) \wedge
-z_1\geq 0,~ \stamp{x}{1} \geq 0 \wedge \stamp{q}{1}(\stamp{x}{1}+z_1)
-\wedge z_1\geq0,~\bot)$. \hfill$\blacksquare$
-\end{example}
-
-A function $\xi : \nat \rightarrow \nat$ is [strictly]
-\emph{monotonic} iff for each $n < m$ we have $\xi(n) \leq \xi(m)$
-     [$\xi(n) < \xi(m)$] and \emph{finite-range} iff for each $n \in
-     \nat$ the set $\set{m \mid \xi(m)=n}$ is finite. If $\xi$ is
-     finite-range, we denote by $\xi_{\max}^{-1}(n) \in \nat$ the
-     maximal value $m$ such that $\xi(m)=n$. It is easy to see that
-     $\xi_{\max}^{-1}$ is strictly monotonic if $\xi$ is monotonic and
-     finite-range. The lemma below gives the proof of correctness for
-     the previous construction of non-local GLI.
-
-\begin{lemma}\label{lemma:subst-quant}
-  Given a non-empty input event sequence $\alpha = a_1 \ldots a_n \in
-  \Sigma^*$, such that $\accform{\alpha}$ is unsatisfiable, let
-  $Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$ be a prenex form of
-  $\quantform{\alpha}$ and let $\xi : [1,m] \rightarrow [n]$ be a
-  monotonic function mapping each transition quantifier to the minimal
-  index from the sequence $\quantformn{\alpha_0}, \ldots,
-  \quantformn{\alpha_n}$ where it occurs. Then the following
-  hold: \begin{compactenum}
-  \item\label{it1:subst-quant} one can effectively build witness terms
-    $\tau_{i_1}, \ldots, \tau_{i_\ell}$, where 
-    \[\begin{array}{l}   
-    \set{i_1, \ldots,i_\ell} = \set{j \in [1,m] \mid Q_j = \forall} \text{ and} \\
-    \voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup \set{x_k
-    \mid k < i_j, Q_k = \exists},~ \forall j \in [1,\ell]
-    \end{array}\]
-    such that $\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable, and
-    %
-  \item\label{it2:subst-quant} one can effectively build a GLI $(I_0,
-    \ldots, I_n)$ for $\alpha$, such that \[\voc{I_k} \subseteq
-    \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
-      \xi^{-1}(k),~ Q_j = \exists},~ \forall k \in [n]\enspace.\]
-  \end{compactenum}
-\end{lemma}
-\proof{(\ref{it1:subst-quant}) If $\accform{\alpha}$ is unsatisfiable,
-  by Lemmas \ref{lemma:quant} and \ref{lemma:subst}, we obtain that,
-  successively $\quantform{\alpha}$ and $\substform{\alpha}$ are
-  unsatisfiable. Let $Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$ and
-  $Q_1x_1 \ldots Q_mx_m ~.~ \overline{\Phi}$ be prenex forms for
-  $\quantform{\alpha}$ and $\substform{\alpha}$, respectively. Since
-  we assumed that the first-order theory of the data domain has
-  witness-producing quantifier elimination, using Theorem
-  \ref{thm:skolem} one can effectively build witness terms
-  $\tau_{i_1}, \ldots, \tau_{i_\ell}$, where
-  $\set{i_1,\ldots,i_\ell}=\set{i \in [1,m] \mid Q_i=\forall}$
-  and: \begin{compactitem}
-  \item $\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup
-    \set{x_k \mid k < i_j, Q_k = \exists}$, for all $j \in [1,\ell]$ and
-    %
-  \item $\overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
-    \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.
-  \end{compactitem}
-  Let $\widehat{\Phi}_0$, \ldots, $\widehat{\Phi}_n$ be the sequence
-  of quantifier-free formulae, defined as follows: \begin{compactitem}
-  \item $\widehat{\Phi}_0$ is the matrix of some prenex form of
-    $\stamp{\iota}{0}$,
-    %
-  \item for all $i=1,\ldots,n$, let $\widehat{\Phi}_i$ be the matrix
-    of some prenex form of:
-    \[\widehat{\Phi}_i \isdef \widehat{\Phi}_{i-1} \wedge
-     \underbrace{\bigwedge_{\begin{array}{l}
-         \scriptstyle{\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \widehat{\Phi}_{i-1}} \\
-        \scriptstyle{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta} 
-     \end{array}} \stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow 
-     \stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]}_{\isdef \phi_i}\]  
-  \end{compactitem}
-  It is easy to see that $\widehat{\Phi}$ is the matrix of some prenex
-  form of:
-  \[\widehat{\Phi}_n \wedge \underbrace{\bigwedge_{\begin{array}{l}
-        \scriptstyle{\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \widehat{\Phi}_n} \\
-        \scriptstyle{q \in Q \setminus F}
-    \end{array}} \stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot}_{\isdef \psi}\]
-  Applying the equivalence from Fact \ref{fact:subst} in the proof of
-  Lemma \ref{lemma:subst}, we obtain a sequence of quantifier-free
-  formulae $\overline{\Phi}_0, \ldots, \overline{\Phi}_n$ such that
-  $\overline{\Phi}_i \equiv \widehat{\Phi}_i$, for all $i \in [n]$ and
-  $\overline{\Phi}$ is obtained from $\overline{\Phi}_n$ by replacing
-  each occurrence of a predicate atom $q(t_1,\ldots,t_{\#(q)})$ in
-  $\overline{\Phi}_n$ by $\bot$ if $q \in Q\setminus F$ and by $\top$
-  if $q \in F$. Clearly $\overline{\Phi} \equiv \widehat{\Phi}$, thus
-  $\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
-    \tau_{i_\ell}/x_{i_\ell}] \equiv
-  \overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
-    \tau_{i_\ell}/x_{i_\ell}] \equiv \bot$.
-
-  \noindent(\ref{it2:subst-quant}) With the notation introduced at
-  point (\ref{it1:subst-quant}), we have $\widehat{\Phi} =
-  \widehat{\Phi}_0 \wedge \bigwedge_{i=1}^n \phi_i \wedge
-  \psi$. Consider the sequence of witness terms $\tau_{i_1}, \ldots,
-  \tau_{i_\ell}$, whose existence is proved by point
-  (\ref{it1:subst-quant}). Because $\voc{\tau_{i_j}} \subseteq
-  \stamp{X}{\leq \xi(i_j)} \cup \set{x_k \mid k < i_j, Q_k =
-    \exists}$, for all $j \in [1,\ell]$, and moreover $\xi^{-1}$ is
-  strictly monotonic, we obtain: \begin{compactitem}
-  \item $\voc{\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{0} \cup
-    \stamp{X}{0} \cup \set{x_j \mid j < \xi_{\max}^{-1}(0), Q_j = \exists}$,
-    %
-  \item $\voc{\phi_i[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{i-1} \cup
-    \stamp{Q}{i} \cup \stamp{X}{\leq i} \cup \set{x_j \mid j <
-      \xi_{\max}^{-1}(i), Q_j = \exists}$, for all $i \in [1,n]$,
-    %
-  \item $\voc{\psi[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{n} \cup
-    \stamp{X}{\leq n} \cup \set{x_j \mid j \in [1,m], Q_j = \exists}$.
-  \end{compactitem}
-  By repeatedly applying Lyndon's Interpolation Theorem, we obtain a
-  sequence of formulae $(I_0, \ldots, I_n)$ such that: \begin{compactitem}
-    \item $\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}] \models I_0$ and $\voc{I_0} \subseteq
-      \stamp{Q}{0} \cup \stamp{X}{0} \cup \set{x_j \mid j <
-        \xi_{\max}^{-1}(0), Q_j = \exists}$,
-      %
-    \item $I_{k-1} \wedge \phi_i[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}] \models I_k$ and $\voc{I_k} \subseteq
-      \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
-       \xi_{\max}^{-1}(k), Q_j = \exists}$, for all $k \in [1,n]$,
-      %
-    \item $I_n \wedge \psi[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.
-  \end{compactitem}
-  To show that $(I_0,\ldots,I_n)$ is a GLI for $a_1\ldots a_n$, it is
-  sufficient to notice that \[\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
-    \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{k}(\vec{y}) \rightarrow \stamp{\psi}{k+1} \models \phi_k\]
-  for all $k \in [1,n]$. Consequently, we obtain: \begin{compactitem}
-    \item $\stamp{\iota}{0} \models \widehat{\Phi}_0 \models I_0$, by
-      Theorem \ref{thm:skolem},
-      %
-    \item $I_{k-1} \wedge \Big(\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
-      \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-      \stamp{q}{k-1}(\vec{y}) \rightarrow \stamp{\psi}{k}\Big) \models
-      I_{k-1} \wedge \phi_k \models I_k$, and 
-      %
-    \item $I_n \wedge \Big(\bigwedge_{q \in Q \setminus F} \forall y_1
-      \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot\Big)
-      \models I_n \wedge \psi \models \bot$,     
-  \end{compactitem}
-  as required by Definition
-  \ref{def:generalized-lyndon-interpolant}. \qed}
-
-In conclusion, under two assumptions about the first-order theory of
-the data domain, namely the \begin{inparaenum}[(i)]
-\item witness-producing quantifier elimination, and
-\item Lyndon interpolation for the quantifier-free fragment with
-  uninterpreted functions,
-\end{inparaenum}
-we developped a rather generic method that produces generalized Lyndon
-interpolants for unfeasible input event sequences. Moreover, each
-formula $I_k$ in the interpolant refers only to the current predicate
-symbols $\stamp{Q}{I_k}$, the current and past input variables
-$\stamp{X}{\leq k}$ and the existentially quantified transition
-variables introduced at the previous steps $\set{x_j \mid j <
-  \xi_{\max}^{-1}(k), Q_j = \exists}$. The remaining question is how
-to use such non-local interpolants to label the unfolding of an
-automaton (Definition \ref{def:unfolding}) and to compute the coverage
-between nodes of the unfolding (Definition \ref{def:coverage}).
-
-\subsection{Unfolding with Non-local Interpolants}
-
-As required by Definition \ref{def:unfolding}, the unfolding $U$ of an
-automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$ is labeled by
-formulae $U(\alpha) \in \posforms(Q,\emptyset)$, with no free symbols,
-other than predicate symbols, such that the labeling is compatible
-with the transition relation of the automaton, according to the point
-(\ref{it3:unfolding}) of Definition \ref{def:unfolding}. The following
-lemma describes the refinement of the labeling of an input sequence
-$\alpha$ of length $n$ by a non-local GLI $(I_0, \ldots, I_n)$, such
-that $\voc{I_k} \subseteq \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup
-\vec{x}_k$, where $\vec{x}_k$ are the existentially quantified
-variables from the prenex normal form of $\quantform{\alpha_k}$.
-
-\begin{lemma}\label{lemma:refinement}
-  Let $U$ be an unfolding of an automaton $\A =
-  \tuple{\Sigma,X,Q,\iota,F,\Delta}$ such that $\alpha = a_1\ldots a_n
-  \in \dom(U)$ and $(I_0, \ldots, I_n)$ be a GLI for $\alpha$. The
-  mapping $U' : \dom(U) \rightarrow \posforms(Q,\emptyset)$ defined as: 
-  \begin{compactitem}
-  \item $U'(\alpha_k) = U(\alpha_k) \wedge J_k$, for all $k \in [n]$,
-    where $J_k$ is the formula obtained from $I_k$ by replacing each
-    time-stamped predicate symbol $\stamp{q}{k}$ by $q$ and
-    existentially quantifying each free variable in $I_k$,
-  %
-\item $U'(\beta) = U(\beta)$ if $\beta \in \dom(U)$ and $\beta
-  \not\prefix \alpha$,
-  \end{compactitem}
-  is an unfolding of $\A$. 
-\end{lemma}
-\proof{
-  The new set of formulae $U'(\alpha_0), \ldots, U'(\alpha_n)$ complies
-  with Definition \ref{def:unfolding}, because: \begin{compactitem}
-  \item $U'(\alpha_0) \equiv \iota$, since, by point
-    \ref{it2:generalized-lyndon-interpolant} of Definition
-    \ref{def:generalized-lyndon-interpolant}, we have $\stamp{\iota}{0}
-    \models I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) =
-    U(\alpha_0) \wedge J_0 \equiv \iota \wedge J_0 \equiv \iota$, and
-    %
-  \item by Definition \ref{def:generalized-lyndon-interpolant}
-    (\ref{it3:generalized-lyndon-interpolant}) we have, for all $k \in
-    [n-1]$:
-    \[I_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
-    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
-    \rightarrow \stamp{\psi}{k+1} \models I_{k+1}\] We write
-    $\overstamp{I}{j}_k$ for the formula in which each predicate
-    symbol $\stamp{q}{k}$ is replaced by $\stamp{q}{j}$. Then the
-    following entailment holds:
-    \[\overstamp{I}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
-    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
-    \rightarrow \stamp{\psi}{1} \models \overstamp{I}{1}_{k+1}\]
-    Because $J_k$ is obtained by removing the time stamps from the predicate symbols and  
-    existentially quantifying all the free variables of $I_k$, we also
-    obtain, applying Fact \ref{fact:ex-entail} below:
-    \[\stamp{J}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
-    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
-    \rightarrow \stamp{\psi}{1} \models \stamp{J}{1}_{k+1}\] Since $U$
-    satisfies the labeling condition of Definition \ref{def:unfolding}
-    (\ref{it3:unfolding}) and $U'(\alpha_k) = U(\alpha_k) \wedge J_k$,
-    we obtain, as required: 
-    \[\stamp{U'(\alpha_k)}{0} \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
-    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
-    \rightarrow \stamp{\psi}{1} \models
-    \stamp{U'(\alpha_{k+1})}{1}\enspace.\]
-  \end{compactitem}
-
-  \begin{fact}\label{fact:ex-entail}
-    Given formulae $\phi(\vec{x}, \vec{y})$ and $\psi(\vec{x})$ such
-    that $\phi(\vec{x}, \vec{y}) \models \psi(\vec{x})$, we also have
-    $\exists \vec{x} ~.~ \phi(\vec{x},\vec{y}) \models \exists \vec{x}
-    ~.~ \psi(\vec{x})$. 
-  \end{fact}
-  \proof{ For each choice of a valuation for the existentially
-    quantified variables on the left-hand side, we chose the same
-    valuation for the variables on the right-hand side. \qed}\qed}
-
-Observe that, by Lemma \ref{lemma:subst-quant}
-(\ref{it2:subst-quant}), the set of free variables of a GLI formula
-$I_k$ consists of \begin{inparaenum}[(i)]
-\item variables $\stamp{X}{\leq k}$ keeping track of data values seen
-  in the input at some earlier moment in time, and
-%
-\item variables that track past choices made within the transition
-  rules.
-\end{inparaenum}
-Basically, it is not important when exactly in the past a certain
-input has been read or when a choice has been made, as only the value
-of the variable determines the future behavior. Intuitively,
-existential quantification of these variables does the job of ignoring
-when in the past these values have been seen.
-
-The last ingredient of the lazy annotation semi-algorithm based on
-unfoldings consist in the implementation of the coverage check, when
-the unfolding of an automaton is labeled with conjunctions of
-existentially quantified formulae with predicate symbols, obtained
-from interpolation. By Definition \ref{def:coverage}, checking whether
-a given node $\alpha \in \dom(U)$ is covered amounts to finding a
-prefix $\alpha' \prefix \alpha$ and a node $\beta \in \dom(U)$ such
-that $U(\alpha') \models U(\beta)$, or equivalently, the formula
-$U(\alpha') \wedge \neg U(\beta)$ is unsatisfiable. However, the
-latter formula, in prenex form, has quantifier prefix in the language
-$\exists^*\forall^*$ and, as previously mentioned, the satisfiability
-problem for such formulae becomes undecidable when the data theory
-subsumes Presburger arithmetic \cite{Halpern91}.
-
-Nevertheless, if we require just a yes/no answer (i.e.\ not an
-interpolant) recently developped quantifier instantiation heuristics
-\cite{ReynoldsKK17} perform rather well in answering a large number of
-queries in this class. Observe, moreover, that coverage does not need
-to rely on a complete decision procedure. If the prover fails in
-answering the above satisfiability query, then the semi-algorithm
-assumes that the node is not covered and continues exploring its
-successors. Failure to compute complete coverage may lead to
-divergence (non-termination) and ultimately, to failure to prove
-emptiness, but does not affect the soundness of the semi-algorithm
-(real counterexamples will still be found).
-
-\section{Applications}
-\label{sec:applications}
-
-The main application of first-order alternating automata is checking
-inclusion between various classes of automata extended with variables
-ranging over infinite domains that recognize languages over infinite
-alphabets. The most widely known such classes are \emph{timed
-  automata} \cite{AlurDill94} and \emph{finite-memory (register)
-  automata} \cite{KaminskiFrancez94}. In both cases, complementation
-is not possible inside the class and inclusion is undecidable. Our
-contribution is providing a systematic semi-algorithm for these
-decision problems. In addition, the method described in
-\S\ref{sec:emptiness} can extend our previous \emph{generic register
-  automata} \cite{HolikIosifRogalewiczVojnar18} inclusion checking
-framework, by allowing monitor (right-hand side) automata to have
-local variables, that are not visible in the language\footnote{We
-  conjecture that hidden computations using invisible variables
-  (clocks) are the reason for which timed and register automata cannot
-  be complemented within their class.}.
-
-Another application is checking safety (mutual exclusion, absence of
-deadlocks, etc.) and liveness (termination, lack of starvation, etc.)
-properties of parameterized concurrent programs, consisting of an
-unbounded number of replicated threads that communicate via a fixed
-set of global variables (locks, counters, etc.). The verification of
-parametric programs has been reduced to checking the emptiness of a
-(possibly infinite) sequence of first-order alternating automata,
-called \emph{predicate automata} \cite{Farzan15,Farzan16}, encoding
-the inclusion of the set of traces of a parametric concurrent program
-into increasingly general \emph{proof spaces}, obtained by
-generalization of counterexamples. The program and the proof spaces
-are first-order alternating automata over the infinite alphabet of
-pairs consisting of program statements and thread identifiers.
-
-\subsection{Timed Automata}
-\label{sec:ta}
-
-The standard definition of a finite \emph{timed word} is a sequence of
-pairs $(a_1,\tau_1), \ldots, (a_n,\tau_n) \in (\Sigma \times
-\real)^*$, where $\real$ is the set of real numbers, such that $0 \leq
-\tau_i < \tau_{i+1}$, for all $i \in [1,n-1]$. Intuitively, $\tau_i$
-is the moment in time where the input event $a_i$ occurs. Given a set
-$C$ of \emph{clocks}, the set $\Phi(C)$ of \emph{clock constraints} is
-defined inductively as the set of formulae $x \leq c$, $x \geq c$,
-$\neg\delta$, $\delta_1 \wedge \delta_2$, where $x \in C$, $c \in
-\rat$ is a rational constant and $\delta,\delta_1,\delta_2 \in
-\Phi(X)$.
-
-A \emph{timed automaton} is a tuple $T = \tuple{\Sigma,S,S_0,F,C,E}$,
-where: $\Sigma$ is a finite set of input events, $S$ is a finite set
-of states, $S_0, F \subseteq S$ are sets of initial and final states,
-respectively, $C$ is a finite set of clocks and $E \subseteq S \times
-\Sigma \times S \times 2^C \times \Phi(C)$ is the set of transitions
-$(s,a,s',\lambda,\delta)$ from state $s$ to state $s'$ with symbol
-$a$, $\lambda$ is the set of clocks to be reset and $\delta$ is a
-clock constraint. A run of $T$ over a timed word $w = (a_1,\tau_1)
-\ldots (a_n,\tau_n)$ is a sequence $(s_0,\gamma_0) \ldots
-(s_n,\gamma_n)$, where $s_i \in S$, $\gamma_i : C \rightarrow \real$
-are clock valuations, for all $i \in [n]$, and: \begin{compactitem}
-\item $s_0 \in S_0$ and $\gamma_0(x)=0$ for all $x \in C$, 
-%
-\item for all $i \in [n]$, there exists a transition
-  $(s_i,a_i,s_{i+1},\lambda_i,\delta_i) \in E$ such that $\gamma_i +
-  \tau_{i+1} - \tau_i \models \delta_i$, and for all $x \in C$,
-  $\gamma_{i+1}(x) = 0$ if $x \in \lambda_i$ and $\gamma_{i+1}(x) = \gamma_i(x)
-  + \tau_{i+1}-\tau_i$, otherwise. Here $\tau_0 \isdef 0$ and $\gamma_i +
-  \tau_{i+1} - \tau_i$ is the valuation mapping each $x \in C$ to
-  $\gamma_i(x) + \tau_{i+1} - \tau_i$.
-\end{compactitem}
-The run is accepting iff $s_n \in F$, in which case $T$ accepts
-$w$. As usual, we denote by $\lang{T}$ the set of finite words
-accepted by $T$. It is well-known that, in general, there is no timed
-automaton accepting the complement language $(\Sigma\times\real)^*
-\setminus \lang{T}$ and, moreover, the language inclusion problem is
-undecidable \cite{AlurDill94}.
-
-Given a timed automaton $T = \tuple{\Sigma,S,S_0,F,C,E}$, we define a
-first-order alternating automaton $\A_T =
-\tuple{\Sigma,\set{t},Q_T,\iota_T,F_T,\Delta_T}$, with a single input
-variable $t$, ranging over $\real$, such that each timed word $w =
-(a_1,\tau_1) \ldots (a_n,\tau_n)$ corresponds to a unique data word
-$d(w) = (a_1,\nu_1) \ldots (a_n,\nu_n)$ such that $\nu_i(t) = \tau_i$
-for all $i \in [1,n]$ and $\lang{\A_T} = \set{d(w) \mid
-  w\in\lang{T}}$. The only difficulty here is capturing the fact that
-all the clocks of $T$ evolve at the same pace, which is easily done
-using a technique from \cite{Fribourg98}, which replaces
-each clock $x_i$ of $T$ by a variable $y_i$ tracking the difference
-between the values of $t$ and $x_i$.
-
-Formally, if $C = \set{x_1,\ldots,x_k}$ and $S =
-\set{s_1,\ldots,s_m}$, we define $Q_T \isdef \set{q_1, \ldots, q_m}$ ,
-where $\#(q_i) = k+1$ for all $i \in [1,m]$, $\iota_T \isdef
-\bigvee_{s_i \in S_0} q_i(0,\ldots,0)$, $F_T \isdef \set{q_i \mid s_i
-  \in F}$ and, for each transition $(s_i,a,s_j,\lambda,\delta) \in E$,
-$\Delta_T$ contains the rule:
-\[q_i(y_1,\ldots,y_k,z) \arrow{a(t)}{} t > z \wedge \delta(z-y_1,\ldots,z-y_k) \wedge q_j(y'_1,\ldots,y'_k,t)\]
-where $y'_i$ stands for $z$ if $x_i \in \lambda$ and for $y_i$, 
-otherwise.  Moreover, nothing else is in $\Delta_T$. We establish the
-following connection between a timed automaton and its corresponding
-first-order alternating automaton.
-
-\begin{proposition}\label{prop:timed-alternating}
-  Given a timed automaton $T = \tuple{\Sigma,S,S_0,F,C,E}$, the
-  first-order alternating automaton $\A_T =
-  \tuple{\Sigma,\set{t},Q_T,\iota_T,F_T,\Delta_T}$ recognizes the
-  language $\lang{\A_T} = \set{d(w) \mid w \in \lang{T}}$. 
-\end{proposition}
-\proof{ ``$\subseteq$'' Let $w = (a_1,\nu_1) \ldots (a_n,\nu_n) \in
-  \lang{\A_T}$ be a data word. We show the existence of a timed word
-  $(a_1,\tau_1) \ldots (a_n,\tau_n) \in \lang{T}$ such that
-  $\nu_i(t)=\tau_i$, for all $i \in [1,n]$, by induction on
-  $n\geq0$. In fact we shall prove the following stronger
-  statements: \begin{compactenum}
-  \item each execution of $\A_T$ over $w$ starting with a cube $c \in
-    \cube{\minsem{\iota_T}}$ is a linear tree, in which each node has
-    at most one child.
-    %
-  \item for each execution $q_{i_0}(d^0_1,\ldots,d^0_k,\tau_0) \ldots
-    q_{i_n}(d^n_1,\ldots,d^n_k,\tau_n)$ of $\A_T$, $T$ has an
-    execution $(s_{i_0}, \gamma_0) \ldots (s_{i_n}, \gamma_n)$ over
-    the timed word $(a_1, \tau_1) \ldots (a_n, \tau_n)$, such that,
-    for all $i \in [1,n]$ and all $\ell \in [1,k]$, we have
-    $\gamma_i(x_\ell) = \tau_{i-1} - d^i_\ell$.
-  \end{compactenum}
-  The first point above is by inspection of $\iota_T = \bigvee_{s_i
-    \in S_0} q_i(0,\ldots,0)$ and of the rules from
-  $\Delta_T$. Indeed, each minimal model of $\iota_T$ corresponds to a
-  cube $q(0,\ldots,0)$ and each rule has exactly one predicate atom on
-  its right-hand side, thus each node of the execution will have at
-  most one successor. The second point is by induction on $n\geq0$.
-
-  ``$\supseteq$'' Let $w = (a_1,\tau_1) \ldots (a_n,\tau_n) \in
-  \lang{T}$ be a timed word. By induction on $n\geq0$, we show that
-  for each run $(s_{i_0}, \gamma_0) \ldots (s_{i_n}, \gamma_n)$ of $T$
-  over $w$, $\A_T$ has a linear execution
-  $q_{i_0}(d^0_1,\ldots,d^0_k,\tau_0) \ldots
-  q_{i_n}(d^n_1,\ldots,d^n_k,\tau_n)$ such that, for all $i \in [1,n]$
-  and all $\ell \in [1,k]$, we have $\gamma_i(x_\ell) = \tau_{i-1} -
-  d^i_\ell$. \qed}
-
-An easy consequence is that the timed language inclusion problem
-``given timed automata $T_1$ and $T_2$, does $\lang{T_1} \subseteq
-\lang{T_2}$ ?'' is reduced in polynomial time to the emptiness problem
-$\lang{\A_{T_1}} \cap \lang{\overline{\A_{T_2}}} = \emptyset$, for
-which (\S\ref{sec:emptiness}) provides a semi-algorithm. Observe,
-moreover, that no transition quantifiers are needed to encode timed
-automata as first-order alternating automata.
-
-\subsection{Register Automata}
-\label{sec:ra}
-
-Finite-memory automata, most commonly referred to as register automata
-\cite{KaminskiFrancez94} are among the first attempts at lifting the
-finite alphabet restriction of classical Rabin-Scott automata. In a
-nutshell, a register automaton is a finite-state automaton equipped
-with a finite set of registers $x_1,\ldots,x_r$ able to copy input
-values and compare them with subsequent input. Consequently, basic
-results from classical automata theory, such as the pumping lemma or
-the closure under complement do not hold in this model and, moreover,
-inclusion of languages recognized by register automata is undecidable
-\cite{NevenSchwentickVianu03}.
-
-Let $\Sigma$ be an infinite alphabet, $\#$ be a symbol not in $\Sigma$
-and $r>0$ be an integer constant, denoting the number of registers. An
-\emph{assignment} is a word $\vec{v} = v_1 \ldots v_r$ such that if
-$v_i = v_j$ and $i \neq j$ then $v_i = \#$, for all $i,j \in
-     [1,r]$. We write $[\vec{v}]$ for the set $\set{v_i \mid i \in
-       [1,r]}$ of values in the assignment $\vec{v}$. A
-     \emph{finite-memory (register) automaton} is a tuple $R =
-     \tuple{S,q_0,\vec{u}, \rho, \mu, F}$, where $S$ is a finite set
-     of states, $q_0 \in S$ is the initial state, $\vec{u} = u_1
-     \ldots u_r$ is the initial assignment, $\rho : S \rightarrow
-            [1,r]$ is the reassignment partial function, $\mu
-            \subseteq S \times [1,r] \times S$ is the transition
-            relation and $F \subseteq S$ is the set of final states. A
-            run of $A$ over an input word $a_1 \ldots a_n \in
-            \Sigma^*$ is a sequence $(s_0,\vec{v}_0) \ldots
-            (s_n,\vec{v}_n)$ such that $\vec{v}_0 = \vec{u}$ and, for
-            all $i \in [1,n]$, exactly one of the following
-            holds: \begin{compactitem}
-\item if there exists $k\in[1,r]$ such that $a_i=(\vec{v}_{i-1})_k$
-  then $\vec{v}_i = \vec{v}_{i-1}$ and $(s_{i-1},k,s_i) \in \mu$, 
-%
-\item otherwise $a_i\not\in[\vec{v}_{i-1}]$, $\rho(s_{i-1})$ is
-  defined, $(\vec{v}_i)_{\rho(s_{i-1})} = a_i$, for each $k \in
-  [1,r] \setminus \set{\rho(s_{i-1})}$, we have $(\vec{v}_i)_k =
-  (\vec{v}_{i-1})_k$ and $(s_{i-1},\rho(s_{i-1}),s_i) \in \mu$.
-\end{compactitem}
-Intuitively, if the input symbol is already stored in some register,
-the automaton moves to the next state if, moreover, the transition
-relation allows it, otherwise it copies the input to the register
-indicated by the reassignment, erasing its the previous value, and
-moves according to the transition relation.
-
-The translation of register automata to first-order alternating
-automata is quite natural, because registers can be encoded as
-arguments of predicate atoms. Formally, given a register automaton $R
-= \tuple{S,s_0,\vec{u},\rho,\mu,F}$, such that $S =
-\set{s_0,\ldots,s_m}$, we define the alternating automaton $\A_R =
-\tuple{\set{\alpha},\set{x},Q_R,\iota_R,F_R,\Delta_R}$, where
-$\alpha \not\in \Sigma$, $Q_R \isdef \set{q_0,\ldots,q_m}$ and
-$\#(q_i) = r$ for all $i \in [m]$, $\iota_R \isdef q_0(\vec{u})$, $F_R
-\isdef \set{q_i \mid s_i \in F}$ and, for each transition $(s_i, k,
-s_j) \in \mu$, $\Delta_T$ contains the rule:
-\[q_i(y_1,\ldots,y_r) \arrow{\alpha(x)}{} y_k=x \wedge q_j(y_1,\ldots,y_r) ~\vee~
-\bigwedge_{i=1}^r x \neq y_i \wedge q_j(y_1,\ldots,y_{k-1},x,y_{k+1},
-\ldots, y_r)\] Moreover, nothing else is in $\Delta_R$. The connection
-between register automata and first-order alternating automata is
-stated below.
-
-\begin{proposition}\label{prop:register-alternating}
-  Given a register automaton $R = \tuple{S,s_0,\vec{u},\rho,\mu,F}$
-  over an infinite alphabet $\Sigma$, the first-order alternating
-  automaton $\A_R = \tuple{\set{\alpha},Q_R,\iota_R,F_R,\Delta_R}$
-  recognizes the languge $\lang{\A_R} = \set{ (\alpha,a_1) \ldots
-    (\alpha,a_n) \mid a_1 \ldots a_n \in \lang{R} }$. 
-\end{proposition}
-\proof{ ``$\subseteq$'' Let $w = (\alpha,a_1) \ldots (\alpha,a_n)
-  \in \lang{\A_R}$. First, it is easy to show that each execution of
-  $\A_R$, that starts in some cube $c \in \cube{\minsem{\iota_R}}$, is
-  a linear tree with labels $q_0(\vec{v}_0), \ldots, q_n(\vec{v}_0)$
-  such that $\vec{v}_0 = \vec{u}$. Second by induction on $n\geq0$, we
-  prove that $\A_R$ has a run as above over $w$ only if $R$ has a run
-  $(q_0,\vec{v}_0), \ldots, (q_n,\vec{v}_n)$ over $a_1 \ldots
-  a_n$. ``$\supseteq$'' Let $w = a_1 \ldots a_n \in \lang{R}$ and
-  $q_0(\vec{v}_0), \ldots, q_n(\vec{v}_0)$ be a run of $R$ over $w$,
-  such that $\vec{v}_0 = \vec{u}$. By induction on $n\geq0$, we can
-  build an execution of $\A_R$ over $(\alpha,a_1) \ldots
-  (\alpha,a_n)$ that is a linear tree with labels $q_0(\vec{v}_0),
-  \ldots, q_n(\vec{v}_n)$. \qed}
-
-Consequently, the language inclusion problem ``given register automata
-$R_1$ and $R_2$, does $\lang{R_1} \subseteq \lang{R_2}$?'' is reduced
-in polynomial time to emptiness problem $\lang{\A_{R_1}} \cap
-\lang{\overline{\A_{R_2}}} = \emptyset$, for which
-(\S\ref{sec:emptiness}) provides a semi-algorithm. Notice further that
-the encoding of register automata as first-order alternating automata
-uses no transition quantifiers.
-
-\subsection{Predicate Automata}
-\label{sec:pa}
-
-The model of \emph{predicate automata} \cite{Farzan15,Farzan16} has
-emerged recently as a tool for checking safety and liveness properties
-of parameterized concurrent programs, in which there is an unbounded
-number of replicated threads that communicate via global
-variables. Predicate automata recognize finite sequences of actions
-that are pairs $(\sigma,i)$, where $\sigma$ is from a finite set
-$\Sigma$ of program statements and $i \in \nat$ ranges over an
-unbounded set of thread identifiers. To avoid clutter, we shall view a
-pair $(\sigma,i)$ as a data symbol $(\sigma,\nu)$ where $\nu(x)=i$,
-for a designated input variable $x$.
-
-Since thread identifiers can only be compared for equality, the data
-theory of predicate automata is the first-order theory of
-equality. Moreover, transition quantifiers are only needed for
-checking termination and, generally, liveness properties
-\cite{Farzan16}.
-
-However, the execution semantics of predicate automata differs from
-that of first-order automata with respect to the following detail:
-initial configurations and successors of predicate automata are
-defined using the entire sets of models of the initial sentence and
-transition rules, not just the minimal ones, as in our case.
-
-Formally, a run of a predicate automaton $P =
-\tuple{\Sigma,\set{x},Q,\iota,F,\Delta}$ over a word $(a_1,\nu_1)
-\ldots$ $(a_n,\nu_n)$ is a sequence of interpretations $\I_0, \ldots,
-\I_n$ such that $\I_0 \in \sem{\iota}$ and for each $i \in [1,n]$,
-each $q \in Q$ and each tuple $\tuple{d_1,\ldots,d_{\#(q)}} \in
-\I_{i-1}(q)$, we have $\I_i \in \sem{\psi}_\mu$, for each rule
-$q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(x)}{} \psi \in \Delta$, where
-$\mu=\nu_i[y_1\leftarrow d_1, \ldots, y_{\#(q)} \leftarrow
-  d_{\#(q)}]$. The run is accepting if and only if $\I(q)=\emptyset$
-for all $q \in Q \setminus F$.
-
-In fact, as shown next, this more simple execution semantics is
-equivalent, from the language point of view, with the semantics given
-by Definitions \ref{def:execution} and \ref{def:accepting}.  We
-believe that the semantics of first-order alternating automata based
-on minimal models is important for its relation to the textbook
-semantics of boolean alternating automata
-\cite{ChandraKozenStockmeyer81}.
-
-\begin{proposition}
-  Given a predicate automaton $P =
-  \tuple{\Sigma,\set{x},Q,\iota,F,\Delta}$, let $\A_P$ be the
-  first-order alternating automaton that has the same description as
-  $P$. Then $\lang{P} = \lang{\A_P}$. 
-\end{proposition}
-\proof{
-\qed}
-As before, this result enables using the semi-algorithm from
-\S\ref{sec:emptiness} for checking emptiness of predicate automata.
-We point out that, although quantifier-free predicate automata with
-predicates of arity one are decidable for emptiness \cite{Farzan15},
-currently there is no method for checking emptiness of predicate
-automata with predicates of arity greater than one, other than the
-explicit enumeration of cubes. Moreover, no method for dealing with
-emptiness in the presence of transition quantifiers is known to exist.
+\input abstract
+\input body
 

### 2018-11-06 quantifiers/body.tex
ok

@@
 
-We describe applications of this semi-algorithmic method to several
-undecidable but recursively enumerable problems, for which no
-effective method exist, to date: inclusion between recognizable timed
+As applications of first order alternating automata, we identified
+several undecidable but recursively enumerable problems, for which no
+effective methods exist, to date: inclusion between recognizable timed
 languages \cite{AlurDill94} and between languages recognized by
@@
 \fi
+
+\section{Experimental Results}

### 2018-11-07 quantifiers/body.tex
ok

@@
 As applications of first order alternating automata, we identified
-several undecidable but recursively enumerable problems, for which no
-effective methods exist, to date: inclusion between recognizable timed
-languages \cite{AlurDill94} and between languages recognized by
-finite-memory automata \cite{KaminskiFrancez94}, as well as emptiness
-of predicate automata, a model of first-order alternating automata
-used to check safety and liveness properties of parameterized
-concurrent programs \cite{Farzan15,Farzan16}.
+several undecidable, for which no semi-algorithmic methods exist:
+inclusion between recognizable timed languages \cite{AlurDill94}, 
+between languages recognized by finite-memory automata
+\cite{KaminskiFrancez94} and emptiness of predicate automata,
+a model of first-order alternating automata used to check safety and
+liveness properties of parameterized concurrent programs
+\cite{Farzan15,Farzan16}.
+
+For reasons of space, all proofs of technical results in this paper
+are given in \cite{IosifXu18ArXiv}.
 

### 2018-11-15 quantifiers/body.tex
201811151930

@@
 \section{Experimental Results}
+
+We have implemented a version of the IMPACT semi-algorithm \cite{McMillan06} in a prototype tool called FOADA \footnote{The implementation is available on \url{https://github.com/cathiec/FOADA}.} that uses Z3 solver \footnote{\url{https://rise4fun.com/z3}} via the Java SMT interface \footnote{\url{https://github.com/sosy-lab/java-smt}} for the satisfiability queries and interpolant generation.
+We applied the tool to some predicate automata \cite{Farzan15} \footnote{\url{https://github.com/zkincaid/duet/tree/ark2/regression/predicateAutomata}} (*.pa), several array logic entailments (array\_rotation, array\_simple, array\_shift), two timed communication  (abp, train), hardware circuits safety proving (hw1, hw2) and a railroad crossing controller with hidden variables (rr-crossing).
+The results \footnote{The results were obtained on MacOS x64 - 1.3 GHz Intel Core i5 - 8 GB 1867 MHz LPDDR3.} are given in Table 1.
+
+\begin{table}[htb]
+\vspace*{-\baselineskip}
+\begin{center}
+{\fontsize{7}{8}\selectfont
+\begin{tabular}{||l|c|c|c|c|c||}
+\hline
+Example & $\len{\A}$ (bytes) & $L(\A)=\emptyset$ ? & Nodes Expanded & Nodes Visited & Times (ms) \\
+\hline
+incdec.pa & 499 & no & 13 & 7 & 273 \\
+\hline
+localdec.pa & 678 & no & 17 & 11 & 560 \\
+\hline
+ticket.pa & 4250 & no & 31 & 17 & 999 \\
+\hline
+count\_thread0.pa & 9767 & no & 100 & 76 & 6055 \\
+\hline
+count\_thread1.pa & 10925 & no & 766 & 692 & 74445 \\
+\hline
+array\_rotation.ada & 1834 & yes & 9 & 8 & 1543 \\
+\hline
+array\_simple.ada & 3440 & yes & 11 & 10 & 6787 \\
+\hline
+array\_shift.ada & 874 & yes & 6 & 5 & 413 \\
+\hline
+abp.ada & 6909 & no & 52 & 47 & 4788 \\
+\hline
+train.ada & 1823 & yes & 68 & 67 & 7319 \\
+\hline
+hw1.ada & 322 & Solver Error & / & / & / \\
+\hline
+hw2.ada & 674 & yes & 20 & 22 & 4974 \\
+\hline
+rr-crossing.foada & 1780 & yes & 67 & 67 & 7574 \\
+\hline
+\end{tabular}
+}
+\caption{}\label{tab:experiments}
+\end{center}
+\vspace*{-2\baselineskip}
+\end{table}

### 2018-11-16 quantifiers/body.tex
ok

--- 2018-11-07 quantifiers/body.tex
@@
 \section{Experimental Results}
+
+We have implemented a version of the IMPACT semi-algorithm
+\cite{McMillan06} in a prototype tool called FOADA, which is avaliable
+online \cite{foada}. The tool is written in Java and uses the Z3 SMT
+solver \cite{z3}, via the JavaSMT interface \cite{javasmt}, for
+spuriousness and coverage queries and also for interpolant generation.
+
+The experimental results, reported in Table \ref{tab:experiments},
+come from several sources, namely predicate automata models (*.pa)
+\cite{Farzan15,Farzan16} available online \cite{pa}, timed automata
+inclusion problems (abp.ada, train.ada, rr-crossing.foada), array
+logic entailments (array\_rotation.ada, array\_simple.ada,
+array\_shift.ada) and hardware circuit verification (hw1.ada,
+hw2.ada), initially considered in \cite{IosifRV16}. The main advantage
+of using FOADA over the INCLUDER \cite{includer} tool used in
+\cite{IosifRV16} is the possibility of checking inclusion between
+infinite alphabet automata with hidden (local) variables, whose values
+are not visible in the input. In particular, this is essential for
+checking inclusion of timed automata that use internal clocks to
+control the computation.
+
+\begin{table}[htb]
+\vspace*{-\baselineskip}
+\begin{center}
+{\fontsize{7}{8}\selectfont
+\begin{tabular}{||l|c|c|c|c|c||}
+\hline
+Example & $\len{\A}$ (bytes) & $L(\A)=\emptyset$ ? & Nodes Expanded & Nodes Visited & Times (ms) \\
+\hline
+incdec.pa & 499 & no & 13 & 7 & 273 \\
+\hline
+localdec.pa & 678 & no & 17 & 11 & 560 \\
+\hline
+ticket.pa & 4250 & no & 31 & 17 & 999 \\
+\hline
+count\_thread0.pa & 9767 & no & 100 & 76 & 6055 \\
+\hline
+count\_thread1.pa & 10925 & no & 766 & 692 & 74445 \\
+\hline
+array\_rotation.ada & 1834 & yes & 9 & 8 & 1543 \\
+\hline
+array\_simple.ada & 3440 & yes & 11 & 10 & 6787 \\
+\hline
+array\_shift.ada & 874 & yes & 6 & 5 & 413 \\
+\hline
+abp.ada & 6909 & no & 52 & 47 & 4788 \\
+\hline
+train.ada & 1823 & yes & 68 & 67 & 7319 \\
+\hline
+hw1.ada & 322 & Solver Error & / & / & / \\
+\hline
+hw2.ada & 674 & yes & 20 & 22 & 4974 \\
+\hline
+rr-crossing.foada & 1780 & yes & 67 & 67 & 7574 \\
+\hline
+\end{tabular}
+}
+\caption{Experiments with First Order Alternating Automata}
+\label{tab:experiments}
+\end{center}
+\vspace*{-2\baselineskip}
+\end{table}
+
+The experiments were carried out on a MacOS x64 - 1.3 GHz
+  Intel Core i5 - 8 GB 1867 MHz LPDDR3 machine.

### 2018-11-16 quantifiers/body.tex
ok

@@
 As applications of first order alternating automata, we identified
-several undecidable, for which no semi-algorithmic methods exist:
-inclusion between recognizable timed languages \cite{AlurDill94}, 
-between languages recognized by finite-memory automata
-\cite{KaminskiFrancez94} and emptiness of predicate automata,
-a model of first-order alternating automata used to check safety and
-liveness properties of parameterized concurrent programs
+several undecidable problems for which no semi-algorithmic methods
+exist: inclusion between recognizable timed languages
+\cite{AlurDill94}, languages recognized by finite-memory automata
+\cite{KaminskiFrancez94} and emptiness of predicate automata, a model
+of first-order alternating automata used to check safety and liveness
+properties of parameterized concurrent programs
 \cite{Farzan15,Farzan16}.

### 2018-11-16 quantifiers/body.tex
ok

@@
 \cite{AlurDill94}, languages recognized by finite-memory automata
-\cite{KaminskiFrancez94} and emptiness of predicate automata, a model
-of first-order alternating automata used to check safety and liveness
-properties of parameterized concurrent programs
+\cite{KaminskiFrancez94} and emptiness of predicate automata, a
+subclass of first-order alternating automata used to check safety and
+liveness properties of parameterized concurrent programs
 \cite{Farzan15,Farzan16}.

### 2018-11-16 quantifiers/body.tex
201811161615
201811161905
ok

@@
 
+\vspace*{-0.5\baselineskip}
 \section{Experimental Results}
+\vspace*{-0.5\baselineskip}
 
@@
 spuriousness and coverage queries and also for interpolant generation.
+The experiments were carried out on a MacOS x64 - 1.3 GHz Intel Core
+i5 - 8 GB 1867 MHz LPDDR3 machine.
 
@@
 \cite{Farzan15,Farzan16} available online \cite{pa}, timed automata
-inclusion problems (abp.ada, train.ada, rr-crossing.foada), array
-logic entailments (array\_rotation.ada, array\_simple.ada,
-array\_shift.ada) and hardware circuit verification (hw1.ada,
-hw2.ada), initially considered in \cite{IosifRV16}. The main advantage
-of using FOADA over the INCLUDER \cite{includer} tool used in
-\cite{IosifRV16} is the possibility of checking inclusion between
-infinite alphabet automata with hidden (local) variables, whose values
-are not visible in the input. In particular, this is essential for
-checking inclusion of timed automata that use internal clocks to
-control the computation.
+inclusion problems ({\tt abp.ada}, {\tt train.ada}, {\tt
+  rr-crossing.foada}), array logic entailments ({\tt
+  array\_rotation.ada}, {\tt array\_simple.ada}, {\tt
+  array\_shift.ada}) and hardware circuit verification ({\tt hw1.ada},
+{\tt hw2.ada}), initially considered in \cite{IosifRV16}. The {\tt
+  train-simpleN.} {\tt foada} and {\tt fischer-mutexN.foada} examples are
+parametric verification problems in which one checks inclusions of the
+form $\bigcap_{i=1}^N\lang{A_i} \subseteq \lang{B}$, where $A_i$ is
+the $i$-th copy of the same template automaton.
+
+The advantage of using FOADA over the INCLUDER \cite{includer} tool
+from \cite{IosifRV16} is the possibility of having infinite alphabet
+automata with hidden (local) variables, whose values are not visible
+in the input. In particular, this is essential for checking inclusion
+of timed automata that use internal clocks to control the computation.
 
@@
 \hline
-Example & $\len{\A}$ (bytes) & $L(\A)=\emptyset$ ? & Nodes Expanded & Nodes Visited & Times (ms) \\
+Example & $\len{\A}$ (bytes) & $L(\A)=\emptyset$ ? & Nodes Expanded & Nodes Visited & Time (ms) \\
 \hline
-incdec.pa & 499 & no & 13 & 7 & 273 \\
+incdec.pa & 499 & no & 21 & 17 & 779 \\
 \hline
-localdec.pa & 678 & no & 17 & 11 & 560 \\
+localdec.pa & 678 & no & 49 & 35 & 1814 \\
 \hline
-ticket.pa & 4250 & no & 31 & 17 & 999 \\
+ticket.pa & 4250 & no & 229 & 91 & 9543 \\
 \hline
-count\_thread0.pa & 9767 & no & 100 & 76 & 6055 \\
+count\_thread0.pa & 9767 & no & 154 & 128 & 8553 \\
 \hline
-count\_thread1.pa & 10925 & no & 766 & 692 & 74445 \\
+count\_thread1.pa & 10925 & no & 766 & 692 & 76771 \\
+\hline
+local0.pa & 10595 & no & 73 & 27 & 1431 \\
+\hline
+local1.pa & 11385 & no & 1135 & 858 & 101042 \\
 \hline
@@
 rr-crossing.foada & 1780 & yes & 67 & 67 & 7574 \\
+% \hline
+% running2.foada & 1217 & yes & 13 & 13 & 788 \\
+\hline
+train-simple1.foada & 5421 & yes & 43 & 44 & 2893 \\
+\hline
+train-simple2.foada & 10177 & yes & 111 & 113 & 8386 \\
+\hline
+train-simple3.foada & 15961 & yes & 196 & 200 & 15041 \\
+% \hline
+% fischer-mutex1.foada & 1566 & no & 31 & 31 & 1552 \\
+\hline
+fischer-mutex2.foada & 3000 & yes & 23 & 23 & 808 \\
+\hline
+fischer-mutex3.foada & 4452 & yes & 33 & 33 & 1154 \\
 \hline
@@
 \end{table}
-
-The experiments were carried out on a MacOS x64 - 1.3 GHz
-  Intel Core i5 - 8 GB 1867 MHz LPDDR3 machine.

### 2018-11-17 quantifiers/body.tex
ok

@@
 languages are defined over finite alphabets. In practice, this
-assumption is not realistic when attempting to model real-time
-hardware and communication protocols or even simple programs, whose
-input and observable output requires taking into account data values,
-ranging over very large domains, better viewed as infinite
-mathematical abstractions.
+assumption is problematic when attempting to use automata as models of
+real-time systems or even simple programs, whose input and observable
+output requires taking into account data values, ranging over very
+large domains, better viewed as infinite mathematical abstractions.
 
@@
 the \impact~procedure, originally developed for checking safety of
-nondeterministic integer programs \cite{McMillan06}. Essentially,
-checking emptiness of first-order alternating automata by lazy
-annotation with interpolants faces two problems: \begin{compactenum}
+nondeterministic integer programs \cite{McMillan06}. However, checking
+emptiness of first-order alternating automata by lazy annotation with
+interpolants faces two problems: \begin{compactenum}
 \item Quantified transition rules make it hard, or even impossible, to
@@
 automaton $B$ must be declared in the automata $A_1, \ldots, A_n$ that
-model the concurrent components of the system under check.
+model the concurrent components of the system under check. Using
+first-order alternating automata allows to bypass this limitation of
+our previous work.
 
@@
 this problem by considering boolean valuations (models) instead of
-formulae.
+formulae. However, describing first-order alternating automata using
+interpretations instead of formulae would be rather hard to follow.
 
@@
 formalize the execution semantics of automata, we relate sets of
-configurations to models of first order sentences, as follows. Given a
-formula $\phi$ and a valuation of its free variables $\nu$, each
-interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
-configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid q
-  \in Q,~ \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a
-\emph{cube}. For a set $\mathcal{S}$ of interpretations, we define
-$\cube{\mathcal{S}} \isdef \set{\cube{\I} \mid \I \in
-  \mathcal{S}}$.
+configurations to models of first order sentences, as follows. Each
+first-order interpretation $\I$ corresponds to a set of configurations
+$\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid q \in Q,~
+  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a \emph{cube}. For a
+set $\mathcal{S}$ of interpretations, we define $\cube{\mathcal{S}}
+\isdef \set{\cube{\I} \mid \I \in \mathcal{S}}$.
 
@@
 
-Given a sequence of input events $\alpha = a_1 \ldots a_n$, the
-\emph{path formula} of $\alpha$ is defined as:
+Given a sequence of input events $\alpha = a_1 \ldots a_n \in
+\Sigma^*$, the \emph{path formula} of $\alpha$ is:
 \begin{equation}\label{eq:pathform}
@@
 low-level configuration-based execution semantics and the symbolic
-path formulae, consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any
-execution forest $\T$ of $\A$ over $w$ is associated an interpretation
-$\I_{\T}$ of the set of time-stamped predicates $\stamp{Q}{\leq n}$,
-defined as:
+path formulae, consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in
+\Sigma[X]^*$. Any execution forest $\T$ of $\A$ over $w$ is associated
+an interpretation $\I_{\T}$ of the set of time-stamped predicates
+$\stamp{Q}{\leq n}$, defined as:
 \[\I_{\T}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
@@
 known decidable theories, supported by standard SMT solvers. There are
-basically two reasons for this, namely \begin{inparaenum}[(i)]
+basically two reasons for this, namely\begin{inparaenum}[(i)]
 \item the presence of predicate symbols, and
@@
 arithmetic, become undecidable if predicate atoms are added to
-them. Therefore, the result below makes a step towards deciding
-whether the automaton accepts a word with a given input sequence, by
-reducing this problem to the satisfiability of a quantified formula
-without predicates.
+them. Therefore, the result below makes a step forward towards
+deciding whether the automaton accepts a word with a given input
+sequence, by reducing this problem to the satisfiability of a
+quantified formula without predicates.
 
@@
 unfolding has been found. For a detailed presentation of the emptiness
-procedure, we refer the reader to \cite{IosifXu18}.
+procedure, we refer to \cite{IosifXu18}.
 
@@
 the unfolded control structure are assertions about the values of the
-program variables in a given control state, at a certain step of a
-given execution. However, in an alternating model of computation, it
-is useful to distinguish between
+program variables in a given control state, at a certain step of an
+execution. However, in an alternating model of computation, it is
+useful to distinguish between
 \begin{inparaenum}[(i)]
@@
 
-\begin{theorem}\label{thm:skolem}
+\begin{theorem}[\cite{BorgerGraedelGurevich97}]\label{thm:skolem}
   Given $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$ a first order sentence,
@@
      finite-range, we denote by $\xi_{\max}^{-1}(n) \in \nat$ the
-     maximal value $m$ such that $\xi(m)=n$. It is easy to see that
-     $\xi_{\max}^{-1}$ is strictly monotonic if $\xi$ is monotonic and
-     finite-range. The lemma below gives the proof of correctness for
-     the previous construction of non-local GLI.
+     maximal value $m$ such that $\xi(m)=n$. The lemma below gives the
+     proof of correctness for the construction of non-local GLI.
 
@@
   index from the sequence $\quantformn{\alpha_0}, \ldots,
-  \quantformn{\alpha_n}$ where it occurs. Then the following
-  hold: \begin{compactenum}
-  \item\label{it1:subst-quant} one can effectively build witness terms
-    $\tau_{i_1}, \ldots, \tau_{i_\ell}$, where \(\set{i_1,
-    \ldots,i_\ell} = \set{j \in [1,m] \mid Q_j = \forall}\) and
-    \(\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup
-    \set{x_k \mid k < i_j, Q_k = \exists},~ \forall j \in [1,\ell]\)
-    such that $\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
+  \quantformn{\alpha_n}$ where it occurs. Then one can effectively build:
+  \begin{compactenum}
+  \item\label{it1:subst-quant} witness terms $\tau_{i_1}, \ldots,
+    \tau_{i_\ell}$, where \(\set{i_1, \ldots,i_\ell} = \set{j \in
+      [1,m] \mid Q_j = \forall}\) and \(\voc{\tau_{i_j}} \subseteq
+    \stamp{X}{\leq \xi(i_j)} \cup \set{x_k \mid k < i_j, Q_k =
+      \exists},~ \forall j \in [1,\ell]\) such that
+    $\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
       \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable, and
     %
-  \item\label{it2:subst-quant} one can effectively build a GLI $(I_0,
-    \ldots, I_n)$ for $\alpha$, such that \(\voc{I_k} \subseteq
-    \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
-      \xi^{-1}(k),~ Q_j = \exists},~ \forall k \in [n]\).
+  \item\label{it2:subst-quant} a GLI $(I_0, \ldots, I_n)$ for
+    $\alpha$, such that \(\voc{I_k} \subseteq \stamp{Q}{k} \cup
+    \stamp{X}{\leq k} \cup \set{x_j \mid j < \xi^{-1}(k),~ Q_j =
+      \exists}\), for all $k \in [n]$. 
   \end{compactenum}
@@
 In conclusion, under two assumptions about the first order theory of
-the data domain, namely the \begin{inparaenum}[(i)]
+the data domain, namely the\begin{inparaenum}[(i)]
 \item witness-producing quantifier elimination, and

### 2019-02-03 quantifiers/body.tex
ok

@@
 
-Many results in formal language theory rely on the assumption that
-languages are defined over finite alphabets. In practice, this
-assumption is problematic when attempting to use automata as models of
-real-time systems or even simple programs, whose input and observable
-output requires taking into account data values, ranging over very
-large domains, better viewed as infinite mathematical abstractions.
-
-Alternating automata are a generalization of nondeterministic automata
-with universal transitions, that create several copies of the
-automaton, which synchronize on the same input word. Alternating
-automata are appealing for verification because they allow encoding of
-problems such as temporal logic model checking in linear time, as
-opposed to the exponential time required by nondeterministic automata
-\cite{Vardi95}. A finite-alphabet alternating automaton is typically
-described by a set of transition rules $q \arrow{a}{} \phi$, where $q$
-is a state, $a$ is an input symbol and $\phi$ is a positive boolean
-combinations of states, viewed as propositional variables.
-
-Here we introduce a generalized alternating automata model in which
-states are predicate symbols $q(y_1,\ldots,y_k)$, the input has
-associated data variables $x_1,\ldots,x_n$, ranging over an infinite
-domain and transitions are of the form $q(y_1,\ldots,y_k)
-\arrow{a(x_1,\ldots,x_n)}{} \phi$, where $\phi$ is any formula in the
-first-order theory of the data domain, in which each state predicate
-occurs under an even number of negations. In this model, the arguments
-of a predicate atom $q(y_1,\ldots,y_k)$ track the values of the
-\emph{internal variables} associated with the state. Together with the
-input values $x_1,\ldots,x_n$, these values are used to compute the
-successor states and are invisible in the input sequence.
-
-Previous attempts to generalize classical Rabin-Scott automata to
+Many results in automata theory rely on the finite alphabet
+hypothesis, which guarantees, in some cases, the existence of
+determinization, complementation and inclusion checking
+methods. However, this hypothesis prevents the use of automata as
+models of real-time systems or even simple programs, whose input and
+output are data values ranging over very large domains, typically
+viewed as infinite mathematical abstractions.
+
+Traditional attempts to generalize classical Rabin-Scott automata to
 infinite alphabets, such as timed automata \cite{AlurDill94} and
-finite-memory (register) automata \cite{KaminskiFrancez94} face the
+finite-memory automata \cite{KaminskiFrancez94} face the
 \emph{complement closure} problem: there exist automata for which the
@@
 
-The solution we adopt here is a tight coupling of internal variables
-to control states, using uninterpreted predicate symbols. As we show,
-this allows for linear-time complementation just as in the case of
-boolean alternating automata. Complementation is, moreover, possible
-when the transition formulae contain first-order quantifiers,
-generating infinitely-branching execution trees. The price to be paid
-for this expressivity is that emptiness of first-order alternating
-automata is undecidable, even for the simplest data theory of equality
+Even when assuming the alphabet to be finite, complementation of a
+finite-state automaton faces an inherent exponential blowup due to
+nondeterminism. However, if we allow universal nondeterminsim, in
+addition to the classical existential nondeterminism, complementation
+is possible is linear time. The automata having both existential and
+universal nondeterminism are called \emph{alternating automata}
+\cite{ChandraKozenStockmeyer81}. A finite-alphabet alternating
+automaton is described by a set of transition rules $q \arrow{a}{}
+\phi$, where $q$ is a state, $a$ is an input symbol and $\phi$ is a
+boolean formula, whose propositional variables denote successor
+states.
+
+\paragraph{\em Our Contribution}
+We extend the alternating automata model to infinite data alphabets,
+by defining a model of computation in which all boolean operations,
+including complementation, can be done in linear time.  The control
+states are given by $k$-ary predicate symbols $q(y_1,\ldots,y_k)$, the
+input consists of an event $a$ from a finite alphabet and a tuple of
+data variables $x_1,\ldots,x_n$, ranging over an infinite domain, and
+transitions are of the form $q(y_1,\ldots,y_k)
+\arrow{a(x_1,\ldots,x_n)}{} \phi(x_1, \ldots, x_n, y_1, \ldots, y_k)$,
+where $\phi$ is a formula in the first-order theory of the data
+domain. In this model, the arguments of a predicate atom
+$q(y_1,\ldots,y_k)$ represent the values of the \emph{internal
+  variables} associated with the state. Together with the input values
+$x_1,\ldots,x_n$, these values define the next configurations, but
+remain invisible in the input sequence.
+
+The tight coupling of internal values and control states, by means of
+uninterpreted predicate symbols, allows for linear-time
+complementation just as in the case of classical propositional
+alternating automata. Complementation is, moreover, possible when the
+transition formulae contain first-order quantifiers, generating
+infinitely-branching execution trees. The price to be paid for this
+expressivity is that emptiness of first-order alternating automata is
+undecidable, even for the simplest data theory of equality
 \cite{Farzan15}.
@@
 semi-algorithm for first-order alternating automata, in the spirit of
-the \impact~procedure, originally developed for checking safety of
-nondeterministic integer programs \cite{McMillan06}. However, checking
-emptiness of first-order alternating automata by lazy annotation with
-interpolants faces two problems: \begin{compactenum}
-\item Quantified transition rules make it hard, or even impossible, to
-  decide if a given symbolic trace is spurious. This is mainly because
-  adding uninterpreted predicate symbols to decidable first-order
-  theories, such as Presburger arithmetic, results in undecidability
+the \impact~lazy annotation procedure, originally developed for
+checking safety of nondeterministic integer programs
+\cite{McMillan06,McMillan14}. In a nutshell, a lazy annotation
+procedure unfolds an automaton $A$ trying to find an execution that
+recognizes a word $w \in \lang{A}$. If a path that reaches a final
+state does not correspond to a concrete run of the automaton, the
+positions on the path are labeled with interpolants from the proof of
+infeasibility, thus marking this path and all continuations as
+infeasible for future searches.
+
+Checking emptiness of first-order alternating automata by lazy
+annotation with interpolants faces two problems: \begin{compactenum}
+\item Quantified transition rules make it hard, if not impossible, to
+  decide if a path is infeasible. This is mainly because adding
+  uninterpreted predicate symbols to decidable first-order theories,
+  such as Presburger arithmetic, results in undecidability
   \cite{Halpern91}. To deal with this problem, we assume that the
@@
 %
-\item The interpolants that prove the spuriousness of a symbolic path
-  are not \emph{local}, as they may refer to input values encountered
-  in the past. However, the future executions are oblivious to
-  \emph{when} these values have been seen in the past and depend only
-  on the data constraints between the values. We use this fact to
+\item The interpolants that prove the infeasibility of a path are not
+  \emph{local}, as they may refer to input values encountered in the
+  past. However, the future executions are oblivious to \emph{when}
+  these values have been seen in the past and depend only on the
+  relation between the past and current values. We use this fact to
   define a labeling of nodes, visited by the lazy annotation
@@
 
-As applications of first order alternating automata, we identified
-several undecidable problems for which no semi-algorithmic methods
-exist: inclusion between recognizable timed languages
-\cite{AlurDill94}, languages recognized by finite-memory automata
+\paragraph{\em Applications}
+We use first order alternating automata to develop practical
+semi-algorithms for a number of known undecidable problems, such as:
+inclusion of regular timed languages \cite{AlurDill94}, inclusion of
+quasi-regular languages recognized by finite-memory automata
 \cite{KaminskiFrancez94} and emptiness of predicate automata, a
-subclass of first-order alternating automata used to check safety and
-liveness properties of parameterized concurrent programs
-\cite{Farzan15,Farzan16}.
-
-For reasons of space, all proofs of technical results in this paper
-are given in \cite{IosifXu18ArXiv}.
-
-\paragraph{\bf Related Work} 
+subclass of first-order alternating automata used to verify
+parameterized concurrent programs \cite{Farzan15,Farzan16}.
+
+Consider, for instance the problem of proving that $\lang{A} \subseteq
+\lang{B}$, where $A$ and $B$ are the timed automata in
+Fig. \ref{fig:ta}. The automaton $A$ ($B$) has a single clock $x$
+($y$) which is reset on the transition from $s_1$ to $s_2$ ($q_1$ to
+$q_2$). Intuititvely, $A$ recognizes the sequences in which all events
+are $a$ and the second event occurs after at least $2$ time units,
+whereas $B$ recognizes the sequences in which there exists an event
+$a$ such that no event occurs after it in exactly $1$ time
+unit. Formally, $\lang{A} = \set{a, t_1, \ldots a, t_n \mid t_1,
+  \ldots, t_n \in \real,~ t_1 \geq 2}$ and $\lang{B} = \set{\sigma_1,
+  t_1 \ldots \sigma_{i-1}, t_{i-1}, a, t_i, \ldots \sigma_n, t_n \mid
+  \sigma_1, \ldots, \sigma_n \in \set{a,b},~ t_1, \ldots, t_n \in
+  \real,~ t_i-t_{i-1} \neq 1}$. It is proved\footnote{See
+  \cite[Theorem 1]{AlurMadhusudan04} for a proof.} that $\clang{B}$ is
+not recognized by a timed automaton, thus the inclusion cannot be
+proved in the framework of timed regular languages, as $\lang{A} \cap
+\clang{B} = \emptyset$.
+
+We encode timed automata as first order alternating automata and uses
+closure of the latter class under intersection and complement to
+compute an alternating automaton whose emptiness proves the inclusion
+$\lang{A} \subseteq \lang{B}$ between timed regular languages. The
+idea is to represent each control state of $A$ and $B$ by a binary
+predicate symbol with the same name. Consider the encoding of $A$ in
+Fig. \ref{fig:ta} (right) inspired by a translation of timed automata
+into Datalog programs \cite{Fribourg98}. The first argument of
+$s_1(x,z)$ is used to track the value of the $x$ clock of $A$, whereas
+$z$ is the current time value. The time stamp of the input event is
+the input variable $t$, which is bigger or equal than the current time
+$z$, since the lapse of time between two consequtive events is always
+positive. The current value of the clock is $z-x$, thus the guard $x
+\geq 2$ on the $s_2 \arrow{}{} s_2$ transition of $A$ from becomes
+$z-x \geq 2$ in the second rule, and the reset of $x$ on the $s_1
+\arrow{}{} s_2$ transition is encoded by passing $z$ as the first
+parameter of $s_2$, in the first rule. Note that the time increment
+equals the increment of the clocks that are not reset.
+
+\begin{figure}[htb]
+\vspace*{-\baselineskip}
+\begin{center}
+\begin{tabular}{ccc}
+\begin{minipage}{3cm}
+\input ta.pdf_t
+\end{minipage} 
+&&
+\begin{minipage}{5cm}
+\small{\[A \left\{\begin{array}{rcl}
+s_1(x,z) & \arrow{a(t)}{} & t \geq z \wedge s_2(z,t) \\
+s_2(x,z) & \arrow{a(t)}{} & t \geq z \wedge z-x \geq 2 \wedge s_2(x,t)
+\end{array}\right.\] 
+\[B \left\{\begin{array}{rcl}
+q_1(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge q_1(y,t) \\
+q_1(y,u) & \arrow{a(t)}{} & t \geq u \wedge q_2(u,t) \\
+q_2(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge u - y \neq 1 \wedge q_2(y,t)
+\end{array}\right.\]}
+\end{minipage} \\
+&& 
+\begin{minipage}{5cm}
+{\small\[\overline{B}\left\{\begin{array}{rcl}
+q_1(y,u) & \arrow{a(t),b(t)}{} & t < u \vee q_1(y,t) \\
+q_1(y,u) & \arrow{a(t)}{} & t < u \vee q_2(u,t) \\
+q_2(y,u) & \arrow{a(t),b(t)}{} & t < u \vee u - y = 1 \vee q_2(y,t)
+\end{array}\right.\]}
+\end{minipage}
+\end{tabular}
+\end{center}
+\vspace*{-\baselineskip}
+\caption{First Order Alternating Automata for Solving Inclusion of Timed Automata}
+\label{fig:ta}
+\vspace*{-\baselineskip}
+\end{figure}
+
+The first order alternating automaton recognizing the (encoding of
+the) $\clang{B}$ language is simply obtained by negating the theory
+atoms $t \geq u$ and $u - y \neq 1$ and interchanging conjunctions
+with disjunctions in the rules of the alternating automaton encoding
+$B$, denoted as $\overline{B}$ in Fig. \ref{fig:ta} (bottom
+right). Finally, we check the emptiness of the alternating automaton
+consisting of the rules of $A$ and $\overline{B}$, started in the
+initial configuration $s_1(0,0) \wedge q_1(0,0)$. Our lazy annotation
+semi-algorithm finds that this automaton is empty in \comment[ri]{...}
+seconds.
+
+\paragraph{\em Related Work} 
 The first order alternating automata model presented in this paper

### 2019-02-03 quantifiers/body.tex
ok

@@
 
-Even when assuming the alphabet to be finite, complementation of a
-finite-state automaton faces an inherent exponential blowup due to
-nondeterminism. However, if we allow universal nondeterminsim, in
-addition to the classical existential nondeterminism, complementation
-is possible is linear time. The automata having both existential and
-universal nondeterminism are called \emph{alternating automata}
+Even for finite alphabets, complementation of finite-state automata
+faces inherent exponential blowup, due to nondeterminism. However, if
+we allow universal nondeterminism, in addition to the classical
+existential nondeterminism, complementation is possible is linear
+time. The automata having both existential and universal
+nondeterminism are called \emph{alternating automata}
 \cite{ChandraKozenStockmeyer81}. A finite-alphabet alternating

### 2019-02-03 quantifiers/body.tex
ok

@@
 existential nondeterminism, complementation is possible is linear
-time. The automata having both existential and universal
-nondeterminism are called \emph{alternating automata}
-\cite{ChandraKozenStockmeyer81}. A finite-alphabet alternating
-automaton is described by a set of transition rules $q \arrow{a}{}
-\phi$, where $q$ is a state, $a$ is an input symbol and $\phi$ is a
-boolean formula, whose propositional variables denote successor
-states.
+time. Having both existential and universal nondeterminism defines the
+\emph{alternating automata} model \cite{ChandraKozenStockmeyer81}. A
+finite-alphabet alternating automaton is described by a set of
+transition rules $q \arrow{a}{} \phi$, where $q$ is a state, $a$ is an
+input symbol and $\phi$ is a boolean formula, whose propositional
+variables denote successor states.
 

### 2019-02-03 quantifiers/body.tex
ok

@@
 \paragraph{\em Our Contribution}
-We extend the alternating automata model to infinite data alphabets,
-by defining a model of computation in which all boolean operations,
-including complementation, can be done in linear time.  The control
-states are given by $k$-ary predicate symbols $q(y_1,\ldots,y_k)$, the
-input consists of an event $a$ from a finite alphabet and a tuple of
-data variables $x_1,\ldots,x_n$, ranging over an infinite domain, and
+We extend alternating automata to infinite data alphabets, by defining
+a model of computation in which all boolean operations, including
+complementation, can be done in linear time.  The control states are
+given by $k$-ary predicate symbols $q(y_1,\ldots,y_k)$, the input
+consists of an event $a$ from a finite alphabet and a tuple of data
+variables $x_1,\ldots,x_n$, ranging over an infinite domain, and
 transitions are of the form $q(y_1,\ldots,y_k)
@@
 \begin{center}
-\begin{tabular}{ccc}
-\begin{minipage}{3cm}
+\begin{minipage}{5cm}
 \input ta.pdf_t
 \end{minipage} 
-&&
 \begin{minipage}{5cm}
-\small{\[A \left\{\begin{array}{rcl}
+\footnotesize{\[A \left\{\begin{array}{rcl}
 s_1(x,z) & \arrow{a(t)}{} & t \geq z \wedge s_2(z,t) \\
@@
 q_2(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge u - y \neq 1 \wedge q_2(y,t)
-\end{array}\right.\]}
-\end{minipage} \\
-&& 
-\begin{minipage}{5cm}
-{\small\[\overline{B}\left\{\begin{array}{rcl}
+\end{array}\right.\]
+\[\overline{B}\left\{\begin{array}{rcl}
 q_1(y,u) & \arrow{a(t),b(t)}{} & t < u \vee q_1(y,t) \\
@@
 \end{minipage}
-\end{tabular}
 \end{center}
@@
 The first order alternating automata model presented in this paper
-stems from our previous work on boolean alternating automata extended
-with variables ranging over infinite data \cite{IosifXu18}. There we
-considered states to be propositional variables, as in the classical
-textbook alternating automata model, and all variables of the
-automaton to be observable in the input. The model in this paper
-overcomes this latter restriction by allowing for internal variables,
-whose variables are not visible in the language.
-
-This solves an older language inclusion problem
-$\bigcap_{i=1}^n\lang{A_i} \subseteq \lang{B}$, between finite-state
-automata with data variables, whose languages are alternating
-sequences of input events and variable valuations
-\cite{IosifRV16}. There, we assumed that all variables of the observer
-automaton $B$ must be declared in the automata $A_1, \ldots, A_n$ that
-model the concurrent components of the system under check. Using
-first-order alternating automata allows to bypass this limitation of
-our previous work.
+extends the alternating automata with variables ranging over infinite
+data considered in \cite{IosifXu18}. There all variables were required
+to be observable in the input. We overcome this restriction by
+allowing internal (invisible) variables. Another closely related work
+\cite{IosifRV16} considers an inclusion between an asynchronous
+product of automata $A_1 \times \ldots \times A_n$, extended with data
+variables, and a monitor automaton $B$. The semi-algorithm defined
+here relies on the assumption that all variables of the observer $B$
+must be declared in the automata $A_1, \ldots, A_n$ under check. First
+order alternating automata bypass this limitation, since the emptiness
+checking semi-algorithm handles invisible variables.
 
 The work probably closest to the one reported here concerns the model
-of predicate automata (PA) \cite{Farzan15,Farzan16,KincaidPhD},
-applied to the verification of parameterized concurrent programs with
-shared memory. In this model, the alphabet consists of pairs of
-program statements and thread identifiers, thus being infinite because
-the number of threads is unbounded. Because thread identifiers can
-only be compared for (dis-)equality, the data theory in PA is the
-theory of equality. Even with this simplification, the emptiness
-problem is undecidable when either the predicates have arity greater
-than one \cite{Farzan15} or quantified transition rules
+of predicate automata (PA) \cite{Farzan15,Farzan16,KincaidPhD}, used
+in the verification of parameterized concurrent programs with shared
+memory. In this model, the alphabet consists of pairs of program
+statements and thread identifiers, thus being infinite because the
+number of threads is unbounded. Because thread identifiers can only be
+compared for (dis-)equality, the data theory in PA is the theory of
+equality. Even with this simplification, the emptiness problem is
+undecidable when either the predicates have arity greater than one
+\cite{Farzan15} or quantified transition rules
 \cite{KincaidPhD}. Checking emptiness of quantifier free PA is
@@
 argument values. However, no semi-algorithm is given for quantified
-PA.  Dealing with quantified transition rules is one of the
-contributions of the work reported in this paper.
+PA.  Dealing with quantified transition rules is one of our
+contributions.
 

### 2019-02-03 quantifiers/body.tex
ok

@@
 \vspace*{-\baselineskip}
-\caption{First Order Alternating Automata for Solving Inclusion of Timed Automata}
+\caption{First Order Alternating Automata Checking Inclusion of Timed Automata}
 \label{fig:ta}
@@
 
-The work probably closest to the one reported here concerns the model
-of predicate automata (PA) \cite{Farzan15,Farzan16,KincaidPhD}, used
-in the verification of parameterized concurrent programs with shared
+The work probably closest to ours concerns the model of predicate
+automata (PA) \cite{Farzan15,Farzan16,KincaidPhD}, used in the
+verification of parameterized concurrent programs with shared
 memory. In this model, the alphabet consists of pairs of program
@@
 
-For two integers $0 \leq i \leq j$, we denote by $[i,j]$ the set
-$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[0,i]$. We consider two
+For two integers $0 \leq i \leq j$, we define $[i,j] \isdef
+\set{i,\ldots,j}$ and $[i] \isdef [0,i]$. We consider two disjoint
 sorts $\Data$ and $\Bool$, where $\Data$ is an infinite domain and
@@
 and false ($\bot$), respectively. The $\Data$ sort is equipped with
-finitely many function symbols $f : \Data^{\#(f)} \rightarrow \Data$,
-where $\#(f)\geq0$ denotes the number of arguments (arity) of
-$f$. When $\#(f)=0$, we say that $f$ is a constant. A \emph{predicate}
-is a function symbol $p : \Data^{\#(p)} \rightarrow \Bool$, denoting a
-relation of arity $\#(p)$ and we write $\preds$ for the set of
-predicates.
-
-In the following, we shall consider that the interpretation of all
-function symbols $f : \Data^{\#(f)} \rightarrow \Data$ that are not
-predicates is fixed by the interpretation of the $\Data$ sort,
-e.g.\ if $\Data$ is the set of integers $\zed$, the function symbols
-are zero, the successor function and the arithmetic operations of
-addition and multiplication. For simplicity, we further blur the
-notational distinction between function symbols and their
-interpretations.
-
-Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
+countably many function symbols $f : \Data^{\#(f)} \rightarrow \Data
+\cup \Bool$, where $\#(f)\geq0$ denotes the number of arguments
+(arity) of $f$. A \emph{predicate} is a function symbol $p :
+\Data^{\#(p)} \rightarrow \Bool$ that is, a $\#(p)$-ary relation.
+
+We consider the interpretation of all function symbols $f :
+\Data^{\#(f)} \rightarrow \Data$ to be fixed by the interpretation of
+the $\Data$ sort, for instance if $\Data$ is the set of integers
+$\zed$, these are zero, the successor function and the arithmetic
+operations of addition and multiplication. We extend this convention
+to several predicates over $\Data$, such as the inequality relation
+over $\zed$, and write $\preds$ for the set of \emph{uninterpreted
+  predicates}.
+
+Let $\vars = \set{x,y,z,\ldots}$ be a countably infinite set of
 variables, ranging over $\Data$. Terms are either constants of sort
@@
 formulae is defined by the syntax below:
-\[\phi := t \teq s \mid p(t_1,\ldots,t_{\#(p)}) 
+\[\phi := t = s \mid p(t_1,\ldots,t_{\#(p)}) 
 \mid \neg \phi_1 \mid \phi_1 \wedge \phi_2 \mid \exists x ~.~
-\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms. We write
-$\phi_1 \vee \phi_2$, $\phi_1 \rightarrow \phi_2$ and $\forall x ~.~
-\phi_1$ for $\neg(\neg\phi_1 \wedge \neg\phi_2)$, $\neg\phi_1 \vee
-\phi_2$ and $\neg\exists x ~.~ \neg\phi_1$, respectively. We denote by
-$\fv{}{\phi}$ the set of free variables in $\phi$. The size
-$\len{\phi}$ of a formula $\phi$ is the number of symbols needed to
-write it down.
-
-A \emph{sentence} is a formula $\phi$ in which each variable occurs
-under the scope of a quantifier, i.e. $\fv{}{\phi} = \emptyset$.  A
-formula is \emph{positive} if each predicate symbol occurs under an
-even number of negations and we denote by $\posforms(Q,X)$ the set of
-positive formulae with predicates from the set $Q \subseteq \preds$
-and free variables from the set $X \subseteq \vars$.
-
-A formula is in \emph{prenex form} if it is of the form $\varphi =
-Q_1x_1 \ldots Q_nx_n ~.~ \phi$, where $\phi$ has no quantifiers. In
-this case we call $\phi$ the \emph{matrix} of $\varphi$. Every
-first order formula can be written in prenex form, by renaming each
-quantified variable to a unique name and moving the quantifiers
-upfront.
+\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms and $p$ is a
+predicate symbol. We write $\phi_1 \vee \phi_2$, $\phi_1 \rightarrow
+\phi_2$ and $\forall x ~.~ \phi_1$ for $\neg(\neg\phi_1 \wedge
+\neg\phi_2)$, $\neg\phi_1 \vee \phi_2$ and $\neg\exists x ~.~
+\neg\phi_1$, respectively. $\fv{}{\phi}$ is the set of free variables
+in $\phi$ and the size $\len{\phi}$ of a formula $\phi$ is the number
+of symbols needed to write it down. A \emph{sentence} is a formula
+$\phi$ with no free variables. A formula is \emph{positive} if each
+uninterpreted predicate symbol occurs under an even number of
+negations and we denote by $\posforms(Q,X)$ the set of positive
+formulae with predicates from the set $Q \subseteq \preds$ and free
+variables from the set $X \subseteq \vars$. A formula is in
+\emph{prenex form} if it is of the form $\varphi = Q_1x_1 \ldots
+Q_nx_n ~.~ \phi$, where $\phi$ has no quantifiers. In this case we
+call $\phi$ the \emph{matrix} of $\varphi$. Every first order formula
+can be written in prenex form, by renaming each quantified variable to
+a unique name and moving the quantifiers upfront.
 
@@
 
-An \emph{interpretation} $\I$ maps each predicate $p$ into a set $p^\I
-\subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element of $\Data$
-if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable $x$ into an
-element of $\Data$. Given a term $t$, we denote by $t^\nu$ the value
-obtained by replacing each variable $x$ by the value $\nu(x)$ and
-evaluating each function application. For a formula $\phi$, we define
-the forcing relation $\I,\nu \models \phi$ recursively on the
-structure of $\phi$, as usual. 
+An \emph{interpretation} $\I$ maps each predicate symbol $p$ into a
+set $p^\I \subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element
+of $\Bool$ if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable
+$x$ into an element of $\Data$. Given a term $t$, we denote by $t^\nu$
+the value obtained by replacing each variable $x$ by the value
+$\nu(x)$ and evaluating each function application. For a formula
+$\phi$, we define the forcing relation $\I,\nu \models \phi$
+recursively on the structure of $\phi$, as usual.  
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \ifLongVersion
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \[
@@
 and behaves like $\nu$ elsewhere.  
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \fi 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 For a formula $\phi$ and a valuation $\nu$, we define $\sem{\phi}_\nu
 \isdef \set{\I \mid \I,\nu \models \phi}$ and drop the $\nu$ subscript
-for sentences. A sentence $\phi$ is \emph{satisfiable}
-(\emph{unsatisfiable}) if $\sem{\phi}\neq\emptyset$
-($\sem{\phi}=\emptyset$). An element of $\sem{\phi}$ is called a
+for sentences. A sentence $\phi$ is \emph{satisfiable} if
+$\sem{\phi}\neq\emptyset$. An element of $\sem{\phi}$ is called a
 \emph{model} of $\phi$. A formula $\phi$ is \emph{valid} if $\I,\nu
 \models \phi$ for every interpretation $\I$ and every valuation
-$\nu$. For two formulae $\phi$ and $\psi$ we write $\phi \models \psi$
-for $\sem{\phi} \subseteq \sem{\psi}$, in which case we say that
-$\phi$ \emph{entails} $\psi$.
+$\nu$. We say that $\phi$ \emph{entails} $\psi$, written $\phi \models
+\psi$ if and only if $\sem{\phi} \subseteq \sem{\psi}$.
 
@@
 defined as $\I_1 \subseteq \I_2$ if and only if $p^{\I_1} \subseteq
-p^{\I_2}$ for each predicate $p \in \preds$. Given a set $\mathcal{S}$
-of interpretations, a minimal element $\I \in \mathcal{S}$ is an
-interpretation such that for no other interpretation $\I' \in
-\mathcal{S} \setminus \set{\I}$ do we have $\I' \subseteq \I$. For a
-formula $\phi$ and a valuation $\nu$, we denote by $\minsem{\phi}_\nu$
-and $\minsem{\phi}$ the set of minimal interpretations from
-$\sem{\phi}_\nu$ and $\sem{\phi}$, respectively.
+p^{\I_2}$ for each predicate symbol $p \in \preds$. Given a formula
+$\phi$ and a valuation $\nu$, we define $\minsem{\phi}_\nu \isdef
+\set{\I \mid \I,\nu \models \phi,~ \forall \I' \subseteq \I ~.~
+  \I',\nu \not\models \phi}$ the set of minimal interpretations that,
+together with $\nu$, form models of $\phi$.
 
@@
 $\data{w}$ the valuation associating each time-stamped variable
-$\stamp{x}{i}$ the value $\nu_i(x)$, for all $x \in \vars$ and
+$\stamp{x}{i}$, where $x \in \vars$, the value $\nu_i(x)$, for all
 $i\in[1,n]$. We denote by $\varepsilon$ the empty sequence, by
-$\Sigma^*$ the set of finite sequences of input events and by
-$\Sigma[X]^*$ the set of data words over the variables $X$.
-
-Formally, a \emph{first order alternating automaton} is a tuple $\A =
+$\Sigma^*$ the set of finite input sequences and by $\Sigma[X]^*$ the
+set of finite data words over the variables $X$.
+
+A \emph{first order alternating automaton} is a tuple $\A =
 \tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is a finite set of
@@
 referred to as \emph{transition quantifiers}. The \emph{size} of $\A$
-is defined as $\len{\A} = \len{\iota} +
-\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
-\len{\psi}$.
+is $\len{\A} \isdef \len{\iota} + \sum \set{\len{\psi} \mid q(\vec{y})
+  \arrow{a(X)}{} \psi \in \Delta}$.
 
@@
 The execution semantics of automata is given in close analogy with the
-case of boolean alternating automata, with transition rules of the
-form $q \arrow{a}{} \phi$, where $q$ is a boolean constant and $\phi$
-a positive boolean combination of such constants. For instance, $q_0
-\arrow{a}{} q_1 \wedge q_2 \vee q_3$ means that the automaton can
-choose to transition in either both $q_1$ and $q_2$ or in $q_3$
-alone. This intuition leads to saying that the steps of the automaton
-are defined by the minimal boolean models of the transition
-formulae. In this case, both $\set{q_1 \leftarrow \top, q_2 \leftarrow
-  \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow \bot, q_2
-  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, however
-$\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
-is a model but is not minimal. The original definition of alternating
-finite-state automata \cite{ChandraKozenStockmeyer81} works around
-this problem by considering boolean valuations (models) instead of
-formulae. However, describing first-order alternating automata using
-interpretations instead of formulae would be rather hard to follow.
-
-Given a predicate $q \in Q$ and a tuple of data values
-$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
-a \emph{configuration}\footnote{Note that a configuration is not a
-  logical term since data values cannot be written in logic.}. To
-formalize the execution semantics of automata, we relate sets of
-configurations to models of first order sentences, as follows. Each
-first-order interpretation $\I$ corresponds to a set of configurations
-$\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid q \in Q,~
-  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a \emph{cube}. For a
-set $\mathcal{S}$ of interpretations, we define $\cube{\mathcal{S}}
-\isdef \set{\cube{\I} \mid \I \in \mathcal{S}}$.
+classical definition of propositional alternating automata, with
+transition rules of the form $q \arrow{a}{} \phi$, where $q$ is a
+propositional variable and $\phi$ a positive boolean combination of
+propositional variables. For instance, $q_0 \arrow{a}{} q_1 \wedge q_2
+\vee q_3$ means that the automaton can choose to transition in either
+both $q_1$ and $q_2$ or in $q_3$ alone. This intuition leads to saying
+that a transition is a \emph{minimal model} of a right hand side
+formula of a rule. In our example, both $\set{q_1 \leftarrow \top, q_2
+  \leftarrow \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow
+  \bot, q_2 \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models,
+however $\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow
+  \top}$, is a model but is not minimal. The original definition of
+alternating finite-state automata \cite{ChandraKozenStockmeyer81}
+works around this problem by considering boolean valuations (models)
+instead of formulae. However, a finite description of a first order
+alternating automaton must use formulae, because there can be
+infinitely many satisfying interpretations that are incomparable
+w.r.t. the pointwise subset ordering.
+
+Given an uninterpreted predicate symbol $q \in Q$ and a tuple of data
+values $d_1,\ldots,d_{\#(q)}$, the tuple $(q, d_1,\ldots,d_{\#(q)})$
+is called a \emph{configuration}, sometimes written
+$q(d_1,\ldots,d_{\#(q)})$, when no confusion arises. A configuration
+is \emph{final} if $q \in F$. An interpretation $\I$ corresponds to a
+set of configurations $\cube{\I} \isdef \set{(q, d_1,\ldots,d_{\#(q)})
+  \mid q \in Q,~ (d_1,\ldots,d_{\#(q)}) \in q^\I}$, called a
+\emph{cube}. This notation is lifted to sets of configurations in the
+usual way.
 
 \begin{definition}\label{def:execution}
-Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$ and a cube $c$, an
-\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$, starting with $c$, is
-a (possibly infinite) forest $\T = \set{T_1,T_2,\ldots}$, where each
-$T_i$ is a tree labeled with configurations, such that:
+Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$ and a
+cube $c$, an \emph{execution} of
+$\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$, starting with $c$, is
+a forest $\T = \set{T_1,T_2,\ldots}$, where each $T_i$ is a tree
+labeled with configurations, such that:
 \begin{compactenum}
@@
 %
-\item\label{it2:execution} if $q(d_1,\ldots,d_{\#(q)})$ labels a node on the level $j \in
-  [n-1]$ in $T_i$, then the labels of its children form a cube from
-  $\cube{\minsem{\psi}_{\eta}}$, where $\eta = \nu_{j+1}[y_1
-    \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$ and
-  \(q(y_1,\ldots,y_{\#(q)}) \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is
-  a transition rule of $\A$.
+\item\label{it2:execution} if $(q,d_1,\ldots,d_{\#(q)})$ labels a node
+  on the level $j \in [n-1]$ in $T_i$, then the labels of its children
+  form a cube from $\cube{\minsem{\psi}_{\eta}}$, where $\eta =
+  \nu_{j+1}[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$
+  and \(q(y_1,\ldots,y_{\#(q)}) \arrow{a_{j+1}(X)}{} \psi \in \Delta\)
+  is a transition rule of $\A$.
 \end{compactenum}
 \end{definition}
-
-\begin{definition}\label{def:accepting}
 An execution $\T$ over $w$, starting with $c$, is \emph{accepting} if
-and only if \begin{compactitem}
-\item\label{it1:accepting} all paths in $\T$ have the same length $n$,
-  and
-  %
-\item\label{it2:accepting} the frontier of each tree $T \in \T$ is
-  labeled with \emph{final configurations} $q(d_1,\ldots,d_{\#(q)})$,
-  where $q \in F$.
-\end{compactitem} 
-If $\A$ has an accepting execution over $w$ starting with a cube $c
-\in \cube{\minsem{\iota}}$, then $\A$ \emph{accepts} $w$ and let
-$\lang{\A}$ be the set of words accepted by $\A$.
-\end{definition}
-In this paper, we address the following questions: \begin{compactenum}
+and only if all paths in $\T$ have the same length and the frontier of
+each tree $T \in \T$ is labeled with final configurations.  If $\A$
+has an accepting execution over $w$ starting with a cube $c \in
+\cube{\minsem{\iota}}$, then $\A$ \emph{accepts} $w$ and let
+$\lang{\A}$ be the set of words accepted by $\A$. We consider the
+following problems: \begin{compactenum}
 \item \emph{boolean closure}: given automata $\A_i =
@@
 \item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
-  \emptyset$?
+  \emptyset$ ?
 \end{compactenum}
-
-\subsection{Symbolic Execution}
+For technical reasons, we address the following \emph{relative
+  acceptance} problem first: given an automaton $\A$ and an input
+sequence $\alpha \in \Sigma^*$, does there exists a word $w \in
+\lang{\A}$ such that $\event{w} = \alpha$ ?  By solving this problem
+first, we develop the machinery required to prove that first order
+alternating automata are closed under all boolean operations and,
+further, set up the ground for a semi-algorithm for the emptiness
+problem.
+
+\subsection{Path Formulae}
 
@@
 $\accform{\alpha}$ is satisfiable if and only if the automaton accepts
-a word $w$ such that $\event{w} = \alpha$.
+a word $w$ such that $\event{w} = \alpha$. We reduce thus the relative
+acceptance problem to the satisfiability of an effectively computable
+formula in the first order theory of data with uninterpreted predicate
+symbols.
 
@@
 = \set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} =
-\set{\stamp{x}{i} \mid x \in X}$ the sets of time-stamped predicates
-and variables, respectively. As a shorthand, we write $\stamp{Q}{\leq
-  n}$ (resp.  $\stamp{X}{\leq n}$) for the set $\set{\stamp{q}{i} \mid
-  q \in Q, i \in [n]}$ (resp. $\set{\stamp{x}{i} \mid x \in X, i \in
-  [n]}$). For a formula $\psi$ and $i \in \nat$, we define
-$\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the
-formula in which all input variables and state predicates (and only
-those symbols) are replaced by their time-stamped counterparts. As a
-shorthand, we shall write $q(\vec{y})$ for $q(y_1,\ldots,y_{\#(q)})$,
-when no confusion arises.
+\set{\stamp{x}{i} \mid x \in X}$ the sets of time-stamped predicate
+symbols and variables, respectively. As a shorthand, we write
+$\stamp{Q}{\leq n}$ (resp.  $\stamp{X}{\leq n}$) for the set
+$\set{\stamp{q}{i} \mid q \in Q, i \in [n]}$ (resp. $\set{\stamp{x}{i}
+  \mid x \in X, i \in [n]}$). For a formula $\psi$ and $i \in \nat$,
+we define $\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$
+the formula in which all input variables and state predicates (and
+only those symbols) are replaced by their time-stamped
+counterparts. As a shorthand, we shall write $q(\vec{y})$ for
+$q(y_1,\ldots,y_{\#(q)})$, when no confusion arises.
 
@@
 \begin{equation}\label{eq:pathform}
+\begin{array}{c}
 \pathform{\alpha} \isdef \stamp{\iota}{0} \wedge 
@@
 \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i}
+\end{array}
 \end{equation}
@@
 be clear from the context. To formalize the relation between the
-low-level configuration-based execution semantics and the symbolic
-path formulae, consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in
-\Sigma[X]^*$. Any execution forest $\T$ of $\A$ over $w$ is associated
-an interpretation $\I_{\T}$ of the set of time-stamped predicates
-$\stamp{Q}{\leq n}$, defined as:
-\[\I_{\T}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid
-  q(d_1,\ldots,d_{\#(q)}) \text{ labels a node on level $i$ in $\T$}} 
+low-level configuration-based execution semantics and path formulae,
+consider a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in
+\Sigma[X]^*$. Any execution $\T$ of $\A$ over $w$ has an associated
+interpretation $\I_{\T}$ of time-stamped predicates $\stamp{Q}{\leq
+  n}$:
+\[\I_{\T}(\stamp{q}{i}) \isdef \set{(d_1,\ldots,d_{\#(q)}) \mid
+  (q,d_1,\ldots,d_{\#(q)}) \text{ labels a node on level $i$ in $\T$}}
 ,~\forall q\in Q ~\forall i \in [n]\]
@@
 \begin{equation}\label{eq:accform}
+\begin{array}{c}
 \accform{\alpha} \isdef \pathform{\alpha} \wedge \bigwedge_{q
@@
 \stamp{q}{n}(\vec{y}) \rightarrow \bot
+\end{array}
 \end{equation}
@@
 transition rule $q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi$ of $\A$.
-
-The acceptance formula $\accform{\A}$ is false in every interpretation
-of the predicates that assigns a non-empty set to a non-final
-predicate occurring on the last level in the execution forest. The
-relation between the words accepted by $\A$ and the acceptance formula
-above, is formally captured by the following lemma:
+The relation between the words accepted by $\A$ and the acceptance
+formula above, is formally captured by the following lemma:
 
@@
 formulae have a particular form, which allows the elimination of path
-quantifiers and of predicates, by a couple of
+quantifiers and uninterpreted predicate symbols, by a couple of
 satisfiability-preserving transformations. The result of applying
@@
 only quantifiers are those introduced by the transition rules of the
-automaton, referred to as transition quantifiers. We shall further
-assume (\S\ref{sec:emptiness}) that the first order theory of the data
-sort $\Data$ has quantifier elimination, which allows to effectively
-decide the satisfiability of such formulae.
+automaton. Next, in \S\ref{sec:emptiness} we assume that the first
+order theory of the data sort $\Data$ has quantifier elimination,
+which allows to effectively decide satisfiability of acceptance
+formulae.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\ifLongVersion
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{definition}\label{def:quantform}
@@
   \stamp{\iota}{0}$ and, for all $i \in [1,n]$:
-  \vspace*{-.5\baselineskip}
   \[\quantformn{\alpha_i} \isdef \quantformn{\alpha_{i-1}} \wedge
@@
       \]  
-      \vspace*{-.5\baselineskip}
       We write
@@
 \end{definition}
-Observe that $\quantform{\alpha}$ contains no path quantifiers, as
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\else
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Let $\quantformn{\alpha_0}, \ldots, \quantformn{\alpha_n}$ be the
+sequence of formulae defined by $\quantformn{\alpha_0} \isdef
+\stamp{\iota}{0}$ and, for all $i \in [1,n]$, let
+$\quantformn{\alpha_i}$ be the conjunction of
+$\quantformn{\alpha_{i-1}}$ with all formulae
+$\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow
+\stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$, such that
+$\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)})$ occurs in
+$\quantformn{\alpha_{i-1}}$. Next, we write $\quantform{\alpha}$ for
+the conjunction of $\quantformn{\alpha_n}$ with all
+$\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot$, such that
+$\stamp{q}{n}(t_1,\ldots,t_{\#(q)})$ occurs in
+$\quantformn{\alpha_n}$, for some $q \in Q \setminus F$.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\fi
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Note that $\quantform{\alpha}$ contains no path quantifiers, as
 required. On the other hand, the scope of the transition quantifiers
@@
 \end{example}
-
-The next lemma establishes a formal relation between the
-satisfiability of an acceptance formula $\accform{\alpha}$ and that of
-the formula $\quantform{\alpha}$, obtained by eliminating the path
-quantifiers from $\accform{\alpha}$.
 
@@
 
-We proceed with the elimination of predicate atoms from
-$\quantform{\alpha}$, defined below.
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\ifLongVersion
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{definition}\label{def:substform}
@@
 \end{definition}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\else
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Further, we eliminate the predicate atoms from $\quantform{\alpha}$,
+by considering the sequence of formulae $\substformn{\alpha_0} \isdef
+\stamp{\iota}{0}$ and, for all $i \in [1,n]$, $\substformn{\alpha_i}$
+is obtained by substituting each predicate atom
+$\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)})$ in $\substformn{\alpha_{i-1}}$
+by $\stamp{\psi}{i}[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$, where
+$q(\vec{y}) \arrow{a_{i}(X)}{} \psi \in \Delta$. We write
+$\substform{\alpha}$ for the formula obtained by replacing, in
+$\substformn{\alpha}$, each occurrence of a predicate $\stamp{q}{n}$,
+such that $q \in Q \setminus F$ (resp. $q \in F$), by $\bot$
+(resp. $\top$).
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\fi
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{example}[Contd. from Example \ref{ex:quant-elim}]\label{ex:pred-elim}
@@
 Finally, we define the acceptance of a word with a given input event
-sequence by means of a formula in which no predicate atom occurs. As
-previously discussed, several decidable theories, such as Presburger
-arithmetic, become undecidable if predicate atoms are added to
-them. Therefore, the result below makes a step forward towards
-deciding whether the automaton accepts a word with a given input
-sequence, by reducing this problem to the satisfiability of a
-quantified formula without predicates.
+sequence by means of a quantifier free formula in which no predicate
+atom occurs.
 
@@
 \[\begin{array}{rclcrclcrcl}
-\dual{(\phi_1 \vee \phi_2)} & = & \dual{\phi_1} \wedge \dual{\phi_2} && 
-\dual{(\phi_1 \wedge \phi_2)} & = & \dual{\phi_1} \vee \dual{\phi_2} &&
-\dual{(t \teq s)} & = & \neg(t \teq s) \\
-\dual{(\exists x ~.~ \phi_1)} & = & \forall x ~.~ \dual{\phi_1} && 
-\dual{(\forall x ~.~ \phi_1)} & = & \exists x ~.~ \dual{\phi_1} && 
-\dual{(\neg(t \teq s))} & = & t \teq s \\
-\dual{(q(x_1,\ldots,x_{\#(q)}))} & = & q(x_1,\ldots,x_{\#(q)})
+\dual{(\phi_1 \vee \phi_2)} & \isdef & \dual{\phi_1} \wedge \dual{\phi_2} && 
+\dual{(\phi_1 \wedge \phi_2)} & \isdef & \dual{\phi_1} \vee \dual{\phi_2} &&
+\dual{(t = s)} & \isdef & t \neq s \\
+\dual{(\exists x ~.~ \phi_1)} & \isdef & \forall x ~.~ \dual{\phi_1} && 
+\dual{(\forall x ~.~ \phi_1)} & \isdef & \exists x ~.~ \dual{\phi_1} && 
+\dual{(t \neq s)} & \isdef & t = s \\
+&&&& \dual{q(x_1,\ldots,x_{\#(q)})} & \isdef & q(x_1,\ldots,x_{\#(q)})
 \end{array}\]
-Observe that, because predicate atoms do not occur negated in $\phi$,
-there is no need to define dualization for formulae of the form $\neg
-q(x_1,\ldots,x_{\#(q)})$. The following theorem shows closure of
-automata under all boolean operations:
+The following theorem shows closure of automata under all boolean
+operations:
 
@@
 \end{theorem}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\ifLongVersion\else
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\paragraph{\em Proof sketch.} (\ref{it1:thm:closure}) The construction of $\A_\cap$ 
+and $\A_\cup$ resembles the standard intersection and union of
+propositional alternating automata
+\cite{Arxiv}. (\ref{it2:thm:closure}) Let $w \in \Sigma[X]^*$ be a
+word. We denote by $\accformA{\A_1}{\event{w}}$ and
+$\substformA{\A_1}{\event{w}}$
+[resp. $\accformA{\overline{\A}_1}{\event{w}}$ and
+  $\substformA{\overline{\A}_1}{\event{w}}$] the formulae
+$\accform{\event{w}}$ and $\substform{\event{w}}$ for $\A_1$ and
+$\overline{\A}_1$, respectively. It is enough to show that
+$\substformA{\overline{\A}_1}{\event{w}} =
+\neg\substformA{\A_1}{\event{w}}$ and apply Lemma
+\ref{lemma:quant-pred-acceptance} to prove that $w \in \lang{\A_1}
+\iff w \not\in \lang{\overline{\A}_1}$. Since the choice of $w$ was
+arbitrary, this proves $\lang{\overline{\A}_1} = \Sigma[X]^* \setminus
+\lang{\A_1}$. By induction on the number of predicate atoms in
+$\accformA{\A_1}{\event{w}}$ that are replaced during the generation
+of $\substformA{\A_1}{\event{w}}$. The proof relies on the following
+fact, proved by induction on the structure of $\phi$:
+  \begin{fact}\label{fact:dual-neg}
+    Let $\phi$ be a positive formula and let $q(t_1,\ldots,t_{\#(q)})$
+    be the only occurrence of a predicate symbol within $\phi$. Then,
+    we have the equivalence:
+    \[\neg\phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]
+    \equiv
+    \dual{\phi}[\neg\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]\]
+  \end{fact}
+\qed
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\fi
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \proof{ (\ref{it1:thm:closure}) ``$\subseteq$'' Let $w \in

### 2019-02-04 quantifiers/body.tex
ok

@@
 complement language cannot be recognized by an automaton in the same
-class. This excludes the possibility of encoding a language inclusion
+class. This makes it impossible to encode a language inclusion
 problem $\lang{A} \subseteq \lang{B}$ as the emptiness of an automaton
@@
 procedure unfolds an automaton $A$ trying to find an execution that
-recognizes a word $w \in \lang{A}$. If a path that reaches a final
+recognizes a word from $\lang{A}$. If a path that reaches a final
 state does not correspond to a concrete run of the automaton, the
@@
 infeasibility, thus marking this path and all continuations as
-infeasible for future searches.
-
-Checking emptiness of first-order alternating automata by lazy
-annotation with interpolants faces two problems: \begin{compactenum}
+infeasible for future searches. Termination of lazy annotation
+procedures is not guaranteed, but having a suitable coverage relation
+between the nodes of the search tree may ensure convergence of some
+real-life examples. Applying lazy annotation to first order
+alternating automata faces two nontrivial
+problems: \begin{compactenum}
 \item Quantified transition rules make it hard, if not impossible, to
@@
 
-We encode timed automata as first order alternating automata and uses
-closure of the latter class under intersection and complement to
-compute an alternating automaton whose emptiness proves the inclusion
-$\lang{A} \subseteq \lang{B}$ between timed regular languages. The
-idea is to represent each control state of $A$ and $B$ by a binary
-predicate symbol with the same name. Consider the encoding of $A$ in
+Our approach is to encode timed automata as first order alternating
+automata and use closure of the latter class under intersection and
+complement to compute an alternating automaton, whose emptiness proves
+that $\lang{A} \subseteq \lang{B}$. Consider the encoding of $A$ in
 Fig. \ref{fig:ta} (right) inspired by a translation of timed automata
-into Datalog programs \cite{Fribourg98}. The first argument of
-$s_1(x,z)$ is used to track the value of the $x$ clock of $A$, whereas
-$z$ is the current time value. The time stamp of the input event is
-the input variable $t$, which is bigger or equal than the current time
-$z$, since the lapse of time between two consequtive events is always
-positive. The current value of the clock is $z-x$, thus the guard $x
+into Datalog programs \cite{Fribourg98}.  Each control state of $A$
+and $B$ is represented by a binary predicate symbol with the same
+name. The first argument of $s_1(x,z)$ tracks the last moment in time
+when the $\mathsf{x}$ clock of $A$ was reset and $z$ is the current
+time value. The time stamp of the input event $a$ is the input
+variable $t$, which must be bigger or equal to $z$, as the lapse of
+time between two consequtive events is always positive. The current
+value of the $\mathsf{x}$ clock is $z-x$, thus the guard $\mathsf{x}
 \geq 2$ on the $s_2 \arrow{}{} s_2$ transition of $A$ from becomes
@@
 \arrow{}{} s_2$ transition is encoded by passing $z$ as the first
-parameter of $s_2$, in the first rule. Note that the time increment
-equals the increment of the clocks that are not reset.
+parameter of $s_2$, in the first rule.
 
@@
 \cite{Farzan15} or quantified transition rules
-\cite{KincaidPhD}. Checking emptiness of quantifier free PA is
+\cite{KincaidPhD}. Checking emptiness of quantifier-free PA is
 possible semi-algorithmically, by explicitly enumerating reachable
@@
 \posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive formula,
-where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$. The quantifiers
-occurring in the right-hand side formula of a transition rule are
-referred to as \emph{transition quantifiers}. The \emph{size} of $\A$
-is $\len{\A} \isdef \len{\iota} + \sum \set{\len{\psi} \mid q(\vec{y})
+where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$.  Without loss
+of generality, we consider, for each predicate $q \in Q$ and each
+input event $a \in \Sigma$, at most one such rule, as two or more
+rules can be joined using disjunction. The quantifiers occurring in
+the right-hand side formula of a transition rule are referred to as
+\emph{transition quantifiers}. The \emph{size} of $\A$ is $\len{\A}
+\isdef \len{\iota} + \sum \set{\len{\psi} \mid q(\vec{y})
   \arrow{a(X)}{} \psi \in \Delta}$.
 
-The intuition of a transition rule \(q(y_1,\ldots,y_{\#(q)})
-\arrow{a(X)}{} \psi\) is the following: $a$ is the input event and $X$
-are the input data values that trigger the transition, whereas $q$ and
-$y_1,\ldots,y_{\#(q)}$ are the current control state and data values
-in that state, respectively. Without loss of generality, we consider,
-for each predicate $q \in Q$ and each input event $a \in \Sigma$, at
-most one such rule, as two or more rules can be joined using
-disjunction.
-
+Given an uninterpreted predicate symbol $q \in Q$ and a tuple of data
+values $d_1,\ldots,d_{\#(q)}$, the tuple $(q, d_1,\ldots,d_{\#(q)})$
+is called a \emph{configuration}, sometimes written
+$q(d_1,\ldots,d_{\#(q)})$, when no confusion arises. A configuration
+is \emph{final} if $q \in F$. An interpretation $\I$ corresponds to a
+set of configurations $\cube{\I} \isdef \set{(q, d_1,\ldots,d_{\#(q)})
+  \mid q \in Q,~ (d_1,\ldots,d_{\#(q)}) \in q^\I}$, called a
+\emph{cube}. This notation is lifted to sets of configurations in the
+usual way.
+
+\begin{definition}\label{def:execution}
+Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$ and a
+cube $c$, an \emph{execution} of
+$\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$, starting with $c$, is
+a forest $\T = \set{T_1,T_2,\ldots}$, where each $T_i$ is a tree
+labeled with configurations, such that:
+\begin{compactenum}
+\item\label{it1:execution} $c = \set{T(\epsilon) \mid T \in \T}$ is the set of
+  configurations labeling the roots of $T_1,T_2,\ldots$ and
+%
+\item\label{it2:execution} if $(q,d_1,\ldots,d_{\#(q)})$ labels a node
+  on the level $j \in [n-1]$ in $T_i$, then the labels of its children
+  form a cube from $\cube{\minsem{\psi}_{\eta}}$, where $\eta =
+  \nu_{j+1}[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$
+  and \(q(y_1,\ldots,y_{\#(q)}) \arrow{a_{j+1}(X)}{} \psi \in \Delta\)
+  is a transition rule of $\A$.
+\end{compactenum}
+\end{definition}
 The execution semantics of automata is given in close analogy with the
@@
 works around this problem by considering boolean valuations (models)
-instead of formulae. However, a finite description of a first order
-alternating automaton must use formulae, because there can be
-infinitely many satisfying interpretations that are incomparable
-w.r.t. the pointwise subset ordering.
-
-Given an uninterpreted predicate symbol $q \in Q$ and a tuple of data
-values $d_1,\ldots,d_{\#(q)}$, the tuple $(q, d_1,\ldots,d_{\#(q)})$
-is called a \emph{configuration}, sometimes written
-$q(d_1,\ldots,d_{\#(q)})$, when no confusion arises. A configuration
-is \emph{final} if $q \in F$. An interpretation $\I$ corresponds to a
-set of configurations $\cube{\I} \isdef \set{(q, d_1,\ldots,d_{\#(q)})
-  \mid q \in Q,~ (d_1,\ldots,d_{\#(q)}) \in q^\I}$, called a
-\emph{cube}. This notation is lifted to sets of configurations in the
-usual way.
-
-\begin{definition}\label{def:execution}
-Given a word $w=(a_1,\nu_1) \ldots (a_n,\nu_n) \in \Sigma[X]^*$ and a
-cube $c$, an \emph{execution} of
-$\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$, starting with $c$, is
-a forest $\T = \set{T_1,T_2,\ldots}$, where each $T_i$ is a tree
-labeled with configurations, such that:
-\begin{compactenum}
-\item\label{it1:execution} $c = \set{T(\epsilon) \mid T \in \T}$ is the set of
-  configurations labeling the roots of $T_1,T_2,\ldots$ and
-%
-\item\label{it2:execution} if $(q,d_1,\ldots,d_{\#(q)})$ labels a node
-  on the level $j \in [n-1]$ in $T_i$, then the labels of its children
-  form a cube from $\cube{\minsem{\psi}_{\eta}}$, where $\eta =
-  \nu_{j+1}[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$
-  and \(q(y_1,\ldots,y_{\#(q)}) \arrow{a_{j+1}(X)}{} \psi \in \Delta\)
-  is a transition rule of $\A$.
-\end{compactenum}
-\end{definition}
+instead of formulae. In contrast, a finite description of a first
+order alternating automaton must be symbolic (in terms of formulae)
+because here, a given configuration can have infinitely many
+successors.
+
 An execution $\T$ over $w$, starting with $c$, is \emph{accepting} if
@@
 \end{compactenum}
-For technical reasons, we address the following \emph{relative
-  acceptance} problem first: given an automaton $\A$ and an input
-sequence $\alpha \in \Sigma^*$, does there exists a word $w \in
-\lang{\A}$ such that $\event{w} = \alpha$ ?  By solving this problem
-first, we develop the machinery required to prove that first order
-alternating automata are closed under all boolean operations and,
-further, set up the ground for a semi-algorithm for the emptiness
-problem.
+For technical reasons, we address the following problem next: given an
+automaton $\A$ and an input sequence $\alpha \in \Sigma^*$, does there
+exists a word $w \in \lang{\A}$ such that $\event{w} = \alpha$ ?  By
+solving this problem first, we develop the machinery required to prove
+that first order alternating automata are closed under all boolean
+operations and, further, set up the ground for a semi-algorithm for
+the emptiness problem.
 
 \subsection{Path Formulae}
+\label{sec:path-formulae}
 
@@
 $\accform{\alpha}$, which defines only accepting executions over words
-that share a given input sequence. Otherwise stated,
-$\accform{\alpha}$ is satisfiable if and only if the automaton accepts
-a word $w$ such that $\event{w} = \alpha$. We reduce thus the relative
-acceptance problem to the satisfiability of an effectively computable
-formula in the first order theory of data with uninterpreted predicate
-symbols.
+that share a given input sequence, i.e.\ $\accform{\alpha}$ is
+satisfiable if and only if the automaton accepts a word $w$ such that
+$\event{w} = \alpha$.
 
@@
 \set{\stamp{x}{i} \mid x \in X}$ the sets of time-stamped predicate
-symbols and variables, respectively. As a shorthand, we write
-$\stamp{Q}{\leq n}$ (resp.  $\stamp{X}{\leq n}$) for the set
-$\set{\stamp{q}{i} \mid q \in Q, i \in [n]}$ (resp. $\set{\stamp{x}{i}
-  \mid x \in X, i \in [n]}$). For a formula $\psi$ and $i \in \nat$,
-we define $\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$
-the formula in which all input variables and state predicates (and
-only those symbols) are replaced by their time-stamped
-counterparts. As a shorthand, we shall write $q(\vec{y})$ for
-$q(y_1,\ldots,y_{\#(q)})$, when no confusion arises.
+symbols and variables, respectively. We also define $\stamp{Q}{\leq n}
+\isdef \set{\stamp{q}{i} \mid q \in Q, i \in [n]}$ and $\stamp{X}{\leq
+  n} \isdef \set{\stamp{x}{i} \mid x \in X, i \in [n]}$. For a formula
+$\psi$ and $i \in \nat$, we define $\stamp{\psi}{i} \isdef
+\psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the formula in which all input
+variables and state predicates (and only those symbols) are replaced
+by their time-stamped counterparts. Moreover, we write $q(\vec{y})$
+for $q(y_1,\ldots,y_{\#(q)})$, when no confusion arises.
 
@@
 known decidable theories, supported by standard SMT solvers. There are
-basically two reasons for this, namely\begin{inparaenum}[(i)]
+basically two reasons for this, namely\ \begin{inparaenum}[(i)]
 \item the presence of predicate symbols, and
@@
 For the time being, let us formally define the elimination of
-transition quantifiers and predicates, respectively. Consider a given
-sequence of input events $\alpha = a_1 \ldots a_n$ and denote by
-$\alpha_i$ the prefix $a_1 \ldots a_i$ of $\alpha$, for $i \in [n]$,
+transition quantifiers and predicates, respectively. Let $\alpha = a_1
+\ldots a_n$ be a given sequence of input events and let
+$\alpha_i$ be the prefix $a_1 \ldots a_i$ of $\alpha$, for $i \in [n]$,
 where $\alpha_0=\epsilon$.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Let $\quantformn{\alpha_0}, \ldots, \quantformn{\alpha_n}$ be the
-sequence of formulae defined by $\quantformn{\alpha_0} \isdef
+We consider the sequence of formulae $\quantformn{\alpha_0}, \ldots,
+\quantformn{\alpha_n}$ defined as $\quantformn{\alpha_0} \isdef
 \stamp{\iota}{0}$ and, for all $i \in [1,n]$, let
@@
   \end{fact}
-  \proof{Assume w.l.o.g. that $\phi$ is quantifier free. The proof can
+  \proof{Assume w.l.o.g. that $\phi$ is quantifier-free. The proof can
     be easily generalized to the case $\phi$ has quantifiers. Let
@@
 Finally, we define the acceptance of a word with a given input event
-sequence by means of a quantifier free formula in which no predicate
+sequence by means of a quantifier-free formula in which no predicate
 atom occurs.
@@
 The following theorem shows closure of automata under all boolean
-operations:
+operations. Note that it is sufficient to show closure under
+intersection and negation because $\lang{\A_1} \cup \lang{\A_2} =
+\Sigma[X]^* (\clang{\A_1} \cap \clang{\A_2})$, for any two automata
+$\A_1$ and $\A_2$ with input event alphabet $\Sigma$ and input
+variables $X$.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph{\em Proof sketch.} (\ref{it1:thm:closure}) The construction of $\A_\cap$ 
-and $\A_\cup$ resembles the standard intersection and union of
-propositional alternating automata
-\cite{Arxiv}. (\ref{it2:thm:closure}) Let $w \in \Sigma[X]^*$ be a
-word. We denote by $\accformA{\A_1}{\event{w}}$ and
+\paragraph{\em Proof sketch.} (\ref{it1:thm:closure})
+The proof for $\A_\cap$ resembles the standard proof for the
+intersection of propositional alternating automata
+\cite{IosifXu18Arxiv} (\ref{it2:thm:closure}) Let $w \in \Sigma[X]^*$
+be a word. We denote by $\accformA{\A_1}{\event{w}}$ and
 $\substformA{\A_1}{\event{w}}$
@@
 arbitrary, this proves $\lang{\overline{\A}_1} = \Sigma[X]^* \setminus
-\lang{\A_1}$. By induction on the number of predicate atoms in
-$\accformA{\A_1}{\event{w}}$ that are replaced during the generation
-of $\substformA{\A_1}{\event{w}}$. The proof relies on the following
-fact, proved by induction on the structure of $\phi$:
+\lang{\A_1}$. The proof is by induction on the number of predicate
+atoms in $\accformA{\A_1}{\event{w}}$ that are replaced during the
+generation of $\substformA{\A_1}{\event{w}}$ and relies on the
+following fact, itself proved by induction on the structure of $\phi$:
   \begin{fact}\label{fact:dual-neg}
@@
   \end{fact}
-\qed
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-The problem of checking emptiness of a given automaton is undecidable,
-even for automata with predicates of arity two, whose transition rules
-use only equalities and disequalities, having no transition
-quantifiers \cite{Farzan15}. Since even such simple classes of
-alternating automata have no general decision procedure for emptiness,
-we use an abstraction-refinement semi-algorithm based on \emph{lazy
-  annotation} \cite{McMillan06,McMillan14}.
+The emptiness problem is undecidable even for automata with predicates
+of arity two, whose transition rules use only equalities and
+disequalities, having no transition quantifiers \cite{Farzan15}. Since
+even such simple classes of alternating automata have no general
+decision procedure for emptiness, we use an abstraction-refinement
+semi-algorithm based on \emph{lazy annotation}
+\cite{McMillan06,McMillan14}.
 
 In a nutshell, a lazy annotation procedure systematically explores the
-set of execution paths (in our case, sequences of input events) in
-search of an accepting execution. Each path has a corresponding path
-formula that defines all words accepted along that path. If the path
-formula is satisfiable, the automaton accepts a word. Otherwise, the
-path is said to be \emph{spurious}. When a spurious path is
-encountered, the search backtracks and the path is annotated with a
-set of learned facts, that marks this path as infeasible. The
-semi-algorithm uses moreover a coverage relation between paths,
-ensuring that the continuations of already covered paths are never
-explored. Sometimes this coverage relation provides a sound
-termination argument, when the automaton is empty.
+set of finite input event sequences searching for an accepting
+execution. For an input sequence, if the path formula is satisfiable,
+we compute a word in the language of the automaton, from the model of
+the path formula. Otherwise, i.e.\ the sequence is \emph{spurious},
+the search backtracks and each position in the sequence is annotated
+with a tuple of interpolants, that marks the entire sequence as
+infeasible. The semi-algorithm uses moreover a coverage relation
+between sequences, ensuring that the continuations of already covered
+sequences are never explored. Sometimes this coverage relation
+provides a sound termination argument, in case when the automaton is
+empty.
 
@@
 of alternating automata, that uses only predicates or arity zero
-(booleans) and no transition quantifiers. For simplicity, we do not
-present the details of this algorithm and shall content ourselves of
-several high-level definitions.
-
-Given a finite input event alphabet $\Sigma$, for two sequences
-$\alpha, \beta \in \Sigma^*$, we say that $\alpha$ is a prefix of
-$\beta$, written $\alpha \prefix \beta$, if $\alpha=\beta\gamma$ for
-some sequence $\gamma\in\Sigma^*$. A set $S$ of sequences
-is: \begin{compactitem}
-\item \emph{prefix-closed} if for each $\alpha \in S$, if $\beta \prefix
-\alpha$ then $\beta \in S$, and
-\item \emph{complete} if for each $\alpha \in S$, there exists $a \in
-  \Sigma$ such that $\alpha a \in S$ if and only if $\alpha b \in S$
-  for all $b \in \Sigma$. 
-\end{compactitem}
-Observe that a prefix-closed set is the backbone of a tree whose edges
-are labeled with input events. If the set is, moreover, complete, then
-every node of the tree has either zero successors, in which case it is
-called a \emph{leaf}, or it has a successor edge labeled with $a$ for
-each input event $a \in \Sigma$.
+(propositional variables) and no transition quantifiers. For
+simplicity, we do not present the details of the semi-algorithm and
+shall content ourselves of several high-level definitions. The
+interested reader is referred to \cite{IosifXu18} for a detailed
+presentation.
+
+For two input event sequences $\alpha, \beta \in \Sigma^*$, we say
+that $\alpha$ is a prefix of $\beta$, written $\alpha \prefix \beta$,
+if $\alpha=\beta\gamma$ for some sequence $\gamma\in\Sigma^*$. A set
+$S$ of sequences is \emph{prefix-closed} if for each $\alpha \in S$,
+if $\beta \prefix \alpha$ then $\beta \in S$, and \emph{complete} if
+for each $\alpha \in S$, there exists $a \in \Sigma$ such that $\alpha
+a \in S$ if and only if $\alpha b \in S$ for all $b \in \Sigma$. A
+prefix-closed set is the backbone of a tree whose edges are labeled
+with input events. If the set is, moreover, complete, then every node
+of the tree has either zero successors, in which case it is called a
+\emph{leaf}, or it has a successor edge labeled with $a$ for each
+input event $a \in \Sigma$.
 
@@
   \Sigma^* \rightharpoonup_{\mathit{fin}} \posforms(Q,\emptyset)$,
-  such that: \begin{compactenum}
-  \item\label{it1:unfolding} $\dom(U)$ is a finite prefix-closed
-    complete set,
-    %
-  \item\label{it2:unfolding} $U(\epsilon) = \iota$, and
-    %
-  \item\label{it3:unfolding} for each sequence $\alpha a \in \dom(U)$,
-    such that $\alpha \in \Sigma^*$ and $a \in \Sigma$:
-    \[\stamp{U(\alpha)}{0} \wedge
+  whose domain $\dom(U)$ is a finite prefix-closed complete set, such
+  that $U(\epsilon) = \iota$, and for each sequence $\alpha a \in
+  \dom(U)$, such that $\alpha \in \Sigma^*$ and $a \in \Sigma$:
+  \[\begin{array}{c}
+  \stamp{U(\alpha)}{0} \wedge
   \bigwedge_{q(\vec{y}) \arrow{a(X)}{} \psi} \forall y_1 \ldots
   \forall y_{\#q} ~.~ \stamp{q}{0}(\vec{y}) \rightarrow
-    \stamp{\psi}{1} \models \stamp{U(\alpha a)}{1}\]
-  \end{compactenum}
+  \stamp{\psi}{1} \models \stamp{U(\alpha a)}{1}
+  \end{array}\]  
   Moreover, $U$ is \emph{safe} if for each $\alpha \in \dom(U)$, the
@@
 As mentioned above, we check emptiness of first order alternating
-automata using the same method previously used to check emptiness of a
-simpler model of alternating automata, which uses boolean constants
-for control states and whose transition rules have no quantifiers
-\cite{IosifXu18}. The higher complexity of the automata model
-considered here, manifests itself within the interpolant generation
-procedure, used to refine the labeling of the unfolding. We discuss
-generation of interpolants in the next section.
+automata using a semi-algorithm previously used to check emptiness of
+a simpler model of alternating automata, which uses propositional
+variables for control states and whose transition rules have no
+quantifiers \cite{IosifXu18}. The higher complexity of the automata
+model considered here, manifests itself within the interpolant
+generation procedure, used to refine the labeling of the unfolding. We
+discuss generation of interpolants in the next section.
 
@@
 program variables in a given control state, at a certain step of an
-execution. However, in an alternating model of computation, it is
-useful to distinguish between
+execution. Because we consider alternating computation trees
+(forests), we must distinguish between
 \begin{inparaenum}[(i)]
@@
 \end{inparaenum} 
-In logical terms, \emph{control-local} interpolants are defined by
-formulae involving a single predicate symbol, whereas
-\emph{time-local} interpolants involve only predicates $\stamp{q}{i}$
-and variables $\stamp{x}{i}$, for a single $i \geq 0$.
-
-\paragraph{Remark}
-When considering an alternating model of computation, control-local
-interpolants are not always enough to prove emptiness, because of the
+In logical terms, \emph{control-local} interpolants are formulae
+involving a single predicate symbol, whereas \emph{time-local}
+interpolants involve only predicates $\stamp{q}{i}$ and variables
+$\stamp{x}{i}$, for a single $i \geq 0$.
+
+\paragraph{\em Remark}
+When considering alternating executions, control-local interpolants
+are not always enough to prove emptiness, because of the
 synchronization of several branches of the computation on the same
-sequence of input values. Consider, for instance, an automaton with
-the following transition rules and final state $q_f$:
+input word. Consider, for instance, an automaton with the following
+transition rules and final state $q_f$:
 \[\begin{array}{rclcrclcrcl}
@@
 q_1(y) & \arrow{a(x)}{} & q_1(y+x) \\ 
-q_2(y) & \arrow{a(x)}{} & y-x > 0 \wedge q_f && 
-q_2(y) & \arrow{a(x)}{} & q_2(y-x)
+&&&& q_2(y) & \arrow{a(x)}{} & y-x > 0 \wedge q_f
+&& q_2(y) & \arrow{a(x)}{} & q_2(y-x)
 \end{array}\]
@@
 \[\begin{array}{c}
-q_0(0) \arrow{(a,\nu_1)}{} \set{q_1(k_1),q_2(-k_1)} \ldots \arrow{(a,\nu_{n-1})}{} 
+q_0(0) \arrow{(a,\nu_1)}{} \set{q_1(k_1),q_2(-k_1)} \ldots
+\arrow{(a,\nu_{n-1})}{} 
 \set{q_1(\sum_{i=1}^{n-1} k_i),q_2(-\sum_{i=1}^{n-1} k_i)}
@@
 An overapproximation of the set of cubes generated after one or more
-steps is defined by the formula: $\exists x_1 \exists x_2 ~.~ q_1(x_1)
-\wedge q_2(x_2) \wedge x_1+x_2 \teq 0$. Observe that a control-local
-formula using one occurrence of a predicate would give a too rough
+steps is defined by the formula $\exists x_1 \exists x_2 ~.~ q_1(x_1)
+\wedge q_2(x_2) \wedge x_1+x_2 = 0$. Note that a control-local formula
+using one occurrence of a predicate would give a too rough
 overapproximation of this set, unable to prove the emptiness of the
@@
 the term \emph{local} to denote time-local interpolants, with no free
-variables.
+variables. The following definition generalizes interpolants from
+unsatisfiable conjunctions to sequences of input events:
 
@@
   sequence $(I_0,\ldots,I_n)$ of formulae such that, for all $k \in
-  [n-1]$: \begin{compactenum}
+  [n-1]$, the following hold: \begin{compactenum}
     \item\label{it1:generalized-lyndon-interpolant} $\pset{-}{I_k} =
@@
   If there exists a Lyndon interpolant for any two formulae $\phi$ and
-  $\psi$, such that $\phi \wedge \psi$ is unsatisfiable, then any
-  sequence of input events $\alpha = a_1 \ldots a_n \in \Sigma^*$,
-  such that $\accform{\alpha}$ is unsatisfiable, has a local GLI
-  $(I_0,\ldots,I_n)$.
+  $\psi$, in the first order theory of data with uninterpreted
+  predicate symbols, such that $\phi \wedge \psi$ is unsatisfiable,
+  then any sequence of input events $\alpha = a_1 \ldots a_n \in
+  \Sigma^*$, such that $\accform{\alpha}$ is unsatisfiable, has a
+  local GLI $(I_0,\ldots,I_n)$.
 \end{proposition}
@@
 
-The main problem with the local GLI construction described in the
-proof of Proposition \ref{prop:local-gli} is that the existence of
+A problematic point of the above proposition is that the existence of
 Lyndon interpolants (Definition \ref{def:lyndon-interpolant}) is
-guaranteed in principle, but the proof is non-constructive. Building
-an interpolant for an unsatisfiable conjunction of formulae $\phi
-\wedge \psi$ is typically the job of the decision procedure that
-proves the unsatisfiability and, in general, there is no such
-procedure, when $\phi$ and $\psi$ contain predicates and have
-non-trivial quantifier alternation. In this case, some provers use
-instantiation heuristics for the universal quantifiers that are
-sufficient for proving unsatisfiability, however these heuristics are
-not always suitable for interpolant generation. Consequently, from now
-on, we assume the existence of an effective Lyndon interpolation
-procedure only for decidable theories, such as the quantifier-free
-linear (integer) arithmetic with uninterpreted functions (UFLIA,
-UFLRA, etc.) \cite{RybalchenkoSofronieStokkermans}.
-
-This is where the predicate-free path formulae (Definition
-\ref{def:substform}) come into play. For a given event sequence
-$\alpha$, the automaton $\A$ accepts a word $w$ such that $\event{w} =
-\alpha$ if and only if $\substform{\alpha}$ is satisfiable. Assuming
-further that the equality atoms in the transition rules of $\A$ are
-written in the language of a decidable first order theory, such as
-Presburger arithmetic, Lemma \ref{lemma:quant-pred-acceptance} gives
-us an effective way of checking emptiness of $\A$, relative to a given
-event sequence. However, this method does not cope well with lazy
+proved in principle, but the proof is non-constructive. In other
+words, the proof of Proposition \ref{prop:local-gli} does not yield an
+algorithm for computing GLIs, for the following reason. Building an
+interpolant for an unsatisfiable conjunction of formulae $\phi \wedge
+\psi$ is typically the job of the decision procedure that proves the
+unsatisfiability and, in general, there is no such procedure, when
+$\phi$ and $\psi$ contain predicates and have non-trivial quantifier
+alternation. In this case, some provers use instantiation heuristics
+for the universal quantifiers that are sufficient for proving
+unsatisfiability, however these heuristics are not always suitable for
+interpolant generation. Consequently, from now on, we assume the
+existence of an effective Lyndon interpolation procedure only for
+decidable theories, such as the quantifier-free linear (integer)
+arithmetic with uninterpreted functions (UFLIA, UFLRA, etc.)
+\cite{RybalchenkoSofronieStokkermans}.
+
+This is where the predicate-free path formulae (defined in
+\S\ref{sec:path-formulae}) come into play. Recall that, for a given
+event sequence $\alpha$, the automaton $\A$ accepts a word $w$ such
+that $\event{w} = \alpha$ if and only if $\substform{\alpha}$ is
+satisfiable (Lemma \ref{lemma:quant-pred-acceptance}). Assuming
+further that the equality and interpreted predicates
+(e.g.\ inequalities for integers) atoms from the transition rules of
+$\A$ belong to a decidable first order theory, such as Presburger
+arithmetic, Lemma \ref{lemma:quant-pred-acceptance} gives us an
+effective way of checking emptiness of $\A$, relative to a given event
+sequence. However, this method does not cope well with lazy
 annotation, because there is no way to extract, from the
@@
 \item\label{it1:gli} the formula $\substform{\alpha}$, obtained by
-  repeated substitutions (Definition \ref{def:substform}) loses track
-  of the steps of the execution, and
+  repeated substitutions loses track of the steps of the execution,
+  and
 %
@@
 both written in prenex form, with matrices $\widehat{\Phi}$ and
-$\overline{\Phi}$, respectively. Because we assumed that the
-first order theory of $\Data$ has quantifier elimination, the
-satisfiability problem for $\substform{\alpha}$ is decidable. If
-$\substform{\alpha}$ is satisfiable, we build a counterexample for
-emptiness $w$ such that $\event{w}=\alpha$ and $\data{w}$ is a
-satisfying assignment for $\substform{\alpha}$. Otherwise,
-$\substform{\alpha}$ is unsatisfiable and there exist witness terms
-$\tau_{i_1} \ldots \tau_{i_\ell}$, where $\set{i_1, \ldots, i_\ell} =
-\set{j \in [1,m] \mid Q_j = \forall}$, such that
-$\overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
+$\overline{\Phi}$, respectively. Because we assumed that the first
+order theory of $\Data$ has quantifier elimination, the satisfiability
+problem for $\substform{\alpha}$ is decidable. If $\substform{\alpha}$
+is satisfiable, we build a counterexample for emptiness $w$ such that
+$\event{w}=\alpha$ and $\data{w}$ is a satisfying assignment for
+$\substform{\alpha}$. Otherwise, $\substform{\alpha}$ is unsatisfiable
+and there exist witness terms $\tau_{i_1} \ldots \tau_{i_\ell}$, where
+$\set{i_1, \ldots, i_\ell} = \set{j \in [1,m] \mid Q_j = \forall}$,
+such that $\overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
   \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable (Theorem
@@
 $\Data$, extended with uninterpreted function symbols, to compute a
-sequence of non-local GLI $(I_0, \ldots, I_n)$ such that $\voc{I_k}
+(not necessarily local) GLI $(I_0, \ldots, I_n)$ such that $\voc{I_k}
 \cap \stamp{Q}{\leq n} \subseteq \stamp{Q}{k}$, for all $k \in [n]$.
@@
 
-\begin{example}[Contd. from Examples \ref{ex:quant-elim} and \ref{ex:pred-elim}]\label{ex:gli}
-The formula $\substform{\alpha}$ (Example \ref{ex:pred-elim}) is
-unsatisfiable and let $\tau_2 = z_1$ be the witness term for the
-universally quantified variable $z_2$. Replacing $z_2$ with $\tau_2$
-in the matrix of $\quantform{\alpha}$ (Example \ref{ex:quant-elim})
-yields the unsatisfiable conjunction: 
-\vspace*{-\baselineskip}
-\[\begin{array}{c}
-z_1 \geq 0 \wedge \stamp{q}{0}(z_1)  ~\wedge~
-\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge (z_1 \geq z_1 \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_1)) ~\wedge~ \\
-\stamp{q}{1}(\stamp{x}{1}+z_1) \rightarrow \stamp{x}{1}+z_1 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_1)
+\begin{example}[Contd. from Examples \ref{ex:quant-elim} and
+    \ref{ex:pred-elim}]\label{ex:gli} The formula $\substform{\alpha}$
+  (Example \ref{ex:pred-elim}) is unsatisfiable and let $\tau_2 \isdef
+  z_1$ be the witness term for the universally quantified variable
+  $z_2$. Replacing $z_2$ with $\tau_2$ in the matrix of
+  $\quantform{\alpha}$ (Example \ref{ex:quant-elim}) yields the
+  unsatisfiable conjunction below, obtained after trivial
+  simplifications:
+\[\begin{array}{lcll}
+           [z_1 \geq 0 \wedge \stamp{q}{0}(z_1)] & \wedge &
+           [\stamp{q}{0}(z_1) \rightarrow \stamp{x}{1} \geq 0 \wedge \stamp{q}{1}(\stamp{x}{1}+z_1)] ~\wedge \\
+           && [\stamp{q}{1}(\stamp{x}{1}+z_1) \rightarrow \stamp{x}{1}+z_1 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_1)]
 \end{array}\]
-A non-local GLI for the above is $(\stamp{q}{0}(z_1) \wedge
-z_1\geq 0,~ \stamp{x}{1} \geq 0 \wedge \stamp{q}{1}(\stamp{x}{1}+z_1)
-\wedge z_1\geq0,~\bot)$. \hfill$\blacksquare$
+A non-local GLI for the above conjunction is the sequence of formulae: 
+\[(\stamp{q}{0}(z_1) \wedge z_1\geq 0,~ \stamp{x}{1} \geq 0 \wedge \stamp{q}{1}(\stamp{x}{1}+z_1)
+\wedge z_1\geq0,~\bot) \hspace*{4cm}\blacksquare\]
 \end{example}
 
-A function $\xi : \nat \rightarrow \nat$ is [strictly]
-\emph{monotonic} iff for each $n < m$ we have $\xi(n) \leq \xi(m)$
-     [$\xi(n) < \xi(m)$] and \emph{finite-range} iff for each $n \in
-     \nat$ the set $\set{m \mid \xi(m)=n}$ is finite. If $\xi$ is
-     finite-range, we denote by $\xi_{\max}^{-1}(n) \in \nat$ the
-     maximal value $m$ such that $\xi(m)=n$. The lemma below gives the
-     proof of correctness for the construction of non-local GLI.
-
+We formalize and prove the correctness for the above construction of
+non-local GLI. A function $\xi : \nat \rightarrow \nat$ is
+\emph{monotonic} iff for each $n < m$ we have $\xi(n) \leq \xi(m)$ and
+\emph{finite-range} iff for each $n \in \nat$ the set $\set{m \mid
+  \xi(m)=n}$ is finite. If $\xi$ is finite-range, we denote by
+$\xi_{\max}^{-1}(n) \in \nat$ the maximal value $m$ such that
+$\xi(m)=n$.
 \begin{lemma}\label{lemma:subst-quant}
@@
     $\alpha$, such that \(\voc{I_k} \subseteq \stamp{Q}{k} \cup
-    \stamp{X}{\leq k} \cup \set{x_j \mid j < \xi^{-1}(k),~ Q_j =
-      \exists}\), for all $k \in [n]$. 
+    \stamp{X}{\leq k} \cup \set{x_j \mid j < \xi_{\max}^{-1}(k),~ Q_j
+      = \exists}\), for all $k \in [n]$.
   \end{compactenum}
@@
 In conclusion, under two assumptions about the first order theory of
-the data domain, namely the\begin{inparaenum}[(i)]
+the data domain, namely the\ \begin{inparaenum}[(i)]
 \item witness-producing quantifier elimination, and
@@
 \end{inparaenum}
-we developped a rather generic method that produces generalized Lyndon
-interpolants for unfeasible input event sequences. Moreover, each
-formula $I_k$ in the interpolant refers only to the current predicate
-symbols $\stamp{Q}{I_k}$, the current and past input variables
-$\stamp{X}{\leq k}$ and the existentially quantified transition
-variables introduced at the previous steps $\set{x_j \mid j <
-  \xi_{\max}^{-1}(k), Q_j = \exists}$. The remaining question is how
-to use such non-local interpolants to label the unfolding of an
-automaton (Definition \ref{def:unfolding}) and to compute the coverage
-between nodes of the unfolding (Definition \ref{def:coverage}).
+we developped a rather generic method that produces GLIs for
+unfeasible input event sequences. Moreover, each formula $I_k$ in the
+interpolant refers only to the current predicate symbols
+$\stamp{Q}{I_k}$, the current and past input variables $\stamp{X}{\leq
+  k}$ and the existentially quantified transition variables introduced
+at the previous steps $\set{x_j \mid j < \xi_{\max}^{-1}(k), Q_j =
+  \exists}$. The remaining question is how to use such non-local GLIs
+to label the sequences in the unfolding of an automaton (Definition
+\ref{def:unfolding}) and to compute the coverage (Definition
+\ref{def:coverage}) between nodes of this unfolding.
 
@@
 other than predicate symbols, such that the labeling is compatible
-with the transition relation of the automaton, according to the point
-(\ref{it3:unfolding}) of Definition \ref{def:unfolding}. The following
-lemma describes the refinement of the labeling of an input sequence
-$\alpha$ of length $n$ by a non-local GLI $(I_0, \ldots, I_n)$, such
-that $\voc{I_k} \subseteq \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup
-\vec{x}_k$, where $\vec{x}_k$ are the existentially quantified
-variables from the prenex normal form of $\quantform{\alpha_k}$.
+with the transition relation of the automaton. Each newly expanded
+input sequence of $\A$ is initially labeled with $\top$ and the labels
+are refined using GLIs computed from proofs of spuriousness. The
+following lemma describes the refinement of the labeling of an input
+sequence by a non-local GLI:
 
@@
   \tuple{\Sigma,X,Q,\iota,F,\Delta}$ such that $\alpha = a_1\ldots a_n
-  \in \dom(U)$ and $(I_0, \ldots, I_n)$ be a GLI for $\alpha$. The
-  mapping $U' : \dom(U) \rightarrow \posforms(Q,\emptyset)$ defined as: 
+  \in \dom(U)$ is an input event sequence and $(I_0, \ldots, I_n)$ be
+  a GLI for $\alpha$. Then the following mapping $U' : \dom(U)
+  \rightarrow \posforms(Q,\emptyset)$ is an unfolding of $\A$, where:
   \begin{compactitem}
@@
   \end{compactitem}
-  is an unfolding of $\A$. 
 \end{lemma}
@@
 Basically, it is not important when exactly in the past a certain
-input has been read or when a choice has been made, as only the value
-of the variable determines the future behavior. Intuitively,
-existential quantification of these variables does the job of ignoring
-when in the past these values have been seen.
+input has been read or when a choice has been made, as only the
+relation between the values of these and the current variables
+determines the future behavior of the automaton. Intuitively,
+quantifying these variables existentially does the job of ignoring
+when exactly in the past these values have been seen.
 
@@
 \end{table}
+
+\section{Conclusions}
+
+We present first order alternating automata, a model of computation
+that generalizes classical boolean alternating automata to first order
+theories. Due to their expressivity, first order alternating automata
+are closed under union, intersection and complement. However the
+emptiness problem is undecidable even in the most simple case, of the
+quantifier-free theory of equality with uninterpreted predicate
+symbols. We deal with the emptiness problem by developping a practical
+semi-algorithm that always terminates, when the automaton is not
+empty. In case of emptiness, termination of the semi-algorithm occurs
+in most practical test cases, as shown by a number of experiments.

### 2019-02-04 quantifiers/abstract.tex
ok

@@
 undecidable, even for the simplest data theory of equality. To cope
-with this limitation, we develop an abstraction refinement
+with the undecidability problem, we develop an abstraction refinement
 semi-algorithm based on lazy annotation of the symbolic execution
@@
 automata, previously used in the verification of parameterized
-concurrent programs, composed of replicated threads, with a
-shared-memory communication model.
+concurrent programs, composed of replicated threads, with shared
+memory.
 \end{abstract}

### 2019-02-04 quantifiers/body.tex
ok

@@
 \begin{figure}[htb]
-\vspace*{-\baselineskip}
+\vspace*{-2\baselineskip}
 \begin{center}
@@
 \end{definition}
+
 The execution semantics of automata is given in close analogy with the
@@
 propositional variable and $\phi$ a positive boolean combination of
-propositional variables. For instance, $q_0 \arrow{a}{} q_1 \wedge q_2
-\vee q_3$ means that the automaton can choose to transition in either
-both $q_1$ and $q_2$ or in $q_3$ alone. This intuition leads to saying
-that a transition is a \emph{minimal model} of a right hand side
-formula of a rule. In our example, both $\set{q_1 \leftarrow \top, q_2
-  \leftarrow \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow
-  \bot, q_2 \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models,
-however $\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow
-  \top}$, is a model but is not minimal. The original definition of
-alternating finite-state automata \cite{ChandraKozenStockmeyer81}
-works around this problem by considering boolean valuations (models)
-instead of formulae. In contrast, a finite description of a first
-order alternating automaton must be symbolic (in terms of formulae)
-because here, a given configuration can have infinitely many
-successors.
+propositional variables. For instance, $q_0 \arrow{a}{} (q_1 \wedge
+q_2) \vee q_3$ means that the automaton can choose to transition in
+either both $q_1$ and $q_2$ or in $q_3$ alone. This intuition leads to
+saying that a transition is a \emph{minimal model} of a right hand
+side formula of a rule. In our example, both $\set{q_1 \leftarrow
+  \top, q_2 \leftarrow \top, q_3 \leftarrow \bot}$ and $\set{q_1
+  \leftarrow \bot, q_2 \leftarrow \bot, q_3 \leftarrow \top}$ are
+minimal models, however $\set{q_1 \leftarrow \top, q_2 \leftarrow
+  \top, q_3 \leftarrow \top}$, is a model but is not minimal. The
+original definition of alternating finite-state automata
+\cite{ChandraKozenStockmeyer81} works around this problem by
+considering boolean valuations (models) instead of formulae. In
+contrast, a finite description of a first order alternating automaton
+must be symbolic (in terms of formulae) because here, a given
+configuration can have infinitely many successors.
 
@@
 \cube{\minsem{\iota}}$, then $\A$ \emph{accepts} $w$ and let
-$\lang{\A}$ be the set of words accepted by $\A$. We consider the
-following problems: \begin{compactenum}
+$\lang{\A}$ be the set of words accepted by $\A$.
+
+\begin{example}\label{ex:foada}
+  Consider the automaton $\A = \tuple{\set{a}, \set{x},
+    \set{q_0,q_1,q_2,q_f}, q_0(0), \set{q_f}, \Delta}$, where $\Delta$
+  consists of the rules below:
+  \[\begin{array}{rclcrclcrcl}
+  q_0(y) & \arrow{a(x)}{} & q_1(y+x) \wedge q_2(y-x) & \hspace*{2mm} &
+  q_1(y) & \arrow{a(x)}{} & q_1(y+x) \hspace*{2mm} &
+  q_1(y) & \arrow{a(x)}{} & y > 0 \wedge q_f \\ 
+  &&&& q_2(y) & \arrow{a(x)}{} & q_2(y-x) &
+  q_2(y) & \arrow{a(x)}{} & y > 0 \wedge q_f
+  \end{array}\]
+  A possible execution tree of this automaton is the following:
+  \begin{center}
+    \input exec.pdf_t
+  \end{center}
+  The execution is not accepting, since its frontier is not labeled
+  with final configurations everywhere. It is not hard to see that
+  this automaton has no accepting execution. \hfill$\blacksquare$
+\end{example}
+
+In the rest of this paper, we are concerned with the following
+problems: \begin{compactenum}
 \item \emph{boolean closure}: given automata $\A_i =
@@
 synchronization of several branches of the computation on the same
-input word. Consider, for instance, an automaton with the following
-transition rules and final state $q_f$:
-\[\begin{array}{rclcrclcrcl}
-q_0(y) & \arrow{a(x)}{} & q_1(y+x) \wedge q_2(y-x) & \hspace*{2mm} & 
-q_1(y) & \arrow{a(x)}{} & y+x > 0 \wedge q_f & \hspace*{2mm} & 
-q_1(y) & \arrow{a(x)}{} & q_1(y+x) \\ 
-&&&& q_2(y) & \arrow{a(x)}{} & y-x > 0 \wedge q_f
-&& q_2(y) & \arrow{a(x)}{} & q_2(y-x)
-\end{array}\]
-Started in an initial configuration $q_0(0)$ with an input word
-$(a,\nu_1) \ldots (a,\nu_{n-1}) (a,\nu_n)$, such that $\nu_i(x)=k_i$,
-the automaton executes as follows:
+input word. Consider, for instance, the automaton from Example
+\ref{ex:foada}, started in an initial configuration $q_0(0)$, on an
+input word $(a,\nu_1) \ldots (a,\nu_n)$, such that $\nu_i(x)=k_i$, for
+all $i \in [1,n]$, the automaton executes as follows:
 \[\begin{array}{c}
-q_0(0) \arrow{(a,\nu_1)}{} \set{q_1(k_1),q_2(-k_1)} \ldots
-\arrow{(a,\nu_{n-1})}{} 
-\set{q_1(\sum_{i=1}^{n-1} k_i),q_2(-\sum_{i=1}^{n-1} k_i)}
-\arrow{(a,\nu_n)}{} \emptyset
+\set{(q_0,0)} \arrow{(a,\nu_1)}{} \set{(q_1,k_1),(q_2,-k_1)} \ldots
+\arrow{(a,\nu_{n})}{} 
+\set{(q_1,\sum_{i=1}^{n} k_i),(q_2,-\sum_{i=1}^{n} k_i)}
 \end{array}\]
@@
 steps is defined by the formula $\exists x_1 \exists x_2 ~.~ q_1(x_1)
-\wedge q_2(x_2) \wedge x_1+x_2 = 0$. Note that a control-local formula
+\wedge q_2(x_2) \wedge x_1+x_2 \geq 0$. Note that a control-local formula
 using one occurrence of a predicate would give a too rough
 overapproximation of this set, unable to prove the emptiness of the
-automaton. \hfill$\blacksquare$
+automaton. This is because any formula with a single occurrence of an
+atom of the form $q(x)$ would lose the constraint $x_1+x_2 \geq 0$
+between the values occurring at the same level, on both branches of
+the execution. \hfill$\blacksquare$
 
@@
 \end{proposition}
-\proof{ By definition, $\accform{\alpha}$ is the formula:
-  \[\stamp{\iota}{0} \wedge 
+\paragraph{\em Proof.} By definition, $\accform{\alpha}$ is the formula:
+\[\begin{array}{c}
+\stamp{\iota}{0} \wedge 
  \bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in
@@
  \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
- ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot\]
+ ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot
+ \end{array}\]
  We define the formulae: 
@@
        \cup \pset{-}{\psi}) = \emptyset$. Continuing in this way, we
-       obtain formulae $I_0, I_1, \ldots, I_{n}$ as required. \qed}
+       obtain formulae $I_0, I_1, \ldots, I_{n}$ as required. \qed
 
@@
   z_1$ be the witness term for the universally quantified variable
-  $z_2$. Replacing $z_2$ with $\tau_2$ in the matrix of
+  $z_2$. Replacing $z_2$ with $\tau_2$ ($z_1$) in the matrix of
   $\quantform{\alpha}$ (Example \ref{ex:quant-elim}) yields the
@@
 \begin{table}[htb]
-\vspace*{-\baselineskip}
+%\vspace*{-\baselineskip}
 \begin{center}

### 2019-02-05 quantifiers/body.tex
ok

@@
   \end{array}\]  
-  Moreover, $U$ is \emph{safe} if for each $\alpha \in \dom(U)$, the
-  formula $U(\alpha) \wedge \bigwedge_{q \in Q \setminus F} \forall
-  y_1 \ldots \forall_{y_{\#(q)}} ~.~ q(\vec{y}) \rightarrow \bot$ is
+  $U$ is \emph{safe} if for each $\alpha \in \dom(U)$, the formula
+  $U(\alpha) \wedge \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots
+  \forall_{y_{\#(q)}} ~.~ q(\vec{y}) \rightarrow \bot$ is
   unsatisfiable.
@@
       \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
-      y_{\#(q)} ~.~ q(\vec{y})$ is unsatisfiable.
+      y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot$ is unsatisfiable.
   \end{compactenum}
@@
 \end{theorem}
-\proof{See \cite[Theorem 2.1.8]{BorgerGraedelGurevich97} and
-  \cite[Lemma 2.1.9]{BorgerGraedelGurevich97}. \qed}
+\paragraph{Proof.} See \cite[Theorem 2.1.8]{BorgerGraedelGurevich97} and
+  \cite[Lemma 2.1.9]{BorgerGraedelGurevich97}. \qed
 
@@
 
-\vspace*{-0.5\baselineskip}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\enlargethispage{6mm}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Experimental Results}
-\vspace*{-0.5\baselineskip}
 
@@
 
-The experimental results, reported in Table \ref{tab:experiments},
-come from several sources, namely predicate automata models (*.pa)
-\cite{Farzan15,Farzan16} available online \cite{pa}, timed automata
-inclusion problems ({\tt abp.ada}, {\tt train.ada}, {\tt
-  rr-crossing.foada}), array logic entailments ({\tt
-  array\_rotation.ada}, {\tt array\_simple.ada}, {\tt
-  array\_shift.ada}) and hardware circuit verification ({\tt hw1.ada},
-{\tt hw2.ada}), initially considered in \cite{IosifRV16}. The {\tt
-  train-simpleN.} {\tt foada} and {\tt fischer-mutexN.foada} examples are
-parametric verification problems in which one checks inclusions of the
-form $\bigcap_{i=1}^N\lang{A_i} \subseteq \lang{B}$, where $A_i$ is
-the $i$-th copy of the same template automaton.
-
-The advantage of using FOADA over the INCLUDER \cite{includer} tool
-from \cite{IosifRV16} is the possibility of having infinite alphabet
-automata with hidden (local) variables, whose values are not visible
-in the input. In particular, this is essential for checking inclusion
-of timed automata that use internal clocks to control the computation.
-
 \begin{table}[htb]
-%\vspace*{-\baselineskip}
+\vspace*{-\baselineskip}
 \begin{center}
@@
 \end{center}
-\vspace*{-2\baselineskip}
+\vspace*{-3\baselineskip}
 \end{table}
 
+The experimental results, reported in Table \ref{tab:experiments},
+come from several sources, namely predicate automata models (*.pa)
+\cite{Farzan15,Farzan16} available online \cite{pa}, timed automata
+inclusion problems ({\tt abp.ada}, {\tt train.ada}, {\tt
+  rr-crossing.foada}), array logic entailments ({\tt
+  array\_rotation.ada}, {\tt array\_simple.ada}, {\tt
+  array\_shift.ada}) and hardware circuit verification ({\tt hw1.ada},
+{\tt hw2.ada}), initially considered in \cite{IosifRV16}. The {\tt
+  train-simpleN.} {\tt foada} and {\tt fischer-mutexN.foada} examples are
+parametric verification problems in which one checks inclusions of the
+form $\bigcap_{i=1}^N\lang{A_i} \subseteq \lang{B}$, where $A_i$ is
+the $i$-th copy of the same template automaton.
+
+The advantage of using FOADA over the INCLUDER \cite{includer} tool
+from \cite{IosifRV16} is the possibility of having infinite alphabet
+automata with hidden (local) variables, whose values are not visible
+in the input. In particular, this is essential for checking inclusion
+of timed automata that use internal clocks to control the computation.
+
+\vspace*{-\baselineskip}
 \section{Conclusions}

### 2019-02-05 quantifiers/body.tex
ok

@@
   \end{array}\]
-  $U$ is \emph{safe} if for each $\alpha \in \dom(U)$, the formula
-  $U(\alpha) \wedge \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots
+  A path $\alpha$ is \emph{safe} in $U$ if and only if $U(\alpha)
+  \wedge \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots
   \forall_{y_{\#(q)}} ~.~ q(\vec{y}) \rightarrow \bot$ is
-  unsatisfiable.
+  unsatisfiable. The unfolding $U$ is safe if and only if every path
+  in $\dom(U)$ is safe in $U$.
 \end{definition}
@@
 \stamp{\iota}{0} \wedge 
- \bigwedge_{i=1}^n \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in
+ \bigwedge_{i=1}^n \underbrace{\bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in
    \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
- \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i} \wedge
- \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
- ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot
+ \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i}}_{\varphi_i} \wedge
+ \underbrace{\bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
+ ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot}_{\phi}
  \end{array}\]
- We define the formulae: 
- \[\begin{array}{rcl}
- \varphi_i & \isdef & \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
- \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i},~ \text{for all}~ i \in [1,n] \\
- \psi & \isdef & \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
- ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot
- \end{array}\]
+ %% We define the formulae: 
+ %% \[\begin{array}{rcl}
+ %% \varphi_i & \isdef & \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
+ %% \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i},~ \text{for all}~ i \in [1,n] \\
+ %% \phi & \isdef & \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
+ %% ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot
+ %% \end{array}\]
  Observe that $\voc{\stamp{\iota}{0}} \subseteq \stamp{Q}{0}$,
@@
  \stamp{Q}{i-1}\cup\stamp{Q}{i}\cup\stamp{X}{i}$, for all $i \in
-       [1,n]$, and $\voc{\psi} \subseteq \stamp{Q}{n}$.  We apply
+       [1,n]$, and $\voc{\phi} \subseteq \stamp{Q}{n}$.  We apply
        Lyndon's Interpolation Theorem for the formulae
        $\stamp{\iota}{0}$ and $\bigwedge_{i=1}^n \varphi_i \wedge
-       \psi$ and obtain a formula $I_0$, such that $\stamp{\iota}{0}
+       \phi$ and obtain a formula $I_0$, such that $\stamp{\iota}{0}
        \models I_0$, $I_0 \wedge \bigwedge_{i=1}^n \varphi_i \wedge
-       \psi$ is unsatisfiable, $\voc{I_0} \subseteq
+       \phi$ is unsatisfiable, $\voc{I_0} \subseteq
        \voc{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n \voc{\varphi_i}
-       \cup \voc{\psi}) \subseteq \stamp{Q}{0}$ and $\pset{-}{I_0}
+       \cup \voc{\phi}) \subseteq \stamp{Q}{0}$ and $\pset{-}{I_0}
        \subseteq \pset{-}{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n
-       \pset{-}{\varphi} \cup \pset{-}{\psi}) = \emptyset$. Repeating
+       \pset{-}{\varphi} \cup \pset{-}{\phi}) = \emptyset$. Repeating
        the reasoning for the formulae $I_0 \wedge \varphi_1$ and
-       $\bigwedge_{i=2}^n \varphi_i \wedge \psi$, we obtain $I_1$,
+       $\bigwedge_{i=2}^n \varphi_i \wedge \phi$, we obtain $I_1$,
        such that $I_0 \wedge \varphi_1 \models I_1$, $I_1 \wedge
-       \bigwedge_{i=2}^n \varphi_i \wedge \psi$ is unsatisfiable,
+       \bigwedge_{i=2}^n \varphi_i \wedge \phi$ is unsatisfiable,
        $\voc{I_1} \subseteq (\voc{I_0} \cup \voc{\varphi_1}) \cap
-       (\bigcup_{i=2}^n \voc{\varphi_i} \cup \voc{\psi}) \subseteq
+       (\bigcup_{i=2}^n \voc{\varphi_i} \cup \voc{\phi}) \subseteq
        \stamp{Q}{1}$ and $\pset{-}{I_1} \subseteq (\pset{-}{I_0} \cup
        \pset{-}{\varphi_1}) \cap (\bigcup_{i=2}^n \pset{-}{\varphi_i}
-       \cup \pset{-}{\psi}) = \emptyset$. Continuing in this way, we
+       \cup \pset{-}{\phi}) = \emptyset$. Continuing in this way, we
        obtain formulae $I_0, I_1, \ldots, I_{n}$ as required. \qed
@@
   \end{compactitem}
+  Moreover, $\alpha$ is safe in $U'$.
 \end{lemma}
@@
 quantifying these variables existentially does the job of ignoring
-when exactly in the past these values have been seen.
+when exactly in the past these values have been seen. Moreover, the
+last point of Lemma \ref{lemma:refinement} ensures that the refined
+path is safe in the new unfolding and thus will stay safe in all
+future unfoldings of $\A$.
 
@@
 \fi
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\enlargethispage{6mm}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\vspace*{-\baselineskip}
 \section{Conclusions}

### 2019-02-07 quantifiers/body.tex
ok
201902070143

@@
 initial configuration $s_1(0,0) \wedge q_1(0,0)$. Our lazy annotation
-semi-algorithm finds that this automaton is empty in \comment[ri]{...}
+semi-algorithm finds that this automaton is empty in 347 ms
 seconds.
@@
 \end{theorem}
-\paragraph{Proof.} See \cite[Theorem 2.1.8]{BorgerGraedelGurevich97} and
+\paragraph{\em Proof.} See \cite[Theorem 2.1.8]{BorgerGraedelGurevich97} and
   \cite[Lemma 2.1.9]{BorgerGraedelGurevich97}. \qed
@@
 \begin{table}[htb]
-\vspace*{-\baselineskip}
+%\vspace*{-\baselineskip}
 \begin{center}
 {\fontsize{7}{8}\selectfont
-\begin{tabular}{||l|c|c|c|c|c||}
+\begin{tabular}{||l|c|c|c|c|c|c|c|c||}
 \hline
-Example & $\len{\A}$ (bytes) & $L(\A)=\emptyset$ ? & Nodes Expanded & Nodes Visited & Time (ms) \\
+Example & $\len{\A}$ (bytes) & Nb of Predicates & Nb of Variables & Nb of Transitions & $L(\A)=\emptyset$ ? & Nodes Expanded & Nodes Visited & Time (ms) \\
 \hline
-incdec.pa & 499 & no & 21 & 17 & 779 \\
+incdec.pa & 499 & 3 & 1 & 12 & no & 21 & 17 & 779 \\
 \hline
-localdec.pa & 678 & no & 49 & 35 & 1814 \\
+localdec.pa & 678 & 4 & 1 & 16 & no & 49 & 35 & 1814 \\
 \hline
-ticket.pa & 4250 & no & 229 & 91 & 9543 \\
+ticket.pa & 4250 & 13 & 1 & 73 & no & 229 & 91 & 9543 \\
 \hline
-count\_thread0.pa & 9767 & no & 154 & 128 & 8553 \\
+count\_thread0.pa & 9767 & 14 & 1 & 126 & no & 154 & 128 & 8553 \\
 \hline
-count\_thread1.pa & 10925 & no & 766 & 692 & 76771 \\
+count\_thread1.pa & 10925 & 15 & 1 & 135 & no & 766 & 692 & 76771 \\
 \hline
-local0.pa & 10595 & no & 73 & 27 & 1431 \\
+local0.pa & 10595 & 13 & 1 & 117 & no & 73 & 27 & 1431 \\
 \hline
-local1.pa & 11385 & no & 1135 & 858 & 101042 \\
+local1.pa & 11385 & 14 & 1 & 126 & no & 1135 & 858 & 101042 \\
 \hline
-array\_rotation.ada & 1834 & yes & 9 & 8 & 1543 \\
+array\_rotation.ada & 1834 & 8 & 7 & 7 & yes & 9 & 8 & 1543 \\
 \hline
-array\_simple.ada & 3440 & yes & 11 & 10 & 6787 \\
+array\_simple.ada & 3440 & 9 & 5 & 8 & yes & 11 & 10 & 6787 \\
 \hline
-array\_shift.ada & 874 & yes & 6 & 5 & 413 \\
+array\_shift.ada & 874 & 6 & 5 & 5 & yes & 6 & 5 & 413 \\
 \hline
-abp.ada & 6909 & no & 52 & 47 & 4788 \\
+abp.ada & 6909 & 16 & 14 & 28 & no & 52 & 47 & 4788 \\
 \hline
-train.ada & 1823 & yes & 68 & 67 & 7319 \\
+train.ada & 1823 & 10 & 4 & 26 & yes & 68 & 67 & 7319 \\
 \hline
-hw1.ada & 322 & Solver Error & / & / & / \\
+hw1.ada & 322 & 3 & 2 & 5 & Solver Error & / & / & / \\
 \hline
-hw2.ada & 674 & yes & 20 & 22 & 4974 \\
+hw2.ada & 674 & 7 & 2 & 8 & yes & 20 & 22 & 4974 \\
 \hline
-rr-crossing.foada & 1780 & yes & 67 & 67 & 7574 \\
+rr-crossing.foada & 1780 & 10 & 1 & 16 & yes & 67 & 67 & 7574 \\
 % \hline
@@
 \hline
-train-simple1.foada & 5421 & yes & 43 & 44 & 2893 \\
+train-simple1.foada & 5421 & 13 & 1 & 61 & yes & 43 & 44 & 2893 \\
 \hline
-train-simple2.foada & 10177 & yes & 111 & 113 & 8386 \\
+train-simple2.foada & 10177 & 16 & 1 & 118 & yes & 111 & 113 & 8386 \\
 \hline
-train-simple3.foada & 15961 & yes & 196 & 200 & 15041 \\
+train-simple3.foada & 15961 & 19 & 1 & 193 & yes & 196 & 200 & 15041 \\
 % \hline
@@
 \hline
-fischer-mutex2.foada & 3000 & yes & 23 & 23 & 808 \\
+fischer-mutex2.foada & 3000 & 11 & 2 & 23 & yes & 23 & 23 & 808 \\
 \hline
-fischer-mutex3.foada & 4452 & yes & 33 & 33 & 1154 \\
+fischer-mutex3.foada & 4452 & 16 & 2 & 34 & yes & 33 & 33 & 1154 \\
 \hline
@@
 \end{center}
-\vspace*{-3\baselineskip}
+%\vspace*{-3\baselineskip}
 \end{table}

### 2019-02-07 quantifiers/abstract.tex
ok

@@
 elimination with witness term generation and (ii) Lyndon interpolation
-in the quantifier-free data theory with uninterpreted predicate
-symbols. This provides a method for checking inclusion of timed and
-finite-memory register automata, and emptiness of quantified predicate
-automata, previously used in the verification of parameterized
-concurrent programs, composed of replicated threads, with shared
-memory.
+in the quantifier-free theory of the data domain, with uninterpreted
+predicate symbols. This provides a method for checking inclusion of
+timed and finite-memory register automata, and emptiness of quantified
+predicate automata, previously used in the verification of
+parameterized concurrent programs, composed of replicated threads,
+with shared memory.
 \end{abstract}

### 2019-02-07 quantifiers/appendix.tex
ok

\section{Emptiness Semi-algorithm for First Order Alternating Automata}
\label{app:algorithm}

\section{Proofs}
\label{app:proofs}

### 2019-02-07 quantifiers/body.tex
ok

@@
 procedures is not guaranteed, but having a suitable coverage relation
-between the nodes of the search tree may ensure convergence of some
-real-life examples. Applying lazy annotation to first order
+between the nodes of the search tree may ensure convergence of many
+real-life examples. However, applying lazy annotation to first order
 alternating automata faces two nontrivial
 problems: \begin{compactenum}
-\item Quantified transition rules make it hard, if not impossible, to
-  decide if a path is infeasible. This is mainly because adding
-  uninterpreted predicate symbols to decidable first-order theories,
-  such as Presburger arithmetic, results in undecidability
+\item Quantified transition rules make it hard, if not impossible, in
+  general, to decide if a path is infeasible. This is mainly because
+  adding uninterpreted predicate symbols to decidable first-order
+  theories, such as Presburger arithmetic, results in undecidability
   \cite{Halpern91}. To deal with this problem, we assume that the
@@
 $q_2$). Intuititvely, $A$ recognizes the sequences in which all events
-are $a$ and the second event occurs after at least $2$ time units,
-whereas $B$ recognizes the sequences in which there exists an event
-$a$ such that no event occurs after it in exactly $1$ time
-unit. Formally, $\lang{A} = \set{a, t_1, \ldots a, t_n \mid t_1,
-  \ldots, t_n \in \real,~ t_1 \geq 2}$ and $\lang{B} = \set{\sigma_1,
-  t_1 \ldots \sigma_{i-1}, t_{i-1}, a, t_i, \ldots \sigma_n, t_n \mid
-  \sigma_1, \ldots, \sigma_n \in \set{a,b},~ t_1, \ldots, t_n \in
-  \real,~ t_i-t_{i-1} \neq 1}$. It is proved\footnote{See
-  \cite[Theorem 1]{AlurMadhusudan04} for a proof.} that $\clang{B}$ is
-not recognized by a timed automaton, thus the inclusion cannot be
-proved in the framework of timed regular languages, as $\lang{A} \cap
-\clang{B} = \emptyset$.
+are $a$ and the second event, if it occurs at all, it occurs after at
+least $2$ time units. On the other hand, $B$ recognizes the sequences
+in which there exists an event $a$ such that no event occurs after it
+in exactly $1$ time unit. Formally, $\lang{A} = \set{a,t \mid t \in
+  \real} \cup \set{a, t_1, \ldots a, t_n \mid t_1, \ldots, t_n \in
+  \real,~ n \geq 2,~ t_1 \geq 2}$ and $\lang{B} = \set{\sigma_1, t_1
+  \ldots \sigma_{i-1}, t_{i-1}, a, t_i, \ldots \sigma_n, t_n \mid
+  \sigma_1, \ldots, \sigma_n \in \set{a,b},~ n \geq 1,~ t_1, \ldots,
+  t_n \in \real,~ t_i-t_{i-1} \neq 1}$. It is proved\footnote{See
+  \cite[Theorem 1]{AlurMadhusudan04} for a proof.}  that $\clang{B}$
+cannot be recognized by a timed automaton, thus the inclusion cannot
+be proved in the framework of timed regular languages, as $\lang{A}
+\cap \clang{B} = \emptyset$.
 
@@
 time value. The time stamp of the input event $a$ is the input
-variable $t$, which must be bigger or equal to $z$, as the lapse of
+variable $t$, which must be bigger or equal to $z$, since the lapse of
 time between two consequtive events is always positive. The current
@@
 \arrow{}{} s_2$ transition is encoded by passing $z$ as the first
-parameter of $s_2$, in the first rule.
+parameter of $s_2$, in the first rule. The initial configuration of
+$A$ [$B$] is given by the atomic propositon $s_1(0,0)$ [$q_1(0,0)$]
+and $s_2$ [$q_2$] marks the final configurations.
 
@@
 with disjunctions in the rules of the alternating automaton encoding
-$B$, denoted as $\overline{B}$ in Fig. \ref{fig:ta} (bottom
-right). Finally, we check the emptiness of the alternating automaton
-consisting of the rules of $A$ and $\overline{B}$, started in the
-initial configuration $s_1(0,0) \wedge q_1(0,0)$. Our lazy annotation
-semi-algorithm finds that this automaton is empty in 347 ms
-seconds.
+$B$, denoted as $\overline{B}$ in Fig. \ref{fig:ta} (bottom right),
+with final configurations marked by $q_1$, instead of $q_2$. Finally,
+we check the emptiness of the alternating automaton consisting of the
+rules of $A$ and $\overline{B}$, started in the initial configuration
+$s_1(0,0) \wedge q_1(0,0)$. Our lazy annotation semi-algorithm proves
+emptiness of this automaton in $\sim\!0.3$ seconds, on an average
+laptop machine. The missing proofs of the technical results are given
+in Appendinx \ref{app:proofs}.
 
@@
 variables, and a monitor automaton $B$. The semi-algorithm defined
-here relies on the assumption that all variables of the observer $B$
-must be declared in the automata $A_1, \ldots, A_n$ under check. First
-order alternating automata bypass this limitation, since the emptiness
-checking semi-algorithm handles invisible variables.
+there was based on the assumption that all variables of the observer
+$B$ must be declared in the automata $A_1, \ldots, A_n$ under
+check. First order alternating automata bypass this limitation, since
+our emptiness checking semi-algorithm can handle invisible variables.
 
@@
 statements and thread identifiers, thus being infinite because the
-number of threads is unbounded. Because thread identifiers can only be
+number of threads is unbounded. Since thread identifiers can only be
 compared for (dis-)equality, the data theory in PA is the theory of
@@
 undecidable when either the predicates have arity greater than one
-\cite{Farzan15} or quantified transition rules
+\cite{Farzan15} or use quantified transition rules
 \cite{KincaidPhD}. Checking emptiness of quantifier-free PA is
@@
 $F \subseteq Q$ is the set of predicates denoting final states, and
-$\Delta$ is a set of \emph{transition rules} of the form
-\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
-predicate, $a \in \Sigma$ is an input event and $\psi \in
+$\Delta$ is a set of \emph{transition rules}. A transition rule is of
+the form \(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in
+Q$ is a predicate, $a \in \Sigma$ is an input event and $\psi \in
 \posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive formula,
@@
 rules can be joined using disjunction. The quantifiers occurring in
-the right-hand side formula of a transition rule are referred to as
+the right-hand side formula of a transition rule are called
 \emph{transition quantifiers}. The \emph{size} of $\A$ is $\len{\A}
@@
 
-Given an uninterpreted predicate symbol $q \in Q$ and a tuple of data
-values $d_1,\ldots,d_{\#(q)}$, the tuple $(q, d_1,\ldots,d_{\#(q)})$
+The semantics of first order alternating automata is analogous to the
+semantics of propositional alternating automata, with rules of the
+form $q \arrow{a}{} \phi$, where $q$ is a propositional variable and
+$\phi$ a positive boolean combination of propositional variables. For
+instance, $q_0 \arrow{a}{} (q_1 \wedge q_2) \vee q_3$ means that the
+automaton can choose to transition in either both $q_1$ and $q_2$ or
+in $q_3$ alone. This leads to defining transitions as the
+\emph{minimal models} of the right hand side of a rule\footnote{Both
+  $\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow
+    \bot}$ and $\set{q_1 \leftarrow \bot, q_2 \leftarrow \bot, q_3
+    \leftarrow \top}$ are minimal models, however $\set{q_1 \leftarrow
+    \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$ is a model but is
+  not minimal.}. The original definition of alternating automata
+\cite{ChandraKozenStockmeyer81} works around this problem and
+considers boolean valuations instead of formulae. In contrast, a
+finite description of a first order alternating automaton cannot be
+given in terms of interpretations, as a first order formula may have
+infinitely many models, corresponding to infinitely many initial or
+successor states occurring within an execution step.
+
+Given an uninterpreted predicate symbol $q \in Q$ and data
+values $d_1,\ldots,d_{\#(q)} \in \Data$, the tuple $(q, d_1,\ldots,d_{\#(q)})$
 is called a \emph{configuration}, sometimes written
@@
 
-The execution semantics of automata is given in close analogy with the
-classical definition of propositional alternating automata, with
-transition rules of the form $q \arrow{a}{} \phi$, where $q$ is a
-propositional variable and $\phi$ a positive boolean combination of
-propositional variables. For instance, $q_0 \arrow{a}{} (q_1 \wedge
-q_2) \vee q_3$ means that the automaton can choose to transition in
-either both $q_1$ and $q_2$ or in $q_3$ alone. This intuition leads to
-saying that a transition is a \emph{minimal model} of a right hand
-side formula of a rule. In our example, both $\set{q_1 \leftarrow
-  \top, q_2 \leftarrow \top, q_3 \leftarrow \bot}$ and $\set{q_1
-  \leftarrow \bot, q_2 \leftarrow \bot, q_3 \leftarrow \top}$ are
-minimal models, however $\set{q_1 \leftarrow \top, q_2 \leftarrow
-  \top, q_3 \leftarrow \top}$, is a model but is not minimal. The
-original definition of alternating finite-state automata
-\cite{ChandraKozenStockmeyer81} works around this problem by
-considering boolean valuations (models) instead of formulae. In
-contrast, a finite description of a first order alternating automaton
-must be symbolic (in terms of formulae) because here, a given
-configuration can have infinitely many successors.
-
 An execution $\T$ over $w$, starting with $c$, is \emph{accepting} if
@@
     \set{q_0,q_1,q_2,q_f}, q_0(0), \set{q_f}, \Delta}$, where $\Delta$
-  consists of the rules below:
-  \[\begin{array}{rclcrclcrcl}
-  q_0(y) & \arrow{a(x)}{} & q_1(y+x) \wedge q_2(y-x) & \hspace*{2mm} &
-  q_1(y) & \arrow{a(x)}{} & q_1(y+x) \hspace*{2mm} &
-  q_1(y) & \arrow{a(x)}{} & y > 0 \wedge q_f \\ 
-  &&&& q_2(y) & \arrow{a(x)}{} & q_2(y-x) &
-  q_2(y) & \arrow{a(x)}{} & y > 0 \wedge q_f
-  \end{array}\]
-  A possible execution tree of this automaton is the following:
+  is the set: \(q_0(y) \arrow{a(x)}{} q_1(y+x) \wedge q_2(y-x)\),
+  \(q_1(y) \arrow{a(x)}{} q_1(y+x) \vee (y > 0 \wedge q_f)\) and
+  \(q_2(y) \arrow{a(x)}{} q_2(y-x) \vee (y > 0 \wedge q_f)\). A
+  possible execution tree of this automaton is the following:
   \begin{center}
@@
   \end{center}
-  The execution is not accepting, since its frontier is not labeled
-  with final configurations everywhere. It is not hard to see that
-  this automaton has no accepting execution. \hfill$\blacksquare$
+  The execution tree is not accepting, since its frontier is not
+  labeled with final configurations everywhere. Incidentally, here we
+  have $\lang{\A} = \emptyset$, which is proved by our tool in
+  $\sim\!0.5$ seconds on an average machine. \hfill$\blacksquare$
 \end{example}
@@
 solving this problem first, we develop the machinery required to prove
-that first order alternating automata are closed under all boolean
-operations and, further, set up the ground for a semi-algorithm for
-the emptiness problem.
+that first order alternating automata are closed under complement and,
+further, set up the ground for developping a practical semi-algorithm
+for the emptiness problem.
 
@@
 $\pathform{\alpha}$ to an \emph{acceptance formula}
-$\accform{\alpha}$, which defines only accepting executions over words
-that share a given input sequence, i.e.\ $\accform{\alpha}$ is
-satisfiable if and only if the automaton accepts a word $w$ such that
-$\event{w} = \alpha$.
+$\accform{\alpha}$, which defines only those executions that are
+accepting among $\pathform{\alpha}$. Consequently, the automaton
+accepts a word $w$ such that $\event{w} = \alpha$ if and only if
+$\accform{\alpha}$ is satisfiable.
 
@@
 monadic predicate symbol to it yields undecidability in the presence
-of non-trivial quantifier alternation \cite{Halpern91}. However the
-quantifier-free fragment of Presburger arithmetic extended with
-predicate symbols can be shown to be decidable, using a Nelson-Oppen
+of non-trivial quantifier alternation \cite{Halpern91}. On the other
+hand, the quantifier-free fragment of Presburger arithmetic extended
+with uninterpreted function symbols is decidable, by a Nelson-Oppen
 style congruence closure argument \cite{NelsonOppen80}.
 
-To tackle this problem, we start from the observation that acceptance
-formulae have a particular form, which allows the elimination of path
-quantifiers and uninterpreted predicate symbols, by a couple of
+To tackle the problem of deciding satisfiability of $\accform{\alpha}$
+formulae, we start from the observation that their form is rather
+particular, which allows the elimination of path quantifiers and
+uninterpreted predicate symbols, by a couple of
 satisfiability-preserving transformations. The result of applying
@@
 only quantifiers are those introduced by the transition rules of the
-automaton. Next, in \S\ref{sec:emptiness} we assume that the first
-order theory of the data sort $\Data$ has quantifier elimination,
-which allows to effectively decide satisfiability of acceptance
-formulae.
+automaton. Next, in \S\ref{sec:emptiness} we shall assume moreover
+that the first order theory of the data sort $\Data$ (without
+uninterpreted predicate symbols) has quantifier elimination, providing
+thus an effective decision procedure.
 
 For the time being, let us formally define the elimination of
-transition quantifiers and predicates, respectively. Let $\alpha = a_1
-\ldots a_n$ be a given sequence of input events and let
-$\alpha_i$ be the prefix $a_1 \ldots a_i$ of $\alpha$, for $i \in [n]$,
-where $\alpha_0=\epsilon$.
+transition quantifiers and predicate symbols. Let $\alpha = a_1 \ldots
+a_n$ be a given sequence of input events and let $\alpha_i$ be the
+prefix $a_1 \ldots a_i$ of $\alpha$, for $i \in [n]$, where
+$\alpha_0=\epsilon$.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 $\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)})$ occurs in
-$\quantformn{\alpha_{i-1}}$. Next, we write $\quantform{\alpha}$ for
-the conjunction of $\quantformn{\alpha_n}$ with all
+$\quantformn{\alpha_{i-1}}$, for some terms
+$t_1,\ldots,t_{\#(q)}$. Next, we write $\quantform{\alpha}$ for the
+conjunction of $\quantformn{\alpha_n}$ with all
 $\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot$, such that
@@
   \[\begin{array}{rcl}
-  \accform{\alpha} & = & \exists z ~.~ z \geq 0 \wedge \stamp{q}{0}(z) ~\wedge \\
-  && \forall y ~.~ \stamp{q}{0}(y) \rightarrow [\stamp{x}{1}\geq0 \wedge \forall z ~.~ z\geq y 
-    \rightarrow \stamp{q}{1}(\stamp{x}{1}+z)] ~\wedge \\
+  \accform{\alpha} & = & \exists z_1 ~.~ z_1 \geq 0 \wedge \stamp{q}{0}(z_1) ~\wedge \\
+  && \forall y ~.~ \stamp{q}{0}(y) \rightarrow [\stamp{x}{1}\geq0 \wedge \forall z_2 ~.~ z_2\geq y 
+    \rightarrow \stamp{q}{1}(\stamp{x}{1}+z_2)] ~\wedge \\
   && \forall y ~.~ \stamp{q}{1}(y) \rightarrow [y<0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+y)]
@@
   && [\stamp{q}{1}(\stamp{x}{1}+z_2) \rightarrow \stamp{x}{1}+z_2 < 0 \wedge \stamp{q_f}{2}(\stamp{x}{2}+\stamp{x}{1}+z_2)]   
-  \enspace\enspace\hfill\blacksquare
   \end{array}\]
+  Notice that the transition quantifiers $\exists z_1$ and $\forall
+  z_2$ from $\accform{\alpha}$ range now over the entire formula
+  $\quantform{\alpha}$. \hfill$\blacksquare$
 \end{example}
@@
   valuation $\nu : \stamp{X}{\leq n} \rightarrow \Data$, the following
-  hold: \begin{compactenum}
-  \item\label{it1:quant} for all interpretations $\I$, if $\I,\nu
-    \models \accform{\alpha}$ then $\I,\nu \models
-    \quantform{\alpha}$. 
-    %
-  \item\label{it2:quant} if there exists an interpretation $\I$ such
-    that $\I,\nu \models \quantform{\alpha}$ then there exists an
-    interpretation $\J \subseteq \I$ such that $\J,\nu \models
-    \accform{\alpha}$. 
+  hold, for every interpretation $\I$: \begin{compactenum}
+  \item\label{it1:quant} if $\I,\nu \models \accform{\alpha}$ then
+    $\I,\nu \models \quantform{\alpha}$, and
+    %
+  \item\label{it2:quant} if $\I,\nu \models \quantform{\alpha}$ there
+    exists an interpretation $\J \subseteq \I$ such that $\J,\nu
+    \models \accform{\alpha}$.
   \end{compactenum}
@@
 by considering the sequence of formulae $\substformn{\alpha_0} \isdef
-\stamp{\iota}{0}$ and, for all $i \in [1,n]$, $\substformn{\alpha_i}$
-is obtained by substituting each predicate atom
+\stamp{\iota}{0}$ and $\substformn{\alpha_i}$ is obtained by
+substituting each predicate atom
 $\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)})$ in $\substformn{\alpha_{i-1}}$
 by $\stamp{\psi}{i}[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$, where
-$q(\vec{y}) \arrow{a_{i}(X)}{} \psi \in \Delta$. We write
-$\substform{\alpha}$ for the formula obtained by replacing, in
-$\substformn{\alpha}$, each occurrence of a predicate $\stamp{q}{n}$,
-such that $q \in Q \setminus F$ (resp. $q \in F$), by $\bot$
-(resp. $\top$).
+$q(\vec{y}) \arrow{a_{i}(X)}{} \psi \in \Delta$, for all $i \in
+[1,n]$. We write $\substform{\alpha}$ for the formula obtained by
+replacing, in $\substformn{\alpha}$, each occurrence of a predicate
+$\stamp{q}{n}$, such that $q \in Q \setminus F$ (resp. $q \in F$), by
+$\bot$ (resp. $\top$).
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 operations. Note that it is sufficient to show closure under
-intersection and negation because $\lang{\A_1} \cup \lang{\A_2} =
-\Sigma[X]^* (\clang{\A_1} \cap \clang{\A_2})$, for any two automata
-$\A_1$ and $\A_2$ with input event alphabet $\Sigma$ and input
-variables $X$.
+intersection and negation because $\lang{\A_1} \cup \lang{\A_2}$ is
+the complement of the language $\clang{\A_1} \cap \clang{\A_2}$, for
+any two automata $\A_1$ and $\A_2$ with the same input event alphabet
+and set of input variables.
 
@@
 \paragraph{\em Proof sketch.} (\ref{it1:thm:closure})
-The proof for $\A_\cap$ resembles the standard proof for the
-intersection of propositional alternating automata
-\cite{IosifXu18Arxiv} (\ref{it2:thm:closure}) Let $w \in \Sigma[X]^*$
-be a word. We denote by $\accformA{\A_1}{\event{w}}$ and
+The proof for $\A_\cap$ is similar to the standard proof for the
+intersection of propositional alternating
+automata. (\ref{it2:thm:closure}) Let $w \in \Sigma[X]^*$ be a
+word. We denote by $\accformA{\A_1}{\event{w}}$ and
 $\substformA{\A_1}{\event{w}}$
@@
   \begin{fact}\label{fact:dual-neg}
-    Let $\phi$ be a positive formula and let $q(t_1,\ldots,t_{\#(q)})$
-    be the only occurrence of a predicate symbol within $\phi$. Then,
-    we have the equivalence:
+    Let $\phi$ be a positive first order formula and let
+    $q(t_1,\ldots,t_{\#(q)})$ be the only occurrence of a predicate
+    symbol within $\phi$. Then, for any formula $\psi$, we have the
+    equivalence:
     \[\neg\phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]
@@
 (propositional variables) and no transition quantifiers. For
-simplicity, we do not present the details of the semi-algorithm and
-shall content ourselves of several high-level definitions. The
-interested reader is referred to \cite{IosifXu18} for a detailed
-presentation.
+simplicity, we defer the details of the semi-algorithm to Appendix
+\ref{app:algorithm} and content ourselves of several high-level
+definitions.
 

### 2019-02-07 quantifiers/body.tex
ok

@@
 (propositional variables) and no transition quantifiers. For
-simplicity, we defer the details of the semi-algorithm to Appendix
-\ref{app:algorithm} and content ourselves of several high-level
-definitions.
+simplicity, we defer the pseudocode of the semi-algorithm to Appendix
+\ref{app:algorithm} and give here several high-level definitions.
 
@@
 unfolding has been found. For a detailed presentation of the emptiness
-procedure, we refer to \cite{IosifXu18}.
+procedure, we refer to Appendix \ref{app:algorithm}.
 
@@
 Typically, when checking the unreachability of a set of program
-configurations \cite{McMillan06}, the interpolants used to annotate
-the unfolded control structure are assertions about the values of the
-program variables in a given control state, at a certain step of an
-execution. Because we consider alternating computation trees
+configurations, the interpolants used to annotate the unfolded control
+structure are assertions about the values of the program variables in
+a given control state, at a certain step of an execution
+\cite{McMillan06}. Because we consider alternating computation trees
 (forests), we must distinguish between
@@
 
-In the rest of this section, let us fix an automaton $\A =
-\tuple{\Sigma,X,Q,\iota,F,\Delta}$. Due to the above observation, none
-of the interpolants considered will be control-local and we shall use
-the term \emph{local} to denote time-local interpolants, with no free
+In the rest of this section, fix an automaton $\A =
+\tuple{\Sigma,X,Q,\iota,F,\Delta}$. Due to the above remark, none of
+the interpolants considered will be control-local and we shall use the
+term \emph{local} to denote time-local interpolants, with no free
 variables. The following definition generalizes interpolants from
-unsatisfiable conjunctions to sequences of input events:
+unsatisfiable conjunctions to input sequences:
 
@@
 unsatisfiability proof of $\substform{\alpha}$, the interpolants
-needed to annotate $\alpha$. This is because \begin{inparaenum}[(i)]
+needed to annotate $\alpha$. This is because \begin{inparaenum}[(I)]
 \item\label{it1:gli} the formula $\substform{\alpha}$, obtained by
@@
   $\substform{\alpha}$ make it difficult to write $\substform{\alpha}$
-  as an unsatisfiable conjunction of formulae from which interpolants
-  are extracted (Definition \ref{def:lyndon-interpolant}).
+  as an unsatisfiable quantifier-free conjunction of formulae from
+  which interpolants are extracted (Definition
+  \ref{def:lyndon-interpolant}).
 \end{inparaenum}
@@
 \cite[Lemma 2.1.9]{BorgerGraedelGurevich97}. \qed
-
+\noindent
 Examples of witness-producing quantifier elimination procedures can be
@@
 
-Under the assumption that witness terms can be effectively built, let
-us describe the generation of a non-local GLI for a given input event
+Under the assumption that witness terms can be effectively built, we
+describe the generation of a non-local GLI for a given input event
 sequence $\alpha = a_1 \ldots a_n$. First, we generate successively
@@
 $\quantform{\alpha}$, is unsatisfiable as well (Lemma
-\ref{lemma:subst-quant}). Because this latter formula is structured as
-a conjunction of formulae $\stamp{\iota}{0} \wedge \phi_1 \ldots
-\wedge \phi_n \wedge \psi$, where $\voc{\phi_k} \cap \stamp{Q}{\leq n}
-\subseteq \stamp{Q}{k-1} \cup \stamp{Q}{k}$ and $\voc{\psi} \cap
-\stamp{Q}{\leq n} \subseteq \stamp{Q}{n}$, it is now possible to use
-an existing interpolation procedure for the quantifier-free theory of
-$\Data$, extended with uninterpreted function symbols, to compute a
-(not necessarily local) GLI $(I_0, \ldots, I_n)$ such that $\voc{I_k}
-\cap \stamp{Q}{\leq n} \subseteq \stamp{Q}{k}$, for all $k \in [n]$.
+\ref{lemma:subst-quant} below). Because this latter formula is
+structured as a conjunction of formulae $\stamp{\iota}{0} \wedge
+\phi_1 \ldots \wedge \phi_n \wedge \psi$, where $\voc{\phi_k} \cap
+\stamp{Q}{\leq n} \subseteq \stamp{Q}{k-1} \cup \stamp{Q}{k}$ and
+$\voc{\psi} \cap \stamp{Q}{\leq n} \subseteq \stamp{Q}{n}$, it is now
+possible to use an existing interpolation procedure for the
+quantifier-free theory of $\Data$, extended with uninterpreted
+function symbols, to compute a (not necessarily local) GLI $(I_0,
+\ldots, I_n)$ such that $\voc{I_k} \cap \stamp{Q}{\leq n} \subseteq
+\stamp{Q}{k}$, for all $k \in [n]$.
 
@@
 
-In conclusion, under two assumptions about the first order theory of
-the data domain, namely the\ \begin{inparaenum}[(i)]
+Consequently, under two assumptions about the first order theory of
+the data domain, namely\ \begin{inparaenum}[(i)]
 \item witness-producing quantifier elimination, and
@@
 \end{inparaenum}
-we developped a rather generic method that produces GLIs for
-unfeasible input event sequences. Moreover, each formula $I_k$ in the
-interpolant refers only to the current predicate symbols
-$\stamp{Q}{I_k}$, the current and past input variables $\stamp{X}{\leq
-  k}$ and the existentially quantified transition variables introduced
-at the previous steps $\set{x_j \mid j < \xi_{\max}^{-1}(k), Q_j =
-  \exists}$. The remaining question is how to use such non-local GLIs
-to label the sequences in the unfolding of an automaton (Definition
-\ref{def:unfolding}) and to compute the coverage (Definition
-\ref{def:coverage}) between nodes of this unfolding.
+we developped a generic method that produces GLIs for unfeasible input
+event sequences. Moreover, each formula $I_k$ in the interpolant
+refers only to the current predicate symbols $\stamp{Q}{I_k}$, the
+current and past input variables $\stamp{X}{\leq k}$ and the
+existentially quantified transition variables introduced at the
+previous steps $\set{x_j \mid j < \xi_{\max}^{-1}(k), Q_j =
+  \exists}$. The remaining questions are how to use these GLIs to
+label the sequences in the unfolding of an automaton (Definition
+\ref{def:unfolding}) and compute coverage (Definition
+\ref{def:coverage}) between nodes of the unfolding.
 
@@
   \tuple{\Sigma,X,Q,\iota,F,\Delta}$ such that $\alpha = a_1\ldots a_n
-  \in \dom(U)$ is an input event sequence and $(I_0, \ldots, I_n)$ be
-  a GLI for $\alpha$. Then the following mapping $U' : \dom(U)
-  \rightarrow \posforms(Q,\emptyset)$ is an unfolding of $\A$, where:
+  \in \dom(U)$ and $(I_0, \ldots, I_n)$ is a GLI for $\alpha$. Then
+  the mapping $U' : \dom(U) \rightarrow \posforms(Q,\emptyset)$ is an
+  unfolding of $\A$, where:
   \begin{compactitem}
   \item $U'(\alpha_k) = U(\alpha_k) \wedge J_k$, for all $k \in [n]$,
-    where $J_k$ is the formula obtained from $I_k$ by replacing each
-    time-stamped predicate symbol $\stamp{q}{k}$ by $q$ and
-    existentially quantifying each free variable in $I_k$,
+    where $J_k$ is the formula obtained from $I_k$ by removing the
+    time stamp of each predicate symbol $\stamp{q}{k}$ and
+    existentially quantifying each free variable in $I, _k$,
   %
@@
 Basically, it is not important when exactly in the past a certain
-input has been read or when a choice has been made, as only the
+input has been read or when a choice has been made, because only the
 relation between the values of these and the current variables
-determines the future behavior of the automaton. Intuitively,
-quantifying these variables existentially does the job of ignoring
-when exactly in the past these values have been seen. Moreover, the
-last point of Lemma \ref{lemma:refinement} ensures that the refined
-path is safe in the new unfolding and thus will stay safe in all
-future unfoldings of $\A$.
+determines the future behavior of the automaton. Quantifying these
+variables existentially does the job of ignoring when exactly in the
+past these values have been seen. Moreover, the last point of Lemma
+\ref{lemma:refinement} ensures that the refined path is safe in the
+new unfolding and will stay safe in all future refinements of this
+unfolding.
 
@@
 We have implemented a version of the IMPACT semi-algorithm
-\cite{McMillan06} in a prototype tool called FOADA, which is avaliable
-online \cite{foada}. The tool is written in Java and uses the Z3 SMT
-solver \cite{z3}, via the JavaSMT interface \cite{javasmt}, for
-spuriousness and coverage queries and also for interpolant generation.
-The experiments were carried out on a MacOS x64 - 1.3 GHz Intel Core
-i5 - 8 GB 1867 MHz LPDDR3 machine.
+\cite{McMillan06} in a prototype tool, avaliable
+online\footnote{\url{http://nts.imag.fr/images/e/ea/Cav19b.tar.gz}}. The
+tool is written in Java and uses the Z3 SMT solver \cite{z3}, via the
+JavaSMT interface \cite{javasmt}, for spuriousness and coverage
+queries and also for interpolant generation. Table
+\ref{tab:experiments} reports the size of the input automaton in
+bytes, the numbers of Predicates, Variables and Transitions, the
+result of emptiness check, the number of Expanded and Visited Nodes
+during the unfolding and the Time in miliseconds.  The experiments
+were carried out on a MacOS x64 - 1.3 GHz Intel Core i5 - 8 GB 1867
+MHz LPDDR3 machine.
 
 \begin{table}[htb]
-%\vspace*{-\baselineskip}
+\vspace*{-\baselineskip}
 \begin{center}
@@
 \hline
-Example & $\len{\A}$ (bytes) & Nb of Predicates & Nb of Variables & Nb of Transitions & $L(\A)=\emptyset$ ? & Nodes Expanded & Nodes Visited & Time (ms) \\
+Example & $\len{\A}$ (bytes) & Predicates & Variables & Transitions & $L(\A)=\emptyset$ ? & Nodes Expanded & Nodes Visited & Time (msec) \\
 \hline
@@
 \end{center}
-%\vspace*{-3\baselineskip}
+\vspace*{-3\baselineskip}
 \end{table}
 
-The experimental results, reported in Table \ref{tab:experiments},
-come from several sources, namely predicate automata models (*.pa)
+The test cases shown in Table \ref{tab:experiments}, come from several
+sources, namely predicate automata models (*.pa)
 \cite{Farzan15,Farzan16} available online \cite{pa}, timed automata
@@
   rr-crossing.foada}), array logic entailments ({\tt
-  array\_rotation.ada}, {\tt array\_simple.ada}, {\tt
+  array\_rota-} {\tt tion.ada}, {\tt array\_simple.ada}, {\tt
   array\_shift.ada}) and hardware circuit verification ({\tt hw1.ada},
 {\tt hw2.ada}), initially considered in \cite{IosifRV16}. The {\tt
-  train-simpleN.} {\tt foada} and {\tt fischer-mutexN.foada} examples are
-parametric verification problems in which one checks inclusions of the
-form $\bigcap_{i=1}^N\lang{A_i} \subseteq \lang{B}$, where $A_i$ is
-the $i$-th copy of the same template automaton.
+  train-simpleN.} {\tt foada} and {\tt fischer-mutexN.foada} examples
+are parametric verification problems in which one checks inclusions of
+the form $\bigcap_{i=1}^N\lang{A_i} \subseteq \lang{B}$, where $A_i$
+is the $i$-th copy of the same template automaton.
 

### 2019-02-07 quantifiers/appendix.tex
ok

@@
-\section{Emptiness Semi-algorithm for First Order Alternating Automata}
+\section{Emptiness Checking for First Order Alternating Automata}
 \label{app:algorithm}
 
-\section{Proofs}
+The execution of Algorithm \ref{alg:impact} consists of three phases,
+corresponding to the \textsc{Close}, \textsc{Refine} and
+\textsc{Expand} of the original \impact\ procedure \cite{McMillan06}.
+
+\begin{algorithm}[htb]
+\begin{algorithmic}[0] \State {\bf input}: a first order
+  alternating automaton $\A = \tuple{\Sigma,X,Q,\iota,F,\Delta}$
+
+  \State {\bf output}: $\top$ if $L(\A)=\emptyset$, or word $w \in
+  L(\A)$, otherwise
+
+  \State {\bf data structures}: $\worklist$ and unfolding tree
+  $\mathcal{U} = \tuple{N,E,\rootNode,U,\lhd}$,
+  where: \begin{compactitem}
+  \item $N$ is a set of nodes,
+  \item $E \subseteq N \times \Sigma \times N$ is a set of edges labeled by input events,
+  \item $U : N \rightarrow \posforms(Q,\emptyset)$ is a labeling of
+    nodes with positive sentences
+  \item $\lhd \subseteq N \times N$ is a coverage relation,
+  \end{compactitem}
+  \State {\bf initially} $\worklist = \set{\rootNode}$ and $N=E=U=\lhd=\emptyset$.
+\end{algorithmic}
+\begin{algorithmic}[1]   
+  \While{$\worklist \neq \emptyset$}
+  \label{ln:impact-while}
+
+  \State dequeue $n$ from $\worklist$ 
+  \label{ln:impact-dequeue}
+
+  \State $N \leftarrow N \cup \set{n}$
+
+  \State let $\alpha(n) = a_1, \ldots, a_k$ be the input event sequence
+  on the path from $\rootNode$ to $n$
+
+  \If{$\substform{\alpha}(\stamp{X}{1}, \ldots, \stamp{X}{k})$ is satisfiable} 
+  \Comment{counterexample is feasible}
+
+  \State get model $\nu$ of $\substform{\alpha}(\stamp{X}{1}, \ldots, \stamp{X}{k})$
+
+  \State {\bf return} $w = (a_1,\nu(\stamp{X}{1})) \ldots (a_k,\nu(\stamp{X}{k}))$
+  \label{ln:impact-nonempty}
+  \Comment{$w \in L(\A)$ by construction}
+
+  \Else \Comment{spurious counterexample}
+
+  \State let $(I_0,\ldots,I_k)$ be a GLI for $\alpha$
+  \label{ln:refine-begin}
+
+  \State $b \leftarrow \bot$
+
+  \For{$i=0,\ldots,k$}
+
+  \If{$U(n_i) \not\models I_i$}
+
+  \State $\lhd \leftarrow \lhd \setminus \set{ m \in N \mid (m,n_i) \in \lhd }$
+  \Comment{uncover the nodes covered by $n_i$}
+
+  \State $U(n_i) \leftarrow U(n_i) \wedge J_i$
+  \Comment{strenghten the label of $n_i$ (Lemma \ref{lemma:refinement})}
+
+  \If{$\neg b$}
+
+  \State $b \leftarrow \Call{Close}{n_i}$
+  \label{ln:refine-end}
+
+  \EndIf % $\neg b$
+
+  \EndIf % $\Lambda(n_i) \not\models I_i$
+
+  \EndFor
+
+  \EndIf % $\mu(n)$ is satisfiable
+
+  \If{$n$ is not covered}
+
+  \For{$a \in \Sigma$} 
+  \Comment{expand $n$}
+  \label{ln:expand-begin}
+
+  \State let $s$ be a fresh node and $e = (n,a,s)$ be a new edge
+
+  \State $E \leftarrow E \cup \set{e}$
+  \label{ln:edge-insert}
+
+  \State $U \leftarrow U \cup \set{(s,\top)}$
+
+  \State enqueue $s$ into $\worklist$
+  \label{ln:expand-end}
+
+  \EndFor
+
+  \EndIf % $n$ is not covered
+
+  \EndWhile
+
+  \State {\bf return} $\top$
+\end{algorithmic}
+\caption{\impact-based Semi-algorithm for First Order Alternating Automata}
+\label{alg:impact}
+\end{algorithm}
+
+\begin{algorithm}[htb]
+\begin{algorithmic}[1]
+\Function{Close}{$x$} {\bf returns} $\Bool$
+
+\For{$y \in N$ such that $\alpha(y) \prec^* \alpha(x)$}
+
+\If{$U(x) \models U(y)$}
+
+\State $\lhd \leftarrow \left(\lhd \setminus \set{ (p,q) \in \lhd \mid
+  \text{$q$ is $x$ or a successor of $x$}}\right) \cup \set{(x,y)}$
+\label{ln:close-uncover}
+
+\State {\bf return} $\top$
+
+\EndIf
+
+\EndFor
+
+\State {\bf return} $\bot$
+
+\EndFunction
+\end{algorithmic}
+\caption{The \textsc{Close} function from Algorithm \ref{alg:impact}}
+\end{algorithm}
+
+Let $n$ be a node removed from the worklist at line
+\ref{ln:impact-dequeue} and let $\alpha(n)$ be the input sequence
+labeling the path from the root node to $n$. If
+$\substform{\alpha(n)}$ is satisfiable, the sequence $\alpha(n)$ is
+feasible, in which case a model of $\substform{\alpha(n)}$ is obtained
+and a word $w \in L(\A)$ is returned. Otherwise, $\alpha(n)$ is an
+infeasible input sequence and the procedure enters the refinement
+phase (lines \ref{ln:refine-begin}-\ref{ln:refine-end}). The GLI for
+$\alpha(n)$ is used to strenghten the labels of all the ancestors of
+$n$, by conjoining the formulae of the interpolant, changed according
+to Lemma \ref{lemma:refinement}, to the existing labels.
+
+In this process, the nodes on the path between $\rootNode$ and $n$,
+including $n$, might become eligible for coverage, therefore we
+attempt to close each ancestor of $n$ that is impacted by the
+refinement (line \ref{ln:refine-end}). Observe that, in this case the
+call to $\Call{Close}{}$ must uncover each node which is covered by a
+successor of $n$ (line \ref{ln:close-uncover} of the $\Call{Close}{}$
+function). This is required because, due to the over-approximation of
+the sets of reachable configurations, the covering relation is not
+transitive, as explained in \cite{McMillan06}. If $\Call{Close}{}$
+adds a covering edge $(n_i,m)$ to $\lhd$, it does not have to be
+called for the successors of $n_i$ on this path, which is handled via
+the boolean flag $b$. Finally, if $n$ is still uncovered (it has not
+been previously covered during the refinement phase) we expand $n$
+(lines \ref{ln:expand-begin}-\ref{ln:expand-end}) by creating a new
+node for each successor $s$ via the input event $a \in \Sigma$ and
+inserting it into the worklist.
+
+\section{Proofs of the Results in the Paper}
 \label{app:proofs}

### 2019-02-07 quantifiers/body.tex
ok

@@
 inclusion problems ({\tt abp.ada}, {\tt train.ada}, {\tt
-  rr-crossing.foada}), array logic entailments ({\tt
-  array\_rota-} {\tt tion.ada}, {\tt array\_simple.ada}, {\tt
-  array\_shift.ada}) and hardware circuit verification ({\tt hw1.ada},
-{\tt hw2.ada}), initially considered in \cite{IosifRV16}. The {\tt
-  train-simpleN.} {\tt foada} and {\tt fischer-mutexN.foada} examples
-are parametric verification problems in which one checks inclusions of
-the form $\bigcap_{i=1}^N\lang{A_i} \subseteq \lang{B}$, where $A_i$
-is the $i$-th copy of the same template automaton.
+  rr-crossing.foada}), array logic entailments ({\tt array\_rota-}
+{\tt tion.ada}, {\tt array\_simple.ada}, {\tt array\_shift.ada}) and
+hardware circuit verification ({\tt hw1.ada}, {\tt hw2.ada}),
+initially considered in \cite{IosifRV16}, with the restriction that
+local variables are made visible in the input. The {\tt
+  train-simpleN.} {\tt foada} and {\tt fischer-mutexN.} {\tt foada}
+examples are parametric verification problems in which one checks
+inclusions of the form $\bigcap_{i=1}^N\lang{A_i} \subseteq \lang{B}$,
+where $A_i$ is the $i$-th copy of the same template automaton.
 
 The advantage of using FOADA over the INCLUDER \cite{includer} tool
-from \cite{IosifRV16} is the possibility of having infinite alphabet
-automata with hidden (local) variables, whose values are not visible
+from \cite{IosifRV16} is the possibility of having automata over
+infinite alphabets with local variables, whose values are not visible
 in the input. In particular, this is essential for checking inclusion

### 2019-02-07 quantifiers/appendix.tex
ok

@@
   \end{compactitem}
-  \State {\bf initially} $\worklist = \set{\rootNode}$ and $N=E=U=\lhd=\emptyset$.
+  \State where, for a node $n$, $\alpha(n)$ is the input event
+  sequence on the path from $\rootNode$ to $n$ in $\mathcal{U}$.
+
+  \State {\bf initially} $\worklist = \set{\rootNode}$ and
+  $N=E=U=\lhd=\emptyset$.
 \end{algorithmic}
@@
 
-  \State let $\alpha(n) = a_1, \ldots, a_k$ be the input event sequence
-  on the path from $\rootNode$ to $n$
+  \State let $\alpha(n)$ be $a_1, \ldots, a_k$
 
@@
 
-\State $\lhd \leftarrow \left(\lhd \setminus \set{ (p,q) \in \lhd \mid
-  \text{$q$ is $x$ or a successor of $x$}}\right) \cup \set{(x,y)}$
+\State $\lhd \leftarrow \left[\lhd \setminus \set{ (p,q) \in \lhd \mid
+  \text{$q$ is $x$ or a successor of $x$}}\right] \cup \set{(x,y)}$
 \label{ln:close-uncover}
@@
 
-\section{Proofs of the Results in the Paper}
+\section{Proofs of the Technical Results in the Paper}
 \label{app:proofs}
+
+\subsection{Proof of Lemma \ref{lemma:path-formula}}
+
+``$\subseteq$'' Let $\I$ be a minimal interpretation such that
+$\I,\data{w} \models \pathform{\event{w}}$. We show that there exists
+an execution $\T$ of $\A$ over $w$ such that $\I = \I_\T$, by
+induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
+$\pathform{\event{w}} = \stamp{\iota}{0}$. Because $\iota$ is a
+sentence, the valuation $\data{w}$ is not important in $\I,\data{w}
+\models \stamp{\iota}{0}$ and, moreover, since $\I$ is minimal, we
+have $\I \in \minsem{\stamp{\iota}{0}}$. We define the interpretation
+$\J(q)=\I(\stamp{q}{0})$, for all $q \in Q$. Then $\cube{\J}$ is an
+execution of $\A$ over $\epsilon$ and $\I = \I_{\cube{\J}}$ is
+immediate. For the inductive case $n>0$, we assume that
+$w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\J$ be the interpretation
+defined as $\I$ for all $\stamp{q}{i}$, with $q \in Q$ and $i \in
+[n-1]$, and $\emptyset$ everywhere else. Then $\J,u_\Data \models
+\pathform{u_\Sigma}$ and $\J$ is moreover minimal. By the induction
+hypothesis, there exists an execution $\G$ of $\A$ over $u$, such that
+$\J = \I_\G$. Consider a leaf of a tree $T \in \G$, labeled with a
+configuration $q(d_1,\ldots,d_{\#(q)})$ and let $\forall y_1 \ldots
+\forall y_{\#(q)} ~.~ \stamp{q}{n-1}(\vec{y}) \rightarrow
+\stamp{\psi}{n}$ be the subformula of $\pathform{\event{w}}$
+corresponding to the application(s) of the transition rule $q(\vec{y})
+\arrow{a_n}{} \psi$ at the $(n-1)$-th step. Let $\nu = \data{w}[y_1
+  \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$. Because $\I,
+\data{w} \models \forall y_1 \ldots \forall y_{\#(q)} ~.~
+\stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$, we have $\I \in
+\sem{\stamp{\psi}{n}}_{\nu}$ and let $\K$ be one of the minimal
+interpretations such that $\K \subseteq \I$ and $\K \in
+\sem{\stamp{\psi}{n}}_{\nu}$. It is not hard to see that $\K$ exists
+and is unique, otherwise we could take the pointwise intersection of
+two or more such interpretations. We define the interpretation
+$\overline{\K}(q) = \overline{\K}(\stamp{q}{n})$ for all $q \in Q$. We
+have that $\overline{\K} \in \minsem{\psi}_{\nu}$ --- if
+$\overline{\K}$ was not minimal, $\K$ was not minimal to start with,
+contradiction. Then we extend the execution $\G$ by appending to each
+node labeled with a configuration $q(d_1, \ldots, d_{\#(q)})$ the cube
+$\cube{\overline{\K}}$. By repeating this step for all leaves of a
+tree in $\G$, we obtain an execution of $\A$ over $w$.
+
+``$\supseteq$'' Let $\T$ be an execution of $\A$ over $w$. We show
+that $\I_{\T}$ is a minimal interpretation such that $\I_{\T},
+\data{w} \models \pathform{\event{w}}$, by induction on $n \geq
+0$. For $n=0$, $\T$ is a cube from $\cube{\minsem{\iota}}$, by
+definition. Then $\I_{\T} \models \stamp{\iota}{0}$ and moreover, it
+is a minimal such interpretation. For the inductive case $n > 0$, let
+$w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\G$ be the restriction of
+$\T$ to $u$. Consequently, $\I_\G$ is the restriction of $\I_\T$ to
+$\stamp{Q}{\leq n-1}$. By the inductive hypothesis, $\I_{\G}$ is a
+minimal interpretation such that \(\I_{\G}, u_{\Data} \models
+\pathform{u_\Sigma}\). Since $\I_{\T}(\stamp{q}{n}) =
+\set{\tuple{d_1,\ldots,d_{\#(q)}} \mid q(d_1,\ldots,d_{\#(q)}) \text{
+    labels a node on the $n$-th level in $\T$}}$, we have $\I_{\T},
+\data{w} \models \varphi$, for each subformula $\varphi = \forall y_1
+\ldots \forall y_{\#(q)} ~.~ \stamp{q}{n-1}(\vec{y}) \rightarrow
+\stamp{\psi}{n}$ of $\pathform{\event{w}}$, by the execution semantics
+of $\A$. This is the case because the children of each node labeled
+with $q(d_1,\ldots,d_{\#(q)})$ on the $(n-1)$-th level of $\T$ form a
+cube from $\cube{\minsem{\psi}_\nu}$, where $\nu$ is a valuation that
+assigns each $y_i$ the value $d_i$ and behaves like $\data{w}$,
+otherwise. Now supppose, for a contradiction, that $\I_{\T}$ is not
+minimal and let $\J \subsetneq \I_{\T}$ be an interpretation such that
+$\J,\data{w} \models \pathform{\event{w}}$. First, we show that the
+restriction $\J'$ of $\J$ to $\bigcup_{i=0}^{n-1} \stamp{Q}{i}$ must
+coincide with $\I_{\G}$. Assuming this is not the case, i.e.\ $\J'
+\subsetneq \I_{\G}$, contradicts the minimality of $\I_{\G}$. Then the
+only possibility is that $\J(\stamp{q}{n}) \subsetneq
+\I_{\T}(\stamp{q}{n})$, for some $q \in Q$. Let
+$p_1(y_1,\ldots,y_{\#(p_1)}) \arrow{a_n}{} \psi_1, \ldots,
+p_k(y_1,\ldots,y_{\#(p_k)}) \arrow{a_n}{} \psi_k$ be the set of
+transition rules in which the predicate symbol $q$ occurs on the
+right-hand side. Then it must be the case that, for some node on the
+$(n-1)$-th level of $\G$, labeled with a configuration
+$p_i(d_1,\ldots,d_{\#(p_i)})$, the set of children does not form a
+minimal cube from $\cube{\minsem{\stamp{\psi_i}{n}}}$, which
+contradicts the execution semantics of $\A$. \qed
+
+\subsection{Proof of Lemma \ref{lemma:acceptance}}
+
+``$(\ref{it1:lemma:acceptance}) \Rightarrow
+(\ref{it2:lemma:acceptance})$'' Let $\I$ be an interpretation such
+that $\I,\data{w} \models \accform{\event{w}}$. By Lemma
+\ref{lemma:path-formula}, $\A$ has an execution $\T$ over $w$ such
+that $\I = \I_{\T}$. To prove that $\T$ is accepting, we show
+that \begin{inparaenum}[(i)]
+\item\label{it1:acceptance} all paths in $\T$ have length $n$ and
+  that
+\item\label{it2:acceptance} the frontier of $\T$ is labeled with final
+  configurations only. \end{inparaenum} First, assume that
+(\ref{it1:acceptance}) there exists a path in $\T$ of length $0 \leq m
+< n$. Then there exists a node on the $m$-th level, labeled with some
+configuration $q(d_1,\ldots,q_{\#(q)})$, that has no children. By the
+definition of the execution semantics of $\A$, we have
+$\cube{\minsem{\psi}_{\eta}} = \emptyset$, where $q(\vec{y})
+\arrow{a_{m+1}(X)}{} \psi$ is the transition rule of $\A$ that applies
+for $q$ and $a_{m+1}$ and $\eta= \data{w}[y_1 \leftarrow
+  d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$. Hence
+$\sem{\psi}_{\eta} = \emptyset$, and because $\I,\data{w} \models
+\accform{\alpha}$, we obtain that $\I,\eta \models q(\vec{y})
+\rightarrow \stamp{\psi}{m+1}$, thus $\tuple{d_1,\ldots,d_{\#(q)}}
+\not\in \I(q)$. However, this contradicts the fact that $\I=\I_{\T}$
+and that $q(d_1,\ldots,d_{\#(q)})$ labels a node of $\T$. Second,
+assume that (\ref{it2:acceptance}), there exists a frontier node of
+$\T$ labeled with a configuration $q(d_1,\ldots,d_{\#(q)})$ such that
+$q \in Q \setminus F$. Since $\I,\data{w} \models \forall y_1 \ldots
+\forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot$, by a similar
+reasoning as in the above case, we obtain that
+$\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$, contradiction.
+
+``$(\ref{it2:lemma:acceptance}) \Rightarrow
+(\ref{it1:lemma:acceptance})$'' Let $\T$ be an accepting execution of
+$\A$ over $w$. We prove that $\I_\T,\data{w} \models
+\accform{\event{w}}$. By Lemma \ref{lemma:path-formula}, we obtain
+$\I_\T,\data{w} \models \pathform{\event{w}}$. Since every path in
+$\T$ is of length $n$ and all nodes on the $n$-th level of $\T$ are
+labeled by final configurations, we obtain that $\I_\T,\data{w}
+\models \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
+y_{\#(q)} ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot$, trivially.
+\qed
+
+\subsection{Proof of Lemma \ref{lemma:quant}}
+
+(\ref{it1:quant}) Trivial, since every subformula
+$q(t_1,\ldots,t_{\#(q)}) \rightarrow
+\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ of $\quantform{\alpha}$ is
+entailed by a subformula $\forall y_1 \ldots \forall y_{\#(q)} ~.~
+q(y_1, \ldots, y_{\#(q)}) \rightarrow \psi$ of $\accform{\alpha}$.
+
+\noindent(\ref{it2:quant}) By repeated applications of the following
+fact:
+
+\begin{fact}
+  Given formulae $\phi$ and $\psi$, such that no predicate atom with
+  predicate symbol $q$ occurs in $\psi(y_1,\ldots,y_{\#(q)})$, for
+  each valuation $\nu$, if there exists an interpretation $\I$ such
+  that $\I,\nu \models \phi \wedge \bigwedge_{q(t_1,\ldots,t_{\#(q)})
+    \text{ occurs in } \phi} q(t_1,\ldots,t_{\#(q)}) \rightarrow
+  \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ then there exists a
+  valuation $\J$ such that $\J(q) \subseteq \I(q)$ and $\J(q') =
+  \I(q')$ for all $q' \in Q \setminus \set{q}$ and $\J,\nu \models
+  \phi \wedge \forall y_1 \ldots \forall y_{\#(q)} ~.~
+  q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$.
+\end{fact}
+\paragraph{\em Proof.}
+Assume w.l.o.g. that $\phi$ is quantifier-free. The proof can be
+easily generalized to the case $\phi$ has quantifiers. Let $\J(q) =
+\set{\langle t_1^\nu, \ldots, t_{\#(q)}^\nu \rangle \in \I(q) \mid
+  q(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \phi}$ and $\J(q') =
+\I(q')$ for all $q' \in Q \setminus \set{q}$. Since $\I,\nu \models
+\phi$, we obtain that also $\J,\nu \models \phi$ because the tuples of
+values in $\I(q) \setminus \J(q)$ are not interpretations of terms
+that occur within subformulae $q(t_1,\ldots,t_{\#(q)})$ of
+$\phi$. Moreover, $\bigwedge_{q(t_1,\ldots,t_{\#(q)}) \text{ occurs in
+  } \phi} q(t_1,\ldots,t_{\#(q)}) \rightarrow
+\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ and $\forall y_1 \ldots
+\forall y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$ are
+equivalent under $\J$, thus $\J,\nu \models \forall y_{\#(q)} ~.~
+q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$, as required. \qed
+
+\noindent This concludes the proof of Lemma \ref{lemma:quant}. \qed
+
+\subsection{Proof of Lemma \ref{lemma:subst}}
+
+By induction on $n\geq0$. The base case $n=0$ is trivial, since
+$\quantform{A} = \substform{A} = \stamp{\iota}{0}$. For the induction
+step, we rely on the following fact:
+
+ \begin{fact}\label{fact:subst}
+   Given formulae $\phi$ and $\psi$, such that $\phi$ is positive
+   $q(t_1,\ldots,t_{\#(q)})$ is the only one occurrence of the
+   predicate symbol $q$ in $\phi$ and no predicate atom with
+   predicate symbol $q$ occurs in $\psi(y_1,\ldots,y_{\#(q)})$, for
+   each interpretation $\I$ and each valuation $\nu$, we have:
+   \[\begin{array}{l} 
+   \I,\nu \models \phi \wedge q(t_1,\ldots,t_{\#(q)}) \rightarrow
+   \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}] \iff \\ \nu \models
+   \phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]\enspace.
+   \end{array}\]
+ \end{fact}
+
+ \paragraph{\em Proof.}
+ We assume w.l.o.g. that $\phi$ is quantifier-free. The proof can be
+ easily generalized to the case $\phi$ has quantifiers.
+
+ \noindent''$\Rightarrow$'' 
+    We distinguish two cases: \begin{compactitem}
+    \item if $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \in \I(q)$ then
+      $\I,\nu \models \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$. Since
+      $\phi$ is positive, replacing $q(t_1,\ldots,t_{\#(q)})$ with
+      $\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ does not change the
+      truth value of $\phi$ under $\nu$, thus $\nu \models
+      \phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$.
+      %
+    \item else, $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \not\in
+      \I(q)$, thus $\nu \models
+      \phi[\bot/q(t_1,\ldots,t_{\#(q)})]$. Since $\phi$ is positive
+      and $\bot$ entails $\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$,
+      we obtain $\nu \models
+      \phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$
+      by monotonicity. 
+    \end{compactitem}
+
+    \noindent``$\Leftarrow$'' Let $\I$ be any interpretation such that
+    $\I(q) = \set{\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \mid \nu
+      \models \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]}$. We
+    distinguish two cases: \begin{compactitem}
+    \item if $\I(q) \not \emptyset$ then $\I,\nu \models
+      q(t_1,\ldots,t_{\#(q)})$ and $\nu \models \psi[t_1/y_1, \ldots,
+        t_{\#(q)}/y_{\#(q)}]$. Thus replacing $\psi[t_1/y_1, \ldots,
+        t_{\#(q)}/y_{\#(q)}]$ by $q(t_1,\ldots,t_{\#(q)})$ does not
+      change the truth value of $\phi$ under $\I$ and $\nu$, and we
+      obtain $\I,\nu \models \phi$. Moreover, $\I,\nu \models
+      \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ implies $\I,\nu
+      \models q(t_1,\ldots,t_{\#(q)}) \rightarrow \psi[t_1/y_1,
+        \ldots, t_{\#(q)}/y_{\#(q)}]$.
+        %
+    \item else $\I(q) = \emptyset$, hence $\nu \not\models
+      \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$, thus $\nu \models
+      \phi[\bot/q(t_1,\ldots,t_{\#(q)})]$. Because $\phi$ is positive,
+      we obtain $\I,\nu \models \phi$ by monotonicity. But $\I,\nu
+      \models q(t_1,\ldots,t_{\#(q)}) \rightarrow \psi[t_1/y_1,
+        \ldots, t_{\#(q)}/y_{\#(q)}]$ trivially, because $\I,\nu
+      \not\models q(t_1,\ldots,t_{\#(q)})$. \qed
+      \end{compactitem}
+
+This concludes the proof of Lemma \ref{lemma:subst}. \qed
+
+\subsection{Proof of Lemma \ref{lemma:quant-pred-acceptance}}
+
+By Lemma \ref{lemma:acceptance}, $w \in \lang{\A}$ if and only if
+$\I,\data{w} \models \accform{\event{w}}$, for some interpretation
+$\I$. By Lemma \ref{lemma:quant}, there exists an interpretation $\I$
+such that $\I,\data{w} \models \accform{\event{w}}$ if and only if
+there exists an interpretation $\J$ such that $\J,\nu \models
+\quantform{\event{w}}$. By Lemma \ref{lemma:subst}, there exists an
+interpretation $\J$ such that $\J,\nu \models \quantform{\event{w}}$
+if and only if $\nu \models \substform{\event{w}}$. \qed
+
+\subsection{Proof of Theorem \ref{thm:soundness}}
+
+Let $U$ be a safe and complete unfolding of $\A$, such that $\dom(U)
+\neq \emptyset$. Suppose, by contradiction, that there exists a word
+$w \in \lang{A}$ and let $\alpha \isdef \event{w}$. Since $w \in
+\lang{A}$, by Lemma \ref{lemma:acceptance}, there exists an
+interpretation $\I$ such that $\I,\data{w} \models
+\accform{\alpha}$. Assume first that $\alpha \in \dom(U)$.  In this
+case, one can show, by induction on the length $n\geq0$ of $w$, that
+$\pathform{\alpha} \models \stamp{U(\alpha)}{n}$, thus $\I,\data{w}
+\models \stamp{U(\alpha)}{n}$. Since $\I,\data{w} \models
+\accform{\alpha}$, we have $\I,\data{w} \models \bigwedge_{q \in Q
+  \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
+\stamp{q}{n}(\vec{y}) \rightarrow \bot$, hence $\stamp{U(\alpha)}{n}
+\wedge \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
+y_{\#(q)} ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot$. By renaming
+$\stamp{q}{n}$ with $q$ in the previous formula, we obtain $U(\alpha)
+\wedge \forall y_1 \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow
+\bot$ is satisfiable, thus $U$ is not safe, contradiction.
+
+We proceed thus under the assumption that $\alpha \not\in
+\dom(U)$. Since $\dom(U)$ is a nonempty prefix-closed set, there
+exists a strict prefix $\alpha'$ of $\alpha$ that is a leaf of
+$\dom(U)$. Since $U$ is closed, the leaf $\alpha'$ must be covered and
+let $\alpha_1 \prefix \alpha' \prefix \alpha$ be a node such that
+$U(\alpha_1) \models U(\beta_1)$, for some uncovered node $\beta_1 \in
+\dom(U)$. Let $\gamma_1$ be the unique sequence such that
+$\alpha_1\gamma_1 = \alpha$. By Definition \ref{def:coverage}, since
+$\alpha_1 \cover \beta_1$ and $\event{w} = \alpha_1\gamma_1 \in
+\lang{\A}$, there exists a word $w_1$ and a cube $c_1 \in
+\cube{\sem{U(\alpha_1)}} \subseteq \cube{\sem{U(\beta_1)}}$, such that
+$\event{w_1} = \gamma_1$ and $\A$ accepts $w_1$ starting with
+$c_1$. If $\beta_1\gamma_1 \in \dom(U)$, we obtain a contradiction by
+a similar argument as above. Hence $\beta_1\gamma_1 \not\in \dom(U)$
+and there exists a leaf of $\dom(U)$ which is also a prefix of
+$\beta_1\gamma_1$. Since $U$ is closed, this leaf is covered by an
+uncovered node $\beta_2 \in \dom(U)$ and let $\alpha_2 \in \dom(U)$ be
+the minimal (in the prefix partial order) node such that $\beta_1
+\prefix \alpha_2 \prefix \beta_1\gamma_1$ and $\alpha_2 \cover
+\beta_2$. Let $\gamma_2$ be the unique sequence such that
+$\alpha_2\gamma_2 = \beta_1\gamma_1$. Since $\beta_1$ is uncovered, we
+have $\beta_1 \neq \alpha_2$ and thus $\len{\gamma_1} >
+\len{\gamma_2}$. By repeating the above reasoning for $\alpha_2$,
+$\beta_2$ and $\gamma_2$, we obtain an infinite sequence
+$\len{\gamma_1} > \len{\gamma_2} > \ldots$, which is again a
+contradiction. \qed
+
+\subsection{Proof of Lemma \ref{lemma:subst-quant}}
+
+(\ref{it1:subst-quant}) If $\accform{\alpha}$ is unsatisfiable, by
+Lemmas \ref{lemma:quant} and \ref{lemma:subst}, we obtain that,
+successively $\quantform{\alpha}$ and $\substform{\alpha}$ are
+unsatisfiable. Let $Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$ and
+$Q_1x_1 \ldots Q_mx_m ~.~ \overline{\Phi}$ be prenex forms for
+$\quantform{\alpha}$ and $\substform{\alpha}$, respectively. Since we
+assumed that the first order theory of the data domain has
+witness-producing quantifier elimination, using Theorem
+\ref{thm:skolem} one can effectively build witness terms $\tau_{i_1},
+\ldots, \tau_{i_\ell}$, where $\set{i_1,\ldots,i_\ell}=\set{i \in
+  [1,m] \mid Q_i=\forall}$ and: \begin{compactitem}
+\item $\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup
+  \set{x_k \mid k < i_j, Q_k = \exists}$, for all $j \in [1,\ell]$ and
+  %
+\item $\overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
+  \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.
+\end{compactitem}
+Let $\widehat{\Phi}_0$, \ldots, $\widehat{\Phi}_n$ be the sequence of
+quantifier-free formulae, defined as follows: \begin{compactitem}
+\item $\widehat{\Phi}_0$ is the matrix of some prenex form of
+  $\stamp{\iota}{0}$,
+  %
+\item for all $i=1,\ldots,n$, let $\widehat{\Phi}_i$ be the matrix of
+  some prenex form of:
+  \[\widehat{\Phi}_i \isdef \widehat{\Phi}_{i-1} \wedge
+  \underbrace{\bigwedge_{\begin{array}{l}
+        \scriptstyle{\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \widehat{\Phi}_{i-1}} \\
+        \scriptstyle{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta} 
+    \end{array}} \stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow 
+    \stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]}_{\isdef \phi_i}\]  
+\end{compactitem}
+It is easy to see that $\widehat{\Phi}$ is the matrix of some prenex
+form of:
+\[\widehat{\Phi}_n \wedge \underbrace{\bigwedge_{\begin{array}{l}
+      \scriptstyle{\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \widehat{\Phi}_n} \\
+      \scriptstyle{q \in Q \setminus F}
+  \end{array}} \stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot}_{\isdef \psi}\]
+Applying the equivalence from Fact \ref{fact:subst} in the proof of
+Lemma \ref{lemma:subst}, we obtain a sequence of quantifier-free
+formulae $\overline{\Phi}_0, \ldots, \overline{\Phi}_n$ such that
+$\overline{\Phi}_i \equiv \widehat{\Phi}_i$, for all $i \in [n]$ and
+$\overline{\Phi}$ is obtained from $\overline{\Phi}_n$ by replacing
+each occurrence of a predicate atom $q(t_1,\ldots,t_{\#(q)})$ in
+$\overline{\Phi}_n$ by $\bot$ if $q \in Q\setminus F$ and by $\top$ if
+$q \in F$. Clearly $\overline{\Phi} \equiv \widehat{\Phi}$, thus
+$\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots, \tau_{i_\ell}/x_{i_\ell}]
+\equiv \overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
+  \tau_{i_\ell}/x_{i_\ell}] \equiv \bot$.
+
+\noindent(\ref{it2:subst-quant}) With the notation introduced at point
+(\ref{it1:subst-quant}), we have $\widehat{\Phi} = \widehat{\Phi}_0
+\wedge \bigwedge_{i=1}^n \phi_i \wedge \psi$. Consider the sequence of
+witness terms $\tau_{i_1}, \ldots, \tau_{i_\ell}$, whose existence is
+proved by point (\ref{it1:subst-quant}). Because $\voc{\tau_{i_j}}
+\subseteq \stamp{X}{\leq \xi(i_j)} \cup \set{x_k \mid k < i_j, Q_k =
+  \exists}$, for all $j \in [1,\ell]$, and moreover $\xi^{-1}$ is
+strictly monotonic, we obtain: \begin{compactitem}
+\item $\voc{\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
+    \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{0} \cup
+  \stamp{X}{0} \cup \set{x_j \mid j < \xi_{\max}^{-1}(0), Q_j = \exists}$,
+  %
+\item $\voc{\phi_i[\tau_{i_1}/x_{i_1}, \ldots,
+    \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{i-1} \cup
+  \stamp{Q}{i} \cup \stamp{X}{\leq i} \cup \set{x_j \mid j <
+    \xi_{\max}^{-1}(i), Q_j = \exists}$, for all $i \in [1,n]$,
+  %
+\item $\voc{\psi[\tau_{i_1}/x_{i_1}, \ldots,
+    \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{n} \cup
+  \stamp{X}{\leq n} \cup \set{x_j \mid j \in [1,m], Q_j = \exists}$.
+\end{compactitem}
+By repeatedly applying Lyndon's Interpolation Theorem, we obtain a
+sequence of formulae $(I_0, \ldots, I_n)$ such
+that: \begin{compactitem}
+\item $\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
+  \tau_{i_\ell}/x_{i_\ell}] \models I_0$ and $\voc{I_0} \subseteq
+  \stamp{Q}{0} \cup \stamp{X}{0} \cup \set{x_j \mid j <
+    \xi_{\max}^{-1}(0), Q_j = \exists}$,
+  %
+\item $I_{k-1} \wedge \phi_i[\tau_{i_1}/x_{i_1}, \ldots,
+  \tau_{i_\ell}/x_{i_\ell}] \models I_k$ and $\voc{I_k} \subseteq
+  \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
+    \xi_{\max}^{-1}(k), Q_j = \exists}$, for all $k \in [1,n]$,
+  %
+\item $I_n \wedge \psi[\tau_{i_1}/x_{i_1}, \ldots,
+  \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.
+\end{compactitem}
+To show that $(I_0,\ldots,I_n)$ is a GLI for $a_1\ldots a_n$, it is
+sufficient to notice that \[\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
+  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
+\stamp{q}{k}(\vec{y}) \rightarrow \stamp{\psi}{k+1} \models \phi_k\]
+for all $k \in [1,n]$. Consequently, we obtain: \begin{compactitem}
+\item $\stamp{\iota}{0} \models \widehat{\Phi}_0 \models I_0$, by
+  Theorem \ref{thm:skolem},
+  %
+\item $I_{k-1} \wedge \Big(\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
+  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
+  \stamp{q}{k-1}(\vec{y}) \rightarrow \stamp{\psi}{k}\Big) \models
+  I_{k-1} \wedge \phi_k \models I_k$, and 
+  %
+\item $I_n \wedge \Big(\bigwedge_{q \in Q \setminus F} \forall y_1
+  \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot\Big)
+  \models I_n \wedge \psi \models \bot$,     
+\end{compactitem}
+as required by Definition
+\ref{def:generalized-lyndon-interpolant}. \qed

### 2019-02-07 quantifiers/body.tex
ok

@@
 in which there exists an event $a$ such that no event occurs after it
-in exactly $1$ time unit. Formally, $\lang{A} = \set{a,t \mid t \in
-  \real} \cup \set{a, t_1, \ldots a, t_n \mid t_1, \ldots, t_n \in
-  \real,~ n \geq 2,~ t_1 \geq 2}$ and $\lang{B} = \set{\sigma_1, t_1
-  \ldots \sigma_{i-1}, t_{i-1}, a, t_i, \ldots \sigma_n, t_n \mid
-  \sigma_1, \ldots, \sigma_n \in \set{a,b},~ n \geq 1,~ t_1, \ldots,
-  t_n \in \real,~ t_i-t_{i-1} \neq 1}$. It is proved\footnote{See
+in exactly $1$ time unit. Formally, $\lang{A} = \set{a, t_1, \ldots a,
+  t_n \mid t_1, \ldots, t_n \in \real,~ n > 1 \Rightarrow t_1 \geq 2}$
+and $\lang{B} = \set{\sigma_1, t_1 \ldots \sigma_{i-1}, t_{i-1}, a,
+  t_i, \ldots \sigma_n, t_n \mid \sigma_1, \ldots, \sigma_n \in
+  \set{a,b},~ n \geq 1,~ t_1, \ldots, t_n \in \real,~ 1 < i < n
+  \Rightarrow t_i-t_{i-1} \neq 1}$. It is proved\footnote{See
   \cite[Theorem 1]{AlurMadhusudan04} for a proof.}  that $\clang{B}$
@@
 \cite{McMillan06} in a prototype tool, avaliable
-online\footnote{\url{http://nts.imag.fr/images/e/ea/Cav19b.tar.gz}}. The
-tool is written in Java and uses the Z3 SMT solver \cite{z3}, via the
-JavaSMT interface \cite{javasmt}, for spuriousness and coverage
-queries and also for interpolant generation. Table
-\ref{tab:experiments} reports the size of the input automaton in
-bytes, the numbers of Predicates, Variables and Transitions, the
-result of emptiness check, the number of Expanded and Visited Nodes
-during the unfolding and the Time in miliseconds.  The experiments
-were carried out on a MacOS x64 - 1.3 GHz Intel Core i5 - 8 GB 1867
-MHz LPDDR3 machine.
+online\footnote{\url{https://gofile.io/?c=7zB3g7}}. The tool is
+written in Java and uses the Z3 SMT solver \cite{z3}, via the JavaSMT
+interface \cite{javasmt}, for spuriousness and coverage queries and
+also for interpolant generation. Table \ref{tab:experiments} reports
+the size of the input automaton in bytes, the numbers of Predicates,
+Variables and Transitions, the result of emptiness check, the number
+of Expanded and Visited Nodes during the unfolding and the Time in
+miliseconds.  The experiments were carried out on a MacOS x64 - 1.3
+GHz Intel Core i5 - 8 GB 1867 MHz LPDDR3 machine.
 

### 2019-02-07 quantifiers/body.tex
ok

@@
 in exactly $1$ time unit. Formally, $\lang{A} = \set{a, t_1, \ldots a,
-  t_n \mid t_1, \ldots, t_n \in \real,~ n > 1 \Rightarrow t_1 \geq 2}$
-and $\lang{B} = \set{\sigma_1, t_1 \ldots \sigma_{i-1}, t_{i-1}, a,
-  t_i, \ldots \sigma_n, t_n \mid \sigma_1, \ldots, \sigma_n \in
-  \set{a,b},~ n \geq 1,~ t_1, \ldots, t_n \in \real,~ 1 < i < n
-  \Rightarrow t_i-t_{i-1} \neq 1}$. It is proved\footnote{See
-  \cite[Theorem 1]{AlurMadhusudan04} for a proof.}  that $\clang{B}$
-cannot be recognized by a timed automaton, thus the inclusion cannot
-be proved in the framework of timed regular languages, as $\lang{A}
-\cap \clang{B} = \emptyset$.
+  t_n \mid t_1, \ldots, t_n \in \real,~ n > 1 \Rightarrow t_2-t_1 \geq
+  2}$ and $\lang{B} = \set{\sigma_1, t_1 \ldots \sigma_{i-1}, t_{i-1},
+  a, t_i, \sigma_{i+1}, t_{i+1}, \ldots \sigma_n, t_n \mid \sigma_1,
+  \ldots, \sigma_n \in \set{a,b},~ n \geq 1,~ t_1, \ldots, t_n \in
+  \real,~ 1 \leq i < n \Rightarrow t_{i+1}-t_i \neq 1}$. It is
+proved\footnote{See \cite[Theorem 1]{AlurMadhusudan04} for a proof.}
+that $\clang{B}$ cannot be recognized by a timed automaton, thus the
+inclusion cannot be proved in the framework of timed regular
+languages, as $\lang{A} \cap \clang{B} = \emptyset$.
 

### 2019-02-07 quantifiers/appendix.tex
ok

@@
 \ref{def:generalized-lyndon-interpolant}. \qed
+
+\subsection{Proof of Lemma \ref{lemma:refinement}}
+
+The new set of formulae $U'(\alpha_0), \ldots, U'(\alpha_n)$ complies
+with Definition \ref{def:unfolding}, because: \begin{compactitem}
+\item $U'(\alpha_0) \equiv \iota$, since, by point
+  \ref{it2:generalized-lyndon-interpolant} of Definition
+  \ref{def:generalized-lyndon-interpolant}, we have $\stamp{\iota}{0}
+  \models I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) =
+  U(\alpha_0) \wedge J_0 \equiv \iota \wedge J_0 \equiv \iota$, and
+    %
+\item by Definition \ref{def:generalized-lyndon-interpolant}
+  (\ref{it3:generalized-lyndon-interpolant}) we have, for all $k \in
+  [n-1]$:
+  \[I_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
+  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
+  \rightarrow \stamp{\psi}{k+1} \models I_{k+1}\] We write
+  $\overstamp{I}{j}_k$ for the formula in which each predicate
+  symbol $\stamp{q}{k}$ is replaced by $\stamp{q}{j}$. Then the
+  following entailment holds:
+  \[\overstamp{I}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
+  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
+  \rightarrow \stamp{\psi}{1} \models \overstamp{I}{1}_{k+1}\]
+  Because $J_k$ is obtained by removing the time stamps from the predicate symbols and  
+  existentially quantifying all the free variables of $I_k$, we also
+  obtain, applying Fact \ref{fact:ex-entail} below:
+  \[\stamp{J}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
+  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
+  \rightarrow \stamp{\psi}{1} \models \stamp{J}{1}_{k+1}\] Since $U$
+  satisfies the labeling condition of Definition \ref{def:unfolding}
+  and $U'(\alpha_k) = U(\alpha_k) \wedge J_k$, we obtain, as required:
+  \[\stamp{U'(\alpha_k)}{0} \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
+  \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
+  \rightarrow \stamp{\psi}{1} \models
+  \stamp{U'(\alpha_{k+1})}{1}\enspace.\]
+\end{compactitem}
+
+\begin{fact}\label{fact:ex-entail}
+  Given formulae $\phi(\vec{x}, \vec{y})$ and $\psi(\vec{x})$ such
+  that $\phi(\vec{x}, \vec{y}) \models \psi(\vec{x})$, we also have
+  $\exists \vec{x} ~.~ \phi(\vec{x},\vec{y}) \models \exists \vec{x}
+  ~.~ \psi(\vec{x})$. 
+\end{fact}
+\paragraph{\em Proof.} For each choice of a valuation for the existentially
+quantified variables on the left-hand side, we chose the same
+valuation for the variables on the right-hand side. \qed \qed

### 2019-02-07 quantifiers/body.tex
ok

@@
 that $\clang{B}$ cannot be recognized by a timed automaton, thus the
-inclusion cannot be proved in the framework of timed regular
-languages, as $\lang{A} \cap \clang{B} = \emptyset$.
+inclusion cannot be proved in the framework of timed automata
+\cite{AlurDill94,AlurMadhusudan04}, by showing that $\lang{A} \cap
+\clang{B} = \emptyset$.
 
@@
 and $B$ is represented by a binary predicate symbol with the same
-name. The first argument of $s_1(x,z)$ tracks the last moment in time
-when the $\mathsf{x}$ clock of $A$ was reset and $z$ is the current
-time value. The time stamp of the input event $a$ is the input
-variable $t$, which must be bigger or equal to $z$, since the lapse of
-time between two consequtive events is always positive. The current
-value of the $\mathsf{x}$ clock is $z-x$, thus the guard $\mathsf{x}
-\geq 2$ on the $s_2 \arrow{}{} s_2$ transition of $A$ from becomes
-$z-x \geq 2$ in the second rule, and the reset of $x$ on the $s_1
-\arrow{}{} s_2$ transition is encoded by passing $z$ as the first
-parameter of $s_2$, in the first rule. The initial configuration of
-$A$ [$B$] is given by the atomic propositon $s_1(0,0)$ [$q_1(0,0)$]
-and $s_2$ [$q_2$] marks the final configurations.
+name. The first argument of $s_1(x,z)$ tracks the (last) time value
+time when the $\mathsf{x}$ clock of $A$ was reset (if it is reset
+more than once) and $z$ is the time value of the last event. The time
+stamp of the current event $a$ is the input variable $t$, which must
+be bigger or equal to $z$, since the lapse of time between two
+consequtive events is always positive. The current value of the
+$\mathsf{x}$ clock is thus $z-x$ and the guard $\mathsf{x} \geq 2$ on
+the $s_2 \arrow{}{} s_2$ transition of $A$ from becomes $z-x \geq 2$
+in the second rule. The reset of $x$ on the $s_1 \arrow{}{} s_2$
+transition is encoded by passing $z$ as the first parameter of $s_2$,
+in the first rule. The initial configuration of $A$ [$B$] is
+$s_1(0,0)$ [$q_1(0,0)$] and $s_2$ [$q_2$] marks the final
+configurations.
 
@@
 emptiness of this automaton in $\sim\!0.3$ seconds, on an average
-laptop machine. The missing proofs of the technical results are given
-in Appendinx \ref{app:proofs}.
+laptop machine.
 
@@
 
-\section{Preliminaries}
+For space reasons, the proofs of the technical results are given in
+Appendinx \ref{app:proofs}.
+
+\vspace*{-.5\baselineskip}
+\subsection{Preliminaries}
 
@@
 
-\subsection{Closure Properties}
+\subsection{Boolean Closure of First Order Alternating Automata}
 \label{sec:closure}

### 2019-03-24 quantifiers/rebuttal.txt
rebuttal

We first wish to thank all reviewers for their time and great
work. Below we answer the concerns raised by Reviewers 1 and 4.

Review 1:

 1.1 Relation of FOAA with s-AFA. Symbolic (alternating) automata are
 an important extension of finite-state automata to infinite alphabets
 and lacking those references in our paper is a minus that can be
 corrected. One explanation is that we focused on infinite alphabet
 extensions that follow an idea of the seminal works on timed (TA) and
 finite-memory (register) automata (RA). Indeed, both TA and RA relate
 values occurring at different positions in the input, either by using
 clocks (with reset and difference guards) or registers (with
 rewriting and equality guards). This expressivity is paid for in that
 inclusion becomes undecidable (unless severe restrictions on the
 number of clocks/registers apply).

 Our original motivation here was not as much finding decidable
 infinite alphabet alternating automata but rather having a very
 general model, (unsurprisingly) closed under complement and
 intersection, in which all undecidable inclusion problems can be
 encoded, then devise a practical semi-algorithm for it (that works on
 reasonably complex examples). Since, to the best of our knowledge,
 there are no published semi-algorithms for TA or RA inclusion, our
 method could contribute to the state-of-the-art in these areas, as
 well as in program verification (we have considered several programs
 handling integer arrays in our examples).

 1.2 Restrictions of FOAA with decidable emptiness. As proved in [5],
 emptiness of predicate automata is undecidable even for
 quantifier-free rules with equalities and disequalities, not using
 the input variables, when predicates have arities two or more
 (equalities and monadic predicates is decidable though). In the light
 of this result (pg. 9, last paragraph), any reasonably interesting
 extension of monadic predicate automata seems to be undecidable for
 emptiness.

 We also compare with an alternating automaton model [11], in which
 states are propositional variables and comparisons are possible only
 between adjacent input values, without hidden variables. This model
 is intermediate between s-AFA and FOAA, nevertheless emptiness
 remains undecidable (reduction from 2-counter machine reachability).

Review 4:

4.1 Relation with Constrained Horn Clauses (CHC). The main difference
between the unfoldings of CHC and those of FOAA is that, in CHC, the
different branches of the unfolding tree are essentially unrelated,
whereas in FOAA, all branches of the computation tree must synchronize
on the same input word. To account for this synchronization, a CHC
encoding of FOADA must use recursive algebraic data types (lists) to
ensure that, at each step, all incoming branches consider exactly the
same list, encoding the continuation of the input from that point on
(CHC traverse the input from the end to the beginning). The encoding
of FOAA in CHC would benefit from the advantage of using off-the-shelf
CHC solvers, but, in our opinion, faces several practical problems:

I. Checking that, in each clause, all incoming branches have the same
input can be done using equalities between list arguments of the
subgoals. However, the body (left-hand side) of the clause must be
given either as a conjunction or, if quantifiers are considered, as
prenex DNF. In principle, this can be done, but the reduction is a bit
more complex than suggested and, moreover, writing each rule in DNF
can be costly.

II. Introducing a list ADT in the encoding means that the emptiness
checking semi-algorithm relies on an interpolation procedure for
ADTs. Although satisfiability checking for quantifier-free ADTs has
reached maturity and integration with many SMT solvers, interpolation
of ADT theories is at its beginnings. In fact, the 2018 technical
report [*] claims to introduce the first interpolation procedure for
ADTs:

* Hossein Hojjat, Philipp Ruemmer. Deciding and Interpolating
Algebraic Data Types by Reduction. ArXiv:1801.02367. Previous version
appeared in SYNASC'17.

Moreover, the availability of such procedures in off-the-shelf CHC
solvers is still an issue --- the integration of the above procedure
in ELDARICA is underway, according to [*]. However, when such
technical problems are solved, it would definitely be interesting to
have an experimental comparison between, e.g. FOAA and ELDARICA.

III. Working with quantified rules in FOADA requires a quantifier
elimination procedure that produces witness terms. According to [15],
witness-producing quantifier elimination is available for first-order
theories and BAPA. However, if quantified variables occur in the ADT
constructors, the theory is most likely undecidable, thus may not have
quantifier elimination, by an argument similar to the one of [8].

Essentially, for the above reasons, we have chosen to design our
emptiness checking semi-algorithm from scratch, trying to put as few
assumptions as possible on the underlying first-order data theory,
namely (a) Lyndon interpolation for the quantifier-free fragment and
(b) witness term generation for quantifier elimination.

### 2019-03-24 quantifiers/rebuttal.txt
ok

@@
 
- 1.1 Relation of FOAA with s-AFA. Symbolic (alternating) automata are
- an important extension of finite-state automata to infinite alphabets
- and lacking those references in our paper is a minus that can be
+ 1.1 Relation of FOAA with s-AFA.
+
+ Symbolic (alternating) finite-state automata are an important
+ extension of finite-state automata to infinite alphabets and lacking
+ those references in our paper is a minus that should be
  corrected. One explanation is that we focused on infinite alphabet
- extensions that follow an idea of the seminal works on timed (TA) and
- finite-memory (register) automata (RA). Indeed, both TA and RA relate
+ extensions that follow an idea of the seminal papers on timed (TA)
+ [1] and finite-memory (register) automata (RA) [13]. Unlike symbolic
+ automata that test only the current input, both TA and RA relate
  values occurring at different positions in the input, either by using
  clocks (with reset and difference guards) or registers (with
- rewriting and equality guards). This expressivity is paid for in that
- inclusion becomes undecidable (unless severe restrictions on the
- number of clocks/registers apply).
+ over-writing and equality guards). The price to be paid for this
+ expressivity is that language inclusion becomes undecidable (unless
+ severe restrictions on the number of clocks/registers apply).
 
  Our original motivation here was not as much finding decidable
- infinite alphabet alternating automata but rather having a very
- general model, (unsurprisingly) closed under complement and
- intersection, in which all undecidable inclusion problems can be
- encoded, then devise a practical semi-algorithm for it (that works on
- reasonably complex examples). Since, to the best of our knowledge,
- there are no published semi-algorithms for TA or RA inclusion, our
- method could contribute to the state-of-the-art in these areas, as
- well as in program verification (we have considered several programs
+ infinite alphabet models but rather having a very general model,
+ (unsurprisingly) closed under complement, union and intersection, in
+ which all undecidable inclusion problems can be encoded, then devise
+ a practical semi-algorithm for it (that works on reasonably complex
+ examples). Since, to the best of our knowledge, there are no
+ published semi-algorithms for TA or RA inclusion (tools such as UPAAL
+ check only for emptiness), our method could contribute to these areas
+ and to program verification (we have considered several programs
  handling integer arrays in our examples).
 
- 1.2 Restrictions of FOAA with decidable emptiness. As proved in [5],
- emptiness of predicate automata is undecidable even for
- quantifier-free rules with equalities and disequalities, not using
- the input variables, when predicates have arities two or more
- (equalities and monadic predicates is decidable though). In the light
- of this result (pg. 9, last paragraph), any reasonably interesting
+ 1.2 Restrictions of FOAA with decidable emptiness problems.
+
+ As proved in [5], emptiness of predicate automata is undecidable even
+ for quantifier-free rules with equalities and disequalities, not
+ using the input variables, when predicates have arities of two or
+ more (having only equalities and monadic predicates yields a
+ decidable model though). In the light of this result (mentioned on
+ pg. 9, last paragraph in the paper), any reasonably interesting
  extension of monadic predicate automata seems to be undecidable for
@@
  is intermediate between s-AFA and FOAA, nevertheless emptiness
- remains undecidable (reduction from 2-counter machine reachability).
+ remains undecidable, by reduction from 2-counter machine
+ reachability.
 
@@
 
-4.1 Relation with Constrained Horn Clauses (CHC). The main difference
-between the unfoldings of CHC and those of FOAA is that, in CHC, the
-different branches of the unfolding tree are essentially unrelated,
-whereas in FOAA, all branches of the computation tree must synchronize
-on the same input word. To account for this synchronization, a CHC
-encoding of FOADA must use recursive algebraic data types (lists) to
-ensure that, at each step, all incoming branches consider exactly the
-same list, encoding the continuation of the input from that point on
-(CHC traverse the input from the end to the beginning). The encoding
-of FOAA in CHC would benefit from the advantage of using off-the-shelf
-CHC solvers, but, in our opinion, faces several practical problems:
+4.1 Relation with Constrained Horn Clauses (CHC).
 
-I. Checking that, in each clause, all incoming branches have the same
-input can be done using equalities between list arguments of the
+ The main difference between the unfoldings of CHC and those of FOAA
+ is that, in CHC, the different branches of the unfolding tree are
+ essentially unrelated, whereas in FOAA, all branches of the
+ computation tree must synchronize on the same input word. To account
+ for this synchronization, a CHC encoding of FOADA must use recursive
+ algebraic data types (ADT), namely lists, to ensure that, at each
+ step, all incoming branches consider exactly the same list, encoding
+ the continuation of the input from that point on (CHC traverse the
+ input from the end to the beginning). An encoding of FOAA in CHC
+ would benefit from the advantage of using off-the-shelf CHC solvers,
+ but, in our opinion, would face several problems (listed in
+ increasing order of difficulty):
+
+I. Checking that, in each clause, all incoming branches read the same
+input can be done using equalities between the list arguments of the
 subgoals. However, the body (left-hand side) of the clause must be
 given either as a conjunction or, if quantifiers are considered, as
-prenex DNF. In principle, this can be done, but the reduction is a bit
-more complex than suggested and, moreover, writing each rule in DNF
-can be costly.
+prenex DNF. This issue can be solved, but the reduction is a bit more
+complex than suggested and, moreover, writing each rule in DNF can be
+costly.
 
-II. Introducing a list ADT in the encoding means that the emptiness
-checking semi-algorithm relies on an interpolation procedure for
-ADTs. Although satisfiability checking for quantifier-free ADTs has
-reached maturity and integration with many SMT solvers, interpolation
-of ADT theories is at its beginnings. In fact, the 2018 technical
-report [*] claims to introduce the first interpolation procedure for
-ADTs:
+II. Introducing a list ADT in the encoding requires the use of an
+ interpolation procedure for ADTs in the emptiness
+ semi-algorithm. Although satisfiability of quantifier-free ADTs is
+ mature and integrated with many SMT solvers, interpolation of ADT
+ theories is at its beginnings. In fact, the technical report [*]
+ claims to introduce the first interpolation procedure for ADTs:
 
 * Hossein Hojjat, Philipp Ruemmer. Deciding and Interpolating
-Algebraic Data Types by Reduction. ArXiv:1801.02367. Previous version
-appeared in SYNASC'17.
+Algebraic Data Types by Reduction. ArXiv:1801.02367. Proc. of 
+SYNASC'17.
 
 Moreover, the availability of such procedures in off-the-shelf CHC
-solvers is still an issue --- the integration of the above procedure
-in ELDARICA is underway, according to [*]. However, when such
-technical problems are solved, it would definitely be interesting to
-have an experimental comparison between, e.g. FOAA and ELDARICA.
+solvers is still an unresolved issue. For instance, the integration of
+the above procedure in ELDARICA is underway, according to
+[*]. However, when such technical problems will be solved, it would
+definitely be interesting to have an experimental comparison between,
+e.g. FOAA and ELDARICA.
 
 III. Working with quantified rules in FOADA requires a quantifier
-elimination procedure that produces witness terms. According to [15],
-witness-producing quantifier elimination is available for first-order
-theories and BAPA. However, if quantified variables occur in the ADT
-constructors, the theory is most likely undecidable, thus may not have
-quantifier elimination, by an argument similar to the one of [8].
+elimination procedure that produces witness terms. However, if
+quantified variables occur in the ADT constructors, the theory is most
+likely undecidable, thus cannot have quantifier elimination, by an
+argument similar to the one of [8].
 
 Essentially, for the above reasons, we have chosen to design our
-emptiness checking semi-algorithm from scratch, trying to put as few
-assumptions as possible on the underlying first-order data theory,
-namely (a) Lyndon interpolation for the quantifier-free fragment and
-(b) witness term generation for quantifier elimination.
+emptiness checking semi-algorithm from scratch, trying to use as few
+assumptions as possible on the underlying first-order data theory. In
+fact, we only assume (a) Lyndon interpolation for quantifier-free
+formulae with uninterpreted predicates and (b) witness term generation
+during quantifier elimination.

### 2019-05-21 quantifiers/cav2019/abstract.tex
added final version

--- 2019-02-07 quantifiers/abstract.tex
@@
 \begin{abstract}
-We introduce first order alternating automata, a generalization of
+We introduce first-order alternating automata, a generalization of
 boolean alternating automata, in which transition rules are described
-by multisorted first order formulae, with states and internal
+by multisorted first-order formulae, with states and internal
 variables given by uninterpreted predicate terms. The model is closed

### 2019-05-21 quantifiers/cav2019/body.tex
added final version

--- 2019-02-07 quantifiers/body.tex
@@
 between the nodes of the search tree may ensure convergence of many
-real-life examples. However, applying lazy annotation to first order
+real-life examples. However, applying lazy annotation to first-order
 alternating automata faces two nontrivial
@@
   \cite{Halpern91}. To deal with this problem, we assume that the
-  first order data theory, without uninterpreted predicate symbols,
+  first-order data theory, without uninterpreted predicate symbols,
   has a quantifier elimination procedure, that instantiates quantifers
@@
 
-\paragraph{\em Applications}
-We use first order alternating automata to develop practical
+We use first-order alternating automata to develop practical
 semi-algorithms for a number of known undecidable problems, such as:
@@
 
-Consider, for instance the problem of proving that $\lang{A} \subseteq
-\lang{B}$, where $A$ and $B$ are the timed automata in
-Fig. \ref{fig:ta}. The automaton $A$ ($B$) has a single clock $x$
-($y$) which is reset on the transition from $s_1$ to $s_2$ ($q_1$ to
-$q_2$). Intuititvely, $A$ recognizes the sequences in which all events
-are $a$ and the second event, if it occurs at all, it occurs after at
-least $2$ time units. On the other hand, $B$ recognizes the sequences
-in which there exists an event $a$ such that no event occurs after it
-in exactly $1$ time unit. Formally, $\lang{A} = \set{a, t_1, \ldots a,
-  t_n \mid t_1, \ldots, t_n \in \real,~ n > 1 \Rightarrow t_2-t_1 \geq
-  2}$ and $\lang{B} = \set{\sigma_1, t_1 \ldots \sigma_{i-1}, t_{i-1},
-  a, t_i, \sigma_{i+1}, t_{i+1}, \ldots \sigma_n, t_n \mid \sigma_1,
-  \ldots, \sigma_n \in \set{a,b},~ n \geq 1,~ t_1, \ldots, t_n \in
-  \real,~ 1 \leq i < n \Rightarrow t_{i+1}-t_i \neq 1}$. It is
-proved\footnote{See \cite[Theorem 1]{AlurMadhusudan04} for a proof.}
-that $\clang{B}$ cannot be recognized by a timed automaton, thus the
-inclusion cannot be proved in the framework of timed automata
-\cite{AlurDill94,AlurMadhusudan04}, by showing that $\lang{A} \cap
-\clang{B} = \emptyset$.
-
-Our approach is to encode timed automata as first order alternating
-automata and use closure of the latter class under intersection and
-complement to compute an alternating automaton, whose emptiness proves
-that $\lang{A} \subseteq \lang{B}$. Consider the encoding of $A$ in
-Fig. \ref{fig:ta} (right) inspired by a translation of timed automata
-into Datalog programs \cite{Fribourg98}.  Each control state of $A$
-and $B$ is represented by a binary predicate symbol with the same
-name. The first argument of $s_1(x,z)$ tracks the (last) time value
-time when the $\mathsf{x}$ clock of $A$ was reset (if it is reset
-more than once) and $z$ is the time value of the last event. The time
-stamp of the current event $a$ is the input variable $t$, which must
-be bigger or equal to $z$, since the lapse of time between two
-consequtive events is always positive. The current value of the
-$\mathsf{x}$ clock is thus $z-x$ and the guard $\mathsf{x} \geq 2$ on
-the $s_2 \arrow{}{} s_2$ transition of $A$ from becomes $z-x \geq 2$
-in the second rule. The reset of $x$ on the $s_1 \arrow{}{} s_2$
-transition is encoded by passing $z$ as the first parameter of $s_2$,
-in the first rule. The initial configuration of $A$ [$B$] is
-$s_1(0,0)$ [$q_1(0,0)$] and $s_2$ [$q_2$] marks the final
-configurations.
-
-\begin{figure}[htb]
-\vspace*{-2\baselineskip}
-\begin{center}
-\begin{minipage}{5cm}
-\input ta.pdf_t
-\end{minipage} 
-\begin{minipage}{5cm}
-\footnotesize{\[A \left\{\begin{array}{rcl}
-s_1(x,z) & \arrow{a(t)}{} & t \geq z \wedge s_2(z,t) \\
-s_2(x,z) & \arrow{a(t)}{} & t \geq z \wedge z-x \geq 2 \wedge s_2(x,t)
-\end{array}\right.\] 
-\[B \left\{\begin{array}{rcl}
-q_1(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge q_1(y,t) \\
-q_1(y,u) & \arrow{a(t)}{} & t \geq u \wedge q_2(u,t) \\
-q_2(y,u) & \arrow{a(t),b(t)}{} & t \geq u \wedge u - y \neq 1 \wedge q_2(y,t)
-\end{array}\right.\]
-\[\overline{B}\left\{\begin{array}{rcl}
-q_1(y,u) & \arrow{a(t),b(t)}{} & t < u \vee q_1(y,t) \\
-q_1(y,u) & \arrow{a(t)}{} & t < u \vee q_2(u,t) \\
-q_2(y,u) & \arrow{a(t),b(t)}{} & t < u \vee u - y = 1 \vee q_2(y,t)
-\end{array}\right.\]}
-\end{minipage}
-\end{center}
-\vspace*{-\baselineskip}
-\caption{First Order Alternating Automata Checking Inclusion of Timed Automata}
-\label{fig:ta}
-\vspace*{-\baselineskip}
-\end{figure}
-
-The first order alternating automaton recognizing the (encoding of
-the) $\clang{B}$ language is simply obtained by negating the theory
-atoms $t \geq u$ and $u - y \neq 1$ and interchanging conjunctions
-with disjunctions in the rules of the alternating automaton encoding
-$B$, denoted as $\overline{B}$ in Fig. \ref{fig:ta} (bottom right),
-with final configurations marked by $q_1$, instead of $q_2$. Finally,
-we check the emptiness of the alternating automaton consisting of the
-rules of $A$ and $\overline{B}$, started in the initial configuration
-$s_1(0,0) \wedge q_1(0,0)$. Our lazy annotation semi-algorithm proves
-emptiness of this automaton in $\sim\!0.3$ seconds, on an average
-laptop machine.
-
 \paragraph{\em Related Work}
-The first order alternating automata model presented in this paper
-extends the alternating automata with variables ranging over infinite
-data considered in \cite{IosifXu18}. There all variables were required
-to be observable in the input. We overcome this restriction by
-allowing internal (invisible) variables. Another closely related work
+Recognizers for languages over infinite alphabets have found various
+applications, ranging from Unicode text recognition
+\cite{DAntoniKincaidWang18} to runtime program monitoring
+\cite{BarringerRydeheardHavelund07}. Extending finite automata to
+infinite alphabets has been considered in the context of
+\emph{symbolic alternating finite automata} (s-AFA), whose transitions
+are labeled with guards taken from a decidable theory of the data
+domain \cite{DAntoniKincaidWang18}. As in our model, s-AFA are closed
+under union, intersection and complement and emptiness is decidable,
+due to the lack of registers. However, s-AFA are strictly less
+expressive than our model, because comparing data at different
+positions in the input word is not possible.
+
+\emph{Constrained Horn clauses} (CHC) are a branching computation
+model widespread in program verification \cite{Grebenshchikov12}. The
+main difference between alternating and bottom-up branching
+computations is that, in an alternating model, all branches of the
+computation must synchronize on the same input word. With this in
+mind, it is possible to express emptiness of first-order alternating
+automata as the existence of solutions of a CHC over a higher-order
+theory of data, extended with algebraic data types (lists). The
+effectiveness of such an encoding depends on the effectiveness of
+interpolation and witness term generation for theories of algebraic
+data types \cite{HojjatRuemmer18}.
+
+The alternating automata model presented in this paper extends the
+alternating automata with variables ranging over infinite data
+considered in \cite{IosifXu18}. There all variables were required to
+be observable in the input. We overcome this restriction by allowing
+internal (invisible) variables. Another closely related work
 \cite{IosifRV16} considers an inclusion between an asynchronous
@@
 $B$ must be declared in the automata $A_1, \ldots, A_n$ under
-check. First order alternating automata bypass this limitation, since
-our emptiness checking semi-algorithm can handle invisible variables.
-
-The work probably closest to ours concerns the model of predicate
-automata (PA) \cite{Farzan15,Farzan16,KincaidPhD}, used in the
-verification of parameterized concurrent programs with shared
-memory. In this model, the alphabet consists of pairs of program
-statements and thread identifiers, thus being infinite because the
-number of threads is unbounded. Since thread identifiers can only be
-compared for (dis-)equality, the data theory in PA is the theory of
-equality. Even with this simplification, the emptiness problem is
+check. This limitation can now be bypassed, since the inclusion
+problem can be encoded as emptiness of a first-order alternating
+automaton and, moreover, the emptiness checking semi-algorithm can
+handle invisible variables.
+
+The work probably closest to ours concerns the model of
+\emph{predicate automata} (PA) \cite{Farzan15,Farzan16,KincaidPhD},
+used in the verification of parameterized concurrent programs with
+shared memory. In this model, the alphabet consists of pairs of
+program statements and thread identifiers and is considered infinite
+because the number of threads is unbounded. Since thread identifiers
+can only be compared for equality, the data theory in PA is the theory
+of equality. Even with this simplification, the emptiness problem is
 undecidable when either the predicates have arity greater than one
@@
 possible semi-algorithmically, by explicitly enumerating reachable
-configurations and checking coverage by looking for permuations of
-argument values. However, no semi-algorithm is given for quantified
-PA.  Dealing with quantified transition rules is one of our
+configurations and checking coverage by looking for permutations of
+argument values. However, no semi-algorithm has been given for
+quantified PA. Dealing with quantified transition rules is one of our
 contributions.
 
-For space reasons, the proofs of the technical results are given in
-Appendinx \ref{app:proofs}.
-
-\vspace*{-.5\baselineskip}
 \subsection{Preliminaries}
@@
 to several predicates over $\Data$, such as the inequality relation
-over $\zed$, and write $\preds$ for the set of \emph{uninterpreted
-  predicates}.
+over $\zed$, and write $\preds$ for the set of remaining
+\emph{uninterpreted predicates}.
 
@@
 $\Data$, variables or function applications $f(t_1,\ldots,t_{\#(f)})$,
-where $t_1,\ldots,t_{\#(f)}$ are terms. The set of first order
+where $t_1,\ldots,t_{\#(f)}$ are terms. The set of first-order
 formulae is defined by the syntax below:
@@
 Q_nx_n ~.~ \phi$, where $\phi$ has no quantifiers. In this case we
-call $\phi$ the \emph{matrix} of $\varphi$. Every first order formula
+call $\phi$ the \emph{matrix} of $\varphi$. Every first-order formula
 can be written in prenex form, by renaming each quantified variable to
@@
 $\phi$, we define the forcing relation $\I,\nu \models \phi$
-recursively on the structure of $\phi$, as usual.  
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\ifLongVersion
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\[
-\begin{array}{rcl}
-\I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
-\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I \\
-\I,\nu \models \neg\phi_1 & \iff & \I,\nu \not\models \phi_1 \\
-\I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
-\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data 
-\end{array}
-\]
-where $\nu[x\leftarrow d]$ is the valuation which assigns $d$ to $x$
-and behaves like $\nu$ elsewhere.  
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\fi 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-For a formula $\phi$ and a valuation $\nu$, we define $\sem{\phi}_\nu
-\isdef \set{\I \mid \I,\nu \models \phi}$ and drop the $\nu$ subscript
-for sentences. A sentence $\phi$ is \emph{satisfiable} if
+recursively on the structure of $\phi$, as usual. For a formula
+$\phi$ and a valuation $\nu$, we define $\sem{\phi}_\nu \isdef \set{\I
+  \mid \I,\nu \models \phi}$ and drop the $\nu$ subscript for
+sentences. A sentence $\phi$ is \emph{satisfiable} if
 $\sem{\phi}\neq\emptyset$. An element of $\sem{\phi}$ is called a
@@
 
-A \emph{first order alternating automaton} is a tuple $\A =
+A \emph{first-order alternating automaton} is a tuple $\A =
 \tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is a finite set of
@@
 \posforms(Q,\emptyset)$ is a sentence defining initial configurations,
-$F \subseteq Q$ is the set of predicates denoting final states, and
+$F \subseteq Q$ is the set of predicates denoting final states and
 $\Delta$ is a set of \emph{transition rules}. A transition rule is of
@@
 
-The semantics of first order alternating automata is analogous to the
+The semantics of first-order alternating automata is analogous to the
 semantics of propositional alternating automata, with rules of the
@@
 considers boolean valuations instead of formulae. In contrast, a
-finite description of a first order alternating automaton cannot be
-given in terms of interpretations, as a first order formula may have
+finite description of a first-order alternating automaton cannot be
+given in terms of interpretations, as a first-order formula may have
 infinitely many models, corresponding to infinitely many initial or
@@
 \cube{\minsem{\iota}}$, then $\A$ \emph{accepts} $w$ and let
-$\lang{\A}$ be the set of words accepted by $\A$.
-
-\begin{example}\label{ex:foada}
-  Consider the automaton $\A = \tuple{\set{a}, \set{x},
+$\lang{\A}$ be the set of words accepted by $\A$. For example,
+consider the automaton $\A = \tuple{\set{a}, \set{x},
   \set{q_0,q_1,q_2,q_f}, q_0(0), \set{q_f}, \Delta}$, where $\Delta$
@@
 \(q_1(y) \arrow{a(x)}{} q_1(y+x) \vee (y > 0 \wedge q_f)\) and
-  \(q_2(y) \arrow{a(x)}{} q_2(y-x) \vee (y > 0 \wedge q_f)\). A
-  possible execution tree of this automaton is the following:
+\(q_2(y) \arrow{a(x)}{} q_2(y-x) \vee (y > 0 \wedge q_f)\). A possible
+execution tree of this automaton is the following:
 \begin{center}
@@
 have $\lang{\A} = \emptyset$, which is proved by our tool in
-  $\sim\!0.5$ seconds on an average machine. \hfill$\blacksquare$
-\end{example}
-
+$\sim\!0.5$ seconds on an average machine.
+
+\noindent
 In the rest of this paper, we are concerned with the following
@@
 solving this problem first, we develop the machinery required to prove
-that first order alternating automata are closed under complement and,
+that first-order alternating automata are closed under complement and,
 further, set up the ground for developping a practical semi-algorithm
@@
 \end{lemma}
-\proof{ ``$\subseteq$'' Let $\I$ be a minimal interpretation such that
-  $\I,\data{w} \models \pathform{\event{w}}$. We show that there exists
-  an execution $\T$ of $\A$ over $w$ such that $\I = \I_\T$, by
-  induction on $n\geq 0$. For $n=0$, we have $w=\epsilon$ and
-  $\pathform{\event{w}} = \stamp{\iota}{0}$. Because $\iota$ is a
-  sentence, the valuation $\data{w}$ is not important in $\I,\data{w}
-  \models \stamp{\iota}{0}$ and, moreover, since $\I$ is minimal, we
-  have $\I \in \minsem{\stamp{\iota}{0}}$. We define the
-  interpretation $\J(q)=\I(\stamp{q}{0})$, for all $q \in Q$. Then
-  $\cube{\J}$ is an execution of $\A$ over $\epsilon$ and $\I =
-  \I_{\cube{\J}}$ is immediate. For the inductive case $n>0$, we
-  assume that $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\J$ be the
-  interpretation defined as $\I$ for all $\stamp{q}{i}$, with $q \in
-  Q$ and $i \in [n-1]$, and $\emptyset$ everywhere else. Then
-  $\J,u_\Data \models \pathform{u_\Sigma}$ and $\J$ is moreover
-  minimal. By the induction hypothesis, there exists an execution $\G$
-  of $\A$ over $u$, such that $\J = \I_\G$. Consider a leaf of a tree
-  $T \in \G$, labeled with a configuration $q(d_1,\ldots,d_{\#(q)})$
-  and let $\forall y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$ be the
-  subformula of $\pathform{\event{w}}$ corresponding to the
-  application(s) of the transition rule $q(\vec{y}) \arrow{a_n}{}
-  \psi$ at the $(n-1)$-th step. Let $\nu = \data{w}[y_1 \leftarrow
-    d_1,\ldots,y_{\#(q)} \leftarrow d_{\#(q)}]$. Because $\I, \data{w}
-  \models \forall y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{n-1}(\vec{y}) \rightarrow \stamp{\psi}{n}$, we have $\I
-  \in \sem{\stamp{\psi}{n}}_{\nu}$ and let $\K$ be one of the minimal
-  interpretations such that $\K \subseteq \I$ and $\K \in
-  \sem{\stamp{\psi}{n}}_{\nu}$. It is not hard to see that $\K$ exists
-  and is unique, otherwise we could take the pointwise intersection of
-  two or more such interpretations. We define the interpretation
-  $\overline{\K}(q) = \overline{\K}(\stamp{q}{n})$ for all $q \in
-  Q$. We have that $\overline{\K} \in \minsem{\psi}_{\nu}$ --- if
-  $\overline{\K}$ was not minimal, $\K$ was not minimal to start with,
-  contradiction. Then we extend the execution $\G$ by appending to
-  each node labeled with a configuration $q(d_1, \ldots, d_{\#(q)})$
-  the cube $\cube{\overline{\K}}$. By repeating this step for all
-  leaves of a tree in $\G$, we obtain an execution of $\A$ over $w$.
-
-  ``$\supseteq$'' Let $\T$ be an execution of $\A$ over $w$. We show
-  that $\I_{\T}$ is a minimal interpretation such that $\I_{\T},
-  \data{w} \models \pathform{\event{w}}$, by induction on $n \geq
-  0$. For $n=0$, $\T$ is a cube from $\cube{\minsem{\iota}}$, by
-  definition. Then $\I_{\T} \models \stamp{\iota}{0}$ and moreover, it
-  is a minimal such interpretation. For the inductive case $n > 0$,
-  let $w=u\cdot(a_n,\nu_n)$ for a word $u$. Let $\G$ be the
-  restriction of $\T$ to $u$. Consequently, $\I_\G$ is the restriction
-  of $\I_\T$ to $\stamp{Q}{\leq n-1}$. By the inductive hypothesis,
-  $\I_{\G}$ is a minimal interpretation such that \(\I_{\G}, u_{\Data}
-  \models \pathform{u_\Sigma}\). Since $\I_{\T}(\stamp{q}{n}) =
-  \set{\tuple{d_1,\ldots,d_{\#(q)}} \mid q(d_1,\ldots,d_{\#(q)})
-    \text{ labels a node on the $n$-th level in $\T$}}$, we have $\I_{\T},
-  \data{w} \models \varphi$, for each subformula $\varphi = \forall y_1
-  \ldots \forall y_{\#(q)} ~.~ \stamp{q}{n-1}(\vec{y})
-  \rightarrow \stamp{\psi}{n}$ of $\pathform{\event{w}}$, by the
-  execution semantics of $\A$. This is the case because the children
-  of each node labeled with $q(d_1,\ldots,d_{\#(q)})$ on the
-  $(n-1)$-th level of $\T$ form a cube from
-  $\cube{\minsem{\psi}_\nu}$, where $\nu$ is a valuation that assigns
-  each $y_i$ the value $d_i$ and behaves like $\data{w}$,
-  otherwise. Now supppose, for a contradiction, that $\I_{\T}$ is not
-  minimal and let $\J \subsetneq \I_{\T}$ be an interpretation such
-  that $\J,\data{w} \models \pathform{\event{w}}$. First, we show that
-  the restriction $\J'$ of $\J$ to $\bigcup_{i=0}^{n-1} \stamp{Q}{i}$
-  must coincide with $\I_{\G}$. Assuming this is not the case,
-  i.e.\ $\J' \subsetneq \I_{\G}$, contradicts the minimality of
-  $\I_{\G}$. Then the only possibility is that $\J(\stamp{q}{n})
-  \subsetneq \I_{\T}(\stamp{q}{n})$, for some $q \in Q$. Let
-  $p_1(y_1,\ldots,y_{\#(p_1)}) \arrow{a_n}{} \psi_1, \ldots,
-  p_k(y_1,\ldots,y_{\#(p_k)}) \arrow{a_n}{} \psi_k$ be the set of
-  transition rules in which the predicate symbol $q$ occurs on the
-  right-hand side. Then it must be the case that, for some node on the
-  $(n-1)$-th level of $\G$, labeled with a configuration
-  $p_i(d_1,\ldots,d_{\#(p_i)})$, the set of children does not form a
-  minimal cube from $\cube{\minsem{\stamp{\psi_i}{n}}}$, which
-  contradicts the execution semantics of $\A$. \qed}
 
@@
   every word $w \in \Sigma[X]^*$, the following are
-  equivalent: \begin{compactenum}
+  equivalent: \begin{inparaenum}[(1)]
   \item\label{it1:lemma:acceptance} there exists an interpretation
-    $\I$ such that $\I,\data{w} \models \accform{\event{w}}$, 
+    $\I$ such that $\I,\data{w} \models \accform{\event{w}}$ and
     %
   \item\label{it2:lemma:acceptance} $w \in \lang{\A}$. 
-  \end{compactenum}
+  \end{inparaenum}
 \end{lemma}
-\proof{ ``$(\ref{it1:lemma:acceptance}) \Rightarrow
-  (\ref{it2:lemma:acceptance})$'' Let $\I$ be an interpretation such
-  that $\I,\data{w} \models \accform{\event{w}}$. By Lemma
-  \ref{lemma:path-formula}, $\A$ has an execution $\T$ over $w$ such
-  that $\I = \I_{\T}$. To prove that $\T$ is accepting, we show
-  that \begin{inparaenum}[(i)]
-    \item\label{it1:acceptance} all paths in $\T$ have length $n$ and
-      that
-    \item\label{it2:acceptance} the frontier of $\T$ is labeled with
-      final configurations only. \end{inparaenum} First, assume that
-  (\ref{it1:acceptance}) there exists a path in $\T$ of length $0 \leq
-  m < n$. Then there exists a node on the $m$-th level, labeled with
-  some configuration $q(d_1,\ldots,q_{\#(q)})$, that has no
-  children. By the definition of the execution semantics of $\A$, we
-  have $\cube{\minsem{\psi}_{\eta}} = \emptyset$, where
-  $q(\vec{y}) \arrow{a_{m+1}(X)}{} \psi$ is the transition
-  rule of $\A$ that applies for $q$ and $a_{m+1}$ and $\eta=
-  \data{w}[y_1 \leftarrow d_1,\ldots,y_{\#(q)} \leftarrow
-    d_{\#(q)}]$. Hence $\sem{\psi}_{\eta} = \emptyset$, and because
-  $\I,\data{w} \models \accform{\alpha}$, we obtain that $\I,\eta \models
-  q(\vec{y}) \rightarrow \stamp{\psi}{m+1}$, thus
-  $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$. However, this
-  contradicts the fact that $\I=\I_{\T}$ and that
-  $q(d_1,\ldots,d_{\#(q)})$ labels a node of $\T$. Second, assume that
-  (\ref{it2:acceptance}), there exists a frontier node of $\T$ labeled
-  with a configuration $q(d_1,\ldots,d_{\#(q)})$ such that $q \in Q
-  \setminus F$. Since $\I,\data{w} \models \forall y_1 \ldots \forall
-  y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot$, by a
-  similar reasoning as in the above case, we obtain that
-  $\tuple{d_1,\ldots,d_{\#(q)}} \not\in \I(q)$, contradiction.
-
-  ``$(\ref{it2:lemma:acceptance}) \Rightarrow
-  (\ref{it1:lemma:acceptance})$'' Let $\T$ be an accepting execution
-  of $\A$ over $w$. We prove that $\I_\T,\data{w} \models
-  \accform{\event{w}}$. By Lemma \ref{lemma:path-formula}, we obtain
-  $\I_\T,\data{w} \models \pathform{\event{w}}$. Since every path in
-  $\T$ is of length $n$ and all nodes on the $n$-th level of $\T$ are
-  labeled by final configurations, we obtain that $\I_\T,\data{w}
-  \models \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
-  y_{\#(q)} ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot$,
-  trivially.  \qed}
 
@@
 automaton. Next, in \S\ref{sec:emptiness} we shall assume moreover
-that the first order theory of the data sort $\Data$ (without
+that the first-order theory of the data sort $\Data$ (without
 uninterpreted predicate symbols) has quantifier elimination, providing
@@
 prefix $a_1 \ldots a_i$ of $\alpha$, for $i \in [n]$, where
-$\alpha_0=\epsilon$.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\ifLongVersion
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{definition}\label{def:quantform}
-  Let $\quantformn{\alpha_0}, \ldots, \quantformn{\alpha_n}$ be the
-  sequence of formulae defined by $\quantformn{\alpha_0} \isdef
-  \stamp{\iota}{0}$ and, for all $i \in [1,n]$:
-  \[\quantformn{\alpha_i} \isdef \quantformn{\alpha_{i-1}} \wedge
-  \bigwedge_{\begin{array}{l}
-      \scriptstyle{\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_{i-1}}} \\
-      \scriptstyle{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta} 
-  \end{array}} \stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow 
-    \stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]
-      \]  
-      We write
-    $\quantform{\alpha}$ for the prenex normal form of the formula: 
-      \[\quantformn{\alpha_n} \wedge \bigwedge_{\begin{array}{l}
-          \scriptstyle{\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \quantformn{\alpha_n}} \\
-          \scriptstyle{q \in Q \setminus F}
-      \end{array}} \stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot \enspace.\]
-\end{definition}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\else
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-We consider the sequence of formulae $\quantformn{\alpha_0}, \ldots,
-\quantformn{\alpha_n}$ defined as $\quantformn{\alpha_0} \isdef
-\stamp{\iota}{0}$ and, for all $i \in [1,n]$, let
-$\quantformn{\alpha_i}$ be the conjunction of
+$\alpha_0=\epsilon$. We consider the sequence of formulae
+$\quantformn{\alpha_0}, \ldots, \quantformn{\alpha_n}$ defined as
+$\quantformn{\alpha_0} \isdef \stamp{\iota}{0}$ and, for all $i \in
+[1,n]$, let $\quantformn{\alpha_i}$ be the conjunction of
 $\quantformn{\alpha_{i-1}}$ with all formulae
@@
 $\stamp{q}{n}(t_1,\ldots,t_{\#(q)})$ occurs in
-$\quantformn{\alpha_n}$, for some $q \in Q \setminus F$.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\fi
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Note that $\quantform{\alpha}$ contains no path quantifiers, as
-required. On the other hand, the scope of the transition quantifiers
-in $\quantform{\alpha}$ exceeds the right-hand side formulae from the
+$\quantformn{\alpha_n}$, for some $q \in Q \setminus F$. Note that
+$\quantform{\alpha}$ contains no path quantifiers, as required. On the
+other hand, the scope of the transition quantifiers in
+$\quantform{\alpha}$ exceeds the right-hand side formulae from the
 transition rules, as shown by the following example.
@@
   Notice that the transition quantifiers $\exists z_1$ and $\forall
-  z_2$ from $\accform{\alpha}$ range now over the entire formula
+  z_2$ from $\accform{\alpha}$ range now over
   $\quantform{\alpha}$. \hfill$\blacksquare$
@@
   valuation $\nu : \stamp{X}{\leq n} \rightarrow \Data$, the following
-  hold, for every interpretation $\I$: \begin{compactenum}
+  hold, for every interpretation $\I$: \begin{inparaenum}[(1)]
   \item\label{it1:quant} if $\I,\nu \models \accform{\alpha}$ then
@@
     \models \accform{\alpha}$.
-  \end{compactenum}
+  \end{inparaenum}
 \end{lemma}
-\proof{(\ref{it1:quant}) Trivial, since every subformula
-  $q(t_1,\ldots,t_{\#(q)}) \rightarrow
-  \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ of $\quantform{\alpha}$ is
-  entailed by a subformula $\forall y_1 \ldots \forall y_{\#(q)} ~.~
-  q(y_1, \ldots, y_{\#(q)}) \rightarrow \psi$ of $\accform{\alpha}$.
-
-  \noindent(\ref{it2:quant}) By repeated applications of the following fact:
-
-  \begin{fact}
-    Given formulae $\phi$ and $\psi$, such that no predicate atom with
-    predicate symbol $q$ occurs in $\psi(y_1,\ldots,y_{\#(q)})$, for
-    each valuation $\nu$, if there exists an interpretation $\I$ such
-    that $\I,\nu \models \phi \wedge
-    \bigwedge_{q(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \phi}
-    q(t_1,\ldots,t_{\#(q)}) \rightarrow
-    \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ then there exists a
-    valuation $\J$ such that $\J(q) \subseteq \I(q)$ and $\J(q') =
-    \I(q')$ for all $q' \in Q \setminus \set{q}$ and $\J,\nu \models
-    \phi \wedge \forall y_1 \ldots \forall y_{\#(q)} ~.~
-    q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$.
-  \end{fact}
-  \proof{Assume w.l.o.g. that $\phi$ is quantifier-free. The proof can
-    be easily generalized to the case $\phi$ has quantifiers. Let
-    $\J(q) = \set{\langle t_1^\nu, \ldots, t_{\#(q)}^\nu \rangle \in
-      \I(q) \mid q(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \phi}$ and
-    $\J(q') = \I(q')$ for all $q' \in Q \setminus \set{q}$. Since
-    $\I,\nu \models \phi$, we obtain that also $\J,\nu \models \phi$
-    because the tuples of values in $\I(q) \setminus \J(q)$ are not
-    interpretations of terms that occur within subformulae
-    $q(t_1,\ldots,t_{\#(q)})$ of $\phi$. Moreover,
-    $\bigwedge_{q(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \phi}
-    q(t_1,\ldots,t_{\#(q)}) \rightarrow
-    \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ and $\forall y_1 \ldots
-    \forall y_{\#(q)} ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$
-    are equivalent under $\J$, thus $\J,\nu \models \forall y_{\#(q)}
-    ~.~ q(y_1,\ldots,y_{\#(q)}) \rightarrow \psi$, as required. \qed}
-
-  \noindent This concludes the proof. \qed}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\ifLongVersion
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{definition}\label{def:substform}
-  Let $\substformn{\alpha_0}, \ldots, \substformn{\alpha_n}$ be the
-  sequence of formulae defined by $\substformn{\alpha_0} \isdef
-  \stamp{\iota}{0}$ and, for all $i \in [1,n]$,
-  $\substformn{\alpha_i}$ is obtained by replacing each occurrence of
-  a predicate atom $\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)})$ in
-  $\substformn{\alpha_{i-1}}$ with the formula
-  $\stamp{\psi}{i}[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$, where
-  $q(\vec{y}) \arrow{a_{i}(X)}{} \psi \in \Delta$. We write
-  $\substform{\alpha}$ for the formula obtained by replacing, in
-  $\substformn{\alpha}$, each occurrence of a predicate
-  $\stamp{q}{n}$, such that $q \in Q \setminus F$ (resp. $q \in F$),
-  by $\bot$ (resp. $\top$).
-\end{definition}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\else
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 Further, we eliminate the predicate atoms from $\quantform{\alpha}$,
@@
 $\bot$ (resp. $\top$).
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\fi
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{example}[Contd. from Example \ref{ex:quant-elim}]\label{ex:pred-elim}
@@
 \end{lemma}
-\proof{ By induction on $n\geq0$. The base case $n=0$ is trivial,
-  since $\quantform{A} = \substform{A} = \stamp{\iota}{0}$. For the
-  induction step, we rely on the following fact:
-
-  \begin{fact}\label{fact:subst}
-    Given formulae $\phi$ and $\psi$, such that $\phi$ is positive
-    $q(t_1,\ldots,t_{\#(q)})$ is the only one occurrence of the
-    predicate symbol $q$ in $\phi$ and no predicate atom with
-    predicate symbol $q$ occurs in $\psi(y_1,\ldots,y_{\#(q)})$, for
-    each interpretation $\I$ and each valuation $\nu$, we have:
-    \[\begin{array}{l} 
-    \I,\nu \models \phi \wedge q(t_1,\ldots,t_{\#(q)}) \rightarrow
-    \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}] \iff \\ \nu \models
-    \phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]\enspace.
-    \end{array}\]
-  \end{fact}
-  \proof{ We assume w.l.o.g. that $\phi$ is quantifier-free. The proof
-    can be easily generalized to the case $\phi$ has quantifiers.
-
-    \noindent''$\Rightarrow$'' 
-    We distinguish two cases: \begin{compactitem}
-    \item if $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \in \I(q)$ then
-      $\I,\nu \models \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$. Since
-      $\phi$ is positive, replacing $q(t_1,\ldots,t_{\#(q)})$ with
-      $\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$ does not change the
-      truth value of $\phi$ under $\nu$, thus $\nu \models
-      \phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$.
-      %
-    \item else, $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \not\in
-      \I(q)$, thus $\nu \models
-      \phi[\bot/q(t_1,\ldots,t_{\#(q)})]$. Since $\phi$ is positive
-      and $\bot$ entails $\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$,
-      we obtain $\nu \models
-      \phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$
-      by monotonicity. 
-    \end{compactitem}
-
-    \noindent``$\Leftarrow$'' Let $\I$ be any interpretation such that
-    $\I(q) = \set{\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \mid \nu
-      \models \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]}$. We
-    distinguish two cases: \begin{compactitem}
-      \item if $\I(q) \not \emptyset$ then $\I,\nu \models
-        q(t_1,\ldots,t_{\#(q)})$ and $\nu \models \psi[t_1/y_1,
-          \ldots, t_{\#(q)}/y_{\#(q)}]$. Thus replacing $\psi[t_1/y_1,
-          \ldots, t_{\#(q)}/y_{\#(q)}]$ by $q(t_1,\ldots,t_{\#(q)})$
-        does not change the truth value of $\phi$ under $\I$ and
-        $\nu$, and we obtain $\I,\nu \models \phi$. Moreover, $\I,\nu
-        \models \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ implies
-        $\I,\nu \models q(t_1,\ldots,t_{\#(q)}) \rightarrow
-        \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$.
-        %
-      \item else $\I(q) = \emptyset$, hence $\nu \not\models
-        \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$, thus $\nu \models
-        \phi[\bot/q(t_1,\ldots,t_{\#(q)})]$. Because $\phi$ is
-        positive, we obtain $\I,\nu \models \phi$ by monotonicity. But
-        $\I,\nu \models q(t_1,\ldots,t_{\#(q)}) \rightarrow
-        \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ trivially, because
-        $\I,\nu \not\models q(t_1,\ldots,t_{\#(q)})$.\qed
-    \end{compactitem}}
-This concludes the proof. \qed}
-
-%% \begin{proposition}\label{prop:subst}
-%%   Let $\phi$ be a positive formula and let $q(t_1,\ldots,t_{\#(q)})$
-%%   be the only predicate atom occurring within $\phi$. Then, for every
-%%   valuation $\nu$ of $\fv{}{\phi}$ and every formula $\psi$ with no
-%%   predicate occurrences, the following are
-%%   equivalent: \begin{compactenum}
-%%   \item\label{it1:prop:subst} $\nu \models \phi[\psi[t_1/y_1, \ldots,
-%%       t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$, 
-%%     %
-%%   \item\label{it2:prop:subst} $\I,\nu \models \phi \wedge \forall y_1
-%%     \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow
-%%     \psi$, for some interpretation $\I$.
-%%   \end{compactenum}
-%%   where $y_1,\ldots,y_{\#(q)} \not\in \fv{}{\phi}$. 
-%% \end{proposition}
-%% \proof{ We give the proof in the case $\phi$ is quantifier-free. The
-%%   generalization to a non-empty quantifier prefix is straightforward.
-%%   ``$(\ref{it1:prop:subst}) \Rightarrow (\ref{it2:prop:subst})$'' We
-%%   distinguish two cases:
-%%   \begin{compactitem}
-%%   \item if $\nu \models \psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$
-%%     then let $\I(q) = \set{\tuple{t_1^\nu,\ldots,t_{\#(q)}^\nu}}$ and
-%%     $\emptyset$ everywhere else. We have $\I,\nu \models
-%%     q(t_1,\ldots,t_{\#(q)})$ and, since $\phi$ is positive, replacing
-%%     $\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ with
-%%     $q(t_1,\ldots,t_{\#(q)})$ in $\phi$ does not change its
-%%     satisfiability under $\I,\nu$, hence $\I,\nu \models
-%%     \phi$. Moreover, $\nu \models \psi[t_1/y_1, \ldots,
-%%       t_{\#(q)}/y_{\#(q)}]$ thus $\I,\nu \models \forall y_1 \ldots
-%%     \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \psi$.
-%%     %
-%%   \item else $\nu \not\models \psi[t_1/y_1, \ldots,
-%%     t_{\#(q)}/y_{\#(q)}]$ and let $\I$ be the interpretation assigning
-%%     each predicate symbol with $\emptyset$. Then replacing
-%%     $\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]$ with
-%%     $q(t_1,\ldots,t_{\#(q)})$ in $\phi$ does not change its
-%%     satisfiability under $\I,\nu$. Moreover, since $\I$ maps $q$ to
-%%     $\emptyset$, we have $\I,\nu \models \forall y_1 \ldots \forall
-%%     y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \psi$, trivially.
-%%   \end{compactitem}
-
-%%   ``$(\ref{it2:prop:subst}) \Rightarrow (\ref{it1:prop:subst})$'' We
-%%   distinguish two cases: \begin{compactitem}
-%%   \item if $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \in \I(q)$, since
-%%     $\I,\nu \models \forall y_1 \ldots \forall y_{\#(q)} ~.~
-%%     q(\vec{y}) \rightarrow \psi$, we have $\nu \models
-%%     \psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]$. Since $\phi$ is
-%%     positive, replacing $q(t_1,\ldots,t_{\#(q)})$ by $\psi$ in $\phi$
-%%     does not change its satisfiability under $\nu$, hence $\nu \models
-%%     \phi[\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]/q(t_1, \ldots,
-%%       t_{\#(q)})]$.
-%%     %
-%%   \item else $\tuple{t_1^\nu, \ldots, t_{\#(q)}^\nu} \not\in \I(q)$,
-%%     that is $\I,\nu \not\models q(t_1,\ldots,t_{\#(q)})$. Then $\nu
-%%     \models \phi[\bot/q(t_1,\ldots,t_{\#(q)})]$ and we obtain $\nu
-%%     \models \phi[\psi[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]/q(t_1,
-%%       \ldots, t_{\#(q)})]$, because $\phi$ is positive, thus
-%%     $\phi[\bot/q(t_1,\ldots,t_{\#(q)})] \models \phi[\psi[t_1/y_1,
-%%         \ldots, t_{\#(q)}/y_{\#(q)}]/q(t_1, \ldots, t_{\#(q)})]$. \qed
-%%   \end{compactitem}}
 
@@
 \end{lemma}
-\proof{ By Lemma \ref{lemma:acceptance}, $w \in \lang{\A}$ if and only
-  if $\I,\data{w} \models \accform{\event{w}}$, for some
-  interpretation $\I$. By Lemma \ref{lemma:quant}, there exists an
-  interpretation $\I$ such that $\I,\data{w} \models
-  \accform{\event{w}}$ if and only if there exists an interpretation
-  $\J$ such that $\J,\nu \models \quantform{\event{w}}$. By Lemma
-  \ref{lemma:subst}, there exists an interpretation $\J$ such that
-  $\J,\nu \models \quantform{\event{w}}$ if and only if $\nu \models
-  \substform{\event{w}}$. \qed}
 
@@
 
-%% The dualization operation has the following property:
-%% \begin{lemma}\label{lemma:dual}
-%%   Given a formula $\phi$ and an automaton
-%%   $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$, we have the equivalence
-%%   $\Accept{F}{\phi} \equiv \neg(\dual{\Accept{Q\setminus F}{\phi}})$. 
-%% \end{lemma}
-%% \proof{ By straightforward induction on the structure of $\phi$. \qed}
-
 \begin{theorem}\label{thm:closure}
@@
       \tuple{\Sigma,X,Q_i,\dual{\iota},Q_i\setminus
-        F_i,\dual{\Delta}_i}$ and, for $i=1,2$: \[\dual{\Delta}_i =
+        F_i,\dual{\Delta}_i}$ and \(\dual{\Delta}_i =
       \set{q(\vec{y}) \arrow{a(X)}{} \dual{\psi} \mid
         q(\vec{y}) \arrow{a(X)}{} \psi \in
-        \Delta_i}\enspace.\]
+        \Delta_i}\), for $i=1,2$. 
   \end{compactenum}
   Moreover, $\len{\A_\cap} = \bigO{\len{\A_1}+\len{\A_2}}$ and
-  $\len{\overline{\A_i}} = \bigO{\len{\A_i}}$, for all $i=1,2$.
+  $\len{\overline{\A_i}} = \bigO{\len{\A_i}}$, for $i=1,2$.
 \end{theorem}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\ifLongVersion\else
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\paragraph{\em Proof sketch.} (\ref{it1:thm:closure})
-The proof for $\A_\cap$ is similar to the standard proof for the
-intersection of propositional alternating
-automata. (\ref{it2:thm:closure}) Let $w \in \Sigma[X]^*$ be a
-word. We denote by $\accformA{\A_1}{\event{w}}$ and
-$\substformA{\A_1}{\event{w}}$
-[resp. $\accformA{\overline{\A}_1}{\event{w}}$ and
-  $\substformA{\overline{\A}_1}{\event{w}}$] the formulae
-$\accform{\event{w}}$ and $\substform{\event{w}}$ for $\A_1$ and
-$\overline{\A}_1$, respectively. It is enough to show that
-$\substformA{\overline{\A}_1}{\event{w}} =
-\neg\substformA{\A_1}{\event{w}}$ and apply Lemma
-\ref{lemma:quant-pred-acceptance} to prove that $w \in \lang{\A_1}
-\iff w \not\in \lang{\overline{\A}_1}$. Since the choice of $w$ was
-arbitrary, this proves $\lang{\overline{\A}_1} = \Sigma[X]^* \setminus
-\lang{\A_1}$. The proof is by induction on the number of predicate
-atoms in $\accformA{\A_1}{\event{w}}$ that are replaced during the
-generation of $\substformA{\A_1}{\event{w}}$ and relies on the
-following fact, itself proved by induction on the structure of $\phi$:
-  \begin{fact}\label{fact:dual-neg}
-    Let $\phi$ be a positive first order formula and let
-    $q(t_1,\ldots,t_{\#(q)})$ be the only occurrence of a predicate
-    symbol within $\phi$. Then, for any formula $\psi$, we have the
-    equivalence:
-    \[\neg\phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]
-    \equiv
-    \dual{\phi}[\neg\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]\]
-  \end{fact}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\fi
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\proof{ (\ref{it1:thm:closure}) ``$\subseteq$'' Let $w \in
-  \lang{\A_\cap}$ be a word and $\T$ be an execution of $\A_\cap$ over
-  $w$. Since $Q_1 \cap Q_2 = \emptyset$, it is possible to partition
-  $\T$ into $\T_1$ and $\T_2$ such that the roots of $\T_i$ form a
-  cube from $\cube{\minsem{\iota_i}}$, for all $i=1,2$. Because
-  $\Delta_1 \cap \Delta_2 = \emptyset$, by induction on
-  $\len{w}\geq0$, one shows that $\T_i$ is an execution of $\A_i$ over
-  $w$, for all $i=1,2$. Finally, because $\T$ is accepting, we obtain
-  that $\T_1$ and $\T_2$ are accepting, respectively, hence $w \in
-  \lang{\A_1} \cap \lang{\A_2}$. ``$\supseteq$'' Let $w \in
-  \lang{\A_1} \cap \lang{\A_2}$ and let $\T_i$ an accepting execution
-  of $\A_i$ over $w$, for all $i=1,2$. We show that $\T_1 \cup \T_2$
-  is an execution of $\A_\cap$ over $w$, by induction on
-  $\len{w}\geq0$. For the base case $\len{w}=0$, we have $\T_i \in
-  \cube{\minsem{\iota_i}}$ for all $i=1,2$ and since $Q_1 \cap Q_2 =
-  \emptyset$, we have $\T_1 \cup \T_2 \in \cube{\minsem{\iota_1 \wedge
-      \iota_2}}$. The induction step follows as a consequence of the
-  fact that $\Delta_1 \cup \Delta_2$ is the set of transition rules of
-  $\A_\cap$. Finally, since both $\T_1$ and $\T_2$ are accepting,
-  $\T_1 \cup \T_2$ is accepting as well. Moreover, we have: 
-  \[
-  \len{\A_\cap} = \len{\iota_1 \wedge \iota_2} + \sum_{q(\vec{y})
-    \arrow{a(X)}{} \psi \in \Delta_1 \cup \Delta_2} \len{\psi} =
-  1 + \len{\iota_1} + \len{\iota_2} + \sum_{q(\vec{y})
-    \arrow{a(X)}{} \psi \in \Delta_1} \len{\psi} + \sum_{q(\vec{y})
-    \arrow{a(X)}{} \psi \in \Delta_2} \len{\psi} \enspace. 
-  \]
-
-  \noindent(\ref{it2:thm:closure}) Let $w \in \Sigma[X]^*$ be a
-  word. We denote by $\accformA{\A_1}{\event{w}}$ and
-  $\substformA{\A_1}{\event{w}}$
-  [resp. $\accformA{\overline{\A}_1}{\event{w}}$ and
-    $\substformA{\overline{\A}_1}{\event{w}}$] the formulae
-  $\accform{\event{w}}$ and $\substform{\event{w}}$ for $\A_1$ and
-  $\overline{\A}_1$, respectively. It is enough to show that
-  $\substformA{\overline{\A}_1}{\event{w}} =
-  \neg\substformA{\A_1}{\event{w}}$ and apply Lemma
-  \ref{lemma:quant-pred-acceptance} to prove that $w \in \lang{\A_1}
-  \iff w \not\in \lang{\overline{\A}_1}$. Since the choice of $w$ was
-  arbitrary, this proves $\lang{\overline{\A}_1} = \Sigma[X]^*
-  \setminus \lang{\A_1}$. By induction on the number of predicate
-  atoms in $\accformA{\A_1}{\event{w}}$ that are replaced during the
-  generation of $\substformA{\A_1}{\event{w}}$. The proof relies on
-  the following fact:
-
-  \begin{fact}\label{fact:dual-neg}
-    Let $\phi$ be a positive formula and let $q(t_1,\ldots,t_{\#(q)})$
-    be the only occurrence of a predicate symbol within $\phi$. Then,
-    every formula $\phi$ with no predicate occurrences:
-    $\neg\phi[\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]
-    \equiv
-    \dual{\phi}[\neg\psi[t_1/y_1,\ldots,t_{\#(q)}/y_{\#(q)}]/q(t_1,\ldots,t_{\#(q)})]$.
-  \end{fact}
-  \proof{By induction on the structure of $\phi$. \qed}}
 
@@
 semi-algorithm based on \emph{lazy annotation}
-\cite{McMillan06,McMillan14}.
-
-In a nutshell, a lazy annotation procedure systematically explores the
-set of finite input event sequences searching for an accepting
-execution. For an input sequence, if the path formula is satisfiable,
-we compute a word in the language of the automaton, from the model of
-the path formula. Otherwise, i.e.\ the sequence is \emph{spurious},
-the search backtracks and each position in the sequence is annotated
-with a tuple of interpolants, that marks the entire sequence as
-infeasible. The semi-algorithm uses moreover a coverage relation
-between sequences, ensuring that the continuations of already covered
-sequences are never explored. Sometimes this coverage relation
-provides a sound termination argument, in case when the automaton is
-empty.
-
-We check emptiness of first order alternating automata using a version
-of the \impact~ lazy annotation semi-algorithm \cite{McMillan06}. An
-analogous procedure is given in \cite{IosifXu18}, for a simpler model
-of alternating automata, that uses only predicates or arity zero
-(propositional variables) and no transition quantifiers. For
-simplicity, we defer the pseudocode of the semi-algorithm to Appendix
-\ref{app:algorithm} and give here several high-level definitions.
+\cite{McMillan06,McMillan14}. In a nutshell, a lazy annotation
+procedure systematically explores the set of finite input event
+sequences searching for an accepting execution. For an input sequence,
+if the path formula is satisfiable, we compute a word in the language
+of the automaton, from the model of the path formula. Otherwise,
+i.e.\ the sequence is \emph{spurious}, the search backtracks and each
+position in the sequence is annotated with an interpolant, thus
+marking the sequence as infeasible. The semi-algorithm uses moreover a
+coverage relation between sequences, ensuring that the continuations
+of already covered sequences are never explored. Sometimes this
+coverage relation provides a sound termination argument, in case when
+the automaton is empty.
 
@@
   instance, using breadth-first search.} from $\Sigma^*$ will suffice
-to discover a word $w \in \lang{\A}$, provided that the first order
+to discover a word $w \in \lang{\A}$, provided that the first-order
 theory of the data domain $\Data$ is decidable (Lemma
@@
 
-%% , for all words $w \in \Sigma[X]^*$ and all cubes $c \in
-%% \cube{\sem{U(\alpha')}}$, if $\A$ accepts $w$ started in $c$ then
-%% there exists a cube $d \in \cube{\sem{U(\beta)}}$ such that $\A$
-%% accepts $w$ started in $d$.
+\input algorithm
 
@@
 labeling. The semi-algorithm stops when (and if) a safe complete
-unfolding has been found. For a detailed presentation of the emptiness
-procedure, we refer to Appendix \ref{app:algorithm}.
+unfolding has been found.
 
@@
 \end{theorem}
-\proof{ Let $U$ be a safe and complete unfolding of $\A$, such that
-  $\dom(U) \neq \emptyset$. Suppose, by contradiction, that there
-  exists a word $w \in \lang{A}$ and let $\alpha \isdef
-  \event{w}$. Since $w \in \lang{A}$, by Lemma \ref{lemma:acceptance},
-  there exists an interpretation $\I$ such that $\I,\data{w} \models
-  \accform{\alpha}$. Assume first that $\alpha \in \dom(U)$.  In this
-  case, one can show, by induction on the length $n\geq0$ of $w$, that
-  $\pathform{\alpha} \models \stamp{U(\alpha)}{n}$, thus $\I,\data{w}
-  \models \stamp{U(\alpha)}{n}$. Since $\I,\data{w} \models
-  \accform{\alpha}$, we have $\I,\data{w} \models \bigwedge_{q \in Q
-    \setminus F} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{n}(\vec{y}) \rightarrow \bot$, hence $\stamp{U(\alpha)}{n}
-  \wedge \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall
-  y_{\#(q)} ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot$. By renaming
-  $\stamp{q}{n}$ with $q$ in the previous formula, we obtain
-  $U(\alpha) \wedge \forall y_1 \ldots \forall y_{\#(q)} ~.~
-  q(\vec{y}) \rightarrow \bot$ is satisfiable, thus $U$ is not safe,
-  contradiction.
-
-  We proceed thus under the assumption that $\alpha \not\in
-  \dom(U)$. Since $\dom(U)$ is a nonempty prefix-closed set, there
-  exists a strict prefix $\alpha'$ of $\alpha$ that is a leaf of
-  $\dom(U)$. Since $U$ is closed, the leaf $\alpha'$ must be covered
-  and let $\alpha_1 \prefix \alpha' \prefix \alpha$ be a node such
-  that $U(\alpha_1) \models U(\beta_1)$, for some uncovered node
-  $\beta_1 \in \dom(U)$.
-  %% for every word $w'$ and every cube $c \in
-  %% \cube{\sem{U(\alpha_1)}}$ such that $\A$ accepts $w'$ starting
-  %% with $c$, there exists a cube $d \in \cube{\sem{U(\beta_1)}}$
-  %% such that $\A$ accepts $w'$ starting with $d$.
-  Let $\gamma_1$ be the unique sequence such that $\alpha_1\gamma_1 =
-  \alpha$. By Definition \ref{def:coverage}, since $\alpha_1 \cover
-  \beta_1$ and $\event{w} = \alpha_1\gamma_1 \in \lang{\A}$, there
-  exists a word $w_1$ and a cube $c_1 \in \cube{\sem{U(\alpha_1)}}
-  \subseteq \cube{\sem{U(\beta_1)}}$, such that $\event{w_1} =
-  \gamma_1$ and $\A$ accepts $w_1$ starting with $c_1$. If
-  $\beta_1\gamma_1 \in \dom(U)$, we obtain a contradiction by a
-  similar argument as above. Hence $\beta_1\gamma_1 \not\in \dom(U)$
-  and there exists a leaf of $\dom(U)$ which is also a prefix of
-  $\beta_1\gamma_1$. Since $U$ is closed, this leaf is covered by an
-  uncovered node $\beta_2 \in \dom(U)$ and let $\alpha_2 \in \dom(U)$
-  be the minimal (in the prefix partial order) node such that $\beta_1
-  \prefix \alpha_2 \prefix \beta_1\gamma_1$ and $\alpha_2 \cover
-  \beta_2$. Let $\gamma_2$ be the unique sequence such that
-  $\alpha_2\gamma_2 = \beta_1\gamma_1$. Since $\beta_1$ is uncovered,
-  we have $\beta_1 \neq \alpha_2$ and thus $\len{\gamma_1} >
-  \len{\gamma_2}$. By repeating the above reasoning for $\alpha_2$,
-  $\beta_2$ and $\gamma_2$, we obtain an infinite sequence
-  $\len{\gamma_1} > \len{\gamma_2} > \ldots$, which is again a
-  contradiction. \qed}
-
-As mentioned above, we check emptiness of first order alternating
-automata using a semi-algorithm previously used to check emptiness of
-a simpler model of alternating automata, which uses propositional
-variables for control states and whose transition rules have no
-quantifiers \cite{IosifXu18}. The higher complexity of the automata
-model considered here, manifests itself within the interpolant
-generation procedure, used to refine the labeling of the unfolding. We
-discuss generation of interpolants in the next section.
+
+We describe the semi-algorithm used to check emptiness of first-order
+alternating automata. The execution of Algorithm \ref{alg:impact}
+consists of three phases, corresponding to the \textsc{Close},
+\textsc{Refine} and \textsc{Expand} of the original \impact\ procedure
+\cite{McMillan06}.  Let $n$ be a node removed from the worklist at
+line \ref{ln:impact-dequeue} and let $\alpha(n)$ be the input sequence
+labeling the path from the root node to $n$. If
+$\substform{\alpha(n)}$ is satisfiable, the sequence $\alpha(n)$ is
+feasible, in which case a model of $\substform{\alpha(n)}$ is obtained
+and a word $w \in L(\A)$ is returned. Otherwise, $\alpha(n)$ is an
+infeasible input sequence and the procedure enters the refinement
+phase (lines \ref{ln:refine-begin}-\ref{ln:refine-end}). The GLI for
+$\alpha(n)$ is used to strenghten the labels of all the ancestors of
+$n$, by conjoining the formulae of the interpolant, changed according
+to Lemma \ref{lemma:refinement}, to the existing labels.
+
+In this process, the nodes on the path between $\rootNode$ and $n$,
+including $n$, might become eligible for coverage, therefore we
+attempt to close each ancestor of $n$ that is impacted by the
+refinement (line \ref{ln:refine-end}). Observe that, in this case the
+call to $\Call{Close}{}$ must uncover each node which is covered by a
+successor of $n$ (line \ref{ln:close-uncover} of the $\Call{Close}{}$
+function). This is required because, due to the over-approximation of
+the sets of reachable configurations, the covering relation is not
+transitive, as explained in \cite{McMillan06}. If $\Call{Close}{}$
+adds a covering edge $(n_i,m)$ to $\lhd$, it does not have to be
+called for the successors of $n_i$ on this path, which is handled via
+the boolean flag $b$. Finally, if $n$ is still uncovered (it has not
+been previously covered during the refinement phase) we expand $n$
+(lines \ref{ln:expand-begin}-\ref{ln:expand-end}) by creating a new
+node for each successor $s$ via the input event $a \in \Sigma$ and
+inserting it into the worklist.
 
@@
 interpolants involve only predicates $\stamp{q}{i}$ and variables
-$\stamp{x}{i}$, for a single $i \geq 0$.
-
-\paragraph{\em Remark}
-When considering alternating executions, control-local interpolants
-are not always enough to prove emptiness, because of the
-synchronization of several branches of the computation on the same
-input word. Consider, for instance, the automaton from Example
-\ref{ex:foada}, started in an initial configuration $q_0(0)$, on an
-input word $(a,\nu_1) \ldots (a,\nu_n)$, such that $\nu_i(x)=k_i$, for
-all $i \in [1,n]$, the automaton executes as follows:
-\[\begin{array}{c}
-\set{(q_0,0)} \arrow{(a,\nu_1)}{} \set{(q_1,k_1),(q_2,-k_1)} \ldots
-\arrow{(a,\nu_{n})}{} 
-\set{(q_1,\sum_{i=1}^{n} k_i),(q_2,-\sum_{i=1}^{n} k_i)}
-\end{array}\]
-An overapproximation of the set of cubes generated after one or more
-steps is defined by the formula $\exists x_1 \exists x_2 ~.~ q_1(x_1)
-\wedge q_2(x_2) \wedge x_1+x_2 \geq 0$. Note that a control-local formula
-using one occurrence of a predicate would give a too rough
-overapproximation of this set, unable to prove the emptiness of the
-automaton. This is because any formula with a single occurrence of an
-atom of the form $q(x)$ would lose the constraint $x_1+x_2 \geq 0$
-between the values occurring at the same level, on both branches of
-the execution. \hfill$\blacksquare$
+$\stamp{x}{i}$, for a single $i \geq 0$. When considering alternating
+executions, control-local interpolants are not always enough to prove
+emptiness, because of the synchronization of several branches of the
+computation on the same input word. For this reason, the interpolants
+considered in this paper will never be control-local and we shall use
+the term \emph{local} to denote time-local interpolants, with no free
+variables.
 
@@
 \end{definition}
-%% The existence of Lyndon interpolants for first order logic, with no
-%% predefined interpretation of function symbols\footnote{A model
-%%   specifies the universe and the interpretation of function symbols.},
-%% was proved by a non-constructive model-theoretic argument
-%% \cite{Lyndon59}. In the following, we shall consider only those
-%% theories of the data domain $\Data$, for which there exists a decision
-%% procedure for satisfiability, able to produce Lyndon interpolants. For
-%% instance, such interpolating decision procedures exist for linear real
-%% and integer arithmetic, with uninterpreted function symbols (UFLRA and
-%% UFLIA).
 
 In the rest of this section, fix an automaton $\A =
-\tuple{\Sigma,X,Q,\iota,F,\Delta}$. Due to the above remark, none of
-the interpolants considered will be control-local and we shall use the
-term \emph{local} to denote time-local interpolants, with no free
-variables. The following definition generalizes interpolants from
-unsatisfiable conjunctions to input sequences:
+\tuple{\Sigma,X,Q,\iota,F,\Delta}$. The following definition
+generalizes interpolants from unsatisfiable conjunctions to input
+sequences:
 
 \begin{definition}\label{def:generalized-lyndon-interpolant}
-  Given a non-empty sequence of input events $\alpha = a_1 \ldots a_n
-  \in \Sigma^*$, a \emph{generalized Lyndon interpolant (GLI)} is a
+  Given a sequence of input events $\alpha = a_1 \ldots a_n \in
+  \Sigma^*$, a \emph{generalized Lyndon interpolant (GLI)} is a
   sequence $(I_0,\ldots,I_n)$ of formulae such that, for all $k \in
-  [n-1]$, the following hold: \begin{compactenum}
+  [n-1]$, the following hold: \begin{inparaenum}[(1)]
     \item\label{it1:generalized-lyndon-interpolant} $\pset{-}{I_k} =
@@
     \item\label{it2:generalized-lyndon-interpolant} $\stamp{\iota}{0}
-      \models I_0$ and \(I_k \wedge \Big(\bigwedge_{q(\vec{y})
+      \models I_0$, \(I_k \wedge \Big(\bigwedge_{q(\vec{y})
         \arrow{a_i(X)}{} \psi \in \Delta} \forall y_1 \ldots \forall
       y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y}) \rightarrow
-      \stamp{\psi}{k+1}\Big) \models I_{k+1}\),
+      \stamp{\psi}{k+1}\Big) \models I_{k+1}\) and
       %
@@
       y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot$ is unsatisfiable.
-  \end{compactenum}
+  \end{inparaenum}
   Moreover, the GLI is \emph{local} if and only if $\voc{I_k}
@@
   If there exists a Lyndon interpolant for any two formulae $\phi$ and
-  $\psi$, in the first order theory of data with uninterpreted
+  $\psi$, in the first-order theory of data with uninterpreted
   predicate symbols, such that $\phi \wedge \psi$ is unsatisfiable,
@@
 \end{proposition}
-\paragraph{\em Proof.} By definition, $\accform{\alpha}$ is the formula:
-\[\begin{array}{c}
-\stamp{\iota}{0} \wedge 
- \bigwedge_{i=1}^n \underbrace{\bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in
-   \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
- \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i}}_{\varphi_i} \wedge
- \underbrace{\bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
- ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot}_{\phi}
- \end{array}\]
- %% We define the formulae: 
- %% \[\begin{array}{rcl}
- %% \varphi_i & \isdef & \bigwedge_{q(\vec{y}) \arrow{a_i(X)}{} \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
- %% \stamp{q}{i-1}(\vec{y}) \rightarrow \stamp{\psi}{i},~ \text{for all}~ i \in [1,n] \\
- %% \phi & \isdef & \bigwedge_{q \in Q \setminus F} \forall y_1 \ldots \forall y_{\#(q)}
- %% ~.~ \stamp{q}{n}(\vec{y}) \rightarrow \bot
- %% \end{array}\]
- Observe that $\voc{\stamp{\iota}{0}} \subseteq \stamp{Q}{0}$,
- $\voc{\varphi_i} \subseteq
- \stamp{Q}{i-1}\cup\stamp{Q}{i}\cup\stamp{X}{i}$, for all $i \in
-       [1,n]$, and $\voc{\phi} \subseteq \stamp{Q}{n}$.  We apply
-       Lyndon's Interpolation Theorem for the formulae
-       $\stamp{\iota}{0}$ and $\bigwedge_{i=1}^n \varphi_i \wedge
-       \phi$ and obtain a formula $I_0$, such that $\stamp{\iota}{0}
-       \models I_0$, $I_0 \wedge \bigwedge_{i=1}^n \varphi_i \wedge
-       \phi$ is unsatisfiable, $\voc{I_0} \subseteq
-       \voc{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n \voc{\varphi_i}
-       \cup \voc{\phi}) \subseteq \stamp{Q}{0}$ and $\pset{-}{I_0}
-       \subseteq \pset{-}{\stamp{\iota}{0}} \cap (\bigcup_{i=1}^n
-       \pset{-}{\varphi} \cup \pset{-}{\phi}) = \emptyset$. Repeating
-       the reasoning for the formulae $I_0 \wedge \varphi_1$ and
-       $\bigwedge_{i=2}^n \varphi_i \wedge \phi$, we obtain $I_1$,
-       such that $I_0 \wedge \varphi_1 \models I_1$, $I_1 \wedge
-       \bigwedge_{i=2}^n \varphi_i \wedge \phi$ is unsatisfiable,
-       $\voc{I_1} \subseteq (\voc{I_0} \cup \voc{\varphi_1}) \cap
-       (\bigcup_{i=2}^n \voc{\varphi_i} \cup \voc{\phi}) \subseteq
-       \stamp{Q}{1}$ and $\pset{-}{I_1} \subseteq (\pset{-}{I_0} \cup
-       \pset{-}{\varphi_1}) \cap (\bigcup_{i=2}^n \pset{-}{\varphi_i}
-       \cup \pset{-}{\phi}) = \emptyset$. Continuing in this way, we
-       obtain formulae $I_0, I_1, \ldots, I_{n}$ as required. \qed
-
 A problematic point of the above proposition is that the existence of
@@
 (e.g.\ inequalities for integers) atoms from the transition rules of
-$\A$ belong to a decidable first order theory, such as Presburger
+$\A$ belong to a decidable first-order theory, such as Presburger
 arithmetic, Lemma \ref{lemma:quant-pred-acceptance} gives us an
@@
 \begin{theorem}[\cite{BorgerGraedelGurevich97}]\label{thm:skolem}
-  Given $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$ a first order sentence,
+  Given $Q_1 x_1 \ldots Q_n x_n ~.~ \phi$ a first-order sentence,
   where $Q_1, \ldots, Q_n \in \set{\exists,\forall}$ and $\phi$ is
@@
 \end{theorem}
-\paragraph{\em Proof.} See \cite[Theorem 2.1.8]{BorgerGraedelGurevich97} and
-\cite[Lemma 2.1.9]{BorgerGraedelGurevich97}. \qed
-\noindent
 Examples of witness-producing quantifier elimination procedures can be
@@
 \cite{KuncakMayerPiskacSuter12}.
-
-%% Furthermore, we describe a cubic-time witness generation for
-%% difference bounds constraints, a simple subset of linear integer
-%% (real) arithmetic (\S\ref{sec:applications}).
 
@@
 
-%% Since each formula $I_k$ of the GLI may refer to the current
-%% predicates $\stamp{Q}{k}$ and to the past and current input variables
-%% $\stamp{X}{\leq k}$, we further give up on the locality of
-%% interpolants and work with non-local GLI that are effectively
-%% computable by a combination of quantifier elimination with
-%% interpolation for quantifier-free data theories with uninterpreted
-%% functions.
-
 \begin{example}[Contd. from Examples \ref{ex:quant-elim} and
@@
   $\quantform{\alpha}$ and let $\xi : [1,m] \rightarrow [n]$ be a
-  monotonic function mapping each transition quantifier to the minimal
-  index from the sequence $\quantformn{\alpha_0}, \ldots,
-  \quantformn{\alpha_n}$ where it occurs. Then one can effectively build:
+  monotonic finite-range function mapping each transition quantifier
+  to the minimal index from the sequence $\quantformn{\alpha_0},
+  \ldots, \quantformn{\alpha_n}$ where it occurs. Then one can
+  effectively build:
   \begin{compactenum}
@@
 \end{lemma}
-\proof{(\ref{it1:subst-quant}) If $\accform{\alpha}$ is unsatisfiable,
-  by Lemmas \ref{lemma:quant} and \ref{lemma:subst}, we obtain that,
-  successively $\quantform{\alpha}$ and $\substform{\alpha}$ are
-  unsatisfiable. Let $Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$ and
-  $Q_1x_1 \ldots Q_mx_m ~.~ \overline{\Phi}$ be prenex forms for
-  $\quantform{\alpha}$ and $\substform{\alpha}$, respectively. Since
-  we assumed that the first order theory of the data domain has
-  witness-producing quantifier elimination, using Theorem
-  \ref{thm:skolem} one can effectively build witness terms
-  $\tau_{i_1}, \ldots, \tau_{i_\ell}$, where
-  $\set{i_1,\ldots,i_\ell}=\set{i \in [1,m] \mid Q_i=\forall}$
-  and: \begin{compactitem}
-  \item $\voc{\tau_{i_j}} \subseteq \stamp{X}{\leq \xi(i_j)} \cup
-    \set{x_k \mid k < i_j, Q_k = \exists}$, for all $j \in [1,\ell]$ and
-    %
-  \item $\overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
-    \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.
-  \end{compactitem}
-  Let $\widehat{\Phi}_0$, \ldots, $\widehat{\Phi}_n$ be the sequence
-  of quantifier-free formulae, defined as follows: \begin{compactitem}
-  \item $\widehat{\Phi}_0$ is the matrix of some prenex form of
-    $\stamp{\iota}{0}$,
-    %
-  \item for all $i=1,\ldots,n$, let $\widehat{\Phi}_i$ be the matrix
-    of some prenex form of:
-    \[\widehat{\Phi}_i \isdef \widehat{\Phi}_{i-1} \wedge
-     \underbrace{\bigwedge_{\begin{array}{l}
-         \scriptstyle{\stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \widehat{\Phi}_{i-1}} \\
-        \scriptstyle{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{} \psi \in \Delta} 
-     \end{array}} \stamp{q}{i-1}(t_1,\ldots,t_{\#(q)}) \rightarrow 
-     \stamp{\psi}{i}[t_1/y_1, \ldots, t_{\#(q)}/y_{\#(q)}]}_{\isdef \phi_i}\]  
-  \end{compactitem}
-  It is easy to see that $\widehat{\Phi}$ is the matrix of some prenex
-  form of:
-  \[\widehat{\Phi}_n \wedge \underbrace{\bigwedge_{\begin{array}{l}
-        \scriptstyle{\stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \text{ occurs in } \widehat{\Phi}_n} \\
-        \scriptstyle{q \in Q \setminus F}
-    \end{array}} \stamp{q}{n}(t_1,\ldots,t_{\#(q)}) \rightarrow \bot}_{\isdef \psi}\]
-  Applying the equivalence from Fact \ref{fact:subst} in the proof of
-  Lemma \ref{lemma:subst}, we obtain a sequence of quantifier-free
-  formulae $\overline{\Phi}_0, \ldots, \overline{\Phi}_n$ such that
-  $\overline{\Phi}_i \equiv \widehat{\Phi}_i$, for all $i \in [n]$ and
-  $\overline{\Phi}$ is obtained from $\overline{\Phi}_n$ by replacing
-  each occurrence of a predicate atom $q(t_1,\ldots,t_{\#(q)})$ in
-  $\overline{\Phi}_n$ by $\bot$ if $q \in Q\setminus F$ and by $\top$
-  if $q \in F$. Clearly $\overline{\Phi} \equiv \widehat{\Phi}$, thus
-  $\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
-    \tau_{i_\ell}/x_{i_\ell}] \equiv
-  \overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
-    \tau_{i_\ell}/x_{i_\ell}] \equiv \bot$.
-
-  \noindent(\ref{it2:subst-quant}) With the notation introduced at
-  point (\ref{it1:subst-quant}), we have $\widehat{\Phi} =
-  \widehat{\Phi}_0 \wedge \bigwedge_{i=1}^n \phi_i \wedge
-  \psi$. Consider the sequence of witness terms $\tau_{i_1}, \ldots,
-  \tau_{i_\ell}$, whose existence is proved by point
-  (\ref{it1:subst-quant}). Because $\voc{\tau_{i_j}} \subseteq
-  \stamp{X}{\leq \xi(i_j)} \cup \set{x_k \mid k < i_j, Q_k =
-    \exists}$, for all $j \in [1,\ell]$, and moreover $\xi^{-1}$ is
-  strictly monotonic, we obtain: \begin{compactitem}
-  \item $\voc{\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{0} \cup
-    \stamp{X}{0} \cup \set{x_j \mid j < \xi_{\max}^{-1}(0), Q_j = \exists}$,
-    %
-  \item $\voc{\phi_i[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{i-1} \cup
-    \stamp{Q}{i} \cup \stamp{X}{\leq i} \cup \set{x_j \mid j <
-      \xi_{\max}^{-1}(i), Q_j = \exists}$, for all $i \in [1,n]$,
-    %
-  \item $\voc{\psi[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}]} \subseteq \stamp{Q}{n} \cup
-    \stamp{X}{\leq n} \cup \set{x_j \mid j \in [1,m], Q_j = \exists}$.
-  \end{compactitem}
-  By repeatedly applying Lyndon's Interpolation Theorem, we obtain a
-  sequence of formulae $(I_0, \ldots, I_n)$ such that: \begin{compactitem}
-    \item $\widehat{\Phi}_0[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}] \models I_0$ and $\voc{I_0} \subseteq
-      \stamp{Q}{0} \cup \stamp{X}{0} \cup \set{x_j \mid j <
-        \xi_{\max}^{-1}(0), Q_j = \exists}$,
-      %
-    \item $I_{k-1} \wedge \phi_i[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}] \models I_k$ and $\voc{I_k} \subseteq
-      \stamp{Q}{k} \cup \stamp{X}{\leq k} \cup \set{x_j \mid j <
-       \xi_{\max}^{-1}(k), Q_j = \exists}$, for all $k \in [1,n]$,
-      %
-    \item $I_n \wedge \psi[\tau_{i_1}/x_{i_1}, \ldots,
-      \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable.
-  \end{compactitem}
-  To show that $(I_0,\ldots,I_n)$ is a GLI for $a_1\ldots a_n$, it is
-  sufficient to notice that \[\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
-    \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-  \stamp{q}{k}(\vec{y}) \rightarrow \stamp{\psi}{k+1} \models \phi_k\]
-  for all $k \in [1,n]$. Consequently, we obtain: \begin{compactitem}
-    \item $\stamp{\iota}{0} \models \widehat{\Phi}_0 \models I_0$, by
-      Theorem \ref{thm:skolem},
-      %
-    \item $I_{k-1} \wedge \Big(\bigwedge_{q(\vec{y}) \arrow{a_k(X)}{}
-      \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
-      \stamp{q}{k-1}(\vec{y}) \rightarrow \stamp{\psi}{k}\Big) \models
-      I_{k-1} \wedge \phi_k \models I_k$, and 
-      %
-    \item $I_n \wedge \Big(\bigwedge_{q \in Q \setminus F} \forall y_1
-      \ldots \forall y_{\#(q)} ~.~ q(\vec{y}) \rightarrow \bot\Big)
-      \models I_n \wedge \psi \models \bot$,     
-  \end{compactitem}
-  as required by Definition
-  \ref{def:generalized-lyndon-interpolant}. \qed}
-
-Consequently, under two assumptions about the first order theory of
+
+Consequently, under two assumptions about the first-order theory of
 the data domain, namely\ \begin{inparaenum}[(i)]
@@
 \end{inparaenum}
-we developped a generic method that produces GLIs for unfeasible input
-event sequences. Moreover, each formula $I_k$ in the interpolant
-refers only to the current predicate symbols $\stamp{Q}{I_k}$, the
-current and past input variables $\stamp{X}{\leq k}$ and the
-existentially quantified transition variables introduced at the
-previous steps $\set{x_j \mid j < \xi_{\max}^{-1}(k), Q_j =
-  \exists}$. The remaining questions are how to use these GLIs to
-label the sequences in the unfolding of an automaton (Definition
+we developed a generic method that produces GLIs for unfeasible input
+event sequences. Moreover, each formula in the interpolant refers only
+to the current predicate symbols, the current and past input variables
+and the existentially quantified transition variables introduced at
+the previous steps. The remaining questions are how to use these GLIs
+to label the sequences in the unfolding of an automaton (Definition
 \ref{def:unfolding}) and compute coverage (Definition
@@
     time stamp of each predicate symbol $\stamp{q}{k}$ and
-    existentially quantifying each free variable in $I, _k$,
+    existentially quantifying each free variable, and
   %
@@
 \end{lemma}
-\proof{
-  The new set of formulae $U'(\alpha_0), \ldots, U'(\alpha_n)$ complies
-  with Definition \ref{def:unfolding}, because: \begin{compactitem}
-  \item $U'(\alpha_0) \equiv \iota$, since, by point
-    \ref{it2:generalized-lyndon-interpolant} of Definition
-    \ref{def:generalized-lyndon-interpolant}, we have $\stamp{\iota}{0}
-    \models I_0$, thus $\iota \models J_0$ and $U'(\alpha_0) =
-    U(\alpha_0) \wedge J_0 \equiv \iota \wedge J_0 \equiv \iota$, and
-    %
-  \item by Definition \ref{def:generalized-lyndon-interpolant}
-    (\ref{it3:generalized-lyndon-interpolant}) we have, for all $k \in
-    [n-1]$:
-    \[I_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
-    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{k}(\vec{y})
-    \rightarrow \stamp{\psi}{k+1} \models I_{k+1}\] We write
-    $\overstamp{I}{j}_k$ for the formula in which each predicate
-    symbol $\stamp{q}{k}$ is replaced by $\stamp{q}{j}$. Then the
-    following entailment holds:
-    \[\overstamp{I}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
-    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
-    \rightarrow \stamp{\psi}{1} \models \overstamp{I}{1}_{k+1}\]
-    Because $J_k$ is obtained by removing the time stamps from the predicate symbols and  
-    existentially quantifying all the free variables of $I_k$, we also
-    obtain, applying Fact \ref{fact:ex-entail} below:
-    \[\stamp{J}{0}_k \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
-    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
-    \rightarrow \stamp{\psi}{1} \models \stamp{J}{1}_{k+1}\] Since $U$
-    satisfies the labeling condition of Definition \ref{def:unfolding}
-    (\ref{it3:unfolding}) and $U'(\alpha_k) = U(\alpha_k) \wedge J_k$,
-    we obtain, as required: 
-    \[\stamp{U'(\alpha_k)}{0} \wedge \bigwedge_{\scriptscriptstyle{q(\vec{y}) \arrow{a_k(X)}{} \psi \in \Delta}} 
-    \forall y_1 \ldots \forall y_{\#(q)} ~.~ \stamp{q}{0}(\vec{y})
-    \rightarrow \stamp{\psi}{1} \models
-    \stamp{U'(\alpha_{k+1})}{1}\enspace.\]
-  \end{compactitem}
-
-  \begin{fact}\label{fact:ex-entail}
-    Given formulae $\phi(\vec{x}, \vec{y})$ and $\psi(\vec{x})$ such
-    that $\phi(\vec{x}, \vec{y}) \models \psi(\vec{x})$, we also have
-    $\exists \vec{x} ~.~ \phi(\vec{x},\vec{y}) \models \exists \vec{x}
-    ~.~ \psi(\vec{x})$. 
-  \end{fact}
-  \proof{ For each choice of a valuation for the existentially
-    quantified variables on the left-hand side, we chose the same
-    valuation for the variables on the right-hand side. \qed}\qed}
-
 Observe that, by Lemma \ref{lemma:subst-quant}
@@
 Nevertheless, if we require just a yes/no answer (i.e.\ not an
-interpolant) recently developped quantifier instantiation heuristics
+interpolant) recently developed quantifier instantiation heuristics
 \cite{ReynoldsKK17} perform rather well in answering a large number of
@@
 
-\ifLongVersion
-\section{Applications}
-\label{sec:applications}
-
-The main application of first order alternating automata is checking
-inclusion between various classes of automata extended with variables
-ranging over infinite domains that recognize languages over infinite
-alphabets. The most widely known such classes are \emph{timed
-  automata} \cite{AlurDill94} and \emph{finite-memory (register)
-  automata} \cite{KaminskiFrancez94}. In both cases, complementation
-is not possible inside the class and inclusion is undecidable. Our
-contribution is providing a systematic semi-algorithm for these
-decision problems. In addition, the method described in
-\S\ref{sec:emptiness} can extend our previous \emph{generic register
-  automata} \cite{IosifRV16} inclusion checking framework, by allowing
-monitor (right-hand side) automata to have local variables, that are
-not visible in the language.
-
-Another application is checking safety (mutual exclusion, absence of
-deadlocks, etc.) and liveness (termination, lack of starvation, etc.)
-properties of parameterized concurrent programs, consisting of an
-unbounded number of replicated threads that communicate via a fixed
-set of global variables (locks, counters, etc.). The verification of
-parametric programs has been reduced to checking the emptiness of a
-(possibly infinite) sequence of first order alternating automata,
-called \emph{predicate automata} \cite{Farzan15,Farzan16}, encoding
-the inclusion of the set of traces of a parametric concurrent program
-into increasingly general \emph{proof spaces}, obtained by
-generalization of counterexamples. The program and the proof spaces
-are first order alternating automata over the infinite alphabet of
-pairs consisting of program statements and thread identifiers.
-
-\subsection{Timed Automata}
-\label{sec:ta}
-
-The standard definition of a finite \emph{timed word} is a sequence of
-pairs $(a_1,\tau_1), \ldots, (a_n,\tau_n) \in (\Sigma \times
-\real)^*$, where $\real$ is the set of real numbers, such that $0 \leq
-\tau_i < \tau_{i+1}$, for all $i \in [1,n-1]$. Intuitively, $\tau_i$
-is the moment in time where the input event $a_i$ occurs. Given a set
-$C$ of \emph{clocks}, the set $\Phi(C)$ of \emph{clock constraints} is
-defined inductively as the set of formulae $x \leq c$, $x \geq c$,
-$\neg\delta$, $\delta_1 \wedge \delta_2$, where $x \in C$, $c \in
-\rat$ is a rational constant and $\delta,\delta_1,\delta_2 \in
-\Phi(X)$.
-
-A \emph{timed automaton} is a tuple $T = \tuple{\Sigma,S,S_0,F,C,E}$,
-where: $\Sigma$ is a finite set of input events, $S$ is a finite set
-of states, $S_0, F \subseteq S$ are sets of initial and final states,
-respectively, $C$ is a finite set of clocks and $E \subseteq S \times
-\Sigma \times S \times 2^C \times \Phi(C)$ is the set of transitions
-$(s,a,s',\lambda,\delta)$ from state $s$ to state $s'$ with symbol
-$a$, $\lambda$ is the set of clocks to be reset and $\delta$ is a
-clock constraint. A run of $T$ over a timed word $w = (a_1,\tau_1)
-\ldots (a_n,\tau_n)$ is a sequence $(s_0,\gamma_0) \ldots
-(s_n,\gamma_n)$, where $s_i \in S$, $\gamma_i : C \rightarrow \real$
-are clock valuations, for all $i \in [n]$, and: \begin{compactitem}
-\item $s_0 \in S_0$ and $\gamma_0(x)=0$ for all $x \in C$, 
-%
-\item for all $i \in [n]$, there exists a transition
-  $(s_i,a_i,s_{i+1},\lambda_i,\delta_i) \in E$ such that $\gamma_i +
-  \tau_{i+1} - \tau_i \models \delta_i$, and for all $x \in C$,
-  $\gamma_{i+1}(x) = 0$ if $x \in \lambda_i$ and $\gamma_{i+1}(x) = \gamma_i(x)
-  + \tau_{i+1}-\tau_i$, otherwise. Here $\tau_0 \isdef 0$ and $\gamma_i +
-  \tau_{i+1} - \tau_i$ is the valuation mapping each $x \in C$ to
-  $\gamma_i(x) + \tau_{i+1} - \tau_i$.
-\end{compactitem}
-The run is accepting iff $s_n \in F$, in which case $T$ accepts
-$w$. As usual, we denote by $\lang{T}$ the set of finite words
-accepted by $T$. It is well-known that, in general, there is no timed
-automaton accepting the complement language $(\Sigma\times\real)^*
-\setminus \lang{T}$ and, moreover, the language inclusion problem is
-undecidable \cite{AlurDill94}.
-
-Given a timed automaton $T = \tuple{\Sigma,S,S_0,F,C,E}$, we define a
-first order alternating automaton $\A_T =
-\tuple{\Sigma,\set{t},Q_T,\iota_T,F_T,\Delta_T}$, with a single input
-variable $t$, ranging over $\real$, such that each timed word $w =
-(a_1,\tau_1) \ldots (a_n,\tau_n)$ corresponds to a unique data word
-$d(w) = (a_1,\nu_1) \ldots (a_n,\nu_n)$ such that $\nu_i(t) = \tau_i$
-for all $i \in [1,n]$ and $\lang{\A_T} = \set{d(w) \mid
-  w\in\lang{T}}$. The only difficulty here is capturing the fact that
-all the clocks of $T$ evolve at the same pace, which is easily done
-using a technique from \cite{Fribourg98}, which replaces
-each clock $x_i$ of $T$ by a variable $y_i$ tracking the difference
-between the values of $t$ and $x_i$.
-
-Formally, if $C = \set{x_1,\ldots,x_k}$ and $S =
-\set{s_1,\ldots,s_m}$, we define $Q_T \isdef \set{q_1, \ldots, q_m}$ ,
-where $\#(q_i) = k+1$ for all $i \in [1,m]$, $\iota_T \isdef
-\bigvee_{s_i \in S_0} q_i(0,\ldots,0)$, $F_T \isdef \set{q_i \mid s_i
-  \in F}$ and, for each transition $(s_i,a,s_j,\lambda,\delta) \in E$,
-$\Delta_T$ contains the rule:
-\[q_i(y_1,\ldots,y_k,z) \arrow{a(t)}{} t > z \wedge \delta(z-y_1,\ldots,z-y_k) \wedge q_j(y'_1,\ldots,y'_k,t)\]
-where $y'_i$ stands for $z$ if $x_i \in \lambda$ and for $y_i$, 
-otherwise.  Moreover, nothing else is in $\Delta_T$. We establish the
-following connection between a timed automaton and its corresponding
-first order alternating automaton.
-
-\begin{proposition}\label{prop:timed-alternating}
-  Given a timed automaton $T = \tuple{\Sigma,S,S_0,F,C,E}$, the
-  first order alternating automaton $\A_T =
-  \tuple{\Sigma,\set{t},Q_T,\iota_T,F_T,\Delta_T}$ recognizes the
-  language $\lang{\A_T} = \set{d(w) \mid w \in \lang{T}}$. 
-\end{proposition}
-\proof{ ``$\subseteq$'' Let $w = (a_1,\nu_1) \ldots (a_n,\nu_n) \in
-  \lang{\A_T}$ be a data word. We show the existence of a timed word
-  $(a_1,\tau_1) \ldots (a_n,\tau_n) \in \lang{T}$ such that
-  $\nu_i(t)=\tau_i$, for all $i \in [1,n]$, by induction on
-  $n\geq0$. In fact we shall prove the following stronger
-  statements: \begin{compactenum}
-  \item each execution of $\A_T$ over $w$ starting with a cube $c \in
-    \cube{\minsem{\iota_T}}$ is a linear tree, in which each node has
-    at most one child.
-    %
-  \item for each execution $q_{i_0}(d^0_1,\ldots,d^0_k,\tau_0) \ldots
-    q_{i_n}(d^n_1,\ldots,d^n_k,\tau_n)$ of $\A_T$, $T$ has an
-    execution $(s_{i_0}, \gamma_0) \ldots (s_{i_n}, \gamma_n)$ over
-    the timed word $(a_1, \tau_1) \ldots (a_n, \tau_n)$, such that,
-    for all $i \in [1,n]$ and all $\ell \in [1,k]$, we have
-    $\gamma_i(x_\ell) = \tau_{i-1} - d^i_\ell$.
-  \end{compactenum}
-  The first point above is by inspection of $\iota_T = \bigvee_{s_i
-    \in S_0} q_i(0,\ldots,0)$ and of the rules from
-  $\Delta_T$. Indeed, each minimal model of $\iota_T$ corresponds to a
-  cube $q(0,\ldots,0)$ and each rule has exactly one predicate atom on
-  its right-hand side, thus each node of the execution will have at
-  most one successor. The second point is by induction on $n\geq0$.
-
-  ``$\supseteq$'' Let $w = (a_1,\tau_1) \ldots (a_n,\tau_n) \in
-  \lang{T}$ be a timed word. By induction on $n\geq0$, we show that
-  for each run $(s_{i_0}, \gamma_0) \ldots (s_{i_n}, \gamma_n)$ of $T$
-  over $w$, $\A_T$ has a linear execution
-  $q_{i_0}(d^0_1,\ldots,d^0_k,\tau_0) \ldots
-  q_{i_n}(d^n_1,\ldots,d^n_k,\tau_n)$ such that, for all $i \in [1,n]$
-  and all $\ell \in [1,k]$, we have $\gamma_i(x_\ell) = \tau_{i-1} -
-  d^i_\ell$. \qed}
-
-An easy consequence is that the timed language inclusion problem
-``given timed automata $T_1$ and $T_2$, does $\lang{T_1} \subseteq
-\lang{T_2}$ ?'' is reduced in polynomial time to the emptiness problem
-$\lang{\A_{T_1}} \cap \lang{\overline{\A_{T_2}}} = \emptyset$, for
-which (\S\ref{sec:emptiness}) provides a semi-algorithm. Observe,
-moreover, that no transition quantifiers are needed to encode timed
-automata as first order alternating automata.
-
-\subsection{Register Automata}
-\label{sec:ra}
-
-Finite-memory automata, most commonly referred to as register automata
-\cite{KaminskiFrancez94} are among the first attempts at lifting the
-finite alphabet restriction of classical Rabin-Scott automata. In a
-nutshell, a register automaton is a finite-state automaton equipped
-with a finite set of registers $x_1,\ldots,x_r$ able to copy input
-values and compare them with subsequent input. Consequently, basic
-results from classical automata theory, such as the pumping lemma or
-the closure under complement do not hold in this model and, moreover,
-inclusion of languages recognized by register automata is undecidable
-\cite{NevenSchwentickVianu03}.
-
-Let $\Sigma$ be an infinite alphabet, $\#$ be a symbol not in $\Sigma$
-and $r>0$ be an integer constant, denoting the number of registers. An
-\emph{assignment} is a word $\vec{v} = v_1 \ldots v_r$ such that if
-$v_i = v_j$ and $i \neq j$ then $v_i = \#$, for all $i,j \in
-     [1,r]$. We write $[\vec{v}]$ for the set $\set{v_i \mid i \in
-       [1,r]}$ of values in the assignment $\vec{v}$. A
-     \emph{finite-memory (register) automaton} is a tuple $R =
-     \tuple{S,q_0,\vec{u}, \rho, \mu, F}$, where $S$ is a finite set
-     of states, $q_0 \in S$ is the initial state, $\vec{u} = u_1
-     \ldots u_r$ is the initial assignment, $\rho : S \rightarrow
-            [1,r]$ is the reassignment partial function, $\mu
-            \subseteq S \times [1,r] \times S$ is the transition
-            relation and $F \subseteq S$ is the set of final states. A
-            run of $A$ over an input word $a_1 \ldots a_n \in
-            \Sigma^*$ is a sequence $(s_0,\vec{v}_0) \ldots
-            (s_n,\vec{v}_n)$ such that $\vec{v}_0 = \vec{u}$ and, for
-            all $i \in [1,n]$, exactly one of the following
-            holds: \begin{compactitem}
-\item if there exists $k\in[1,r]$ such that $a_i=(\vec{v}_{i-1})_k$
-  then $\vec{v}_i = \vec{v}_{i-1}$ and $(s_{i-1},k,s_i) \in \mu$, 
-%
-\item otherwise $a_i\not\in[\vec{v}_{i-1}]$, $\rho(s_{i-1})$ is
-  defined, $(\vec{v}_i)_{\rho(s_{i-1})} = a_i$, for each $k \in
-  [1,r] \setminus \set{\rho(s_{i-1})}$, we have $(\vec{v}_i)_k =
-  (\vec{v}_{i-1})_k$ and $(s_{i-1},\rho(s_{i-1}),s_i) \in \mu$.
-\end{compactitem}
-Intuitively, if the input symbol is already stored in some register,
-the automaton moves to the next state if, moreover, the transition
-relation allows it, otherwise it copies the input to the register
-indicated by the reassignment, erasing its the previous value, and
-moves according to the transition relation.
-
-The translation of register automata to first order alternating
-automata is quite natural, because registers can be encoded as
-arguments of predicate atoms. Formally, given a register automaton $R
-= \tuple{S,s_0,\vec{u},\rho,\mu,F}$, such that $S =
-\set{s_0,\ldots,s_m}$, we define the alternating automaton $\A_R =
-\tuple{\set{\alpha},\set{x},Q_R,\iota_R,F_R,\Delta_R}$, where
-$\alpha \not\in \Sigma$, $Q_R \isdef \set{q_0,\ldots,q_m}$ and
-$\#(q_i) = r$ for all $i \in [m]$, $\iota_R \isdef q_0(\vec{u})$, $F_R
-\isdef \set{q_i \mid s_i \in F}$ and, for each transition $(s_i, k,
-s_j) \in \mu$, $\Delta_T$ contains the rule:
-\[q_i(y_1,\ldots,y_r) \arrow{\alpha(x)}{} y_k=x \wedge q_j(y_1,\ldots,y_r) ~\vee~
-\bigwedge_{i=1}^r x \neq y_i \wedge q_j(y_1,\ldots,y_{k-1},x,y_{k+1},
-\ldots, y_r)\] Moreover, nothing else is in $\Delta_R$. The connection
-between register automata and first order alternating automata is
-stated below.
-
-\begin{proposition}\label{prop:register-alternating}
-  Given a register automaton $R = \tuple{S,s_0,\vec{u},\rho,\mu,F}$
-  over an infinite alphabet $\Sigma$, the first order alternating
-  automaton $\A_R = \tuple{\set{\alpha},Q_R,\iota_R,F_R,\Delta_R}$
-  recognizes the languge $\lang{\A_R} = \set{ (\alpha,a_1) \ldots
-    (\alpha,a_n) \mid a_1 \ldots a_n \in \lang{R} }$. 
-\end{proposition}
-\proof{ ``$\subseteq$'' Let $w = (\alpha,a_1) \ldots (\alpha,a_n)
-  \in \lang{\A_R}$. First, it is easy to show that each execution of
-  $\A_R$, that starts in some cube $c \in \cube{\minsem{\iota_R}}$, is
-  a linear tree with labels $q_0(\vec{v}_0), \ldots, q_n(\vec{v}_0)$
-  such that $\vec{v}_0 = \vec{u}$. Second by induction on $n\geq0$, we
-  prove that $\A_R$ has a run as above over $w$ only if $R$ has a run
-  $(q_0,\vec{v}_0), \ldots, (q_n,\vec{v}_n)$ over $a_1 \ldots
-  a_n$. ``$\supseteq$'' Let $w = a_1 \ldots a_n \in \lang{R}$ and
-  $q_0(\vec{v}_0), \ldots, q_n(\vec{v}_0)$ be a run of $R$ over $w$,
-  such that $\vec{v}_0 = \vec{u}$. By induction on $n\geq0$, we can
-  build an execution of $\A_R$ over $(\alpha,a_1) \ldots
-  (\alpha,a_n)$ that is a linear tree with labels $q_0(\vec{v}_0),
-  \ldots, q_n(\vec{v}_n)$. \qed}
-
-Consequently, the language inclusion problem ``given register automata
-$R_1$ and $R_2$, does $\lang{R_1} \subseteq \lang{R_2}$?'' is reduced
-in polynomial time to emptiness problem $\lang{\A_{R_1}} \cap
-\lang{\overline{\A_{R_2}}} = \emptyset$, for which
-(\S\ref{sec:emptiness}) provides a semi-algorithm. Notice further that
-the encoding of register automata as first order alternating automata
-uses no transition quantifiers.
-
-\subsection{Predicate Automata}
-\label{sec:pa}
-
-The model of \emph{predicate automata} \cite{Farzan15,Farzan16} has
-emerged recently as a tool for checking safety and liveness properties
-of parameterized concurrent programs, in which there is an unbounded
-number of replicated threads that communicate via global
-variables. Predicate automata recognize finite sequences of actions
-that are pairs $(\sigma,i)$, where $\sigma$ is from a finite set
-$\Sigma$ of program statements and $i \in \nat$ ranges over an
-unbounded set of thread identifiers. To avoid clutter, we shall view a
-pair $(\sigma,i)$ as a data symbol $(\sigma,\nu)$ where $\nu(x)=i$,
-for a designated input variable $x$.
-
-Since thread identifiers can only be compared for equality, the data
-theory of predicate automata is the first order theory of
-equality. Moreover, transition quantifiers are only needed for
-checking termination and, generally, liveness properties
-\cite{Farzan16}.
-
-However, the execution semantics of predicate automata differs from
-that of first order automata with respect to the following detail:
-initial configurations and successors of predicate automata are
-defined using the entire sets of models of the initial sentence and
-transition rules, not just the minimal ones, as in our case.
-
-Formally, a run of a predicate automaton $P =
-\tuple{\Sigma,\set{x},Q,\iota,F,\Delta}$ over a word $(a_1,\nu_1)
-\ldots$ $(a_n,\nu_n)$ is a sequence of interpretations $\I_0, \ldots,
-\I_n$ such that $\I_0 \in \sem{\iota}$ and for each $i \in [1,n]$,
-each $q \in Q$ and each tuple $\tuple{d_1,\ldots,d_{\#(q)}} \in
-\I_{i-1}(q)$, we have $\I_i \in \sem{\psi}_\nu$, for each rule
-$q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(x)}{} \psi \in \Delta$, where
-$\nu=\nu_i[y_1\leftarrow d_1, \ldots, y_{\#(q)} \leftarrow
-  d_{\#(q)}]$. The run is accepting if and only if $\I(q)=\emptyset$
-for all $q \in Q \setminus F$.
-
-In fact, as shown next, this more simple execution semantics is
-equivalent, from the language point of view, with the semantics given
-by Definitions \ref{def:execution} and \ref{def:accepting}.  We
-believe that the semantics of first order alternating automata based
-on minimal models is important for its relation to the textbook
-semantics of boolean alternating automata
-\cite{ChandraKozenStockmeyer81}.
-
-\begin{proposition}
-  Given a predicate automaton $P =
-  \tuple{\Sigma,\set{x},Q,\iota,F,\Delta}$, let $\A_P$ be the
-  first order alternating automaton that has the same description as
-  $P$. Then $\lang{P} = \lang{\A_P}$. 
-\end{proposition}
-\proof{ ``$\subseteq$'' Let $w = (a_1,\nu_1) \ldots (a_n,\nu_n) \in
-  \lang{P}$ be a word and $\I_0, \ldots, \I_n$ be an accepting
-  execution of $P$ over $w$. Let $\stamp{\I}{i}_j$ be the
-  interpretation that associates each predicate $\stamp{q}{i}$ the set
-  $\I_j(q)$, for $i,j \in [n]$. Then one builds, by induction on
-  $n\geq0$, an execution $\T$ of $\A_P$ such that $\I_\T \subseteq
-  \bigcup_{i=0}^n \stamp{\I}{i}_i$, where $\I_\T$ is the unique
-  interpretation associated with $\T$. Since $\I_0, \ldots, \I_n$ is
-  accepting, we have $\stamp{\I}{n}_n(\stamp{q}{n}) = \emptyset$, for
-  all $q \in Q \setminus F$ and hence $\I_\T(\stamp{q}{n}) =
-  \emptyset$, for all $q \in Q \setminus F$ and, consequently $w \in
-  \lang{\A_P}$. ``$\supseteq$'' Let $w = (a_1,\nu_1) \ldots
-  (a_n,\nu_n) \in \lang{\A_P}$ be a word and $\T$ be an accepting
-  execution of $\A_P$ over $w$. We define the sequence of
-  interpretations $\I_0, \ldots, \I_n$ as $\I_i(q) =
-  \I_\T(\stamp{q}{i})$, for each $i \in [n]$ and each $q \in Q$. By
-  induction on $n\geq0$ one shows that $\I_0, \ldots, \I_n$ is an
-  execution $P$. Moreover, since $\T$ is accepting, we have $\I_n(q) =
-  \I_\T(\stamp{q}{n}) = \emptyset$, for each $q \in Q \setminus F$,
-  thus $w \in \lang{P}$. \qed}
-
-As before, this result enables using the
-semi-algorithm from \S\ref{sec:emptiness} for checking emptiness of
-predicate automata.  We point out that, although quantifier-free
-predicate automata with predicates of arity one are decidable for
-emptiness \cite{Farzan15}, currently there is no method for checking
-emptiness of predicate automata with predicates of arity greater than
-one, other than the explicit enumeration of cubes. Moreover, no method
-for dealing with emptiness in the presence of transition quantifiers
-is known to exist.
-\fi
-
 \section{Experimental Results}
@@
 \cite{McMillan06} in a prototype tool, avaliable
-online\footnote{\url{https://gofile.io/?c=7zB3g7}}. The tool is
+online \cite{foada}. The tool is
 written in Java and uses the Z3 SMT solver \cite{z3}, via the JavaSMT
@@
 inclusions of the form $\bigcap_{i=1}^N\lang{A_i} \subseteq \lang{B}$,
-where $A_i$ is the $i$-th copy of the same template automaton.
+where $A_i$ is the $i$-th copy of the template automaton.
 
@@
 
-We present first order alternating automata, a model of computation
-that generalizes classical boolean alternating automata to first order
-theories. Due to their expressivity, first order alternating automata
+We present first-order alternating automata, a model of computation
+that generalizes classical boolean alternating automata to first-order
+theories. Due to their expressivity, first-order alternating automata
 are closed under union, intersection and complement. However the

### 2019-07-04 quantifiers/rebuttal.txt
ok

@@
  those references in our paper is a minus that should be
- corrected. One explanation is that we focused on infinite alphabet
- extensions that follow an idea of the seminal papers on timed (TA)
- [1] and finite-memory (register) automata (RA) [13]. Unlike symbolic
- automata that test only the current input, both TA and RA relate
- values occurring at different positions in the input, either by using
- clocks (with reset and difference guards) or registers (with
- over-writing and equality guards). The price to be paid for this
- expressivity is that language inclusion becomes undecidable (unless
- severe restrictions on the number of clocks/registers apply).
+ corrected. The reason is our original interest in language
+ recognizers relating values occurring at different positions in the
+ input, by means of internal variables: clocks for timed automata (TA)
+ and registers for finite-memory automata [13]. Our motivation is not
+ as much finding decidable infinite alphabet models but rather having
+ a very general model, closed under complement, union and intersection
+ (boolean closure is merely a consequence of its expressive power), in
+ which all undecidable inclusion problems can be encoded, then devise
+ a practical semi-algorithm for it.
 
- Our original motivation here was not as much finding decidable
- infinite alphabet models but rather having a very general model,
- (unsurprisingly) closed under complement, union and intersection, in
- which all undecidable inclusion problems can be encoded, then devise
- a practical semi-algorithm for it (that works on reasonably complex
- examples). Since, to the best of our knowledge, there are no
- published semi-algorithms for TA or RA inclusion (tools such as UPAAL
- check only for emptiness), our method could contribute to these areas
- and to program verification (we have considered several programs
- handling integer arrays in our examples).
+ Although the finite event alphabet \Sigma can be encoded in the input
+ data (it is introduced for technical convenience), modeling control
+ by uninterpreted predicates with internal variables as arguments is
+ key to capturing computations using hidden variables (pg 2, first
+ paragraph). Note that Definition 1 (point 2) requires synchronization
+ of all execution branches on all input events and values.
+
+ To the best of our knowledge, there are no published semi-algorithms
+ for TA or RA inclusion (tools such as UPAAL check only for
+ emptiness). Our method contributes to these areas and to program
+ verification (Section 5 considers several programs handling integer
+ arrays).
 
@@
  more (having only equalities and monadic predicates yields a
- decidable model though). In the light of this result (mentioned on
- pg. 9, last paragraph in the paper), any reasonably interesting
- extension of monadic predicate automata seems to be undecidable for
- emptiness.
+ decidable model though). In the light of this result (pg. 9, last
+ paragraph), any reasonably interesting extension of monadic predicate
+ automata is likely to have an undecidable emptiness problem.
 
  We also compare with an alternating automaton model [11], in which
- states are propositional variables and comparisons are possible only
- between adjacent input values, without hidden variables. This model
- is intermediate between s-AFA and FOAA, nevertheless emptiness
- remains undecidable, by reduction from 2-counter machine
- reachability.
+ states are propositional variables (no uninterpreted predicates) and
+ comparisons are possible only between adjacent input values, without
+ hidden variables. This model is intermediate between s-AFA and FOAA,
+ nevertheless emptiness remains undecidable.
 
@@
  is that, in CHC, the different branches of the unfolding tree are
- essentially unrelated, whereas in FOAA, all branches of the
- computation tree must synchronize on the same input word. To account
- for this synchronization, a CHC encoding of FOADA must use recursive
- algebraic data types (ADT), namely lists, to ensure that, at each
- step, all incoming branches consider exactly the same list, encoding
- the continuation of the input from that point on (CHC traverse the
- input from the end to the beginning). An encoding of FOAA in CHC
- would benefit from the advantage of using off-the-shelf CHC solvers,
- but, in our opinion, would face several problems (listed in
- increasing order of difficulty):
+ unrelated (at arbitrary depths), whereas in FOAA, all branches of the
+ computation tree must synchronize on the same input word. As
+ suggested, to reflect this synchronization, a CHC encoding of FOADA
+ must use data lists to ensure that, at each step, all incoming
+ branches consider exactly the same list, encoding the continuation of
+ the input from that point on (CHC traverse the input from the end to
+ the beginning). An encoding of FOAA in CHC would benefit from the
+ advantage of using off-the-shelf CHC solvers, but, in our opinion,
+ would face several problems, listed below in increasing order of
+ difficulty:
 
@@
  input can be done using equalities between the list arguments of the
-subgoals. However, the body (left-hand side) of the clause must be
-given either as a conjunction or, if quantifiers are considered, as
-prenex DNF. This issue can be solved, but the reduction is a bit more
-complex than suggested and, moreover, writing each rule in DNF can be
-costly.
+ subgoal predicates. However, the body of the clause must be given
+ either as a conjunction or, if quantifiers are considered, as prenex
+ DNF. However, turning each rule into DNF would potentially add to the
+ complexity.
 
-II. Introducing a list ADT in the encoding requires the use of an
- interpolation procedure for ADTs in the emptiness
+II. Introducing recursive algebraic data types (ADT) in the encoding
+ requires an interpolation procedure for ADTs in the emptiness
  semi-algorithm. Although satisfiability of quantifier-free ADTs is
@@
  * Hossein Hojjat, Philipp Ruemmer. Deciding and Interpolating
-Algebraic Data Types by Reduction. ArXiv:1801.02367. Proc. of 
+ Algebraic Data Types by Reduction. ArXiv:1801.02367. Proc.
  SYNASC'17.
@@
  Moreover, the availability of such procedures in off-the-shelf CHC
-solvers is still an unresolved issue. For instance, the integration of
-the above procedure in ELDARICA is underway, according to
+ solvers is still an unresolved issue. For instance, the integration
+ of the above procedure in ELDARICA is underway, according to
  [*]. However, when such technical problems will be solved, it would
-definitely be interesting to have an experimental comparison between,
-e.g. FOAA and ELDARICA.
+ definitely be interesting to have an experimental comparison between
+ FOAA and ELDARICA (running on an encoding of our examples using
+ ADTs).
 
-III. Working with quantified rules in FOADA requires a quantifier
-elimination procedure that produces witness terms. However, if
-quantified variables occur in the ADT constructors, the theory is most
-likely undecidable, thus cannot have quantifier elimination, by an
-argument similar to the one of [8].
+III. Having quantified rules in FOADA requires a quantifier
+ elimination procedure that produces witness terms. However, combining
+ quantifiers and ADTs results most likely in an undecidable theory,
+ for which no quantifier elimination exists (by an argument similar to
+ the one of [8]).
 
  Essentially, for the above reasons, we have chosen to design our
-emptiness checking semi-algorithm from scratch, trying to use as few
+ emptiness checking semi-algorithm from scratch, imposing as few
  assumptions as possible on the underlying first-order data theory. In
  fact, we only assume (a) Lyndon interpolation for quantifier-free
-formulae with uninterpreted predicates and (b) witness term generation
-during quantifier elimination.
+ formulae with uninterpreted predicates (without ADTs) and (b) witness
+ term generation during quantifier elimination. Both assumptions are
+ fairly reasonable for standard SMT theories (e.g. UFLIA, UFLRA).
